### 2013-06-13 minipaper/renderers.tex
first commit for all the SML and haskell work

\documentclass{article}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{qtree}
\usepackage{amssymb,amsthm}
\usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
\parindent 0pt
\parskip .75em

%utils
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}

%source nodes 
\newcommand {\oneS}{\mathtt{1S}}
\newcommand {\oneG}{\mathtt{1G}}
\newcommand {\comp}{\mathtt{>=>}}
\newcommand {\chain}{\mathtt{\ggg}}
\newcommand {\fix}{\mathtt{fix}}
\newcommand {\call}{\mathtt{call}}
\newcommand {\id}{\mathtt{id}}
\newcommand {\sampP}{\mathtt{16^2\_SP}}
\newcommand {\twoGP}{\mathtt{2\_GP}}

% monophasic work items
\newcommand {\sbag}{\mathtt{SBag}}
\newcommand {\gbag}{\mathtt{GBag}}
\newcommand {\twoHolder}{\mathtt{2Hold}}
\newcommand {\filt}{\mathtt{Filt}}

%structs
\newcommand {\sleft}{\mathtt{Left}}
\newcommand {\sright}{\mathtt{Right}}
\newcommand {\bound}{\mathit{bound}}
\newcommand {\bag}{\mathtt{Bag}}
\newcommand {\twoBox}{\mathtt{OpenNode}}
\newcommand {\plus}{\mathtt{Plus}}

% other relations
\newcommand{\isect}{\between}
\newcommand{\nisect}{\not \between}
\newcommand{\ccup}{\circled{$\cup$}}
\newcommand{\frust}{\triangledown}

% Equation spacing
\setlength{\jot}{3ex}

\title{\Large\textbf{A Representation of Renderers}}
\author{Nicolas Feltman}
\begin{document}
\maketitle

\section{Definitions}

\subsection{Entities}

Our representation uses a point-free style, meaning that the entities that our algorithm is processing are not explicitly represented with variables.  Nonetheless, those entities do exist, and to understand our representation one must first understand the possible types of those entities.  Those are:

\begin{itemize}

\item {\bf GS-Tuple} (description)
\item {\bf Hits} (description)
\item {\bf Fragment} (description)

\end{itemize}

\subsection{Nodes}

In this section, we present all of the node building blocks from which our representation is built.

The first things to cover are the transitional nodes, namely Hit and Shade.  These are the only nodes that allows entity type change, and so they must be present in order for the system to do any work.  The transition nodes are nullary, meaning that they don't require any child nodes.  The Shade node always takes in a single Hit entity, and outputs a Fragment entity with the same key.  The hit entity takes in a GS tuple containing a single geometry element and a single sample element, and returns their intersection point if and only if that intersection exists.  If no intersection exists, then the Hit node has no output.

The decomposition nodes ({\em eg.} 1G, 1S, 1H, 16$^2$-SP, 2-GP), also nullary, tell the algorithm how to divide the entities being operated on at that moment.  For instance, 2-GP indicates that the algorithm should split the current GS-tuple into two others, based on the geometry.  That is, half the geometry goes to one GS tuple, half goes to the other, and all samples go to both tuples.  Similarly, $16^2$-SP means that the samples should be divided into 256 bins, using some 16$\times$16 grid, and that all geometry should go to each bin.  The 1G and 1S nodes indicate that the algorithm should split up the geometry and samples into their individual units, but still paired with the full set of the other.  1H similarly indicates to break up the hits fully.  All of these decompositional nodes can be thought of as having one input and many outputs.

The filter node, which is nullary, allows a renderer to discard GS-tuples which will not produce any hits.  They do this by by forming bounds around all of the geometry and all of the samples, according to the bounding arguments passed in.  If these bounds have any intersection, then the whole GS-tuple passes, otherwise it is dropped.  Thus the filter node can be thought of as having one input, and zero or one output.

The chain node, $\ggg$, is our first node with children.  Specifically, it is binary.  We can interpret $\ggg$ to mean ``do the thing on the left, then do the thing on the right,'' where the child on the left has no more than one output ({\em eg.} a transition node or a filter).  The number of outputs of $\ggg$ is equal to the number of outputs of the child on the right. Of course, if the item on the left ends up having no outputs, then the item on the right never executes, and the $\ggg$ itself has no outputs.

The $\oplus$ and $\cup$ nodes are binary combinator nodes, like $\ggg$, but they permit children on the left with multiple outputs.  The item on the right must have zero or one outputs.  Thus $\oplus$ and $\cup$ operate by running the left child on the current input, and then for all outputs running the right child.  These outputs are then merged together in a way unique to $\oplus$ and $\cup$.  For this reason, these two are called merge nodes.  In particular, $\oplus$ merges by comparing keys and taking closer hits for those keys.

\subsection{Bounds}

Cover bounding.

\section{Example Trees}

\subsection{Tiled Forward}
Here is the basic tiled rasterizer.
\begin{center}
\Tree [. [.$\oplus$ 1G [.$\cup$ 16^2-SP [.$\chain$ filt(1,$\frust$) [.trimL($\infty$,$\frust$) [.$\cup$ 1S [.$\chain$ Hit Shade ]]]]]]]
\end{center}

\subsection{Tiled Defered Shading}

\begin{center}
\Tree [. [.$\cup$ 16^2-SP [.$\chain$ [.$\oplus$ 1G [.$\chain$ filt(1,$\frust$) [.$\cup$ 1S Hit ]]] [.trimL($\frust$) [.$\cup$ 1H Shade ]]]]]
\end{center}

\subsection{Prepass With Double-Bounded Z Pyramid}

\begin{center}
\Tree [. [.$\chain$ [.Prepass [.$\oplus$ Brute Z-Hint ]] [.$\oplus$ 1G [.$\cup$ 4^2-SP [.$\chain$ filt(1,$\frust$) [.$\cup$ 4^2-SP [.$\chain$ filt(1,$\frust$) [.$\cup$ 1S [.$\chain$ Hit Shade ]]]]]]]]]
\end{center}

\subsection{Tiled Forward+ Shading}

\begin{center}
\Tree [. [.$\chain$ [.Prepass [.$\oplus$ Brute Z-Hint ]] [.$\oplus$ 1G [.$\cup$ 16^2-SP [.$\chain$ filt(1,$\frust$) [.trimL($\infty$,$\frust$) [.$\cup$ 1S [.$\chain$ Hit Shade ]]]]] ]]]
\end{center}

\subsection{Quad-Tree over lights}

\begin{center}
\Tree [. [.$\chain$ [.Prepass [.$\oplus$ Brute Z-Hint ]] [.$\oplus$ 1G [.$\cup$ 4^2-SP [.trimL($\infty$,$\frust$) [.$\cup$ 4^2-SP [.$\chain$ filt(1,$\frust$) [.trimL($\infty$,$\frust$) [.$\cup$ 1S [.$\chain$ Hit Shade ]]]]]]]]]]
\end{center}

\end{document}

### 2013-06-13 oldSemantics/renderers.tex
first commit for all the SML and haskell work

\documentclass{article}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{qtree}
\usepackage{amssymb,amsthm}
\usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
\parindent 0pt
\parskip .75em

%source nodes 
\newcommand {\oneS}{\mathtt{1S}}
\newcommand {\oneG}{\mathtt{1G}}
\newcommand {\comp}{\mathtt{>=>}}
\newcommand {\fix}{\mathtt{fix}}
\newcommand {\call}{\mathtt{call}}
\newcommand {\id}{\mathtt{id}}
\newcommand {\sampP}{\mathtt{16^2\_SP}}
\newcommand {\twoGP}{\mathtt{2\_GP}}

% monophasic work items
\newcommand {\sbag}{\mathtt{SBag}}
\newcommand {\gbag}{\mathtt{GBag}}
\newcommand {\twoHolder}{\mathtt{2Hold}}
\newcommand {\filt}{\mathtt{Filt}}

%structs
\newcommand {\sleft}{\mathtt{Left}}
\newcommand {\sright}{\mathtt{Right}}
\newcommand {\bound}{\mathit{bound}}
\newcommand {\bag}{\mathtt{Bag}}
\newcommand {\twoBox}{\mathtt{OpenNode}}
\newcommand {\plus}{\mathtt{Plus}}

% other relations
\newcommand{\isect}{\between}
\newcommand{\nisect}{\not \between}

% Equation spacing
\setlength{\jot}{3ex}

\title{\Large\textbf{ RenderGen Notes}}
\author{Nicolas Feltman, Yong He, Kayvon Fatahalian}
\begin{document}
\maketitle

\section{Fragment Calculus}

A fragment group is represented as a partial function from samples to geometry.  Let $\oplus$ be a function which merges two fragment groups.

\section{The Monophasic Renderer Model}

This model takes in both the geometry and the samples at the same time and computes the results directly.

\subsection{Utility Nodes}
\subsubsection{Fixed Points}
\begin{gather}
\infer{\langle \fix(N,l), \kappa, \Phi, G, S\rangle \to\langle N, \kappa, [\Phi, l:N], \Phi, G, S \rangle}{ \cdot }
\\
\infer{\langle \call(l), \kappa, \Phi, G, S \rangle \to \langle \Phi(l), \kappa, \Phi, \Phi, G, S \rangle }{\cdot}
\end{gather}

\subsection{Decompositional Nodes}

\subsection{Filter Work Item}
\begin{gather}
\infer {\filt(\gamma, \sigma, A) \to A}{\gamma \isect \sigma}
\\
\infer {\filt(\gamma, \sigma, A) \to \bot}{\gamma \nisect \sigma}
\end{gather}

\subsubsection{Identity Node}
\begin{gather}
\infer {\langle\id, \kappa, \Phi,G, S \rangle \to \langle \kappa, \emptyset, \Phi, G, S \rangle}{\cdot}
\end{gather}

\subsubsection{Full Decomposition Nodes}
\begin{gather}
\infer{\bag(\ldots, A,\ldots, B, \ldots) \to \bag(\ldots, B,\ldots, A, \ldots)}{\cdot}
\\
\infer{\bag(A, \ldots) \to \bag(A', \ldots)}{A \to A'}
\\
\infer{\bag(\bot, A, \ldots) \to \bag(A, \ldots)}{\cdot}
\\
\infer{\bag(\bot) \to \bot}{\cdot}
\\
\infer{\bag(f_1, \ldots, f_n) \to f_1 \oplus \cdots \oplus f_n}{\cdot} 
\end{gather}
First, $\oneS$:
\begin{gather}
\infer{\langle\oneS, \kappa, \Phi, G, S  \rangle \to \bag(
\langle \kappa, \emptyset, \Phi, G, s_1 \rangle, \ldots, 
\langle \kappa, \emptyset, \Phi, G, s_n \rangle)}
{S = \{s_1, \ldots, s_n\}}
\end{gather}
And now $\oneG$:
\begin{gather}
\infer{\langle\oneS, \kappa, \Phi, G, S \rangle \to \bag(
\langle \kappa, \emptyset, \Phi, g_1, S \rangle, \ldots, 
\langle \kappa, \emptyset, \Phi, g_n, S \rangle)}
{G = \{g_1, \ldots, g_n\}}
\end{gather}

\subsubsection{Geometry Partitions}
\begin{gather}
\infer{
\langle\twoGP(p), \kappa, \Phi, G^\gamma, S^\sigma \rangle \to \bag(
\langle \kappa, \emptyset, \Phi, G_1^{\gamma_1}, S^\sigma \rangle,
\langle \kappa, \emptyset, \Phi, G_2^{\gamma_2}, S^\sigma \rangle)}
{(G_1,G_2) = p(G) & \gamma_1 = \bound(G_1) & \gamma_2 = \bound(G_2)}
\end{gather}

\subsubsection{Samples Partitions (TODO)}
I next present the basic sample partition.
\begin{gather}
\infer{\langle\twoGP(p, \gamma), \kappa, \Phi, G \rangle \to \bag(G)}{}
\\
\infer{\langle\sampP(G), S, t\rangle \to \langle \lambda(F).\bigcup F, \{(G, b, S_i, t_i) : b \isect t_i \}\rangle}{\{t_i\} = \mathit{split}(t)} 
\end{gather}

\subsection{Modifier and Combinator Nodes}
The following nodes do not directly represent decomposition, but instead denote ways to combine or modify the basic nodes.

\subsubsection{Branching}
\begin{gather}
\infer{\langle N +_p M, \kappa, \Phi, G^\gamma, S^\sigma\rangle \to \langle N, \kappa, \Phi, G^\gamma, S^\sigma \rangle}{p(G,S) = \mathrm{left}}
\\
\infer{\langle N +_p M, \kappa, \Phi, G^\gamma, S^\sigma\rangle \to \langle M, \kappa, \Phi, G^\gamma, S^\sigma \rangle}{p(G,S) = \mathrm{right}}
\end{gather}

\subsubsection{Composition (TODO)}
The composition of decompositions is denoted with a $\comp$. 
\begin{gather}
\infer{\langle N \comp M, \emptyset, \Phi, G^\gamma, S^\sigma \rangle \to \langle  N, M, \Phi, G^\gamma, S^\sigma \rangle}{\cdot}
\\
\infer{\langle N \comp M, \kappa, \Phi, G^\gamma, S^\sigma \rangle \to \langle  N, M \comp \kappa, \Phi, G^\gamma, S^\sigma \rangle}{ \kappa \not = \emptyset}
\end{gather}

\subsubsection{Temporary Decompositions}

forthcoming

\section{The Biphasic Renderer Model (ALL TODO)}

A renderer is entirely specified by a tree of source nodes. The renderer operates in two phases:
\begin{itemize}
\item {\bf Precompute Phase:} This phase takes the source nodes and a set of geometry and produces a structure which holds all of the geometry.
\item {\bf Evaluation Phase:} The input to the evaluation phase is the samples, as well as the structure that was build by the precomput phase. (Lazy BVH build here)
\end{itemize}

\subsection{Node Summary}

(NOTE: ignore the content of these tables; they are not being synced)
\\ \\
At a glance, here are all of the source nodes.
\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
Name & Node Arity & Special Arguments & Notes \\ \hline
$\id$ & 0 & - & -\\ \hline
$\sampP$ & 0 & $\sigma$ & -\\ \hline
$\twoGP$ & 0 & $p, \gamma$ & $\gamma$ is a bound type argument \\ \hline
$\oneS$ & 0 & - & -\\ \hline
$\comp$ & 2 & - & - \\ \hline
$\plus$ & 2 & $p, m$ & mode $m$ determine's $p$'s input \\ \hline
$\fix$ & 1 & $l$ & - \\ \hline
\end{tabular}
\end{center}

Here is a table of all of the geometry structure elements.  Obviously, elements like $\sleft$ and $\sright$ could be combined into a single node with a binary special argument, if one preferred that representation.

\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
Name & Node Arity & Special Arguments & Notes\\ \hline
$\bag$ & 0 & - & -\\ \hline
$\twoBox$ & 2 & $b_1 : \gamma$, $b_2 : \gamma$ & -\\ \hline
$\sleft$ & 1 & - &  - \\ \hline
$\sright$ & 1 & - &  - \\ \hline
$\plus$ & 2 & - & - \\ \hline
\end{tabular}
\end{center}

We also have some sample structure elements.

\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
Name & Node Arity & Special Arguments & Notes\\ \hline
$\bag$ & 0 & S & - \\ \hline
\end{tabular}
\end{center}

The input to the first phase

\subsection{Utility Nodes}

\subsubsection{Fixed Points}
Recursion requires two nodes: one to set a fixed-point label, and another to call it.  The nodes are called $\fix$ and $\call$, respectively.  The former is node-unary, the latter is node-nullary, and both require a single special argument, which is the name of the label in use.  Note that a label introduced at a $\fix$ node is only in scope in the subtree beneath that node.  Also, when drawing the tree graphically, $\call(l)$ is usually represented with just the label $l$.

Semantically, this behavior is handled by keeping about an environment variable $\Phi$, which maps labels to nodes.  The environment is untouched by all nodes except $\fix$ and $\call$.  In particular, $\fix$ introduces labels to the environment, and $\call$ reads labels from the environment.  Noting that $[\Phi, l:N]$ denotes the function $\Phi$ extended with the output $N$ at input $l$, we have:
\begin{gather}
\infer{\langle \fix(N,l), \kappa, \Phi, G \rangle \to\langle N, \kappa, [\Phi, l:N], G \rangle}{ \cdot }
\\
\infer{\langle \call(l), \kappa, \Phi, G \rangle \to \langle \Phi(l), \kappa, \Phi, G \rangle }{\cdot}
\end{gather}

Then during the evaluation phase, we have,
\begin{gather}
\infer{\langle \fix(N,l), \Phi, R, S \rangle \to \langle m, F \rangle}{\langle N, [\Phi, l:N], R, S \rangle \to \langle m, F \rangle}
\\
\infer{\langle \call(l), \Phi, R, S \rangle \to \langle m, F \rangle}{\langle \Phi(l), \Phi, R, S \rangle \to \langle m, F \rangle}
\end{gather}

\subsection{Decompositional Nodes}
In this subsection, I'll cover all of the nodes which perform decompositions.  All of these nodes are nullary, meaning that they do not take any arguments.

\subsubsection{Identity Node}
The identity node, noted $\id$, represents the trivial decomposition.
\begin{gather}
\infer[\text{(Precompute Phase)}] {\langle\id, \kappa, \Phi, G \rangle \to \langle \kappa, \emptyset, \Phi, G \rangle}{\cdot} \\
\infer[\text{(Eval Phase)}] {\langle\id, R, S \rangle \to \langle \cdot, [(R,S)] \rangle}{\cdot} 
\end{gather}

\subsubsection{Full Decomposition Nodes}
I next note the full decomposition nodes, $\oneS$ and $\oneG$.  These decompositions completely decompose their respective sets.  First we cover $\oneS$:
\begin{gather}
\infer{\langle\oneS, \kappa, \Phi, G \rangle \to \langle \kappa, \emptyset, \Phi, G \rangle}{\cdot}
\\
\infer{\langle\oneS, R, S\rangle \to \langle \lambda(F).\bigcup F, \{(R,s) : s \in S\} \rangle}{\cdot} 
\end{gather}
And now $\oneG$:
\begin{gather}
\infer{\langle\oneG, \emptyset, \Phi, G \rangle \to \bag(G)}{\cdot}
\\
\infer{\langle\oneG, \bag(G), S\rangle \to \langle \ldots, \{(g,S) : g \in G\} \rangle}{\cdot} 
\end{gather}

\subsubsection{Geometry Partitions}
I next present the basic geometry partition. In this case, the node partitions into two sets, but the generalization to more is obvious.  The parameter $\gamma$ identifies a module for producing bounds (for examples, an axis aligned bounding box, or half-space bounds, etc.).
\begin{gather}
\infer{\langle\twoGP(p, \gamma), \kappa, \Phi, G \rangle \to \twoBox(\langle \kappa, \emptyset, \Phi, G_1 \rangle,\langle \kappa, \emptyset, \Phi, G_2 \rangle,b_1,b_2)}
{(G_1,G_2) = p(G) & b_1 = \bound_\gamma(G_1) & b_2 = \bound_\gamma(G_2)}
\\
\infer{\twoBox(A,B,b_1,b_2) \to \twoBox(A',B,b_1,b_2)}{A \to A'}
\\
\infer{\twoBox(A,B,b_1,b_2) \to \twoBox(A,B',b_1,b_2)}{B \to B'}
\end{gather}
The resulting $\twoBox$ structure contains the bounds of the two children, and pointers to substructures ({\it i.e.}, one node in a BVH tree).  This kind of node assumes that its bounds have been checked by the parent structure, or that bounds checking is not important.

In the following nodes, the symbol $\isect$ means that two bounds intersect, and $\nisect$ means they do not intersect.
\begin{gather}
\infer{\langle\twoGP(p, \gamma),\twoBox(n_1,n_2,b_1,b_2), S, t\rangle \to \langle \{\}, \{\} \rangle}{b_1 \nisect t & b_2 \nisect t} 
\\
\infer{\langle\twoGP(p, \gamma),\twoBox(n_1,n_2,b_1,b_2), S, t\rangle \to \langle f, P \rangle}{b_1 \isect t & b_2 \nisect t & \langle n_1, S, t\rangle \to \langle f, P\rangle} 
\\
\infer{\langle\twoGP(p, \gamma),\twoBox(n_1,n_2,b_1,b_2), S, t\rangle \to \langle f, P \rangle}{b_1 \nisect t & b_2 \isect t & \langle n_2, S, t\rangle \to \langle f, P\rangle} 
\end{gather}
TODO: figure out how to carry around sample metadata
\subsubsection{Samples Partitions}
I next present the basic sample partition.
\begin{gather}
\infer{\langle\twoGP(p, \gamma), \kappa, \Phi, G \rangle \to \bag(G)}{}
\\
\infer{\langle\sampP(G), S, t\rangle \to \langle \lambda(F).\bigcup F, \{(G, b, S_i, t_i) : b \isect t_i \}\rangle}{\{t_i\} = \mathit{split}(t)} 
\end{gather}

\subsection{Modifier and Combinator Nodes}
The following nodes do not directly represent decomposition, but instead denote ways to combine or modify the basic nodes.

\subsubsection{Branching}
Sometimes, the choice of what decomposition you want to perform is dependent on data.  To express this situation, we have the branch operator.    The expression $N \plus_p^m M$ denotes a conditional choice between the nodes $N$ and $M$, based on predicate $p$, evaluating under one of three modes $m$ (geometry, sample-duplicate, and sample-defer).  The particular mode $m$ determines the type of $p$.  The first mode is for when the branch choice depends only on geometry, meaning that the decision can be made during the build phase.  Note the rightmost judgement above the bar means ``the node $N$, for samples $G$, will produce a constructor $f$ to operate on the results of the breakdown $C$".
\begin{gather}
\infer{\langle N +_p^m M, \kappa, \Phi, G \rangle \to \sleft(\langle N, \kappa, \Phi, G \rangle)}{m={\rm geom} & p(G) = \mathrm{left}}
\\
\infer{\langle N +_p^m M, \kappa, \Phi, G \rangle \to \sright(\langle M, \kappa, \Phi, G \rangle)}{m={\rm geom} & p(G) = \mathrm{right}}
\\
\infer{\sleft(A) \to \sleft(A')}{A \to A'}
\\
\infer{\sright(A) \to \sright(A')}{A \to A'}
\end{gather}
Next there is the case where the branch decision depends on samples and no action is taken during the build phase, probably because the user doesn't want to risk doing unnecessary decompositions.
\begin{gather}
\infer{\langle N +_p^m M, \kappa, \Phi, G \rangle \to \bag(G)}{m={\rm sDef}}
\end{gather}
Alternatively, the user may want to perform both sub-decompositions proactively.  This may be wasted work and wasted storage if no samples end up requiring the sub-expressions to be evaluated.  However, if it's likely that both will end up evaluated, we'd like to perform the appropriate precomputation.  The best example of this so far is Warren's frustrum rasterizer.  In this case, the duplicate represntation down both sides of the tree has little memory overhead because a BVH subtree can be represented as a partition of an array. 
\begin{gather}
\infer{\langle N +_p^m M, \kappa, \Phi, G \rangle \to \plus(\langle N, \kappa, \Phi, G \rangle, \langle M, \kappa, \Phi, G \rangle)}{m={\rm sDup}}
\\
\infer{\plus(A, B) \to \plus(A', B)}{A \to A'}
\\
\infer{\plus(A, B) \to \plus(A, B')}{B \to B'}
\end{gather}

\subsubsection{Composition}
The composition of decompositions is denoted with a $\comp$. Note that previously, I have used variables like $C_N$ to denote lists of sets of geometry that are output from the $N$ decomposition.  Below, I instead deconstruct $C_N$ into the list $[G_1,\ldots,G_n]$, because I need to pass each $G_i$ independently to the right-hand-side $M$ decomposition.
\begin{gather}
\infer{\langle N \comp M, \emptyset, \Phi, G \rangle \to \langle  N, M, \Phi, G \rangle}{\cdot}
\\
\infer{\langle N \comp M, \kappa, \Phi, G \rangle \to \langle  N, M \comp \kappa, \Phi, G \rangle}{ \kappa \not = \emptyset}
\end{gather}

\subsubsection{Temporary Decompositions}

forthcoming

\subsubsection{Phase-Adjustment Node}

forthcoming

\section{Example Renderers}

\subsection{Basic BVH Ray-Tracer}

\begin{center}
\Tree [. [.$\comp$ $\oneS$ [.VP [.V is ] [.NP fun ]]]]
\end{center}

\end{document}

### 2013-06-13 paper/renderers.tex
first commit for all the SML and haskell work

\documentclass{article}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{qtree}
\usepackage{amssymb,amsthm}
\usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
\parindent 0pt

\title{\Large\textbf{A Generative Representation of Sampling-Based Visibility Algorithms}}
\author{Nicolas Feltman et al.}
\begin{document}
\maketitle

\begin{center}
There has been considerable research activity in the field of sampling-based visibility algorithms for the past 25 years.  To achieve high performance, these algorithms have become very specialized to workloads and hardware domains.  [why is this a problem]. We present a novel representation of the space of sampling-based visibility algorithms using notions of decomposition, bounds formation, bounds testing, and map-reduction as primitives.  We also present a compiler capable of translating our representation to single-core code. 
\end{center}

\end{document}

### 2013-06-13 MockingbirdHaskell/src/Geometry.hs

@@ -1,0 +1,31 @@
+
+module Geometry where
+import MbVectors
+import MbUtilities
+
+type Tri3 = X3 Vec3
+type Seg3 = (Vec3, Vec3, Interval)
+type Plane3 = (Vec3, Vec3)
+type BBox3 = (Vec3, Vec3)
+
+triToPlane :: Tri3 -> Plane3
+triToPlane (X3 p1 p2 p3) = (p1, cross (vsub p2 p1) (vsub p3 p1))
+
+intersectSegPlane :: Seg3 -> Plane3 -> Maybe R
+intersectSegPlane (o,d,i) (p,n) = 
+    let denom = dot d n in
+    if denom == 0.0 then Nothing else
+        let t = (dot n (vsub p o)) / denom in
+        if contains i t then Just t else Nothing
+
+sideTest :: Vec3 -> Vec3 -> Vec3 -> Bool
+sideTest d v1 v2 = (>0.0).(dot d).(cross v1) $ (vsub v2 v1)
+
+lineTest :: Vec3 -> Tri3 -> Bool
+lineTest d tri = let X3 a b c = (cycleX3 (\ v1 v2 _ -> sideTest d v1 v2)) $ tri in a==b && b == c 
+        
+intersectSegTri :: Seg3 -> Tri3 -> Maybe R
+intersectSegTri seg@(o,d,_) tri = 
+    do  t <- intersectSegPlane seg (triToPlane tri)
+        if lineTest d (xzip3 vsub tri (X3 o o o)) then Just t else Nothing
+

### 2013-06-13 MockingbirdHaskell/src/Main.hs

@@ -1,0 +1,29 @@
+module Main where
+import MbSyntax
+import MbEntities
+import MbVectors
+import MbKeys
+import MbUtilities
+import Geometry
+import Data.List
+import Data.Vector.Storable
+import Codec.Picture
+import Codec.Picture.Png
+import qualified Data.ByteString as B
+
+
+fragsmentsToImage :: Int -> Int -> FragList t -> Image PixelRGB8
+fragsmentsToImage x y (KeyedList frags) = (Image x y).fromList $ frags >>= \(_,Depthed v) -> maybe [0, 0, 0] (\_->[255, 255, 255]) v
+
+--fragmentsToASCII :: Int -> Int -> [Frag] -> String
+--fragmentsToASCII x y = (concat $ (map $ (++"\n").(map $ maybe '.' (\_->'X')) )).(arrangeGrid x y)
+
+pixX = 300
+pixY = 250
+oneTri = X3 (vec3 0 1 1) (vec3 1 0 1) (vec3 0 0 1)
+rays = generateSamples (pi*0.66) 1.2 pixX pixY 
+tree = (MergeF2 General OneG (MergeF2 Union OneS (ChainGHF HitNode Shade)))
+frags = evalGS2F (Shaders (\_ -> ())) tree ([(oneTri,())],rays)
+
+main::IO()
+main = B.writeFile "output.png" $ encodePng $ fragsmentsToImage pixX pixY frags

### 2013-06-13 MockingbirdHaskell/src/MbEntities.hs

@@ -1,0 +1,49 @@
+{-# LANGUAGE DeriveFunctor #-}
+
+module MbEntities where
+import MbUtilities
+import MbVectors
+import MbKeys
+import Geometry
+import Data.List
+import Data.Monoid
+
+type Geom m = (Tri3,m)
+type GS m = ([Geom m], KeyedList Seg3)
+data Depthed t = Depthed (Maybe (Float, t)) deriving Functor
+type Hit m = Depthed (Geom m)
+type HitList m = KeyedList (Hit m)
+type FragList t = KeyedList (Depthed t)
+
+data Shaders m t = Shaders (Geom m -> t)
+
+
+instance Monoid (Depthed a) where
+    mempty = Depthed Nothing
+    mappend a (Depthed Nothing) = a
+    mappend (Depthed Nothing) b = b
+    mappend a@(Depthed (Just(da,_))) b@(Depthed (Just(db,_))) = if da < db then a else b
+
+
+hitDyn :: GS m -> (Int, Hit m)
+hitDyn (geom@(tri,_):[], KeyedList ((k,samp):[])) = (k, Depthed $ fmap (\d -> (d,geom)) $ intersectSegTri samp tri )
+hitDyn (_,_) = error "Hit node requires single sample and single geometry."
+
+shadeDyn :: Shaders m t -> HitList m -> (Int, Depthed t)
+shadeDyn (Shaders shadeFunc) (KeyedList ((k,Depthed h):[])) = (k, Depthed $ fmap (\(dep,g) -> (dep, shadeFunc g)) h)
+shadeDyn _ _ = error "Shade node requires a single hit."
+
+buildMatrix :: Int -> Int -> [(Int, Float, Float)]
+buildMatrix w h = 
+    do  y <- [0..(h-1)]
+        x <- [0..(w-1)] 
+        let fi = fromIntegral in [(x+y*w, (fi x)/(fi w-1)*2-1, (fi (h-1-y))/(fi h-1)*2-1)]
+
+generateSamples :: R -> R -> Int -> Int -> KeyedList Seg3
+generateSamples fovX aspectRatio pw ph = 
+    let w = tan(fovX/2.0) in
+    let h = w / aspectRatio in
+    let m = buildMatrix pw ph in
+    KeyedList [(k, (vec3 0.0 0.0 0.0, normalize (vec3 (x*w) (y*h) 1), positives)) | (k, x, y) <- m]
+
+                

### 2013-06-13 MockingbirdHaskell/src/MbKeys.hs

@@ -1,0 +1,37 @@
+{-# LANGUAGE DeriveFunctor #-}
+
+module MbKeys where
+import MbUtilities
+import Data.Monoid
+import Data.List
+import Data.Ord
+
+data KeyedList t = KeyedList { backing :: [(Int, t)]} deriving Functor
+
+-- MERGING
+
+pickKeyed :: Monoid t => (Int,t) -> (Int,t) -> (Int,t)
+pickKeyed (k1,d1) (k2,d2) = if k1 == k2 then (k1, d1 `mappend` d2) else error "Trying to merge things with different key."
+
+mergeByKey :: Monoid t => [KeyedList t] -> KeyedList t
+mergeByKey = (KeyedList).(foldl1 (angryZipWith pickKeyed)).(map backing)
+
+upwrap (KeyedList a) = a
+
+mergeDisjoint :: [KeyedList t] -> KeyedList t
+mergeDisjoint = (KeyedList).(sortBy $ comparing fst).concat.(map backing)
+
+
+-- GRID ARRANGEMENT
+                
+splitMaybe n xs@(_:_) = Just (splitAt n xs) 
+splitMaybe _ [] = Nothing
+
+chunkify :: Int -> [t] -> [[t]]
+chunkify n xs = unfoldr (splitMaybe n) xs
+
+arrangeGrid :: Int -> Int -> KeyedList t -> [[t]]
+arrangeGrid w h (KeyedList xs) = 
+    if length xs == w*h 
+    then (chunkify w).(map snd) $ sortBy (comparing fst) xs
+    else error "Wrong size."

### 2013-06-13 MockingbirdHaskell/src/MbSyntax.hs

@@ -1,0 +1,68 @@
+
+module MbSyntax where
+import MbEntities
+import MbUtilities
+import MbKeys
+import Data.Monoid
+
+data DecompGS = OneG | TwoGP | OneS | SixteenSqSP
+data DecompH = OneH
+
+data GeomBound = SingleG | BBboxG
+data SampleBound = SingleS | BBboxS
+data MergeImp = General | Union
+
+data GS2GS = Filt | ChainGGG GS2GS GS2GS
+-- data H2H
+-- data F2F
+
+data GS2H = MergeH MergeImp DecompGS GS2H | ChainGGH GS2GS GS2H | HitNode
+data H2F = MergeF1 MergeImp DecompH H2F | Shade
+data GS2F = MergeF2 MergeImp DecompGS GS2F | ChainGGF GS2GS GS2F | ChainGHF GS2H H2F
+
+
+-- DECOMPOSITION
+
+evalDecompGS :: DecompGS -> GS m -> [GS m]
+evalDecompGS OneG (g,s) = map (\e -> ([e],s)) g
+evalDecompGS TwoGP (g,s) = undefined
+evalDecompGS OneS (g, KeyedList s) = map (\e -> (g, KeyedList [e])) s
+evalDecompGS SixteenSqSP (g,s) = undefined
+
+evalDecompH :: DecompH -> HitList m -> [HitList m]
+evalDecompH OneH (KeyedList l) = map ((KeyedList).(:[])) l
+
+-- MERGING
+
+evalMerge :: MergeImp -> [KeyedList (Depthed t)] -> KeyedList (Depthed t)
+evalMerge General = mergeByKey
+evalMerge Union = mergeDisjoint
+
+
+-- HELPERS
+
+chainHelp :: (a -> Maybe b) -> (a -> c) -> (b -> c) -> a -> c
+chainHelp f l r a = maybe (l a) r (f a)
+
+allMiss :: KeyedList t -> KeyedList (Depthed u)
+allMiss = fmap (\_ -> mempty)
+
+-- ALL EVALUATORS
+
+evalGS2GS :: GS2GS -> GS m -> Maybe (GS m)
+evalGS2GS Filt gst = undefined -- if f gst then Just gst else Nothing
+evalGS2GS (ChainGGG lhs rhs) gst = evalGS2GS lhs gst >>= (evalGS2GS rhs) 
+
[…]

### 2013-06-13 MockingbirdHaskell/src/MbUtilities.hs

@@ -1,0 +1,26 @@
+ {-# LANGUAGE DeriveFunctor #-}
+
+module MbUtilities where
+
+
+infF :: Float
+infF = 1.0/0.0
+
+angryZipWith :: (a -> b -> c) -> [a] -> [b] -> [c]
+angryZipWith f [] [] = []
+angryZipWith f [] (_:_) = error "First list shorter than second."
+angryZipWith f (_:_) [] = error "Second list shorter than first."
+angryZipWith f (x:xs) (y:ys) = (f x y):(angryZipWith f xs ys)
+
+-- X3 stuff
+
+data X3 t = X3 !t !t !t deriving (Functor, Show)
+
+cycleX3 :: (a -> a -> a-> b) -> X3 a -> X3 b
+cycleX3 f (X3 x y z) =  X3 (f x y z) (f y z x) (f z x y)
+
+xzip3 :: (a -> b -> c) -> X3 a -> X3 b -> X3 c
+xzip3 f (X3 x y z) (X3 a b c) = X3 (f x a) (f y b) (f z c)
+
+x3ToList :: X3 a -> [a]
+x3ToList (X3 a b c) = [a,b,c]

### 2013-06-13 MockingbirdHaskell/src/MbVectors.hs

@@ -1,0 +1,53 @@
+
+module MbVectors where
+import MbUtilities
+
+type R = Float
+
+-- VECTORS
+
+type Vec3 = X3 R
+
+vec3 :: Float -> Float -> Float -> Vec3
+vec3 x y z = X3 x y z
+
+vneg :: Vec3 -> Vec3
+vneg = fmap (0-)
+
+vadd :: Vec3 -> Vec3 -> Vec3
+vadd = xzip3 (+)
+
+vsub :: Vec3 -> Vec3 -> Vec3
+vsub = xzip3 (-)
+
+vmul :: Vec3 -> R -> Vec3
+vmul v c = fmap (*c) v
+
+vdiv :: Vec3 -> R -> Vec3
+vdiv v c = fmap (/c) v
+
+dot :: Vec3 -> Vec3 -> R
+dot (X3 a b c) (X3 x y z) = a*x + b*y + c*z
+
+normSq :: Vec3 -> R
+normSq v = dot v v
+
+norm :: Vec3 -> R
+norm = sqrt.normSq
+
+normalize :: Vec3 -> Vec3
+normalize v = vdiv v (norm v)
+
+cross :: Vec3 -> Vec3 -> Vec3
+cross (X3 a b c) (X3 x y z) = X3 (b*z-c*y) (c*x-z*a) (a*y-b*x)
+
+-- INTERVALS
+
+type Interval = (R,R)
+data EInterval = Empty | NonEmpty Interval
+
+contains :: Interval -> R -> Bool
+contains (s,e) r = s <= r && r <= e
+
+positives :: Interval
+positives = (0.0, infF)

### 2013-06-13 MockingbirdHaskell/src/ObjParser.hs

@@ -1,0 +1,9 @@
+
+-- module ObjParser where
+-- import Data.List.Split
+-- import Geometry
+
+-- parseObj :: String -> [Tri3]
+-- parseObj = 
+    -- (foldl accum ([],[])).(splitOn "\n") where
+        -- accum (vs,fs) line = if line == "" then 

### 2013-06-13 MockingbirdML/src/runtests.py

@@ -1,0 +1,38 @@
+import os
+
+testrange = [
+(1,30),
+(2,10),
+(3,50),
+]
+
+print "COMPILING ALL RENDERERS"
+for (i,_) in testrange :
+	s = str(i)
+	outfile = "gen"+s+".sml"
+	makefile = "tests/make"+s+".cm"
+	
+	os.system("sml @SMLload=bin/compiler-image.x86-win32 "+s+" tests/"+outfile)
+	
+	f = open(makefile, 'w')
+	cmContents = "Group is \n\t"+outfile+"\n\t../runtime/runtime.cm\n"
+	f.write(cmContents)
+	f.close()
+
+print "LINKING ALL RENDERERS"
+os.chdir("tests") # safer to avoid path seperators
+os.environ["CM_VERBOSE"] = "false"
+for (i,_) in testrange :
+	print "Linking renderer "+str(i)+"..."
+	makefile = "make"+str(i)+".cm"
+	executable = "heap"+str(i)
+	cmd = "ml-build.bat "+makefile+" GeneratedRenderer.render "+executable
+	os.system(cmd)
+
+print "RUNNING ALL RENDERERS"
+for (i,p) in testrange :
+	print "Running renderer "+str(i)+"..."
+	s = str(i)
+	outfile = "pic"+s+".ppm"
+	cmd = "sml @SMLload=heap"+s+".x86-win32 ../../models/teapot.obj "+str(p)+" "+outfile
+	os.system(cmd)

### 2013-06-14 MockingbirdML/src/runtests.py
Stripped out the keys.  Repeat-work raytracer works now.

@@ -1,9 +1,10 @@
 import os
 
 testrange = [
-(1,30),
-(2,10),
-(3,50),
+#(1,30),
+#(2,10),
+#(3,50),
+(5,30),
 ]
 
 print "COMPILING ALL RENDERERS"

### 2013-08-30 MockingbirdML/src/runtests.py
Old changes.  Not sure what they are.

@@ -7,7 +7,7 @@ testrange = [
 #(2,10),
 #(3,50),
 #(5,30),
-(6,20),
+(6,150),
 ]
 
 print "COMPILING ALL RENDERERS"

### 2013-12-10 paper/paper.tex
changed the name of the paper files

\documentclass{article}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{qtree}
\usepackage{amssymb,amsthm}
\usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
\parindent 0pt

\title{\Large\textbf{General Automatic Stage Separation with Applications to Algorithm Derivation}}
\author{Nicolas Feltman et al.}
\begin{document}
\maketitle
\section{Introduction}
Often when programming high-performance systems, a programmer will wish to specialize a bivariate function to one of its inputs.  This need may arise because one input varies at a higher frequency than the other, and we wish to avoid repeated calculation. It may also be that one input is available at a time when computational resources are less expensive.  It's even possible that knowledge of an input could allow further code optimizations that would have otherwise been unwise to speculate.  Broadly, the solutions to this specialization problem can be called {\it staging}, in that different parts of the code are intended to run at different points in time, {\it i.e.} stages.  The general theme is to do some work with the {\em immediate input}, producing an intermediate result, and then finish the rest of the computation once the {\em delayed input} is available.  In particular, these staging techniques include:
\begin{itemize}
\item {\bf Program Generation} A common solution is to generate code in some target language that itself will take in the second stage input.  Using the equational notation of [Jones 96], we say that a programmer defines $\mathit{genf}$ in a language $C$ so that \[[[\mathit{genf}]_C~x]_T~y = [f]_L~(x,y)\] for all $x$ and $y$. Examples of this technique include most compilers (where $f$ is a hypothetical interpretter, and the first stage input is the source program), or even specialized compilers like Spiral (where the first stage is simply the dimension of the DFT).  The benefit of this technique is full control, and the fact that implementations of the target language $T$ may be inherently faster than those of $L$.  The drawbacks of this technique are that it is essentially a manual operation, and that correctness of $\mathit{genf}$ with respect to $\mathit{f}$ is difficult to prove.
\item {\bf Partial Evaluation} The idea of partial evaluation is to automatically specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.  In particular, this means defining a program in language $C$ called $pe$, which we call a partial evaluator for language $L$.  This means that we can pass to $pe$ the implementation of a function $f$ along with $f$'s immediate input $x$, and get back a version of $f$ that is {\em specialized} to $x$.  This specialized version, also called a {\em residual}, is often denoted as $f_x$.  Equationally, we have
\begin{align*}
[pe]_C~(f,x) &= f_x \\
[f_x]_L~y &= [f]_L~(x,y)
\end{align*}
 for all $f$, $x$, and $y$.  As seminally observed by [Futamura 71], a partial evaluator and interpretter can be used together to achieve the same result as manual code generation, automatically!  Using the $\mathtt{fftw}$ example, we see that $\mathtt{dft} = pe~(\mathtt{fft},x)$, where $x$ is the dimension of the dft.  This has the benefit of being an automatic process, although the generated compiler is only as good as the partial evaluator.  Much of the subsequent work in partial evaluators has been with the purpose of chasing this goal.
\item {\bf Stage Separation} The final technique, stage separation, is conceptually the simplest.  The idea is to define in $C$ a program called $ps$, which is a path separator for language $L$. It works by cleaving the $L$-function $f$ into two $L$-functions $f_1$ and $f_2$, where $f_1$ builds a data structure from the available input, and $f_2$ consumes that data structure as well as the delayed input to produce the standard ouput.  Equationally, this is
\begin{align*}
[pe]_C~f &= (f_1,f_2) \\
[f_2]_L~([f_1]_L~x,y) &= [f]_L~(x,y)
\end{align*}
for all $f$, $x$, and $y$.  This technique, performed manually, is a common exercise for every programmer.  In can be considered the general form of many common compiler optimizations such as loop hoisting, an example we will consider later.  The key difference between stage separation and partial evaluation is that the former is performed in absence of the first input.  This suggests that stage seperation is applicable in more scenarios than partial evaluation, at the cost of being less aggressive.  Also, whereas stage seperation creates a data structure to represent the intermediate progress of a computation, partial evaluation usually embeds that datastructure in the code of the residual [some related work says it can alleviate this].  For these reasons, stage splitting should be thought of as a mechanism for {\em algorithmic} optimization, rather than {\em systematic} optimization.
\end{itemize}
In this work we present a stage separation algorithm for a fragment of ML.  Specifically, our formulation can separate terms containing sums and recursion, which have not appeared previously in the stage seperation literature.  We anticipate that the algorithm will also be able to also separate this language extended with with first-class functions, although this is left as future work.  Additionally, our presentation is more explicitly type-motivated than those that have come before.

\section{Binding Time Analysis}
Previous research has observed that the stage separation problem, as defined above, is inherently ambiguous (indeed, this is true of all the staging techniques).  That is, there can exist multiple ways to partition $f$ into $f_1$ and $f_2$ that satisfy $f_2(f_1(x),y) = f(x,y)$.  For a trivial example, we can set $f_1$ to be the identity and $f_2$ to be $f$.  We can resolve this ambiguity by requiring that $f$ be given to us in a form which unambiguously specifies the part of the computation to be considered first-stage, and the part of the computation to be considered second-stage.  The particular form that we employ, introduced later, is suggested by [Pfenning and Davies 2001], which explores the connection between staging phenomena and modal type systems.  That work also shows the equivalence of the form we use and others, such as the two-level languages of [nielson$^2$].

The process of adding staging annotations to an otherwise ambiguous computation is called {\it binding time analysis}.  We do not explore binding time analysis in this work, and instead assume that complete annotations are given by the programmer or by some analyzer.

\section{$\lambda^{12}$}

(Introduce the language, talk about its type system and maybe its evaluation.)

\section{Examples}

Some examples to get us comfortable with the language.

\subsection{loop hoisting}

\section{Stage Seperaion Algorithm}

Talk about the goal of the transformation.  Introduce the general form with open variables and a context, which is {\em not} the simplistic $f_1$, $f_2$ thing.  Stress the first image/second image/boundary type stuff.

\subsection {Type translation}

\subsection{Term translation}

State the main theorems we want to be true.  Refer to the

\subsection{Functions}

Talk about why the current set up for functions alleviates the boundary-type guessing problem.

\subsection{Issue with sums}

Sums, as done, are bad.

\section {Implementation}

Not much to discuss.  It exsists.  Talk about how crufty the naive transformation is.

\section{Related Work in Stage Seperation}
Stage seperation was introduced in [Jorring and Scherlis 86], then called ``pass seperation."  They gave an algorithm that simultaneously performed a binding time analysis and stage seperation for a recursive first-order language with if statements and primitive operations (essentially, the core of Java).

[Hannan 94] gives an algorithm for simultaneous binding time analysis and stage seperation for a class of languages called "abstract machines".

[Knoblock and Ruf 96] does BTA and PS for non-recursive imperative shader programs.  They pay a lot of attention to the size of the intermediate data structure, since memory is at a premium in their shaders.

\end{document}

### 2013-12-11 paper/paper.tex
Made lots of additions.  Related work still needs work.

@@
 \usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
-\parindent 0pt
 
-\title{\Large\textbf{General Automatic Stage Separation with Applications to Algorithm Derivation}}
+\title{\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}}
 \author{Nicolas Feltman et al.}
@@
 \section{Introduction}
-Often when programming high-performance systems, a programmer will wish to specialize a bivariate function to one of its inputs.
-This need may arise because one input varies at a higher frequency than the other, and we wish to avoid repeated calculation.
-It may also be that one input is available at a time when computational resources are less expensive.
-It's even possible that knowledge of an input could allow further code optimizations that would have otherwise been unwise to speculate.
-Broadly, the solutions to this specialization problem can be called {\it staging}, in that different parts of the code are intended to run at different points in time, {\it i.e.} stages.
-The general theme is to do some work with the {\em immediate input}, producing an intermediate result, and then finish the rest of the computation once the {\em delayed input} is available.
-In particular, these staging techniques include:
-\begin{itemize}
-\item {\bf Program Generation} A common solution is to generate code in some target language that itself will take in the second stage input.
-Using the equational notation of [Jones 96], we say that a programmer defines $\mathit{genf}$ in a language $C$ so that \[[[\mathit{genf}]_C~x]_T~y = [f]_L~(x,y)\] for all $x$ and $y$.
-Examples of this technique include most compilers (where $f$ is a hypothetical interpretter, and the first stage input is the source program), or even specialized compilers like Spiral (where the first stage is simply the dimension of the DFT).
-The benefit of this technique is full control, and the fact that implementations of the target language $T$ may be inherently faster than those of $L$.
-The drawbacks of this technique are that it is essentially a manual operation, and that correctness of $\mathit{genf}$ with respect to $\mathit{f}$ is difficult to prove.
-\item {\bf Partial Evaluation} The idea of partial evaluation is to automatically specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.
-In particular, this means defining a program in language $C$ called $pe$, which we call a partial evaluator for language $L$.
-This means that we can pass to $pe$ the implementation of a function $f$ along with $f$'s immediate input $x$, and get back a version of $f$ that is {\em specialized} to $x$.
+Often when programming high-performance systems, a programmer will wish to specialize a bivariate function $f : \alpha \times \beta \to \gamma$ to one of its inputs, $x : \alpha$, while leaving the other, $y : \beta$, for later.
+This need may arise because $x$ varies at a higher frequency than $y$, and we would wish to avoid repeated calculation.
+It may also be the case that $x$ is available at a time when computational resources are less expensive than when $y$ becomes available.
+It's even possible that knowledge of $x$ allows further code optimizations that would have otherwise been unwise to speculate.
+
+Consider two examples: the discrete Fourier transform (DFT) and a raytracer.
+In the DFT case we say that $x : \mathbb{N}$ is the dimension of the transformation, $y : \mathbb{R}^x $ is the data to be transformed, and $f$ simply computes the DFT.
+Much of the structure of modern DFT algorithms depends on $x$ (for both mathematical decomposition and cache performance reasons) but not on $y$, so a great deal of work could be done to specialize the algorithm to a particular number of dimensions.
+In the raytracer case, we say that $x : \mathtt{Geom}$ is a specification of the scene geometry (perhaps a list of triangles), $y : \mathtt{Ray}$ is a ray cast from some point in the scene, and $f$ computes the first intersection of the ray and the geometry.
+In modern raytracers, we may cast millions of rays without varying the scene geometry, so it could be advantageous to precompute as much of the algorithm as possible based on only scene geometry.
+
+We say that the problems above exhibit a {\em staged} structure, in that they can be broken down into parts that run at different points in time ({\em i.e.}~stages).
+There are a variety of techniques to take advantage of this staged structure.
+The general theme is to do some work with the immedaite input ($x$), producing an intermediate result, and then finish the rest of the computation once the delayed input ($y$) is available.
+We consider three techniques: program generation, partial evaluation, and pass separation.
+Each technique has broader applicability than the previous, but is less aggressive of an optimization.
+
+\subsection {Program Generation} 
+A common solution is to write code that writes code.
+Essentially, second-stage computations are represented by second-stage code, and second-stage code is a value that can be passed around and manipulated by first-stage code.
+To use the terminology from above: this first stage code is a function that takes in $x$ and outputs second stage function, which itself takes in $y$ and outputs the final answer.
+Note that this is essentially a manual process, in that the programmer must know the function $f$ which is being specialized.
+Using the equational notation of [Jones 96], where ``$\llbracket h \rrbracket~z$'' means the ``the code $h$ interpreted as a function and applied to input $z$," we say that a programmer defines a program generator $g$ that satisfies
+\[
+\llbracket \llbracket g \rrbracket~x\rrbracket~y =\llbracket f \rrbracket~(x,y)
+\] 
+for all $x$ and $y$.
+When well supported by the language (see terra, metaml) this technique is often called {\em metaprogramming}.
+Of course, even without first class support, one can represent the second stage code using strings or other data types.
+Defined so broadly, program generation includes all compilers, where $f$ is a hypothetical interpretter, $g$ is the actual compiler, $x$ is a program in the source language, and $y$ is the input to $x$.
+Program generation has also been used to implement the DFT example in the form of FFTW [cite this].
+
+The benefit of the program generation technique is full control, even up to the ability to use domain-specific optimizations.  This comes at the cost of essentially being a manual operation.
+
+\subsection {Partial Evaluation} 
+The idea of partial evaluation is to {\em automatically} specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.
+In particular this means defining a program $pe$, called a partial evaluator, to which we can pass the implementation of a function $f$ along with $f$'s immediate input $x$, and get back a version of $f$ that is {\em specialized} to $x$.
 This specialized version, also called a {\em residual}, is often denoted as $f_x$.
@@
 \begin{align*}
-[pe]_C~(f,x) &= f_x \\
-[f_x]_L~y &= [f]_L~(x,y)
+\llbracket pe\rrbracket~(f,x) &= f_x \\
+\llbracket f_x\rrbracket~y &= \llbracket f\rrbracket~(x,y)
 \end{align*}
  for all $f$, $x$, and $y$.
-As seminally observed by [Futamura 71], a partial evaluator and interpretter can be used together to achieve the same result as manual code generation, automatically!
-Using the $\mathtt{fftw}$ example, we see that $\mathtt{dft} = pe~(\mathtt{fft},x)$, where $x$ is the dimension of the dft.
-This has the benefit of being an automatic process, although the generated compiler is only as good as the partial evaluator.
+As seminally observed by [Futamura 71], partially evaluating an interpretter on a source program is equivalent to compiling that program to whatever language the interpretter was written in!
+This has the benefit of being an automatic process, although the effective compiler is only as good as the partial evaluator.
 Much of the subsequent work in partial evaluators has been with the purpose of chasing this goal.
-\item {\bf Stage Separation} The final technique, stage separation, is conceptually the simplest.
-The idea is to define in $C$ a program called $ps$, which is a path separator for language $L$.
-It works by cleaving the $L$-function $f$ into two $L$-functions $f_1$ and $f_2$, where $f_1$ builds a data structure from the available input, and $f_2$ consumes that data structure as well as the delayed input to produce the standard ouput.
+
+Partial evaluation is a more general technique than program generation, in that the partial evaluator is defined once and works for all $f$s that need to be specialized.
+Of course, the partial evaluator has no domain knowledge of the program it is splitting, so it cannot be nearly as aggressive as a program generator might be.
+The two techniques are similar in that they both require the first input $x$ to do their specializing action.
+
+\subsection {Pass Separation} 
+The final technique, pass separation, is conceptually the simplest.
+The idea is to define a program called $ps$, for {\em pass separator}.
+It works by cleaving the function $f:\alpha \times \beta \to \gamma$ into two functions $f_1 : \alpha \to \tau$ and $f_2 : \beta \times \tau \to \gamma$ for some type $\tau$, where $f_1$ builds a data structure (of type $\tau$) from $x$, and $f_2$ consumes that data structure as well as $y$ to produce the standard ouput.
+In analogy to partial evaluation, we call $f_2$ the {\em residual}.
 Equationally, this is
 \begin{align*}
-[pe]_C~f &= (f_1,f_2) \\
-[f_2]_L~([f_1]_L~x,y) &= [f]_L~(x,y)
+\llbracket pe\rrbracket~f &= (f_1,f_2) \\
+\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) &= \llbracket f\rrbracket~(x,y)
 \end{align*}
@@
 This technique, performed manually, is a common exercise for every programmer.
-In can be considered the general form of many common compiler optimizations such as loop hoisting, an example we will consider later.
-The key difference between stage separation and partial evaluation is that the former is performed in absence of the first input.
-This suggests that stage seperation is applicable in more scenarios than partial evaluation, at the cost of being less aggressive.
-Also, whereas stage seperation creates a data structure to represent the intermediate progress of a computation, partial evaluation usually embeds that datastructure in the code of the residual [some related work says it can alleviate this].
-For these reasons, stage splitting should be thought of as a mechanism for {\em algorithmic} optimization, rather than {\em systematic} optimization.
-\end{itemize}
-In this work we present a stage separation algorithm for a fragment of ML.
+In can be considered the general form of common compiler optimizations such as loop hoisting, an example considered later.
+Like partial evaluation, pass separation is an automatic operation, but unlike both partial evaluation and program generation, pass separation does not require the value of the first argument, $x$, to properly specialize $f$.
+This makes pass separation the most widely applicable of the staging techniques, at the cost of being able to perform aggressive optimizations in the residual that might depend on the first argument.
+
+In this work we present a pass separation algorithm for a fragment of ML.
 Specifically, our formulation can separate terms containing sums and recursion, which have not appeared previously in the stage seperation literature.
@@
 
-\section{Binding Time Analysis}
-Previous research has observed that the stage separation problem, as defined above, is inherently ambiguous (indeed, this is true of all the staging techniques).
-That is, there can exist multiple ways to partition $f$ into $f_1$ and $f_2$ that satisfy $f_2(f_1(x),y) = f(x,y)$.
+\section{Binding-Time Analysis and Stage Untangling}
+
+The pass separation problem can naturally be decomposed into two subproblems:
+\begin{itemize}
+\item Decide which parts of the definition of $f$ belong to which stage and produce an annotated version of $f$.
+\item Use the annotated version of $f$ to produce $f_1$ and $f_2$.  
+\end{itemize}
+
+An analogous form of this decomposition exists in the partial evaluation literature, wherein the first subproblem is known as {\em binding-time analysis}.
+The second subproblem, in which the difference between pass separation and partial evaluation is more manifest, does not seem to yet have a name.
+We call it {\em stage untangling}.
+Unlike previous work in pass separation, we focuses entirely on the stage untangling problem by assuming that $f$ is given to us in an annotated form.
+
+Previous research has observed that the pass separation problem is inherently ambiguous in that the equations defined above do not fully specify the definition of $pe$.
+That is, there can exist multiple ways to partition $f$ into $f_1$ and $f_2$ that satisfy $\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) = \llbracket f\rrbracket~(x,y)$.
 For a trivial example, we can set $f_1$ to be the identity and $f_2$ to be $f$.
-We can resolve this ambiguity by requiring that $f$ be given to us in a form which unambiguously specifies the part of the computation to be considered first-stage, and the part of the computation to be considered second-stage.
-The particular form that we employ, introduced later, is suggested by [Pfenning and Davies 2001], which explores the connection between staging phenomena and modal type systems.
-That work also shows the equivalence of the form we use and others, such as the two-level languages of [nielson$^2$].
-
-The process of adding staging annotations to an otherwise ambiguous computation is called {\it binding time analysis}.
-We do not explore binding time analysis in this work, and instead assume that complete annotations are given by the programmer or by some analyzer.
+Fortunately, this ambiguity is contained entirely within the binding-time analysis portion of pass separation; stage untangling is entirely determined.
 
@@
 \section{Related Work in Stage Seperation}
-Stage seperation was introduced in [Jorring and Scherlis 86], then called ``pass seperation."  They gave an algorithm that simultaneously performed a binding time analysis and stage seperation for a recursive first-order language with if statements and primitive operations (essentially, the core of Java).
-
-[Hannan 94] gives an algorithm for simultaneous binding time analysis and stage seperation for a class of languages called "abstract machines".
+Pass seperation was introduced in [Jorring and Scherlis 86].
+They motivating pass separation as a technique using hand-separated examples, butstopped short of providing an algorithm to do this automatically.
 
@@
 
+They gave an algorithm that simultaneously performed a binding time analysis and stage seperation for a recursive first-order language with if statements and primitive operations (essentially, the core of Java).
+
+[Hannan 94] gives an algorithm for simultaneous binding time analysis and stage seperation for a class of languages called "abstract machines".
+
 \end{document}

### 2013-12-12 paper/paper.tex
beefed up related work; added a bibliography

@@
 \usepackage{amssymb,amsthm}
+\usepackage{cite}
 \usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
@@
 Note that this is essentially a manual process, in that the programmer must know the function $f$ which is being specialized.
-Using the equational notation of [Jones 96], where ``$\llbracket h \rrbracket~z$'' means the ``the code $h$ interpreted as a function and applied to input $z$," we say that a programmer defines a program generator $g$ that satisfies
+Using the equational notation of \cite{jones96}, where ``$\llbracket h \rrbracket~z$'' means the ``the code $h$ interpreted as a function and applied to input $z$," we say that a programmer defines a program generator $g$ that satisfies
 \[
@@
 for all $x$ and $y$.
-When well supported by the language (see terra, metaml) this technique is often called {\em metaprogramming}.
+When well supported by the language (\cite{devito13}, [some metaml citation]) this technique is often called {\em metaprogramming}.
 Of course, even without first class support, one can represent the second stage code using strings or other data types.
 Defined so broadly, program generation includes all compilers, where $f$ is a hypothetical interpretter, $g$ is the actual compiler, $x$ is a program in the source language, and $y$ is the input to $x$.
-Program generation has also been used to implement the DFT example in the form of FFTW [cite this].
+Program generation has also been used to implement the DFT example in the form of FFTW \cite{FFTW05}.
 
@@
  for all $f$, $x$, and $y$.
-As seminally observed by [Futamura 71], partially evaluating an interpretter on a source program is equivalent to compiling that program to whatever language the interpretter was written in!
+As seminally observed by \cite{futamura71}, partially evaluating an interpretter on a source program is equivalent to compiling that program to whatever language the interpretter was written in!
 This has the benefit of being an automatic process, although the effective compiler is only as good as the partial evaluator.
@@
 
-Sums, as done, are bad.
+Sums result in dependant types, which is cool.
 
@@
 \section{Related Work in Stage Seperation}
-Pass seperation was introduced in [Jorring and Scherlis 86].
-They motivating pass separation as a technique using hand-separated examples, butstopped short of providing an algorithm to do this automatically.
 
-[Knoblock and Ruf 96] does BTA and PS for non-recursive imperative shader programs.
-They pay a lot of attention to the size of the intermediate data structure, since memory is at a premium in their shaders.
+Our starting language, $\lambda^{12}$, is essentially a two-stage version of $\lambda^\bigcirc$ from \cite{davies96}, but beefed up with products, sums, and fixed-points.
+That work motivates $\lambda^\bigcirc$ as an image of linear temporal logic under the Curry-Howard correspondence, and it makes the argument that $\lambda^\bigcirc$ is a good model for binding time analysis.
+Although $\lambda^\bigcirc$ was designed with partial evaluation in mind, it turns out to be a good model for pass separation as well.
+In contrast, \cite{davies01} presents $\lambda^\Box$, which is based off of branching temporal logic.
+As that work shows, $\lambda^\Box$ is a good system for program generation, although we find the closed-code requirement to be more restricting than we need.
 
-They gave an algorithm that simultaneously performed a binding time analysis and stage seperation for a recursive first-order language with if statements and primitive operations (essentially, the core of Java).
+Pass seperation was introduced in \cite{jorring}.
+They focused on motivating pass separation as a technique for compiler generation, and hinted at wider applicability.
+Their approach was not mechanized or automated in any sense, and their examples were separated by hand.
+We believe that out work represents the fulfillment of their prediction that ``the [pass separation] approach will elude full automation for some time."
 
-[Hannan 94] gives an algorithm for simultaneous binding time analysis and stage seperation for a class of languages called "abstract machines".
+\cite{hannan94} also uses a pass separation technique for generating compiler/evaluator pairs from interpretters specified as term-rewrite systems.
+This heavy restriction on the form of the input prevents generalization of their style of pass separation.
+[I don't know enough about term-rewrite systems yet to be able to comment further.]  Their method, while amenable to mechanization, was not actually implemented.
+
+\cite{knoblock} represents the first attempt at fully automatic pass separation for code approaching general-purpose.
+In particular, they implemented pass separation for a C-like shading language including basic arithmetic and if statements.
+The main goal of their work, like ours, was to minimize recomputation.
+That said, their starting language was sufficiently restricted for recursive boundary types to be inexpressible.
+Given that memory is at a premium in shading languages, so much of that work was also focused on minimizing the memory-footprint of the boundary type.
+
+\bibliography{paper}
+\bibliographystyle{plain}
 

### 2014-02-05 paper/paper.tex
Redid intro.  Added erasure semantics.  Added splitting.

@@
 \documentclass{article}
-\usepackage{tikz}
 \usepackage{amsmath}
+\usepackage{bbm}
 \usepackage{stmaryrd}
 \usepackage{proof}
-\usepackage{qtree}
 \usepackage{amssymb,amsthm}
 \usepackage{cite}
+\usepackage{mathpartir}
 \usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
+
+% core conventions and language names
+\newcommand {\bbone} {\mathbbm 1}
+\newcommand {\bbtwo} {\mathbbm 2}
+\newcommand {\corelang} {$\mathrm{L}^{\bbone\bbtwo}$}
+\newcommand {\lang} {$\lambda^{\bbone\bbtwo}$}
+\newcommand {\coremono} {L}
+\newcommand {\mono} {$\lambda$}
+
+% bnf stuff
+\newcommand {\myit} [1]{\operatorname{\it{#1}}}
+\newcommand {\stage} {\langle\mathit{stage}\rangle}
+\newcommand {\type} {\langle\mathit{type}\rangle}
+\newcommand {\typeo} {\langle\bbone\text{-}\mathit{type}\rangle}
+\newcommand {\typet} {\langle\bbtwo\text{-}\mathit{type}\rangle}
+\newcommand {\expr} {\langle\mathit{exp}\rangle}
+\newcommand {\expro} {\langle\bbone\text{-}\mathit{exp}\rangle}
+\newcommand {\exprt} {\langle\bbtwo\text{-}\mathit{exp}\rangle}
+\newcommand {\val} {\langle\mathit{val}\rangle}
+\newcommand {\valo} {\langle\bbone\text{-}\mathit{val}\rangle}
+\newcommand {\valt} {\langle\bbtwo\text{-}\mathit{val}\rangle}
+\newcommand {\var} {\langle\mathit{var}\rangle}
+\newcommand {\context} {\langle\mathit{cont}\rangle}
+\newcommand {\inte} {\langle\mathit{int}\rangle}
+\newcommand {\resi} {\langle\mathit{res}\rangle}
+\newcommand {\gbar} {~~|~~}
+
+% nodes
+\newcommand {\pause} {{\tt hold}}
+\newcommand {\next} {{\tt next}}
+\newcommand {\prev} {{\tt prev}}
+\newcommand {\fut} {\bigcirc}
+\newcommand {\transfers} {\nearrow}
+\newcommand {\gds} {{\Gamma \vdash^\sigma}}
+\newcommand {\gdo} {{\Gamma \vdash^\bbone}}
+\newcommand {\gdt} {{\Gamma \vdash^\bbtwo}}
+\newcommand {\letin} [3] {{\tt let}~{#1} = {#2}~{\tt in}~{#3}}
+\newcommand {\caseof} [3] {{\tt case}~{#1} ~{\tt of}~{#2}~{\tt |}~{#3}}
+\newcommand {\tallcase} [3] {\begin{array}{l} {\tt case}~{#1} ~{\tt of}\\~~{#2}\\{\tt |}~{#3} \end{array}}
+\newcommand {\ifthen} [3] {{\tt if}~{#1} ~{\tt then}~{#2}~{\tt else}~{#3}}
+\newcommand {\tallif} [3] {\begin{array}{l} {\tt if}~{#1} \\{\tt then}~{#2}\\{\tt else}~{#3} \end{array}}
+\newcommand {\lam} [3] {\lambda{#1}\mathrm{:}{#2}.{#3}}
+\newcommand {\valprod} [2] {{\bf (}{#1},{#2}{\bf )}}
+\newcommand {\valleft} {{\bf \iota}_1}
+\newcommand {\valright} {{\bf \iota}_2}
+\newcommand {\vallam} [2] {{\bf \lambda}{#1}.{#2}}
+\newcommand {\valdum} {{\bf dummy}}
+
+%judgements
+\newcommand {\splito} {\overset{\bbone}\rightsquigarrow}
+\newcommand {\splits} {\overset{\bbtwo}\rightsquigarrow}
 
@@
 \section{Introduction}
+
+Many computational tasks are premised on the notion of multiple domains, places and times at which computation occurs.
+Yet programming languages traditionally view their execution environments as monolithic, specifying the What and the How, but not the When nor the Where.
+As always, where the language falls short, a programmer instead must bear the burden.
+Inevitably, this comes at the cost of expressivity, safety, or both.
+In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expresivity nor the safety that programmers have come to expect in single-domain languages.
+
+Consider the programming of a pipeline system.
+Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reording in between.
+We can view an algorithm for such as system as one computation, split accross two spatial domains (processors).
+For a concrete example, we draw inspiration from graphics systems.
+Lets say that we want to map a function $f : \mathrm{PixelCoord} \to \mathrm{Color}$ over every pixel on a 1024x768 screen.
+Due to geometric coherence inherent in the problem, it's often advantageous to compute some information at the level of coarse 32x32 tiles (e.g.
+a list of the scene primitives which interact with that part of the screen) whose results can be used to accelerate the final per-pixel calculation.
+In adition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.
+
 Often when programming high-performance systems, a programmer will wish to specialize a bivariate function $f : \alpha \times \beta \to \gamma$ to one of its inputs, $x : \alpha$, while leaving the other, $y : \beta$, for later.
@@
 It may also be the case that $x$ is available at a time when computational resources are less expensive than when $y$ becomes available.
-It's even possible that knowledge of $x$ allows further code optimizations that would have otherwise been unwise to speculate.
-
-Consider two examples: the discrete Fourier transform (DFT) and a raytracer.
-In the DFT case we say that $x : \mathbb{N}$ is the dimension of the transformation, $y : \mathbb{R}^x $ is the data to be transformed, and $f$ simply computes the DFT.
-Much of the structure of modern DFT algorithms depends on $x$ (for both mathematical decomposition and cache performance reasons) but not on $y$, so a great deal of work could be done to specialize the algorithm to a particular number of dimensions.
-In the raytracer case, we say that $x : \mathtt{Geom}$ is a specification of the scene geometry (perhaps a list of triangles), $y : \mathtt{Ray}$ is a ray cast from some point in the scene, and $f$ computes the first intersection of the ray and the geometry.
+Here we say that the calculation is split accross two temporal domains.
+Pulling another example from graphics, we consider a raytracer wherein $x : \mathtt{Geom}$ is a specification of the scene geometry (perhaps a list of triangles), $y : \mathtt{Ray}$ is a ray cast from some point in the scene, and $f$ computes the first intersection of the ray and the geometry.
 In modern raytracers, we may cast millions of rays without varying the scene geometry, so it could be advantageous to precompute as much of the algorithm as possible based on only scene geometry.
 
-We say that the problems above exhibit a {\em staged} structure, in that they can be broken down into parts that run at different points in time ({\em i.e.}~stages).
-There are a variety of techniques to take advantage of this staged structure.
-The general theme is to do some work with the immedaite input ($x$), producing an intermediate result, and then finish the rest of the computation once the delayed input ($y$) is available.
-We consider three techniques: program generation, partial evaluation, and pass separation.
-Each technique has broader applicability than the previous, but is less aggressive of an optimization.
+In addition to the partitioning of a calculation across multiple domains, the examples above exhibit one further similarity: communication between domains is one-directional.
+In the spatial case, information always flows from one processor to the next, but never backwards.
+In the temporal case too, information made available at the second time cannot be observed earlier (indeed this seems to be a defining feature of time, generally).
+For this reason, we refer to the domains as stages.
+This property, that information flow is acyclic, informs the rest of this paper.
 
-\subsection {Program Generation} 
-A common solution is to write code that writes code.
+[Talk about how languages without inherent support for domains must instead tackle the problem by writing various passes/shaders.
+This is bad because it's not compositional.
+For instance, I cannot write a function that talks about multiple stages at once.]
+
+[Because of the acyclic nature of the problem, we can take algorithms specified in our multi-domain language and tease them apart, essentially recovering the old way of doing things.
+This is called stage untangling.
+It gets us all the safety and expression without messing up the rest of the compiler.]
+
+[Talk about standard partial evaluation and pass separation, then place our work in the second, modulo binding time analysis.]
+
+%\subsection {Program Generation} 
+%A common solution is to write code that writes code.
 Essentially, second-stage computations are represented by second-stage code, and second-stage code is a value that can be passed around and manipulated by first-stage code.
@@
 Using the equational notation of \cite{jones96}, where ``$\llbracket h \rrbracket~z$'' means the ``the code $h$ interpreted as a function and applied to input $z$," we say that a programmer defines a program generator $g$ that satisfies
-\[
-\llbracket \llbracket g \rrbracket~x\rrbracket~y =\llbracket f \rrbracket~(x,y)
-\] 
-for all $x$ and $y$.
+%\[
+%\llbracket \llbracket g \rrbracket~x\rrbracket~y =\llbracket f \rrbracket~(x,y)
+%\] 
+%for all $x$ and $y$.
 When well supported by the language (\cite{devito13}, [some metaml citation]) this technique is often called {\em metaprogramming}.
@@
 Program generation has also been used to implement the DFT example in the form of FFTW \cite{FFTW05}.
-
-The benefit of the program generation technique is full control, even up to the ability to use domain-specific optimizations.  This comes at the cost of essentially being a manual operation.
-
-\subsection {Partial Evaluation} 
-The idea of partial evaluation is to {\em automatically} specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.
+%
+%The benefit of the program generation technique is full control, even up to the ability to use domain-specific optimizations.  This comes at the cost of essentially being a manual operation.  
+%
+%\subsection {Partial Evaluation} 
+%The idea of partial evaluation is to {\em automatically} specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.
 In particular this means defining a program $pe$, called a partial evaluator, to which we can pass the implementation of a function $f$ along with $f$'s immediate input $x$, and get back a version of $f$ that is {\em specialized} to $x$.
@@
 Equationally, we have
-\begin{align*}
-\llbracket pe\rrbracket~(f,x) &= f_x \\
-\llbracket f_x\rrbracket~y &= \llbracket f\rrbracket~(x,y)
-\end{align*}
- for all $f$, $x$, and $y$.
+%\begin{align*}
+%\llbracket pe\rrbracket~(f,x) &= f_x \\
+%\llbracket f_x\rrbracket~y &= \llbracket f\rrbracket~(x,y)
+%\end{align*}
+% for all $f$, $x$, and $y$.
 As seminally observed by \cite{futamura71}, partially evaluating an interpretter on a source program is equivalent to compiling that program to whatever language the interpretter was written in!
@@
 Much of the subsequent work in partial evaluators has been with the purpose of chasing this goal.
-
-Partial evaluation is a more general technique than program generation, in that the partial evaluator is defined once and works for all $f$s that need to be specialized.
+%
+%Partial evaluation is a more general technique than program generation, in that the partial evaluator is defined once and works for all $f$s that need to be specialized.
 Of course, the partial evaluator has no domain knowledge of the program it is splitting, so it cannot be nearly as aggressive as a program generator might be.
 The two techniques are similar in that they both require the first input $x$ to do their specializing action.
-
-\subsection {Pass Separation} 
-The final technique, pass separation, is conceptually the simplest.
+%
+%\subsection {Pass Separation} 
+%The final technique, pass separation, is conceptually the simplest.
 The idea is to define a program called $ps$, for {\em pass separator}.
@@
 Equationally, this is
-\begin{align*}
-\llbracket pe\rrbracket~f &= (f_1,f_2) \\
-\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) &= \llbracket f\rrbracket~(x,y)
-\end{align*}
-for all $f$, $x$, and $y$.
+%\begin{align*}
+%\llbracket pe\rrbracket~f &= (f_1,f_2) \\
+%\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) &= \llbracket f\rrbracket~(x,y)
+%\end{align*}
+%for all $f$, $x$, and $y$.
 This technique, performed manually, is a common exercise for every programmer.
@@
 This makes pass separation the most widely applicable of the staging techniques, at the cost of being able to perform aggressive optimizations in the residual that might depend on the first argument.
-
-In this work we present a pass separation algorithm for a fragment of ML.
+%
+%In this work we present a pass separation algorithm for a fragment of ML.
 Specifically, our formulation can separate terms containing sums and recursion, which have not appeared previously in the stage seperation literature.
@@
 Additionally, our presentation is more explicitly type-motivated than those that have come before.
-
-\section{Binding-Time Analysis and Stage Untangling}
-
-The pass separation problem can naturally be decomposed into two subproblems:
-\begin{itemize}
-\item Decide which parts of the definition of $f$ belong to which stage and produce an annotated version of $f$.
-\item Use the annotated version of $f$ to produce $f_1$ and $f_2$.  
-\end{itemize}
-
-An analogous form of this decomposition exists in the partial evaluation literature, wherein the first subproblem is known as {\em binding-time analysis}.
+%
+%\section{Binding-Time Analysis and Stage Untangling}
+%
+%The pass separation problem can naturally be decomposed into two subproblems:
+%\begin{itemize}
+%\item Decide which parts of the definition of $f$ belong to which stage and produce an annotated version of $f$.
+%\item Use the annotated version of $f$ to produce $f_1$ and $f_2$.  
+%\end{itemize}
+%
+%An analogous form of this decomposition exists in the partial evaluation literature, wherein the first subproblem is known as {\em binding-time analysis}.
 The second subproblem, in which the difference between pass separation and partial evaluation is more manifest, does not seem to yet have a name.
@@
 Unlike previous work in pass separation, we focuses entirely on the stage untangling problem by assuming that $f$ is given to us in an annotated form.
-
-Previous research has observed that the pass separation problem is inherently ambiguous in that the equations defined above do not fully specify the definition of $pe$.
+%
+%Previous research has observed that the pass separation problem is inherently ambiguous in that the equations defined above do not fully specify the definition of $pe$.
 That is, there can exist multiple ways to partition $f$ into $f_1$ and $f_2$ that satisfy $\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) = \llbracket f\rrbracket~(x,y)$.
@@
 
-\section{$\lambda^{12}$}
+\section{A Simple Two-Stage Language}
 
-(Introduce the language, talk about its type system and maybe its evaluation.)
+We'll start by analyzing a simple two-stage language called \corelang.
+We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.
+Notably, we leave out functions, fixed point operators, and full sum types.
 
-\section{Examples}
+The grammar for \corelang is given in Figure \ref{fig:coreSyntax}
 
-Some examples to get us comfortable with the language.
+\begin{figure}
+\caption{\corelang~Syntax}
+\label{fig:coreSyntax}
+\centering
+\begin{tabular}{ll} 
+$\begin{aligned}
+\typeo &::= \text{unit}~|~\text{int}~|~\text{bool} \\
+&\gbar \typeo \times \typeo \\
+&\gbar \fut \typet 
+\end{aligned} $  
+& 
+$\begin{aligned}
+\typet &::=  \text{unit}~|~\text{int}~|~\text{bool} \\
+&\gbar \typet \times \typet \\
+&
+\end{aligned} $  
+\\ 
+$\begin{aligned}
+\expro &::= ()~|~\inte~|~\mathtt{true}~|~\mathtt{false}  \\
+&\gbar \letin{\var}{\expro}{\expro} \\
+&\gbar \var \\
+&\gbar (\expro, \expro) \\
+&\gbar \pi_1~\expro \gbar \pi_2~\expro \\
+&\gbar \ifthen {\expro}{\expro}{\expro} \\
+&\gbar \next~\exprt 
+\end{aligned} $ 
+& 
+$\begin{aligned}
+\exprt &::= ()~|~\inte~|~\mathtt{true}~|~\mathtt{false} \\
+&\gbar \letin{\var}{\exprt}{\exprt} \\
+&\gbar \var \\
+&\gbar (\exprt, \exprt) \\
+&\gbar \pi_1~\exprt \gbar \pi_2~\exprt \\
+&\gbar \ifthen {\exprt}{\exprt}{\exprt} \\
+&\gbar \prev~\expro
+\end{aligned} $
+\\ 
+$\begin{aligned}
+\valo &::= \valdum \\
+&\gbar ()~|~\inte~|~\mathbf{true}~|~\mathbf{false} \\
+&\gbar \valprod {\valo} {\valo}
+\end{aligned} $  
+& 
+$\begin{aligned}
+\context &::= \mathrm{empty} \\
+&\gbar \context, \var : \typeo ^\bbone \\
+&\gbar \context, \var : \typet ^\bbtwo
+\end{aligned} $
+\end{tabular}
+\end{figure}
 
-\subsection{loop hoisting}
+\subsection{Static Semantics}
 
-\section{Stage Seperaion Algorithm}
+Introduce the type system.  We got it from \cite{davies96}.
 
-Talk about the goal of the transformation.
-Introduce the general form with open variables and a context, which is {\em not} the simplistic $f_1$, $f_2$ thing.
-Stress the first image/second image/boundary type stuff.
+\begin{figure}
+\caption{\corelang~Static Semantics}
+\label{fig:coreStatics}
+\begin{mathpar}
+\infer [\mathrm{unit}] {\gds () : \text{unit}}{\cdot} \and
+\infer [\mathrm{int}] {\gds i : \text{int}}{\cdot} \and
+\infer [\mathrm{bool_1}] {\gds \mathtt{true} : \text{bool}}{\cdot} \and
+\infer [\mathrm{bool_2}] {\gds \mathtt{false} : \text{bool}}{\cdot} \and
+\infer [\mathrm{let}] {\letin{x}{e_1}{e_2} : B}{\gds e_1 : A & \Gamma,x:A^\sigma \vdash e_2 : B} \and
+\infer [\mathrm{hyp}] {\gds x : A}{x : A^\sigma \in \Gamma} \and
+\infer [\times\mathrm{I}] {\gds (e_1,e_2) : A\times B}{\gds e_1 : A & \gds e_2 : B} \and
+\infer [\times\mathrm{E_1}] {\gds \pi_1~e : A}{\gds e : A\times B} \and
+\infer [\times\mathrm{E_2}] {\gds \pi_2~e : B}{\gds e : A\times B} \and
+\infer [\mathrm{if}] {\gds \ifthen{e_1}{e_2}{e_3} : A}{\gds e_1 : \mathrm{bool} & \gds e_2 : A & \gds e_3 : A} \and
+\infer [\fut\mathrm{I}] {\Gamma \vdash^\bbone \next~e : \fut A}{\Gamma \vdash^\bbtwo e : A} \and
+\infer [\fut\mathrm{E}] {\Gamma \vdash^\bbtwo \prev~e : A}{\Gamma \vdash^\bbone e : \fut A} \and
+\end{mathpar}
+\end{figure}
 
-\subsection {Type translation}
+\subsection{Erasure Dynamic Semantics}
 
-\subsection{Term translation}
+Check Figure \ref{fig:erasureSemantics}
 
-State the main theorems we want to be true.  Refer to the
+\newcommand{\downer}{\Downarrow}
 
-\subsection{Functions}
+\begin{figure}
+\caption{\corelang~Erasure Dynamic Semantics}
+\label{fig:erasureSemantics}
+\begin{mathpar}
+\infer [\mathrm{unit}] { () \downer {\bf ()}}{\cdot} \and
+\infer [\mathrm{int}] {i \downer i}{\cdot} \and
+\infer [\mathrm{bool_1}] {\mathtt{true} \downer {\bf true}}{\cdot} \and
+\infer [\mathrm{bool_2}] {\mathtt{false}\downer {\bf false}}{\cdot} \and
+\infer [\mathrm{let}] {\letin{x}{e_1}{e_2} \downer v_2}{e_1 \downer v_1 & [v_1/x]e_2 \downer v_2} \and
+\infer [\times\mathrm{I}] {(e_1,e_2) \downer (v_1,v_2)}{e_1 \downer v_1 & e_2 \downer v_2} \and
+\infer [\times\mathrm{E_1}] {\pi_1~e \downer v_1}{e \downer (v_1,v_2)} \and
+\infer [\times\mathrm{E_2}] {\pi_2~e \downer v_2}{e\downer (v_1,v_2)} \and
+\infer [\mathrm{if_1}] {\ifthen{e_1}{e_2}{e_3} \downer v}{e_1 \downer {\bf true} & e_2 \downer v} \and
+\infer [\mathrm{if_2}] {\ifthen{e_1}{e_2}{e_3} \downer v}{e_1 \downer {\bf false} & e_3 \downer v} \\
+\infer [\fut\mathrm{I}] {\next~e \downer v}{e \downer v} \and
+\infer [\fut\mathrm{E}] {\prev~e \downer v}{e \downer v} 
+\end{mathpar}
+\end{figure}
 
-Talk about why the current set up for functions alleviates the boundary-type guessing problem.
+%\subsection{Better Dynamic Semantics}
+%
+%Introduce the dynamics semantics.  Figure \ref{fig:coreDynamics}.  It's small step.  We figured this out on our own, but it really should exist in the partial evaluation literature somewhere.
+%
+%\newcommand {\stepo} {\hookrightarrow_\bbone} 
+%\newcommand {\stept} {\hookrightarrow_\bbtwo} 
+%\newcommand {\pco} {~\mathrm{pc}\bbone} 
+%\newcommand {\fco} {~\mathrm{fc}\bbone} 
+%\newcommand {\pct} {~\mathrm{pc}\bbtwo} 
+%\newcommand {\fct} {~\mathrm{fc}\bbtwo} 
+%
+%\begin{figure}
+%\caption{\corelang~Partially Computed Values (pc$\bbone$ and pc$\bbtwo$)}
+%\label{fig:coreDynamics}
+%\begin{mathpar}
+%\infer {() \pco}{\cdot} \and
+%\infer {i \pco}{\cdot} \and
+%\infer {b \pco}{\cdot} \and
+%\infer {(v_1, v_2) \pco}{v_1 \pco & v_2 \pco} \and
+%\infer {\next~e \pco}{e \pct} \\
+%\infer {() \pct}{\cdot} \and
+%\infer {i \pct}{\cdot} \and
+%\infer {b \pct}{\cdot} \and
+%\infer {(v_1, v_2) \pct}{v_1 \pct & v_2 \pct} \and
+%\infer {\pi_i e \pct}{e \pct} \and
+%\infer {x \pct}{\cdot} \and
+%\infer {\letin{x}{e_1}{e_2} \pct}{e_1 \pct & e_2 \pct} \and
+%\infer {\ifthen{e_1}{e_2}{e_3} \pct}{e_1 \pct & e_2 \pct & e_3 \pct} \and
+%\infer {\prev~e \pct}{e \pco} 
+%\end{mathpar}
+%\end{figure}
+%
+%\begin{figure}
+%\caption{\corelang~Fully Computed Values (fc$\bbone$ and fc$\bbtwo$)}
+%\label{fig:coreDynamics}
+%\begin{mathpar}
+%\infer {() \fco}{\cdot} \and
+%\infer {i \fco}{\cdot} \and
+%\infer {b \fco}{\cdot} \and
+%\infer {(v_1, v_2) \fco}{v_1 \fco & v_2 \fco} \and
+%\infer {\next~e \fco}{e \fct} \\
+%\infer {() \fct}{\cdot} \and
+%\infer {i \fct}{\cdot} \and
+%\infer {b \fct}{\cdot} \and
+%\infer {(v_1, v_2) \fct}{v_1 \fct & v_2 \fct} 
+%\end{mathpar}
+%\end{figure}
+%
+%\begin{figure}
+%\caption{\corelang~Transitions}
+%\label{fig:coreDynamics}
+%\begin{mathpar}
+%\infer {\ifthen{e_1}{e_2}{e_3} \stepo \ifthen{e_1'}{e_2}{e_3}}{e_1 \stepo e_1'}  \\
+%\infer {\ifthen{\tt true}{e_2}{e_3} \stepo e_2}{\cdot} \and 
+%\infer {\ifthen{\tt false}{e_2}{e_3} \stepo e_3}{\cdot}  \\
+%\infer {\ifthen{e_1}{e_2}{e_3} \stept \ifthen{e_1'}{e_2}{e_3}}{e_1 \stept e_1' & e_2 \pct & e_3 \pct}  \\
+%\infer {\ifthen{\tt true}{e_2}{e_3} \stept e_2}{e_2 \pct & e_3 \pct} \and 
+%\infer {\ifthen{\tt false}{e_2}{e_3} \stept e_3}{e_2 \pct & e_3 \pct} \\
+%\infer {\letin{x}{e_1}{e_2} \stepo \letin{x}{e_1'}{e_2}}{e_1 \stepo e_1'} \and
+%\infer {\letin{x}{e_1}{e_2} \stepo [e_1/x]e_2}{e_1 \fco} \and
+%\infer {\letin{x}{e_1}{e_2} \stept \letin{x}{e_1'}{e_2}}{e_1 \stept e_1'} \and
+%\infer {\letin{x}{e_1}{e_2} \stept [e_1/x]e_2}{e_1 \fct} \and
+%\infer {x \downtwo x}{\cdot} \and
+%\infer {\gds (e_1,e_2) : A\times B}{\gds e_1 : A & \gds e_2 : B} \and
+%\infer {\ifthen{e_1}{e_2}{e_3} \downone [v,r]}{e_1 \downone {\bf true} & e_2 \downone [v,r]} \and
+%\infer {\ifthen{e_1}{e_2}{e_3} \downone [v,r]}{e_1 \downone {\bf false} & e_3 \downone [v,r]} \and
+%\infer {\ifthen{e_1}{e_2}{e_3} \downtwo [v, \ifthen{r_1}{r_2}{r_3}]}{e_1 \downtwo [\valdum,r_1] & e_2 \downtwo [v_2,r_2] & e_3 \downtwo [v_3,r_3]} \and
+%\infer {\next~e \downone [\valdum,r]}{e \downtwo r} \and
+%\infer {\prev~e \downtwo r}{e \downone [\valdum, r]} \and
+%\end{mathpar}
+%\end{figure}
 
-\subsection{Issue with sums}
+%\subsection{Good Theorems About the Language Alone}
+%
+%\begin{enumerate}
+%\item Type Safety.
+%\item Unidirectional Flow. [Essentially, you cannot write a program that goes back in time.]
+%\item Erasure property.  [If you ignore the nexts and prevs, the result is weakly equivalent.] [This was proven for a denotational semantics in some Jones paper that I need to find...]
+%\end{enumerate}
 
-Sums result in dependant types, which is cool.
+\subsection{Splitting}
+
+Introduce \coremono, in figure \ref{fig:baseSyntax}.  We're gonna split into two terms of that.
+
+\begin{figure}
+\caption{\coremono~Syntax}
+\label{fig:baseSyntax}
+\centering
+\begin{tabular}{ll} 
+$\begin{aligned}
+\expr &::= ()~|~\inte~|~\mathtt{true}~|~\mathtt{false} \\
+&\gbar \letin{\var}{\expr}{\expr} \\
+&\gbar \var \\
+&\gbar (\expr, \expr) \\
+&\gbar \pi_1~\expr \gbar \pi_2~\expr \\
+&\gbar \ifthen {\expr}{\expr}{\expr} 
+\end{aligned} $
+& 
+$\begin{aligned}
+\type &::=  \text{unit}~|~\text{int}~|~\text{bool} \\
+&\gbar \type \times \type 
+\\
+\val &::= ()~|~\inte~|~\mathbf{true}~|~\mathbf{false} \\
+&\gbar \valprod {\val} {\val} 
+\\
+\context &::= \mathrm{empty} \\
+&\gbar \context, \var : \type
+\end{aligned} $
+\end{tabular}
+\end{figure}
+
+\begin{figure*}
+\caption{Splitting}
+\label{fig:split}
+\begin{mathpar}
+\infer [\mathrm{unit}\splito] 
+	{\gdo () : \text{unit} \splito [((),()),\_.()]}{\cdot} \and
+\infer [\mathrm{unit}\splits]
+	 {\gdt () : \text{unit} \splits [(),\_.()]}{\cdot} \and
+\infer [\times\mathrm{I}\splito] 
+	{\gdo (e_1,e_2) : A\times B \splito [((\pi_1 c, \pi_1 c),(\pi_2 c_1, \pi_2 c_2)), l.(\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})] }
+	{ \gdo e_1 : A \splito [c_1,l_1.r_1] 
+	&\gdo e_2 : B \splito [c_2,l_2.r_2]} \and
+\infer [\times\mathrm{I}\splits] 
+	{\gdt (e_1,e_2) : A\times B \splits [(p_1,p_2), l.(\letin{l_1}{\pi_1~l}{r_1},\letin{l_2}{\pi_2~l}{r_2})] }
+	{ \gdt e_1 : A \splits [p_1,l_1.r_1] 
+	&\gdt e_2 : B \splits [p_2,l_2.r_2]} \and
+\infer [\times\mathrm{E_1}\splito] 
+	{\gdo \pi_1~e : A \splito [(\pi_1(\pi_1c),\pi_2 c),l.\pi_1~r] }{\gdo e : A\times B \splito [c,l.r] } \and
+\infer [\times\mathrm{E_1}\splits] 
+	{\gdt \pi_1~e : A \splits [p,l.\pi_1~r] }{\gdt e : A\times B \splits [p,l.r] } \and
+\infer [\times\mathrm{E_2}\splito] 
+	{\gdo \pi_2~e : B \splito [(\pi_2(\pi_1c),\pi_2 c),l.\pi_2~r] }{\gdo e : A\times B \splito [c,l.r]} \and
+\infer [\times\mathrm{E_2}\splits] 
+	{\gdt \pi_2~e : B \splits [p,l.\pi_2~r] }{\gdt e : A\times B \splits [p,l.r]} \and
+\infer [\fut\mathrm{I}\splito] {\gdo \next~e : \fut A \splito [((),p),l.r]}{\gdt e : A\splits [p,l.r] } \and
+\infer [\fut\mathrm{E}\splits] {\gdt \prev~e : A\splits [\pi_2~c,l.r] }{\gdo e : \fut A \splito [c,l.r]} \and
+\infer [+\mathrm{E}\splito]
+	{\gdo \left( \tallif {e_1}{e_2}{e_3} \right) :A \splito 
+	\left[
+		\begin{array}{l}
+		\left(\tallif {\pi_1c_1}
+			{\letin{y}{c_2}{(\pi_1 y, \iota_1(\pi_2 y))}}
+			{\letin{y}{c_3}{(\pi_1 y, \iota_2(\pi_2 y))}}
+		\right), \\
+		l.\left(\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3}
+		\right)
+		\end{array}
+	\right]}
+	{\gdo e_1 : {\rm bool} \splito [c_1, l_1.r_1]
+	& \gdo e_2 : A \splito [c_2,l_2.r_2]
+	& \gdo e_3 : A \splito [c_3,l_3.r_3]} \and
+%\infer [+\mathrm{E}\splito]
+%	{\gdo \left( \tallif {e_1}{e_2}{e_3} \right) :A \splito 
+%	\left[
+%		\begin{array}{l}
+%		\letin{c}{c_1}{\left(\tallif {\pi_1c}
+%			{\letin{y}{c_2}{(\pi_1 y, (\iota_1(\pi_2 y), \pi_2 c))}}
+%			{\letin{y}{c_3}{(\pi_1 y, (\iota_2(\pi_2 y), \pi_2 c))}}
+%		\right)}, \\
+%		l.\left(\letin{c_1}{\pi_2 l}{r_1}; 
+%			\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3}
+%		\right)
+%		\end{array}
+%	\right]}
+%	{\gdo e_1 : {\rm bool} \splito [c_1, l_1.r_1]
+%	& \gdo e_2 : A \splito [c_2,l_2.r_2]
+%	& \gdo e_3 : A \splito [c_3,l_3.r_3]} \and
+\infer [+\mathrm{E}\splits]
+	{\gdt \left( \tallif {e_1}{e_2}{e_3}\right) : A \splits 
+	\left[(p_1,p_2,p_3), 
+	l.\left(\tallif{(\letin {l_1}{\pi_1 l}{r_1})}{(\letin {l_2}{\pi_2 l}{r_2})} {(\letin {l_3}{\pi_3 l}{r_3})} \right)
+	\right] }
+	{ \gdt e_1 : {\rm bool} \splits [p_1,l_1.r_1]
+	& \gdt e_2 : A \splits [p_2,l_2.r_2] 
+	& \gdt e_3 : A \splits [p_3,l_3.r_3]} 
+\end{mathpar}
+\end{figure*}
+
+\subsection {Splitting Theorems}
+
+\begin{enumerate}
+\item It always pops out a well typed term.
+\item Meaning is preserved, up to some translation.
+\end{enumerate}
 

### 2014-02-14 paper/paper.tex
added value splitting
Pulled out macros.  Refactored judgements.  Added words.

@@
 \usepackage{proof}
-\usepackage{amssymb,amsthm}
+\usepackage{amssymb}
+\usepackage{mathabx}
 \usepackage{cite}
@@
 
-% core conventions and language names
-\newcommand {\bbone} {\mathbbm 1}
-\newcommand {\bbtwo} {\mathbbm 2}
-\newcommand {\corelang} {$\mathrm{L}^{\bbone\bbtwo}$}
-\newcommand {\lang} {$\lambda^{\bbone\bbtwo}$}
-\newcommand {\coremono} {L}
-\newcommand {\mono} {$\lambda$}
-
-% bnf stuff
-\newcommand {\myit} [1]{\operatorname{\it{#1}}}
-\newcommand {\stage} {\langle\mathit{stage}\rangle}
-\newcommand {\type} {\langle\mathit{type}\rangle}
-\newcommand {\typeo} {\langle\bbone\text{-}\mathit{type}\rangle}
-\newcommand {\typet} {\langle\bbtwo\text{-}\mathit{type}\rangle}
-\newcommand {\expr} {\langle\mathit{exp}\rangle}
-\newcommand {\expro} {\langle\bbone\text{-}\mathit{exp}\rangle}
-\newcommand {\exprt} {\langle\bbtwo\text{-}\mathit{exp}\rangle}
-\newcommand {\val} {\langle\mathit{val}\rangle}
-\newcommand {\valo} {\langle\bbone\text{-}\mathit{val}\rangle}
-\newcommand {\valt} {\langle\bbtwo\text{-}\mathit{val}\rangle}
-\newcommand {\var} {\langle\mathit{var}\rangle}
-\newcommand {\context} {\langle\mathit{cont}\rangle}
-\newcommand {\inte} {\langle\mathit{int}\rangle}
-\newcommand {\resi} {\langle\mathit{res}\rangle}
-\newcommand {\gbar} {~~|~~}
-
-% nodes
-\newcommand {\pause} {{\tt hold}}
-\newcommand {\next} {{\tt next}}
-\newcommand {\prev} {{\tt prev}}
-\newcommand {\fut} {\bigcirc}
-\newcommand {\transfers} {\nearrow}
-\newcommand {\gds} {{\Gamma \vdash^\sigma}}
-\newcommand {\gdo} {{\Gamma \vdash^\bbone}}
-\newcommand {\gdt} {{\Gamma \vdash^\bbtwo}}
-\newcommand {\letin} [3] {{\tt let}~{#1} = {#2}~{\tt in}~{#3}}
-\newcommand {\caseof} [3] {{\tt case}~{#1} ~{\tt of}~{#2}~{\tt |}~{#3}}
-\newcommand {\tallcase} [3] {\begin{array}{l} {\tt case}~{#1} ~{\tt of}\\~~{#2}\\{\tt |}~{#3} \end{array}}
-\newcommand {\ifthen} [3] {{\tt if}~{#1} ~{\tt then}~{#2}~{\tt else}~{#3}}
-\newcommand {\tallif} [3] {\begin{array}{l} {\tt if}~{#1} \\{\tt then}~{#2}\\{\tt else}~{#3} \end{array}}
-\newcommand {\lam} [3] {\lambda{#1}\mathrm{:}{#2}.{#3}}
-\newcommand {\valprod} [2] {{\bf (}{#1},{#2}{\bf )}}
-\newcommand {\valleft} {{\bf \iota}_1}
-\newcommand {\valright} {{\bf \iota}_2}
-\newcommand {\vallam} [2] {{\bf \lambda}{#1}.{#2}}
-\newcommand {\valdum} {{\bf dummy}}
-
-%judgements
-\newcommand {\splito} {\overset{\bbone}\rightsquigarrow}
-\newcommand {\splits} {\overset{\bbtwo}\rightsquigarrow}
+\input {macros}
 
@@
 We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.
-Notably, we leave out functions, fixed point operators, and full sum types.
+Notably, we leave out functions, fixed point operators, and full sum types, which will be added later.
 
-The grammar for \corelang is given in Figure \ref{fig:coreSyntax}
+The grammar for types, terms, values, and contexts in \corelang are given in Figure \ref{fig:coreSyntax}.
+The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.
+It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.
+Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.
+Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.
+The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.
+The reasons for this will become clear later.
+
+At the type and value levels, stage $\bbone$ can refer to stage $\bbtwo$, but stage $\bbtwo$ cannot refer back to stage $\bbone$.
+Alternatively, the two stages are mutually recursive at the term level.
+These two patterns together fulfill the promise of {\em mixed code, staged evalutation}.
+
+In the rest of this document, I will tend to use $\{A,B,C\}$ for types, $\{v,u,w,t\}$ for values, $\{x,y\}$ for variables, $e$ for terms, $i$ for integers, and $b$ for booleans.
 
@@
 $\begin{aligned}
-\expro &::= ()~|~\inte~|~\mathtt{true}~|~\mathtt{false}  \\
+\expro &::= ()~|~\inte~|~\bool  \\
 &\gbar \letin{\var}{\expro}{\expro} \\
@@
 $\begin{aligned}
-\exprt &::= ()~|~\inte~|~\mathtt{true}~|~\mathtt{false} \\
+\exprt &::= ()~|~\inte~|~\bool \\
 &\gbar \letin{\var}{\exprt}{\exprt} \\
@@
 $\begin{aligned}
-\valo &::= \valdum \\
-&\gbar ()~|~\inte~|~\mathbf{true}~|~\mathbf{false} \\
-&\gbar \valprod {\valo} {\valo}
+\valo &::= ()~|~\inte~|~\bool \\
+&\gbar \valprod {\valo} {\valo} \\
+&\gbar \valnext~\valt
 \end{aligned} $
@@
 $\begin{aligned}
-\context &::= \mathrm{empty} \\
-&\gbar \context, \var : \typeo ^\bbone \\
-&\gbar \context, \var : \typet ^\bbtwo
+\valt &::= ()~|~\inte~|~\bool \\
+&\gbar \valprod {\valt} {\valt} \\
+& 
+\end{aligned} $
+\\
+$\begin{aligned}
+\contextot &::=\emptyC \\
+&\gbar \contextot, \var : \typeo ^\bbone \\
+&\gbar \contextot, \var : \typet ^\bbtwo
 \end{aligned} $
@@
 
-Introduce the type system.  We got it from \cite{davies96}.
+The static semantics for \corelang~is given in Figure \ref{fig:coreStatics}.
+Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).
+The static semantics comprises two judgements.
+Those are:
+
+\begin{center}
+\begin{tabular}{|l|l|} \hline
+Name & Pattern \\ \hline
+Typing at $\bbone$ & $\typesone[\contextot]{\expro}{\typeo}$ \\  \hline
+Typing at $\bbtwo$ & $\typestwo[\contextot]{\exprt}{\typet}$ \\ \hline
+\end{tabular}
+\end{center}
+
+We see that the typing judgements for both stages are each what one would expect for an unstaged language.
+That is, there are introduction and elimination forms for base types and products at both stage $\bbone$ and stage $\bbtwo$.
+Furthermore, these rules preserve stage.
+For these reasons, we can think of the base type and product features as essentially being orthogonal to the staging features.
+The terms $\next$ and $\prev$, which are the only terms that allow us to change stage, are respectively the introduction and elimination forms for $\fut$.
 
@@
 \begin{mathpar}
-\infer [\mathrm{unit}] {\gds () : \text{unit}}{\cdot} \and
-\infer [\mathrm{int}] {\gds i : \text{int}}{\cdot} \and
-\infer [\mathrm{bool_1}] {\gds \mathtt{true} : \text{bool}}{\cdot} \and
-\infer [\mathrm{bool_2}] {\gds \mathtt{false} : \text{bool}}{\cdot} \and
-\infer [\mathrm{let}] {\letin{x}{e_1}{e_2} : B}{\gds e_1 : A & \Gamma,x:A^\sigma \vdash e_2 : B} \and
-\infer [\mathrm{hyp}] {\gds x : A}{x : A^\sigma \in \Gamma} \and
-\infer [\times\mathrm{I}] {\gds (e_1,e_2) : A\times B}{\gds e_1 : A & \gds e_2 : B} \and
-\infer [\times\mathrm{E_1}] {\gds \pi_1~e : A}{\gds e : A\times B} \and
-\infer [\times\mathrm{E_2}] {\gds \pi_2~e : B}{\gds e : A\times B} \and
-\infer [\mathrm{if}] {\gds \ifthen{e_1}{e_2}{e_3} : A}{\gds e_1 : \mathrm{bool} & \gds e_2 : A & \gds e_3 : A} \and
-\infer [\fut\mathrm{I}] {\Gamma \vdash^\bbone \next~e : \fut A}{\Gamma \vdash^\bbtwo e : A} \and
-\infer [\fut\mathrm{E}] {\Gamma \vdash^\bbtwo \prev~e : A}{\Gamma \vdash^\bbone e : \fut A} \and
+\infertypesone [\rmunit] 			{\typesone {()}{\rm unit}}{\cdot} \and
+\infertypesone [\mathrm{int}]		{\typesone {i}{\rm int}}{\cdot} \and
+\infertypesone [\mathrm{bool}]	{\typesone {b} {\rm bool}}{\cdot} \and
+\infertypesone [\mathrm{hyp}]		{\typesone x A}{x : A^\bbone \in \Gamma} \and
+\infertypestwo [\rmunit]			{\typestwo {()}{\rm unit}}{\cdot} \and
+\infertypestwo [\mathrm{int}]		{\typestwo {i}{\rm int}}{\cdot} \and
+\infertypestwo [\mathrm{bool}]	{\typestwo {b} {\rm bool}}{\cdot} \and
+\infertypestwo [\mathrm{hyp}]		{\typestwo x A}{x : A^\bbtwo \in \Gamma} \and
+\infertypesone [\mathrm{let}]		{\typesone{\letin{x}{e_1}{e_2}}{B}}{\typesone{e_1} A & \typesone [\Gamma,x:A^\bbone] {e_2} B} \and
+\infertypestwo [\mathrm{let}]		{\typestwo{\letin{x}{e_1}{e_2}}{B}}{\typestwo{e_1} A & \typestwo [\Gamma,x:A^\bbtwo] {e_2} B} \and
+\infertypesone [\times\mathrm{I}]		{\typesone {(e_1,e_2)} {A\times B}}{\typesone {e_1} A & \typesone {e_2} B} \and
+\infertypesone [\times\mathrm{E_1}]	{\typesone{\pi_1~e} A}{\typesone e {A\times B}} \and
+\infertypesone [\times\mathrm{E_2}]	{\typesone{\pi_2~e} B}{\typesone e {A\times B}} \and
+\infertypestwo [\times\mathrm{I}]		{\typestwo {(e_1,e_2)} {A\times B}}{\typestwo {e_1} A & \typestwo {e_2} B} \and
+\infertypestwo [\times\mathrm{E_1}]	{\typestwo{\pi_1~e} A}{\typestwo e {A\times B}} \and
+\infertypestwo [\times\mathrm{E_2}]	{\typestwo{\pi_2~e} B}{\typestwo e {A\times B}} \and
+\infertypesone [\mathrm{if}] {\typesone{\ifthen{e_1}{e_2}{e_3}} A}{\typesone{e_1} {\rm bool} & \typesone {e_2} A & \typesone {e_3} A} \and
+\infertypestwo [\mathrm{if}] {\typestwo{\ifthen{e_1}{e_2}{e_3}} A}{\typestwo{e_1} {\rm bool} & \typestwo {e_2} A & \typestwo {e_3} A} \and
+\infertypesone [\fut\mathrm{I}]	{\typesone{\next~e}{\fut A}}{\typestwo e A} \and
+\infertypestwo [\fut\mathrm{E}]	{\typestwo {\prev~e} A}{\typesone e {\fut A}} \and
 \end{mathpar}
@@
 
-Check Figure \ref{fig:erasureSemantics}
+The dynamics semantics for \corelang~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:
 
-\newcommand{\downer}{\Downarrow}
+\begin{center}
+\begin{tabular}{|l|l|} \hline
+Name & Pattern \\ \hline
+Reduction at $\bbone$ & $\reduceone{\expro}{\valo}$ \\  \hline
+Reduction at $\bbtwo$ & $\reducetwo{\exprt}{\valt}$ \\ \hline
+\end{tabular}
+\end{center}
+
+The primary goal for this semantics was simplicity.
+This was achieved in that, like the static semantics, the judgements for each level are extremely similar to those for an unstaged language.
+This simplicity came at the cost of some descriptiveness, however.
+The inadequacy of this semantics will become apparent shortly.
+
+First, we discuss a few desirable theorems.
+Chiefly among them is type preservation (I haven't defined the judgements for the types of values, but it should be rather clear).
+As will be the pattern for many theorems to come, our inductive hypothesis comprises two mutually dependent lemmas, one for stage $\bbone$, and the other for stage $\bbtwo$:
+\begin{center}
+\begin{tabular}{l}
+If $\typesone [\emptyC]{e}{A}$ \\
+and $\reduceone{e}{v}$ \\
+then $v : A$
+\end{tabular}
+~~~
+\begin{tabular}{l}
+If $\typestwo [\emptyC]{e}{A}$ \\
+and $\reducetwo{e}{v}$ \\
+then $v : A$
+\end{tabular}
+\end{center}
+[Proof not done, but it looks trivial in every case. [Wait no, substitutions kill me again.]]
+
+We also want some formalization of the idea that stage two computation cannot affect stage one computation.  At a minimum, the following lemma should be true [this needs lots of work],
+
+\begin{center}
+\begin{tabular}{l}
+If $\exists v\in \valt$ s.t. $\reduceone {[v/x]e}{u}$ \\
+then $\forall v\in \valt$ s.t. $\reduceone {[v/x]e}{u'}$, $u \simeq u'$
+\end{tabular}
+\end{center}
+
+As it turns out, this is not true! [Example forthcoming.]
 
@@
 \begin{mathpar}
-\infer [\mathrm{unit}] { () \downer {\bf ()}}{\cdot} \and
-\infer [\mathrm{int}] {i \downer i}{\cdot} \and
-\infer [\mathrm{bool_1}] {\mathtt{true} \downer {\bf true}}{\cdot} \and
-\infer [\mathrm{bool_2}] {\mathtt{false}\downer {\bf false}}{\cdot} \and
-\infer [\mathrm{let}] {\letin{x}{e_1}{e_2} \downer v_2}{e_1 \downer v_1 & [v_1/x]e_2 \downer v_2} \and
-\infer [\times\mathrm{I}] {(e_1,e_2) \downer (v_1,v_2)}{e_1 \downer v_1 & e_2 \downer v_2} \and
-\infer [\times\mathrm{E_1}] {\pi_1~e \downer v_1}{e \downer (v_1,v_2)} \and
-\infer [\times\mathrm{E_2}] {\pi_2~e \downer v_2}{e\downer (v_1,v_2)} \and
-\infer [\mathrm{if_1}] {\ifthen{e_1}{e_2}{e_3} \downer v}{e_1 \downer {\bf true} & e_2 \downer v} \and
-\infer [\mathrm{if_2}] {\ifthen{e_1}{e_2}{e_3} \downer v}{e_1 \downer {\bf false} & e_3 \downer v} \\
-\infer [\fut\mathrm{I}] {\next~e \downer v}{e \downer v} \and
-\infer [\fut\mathrm{E}] {\prev~e \downer v}{e \downer v} 
+\inferreduceone [\rmunit]				{\reduceone {()}{()}}{\cdot} \and
+\inferreduceone [\mathrm{int}]		{\reduceone{i}{i}}{\cdot} \and
+\inferreduceone [\mathrm{bool}]		{\reduceone{b}{b}}{\cdot} \and
+\inferreduceone [\mathrm{let}]			{\reduceone{\letin{x}{e_1}{e_2}}{v_2}}{\reduceonesub [1] & \reduceone{[v_1/x]e_2}{v_2}} \and
+\inferreducetwo [\rmunit]				{\reducetwo {()}{()}}{\cdot} \and
+\inferreducetwo [\mathrm{int}]		{\reducetwo{i}{i}}{\cdot} \and
+\inferreducetwo [\mathrm{bool}]		{\reducetwo{b}{b}}{\cdot} \and
+\inferreducetwo [\mathrm{let}]			{\reducetwo{\letin{x}{e_1}{e_2}}{v_2}}{\reducetwosub [1] & \reducetwo{[v_1/x]e_2}{v_2}} \and
+\inferreduceone [\times\mathrm{I}]	{\reduceone{(e_1,e_2)}{\valprod{v_1}{v_2}}}{\reduceonesub [1] & \reduceonesub [2]} \and
+\inferreduceone [\times\mathrm{E_1}]	{\reduceone{\pi_1~e}{v_1}}{\reduceone{e}{\valprod{v_1}{v_2}}} \and
+\inferreduceone [\times\mathrm{E_2}]	{\reduceone{\pi_2~e}{v_2}}{\reduceone{e}{\valprod{v_1}{v_2}} } \and
+\inferreducetwo [\times\mathrm{I}]	{\reducetwo{(e_1,e_2)}{\valprod{v_1}{v_2}}}{\reduceonesub [1] & \reduceonesub [2]} \and
+\inferreducetwo [\times\mathrm{E_1}]	{\reducetwo{\pi_1~e}{v_1}}{\reduceone{e}{\valprod{v_1}{v_2}}} \and
+\inferreducetwo [\times\mathrm{E_2}]	{\reducetwo{\pi_2~e}{v_2}}{\reduceone{e}{\valprod{v_1}{v_2}}} \and
+\inferreduceone [\mathrm{if_1}] 		{\reduceone{\ifthen{e_1}{e_2}{e_3}}{v}}{\reduceone{e_1}{\bf true} & \reduceone{e_2}{v}} \and
+\inferreduceone [\mathrm{if_2}] 		{\reduceone{\ifthen{e_1}{e_2}{e_3}}{v}}{\reduceone{e_1}{\bf false} & \reduceone{e_3}{v}} \\
+\inferreducetwo [\mathrm{if_1}] 		{\reducetwo{\ifthen{e_1}{e_2}{e_3}}{v}}{\reducetwo{e_1}{\bf true} & \reducetwo{e_2}{v}} \and
+\inferreducetwo [\mathrm{if_2}] 		{\reducetwo{\ifthen{e_1}{e_2}{e_3}}{v}}{\reducetwo{e_1}{\bf false} & \reducetwo{e_3}{v}} \\
+\inferreduceone [\fut\mathrm{I}]		{\reduceone{\next~e}{\valnext~v}}{\reducetwosub} \and
+\inferreducetwo [\fut\mathrm{E}]		{\reducetwo{\prev~e} v}{\reduceone e {\valnext~v}} 
 \end{mathpar}
@@
 
-Introduce \coremono, in figure \ref{fig:baseSyntax}.  We're gonna split into two terms of that.
+We now cover splitting, the process by which mixed code is teased apart into two individual chunks of code, one for each stage.
+To begin, we first introduce the target lagnuage (for both stage), \coremono.
+The grammar for \coremono~is given in figure \ref{fig:baseSyntax}.
+We see that \coremono~is (as one would hope) a traditional single-stage language, with all the same basetypes as \corelang, in addition to products, if expressions, and full sums (injections and case expressions).
+The presence of this last feature is a bit curious, since \corelang~did not support full sums.
+The necessity for sums should become apparent in time.
+The static and dynamic semantics of \coremono~are not given, but should be obvious.
+They are given by the judgements:
+
+\begin{center}
+\begin{tabular}{|l|l|} \hline
+Name & Pattern \\ \hline
+Typing & $\types{\expr}{\val}$ \\  \hline
+Reduction & $\reduce{\expr}{\val}$ \\ \hline
+\end{tabular}
+\end{center}
+
+The splitting judgements come in seven parts, two each for types, terms, and values, and one for contexts.  In tabular form, they are:
+\begin{center}
+\begin{tabular}{|l|l|} \hline
+Name & Pattern \\ \hline
+Value Splitting at $\bbone$& $\valo \vsplito [\val,\val]$ \\  \hline
+Value Splitting at $\bbtwo$& $\valt \vsplits \val$ \\  \hline
+Type Splitting at $\bbone$& $\typeo \tsplito [\type,\type]$ \\  \hline
+Type Splitting at $\bbtwo$& $\typet \tsplits \type$ \\  \hline
+Context Splitting & $\contextot \csplit [\context,\context]$ \\  \hline
+Term Splitting at $\bbone$& $\splitone{\expro}{\typeo}{\expr,\var.\expr}$ \\  \hline
+Term Splitting at $\bbtwo$& $\splittwo{\exprt}{\typet}{\expr,\var.\expr}$ \\  \hline
+\end{tabular}
+\end{center}
+
+[...more...]
 
@@
 $\begin{aligned}
-\expr &::= ()~|~\inte~|~\mathtt{true}~|~\mathtt{false} \\
+\expr &::= ()~|~\inte~|~\bool \\
 &\gbar \letin{\var}{\expr}{\expr} \\
@@
 &\gbar \pi_1~\expr \gbar \pi_2~\expr \\
-&\gbar \ifthen {\expr}{\expr}{\expr} 
+&\gbar \inl~\expr \gbar \inr~\expr \\
+&\gbar \ifthen {\expr}{\expr}{\expr}  \\
+&\gbar \caseof {\expr}{x_1.\expr}{x_2.\expr} 
 \end{aligned} $
@@
 $\begin{aligned}
-\type &::=  \text{unit}~|~\text{int}~|~\text{bool} \\
-&\gbar \type \times \type 
+\type &::=  \rmunit~|~\text{int}~|~\text{bool} \\
+&\gbar \type \times \type  \\
+&\gbar \type + \type 
 \\
-\val &::= ()~|~\inte~|~\mathbf{true}~|~\mathbf{false} \\
-&\gbar \valprod {\val} {\val} 
+\val &::= ()~|~\inte~|~\bool \\
+&\gbar \valprod {\val} {\val}  \\
+&\gbar \valinl~\expr \gbar \valinr~\expr
 \\
-\context &::= \mathrm{empty} \\
+\context &::= \emptyC \\
 &\gbar \context, \var : \type
@@
 \begin{figure*}
-\caption{Splitting}
-\label{fig:split}
+\caption{Type Splitting}
+\label{fig:typeSplit}
 \begin{mathpar}
-\infer [\mathrm{unit}\splito] 
-	{\gdo () : \text{unit} \splito [((),()),\_.()]}{\cdot} \and
-\infer [\mathrm{unit}\splits]
-	 {\gdt () : \text{unit} \splits [(),\_.()]}{\cdot} \and
-\infer [\times\mathrm{I}\splito] 
-	{\gdo (e_1,e_2) : A\times B \splito [((\pi_1 c, \pi_1 c),(\pi_2 c_1, \pi_2 c_2)), l.(\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})] }
-	{ \gdo e_1 : A \splito [c_1,l_1.r_1] 
-	&\gdo e_2 : B \splito [c_2,l_2.r_2]} \and
-\infer [\times\mathrm{I}\splits] 
-	{\gdt (e_1,e_2) : A\times B \splits [(p_1,p_2), l.(\letin{l_1}{\pi_1~l}{r_1},\letin{l_2}{\pi_2~l}{r_2})] }
-	{ \gdt e_1 : A \splits [p_1,l_1.r_1] 
-	&\gdt e_2 : B \splits [p_2,l_2.r_2]} \and
-\infer [\times\mathrm{E_1}\splito] 
-	{\gdo \pi_1~e : A \splito [(\pi_1(\pi_1c),\pi_2 c),l.\pi_1~r] }{\gdo e : A\times B \splito [c,l.r] } \and
-\infer [\times\mathrm{E_1}\splits] 
-	{\gdt \pi_1~e : A \splits [p,l.\pi_1~r] }{\gdt e : A\times B \splits [p,l.r] } \and
-\infer [\times\mathrm{E_2}\splito] 
-	{\gdo \pi_2~e : B \splito [(\pi_2(\pi_1c),\pi_2 c),l.\pi_2~r] }{\gdo e : A\times B \splito [c,l.r]} \and
-\infer [\times\mathrm{E_2}\splits] 
-	{\gdt \pi_2~e : B \splits [p,l.\pi_2~r] }{\gdt e : A\times B \splits [p,l.r]} \and
-\infer [\fut\mathrm{I}\splito] {\gdo \next~e : \fut A \splito [((),p),l.r]}{\gdt e : A\splits [p,l.r] } \and
-\infer [\fut\mathrm{E}\splits] {\gdt \prev~e : A\splits [\pi_2~c,l.r] }{\gdo e : \fut A \splito [c,l.r]} \and
-\infer [+\mathrm{E}\splito]
-	{\gdo \left( \tallif {e_1}{e_2}{e_3} \right) :A \splito 
-	\left[
+\infer [\mathrm{unit}\tsplito]	{\rmunit \tsplito [\rmunit,\rmunit]}{\cdot} \and
+\infer [\mathrm{int}\tsplito]	{{\rm int} \tsplito [{\rm int}, \rmunit]}{\cdot} \and
+\infer [\mathrm{bool}\tsplito]	{{\rm bool} \tsplito [{\rm bool}, \rmunit]}{\cdot} \and
+\infer [\times\tsplito]	
+	{A\times B \tsplito [A_1 \times B_1, A_2 \times B_2]}
+	{A \tsplito [A_1, A_2] & B \tsplito [B_1,B_2]} \and
+\infer [\fut\tsplito]
+	{\fut~A \tsplito [{\rm unit}, A']}
+	{A \tsplits A'} \and
+\end{mathpar}
+\end{figure*}
+
+\begin{figure*}
+\caption{Value Splitting}
+\label{fig:valSplit}
+\begin{mathpar}
+\infer [\rmunit\vsplito]		{() \vsplito [(), ()]}{\cdot} \and
+\infer [\mathrm{int}\vsplito]	{i \vsplito [i,()]}{\cdot} \and
+\infer [\mathrm{bool}\vsplito]	{b \vsplito [b,()]}{\cdot} \and
+\infer [\times\vsplito]	
+	{\valprod{v_1}{v_2} \vsplito [\valprod{u_1}{u_2}, \valprod{w_1}{w_2}]}
+	{v_1 \vsplito [u_1,w_1] & v_2 \vsplito [u_2,w_2]} \and
+\infer [\fut\vsplito]
+	{\valnext~v \vsplito [(),v']}
+	{v \vsplits v'} 
+\end{mathpar}
+\end{figure*}
+
+\begin{figure*}
+\caption{Context Splitting}
+\label{fig:contSplit}
+\begin{mathpar}
+\infer {\emptyC \csplit [\emptyC~;~\emptyC]}{\cdot} \and
+\infer {\Gamma, v : A^\bbone \csplit [\Gamma_1,v:A_1; \Gamma_2,v:A_2]}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplito [A_1, A_2]} \and
+\infer {\Gamma, v : A^\bbtwo \csplit [\Gamma_1; \Gamma_2,v:A']}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplits A'} 
+\end{mathpar}
+\end{figure*}
+
+\begin{figure*}
+\caption{Term Splitting}
+\label{fig:termSplit}
+\begin{mathpar}
+\infersplitone [\rmunit] 
+	{\splitone{()}{\rmunit}{((),()),\_.()}} {\cdot} \and
+\infersplittwo [\rmunit]
+	{\splittwo{()}{\rmunit}{(),\_.()}} {\cdot} \and
+\infersplitone [{\rm int}] 
+	{\splitone{i}{\rm int}{(i,()),\_.()}} {\cdot} \and
+\infersplittwo [{\rm int}]
+	{\splittwo{i}{\rm int}{(),\_.i}} {\cdot} \and
+\infersplitone [\times\mathrm{I}] 
+	{\splitone {(e_1,e_2)}{A\times B}
+		{ \begin{array}{l}
+		\left(\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}\right), \\
+		l.(\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})
+		\end{array}}}
+	{ \splitonesub [1] A & \splitonesub [2] B} \and
+\infersplittwo [\times\mathrm{I}] 
+	{\splittwo {(e_1,e_2)}{A\times B}{(p_1,p_2), l.(\letin{l_1}{\pi_1~l}{r_1},\letin{l_2}{\pi_2~l}{r_2})}}
+	{ \splittwosub [1] A & \splittwosub [2] B} \and
+\infersplitone [\times\mathrm{E_1}] 
+	{\splitone {\pi_1~e}{A}{(\pi_1(\pi_1c),\pi_2 c),l.\pi_1~r}}{\splitonesub{A\times B}} \and
+\infersplittwo [\times\mathrm{E_1}] 
+	{\splittwo{\pi_1~e}{A}{p,l.\pi_1~r} }{\splittwosub{A\times B}} \and
+\infersplitone [\times\mathrm{E_2}] 
+	{\splitone {\pi_2~e}{B}{(\pi_2(\pi_1c),\pi_2 c),l.\pi_2~r}}{\splitonesub{A\times B}} \and
+\infersplittwo [\times\mathrm{E_2}] 
+	{\splittwo{\pi_2~e}{B}{p,l.\pi_2~r} }{\splittwosub{A\times B}} \and
+\infersplitone [\mathrm{unit}] 
+	{\splitone {v}{A}{(x,()),\_.()}}{x:A^\bbone \in \Gamma} \and
+\infersplittwo [\mathrm{unit}]
+	 {\splittwo{v}{A}{(),\_.x}}{x:A^\bbtwo \in \Gamma} \\
+\infersplitone [\mathrm{let}] 
+	{\splitone{\letin{x}{e_1}{e_2}}{B}{\cdots}}
+	{\splitonesub [1] A & \splitone [\Gamma,x:A^\bbone] {e_2}{B}{c_2,l_2.r_2}} \and
+\infersplittwo [\mathrm{let}] 
+	{\splittwo{\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\letin{l_1}{\pi_1 l}{r_1})}{\letin{l_2}{\pi_2 l}{r_2}}}}
+	{\splittwosub [1] A & \splittwo [\Gamma,x:A^\bbtwo] {e_2}{B}{p_2,l_2.r_2}} \and
+\infersplitone [\fut\mathrm{I}] 
+	{\splitone{\next~e}{\fut A}{((),p),l.r}}{\splittwosub A} \and
+\infersplittwo [\fut\mathrm{E}] 
+	{\splittwo{\prev~e}{A} {\pi_2~c,l.r}}{\splitonesub {\fut A}} \and
+\infersplitone [+\mathrm{E}]
+	{\splitone {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
+	{
 		\begin{array}{l}
 		\left(\tallif {\pi_1c_1}
-			{\letin{y}{c_2}{(\pi_1 y, \iota_1(\pi_2 y))}}
-			{\letin{y}{c_3}{(\pi_1 y, \iota_2(\pi_2 y))}}
+			{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
+			{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
 		\right), \\
@@
 		\end{array}
-	\right]}
-	{\gdo e_1 : {\rm bool} \splito [c_1, l_1.r_1]
-	& \gdo e_2 : A \splito [c_2,l_2.r_2]
-	& \gdo e_3 : A \splito [c_3,l_3.r_3]} \and
-%\infer [+\mathrm{E}\splito]
-%	{\gdo \left( \tallif {e_1}{e_2}{e_3} \right) :A \splito 
-%	\left[
-%		\begin{array}{l}
-%		\letin{c}{c_1}{\left(\tallif {\pi_1c}
-%			{\letin{y}{c_2}{(\pi_1 y, (\iota_1(\pi_2 y), \pi_2 c))}}
-%			{\letin{y}{c_3}{(\pi_1 y, (\iota_2(\pi_2 y), \pi_2 c))}}
-%		\right)}, \\
-%		l.\left(\letin{c_1}{\pi_2 l}{r_1}; 
-%			\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3}
-%		\right)
-%		\end{array}
-%	\right]}
-%	{\gdo e_1 : {\rm bool} \splito [c_1, l_1.r_1]
-%	& \gdo e_2 : A \splito [c_2,l_2.r_2]
-%	& \gdo e_3 : A \splito [c_3,l_3.r_3]} \and
-\infer [+\mathrm{E}\splits]
-	{\gdt \left( \tallif {e_1}{e_2}{e_3}\right) : A \splits 
-	\left[(p_1,p_2,p_3), 
+	}}
+	{\splitonesub [1] {\rm bool} & \splitonesub [2] A & \splitonesub [3] A } \and
+\infersplittwo [+\mathrm{E}]
+	{\splittwo {\left( \tallif {e_1}{e_2}{e_3}\right)}{A} 
+	{(p_1,p_2,p_3), 
 	l.\left(\tallif{(\letin {l_1}{\pi_1 l}{r_1})}{(\letin {l_2}{\pi_2 l}{r_2})} {(\letin {l_3}{\pi_3 l}{r_3})} \right)
-	\right] }
-	{ \gdt e_1 : {\rm bool} \splits [p_1,l_1.r_1]
-	& \gdt e_2 : A \splits [p_2,l_2.r_2] 
-	& \gdt e_3 : A \splits [p_3,l_3.r_3]} 
+	}}
+	{\splittwosub [1] {\rm bool} & \splittwosub [2] A & \splittwosub [3] A } \and
 \end{mathpar}
@@
 
+\subsubsection{It preserves types}
+
+The theorem we want to show has two parts, one for the $\bbone$ code, and one for the $\bbtwo$ code.  
+\begin{center}
+\begin{tabular}{l}
+If $\splitone {e}{A}{c,l.r}$ \\
+then $\Gamma \csplit [\Gamma_1;\Gamma_2]$ \\
+and $A \tsplito [A_1,A_2]$ \\
+and $ \types [\Gamma_1]{c} {A_1 \times \tau}$ \\
+and $ \types [\Gamma_2, p:\tau]{r} {A_2}$ 
+\end{tabular}
+~~~
+\begin{tabular}{l}
+If $\splittwo {e}{A}{p,l.r}$ \\
+then $\Gamma \csplit [\Gamma_1;\Gamma_2]$ \\
+and $A \tsplits A'$ \\
+and $ \types [\Gamma_1]{p} {\tau}$ \\
+and $ \types [\Gamma_2, p:\tau]{r} {A'}$ 
+\end{tabular}
+\end{center}
+
+\subsubsection{It preserves meaning}
+
+The theorem we want to show has two parts, one for the $\bbone$ code, and one for the $\bbtwo$ code.  
+\begin{center}
+\begin{tabular}{l}
+If $\splitone [\emptyC]{e}{A}{c,l.r}$ \\
+and $\reduceone{e}{v}$ \\
+and $c \Downarrow \valprod{t}{u}$ \\
+and $[u/l]r \Downarrow w$ \\
+then $v \vsplito [t,w]$
+\end{tabular}
+~~~
+\begin{tabular}{l}
+If $\splittwo[\emptyC]{e}{B}{p,l.r}$ \\
+and $\reducetwo{e}{v}$ \\
+and $p \Downarrow u$ \\
+and $[u/l]r \Downarrow w$ \\
+then $v \vsplits w$
+\end{tabular}
+\end{center}
+
+We proceed by cases of the term splitting judgement.
+
+\begin{itemize}
+\item {\bf Int.} Suppose that
 \begin{enumerate}
-\item It always pops out a well typed term.
-\item Meaning is preserved, up to some translation.
+\item $\splitone{i}{\rm int}{(i,()),\_.()}$
 \end{enumerate}
+Since reduction is a function, we also have
+\begin{enumerate}
+\item $\reduceone{i}{i}$
+\item $\reduceone{(i,())}{\valprod {i}{()}}$
+\item $\reduceone{[()/\_]()}{()}$
+\end{enumerate}
+And since value splitting is a function,
+\begin{enumerate}
+\item $i \vsplito [i,()]$
+\end{enumerate}
+Which is what we need to show.
+
+\item {\bf Tuple.} Suppose that
+\begin{enumerate}
+\item $\splitone {(e_1,e_2)}{A\times B}
+		{ \begin{array}{l}
+		(\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}), \\
+		l.(\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})
+		\end{array}} $
+\item $\reduceone{(e_1,e_2)}{v}$
+\item $\reduce{\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}}{\valprod tu}$
+\item $\reduce{[u/l](\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})}{w}$
+\end{enumerate}
+Since the splitting and evaluation rules are syntax-directed, it follows that
+\begin{enumerate}
+\item $\splitone {e_1}{A}{c_1,l_1.r_1}$
+\item $\splitone {e_2}{B}{c_2,l_2.r_2}$
+\item $v$ has the form $\valprod{v_1}{v_2}$, by inversion of tuple evaluation
+\item $\reduceone{(e_1,e_2)}{\valprod{v_1}{v_2}}$
+\item $\reduceone{e_1}{v_1}$
+\item $\reduceone{e_2}{v_2}$
+\item By inversion of evaluation, $t$ has the form $\valprod{t_1}{t_2}$  and $u$ has the form $\valprod{u_1}{u_2}$
+\item $\reduceone{[\valprod{t_1}{u_1},\valprod{t_2}{u_2}/y_1,y_2]((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}{\valprod tu}$
+\item $\reduceone{c_1}{\valprod {t_1}{u_1}}$, by inversion of let eval
+\item $\reduceone{c_2}{\valprod {t_2}{u_2}}$, by inversion of let eval
+\item By inversion of evaluation, $w$ has the form $\valprod{w_1}{w_2}$
+\item $\reduce{[\valprod{u_1}{u_2}/l](\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})}{\valprod{w_1}{w_2}}$, substituting for $u$, $w$
+\item $\reduce{[u_1/l_1]r_1}{w_1}$, inversion of evaluation
+\item $\reduce{[u_2/l_2]r_2}{w_2}$, inversion of evaluation
+\end{enumerate}
+We now have enough to invoke the inductive hypothesis,
+\begin{enumerate}
+\item $v_1 \vsplito [t_1,w_1]$ 
+\item $v_2 \vsplito [t_2,w_2]$ 
+\item $\valprod{v_1}{v_2} \vsplito [\valprod{t_1}{t_2},\valprod{w_1}{w_2}]$ 
+\item $v \vsplito [t,w]$
+\end{enumerate}
+Which is what we need to show.
+\end{itemize}
 

### 2014-02-21 paper/paper.tex
Added a better semantics.  Refactored some figures using fancy latex macros.
Refactored splitting rules.  Fixed up some proofs.

@@
 \begin{mathpar}
-\infertypesone [\rmunit] 			{\typesone {()}{\rm unit}}{\cdot} \and
-\infertypesone [\mathrm{int}]		{\typesone {i}{\rm int}}{\cdot} \and
-\infertypesone [\mathrm{bool}]	{\typesone {b} {\rm bool}}{\cdot} \and
-\infertypesone [\mathrm{hyp}]		{\typesone x A}{x : A^\bbone \in \Gamma} \and
-\infertypestwo [\rmunit]			{\typestwo {()}{\rm unit}}{\cdot} \and
-\infertypestwo [\mathrm{int}]		{\typestwo {i}{\rm int}}{\cdot} \and
-\infertypestwo [\mathrm{bool}]	{\typestwo {b} {\rm bool}}{\cdot} \and
-\infertypestwo [\mathrm{hyp}]		{\typestwo x A}{x : A^\bbtwo \in \Gamma} \and
-\infertypesone [\mathrm{let}]		{\typesone{\letin{x}{e_1}{e_2}}{B}}{\typesone{e_1} A & \typesone [\Gamma,x:A^\bbone] {e_2} B} \and
-\infertypestwo [\mathrm{let}]		{\typestwo{\letin{x}{e_1}{e_2}}{B}}{\typestwo{e_1} A & \typestwo [\Gamma,x:A^\bbtwo] {e_2} B} \and
-\infertypesone [\times\mathrm{I}]		{\typesone {(e_1,e_2)} {A\times B}}{\typesone {e_1} A & \typesone {e_2} B} \and
-\infertypesone [\times\mathrm{E_1}]	{\typesone{\pi_1~e} A}{\typesone e {A\times B}} \and
-\infertypesone [\times\mathrm{E_2}]	{\typesone{\pi_2~e} B}{\typesone e {A\times B}} \and
-\infertypestwo [\times\mathrm{I}]		{\typestwo {(e_1,e_2)} {A\times B}}{\typestwo {e_1} A & \typestwo {e_2} B} \and
-\infertypestwo [\times\mathrm{E_1}]	{\typestwo{\pi_1~e} A}{\typestwo e {A\times B}} \and
-\infertypestwo [\times\mathrm{E_2}]	{\typestwo{\pi_2~e} B}{\typestwo e {A\times B}} \and
-\infertypesone [\mathrm{if}] {\typesone{\ifthen{e_1}{e_2}{e_3}} A}{\typesone{e_1} {\rm bool} & \typesone {e_2} A & \typesone {e_3} A} \and
-\infertypestwo [\mathrm{if}] {\typestwo{\ifthen{e_1}{e_2}{e_3}} A}{\typestwo{e_1} {\rm bool} & \typestwo {e_2} A & \typestwo {e_3} A} \and
-\infertypesone [\fut\mathrm{I}]	{\typesone{\next~e}{\fut A}}{\typestwo e A} \and
-\infertypestwo [\fut\mathrm{E}]	{\typestwo {\prev~e} A}{\typesone e {\fut A}} \and
+\infertypesone [\rmunit] 	{\ty {()}\rmunit}						{\cdot} \and
+\infertypesone [int]			{\ty {i}{\rm int}}					{\cdot} \and
+\infertypesone [bool]			{\ty {b} \rmbool}						{\cdot} \and
+\infertypesone [hyp]			{\ty x A}								{x : A^\bbone \in \Gamma} \and
+\infertypestwo [\rmunit]		{\ty {()}{\rmunit}}					{\cdot} \and
+\infertypestwo [int]			{\ty {i}{\rm int}}					{\cdot} \and
+\infertypestwo [bool]			{\ty {b} \rmbool}						{\cdot} \and
+\infertypestwo [hyp]			{\ty x A}								{x : A^\bbtwo \in \Gamma} \and
+\infertypesone [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbone] {e_2} B} \and
+\infertypestwo [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbtwo] {e_2} B} \and
+\infertypesone [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \ty {e_2} B} \and
+\infertypesone [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
+\infertypesone [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
+\infertypestwo [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \typestwo {e_2} B} \and
+\infertypestwo [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
+\infertypestwo [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
+\infertypesone [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
+\infertypestwo [if]			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
+\infertypesone [\fut I]		{\ty {\next~e}{\fut A}}				{\typestwo e A} \and
+\infertypestwo [\fut E]		{\ty {\prev~e} A}						{\typesone e {\fut A}} \and
 \end{mathpar}
@@
 \begin{mathpar}
-\inferreduceone [\rmunit]				{\reduceone {()}{()}}{\cdot} \and
-\inferreduceone [\mathrm{int}]		{\reduceone{i}{i}}{\cdot} \and
-\inferreduceone [\mathrm{bool}]		{\reduceone{b}{b}}{\cdot} \and
-\inferreduceone [\mathrm{let}]			{\reduceone{\letin{x}{e_1}{e_2}}{v_2}}{\reduceonesub [1] & \reduceone{[v_1/x]e_2}{v_2}} \and
-\inferreducetwo [\rmunit]				{\reducetwo {()}{()}}{\cdot} \and
-\inferreducetwo [\mathrm{int}]		{\reducetwo{i}{i}}{\cdot} \and
-\inferreducetwo [\mathrm{bool}]		{\reducetwo{b}{b}}{\cdot} \and
-\inferreducetwo [\mathrm{let}]			{\reducetwo{\letin{x}{e_1}{e_2}}{v_2}}{\reducetwosub [1] & \reducetwo{[v_1/x]e_2}{v_2}} \and
-\inferreduceone [\times\mathrm{I}]	{\reduceone{(e_1,e_2)}{\valprod{v_1}{v_2}}}{\reduceonesub [1] & \reduceonesub [2]} \and
-\inferreduceone [\times\mathrm{E_1}]	{\reduceone{\pi_1~e}{v_1}}{\reduceone{e}{\valprod{v_1}{v_2}}} \and
-\inferreduceone [\times\mathrm{E_2}]	{\reduceone{\pi_2~e}{v_2}}{\reduceone{e}{\valprod{v_1}{v_2}} } \and
-\inferreducetwo [\times\mathrm{I}]	{\reducetwo{(e_1,e_2)}{\valprod{v_1}{v_2}}}{\reduceonesub [1] & \reduceonesub [2]} \and
-\inferreducetwo [\times\mathrm{E_1}]	{\reducetwo{\pi_1~e}{v_1}}{\reduceone{e}{\valprod{v_1}{v_2}}} \and
-\inferreducetwo [\times\mathrm{E_2}]	{\reducetwo{\pi_2~e}{v_2}}{\reduceone{e}{\valprod{v_1}{v_2}}} \and
-\inferreduceone [\mathrm{if_1}] 		{\reduceone{\ifthen{e_1}{e_2}{e_3}}{v}}{\reduceone{e_1}{\bf true} & \reduceone{e_2}{v}} \and
-\inferreduceone [\mathrm{if_2}] 		{\reduceone{\ifthen{e_1}{e_2}{e_3}}{v}}{\reduceone{e_1}{\bf false} & \reduceone{e_3}{v}} \\
-\inferreducetwo [\mathrm{if_1}] 		{\reducetwo{\ifthen{e_1}{e_2}{e_3}}{v}}{\reducetwo{e_1}{\bf true} & \reducetwo{e_2}{v}} \and
-\inferreducetwo [\mathrm{if_2}] 		{\reducetwo{\ifthen{e_1}{e_2}{e_3}}{v}}{\reducetwo{e_1}{\bf false} & \reducetwo{e_3}{v}} \\
-\inferreduceone [\fut\mathrm{I}]		{\reduceone{\next~e}{\valnext~v}}{\reducetwosub} \and
-\inferreducetwo [\fut\mathrm{E}]		{\reducetwo{\prev~e} v}{\reduceone e {\valnext~v}} 
+\inferreduceone [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
+\inferreduceone [int]				{\red {i}{i}}								{\cdot} 											\and
+\inferreduceone [bool]			{\red {b}{b}}								{\cdot} 											\and
+\inferreduceone [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
+\inferreducetwo [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
+\inferreducetwo [int]				{\red {i}{i}}								{\cdot} 											\and
+\inferreducetwo [bool]			{\red {b}{b}}								{\cdot} 											\and
+\inferreducetwo [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
+\inferreduceone [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]} 							\and
+\inferreduceone [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
+\inferreduceone [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
+\inferreducetwo [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]}								\and
+\inferreducetwo [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
+\inferreducetwo [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
+\inferreduceone [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
+\inferreduceone [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
+\inferreducetwo [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
+\inferreducetwo [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
+\inferreduceone [\fut I]			{\red {\next~e}{\valnext~v}}				{\reducetwosub}									\and
+\inferreducetwo [\fut E]			{\red {\prev~e} v}						{\reduceone e {\valnext~v}} 
 \end{mathpar}
@@
 
-%\subsection{Better Dynamic Semantics}
-%
-%Introduce the dynamics semantics.  Figure \ref{fig:coreDynamics}.  It's small step.  We figured this out on our own, but it really should exist in the partial evaluation literature somewhere.
-%
-%\newcommand {\stepo} {\hookrightarrow_\bbone} 
-%\newcommand {\stept} {\hookrightarrow_\bbtwo} 
-%\newcommand {\pco} {~\mathrm{pc}\bbone} 
-%\newcommand {\fco} {~\mathrm{fc}\bbone} 
-%\newcommand {\pct} {~\mathrm{pc}\bbtwo} 
-%\newcommand {\fct} {~\mathrm{fc}\bbtwo} 
-%
-%\begin{figure}
-%\caption{\corelang~Partially Computed Values (pc$\bbone$ and pc$\bbtwo$)}
-%\label{fig:coreDynamics}
-%\begin{mathpar}
-%\infer {() \pco}{\cdot} \and
-%\infer {i \pco}{\cdot} \and
-%\infer {b \pco}{\cdot} \and
-%\infer {(v_1, v_2) \pco}{v_1 \pco & v_2 \pco} \and
-%\infer {\next~e \pco}{e \pct} \\
-%\infer {() \pct}{\cdot} \and
-%\infer {i \pct}{\cdot} \and
-%\infer {b \pct}{\cdot} \and
-%\infer {(v_1, v_2) \pct}{v_1 \pct & v_2 \pct} \and
-%\infer {\pi_i e \pct}{e \pct} \and
-%\infer {x \pct}{\cdot} \and
-%\infer {\letin{x}{e_1}{e_2} \pct}{e_1 \pct & e_2 \pct} \and
-%\infer {\ifthen{e_1}{e_2}{e_3} \pct}{e_1 \pct & e_2 \pct & e_3 \pct} \and
-%\infer {\prev~e \pct}{e \pco} 
-%\end{mathpar}
-%\end{figure}
-%
-%\begin{figure}
-%\caption{\corelang~Fully Computed Values (fc$\bbone$ and fc$\bbtwo$)}
-%\label{fig:coreDynamics}
-%\begin{mathpar}
-%\infer {() \fco}{\cdot} \and
-%\infer {i \fco}{\cdot} \and
-%\infer {b \fco}{\cdot} \and
-%\infer {(v_1, v_2) \fco}{v_1 \fco & v_2 \fco} \and
-%\infer {\next~e \fco}{e \fct} \\
-%\infer {() \fct}{\cdot} \and
-%\infer {i \fct}{\cdot} \and
-%\infer {b \fct}{\cdot} \and
-%\infer {(v_1, v_2) \fct}{v_1 \fct & v_2 \fct} 
-%\end{mathpar}
-%\end{figure}
-%
-%\begin{figure}
-%\caption{\corelang~Transitions}
-%\label{fig:coreDynamics}
-%\begin{mathpar}
-%\infer {\ifthen{e_1}{e_2}{e_3} \stepo \ifthen{e_1'}{e_2}{e_3}}{e_1 \stepo e_1'}  \\
-%\infer {\ifthen{\tt true}{e_2}{e_3} \stepo e_2}{\cdot} \and 
-%\infer {\ifthen{\tt false}{e_2}{e_3} \stepo e_3}{\cdot}  \\
-%\infer {\ifthen{e_1}{e_2}{e_3} \stept \ifthen{e_1'}{e_2}{e_3}}{e_1 \stept e_1' & e_2 \pct & e_3 \pct}  \\
-%\infer {\ifthen{\tt true}{e_2}{e_3} \stept e_2}{e_2 \pct & e_3 \pct} \and 
-%\infer {\ifthen{\tt false}{e_2}{e_3} \stept e_3}{e_2 \pct & e_3 \pct} \\
-%\infer {\letin{x}{e_1}{e_2} \stepo \letin{x}{e_1'}{e_2}}{e_1 \stepo e_1'} \and
-%\infer {\letin{x}{e_1}{e_2} \stepo [e_1/x]e_2}{e_1 \fco} \and
-%\infer {\letin{x}{e_1}{e_2} \stept \letin{x}{e_1'}{e_2}}{e_1 \stept e_1'} \and
-%\infer {\letin{x}{e_1}{e_2} \stept [e_1/x]e_2}{e_1 \fct} \and
-%\infer {x \downtwo x}{\cdot} \and
-%\infer {\gds (e_1,e_2) : A\times B}{\gds e_1 : A & \gds e_2 : B} \and
-%\infer {\ifthen{e_1}{e_2}{e_3} \downone [v,r]}{e_1 \downone {\bf true} & e_2 \downone [v,r]} \and
-%\infer {\ifthen{e_1}{e_2}{e_3} \downone [v,r]}{e_1 \downone {\bf false} & e_3 \downone [v,r]} \and
-%\infer {\ifthen{e_1}{e_2}{e_3} \downtwo [v, \ifthen{r_1}{r_2}{r_3}]}{e_1 \downtwo [\valdum,r_1] & e_2 \downtwo [v_2,r_2] & e_3 \downtwo [v_3,r_3]} \and
-%\infer {\next~e \downone [\valdum,r]}{e \downtwo r} \and
-%\infer {\prev~e \downtwo r}{e \downone [\valdum, r]} \and
-%\end{mathpar}
-%\end{figure}
-
-%\subsection{Good Theorems About the Language Alone}
-%
-%\begin{enumerate}
-%\item Type Safety.
-%\item Unidirectional Flow. [Essentially, you cannot write a program that goes back in time.]
-%\item Erasure property.  [If you ignore the nexts and prevs, the result is weakly equivalent.] [This was proven for a denotational semantics in some Jones paper that I need to find...]
-%\end{enumerate}
+\begin{figure}
+\caption{\corelang~Reference Semantics}
+\label{fig:refSemantics}
+\begin{mathpar}
+\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
+\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
+\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\\
+\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
+\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
+\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\\
+\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
+\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\\
+\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
+\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
+\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
+\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
+\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
+\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
+\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
+\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{r_2}}							{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
+\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{r_3}}							{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\\
+\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
+\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
+\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
+\end{mathpar}
+\end{figure}
 
@@
 \label{fig:termSplit}
+\newcommand {\projbind} [1] {\letin{l_{#1}}{\pi_{#1} l}{r_{#1}}}
 \begin{mathpar}
-\infersplitone [\rmunit] 
-	{\splitone{()}{\rmunit}{((),()),\_.()}} {\cdot} \and
-\infersplittwo [\rmunit]
-	{\splittwo{()}{\rmunit}{(),\_.()}} {\cdot} \and
-\infersplitone [{\rm int}] 
-	{\splitone{i}{\rm int}{(i,()),\_.()}} {\cdot} \and
-\infersplittwo [{\rm int}]
-	{\splittwo{i}{\rm int}{(),\_.i}} {\cdot} \and
-\infersplitone [\times\mathrm{I}] 
-	{\splitone {(e_1,e_2)}{A\times B}
+\infersplitone [unit]			{\spl {()}{\rmunit}{((),()),\_.()}}															{\cdot} \and
+\infersplittwo [unit]			{\spl {()}{\rmunit}{(),\_.()}}																{\cdot} \and
+\infersplitone [int]			{\spl {i}{\rm int}{(i,()),\_.()}}															{\cdot} \and
+\infersplittwo [int]			{\spl {i}{\rm int}{(),\_.i}}																	{\cdot} \and
+\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
 									{ \begin{array}{l}
-		\left(\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}\right), \\
-		l.(\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})
-		\end{array}}}
-	{ \splitonesub [1] A & \splitonesub [2] B} \and
-\infersplittwo [\times\mathrm{I}] 
-	{\splittwo {(e_1,e_2)}{A\times B}{(p_1,p_2), l.(\letin{l_1}{\pi_1~l}{r_1},\letin{l_2}{\pi_2~l}{r_2})}}
-	{ \splittwosub [1] A & \splittwosub [2] B} \and
-\infersplitone [\times\mathrm{E_1}] 
-	{\splitone {\pi_1~e}{A}{(\pi_1(\pi_1c),\pi_2 c),l.\pi_1~r}}{\splitonesub{A\times B}} \and
-\infersplittwo [\times\mathrm{E_1}] 
-	{\splittwo{\pi_1~e}{A}{p,l.\pi_1~r} }{\splittwosub{A\times B}} \and
-\infersplitone [\times\mathrm{E_2}] 
-	{\splitone {\pi_2~e}{B}{(\pi_2(\pi_1c),\pi_2 c),l.\pi_2~r}}{\splitonesub{A\times B}} \and
-\infersplittwo [\times\mathrm{E_2}] 
-	{\splittwo{\pi_2~e}{B}{p,l.\pi_2~r} }{\splittwosub{A\times B}} \and
-\infersplitone [\mathrm{unit}] 
-	{\splitone {v}{A}{(x,()),\_.()}}{x:A^\bbone \in \Gamma} \and
-\infersplittwo [\mathrm{unit}]
-	 {\splittwo{v}{A}{(),\_.x}}{x:A^\bbtwo \in \Gamma} \\
-\infersplitone [\mathrm{let}] 
-	{\splitone{\letin{x}{e_1}{e_2}}{B}{\cdots}}
-	{\splitonesub [1] A & \splitone [\Gamma,x:A^\bbone] {e_2}{B}{c_2,l_2.r_2}} \and
-\infersplittwo [\mathrm{let}] 
-	{\splittwo{\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\letin{l_1}{\pi_1 l}{r_1})}{\letin{l_2}{\pi_2 l}{r_2}}}}
-	{\splittwosub [1] A & \splittwo [\Gamma,x:A^\bbtwo] {e_2}{B}{p_2,l_2.r_2}} \and
-\infersplitone [\fut\mathrm{I}] 
-	{\splitone{\next~e}{\fut A}{((),p),l.r}}{\splittwosub A} \and
-\infersplittwo [\fut\mathrm{E}] 
-	{\splittwo{\prev~e}{A} {\pi_2~c,l.r}}{\splitonesub {\fut A}} \and
-\infersplitone [+\mathrm{E}]
-	{\splitone {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
-	{
-		\begin{array}{l}
-		\left(\tallif {\pi_1c_1}
+									\left(
+										\letin{y_1}{c_1}{\letin{y_2}{c_2}{
+										((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}
+									\right), \\
+									l.(\projbind 1,\projbind 2)
+									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
+\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),l.(\projbind 1,\projbind 2)}}					{\sub 1 A & \sub 2 B} \and
+\infersplitone [\times E_1]	{\spl {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}}					{\sub {} {A\times B}} \and
+\infersplitone [\times E_2]	{\spl {\pi_2~e}{B}{\letin{y}{c}{(\pi_2(\pi_1 y),\pi_2 y)},l.\pi_2~r}}					{\sub {} {A\times B}} \and
+\infersplittwo [\times E_1]	{\spl {\pi_1~e}{A}{p,l.\pi_1~r}}															{\sub {} {A\times B}} \and
+\infersplittwo [\times E_2]	{\spl {\pi_2~e}{B}{p,l.\pi_2~r}}															{\sub {} {A\times B}} \and
+\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{x:A^\bbone \in \Gamma} \and
+\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
+\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{\cdots}}														{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
+\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
+\infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} \and
+\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pi_2~c,l.r}}															{\splitonesub {} {\fut A}} \and
+\infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
+								{\begin{array}{l}
+									\left(\tallif {\pi_1 c_1}
 										{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
@@
 									\right), \\
-		l.\left(\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3}
-		\right)
-		\end{array}
-	}}
-	{\splitonesub [1] {\rm bool} & \splitonesub [2] A & \splitonesub [3] A } \and
-\infersplittwo [+\mathrm{E}]
-	{\splittwo {\left( \tallif {e_1}{e_2}{e_3}\right)}{A} 
-	{(p_1,p_2,p_3), 
-	l.\left(\tallif{(\letin {l_1}{\pi_1 l}{r_1})}{(\letin {l_2}{\pi_2 l}{r_2})} {(\letin {l_3}{\pi_3 l}{r_3})} \right)
-	}}
-	{\splittwosub [1] {\rm bool} & \splittwosub [2] A & \splittwosub [3] A } \and
+									l.(\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3})
+								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
+\infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
+								{(p_1,p_2,p_3),l.\left(\tallif{(\projbind 1)}{(\projbind 2)}{(\projbind 3)}\right)}}	{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
 \end{mathpar}
@@
 If $\splitone [\emptyC]{e}{A}{c,l.r}$ \\
-and $\reduceone{e}{v}$ \\
-and $c \Downarrow \valprod{t}{u}$ \\
-and $[u/l]r \Downarrow w$ \\
-then $v \vsplito [t,w]$
+and $\rsone{e}{v}{q}$ \\
+and $\reducetwo{q}{w}$ \\
+then $\reduce{c}{\valprod{v}{u}}$ \\
+and $\reduce{[u/l]r}{w}$
 \end{tabular}
@@
 If $\splittwo[\emptyC]{e}{B}{p,l.r}$ \\
-and $\reducetwo{e}{v}$ \\
-and $p \Downarrow u$ \\
-and $[u/l]r \Downarrow w$ \\
-then $v \vsplits w$
+and $\rstwo{e}{q}$ \\
+and $\reducetwo{q}{w}$ \\
+then $\reduce{p}{u}$ \\
+and $\reduce{[u/l]r}{w}$ 
 \end{tabular}
@@
 \item $\splitone{i}{\rm int}{(i,()),\_.()}$
+\item $\rsone{i}{v}{q}$
+\item $\reducetwo{r}{w}$
 \end{enumerate}
-Since reduction is a function, we also have
+By inversion of reduction relations, we know
 \begin{enumerate}
-\item $\reduceone{i}{i}$
-\item $\reduceone{(i,())}{\valprod {i}{()}}$
-\item $\reduceone{[()/\_]()}{()}$
+\item $v$ has the form $i$
+\item $q$ has the form $()$
+\item $w$ has the form $()$
 \end{enumerate}
-And since value splitting is a function,
+It is directly provable that,
 \begin{enumerate}
-\item $i \vsplito [i,()]$
+\item $\reduce{(i,())}{\valprod{i}{()}}$
+\item $\reduce{[()/l]()}{()}$
 \end{enumerate}
@@
 		\end{array}} $
-\item $\reduceone{(e_1,e_2)}{v}$
-\item $\reduce{\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}}{\valprod tu}$
-\item $\reduce{[u/l](\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})}{w}$
+\item $\rsone{(e_1,e_2)}{v}{q}$
+\item $\reducetwo{q}{w}$
 \end{enumerate}
-Since the splitting and evaluation rules are syntax-directed, it follows that
+By inversion of the splitting and reduction rules, we know
 \begin{enumerate}
-\item $\splitone {e_1}{A}{c_1,l_1.r_1}$
-\item $\splitone {e_2}{B}{c_2,l_2.r_2}$
-\item $v$ has the form $\valprod{v_1}{v_2}$, by inversion of tuple evaluation
-\item $\reduceone{(e_1,e_2)}{\valprod{v_1}{v_2}}$
-\item $\reduceone{e_1}{v_1}$
-\item $\reduceone{e_2}{v_2}$
-\item By inversion of evaluation, $t$ has the form $\valprod{t_1}{t_2}$  and $u$ has the form $\valprod{u_1}{u_2}$
-\item $\reduceone{[\valprod{t_1}{u_1},\valprod{t_2}{u_2}/y_1,y_2]((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}{\valprod tu}$
-\item $\reduceone{c_1}{\valprod {t_1}{u_1}}$, by inversion of let eval
-\item $\reduceone{c_2}{\valprod {t_2}{u_2}}$, by inversion of let eval
-\item By inversion of evaluation, $w$ has the form $\valprod{w_1}{w_2}$
-\item $\reduce{[\valprod{u_1}{u_2}/l](\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})}{\valprod{w_1}{w_2}}$, substituting for $u$, $w$
-\item $\reduce{[u_1/l_1]r_1}{w_1}$, inversion of evaluation
-\item $\reduce{[u_2/l_2]r_2}{w_2}$, inversion of evaluation
+\item $\splitone{e_1}{A}{c_1,l_1.r_1}$ and $\splitone{e_2}{B}{c_2,l_2.r_2}$
+\item $v$ has the form $\valprod{v_1}{v_2}$
+\item $q$ has the form $(q_1,q_2)$
+\item $\rsone{e_1}{v_1}{q_1}$ and $\rsone{e_2}{v_2}{q_2}$
+\item $w$ has the form $\valprod{w_1}{w_2}$
+\item $\reducetwo{q_1}{w_1}$ and $\reducetwo{q_2}{w_2}$
 \end{enumerate}
-We now have enough to invoke the inductive hypothesis,
+This gives us enough to invoke the inductive hypothesis for the each side, yielding
 \begin{enumerate}
-\item $v_1 \vsplito [t_1,w_1]$ 
-\item $v_2 \vsplito [t_2,w_2]$ 
-\item $\valprod{v_1}{v_2} \vsplito [\valprod{t_1}{t_2},\valprod{w_1}{w_2}]$ 
-\item $v \vsplito [t,w]$
+\item $\reduce{c_1}{\valprod{v_1}{u_1}}$ 
+\item $\reduce{[u_1/l_1]r_1}{w_1}$ 
+\item $\reduce{c_2}{\valprod{v_2}{u_2}}$ 
+\item $\reduce{[u_2/l_2]r_2}{w_2}$ 
 \end{enumerate}
-Which is what we need to show.
+It is now directly provable from the evaluation rules that
+\begin{enumerate}
+\item $\reduce{\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}}{\valprod {\valprod{v_1}{v_2}}{\valprod{u_1}{u_2}}}$
+\item $\reduce{[\valprod{u_1}{u_2}/l](\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})}{\valprod{w_1}{w_2}}$
+\end{enumerate}
+Which is what we wanted to show.
+
+\item {\bf Projection.} Suppose that
+\begin{enumerate}
+\item $\splitone {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}$
+\item $\rsone{\pi_1~e}{v}{q}$
+\item $\reducetwo{q}{w}$
+\end{enumerate}
+By inversion of the splitting and reduction rules, we have
+\begin{enumerate}
+\item $\splitone {e}{A \times B}{c,l.r}$
+\item $q$ has the form $\pi_1~q'$
+\item $\rsone{e}{\valprod{v}{v_2}}{q'}$
+\item $\reducetwo{q'}{\valprod{w}{w_2}}$
+\end{enumerate}
+This is enough to invoke our inductive hypothesis, yielding
+\begin{enumerate}
+\item $\reduce{c}{\valprod{\valprod{v}{v_2}}{u}}$
+\item $\reduce{[u/l]r}{\valprod{w}{w_2}}$
+\end{enumerate}
+From this, it can be directly proven that
+\begin{enumerate}
+\item $\reduce{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)}}{\valprod{v}{u}}$
+\item $\reduce{[u/l](\pi_1~r)}{w}$
+\end{enumerate}
+Which is what we wanted to show.
+
+\item {\bf If.} Suppose that
+\begin{enumerate}
+\item $\splitone {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
+								{\begin{array}{l}
+									\left(\tallif {\pi_1c_1}
+										{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
+										{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
+									\right), \\
+									l.\left(\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3}
+									\right)
+								\end{array}}$
+\item $\rsone{\ifthen {e_1}{e_2}{e_3}}{v}{q}$
+\item $\reducetwo{q}{w}$
+\end{enumerate}
+By inversion of the splitting rule, we have
+\begin{enumerate}
+\item $\splitonesub 1 \rmbool$ and $\splitonesub 2 A$ and $\splitonesub 3 A$
+\end{enumerate}
+There are two rules for evaluation of an if.  We proceed with the true case first:
+\begin{enumerate}
+\item $\rsone{e_2}{\bf true}{q_1}$ and (something about $q_1$) and $\rsone{e_2}{v}{q}$
+\item $\reduce{c_1}{\valprod{\bf true}{u_1}}$, by inductive hypothesis for 1
+\item $\reduce{[u_1/l_1]1_2}{w}$, by inductive hypothesis for 1
+\item $\reduce{c}{\valprod{v}{u_2}}$, by inductive hypothesis for 2
+\item $\reduce{[u_2/l_2]r_2}{w}$, by inductive hypothesis for 2
+\item $\reduce{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)}}{\valprod{v}{u}}$
+\end{enumerate}
+From this, it can be directly proven that
+\begin{enumerate}
+\item $\reduce{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)}}{\valprod{v}{u}}$
+\item $\reduce{[u/l](\pi_1~r)}{w}$
+\end{enumerate}
+Which is what we wanted to show.
+
 \end{itemize}
@@
 
-Not much to discuss.  It exsists.  Talk about how crufty the naive transformation is.
+Not much to discuss.  It exists.  Talk about how crufty the naive transformation is.
 

### 2014-02-21 paper/paper.tex
Worked out most of the proof for the if case.

@@
 \begin{figure}
-\caption{\corelang~Reference Semantics}
+\caption{\corelang~Diagonal Semantics}
 \label{fig:refSemantics}
@@
 \inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
-\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\\
+\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
+\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
 \inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
 \inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
-\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\\
+\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
+\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
 \inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
-\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\\
-\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
+\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
 \inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
+\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
+\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
 \inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
-\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
 \inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
-\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
 \inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
-\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{r_2}}							{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
-\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{r_3}}							{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\\
+\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
+\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
 \inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
@@
 									\right), \\
-									l.(\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3})
+									l.(\caseof{l}{l_2.r_2}{l_3.r_3})
 								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
@@
 									\right), \\
-									l.\left(\caseof{\pi_1 l}{l_2.r_2}{l_3.r_3}
+									l.\left(\caseof{l}{l_2.r_2}{l_3.r_3}
 									\right)
@@
 \end{enumerate}
-There are two rules for evaluation of an if.  We proceed with the true case first:
+There are two rules for evaluation of an if.  We proceed by inverting the if-true and evaluation rules first:
 \begin{enumerate}
-\item $\rsone{e_2}{\bf true}{q_1}$ and (something about $q_1$) and $\rsone{e_2}{v}{q}$
-\item $\reduce{c_1}{\valprod{\bf true}{u_1}}$, by inductive hypothesis for 1
-\item $\reduce{[u_1/l_1]1_2}{w}$, by inductive hypothesis for 1
-\item $\reduce{c}{\valprod{v}{u_2}}$, by inductive hypothesis for 2
-\item $\reduce{[u_2/l_2]r_2}{w}$, by inductive hypothesis for 2
-\item $\reduce{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)}}{\valprod{v}{u}}$
+\item $q$ has the form $(q_1;q_2)$
+\item $\reducetwo{q_1}{w_1}$ and $\reducetwo{q_2}{w}$
+\item $\rsone{e_1}{\bf true}{q_1}$ and $\rsone{e_2}{v}{q_2}$
+\end{enumerate}
+We can now invoke the inductive hypothesis for the predicate to yield
+\begin{enumerate}
+\item $\reduce{c_1}{\valprod{\bf true}{u_1}}$
+\item $\reduce{[u_1/l_1]q_1}{w_1}$
+\end{enumerate}
+We can also invoke the inductive hypothesis for the branch to yield
+\begin{enumerate}
+\item $\reduce{c_2}{\valprod{v}{u_2}}$
+\item $\reduce{[u_2/l_2]r_2}{w}$
 \end{enumerate}
@@
 \begin{enumerate}
-\item $\reduce{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)}}{\valprod{v}{u}}$
-\item $\reduce{[u/l](\pi_1~r)}{w}$
+\item $\reduce{
+		\left(\tallif {\pi_1c_1}
+			{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
+			{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
+		\right)}{\valprod{v}{\inl~u_2}}$
+\item $\reduce{[\inl~u_2/l](\caseof{l}{l_2.r_2}{l_3.r_3})}{w}$
 \end{enumerate}
-Which is what we wanted to show.
+Everything works similarly for the if-false case.  Either way, we have what we want to show.
 

### 2014-03-07 paper/paper.tex
Fixed up first stage splitting for let and if, in the theory.
Added rules for sums and functions.

@@
 \maketitle
+
+\input {grammars}
+
 \section{Introduction}
@@
 As always, where the language falls short, a programmer instead must bear the burden.
-Inevitably, this comes at the cost of expressivity, safety, or both.
-In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expresivity nor the safety that programmers have come to expect in single-domain languages.
+Inevitably, this comes at the cost of expressiveness, safety, or both.
+In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expressiveness nor the safety that programmers have come to expect in single-domain languages.
 
 Consider the programming of a pipeline system.
-Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reording in between.
-We can view an algorithm for such as system as one computation, split accross two spatial domains (processors).
+Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reordering in between.
+We can view an algorithm for such as system as one computation, split across two spatial domains (processors).
 For a concrete example, we draw inspiration from graphics systems.
@@
 a list of the scene primitives which interact with that part of the screen) whose results can be used to accelerate the final per-pixel calculation.
-In adition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.
+In addition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.
 
@@
 
-We'll start by analyzing a simple two-stage language called \corelang.
+We'll start by analyzing a simple two-stage language called \ellStaged.
 We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.
@@
 
-The grammar for types, terms, values, and contexts in \corelang are given in Figure \ref{fig:coreSyntax}.
+The grammar for types, terms, values, and contexts in \ellStaged are given in Figure \ref{fig:coreSyntax}.
 The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.
@@
 
-\begin{figure}
-\caption{\corelang~Syntax}
-\label{fig:coreSyntax}
-\centering
-\begin{tabular}{ll} 
-$\begin{aligned}
-\typeo &::= \text{unit}~|~\text{int}~|~\text{bool} \\
-&\gbar \typeo \times \typeo \\
-&\gbar \fut \typet 
-\end{aligned} $  
-& 
-$\begin{aligned}
-\typet &::=  \text{unit}~|~\text{int}~|~\text{bool} \\
-&\gbar \typet \times \typet \\
-&
-\end{aligned} $  
-\\ 
-$\begin{aligned}
-\expro &::= ()~|~\inte~|~\bool  \\
-&\gbar \letin{\var}{\expro}{\expro} \\
-&\gbar \var \\
-&\gbar (\expro, \expro) \\
-&\gbar \pi_1~\expro \gbar \pi_2~\expro \\
-&\gbar \ifthen {\expro}{\expro}{\expro} \\
-&\gbar \next~\exprt 
-\end{aligned} $ 
-& 
-$\begin{aligned}
-\exprt &::= ()~|~\inte~|~\bool \\
-&\gbar \letin{\var}{\exprt}{\exprt} \\
-&\gbar \var \\
-&\gbar (\exprt, \exprt) \\
-&\gbar \pi_1~\exprt \gbar \pi_2~\exprt \\
-&\gbar \ifthen {\exprt}{\exprt}{\exprt} \\
-&\gbar \prev~\expro
-\end{aligned} $
-\\ 
-$\begin{aligned}
-\valo &::= ()~|~\inte~|~\bool \\
-&\gbar \valprod {\valo} {\valo} \\
-&\gbar \valnext~\valt
-\end{aligned} $
-& 
-$\begin{aligned}
-\valt &::= ()~|~\inte~|~\bool \\
-&\gbar \valprod {\valt} {\valt} \\
-& 
-\end{aligned} $
-\\
-$\begin{aligned}
-\contextot &::=\emptyC \\
-&\gbar \contextot, \var : \typeo ^\bbone \\
-&\gbar \contextot, \var : \typet ^\bbtwo
-\end{aligned} $
-\end{tabular}
-\end{figure}
-
 \subsection{Static Semantics}
 
-The static semantics for \corelang~is given in Figure \ref{fig:coreStatics}.
+The static semantics for \ellStaged~is given in Figure \ref{fig:coreStatics}.
 Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).
@@
 \begin{figure}
-\caption{\corelang~Static Semantics}
+\caption{\ellStaged~Static Semantics}
 \label{fig:coreStatics}
@@
 
-The dynamics semantics for \corelang~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:
+The dynamics semantics for \ellStaged~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:
 
@@
 \begin{figure}
-\caption{\corelang~Erasure Dynamic Semantics}
+\caption{\ellStaged~Erasure Dynamic Semantics}
 \label{fig:erasureSemantics}
@@
 \begin{figure}
-\caption{\corelang~Diagonal Semantics}
+\caption{\ellStaged~Diagonal Semantics}
 \label{fig:refSemantics}
@@
 We now cover splitting, the process by which mixed code is teased apart into two individual chunks of code, one for each stage.
-To begin, we first introduce the target lagnuage (for both stage), \coremono.
-The grammar for \coremono~is given in figure \ref{fig:baseSyntax}.
-We see that \coremono~is (as one would hope) a traditional single-stage language, with all the same basetypes as \corelang, in addition to products, if expressions, and full sums (injections and case expressions).
-The presence of this last feature is a bit curious, since \corelang~did not support full sums.
+To begin, we first introduce the target lagnuage (for both stage), \ellTarget.
+The grammar for \ellTarget~is given in figure \ref{fig:baseSyntax}.
+We see that \ellTarget~is (as one would hope) a traditional single-stage language, with all the same basetypes as \ellStaged, in addition to products, if expressions, and full sums (injections and case expressions).
+The presence of this last feature is a bit curious, since \ellStaged~did not support full sums.
 The necessity for sums should become apparent in time.
-The static and dynamic semantics of \coremono~are not given, but should be obvious.
+The static and dynamic semantics of \ellTarget~are not given, but should be obvious.
 They are given by the judgements:
@@
 [...more...]
-
-\begin{figure}
-\caption{\coremono~Syntax}
-\label{fig:baseSyntax}
-\centering
-\begin{tabular}{ll} 
-$\begin{aligned}
-\expr &::= ()~|~\inte~|~\bool \\
-&\gbar \letin{\var}{\expr}{\expr} \\
-&\gbar \var \\
-&\gbar (\expr, \expr) \\
-&\gbar \pi_1~\expr \gbar \pi_2~\expr \\
-&\gbar \inl~\expr \gbar \inr~\expr \\
-&\gbar \ifthen {\expr}{\expr}{\expr}  \\
-&\gbar \caseof {\expr}{x_1.\expr}{x_2.\expr} 
-\end{aligned} $
-& 
-$\begin{aligned}
-\type &::=  \rmunit~|~\text{int}~|~\text{bool} \\
-&\gbar \type \times \type  \\
-&\gbar \type + \type 
-\\
-\val &::= ()~|~\inte~|~\bool \\
-&\gbar \valprod {\val} {\val}  \\
-&\gbar \valinl~\expr \gbar \valinr~\expr
-\\
-\context &::= \emptyC \\
-&\gbar \context, \var : \type
-\end{aligned} $
-\end{tabular}
-\end{figure}
 
@@
 \label{fig:termSplit}
-\newcommand {\projbind} [1] {\letin{l_{#1}}{\pi_{#1} l}{r_{#1}}}
 \begin{mathpar}
@@
 \infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
-\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{\cdots}}														{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
+\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{
+								\begin{array}{l}
+									\left(\talllet{y_1}{c_1}{
+									\talllet{y_2}{(\letin{x}{\pi_1 y_1}{c_2})}{
+									(\pi_1 y_2, (\pi_2 y_1, \pi_2 y_2))}}\right), \\
+									l.\letin{x}{(\projbind 1)}{\projbind 2}
+								\end{array}}}																					{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
 \infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
@@
 								{\begin{array}{l}
-									\left(\tallif {\pi_1 c_1}
-										{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
-										{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
+									\left(\talllet{z}{c_1}{\tallif {\pi_1 z}
+										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
+										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}}
 									\right), \\
-									l.(\caseof{l}{l_2.r_2}{l_3.r_3})
+									l.(\projbind 1;\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
 								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
@@
 
-\section {Implementation}
+\section {Adding Full Sums and Functions}
 
-Not much to discuss.  It exists.  Talk about how crufty the naive transformation is.
+In this section, we consider adding two more features to our source language: sums and functions.
+We will call this augmented language \lamStaged.
+The grammar for \lamStaged is given in \ref{fig:lamStagedSyntax}.
 
-\section{Related Work in Stage Seperation}
+\begin{figure*}
+\caption{Sum and Function Typing}
+\label{fig:advancedTypes}
+\begin{mathpar}
+\infertypesone [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbone] e B} 																\and
+\infertypesone [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
+\infertypestwo [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbtwo] e B} 																\and
+\infertypestwo [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
+\infertypesone [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
+\infertypesone [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\\
+\infertypestwo [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
+\infertypestwo [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\and
+\infertypesone [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbone]{e_2} C & \ty[\Gamma,x_3:A^\bbone]{e_3} C} 	\and
+\infertypestwo [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbtwo]{e_2} C & \ty[\Gamma,x_3:A^\bbtwo]{e_3} C} 	\and
+\end{mathpar}
+\end{figure*}
+
+\begin{figure*}
+\caption{Sum and Function Erasure Evaluation}
+\label{fig:advancedTypes}
+\begin{mathpar}
+\inferreduceone [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
+\inferreduceone [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
+\inferreducetwo [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
+\inferreducetwo [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
+\inferreduceone [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
+\inferreduceone [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
+\inferreducetwo [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
+\inferreducetwo [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
+\inferreduceone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
+\inferreduceone [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
+\inferreducetwo [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
+\inferreducetwo [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
+\end{mathpar}
+\end{figure*}
+
+\begin{figure*}
+\caption{Sum and Function Diagonal Evaluation}
+\label{fig:advancedTypes}
+\begin{mathpar}
+\inferRedsplitone [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{e}}{()}}			{\sub} 																	\and
+\inferSpeculation [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{r}}}				{\sub} 																	\and
+\inferRedsplitone [\to E]	{\red {e_1~e_2}{v'}{(r_1;\letin{x}{r_2}{r'})}}		{\red {e_1}{\vallam{x}{M}}{r_1} 
+																							& \sub [2] & \red{[[e,x]/x]M}{v'}{r'}} 							\and
+\inferSpeculation [\to E]	{\red {e_1~e_2}{r_1~r_2}}							{\sub [1] & \sub[2]}													\and
+\inferRedsplitone [+ I_1]	{\red {\inl~e} {\valinl~v}{r}}						{\sub} 																	\and
+\inferSpeculation [+ I_1]	{\red {\inl~e} {\valinl~r}}							{\sub} 																	\\
+\inferRedsplitone [+ I_2]	{\red {\inr~e} {\valinr~v}{r}}						{\sub} 																	\and
+\inferSpeculation [+ I_2]	{\red {\inr~e} {\valinr~r}}							{\sub} 																	\\
+\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
+								{\letin{x_2}{r_1}{r_2}}}								{\red {e_1}{\valinl~v}{r_1} & \red {[[v,x_2]/x_2]e_2}{v'}{r_2}} \and
+\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
+								{\letin{x_3}{r_1}{r_3}}}								{\red {e_1}{\valinr~v}{r_1} & \red {[[v,x_3]/x_3]e_3}{v'}{r_3}} \and
+\inferSpeculation [+ E]		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
+								{\caseof{r_1}{x_2.r_2}{x_3.r_3}}}				{\sub [1] & \sub [2] & \sub[3]}					 						\and
+\end{mathpar}
+\end{figure*}
+
+\begin{figure*}
+\caption{Sum and Function Splitting}
+\label{fig:advancedTypes}
+\begin{mathpar}
+\infersplitone	 [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{\vallam{x}{c},
+							\_.(\vallam{y}{\letin{x}{\pi_1 y}{\letin{l}{\pi_2 y}{r}}})}}		{\sub [\Gamma,x:A^\bbone] {} B} 						\and
+\infersplittwo [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{p,l.(\vallam{x}{M})}}				{\sub [\Gamma,x:A^\bbtwo] {} B} 						\and
+\infersplitone [\to E]	{\spl {e_1~e_2}{B}{\ldots}}											{\sub 1 {A \to B} & \sub 2 A}							\and
+\infersplittwo [\to E]	{\spl {e_1~e_2}{B}{(p_1,p_2),l.(\projbind 1)(\projbind 2)}}		{\sub 1 {A \to B} & \sub 2 A}							\and
+\infersplitone [+ I_1]	{\spl {\inl~e}{A+B}{\letin{y}{c}{(\inl(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
+\infersplitone [+ I_2]	{\spl {\inr~e}{A+B}{\letin{y}{c}{(\inr(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
+\infersplittwo [+ I_1]	{\spl {\inl~e}{A+B}{l.\inl~r}}										{\sub {} A}												\and
+\infersplittwo [+ I_1]	{\spl {\inr~e}{A+B}{l.\inr~r}}										{\sub {} A}												\and
+\infersplitone [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
+								{\begin{array}{l}\left(
+								\talllet{y_1}{c_1}{
+									\tallcase{\pi_1y_1}
+									{x_2.\letin{y_2}{c_2}{(\pi_1y_2,(\pi_2y_1,\inl(\pi_2y_2)))}}
+									{x_3.\letin{y_3}{c_3}{(\pi_1y_3,(\pi_2y_1,\inr(\pi_2y_3)))}}
+								}\right), \\
+								l.\left(\talllet{z}{(\projbind 1)}{
+									\tallcase{\pi_2 l}
+									{l_2.\letin{x_2}{z}{r_2}}
+									{l_3.\letin{x_3}{z}{r_3}}}\right)\end{array}}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbone] 2 C 
+																									& \sub [\Gamma,x_3:A^\bbone] 3 C} 						\and
+\infersplittwo [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
+								{(p_1,p_2,p_3), l.\left(\tallcase{(\projbind 1)}
+									{x_2.\projbind 2}{x_3.\projbind 3}\right)}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbtwo] 2 C 
+																									& \sub [\Gamma,x_3:A^\bbtwo] 3 C} 						\and
+\end{mathpar}
+\end{figure*}
+
+\section{Related Work in Stage Separation}
 

### 2014-03-12 paper/paper.tex
split stuff into more files; iterated on proofs

@@
 \usepackage{mathabx}
+\usepackage{float}
 \usepackage{cite}
@@
 \input {grammars}
+\input {semantics}
+\input {splitting}
 
@@
 
-\begin{figure}
-\caption{\ellStaged~Static Semantics}
-\label{fig:coreStatics}
-\begin{mathpar}
-\infertypesone [\rmunit] 	{\ty {()}\rmunit}						{\cdot} \and
-\infertypesone [int]			{\ty {i}{\rm int}}					{\cdot} \and
-\infertypesone [bool]			{\ty {b} \rmbool}						{\cdot} \and
-\infertypesone [hyp]			{\ty x A}								{x : A^\bbone \in \Gamma} \and
-\infertypestwo [\rmunit]		{\ty {()}{\rmunit}}					{\cdot} \and
-\infertypestwo [int]			{\ty {i}{\rm int}}					{\cdot} \and
-\infertypestwo [bool]			{\ty {b} \rmbool}						{\cdot} \and
-\infertypestwo [hyp]			{\ty x A}								{x : A^\bbtwo \in \Gamma} \and
-\infertypesone [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbone] {e_2} B} \and
-\infertypestwo [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbtwo] {e_2} B} \and
-\infertypesone [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \ty {e_2} B} \and
-\infertypesone [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
-\infertypesone [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
-\infertypestwo [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \typestwo {e_2} B} \and
-\infertypestwo [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
-\infertypestwo [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
-\infertypesone [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
-\infertypestwo [if]			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
-\infertypesone [\fut I]		{\ty {\next~e}{\fut A}}				{\typestwo e A} \and
-\infertypestwo [\fut E]		{\ty {\prev~e} A}						{\typesone e {\fut A}} \and
-\end{mathpar}
-\end{figure}
-
 \subsection{Erasure Dynamic Semantics}
@@
 
-\begin{figure}
-\caption{\ellStaged~Erasure Dynamic Semantics}
-\label{fig:erasureSemantics}
-\begin{mathpar}
-\inferreduceone [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
-\inferreduceone [int]				{\red {i}{i}}								{\cdot} 											\and
-\inferreduceone [bool]			{\red {b}{b}}								{\cdot} 											\and
-\inferreduceone [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
-\inferreducetwo [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
-\inferreducetwo [int]				{\red {i}{i}}								{\cdot} 											\and
-\inferreducetwo [bool]			{\red {b}{b}}								{\cdot} 											\and
-\inferreducetwo [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
-\inferreduceone [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]} 							\and
-\inferreduceone [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreduceone [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreducetwo [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]}								\and
-\inferreducetwo [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreducetwo [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreduceone [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
-\inferreduceone [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
-\inferreducetwo [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
-\inferreducetwo [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
-\inferreduceone [\fut I]			{\red {\next~e}{\valnext~v}}				{\reducetwosub}									\and
-\inferreducetwo [\fut E]			{\red {\prev~e} v}						{\reduceone e {\valnext~v}} 
-\end{mathpar}
-\end{figure}
-
-\begin{figure}
-\caption{\ellStaged~Diagonal Semantics}
-\label{fig:refSemantics}
-\begin{mathpar}
-\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
-\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
-\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
-\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
-\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
-\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
-\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
-\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
-\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
-\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
-\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
-\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
-\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
-\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
-\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
-\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
-\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
-\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
-\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
-\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
-\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
-\end{mathpar}
-\end{figure}
-
-\subsection{Splitting}
+\section{The Stage-Splitting Transformation}
 
@@
 [...more...]
-
-\begin{figure*}
-\caption{Type Splitting}
-\label{fig:typeSplit}
-\begin{mathpar}
-\infer [\mathrm{unit}\tsplito]	{\rmunit \tsplito [\rmunit,\rmunit]}{\cdot} \and
-\infer [\mathrm{int}\tsplito]	{{\rm int} \tsplito [{\rm int}, \rmunit]}{\cdot} \and
-\infer [\mathrm{bool}\tsplito]	{{\rm bool} \tsplito [{\rm bool}, \rmunit]}{\cdot} \and
-\infer [\times\tsplito]
-	{A\times B \tsplito [A_1 \times B_1, A_2 \times B_2]}
-	{A \tsplito [A_1, A_2] & B \tsplito [B_1,B_2]} \and
-\infer [\fut\tsplito]
-	{\fut~A \tsplito [{\rm unit}, A']}
-	{A \tsplits A'} \and
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Value Splitting}
-\label{fig:valSplit}
-\begin{mathpar}
-\infer [\rmunit\vsplito]		{() \vsplito [(), ()]}{\cdot} \and
-\infer [\mathrm{int}\vsplito]	{i \vsplito [i,()]}{\cdot} \and
-\infer [\mathrm{bool}\vsplito]	{b \vsplito [b,()]}{\cdot} \and
-\infer [\times\vsplito]	
-	{\valprod{v_1}{v_2} \vsplito [\valprod{u_1}{u_2}, \valprod{w_1}{w_2}]}
-	{v_1 \vsplito [u_1,w_1] & v_2 \vsplito [u_2,w_2]} \and
-\infer [\fut\vsplito]
-	{\valnext~v \vsplito [(),v']}
-	{v \vsplits v'} 
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Context Splitting}
-\label{fig:contSplit}
-\begin{mathpar}
-\infer {\emptyC \csplit [\emptyC~;~\emptyC]}{\cdot} \and
-\infer {\Gamma, v : A^\bbone \csplit [\Gamma_1,v:A_1; \Gamma_2,v:A_2]}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplito [A_1, A_2]} \and
-\infer {\Gamma, v : A^\bbtwo \csplit [\Gamma_1; \Gamma_2,v:A']}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplits A'} 
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Term Splitting}
-\label{fig:termSplit}
-\begin{mathpar}
-\infersplitone [unit]			{\spl {()}{\rmunit}{((),()),\_.()}}															{\cdot} \and
-\infersplittwo [unit]			{\spl {()}{\rmunit}{(),\_.()}}																{\cdot} \and
-\infersplitone [int]			{\spl {i}{\rm int}{(i,()),\_.()}}															{\cdot} \and
-\infersplittwo [int]			{\spl {i}{\rm int}{(),\_.i}}																	{\cdot} \and
-\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
-									{ \begin{array}{l}
-									\left(
-										\letin{y_1}{c_1}{\letin{y_2}{c_2}{
-										((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}
-									\right), \\
-									l.(\projbind 1,\projbind 2)
-									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
-\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),l.(\projbind 1,\projbind 2)}}					{\sub 1 A & \sub 2 B} \and
-\infersplitone [\times E_1]	{\spl {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}}					{\sub {} {A\times B}} \and
-\infersplitone [\times E_2]	{\spl {\pi_2~e}{B}{\letin{y}{c}{(\pi_2(\pi_1 y),\pi_2 y)},l.\pi_2~r}}					{\sub {} {A\times B}} \and
-\infersplittwo [\times E_1]	{\spl {\pi_1~e}{A}{p,l.\pi_1~r}}															{\sub {} {A\times B}} \and
-\infersplittwo [\times E_2]	{\spl {\pi_2~e}{B}{p,l.\pi_2~r}}															{\sub {} {A\times B}} \and
-\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{x:A^\bbone \in \Gamma} \and
-\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
-\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{
-								\begin{array}{l}
-									\left(\talllet{y_1}{c_1}{
-									\talllet{y_2}{(\letin{x}{\pi_1 y_1}{c_2})}{
-									(\pi_1 y_2, (\pi_2 y_1, \pi_2 y_2))}}\right), \\
-									l.\letin{x}{(\projbind 1)}{\projbind 2}
-								\end{array}}}																					{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
-\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
-\infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} \and
-\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pi_2~c,l.r}}															{\splitonesub {} {\fut A}} \and
-\infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
-								{\begin{array}{l}
-									\left(\talllet{z}{c_1}{\tallif {\pi_1 z}
-										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
-										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}}
-									\right), \\
-									l.(\projbind 1;\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
-								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
-\infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
-								{(p_1,p_2,p_3),l.\left(\tallif{(\projbind 1)}{(\projbind 2)}{(\projbind 3)}\right)}}	{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
-\end{mathpar}
-\end{figure*}
 
@@
 
+\section {Adding Full Sums and Functions}
+
+In this section, we consider adding two more features to our source language: sums and functions.
+We will call this augmented language \lamStaged.
+The grammar for \lamStaged is given in \ref{fig:lamStagedSyntax}.
+
+\section{Related Work in Stage Separation}
+
+Our starting language, $\lambda^{12}$, is essentially a two-stage version of $\lambda^\bigcirc$ from \cite{davies96}, but beefed up with products, sums, and fixed-points.
+That work motivates $\lambda^\bigcirc$ as an image of linear temporal logic under the Curry-Howard correspondence, and it makes the argument that $\lambda^\bigcirc$ is a good model for binding time analysis.
+Although $\lambda^\bigcirc$ was designed with partial evaluation in mind, it turns out to be a good model for pass separation as well.
+In contrast, \cite{davies01} presents $\lambda^\Box$, which is based off of branching temporal logic.
+As that work shows, $\lambda^\Box$ is a good system for program generation, although we find the closed-code requirement to be more restricting than we need.
+
+Pass seperation was introduced in \cite{jorring}.
+They focused on motivating pass separation as a technique for compiler generation, and hinted at wider applicability.
+Their approach was not mechanized or automated in any sense, and their examples were separated by hand.
+We believe that out work represents the fulfillment of their prediction that ``the [pass separation] approach will elude full automation for some time."
+
+\cite{hannan94} also uses a pass separation technique for generating compiler/evaluator pairs from interpretters specified as term-rewrite systems.
+This heavy restriction on the form of the input prevents generalization of their style of pass separation.
+[I don't know enough about term-rewrite systems yet to be able to comment further.]  Their method, while amenable to mechanization, was not actually implemented.
+
+\cite{knoblock} represents the first attempt at fully automatic pass separation for code approaching general-purpose.
+In particular, they implemented pass separation for a C-like shading language including basic arithmetic and if statements.
+The main goal of their work, like ours, was to minimize recomputation.
+That said, their starting language was sufficiently restricted for recursive boundary types to be inexpressible.
+Given that memory is at a premium in shading languages, so much of that work was also focused on minimizing the memory-footprint of the boundary type.
+
+\section{Appendix}
+
+\subsection {Equivalence of Diagonal Evaluation and Splitting (for \ellStaged)}
+
 We proceed by cases of the term splitting judgement.
@@
 \begin{itemize}
+% INT, BOOL, UNIT
 \item {\bf Int.} Suppose that
@@
 \item $\rsone{i}{v}{q}$
-\item $\reducetwo{r}{w}$
+\item $\reducetwo{q}{w}$
 \end{enumerate}
@@
 \item $\reduce{(i,())}{\valprod{i}{()}}$
-\item $\reduce{[()/l]()}{()}$
+\item $\reduce{[()/\_]()}{()}$
 \end{enumerate}
@@
 
+\item {\bf Boolean.} Same as int.
+\item {\bf Unit.} Same as int.
+
+% TUPLE
 \item {\bf Tuple.} Suppose that
@@
 
-\item {\bf Projection.} Suppose that
+% PROJECTIONS
+\item {\bf Left Projection.} Suppose that
 \begin{enumerate}
@@
 
+\item {\bf Right Projection.} Same as left projection.
+
+% IF
 \item {\bf If.} Suppose that
@@
 								{\begin{array}{l}
-									\left(\tallif {\pi_1c_1}
-										{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
-										{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
-									\right), \\
-									l.\left(\caseof{l}{l_2.r_2}{l_3.r_3}
-									\right)
+									\left(
+										\talllet{z}{c_1}{
+										\tallif {\pi_1 z}
+										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
+										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}
+									}\right), \\
+									l.(\letin{l_1}{\pi_1 l}{r_1};\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
 								\end{array}}$
@@
 \end{enumerate}
-There are two rules for evaluation of an if.  We proceed by inverting the if-true and evaluation rules first:
+There are two rules for evaluation of an if.  We proceed by inverting the if-true evaluation rules first:
 \begin{enumerate}
 \item $q$ has the form $(q_1;q_2)$
+\item $\rsone{e_1}{\bf true}{q_1}$ and $\rsone{e_2}{v}{q_2}$
 \item $\reducetwo{q_1}{w_1}$ and $\reducetwo{q_2}{w}$
-\item $\rsone{e_1}{\bf true}{q_1}$ and $\rsone{e_2}{v}{q_2}$
 \end{enumerate}
@@
 \item $\reduce{c_1}{\valprod{\bf true}{u_1}}$
-\item $\reduce{[u_1/l_1]q_1}{w_1}$
+\item $\reduce{[u_1/l_1]r_1}{w_1}$
 \end{enumerate}
@@
 \item $\reduce{
-		\left(\tallif {\pi_1c_1}
-			{\letin{y}{c_2}{(\pi_1 y, \inl(\pi_2 y))}}
-			{\letin{y}{c_3}{(\pi_1 y, \inr(\pi_2 y))}}
-		\right)}{\valprod{v}{\inl~u_2}}$
-\item $\reduce{[\inl~u_2/l](\caseof{l}{l_2.r_2}{l_3.r_3})}{w}$
+		\left(
+			\talllet{z}{c_1}{
+			\tallif {\pi_1 z}
+			{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
+			{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}
+		}\right)}{\valprod{v}{\valprod{u_1}{\inl~u_2}}}$
+\item $\reduce{[\valprod{u_1}{\inl~u_2}/l](\letin{l_1}{\pi_1 l}{r_1};\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})}{w}$
 \end{enumerate}
@@
 
+% NEXT
+\item {\bf Next.} Suppose that
+\begin{enumerate}
+\item $\splitone{\next~e}{\fut A}{((),p),l.r}$
+\item $\rsone{\next~e}{v}{q}$
+\item $\reducetwo{q}{w}$
+\end{enumerate}
+By inversion of the splitting and reduction rules, we have
+\begin{enumerate}
+\item $\splittwo {e}{A}{p,l.r}$
+\item $v$ has the form $()$
+\item $\rstwo{e}{q}$
+\end{enumerate}
+We can now invoke the stage $\bbtwo$ part of the inductive hypothesis to yield,
+\begin{enumerate}
+\item $\reduce {p}{u}$
+\item $\reduce {[u/l]r}{w}$
+\end{enumerate}
+It can then be directly proven that,
+\begin{enumerate}
+\item $\reduce {((),p)}{((),u)}$
+\end{enumerate}
+Which (along with the previous line) is what we wanted to show.
+
 \end{itemize}
 
-\section {Adding Full Sums and Functions}
-
-In this section, we consider adding two more features to our source language: sums and functions.
-We will call this augmented language \lamStaged.
-The grammar for \lamStaged is given in \ref{fig:lamStagedSyntax}.
-
-\begin{figure*}
-\caption{Sum and Function Typing}
-\label{fig:advancedTypes}
-\begin{mathpar}
-\infertypesone [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbone] e B} 																\and
-\infertypesone [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
-\infertypestwo [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbtwo] e B} 																\and
-\infertypestwo [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
-\infertypesone [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
-\infertypesone [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\\
-\infertypestwo [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
-\infertypestwo [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\and
-\infertypesone [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbone]{e_2} C & \ty[\Gamma,x_3:A^\bbone]{e_3} C} 	\and
-\infertypestwo [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbtwo]{e_2} C & \ty[\Gamma,x_3:A^\bbtwo]{e_3} C} 	\and
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Sum and Function Erasure Evaluation}
-\label{fig:advancedTypes}
-\begin{mathpar}
-\inferreduceone [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
-\inferreduceone [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
-\inferreducetwo [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
-\inferreducetwo [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
-\inferreduceone [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
-\inferreduceone [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
-\inferreducetwo [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
-\inferreducetwo [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
-\inferreduceone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
-\inferreduceone [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
-\inferreducetwo [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
-\inferreducetwo [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Sum and Function Diagonal Evaluation}
-\label{fig:advancedTypes}
-\begin{mathpar}
-\inferRedsplitone [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{e}}{()}}			{\sub} 																	\and
-\inferSpeculation [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{r}}}				{\sub} 																	\and
-\inferRedsplitone [\to E]	{\red {e_1~e_2}{v'}{(r_1;\letin{x}{r_2}{r'})}}		{\red {e_1}{\vallam{x}{M}}{r_1} 
-																							& \sub [2] & \red{[[e,x]/x]M}{v'}{r'}} 							\and
-\inferSpeculation [\to E]	{\red {e_1~e_2}{r_1~r_2}}							{\sub [1] & \sub[2]}													\and
-\inferRedsplitone [+ I_1]	{\red {\inl~e} {\valinl~v}{r}}						{\sub} 																	\and
-\inferSpeculation [+ I_1]	{\red {\inl~e} {\valinl~r}}							{\sub} 																	\\
-\inferRedsplitone [+ I_2]	{\red {\inr~e} {\valinr~v}{r}}						{\sub} 																	\and
-\inferSpeculation [+ I_2]	{\red {\inr~e} {\valinr~r}}							{\sub} 																	\\
-\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
-								{\letin{x_2}{r_1}{r_2}}}								{\red {e_1}{\valinl~v}{r_1} & \red {[[v,x_2]/x_2]e_2}{v'}{r_2}} \and
-\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
-								{\letin{x_3}{r_1}{r_3}}}								{\red {e_1}{\valinr~v}{r_1} & \red {[[v,x_3]/x_3]e_3}{v'}{r_3}} \and
-\inferSpeculation [+ E]		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
-								{\caseof{r_1}{x_2.r_2}{x_3.r_3}}}				{\sub [1] & \sub [2] & \sub[3]}					 						\and
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Sum and Function Splitting}
-\label{fig:advancedTypes}
-\begin{mathpar}
-\infersplitone	 [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{\vallam{x}{c},
-							\_.(\vallam{y}{\letin{x}{\pi_1 y}{\letin{l}{\pi_2 y}{r}}})}}		{\sub [\Gamma,x:A^\bbone] {} B} 						\and
-\infersplittwo [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{p,l.(\vallam{x}{M})}}				{\sub [\Gamma,x:A^\bbtwo] {} B} 						\and
-\infersplitone [\to E]	{\spl {e_1~e_2}{B}{\ldots}}											{\sub 1 {A \to B} & \sub 2 A}							\and
-\infersplittwo [\to E]	{\spl {e_1~e_2}{B}{(p_1,p_2),l.(\projbind 1)(\projbind 2)}}		{\sub 1 {A \to B} & \sub 2 A}							\and
-\infersplitone [+ I_1]	{\spl {\inl~e}{A+B}{\letin{y}{c}{(\inl(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
-\infersplitone [+ I_2]	{\spl {\inr~e}{A+B}{\letin{y}{c}{(\inr(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
-\infersplittwo [+ I_1]	{\spl {\inl~e}{A+B}{l.\inl~r}}										{\sub {} A}												\and
-\infersplittwo [+ I_1]	{\spl {\inr~e}{A+B}{l.\inr~r}}										{\sub {} A}												\and
-\infersplitone [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
-								{\begin{array}{l}\left(
-								\talllet{y_1}{c_1}{
-									\tallcase{\pi_1y_1}
-									{x_2.\letin{y_2}{c_2}{(\pi_1y_2,(\pi_2y_1,\inl(\pi_2y_2)))}}
-									{x_3.\letin{y_3}{c_3}{(\pi_1y_3,(\pi_2y_1,\inr(\pi_2y_3)))}}
-								}\right), \\
-								l.\left(\talllet{z}{(\projbind 1)}{
-									\tallcase{\pi_2 l}
-									{l_2.\letin{x_2}{z}{r_2}}
-									{l_3.\letin{x_3}{z}{r_3}}}\right)\end{array}}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbone] 2 C 
-																									& \sub [\Gamma,x_3:A^\bbone] 3 C} 						\and
-\infersplittwo [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
-								{(p_1,p_2,p_3), l.\left(\tallcase{(\projbind 1)}
-									{x_2.\projbind 2}{x_3.\projbind 3}\right)}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbtwo] 2 C 
-																									& \sub [\Gamma,x_3:A^\bbtwo] 3 C} 						\and
-\end{mathpar}
-\end{figure*}
-
-\section{Related Work in Stage Separation}
-
-Our starting language, $\lambda^{12}$, is essentially a two-stage version of $\lambda^\bigcirc$ from \cite{davies96}, but beefed up with products, sums, and fixed-points.
-That work motivates $\lambda^\bigcirc$ as an image of linear temporal logic under the Curry-Howard correspondence, and it makes the argument that $\lambda^\bigcirc$ is a good model for binding time analysis.
-Although $\lambda^\bigcirc$ was designed with partial evaluation in mind, it turns out to be a good model for pass separation as well.
-In contrast, \cite{davies01} presents $\lambda^\Box$, which is based off of branching temporal logic.
-As that work shows, $\lambda^\Box$ is a good system for program generation, although we find the closed-code requirement to be more restricting than we need.
-
-Pass seperation was introduced in \cite{jorring}.
-They focused on motivating pass separation as a technique for compiler generation, and hinted at wider applicability.
-Their approach was not mechanized or automated in any sense, and their examples were separated by hand.
-We believe that out work represents the fulfillment of their prediction that ``the [pass separation] approach will elude full automation for some time."
-
-\cite{hannan94} also uses a pass separation technique for generating compiler/evaluator pairs from interpretters specified as term-rewrite systems.
-This heavy restriction on the form of the input prevents generalization of their style of pass separation.
-[I don't know enough about term-rewrite systems yet to be able to comment further.]  Their method, while amenable to mechanization, was not actually implemented.
-
-\cite{knoblock} represents the first attempt at fully automatic pass separation for code approaching general-purpose.
-In particular, they implemented pass separation for a C-like shading language including basic arithmetic and if statements.
-The main goal of their work, like ours, was to minimize recomputation.
-That said, their starting language was sufficiently restricted for recursive boundary types to be inexpressible.
-Given that memory is at a premium in shading languages, so much of that work was also focused on minimizing the memory-footprint of the boundary type.
+\subsection {Equivalence of Diagonal Evaluation and Splitting (extra \lamStaged~cases)}
 

### 2014-03-12 paper/semantics.tex
split stuff into more files; iterated on proofs

\begin{figure}
\caption{\ellStaged~Static Semantics}
\label{fig:coreStatics}
\begin{mathpar}
\infertypesone [\rmunit] 	{\ty {()}\rmunit}						{\cdot} \and
\infertypesone [int]			{\ty {i}{\rm int}}					{\cdot} \and
\infertypesone [bool]			{\ty {b} \rmbool}						{\cdot} \and
\infertypesone [hyp]			{\ty x A}								{x : A^\bbone \in \Gamma} \and
\infertypestwo [\rmunit]		{\ty {()}{\rmunit}}					{\cdot} \and
\infertypestwo [int]			{\ty {i}{\rm int}}					{\cdot} \and
\infertypestwo [bool]			{\ty {b} \rmbool}						{\cdot} \and
\infertypestwo [hyp]			{\ty x A}								{x : A^\bbtwo \in \Gamma} \and
\infertypesone [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbone] {e_2} B} \and
\infertypestwo [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbtwo] {e_2} B} \and
\infertypesone [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \ty {e_2} B} \and
\infertypesone [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
\infertypesone [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
\infertypestwo [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \typestwo {e_2} B} \and
\infertypestwo [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
\infertypestwo [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
\infertypesone [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
\infertypestwo [if]			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
\infertypesone [\fut I]		{\ty {\next~e}{\fut A}}				{\typestwo e A} \and
\infertypestwo [\fut E]		{\ty {\prev~e} A}						{\typesone e {\fut A}} \and
\end{mathpar}
\end{figure}

\begin{figure*}
\caption{Sum and Function Typing}
\label{fig:advancedTypes}
\begin{mathpar}
\infertypesone [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbone] e B} 																\and
\infertypesone [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
\infertypestwo [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbtwo] e B} 																\and
\infertypestwo [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
\infertypesone [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
\infertypesone [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\\
\infertypestwo [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
\infertypestwo [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\and
\infertypesone [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbone]{e_2} C & \ty[\Gamma,x_3:A^\bbone]{e_3} C} 	\and
\infertypestwo [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbtwo]{e_2} C & \ty[\Gamma,x_3:A^\bbtwo]{e_3} C} 	\and
\end{mathpar}
\end{figure*}

\begin{figure}
\caption{\ellStaged~Erasure Dynamic Semantics}
\label{fig:erasureSemantics}
\begin{mathpar}
\inferreduceone [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
\inferreduceone [int]				{\red {i}{i}}								{\cdot} 											\and
\inferreduceone [bool]			{\red {b}{b}}								{\cdot} 											\and
\inferreduceone [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
\inferreducetwo [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
\inferreducetwo [int]				{\red {i}{i}}								{\cdot} 											\and
\inferreducetwo [bool]			{\red {b}{b}}								{\cdot} 											\and
\inferreducetwo [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
\inferreduceone [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]} 							\and
\inferreduceone [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreduceone [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreducetwo [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]}								\and
\inferreducetwo [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreducetwo [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
\inferreduceone [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
\inferreduceone [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
\inferreducetwo [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
\inferreducetwo [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
\inferreduceone [\fut I]			{\red {\next~e}{\valnext~v}}				{\reducetwosub}									\and
\inferreducetwo [\fut E]			{\red {\prev~e} v}						{\reduceone e {\valnext~v}} 
\end{mathpar}
\end{figure}

\begin{figure*}
\caption{Sum and Function Erasure Evaluation}
\label{fig:advancedTypes}
\begin{mathpar}
\inferreduceone [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
\inferreduceone [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
\inferreducetwo [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
\inferreducetwo [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
\inferreduceone [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
\inferreduceone [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
\inferreducetwo [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
\inferreducetwo [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
\inferreduceone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
\inferreduceone [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
\inferreducetwo [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
\inferreducetwo [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
\end{mathpar}
\end{figure*}

\begin{figure}
\caption{\ellStaged~Diagonal Semantics}
\label{fig:refSemantics}
\begin{mathpar}
\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
\end{mathpar}
\end{figure}

\begin{figure*}
\caption{Sum and Function Diagonal Evaluation}
\label{fig:advancedTypes}
\begin{mathpar}
\inferRedsplitone [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{e}}{()}}			{\cdot} 																\and
\inferSpeculation [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{r}}}				{\sub} 																	\and
\inferRedsplitone [\to E]	{\red {e_1~e_2}{v'}{(r_1;\letin{x}{r_2}{r'})}}		{\red {e_1}{\vallam{x}{M}}{r_1} 
																							& \sub [2] & \red{[[e,x]/x]M}{v'}{r'}} 							\and
\inferSpeculation [\to E]	{\red {e_1~e_2}{r_1~r_2}}							{\sub [1] & \sub[2]}													\and
\inferRedsplitone [+ I_1]	{\red {\inl~e} {\valinl~v}{r}}						{\sub} 																	\and
\inferSpeculation [+ I_1]	{\red {\inl~e} {\valinl~r}}							{\sub} 																	\\
\inferRedsplitone [+ I_2]	{\red {\inr~e} {\valinr~v}{r}}						{\sub} 																	\and
\inferSpeculation [+ I_2]	{\red {\inr~e} {\valinr~r}}							{\sub} 																	\\
\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
								{\letin{x_2}{r_1}{r_2}}}								{\red {e_1}{\valinl~v}{r_1} & \red {[[v,x_2]/x_2]e_2}{v'}{r_2}} \and
\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
								{\letin{x_3}{r_1}{r_3}}}								{\red {e_1}{\valinr~v}{r_1} & \red {[[v,x_3]/x_3]e_3}{v'}{r_3}} \and
\inferSpeculation [+ E]		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
								{\caseof{r_1}{x_2.r_2}{x_3.r_3}}}				{\sub [1] & \sub [2] & \sub[3]}					 						\and
\end{mathpar}
\end{figure*}

### 2014-03-12 paper/splitting.tex
split stuff into more files; iterated on proofs

\begin{figure*}
\caption{Type Splitting}
\label{fig:typeSplit}
\begin{mathpar}
\infer [\mathrm{unit}\tsplito]	{\rmunit \tsplito [\rmunit,\rmunit]}{\cdot} \and
\infer [\mathrm{int}\tsplito]	{{\rm int} \tsplito [{\rm int}, \rmunit]}{\cdot} \and
\infer [\mathrm{bool}\tsplito]	{{\rm bool} \tsplito [{\rm bool}, \rmunit]}{\cdot} \and
\infer [\times\tsplito]
	{A\times B \tsplito [A_1 \times B_1, A_2 \times B_2]}
	{A \tsplito [A_1, A_2] & B \tsplito [B_1,B_2]} \and
\infer [\fut\tsplito]
	{\fut~A \tsplito [{\rm unit}, A']}
	{A \tsplits A'} \and
\end{mathpar}
\end{figure*}

\begin{figure*}
\caption{Value Splitting}
\label{fig:valSplit}
\begin{mathpar}
\infer [\rmunit\vsplito]		{() \vsplito [(), ()]}{\cdot} \and
\infer [\mathrm{int}\vsplito]	{i \vsplito [i,()]}{\cdot} \and
\infer [\mathrm{bool}\vsplito]	{b \vsplito [b,()]}{\cdot} \and
\infer [\times\vsplito]	
	{\valprod{v_1}{v_2} \vsplito [\valprod{u_1}{u_2}, \valprod{w_1}{w_2}]}
	{v_1 \vsplito [u_1,w_1] & v_2 \vsplito [u_2,w_2]} \and
\infer [\fut\vsplito]
	{\valnext~v \vsplito [(),v']}
	{v \vsplits v'} 
\end{mathpar}
\end{figure*}

\begin{figure*}
\caption{Context Splitting}
\label{fig:contSplit}
\begin{mathpar}
\infer {\emptyC \csplit [\emptyC~;~\emptyC]}{\cdot} \and
\infer {\Gamma, v : A^\bbone \csplit [\Gamma_1,v:A_1; \Gamma_2,v:A_2]}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplito [A_1, A_2]} \and
\infer {\Gamma, v : A^\bbtwo \csplit [\Gamma_1; \Gamma_2,v:A']}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplits A'} 
\end{mathpar}
\end{figure*}

\begin{figure*}
\caption{Term Splitting}
\label{fig:termSplit}
\begin{mathpar}
\infersplitone [unit]			{\spl {()}{\rmunit}{((),()),\_.()}}															{\cdot} \and
\infersplittwo [unit]			{\spl {()}{\rmunit}{(),\_.()}}																{\cdot} \and
\infersplitone [int]			{\spl {i}{\rm int}{(i,()),\_.()}}															{\cdot} \and
\infersplittwo [int]			{\spl {i}{\rm int}{(),\_.i}}																	{\cdot} \and
\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
									{ \begin{array}{l}
									\left(
										\letin{y_1}{c_1}{\letin{y_2}{c_2}{
										((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}
									\right), \\
									l.(\projbind 1,\projbind 2)
									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),l.(\projbind 1,\projbind 2)}}					{\sub 1 A & \sub 2 B} \and
\infersplitone [\times E_1]	{\spl {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}}					{\sub {} {A\times B}} \and
\infersplitone [\times E_2]	{\spl {\pi_2~e}{B}{\letin{y}{c}{(\pi_2(\pi_1 y),\pi_2 y)},l.\pi_2~r}}					{\sub {} {A\times B}} \and
\infersplittwo [\times E_1]	{\spl {\pi_1~e}{A}{p,l.\pi_1~r}}															{\sub {} {A\times B}} \and
\infersplittwo [\times E_2]	{\spl {\pi_2~e}{B}{p,l.\pi_2~r}}															{\sub {} {A\times B}} \and
\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{x:A^\bbone \in \Gamma} \and
\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{
								\begin{array}{l}
									\left(\talllet{y_1}{c_1}{
									\talllet{y_2}{(\letin{x}{\pi_1 y_1}{c_2})}{
									(\pi_1 y_2, (\pi_2 y_1, \pi_2 y_2))}}\right), \\
									l.\letin{x}{(\projbind 1)}{\projbind 2}
								\end{array}}}																					{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
\infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} \and
\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pi_2~c,l.r}}															{\splitonesub {} {\fut A}} \and
\infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
								{\begin{array}{l}
									\left(\talllet{z}{c_1}{\tallif {\pi_1 z}
										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}}
									\right), \\
									l.(\projbind 1;\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
\infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
								{(p_1,p_2,p_3),l.\left(\tallif{(\projbind 1)}{(\projbind 2)}{(\projbind 3)}\right)}}	{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
\end{mathpar}
\end{figure*}

\begin{figure*}
\caption{Sum and Function Splitting}
\label{fig:advancedTypes}
\begin{mathpar}
\infersplitone	 [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{\vallam{x}{c},
							\_.(\vallam{y}{\letin{x}{\pi_1 y}{\letin{l}{\pi_2 y}{r}}})}}		{\sub [\Gamma,x:A^\bbone] {} B} 						\and
\infersplittwo [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{p,l.(\vallam{x}{M})}}				{\sub [\Gamma,x:A^\bbtwo] {} B} 						\and
\infersplitone [\to E]	{\spl {e_1~e_2}{B}{\ldots}}											{\sub 1 {A \to B} & \sub 2 A}							\and
\infersplittwo [\to E]	{\spl {e_1~e_2}{B}{(p_1,p_2),l.(\projbind 1)(\projbind 2)}}		{\sub 1 {A \to B} & \sub 2 A}							\and
\infersplitone [+ I_1]	{\spl {\inl~e}{A+B}{\letin{y}{c}{(\inl(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
\infersplitone [+ I_2]	{\spl {\inr~e}{A+B}{\letin{y}{c}{(\inr(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
\infersplittwo [+ I_1]	{\spl {\inl~e}{A+B}{l.\inl~r}}										{\sub {} A}												\and
\infersplittwo [+ I_1]	{\spl {\inr~e}{A+B}{l.\inr~r}}										{\sub {} A}												\and
\infersplitone [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
								{\begin{array}{l}\left(
								\talllet{y_1}{c_1}{
									\tallcase{\pi_1y_1}
									{x_2.\letin{y_2}{c_2}{(\pi_1y_2,(\pi_2y_1,\inl(\pi_2y_2)))}}
									{x_3.\letin{y_3}{c_3}{(\pi_1y_3,(\pi_2y_1,\inr(\pi_2y_3)))}}
								}\right), \\
								l.\left(\talllet{z}{(\projbind 1)}{
									\tallcase{\pi_2 l}
									{l_2.\letin{x_2}{z}{r_2}}
									{l_3.\letin{x_3}{z}{r_3}}}\right)\end{array}}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbone] 2 C 
																									& \sub [\Gamma,x_3:A^\bbone] 3 C} 						\and
\infersplittwo [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
								{(p_1,p_2,p_3), l.\left(\tallcase{(\projbind 1)}
									{x_2.\projbind 2}{x_3.\projbind 3}\right)}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbtwo] 2 C 
																									& \sub [\Gamma,x_3:A^\bbtwo] 3 C} 						\and
\end{mathpar}
\end{figure*}

### 2014-05-10 paper/sigplanconf-template.tex
huge backlog; pattern stuff and a new paper tex file

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Stage-Splitting a Modal Language}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

Introduction

\section{A Two-Stage Modal Language}

Introduce the next and prev concepts, along with typesystem.  Show some examples.  Introduce a hold operation.

\section{Straw Semantics}

Introduce the straw semantics, previously known as the ``erasure" semantics.

\section{Data and Control Dependency}

State a data-dependency theorem (that stage 1 values cannot depend on stage 2 values).  It should be true for the straw semantics!  Also state a control dependency theorem (that the execution of stage 1 code can't depend on stage 2 values).  This won't be true for the straw semantics, and show counterexample.

\section{Diagonal Semantics}

Present another semantics which obeys non-backwards control dependency.  This might be called the diagonal semantics.

\section{Splitting Algorithm}

Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.

\section{Examples for Staged Pipelines}

Give the gist of one-to-one pipeline example (like client/server).
Then talk about a one-to-many pipeline.
Then talk about a many-to-one pipeline like spark.  It clear how to target something like this for known base types on the boundary, and for product types.  But sums on the boundary are hard!  We leave many-to-one as future-work.

\section{Examples of Algorithm Derivation}

Fast exponent example.  Quickselect example.

Things to try: an interpreter which, partially evaluated, does cps or something.

\section{Related Work}

Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.  They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.  They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's nondeterminism.  The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.  Accordingly, their approach was entirely informal, with no implementation realized.  Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."

Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.  The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.  Their shaders are written in a C-like language with basic arithmetic and if statements.  Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.

Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.  Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.  Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.  Spark does not clearly identify this conflict.

[Discuss Yong's recent paper here.  It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.  The difference between this work and everything mention prior, is that first stage results aren't exactly  .  Syntactic restrictions prevent sum types on the boundary, sort of.  If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]

In addition to introducing pass separation, \cite{jorring86} also distinguishes it from two other staging techniques: partial evaluation and meta-programming.

[In brief, the three approaches can be distinguished based on when the first stage evaluation occurs, relative to when the compiler takes action.  In our work, the splitting occurs entirely before the first stage values are known.  In partial evaluation, these things happen essentially simultaneously.  For this reason, partial evaluation can do things like branch elimination for first-stage cases, which we cannot do.]

[Our type system is adapted directly from \cite{davies96}.  It was originally developed to describe binding time analysis for use in partial evaluation.]

[Metaprogramming is more different still.  It involves the creation of values that correspond to second stage code, often also with a primitive to allow the evaluation of second stage code to produce a first stage result.  For a type system that does that, you want \cite{davies01}.  Notice the closed-code restriction.  Meta-ML eases off on this restriction but does not (I think?) eliminate it.]

[What's going on with names and necessity?]

[Our work bears a lot of similarity to ML5, which also uses a modal type system.  The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others). The type systems reflect this directly in the world accessibility relation.  There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-05-13 paper/sigplanconf-template.tex
added comparison to PE and more related work

@@
 \usepackage{amsmath}
+\usepackage{stmaryrd}
 
@@
 
-Introduction
+Staged computation has existed as a programming technique for over three decades.
+Jorring et al.
+(\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.
+The first two of these have received significantly more attention than the third.
+
+We call these techniques {\em staged} in that part of their input comes in the first stage, and part comes in the second stage.  Likewise, some computation occurs at each stage as well.
+
+Countless meta-programming systems exist (...twelve thousand citations...), and their background theory and type-systems are well understood (\cite{davies01}).
+Partial evaluation, too, is well-understood.
+Partial evaluation systems exist...
+.
+This paper explores both theory and applications for stage-splitting.
+
+\section{Stage-Splitting Definition and Comparison to Partial Evaluation}
+
+First, we review the definition of partial evaluation.
+Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.
+This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.
+The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.
+We now state this theorem more formally: a partial evaluator is some function $p$ such that,
+\[
+	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
+\]
+where (borrowing notation from \cite{jones94}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
+
+Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.
+Again, more formally, a stage-splitter is some $s$ such that,
+\[
+	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
+	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
+\]
+
+We first discuss a few similarities between partial-evaluation and stage-splitting.
+First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.
+
+Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.
+Indeed, both admit completely trivial definitions.
+Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available.
+The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.
+Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.
+
+The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.
+For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.
+Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.
+Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.
+Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.
+
+We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
+However, one {\em cannot} construct a valid stage-splitter out of a partial evaluator, since a partial evaluator by definition depends on the first-stage input, whereas a stage-splitter does not have that available.
+
+Note that this does not mean that stage-splitting is a strict generalization of partial-evaluation.
+In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
+It is best to think of stage-splitting
 
@@
 
-Introduce the next and prev concepts, along with typesystem.  Show some examples.  Introduce a hold operation.
+Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
 
@@
 
-Present another semantics which obeys non-backwards control dependency.  This might be called the diagonal semantics.
+Present another semantics which obeys non-backwards control dependency.
+This might be called the diagonal semantics.
+Go through an example such as currying to build intuition for why this is really necessary (i.e., if we don't have these weird rules, currying breaks down!).
 
@@
 
+For each of these examples, talk about what partial evaluation would do and why that might be bad.
+
 \section{Related Work}
@@
 They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
-They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's nondeterminism.
+They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
 The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
@@
 Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
-Spark does not clearly identify this conflict.
+Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
+
+[RTSL and SH]
 
@@
 It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
-The difference between this work and everything mention prior, is that first stage results aren't exactly  .
-Syntactic restrictions prevent sum types on the boundary, sort of.
+They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
 If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
 
-In addition to introducing pass separation, \cite{jorring86} also distinguishes it from two other staging techniques: partial evaluation and meta-programming.
+Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
+That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
+Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
+The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
+They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
 
-[In brief, the three approaches can be distinguished based on when the first stage evaluation occurs, relative to when the compiler takes action.
-In our work, the splitting occurs entirely before the first stage values are known.
-In partial evaluation, these things happen essentially simultaneously.
-For this reason, partial evaluation can do things like branch elimination for first-stage cases, which we cannot do.]
-
-[Our type system is adapted directly from \cite{davies96}.  It was originally developed to describe binding time analysis for use in partial evaluation.]
-
-[Metaprogramming is more different still.
-It involves the creation of values that correspond to second stage code, often also with a primitive to allow the evaluation of second stage code to produce a first stage result.
-For a type system that does that, you want \cite{davies01}.
-Notice the closed-code restriction.
-Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
+[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
 

### 2014-05-13 paper/sigplanconf-template.tex
Got the bibliography mostly working.

@@
 
-Countless meta-programming systems exist (...twelve thousand citations...), and their background theory and type-systems are well understood (\cite{davies01}).
+Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).
 Partial evaluation, too, is well-understood.
@@
 \]
-where (borrowing notation from \cite{jones94}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
+where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
 
@@
 
-Note that this does not mean that stage-splitting is a strict generalization of partial-evaluation.
+Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
 In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
-It is best to think of stage-splitting
+It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer.
+[Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
 
@@
 
+\bibliography{paper}
 \bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
 

### 2014-05-24 paper/sigplanconf-template.tex
Added figures for grammar, values, types, statics, and dynamics.
Added a section about PE->SS.  Also a splitting figure.

@@
 \usepackage{amsmath}
+%\usepackage{amssymb}
+\usepackage{bbm}
 \usepackage{stmaryrd}
+\usepackage{proof}
+\usepackage{mathpartir}
+\usepackage{mathabx}
 
 \begin{document}
+
+\input {macros}
+\input {grammars}
+\input {semantics}
+\input {splitting}
 
@@
 
+\subsection{Partial Evaluator from Stage-Splitter}
+
 We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
-However, one {\em cannot} construct a valid stage-splitter out of a partial evaluator, since a partial evaluator by definition depends on the first-stage input, whereas a stage-splitter does not have that available.
-
 Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
@@
 [Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
+
+\subsection{Stage-Splitter from Partial Evaluator}
+
+Likewise, we can easily recover a stage-splitter from a partial evaluator.  If $p$ is a valid partial evaluator, then we can define a stage splitter $s$ such that $s(f)=(f_1,f_2)$, where
+\begin{align*}
+[f_1](x) &= p (f,x) \\
+[f_2](l,y) &= [l] (y)
+\end{align*}
+This implicitly requires that the languages in which $f_1$ and $f_2$ are expressed are strong enough to write a partial evaluator, but that is the case in this paper.
+A stage-splitter defined this way leaves much to be desired.
+Firstly, partial evaluation of $f$ may be too expensive for the context in which $f_1$ needs to run.
+Additionally, the intermediate data structure created this way may be much larger than necessary, as it would contain all of the residual code.
 
@@
 
-Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.
+[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
+
+The theorem that we want to be true is...
+
+\begin{center}
+\begin{tabular}{l}
+If $\splittwo{e}{A}{p,l.r}$ \\
+and $\reducetwo{e}{v}$ \\
+then $\reduce{p}{u}$ \\
+and $\reduce{[u/l]r}{w}$ \\
+and $v = w$ 
+\end{tabular}
+\end{center}
+
+%\begin{center}
+%\begin{tabular}{l}
+%If $\splitone {e}{A}{c,l.r}$ \\
+%and $\colone {\gamma_1}{\Gamma}$ \\
+%and $\coltwo {\gamma_2}{\Gamma}$\\
+%and $\rsone{\gamma_1(e)}{v}{q}$ \\
+%and $\reducetwo{\gamma_2(q)}{w}$ \\
+%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
+%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
+%\end{tabular}
+%~~~
+%\begin{tabular}{l}
+%If $\splittwo{e}{B}{p,l.r}$ \\
+%and $\colone {\gamma_1}{\Gamma}$ \\
+%and $\coltwo {\gamma_2}{\Gamma}$\\
+%and $\rstwo{\gamma_1(e)}{q}$ \\
+%and $\reducetwo{\gamma_2(q)}{w}$ \\
+%then $\reduce{|\gamma_1|(p)}{u}$ \\
+%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
+%\end{tabular}
+%\end{center}
 

### 2014-05-27 paper/sigplanconf-template.tex
first draft of reference semantics in
Added raw code for two examples.

@@
 \usepackage{mathabx}
+\usepackage{listings}
 
@@
 
-\begin{center}
-\begin{tabular}{l}
-If $\splittwo{e}{A}{p,l.r}$ \\
-and $\reducetwo{e}{v}$ \\
-then $\reduce{p}{u}$ \\
-and $\reduce{[u/l]r}{w}$ \\
-and $v = w$ 
-\end{tabular}
-\end{center}
+%\begin{center}
+%\begin{tabular}{l}
+%If $\splittwo{e}{A}{p,l.r}$ \\
+%and $\reducetwo{e}{v}$ \\
+%then $\reduce{p}{u}$ \\
+%and $\reduce{[u/l]r}{w}$ \\
+%and $v = w$ 
+%\end{tabular}
+%\end{center}
 
@@
 
-Fast exponent example.  Quickselect example.
+Fast exponent example.
+
+\lstset{language=ML, basicstyle=\ttfamily, tabsize=4, escapeinside={"*}{*"}}
+\begin{lstlisting} 
+let exp (b : $int, e : int) : $int =
+	if e == 0 then
+		next{1}
+	else if (e mod 2) == 0 then
+		next{let x = prev{exp(b,e/2)} in x*x}
+	else
+		next{prev{b} * prev{exp (b,e-1)}}		
+\end{lstlisting}
+
+splits into
+
+\begin{lstlisting} 
+let exp (b, e) =
+	((), roll (
+		if e == 0 then
+			inL ()
+		else 
+			inR (
+				if (e mod 2) == 0 then
+					inL (#2 (exp (b,e/2)))
+				else
+					inR (#2 (exp (b,e-1)))
+			)
+	))
+\end{lstlisting}
+
+and
+
+\begin{lstlisting} 
+let exp ((b, e), p) =
+	case unroll p of
+	  () => 1
+	| d  =>
+		case d of
+		  r => let x = exp ((b,()),r) in x*x
+		| r => b * exp ((b,()),r)
+\end{lstlisting}
+
+Quickselect example.
+
+\begin{lstlisting} 
+let qs (l : "*$\mathtt{\mu \alpha. }$*"() + "*$\mathtt{int*\alpha}$*", i: $int) = 
+	case unroll l of
+	  () => next {0}
+	| (h,t) => 
+		let (left,right,n) = partition h t in
+		next{
+			let n = prev{hold n} in
+			case compare prev{i} n of
+			  () (*LT*) => prev {qs left i}
+			| () (*EQ*) => hold h
+			| () (*GT*) => 
+				prev {qs right next{prev{i}-n-1}}
+		}	
+\end{lstlisting}
 

### 2014-06-02 paper/staging.tex
rename the pdfs

--- 2014-03-12 paper/paper.tex
@@
 \begin{tabular}{l}
-If $\splitone [\emptyC]{e}{A}{c,l.r}$ \\
-and $\rsone{e}{v}{q}$ \\
-and $\reducetwo{q}{w}$ \\
-then $\reduce{c}{\valprod{v}{u}}$ \\
-and $\reduce{[u/l]r}{w}$
+If $\splitone {e}{A}{c,l.r}$ \\
+and $\colone {\gamma_1}{\Gamma}$ \\
+and $\coltwo {\gamma_2}{\Gamma}$\\
+and $\rsone{\gamma_1(e)}{v}{q}$ \\
+and $\reducetwo{\gamma_2(q)}{w}$ \\
+then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
+and $\reduce{[u/l]|\gamma_2|(r)}{w}$
 \end{tabular}
@@
 \begin{tabular}{l}
-If $\splittwo[\emptyC]{e}{B}{p,l.r}$ \\
-and $\rstwo{e}{q}$ \\
-and $\reducetwo{q}{w}$ \\
-then $\reduce{p}{u}$ \\
-and $\reduce{[u/l]r}{w}$ 
+If $\splittwo{e}{B}{p,l.r}$ \\
+and $\colone {\gamma_1}{\Gamma}$ \\
+and $\coltwo {\gamma_2}{\Gamma}$\\
+and $\rstwo{\gamma_1(e)}{q}$ \\
+and $\reducetwo{\gamma_2(q)}{w}$ \\
+then $\reduce{|\gamma_1|(p)}{u}$ \\
+and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
 \end{tabular}
@@
 
+% LET
+\item {\bf Let (REALLY REALLY NOT DONE).} Suppose that
+\begin{enumerate}
+\item $\splitone{\next~e}{\fut A}{((),p),l.r}$
+\item $\rsone{\next~e}{v}{q}$
+\item $\reducetwo{q}{w}$
+\end{enumerate}
+By inversion of the splitting and reduction rules, we have
+\begin{enumerate}
+\item $\splittwo {e}{A}{p,l.r}$
+\item $v$ has the form $()$
+\item $\rstwo{e}{q}$
+\end{enumerate}
+We can now invoke the stage $\bbtwo$ part of the inductive hypothesis to yield,
+\begin{enumerate}
+\item $\reduce {p}{u}$
+\item $\reduce {[u/l]r}{w}$
+\end{enumerate}
+It can then be directly proven that,
+\begin{enumerate}
+\item $\reduce {((),p)}{((),u)}$
+\end{enumerate}
+Which (along with the previous line) is what we wanted to show.
+
 \end{itemize}

### 2014-06-05 paper/paper.tex
added a few words about the semantics

@@
-\documentclass{article}
+\documentclass{sigplanconf}
+
+% The following \documentclass options may be useful:
+
+% preprint      Remove this option only once the paper is in final form.
+% 10pt          To set in 10-point type instead of 9-point.
+% 11pt          To set in 11-point type instead of 9-point.
+% authoryear    To obtain author/year citation style instead of numeric.
+
 \usepackage{amsmath}
+%\usepackage{amssymb}
 \usepackage{bbm}
@@
 \usepackage{proof}
-\usepackage{amssymb}
+\usepackage{mathpartir}
 \usepackage{mathabx}
-\usepackage{float}
-\usepackage{cite}
-\usepackage{mathpartir}
-\usepackage[left=3cm,top=3cm,right=4cm,nohead,bottom=3cm]{geometry}
+\usepackage{listings}
+
+\begin{document}
 
 \input {macros}
-
-\title{\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}}
-\author{Nicolas Feltman et al.}
-\begin{document}
-\maketitle
-
 \input {grammars}
@@
 
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
+\copyrightyear{20yy} 
+\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
+\doi{nnnnnnn.nnnnnnn}
+
+% Uncomment one of the following two, if you are not going for the 
+% traditional copyright transfer agreement.
+
+%\exclusivelicense                % ACM gets exclusive license to publish, 
+                                  % you retain copyright
+
+%\permissiontopublish             % ACM gets nonexclusive license to publish
+                                  % (paid open-access papers, 
+                                  % short abstracts)
+
+\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{short description of paper}   % 'preprint' option specified.
+
+\title{Stage-Splitting a Modal Language}
+
+\authorinfo{Name1}
+           {Affiliation1}
+           {Email1}
+\authorinfo{Name2\and Name3}
+           {Affiliation2/3}
+           {Email2/3}
+
+\maketitle
+
+\begin{abstract}
+This is the text of the abstract.
+\end{abstract}
+
+\category{CR-number}{subcategory}{third-level}
+
+% general terms are not compulsory anymore, 
+% you may leave them out
+\terms
+term1, term2
+
+\keywords
+keyword1, keyword2
+
 \section{Introduction}
 
-Many computational tasks are premised on the notion of multiple domains, places and times at which computation occurs.
-Yet programming languages traditionally view their execution environments as monolithic, specifying the What and the How, but not the When nor the Where.
-As always, where the language falls short, a programmer instead must bear the burden.
-Inevitably, this comes at the cost of expressiveness, safety, or both.
-In this work, we present a language for specifying certain multi-domain computations, sacrificing neither the expressiveness nor the safety that programmers have come to expect in single-domain languages.
+Staged computation has existed as a programming technique for over three decades.
+Jorring et al.
+(\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.
+The first two of these have received significantly more attention than the third.
 
-Consider the programming of a pipeline system.
-Generally in such a system, data comes in to one processor, is transformed to an intermediate result, and then fed to later processors, possibly with some buffering, duplication, and reordering in between.
-We can view an algorithm for such as system as one computation, split across two spatial domains (processors).
-For a concrete example, we draw inspiration from graphics systems.
-Lets say that we want to map a function $f : \mathrm{PixelCoord} \to \mathrm{Color}$ over every pixel on a 1024x768 screen.
-Due to geometric coherence inherent in the problem, it's often advantageous to compute some information at the level of coarse 32x32 tiles (e.g.
-a list of the scene primitives which interact with that part of the screen) whose results can be used to accelerate the final per-pixel calculation.
-In addition to yielding algorithmic benefits, this decomposition strategy maps well to parallel hardware.
+We call these techniques {\em staged} in that part of their input comes in the first stage, and part comes in the second stage.  Likewise, some computation occurs at each stage as well.
 
-Often when programming high-performance systems, a programmer will wish to specialize a bivariate function $f : \alpha \times \beta \to \gamma$ to one of its inputs, $x : \alpha$, while leaving the other, $y : \beta$, for later.
-This need may arise because $x$ varies at a higher frequency than $y$, and we would wish to avoid repeated calculation.
-It may also be the case that $x$ is available at a time when computational resources are less expensive than when $y$ becomes available.
-Here we say that the calculation is split accross two temporal domains.
-Pulling another example from graphics, we consider a raytracer wherein $x : \mathtt{Geom}$ is a specification of the scene geometry (perhaps a list of triangles), $y : \mathtt{Ray}$ is a ray cast from some point in the scene, and $f$ computes the first intersection of the ray and the geometry.
-In modern raytracers, we may cast millions of rays without varying the scene geometry, so it could be advantageous to precompute as much of the algorithm as possible based on only scene geometry.
+Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).
+Partial evaluation, too, is well-understood.
+Partial evaluation systems exist...
+.
+This paper explores both theory and applications for stage-splitting.
 
-In addition to the partitioning of a calculation across multiple domains, the examples above exhibit one further similarity: communication between domains is one-directional.
-In the spatial case, information always flows from one processor to the next, but never backwards.
-In the temporal case too, information made available at the second time cannot be observed earlier (indeed this seems to be a defining feature of time, generally).
-For this reason, we refer to the domains as stages.
-This property, that information flow is acyclic, informs the rest of this paper.
+\section{Stage-Splitting Definition and Comparison to Partial Evaluation}
 
-[Talk about how languages without inherent support for domains must instead tackle the problem by writing various passes/shaders.
-This is bad because it's not compositional.
-For instance, I cannot write a function that talks about multiple stages at once.]
+First, we review the definition of partial evaluation.
+Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.
+This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.
+The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.
+We now state this theorem more formally: a partial evaluator is some function $p$ such that,
+\[
+	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
+\]
+where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
 
-[Because of the acyclic nature of the problem, we can take algorithms specified in our multi-domain language and tease them apart, essentially recovering the old way of doing things.
-This is called stage untangling.
-It gets us all the safety and expression without messing up the rest of the compiler.]
+Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.
+Again, more formally, a stage-splitter is some $s$ such that,
+\[
+	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
+	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
+\]
 
-[Talk about standard partial evaluation and pass separation, then place our work in the second, modulo binding time analysis.]
+We first discuss a few similarities between partial-evaluation and stage-splitting.
+First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.
 
-%\subsection {Program Generation} 
-%A common solution is to write code that writes code.
-Essentially, second-stage computations are represented by second-stage code, and second-stage code is a value that can be passed around and manipulated by first-stage code.
-To use the terminology from above: this first stage code is a function that takes in $x$ and outputs second stage function, which itself takes in $y$ and outputs the final answer.
-Note that this is essentially a manual process, in that the programmer must know the function $f$ which is being specialized.
-Using the equational notation of \cite{jones96}, where ``$\llbracket h \rrbracket~z$'' means the ``the code $h$ interpreted as a function and applied to input $z$," we say that a programmer defines a program generator $g$ that satisfies
-%\[
-%\llbracket \llbracket g \rrbracket~x\rrbracket~y =\llbracket f \rrbracket~(x,y)
-%\] 
-%for all $x$ and $y$.
-When well supported by the language (\cite{devito13}, [some metaml citation]) this technique is often called {\em metaprogramming}.
-Of course, even without first class support, one can represent the second stage code using strings or other data types.
-Defined so broadly, program generation includes all compilers, where $f$ is a hypothetical interpretter, $g$ is the actual compiler, $x$ is a program in the source language, and $y$ is the input to $x$.
-Program generation has also been used to implement the DFT example in the form of FFTW \cite{FFTW05}.
-%
-%The benefit of the program generation technique is full control, even up to the ability to use domain-specific optimizations.  This comes at the cost of essentially being a manual operation.  
-%
-%\subsection {Partial Evaluation} 
-%The idea of partial evaluation is to {\em automatically} specialize the code of $f$ to the supplied input, in a manner similar to standard evaluation.
-In particular this means defining a program $pe$, called a partial evaluator, to which we can pass the implementation of a function $f$ along with $f$'s immediate input $x$, and get back a version of $f$ that is {\em specialized} to $x$.
-This specialized version, also called a {\em residual}, is often denoted as $f_x$.
-Equationally, we have
-%\begin{align*}
-%\llbracket pe\rrbracket~(f,x) &= f_x \\
-%\llbracket f_x\rrbracket~y &= \llbracket f\rrbracket~(x,y)
-%\end{align*}
-% for all $f$, $x$, and $y$.
-As seminally observed by \cite{futamura71}, partially evaluating an interpretter on a source program is equivalent to compiling that program to whatever language the interpretter was written in!
-This has the benefit of being an automatic process, although the effective compiler is only as good as the partial evaluator.
-Much of the subsequent work in partial evaluators has been with the purpose of chasing this goal.
-%
-%Partial evaluation is a more general technique than program generation, in that the partial evaluator is defined once and works for all $f$s that need to be specialized.
-Of course, the partial evaluator has no domain knowledge of the program it is splitting, so it cannot be nearly as aggressive as a program generator might be.
-The two techniques are similar in that they both require the first input $x$ to do their specializing action.
-%
-%\subsection {Pass Separation} 
-%The final technique, pass separation, is conceptually the simplest.
-The idea is to define a program called $ps$, for {\em pass separator}.
-It works by cleaving the function $f:\alpha \times \beta \to \gamma$ into two functions $f_1 : \alpha \to \tau$ and $f_2 : \beta \times \tau \to \gamma$ for some type $\tau$, where $f_1$ builds a data structure (of type $\tau$) from $x$, and $f_2$ consumes that data structure as well as $y$ to produce the standard ouput.
-In analogy to partial evaluation, we call $f_2$ the {\em residual}.
-Equationally, this is
-%\begin{align*}
-%\llbracket pe\rrbracket~f &= (f_1,f_2) \\
-%\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) &= \llbracket f\rrbracket~(x,y)
-%\end{align*}
-%for all $f$, $x$, and $y$.
-This technique, performed manually, is a common exercise for every programmer.
-In can be considered the general form of common compiler optimizations such as loop hoisting, an example considered later.
-Like partial evaluation, pass separation is an automatic operation, but unlike both partial evaluation and program generation, pass separation does not require the value of the first argument, $x$, to properly specialize $f$.
-This makes pass separation the most widely applicable of the staging techniques, at the cost of being able to perform aggressive optimizations in the residual that might depend on the first argument.
-%
-%In this work we present a pass separation algorithm for a fragment of ML.
-Specifically, our formulation can separate terms containing sums and recursion, which have not appeared previously in the stage seperation literature.
-We anticipate that the algorithm will also be able to also separate this language extended with with first-class functions, although this is left as future work.
-Additionally, our presentation is more explicitly type-motivated than those that have come before.
-%
-%\section{Binding-Time Analysis and Stage Untangling}
-%
-%The pass separation problem can naturally be decomposed into two subproblems:
-%\begin{itemize}
-%\item Decide which parts of the definition of $f$ belong to which stage and produce an annotated version of $f$.
-%\item Use the annotated version of $f$ to produce $f_1$ and $f_2$.  
-%\end{itemize}
-%
-%An analogous form of this decomposition exists in the partial evaluation literature, wherein the first subproblem is known as {\em binding-time analysis}.
-The second subproblem, in which the difference between pass separation and partial evaluation is more manifest, does not seem to yet have a name.
-We call it {\em stage untangling}.
-Unlike previous work in pass separation, we focuses entirely on the stage untangling problem by assuming that $f$ is given to us in an annotated form.
-%
-%Previous research has observed that the pass separation problem is inherently ambiguous in that the equations defined above do not fully specify the definition of $pe$.
-That is, there can exist multiple ways to partition $f$ into $f_1$ and $f_2$ that satisfy $\llbracket f_2\rrbracket~(\llbracket f_1\rrbracket~x,y) = \llbracket f\rrbracket~(x,y)$.
-For a trivial example, we can set $f_1$ to be the identity and $f_2$ to be $f$.
-Fortunately, this ambiguity is contained entirely within the binding-time analysis portion of pass separation; stage untangling is entirely determined.
+Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.
+Indeed, both admit completely trivial definitions.
+Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available.
+The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.
+Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.
 
-\section{A Simple Two-Stage Language}
+The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.
+For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.
+Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.
+Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.
+Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.
 
-We'll start by analyzing a simple two-stage language called \ellStaged.
-We say it is simple in that it supports boolean, integer, and product types, as well as let bindings, if expressions, and various primitive operations.
-Notably, we leave out functions, fixed point operators, and full sum types, which will be added later.
+\subsection{Partial Evaluator from Stage-Splitter}
 
-The grammar for types, terms, values, and contexts in \ellStaged are given in Figure \ref{fig:coreSyntax}.
-The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.
-It would be tempting to merge the two stages into one syntactic class and maintain the stage invariants with the typing judgements.
-Such a trick would reduce the total number of rules presented later, but at the cost of making those rules type derivation-directed, rather than syntax directed.
-Alternatively, we chose to have one context merged accross both stages, rather than one context for each stage.
-The intended effect of this decision is for bindings in one stage to shadow binding to the same variables in the other stage.
-The reasons for this will become clear later.
+We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
+Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
+In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
+It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer.
+[Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
 
-At the type and value levels, stage $\bbone$ can refer to stage $\bbtwo$, but stage $\bbtwo$ cannot refer back to stage $\bbone$.
-Alternatively, the two stages are mutually recursive at the term level.
-These two patterns together fulfill the promise of {\em mixed code, staged evalutation}.
+\subsection{Stage-Splitter from Partial Evaluator}
 
-In the rest of this document, I will tend to use $\{A,B,C\}$ for types, $\{v,u,w,t\}$ for values, $\{x,y\}$ for variables, $e$ for terms, $i$ for integers, and $b$ for booleans.
+Likewise, we can easily recover a stage-splitter from a partial evaluator.  If $p$ is a valid partial evaluator, then we can define a stage splitter $s$ such that $s(f)=(f_1,f_2)$, where
+\begin{align*}
+[f_1](x) &= p (f,x) \\
+[f_2](l,y) &= [l] (y)
+\end{align*}
+This implicitly requires that the languages in which $f_1$ and $f_2$ are expressed are strong enough to write a partial evaluator, but that is the case in this paper.
+A stage-splitter defined this way leaves much to be desired.
+Firstly, partial evaluation of $f$ may be too expensive for the context in which $f_1$ needs to run.
+Additionally, the intermediate data structure created this way may be much larger than necessary, as it would contain all of the residual code.
 
-\subsection{Static Semantics}
+\section{A Two-Stage Modal Language}
 
-The static semantics for \ellStaged~is given in Figure \ref{fig:coreStatics}.
-Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).
-The static semantics comprises two judgements.
-Those are:
+Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
+
+\section{Diagonal Semantics}
+
+We desire a semantics that meets the following goals,
+
+\begin{enumerate}
+\item In absence of $\next$ expressions, it should be equivalent to standard call-by-value evaluation.
+\item $\next$ and $\prev$ should be inverses of the other.
+\item All of the first stage work should be completed before second stage work.
+\item The result for a stage-2 expression should be the same as if all of the $\next$s and $\prev$s were interpreted as the identity function.
+\end{enumerate}
+
+In this section, we introduce a dynamic semantics for our language.
+Although our type system is virtually identical to that in \cite{davies96}, the semantics is different in terms of cost and termination behavior.
+The semantics comprises the following judgments:
 
 \begin{center}
-\begin{tabular}{|l|l|} \hline
-Name & Pattern \\ \hline
-Typing at $\bbone$ & $\typesone[\contextot]{\expro}{\typeo}$ \\  \hline
-Typing at $\bbtwo$ & $\typestwo[\contextot]{\exprt}{\typet}$ \\ \hline
+\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
+Judgment & Reads & Conditions \\ \hline 
+$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
+$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\gamma$ and value $v$'' 
+& $\typesone e A$ 
+	\newline $\isvalone v$ 
+	\newline ... \\ \hline 
+$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
 \end{tabular}
@@
 
-We see that the typing judgements for both stages are each what one would expect for an unstaged language.
-That is, there are introduction and elimination forms for base types and products at both stage $\bbone$ and stage $\bbtwo$.
-Furthermore, these rules preserve stage.
-For these reasons, we can think of the base type and product features as essentially being orthogonal to the staging features.
-The terms $\next$ and $\prev$, which are the only terms that allow us to change stage, are respectively the introduction and elimination forms for $\fut$.
+\subsection{Adherence to Goals}
 
-\subsection{Erasure Dynamic Semantics}
+We meet our first and second goals (show examples).
 
-The dynamics semantics for \ellStaged~is given in Figure \ref{fig:erasureSemantics} in big-step style.  Predictably, it comprises two judgements:
+That we meet our third goal is clear from just the structure of our judgments (...).
 
-\begin{center}
-\begin{tabular}{|l|l|} \hline
-Name & Pattern \\ \hline
-Reduction at $\bbone$ & $\reduceone{\expro}{\valo}$ \\  \hline
-Reduction at $\bbtwo$ & $\reducetwo{\exprt}{\valt}$ \\ \hline
-\end{tabular}
-\end{center}
+Under the strictest interpretation, we do not meet our third goal.  Consider the following examples (show speculating down an if).
 
-The primary goal for this semantics was simplicity.
-This was achieved in that, like the static semantics, the judgements for each level are extremely similar to those for an unstaged language.
-This simplicity came at the cost of some descriptiveness, however.
-The inadequacy of this semantics will become apparent shortly.
+But as it turns out, a semantics which had the erasure behavior would not meet goal 3! (explain using the same example)
 
-First, we discuss a few desirable theorems.
-Chiefly among them is type preservation (I haven't defined the judgements for the types of values, but it should be rather clear).
-As will be the pattern for many theorems to come, our inductive hypothesis comprises two mutually dependent lemmas, one for stage $\bbone$, and the other for stage $\bbtwo$:
-\begin{center}
-\begin{tabular}{l}
-If $\typesone [\emptyC]{e}{A}$ \\
-and $\reduceone{e}{v}$ \\
-then $v : A$
-\end{tabular}
-~~~
-\begin{tabular}{l}
-If $\typestwo [\emptyC]{e}{A}$ \\
-and $\reducetwo{e}{v}$ \\
-then $v : A$
-\end{tabular}
-\end{center}
-[Proof not done, but it looks trivial in every case. [Wait no, substitutions kill me again.]]
+\section{Splitting Algorithm}
 
-We also want some formalization of the idea that stage two computation cannot affect stage one computation.  At a minimum, the following lemma should be true [this needs lots of work],
+[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
 
-\begin{center}
-\begin{tabular}{l}
-If $\exists v\in \valt$ s.t. $\reduceone {[v/x]e}{u}$ \\
-then $\forall v\in \valt$ s.t. $\reduceone {[v/x]e}{u'}$, $u \simeq u'$
-\end{tabular}
-\end{center}
+The theorem that we want to be true is...
 
-As it turns out, this is not true! [Example forthcoming.]
+%\begin{center}
+%\begin{tabular}{l}
+%If $\splittwo{e}{A}{p,l.r}$ \\
+%and $\reducetwo{e}{v}$ \\
+%then $\reduce{p}{u}$ \\
+%and $\reduce{[u/l]r}{w}$ \\
+%and $v = w$ 
+%\end{tabular}
+%\end{center}
 
-\section{The Stage-Splitting Transformation}
+%\begin{center}
+%\begin{tabular}{l}
+%If $\splitone {e}{A}{c,l.r}$ \\
+%and $\colone {\gamma_1}{\Gamma}$ \\
+%and $\coltwo {\gamma_2}{\Gamma}$\\
+%and $\rsone{\gamma_1(e)}{v}{q}$ \\
+%and $\reducetwo{\gamma_2(q)}{w}$ \\
+%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
+%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
+%\end{tabular}
+%~~~
+%\begin{tabular}{l}
+%If $\splittwo{e}{B}{p,l.r}$ \\
+%and $\colone {\gamma_1}{\Gamma}$ \\
+%and $\coltwo {\gamma_2}{\Gamma}$\\
+%and $\rstwo{\gamma_1(e)}{q}$ \\
+%and $\reducetwo{\gamma_2(q)}{w}$ \\
+%then $\reduce{|\gamma_1|(p)}{u}$ \\
+%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
+%\end{tabular}
+%\end{center}
 
-We now cover splitting, the process by which mixed code is teased apart into two individual chunks of code, one for each stage.
-To begin, we first introduce the target lagnuage (for both stage), \ellTarget.
-The grammar for \ellTarget~is given in figure \ref{fig:baseSyntax}.
-We see that \ellTarget~is (as one would hope) a traditional single-stage language, with all the same basetypes as \ellStaged, in addition to products, if expressions, and full sums (injections and case expressions).
-The presence of this last feature is a bit curious, since \ellStaged~did not support full sums.
-The necessity for sums should become apparent in time.
-The static and dynamic semantics of \ellTarget~are not given, but should be obvious.
-They are given by the judgements:
+\section{Examples for Staged Pipelines}
 
-\begin{center}
-\begin{tabular}{|l|l|} \hline
-Name & Pattern \\ \hline
-Typing & $\types{\expr}{\val}$ \\  \hline
-Reduction & $\reduce{\expr}{\val}$ \\ \hline
-\end{tabular}
-\end{center}
+Give the gist of one-to-one pipeline example (like client/server).
+Then talk about a one-to-many pipeline.
+Then talk about a many-to-one pipeline like spark.
+It clear how to target something like this for known base types on the boundary, and for product types.
+But sums on the boundary are hard!
+We leave many-to-one as future-work.
 
-The splitting judgements come in seven parts, two each for types, terms, and values, and one for contexts.  In tabular form, they are:
-\begin{center}
-\begin{tabular}{|l|l|} \hline
-Name & Pattern \\ \hline
-Value Splitting at $\bbone$& $\valo \vsplito [\val,\val]$ \\  \hline
-Value Splitting at $\bbtwo$& $\valt \vsplits \val$ \\  \hline
-Type Splitting at $\bbone$& $\typeo \tsplito [\type,\type]$ \\  \hline
-Type Splitting at $\bbtwo$& $\typet \tsplits \type$ \\  \hline
-Context Splitting & $\contextot \csplit [\context,\context]$ \\  \hline
-Term Splitting at $\bbone$& $\splitone{\expro}{\typeo}{\expr,\var.\expr}$ \\  \hline
-Term Splitting at $\bbtwo$& $\splittwo{\exprt}{\typet}{\expr,\var.\expr}$ \\  \hline
-\end{tabular}
-\end{center}
+\section{Examples of Algorithm Derivation}
 
-[...more...]
+Fast exponent example.
 
-\subsection {Splitting Theorems}
+\lstset{language=ML, columns=fullflexible, basicstyle=\ttfamily, tabsize=4, escapeinside={"*}{*"}}
+\begin{lstlisting} 
+let exp (b : $int, e : int) : $int =
+	if e == 0 then
+		next{1}
+	else if (e mod 2) == 0 then
+		next{let x = prev{exp(b,e/2)} in x*x}
+	else
+		next{prev{b} * prev{exp (b,e-1)}}		
+\end{lstlisting}
 
-\subsubsection{It preserves types}
+splits into
 
-The theorem we want to show has two parts, one for the $\bbone$ code, and one for the $\bbtwo$ code.  
-\begin{center}
-\begin{tabular}{l}
-If $\splitone {e}{A}{c,l.r}$ \\
-then $\Gamma \csplit [\Gamma_1;\Gamma_2]$ \\
-and $A \tsplito [A_1,A_2]$ \\
-and $ \types [\Gamma_1]{c} {A_1 \times \tau}$ \\
-and $ \types [\Gamma_2, p:\tau]{r} {A_2}$ 
-\end{tabular}
-~~~
-\begin{tabular}{l}
-If $\splittwo {e}{A}{p,l.r}$ \\
-then $\Gamma \csplit [\Gamma_1;\Gamma_2]$ \\
-and $A \tsplits A'$ \\
-and $ \types [\Gamma_1]{p} {\tau}$ \\
-and $ \types [\Gamma_2, p:\tau]{r} {A'}$ 
-\end{tabular}
-\end{center}
+\begin{lstlisting} 
+let exp (b, e) =
+	((), roll (
+		if e == 0 then
+			inL ()
+		else 
+			inR (
+				if (e mod 2) == 0 then
+					inL (#2 (exp (b,e/2)))
+				else
+					inR (#2 (exp (b,e-1)))
+			)
+	))
+\end{lstlisting}
 
-\subsubsection{It preserves meaning}
+and
 
-The theorem we want to show has two parts, one for the $\bbone$ code, and one for the $\bbtwo$ code.  
-\begin{center}
-\begin{tabular}{l}
-If $\splitone [\emptyC]{e}{A}{c,l.r}$ \\
-and $\rsone{e}{v}{q}$ \\
-and $\reducetwo{q}{w}$ \\
-then $\reduce{c}{\valprod{v}{u}}$ \\
-and $\reduce{[u/l]r}{w}$
-\end{tabular}
-~~~
-\begin{tabular}{l}
-If $\splittwo[\emptyC]{e}{B}{p,l.r}$ \\
-and $\rstwo{e}{q}$ \\
-and $\reducetwo{q}{w}$ \\
-then $\reduce{p}{u}$ \\
-and $\reduce{[u/l]r}{w}$ 
-\end{tabular}
-\end{center}
+\begin{lstlisting} 
+let exp ((b, e), p) =
+	case unroll p of
+	  () => 1
+	| d  =>
+		case d of
+		  r => let x = exp ((b,()),r) in x*x
+		| r => b * exp ((b,()),r)
+\end{lstlisting}
 
-\section {Adding Full Sums and Functions}
+Quickselect example.
 
-In this section, we consider adding two more features to our source language: sums and functions.
-We will call this augmented language \lamStaged.
-The grammar for \lamStaged is given in \ref{fig:lamStagedSyntax}.
+\begin{lstlisting} 
+let qs (l : "*$\mathtt{\mu \alpha. }$*"() + "*$\mathtt{int*\alpha}$*", i: $int) = 
+	case unroll l of
+	  () => next {0}
+	| (h,t) => 
+		let (left,right,n) = partition h t in
+		next{
+			let n = prev{hold n} in
+			case compare prev{i} n of
+			  () (*LT*) => prev {qs left i}
+			| () (*EQ*) => hold h
+			| () (*GT*) => 
+				prev {qs right next{prev{i}-n-1}}
+		}	
+\end{lstlisting}
 
-\section{Related Work in Stage Separation}
+Things to try: an interpreter which, partially evaluated, does cps or something.
 
-Our starting language, $\lambda^{12}$, is essentially a two-stage version of $\lambda^\bigcirc$ from \cite{davies96}, but beefed up with products, sums, and fixed-points.
-That work motivates $\lambda^\bigcirc$ as an image of linear temporal logic under the Curry-Howard correspondence, and it makes the argument that $\lambda^\bigcirc$ is a good model for binding time analysis.
-Although $\lambda^\bigcirc$ was designed with partial evaluation in mind, it turns out to be a good model for pass separation as well.
-In contrast, \cite{davies01} presents $\lambda^\Box$, which is based off of branching temporal logic.
-As that work shows, $\lambda^\Box$ is a good system for program generation, although we find the closed-code requirement to be more restricting than we need.
+For each of these examples, talk about what partial evaluation would do and why that might be bad.
 
-Pass seperation was introduced in \cite{jorring}.
-They focused on motivating pass separation as a technique for compiler generation, and hinted at wider applicability.
-Their approach was not mechanized or automated in any sense, and their examples were separated by hand.
-We believe that out work represents the fulfillment of their prediction that ``the [pass separation] approach will elude full automation for some time."
+\section{Related Work}
 
-\cite{hannan94} also uses a pass separation technique for generating compiler/evaluator pairs from interpretters specified as term-rewrite systems.
-This heavy restriction on the form of the input prevents generalization of their style of pass separation.
-[I don't know enough about term-rewrite systems yet to be able to comment further.]  Their method, while amenable to mechanization, was not actually implemented.
+Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
+They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
+They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
+The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
+Accordingly, their approach was entirely informal, with no implementation realized.
+Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
 
-\cite{knoblock} represents the first attempt at fully automatic pass separation for code approaching general-purpose.
-In particular, they implemented pass separation for a C-like shading language including basic arithmetic and if statements.
-The main goal of their work, like ours, was to minimize recomputation.
-That said, their starting language was sufficiently restricted for recursive boundary types to be inexpressible.
-Given that memory is at a premium in shading languages, so much of that work was also focused on minimizing the memory-footprint of the boundary type.
+Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
+The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
+Their shaders are written in a C-like language with basic arithmetic and if statements.
+Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
 
-\section{Appendix}
+Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
+But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
+Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
+Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
+Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
 
-\subsection {Equivalence of Diagonal Evaluation and Splitting (for \ellStaged)}
+[RTSL and SH]
 
-We proceed by cases of the term splitting judgement.
+[Discuss Yong's recent paper here.
+It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
+They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
+If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
 
-\begin{itemize}
-% INT, BOOL, UNIT
-\item {\bf Int.} Suppose that
-\begin{enumerate}
-\item $\splitone{i}{\rm int}{(i,()),\_.()}$
-\item $\rsone{i}{v}{q}$
-\item $\reducetwo{q}{w}$
-\end{enumerate}
-By inversion of reduction relations, we know
-\begin{enumerate}
-\item $v$ has the form $i$
-\item $q$ has the form $()$
-\item $w$ has the form $()$
-\end{enumerate}
-It is directly provable that,
-\begin{enumerate}
-\item $\reduce{(i,())}{\valprod{i}{()}}$
-\item $\reduce{[()/\_]()}{()}$
-\end{enumerate}
-Which is what we need to show.
+Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
+That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
+Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
+The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
+They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
 
-\item {\bf Boolean.} Same as int.
-\item {\bf Unit.} Same as int.
+[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
 
-% TUPLE
-\item {\bf Tuple.} Suppose that
-\begin{enumerate}
-\item $\splitone {(e_1,e_2)}{A\times B}
-		{ \begin{array}{l}
-		(\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}), \\
-		l.(\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})
-		\end{array}} $
-\item $\rsone{(e_1,e_2)}{v}{q}$
-\item $\reducetwo{q}{w}$
-\end{enumerate}
-By inversion of the splitting and reduction rules, we know
-\begin{enumerate}
-\item $\splitone{e_1}{A}{c_1,l_1.r_1}$ and $\splitone{e_2}{B}{c_2,l_2.r_2}$
-\item $v$ has the form $\valprod{v_1}{v_2}$
-\item $q$ has the form $(q_1,q_2)$
-\item $\rsone{e_1}{v_1}{q_1}$ and $\rsone{e_2}{v_2}{q_2}$
-\item $w$ has the form $\valprod{w_1}{w_2}$
-\item $\reducetwo{q_1}{w_1}$ and $\reducetwo{q_2}{w_2}$
-\end{enumerate}
-This gives us enough to invoke the inductive hypothesis for the each side, yielding
-\begin{enumerate}
-\item $\reduce{c_1}{\valprod{v_1}{u_1}}$ 
-\item $\reduce{[u_1/l_1]r_1}{w_1}$ 
-\item $\reduce{c_2}{\valprod{v_2}{u_2}}$ 
-\item $\reduce{[u_2/l_2]r_2}{w_2}$ 
-\end{enumerate}
-It is now directly provable from the evaluation rules that
-\begin{enumerate}
-\item $\reduce{\letin{y_1}{c_1}{\letin{y_2}{c_2}{((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}}{\valprod {\valprod{v_1}{v_2}}{\valprod{u_1}{u_2}}}$
-\item $\reduce{[\valprod{u_1}{u_2}/l](\letin{l_1}{\pi_1 l}{r_1},\letin{l_2}{\pi_2 l}{r_2})}{\valprod{w_1}{w_2}}$
-\end{enumerate}
-Which is what we wanted to show.
+[What's going on with names and necessity?]
 
-% PROJECTIONS
-\item {\bf Left Projection.} Suppose that
-\begin{enumerate}
-\item $\splitone {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}$
-\item $\rsone{\pi_1~e}{v}{q}$
-\item $\reducetwo{q}{w}$
-\end{enumerate}
-By inversion of the splitting and reduction rules, we have
-\begin{enumerate}
-\item $\splitone {e}{A \times B}{c,l.r}$
-\item $q$ has the form $\pi_1~q'$
-\item $\rsone{e}{\valprod{v}{v_2}}{q'}$
-\item $\reducetwo{q'}{\valprod{w}{w_2}}$
-\end{enumerate}
-This is enough to invoke our inductive hypothesis, yielding
-\begin{enumerate}
-\item $\reduce{c}{\valprod{\valprod{v}{v_2}}{u}}$
-\item $\reduce{[u/l]r}{\valprod{w}{w_2}}$
-\end{enumerate}
-From this, it can be directly proven that
-\begin{enumerate}
-\item $\reduce{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)}}{\valprod{v}{u}}$
-\item $\reduce{[u/l](\pi_1~r)}{w}$
-\end{enumerate}
-Which is what we wanted to show.
+[Our work bears a lot of similarity to ML5, which also uses a modal type system.
+The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others).
+The type systems reflect this directly in the world accessibility relation.
+There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]
 
-\item {\bf Right Projection.} Same as left projection.
+\appendix
+\section{Appendix Title}
 
-% IF
-\item {\bf If.} Suppose that
-\begin{enumerate}
-\item $\splitone {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
-								{\begin{array}{l}
-									\left(
-										\talllet{z}{c_1}{
-										\tallif {\pi_1 z}
-										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
-										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}
-									}\right), \\
-									l.(\letin{l_1}{\pi_1 l}{r_1};\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
-								\end{array}}$
-\item $\rsone{\ifthen {e_1}{e_2}{e_3}}{v}{q}$
-\item $\reducetwo{q}{w}$
-\end{enumerate}
-By inversion of the splitting rule, we have
-\begin{enumerate}
-\item $\splitonesub 1 \rmbool$ and $\splitonesub 2 A$ and $\splitonesub 3 A$
-\end{enumerate}
-There are two rules for evaluation of an if.  We proceed by inverting the if-true evaluation rules first:
-\begin{enumerate}
-\item $q$ has the form $(q_1;q_2)$
-\item $\rsone{e_1}{\bf true}{q_1}$ and $\rsone{e_2}{v}{q_2}$
-\item $\reducetwo{q_1}{w_1}$ and $\reducetwo{q_2}{w}$
-\end{enumerate}
-We can now invoke the inductive hypothesis for the predicate to yield
-\begin{enumerate}
-\item $\reduce{c_1}{\valprod{\bf true}{u_1}}$
-\item $\reduce{[u_1/l_1]r_1}{w_1}$
-\end{enumerate}
-We can also invoke the inductive hypothesis for the branch to yield
-\begin{enumerate}
-\item $\reduce{c_2}{\valprod{v}{u_2}}$
-\item $\reduce{[u_2/l_2]r_2}{w}$
-\end{enumerate}
-From this, it can be directly proven that
-\begin{enumerate}
-\item $\reduce{
-		\left(
-			\talllet{z}{c_1}{
-			\tallif {\pi_1 z}
-			{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
-			{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}
-		}\right)}{\valprod{v}{\valprod{u_1}{\inl~u_2}}}$
-\item $\reduce{[\valprod{u_1}{\inl~u_2}/l](\letin{l_1}{\pi_1 l}{r_1};\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})}{w}$
-\end{enumerate}
-Everything works similarly for the if-false case.  Either way, we have what we want to show.
+This is the text of the appendix, if you need one.
 
-% NEXT
-\item {\bf Next.} Suppose that
-\begin{enumerate}
-\item $\splitone{\next~e}{\fut A}{((),p),l.r}$
-\item $\rsone{\next~e}{v}{q}$
-\item $\reducetwo{q}{w}$
-\end{enumerate}
-By inversion of the splitting and reduction rules, we have
-\begin{enumerate}
-\item $\splittwo {e}{A}{p,l.r}$
-\item $v$ has the form $()$
-\item $\rstwo{e}{q}$
-\end{enumerate}
-We can now invoke the stage $\bbtwo$ part of the inductive hypothesis to yield,
-\begin{enumerate}
-\item $\reduce {p}{u}$
-\item $\reduce {[u/l]r}{w}$
-\end{enumerate}
-It can then be directly proven that,
-\begin{enumerate}
-\item $\reduce {((),p)}{((),u)}$
-\end{enumerate}
-Which (along with the previous line) is what we wanted to show.
+\acks
 
-\end{itemize}
+Acknowledgments, if needed.
 
-\subsection {Equivalence of Diagonal Evaluation and Splitting (extra \lamStaged~cases)}
+% We recommend abbrvnat bibliography style.
 
 \bibliography{paper}
-\bibliographystyle{plain}
+\bibliographystyle{abbrvnat}
 

### 2014-06-05 paper/paper.tex
more words about semantics.

@@
 
-We desire a semantics that meets the following goals,
+We desire a semantics that meets the following goals:
 
@@
 $\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
+...&...&... \\ \hline
 \end{tabular}
@@
 
+Evaluation and speculation are the main judgments here, the rest being largely administrative.
+The evaluation judgment operates on stage-1 code whereas the speculation operates on stage-2 code.
+Since $\next$ and $\prev$ are the crossover points between 1-code and 2-code, they are correspondingly the only places where the evaluation and speculation judgments depend on the other.
+
+We can think of our evaluation as proceeding in the standard way for stage-1 code.
+But when the evaluator encounters stage-2 code, it places that code off to the side in a table, and then keeps a reference to the table entry.
+These references are then passed around as stage-1 values, and the tables are managed accordingly.
+
+The evaluation judgment is very similar to standard call-by-value evaluation, as goal 1 would suggest.
+The input to evaluation is a stage-1 expression (usually $e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered later.
+Evaluation has two outputs: the {\em future table} (usually $\xi$) and the {\em partial value} (usually $v$).
+We cover the latter first.
+The partial value is essentially the result of the first-stage portion of $e$, and must be a 1-value of the same type as $e$.
+As usual, this means that $v$ must be composed only of base primitives, tuples, injections, and lambdas (corresponding to base types, products, sums, and functions).
+Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.
+Those stage-2 computations are held in
+
 \subsection{Adherence to Goals}
 
-We meet our first and second goals (show examples).
+We meet our first (discuss) and second goals (show examples).
 

### 2014-06-05 paper/paper.tex
some more text on semantics

@@
 Although our type system is virtually identical to that in \cite{davies96}, the semantics is different in terms of cost and termination behavior.
-The semantics comprises the following judgments:
+
+Abstractly, we can think of our evaluation as proceeding in the standard way for stage-1 code.
+When the evaluator encounters a $\next \{e\}$ expression, it places $e$ off to the side in a table and replaces the whole expression with a reference to the table entry.
+These references are then passed around as stage-1 values for $\fut$ types.
+But what if $e$ contains $\prev$ expressions?
+To ensure that all stage-1 code is evaluated before any stage-2 code, we must evaluate all of the 1-code contained in $e$ before inserting it into the table.
+This entails searching $e$ for all contained $\prev$s and evaluating them in place.
+
+More formally, evaluation comprises the folowing judgments:
 
@@
 Evaluation and speculation are the main judgments here, the rest being largely administrative.
-The evaluation judgment operates on stage-1 code whereas the speculation operates on stage-2 code.
+The evaluation judgment operates on stage-1 code, whereas the speculation operates on stage-2 code.
 Since $\next$ and $\prev$ are the crossover points between 1-code and 2-code, they are correspondingly the only places where the evaluation and speculation judgments depend on the other.
-
-We can think of our evaluation as proceeding in the standard way for stage-1 code.
-But when the evaluator encounters stage-2 code, it places that code off to the side in a table, and then keeps a reference to the table entry.
-These references are then passed around as stage-1 values, and the tables are managed accordingly.
 
@@
 Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.
-Those stage-2 computations are held in
+
+...
 
@@
 But as it turns out, a semantics which had the erasure behavior would not meet goal 3! (explain using the same example)
+
+\subsection{Comparison to Other Semantics}
+
+The differences between our diagonal semantics, the erasure semantics, and Davies' semantics can be characterized entirely by how they behave at $\next$ expressions:
+\begin{itemize}
+\item {\bf Erasure Semantics.} Evaluate contents of $\next$ immediately.
+\item {\bf Diagonal Semantics.} Speculate to find the $\prev$s, evaluate them in place, then place the contents of $\next$ into a table and create a hatted variable to act as a value.
+\item {\bf Davies' Semantics.} Speculate to find the $\prev$s, evaluate them in place, then treat the reduced $\next$ as a value.
+\end{itemize}
 

### 2014-06-06 paper/paper.tex
starting metatheory notes
added more to semantics discussion

@@
 %\usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{bbm}
@@
 \usepackage{listings}
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{definition}[theorem]{Definition}
 
@@
 
-\section{Diagonal Semantics}
+\section{Semantics}
+
+There's more than one way to define a semantics for our language.  We start this section by considering two proposals, characterizing them mainly by how they handle values of $\fut$ type.
+
+\subsection{The Erasure Semantics}
+
+The first proposal we call the {\em erasure semantics}, because it corresponds to what one would get by simply {\em erasing} all of the $\next$ and $\prev$ terms that appear in \lamStaged.
+In particular, we say that $\next v$ is a first-stage value of type $\fut A$ if and only if $v$ is a second-stage value of type $A$, (...)
+
+...
+
+We say that erasure semantics gives us a notion of multi-stage evaluation.
+One judgement is multistage evaluation at 1, and the other is multistage evaluation at 2.
+The downside of this is that the erasure semantics doesn't seem to do all of the stage-1 work before the stage-2 work, for some notion of ``before''.
+Of course, the erasure semantics might be equivalent to some other semantics which does have that property.
+Unfortunately, this is false (show second-stage if example).
+
+...
+
+The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.
+It also {\em obviously} produces the correct answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
+
+\subsection{Meta Semantics}
+
+A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.
+
+In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.
+But in the meta semantics, $\next e$ is a value only if $e$ has no $\prev$ terms; $v$ is allowed to have unreduced stage-2 computation.
+
+...
+
+The Davies semantics is comprised of two mutually recursive judgements: $\daviesz$ and $\davieso$.
+For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.
+(...
+This essentially gives us partial evaluation, and we're left to just evaluate the residual normally to get multi-stage evaluation...)
+
+The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.
+This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins.
+
+The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.
+This makes reasoning about second-stage cost rather difficult.
+
+\subsection{Our Semantics}
 
@@
 \begin{enumerate}
-\item In absence of $\next$ expressions, it should be equivalent to standard call-by-value evaluation.
-\item $\next$ and $\prev$ should be inverses of the other.
-\item All of the first stage work should be completed before second stage work.
-\item The result for a stage-2 expression should be the same as if all of the $\next$s and $\prev$s were interpreted as the identity function.
+\item Modulo termination, it should be equivalent to the erasure semantics.
+\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
+\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
 \end{enumerate}
+
+We meet all of these goals.
 
@@
 $\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
-$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\gamma$ and value $v$'' 
+$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
 & $\typesone e A$ 
@@
 
-\subsection{Adherence to Goals}
+\subsection{Other nice properties}
 
-We meet our first (discuss) and second goals (show examples).
+\begin{enumerate}
+\item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
+\end{enumerate}
 
-That we meet our third goal is clear from just the structure of our judgments (...).
+\subsection{Metatheory}
 
-Under the strictest interpretation, we do not meet our third goal.  Consider the following examples (show speculating down an if).
+\TODO unfinished; where should this go? combine with table above
 
-But as it turns out, a semantics which had the erasure behavior would not meet goal 3! (explain using the same example)
+\noindent\TODO terminology for $\Xi,\Gamma,\xi,v,q$?
 
-\subsection{Comparison to Other Semantics}
+\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2
 
-The differences between our diagonal semantics, the erasure semantics, and Davies' semantics can be characterized entirely by how they behave at $\next$ expressions:
-\begin{itemize}
-\item {\bf Erasure Semantics.} Evaluate contents of $\next$ immediately.
-\item {\bf Diagonal Semantics.} Speculate to find the $\prev$s, evaluate them in place, then place the contents of $\next$ into a table and create a hatted variable to act as a value.
-\item {\bf Davies' Semantics.} Speculate to find the $\prev$s, evaluate them in place, then treat the reduced $\next$ as a value.
-\end{itemize}
+\noindent\TODO put $\Xi$ and $\Gamma$ in the judgments where appropriate
+
+% local macros; put them in macros.tex once they are more finalized
+\newcommand{\wf}{\ \mathsf{wf}}
+\newcommand{\dom}[1]{\mathsf{dom}(#1)}
+\newcommand{\valoo}{\ \bbone\textsf{-}\mathsf{val} @ \bbone}
+\newcommand{\valot}{\ \bbone\textsf{-}\mathsf{val} @ \bbtwo}
+\newcommand{\valtt}{\ \bbtwo\textsf{-}\mathsf{val} @ \bbtwo}
+
+\begin{definition}
+Contexts $\Xi$ and $\Gamma$ are well-formed ($\Xi\wf$, $\Gamma\wf$) if they
+contain only stage-2 variables.
+\end{definition}
+
+\begin{theorem}
+If $\Xi;\typesone e A$ then $\Xi\wf$, $\Gamma\wf$, and $A\istypeone$.
+\end{theorem}
+
+\begin{theorem}
+If $\Xi;\Gamma\vdash\diaonesub$ then:
+\begin{enumerate}
+\item $\typesone e A$ for some 1-type $A$;
+\item $v\valoo$;
+\item \TODO unfinished
+\end{enumerate}
+\end{theorem}
 

### 2014-06-06 paper/intro.tex
Refactor paper into multiple files for cleanliness.
Probably broke build, so will look over Nico's shoulder to fix.

--- 2014-05-13 paper/sigplanconf-template.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-\usepackage{stmaryrd}
-
-\begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Stage-Splitting a Modal Language}
-
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
-
-\maketitle
-
-\begin{abstract}
-This is the text of the abstract.
-\end{abstract}
-
-\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-\terms
-term1, term2
-
-\keywords
-keyword1, keyword2
-
 \section{Introduction}
@@
 This paper explores both theory and applications for stage-splitting.
-
-\section{Stage-Splitting Definition and Comparison to Partial Evaluation}
-
-First, we review the definition of partial evaluation.
-Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.
-This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.
-The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.
-We now state this theorem more formally: a partial evaluator is some function $p$ such that,
-\[
-	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
-\]
-where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
-
-Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.
-Again, more formally, a stage-splitter is some $s$ such that,
-\[
-	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
-	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
-\]
-
-We first discuss a few similarities between partial-evaluation and stage-splitting.
-First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.
-
-Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.
-Indeed, both admit completely trivial definitions.
-Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available.
-The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.
-Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.
-
-The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.
-For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.
-Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.
-Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.
-Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.
-
-We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
-However, one {\em cannot} construct a valid stage-splitter out of a partial evaluator, since a partial evaluator by definition depends on the first-stage input, whereas a stage-splitter does not have that available.
-
-Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
-In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
-It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer.
-[Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
-
-\section{A Two-Stage Modal Language}
-
-Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
-
-\section{Straw Semantics}
-
-Introduce the straw semantics, previously known as the ``erasure" semantics.
-
-\section{Data and Control Dependency}
-
-State a data-dependency theorem (that stage 1 values cannot depend on stage 2 values).
-It should be true for the straw semantics!
-Also state a control dependency theorem (that the execution of stage 1 code can't depend on stage 2 values).
-This won't be true for the straw semantics, and show counterexample.
-
-\section{Diagonal Semantics}
-
-Present another semantics which obeys non-backwards control dependency.
-This might be called the diagonal semantics.
-Go through an example such as currying to build intuition for why this is really necessary (i.e., if we don't have these weird rules, currying breaks down!).
-
-\section{Splitting Algorithm}
-
-Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.
-
-\section{Examples for Staged Pipelines}
-
-Give the gist of one-to-one pipeline example (like client/server).
-Then talk about a one-to-many pipeline.
-Then talk about a many-to-one pipeline like spark.
-It clear how to target something like this for known base types on the boundary, and for product types.
-But sums on the boundary are hard!
-We leave many-to-one as future-work.
-
-\section{Examples of Algorithm Derivation}
-
-Fast exponent example.  Quickselect example.
-
-Things to try: an interpreter which, partially evaluated, does cps or something.
-
-For each of these examples, talk about what partial evaluation would do and why that might be bad.
-
-\section{Related Work}
-
-Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
-They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
-They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
-The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
-Accordingly, their approach was entirely informal, with no implementation realized.
-Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
-
-Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
-The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
-Their shaders are written in a C-like language with basic arithmetic and if statements.
-Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
-
-Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
-But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
-Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
-Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
-Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
-
-[RTSL and SH]
-
-[Discuss Yong's recent paper here.
-It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
-They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
-If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
-
-Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
-That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
-Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
-The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
-They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
-
-[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
-
-[What's going on with names and necessity?]
-
-[Our work bears a lot of similarity to ML5, which also uses a modal type system.
-The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others).
-The type systems reflect this directly in the world accessibility relation.
-There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]
-
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
-
-\acks
-
-Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliography{paper}
-\bibliographystyle{abbrvnat}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-06-06 paper/paper.tex
Refactor paper into multiple files for cleanliness.
Probably broke build, so will look over Nico's shoulder to fix.

--- 2014-06-05 paper/paper.tex
@@
 
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{definition}[theorem]{Definition}
+
 \begin{document}
@@
 \input {grammars}
+\input {abstract}
+\input {intro}
+\input {relatedwork}
 \input {semantics}
@@
 \begin{abstract}
-This is the text of the abstract.
+\input{abstract}
 \end{abstract}
@@
 
-\section{Introduction}
+\input{intro}
+\input{relatedwork}
+\input{semantics}
+\input{splitting}
 
-Staged computation has existed as a programming technique for over three decades.
-Jorring et al.
-(\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.
-The first two of these have received significantly more attention than the third.
+%\appendix
+%\section{Appendix Title}
+%This is the text of the appendix, if you need one.
 
-We call these techniques {\em staged} in that part of their input comes in the first stage, and part comes in the second stage.  Likewise, some computation occurs at each stage as well.
-
-Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).
-Partial evaluation, too, is well-understood.
-Partial evaluation systems exist...
-.
-This paper explores both theory and applications for stage-splitting.
-
-\section{Stage-Splitting Definition and Comparison to Partial Evaluation}
-
-First, we review the definition of partial evaluation.
-Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.
-This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.
-The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.
-We now state this theorem more formally: a partial evaluator is some function $p$ such that,
-\[
-	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
-\]
-where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
-
-Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.
-Again, more formally, a stage-splitter is some $s$ such that,
-\[
-	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
-	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
-\]
-
-We first discuss a few similarities between partial-evaluation and stage-splitting.
-First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.
-
-Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.
-Indeed, both admit completely trivial definitions.
-Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available.
-The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.
-Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.
-
-The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.
-For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.
-Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.
-Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.
-Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.
-
-\subsection{Partial Evaluator from Stage-Splitter}
-
-We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
-Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
-In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
-It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer.
-[Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
-
-\subsection{Stage-Splitter from Partial Evaluator}
-
-Likewise, we can easily recover a stage-splitter from a partial evaluator.  If $p$ is a valid partial evaluator, then we can define a stage splitter $s$ such that $s(f)=(f_1,f_2)$, where
-\begin{align*}
-[f_1](x) &= p (f,x) \\
-[f_2](l,y) &= [l] (y)
-\end{align*}
-This implicitly requires that the languages in which $f_1$ and $f_2$ are expressed are strong enough to write a partial evaluator, but that is the case in this paper.
-A stage-splitter defined this way leaves much to be desired.
-Firstly, partial evaluation of $f$ may be too expensive for the context in which $f_1$ needs to run.
-Additionally, the intermediate data structure created this way may be much larger than necessary, as it would contain all of the residual code.
-
-\section{A Two-Stage Modal Language}
-
-Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
-
-\section{Diagonal Semantics}
-
-We desire a semantics that meets the following goals:
-
-\begin{enumerate}
-\item In absence of $\next$ expressions, it should be equivalent to standard call-by-value evaluation.
-\item $\next$ and $\prev$ should be inverses of the other.
-\item All of the first stage work should be completed before second stage work.
-\item The result for a stage-2 expression should be the same as if all of the $\next$s and $\prev$s were interpreted as the identity function.
-\end{enumerate}
-
-In this section, we introduce a dynamic semantics for our language.
-Although our type system is virtually identical to that in \cite{davies96}, the semantics is different in terms of cost and termination behavior.
-The semantics comprises the following judgments:
-
-\begin{center}
-\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
-Judgment & Reads & Conditions \\ \hline 
-$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
-$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\gamma$ and value $v$'' 
-& $\typesone e A$ 
-	\newline $\isvalone v$ 
-	\newline ... \\ \hline 
-$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
-...&...&... \\ \hline
-\end{tabular}
-\end{center}
-
-Evaluation and speculation are the main judgments here, the rest being largely administrative.
-The evaluation judgment operates on stage-1 code whereas the speculation operates on stage-2 code.
-Since $\next$ and $\prev$ are the crossover points between 1-code and 2-code, they are correspondingly the only places where the evaluation and speculation judgments depend on the other.
-
-We can think of our evaluation as proceeding in the standard way for stage-1 code.
-But when the evaluator encounters stage-2 code, it places that code off to the side in a table, and then keeps a reference to the table entry.
-These references are then passed around as stage-1 values, and the tables are managed accordingly.
-
-The evaluation judgment is very similar to standard call-by-value evaluation, as goal 1 would suggest.
-The input to evaluation is a stage-1 expression (usually $e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered later.
-Evaluation has two outputs: the {\em future table} (usually $\xi$) and the {\em partial value} (usually $v$).
-We cover the latter first.
-The partial value is essentially the result of the first-stage portion of $e$, and must be a 1-value of the same type as $e$.
-As usual, this means that $v$ must be composed only of base primitives, tuples, injections, and lambdas (corresponding to base types, products, sums, and functions).
-Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.
-Those stage-2 computations are held in
-
-\subsection{Adherence to Goals}
-
-We meet our first (discuss) and second goals (show examples).
-
-That we meet our third goal is clear from just the structure of our judgments (...).
-
-Under the strictest interpretation, we do not meet our third goal.  Consider the following examples (show speculating down an if).
-
-But as it turns out, a semantics which had the erasure behavior would not meet goal 3! (explain using the same example)
-
-\section{Splitting Algorithm}
-
-[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
-
-The theorem that we want to be true is...
-
-%\begin{center}
-%\begin{tabular}{l}
-%If $\splittwo{e}{A}{p,l.r}$ \\
-%and $\reducetwo{e}{v}$ \\
-%then $\reduce{p}{u}$ \\
-%and $\reduce{[u/l]r}{w}$ \\
-%and $v = w$ 
-%\end{tabular}
-%\end{center}
-
-%\begin{center}
-%\begin{tabular}{l}
-%If $\splitone {e}{A}{c,l.r}$ \\
-%and $\colone {\gamma_1}{\Gamma}$ \\
-%and $\coltwo {\gamma_2}{\Gamma}$\\
-%and $\rsone{\gamma_1(e)}{v}{q}$ \\
-%and $\reducetwo{\gamma_2(q)}{w}$ \\
-%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
-%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
-%\end{tabular}
-%~~~
-%\begin{tabular}{l}
-%If $\splittwo{e}{B}{p,l.r}$ \\
-%and $\colone {\gamma_1}{\Gamma}$ \\
-%and $\coltwo {\gamma_2}{\Gamma}$\\
-%and $\rstwo{\gamma_1(e)}{q}$ \\
-%and $\reducetwo{\gamma_2(q)}{w}$ \\
-%then $\reduce{|\gamma_1|(p)}{u}$ \\
-%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
-%\end{tabular}
-%\end{center}
-
-\section{Examples for Staged Pipelines}
-
-Give the gist of one-to-one pipeline example (like client/server).
-Then talk about a one-to-many pipeline.
-Then talk about a many-to-one pipeline like spark.
-It clear how to target something like this for known base types on the boundary, and for product types.
-But sums on the boundary are hard!
-We leave many-to-one as future-work.
-
-\section{Examples of Algorithm Derivation}
-
-Fast exponent example.
-
-\lstset{language=ML, columns=fullflexible, basicstyle=\ttfamily, tabsize=4, escapeinside={"*}{*"}}
-\begin{lstlisting} 
-let exp (b : $int, e : int) : $int =
-	if e == 0 then
-		next{1}
-	else if (e mod 2) == 0 then
-		next{let x = prev{exp(b,e/2)} in x*x}
-	else
-		next{prev{b} * prev{exp (b,e-1)}}		
-\end{lstlisting}
-
-splits into
-
-\begin{lstlisting} 
-let exp (b, e) =
-	((), roll (
-		if e == 0 then
-			inL ()
-		else 
-			inR (
-				if (e mod 2) == 0 then
-					inL (#2 (exp (b,e/2)))
-				else
-					inR (#2 (exp (b,e-1)))
-			)
-	))
-\end{lstlisting}
-
-and
-
-\begin{lstlisting} 
-let exp ((b, e), p) =
-	case unroll p of
-	  () => 1
-	| d  =>
-		case d of
-		  r => let x = exp ((b,()),r) in x*x
-		| r => b * exp ((b,()),r)
-\end{lstlisting}
-
-Quickselect example.
-
-\begin{lstlisting} 
-let qs (l : "*$\mathtt{\mu \alpha. }$*"() + "*$\mathtt{int*\alpha}$*", i: $int) = 
-	case unroll l of
-	  () => next {0}
-	| (h,t) => 
-		let (left,right,n) = partition h t in
-		next{
-			let n = prev{hold n} in
-			case compare prev{i} n of
-			  () (*LT*) => prev {qs left i}
-			| () (*EQ*) => hold h
-			| () (*GT*) => 
-				prev {qs right next{prev{i}-n-1}}
-		}	
-\end{lstlisting}
-
-Things to try: an interpreter which, partially evaluated, does cps or something.
-
-For each of these examples, talk about what partial evaluation would do and why that might be bad.
-
-\section{Related Work}
-
-Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
-They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
-They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
-The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
-Accordingly, their approach was entirely informal, with no implementation realized.
-Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
-
-Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
-The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
-Their shaders are written in a C-like language with basic arithmetic and if statements.
-Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
-
-Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
-But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
-Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
-Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
-Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
-
-[RTSL and SH]
-
-[Discuss Yong's recent paper here.
-It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
-They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
-If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
-
-Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
-That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
-Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
-The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
-They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
-
-[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
-
-[What's going on with names and necessity?]
-
-[Our work bears a lot of similarity to ML5, which also uses a modal type system.
-The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others).
-The type systems reflect this directly in the world accessibility relation.
-There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]
-
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
-
-\acks
-
-Acknowledgments, if needed.
+%\acks
+%Acknowledgments, if needed.
 

### 2014-06-06 paper/relatedwork.tex
Refactor paper into multiple files for cleanliness.
Probably broke build, so will look over Nico's shoulder to fix.

--- 2014-05-24 paper/sigplanconf-template.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-%\usepackage{amssymb}
-\usepackage{bbm}
-\usepackage{stmaryrd}
-\usepackage{proof}
-\usepackage{mathpartir}
-\usepackage{mathabx}
-
-\begin{document}
-
-\input {macros}
-\input {grammars}
-\input {semantics}
-\input {splitting}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Stage-Splitting a Modal Language}
-
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
-
-\maketitle
-
-\begin{abstract}
-This is the text of the abstract.
-\end{abstract}
-
-\category{CR-number}{subcategory}{third-level}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-\terms
-term1, term2
-
-\keywords
-keyword1, keyword2
-
-\section{Introduction}
-
-Staged computation has existed as a programming technique for over three decades.
-Jorring et al.
-(\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.
-The first two of these have received significantly more attention than the third.
-
-We call these techniques {\em staged} in that part of their input comes in the first stage, and part comes in the second stage.  Likewise, some computation occurs at each stage as well.
-
-Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).
-Partial evaluation, too, is well-understood.
-Partial evaluation systems exist...
-.
-This paper explores both theory and applications for stage-splitting.
-
 \section{Stage-Splitting Definition and Comparison to Partial Evaluation}
@@
 Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
-
-\section{Straw Semantics}
-
-Introduce the straw semantics, previously known as the ``erasure" semantics.
-
-\section{Data and Control Dependency}
-
-State a data-dependency theorem (that stage 1 values cannot depend on stage 2 values).
-It should be true for the straw semantics!
-Also state a control dependency theorem (that the execution of stage 1 code can't depend on stage 2 values).
-This won't be true for the straw semantics, and show counterexample.
-
-\section{Diagonal Semantics}
-
-Present another semantics which obeys non-backwards control dependency.
-This might be called the diagonal semantics.
-Go through an example such as currying to build intuition for why this is really necessary (i.e., if we don't have these weird rules, currying breaks down!).
-
-\section{Splitting Algorithm}
-
-[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
-
-The theorem that we want to be true is...
-
-\begin{center}
-\begin{tabular}{l}
-If $\splittwo{e}{A}{p,l.r}$ \\
-and $\reducetwo{e}{v}$ \\
-then $\reduce{p}{u}$ \\
-and $\reduce{[u/l]r}{w}$ \\
-and $v = w$ 
-\end{tabular}
-\end{center}
-
-%\begin{center}
-%\begin{tabular}{l}
-%If $\splitone {e}{A}{c,l.r}$ \\
-%and $\colone {\gamma_1}{\Gamma}$ \\
-%and $\coltwo {\gamma_2}{\Gamma}$\\
-%and $\rsone{\gamma_1(e)}{v}{q}$ \\
-%and $\reducetwo{\gamma_2(q)}{w}$ \\
-%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
-%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
-%\end{tabular}
-%~~~
-%\begin{tabular}{l}
-%If $\splittwo{e}{B}{p,l.r}$ \\
-%and $\colone {\gamma_1}{\Gamma}$ \\
-%and $\coltwo {\gamma_2}{\Gamma}$\\
-%and $\rstwo{\gamma_1(e)}{q}$ \\
-%and $\reducetwo{\gamma_2(q)}{w}$ \\
-%then $\reduce{|\gamma_1|(p)}{u}$ \\
-%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
-%\end{tabular}
-%\end{center}
-
-\section{Examples for Staged Pipelines}
-
-Give the gist of one-to-one pipeline example (like client/server).
-Then talk about a one-to-many pipeline.
-Then talk about a many-to-one pipeline like spark.
-It clear how to target something like this for known base types on the boundary, and for product types.
-But sums on the boundary are hard!
-We leave many-to-one as future-work.
-
-\section{Examples of Algorithm Derivation}
-
-Fast exponent example.  Quickselect example.
-
-Things to try: an interpreter which, partially evaluated, does cps or something.
-
-For each of these examples, talk about what partial evaluation would do and why that might be bad.
-
-\section{Related Work}
-
-Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
-They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
-They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
-The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
-Accordingly, their approach was entirely informal, with no implementation realized.
-Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
-
-Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
-The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
-Their shaders are written in a C-like language with basic arithmetic and if statements.
-Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
-
-Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
-But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
-Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
-Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
-Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
-
-[RTSL and SH]
-
-[Discuss Yong's recent paper here.
-It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
-They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
-If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
-
-Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
-That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
-Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
-The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
-They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
-
-[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
-
-[What's going on with names and necessity?]
-
-[Our work bears a lot of similarity to ML5, which also uses a modal type system.
-The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others).
-The type systems reflect this directly in the world accessibility relation.
-There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]
-
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
-
-\acks
-
-Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliography{paper}
-\bibliographystyle{abbrvnat}
-
-\end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2014-06-06 paper/semantics.tex
Stylistic changes, pending completion of value proof.
Added figures for grammar, values, types, statics, and dynamics.
first draft of reference semantics in
Fixed up the semantics some more.
Quick fixes from our meeting.
Reorganized semantics.
changed all figures to have captions at the bottom
various cleanup
Hark! The diagonal semantics returns!
added a few words about the semantics
more words about semantics.
added contexts to all the rules
quick typo
another typo
Refactor paper into multiple files for cleanliness.
Probably broke build, so will look over Nico's shoulder to fix.

@@
 \begin{figure}
+\begin{mathpar}
+\infertypeswor [\rmunit] 	{{\tt unit} \istypewor}								{\cdot} 																			\and
+\infertypeswor [int]			{{\tt int} \istypewor}								{\cdot} 																			\and
+\infertypeswor [bool]			{{\tt bool} \istypewor}								{\cdot} 																			\and
+\infertypeswor [\times]		{A \times B \istypewor}								{A \istypewor & B \istypewor} 													\and
+\infertypeswor [+]			{A + B \istypewor}									{A \istypewor & B \istypewor} 													\and
+\infertypeswor [\to]			{A \to B \istypewor}									{A \istypewor & B \istypewor} 													\and
+\infertypeswor [\fut]			{\fut A \istypeone}									{A \istypetwo} 													
+\end{mathpar}
+\caption{\ellStaged~Valid Types}
+\label{fig:validTypes}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\infertypeswor [\rmunit] 	{() \isvalwor}											{\cdot} 																			\and
+\infertypeswor [int]			{i \isvalwor}											{\cdot} 																			\and
+\infertypeswor [bool]			{b \isvalwor}											{\cdot} 																			\and
+\infertypeswor [\times]		{(v_1, v_2) \isvalwor}								{v_1 \isvalwor & v_2 \isvalwor} 												\and
+\infertypeswor [+_1]			{\inl~v \isvalwor}									{v \isvalwor} 																		\and
+\infertypeswor [+_2]			{\inr~v \isvalwor}									{v \isvalwor} 																		\and
+\infertypeswor [\to]			{\lam x A e \isvalwor}								{\cdot} 																			\and
+\infertypeswor [\fut]			{\isvalone {\hat y}}									{\cdot} 													
+\end{mathpar}
+\caption{\ellStaged~Valid Values}
+\label{fig:validValues}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\infertypeswor [\rmunit] 	{\ty {()}\rmunit}								{\cdot} 																			\and
+\infertypeswor [int]			{\ty {i}{\rm int}}							{\cdot} 																			\and
+\infertypeswor [bool]			{\ty {b} \rmbool}								{\cdot} 																			\and
+\infertypeswor [hyp]			{\ty x A}										{\col x A \in \Gamma} 															\and
+\infertypeswor [\to I]		{\ty {\lam{x}{A}{e}} {A \to B}}				{A \istypewor & \ty [,\col x A] e B} 											\and
+\infertypeswor [\to E]		{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 											\and
+\infertypeswor [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}				{\ty {e_1} A & \ty [,\col x A] {e_2} B} 										\and
+\infertypeswor [\times I]	{\ty {(e_1,e_2)}{A\times B}}					{\ty {e_1} A & \ty {e_2} B} 														\and
+\infertypeswor [\times E_1]	{\ty {\pio~e} A}								{\ty e {A\times B}} 																\and
+\infertypeswor [\times E_2]	{\ty {\pit~e} B}								{\ty e {A\times B}} 																\and
+\infertypeswor [+ I_1]		{\ty {\inl~e} {A + B}}						{\ty e A} 																			\and
+\infertypeswor [+ I_2]		{\ty {\inr~e} {A + B}}						{\ty e B} 																			\and
+\infertypeswor [+ E]			{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C} 	\and
+\infertypeswor [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}			{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} 								\and
+\infertypeswor [\fut I]		{\typesone {\next~e}{\fut A}}				{\typestwo e A} 																	\and
+\infertypeswor [\fut E]		{\typestwo {\prev~e} A}						{\typesone e {\fut A}} 															\and
+\end{mathpar}
 \caption{\ellStaged~Static Semantics}
 \label{fig:coreStatics}
-\begin{mathpar}
-\infertypesone [\rmunit] 	{\ty {()}\rmunit}						{\cdot} \and
-\infertypesone [int]			{\ty {i}{\rm int}}					{\cdot} \and
-\infertypesone [bool]			{\ty {b} \rmbool}						{\cdot} \and
-\infertypesone [hyp]			{\ty x A}								{x : A^\bbone \in \Gamma} \and
-\infertypestwo [\rmunit]		{\ty {()}{\rmunit}}					{\cdot} \and
-\infertypestwo [int]			{\ty {i}{\rm int}}					{\cdot} \and
-\infertypestwo [bool]			{\ty {b} \rmbool}						{\cdot} \and
-\infertypestwo [hyp]			{\ty x A}								{x : A^\bbtwo \in \Gamma} \and
-\infertypesone [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbone] {e_2} B} \and
-\infertypestwo [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}		{\ty {e_1} A & \ty [\Gamma,x:A^\bbtwo] {e_2} B} \and
-\infertypesone [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \ty {e_2} B} \and
-\infertypesone [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
-\infertypesone [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
-\infertypestwo [\times I]	{\ty {(e_1,e_2)}{A\times B}}			{\ty {e_1} A & \typestwo {e_2} B} \and
-\infertypestwo [\times E_1]	{\ty {\pi_1~e} A}						{\ty e {A\times B}} \and
-\infertypestwo [\times E_2]	{\ty {\pi_2~e} B}						{\ty e {A\times B}} \and
-\infertypesone [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
-\infertypestwo [if]			{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} \and
-\infertypesone [\fut I]		{\ty {\next~e}{\fut A}}				{\typestwo e A} \and
-\infertypestwo [\fut E]		{\ty {\prev~e} A}						{\typesone e {\fut A}} \and
-\end{mathpar}
 \end{figure}
 
-\begin{figure*}
-\caption{Sum and Function Typing}
-\label{fig:advancedTypes}
+\begin{figure}
 \begin{mathpar}
-\infertypesone [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbone] e B} 																\and
-\infertypesone [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
-\infertypestwo [\to I]	{\ty {\lam{x}{A}{e}} {A \to B}}				{\ty [\Gamma,x:A^\bbtwo] e B} 																\and
-\infertypestwo [\to E]	{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 														\and
-\infertypesone [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
-\infertypesone [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\\
-\infertypestwo [+ I_1]	{\ty {\inl~e} {A + B}}						{\ty e A} 																						\and
-\infertypestwo [+ I_2]	{\ty {\inr~e} {A + B}}						{\ty e B} 																						\and
-\infertypesone [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbone]{e_2} C & \ty[\Gamma,x_3:A^\bbone]{e_3} C} 	\and
-\infertypestwo [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[\Gamma,x_2:A^\bbtwo]{e_2} C & \ty[\Gamma,x_3:A^\bbtwo]{e_3} C} 	\and
+\inferdiaone [\to I]  		{\red {\lam{x}{A}{e}} {\cdot,\vallam{x}{e}}}					{\cdot} 																					\and
+\inferdiaone [\to E]  		{\red {e_1~e_2} {\gcomp 1 2 \circ \xi',v'}}					{\red {e_1} {\xi_1,\vallam{x}{e'}} & \sub [2] & \red [\Xi,dom(\xi_1),dom(\xi_2);\Gamma] {[v_1/x]e'}{\xi',v'}}	\and
+%
+\inferdiaone [\rmunit]   	{\red {()}{\cdot,()}}												{\cdot} 																					\and
+\inferdiaone [\times I]  	{\red {(e_1,e_2)}{\gcomp 1 2,\valprod{v_1}{v_2}}}				{\sub [1] & \sub [2]} 																	\and
+\inferdiaone [\times E_1]	{\red {\pio~e}{\xi,v_1}}											{\red{e}{\xi,\valprod{v_1}{v_2}}} 													\and
+\inferdiaone [\times E_2]	{\red {\pit~e}{\xi,v_2}}											{\red{e}{\xi,\valprod{v_1}{v_2}}} 													\and
+\inferdiaone [+ I_1]  		{\red {\inl~e} {\xi,\inl~v}}										{\sub} 																						\and
+\inferdiaone [+ I_2]  		{\red {\inr~e} {\xi,\inr~v}}										{\sub} 																						\and
+%
+\inferdiaone [+ E_1]  		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{\gcomp 1 2,v_2}}		{\red {e_1}{\xi_1,\valinl~v} & \red [\Xi,dom(\xi_1);\Gamma]{[v/x_2]e_2}{\xi_2,v_2}} 					\and
+\inferdiaone [+ E_2]  		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{\gcomp 1 2,v_3}}		{\red {e_1}{\xi_1,\valinr~v} & \red [\Xi,dom(\xi_1);\Gamma]{[v/x_3]e_3}{\xi_3,v_3}}					\and
+%
+\inferdiaone [int]			{\red {i}{\cdot,i}}												{\cdot} 																					\and
+\inferdiaone [bool]   		{\red {b}{\cdot,b}}												{\cdot} 																					\and
+\inferdiaone [let]			{\red {\letin{x}{e_1}{e_2}}{\gcomp 1 2, v_2}}					{\sub [1] & \red [\Xi,dom(\xi_1);\Gamma] {[v_1/x]e_2}{\xi_2,v_2}}											\and
+\inferdiaone [if_T]   		{\red {\ifthen{e_1}{e_2}{e_3}}{\gcomp 1 2,v}}					{\red {e_1}{\xi_1,\bf true}  & \red {e_2}{\xi_2,v}} 							\and
+\inferdiaone [if_F]   		{\red {\ifthen{e_1}{e_2}{e_3}}{\gcomp 1 3,v}}					{\red {e_1}{\xi_1,\bf false} & \red {e_3}{\xi_3,v}}								
 \end{mathpar}
-\end{figure*}
+\caption{\ellStaged~Diagonal Semantics (core)}
+\label{fig:diaSemanticsCore}
+\end{figure}
 
 \begin{figure}
-\caption{\ellStaged~Erasure Dynamic Semantics}
-\label{fig:erasureSemantics}
 \begin{mathpar}
-\inferreduceone [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
-\inferreduceone [int]				{\red {i}{i}}								{\cdot} 											\and
-\inferreduceone [bool]			{\red {b}{b}}								{\cdot} 											\and
-\inferreduceone [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
-\inferreducetwo [\rmunit]		{\red {()}{()}}							{\cdot} 											\and
-\inferreducetwo [int]				{\red {i}{i}}								{\cdot} 											\and
-\inferreducetwo [bool]			{\red {b}{b}}								{\cdot} 											\and
-\inferreducetwo [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}		{\sub [1] & \red{[v_1/x]e_2}{v_2}} 				\and
-\inferreduceone [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]} 							\and
-\inferreduceone [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreduceone [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreducetwo [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}	{\sub [1] & \sub [2]}								\and
-\inferreducetwo [\times E_1]		{\red {\pi_1~e}{v_1}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreducetwo [\times E_2]		{\red {\pi_2~e}{v_2}}						{\red{e}{\valprod{v_1}{v_2}}} 					\and
-\inferreduceone [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
-\inferreduceone [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
-\inferreducetwo [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf true}  & \red{e_2}{v}} 			\and
-\inferreducetwo [if_F] 			{\red {\ifthen{e_1}{e_2}{e_3}}{v}}		{\red{e_1}{\bf false} & \red{e_3}{v}}			\\
-\inferreduceone [\fut I]			{\red {\next~e}{\valnext~v}}				{\reducetwosub}									\and
-\inferreducetwo [\fut E]			{\red {\prev~e} v}						{\reduceone e {\valnext~v}} 
+\inferdiaone [\fut I]			{\red {\next~e}{y \mapsto q,\hat y}}				{\diatwo e q}												\and
+\inferdiaspc [\fut E]			{\red {\prev~e} {\xi (y)}}						{\diaone e {\xi, \hat y}} 									
 \end{mathpar}
+\caption{\ellStaged~Diagonal Semantics ($\next$ and $\prev$)}
+\label{fig:diaSemanticsNP}
 \end{figure}
 
-\begin{figure*}
-\caption{Sum and Function Erasure Evaluation}
-\label{fig:advancedTypes}
+\begin{figure}
 \begin{mathpar}
-\inferreduceone [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
-\inferreduceone [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
-\inferreducetwo [\to I]	{\red {\lam{x}{A}{e}} {\vallam{x}{M}}}			{\cdot} 																		\and
-\inferreducetwo [\to E]	{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
-\inferreduceone [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
-\inferreduceone [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
-\inferreducetwo [+ I_1]	{\red {\inl~e} {\valinl~v}}						{\sub} 																			\and
-\inferreducetwo [+ I_2]	{\red {\inr~e} {\valinr~v}}						{\sub} 																			\\
-\inferreduceone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
-\inferreduceone [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
-\inferreducetwo [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinl~v} & \red {[v/x_2]e_2}{v'}} 							\and
-\inferreducetwo [+ E_2]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}	{\red {e_1}{\valinr~v} & \red {[v/x_3]e_3}{v'}} 							\and
+\inferdiaspc [\to I]			{\red {\lam{x}{A}{e}}{\lam{x}{A}{q}}}						{\diatwo [\Xi;\Gamma,x] e q} 			\and
+\inferdiaspc [\to E]			{\red {e_1~e_2}{q_1,q_2}}									{\sub [1] & \sub [2]}						\and		
+\inferdiaspc [\rmunit]		{\red {()}{()}}												{\cdot} 									\and
+\inferdiaspc [\times I]		{\red {(e_1,e_2)}{(q_1,q_2)}}								{\sub [1] & \sub [2]} 					\and
+\inferdiaspc [\times E_1]	{\red {\pio~e}{\pio~q}}										{\sub} 										\and
+\inferdiaspc [\times E_2]	{\red {\pit~e}{\pit~q}}										{\sub} 										\and
+\inferdiaspc [+ I_1]			{\red {\inl~e}{\inl~q}}										{\sub} 										\and
+\inferdiaspc [+ I_2]			{\red {\inr~e}{\inr~q}}										{\sub} 										\and
+\inferdiaspc [+ E_1]			{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
+									{\caseof{q_1}{x_2.q_2}{x_3.q_3}}}						{\sub [1] & \diatwo [\Xi;\Gamma,x_2] {e_2} {q_2} & \diatwo [\Xi;\Gamma,x_3] {e_3} {q_3}} 		\and
+\inferdiaspc [int]			{\red {i}{i}}													{\cdot} 									\and
+\inferdiaspc [bool]			{\red {b}{b}}													{\cdot} 									\and
+\inferdiaspc [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{q_1}{q_2}}}			{\sub [1] & \diatwo [\Xi;\Gamma,x] {e_2} {q_2}} 					\and
+\inferdiaspc [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{q_1}{q_2}{q_3}}}	{\sub [1] & \sub [2] & \sub [3]} 			
 \end{mathpar}
-\end{figure*}
+\caption{\ellStaged~New Speculation}
+\label{fig:diaSemanticsSpec}
+\end{figure}
+%
+%\begin{figure}
+%\begin{mathpar}
+%\inferreducewor [\to I]			{\redone {\lam{x}{A}{e}} {\vallam{x}{e}}}		{\cdot} 																		\and
+%\inferreducewor [\to I]			{\redtwo {\lam{x}{A}{e}} {\vallam{x}{e}}}		{\specwor e} 																	\and
+%\inferreducewor [\to E]			{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
+%%
+%\inferreducewor [\rmunit]		{\red {()}{()}}									{\cdot} 																		\and
+%\inferreducewor [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}			{\sub [1] & \sub [2]} 														\and
+%\inferreducewor [\times E_1]		{\red {\pio~e}{v_1}}								{\red{e}{\valprod{v_1}{v_2}}} 												\and
+%\inferreducewor [\times E_2]		{\red {\pit~e}{v_2}}								{\red{e}{\valprod{v_1}{v_2}}} 												\and
+%\inferreducewor [+ I_1]			{\red {\inl~e} {\inl~v}}							{\sub} 																			\and
+%\inferreducewor [+ I_2]			{\red {\inr~e} {\inr~v}}							{\sub} 																			\and
+%%
+%\inferreducewor [+ E_1]			{\redone {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redone {e_1}{\valinl~v} & \redone {[v/x_2]e_2}{v'}} 					\and
+%\inferreducewor [+ E_2]			{\redone {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redone {e_1}{\valinr~v} & \redone {[v/x_3]e_3}{v'}}						\and
+%\inferreducewor [+ E_1]			{\redtwo {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redtwo {e_1}{\valinl~v} & \redtwo {[v/x_2]e_2}{v'} & \specwor {e_3}}	\and
+%\inferreducewor [+ E_2]			{\redtwo {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redtwo {e_1}{\valinr~v} & \redtwo {[v/x_3]e_3}{v'} & \specwor {e_2}}	\and
+%%
+%\inferreducewor [int]				{\red {i}{i}}											{\cdot} 																		\and
+%\inferreducewor [bool]			{\red {b}{b}}											{\cdot} 																		\and
+%\inferreducewor [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}					{\sub [1] & \red{[v_1/x]e_2}{v_2}} 											\and
+%\inferreducewor [if_T] 			{\redone {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redone {e_1}{\bf true}  & \redone {e_2}{v}} 								\and
+%\inferreducewor [if_F] 			{\redone {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redone {e_1}{\bf false} & \redone {e_3}{v}}								\and
+%\inferreducewor [if_T] 			{\redtwo {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redtwo {e_1}{\bf true}  & \redtwo {e_2}{v} & \specwor {e_3}}			\and
+%\inferreducewor [if_F] 			{\redtwo {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redtwo {e_1}{\bf false} & \redtwo {e_3}{v} & \specwor {e_2}}
+%\end{mathpar}
+%\caption{\ellStaged~Old Semantics (core)}
+%\label{fig:erasureSemanticsCore}
+%\end{figure}
+%
+%\begin{figure}
+%\begin{mathpar}
+%\inferreducewor [\fut I]			{\reducexpl \bbone \bbtwo {\next~e}{\next~v}}	{\reducexpl \bbtwo \bbtwo e v}												\and
+%\inferreducewor [\fut I]			{\reducexpl \bbone \bbone {\next~e}{DUM}}		{\specwor e }																	\and
+%\inferreducewor [\fut E]			{\reducexpl \bbtwo \bbtwo {\prev~e} v}			{\reducexpl \bbone \bbtwo e {\valnext~v}} 									\and
+%\inferreducespc [\fut E]			{\red {\prev~e}}									{\reducexpl \bbone \bbone e {DUM}} 								
+%\end{mathpar}
+%\caption{\ellStaged~Old Semantics ($\next$ and $\prev$)}
+%\label{fig:erasureSemanticsNextPrev}
+%\end{figure}	
+%
+%\begin{figure}
+%\begin{mathpar}
+%\inferreducespc [\to I]			{\red {\lam{x}{A}{e}}}							{\sub} 										\and
+%\inferreducespc [\to E]			{\red {e_1~e_2}}									{\sub [1] & \sub [2]}						\and		
+%\inferreducespc [\rmunit]		{\red {()}}										{\cdot} 									\and
+%\inferreducespc [\times I]		{\red {(e_1,e_2)}}								{\sub [1] & \sub [2]} 					\and
+%\inferreducespc [\times E_1]		{\red {\pio~e}}									{\sub [1]} 								\and
+%\inferreducespc [\times E_2]		{\red {\pit~e}}									{\sub [1]} 								\and
+%\inferreducespc [+ I_1]			{\red {\inl~e}}									{\sub} 										\and
+%\inferreducespc [+ I_2]			{\red {\inr~e}}									{\sub} 										\and
+%\inferreducespc [+ E_1]			{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}}			{\sub [1] & \sub [2] & \sub [3]} 		\and
+%\inferreducespc [int]				{\red {i}}											{\cdot} 									\and
+%\inferreducespc [bool]			{\red {b}}											{\cdot} 									\and
+%\inferreducespc [let]				{\red {\letin{x}{e_1}{e_2}}}						{\sub [1] & \sub [2]} 					\and
+%\inferreducespc [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}}					{\sub [1] & \sub [2] & \sub [3]} 			
+%\end{mathpar}
+%\caption{\ellStaged~Old Speculation}
+%\label{fig:erasureSemanticsSpec}
+%\end{figure}
+%
+%\begin{figure}
+%\caption{\ellStaged~Diagonal Semantics}
+%\label{fig:refSemantics}
+%\begin{mathpar}
+%\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
+%\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
+%\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
+%\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
+%\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
+%\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
+%\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
+%\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
+%\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
+%\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
+%\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
+%\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
+%\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
+%\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
+%\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
+%\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
+%\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
+%\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
+%\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
+%\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
+%\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
+%\end{mathpar}
+%\end{figure}
+%
+%\begin{figure*}
+%\caption{Sum and Function Diagonal Evaluation}
+%\label{fig:advancedTypes}
+%\begin{mathpar}
+%\inferRedsplitone [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{e}}{()}}			{\cdot} 																\and
+%\inferSpeculation [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{r}}}				{\sub} 																	\and
+%\inferRedsplitone [\to E]	{\red {e_1~e_2}{v'}{(r_1;\letin{x}{r_2}{r'})}}		{\red {e_1}{\vallam{x}{M}}{r_1} 
+%																							& \sub [2] & \red{[[e,x]/x]M}{v'}{r'}} 							\and
+%\inferSpeculation [\to E]	{\red {e_1~e_2}{r_1~r_2}}							{\sub [1] & \sub[2]}													\and
+%\inferRedsplitone [+ I_1]	{\red {\inl~e} {\valinl~v}{r}}						{\sub} 																	\and
+%\inferSpeculation [+ I_1]	{\red {\inl~e} {\valinl~r}}							{\sub} 																	\\
+%\inferRedsplitone [+ I_2]	{\red {\inr~e} {\valinr~v}{r}}						{\sub} 																	\and
+%\inferSpeculation [+ I_2]	{\red {\inr~e} {\valinr~r}}							{\sub} 																	\\
+%\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
+%								{\letin{x_2}{r_1}{r_2}}}								{\red {e_1}{\valinl~v}{r_1} & \red {[[v,x_2]/x_2]e_2}{v'}{r_2}} \and
+%\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
+%								{\letin{x_3}{r_1}{r_3}}}								{\red {e_1}{\valinr~v}{r_1} & \red {[[v,x_3]/x_3]e_3}{v'}{r_3}} \and
+%\inferSpeculation [+ E]		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
+%								{\caseof{r_1}{x_2.r_2}{x_3.r_3}}}				{\sub [1] & \sub [2] & \sub[3]}					 						\and
+%\end{mathpar}
+%\end{figure*}
 
-\begin{figure}
-\caption{\ellStaged~Diagonal Semantics}
-\label{fig:refSemantics}
-\begin{mathpar}
-\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
-\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
-\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
-\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
-\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
-\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
-\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
-\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
-\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
-\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
-\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
-\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
-\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
-\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
-\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
-\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
-\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
-\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
-\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
-\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
-\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
-\end{mathpar}
-\end{figure}
+\section{Semantics}
 
-\begin{figure*}
-\caption{Sum and Function Diagonal Evaluation}
-\label{fig:advancedTypes}
-\begin{mathpar}
-\inferRedsplitone [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{e}}{()}}			{\cdot} 																\and
-\inferSpeculation [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{r}}}				{\sub} 																	\and
-\inferRedsplitone [\to E]	{\red {e_1~e_2}{v'}{(r_1;\letin{x}{r_2}{r'})}}		{\red {e_1}{\vallam{x}{M}}{r_1} 
-																							& \sub [2] & \red{[[e,x]/x]M}{v'}{r'}} 							\and
-\inferSpeculation [\to E]	{\red {e_1~e_2}{r_1~r_2}}							{\sub [1] & \sub[2]}													\and
-\inferRedsplitone [+ I_1]	{\red {\inl~e} {\valinl~v}{r}}						{\sub} 																	\and
-\inferSpeculation [+ I_1]	{\red {\inl~e} {\valinl~r}}							{\sub} 																	\\
-\inferRedsplitone [+ I_2]	{\red {\inr~e} {\valinr~v}{r}}						{\sub} 																	\and
-\inferSpeculation [+ I_2]	{\red {\inr~e} {\valinr~r}}							{\sub} 																	\\
-\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
-								{\letin{x_2}{r_1}{r_2}}}								{\red {e_1}{\valinl~v}{r_1} & \red {[[v,x_2]/x_2]e_2}{v'}{r_2}} \and
-\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
-								{\letin{x_3}{r_1}{r_3}}}								{\red {e_1}{\valinr~v}{r_1} & \red {[[v,x_3]/x_3]e_3}{v'}{r_3}} \and
-\inferSpeculation [+ E]		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
-								{\caseof{r_1}{x_2.r_2}{x_3.r_3}}}				{\sub [1] & \sub [2] & \sub[3]}					 						\and
-\end{mathpar}
-\end{figure*}
+There's more than one way to define a semantics for our language.  We start this section by considering two proposals, characterizing them mainly by how they handle values of $\fut$ type.
+
+\subsection{The Erasure Semantics}
+
+The first proposal we call the {\em erasure semantics}, because it corresponds to what one would get by simply {\em erasing} all of the $\next$ and $\prev$ terms that appear in \lamStaged.
+In particular, we say that $\next v$ is a first-stage value of type $\fut A$ if and only if $v$ is a second-stage value of type $A$, (...)
+
+...
+
+We say that erasure semantics gives us a notion of multi-stage evaluation.
+One judgement is multistage evaluation at 1, and the other is multistage evaluation at 2.
+The downside of this is that the erasure semantics doesn't seem to do all of the stage-1 work before the stage-2 work, for some notion of ``before''.
+Of course, the erasure semantics might be equivalent to some other semantics which does have that property.
+Unfortunately, this is false (show second-stage if example).
+
+...
+
+The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.
+It also {\em obviously} produces the correct answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
+
+\subsection{Meta Semantics}
+
+A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.
+
+In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.
+But in the meta semantics, $\next e$ is a value only if $e$ has no $\prev$ terms; $v$ is allowed to have unreduced stage-2 computation.
+
+...
+
+The Davies semantics is comprised of two mutually recursive judgements: $\daviesz$ and $\davieso$.
+For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.
+(...
+This essentially gives us partial evaluation, and we're left to just evaluate the residual normally to get multi-stage evaluation...)
+
+The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.
+This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins.
+
+The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.
+This makes reasoning about second-stage cost rather difficult.
+
+\subsection{Our Semantics}
+
+We desire a semantics that meets the following goals:
+
+\begin{enumerate}
+\item Modulo termination, it should be equivalent to the erasure semantics.
+\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
+\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
+\end{enumerate}
+
+We meet all of these goals.
+
+In this section, we introduce a dynamic semantics for our language.
+Although our type system is virtually identical to that in \cite{davies96}, the semantics is different in terms of cost and termination behavior.
+
+Abstractly, we can think of our evaluation as proceeding in the standard way for stage-1 code.
+When the evaluator encounters a $\next \{e\}$ expression, it places $e$ off to the side in a table and replaces the whole expression with a reference to the table entry.
+These references are then passed around as stage-1 values for $\fut$ types.
+But what if $e$ contains $\prev$ expressions?
+To ensure that all stage-1 code is evaluated before any stage-2 code, we must evaluate all of the 1-code contained in $e$ before inserting it into the table.
+This entails searching $e$ for all contained $\prev$s and evaluating them in place.
+
+More formally, evaluation comprises the folowing judgments:
+
+\begin{center}
+\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
+Judgment & Reads & Conditions \\ \hline 
+$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
+$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
+& $\typesone e A$ 
+	\newline $\isvalone v$ 
+	\newline ... \\ \hline 
+$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
+...&...&... \\ \hline
+\end{tabular}
+\end{center}
+
+Evaluation and speculation are the main judgments here, the rest being largely administrative.
+The evaluation judgment operates on stage-1 code, whereas the speculation operates on stage-2 code.
+Since $\next$ and $\prev$ are the crossover points between 1-code and 2-code, they are correspondingly the only places where the evaluation and speculation judgments depend on the other.
+
+The evaluation judgment is very similar to standard call-by-value evaluation, as goal 1 would suggest.
+The input to evaluation is a stage-1 expression (usually $e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered later.
+Evaluation has two outputs: the {\em future table} (usually $\xi$) and the {\em partial value} (usually $v$).
+We cover the latter first.
+The partial value is essentially the result of the first-stage portion of $e$, and must be a 1-value of the same type as $e$.
+As usual, this means that $v$ must be composed only of base primitives, tuples, injections, and lambdas (corresponding to base types, products, sums, and functions).
+Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.
+
+...
+
+\subsection{Other nice properties}
+
+\begin{enumerate}
+\item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
+\end{enumerate}
+
+\subsection{Metatheory}
+
+\TODO unfinished; where should this go? combine with table above
+
+\noindent\TODO terminology for $\Xi,\Gamma,\xi,v,q$?
+
+\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2
+
+\noindent\TODO put $\Xi$ and $\Gamma$ in the judgments where appropriate
+
+% local macros; put them in macros.tex once they are more finalized
+\newcommand{\wf}{\ \mathsf{wf}}
+\newcommand{\dom}[1]{\mathsf{dom}(#1)}
+\newcommand{\valoo}{\ \bbone\textsf{-}\mathsf{val} @ \bbone}
+\newcommand{\valot}{\ \bbone\textsf{-}\mathsf{val} @ \bbtwo}
+\newcommand{\valtt}{\ \bbtwo\textsf{-}\mathsf{val} @ \bbtwo}
+
+\begin{definition}
+Contexts $\Xi$ and $\Gamma$ are well-formed ($\Xi\wf$, $\Gamma\wf$) if they
+contain only stage-2 variables.
+\end{definition}
+
+\begin{theorem}
+If $\Xi;\typesone e A$ then $\Xi\wf$, $\Gamma\wf$, and $A\istypeone$.
+\end{theorem}
+
+\begin{theorem}
+If $\Xi;\Gamma\vdash\diaonesub$ then:
+\begin{enumerate}
+\item $\typesone e A$ for some 1-type $A$;
+\item $v\valoo$;
+\item \TODO unfinished
+\end{enumerate}
+\end{theorem}

### 2014-06-06 paper/splitting.tex
Stylistic changes, pending completion of value proof.
Added a section about PE->SS.  Also a splitting figure.
Added patterns to splitting figures.
changed all figures to have captions at the bottom
Refactor paper into multiple files for cleanliness.
Probably broke build, so will look over Nico's shoulder to fix.

@@
-\begin{figure*}
-\caption{Type Splitting}
-\label{fig:typeSplit}
-\begin{mathpar}
-\infer [\mathrm{unit}\tsplito]	{\rmunit \tsplito [\rmunit,\rmunit]}{\cdot} \and
-\infer [\mathrm{int}\tsplito]	{{\rm int} \tsplito [{\rm int}, \rmunit]}{\cdot} \and
-\infer [\mathrm{bool}\tsplito]	{{\rm bool} \tsplito [{\rm bool}, \rmunit]}{\cdot} \and
-\infer [\times\tsplito]
-	{A\times B \tsplito [A_1 \times B_1, A_2 \times B_2]}
-	{A \tsplito [A_1, A_2] & B \tsplito [B_1,B_2]} \and
-\infer [\fut\tsplito]
-	{\fut~A \tsplito [{\rm unit}, A']}
-	{A \tsplits A'} \and
-\end{mathpar}
-\end{figure*}
+%\begin{figure*}
+%\caption{Type Splitting}
+%\label{fig:typeSplit}
+%\begin{mathpar}
+%\infer [\mathrm{unit}\tsplito]	{\rmunit \tsplito [\rmunit,\rmunit]}{\cdot} \and
+%\infer [\mathrm{int}\tsplito]	{{\rm int} \tsplito [{\rm int}, \rmunit]}{\cdot} \and
+%\infer [\mathrm{bool}\tsplito]	{{\rm bool} \tsplito [{\rm bool}, \rmunit]}{\cdot} \and
+%\infer [\times\tsplito]
+%	{A\times B \tsplito [A_1 \times B_1, A_2 \times B_2]}
+%	{A \tsplito [A_1, A_2] & B \tsplito [B_1,B_2]} \and
+%\infer [\fut\tsplito]
+%	{\fut~A \tsplito [{\rm unit}, A']}
+%	{A \tsplits A'} \and
+%\end{mathpar}
+%\end{figure*}
+%
+%\begin{figure*}
+%\caption{Value Splitting}
+%\label{fig:valSplit}
+%\begin{mathpar}
+%\infer [\rmunit\vsplito]		{() \vsplito [(), ()]}{\cdot} \and
+%\infer [\mathrm{int}\vsplito]	{i \vsplito [i,()]}{\cdot} \and
+%\infer [\mathrm{bool}\vsplito]	{b \vsplito [b,()]}{\cdot} \and
+%\infer [\times\vsplito]	
+%	{\valprod{v_1}{v_2} \vsplito [\valprod{u_1}{u_2}, \valprod{w_1}{w_2}]}
+%	{v_1 \vsplito [u_1,w_1] & v_2 \vsplito [u_2,w_2]} \and
+%\infer [\fut\vsplito]
+%	{\valnext~v \vsplito [(),v']}
+%	{v \vsplits v'} 
+%\end{mathpar}
+%\end{figure*}
+%
+%\begin{figure*}
+%\caption{Context Splitting}
+%\label{fig:contSplit}
+%\begin{mathpar}
+%\infer {\emptyC \csplit [\emptyC~;~\emptyC]}{\cdot} \and
+%\infer {\Gamma, v : A^\bbone \csplit [\Gamma_1,v:A_1; \Gamma_2,v:A_2]}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplito [A_1, A_2]} \and
+%\infer {\Gamma, v : A^\bbtwo \csplit [\Gamma_1; \Gamma_2,v:A']}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplits A'} 
+%\end{mathpar}
+%\end{figure*}
 
 \begin{figure*}
-\caption{Value Splitting}
-\label{fig:valSplit}
-\begin{mathpar}
-\infer [\rmunit\vsplito]		{() \vsplito [(), ()]}{\cdot} \and
-\infer [\mathrm{int}\vsplito]	{i \vsplito [i,()]}{\cdot} \and
-\infer [\mathrm{bool}\vsplito]	{b \vsplito [b,()]}{\cdot} \and
-\infer [\times\vsplito]	
-	{\valprod{v_1}{v_2} \vsplito [\valprod{u_1}{u_2}, \valprod{w_1}{w_2}]}
-	{v_1 \vsplito [u_1,w_1] & v_2 \vsplito [u_2,w_2]} \and
-\infer [\fut\vsplito]
-	{\valnext~v \vsplito [(),v']}
-	{v \vsplits v'} 
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Context Splitting}
-\label{fig:contSplit}
-\begin{mathpar}
-\infer {\emptyC \csplit [\emptyC~;~\emptyC]}{\cdot} \and
-\infer {\Gamma, v : A^\bbone \csplit [\Gamma_1,v:A_1; \Gamma_2,v:A_2]}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplito [A_1, A_2]} \and
-\infer {\Gamma, v : A^\bbtwo \csplit [\Gamma_1; \Gamma_2,v:A']}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplits A'} 
-\end{mathpar}
-\end{figure*}
-
-\begin{figure*}
-\caption{Term Splitting}
-\label{fig:termSplit}
 \begin{mathpar}
@@
 									\left(
-										\letin{y_1}{c_1}{\letin{y_2}{c_2}{
-										((\pi_1 y_1, \pi_1 y_2),(\pi_2 y_1, \pi_2 y_2))}}
+										\letin{(y_1,z_1)}{c_1}{\letin{(y_2,z_2)}{c_2}{
+										((y_1, y_2),(z_1, z_2))}}
 									\right), \\
-									l.(\projbind 1,\projbind 2)
+									(l_1,l_2).(r_1,r_2)
 									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
-\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),l.(\projbind 1,\projbind 2)}}					{\sub 1 A & \sub 2 B} \and
-\infersplitone [\times E_1]	{\spl {\pi_1~e}{A}{\letin{y}{c}{(\pi_1(\pi_1 y),\pi_2 y)},l.\pi_1~r}}					{\sub {} {A\times B}} \and
-\infersplitone [\times E_2]	{\spl {\pi_2~e}{B}{\letin{y}{c}{(\pi_2(\pi_1 y),\pi_2 y)},l.\pi_2~r}}					{\sub {} {A\times B}} \and
-\infersplittwo [\times E_1]	{\spl {\pi_1~e}{A}{p,l.\pi_1~r}}															{\sub {} {A\times B}} \and
-\infersplittwo [\times E_2]	{\spl {\pi_2~e}{B}{p,l.\pi_2~r}}															{\sub {} {A\times B}} \and
-\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{x:A^\bbone \in \Gamma} \and
-\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{x:A^\bbtwo \in \Gamma} \\
+\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),(l_1,l_2).(r_1,r_2)}}								{\sub 1 A & \sub 2 B} \and
+\infersplitone [\times E_1]	{\spl {\pio~e}{A}{\letin{(y,z)}{c}{(\pio~y,z)},l.\pio~r}}								{\sub {} {A\times B}} \and
+\infersplitone [\times E_2]	{\spl {\pit~e}{B}{\letin{(y,z)}{c}{(\pit~y,z)},l.\pit~r}}								{\sub {} {A\times B}} \and
+\infersplittwo [\times E_1]	{\spl {\pio~e}{A}{p,l.\pio~r}}																{\sub {} {A\times B}} \and
+\infersplittwo [\times E_2]	{\spl {\pit~e}{B}{p,l.\pit~r}}																{\sub {} {A\times B}} \and
+\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{\col x A \in \Gamma} \and
+\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{\col x A \in \Gamma} \and
 \infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{
 								\begin{array}{l}
-									\left(\talllet{y_1}{c_1}{
-									\talllet{y_2}{(\letin{x}{\pi_1 y_1}{c_2})}{
-									(\pi_1 y_2, (\pi_2 y_1, \pi_2 y_2))}}\right), \\
-									l.\letin{x}{(\projbind 1)}{\projbind 2}
-								\end{array}}}																					{\sub 1 A & \sub [\Gamma,x:A^\bbone] 2 B} \and
-\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), l.\letin{x}{(\projbind 1)}{\projbind 2}}}	{\sub 1 A & \sub [\Gamma,x:A^\bbtwo] 2 B} \and
+									\left(\talllet{(x,z_1)}{c_1}{
+									\talllet{(y,z_2)}{c_2}{
+									(y, (z_1, z_2))}}\right), \\
+									(l_1,l_2).\letin{x}{r_1}{r_2}
+								\end{array}}}																					{\sub 1 A & \sub [,\col{x}{A}] 2 B} 		\and
+\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), (l_1,l_2).\letin{x}{r_1}{r_2}}}				{\sub 1 A & \sub [,\col{x}{A}] 2 B} 		\and
 \infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} 							\and
-\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pi_2~c,l.r}}															{\splitonesub {} {\fut A}} \and
+\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pit~c,l.r}}															{\splitonesub {} {\fut A}} 					\and
 \infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
 								{\begin{array}{l}
-									\left(\talllet{z}{c_1}{\tallif {\pi_1 z}
-										{\letin{y}{c_2}{(\pi_1 y, (\pi_2 z, \inl(\pi_2 y)))}}
-										{\letin{y}{c_3}{(\pi_1 y, (\pi_2 z, \inr(\pi_2 y)))}}}
+									\left(\talllet{(y_1,z_1)}{c_1}{\tallif {y_1}
+										{\letin{(y_2,z_2)}{c_2}{(y_2, (z_1, \inl~z_2))}}
+										{\letin{(y_3,z_3)}{c_2}{(y_3, (z_1, \inl~z_3))}}}
 									\right), \\
-									l.(\projbind 1;\caseof{\pi_2 l}{l_2.r_2}{l_3.r_3})
+									(l_1,l_b).(r_1;\caseof{l_b}{l_2.r_2}{l_3.r_3})
 								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A }		\and
 \infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
-								{(p_1,p_2,p_3),l.\left(\tallif{(\projbind 1)}{(\projbind 2)}{(\projbind 3)}\right)}}	{\sub 1 \rmbool & \sub 2 A & \sub 3 A } \and
+								{(p_1,p_2,p_3),(l_1,l_2,l_3).\left(\tallif{r_1}{r_2}{r_2}\right)}}						{\sub 1 \rmbool & \sub 2 A & \sub 3 A }		\and
 \end{mathpar}
+\caption{Term Splitting}
+\label{fig:termSplit}
 \end{figure*}
@@
 \begin{figure*}
-\caption{Sum and Function Splitting}
-\label{fig:advancedTypes}
 \begin{mathpar}
 \infersplitone	 [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{\vallam{x}{c},
-							\_.(\vallam{y}{\letin{x}{\pi_1 y}{\letin{l}{\pi_2 y}{r}}})}}		{\sub [\Gamma,x:A^\bbone] {} B} 						\and
-\infersplittwo [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{p,l.(\vallam{x}{M})}}				{\sub [\Gamma,x:A^\bbtwo] {} B} 						\and
+							\_.(\vallam{(x,l)}{r})}}												{\sub [,\col x A] {} B} 									\and
+\infersplittwo [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{p,l.(\vallam{x}{M})}}				{\sub [,\col x A] {} B} 									\and
 \infersplitone [\to E]	{\spl {e_1~e_2}{B}{\ldots}}											{\sub 1 {A \to B} & \sub 2 A}							\and
-\infersplittwo [\to E]	{\spl {e_1~e_2}{B}{(p_1,p_2),l.(\projbind 1)(\projbind 2)}}		{\sub 1 {A \to B} & \sub 2 A}							\and
-\infersplitone [+ I_1]	{\spl {\inl~e}{A+B}{\letin{y}{c}{(\inl(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
-\infersplitone [+ I_2]	{\spl {\inr~e}{A+B}{\letin{y}{c}{(\inr(\pi_1y),\pi_2 y)},l.r}}	{\sub {} A}												\and
+\infersplittwo [\to E]	{\spl {e_1~e_2}{B}{(p_1,p_2),(l_1,l_2).r_1~r_2}}					{\sub 1 {A \to B} & \sub 2 A}							\and
+\infersplitone [+ I_1]	{\spl {\inl~e}{A+B}{\letin{(y,z)}{c}{(\inl~y,z)},l.r}}			{\sub {} A}												\and
+\infersplitone [+ I_2]	{\spl {\inr~e}{A+B}{\letin{(y,z)}{c}{(\inr~y,z)},l.r}}			{\sub {} A}												\and
 \infersplittwo [+ I_1]	{\spl {\inl~e}{A+B}{l.\inl~r}}										{\sub {} A}												\and
@@
 								{\begin{array}{l}\left(
-								\talllet{y_1}{c_1}{
-									\tallcase{\pi_1y_1}
-									{x_2.\letin{y_2}{c_2}{(\pi_1y_2,(\pi_2y_1,\inl(\pi_2y_2)))}}
-									{x_3.\letin{y_3}{c_3}{(\pi_1y_3,(\pi_2y_1,\inr(\pi_2y_3)))}}
+								\talllet{(y_1,z_1)}{c_1}{
+									\tallcase{y_1}
+									{x_2.\letin{(y_2,z_2)}{c_2}{(y_2,(z_1,\inl(z_2)))}}
+									{x_3.\letin{(y_3,z_3)}{c_3}{(y_3,(z_1,\inr(z_3)))}}
 								}\right), \\
-								l.\left(\talllet{z}{(\projbind 1)}{
-									\tallcase{\pi_2 l}
+								(l_1,l_b).\left(\talllet{z}{r_1}{
+									\tallcase{l_b}
 									{l_2.\letin{x_2}{z}{r_2}}
-									{l_3.\letin{x_3}{z}{r_3}}}\right)\end{array}}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbone] 2 C 
-																									& \sub [\Gamma,x_3:A^\bbone] 3 C} 						\and
+									{l_3.\letin{x_3}{z}{r_3}}}\right)\end{array}}}				{\sub 1 {A+B} & \sub [,\col{x_2} A] 2 C 
+																									& \sub [,\col{x_3} B] 3 C} 								\and
 \infersplittwo [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
-								{(p_1,p_2,p_3), l.\left(\tallcase{(\projbind 1)}
-									{x_2.\projbind 2}{x_3.\projbind 3}\right)}}				{\sub 1 {A+B} & \sub [\Gamma,x_2:A^\bbtwo] 2 C 
-																									& \sub [\Gamma,x_3:A^\bbtwo] 3 C} 						\and
+								{(p_1,p_2,p_3), (l_1,l_2,l_3).\left(\tallcase{r_1}
+									{x_2.r_2}{x_3.r_3}\right)}}				{\sub 1 {A+B} & \sub [,\col{x_2} A] 2 C 
+																									& \sub [,\col{x_3} B] 3 C} 								\and
 \end{mathpar}
+\caption{Sum and Function Splitting}
+\label{fig:advancedTypes}
 \end{figure*}
+
+\section{Splitting Algorithm}
+
+[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
+
+The theorem that we want to be true is...
+
+%\begin{center}
+%\begin{tabular}{l}
+%If $\splittwo{e}{A}{p,l.r}$ \\
+%and $\reducetwo{e}{v}$ \\
+%then $\reduce{p}{u}$ \\
+%and $\reduce{[u/l]r}{w}$ \\
+%and $v = w$ 
+%\end{tabular}
+%\end{center}
+
+%\begin{center}
+%\begin{tabular}{l}
+%If $\splitone {e}{A}{c,l.r}$ \\
+%and $\colone {\gamma_1}{\Gamma}$ \\
+%and $\coltwo {\gamma_2}{\Gamma}$\\
+%and $\rsone{\gamma_1(e)}{v}{q}$ \\
+%and $\reducetwo{\gamma_2(q)}{w}$ \\
+%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
+%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
+%\end{tabular}
+%~~~
+%\begin{tabular}{l}
+%If $\splittwo{e}{B}{p,l.r}$ \\
+%and $\colone {\gamma_1}{\Gamma}$ \\
+%and $\coltwo {\gamma_2}{\Gamma}$\\
+%and $\rstwo{\gamma_1(e)}{q}$ \\
+%and $\reducetwo{\gamma_2(q)}{w}$ \\
+%then $\reduce{|\gamma_1|(p)}{u}$ \\
+%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
+%\end{tabular}
+%\end{center}
+
+\section{Examples for Staged Pipelines}
+
+Give the gist of one-to-one pipeline example (like client/server).
+Then talk about a one-to-many pipeline.
+Then talk about a many-to-one pipeline like spark.
+It clear how to target something like this for known base types on the boundary, and for product types.
+But sums on the boundary are hard!
+We leave many-to-one as future-work.
+
+\section{Examples of Algorithm Derivation}
+
+Fast exponent example.
+
+\lstset{language=ML, columns=fullflexible, basicstyle=\ttfamily, tabsize=4, escapeinside={"*}{*"}}
+\begin{lstlisting} 
+let exp (b : $int, e : int) : $int =
+	if e == 0 then
+		next{1}
+	else if (e mod 2) == 0 then
+		next{let x = prev{exp(b,e/2)} in x*x}
+	else
+		next{prev{b} * prev{exp (b,e-1)}}		
+\end{lstlisting}
+
+splits into
+
+\begin{lstlisting} 
+let exp (b, e) =
+	((), roll (
+		if e == 0 then
+			inL ()
+		else 
+			inR (
+				if (e mod 2) == 0 then
+					inL (#2 (exp (b,e/2)))
+				else
+					inR (#2 (exp (b,e-1)))
+			)
+	))
+\end{lstlisting}
+
+and
+
+\begin{lstlisting} 
+let exp ((b, e), p) =
+	case unroll p of
+	  () => 1
+	| d  =>
+		case d of
+		  r => let x = exp ((b,()),r) in x*x
+		| r => b * exp ((b,()),r)
+\end{lstlisting}
+
+Quickselect example.
+
+\begin{lstlisting} 
+let qs (l : "*$\mathtt{\mu \alpha. }$*"() + "*$\mathtt{int*\alpha}$*", i: $int) = 
+	case unroll l of
+	  () => next {0}
+	| (h,t) => 
+		let (left,right,n) = partition h t in
+		next{
+			let n = prev{hold n} in
+			case compare prev{i} n of
+			  () (*LT*) => prev {qs left i}
+			| () (*EQ*) => hold h
+			| () (*GT*) => 
+				prev {qs right next{prev{i}-n-1}}
+		}	
+\end{lstlisting}
+
+Things to try: an interpreter which, partially evaluated, does cps or something.
+
+For each of these examples, talk about what partial evaluation would do and why that might be bad.
+
+\section{Related Work}
+
+Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
+They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
+They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
+The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
+Accordingly, their approach was entirely informal, with no implementation realized.
+Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
+
+Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
+The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
+Their shaders are written in a C-like language with basic arithmetic and if statements.
+Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
+
+Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
+But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
+Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
+Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
+Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
+
+[RTSL and SH]
+
+[Discuss Yong's recent paper here.
+It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
+They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
+If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
+
+Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
+That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
+Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
+The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
+They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
+
+[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
+
+[What's going on with names and necessity?]
+
+[Our work bears a lot of similarity to ML5, which also uses a modal type system.
+The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others).
+The type systems reflect this directly in the world accessibility relation.
+There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]

### 2014-06-07 paper/semantics.tex
fixed the build and made figures files
added new code examples for future explanatory use
finish stating metatheorems about the diagonal semantics (very rough)

@@
-\begin{figure}
-\begin{mathpar}
-\infertypeswor [\rmunit] 	{{\tt unit} \istypewor}								{\cdot} 																			\and
-\infertypeswor [int]			{{\tt int} \istypewor}								{\cdot} 																			\and
-\infertypeswor [bool]			{{\tt bool} \istypewor}								{\cdot} 																			\and
-\infertypeswor [\times]		{A \times B \istypewor}								{A \istypewor & B \istypewor} 													\and
-\infertypeswor [+]			{A + B \istypewor}									{A \istypewor & B \istypewor} 													\and
-\infertypeswor [\to]			{A \to B \istypewor}									{A \istypewor & B \istypewor} 													\and
-\infertypeswor [\fut]			{\fut A \istypeone}									{A \istypetwo} 													
-\end{mathpar}
-\caption{\ellStaged~Valid Types}
-\label{fig:validTypes}
-\end{figure}
-
-\begin{figure}
-\begin{mathpar}
-\infertypeswor [\rmunit] 	{() \isvalwor}											{\cdot} 																			\and
-\infertypeswor [int]			{i \isvalwor}											{\cdot} 																			\and
-\infertypeswor [bool]			{b \isvalwor}											{\cdot} 																			\and
-\infertypeswor [\times]		{(v_1, v_2) \isvalwor}								{v_1 \isvalwor & v_2 \isvalwor} 												\and
-\infertypeswor [+_1]			{\inl~v \isvalwor}									{v \isvalwor} 																		\and
-\infertypeswor [+_2]			{\inr~v \isvalwor}									{v \isvalwor} 																		\and
-\infertypeswor [\to]			{\lam x A e \isvalwor}								{\cdot} 																			\and
-\infertypeswor [\fut]			{\isvalone {\hat y}}									{\cdot} 													
-\end{mathpar}
-\caption{\ellStaged~Valid Values}
-\label{fig:validValues}
-\end{figure}
-
-\begin{figure}
-\begin{mathpar}
-\infertypeswor [\rmunit] 	{\ty {()}\rmunit}								{\cdot} 																			\and
-\infertypeswor [int]			{\ty {i}{\rm int}}							{\cdot} 																			\and
-\infertypeswor [bool]			{\ty {b} \rmbool}								{\cdot} 																			\and
-\infertypeswor [hyp]			{\ty x A}										{\col x A \in \Gamma} 															\and
-\infertypeswor [\to I]		{\ty {\lam{x}{A}{e}} {A \to B}}				{A \istypewor & \ty [,\col x A] e B} 											\and
-\infertypeswor [\to E]		{\ty {e_1~e_2} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 											\and
-\infertypeswor [let]			{\ty {\letin{x}{e_1}{e_2}}{B}}				{\ty {e_1} A & \ty [,\col x A] {e_2} B} 										\and
-\infertypeswor [\times I]	{\ty {(e_1,e_2)}{A\times B}}					{\ty {e_1} A & \ty {e_2} B} 														\and
-\infertypeswor [\times E_1]	{\ty {\pio~e} A}								{\ty e {A\times B}} 																\and
-\infertypeswor [\times E_2]	{\ty {\pit~e} B}								{\ty e {A\times B}} 																\and
-\infertypeswor [+ I_1]		{\ty {\inl~e} {A + B}}						{\ty e A} 																			\and
-\infertypeswor [+ I_2]		{\ty {\inr~e} {A + B}}						{\ty e B} 																			\and
-\infertypeswor [+ E]			{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}	{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C} 	\and
-\infertypeswor [if] 			{\ty {\ifthen{e_1}{e_2}{e_3}} A}			{\ty {e_1} \rmbool & \ty {e_2} A & \ty {e_3} A} 								\and
-\infertypeswor [\fut I]		{\typesone {\next~e}{\fut A}}				{\typestwo e A} 																	\and
-\infertypeswor [\fut E]		{\typestwo {\prev~e} A}						{\typesone e {\fut A}} 															\and
-\end{mathpar}
-\caption{\ellStaged~Static Semantics}
-\label{fig:coreStatics}
-\end{figure}
-
-\begin{figure}
-\begin{mathpar}
-\inferdiaone [\to I]  		{\red {\lam{x}{A}{e}} {\cdot,\vallam{x}{e}}}					{\cdot} 																					\and
-\inferdiaone [\to E]  		{\red {e_1~e_2} {\gcomp 1 2 \circ \xi',v'}}					{\red {e_1} {\xi_1,\vallam{x}{e'}} & \sub [2] & \red [\Xi,dom(\xi_1),dom(\xi_2);\Gamma] {[v_1/x]e'}{\xi',v'}}	\and
-%
-\inferdiaone [\rmunit]   	{\red {()}{\cdot,()}}												{\cdot} 																					\and
-\inferdiaone [\times I]  	{\red {(e_1,e_2)}{\gcomp 1 2,\valprod{v_1}{v_2}}}				{\sub [1] & \sub [2]} 																	\and
-\inferdiaone [\times E_1]	{\red {\pio~e}{\xi,v_1}}											{\red{e}{\xi,\valprod{v_1}{v_2}}} 													\and
-\inferdiaone [\times E_2]	{\red {\pit~e}{\xi,v_2}}											{\red{e}{\xi,\valprod{v_1}{v_2}}} 													\and
-\inferdiaone [+ I_1]  		{\red {\inl~e} {\xi,\inl~v}}										{\sub} 																						\and
-\inferdiaone [+ I_2]  		{\red {\inr~e} {\xi,\inr~v}}										{\sub} 																						\and
-%
-\inferdiaone [+ E_1]  		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{\gcomp 1 2,v_2}}		{\red {e_1}{\xi_1,\valinl~v} & \red [\Xi,dom(\xi_1);\Gamma]{[v/x_2]e_2}{\xi_2,v_2}} 					\and
-\inferdiaone [+ E_2]  		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{\gcomp 1 2,v_3}}		{\red {e_1}{\xi_1,\valinr~v} & \red [\Xi,dom(\xi_1);\Gamma]{[v/x_3]e_3}{\xi_3,v_3}}					\and
-%
-\inferdiaone [int]			{\red {i}{\cdot,i}}												{\cdot} 																					\and
-\inferdiaone [bool]   		{\red {b}{\cdot,b}}												{\cdot} 																					\and
-\inferdiaone [let]			{\red {\letin{x}{e_1}{e_2}}{\gcomp 1 2, v_2}}					{\sub [1] & \red [\Xi,dom(\xi_1);\Gamma] {[v_1/x]e_2}{\xi_2,v_2}}											\and
-\inferdiaone [if_T]   		{\red {\ifthen{e_1}{e_2}{e_3}}{\gcomp 1 2,v}}					{\red {e_1}{\xi_1,\bf true}  & \red {e_2}{\xi_2,v}} 							\and
-\inferdiaone [if_F]   		{\red {\ifthen{e_1}{e_2}{e_3}}{\gcomp 1 3,v}}					{\red {e_1}{\xi_1,\bf false} & \red {e_3}{\xi_3,v}}								
-\end{mathpar}
-\caption{\ellStaged~Diagonal Semantics (core)}
-\label{fig:diaSemanticsCore}
-\end{figure}
-
-\begin{figure}
-\begin{mathpar}
-\inferdiaone [\fut I]			{\red {\next~e}{y \mapsto q,\hat y}}				{\diatwo e q}												\and
-\inferdiaspc [\fut E]			{\red {\prev~e} {\xi (y)}}						{\diaone e {\xi, \hat y}} 									
-\end{mathpar}
-\caption{\ellStaged~Diagonal Semantics ($\next$ and $\prev$)}
-\label{fig:diaSemanticsNP}
-\end{figure}
-
-\begin{figure}
-\begin{mathpar}
-\inferdiaspc [\to I]			{\red {\lam{x}{A}{e}}{\lam{x}{A}{q}}}						{\diatwo [\Xi;\Gamma,x] e q} 			\and
-\inferdiaspc [\to E]			{\red {e_1~e_2}{q_1,q_2}}									{\sub [1] & \sub [2]}						\and		
-\inferdiaspc [\rmunit]		{\red {()}{()}}												{\cdot} 									\and
-\inferdiaspc [\times I]		{\red {(e_1,e_2)}{(q_1,q_2)}}								{\sub [1] & \sub [2]} 					\and
-\inferdiaspc [\times E_1]	{\red {\pio~e}{\pio~q}}										{\sub} 										\and
-\inferdiaspc [\times E_2]	{\red {\pit~e}{\pit~q}}										{\sub} 										\and
-\inferdiaspc [+ I_1]			{\red {\inl~e}{\inl~q}}										{\sub} 										\and
-\inferdiaspc [+ I_2]			{\red {\inr~e}{\inr~q}}										{\sub} 										\and
-\inferdiaspc [+ E_1]			{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
-									{\caseof{q_1}{x_2.q_2}{x_3.q_3}}}						{\sub [1] & \diatwo [\Xi;\Gamma,x_2] {e_2} {q_2} & \diatwo [\Xi;\Gamma,x_3] {e_3} {q_3}} 		\and
-\inferdiaspc [int]			{\red {i}{i}}													{\cdot} 									\and
-\inferdiaspc [bool]			{\red {b}{b}}													{\cdot} 									\and
-\inferdiaspc [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{q_1}{q_2}}}			{\sub [1] & \diatwo [\Xi;\Gamma,x] {e_2} {q_2}} 					\and
-\inferdiaspc [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{q_1}{q_2}{q_3}}}	{\sub [1] & \sub [2] & \sub [3]} 			
-\end{mathpar}
-\caption{\ellStaged~New Speculation}
-\label{fig:diaSemanticsSpec}
-\end{figure}
-%
-%\begin{figure}
-%\begin{mathpar}
-%\inferreducewor [\to I]			{\redone {\lam{x}{A}{e}} {\vallam{x}{e}}}		{\cdot} 																		\and
-%\inferreducewor [\to I]			{\redtwo {\lam{x}{A}{e}} {\vallam{x}{e}}}		{\specwor e} 																	\and
-%\inferreducewor [\to E]			{\red {e_1~e_2} {v'}}								{\red {e_1} {\vallam{x}{M}} & \red{e_2}v & \red {[v/x]M} {v'}}			\and
-%%
-%\inferreducewor [\rmunit]		{\red {()}{()}}									{\cdot} 																		\and
-%\inferreducewor [\times I]		{\red {(e_1,e_2)}{\valprod{v_1}{v_2}}}			{\sub [1] & \sub [2]} 														\and
-%\inferreducewor [\times E_1]		{\red {\pio~e}{v_1}}								{\red{e}{\valprod{v_1}{v_2}}} 												\and
-%\inferreducewor [\times E_2]		{\red {\pit~e}{v_2}}								{\red{e}{\valprod{v_1}{v_2}}} 												\and
-%\inferreducewor [+ I_1]			{\red {\inl~e} {\inl~v}}							{\sub} 																			\and
-%\inferreducewor [+ I_2]			{\red {\inr~e} {\inr~v}}							{\sub} 																			\and
-%%
-%\inferreducewor [+ E_1]			{\redone {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redone {e_1}{\valinl~v} & \redone {[v/x_2]e_2}{v'}} 					\and
-%\inferreducewor [+ E_2]			{\redone {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redone {e_1}{\valinr~v} & \redone {[v/x_3]e_3}{v'}}						\and
-%\inferreducewor [+ E_1]			{\redtwo {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redtwo {e_1}{\valinl~v} & \redtwo {[v/x_2]e_2}{v'} & \specwor {e_3}}	\and
-%\inferreducewor [+ E_2]			{\redtwo {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}}		{\redtwo {e_1}{\valinr~v} & \redtwo {[v/x_3]e_3}{v'} & \specwor {e_2}}	\and
-%%
-%\inferreducewor [int]				{\red {i}{i}}											{\cdot} 																		\and
-%\inferreducewor [bool]			{\red {b}{b}}											{\cdot} 																		\and
-%\inferreducewor [let]				{\red {\letin{x}{e_1}{e_2}}{v_2}}					{\sub [1] & \red{[v_1/x]e_2}{v_2}} 											\and
-%\inferreducewor [if_T] 			{\redone {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redone {e_1}{\bf true}  & \redone {e_2}{v}} 								\and
-%\inferreducewor [if_F] 			{\redone {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redone {e_1}{\bf false} & \redone {e_3}{v}}								\and
-%\inferreducewor [if_T] 			{\redtwo {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redtwo {e_1}{\bf true}  & \redtwo {e_2}{v} & \specwor {e_3}}			\and
-%\inferreducewor [if_F] 			{\redtwo {\ifthen{e_1}{e_2}{e_3}}{v}}				{\redtwo {e_1}{\bf false} & \redtwo {e_3}{v} & \specwor {e_2}}
-%\end{mathpar}
-%\caption{\ellStaged~Old Semantics (core)}
-%\label{fig:erasureSemanticsCore}
-%\end{figure}
-%
-%\begin{figure}
-%\begin{mathpar}
-%\inferreducewor [\fut I]			{\reducexpl \bbone \bbtwo {\next~e}{\next~v}}	{\reducexpl \bbtwo \bbtwo e v}												\and
-%\inferreducewor [\fut I]			{\reducexpl \bbone \bbone {\next~e}{DUM}}		{\specwor e }																	\and
-%\inferreducewor [\fut E]			{\reducexpl \bbtwo \bbtwo {\prev~e} v}			{\reducexpl \bbone \bbtwo e {\valnext~v}} 									\and
-%\inferreducespc [\fut E]			{\red {\prev~e}}									{\reducexpl \bbone \bbone e {DUM}} 								
-%\end{mathpar}
-%\caption{\ellStaged~Old Semantics ($\next$ and $\prev$)}
-%\label{fig:erasureSemanticsNextPrev}
-%\end{figure}	
-%
-%\begin{figure}
-%\begin{mathpar}
-%\inferreducespc [\to I]			{\red {\lam{x}{A}{e}}}							{\sub} 										\and
-%\inferreducespc [\to E]			{\red {e_1~e_2}}									{\sub [1] & \sub [2]}						\and		
-%\inferreducespc [\rmunit]		{\red {()}}										{\cdot} 									\and
-%\inferreducespc [\times I]		{\red {(e_1,e_2)}}								{\sub [1] & \sub [2]} 					\and
-%\inferreducespc [\times E_1]		{\red {\pio~e}}									{\sub [1]} 								\and
-%\inferreducespc [\times E_2]		{\red {\pit~e}}									{\sub [1]} 								\and
-%\inferreducespc [+ I_1]			{\red {\inl~e}}									{\sub} 										\and
-%\inferreducespc [+ I_2]			{\red {\inr~e}}									{\sub} 										\and
-%\inferreducespc [+ E_1]			{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}}			{\sub [1] & \sub [2] & \sub [3]} 		\and
-%\inferreducespc [int]				{\red {i}}											{\cdot} 									\and
-%\inferreducespc [bool]			{\red {b}}											{\cdot} 									\and
-%\inferreducespc [let]				{\red {\letin{x}{e_1}{e_2}}}						{\sub [1] & \sub [2]} 					\and
-%\inferreducespc [if_T] 			{\red {\ifthen{e_1}{e_2}{e_3}}}					{\sub [1] & \sub [2] & \sub [3]} 			
-%\end{mathpar}
-%\caption{\ellStaged~Old Speculation}
-%\label{fig:erasureSemanticsSpec}
-%\end{figure}
-%
-%\begin{figure}
-%\caption{\ellStaged~Diagonal Semantics}
-%\label{fig:refSemantics}
-%\begin{mathpar}
-%\inferRedsplitone [\rmunit]		{\red {()}{()}{()}}												{\cdot} 												\and
-%\inferRedsplitone [int]			{\red {i}{i}{()}}													{\cdot} 												\and
-%\inferRedsplitone [bool]			{\red {b}{b}{()}}													{\cdot} 												\and
-%\inferRedsplitone [hyp]			{\red {[u,x]}{u}{x}}												{\cdot}													\and
-%\inferSpeculation [\rmunit]		{\red {()}{()}}													{\cdot} 												\and
-%\inferSpeculation [int]			{\red {i}{i}}														{\cdot} 												\and
-%\inferSpeculation [bool]			{\red {b}{b}}														{\cdot} 												\and
-%\inferSpeculation [hyp]			{\red {x}{x}}														{\cdot}													\and
-%\inferRedsplitone [let]			{\red {\letin{x}{e_1}{e_2}}{u_2}{\letin {x}{r_1}{r_2}}}		{\sub [1] & \red{[[u_1,x]/x]e_2}{u_2}{r_2}} 		\and
-%\inferSpeculation [let]			{\red {\letin{x}{e_1}{e_2}}{\letin{x}{r_1}{r_2}}}				{\sub [1] & \sub [2]}									\and
-%\inferRedsplitone [\times I]		{\red {(e_1,e_2)}{\valprod{u_1}{u_2}}{\valprod{r_1}{r_2}}}	{\sub [1] & \sub [2]}									\and
-%\inferRedsplitone [\times E_1]	{\red {\pi_1~e}{u_1}{\pi_1 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}}					\and
-%\inferRedsplitone [\times E_2]	{\red {\pi_2~e}{u_2}{\pi_2 r}}									{\red {e}{\valprod{u_1}{u_2}}{r}} 					\and
-%\inferSpeculation [\times I]		{\red {(e_1,e_2)}{(r_1,r_2)}}									{\sub [1] & \sub [2]}									\and
-%\inferSpeculation [\times E_1]	{\red {\pi_1~e}{\pi_1 r}}										{\red {e}{r}}											\and
-%\inferSpeculation [\times E_2]	{\red {\pi_2~e}{\pi_2 r}}										{\red {e}{r}} 											\and
-%\inferRedsplitone [if_T] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_2)}}					{\red{e_1}{\bf true }{r_1} & \red{e_2}{u}{r_2}}	\and
-%\inferRedsplitone [if_F] 		{\red {\ifthen{e_1}{e_2}{e_3}}{u}{(r_1;r_3)}}					{\red{e_1}{\bf false}{r_1} & \red{e_3}{u}{r_3}}	\and
-%\inferSpeculation [if] 			{\red {\ifthen{e_1}{e_2}{e_3}}{\ifthen{r_1}{r_2}{r_3}}}		{\sub [1] & \sub [2] & \sub [3]}					\\
-%\inferRedsplitone [\fut I]		{\red {\next~e}{()}{r}}											{\rstwosub}											\and
-%\inferSpeculation [\fut E]		{\red {\prev~e}{r}}												{\rsone {e}{()}{r}}
-%\end{mathpar}
-%\end{figure}
-%
-%\begin{figure*}
-%\caption{Sum and Function Diagonal Evaluation}
-%\label{fig:advancedTypes}
-%\begin{mathpar}
-%\inferRedsplitone [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{e}}{()}}			{\cdot} 																\and
-%\inferSpeculation [\to I]	{\red {\lam{x}{A}{e}}{\vallam{x}{r}}}				{\sub} 																	\and
-%\inferRedsplitone [\to E]	{\red {e_1~e_2}{v'}{(r_1;\letin{x}{r_2}{r'})}}		{\red {e_1}{\vallam{x}{M}}{r_1} 
-%																							& \sub [2] & \red{[[e,x]/x]M}{v'}{r'}} 							\and
-%\inferSpeculation [\to E]	{\red {e_1~e_2}{r_1~r_2}}							{\sub [1] & \sub[2]}													\and
-%\inferRedsplitone [+ I_1]	{\red {\inl~e} {\valinl~v}{r}}						{\sub} 																	\and
-%\inferSpeculation [+ I_1]	{\red {\inl~e} {\valinl~r}}							{\sub} 																	\\
-%\inferRedsplitone [+ I_2]	{\red {\inr~e} {\valinr~v}{r}}						{\sub} 																	\and
-%\inferSpeculation [+ I_2]	{\red {\inr~e} {\valinr~r}}							{\sub} 																	\\
-%\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
-%								{\letin{x_2}{r_1}{r_2}}}								{\red {e_1}{\valinl~v}{r_1} & \red {[[v,x_2]/x_2]e_2}{v'}{r_2}} \and
-%\inferRedsplitone [+ E_1]	{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}{v'}
-%								{\letin{x_3}{r_1}{r_3}}}								{\red {e_1}{\valinr~v}{r_1} & \red {[[v,x_3]/x_3]e_3}{v'}{r_3}} \and
-%\inferSpeculation [+ E]		{\red {\caseof{e_1}{x_2.e_2}{x_3.e_3}}
-%								{\caseof{r_1}{x_2.r_2}{x_3.r_3}}}				{\sub [1] & \sub [2] & \sub[3]}					 						\and
-%\end{mathpar}
-%\end{figure*}
-
 \section{Semantics}
@@
 
-...
+\begin{lstlisting} 
+next {
+	if 5*4*3*2 > 111 then
+		prev{hold (2+4)}
+	else
+		prev{hold (4*5)}
+}
+\end{lstlisting}
 
@@
 This makes reasoning about second-stage cost rather difficult.
+
+\begin{lstlisting} 
+let x = next {4+5} in
+next{prev{x} * prev{x}}
+\end{lstlisting}
 
@@
 
-\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2
+\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2. what is the
+terminology? (?, residuals, results?)
 
@@
 
+\noindent\TODO be uniform about which judgments have which contexts, etc
+
 % local macros; put them in macros.tex once they are more finalized
 \newcommand{\wf}{\ \mathsf{wf}}
-\newcommand{\dom}[1]{\mathsf{dom}(#1)}
 \newcommand{\valoo}{\ \bbone\textsf{-}\mathsf{val} @ \bbone}
@@
 
+\noindent\TODO don't call it a table
+
+\noindent\TODO what is the notation for $\xi$s? how do you build them?
+
+\noindent\TODO define reification explicitly as a metafunction
+
+\begin{definition}
+A table $\xi$ is well-formed ($\xi\wf$) if either:
+\begin{enumerate}
+\item $\xi = \cdot$; or
+\item $\xi = \xi',x\mapsto e$ where
+$\Xi,\dom{\xi'};\Gamma\vdash e : B @ \bbtwo$ and
+$\Xi,\dom{\xi'};\Gamma\vdash e \valot$.
+\end{enumerate}
+\end{definition}
+
 \begin{theorem}
@@
 \begin{theorem}
-If $\Xi;\Gamma\vdash\diaonesub$ then:
+If $\diaonesub$ then:
 \begin{enumerate}
 \item $\typesone e A$ for some 1-type $A$;
-\item $v\valoo$;
-\item \TODO unfinished
+\item $\xi\wf$;
+\item $\Xi,\dom\xi;\cdot\vdash v : A @ \bbone$; and
+\item $\Xi,\dom\xi;\cdot\vdash v\valoo$.
 \end{enumerate}
 \end{theorem}
+
+\begin{theorem}
+If $\diatwosub$ then:
+\begin{enumerate}
+\item $\typestwo e A$ for some 2-type $A$;
+\item $\Xi;\Gamma\vdash q : A @ \bbtwo$; and
+\item $\Xi;\Gamma\vdash q\valtt$.
+\end{enumerate}
+\end{theorem}

### 2014-06-08 paper/splitting.tex
fixed the build and made figures files
added new code examples for future explanatory use
added missing rules
fixed example (no pdf commit)

@@
-%\begin{figure*}
-%\caption{Type Splitting}
-%\label{fig:typeSplit}
-%\begin{mathpar}
-%\infer [\mathrm{unit}\tsplito]	{\rmunit \tsplito [\rmunit,\rmunit]}{\cdot} \and
-%\infer [\mathrm{int}\tsplito]	{{\rm int} \tsplito [{\rm int}, \rmunit]}{\cdot} \and
-%\infer [\mathrm{bool}\tsplito]	{{\rm bool} \tsplito [{\rm bool}, \rmunit]}{\cdot} \and
-%\infer [\times\tsplito]
-%	{A\times B \tsplito [A_1 \times B_1, A_2 \times B_2]}
-%	{A \tsplito [A_1, A_2] & B \tsplito [B_1,B_2]} \and
-%\infer [\fut\tsplito]
-%	{\fut~A \tsplito [{\rm unit}, A']}
-%	{A \tsplits A'} \and
-%\end{mathpar}
-%\end{figure*}
-%
-%\begin{figure*}
-%\caption{Value Splitting}
-%\label{fig:valSplit}
-%\begin{mathpar}
-%\infer [\rmunit\vsplito]		{() \vsplito [(), ()]}{\cdot} \and
-%\infer [\mathrm{int}\vsplito]	{i \vsplito [i,()]}{\cdot} \and
-%\infer [\mathrm{bool}\vsplito]	{b \vsplito [b,()]}{\cdot} \and
-%\infer [\times\vsplito]	
-%	{\valprod{v_1}{v_2} \vsplito [\valprod{u_1}{u_2}, \valprod{w_1}{w_2}]}
-%	{v_1 \vsplito [u_1,w_1] & v_2 \vsplito [u_2,w_2]} \and
-%\infer [\fut\vsplito]
-%	{\valnext~v \vsplito [(),v']}
-%	{v \vsplits v'} 
-%\end{mathpar}
-%\end{figure*}
-%
-%\begin{figure*}
-%\caption{Context Splitting}
-%\label{fig:contSplit}
-%\begin{mathpar}
-%\infer {\emptyC \csplit [\emptyC~;~\emptyC]}{\cdot} \and
-%\infer {\Gamma, v : A^\bbone \csplit [\Gamma_1,v:A_1; \Gamma_2,v:A_2]}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplito [A_1, A_2]} \and
-%\infer {\Gamma, v : A^\bbtwo \csplit [\Gamma_1; \Gamma_2,v:A']}{\Gamma \csplit [\Gamma_1; \Gamma_2] & A \tsplits A'} 
-%\end{mathpar}
-%\end{figure*}
-
-\begin{figure*}
-\begin{mathpar}
-\infersplitone [unit]			{\spl {()}{\rmunit}{((),()),\_.()}}															{\cdot} \and
-\infersplittwo [unit]			{\spl {()}{\rmunit}{(),\_.()}}																{\cdot} \and
-\infersplitone [int]			{\spl {i}{\rm int}{(i,()),\_.()}}															{\cdot} \and
-\infersplittwo [int]			{\spl {i}{\rm int}{(),\_.i}}																	{\cdot} \and
-\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
-									{ \begin{array}{l}
-									\left(
-										\letin{(y_1,z_1)}{c_1}{\letin{(y_2,z_2)}{c_2}{
-										((y_1, y_2),(z_1, z_2))}}
-									\right), \\
-									(l_1,l_2).(r_1,r_2)
-									\end{array}}}																				{\sub 1 A & \sub 2 B} \and
-\infersplittwo [\times I]	{\spl {(e_1,e_2)}{A\times B}{(p_1,p_2),(l_1,l_2).(r_1,r_2)}}								{\sub 1 A & \sub 2 B} \and
-\infersplitone [\times E_1]	{\spl {\pio~e}{A}{\letin{(y,z)}{c}{(\pio~y,z)},l.\pio~r}}								{\sub {} {A\times B}} \and
-\infersplitone [\times E_2]	{\spl {\pit~e}{B}{\letin{(y,z)}{c}{(\pit~y,z)},l.\pit~r}}								{\sub {} {A\times B}} \and
-\infersplittwo [\times E_1]	{\spl {\pio~e}{A}{p,l.\pio~r}}																{\sub {} {A\times B}} \and
-\infersplittwo [\times E_2]	{\spl {\pit~e}{B}{p,l.\pit~r}}																{\sub {} {A\times B}} \and
-\infersplitone [hyp]			{\spl {v}{A}{(x,()),\_.()}}																	{\col x A \in \Gamma} \and
-\infersplittwo [hyp]			{\spl {v}{A}{(),\_.x}}																		{\col x A \in \Gamma} \and
-\infersplitone [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{
-								\begin{array}{l}
-									\left(\talllet{(x,z_1)}{c_1}{
-									\talllet{(y,z_2)}{c_2}{
-									(y, (z_1, z_2))}}\right), \\
-									(l_1,l_2).\letin{x}{r_1}{r_2}
-								\end{array}}}																					{\sub 1 A & \sub [,\col{x}{A}] 2 B} 		\and
-\infersplittwo [let]			{\spl {\letin{x}{e_1}{e_2}}{B}{(p_1,p_2), (l_1,l_2).\letin{x}{r_1}{r_2}}}				{\sub 1 A & \sub [,\col{x}{A}] 2 B} 		\and
-\infersplitone [\fut I]		{\spl {\next~e}{\fut A}{((),p),l.r}}														{\splittwosub {} A} 							\and
-\infersplittwo [\fut E]		{\spl {\prev~e}{A} {\pit~c,l.r}}															{\splitonesub {} {\fut A}} 					\and
-\infersplitone [if]			{\spl {\left( \tallif {e_1}{e_2}{e_3} \right)}{A} 
-								{\begin{array}{l}
-									\left(\talllet{(y_1,z_1)}{c_1}{\tallif {y_1}
-										{\letin{(y_2,z_2)}{c_2}{(y_2, (z_1, \inl~z_2))}}
-										{\letin{(y_3,z_3)}{c_2}{(y_3, (z_1, \inl~z_3))}}}
-									\right), \\
-									(l_1,l_b).(r_1;\caseof{l_b}{l_2.r_2}{l_3.r_3})
-								\end{array}}}																					{\sub 1 \rmbool & \sub 2 A & \sub 3 A }		\and
-\infersplittwo [if]			{\spl {\left(\tallif {e_1}{e_2}{e_3}\right)}{A} 
-								{(p_1,p_2,p_3),(l_1,l_2,l_3).\left(\tallif{r_1}{r_2}{r_2}\right)}}						{\sub 1 \rmbool & \sub 2 A & \sub 3 A }		\and
-\end{mathpar}
-\caption{Term Splitting}
-\label{fig:termSplit}
-\end{figure*}
-
-\begin{figure*}
-\begin{mathpar}
-\infersplitone	 [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{\vallam{x}{c},
-							\_.(\vallam{(x,l)}{r})}}												{\sub [,\col x A] {} B} 									\and
-\infersplittwo [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{p,l.(\vallam{x}{M})}}				{\sub [,\col x A] {} B} 									\and
-\infersplitone [\to E]	{\spl {e_1~e_2}{B}{\ldots}}											{\sub 1 {A \to B} & \sub 2 A}							\and
-\infersplittwo [\to E]	{\spl {e_1~e_2}{B}{(p_1,p_2),(l_1,l_2).r_1~r_2}}					{\sub 1 {A \to B} & \sub 2 A}							\and
-\infersplitone [+ I_1]	{\spl {\inl~e}{A+B}{\letin{(y,z)}{c}{(\inl~y,z)},l.r}}			{\sub {} A}												\and
-\infersplitone [+ I_2]	{\spl {\inr~e}{A+B}{\letin{(y,z)}{c}{(\inr~y,z)},l.r}}			{\sub {} A}												\and
-\infersplittwo [+ I_1]	{\spl {\inl~e}{A+B}{l.\inl~r}}										{\sub {} A}												\and
-\infersplittwo [+ I_1]	{\spl {\inr~e}{A+B}{l.\inr~r}}										{\sub {} A}												\and
-\infersplitone [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
-								{\begin{array}{l}\left(
-								\talllet{(y_1,z_1)}{c_1}{
-									\tallcase{y_1}
-									{x_2.\letin{(y_2,z_2)}{c_2}{(y_2,(z_1,\inl(z_2)))}}
-									{x_3.\letin{(y_3,z_3)}{c_3}{(y_3,(z_1,\inr(z_3)))}}
-								}\right), \\
-								(l_1,l_b).\left(\talllet{z}{r_1}{
-									\tallcase{l_b}
-									{l_2.\letin{x_2}{z}{r_2}}
-									{l_3.\letin{x_3}{z}{r_3}}}\right)\end{array}}}				{\sub 1 {A+B} & \sub [,\col{x_2} A] 2 C 
-																									& \sub [,\col{x_3} B] 3 C} 								\and
-\infersplittwo [+ E]		{\spl {\left(\tallcase{e_1}{x_2.e_2}{x_3.e_3}\right)}{C}
-								{(p_1,p_2,p_3), (l_1,l_2,l_3).\left(\tallcase{r_1}
-									{x_2.r_2}{x_3.r_3}\right)}}				{\sub 1 {A+B} & \sub [,\col{x_2} A] 2 C 
-																									& \sub [,\col{x_3} B] 3 C} 								\and
-\end{mathpar}
-\caption{Sum and Function Splitting}
-\label{fig:advancedTypes}
-\end{figure*}
-
 \section{Splitting Algorithm}
@@
 [Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
-
-The theorem that we want to be true is...
-
-%\begin{center}
-%\begin{tabular}{l}
-%If $\splittwo{e}{A}{p,l.r}$ \\
-%and $\reducetwo{e}{v}$ \\
-%then $\reduce{p}{u}$ \\
-%and $\reduce{[u/l]r}{w}$ \\
-%and $v = w$ 
-%\end{tabular}
-%\end{center}
-
-%\begin{center}
-%\begin{tabular}{l}
-%If $\splitone {e}{A}{c,l.r}$ \\
-%and $\colone {\gamma_1}{\Gamma}$ \\
-%and $\coltwo {\gamma_2}{\Gamma}$\\
-%and $\rsone{\gamma_1(e)}{v}{q}$ \\
-%and $\reducetwo{\gamma_2(q)}{w}$ \\
-%then $\reduce{|\gamma_1|(c)}{\valprod{v}{u}}$ \\
-%and $\reduce{[u/l]|\gamma_2|(r)}{w}$
-%\end{tabular}
-%~~~
-%\begin{tabular}{l}
-%If $\splittwo{e}{B}{p,l.r}$ \\
-%and $\colone {\gamma_1}{\Gamma}$ \\
-%and $\coltwo {\gamma_2}{\Gamma}$\\
-%and $\rstwo{\gamma_1(e)}{q}$ \\
-%and $\reducetwo{\gamma_2(q)}{w}$ \\
-%then $\reduce{|\gamma_1|(p)}{u}$ \\
-%and $\reduce{[u/l]|\gamma_2|(r)}{w}$ 
-%\end{tabular}
-%\end{center}
 
@@
 
-\lstset{language=ML, columns=fullflexible, basicstyle=\ttfamily, tabsize=4, escapeinside={"*}{*"}}
 \begin{lstlisting} 
@@
 			  () (*LT*) => prev {qs left i}
-			| () (*EQ*) => hold h
+			| () (*EQ*) => prev {hold h}
 			| () (*GT*) => 

### 2014-06-11 paper/intro.tex
Added a few notes.

@@
 
-Staged computation has existed as a programming technique for over three decades.
+We often find ourselves in situations where the input to a computation arrives in two parts, at different points in time.
+
+The natural response is to try and do some part of the computation early, and the rest of the computation once the second input arrives.
+
+Since the computation runs in two parts, the total cost can be split as well, into $m+n$.
+
+Additionally, we can run the second stage multiple times.  Cost is now $m+bn$.
+
 Jorring et al.
@@
 The first two of these have received significantly more attention than the third.
-
-We call these techniques {\em staged} in that part of their input comes in the first stage, and part comes in the second stage.  Likewise, some computation occurs at each stage as well.
-
 Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).

### 2014-06-11 paper/intro.tex
Reworked file structure a little.

@@
 This paper explores both theory and applications for stage-splitting.
+
+\section{Stage-Splitting Definition and Comparison to Partial Evaluation}
+
+First, we review the definition of partial evaluation.
+Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.
+This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.
+The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.
+We now state this theorem more formally: a partial evaluator is some function $p$ such that,
+\[
+	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
+\]
+where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
+
+Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.
+Again, more formally, a stage-splitter is some $s$ such that,
+\[
+	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
+	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
+\]
+
+We first discuss a few similarities between partial-evaluation and stage-splitting.
+First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.
+
+Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.
+Indeed, both admit completely trivial definitions.
+Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available.
+The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.
+Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.
+
+The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.
+For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.
+Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.
+Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.
+Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.
+
+\subsection{Partial Evaluator from Stage-Splitter}
+
+We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
+Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
+In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
+It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer.
+[Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
+
+\subsection{Stage-Splitter from Partial Evaluator}
+
+Likewise, we can easily recover a stage-splitter from a partial evaluator.  If $p$ is a valid partial evaluator, then we can define a stage splitter $s$ such that $s(f)=(f_1,f_2)$, where
+\begin{align*}
+[f_1](x) &= p (f,x) \\
+[f_2](l,y) &= [l] (y)
+\end{align*}
+This implicitly requires that the languages in which $f_1$ and $f_2$ are expressed are strong enough to write a partial evaluator, but that is the case in this paper.
+A stage-splitter defined this way leaves much to be desired.
+Firstly, partial evaluation of $f$ may be too expensive for the context in which $f_1$ needs to run.
+Additionally, the intermediate data structure created this way may be much larger than necessary, as it would contain all of the residual code.

### 2014-06-11 paper/relatedwork.tex
Reworked file structure a little.

@@
-\section{Stage-Splitting Definition and Comparison to Partial Evaluation}
+\section{Related Work}
 
-First, we review the definition of partial evaluation.
-Informally, a partial evaluator takes the code for a function $f$, as well as the first-stage input $x$ to that function, and produces the code for a version of that function {\em specialized} to the first input, often called $f_x$.
-This $f_x$ function can then be evaluated with the second-stage input to produce the same final answer that $f$ would have.
-The goal of the process is that $f_x$ should be cheaper to evaluate than $f$, although this can't be guaranteed for all inputs.
-We now state this theorem more formally: a partial evaluator is some function $p$ such that,
-\[
-	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
-\]
-where (borrowing notation from \cite{jones96}), $\llbracket f \rrbracket$ means the mathematical function corresponding to the code given by $f$.
+Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
+They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
+They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
+The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
+Accordingly, their approach was entirely informal, with no implementation realized.
+Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
 
-Informally, we define stage-splitting to be the process of taking some function $f$ into two other functions, $f_1$ and $f_2$, where $f_1$ computes a partial result from the first-stage input, and $f_2$ uses that partial result and the second-stage input to compute a final result which is the same as if we had just run the original $f$ on both inputs.
-Again, more formally, a stage-splitter is some $s$ such that,
-\[
-	\forall f. \exists f_1,f_2. [s(f) = (f_1,f_2) \text{ and } 
-	\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)]
-\]
+Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
+The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
+Their shaders are written in a C-like language with basic arithmetic and if statements.
+Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
 
-We first discuss a few similarities between partial-evaluation and stage-splitting.
-First off, both techniques have the same form of input, namely a bivariate function where the first input comes at stage one, and the second input comes at stage two.
+Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
+But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
+Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
+Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
+Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
 
-Again in both cases, the governing equations are too weak to fully determine the definitions of $p$ and $s$.
-Indeed, both admit completely trivial definitions.
-Consider the stage-splitter which always returns the identity for $f_1$ and $f$ for $f_2$, or analogously the partial evaluator which always returns an $f_x$ that just closes over the input $x$ and internally calls $f$ once $y$ is available.
-The ambiguity of these equations (modulo standard program equivalence of the outputs) can be resolved by adding annotations to $f$ to clearly specify the parts of the computation that are first stage and the parts that are second stage.
-Later, we show that the same annotations suffice for both partial evaluation and stage-splitting.
+[RTSL and SH]
 
-The differences between stage-splitting and partial evaluation are likewise evident from these governing equations.
-For instance in partial evaluation, the existential $f_x$ depends on $x$, which means that the partial evaluator cannot be run until $x$ is known.
-Moreover, if one wishes to specialize $f$ for multiple $x$'s, then the partial evaluator must be run several times.
-Depending on the use case and cost of partial evaluation, this may be prohibitively expensive.
-Alternatively, a stage-splitter need only be run once, and this can be done entirely before any $x$ is known.
+[Discuss Yong's recent paper here.
+It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
+They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
+If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
 
-\subsection{Partial Evaluator from Stage-Splitter}
-
-We can recover a valid partial evaluator from a stage-splitter by stage-splitting the input function $f$ into $f_1$ and $f_2$, computing $\llbracket f_1 \rrbracket(x)$ to obtain $\bar x$, and then returning an $f_x$ such that $\llbracket f_x \rrbracket(y) = \llbracket f_2 \rrbracket(\bar x, y)$.
-Note that this does not mean that stage-splitting is a strict generalization of partial evaluation.
-In practice, partial evaluators easily perform optimizations (such as branch elimination, discussed later) which are beyond the scope of stage-splitting, and would require further technology than has been developed here.
-It is best to think of stage-splitting as simply the first half of partial evaluation, where the back half is an optimizer.
-[Might be able to come up with a futamura projection-like statement here, which would be really really really cool.]
-
-\subsection{Stage-Splitter from Partial Evaluator}
-
-Likewise, we can easily recover a stage-splitter from a partial evaluator.  If $p$ is a valid partial evaluator, then we can define a stage splitter $s$ such that $s(f)=(f_1,f_2)$, where
-\begin{align*}
-[f_1](x) &= p (f,x) \\
-[f_2](l,y) &= [l] (y)
-\end{align*}
-This implicitly requires that the languages in which $f_1$ and $f_2$ are expressed are strong enough to write a partial evaluator, but that is the case in this paper.
-A stage-splitter defined this way leaves much to be desired.
-Firstly, partial evaluation of $f$ may be too expensive for the context in which $f_1$ needs to run.
-Additionally, the intermediate data structure created this way may be much larger than necessary, as it would contain all of the residual code.
-
-\section{A Two-Stage Modal Language}
-
-Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
+Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
+That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
+Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
+The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
+They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.

### 2014-06-11 paper/semantics.tex
Added a few notes.
Reworked file structure a little.

@@
+\section{A Two-Stage Modal Language}
+
+Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
+
 \section{Semantics}
@@
 \item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
+\item Second-stage cost is identical to the erasure semantics.
+\item First stage cost is different that erasure, do to speculation.
+We always evaluate stage-1 code once, which increases cost, except when that stage-1 code was inside a 2-lambda that got evaluated many times.
 \end{enumerate}

### 2014-06-11 paper/splitting.tex
Reworked file structure a little.

@@
 
-\section{Related Work}
-
-Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
-They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
-They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
-The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
-Accordingly, their approach was entirely informal, with no implementation realized.
-Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
-
-Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
-The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
-Their shaders are written in a C-like language with basic arithmetic and if statements.
-Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
-
-Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
-But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
-Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
-Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
-Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
-
-[RTSL and SH]
-
-[Discuss Yong's recent paper here.
-It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
-They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
-If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
-
-Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
-That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
-Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
-The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
-They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.
-
 [Meta-ML eases off on this restriction but does not (I think?) eliminate it.]

### 2014-06-12 paper/semantics.tex
Added text about the erasure semantics.

@@
 
-There's more than one way to define a semantics for our language.  We start this section by considering two proposals, characterizing them mainly by how they handle values of $\fut$ type.
+Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.
+In this section, we'll consider a few proposals for our language before settling on one we like.
+All of the proposals are call-by-value, differing primarily in how they handle values of $\fut$ type.
 
@@
 
-The first proposal we call the {\em erasure semantics}, because it corresponds to what one would get by simply {\em erasing} all of the $\next$ and $\prev$ terms that appear in \lamStaged.
-In particular, we say that $\next v$ is a first-stage value of type $\fut A$ if and only if $v$ is a second-stage value of type $A$, (...)
+We first consider the {\em erasure semantics}, so called because it corresponds to what one would get by interpreting \lamStaged as a single-stage language, ignoring all of the $\next$ and $\prev$ terms.
+This gives us two judgements, $\erasone$ and $\erastwo$, corresponding to {\em multistage evaluation} at \bbone and at \bbtwo.
+We call these judgments  ``multistage'' because they cause work to happen at both stages.
 
-...
+Both judgments behave normally at non-staged features.  We cover their behavior at staged features below:
 
-We say that erasure semantics gives us a notion of multi-stage evaluation.
-One judgement is multistage evaluation at 1, and the other is multistage evaluation at 2.
-The downside of this is that the erasure semantics doesn't seem to do all of the stage-1 work before the stage-2 work, for some notion of ``before''.
-Of course, the erasure semantics might be equivalent to some other semantics which does have that property.
-Unfortunately, this is false (show second-stage if example).
+\begin{mathpar}
+\infer {\next~e \erasone \next~v} {e \erastwo v} \and
+\infer {\prev~e \erastwo \next~v} {e \erasone v} \and
+\infer {\pause~e \erasone \next~i} {e \erasone i}
+\end{mathpar}
+
+Essentially, we immediately evaluate under the $\next$, yielding a value for $\fut$ types.
+The $\prev$ terms remove this wrapper.
+As expected, $\pause$ also gives us a way to inject into the wrapper at integers.
+
+The erasure semantics has some undesirable properties.
+By intention, it interleaves the execution of stage-\bbone~and stage-\bbtwo~code, so the evaluation can't really be said to be staged (i.e.
+stage-\bbone~work is done before stage-\bbtwo~work).
+Moreover, the erasure semantics cannot be equivalent to any semantics which does have this property!
+To see why, consider the following code, which types to int at \bbtwo:
 
 \begin{lstlisting} 
-next {
 if 5*4*3*2 > 111 then
-		prev{hold (2+4)}
+	prev{ hold (2+4) }
 else
-		prev{hold (4*5)}
-}
+	prev{ loopForever () (* does not terminate *) }
 \end{lstlisting}
 
+Under the erasure semantics (using $\erastwo$), this code takes the top branch and evaluates to 6.
+But in order know that the {\tt loopForever} function need not be called, the predicate had to be evaluated prior.
+But the predicate is stage-\bbtwo, whereas {\tt loopForever} is stage-\bbone.
+To borrow terminology from \cite{cave14}, this violates causality.
+In order to avoid this problem, a valid semantics must {\em speculate} down the branches of any stage-\bbtwo~if or case statement (or similarly into the body of a stage-\bbtwo function) to find and evaluate all of the stage-\bbone~code.
+Both of the other semantics we will consider have this property.
+
 The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.
-It also {\em obviously} produces the correct answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
+It also {\em obviously} produces the ``correct'' answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
 
@@
 $\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
-$\Xi \vdash \diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
+$\diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
 & $\typesone e A$ 

### 2014-06-13 paper/semantics.tex
changed up the semantics section a bit.  more forthcoming.

--- 2014-06-11 paper/semantics.tex
@@
 
-There's more than one way to define a semantics for our language.  We start this section by considering two proposals, characterizing them mainly by how they handle values of $\fut$ type.
+%Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.
+In this section, we'll consider a few proposals for our language before settling on one we like.
+All of the proposals are call-by-value, differing primarily in how they handle values of $\fut$ type.
 
-\subsection{The Erasure Semantics}
-
-The first proposal we call the {\em erasure semantics}, because it corresponds to what one would get by simply {\em erasing} all of the $\next$ and $\prev$ terms that appear in \lamStaged.
-In particular, we say that $\next v$ is a first-stage value of type $\fut A$ if and only if $v$ is a second-stage value of type $A$, (...)
-
-...
-
-We say that erasure semantics gives us a notion of multi-stage evaluation.
-One judgement is multistage evaluation at 1, and the other is multistage evaluation at 2.
-The downside of this is that the erasure semantics doesn't seem to do all of the stage-1 work before the stage-2 work, for some notion of ``before''.
-Of course, the erasure semantics might be equivalent to some other semantics which does have that property.
-Unfortunately, this is false (show second-stage if example).
-
-\begin{lstlisting} 
-next {
-	if 5*4*3*2 > 111 then
-		prev{hold (2+4)}
-	else
-		prev{hold (4*5)}
-}
-\end{lstlisting}
-
-The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.
-It also {\em obviously} produces the correct answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
-
-\subsection{Meta Semantics}
-
-A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.
-
-In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.
+%\subsection{The Erasure Semantics}
+%
+%We first consider the {\em erasure semantics}, so called because it corresponds to what one would get by interpreting \lamStaged as a single-stage language, ignoring all of the $\next$ and $\prev$ terms.
+This gives us two judgements, $\erasone$ and $\erastwo$, corresponding to {\em multistage evaluation} at \bbone and at \bbtwo.
+We call these judgments  ``multistage'' because they cause work to happen at both stages.
+%
+%Both judgments behave normally at non-staged features.  We cover their behavior at staged features below:
+%
+%\begin{mathpar}
+%\infer {\next~e \erasone \next~v} {e \erastwo v} \and
+%\infer {\prev~e \erastwo \next~v} {e \erasone v} \and
+%\infer {\pause~e \erasone \next~i} {e \erasone i}
+%\end{mathpar}
+%
+%Essentially, we immediately evaluate under the $\next$, yielding a value for $\fut$ types.
+The $\prev$ terms remove this wrapper.
+As expected, $\pause$ also gives us a way to inject into the wrapper at integers.
+%
+%The erasure semantics has some undesirable properties.
+By intention, it interleaves the execution of stage-\bbone~and stage-\bbtwo~code, so the evaluation can't really be said to be staged (i.e.
+stage-\bbone~work is done before stage-\bbtwo~work).
+Moreover, the erasure semantics cannot be equivalent to any semantics which does have this property!
+To see why, consider the following code, which types to int at \bbtwo:
+%
+%\begin{lstlisting} 
+%if 5*4*3*2 > 111 then
+%	prev{ hold (2+4) }
+%else
+%	prev{ loopForever () (* does not terminate *) }
+%\end{lstlisting}
+%
+%Under the erasure semantics (using $\erastwo$), this code takes the top branch and evaluates to 6.
+But in order know that the {\tt loopForever} function need not be called, the predicate had to be evaluated prior.
+But the predicate is stage-\bbtwo, whereas {\tt loopForever} is stage-\bbone.
+To borrow terminology from \cite{cave14}, this violates causality.
+In order to avoid this problem, a valid semantics must {\em speculate} down the branches of any stage-\bbtwo~if or case statement (or similarly into the body of a stage-\bbtwo function) to find and evaluate all of the stage-\bbone~code.
+Both of the other semantics we will consider have this property.
+%
+%The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.
+It also {\em obviously} produces the ``correct'' answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
+%
+%\subsection{Meta Semantics}
+%
+%A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.
+%
+%In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.
 But in the meta semantics, $\next e$ is a value only if $e$ has no $\prev$ terms; $v$ is allowed to have unreduced stage-2 computation.
-
-...
-
-The Davies semantics is comprised of two mutually recursive judgements: $\daviesz$ and $\davieso$.
+%
+%...
+%
+%The Davies semantics is comprised of two mutually recursive judgements: $\daviesz$ and $\davieso$.
 For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.
@@
 This essentially gives us partial evaluation, and we're left to just evaluate the residual normally to get multi-stage evaluation...)
+%
+%The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.
+This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins.
+%
+%The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.
+This makes reasoning about second-stage cost rather difficult.
+%
+%\begin{lstlisting} 
+%let x = next {4+5} in
+%next{prev{x} * prev{x}}
+%\end{lstlisting}
+%
+%\subsection{Our Semantics}
+%
+%We desire a semantics that meets the following goals:
+%
+%\begin{enumerate}
+%\item Modulo termination, it should be equivalent to the erasure semantics.
+%\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
+%\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
+%\end{enumerate}
+%
+%We meet all of these goals.
 
-The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.
-This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins.
+In this section, we discuss the dynamic semantics of our language.
+Before we can get into the details of the semantics, we have to first make some decisions about the behavior we want it to have.
+Similar to \cite{davies96}, we want our semantics to have the form of a judgement that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.
+Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.
+The only significant freedom that this leaves is how to handle $\next$, $\prev$, and $\pause$ constructs.
 
-The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.
-This makes reasoning about second-stage cost rather difficult.
-
+There are two feasible ways to partially evaluate the following stage-\bbtwo~expression:
 \begin{lstlisting} 
-let x = next {4+5} in
-next{prev{x} * prev{x}}
+let x = (next {4+5}, 7+8) in
+next{
+	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
+}
+\end{lstlisting}
+We can either duplicate the contents of the first $\next$ expression to produce,
+\begin{lstlisting}
+(4+5) * (4+5) * 15
+\end{lstlisting}
+or we can bind the contents of the $\next$ expression, and simply duplicate the variable,
+\begin{lstlisting} 
+let x1 = 4+5 in
+x1 * x1 * 15
 \end{lstlisting}
 
-\subsection{Our Semantics}
-
-We desire a semantics that meets the following goals:
-
-\begin{enumerate}
-\item Modulo termination, it should be equivalent to the erasure semantics.
-\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
-\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
-\end{enumerate}
-
-We meet all of these goals.
-
-In this section, we introduce a dynamic semantics for our language.
-Although our type system is virtually identical to that in \cite{davies96}, the semantics is different in terms of cost and termination behavior.
+Davies chose the first, whereas we choose the second technique.
+We think it makes more sense from the point of view of cost.
+That said, it's not immediately clear how to get this behavior.
+We first present an informal example to give an intuition for what our semantics does.
+We start with the same example as before, but wrapped with a $\prev$:
+\begin{lstlisting}
+prev {
+let x = (next {4+5}, 7+8) in
+next{
+	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
+}
+}
+\end{lstlisting}
+The first thing we need to do is evaluate the binding for $x$.
+Of course, we can't evaluate inside the $\next$ statement, but we still need to produce a value that will get substituted.
+What we do instead is replace the body of the $\next$ by a variable and make an explicit substitution, which then floats up to the top of the containing $\prev$:
+\begin{lstlisting} 
+prev{
+[4+5/x1]
+let x = (next{x1}, 15) in
+next{
+	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
+}
+}
+\end{lstlisting}
+We're now free to perform the stage-\bbone~substitution for $x$.  
+\begin{lstlisting} 
+prev{
+[4+5/x1]
+next{
+	prev{#1 (next{x1}, 15)} 
+	* prev{#1 (next{x1}, 15)} 
+	* prev{hold (#2 (next{x1}, 15))}
+}
+}
+\end{lstlisting}
+To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
+\begin{lstlisting} 
+prev{
+[4+5/x1]
+	next{ x1 * x1 * 15 }
+}
+\end{lstlisting}
+The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
+\begin{lstlisting} 
+let x1 = 4+5 in
+ x1 * x1 * 15
+\end{lstlisting}
 

### 2014-06-13 paper/semantics.tex
added more careful reification
changed 'table' to 'context' in a few spots

@@
 Abstractly, we can think of our evaluation as proceeding in the standard way for stage-1 code.
-When the evaluator encounters a $\next \{e\}$ expression, it places $e$ off to the side in a table and replaces the whole expression with a reference to the table entry.
+When the evaluator encounters a $\next \{e\}$ expression, it places $e$ off to the side in a context and replaces the whole expression with a reference to the context entry.
 These references are then passed around as stage-1 values for $\fut$ types.
@@
 
-More formally, evaluation comprises the folowing judgments:
+More formally, evaluation comprises the following judgments:
 
@@
 The input to evaluation is a stage-1 expression (usually $e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered later.
-Evaluation has two outputs: the {\em future table} (usually $\xi$) and the {\em partial value} (usually $v$).
+Evaluation has two outputs: the {\em future context} (usually $\xi$) and the {\em partial value} (usually $v$).
 We cover the latter first.
-The partial value is essentially the result of the first-stage portion of $e$, and must be a 1-value of the same type as $e$.
+The partial value is essentially the result of the first-stage portion of $e$, and must be a \bbone-value of the same type as $e$.
 As usual, this means that $v$ must be composed only of base primitives, tuples, injections, and lambdas (corresponding to base types, products, sums, and functions).
 Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.
+The mapping between these hatted variables and stage-two expressions is
 
@@
 \item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
-\item Second-stage cost is identical to the erasure semantics.
-\item First stage cost is different that erasure, do to speculation.
-We always evaluate stage-1 code once, which increases cost, except when that stage-1 code was inside a 2-lambda that got evaluated many times.
 \end{enumerate}

### 2014-06-15 paper/semantics.tex
splitting up figures into separate files
Clarifying the semantics.  Still more needs to be done.

@@
 Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.
-The only significant freedom that this leaves is how to handle $\next$, $\prev$, and $\pause$ constructs.
+The only significant freedom that this leaves is how to handle the staging constructs ($\next$, $\prev$, and $\pause$).
 
-There are two feasible ways to partially evaluate the following stage-\bbtwo~expression:
+We present here two reasonable interpretations of these staging constructs, 
+in the context of the following stage-\bbtwo~example:
 \begin{lstlisting} 
+prev{
 let x = (next {4+5}, 7+8) in
@@
 }
+}
 \end{lstlisting}
-We can either duplicate the contents of the first $\next$ expression to produce,
+The first option is to duplicate the contents of the first $\next$ expression to produce,
 \begin{lstlisting}
@@
 \end{lstlisting}
-or we can bind the contents of the $\next$ expression, and simply duplicate the variable,
+Essentially, we're treating $\next$ as a value a stage-\bbone, so long as all of its contained $\prev$s are reduced.  
+This is the technique of \cite{davies96}.
+Instead of this, we take a second, nearly equivalent approach.  
+That is to bind the contents of the first $\next$ expression to some variable (here, $x1$) and then duplicate that variable,
 \begin{lstlisting} 
@@
 
-Davies chose the first, whereas we choose the second technique.
-We think it makes more sense from the point of view of cost.
-That said, it's not immediately clear how to get this behavior.
-We first present an informal example to give an intuition for what our semantics does.
-We start with the same example as before, but wrapped with a $\prev$:
+We think that this interpretation makes more sense from the point of view of cost, 
+since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
+This is {\em nearly} equivalent to the first expression, because they have different termination behavior 
+when a non-terminating stage-\bbtwo~expression is declared but never used.
+
+Now that we know {\em what} the behavior is that we want, we can explain {\em how} our semantics achieves that behavior.  We start with the same example:
 \begin{lstlisting}
@@
 \end{lstlisting}
+
+Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
+Instead we lifted it out into a substitution, leaving a hatted variable [note: add hats above] in its place,
+and we reified the substitution into a let-binding at the enclosing $\prev$.
+
+\subsection{more formally...}
 

### 2014-06-15 paper/semantics.tex
braindumped a bunch of words about the static behavior

@@
 
-Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
+\input {figures/grammars}
+\input {figures/statics}
+
+The grammar and type system for our language, \lang, are given in \ref{fig:grammar} and \ref{fig:staticSemantics}.
+At the type level, the only difference between our language and that of \cite{davies96} is that we restrict ours to two stages.
+The features of \lang can be split into two categories: the staging features ($\next$, $\prev$, and $\pause$), and the non-staging features.
+
+All of the non-stage features have the standard type-level behavior, 
+with the small caveat that variables can only be referenced at the stage they were declared.
+
+The $\next$ expression allows us to embed [kayvon wants a different word here] stage-\bbtwo~code in stage-\bbone~code, 
+creating a $\fut$ type at stage-\bbone.  
+This $\fut$ wrapper can only be removed by the $\prev$ construct, which ensures that the result is used only at stage \bbtwo.
+
+The $\pause$ construct gives us a way to move integers from stage \bbone~to stage \bbtwo.  
+It is technically derivable (somewhat inefficiently) from the other features, 
+but we provide it as a primitive for the sake of simplifying our examples.
+
+We formulated our typing judgements in the style of \ref{davies96}, where the whole judgement is annotated with a stage.  
+Variables in the context are annotated with a stage and must be used at the same stage they were declared at.
+
+We now present an example of multi-stage fast-exponent.  
+The exponent ($e$) is provided at the first stage, whereas the base ($b$) is provided at the second stage.
+The code works by squaring if the exponent is even, and multiplication by the base if the exponent is odd.
+We see that all of the exponent math can be done at the first stage.
+
+\begin{lstlisting} 
+let exp (b : $int, e : int) : $int =
+	if e == 0 then
+		next{1}
+	else if (e mod 2) == 0 then
+		next{let x = prev{exp(b,e/2)} in x*x}
+	else
+		next{prev{b} * prev{exp (b,e-1)}}		
+\end{lstlisting}
 
 \section{Semantics}
+
+\input {figures/dynamics}
 
@@
 Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.
-The only significant freedom that this leaves is how to handle the staging constructs ($\next$, $\prev$, and $\pause$).
+The only significant freedom that this leaves is how to handle the staging constructs.
 

### 2014-06-16 paper/semantics.tex
tried a more types-first introduction to the language section

--- 2014-06-15 paper/semantics.tex
@@
 
-Introduce the next and prev concepts, along with typesystem.  Introduce binding time analysis here, and explain that we don't care about it.  Show some examples.  Introduce a hold operation.
+\input {figures/grammars}
+\input {figures/statics}
 
-\section{Semantics}
+Every expression in \lang can be understood as having a {\em stage},
+namely either stage-\bbone~a stage-\bbtwo, in addition to a type.
+Just as the type tells us {\em what} an expression will reduce to,
+a stage tells us {\em when} it will fully reduce. 
+The reason for a two-stage restriction (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
+
+The only types that exist at stage \bbtwo~are the familiar products, sums, functions, and base types.
+In addition to this usual set, stage~\bbone~also has one more type: $\fut A$, where $A$ is a stage-\bbtwo~type.
+This can be understood as the {\em promise} of a value of type $A$ at stage \bbtwo.
+We'll see shortly that the promise given by $\fut A$ can only be treated opaquely at stage one; 
+it cannot be ``cashed out" for an actual $A$ until the second stage.
+
+Although there is a strict one-way dependence between the two stages at the type level, they are mutually dependent at the term level.
+Specifically, $\next$ allows us to embed a stage-\bbtwo~expression into stage \bbone,
+whereas $\prev$ allows us to embed a stage-\bbone~expression into stage \bbtwo.
+These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
+and our concrete syntax, we surround their arguments with \{braces\} to clearly indicate the stage boundary.
+The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
+We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.
+
+The grammar and type system for our language, \lang, are given in \ref{fig:grammar} and \ref{fig:staticSemantics}.
+We formulated our typing judgements in the style of \ref{davies96}, where the whole judgement is annotated with a stage.  
+Variables in the context are annotated with a stage and must be used at the same stage at which they were declared.
+All of the non-stage constructs have the expected type-level behavior.  
+This is made manifest as rules which are entirely abstract over stage.
+
+[still working below this line...]
+
+The $\next$ expression allows us to embed [kayvon wants a different word here] stage-\bbtwo~expression in stage-\bbone~expression, 
+creating a $\fut$ type at stage-\bbone.  
+This $\fut$ wrapper can only be removed by the $\prev$ construct, which ensures that the result is used only at stage \bbtwo.
+
+The $\pause$ construct is a mechnism to wrap stage \bbone~integers so that they can be used at stage \bbtwo.  
+Although $\pause$ can be implemented from our other features, 
+we provide it as a primitive to simplify our examples.
+
+Below is an example of fast-exponent written in \lang.  
+The function excepts an exponent integer at the first stage, 
+a base ($b$) has type $\fut int$, indicated that it's content is accessible only at the second stage,
+and produces a circle int at stage two.
+The code works by squaring if the exponent is even, and multiplication by the base if the exponent is odd.
+The type system here is sufficient to prove that all of the decomposition of the exponent can be done at the first stage.
+
+\begin{lstlisting} 
+let exp (b : $int, e : int) : $int =
+	if e == 0 then
+		next{1}
+	else if (e mod 2) == 0 then
+		next{let x = prev{exp(b,e/2)} in x*x}
+	else
+		next{prev{b} * prev{exp (b,e-1)}}		
+\end{lstlisting}
+
+...segue into semantics here... talk about residual generation...
+
+\subsection{Semantics}
+
+\input {figures/dynamics}
 
@@
 
+% add concept of residuals
 In this section, we discuss the dynamic semantics of our language.  

### 2014-06-16 paper/semantics.tex
fleshed out the exponent example

@@
 Every expression in \lang can be understood as having a {\em stage},
-namely either stage-\bbone~a stage-\bbtwo, in addition to a type.
+namely either \bbone~or \bbtwo, in addition to a type.
 Just as the type tells us {\em what} an expression will reduce to,
 a stage tells us {\em when} it will fully reduce. 
-The reason for a two-stage restriction (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
+The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
 
@@
 These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
-and our concrete syntax, we surround their arguments with \{braces\} to clearly indicate the stage boundary.
+and we surround their arguments with \{braces\} in our concrete syntax to clearly indicate the stage boundaries with in a program.
 The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
@@
 
-[still working below this line...]
-
-The $\next$ expression allows us to embed [kayvon wants a different word here] stage-\bbtwo~expression in stage-\bbone~expression, 
-creating a $\fut$ type at stage-\bbone.  
-This $\fut$ wrapper can only be removed by the $\prev$ construct, which ensures that the result is used only at stage \bbtwo.
+Note that $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
+Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
+That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
+This promise can only be ``cashed in" be the $\prev$ construct.  
+Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
 
@@
 
-Below is an example of fast-exponent written in \lang.  
-The function excepts an exponent integer at the first stage, 
-a base ($b$) has type $\fut int$, indicated that it's content is accessible only at the second stage,
-and produces a circle int at stage two.
-The code works by squaring if the exponent is even, and multiplication by the base if the exponent is odd.
-The type system here is sufficient to prove that all of the decomposition of the exponent can be done at the first stage.
+\subsection {Staged Fast Exponentiation}
+
+Fast exponentiation is a method for calculating $b^p$ in $\log p$ time:
+\[
+	\mathit{fe}(b,p) = \left \{ \begin{array}{ll} 
+		1 &  e = 0 \\ 
+		\mathit{fe}(b,p/2)^2 & p \text{ even} \\ 
+		b \cdot \mathit{fe}(b,p-1) & p \text{ odd} \end{array}
+	\right .
+\]
+It is often observed that the decomposition of the power depends only on the power at prior recursive levels.
+For this reason, we can stage an implemention of fast exponentiation so that the power is available at the first stage and the base is available at the second stage.
+This implies that the result is then available only at the second stage.
+
+We implemented staged fast exponentiation in \lang~below:
 
@@
 
-...segue into semantics here... talk about residual generation...
+The type of the resursive function is $\fut{\rm int} \times {\rm int} \to \fut{\rm int}$.
+Note how the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the first stage.
+
+Given this definition of fast exponent, the question the remains how it can be run.  For this, we need a semantics for \lang.
 
@@
 
-% add concept of residuals
 In this section, we discuss the dynamic semantics of our language.  

### 2014-06-16 paper/semantics.tex
miscellaneous cleanup

@@
 Every expression in \lang can be understood as having a {\em stage},
-namely either \bbone~or \bbtwo, in addition to a type.
-Just as the type tells us {\em what} an expression will reduce to,
-a stage tells us {\em when} it will fully reduce. 
+either \bbone~or \bbtwo, in addition to a type.
+Just as a type describes \emph{which} values an expression reduces to,
+the stage tells us \emph{when} it will be fully reduced. 
+% FIXME why?
 The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
 
-The only types that exist at stage \bbtwo~are the familiar products, sums, functions, and base types.
-In addition to this usual set, stage~\bbone~also has one more type: $\fut A$, where $A$ is a stage-\bbtwo~type.
-This can be understood as the {\em promise} of a value of type $A$ at stage \bbtwo.
+The only types that exist at stage \bbtwo{} are products, sums, functions, and base types.
+In addition to this usual set, stage \bbone{} also has one more type: $\fut A$,
+where $A$ is a stage-\bbtwo{} type.
+This can be understood as \emph{promising} a value of type $A$ at stage \bbtwo.
 We'll see shortly that the promise given by $\fut A$ can only be treated opaquely at stage one; 
-it cannot be ``cashed out" for an actual $A$ until the second stage.
+it cannot be redeemed for an actual $A$ until the second stage.
 
-Although there is a strict one-way dependence between the two stages at the type level, they are mutually dependent at the term level.
-Specifically, $\next$ allows us to embed a stage-\bbtwo~expression into stage \bbone,
-whereas $\prev$ allows us to embed a stage-\bbone~expression into stage \bbtwo.
+Although there is a strict one-way dependence between the two stages at the type
+level, their terms are defined mutually recursively:
+$\next$ embeds stage-\bbtwo{} expressions into stage \bbone,
+whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
 These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
@@
 
-The grammar and type system for our language, \lang, are given in \ref{fig:grammar} and \ref{fig:staticSemantics}.
-We formulated our typing judgements in the style of \ref{davies96}, where the whole judgement is annotated with a stage.  
+The grammar and type system for our language, \lang, are given in
+\ref{fig:lamStagedSyntax} and \ref{fig:coreStatics}.
+We formulated our typing judgments in the style of \cite{davies96}, where the whole judgment is annotated with a stage.  
 Variables in the context are annotated with a stage and must be used at the same stage at which they were declared.
@@
 
-The $\pause$ construct is a mechnism to wrap stage \bbone~integers so that they can be used at stage \bbtwo.  
+The $\pause$ construct is a mechanism to wrap stage \bbone~integers so that they can be used at stage \bbtwo.  
 Although $\pause$ can be implemented from our other features, 
@@
 It is often observed that the decomposition of the power depends only on the power at prior recursive levels.
-For this reason, we can stage an implemention of fast exponentiation so that the power is available at the first stage and the base is available at the second stage.
+For this reason, we can stage an implementation of fast exponentiation so that the power is available at the first stage and the base is available at the second stage.
 This implies that the result is then available only at the second stage.
@@
 
-The type of the resursive function is $\fut{\rm int} \times {\rm int} \to \fut{\rm int}$.
-Note how the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the first stage.
+The type of the recursive function is $\fut{\rm int} \times {\rm int} \to \fut{\rm int}$.
+Note that the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the first stage.
 
+% FIXME but do we give a cost semantics?
 Given this definition of fast exponent, the question the remains how it can be run.  For this, we need a semantics for \lang.
@@
 %We first consider the {\em erasure semantics}, so called because it corresponds to what one would get by interpreting \lamStaged as a single-stage language, ignoring all of the $\next$ and $\prev$ terms.
-This gives us two judgements, $\erasone$ and $\erastwo$, corresponding to {\em multistage evaluation} at \bbone and at \bbtwo.
+This gives us two judgments, $\erasone$ and $\erastwo$, corresponding to {\em multistage evaluation} at \bbone and at \bbtwo.
 We call these judgments  ``multistage'' because they cause work to happen at both stages.
@@
 %
-%The Davies semantics is comprised of two mutually recursive judgements: $\daviesz$ and $\davieso$.
+%The Davies semantics is comprised of two mutually recursive judgments: $\daviesz$ and $\davieso$.
 For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.
@@
 Before we can get into the details of the semantics, we have to first make some decisions about the behavior we want it to have.  
-Similar to \cite{davies96}, we want our semantics to have the form of a judgement that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.
+Similar to \cite{davies96}, we want our semantics to have the form of a judgment that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.
 Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.
@@
 
+% FIXME we need a real name for hatted variables
 The evaluation judgment is very similar to standard call-by-value evaluation, as goal 1 would suggest.

### 2014-06-17 paper/semantics.tex
made a bunch of small fixes

@@
 The grammar and type system for our language, \lang, are given in
-\ref{fig:lamStagedSyntax} and \ref{fig:coreStatics}.
+\ref{fig:grammar} and \ref{fig:statics}.
 We formulated our typing judgments in the style of \cite{davies96}, where the whole judgment is annotated with a stage.  
@@
 
-Note that $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
+In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
 Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
 That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
-This promise can only be ``cashed in" be the $\prev$ construct.  
+This promise can only be redeemed be the $\prev$ construct.  
 Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
@@
 \[
-	\mathit{fe}(b,p) = \left \{ \begin{array}{ll} 
-		1 &  e = 0 \\ 
-		\mathit{fe}(b,p/2)^2 & p \text{ even} \\ 
-		b \cdot \mathit{fe}(b,p-1) & p \text{ odd} \end{array}
+	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
+		1 &  p = 0 \\ 
+		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
+		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
 	\right .
@@
 \begin{lstlisting} 
-let exp (b : $int, e : int) : $int =
+let fexp (b : $int, e : int) : $int =
 	if e == 0 then
@@
 	else if (e mod 2) == 0 then
-		next{let x = prev{exp(b,e/2)} in x*x}
+		next{let x = prev{fexp(b,e/2)} in x*x}
 	else
-		next{prev{b} * prev{exp (b,e-1)}}		
+		next{prev{b} * prev{fexp(b,e-1)}}		
 \end{lstlisting}
@@
 
-% FIXME but do we give a cost semantics?
+% FIXME but do we give a cost semantics? no
 Given this definition of fast exponent, the question the remains how it can be run.  For this, we need a semantics for \lang.

### 2014-06-17 paper/semantics.tex
Added more and better discussion of the fexp example.

@@
 \]
-It is often observed that the decomposition of the power depends only on the power at prior recursive levels.
-For this reason, we can stage an implementation of fast exponentiation so that the power is available at the first stage and the base is available at the second stage.
-This implies that the result is then available only at the second stage.
 
-We implemented staged fast exponentiation in \lang~below:
+Without going into details of how recursion is handled, we implement staged fast exponentiation in \lang~below.  
+We use {\tt \$} in place of $\fut$ in this an other examples.
 
@@
 	else if (e mod 2) == 0 then
-		next{let x = prev{fexp(b,e/2)} in x*x}
+		next{let x = prev{fexp(b,p/2)} in x*x}
 	else
-		next{prev{b} * prev{fexp(b,e-1)}}		
+		next{prev{b} * prev{fexp(b,p-1)}}		
 \end{lstlisting}
 
-The type of the recursive function is $\fut{\rm int} \times {\rm int} \to \fut{\rm int}$.
-Note that the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the first stage.
+Although the code for {\tt fexp} looks very much like the stage-less mathematical definition above, it is in fact a staged program.  
+See how the {\tt if}s and exponent decomposition are all at stage \bbone, since they occur within $\prev$ blocks.  
+Essentially, the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the stage \bbone~without requiring us to disrupt the elegant functional structure of the code.
 
-% FIXME but do we give a cost semantics? no
-Given this definition of fast exponent, the question the remains how it can be run.  For this, we need a semantics for \lang.
+This example also illustrates how stage \bbone~and stage \bbtwo~code can be arbitrarily nested.
+Normal notions of eager evaluation reduce code in a leaf-to-root way, but this conflicts with the precept that we evaluate stage \bbone~before stage \bbtwo.
+To resolve this conflict, we need a formal semantics for \lang.
 

### 2014-06-19 paper/semantics.tex
major reorganizing/rewriting of section 3.

@@
 These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
-and we surround their arguments with \{braces\} in our concrete syntax to clearly indicate the stage boundaries with in a program.
+and we surround their arguments with \{braces\} in the concrete syntax to clearly indicate the stage boundaries with in a program.
 The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
@@
 
+%We formulated our typing judgments in the style of \cite{davies96}, where the
+%whole judgment is annotated with a stage.
+
 The grammar and type system for our language, \lang, are given in
-\ref{fig:grammar} and \ref{fig:statics}.
-We formulated our typing judgments in the style of \cite{davies96}, where the whole judgment is annotated with a stage.  
-Variables in the context are annotated with a stage and must be used at the same stage at which they were declared.
-All of the non-stage constructs have the expected type-level behavior.  
-This is made manifest as rules which are entirely abstract over stage.
+\ref{fig:grammar,fig:statics}.
+We annotate the typing judgment and context variables with stages; only $\fut$
+and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
+of a term or type.
+%This is made manifest as rules which are entirely abstract over stage.
+%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
 
-In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
 Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
@@
 
-\subsection {Staged Fast Exponentiation}
+\subsection{Staged Evaluation}
 
@@
 
-Without going into details of how recursion is handled, we implement staged fast exponentiation in \lang~below.  
-We use {\tt \$} in place of $\fut$ in this an other examples.
+Ignoring for the moment how to define recursive functions in \lang, we can
+implement staged fast exponentiation as follows:%
+\footnote{We render $\fut$ as {\tt \$} in code examples.}
 
@@
 
-Although the code for {\tt fexp} looks very much like the stage-less mathematical definition above, it is in fact a staged program.  
-See how the {\tt if}s and exponent decomposition are all at stage \bbone, since they occur within $\prev$ blocks.  
-Essentially, the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the stage \bbone~without requiring us to disrupt the elegant functional structure of the code.
+Although the code for {\tt fexp} looks very much like the unstaged mathematical
+definition above, it is in fact a staged program: the {\tt if}s and exponent
+decomposition are all at stage \bbone, since they occur within $\prev$ blocks.
+Essentially, the type system is sufficient to prove the observation that all of
+the decomposition of the exponent can be done at the stage \bbone~without
+requiring us to disrupt the elegant functional structure of the code.
 
-This example also illustrates how stage \bbone~and stage \bbtwo~code can be arbitrarily nested.
-Normal notions of eager evaluation reduce code in a leaf-to-root way, but this conflicts with the precept that we evaluate stage \bbone~before stage \bbtwo.
-To resolve this conflict, we need a formal semantics for \lang.
+This example also illustrates how stage \bbone{} and stage \bbtwo{} code can be
+arbitrarily nested.  Ordinary evaluation always eliminates outermost redexes
+first, but this conflicts with the precept that we evaluate all stage \bbone{}
+code before stage \bbtwo{} code, in the case that stage \bbone{} code is inside
+stage \bbtwo{} code.
 
-\subsection{Semantics}
+A suitably staged dynamic semantics for \lang{} must therefore evaluate
+\emph{all} the stage \bbone{} subexpressions of a term before its stage
+\bbtwo{} subexpressions.
+
+Consider how we could achieve this in the following stage \bbtwo{} term:
+\begin{lstlisting} 
+prev{
+let x = (next {4+5}, 7+8) in
+next{
+	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
+}
+}
+\end{lstlisting}
+One option is to duplicate the contents of the first $\next$ expression, to produce
+\begin{lstlisting}
+(4+5) * (4+5) * 15
+\end{lstlisting}
+
+Here we are essentially treating $\next\{e\}$ as a stage \bbone{} value, so long
+as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
+\cite{davies96}.
+
+We take a different approach in our dynamic semantics, which is to let-bind the
+contents of the first $\next$ expression to some variable (here, $x1$) and
+duplicate that variable,
+\begin{lstlisting} 
+let x1 = 4+5 in
+x1 * x1 * 15
+\end{lstlisting}
+
+We think that this interpretation makes more sense from the point of view of cost, 
+since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
+This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
+when a non-terminating stage-\bbtwo~expression is declared but never used.
+
+How do our semantics achieve this behavior? In the original term,
+\begin{lstlisting}
+prev {
+let x = (next {4+5}, 7+8) in
+next{
+	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
+}
+}
+\end{lstlisting}
+we must start by evaluating the binding for $x$.  We can't evaluate inside the
+$\next$ statement, but must still substitute some result into the body of the
+\texttt{let}. We replace the contents of $\next$ with a variable and make an
+explicit substitution, which then floats up to the top of the containing
+$\prev$:
+\begin{lstlisting} 
+prev{
+[4+5/x1]
+let x = (next{x1}, 15) in
+next{
+	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
+}
+}
+\end{lstlisting}
+We're now free to perform the stage-\bbone~substitution for $x$.  
+\begin{lstlisting} 
+prev{
+[4+5/x1]
+next{
+	prev{#1 (next{x1}, 15)} 
+	* prev{#1 (next{x1}, 15)} 
+	* prev{hold (#2 (next{x1}, 15))}
+}
+}
+\end{lstlisting}
+To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
+\begin{lstlisting} 
+prev{
+[4+5/x1]
+	next{ x1 * x1 * 15 }
+}
+\end{lstlisting}
+The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
+\begin{lstlisting} 
+let x1 = 4+5 in
+ x1 * x1 * 15
+\end{lstlisting}
+
+Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
+Instead we lifted it out into a substitution, leaving a hatted variable [\TODO add hats above] in its place,
+and we reified the substitution into a let-binding at the enclosing $\prev$.
+
+\subsection{Dynamics}
 
@@
 
+%{{{ stuff nico commented out
 %Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.
@@
 %We meet all of these goals.
+%}}}
+%{{{ stuff carlo commented out
+%Abstractly, we can think of our evaluation as proceeding in the standard way for
+%stage-1 code. When the evaluator encounters a $\next \{e\}$ expression, it
+%places $e$ off to the side in a context and replaces the whole expression with a
+%reference to the context entry.  These references are then passed around as
+%stage-1 values for $\fut$ types.  But what if $e$ contains $\prev$ expressions?
+%To ensure that all stage-1 code is evaluated before any stage-2 code, we must
+%evaluate all of the 1-code contained in $e$ before inserting it into the table.
+%This entails searching $e$ for all contained $\prev$s and evaluating them in
+%place.
 
-In this section, we discuss the dynamic semantics of our language.  
-Before we can get into the details of the semantics, we have to first make some decisions about the behavior we want it to have.  
-Similar to \cite{davies96}, we want our semantics to have the form of a judgment that partially evaluates an expression to create a residual that can be evaluated at the second stage in the standard way.
-Again like Davies, we choose a call-by-value paradigm as our baseline at both stages.
-The only significant freedom that this leaves is how to handle the staging constructs.
+%The evaluation judgment operates on stage-1 code, whereas the
+%speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the
+%crossover points between 1-code and 2-code, they are correspondingly the only
+%places where the evaluation and speculation judgments depend on the other.
 
-We present here two reasonable interpretations of these staging constructs, 
-in the context of the following stage-\bbtwo~example:
-\begin{lstlisting} 
-prev{
-let x = (next {4+5}, 7+8) in
-next{
-	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
-}
-}
-\end{lstlisting}
-The first option is to duplicate the contents of the first $\next$ expression to produce,
-\begin{lstlisting}
-(4+5) * (4+5) * 15
-\end{lstlisting}
-Essentially, we're treating $\next$ as a value a stage-\bbone, so long as all of its contained $\prev$s are reduced.  
-This is the technique of \cite{davies96}.
-Instead of this, we take a second, nearly equivalent approach.  
-That is to bind the contents of the first $\next$ expression to some variable (here, $x1$) and then duplicate that variable,
-\begin{lstlisting} 
-let x1 = 4+5 in
-x1 * x1 * 15
-\end{lstlisting}
+%The evaluation judgment is very similar to standard call-by-value evaluation.
+%The input to evaluation is a stage-1 expression (usually
+%$e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered
+%later.  Evaluation has two outputs: the {\em future context} (usually $\xi$) and
+%the {\em partial value} (usually $v$).  We cover the latter first.  The partial
+%value is essentially the result of the first-stage portion of $e$, and must be a
+%\bbone-value of the same type as $e$.  As usual, this means that $v$ must be
+%composed only of base primitives, tuples, injections, and lambdas (corresponding
+%to base types, products, sums, and functions).  Analogously, the value
+%corresponding to $\fut$ is a construct called a {\em hatted variable} (written
+%$\hat y$), which signifies a reference to some stage-2 computation.  The mapping
+%between these hatted variables and stage-two expressions is
+%}}}
 
-We think that this interpretation makes more sense from the point of view of cost, 
-since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
-This is {\em nearly} equivalent to the first expression, because they have different termination behavior 
-when a non-terminating stage-\bbtwo~expression is declared but never used.
+% TODO move these to macros.tex once they're finalized
+\newcommand{\pval}{\ensuremath{\mathsf{pval}}}
+\newcommand{\res}{\ensuremath{\mathsf{res}}}
+\renewcommand{\val}{\ensuremath{\mathsf{val}}}
 
-Now that we know {\em what} the behavior is that we want, we can explain {\em how} our semantics achieves that behavior.  We start with the same example:
-\begin{lstlisting}
-prev {
-let x = (next {4+5}, 7+8) in
-next{
-	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
-}
-}
-\end{lstlisting}
-The first thing we need to do is evaluate the binding for $x$.
-Of course, we can't evaluate inside the $\next$ statement, but we still need to produce a value that will get substituted.
-What we do instead is replace the body of the $\next$ by a variable and make an explicit substitution, which then floats up to the top of the containing $\prev$:
-\begin{lstlisting} 
-prev{
-[4+5/x1]
-let x = (next{x1}, 15) in
-next{
-	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
-}
-}
-\end{lstlisting}
-We're now free to perform the stage-\bbone~substitution for $x$.  
-\begin{lstlisting} 
-prev{
-[4+5/x1]
-next{
-	prev{#1 (next{x1}, 15)} 
-	* prev{#1 (next{x1}, 15)} 
-	* prev{hold (#2 (next{x1}, 15))}
-}
-}
-\end{lstlisting}
-To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
-\begin{lstlisting} 
-prev{
-[4+5/x1]
-	next{ x1 * x1 * 15 }
-}
-\end{lstlisting}
-The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
-\begin{lstlisting} 
-let x1 = 4+5 in
- x1 * x1 * 15
-\end{lstlisting}
+The algorithm described above creates three different kinds of expressions which
+cannot be evaluated further at a particular stage:
+\begin{itemize}
+\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
+subexpressions have all been fully evaluated. In the example above,
+\verb|next {x1 * x1 * 15}|.
+\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
+subexpressions have all been fully evaluated. In the example above,
+\verb|4+5|.
+\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
+are the results of a computation after both stages have been completed. In the
+example above, the term evaluates to \verb|1215|.
+\end{itemize}
+The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
+environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
+and a partial value $v$ which may refer to those variables. This judgment
+necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
+\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
+implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
+residual.
 
-Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
-Instead we lifted it out into a substitution, leaving a hatted variable [note: add hats above] in its place,
-and we reified the substitution into a let-binding at the enclosing $\prev$.
+\TODO finish explaining how the nested evaluation actually works.
+explain reification, hatted variables, and contexts.
 
-\subsection{more formally...}
+A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
+given a type of the form $A\to\fut(B\to C)$.%
+\footnote{We can rewrite \texttt{fexp} in this form, or simply apply the following
+higher-order function which makes the adjustment:
 
-Abstractly, we can think of our evaluation as proceeding in the standard way for stage-1 code.
-When the evaluator encounters a $\next \{e\}$ expression, it places $e$ off to the side in a context and replaces the whole expression with a reference to the context entry.
-These references are then passed around as stage-1 values for $\fut$ types.
-But what if $e$ contains $\prev$ expressions?
-To ensure that all stage-1 code is evaluated before any stage-2 code, we must evaluate all of the 1-code contained in $e$ before inserting it into the table.
-This entails searching $e$ for all contained $\prev$s and evaluating them in place.
+\TODO write the function here}
+Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
+function:
+$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
+The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
+C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
+enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
+ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
+Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
+the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.
 
-More formally, evaluation comprises the following judgments:
+\TODO make sure this is true (haven't written the canonical forms theorems yet)
 
-\begin{center}
-\begin{tabular}{|c|p{2.5cm}|p{2.5cm}|} \hline
-Judgment & Reads & Conditions \\ \hline 
-$\isvalone e$ & ``$e$ is a $\bbone$-value under context $\Xi$'' & ... \\ \hline 
-$\diaonesub$ & ``$e$ evaluates to future table $\xi$ and value $v$'' 
-& $\typesone e A$ 
-	\newline $\isvalone v$ 
-	\newline ... \\ \hline 
-$\diatwosub$ & ``$e$ speculates to $q$'' & ... \\ \hline 
-...&...&... \\ \hline
-\end{tabular}
-\end{center}
+That this sequence of evaluations is in fact staged follows from our
+characterizations of partial values, residuals, and values, that $\redonesym$
+outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
 
-Evaluation and speculation are the main judgments here, the rest being largely administrative.
-The evaluation judgment operates on stage-1 code, whereas the speculation operates on stage-2 code.
-Since $\next$ and $\prev$ are the crossover points between 1-code and 2-code, they are correspondingly the only places where the evaluation and speculation judgments depend on the other.
+\begin{remark}
+For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
+always compute an empty environment, and a partial value identical to the result
+of call-by-value evaluation of $e$.
+%derivationally equivalent to standard call-by-value evaluation.
+\end{remark}
 
-% FIXME we need a real name for hatted variables
-The evaluation judgment is very similar to standard call-by-value evaluation, as goal 1 would suggest.
-The input to evaluation is a stage-1 expression (usually $e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered later.
-Evaluation has two outputs: the {\em future context} (usually $\xi$) and the {\em partial value} (usually $v$).
-We cover the latter first.
-The partial value is essentially the result of the first-stage portion of $e$, and must be a \bbone-value of the same type as $e$.
-As usual, this means that $v$ must be composed only of base primitives, tuples, injections, and lambdas (corresponding to base types, products, sums, and functions).
-Analogously, the value corresponding to $\fut$ is a construct called a {\em hatted variable} (written $\hat y$), which signifies a reference to some stage-2 computation.
-The mapping between these hatted variables and stage-two expressions is
-
-...
-
-\subsection{Other nice properties}
-
-\begin{enumerate}
-\item For some $\colone{e}{A}$ containing no $\next$ expressions, the semantics is derivationally equivalent to standard call-by-value evaluation.
-\end{enumerate}
+\TODO 
+If we change the highlighted rules to something slightly different, then we get
+precisely the semantics from \cite{davies96}. This essentially bypasses the
+environment bookkeeping in $\redonesym$, by inlining residuals instead of
+hoisting them in \verb|let|-bindings.
 
@@
 
-\TODO unfinished; where should this go? combine with table above
-
-\noindent\TODO terminology for $\Xi,\Gamma,\xi,v,q$?
-
-\noindent\TODO need to define 1-vals at 1 and 2, and 2-vals at 2. what is the
-terminology? (?, residuals, results?)
-
-\noindent\TODO put $\Xi$ and $\Gamma$ in the judgments where appropriate
-
-\noindent\TODO be uniform about which judgments have which contexts, etc
+\noindent\TODO terminology for the contexts?
 
@@
 
-\noindent\TODO don't call it a table
-
-\noindent\TODO what is the notation for $\xi$s? how do you build them?
-
-\noindent\TODO define reification explicitly as a metafunction
-
 \begin{definition}
-A table $\xi$ is well-formed ($\xi\wf$) if either:
+An environment $\xi$ is well-formed ($\xi\wf$) if either:
 \begin{enumerate}

### 2014-06-19 paper/staging.tex
major reorganizing/rewriting of section 3.

@@
 
-The grammar for types, terms, values, and contexts in \ellStaged are given in Figure \ref{fig:coreSyntax}.
+The grammar for types, terms, values, and contexts in \ellStaged are given in \ref{fig:coreSyntax}.
 The first thing to notice is that stage $\bbone$ and stage $\bbtwo$ code are syntactically seperated at the type, term, and value levels, despite heavy overlap.
@@
 
-The static semantics for \ellStaged~is given in Figure \ref{fig:coreStatics}.
+The static semantics for \ellStaged~is given in \ref{fig:coreStatics}.
 Modulo the two-stage restriction and current absence of functions, it's based off of the system in \cite{davies96}, which was inspired by temporal logic (the source of our $\fut$ modality).

### 2014-06-19 paper/semantics.tex
As per discussion with Carlo today, beginning to tighten up the text at the beginning of section 3.  The goal is to be more concise, as the reader need not be handheld through staging itself

@@
 
-Every expression in \lang can be understood as having a {\em stage},
-either \bbone~or \bbtwo, in addition to a type.
-Just as a type describes \emph{which} values an expression reduces to,
-the stage tells us \emph{when} it will be fully reduced. 
+In this section, we describe \lang, a simple two-stage language
+used as the basis for defining and analyzing stage splitting.
+%The grammar and type system for \lang\ is given in
+%\ref{fig:grammar,fig:statics}.
+Every expression in \lang\ has a statically known {\em stage},
+either \bbone~or \bbtwo, which defines when it is fully reduced.
+%\footnote{Typing judgments and context variables are annotated with stages.}
+%
+% in addition to a type.
+% Just as a type describes \emph{which} values an expression reduces to,
+%the expression's stage tells us \emph{when} it will be fully reduced. 
 % FIXME why?
-The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
-
-The only types that exist at stage \bbtwo{} are products, sums, functions, and base types.
-In addition to this usual set, stage \bbone{} also has one more type: $\fut A$,
-where $A$ is a stage-\bbtwo{} type.
-This can be understood as \emph{promising} a value of type $A$ at stage \bbtwo.
-We'll see shortly that the promise given by $\fut A$ can only be treated opaquely at stage one; 
-it cannot be redeemed for an actual $A$ until the second stage.
+%The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
+Stage-\bbone{} and stage-\bbtwo{} types include products, sums, functions, and base types.
+In addition to these types, stage-\bbone{} expressions may also be of type $\fut A$,
+representing an enclosed stage-\bbtwo{} expression of type $A$.
+$\fut A$ terms must be treated opaquely by stage-\bbone{} code
+as they cannot be reduced (to values with type $A$) until stage \bbtwo.
 
 Although there is a strict one-way dependence between the two stages at the type
-level, their terms are defined mutually recursively:
-$\next$ embeds stage-\bbtwo{} expressions into stage \bbone,
+level, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
+the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
 whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
-These $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
-and we surround their arguments with \{braces\} in the concrete syntax to clearly indicate the stage boundaries with in a program.
-The problem of how to turn an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
-We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.
+The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
+and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.
 
+The grammar and type system for \lang\ is given in
+\ref{fig:grammar,fig:statics}. (The figures annotate typing judgments and context variables with stages.)
+Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
+of a term or type.
 %We formulated our typing judgments in the style of \cite{davies96}, where the
 %whole judgment is annotated with a stage.  
-
-The grammar and type system for our language, \lang, are given in
-\ref{fig:grammar,fig:statics}.
-We annotate the typing judgment and context variables with stages; only $\fut$
-and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
-of a term or type.
+%The grammar and type system for \lang\ is given in
+%\ref{fig:grammar,fig:statics}.
+% We annotate typing judgments and context variables with stages;
 %This is made manifest as rules which are entirely abstract over stage.
 %In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
+Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
+%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
+Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
+Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
+The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
+Although it is possible to implement $\pause$ from other \lang\ features, 
+we provide it as a core primitive to simplify our examples.
 
-Specifically, $\next$, given an argument with type $A$ at stage \bbtwo, forms a $\fut A$ at stage \bbone.  
-That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
-This promise can only be redeemed be the $\prev$ construct.  
-Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
+% As our concern is stage splitting of explicitly staged programs expressed in \lang.
+We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
+We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent spitting.
 
-The $\pause$ construct is a mechanism to wrap stage \bbone~integers so that they can be used at stage \bbtwo.  
-Although $\pause$ can be implemented from our other features, 
-we provide it as a primitive to simplify our examples.
+%The problem of how to convert an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
+%We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.
 
@@
 
-Fast exponentiation is a method for calculating $b^p$ in $\log p$ time:
+As a simple example of expressing staged programs in \lang\, consider the following fast exponentiation algorithm below, which 
+calculates $b^p$ in $\log p$ time:
 \[
@@
 
-Ignoring for the moment how to define recursive functions in \lang, we can
-implement staged fast exponentiation as follows:%
+% Ignoring for the moment how to define recursive functions in \lang,
+\noindent
+The algorithm is expressed in \lang\ below:%
 \footnote{We render $\fut$ as {\tt \$} in code examples.}

### 2014-06-19 paper/semantics.tex
Minor edits to the semantics section 3.1.
-- Highlight in red statements that need to be changed.
-- Fix choice of numbers in the examples.

@@
 
-Although there is a strict one-way dependence between the two stages at the type
-level, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
+\needsfix{Although there is a strict one-way dependence between the two stages at the type
+level}, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
 the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
@@
 
-% As our concern is stage splitting of explicitly staged programs expressed in \lang.
-We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
-We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent spitting.
-
-%The problem of how to convert an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
-%We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.
+We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
+We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
 
@@
 
-As a simple example of expressing staged programs in \lang\, consider the following fast exponentiation algorithm below, which 
+As a simple example of expressing staged programs in \lang, consider the following fast exponentiation algorithm, which 
 calculates $b^p$ in $\log p$ time:
@@
 \noindent
-The algorithm is expressed in \lang\ below:%
+The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
 \footnote{We render $\fut$ as {\tt \$} in code examples.}
@@
 Although the code for {\tt fexp} looks very much like the unstaged mathematical
-definition above, it is in fact a staged program: the {\tt if}s and exponent
-decomposition are all at stage \bbone, since they occur within $\prev$ blocks.
-Essentially, the type system is sufficient to prove the observation that all of
-the decomposition of the exponent can be done at the stage \bbone~without
-requiring us to disrupt the elegant functional structure of the code.
+definition above, it is in fact a staged program: the {\tt if} predicates and exponent
+decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
+\needsfix{Essentially, the type system is sufficient to prove the observation that all of
+the decomposition of the exponent can be done at the stage \bbone\ without
+requiring disruption of the elegant functional structure of the code.}
 
-This example also illustrates how stage \bbone{} and stage \bbtwo{} code can be
-arbitrarily nested.  Ordinary evaluation always eliminates outermost redexes
-first, but this conflicts with the precept that we evaluate all stage \bbone{}
-code before stage \bbtwo{} code, in the case that stage \bbone{} code is inside
-stage \bbtwo{} code.
+A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation always eliminates outermost redexes first, 
+however in the case that stage \bbone\ expressions are contained inside
+stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code.
+Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
+\emph{all} the stage \bbone\ subexpressions of a term before its stage
+\bbtwo\ subexpressions.
 
-A suitably staged dynamic semantics for \lang{} must therefore evaluate
-\emph{all} the stage \bbone{} subexpressions of a term before its stage
-\bbtwo{} subexpressions.
-
-Consider how we could achieve this in the following stage \bbtwo{} term:
+To understand how this behavior can be achieved, consider the following stage-\bbone\ term:
 \begin{lstlisting} 
-prev{
-let x = (next {4+5}, 7+8) in
-next{
-	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
-}
-}
+let x = (next {1+2}, 3+4) in
+  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
 \end{lstlisting}
@@
 \begin{lstlisting}
-(4+5) * (4+5) * 15
+(1+2) * (1+2) * 7
 \end{lstlisting}
@@
 \begin{lstlisting} 
-let x1 = 4+5 in
-x1 * x1 * 15
+let x1 = 1+2 in
+  x1 * x1 * 7
 \end{lstlisting}
@@
 \begin{lstlisting}
-prev {
-let x = (next {4+5}, 7+8) in
-next{
-	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
-}
-}
+let x = (next {1+2}, 3+4) in
+  next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev{
-[4+5/x1]
-let x = (next{x1}, 15) in
-next{
-	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
-}
-}
+[1+2/x1]
+let x = (next{x1}, 7) in
+next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev{
-[4+5/x1]
+[1+2/x1]
 next{
-	prev{#1 (next{x1}, 15)} 
-	* prev{#1 (next{x1}, 15)} 
-	* prev{hold (#2 (next{x1}, 15))}
-}
+	prev{#1 (next{x1}, 7)} *
+	prev{#1 (next{x1}, 7)} *
+	prev{hold (#2 (next{x1}, 7))}
 }
@@
 \begin{lstlisting} 
-prev{
-[4+5/x1]
-	next{ x1 * x1 * 15 }
-}
+[1+2/x1]
+	next{ x1 * x1 * 7 }
 \end{lstlisting}
@@
 \begin{lstlisting} 
-let x1 = 4+5 in
- x1 * x1 * 15
+let x1 = 1+2 in
+  x1 * x1 * 7
 \end{lstlisting}
 
-Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
+Once again, the point here is that we did not duplicate the $1+2$ in our residual.  
 Instead we lifted it out into a substitution, leaving a hatted variable [\TODO add hats above] in its place,
@@
 subexpressions have all been fully evaluated. In the example above,
-\verb|next {x1 * x1 * 15}|.
+\verb|next {x1 * x1 * 7}|.
 \item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
 subexpressions have all been fully evaluated. In the example above,
-\verb|4+5|.
+\verb|1+2|.
 \item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these

### 2014-06-19 paper/semantics.tex
checkin at end of Kayvon's edits for the day.  Handing over to Nico.

@@
-\section{A Two-Stage Modal Language}
+\section{The \lang\ Language}
 
@@
 %The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
-Stage-\bbone{} and stage-\bbtwo{} types include products, sums, functions, and base types.
-In addition to these types, stage-\bbone{} expressions may also be of type $\fut A$,
-representing an enclosed stage-\bbtwo{} expression of type $A$.
-$\fut A$ terms must be treated opaquely by stage-\bbone{} code
-as they cannot be reduced (to values with type $A$) until stage \bbtwo.
+Stage-\bbone\ and stage-\bbtwo\ types include products, sums, functions, and base types.
+In addition to these types, stage-\bbone\ expressions may also be of type $\fut A$,
+representing an enclosed stage-\bbtwo\ expression of type $A$.
+(The type system of \lang\ is adopted directly from that of the modal language \lamCircle\,\cite{davies96}, but is restricted to two stages for simplicity.) $\fut A$ terms must be treated opaquely by stage-\bbone\ code
+as they cannot be reduced (to values with type $A$) until stage \bbtwo. Like in \lamCircle\, there is no reciprocal modality.
 
-\needsfix{Although there is a strict one-way dependence between the two stages at the type
-level}, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
-the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
-whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
+Although there is a strict one-way dependence between the two stages at the type
+level, stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
+the $\next$ construct embeds stage-\bbtwo\ expressions into stage \bbone,
+whereas $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.
 The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
@@
 The grammar and type system for \lang\ is given in
-\ref{fig:grammar,fig:statics}. (The figures annotate typing judgments and context variables with stages.)
+\ref{fig:grammar,fig:statics}. (Typing judgments and context variables are annotated with stages.)
 Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
@@
 Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
-Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
+Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality\,\cite{cave14}.
 The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
@@
 we provide it as a core primitive to simplify our examples.
-
-We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
-We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
 
@@
 decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
-\needsfix{Essentially, the type system is sufficient to prove the observation that all of
-the decomposition of the exponent can be done at the stage \bbone\ without
-requiring disruption of the elegant functional structure of the code.}
+%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
+%the decomposition of the exponent can be done at the stage \bbone\ without
+%requiring disruption of the elegant functional structure of the code.}
+We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
+We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
 
-A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation always eliminates outermost redexes first, 
+A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation eliminates outermost redexes first, 
 however in the case that stage \bbone\ expressions are contained inside
@@
 
-To understand how this behavior can be achieved, consider the following stage-\bbone\ term:
+To understand how this behavior can be achieved, consider evaluation of the stage-\bbone\ term below. (The term contains nested $\next$ and $\prev$ blocks.):
 \begin{lstlisting} 
@@
 
-Here we are essentially treating $\next\{e\}$ as a stage \bbone{} value, so long
+Here we are essentially treating $\next\{e\}$ as a stage-\bbone\ value, so long
 as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
@@
 
-We think that this interpretation makes more sense from the point of view of cost, 
-since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
-This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
-when a non-terminating stage-\bbtwo~expression is declared but never used.
+% We think that this interpretation makes more sense from the point of view of cost, 
+% since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
+% This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
+% when a non-terminating stage-\bbtwo~expression is declared but never used.
 
-How do our semantics achieve this behavior? In the original term,
-\begin{lstlisting}
-let x = (next {1+2}, 3+4) in
-  next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
-\end{lstlisting}
-we must start by evaluating the binding for $x$.  We can't evaluate inside the
+To achieve this behavior, we must start by evaluating the binding for $x$.  We can't evaluate inside the
 $\next$ statement, but must still substitute some result into the body of the

### 2014-06-19 paper/semantics.tex
Forgot to commit in previous checkin.
put the right value in
adding some code formatting

@@
 The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
-\footnote{We render $\fut$ as {\tt \$} in code examples.}
+%\footnote{We render $\fut$ as {\tt \$} in code examples.}
 
@@
 \begin{lstlisting} 
-[1+2/x1]
+[x1|->1+2]
 let x = (next{x1}, 7) in
@@
 \begin{lstlisting} 
-[1+2/x1]
+[x1|->1+2]
 	next{ x1 * x1 * 7 }
@@
 are the results of a computation after both stages have been completed. In the
-example above, the term evaluates to \verb|1215|.
+example above, the term evaluates to \verb|63|.
 \end{itemize}

### 2014-06-19 paper/paper.tex
fixed the build and made figures files
added new code examples for future explanatory use
Reworked file structure a little.
move figures to a subdirectory
splitting up figures into separate files
miscellaneous cleanup
major reorganizing/rewriting of section 3.
adding some code formatting
adding hats

@@
-\documentclass{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 \usepackage{listings}
+\usepackage{hyperref}
+
+\usepackage{cleveref}
+\AtBeginDocument{\renewcommand\ref\cref}
+\crefname{figure}{Figure}{Figures}
 
@@
 \newtheorem{definition}[theorem]{Definition}
+\newtheorem{remark}[theorem]{Remark}
+
+\input {macros}
 
@@
 
-\input {macros}
-\input {grammars}
-\input {abstract}
-\input {intro}
-\input {relatedwork}
-\input {semantics}
-\input {splitting}
+\lstset{language=ML,
+        columns=fullflexible,
+        basicstyle=\ttfamily,
+        tabsize=4,
+        escapeinside={"*}{*"},
+        literate={|->}{{\ $\mapsto$\ }}1
+                 {xhat}{{$\hat{\texttt{x}}$}}1
+                 {yhat}{{$\hat{\texttt{y}}$}}1
+                 {zhat}{{$\hat{\texttt{z}}$}}1
+                 {$}{{$\fut$}}1}
 
@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
+%\titlebanner{banner above paper title}        % These are ignored unless
+%\preprintfooter{short description of paper}   % 'preprint' option specified.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \input{intro}
+
+\input{semantics}
+
+\vfill\eject\pagebreak
+\input{splitting}
+
 \input{relatedwork}
-\input{semantics}
-\input{splitting}
 

### 2014-06-20 paper/semantics.tex
adding hats
iterating semantics explanation

@@
 stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. 
-Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
+Accordingly, a suitably staged dynamic semantics for \lang\ must evaluate
 \emph{all} the stage \bbone\ subexpressions of a term before its stage
@@
 
-To understand how this behavior can be achieved, consider evaluation of the stage-\bbone\ term below. (The term contains nested $\next$ and $\prev$ blocks.):
+\subsection{Non-Duplicating First-Stage Evaluation}
+
+To understand how this behavior can be achieved, consider evaluation of the stage-\bbtwo\ term below, which contains nested $\next$ and $\prev$ blocks:
 \begin{lstlisting} 
+prev{
   let x = (next {1+2}, 3+4) in
   next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
+}
 \end{lstlisting}
-One option is to duplicate the contents of the first $\next$ expression, to produce
+Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
+One option, taken by \cite{davies96}, is to treat $\next\{\mathtt{1+2}\}$ as a value for the purposes of stage-\bbone~evaluation, 
+consequently duplicating it during substitution for {\tt x}.  
+This produces:
 \begin{lstlisting}
@@
 
-Here we are essentially treating $\next\{e\}$ as a stage-\bbone\ value, so long
-as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
-\cite{davies96}.
-
-We take a different approach in our dynamic semantics, which is to let-bind the
-contents of the first $\next$ expression to some variable (here, $x1$) and
-duplicate that variable,
+We take a different approach.  
+Instead of duplicating the contents of the first $\next$ expression, we bind them to some variable (here, {\tt y}) and
+duplicate only a reference to that variable.  This produces:
 \begin{lstlisting} 
-let x1 = 1+2 in
-  x1 * x1 * 7
+let y = 1+2 in
+  y * y * 7
 \end{lstlisting}
@@
 
-To achieve this behavior, we must start by evaluating the binding for $x$.  We can't evaluate inside the
-$\next$ statement, but must still substitute some result into the body of the
-\texttt{let}. We replace the contents of $\next$ with a variable and make an
-explicit substitution, which then floats up to the top of the containing
-$\prev$:
+Achieving this behavior mechanically requires us to resolve a contradiction:
+we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the
+$\next$ block bound to it. Our solution is to replace the $\next$ with a new variable and create an
+explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s contents.  
+This substitution then floats up to the top of the containing $\prev$ block:
+\begin{lstlisting} 
+[y|->1+2]
+let x = (yhat, 7) in
+next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
+\end{lstlisting}
+Note that the variable left behind is rendered with a hat.  By treating this {\em hatted variable} as a value, we're now free to perform the stage-\bbone~substitution for {\tt x}.  
 \begin{lstlisting} 
 [x1|->1+2]
-let x = (next{x1}, 7) in
-next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
-\end{lstlisting}
-We're now free to perform the stage-\bbone~substitution for $x$.  
-\begin{lstlisting} 
-[1+2/x1]
 next{
-	prev{#1 (next{x1}, 7)} *
-	prev{#1 (next{x1}, 7)} *
+	prev{#1 (yhat, 7)} *
+	prev{#1 (yhat, 7)} *
 	prev{hold (#2 (next{x1}, 7))}
@@
 
-\TODO write the function here}
+%\begin{lstlisting} 
+%let adjust (f : $int * int -> $int) =
+%	fn (p : int) => 
+%		next{
+%			fn (b : int) => 
+%				prev{f (next {b}, p)}
+%		}	
+%\end{lstlisting}
+\TODO code written, but listings don't work in footnotes! commented out in source
+}
 Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied

### 2014-06-20 paper/semantics.tex
Finished up the example.
I had to add the prevs back in, making it stage 2.  Sorry, Kayvon.

@@
 \begin{lstlisting} 
-let y = 1+2 in
-  y * y * 7
+let y = 1+2 in y * y * 7
 \end{lstlisting}
@@
 Achieving this behavior mechanically requires us to resolve a contradiction:
-we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the
-$\next$ block bound to it. Our solution is to replace the $\next$ with a new variable and create an
-explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s contents.  
+we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
+Our solution is to replace the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s contents.  
 This substitution then floats up to the top of the containing $\prev$ block:
 \begin{lstlisting} 
+prev {
 [y|->1+2]
   let x = (yhat, 7) in
-next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
+  next{prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}}
+}
 \end{lstlisting}
@@
 \begin{lstlisting} 
-[x1|->1+2]
+prev {
+[y|->1+2]
   next{
-	prev{#1 (yhat, 7)} *
-	prev{#1 (yhat, 7)} *
-	prev{hold (#2 (next{x1}, 7))}
+    prev{#1 (yhat, 7)} * prev{#1 (yhat, 7)} *
+    prev{hold (#2 (yhat, 7))}
+  }
 }
 \end{lstlisting}
-To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
+To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them. 
+As a rule, these will reduce to hatted variables, and $\prev$ removes the hat, leaving the variable in place:
 \begin{lstlisting} 
-[x1|->1+2]
-	next{ x1 * x1 * 7 }
+prev {
+[y|->1+2]
+    next{ y * y * 7 }
+}
 \end{lstlisting}
-The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
+Once again, we lift the contents of the $\next$ into a substitution:
 \begin{lstlisting} 
-let x1 = 1+2 in
-  x1 * x1 * 7
+prev {
+[y|->1+2]
+[z|->y*y*7]
+    zhat
+}
+\end{lstlisting}
+Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
+\begin{lstlisting} 
+let y = 1+2 in
+let z = y * y * 7 in z
 \end{lstlisting}
 
-Once again, the point here is that we did not duplicate the $1+2$ in our residual.  
-Instead we lifted it out into a substitution, leaving a hatted variable [\TODO add hats above] in its place,
-and we reified the substitution into a let-binding at the enclosing $\prev$.
+Thus we have evaluated all of the stage \bbone\ expressions of this program without duplicating the contents of $\next$ blocks.
 

### 2014-06-20 paper/semantics.tex
More on judgements.

@@
 Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
-One option, taken by \cite{davies96}, is to treat $\next\{\mathtt{1+2}\}$ as a value for the purposes of stage-\bbone~evaluation, 
+One option, taken by \cite{davies96}, is to treat \verb|next{1+2}| as a value for the purposes of stage-\bbone~evaluation, 
 consequently duplicating it during substitution for {\tt x}.  
@@
 \begin{itemize}
-\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
+\item 
+Partial values ($\pval$s) are stage \bbone\ terms that have been fully evaluated, 
+but which may contain hatted variables. In the example above, 
+\verb|(|$\hat {\tt y}$\verb|,7)|.
+\item Residuals ($\res$es) are stage \bbtwo\ terms whose stage \bbone\
 subexpressions have all been fully evaluated. In the example above,
-\verb|next {x1 * x1 * 7}|.
-\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
-subexpressions have all been fully evaluated. In the example above,
-\verb|1+2|.
-\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
+\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
+\item Values ($\val$s) are stage \bbtwo\ terms which are fully evaluated; these
 are the results of a computation after both stages have been completed. In the
@@
 \end{itemize}
-The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
-environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
-and a partial value $v$ which may refer to those variables. This judgment
-necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
-\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
-implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
-residual.
+The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
+The future environment is a mapping from newly-created variables 
+(which then may appear---hatted---in $v$)
+to residuals.  It is the formal manifestation of the floating substitutions in our example.
+This judgment, upon encountering a stage \bbtwo\ subexpression, must necessarily search into it in order to find its stage
+\bbone\ subexpressions; we call this process \emph{speculation}. 
+Speculation is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
 
-\TODO finish explaining how the nested evaluation actually works.
-explain reification, hatted variables, and contexts.
+\TODO next
+
+\TODO prev
+
+\TODO reification
+
+\TODO hold
+
+\TODO contexts
 
 A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
-given a type of the form $A\to\fut(B\to C)$.%
+given a type of the form $A\to\fut(B\to C)$%
 \footnote{We can rewrite \texttt{fexp} in this form, or simply apply the following
@@
 %\end{lstlisting}
-\TODO code written, but listings don't work in footnotes! commented out in source
-}
+\TODO code written, but listings don't work in footnotes! commented out in source until we get working
+}.
 Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
@@
 hoisting them in \verb|let|-bindings.
+From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
+But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
+our semantics terminates strictly less often than that of \cite{davies96}.
 

### 2014-06-20 paper/semantics.tex
finished a rough draft of everything before metatheory

@@
 to residuals.  It is the formal manifestation of the floating substitutions in our example.
-This judgment, upon encountering a stage \bbtwo\ subexpression, must necessarily search into it in order to find its stage
-\bbone\ subexpressions; we call this process \emph{speculation}. 
-Speculation is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
+For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
+and the final future environment is gotten by merging the future environments of the subterms.
 
-\TODO next
+When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it searches into the block's stage \bbtwo\ content to find any contained stage
+\bbone\ subexpressions and evaluate them in place.  
+We call this search process \emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
+Once the contents of the $\next$ block are speculated into a residual ($q$), 
+we then return a fresh hatted variable ($\hat y$), 
+along with a future environment which maps that variable to our residual ($y \mapsto q$).
 
-\TODO prev
+At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does nothing but recursively speculate into every subexpression.
+Once speculation finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which produces a future context and (by canonical forms) a hatted variable ($\hat y$).
+The context is then reified (using the $\reifysym$ judgment) into a series of let bindings, with $y$---stripped of its hat---at the bottom.
 
-\TODO reification
+Because the inputs to evaluation and speculation are sometimes open on variables, we adorn our judgments with two contexts.
+The first of these ($\Xi$) keeps track of those hatted variables that may appear in the input.
+Since hatted variables can only be introduced to terms via substitution ({\em i.e.} they do not appear in the original program), 
+this context is augmented only at stage \bbone\ substitution sites.
+The second context ($\Gamma$) keeps track of stage \bbtwo\ variables, which are declared in original program.
+% We chose to keep these as two separate contexts, because only $\Xi$ is needed for well-formedness of partial values.
 
-\TODO hold
+If we change the $\next$ and $\prev$ rules to 
+\begin{mathpar}
+\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
+\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
+\end{mathpar}
+and treat $\next~q$ as a partial value, where $q$ is a residual,
+then we get precisely the semantics from \cite{davies96}. This essentially bypasses the
+environment bookkeeping in $\redonesym$, by inlining residuals instead of
+hoisting them in \verb|let|-bindings.
+From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
+But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
+our semantics terminates strictly less often than that of \cite{davies96}.
 
-\TODO contexts
+\subsection {A Partial Evaluation System}
 
@@
 }.
-Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
+Given its stage \bbone\ argument $a:A$, we can evaluate the partially-applied
 function:
@@
 enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
-ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
+ensures that the resulting $f_a$ has type $B\to C$ in the residual language.
 Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
@@
 \end{remark}
-
-\TODO 
-If we change the highlighted rules to something slightly different, then we get
-precisely the semantics from \cite{davies96}. This essentially bypasses the
-environment bookkeeping in $\redonesym$, by inlining residuals instead of
-hoisting them in \verb|let|-bindings.
-From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
-our semantics terminates strictly less often than that of \cite{davies96}.
 

### 2014-06-20 paper/paper.tex
adding listings stuff

@@
                  {zhat}{{$\hat{\texttt{z}}$}}1
+                 {murec}{{$\mu$}}1
+                 {alpha}{{$\alpha$}}1
+                 {beta}{{$\beta$}}1
                  {$}{{$\fut$}}1}

### 2014-06-22 paper/splitting.tex
Started adding some arguments regarding splitting.

@@
 \section{Splitting Algorithm}
+
+Motivate splitting here.  (Depends on intro)
+Consider splitting at \bbtwo. 
+The basic idea is that, for a given input program, we want to find all of the stage \bbone\ content, pull that out into a single term, 
+and also produce a stage \bbtwo\ term, which takes the results of the stage \bbone\ part and produces the final answer.
+We call these two parts the {\em precomputation} and the {\em residual}, respectively.
+More formally, we want to take some $\coltwo e A$ and produce a $p : \tau$ and a $r : \tau \to A$ such that $e$ and $(r~p)$ reduce to the same value.
+Consider the following stage \bbtwo\ example:
+\begin{lstlisting}
+if 6*7 > 41 then
+	hold{1+2} * 3
+else 
+	hold{2*3} + hold {4*5}
+\end{lstlisting}
+Under the definitions above, this should split into:
+\begin{lstlisting}
+(1+2,2*3,4*5)
+\end{lstlisting}
+and
+\begin{lstlisting}
+fn (l1,l2,l3) => 
+    if 6*7 > 41 then l1 * 3 else l2 + l3
+\end{lstlisting}
+Essentially what we've done is find all of the $\pause$ expressions, 
+bundle them into a tuple in then precomputation, 
+and then unbundle them into the correct place in the residual.
+Note how splitting has the same speculative behavior as our \lang\ dynamics,
+wherein we run the stage \bbone\ code in {\em both} branches of the stage \bbtwo\ {\tt if} expression.
+
+Now consider splitting the following expression, where the {\tt if} is at stage \bbone, rather than stage \bbtwo:
+\begin{lstlisting}
+prev {
+  if 6*7 > 41 then
+    next{1+hold{2*3}}
+  else 
+    next{4+5}
+}
+\end{lstlisting}
+This splits into
+\begin{lstlisting}
+if 6*7 > 41 then inL (2*3) else inR ()
+\end{lstlisting}
+and
+\begin{lstlisting}
+fn l => 
+  case l of
+    l1 => 1+l1
+  | () => 4+5
+\end{lstlisting}
+In this example, we
+
+[develope the more formal version...]
+
+  $\diatwo [\cdot;\cdot] e q$ and $\reduce q v$ iff $\reduce {r~p} {v}$.
+
+\input{figures/splitting-term}
+\input{figures/splitting-sumfun}
 

### 2014-06-23 paper/splitting.tex
factored example section into its own file

@@
 We leave many-to-one as future-work.
-
-\section{Examples of Algorithm Derivation}
-
-Fast exponent example.
-
-\begin{lstlisting} 
-let exp (b : $int, e : int) : $int =
-	if e == 0 then
-		next{1}
-	else if (e mod 2) == 0 then
-		next{let x = prev{exp(b,e/2)} in x*x}
-	else
-		next{prev{b} * prev{exp (b,e-1)}}		
-\end{lstlisting}
-
-splits into
-
-\begin{lstlisting} 
-let exp (b, e) =
-	((), roll (
-		if e == 0 then
-			inL ()
-		else 
-			inR (
-				if (e mod 2) == 0 then
-					inL (#2 (exp (b,e/2)))
-				else
-					inR (#2 (exp (b,e-1)))
-			)
-	))
-\end{lstlisting}
-
-and
-
-\begin{lstlisting} 
-let exp ((b, e), p) =
-	case unroll p of
-	  () => 1
-	| d  =>
-		case d of
-		  r => let x = exp ((b,()),r) in x*x
-		| r => b * exp ((b,()),r)
-\end{lstlisting}
-
-Quickselect example.
-
-\begin{lstlisting} 
-let qs (l : "*$\mathtt{\mu \alpha. }$*"() + "*$\mathtt{int*\alpha}$*", i: $int) = 
-	case unroll l of
-	  () => next {0}
-	| (h,t) => 
-		let (left,right,n) = partition h t in
-		next{
-			let n = prev{hold n} in
-			case compare prev{i} n of
-			  () (*LT*) => prev {qs left i}
-			| () (*EQ*) => prev {hold h}
-			| () (*GT*) => 
-				prev {qs right next{prev{i}-n-1}}
-		}	
-\end{lstlisting}
-
-Things to try: an interpreter which, partially evaluated, does cps or something.
-
-For each of these examples, talk about what partial evaluation would do and why that might be bad.
-
-[Meta-ML eases off on this restriction but does not (I think?) eliminate it.]
-
-[What's going on with names and necessity?]
-
-[Our work bears a lot of similarity to ML5, which also uses a modal type system.
-The difference is that we target stages systems (each stage talks to the next), whereas they target distributed ones (all stages talk to all others).
-The type systems reflect this directly in the world accessibility relation.
-There might be some analogue of stage-splitting in the ML5 work, but I have not yet isolated it (might be buried in CPS conversion).]

### 2014-06-23 paper/splitting.tex
Added more about splitting.

@@
 More formally, we want to take some $\coltwo e A$ and produce a $p : \tau$ and a $r : \tau \to A$ such that $e$ and $(r~p)$ reduce to the same value.
+
+\subsection {Splitting If Statements}
+
 Consider the following stage \bbtwo\ example:
@@
 \end{lstlisting}
-In this example, we
+Here, there's no need to speculate.
+Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject the result into a sum type.
+Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.
 
-[develope the more formal version...]
+\subsection {Splitting at the First Stage}
+
+Consider the example from section \ref{nonDup}:
+
+\begin{lstlisting} 
+prev{
+  let x = (next {1+2}, 3+4) in
+  next{prev{#1 x} * prev{#1 x} * prev{hold(#2 x)}}
+}
+\end{lstlisting}
+
+It's not clear whether the {\tt x} binding should be thought of as happening at stage \bbone\ or stage \bbtwo.
+The variable itself is of course at stage \bbone, but the expression that is bound to it has some stage \bbtwo\ content, namely \verb|1+2|.
+The solution is to think of {\tt x} as existing at both stages, and so it appears in the precomputation {\em and} the residual.
+Here's the precomputation:
+\begin{lstlisting} 
+let x = ((), 3+4) in 
+#2 x
+\end{lstlisting}
+and the residual:
+\begin{lstlisting} 
+fn l =>
+  let x = (1+2, ()) in 
+  (#1 x) * (#1 x) * l
+\end{lstlisting}
+What we see is that {\tt x}, which has type $\rmint \times \fut \rmint$ in the original program, 
+has types $\rmint \times \rmint$ in the precomputation and $\rmunit \times \rmint$ in the residual.
+
+\subsection {Splitting Functions}
+
+\TODO functions
+
+\subsection {Formal Setup}
+
+[develop the more formal version...]
 
@@
 
+[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
+
+\subsection {Boundary Type Worst Case}
+
+\TODO add the example which is the worst case for figuring out boundary types
+
 \input{figures/splitting-term}
 \input{figures/splitting-sumfun}
-
-[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
 

### 2014-06-24 paper/splitting.tex
Few more notes on splitting.

@@
 \end{lstlisting}
-What we see is that {\tt x}, which has type $\rmint \times \fut \rmint$ in the original program, 
-has types $\rmint \times \rmint$ in the precomputation and $\rmunit \times \rmint$ in the residual.
+What we see is that {\tt x}, which has type $A = \rmint \times \fut \rmint$ in the original program, 
+has types $A_1 = \rmint \times \rmint$ in the precomputation and $A_2 = \rmunit \times \rmint$ in the residual.
+We call $A_1$ and $A_2$ the first and second {\em images} of $A$.
 
@@
 
-\TODO functions
+The trick when splitting functions is that the contents themselves may be multi-stage.
+Thus the boundary type appears as output of the first stage and input of the second stage.
+This makes it hard to type the output, because the boundary types are not represented in the original function type.
 
 \subsection {Formal Setup}
+
+We express residuals as functions in the informal development above.
+For reasons of concision, we equivalently represent residuals as terms open on a variable in the formal judgements.
+
+The $\splitonesym$ judgment takes an open stage \bbone\ term to a combined term and a residual.
+The $\splittwosym$ judgment takes an open stage \bbtwo\ term to a precomputation and a residual.
 

### 2014-06-26 paper/semantics.tex
changed defn of hold

@@
   let x = (next {1+2}, 3+4) in
-  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
+  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
 }
@@
   let x = (yhat, 7) in
-  next{prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}}
+  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
 }
@@
     prev{#1 (yhat, 7)} * prev{#1 (yhat, 7)} *
-    prev{hold (#2 (yhat, 7))}
+    hold{#2 (yhat, 7)}
   }
@@
 %if 5*4*3*2 > 111 then
-%	prev{ hold (2+4) }
+%	hold{2+4}
 %else

### 2014-06-28 paper/splitting.tex
Bunch of changes to figures.
changed defn of hold
worked on intro to splitting. need to merge with uncommitted stuff at home

@@
 
-Motivate splitting here.  (Depends on intro)
-Consider splitting at \bbtwo. 
-The basic idea is that, for a given input program, we want to find all of the stage \bbone\ content, pull that out into a single term, 
-and also produce a stage \bbtwo\ term, which takes the results of the stage \bbone\ part and produces the final answer.
-We call these two parts the {\em precomputation} and the {\em residual}, respectively.
-More formally, we want to take some $\coltwo e A$ and produce a $p : \tau$ and a $r : \tau \to A$ such that $e$ and $(r~p)$ reduce to the same value.
+The basic idea is that we want to take a term which contains interleaved stage \bbone\ and stage \bbtwo\ code
+and untangle it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.
+To facilitate communication between the stages, there is a data structure passed between the stages, 
+which is an output of the stage \bbone\ term and and input to the stage \bbtwo\ term.
+
+As with the dynamics, the form of the splitting statement depends on the external stage of the term.
+We consider splitting for stage \bbtwo\ terms first, as it has a simpler form.  
+For example, take \verb|hold{1+2} < 5|.
+We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
+which are called the {\em precomputation} and {\em residual}, respectively.  
+For this splitting to be correct, the residual applied to the precomputation should yield the same result as the original term.
+
+The form of splitting for stage \bbone\ terms is more complicated.
 
@@
   let x = (next {1+2}, 3+4) in
-  next{prev{#1 x} * prev{#1 x} * prev{hold(#2 x)}}
+  next{prev{#1 x} * prev{#1 x} * hold{#2 x}}
 }
@@
 \input{figures/splitting-term}
-\input{figures/splitting-sumfun}
 

### 2014-06-30 paper/splitting.tex
reorganized splitting section

@@
 The form of splitting for stage \bbone\ terms is more complicated.
-
-\subsection {Splitting If Statements}
-
-Consider the following stage \bbtwo\ example:
+To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
+As before, the comparison operation is part of stage \bbtwo, 
+and the addition operation is a stage \bbone\ precomputation, the result of which will eventually become input to the residual.
+The multiplication operation, however, is neither part of the residual nor a precomputation to support it.
+Instead it is the {\em immediate result} of stage \bbone, 
+because it is available to the stage \bbone\ context around our original term.
+For example:
 \begin{lstlisting}
-if 6*7 > 41 then
-	hold{1+2} * 3
-else 
-	hold{2*3} + hold {4*5}
+let x = (next{hold{1+2} < 5},3*4) in
+next{ if prev{#1 x} then 7 else hold {#2 x} }
 \end{lstlisting}
-Under the definitions above, this should split into:
-\begin{lstlisting}
-(1+2,2*3,4*5)
-\end{lstlisting}
-and
-\begin{lstlisting}
-fn (l1,l2,l3) => 
-    if 6*7 > 41 then l1 * 3 else l2 + l3
-\end{lstlisting}
-Essentially what we've done is find all of the $\pause$ expressions, 
-bundle them into a tuple in then precomputation, 
-and then unbundle them into the correct place in the residual.
-Note how splitting has the same speculative behavior as our \lang\ dynamics,
-wherein we run the stage \bbone\ code in {\em both} branches of the stage \bbtwo\ {\tt if} expression.
-
-Now consider splitting the following expression, where the {\tt if} is at stage \bbone, rather than stage \bbtwo:
-\begin{lstlisting}
-prev {
-  if 6*7 > 41 then
-    next{1+hold{2*3}}
-  else 
-    next{4+5}
-}
-\end{lstlisting}
-This splits into
-\begin{lstlisting}
-if 6*7 > 41 then inL (2*3) else inR ()
-\end{lstlisting}
-and
-\begin{lstlisting}
-fn l => 
-  case l of
-    l1 => 1+l1
-  | () => 4+5
-\end{lstlisting}
-Here, there's no need to speculate.
-Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject the result into a sum type.
-Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.
-
-\subsection {Splitting at the First Stage}
-
-Consider the example from section \ref{nonDup}:
-
-\begin{lstlisting} 
-prev{
-  let x = (next {1+2}, 3+4) in
-  next{prev{#1 x} * prev{#1 x} * hold{#2 x}}
-}
-\end{lstlisting}
-
-It's not clear whether the {\tt x} binding should be thought of as happening at stage \bbone\ or stage \bbtwo.
-The variable itself is of course at stage \bbone, but the expression that is bound to it has some stage \bbtwo\ content, namely \verb|1+2|.
-The solution is to think of {\tt x} as existing at both stages, and so it appears in the precomputation {\em and} the residual.
-Here's the precomputation:
-\begin{lstlisting} 
-let x = ((), 3+4) in 
-#2 x
-\end{lstlisting}
-and the residual:
-\begin{lstlisting} 
-fn l =>
-  let x = (1+2, ()) in 
-  (#1 x) * (#1 x) * l
-\end{lstlisting}
-What we see is that {\tt x}, which has type $A = \rmint \times \fut \rmint$ in the original program, 
-has types $A_1 = \rmint \times \rmint$ in the precomputation and $A_2 = \rmunit \times \rmint$ in the residual.
-We call $A_1$ and $A_2$ the first and second {\em images} of $A$.
-
-\subsection {Splitting Functions}
-
-The trick when splitting functions is that the contents themselves may be multi-stage.
-Thus the boundary type appears as output of the first stage and input of the second stage.
-This makes it hard to type the output, because the boundary types are not represented in the original function type.
+To support this, stage \bbone\ splitting produces two outputs: the {\em combined result} and the residual,
+where the combined result reduces to a tuple containing the immediate result and the precomputation.
 
@@
 
-We express residuals as functions in the informal development above.
-For reasons of concision, we equivalently represent residuals as terms open on a variable in the formal judgements.
+We now discuss the formal definition of splitting.
+The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$.
+The $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
+while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
+For concision, we actually represent residuals as a term open on a single variable, rather than as functions.
+For both judgments, we use a context ($\Gamma$) to keep track of the open variables of $e$.
 
-The $\splitonesym$ judgment takes an open stage \bbone\ term to a combined term and a residual.
-The $\splittwosym$ judgment takes an open stage \bbtwo\ term to a precomputation and a residual.
+The rules\footnote{The rules are written using patterns, including the open variable of the residual.} 
+for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
+The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
+while the rule for $\prev$ projects the combined result of its subexpression to just the precomputation.
+The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
+and projects out the integer result in the residual\footnote{The residual of an integer expression is usually trivial, 
+but we have to include it here for termination purposes.}
 
-[develop the more formal version...]
+The rules of stage \bbtwo\ splitting are given in \ref{fig:termSplitTwo}.  
+Every rule works by bundling the precomputations of the constituent parts, and then unbundling them with a pattern.
+The rules of stage \bbone\ splitting are given in \ref{fig:termSplitOne}.
 
-  $\diatwo [\cdot;\cdot] e q$ and $\reduce q v$ iff $\reduce {r~p} {v}$.
+\subsection {Speculation}
 
-[Present the splitting judgement.  Give statements of type and value correctness for splitting.  Give all of the splitting rules.  Talk through a few of them.]
+Notice in the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the precomputation is lifted out from within branches.
+This is the manifestation of the speculation behavior from the semantics.
+
+\subsection {Stage \bbone\ Divergence}
+Consider the case of stage \bbone\ {\tt if} and {\tt case} expressions.
+In both cases we have enough information at stage \bbone\ to know what branch to take, so there's no need to speculate.
+Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject precomputation into a sum type.
+Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.
+
+\subsection {Stage \bbone\ Functions}
+
+The trick when splitting stage \bbone\ functions is that the contents themselves may be multi-stage.
+We handle this by splitting them into two functions:
+one in the immediate result which handles all the stage \bbone\ content of the original,
+and one in the residual which handles all of the stage \bbtwo\ content.
+Note that stage \bbone\ $\lambda$-expressions themselves have only a trivial precomputation.
+
+Thus the boundary type appears as output of the first stage and input of the second stage.
+This makes it hard to type the output, because the boundary types are not represented in the original function type.
 

### 2014-07-01 paper/semantics.tex
start work on metatheory

@@
 
-% TODO move these to macros.tex once they're finalized
-\newcommand{\pval}{\ensuremath{\mathsf{pval}}}
-\newcommand{\res}{\ensuremath{\mathsf{res}}}
-\renewcommand{\val}{\ensuremath{\mathsf{val}}}
-
 The algorithm described above creates three different kinds of expressions which
@@
 \item 
-Partial values ($\pval$s) are stage \bbone\ terms that have been fully evaluated, 
+Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
 but which may contain hatted variables. In the example above, 
 \verb|(|$\hat {\tt y}$\verb|,7)|.
-\item Residuals ($\res$es) are stage \bbtwo\ terms whose stage \bbone\
+\item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
 subexpressions have all been fully evaluated. In the example above,
 \verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
-\item Values ($\val$s) are stage \bbtwo\ terms which are fully evaluated; these
+\item Values ($\valsym$s) are stage \bbtwo\ terms which are fully evaluated; these
 are the results of a computation after both stages have been completed. In the
@@
 
-% local macros; put them in macros.tex once they are more finalized
-\newcommand{\wf}{\ \mathsf{wf}}
-\newcommand{\valoo}{\ \bbone\textsf{-}\mathsf{val} @ \bbone}
-\newcommand{\valot}{\ \bbone\textsf{-}\mathsf{val} @ \bbtwo}
-\newcommand{\valtt}{\ \bbtwo\textsf{-}\mathsf{val} @ \bbtwo}
-
 \begin{definition}
@@
 \item $\xi = \xi',x\mapsto e$ where
-$\Xi,\dom{\xi'};\Gamma\vdash e : B @ \bbtwo$ and
-$\Xi,\dom{\xi'};\Gamma\vdash e \valot$.
+$\xi'\wf$,
+$\Xi,\dom{\xi'};\Gamma\vdash \coltwo{e}{B}$, and
+$\Xi,\dom{\xi'};\Gamma\vdash e \res$.
 \end{enumerate}
@@
 \begin{theorem}
-If $\diaonesub$ then:
+If $\diaonesub$ and $\typesone e A$ then
 \begin{enumerate}
-\item $\typesone e A$ for some 1-type $A$;
 \item $\xi\wf$;
-\item $\Xi,\dom\xi;\cdot\vdash v : A @ \bbone$; and
-\item $\Xi,\dom\xi;\cdot\vdash v\valoo$.
+\item $\Xi,\dom\xi;\cdot\vdash \colone{v}{A}$; and
+\item $\Xi,\dom\xi;\cdot\vdash v\pval$.
 \end{enumerate}
@@
 \begin{theorem}
-If $\diatwosub$ then:
+If $\diatwosub$ and $\typestwo e A$ then
 \begin{enumerate}
-\item $\typestwo e A$ for some 2-type $A$;
-\item $\Xi;\Gamma\vdash q : A @ \bbtwo$; and
-\item $\Xi;\Gamma\vdash q\valtt$.
+\item $\Xi;\Gamma\vdash \coltwo{q}{A}$; and
+\item $\Xi;\Gamma\vdash q\val$.
 \end{enumerate}

### 2014-07-01 paper/semantics.tex
fix the issue with listings in footnotes

@@
 A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
-given a type of the form $A\to\fut(B\to C)$%
-\footnote{We can rewrite \texttt{fexp} in this form, or simply apply the following
-higher-order function which makes the adjustment:
-
-%\begin{lstlisting} 
-%let adjust (f : $int * int -> $int) =
-%	fn (p : int) => 
-%		next{
-%			fn (b : int) => 
-%				prev{f (next {b}, p)}
-%		}	
-%\end{lstlisting}
-\TODO code written, but listings don't work in footnotes! commented out in source until we get working
-}.
+given a type of the form $A\to\fut(B\to C)$.%
+\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
+the following higher-order function which makes the adjustment:
+\begin{lstlisting} 
+let adjust (f : $int * int -> $int) =
+  fn (p : int) => 
+    next{
+      fn (b : int) => 
+        prev{f (next {b}, p)}
+    }
+\end{lstlisting}}
+%
 Given its stage \bbone\ argument $a:A$, we can evaluate the partially-applied

### 2014-07-03 paper/intro.tex
fixed brackets
wrote a bit of intro

@@
 
-We often find ourselves in situations where the input to a computation arrives in two parts, at different points in time.
+It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).
+This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).
+Examples include,
+\begin{itemize}
+\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
+\item Computing the shading of a surfacing in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over $4\times4$ pixel blocks.  
+\item Running a language interpreter on a single program for multiple inputs.
+\end{itemize}
 
-The natural response is to try and do some part of the computation early, and the rest of the computation once the second input arrives.
+The goal is to run programs in a way that doesn't repeat work unless it depends on the varying input.
+There are two separate problems at play here.
+The first is {\em binding-time analysis}, the problem of determining {\em what} portions of the program to run statically versus dynamically.
+After that is {\em staged execution}, the problem of {\em how} to execute in a staged way.
+We  consider only the latter in this paper.
+That is, we assume that programs are given in a representation, covered later, that fully determines staged execution.
 
-Since the computation runs in two parts, the total cost can be split as well, into $m+n$.
+We consider two possible models of staged execution,
 
-Additionally, we can run the second stage multiple times.  Cost is now $m+bn$.
+\TODO partial eval
+
+\TODO stage-splitting
 
@@
 \begin{align*}
-[f_1](x) &= p (f,x) \\
-[f_2](l,y) &= [l] (y)
+\llbracket f_1 \rrbracket (x) &= p (f,x) \\
+\llbracket f_2 \rrbracket (l,y) &= [l] (y)
 \end{align*}

### 2014-07-18 paper/semantics.tex
A few changes to metatheory sections.

@@
 %The input to evaluation is a stage-1 expression (usually
-%$e$), as well as two administrative contexts ($\Gamma$ and $\Xi$), covered
+%$e$), as well as two administrative contexts ($\Gamma$), covered
 %later.  Evaluation has two outputs: the {\em future context} (usually $\xi$) and
@@
 
-Because the inputs to evaluation and speculation are sometimes open on variables, we adorn our judgments with two contexts.
-The first of these ($\Xi$) keeps track of those hatted variables that may appear in the input.
+Because the inputs to evaluation and speculation are sometimes open on variables, we adorn our judgments with a context.
 Since hatted variables can only be introduced to terms via substitution ({\em i.e.} they do not appear in the original program), 
@@
 The second context ($\Gamma$) keeps track of stage \bbtwo\ variables, which are declared in original program.
-% We chose to keep these as two separate contexts, because only $\Xi$ is needed for well-formedness of partial values.
 
@@
 
-\noindent\TODO terminology for the contexts?
-
 \begin{definition}
-Contexts $\Xi$ and $\Gamma$ are well-formed ($\Xi\wf$, $\Gamma\wf$) if they
-contain only stage-2 variables.
+Contexts $\Gamma$ is well-formed ($\Gamma\wf$) if it
+contains only stage-2 variables.
 \end{definition}
@@
 \begin{definition}
-An environment $\xi$ is well-formed ($\xi\wf$) if either:
+An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
 \begin{enumerate}
@@
 \item $\xi = \xi',x\mapsto e$ where
-$\xi'\wf$,
-$\Xi,\dom{\xi'};\Gamma\vdash \coltwo{e}{B}$, and
-$\Xi,\dom{\xi'};\Gamma\vdash e \res$.
+$\Gamma\vdash\xi'\wf$ and
+$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
+$\Gamma,\dom{\xi'}\vdash e \res$.
 \end{enumerate}
@@
 \begin{theorem}
-If $\Xi;\typesone e A$ then $\Xi\wf$, $\Gamma\wf$, and $A\istypeone$.
+If $\typesone e A$ then $\Gamma\wf$ and $A\istypeone$.
 \end{theorem}
@@
 \begin{enumerate}
-\item $\xi\wf$;
-\item $\Xi,\dom\xi;\cdot\vdash \colone{v}{A}$; and
-\item $\Xi,\dom\xi;\cdot\vdash v\pval$.
+\item $\Gamma\vdash\xi\wf$;
+\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
+\item $\Gamma,\dom\xi\vdash v\pval$.
 \end{enumerate}
@@
 \begin{enumerate}
-\item $\Xi;\Gamma\vdash \coltwo{q}{A}$; and
-\item $\Xi;\Gamma\vdash q\val$.
+\item $\Gamma\vdash \coltwo{q}{A}$; and
+\item $\Gamma\vdash q\val$.
 \end{enumerate}

### 2014-07-23 sematics/boundary/boundary.tex
Adding a few files for boundary types.
Will fold these into the paper when we decide that we want to include
them.
more adjustments to boundary types; need to consult
added some notes about boundary types

--- 2014-06-06 paper/paper.tex
@@
-\documentclass{sigplanconf}
+\documentclass[]{article}
 
@@
 \usepackage{amsthm}
-\usepackage{bbm}
 \usepackage{stmaryrd}
@@
 \usepackage{listings}
+\usepackage{graphicx}
 
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{definition}[theorem]{Definition}
+\title{Boundary Types}
+\author{Nico}
+
+\newcommand {\next}{asdlfkj}
+\input {../../paper/macros}
 
 \begin{document}
-
-\input {macros}
-\input {grammars}
-\input {abstract}
-\input {intro}
-\input {relatedwork}
-\input {semantics}
-\input {splitting}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Stage-Splitting a Modal Language}
-
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
-
 \maketitle
 
-\begin{abstract}
-\input{abstract}
-\end{abstract}
+\newcommand {\ktype} {\rm Type}
+\newcommand {\ibk} [2] {{#1}~\rotatebox[origin=c]{270}{$\therefore$}~{#2}}
+\newcommand {\btcont} [3] {{#1}~\triangleright~[{#2},{#3}]}
+\newcommand {\bty} [4] [\Gamma] {{#1}\vdash {#2}\triangleright[{#3},{#4}]}
+\newcommand {\btysub} [2] [] {\bty{e_{#1}:{#2}}{\tau_{#1}}{\sigma_{#1}}}
 
-\category{CR-number}{subcategory}{third-level}
+Consider a language with products and functions.  The internal boundary kind is given by:
 
-% general terms are not compulsory anymore, 
-% you may leave them out
-\terms
-term1, term2
+\begin{mathpar}
+\infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
+\infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
+\infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
+\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
+\end{mathpar}
 
-\keywords
-keyword1, keyword2
+For a statement $e$, the judgment $\btcont {e:A} \tau \sigma$ means that $e$ has the internal boundary type $\tau$ and the external boundary type-constructor $\sigma$.
+Invariant: for $\ibk A \kappa$, if $\btcont {e:A} \tau \sigma$ then $\sigma : \kappa$.
 
-\input{intro}
-\input{relatedwork}
-\input{semantics}
-\input{splitting}
-
-%\appendix
-%\section{Appendix Title}
-%This is the text of the appendix, if you need one.
-
-%\acks
-%Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\bibliography{paper}
-\bibliographystyle{abbrvnat}
+\begin{mathpar}
+\infer {\bty {i:\rm int}\rmunit {()}} {\cdot} \and
+\infer {\bty {b:\rm bool}\rmunit {()}} {\cdot} \and
+\infer {\bty {\pause~e:\fut\rm int}{\tau\times{\rm int}} {()}} {\bty {e:\rm int}\tau {()}} \and
+\infer {\bty [\Gamma,\btcont {x:A} \rmunit \sigma] {x:A}\rmunit\sigma} {\cdot} \and
+\infer 
+	{\bty {(e_1,e_2):A\times B} {\tau_1 \times \tau_2} {(\sigma_1, \sigma_2)}} 
+	{\btysub [1] A & \btysub [2] B} 
+	\\
+\infer 
+	{\bty {\pio~e:A} \tau {(\pi_1\sigma)}} 
+	{\btysub {A\times B}} \and
+\infer 
+	{\bty {\pit~e:B} \tau {(\pi_2\sigma)}} 
+	{\btysub {A\times B}} 
+	\\
+\infer 
+	{\bty {\lam x A e : A \to B} \rmunit {(\lambda \alpha:\kappa.(\tau,\sigma))}} 
+	{\ibk A \kappa & \bty [\Gamma, \btcont {x:A} \rmunit \alpha] {e:B} \tau \sigma} 
+	\and
+\infer 
+	{\bty {e_1~e_2 : B}
+		{\tau_1 \times \tau_2 \times \pi_1 (\sigma_1~\sigma_2)} 
+		{(\pi_2(\sigma_1~\sigma_2))} }
+	{\btysub [1] {A \to B} & \btysub [2] A } 
+\end{mathpar}
 

### 2014-07-24 paper/semantics.tex
increased exposition in intro

@@
 used as the basis for defining and analyzing stage splitting.
+
 %The grammar and type system for \lang\ is given in
 %\ref{fig:grammar,fig:statics}.
-Every expression in \lang\ has a statically known {\em stage},
-either \bbone~or \bbtwo, which defines when it is fully reduced.
-%\footnote{Typing judgments and context variables are annotated with stages.}
-%
-% in addition to a type.
-% Just as a type describes \emph{which} values an expression reduces to,
-%the expression's stage tells us \emph{when} it will be fully reduced. 
-% FIXME why?
-%The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
-Stage-\bbone\ and stage-\bbtwo\ types include products, sums, functions, and base types.
+Every expression in \lang\ has both a type and a {\em stage}, either \bbone~or \bbtwo.
+Just as a type describes \emph{what} sort of value an expression reduces to, the stage tells us \emph{when} an expression will be fully reduced.  
+Within each stage, \lang\ reduces to the simply-typed lambda calculus augmented with products, sums, general recursion, and base types
+\footnote{It needn't be the case that we choose the same set of features at both levels, but this does simplify the exposition.}.
+
 In addition to these types, stage-\bbone\ expressions may also be of type $\fut A$,
-representing an enclosed stage-\bbtwo\ expression of type $A$.
-(The type system of \lang\ is adopted directly from that of the modal language \lamCircle\,\cite{davies96}, but is restricted to two stages for simplicity.) $\fut A$ terms must be treated opaquely by stage-\bbone\ code
-as they cannot be reduced (to values with type $A$) until stage \bbtwo. Like in \lamCircle\, there is no reciprocal modality.
+representing an enclosed stage-\bbtwo\ expression of type $A$. $\fut A$ terms must be treated opaquely by stage-\bbone\ code
+as they cannot be reduced (to values with type $A$) until stage \bbtwo.
+While $\fut$ allows use to embed stage-\bbtwo\ types within stage-\bbtwo\ types, 
+there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
 
-Although there is a strict one-way dependence between the two stages at the type
-level, stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
+Despite this one-way dependence between the two stages at the type level, 
+stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
 the $\next$ construct embeds stage-\bbtwo\ expressions into stage \bbone,

### 2014-07-24 paper/semantics.tex
Fixed up exposition example to reflect new hatted variable style.

@@
 One option, taken by \cite{davies96}, is to treat \verb|next{1+2}| as a value for the purposes of stage-\bbone~evaluation, 
-consequently duplicating it during substitution for {\tt x}.  
+consequently duplicating it during substitution for {\tt x}.  The two $\prev$ constructs eliminate the $\next$ wrappers,
+leaving the contained stage-\bbtwo\ code in place.
 This produces:
@@
 We take a different approach.  
-Instead of duplicating the contents of the first $\next$ expression, we bind them to some variable (here, {\tt y}) and
+Instead of duplicating the contents of the first $\next$ expression, we bind them to some variable (here, $\mathtt{y}$) and
 duplicate only a reference to that variable.  This produces:
@@
 
-% We think that this interpretation makes more sense from the point of view of cost, 
-% since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
-% This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
-% when a non-terminating stage-\bbtwo~expression is declared but never used.
-
 Achieving this behavior mechanically requires us to resolve a contradiction:
 we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
-Our solution is to replace the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s contents.  
+Our solution is to replace the contents of the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s old contents.  
 This substitution then floats up to the top of the containing $\prev$ block:
@@
 prev {
-[y|->1+2]
-  let x = (yhat, 7) in
+[yhat|->1+2]
+  let x = (next{yhat}, 7) in
   next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
@@
 \end{lstlisting}
-Note that the variable left behind is rendered with a hat.  By treating this {\em hatted variable} as a value, we're now free to perform the stage-\bbone~substitution for {\tt x}.  
+As a convention, we render the new variable with a %stylish and fashionable
+hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
 \begin{lstlisting} 
@@
   next{
-    prev{#1 (yhat, 7)} * prev{#1 (yhat, 7)} *
-    hold{#2 (yhat, 7)}
+    prev{#1 (next {yhat}, 7)} * 
+    prev{#1 (next {yhat}, 7)} *
+    hold{#2 (next {yhat}, 7)}
   }
@@
 To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them. 
-As a rule, these will reduce to hatted variables, and $\prev$ removes the hat, leaving the variable in place:
+As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
 \begin{lstlisting} 
 prev {
-[y|->1+2]
-    next{ y * y * 7 }
+[yhat|->1+2]
+    next{ yhat * yhat * 7 }
 }
@@
 prev {
-[y|->1+2]
-[z|->y*y*7]
-    zhat
+[yhat|->1+2]
+[zhat|->yhat*yhat*7]
+    next{ zhat }
 }
@@
 \begin{lstlisting} 
-let y = 1+2 in
-let z = y * y * 7 in z
+let yhat = 1+2 in
+let zhat = yhat * yhat * 7 in z
 \end{lstlisting}

### 2014-07-24 paper/semantics.tex
work on intro of section 3

--- 2014-07-24 paper/semantics.tex
@@
 In this section, we describe \lang, a simple two-stage language
-used as the basis for defining and analyzing stage splitting.
+for which we define and analyze stage splitting.
 
@@
 %\ref{fig:grammar,fig:statics}.
-Every expression in \lang\ has both a type and a {\em stage}, either \bbone~or \bbtwo.
-Just as a type describes \emph{what} sort of value an expression reduces to, the stage tells us \emph{when} an expression will be fully reduced.  
-Within each stage, \lang\ reduces to the simply-typed lambda calculus augmented with products, sums, general recursion, and base types
-\footnote{It needn't be the case that we choose the same set of features at both levels, but this does simplify the exposition.}.
+The language \lang\ is essentially two copies of a typed lambda calculus with
+products, sums, and isorecursive types.%
+\footnote{It is possible for different stages to have different sets of
+features, but for simplicity we do not consider this.}
+Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
+expressing in which copy of the language the term resides. Intuitively, the
+stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
+subexpressions are evaluated before stage-\bbtwo\ ones.
 
-In addition to these types, stage-\bbone\ expressions may also be of type $\fut A$,
-representing an enclosed stage-\bbtwo\ expression of type $A$. $\fut A$ terms must be treated opaquely by stage-\bbone\ code
-as they cannot be reduced (to values with type $A$) until stage \bbtwo.
-While $\fut$ allows use to embed stage-\bbtwo\ types within stage-\bbtwo\ types, 
+In addition, there is a stage-\bbone\ type $\fut A$, containing encapsulated
+stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
+opaquely by stage-\bbone\ code, as they cannot be evaluated until stage \bbtwo.
+While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbtwo\ types,
 there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
-
-Despite this one-way dependence between the two stages at the type level, 
-stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
-the $\next$ construct embeds stage-\bbtwo\ expressions into stage \bbone,
-whereas $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.
-The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
-and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.
+The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
+while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  These are the
+only ways in \lang\ to alter the stage of a term; we surround their arguments
+with braces in \lang\ syntax to clearly indicate stage boundaries within a
+program.
 

### 2014-07-24 paper/semantics.tex
informal discussion

@@
 
-To understand how this behavior can be achieved, consider evaluation of the stage-\bbtwo\ term below, which contains nested $\next$ and $\prev$ blocks:
+\TODO clean up the next few paragraphs; unclear how much of it should go in the
+paper
+
+To understand the difficulty of staged evaluation, consider the following
+example:
+\begin{lstlisting}
+#2 (next {1+2}, 3+4)
+\end{lstlisting}
+This is a stage-\bbone\ expression of type $\rmint$; the pair is a stage-\bbone\
+expression of type $(\fut\rmint)\times\rmint$. A conventional call-by-value
+evaluation strategy demands that we evaluate both components of the pair to
+values before we project from the pair. However, at stage \bbone\ we are not
+allowed to evaluate under the \verb|next|; on the other hand, \verb|next {1+2}|
+is not a value, in the sense that it requires additional computation steps.
+
+Intuitively, the solution is to say that \verb|next {1+2}| is a value \emph{at
+stage \bbone}, even though once we proceed to stage-\bbtwo\ evaluation, it will
+no longer be a value. Therefore, we evaluate the pair to
+\begin{lstlisting}
+#2 (next {1+2}, 7)
+\end{lstlisting}
+and get \verb|7| as the answer.
+
+To complicate matters, consider a situation where we must substitute a stage
+\bbone\ value into some other expression. In the stage-\bbtwo\ term
 \begin{lstlisting} 
@@
 \end{lstlisting}
+our stage-\bbone\ evaluation will clearly yield a term which is not fully
+reduced---since \verb|1+2| is needed to compute the end result, but we cannot
+reduce it to \verb|3| until stage \bbtwo.
+
+On the other hand, if we simply treat \verb|(next {1+2}, 7)| as a value, then we
+will duplicate the \verb|1+2| computation in the term, and in fact we will
+evaluate it twice, in each computation of \verb|#1 (next {1+2}, 7)|.
+
+Our solution is to hoist these residuals, etc.
+
+\vspace{2em}
+\TODO old stuff below here. merge with above text:
+
 Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.

### 2014-07-24 paper/semantics.tex
Fixed up more text to reflect new hatted vars.  Added pval fig.

@@
 but which may contain hatted variables. In the example above, 
-\verb|(|$\hat {\tt y}$\verb|,7)|.
+\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
 \item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
@@
 At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does nothing but recursively speculate into every subexpression.
-Once speculation finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which produces a future context and (by canonical forms) a hatted variable ($\hat y$).
-The context is then reified (using the $\reifysym$ judgment) into a series of let bindings, with $y$---stripped of its hat---at the bottom.
+Once speculation finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which produces a future context and (by canonical forms) a $\next$-wrapped variable ($\next\{\hat y\}$).
+The context is then reified (using the $\reifysym$ judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of its $\next$---at the bottom.
 
-Because the inputs to evaluation and speculation are sometimes open on variables, we adorn our judgments with a context.
-Since hatted variables can only be introduced to terms via substitution ({\em i.e.} they do not appear in the original program), 
-this context is augmented only at stage \bbone\ substitution sites.
-The second context ($\Gamma$) keeps track of stage \bbtwo\ variables, which are declared in original program.
+The context ($\Gamma$) keeps track of stage \bbtwo\ variables in the input term. 
+These both appear in the original program at stage \bbtwo\ and are inserted by the semantics.
+
+As an optimization, we can include the special-case rule,
+\begin{mathpar}
+\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
+\end{mathpar}
+to avoid one-for-one variable bindings in the residual.
+We used this implicitly in the example in the previous section.
 

### 2014-07-24 paper/semantics.tex
few more hat-related fixes

@@
 %places where the evaluation and speculation judgments depend on the other. 
-
-%The evaluation judgment is very similar to standard call-by-value evaluation.
-%The input to evaluation is a stage-1 expression (usually
-%$e$), as well as two administrative contexts ($\Gamma$), covered
-%later.  Evaluation has two outputs: the {\em future context} (usually $\xi$) and
-%the {\em partial value} (usually $v$).  We cover the latter first.  The partial
-%value is essentially the result of the first-stage portion of $e$, and must be a
-%\bbone-value of the same type as $e$.  As usual, this means that $v$ must be
-%composed only of base primitives, tuples, injections, and lambdas (corresponding
-%to base types, products, sums, and functions).  Analogously, the value
-%corresponding to $\fut$ is a construct called a {\em hatted variable} (written
-%$\hat y$), which signifies a reference to some stage-2 computation.  The mapping
-%between these hatted variables and stage-two expressions is
 %}}}
@@
 Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
-but which may contain hatted variables. In the example above, 
+but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
+In the example above, 
 \verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
@@
 The future environment is a mapping from newly-created variables 
-(which then may appear---hatted---in $v$)
+(which then may appear within $\next$ blocks in $v$)
 to residuals.  It is the formal manifestation of the floating substitutions in our example.
@@
 Once the contents of the $\next$ block are speculated into a residual ($q$), 
-we then return a fresh hatted variable ($\hat y$), 
-along with a future environment which maps that variable to our residual ($y \mapsto q$).
+we then return a fresh variable wrapped in a $\next$ block ($\next~\hat y$), 
+along with a future environment which maps that variable to our residual ($\hat y \mapsto q$).
 
@@
 function:
-$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
+$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
 The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to

### 2014-07-24 paper/semantics.tex
temp commit

--- 2014-07-24 paper/semantics.tex
@@
 \end{lstlisting}
-This is a stage-\bbone\ expression of type $\rmint$; the pair is a stage-\bbone\
-expression of type $(\fut\rmint)\times\rmint$. A conventional call-by-value
-evaluation strategy demands that we evaluate both components of the pair to
-values before we project from the pair. However, at stage \bbone\ we are not
-allowed to evaluate under the \verb|next|; on the other hand, \verb|next {1+2}|
-is not a value, in the sense that it requires additional computation steps.
+This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
+stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
+call-by-value evaluation strategy demands that we evaluate both components of
+the pair to values before we project from the pair. However, at stage \bbone\ we
+are not allowed to evaluate under the \verb|next|; on the other hand, 
+\verb|next {1+2}| is not a value, in the sense that it requires additional
+computation steps.
 

### 2014-07-24 paper/semantics.tex
work on 3.2

@@
 
-\TODO clean up the next few paragraphs; unclear how much of it should go in the
-paper
-
 To understand the difficulty of staged evaluation, consider the following
@@
 Intuitively, the solution is to say that \verb|next {1+2}| is a value \emph{at
-stage \bbone}, even though once we proceed to stage-\bbtwo\ evaluation, it will
-no longer be a value. Therefore, we evaluate the pair to
+stage \bbone}, even though it requires additional evaluation at stage \bbtwo.
+Therefore, we evaluate the pair to
 \begin{lstlisting}
@@
 
-To complicate matters, consider a situation where we must substitute a stage
-\bbone\ value into some other expression. In the stage-\bbtwo\ term
+To complicate matters, consider a situation where we must substitute such a
+incompletely-evaluated expression. The following stage-\bbtwo\ term has type
+$\rmint$:
 \begin{lstlisting} 
@@
 \end{lstlisting}
-our stage-\bbone\ evaluation will clearly yield a term which is not fully
-reduced---since \verb|1+2| is needed to compute the end result, but we cannot
-reduce it to \verb|3| until stage \bbtwo.
+Stage-\bbone\ evaluation will not fully reduce this term, because the answer
+depends on the value of \verb|1+2|, but we cannot reduce this to \verb|3| until
+stage \bbtwo.
 
-On the other hand, if we simply treat \verb|(next {1+2}, 7)| as a value, then we
-will duplicate the \verb|1+2| computation in the term, and in fact we will
-evaluate it twice, in each computation of \verb|#1 (next {1+2}, 7)|.
+If we simply treat \verb|(next {1+2}, 7)| as a value during stage \bbone, then
+we will substitute it for the three occurrences of \verb|x| in the body of the
+\verb|let| expression. Thus, the result of stage \bbone\ computation is
+\begin{lstlisting} 
+prev{
+  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
+}
+\end{lstlisting}
+When we evaluate this at stage \bbtwo, we will compute \verb|1+2| twice. Of
+course, this is not particularly expensive, but in general, the duplicated
+computation could be anything, even a recursive call to the function!
 
-Our solution is to hoist these residuals, etc.
-
-\vspace{2em}
-\TODO old stuff below here. merge with above text:
-
-Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
-One option, taken by \cite{davies96}, is to treat \verb|next{1+2}| as a value for the purposes of stage-\bbone~evaluation, 
-consequently duplicating it during substitution for {\tt x}.  The two $\prev$ constructs eliminate the $\next$ wrappers,
-leaving the contained stage-\bbtwo\ code in place.
+To avoid duplicating computations, we take a different approach.  Instead of
+duplicating the contents of the first $\next$ expression, we bind them to some
+variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
 This produces:
-\begin{lstlisting}
-(1+2) * (1+2) * 7
-\end{lstlisting}
-
-We take a different approach.  
-Instead of duplicating the contents of the first $\next$ expression, we bind them to some variable (here, $\mathtt{y}$) and
-duplicate only a reference to that variable.  This produces:
 \begin{lstlisting} 

### 2014-07-24 paper/semantics.tex
added more explanation of partial evaluation system

@@
 
-A function $f$ which accepts inputs at both stages \bbone{} and \bbtwo{} can be
-given a type of the form $A\to\fut(B\to C)$.%
+We now consider how to use this formal system to encode inputs to partial evaluation and stage-splitting systems.
+Recall that in partial evaluation systems, we have a function $f$ for which some input is labeled {\em static} and 
+some labeled {\em dynamic}.  Once the static input is provided, we can partially evaluate the function to produce
+a residual that depends on only the dynamic input.
+
+By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
+we can encode this sort of system with a function type of the form $A\to\fut(B\to C)$.
 \cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
@@
 %
-Given its stage \bbone\ argument $a:A$, we can evaluate the partially-applied
+Here $A$ represents the static input, $B$ represents the dynamic input, and $C$ represents the output.
+
+Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
 function:

### 2014-07-24 paper/splitting.tex
minor changes after meeting with nico
A few changes to metatheory sections.
beginning to generalize the splitting correctness theorems

@@
 
+\subsection{Value Splitting, Etc}
+
+...
+
+\subsection{Metatheory}
+
+\begin{definition}
+For residuals $q_1$ and $q_2$, define $q_1 \le_\beta q_2$ (``$q_1$ is $\beta$-smaller than $q_2$'') if they are
+identical modulo internal beta reductions of $q_2$.
+\end{definition}
+
+The theorem we really want is this:
+
+\begin{theorem}
+If $\cdot\vdash e : A \splittwosym [p,l.r]$ then,
+\begin{itemize}
+\item $\cdot \vdash e : A~@~\bbtwo$ and
+\item $\types [l:\tau] r A$ and
+\item $\diatwo [\cdot] e q$ iff $\reduce p u$ and if so
+\item $q \le_\beta (\letin{l}{u}{r})$
+\end{itemize}
+\end{theorem}
+
+But this isn't strong enough to prove on its own.  
+Firstly, it only considers empty contexts, which is not enough.
+Second, it is only at the second stage.
+
+Here's an empty-context version for the first stage.
+
+\TODO fix below
+\begin{theorem}
+If $\cdot\vdash e : A \splitonesym [c,l.r]$ then,
+\begin{itemize}
+\item $\cdot \vdash e : A~@~\bbtwo$ and
+\item $\types [l:\tau] r A$ and
+\item $\diatwo [\cdot] e q$ iff $\reduce p u$ and if so
+\item $q \le_\beta (\letin{l}{u}{r})$
+\end{itemize}
+\end{theorem}
+
+\input{figures/splitting-misc}
+
 \section{Examples for Staged Pipelines}

### 2014-07-25 paper/semantics.tex
two quick fixes noticed in meeting

@@
 opaquely by stage-\bbone\ code, as they cannot be evaluated until stage \bbtwo.
-While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbtwo\ types,
+While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
 there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.

### 2014-07-25 paper/semantics.tex
Making an editing pass through 3.2.
My focus is eliminating some ambiguous wording in the exposition.  In
particular, I fear there is conflation of the idea of a lambda12
expression at stage 2), and the concept of the residual.

Secondly, I'd like to canonicalize the residual depiction in the
no-duplicate and duplicate cases.

@@
 The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
-Although it is possible to implement $\pause$ from other \lang\ features, 
+It is possible to implement $\pause$ from other \lang\ features, but
 we provide it as a core primitive to simplify our examples.
@@
 
-Although the code for {\tt fexp} looks very much like the unstaged mathematical
+Although the code for {\tt fexp} appears very similar to the unstaged mathematical
 definition above, it is in fact a staged program: the {\tt if} predicates and exponent
@@
 %requiring disruption of the elegant functional structure of the code.}
-We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
+%
+% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
+%
+%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
 We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
-
-A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation eliminates outermost redexes first, 
+%
+A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
 however in the case that stage \bbone\ expressions are contained inside
-stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. 
-Accordingly, a suitably staged dynamic semantics for \lang\ must evaluate
+stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.
+Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
 \emph{all} the stage \bbone\ subexpressions of a term before its stage
@@
 
-To understand the difficulty of staged evaluation, consider the following
+To gain intuition about the challenges of ensuring staged evaluation of \lang, consider the following
 example:
@@
 call-by-value evaluation strategy demands that we evaluate both components of
-the pair to values before we project from the pair. However, at stage \bbone\ we
-are not allowed to evaluate under the \verb|next|; on the other hand, 
-\verb|next {1+2}| is not a value, in the sense that it requires additional
-computation steps.
-
-Intuitively, the solution is to say that \verb|next {1+2}| is a value \emph{at
-stage \bbone}, even though it requires additional evaluation at stage \bbtwo.
+the pair to values before we project from the pair. The problem is that
+while \verb|next {1+2}| is not a value (in the sense that additional
+stage-\bbtwo\ computation steps are necessary to produce the result \verb|3|), evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
+Intuitively, the solution is to designate \verb|next {1+2}| as a value \emph{in
+stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
 Therefore, we evaluate the pair to
@@
 \end{lstlisting}
-and get \verb|7| as the answer.
+then obtain \verb|7| as the result of projection.
 
-To complicate matters, consider a situation where we must substitute such a
-incompletely-evaluated expression. The following stage-\bbtwo\ term has type
-$\rmint$:
+Now consider a more complex example where 
+stage-\bbone\ evaluation must substitute such a incompletely-evaluated
+expression. The following stage-\bbtwo\ term has type $\rmint$:
 \begin{lstlisting} 
@@
 \end{lstlisting}
-Stage-\bbone\ evaluation will not fully reduce this term, because the answer
-depends on the value of \verb|1+2|, but we cannot reduce this to \verb|3| until
+Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
+depends on the value of \verb|1+2|, which is not reduced to \verb|3| until
 stage \bbtwo.
 
-If we simply treat \verb|(next {1+2}, 7)| as a value during stage \bbone, then
-we will substitute it for the three occurrences of \verb|x| in the body of the
-\verb|let| expression. Thus, the result of stage \bbone\ computation is
+If we simply treat \verb|(next {1+2}, 7)| as a value during stage~\bbone, and
+substitute it for the three occurrences of \verb|x| in the body of the
+\verb|let| expression, the result of stage \bbone\ computation is
 \begin{lstlisting} 
@@
 \end{lstlisting}
-When we evaluate this at stage \bbtwo, we will compute \verb|1+2| twice. Of
+Stage-\bbtwo\ evaluation of this expression will compute \verb|1+2| twice. Of
 course, this is not particularly expensive, but in general, the duplicated
-computation could be anything, even a recursive call to the function!
-
-To avoid duplicating computations, we take a different approach.  Instead of
-duplicating the contents of the first $\next$ expression, we bind them to some
-variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
+computation could be anything, even a recursive call to the function! To avoid duplicating
+computations, we take a different approach.  Instead of
+duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
+a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
 This produces:
@@
 Achieving this behavior mechanically requires us to resolve a contradiction:
-we must substitute for \texttt{x} at stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
+we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
 Our solution is to replace the contents of the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s old contents.  

### 2014-07-25 paper/semantics.tex
Clarify stage-2 variable.

@@
 we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
-Our solution is to replace the contents of the $\next$ with a new variable and create an explicit substitution (shown with a $\mapsto$) binding that variable to the $\next$'s old contents.  
+Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.
 This substitution then floats up to the top of the containing $\prev$ block:

### 2014-07-30 paper/splitting.tex
added theorems
started to expand definition of equiv (it's clever); going home
some additions to metatheory of splitting

--- 2014-06-30 paper/splitting.tex
@@
 
+\subsection{Value Splitting, Etc}
+
+...
+
+\subsection{Metatheory}
+
+We start with two careful definitions of equivalence ($\equiv$ on terms below, $\sim$ on values in \ref{fig:auxSplit}).  They are mutually dependent.
+
+There are two problems with these definitions:
+\begin{enumerate}
+\item I'm not sure how to encode $\equiv$ with relations, since the iff requires inversion rather than direct derivation.
+\item In the rule labeled PROBLEM, I have a sudden turnstile out of nowhere.  
+  What I'm trying to say with that premise is ``for any equivalent arguments, the bodies are equivalent".
+\end{enumerate}
+
+\begin{definition}
+For residuals $q$ and $r$, define $q \equiv r : A$ to mean that 
+$e_1 \tworedsym v_1$ iff $\reduce {e_2} {v_2}$ where $v_1 \sim v_2 : A$
+\end{definition}
+
+All I'm trying to say here is that two terms are equivalent if they evaluate to the same value,
+where "same" for functions means that those functions always evaluate to the same thing.
+
+We give the following end-to-end correctness lemmas for closed terms (open term versions forthcoming).
+
+\begin{lemma}
+If $\cdot\vdash e : A \splittwosym [p,l.r]$ then,
+\begin{itemize}
+\item $\cdot \vdash e : A~@~\bbtwo$ 
+\item $\diatwo [\cdot] e q$ iff $\reduce p u$ where
+\item $q \equiv (\letin{l}{u}{r})$
+\end{itemize}
+\end{lemma}
+
+\begin{lemma}
+If $\cdot\vdash e : A \splitonesym [c,l.r]$ then,
+\begin{itemize}
+\item $\typesone [\cdot] e A$ 
+\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ where
+\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+\item $\reify \xi {v_2} q$
+\item $q \equiv (\letin{l}{u}{r})$
+\end{itemize}
+\end{lemma}
+
+%
+%\subsubsection{Simple Types}
+%
+%\begin{theorem}
+%If $\cdot\vdash e : A \splittwosym [p,l:\tau.r]$ then,
+%\begin{itemize}
+%\item $\cdot \vdash e : A~@~\bbtwo$ 
+%\item $\types [\cdot] p \tau$ and $\types [l:\tau] r A$ 
+%\item $\diatwo [\cdot] e q$ iff $\reduce p u$ and if so
+%\item $q \equiv (\letin{l}{u}{r})$
+%\end{itemize}
+%\end{theorem}
+%
+%\begin{theorem}
+%If $\cdot\vdash e : A \splitonesym [c,l:\tau.r]$ then,
+%\begin{itemize}
+%\item $\typesone [\cdot] e A$ 
+%\item $A \tsplito [A_1,A_2]$
+%\item $\types [\cdot] c {A_1 \times \tau}$ and $\types [l:\tau] r A_2$ 
+%\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ and if so
+%\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+%\item $\reify \xi {v_2} q$
+%\item $q \equiv (\letin{l}{u}{r}) : A_2$
+%\end{itemize}
+%\end{theorem}
+%
+%\subsubsection{\bbone-Dependent Types}
+%\begin{theorem}
+%If $\cdot\vdash e : A \splitonesym [c,l:\tau.r]$ then,
+%\begin{itemize}
+%\item $\typesone [\cdot] e A$ 
+%\item $A \tsplito [A_1,a.A_2]$
+%\item $\types [\cdot] c {A_1 \times \tau}$ and $\types [l:\tau,a:A_1] r A_2$ 
+%\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ and if so
+%\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+%\item $\reify \xi {v_2} q$
+%\item $q \equiv (\letin{l}{u}{r}) : A_2$
+%\end{itemize}
+%\end{theorem}
+
+\input{figures/splitting-misc}
+
 \section{Examples for Staged Pipelines}

### 2014-07-30 paper/splitting.tex
added two more notes

@@
 
-We give the following end-to-end correctness lemmas for closed terms (open term versions forthcoming).
+We give the following end-to-end correctness lemmas for closed terms. 
+Open term versions forthcoming.
+The jury is still out on how to make these strong enough to prove that the partitioning between stages is correct.
 
@@
 \end{lemma}
+
+You should think of these theorems as saying that 
+splitting commutes with evaluation.
 

### 2014-07-31 paper/splitting.tex
changed theorems to support opens terms

@@
 
-We give the following end-to-end correctness lemmas for closed terms. 
-Open term versions forthcoming.
+We give the following end-to-end correctness lemmas for open terms. 
+It's a bit of a mess currently, but the $\Gamma$ is supposed to be all of the stage \bbone\ bindings, 
+whereas $\Gamma'$ is the stage \bbtwo\ bindings.
+Substitution splitting works just like value splitting, which is why they use the same symbol.
+
 The jury is still out on how to make these strong enough to prove that the partitioning between stages is correct.
@@
 \begin{lemma}
-If $\cdot\vdash e : A \splittwosym [p,l.r]$ then,
+If $\Gamma, \Gamma'\vdash e : A \splittwosym [p,l.r]$ then for all substitutions $\gamma : \Gamma$,
 \begin{itemize}
-\item $\cdot \vdash e : A~@~\bbtwo$ 
-\item $\diatwo [\cdot] e q$ iff $\reduce p u$ where
-\item $q \equiv (\letin{l}{u}{r})$
+\item $\Gamma,\Gamma' \vdash e : A~@~\bbtwo$ 
+\item $\gamma \vsplito [\gamma_1, \gamma_2]$
+\item $\diatwo [\Gamma'] {\gamma(e)} q$ iff $\reduce {\gamma_1(p)} u$ where
+\item $\Gamma' \vdash q \equiv (\letin{l}{u}{\gamma_2(r)})$
 \end{itemize}
@@
 \begin{lemma}
-If $\cdot\vdash e : A \splitonesym [c,l.r]$ then,
+If $\Gamma, \Gamma'\vdash e : A \splitonesym [c,l.r]$ then for all $\gamma : \Gamma$,
 \begin{itemize}
-\item $\typesone [\cdot] e A$ 
-\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ where
-\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+\item $\typesone [\Gamma, \Gamma'] e A$ 
+\item $\gamma \vsplito [\gamma_1, \gamma_2]$
+\item $\diaone [\Gamma'] {\gamma(e)} {\xi;v}$ iff $\reduce {\gamma_1(c)} {(v_1,u)}$ where
+\item $\Gamma',\dom \xi \vdash v \vsplito [v_1,v_2]$
 \item $\reify \xi {v_2} q$
-\item $q \equiv (\letin{l}{u}{r})$
+\item $\Gamma'\vdash q \equiv (\letin{l}{u}{\gamma_2(r)})$
 \end{itemize}

### 2014-07-31 paper/semantics.tex
added last step in rowan's semantics example
be clear about which language residuals and values are in

@@
 }
+\end{lstlisting}
+Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
+\begin{lstlisting} 
+(1+2) * (1+2) * 7
 \end{lstlisting}
@@
 
+To fully evaluate an expression in \lang, we first evaluate all of its stage
+\bbone\ subexpressions as described above. This results in a stage \bbtwo\
+expression with no stage \bbone\ content; we say that it is a term in a
+monostaged%
+\footnote{In general, \lang could have more than two stages, in which case the
+expression would be in a language with one fewer stage.}
+language called \langTwo. Ordinary \langTwo\ evaluation, implemented by the
+$\tworedsym$ judgment, completes the evaluation of this term.
+
 The algorithm described above creates three different kinds of expressions which
@@
 \verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
-\item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
-subexpressions have all been fully evaluated. In the example above,
+\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
+stage \bbone\ subexpressions have all been fully evaluated. In the example
+above,
 \verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
-\item Values ($\valsym$s) are stage \bbtwo\ terms which are fully evaluated; these
+\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
 are the results of a computation after both stages have been completed. In the
@@
 \end{itemize}
-The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
-The future environment is a mapping from newly-created variables 
-(which then may appear within $\next$ blocks in $v$)
-to residuals.  It is the formal manifestation of the floating substitutions in our example.
-For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
-and the final future environment is gotten by merging the future environments of the subterms.
+The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
+environment} $\xi$ and a partial value $v$.  The future environment is a mapping
+from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
+$v$) to residuals---in our example, we represented this environment with
+explicit substitutions. For all of the normal features of \lang
+(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
+effect on values as standard evaluation, and the final future environment is
+gotten by merging the future environments of the subterms.
 
-When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it searches into the block's stage \bbtwo\ content to find any contained stage
-\bbone\ subexpressions and evaluate them in place.  
-We call this search process \emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
-Once the contents of the $\next$ block are speculated into a residual ($q$), 
-we then return a fresh variable wrapped in a $\next$ block ($\next~\hat y$), 
-along with a future environment which maps that variable to our residual ($\hat y \mapsto q$).
+When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
+searches into the block's stage \bbtwo\ content to find any contained stage
+\bbone\ subexpressions and evaluate them in place.  We call this search process
+\emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes
+a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
+speculated into a residual ($q$), we then return a fresh variable wrapped in a
+$\next$ block ($\next~\hat y$), along with a future environment which maps that
+variable to our residual ($\hat y \mapsto q$).
 
-At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does nothing but recursively speculate into every subexpression.
-Once speculation finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which produces a future context and (by canonical forms) a $\next$-wrapped variable ($\next\{\hat y\}$).
-The context is then reified (using the $\reifysym$ judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of its $\next$---at the bottom.
+At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
+nothing but recursively speculate into every subexpression.  Once speculation
+finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which
+produces a future context and (by canonical forms) a $\next$-wrapped variable
+($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
+judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of
+its $\next$---at the bottom.
 

### 2014-08-05 paper/semantics.tex
moving around text in semantics section

@@
 stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.
-Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
-\emph{all} the stage \bbone\ subexpressions of a term before its stage
-\bbtwo\ subexpressions.
+
+Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
+subexpressions before any of its stage \bbtwo\ subexpressions. This results in a
+stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
+monostaged%
+\footnote{In general, \lang could have more than two stages, in which case the
+expression would be in a language with one fewer stage.}
+language called \langTwo. Then, we perform the remainder of the evaluation with 
+$\tworedsym$, an ordinary dynamic semantics for \langTwo.
 
@@
 
-To gain intuition about the challenges of ensuring staged evaluation of \lang, consider the following
-example:
+To gain intuition about the challenges of implementing this staged dynamic
+semantics, consider the following example:
 \begin{lstlisting}
@@
 \subsection{Dynamics}
+\label{ssec:dynamics}
 
@@
 %}}}
-
-To fully evaluate an expression in \lang, we first evaluate all of its stage
-\bbone\ subexpressions as described above. This results in a stage \bbtwo\
-expression with no stage \bbone\ content; we say that it is a term in a
-monostaged%
-\footnote{In general, \lang could have more than two stages, in which case the
-expression would be in a language with one fewer stage.}
-language called \langTwo. Ordinary \langTwo\ evaluation, implemented by the
-$\tworedsym$ judgment, completes the evaluation of this term.
 

### 2014-08-05 paper/splitting.tex
first cut at splitting correctness theorem

@@
 
-You should think of these theorems as saying that 
-splitting commutes with evaluation.
+\TODO what about totality of splitting?
+
+%You should think of these theorems as saying that 
+%splitting commutes with evaluation.
+While these theorems are rather technical in nature, they imply that splitting a
+program and running it is the same as using the semantics from
+\ref{ssec:dynamics}.
+
+For example, if we have some multi-stage function $f:A\to\fut(B\to C)~@~\bbone$,
+then we expect \verb|prev{f a} b| will evaluate to the same value under our
+staged dynamic semantics and when split.
+
+Consider a term $\vdash e:\fut A~@~\bbone$ (like \verb|next{prev{f a} b}|) which
+might contain computation at both stages. The correctness theorem here states
+that:
+\begin{theorem}
+If $\vdash e:\fut A~@~\bbone$, then (by splitting)
+\begin{itemize}
+\item $\diaone [] e {\xi;v}$
+\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+\item $\reify \xi {v_2} q_S$
+\item $q_S \mathbin{\tworedsym} v_S$
+\end{itemize}
+if and only if (by the staged dynamic semantics)
+\begin{itemize}
+\item $\diaone [] e {\xi;v}$
+\item $\reify \xi v q_D$
+\item $q_D \mathbin{\tworedsym} v_D$
+\end{itemize}
+and if so, then $v_S = v_D$.
+\end{theorem}
+
+\TODO why does splitting not show up in this theorem?
 

### 2014-08-05 paper/splitting.tex
updated splitting theorems.

--- 2014-07-31 paper/splitting.tex
@@
 
-You should think of these theorems as saying that 
-splitting commutes with evaluation.
+\TODO what about totality of splitting?
+
+%You should think of these theorems as saying that 
+%splitting commutes with evaluation.
+These lemmas are rather technical, but they ultimately imply that evaluating a
+closed term by splitting or by the dynamic semantics of \ref{ssec:dynamics} are
+equivalent. 
+We state this result for closed terms at each stage.
+
+\begin{theorem}[Correctness of splitting at $\bbone$]
+If $\vdash e:A~@~\bbone$, then (by splitting)
+\begin{itemize}
+\item $\vdash e : A \splitonesym [c,l.r]$
+\item $\reduce {c} {(v_1,u)}$
+\item $\reduce {(\letin{l}{u}{r})} v_S$
+\end{itemize}
+if and only if (by the staged dynamic semantics)
+\begin{itemize}
+\item $\diaone [] e {\xi;v}$
+\item $\reify \xi v q$
+\item $q \mathbin{\tworedsym} v_D$
+\end{itemize}
+and if so, then $v_S = v_D$.
+\end{theorem}
+
+\begin{theorem}[Correctness of splitting at $\bbtwo$]
+If $\vdash e:A~@~\bbtwo$, then (by splitting)
+\begin{itemize}
+\item $\vdash e : A \splittwosym [p,l.r]$
+\item $\reduce p u$
+\item $(\letin{l}{u}{r}) \mathbin{\tworedsym} v_S$
+\end{itemize}
+if and only if (by the staged dynamic semantics)
+\begin{itemize}
+\item $\diatwo [] e q$
+\item $q \mathbin{\tworedsym} v_D$
+\end{itemize}
+and if so, then $v_S = v_D$.
+\end{theorem}
+
+If we apply the former theorem to \verb|next{e}| of type $\fut A$, we
+essentially obtain the latter theorem at \verb|e|.
+
+The latter theorem implies that, given a multi-stage function $f:A\to\fut(B\to
+C)~@~\bbone$, the two methods of evaluating \verb|prev{f a} b| agree.
+However, we also expect that splitting $f$ directly will give us two functions,
+one which accepts an $A$ and outputs a value and $r$-residual, and another which
+takes in an $r$-residual and a $B$ and outputs a $C$. Moreover, the composition
+of these two functions should be extensionally equal to the staged dynamic
+semantics.
+
+\TODO write the theorem for $\vdash f:A\to\fut(B\to C)~@~\bbone$.
 

### 2014-08-06 paper/splitting.tex
added a paragraph and figure on the output language

@@
 The basic idea is that we want to take a term which contains interleaved stage \bbone\ and stage \bbtwo\ code
-and untangle it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.
+and split it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.
 To facilitate communication between the stages, there is a data structure passed between the stages, 
 which is an output of the stage \bbone\ term and and input to the stage \bbtwo\ term.
+
+The input to our splitting algorithm is a term in \lang, and the output is two terms in \langmono, an
+unstaged language.  The grammar for \langmono is given in \ref{fig:monoGrammar}.  It has no
+staging features (we say that implicitly it has only one stage), and it is untyped.  
+Modulo these differences, \langmono has all of the same features as \lang, 
+although this is a matter of taste
+\footnote{Since \lang is untyped, we could encode everything with just functions.}.
 

### 2014-08-08 paper/splitting.tex
improved equivalence definitions a little

@@
 
-We start with two careful definitions of equivalence ($\equiv$ on terms below, $\sim$ on values in \ref{fig:auxSplit}).  They are mutually dependent.
-
-There are two problems with these definitions:
-\begin{enumerate}
-\item I'm not sure how to encode $\equiv$ with relations, since the iff requires inversion rather than direct derivation.
-\item In the rule labeled PROBLEM, I have a sudden turnstile out of nowhere.  
-  What I'm trying to say with that premise is ``for any equivalent arguments, the bodies are equivalent".
-\end{enumerate}
+We start with two careful definitions of equivalence ($\equiv$ on terms, $\sim$ on values).  They are mutually dependent.
 
 \begin{definition}
-For residuals $q$ and $r$, define $q \equiv r : A$ to mean that 
-$e_1 \tworedsym v_1$ iff $\reduce {e_2} {v_2}$ where $v_1 \sim v_2 : A$
+For evaluation-residual $q$ and splitting-residual $r$, define $q \equiv r : A$ to mean that 
+$q \tworedsym v_q$ iff $\reduce {r} {v_r}$ where $v_q \sim v_r : A$, 
+\end{definition}
+
+\begin{definition}
+For evaluation-residual value $v_1$ and splitting-residual value $v_2$, define $v_1 \sim v_2 : A$ by the following cases:
+\begin{itemize}
+\item $i \sim i : \rmint$
+\item $(v_1,u_1) \sim (v_2,u_2) : A \times B$ where $v_1 \sim v_2 : A$ and $u_1 \sim u_2 : B$
+\item $\inl~v_1 \sim \inl~v_2 : A + B$ where $v_1 \sim v_2 : A$
+\item $\inr~v_1 \sim \inr~v_2 : A + B$ where $v_1 \sim v_2 : B$
+\item $\lam {x_1} A {e_1} \sim \lambda x_2.e_2 : A \to B$ where \\ $\forall (v_1 \sim v_2 : A). [v_1/x_1]e_1 \equiv [v_2/x_2]e_2 : B$
+\end{itemize}
 \end{definition}

### 2014-08-08 paper/splitting.tex
added more text to splitting

@@
 
-The input to our splitting algorithm is a term in \lang, and the output is two terms in \langmono, an
-unstaged language.  The grammar for \langmono is given in \ref{fig:monoGrammar}.  It has no
-staging features (we say that implicitly it has only one stage), and it is untyped.  
-Modulo these differences, \langmono has all of the same features as \lang, 
-although this is a matter of taste
-\footnote{Since \lang is untyped, we could encode everything with just functions.}.
-
 As with the dynamics, the form of the splitting statement depends on the external stage of the term.
@@
 We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
-which are called the {\em precomputation} and {\em residual}, respectively.  
-For this splitting to be correct, the residual applied to the precomputation should yield the same result as the original term.
+which are called the {\em precomputation} and {\em residual}
+\footnote{We also used the term {\em residual} for the result of partial evaluation.
+Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
+and {\em evaluation-residual}.}, respectively.
 
@@
 
+We must also develop a notion of what it means for splitting to be correct.  
+Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
+In general, if a multi-stage program $P$ splits into stage-\bbone\ part $P_1$ and stage-\bbtwo\ part $P_2$, then
+\begin{enumerate}
+\item Partial evaluation of $P$ will terminate if and only if evaluation of $P_1$ terminates, and if so then
+\item the immediate results of each method will be equivalent, and
+\item the evaluation-residual and the splitting-residual ($P_2$ bound under the precomputation) are equivalent.
+\end{enumerate}
+
+This is straight-forward, given a careful definition of equivalence.
+
 \subsection {Formal Setup}
 
-We now discuss the formal definition of splitting.
 The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$.
-The $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
+For both, the input is a term in \lang, and the output is two terms in \langmono, an
+unstaged language.  The grammar for \langmono is given in \ref{fig:monoGrammar}.  It has no
+staging features (we say that implicitly it has only one stage), and it is untyped.  
+Modulo these differences, \langmono has all of the same features as \lang, 
+although this is a matter of taste
+\footnote{Since \lang is untyped, we could encode everything with just functions.}.
+
+Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
 while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
-For concision, we actually represent residuals as a term open on a single variable, rather than as functions.
+For concision, we represent residuals as a term open on a single variable, rather than as functions.
 For both judgments, we use a context ($\Gamma$) to keep track of the open variables of $e$.
@@
 
-We start with two careful definitions of equivalence ($\equiv$ on terms, $\sim$ on values).  They are mutually dependent.
+We start with two mutually dependent definitions of equivalence.  
+Both relate evaluation-residuals on the left with splitting residuals on the right,
+but $\equiv$ operates on terms, whereas $\sim$ on values.
 
@@
 
-All I'm trying to say here is that two terms are equivalent if they evaluate to the same value,
-where "same" for functions means that those functions always evaluate to the same thing.
+Essentially, we can read these as saying that two terms are equivalent if they evaluate to the same value,
+where "same" for functions means that those functions always evaluate to the same thing given equivalent inputs.
 

### 2014-08-08 paper/splitting.tex
minor changes discussed today

@@
 \begin{lemma}
+If $\typesone e A$ then
+$\Gamma\vdash e : A \splitonesym [c,l.r]$.
+If $\typestwo e A$ then
+$\Gamma\vdash e : A \splittwosym [p,l.r]$.
+\end{lemma}
+
+\begin{lemma}
 If $\Gamma, \Gamma'\vdash e : A \splittwosym [p,l.r]$ then for all substitutions $\gamma : \Gamma$,
 \begin{itemize}
-\item $\Gamma,\Gamma' \vdash e : A~@~\bbtwo$ 
 \item $\gamma \vsplito [\gamma_1, \gamma_2]$
@@
 \begin{itemize}
-\item $\typesone [\Gamma, \Gamma'] e A$ 
 \item $\gamma \vsplito [\gamma_1, \gamma_2]$
@@
 \end{lemma}
-
-\TODO what about totality of splitting?
 
@@
 \item $\diaone [] e {\xi;v}$
-\item $\reify \xi v q$
+\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+\item $\reify \xi{v_2}q$
 \item $q \mathbin{\tworedsym} v_D$
 \end{itemize}
-and if so, then $v_S = v_D$.
+and if so, then $v_D \sim v_S$.
 \end{theorem}
@@
 \item $\reduce p u$
-\item $(\letin{l}{u}{r}) \mathbin{\tworedsym} v_S$
+\item $\reduce{(\letin{l}{u}{r})}{v_S}$
 \end{itemize}
@@
 \end{itemize}
-and if so, then $v_S = v_D$.
+and if so, then $v_D \sim v_S$.
 \end{theorem}
@@
 However, we also expect that splitting $f$ directly will give us two functions,
-one which accepts an $A$ and outputs a value and $r$-residual, and another which
-takes in an $r$-residual and a $B$ and outputs a $C$. Moreover, the composition
-of these two functions should be extensionally equal to the staged dynamic
-semantics.
+one which accepts an $A$ and outputs an intermediate value and boundary data,
+and another which takes in that boundary data and a $B$ and outputs a $C$.
+Moreover, the composition of these two functions should be extensionally equal
+to the staged dynamic semantics.
 

### 2014-08-12 paper/splitting.tex
fixing a typo kayvon noticed

@@
 For both, the input is a term in \lang, and the output is two terms in \langmono, an
-unstaged language.  The grammar for \langmono is given in \ref{fig:monoGrammar}.  It has no
+unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
 staging features (we say that implicitly it has only one stage), and it is untyped.  
-Modulo these differences, \langmono has all of the same features as \lang, 
+Modulo these differences, \langmono\ has all of the same features as \lang, 
 although this is a matter of taste
-\footnote{Since \lang is untyped, we could encode everything with just functions.}.
+\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
 

### 2014-08-12 paper/semantics.tex
added more text to splitting
fixing up metatheory

@@
 \section{The \lang\ Language}
+\label{sec:semantics}
 
@@
 
-\begin{definition}
-Contexts $\Gamma$ is well-formed ($\Gamma\wf$) if it
-contains only stage-2 variables.
-\end{definition}
+\TODO explain that the $\vdash_\bbtwo$ judgments are for \langTwo.
+
+%\begin{definition}
+%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
+%contains only stage-2 variables.
+%\end{definition}
 
@@
 \item $\xi = \cdot$; or
-\item $\xi = \xi',x\mapsto e$ where
+\item $\xi = \xi',x:B\mapsto e$ where
 $\Gamma\vdash\xi'\wf$ and
-$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
-$\Gamma,\dom{\xi'}\vdash e \res$.
+$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
+%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
+%$\Gamma,\dom{\xi'}\vdash e \res$.
 \end{enumerate}
@@
 \begin{theorem}
-If $\typesone e A$ then $\Gamma\wf$ and $A\istypeone$.
+If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
 \end{theorem}
@@
 \begin{enumerate}
-\item $\Gamma\vdash \coltwo{q}{A}$; and
-\item $\Gamma\vdash q\val$.
+\item $\typeslangTwo q A$; and
+\item $\Gamma\vdash_\bbtwo q\val$.
 \end{enumerate}
 \end{theorem}
+
+\TODO fix the next theorem:
+
+\begin{theorem}
+If $\Gamma\vdash\xi\wf$ and
+$\Gamma,\dom\xi\vdash \colone{v}{A}$ and
+$\Gamma,\dom\xi\vdash v\pval$
+then 
+$\reify{\xi}{v}{q}$ and
+$\typeslangTwo q A$.
+\end{theorem}

### 2014-08-14 paper/semantics.tex
refine the metatheorems

@@
 
-\TODO explain that the $\vdash_\bbtwo$ judgments are for \langTwo.
+Recall that residuals live in \langTwo; we will indicate typing judgments in
+\langTwo\ with $\vdash_\bbtwo$.
 
@@
 
-\TODO fix the next theorem:
+\begin{theorem}
+If $\Gamma\vdash\xi\wf$ and
+$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
+then 
+$\reify{\xi}{\hat y}{q}$ and
+$\typeslangTwo q A$.
+\end{theorem}
+
+\TODO
+We want to generalize the previous theorem to all 1-types, not just $\fut A$,
+but then we must value-split the partial value, and then we don't know the type
+of the output. (?)
 
@@
 then
-$\reify{\xi}{v}{q}$ and
-$\typeslangTwo q A$.
+$v \vsplito [v_1,v_2]$ and
+$A$ type-splits to $A_1,A_2$ and
+$\reify{\xi}{v_2}{q}$ and
+$\typeslangTwo q A_2$.
 \end{theorem}

### 2014-08-15 paper/semantics.tex
some text in section 3 about top-level operational semantics

--- 2014-08-12 paper/semantics.tex
@@
 C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
-enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
-ensures that the resulting $f_a$ has type $B\to C$ in the residual language.
-Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
-the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.
-
-\TODO make sure this is true (haven't written the canonical forms theorems yet)
+enclosing $v$, via $\reify\xi{v}{f_a}$. By \ref{thm:reify-type}, the resulting
+residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
+and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
 
@@
 
-\TODO fix the next theorem:
-
-\begin{theorem}
+\begin{theorem}\label{thm:reify-type}
 If $\Gamma\vdash\xi\wf$ and
-$\Gamma,\dom\xi\vdash \colone{v}{A}$ and
-$\Gamma,\dom\xi\vdash v\pval$
+$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
 then 
-$\reify{\xi}{v}{q}$ and
+$\reify{\xi}{\hat y}{q}$ and
 $\typeslangTwo q A$.
 \end{theorem}
+
+\TODO
+Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
+integer term is guaranteed not to depend on the table, although one might be
+produced. One may either discard the table, or evaluate everything in the table
+(and terminating with the partial value iff everything in the table terminates).

### 2014-08-26 paper/semantics.tex
Added tech report.  Changes to semantics section.

@@
 stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
-monostaged%
+monostaged
 \footnote{In general, \lang could have more than two stages, in which case the
@@
 
-\subsection{Metatheory}
-
-Recall that residuals live in \langTwo; we will indicate typing judgments in
-\langTwo\ with $\vdash_\bbtwo$.
-
+%\subsection{Metatheory}
+%
+%Recall that residuals live in \langTwo; we will indicate typing judgments in
+%\langTwo\ with $\vdash_\bbtwo$.
+%
+%%\begin{definition}
+%%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
+%%contains only stage-2 variables.
+%%\end{definition}
+%
 %\begin{definition}
-%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
-%contains only stage-2 variables.
+%An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
+%\begin{enumerate}
+%\item $\xi = \cdot$; or
+%\item $\xi = \xi',x:B\mapsto e$ where
+%$\Gamma\vdash\xi'\wf$ and
+%$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
+%%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
+%%$\Gamma,\dom{\xi'}\vdash e \res$.
+%\end{enumerate}
 %\end{definition}
-
-\begin{definition}
-An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
-\begin{enumerate}
-\item $\xi = \cdot$; or
-\item $\xi = \xi',x:B\mapsto e$ where
-$\Gamma\vdash\xi'\wf$ and
-$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
-%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
-%$\Gamma,\dom{\xi'}\vdash e \res$.
-\end{enumerate}
-\end{definition}
-
-\begin{theorem}
-If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
-\end{theorem}
-
-\begin{theorem}
-If $\diaonesub$ and $\typesone e A$ then
-\begin{enumerate}
-\item $\Gamma\vdash\xi\wf$;
-\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
-\item $\Gamma,\dom\xi\vdash v\pval$.
-\end{enumerate}
-\end{theorem}
-
-\begin{theorem}
-If $\diatwosub$ and $\typestwo e A$ then
-\begin{enumerate}
-\item $\typeslangTwo q A$; and
-\item $\Gamma\vdash_\bbtwo q\val$.
-\end{enumerate}
-\end{theorem}
-
-\begin{theorem}\label{thm:reify-type}
-If $\Gamma\vdash\xi\wf$ and
-$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
-then 
-$\reify{\xi}{\hat y}{q}$ and
-$\typeslangTwo q A$.
-\end{theorem}
+%
+%\begin{theorem}
+%If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
+%\end{theorem}
+%
+%\begin{theorem}
+%If $\diaonesub$ and $\typesone e A$ then
+%\begin{enumerate}
+%\item $\Gamma\vdash\xi\wf$;
+%\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
+%\item $\Gamma,\dom\xi\vdash v\pval$.
+%\end{enumerate}
+%\end{theorem}
+%
+%\begin{theorem}
+%If $\diatwosub$ and $\typestwo e A$ then
+%\begin{enumerate}
+%\item $\typeslangTwo q A$; and
+%\item $\Gamma\vdash_\bbtwo q\val$.
+%\end{enumerate}
+%\end{theorem}
+%
+%\begin{theorem}\label{thm:reify-type}
+%If $\Gamma\vdash\xi\wf$ and
+%$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
+%then 
+%$\reify{\xi}{\hat y}{q}$ and
+%$\typeslangTwo q A$.
+%\end{theorem}
 

### 2014-08-26 paper/splitting.tex
Added tech report.  Changes to semantics section.

@@
 \section{Splitting Algorithm}
+\label{sec:splitting}
 
@@
 \input{figures/splitting-misc}
-
-\section{Examples for Staged Pipelines}
-
-Give the gist of one-to-one pipeline example (like client/server).
-Then talk about a one-to-many pipeline.
-Then talk about a many-to-one pipeline like spark.
-It clear how to target something like this for known base types on the boundary, and for product types.
-But sums on the boundary are hard!
-We leave many-to-one as future-work.

### 2014-08-26 paper/techreport.tex
Added tech report.  Changes to semantics section.

--- 2014-06-20 paper/paper.tex
@@
 \usepackage{hyperref}
+\usepackage{cprotect}
 
@@
 \crefname{figure}{Figure}{Figures}
+\crefname{theorem}{Theorem}{Theorems}
 
@@
 
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
+\authorinfo{Nicolas Feltman \and Carlo Angiuli \and Umut Acar \and Kayvon Fatahalian}
+           {Carnegie Mellon University}
+           {\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}
 
@@
 
-\input{intro}
+\section{Introduction}
+
+This document is a tech report on current research into a program transformation
+technique which we call {\em splitting}.  This report has three parts, with
+a section dedicated to each,
+\begin{itemize}
+\item \Cref{sec:semantics} describes a language, called \lang, which unambiguously specifies the inputs to 
+our splitting transformation.  It includes both a type system, which is unimaginatively based on previous work, 
+and a semantics, which is more novel.
+\item \Cref{sec:splitting} describes the splitting transformation itself.
+This takes programs, specified in \lang, and converts them to terms in a standard unstaged language.
+This section also describes the more formal definition of the correctness for this process.
+\item \Cref{sec:examples} gives examples of how this splitting transformation can be used
+to derive data structures, and discusses the effectiveness and limitations of this technique.
+\end{itemize}
 
 \input{semantics}
-
-\vfill\eject\pagebreak
 \input{splitting}
-
-\input{relatedwork}
+\input{algoexamples}
 

### 2014-08-27 paper/semantics.tex
Added section on value splitting and lots of exmaples.

@@
 produces a future context and (by canonical forms) a $\next$-wrapped variable
-($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
+($\next\{\hat y\}$)\footnote{
+Importantly, speculation will search into both branches of a stage-\bbtwo\ {\tt if}
+statement to evaluate the contained stage-\bbone\ code.  Thus the name.
+}.  The context is then reified (using the $\reifysym$
 judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of

### 2014-08-27 paper/splitting.tex
Added section on value splitting and lots of exmaples.

@@
 
-The basic idea is that we want to take a term which contains interleaved stage \bbone\ and stage \bbtwo\ code
+The basic idea is that we want to take a \lang\ term which contains interleaved stage \bbone\ and stage \bbtwo\ code
 and split it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.
@@
 We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
-which are called the {\em precomputation} and {\em residual}
-\footnote{We also used the term {\em residual} for the result of partial evaluation.
+which are called the {\em precomputation} and {\em residual}\footnote{We also used the term {\em residual} for the result of partial evaluation.
 Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
@@
 This is straight-forward, given a careful definition of equivalence.
+
+\subsection{Value Splitting}
+
+Before getting into the formal presentation, 
+we first consider the problem of splitting stage-\bbone\ partial values.
+This serves as a didactic stepping-stone to full term-splitting, 
+and it is a necessary component of the correctness theorem.
+
+The rules for value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
+which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
+Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original value.  
+At base types, value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
+in accordance with the precept that base types have only trivial content at stage \bbtwo.
+Splitting essentially distributes into values at product, sum, and recursive types.
+Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 
+general term splitting, which is covered momentarily.
+
+Lastly, we consider value splitting at the $\fut$ type.  The only partial value with this type
+is a $\next$ block containing a reference into the future environment.  This splits into a 
+nullary tuple and the same variable, again aligning with the precept that $\fut$ types contain trivial stage-\bbone\
+content.
 
@@
 \subsection {Stage \bbone\ Divergence}
-Consider the case of stage \bbone\ {\tt if} and {\tt case} expressions.
+Consider the splitting rules for stage-\bbone\ {\tt if} and {\tt case} expressions.
 In both cases we have enough information at stage \bbone\ to know what branch to take, so there's no need to speculate.
@@
 
-Thus the boundary type appears as output of the first stage and input of the second stage.
-This makes it hard to type the output, because the boundary types are not represented in the original function type.
-
-\subsection {Boundary Type Worst Case}
-
-\TODO add the example which is the worst case for figuring out boundary types
+%\subsection {Boundary Type Worst Case}
+%
+%\TODO add the example which is the worst case for figuring out boundary types
+%
 
 \input{figures/splitting-term}
-
-\subsection{Value Splitting, Etc}
-
-...
 

### 2014-08-29 paper/techreport.tex
Add introduction to mini-report.

--- 2014-06-20 paper/paper.tex
@@
 
-\begin{abstract}
-\input{abstract}
-\end{abstract}
+%\begin{abstract}
+%\input{abstract}
+%\end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+%\category{CR-number}{subcategory}{third-level}
 
@@
 % you may leave them out
-\terms
-term1, term2
+%\terms
+%term1, term2
 
-\keywords
-keyword1, keyword2
+%\keywords
+%keyword1, keyword2
 
@@
 
-\input{intro}
+\section{Introduction}
+
+This document is a tech report on ongoing research into a program transformation
+technique which we call {\em splitting}.  The splitting operation statically transforms a program written in a modal language into multiple unstaged (monostage) terms.
+
+The idea of splitting has previously been explored as an optimization technique that improves performance in situations where multi-variate functions are invoked many times, with some inputs being invariant across invocations and others being dynamic (varying across invocations).
+Prior examples of splitting simple languages include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96}, as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+In this document we define the splitting transformation for the two-stage language \lang, a typed lambda calculus featuring products, sums, and isorecursive types.
+
+\begin{itemize}
+\item \Cref{sec:semantics} describes \lang which unambiguously specifies the inputs to 
+our splitting transformation.  The type system of \lang\ is drawn from prior work, but its semantics are uniquely defined in anticipation of splitting.
+
+\item \Cref{sec:splitting} describes the splitting transformation on \lang.
+As stated above, splitting accepts terms in \lang\ as input, and converts them to terms in a standard unstaged language.
+This section also formally defines the correctness for the splitting transformation.
+
+\item \Cref{sec:examples} gives several examples of how splitting terms in \lang\ can be used
+to derive data structures that are communicated between the two unstaged terms, and discusses the effectiveness and limitations of this technique.
+Unlike prior work, where the data communicated between the resultant unstaged terms was a basic tuple of values, splitting \lang\ terms can generate non-trivial intermediate structures, such as trees.
+\end{itemize}
 
 \input{semantics}
-
-\vfill\eject\pagebreak
 \input{splitting}
-
-\input{relatedwork}
+\input{algoexamples}
 
@@
 
-\bibliography{paper}
+\bibliography{techreport}
 \bibliographystyle{abbrvnat}

### 2014-08-29 paper/semantics.tex
Front the algo examples section with a bit of explanitory text.
Rest of the edits are small nits when reading.

@@
 The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
-while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  These are the
+while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  $\next$ and $\prev$ are the
 only ways in \lang\ to alter the stage of a term; we surround their arguments
@@
 \subsection{Staged Evaluation}
+\label{sec:stagedsemantics}
 
@@
 Although the code for {\tt fexp} appears very similar to the unstaged mathematical
-definition above, it is in fact a staged program: the {\tt if} predicates and exponent
+definition above, it is in fact a staged program.
+The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns stage-\bbtwo\ result.
+The {\tt if} predicates and exponent
 decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
@@
 \end{itemize}
+
+% would like some kind of intro here that anticipates all the complexity.
+% The dynamics of \lang\ consists of three types of judgements: $\redonesym$ (stage-\bbone reduction), $\redtwosym$ (speculation), and $\reifysym$ (reification).
+
 The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future

### 2014-08-29 paper/splitting.tex
Front the algo examples section with a bit of explanitory text.
Rest of the edits are small nits when reading.

@@
 
-The basic idea is that we want to take a \lang\ term which contains interleaved stage \bbone\ and stage \bbtwo\ code
-and split it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.
+The stage splitting operation takes as input a \lang\ term which contains interleaved stage \bbone\ and stage \bbtwo\ code
+and splits it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.  
 To facilitate communication between the stages, there is a data structure passed between the stages, 
@@
 As with the dynamics, the form of the splitting statement depends on the external stage of the term.
-We consider splitting for stage \bbtwo\ terms first, as it has a simpler form.  
-For example, take \verb|hold{1+2} < 5|.
+We consider splitting of stage \bbtwo\ terms first, as it has a simpler form.  
+For example, consider \verb|hold{1+2} < 5|.
 We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 

### 2014-08-29 paper/techreport.tex
Front the algo examples section with a bit of explanitory text.
Rest of the edits are small nits when reading.

@@
 This document is a tech report on ongoing research into a program transformation
-technique which we call {\em splitting}.  The splitting operation statically transforms a program written in a modal language into multiple unstaged (monostage) terms.
+technique which we call {\em stage splitting}.  The stage splitting operation statically transforms a program written in a modal language into multiple unstaged (monostage) terms.
 
 The idea of splitting has previously been explored as an optimization technique that improves performance in situations where multi-variate functions are invoked many times, with some inputs being invariant across invocations and others being dynamic (varying across invocations).
-Prior examples of splitting simple languages include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96}, as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+Prior examples of splitting include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
 In this document we define the splitting transformation for the two-stage language \lang, a typed lambda calculus featuring products, sums, and isorecursive types.

### 2014-08-29 paper/semantics.tex
added more about a partial evaluation system

@@
 
-%The grammar and type system for \lang\ is given in
-%\ref{fig:grammar,fig:statics}.
+\lang\ is largely based off of a language defined in \cite{davies96} 
+which was used as a model for partial evaluation.  
+Our version, despite a slightly different semantics, can be used for the same purpose.
+The details of the partial evaluation view are shown in \cref{sec:partialeval}.
+
 The language \lang\ is essentially two copies of a typed lambda calculus with
@@
 \subsection {A Partial Evaluation System}
+\label{sec:partialeval}
 
 We now consider how to use this formal system to encode inputs to partial evaluation and stage-splitting systems.
-Recall that in partial evaluation systems, we have a function $f$ for which some input is labeled {\em static} and 
+Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
 some labeled {\em dynamic}.  Once the static input is provided, we can partially evaluate the function to produce
-a residual that depends on only the dynamic input.
+a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
+\[
+	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
+\]
+where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
+Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is residual, also called ``f specialized to x."
+The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must evaluate it many times.
 
 By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
-we can encode this sort of system with a function type of the form $A\to\fut(B\to C)$.
+our dynamics can serve as a partial evaluator.   
+Specifically, we encode $f$ as a \lang\ expression with a function type of the form $A\to\fut(B\to C)$
 \cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
@@
     }
-\end{lstlisting}}
+\end{lstlisting}}.
 %

### 2014-08-29 paper/splitting.tex
added more about a partial evaluation system

@@
 
-The rules for value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
+The rules for partial value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
 which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
-Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original value.  
-At base types, value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
+Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original partial value.  
+At base types, partial value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
 in accordance with the precept that base types have only trivial content at stage \bbtwo.
-Splitting essentially distributes into values at product, sum, and recursive types.
+Splitting acts distributively at product, sum, and recursive types.
 Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 

### 2014-08-29 paper/splitting.tex
worked on intro to splitting

@@
 
-The stage splitting operation takes as input a \lang\ term which contains interleaved stage \bbone\ and stage \bbtwo\ code
-and splits it into two separate terms: one with all of the stage \bbone\ code and one with all the stage \bbtwo\ code.  
-To facilitate communication between the stages, there is a data structure passed between the stages, 
-which is an output of the stage \bbone\ term and and input to the stage \bbtwo\ term.
+Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
+Despite this similarity in purpose, the form of stage-splitting is rather different. 
+Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
+where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
+and $f_2$ uses that data structure and the stage-\bbtwo\ input to produce the final output.  
+More precisely, a stage splitter is any $s$ such that
+\[
+	\forall f. \exists f_1,f_2. 
+	\left[
+		\begin{array}{l}
+		s(f) = (f_1,f_2) \text{ and } \\
+		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
+		\end{array}
+	\right]
+\]
+with the same $\llbracket \cdot \rrbracket$ notation as before.
+Importantly, the stage splitting operation does not depend on the stage-\bbone\ input, unlike partial evaluation.
 
-As with the dynamics, the form of the splitting statement depends on the external stage of the term.
+As with the dynamics, our stage-splitting algorithm for expressions in \lang\ take 
+the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$. 
+Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
+
+\subsection{Partial Value Splitting}
+
+Before getting into the formal presentation, 
+we first consider the problem of splitting stage-\bbone\ partial values.
+This serves as a didactic stepping-stone to full term-splitting, 
+and it is a necessary component of the correctness theorem.
+
+The rules for partial value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
+which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
+Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original partial value.  
+At base types, partial value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
+in accordance with the precept that base types have only trivial content at stage \bbtwo.
+Splitting acts distributively at product, sum, and recursive types.
+Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 
+general term splitting, which is covered momentarily.
+
+Lastly, we consider value splitting at the $\fut$ type.  The only partial value with this type
+is a $\next$ block containing a reference into the future environment.  This splits into a 
+nullary tuple and the same variable, again aligning with the precept that $\fut$ types contain trivial stage-\bbone\
+content.
+
+\subsection{Full Expression Splitting}
+
 We consider splitting of stage \bbtwo\ terms first, as it has a simpler form.  
@@
 This is straight-forward, given a careful definition of equivalence.
-
-\subsection{Value Splitting}
-
-Before getting into the formal presentation, 
-we first consider the problem of splitting stage-\bbone\ partial values.
-This serves as a didactic stepping-stone to full term-splitting, 
-and it is a necessary component of the correctness theorem.
-
-The rules for partial value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
-which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
-Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original partial value.  
-At base types, partial value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
-in accordance with the precept that base types have only trivial content at stage \bbtwo.
-Splitting acts distributively at product, sum, and recursive types.
-Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 
-general term splitting, which is covered momentarily.
-
-Lastly, we consider value splitting at the $\fut$ type.  The only partial value with this type
-is a $\next$ block containing a reference into the future environment.  This splits into a 
-nullary tuple and the same variable, again aligning with the precept that $\fut$ types contain trivial stage-\bbone\
-content.
 

### 2014-08-30 paper/splitting.tex
fixed up the partial value splitting section

@@
 Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
+After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.
 
@@
 
-Before getting into the formal presentation, 
-we first consider the problem of splitting stage-\bbone\ partial values.
-This serves as a didactic stepping-stone to full term-splitting, 
-and it is a necessary component of the correctness theorem.
+We now consider the problem of splitting stage-\bbone\ partial values.
+This serves as a simple didactic stepping-stone to the full formulation, 
+and we will see later that it is a necessary component of the correctness theorem.
 
-The rules for partial value splitting are given in \cref{fig:valSplit}.  It consists of a single judgment, $v \vsplito [v_1,v_2]$,
-which sends a partial value $v$ to two single-stage values $v_1$ and $v_2$. 
-Those respectively represent the stage-\bbone\ and stage-\bbtwo\ content of the original partial value.  
-At base types, partial value splitting assigns all content to stage \bbone, and puts in a nullary tuple for stage two,
-in accordance with the precept that base types have only trivial content at stage \bbtwo.
-Splitting acts distributively at product, sum, and recursive types.
-Functions split into two other functions.  Since the body of a function is an open term, this rule must rely on 
-general term splitting, which is covered momentarily.
+The basic goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
+where $v_1$ contains only the information from $v$ that is relevant to the first stage, 
+and $v_2$ contains only the information from $v$ that is relevant to the second stage.
+There are many reasonable ways to do this, 
+and we simply chose one that preserves as much structure as possible from $v$.
+Formally, partial value splitting comprises a single judgment, $v \vsplito [v_1,v_2]$, the rules for which are given in \cref{fig:valSplit}.
+It essentially acts distributively at product, sum, and recursive types, 
+doing little more than recuring into subvalues and duplicating the structure to both results.
+At base types, partial value splitting assign all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
+Conversely at $\fut$ types, splitting assigns all content ({\em i.e.} the environment reference) to stage \bbtwo, and uses a nullary tuple for stage \bbone.
 
-Lastly, we consider value splitting at the $\fut$ type.  The only partial value with this type
-is a $\next$ block containing a reference into the future environment.  This splits into a 
-nullary tuple and the same variable, again aligning with the precept that $\fut$ types contain trivial stage-\bbone\
-content.
+Functions are inherently more complicated. 
+Since the body of a function is an open term, the partial value splitting rule for functions must rely on general expression splitting, which is covered momentarily.
+The most we can say here is a function splits into two other functions, 
+whose bodies correspond to the stage \bbone\ and stage \bbtwo\ components of the original function's body.
 

### 2014-08-30 paper/semantics.tex
Minor changes on partial evaluation.

@@
 
-\lang\ is largely based off of a language defined in \cite{davies96} 
-which was used as a model for partial evaluation.  
-Our version, despite a slightly different semantics, can be used for the same purpose.
-The details of the partial evaluation view are shown in \cref{sec:partialeval}.
+% \lang\ is largely based off the
+% language \lamCircle\ by \cite{davies96}.  
+% which was used as a model for partial evaluation.  
+% Our version, despite a slightly different semantics, can be used for the same purpose.
+% The details of the partial evaluation view are shown in \cref{sec:partialeval}.
 
-The language \lang\ is essentially two copies of a typed lambda calculus with
+\lang\ is essentially two copies of a typed lambda calculus with
 products, sums, and isorecursive types.%
@@
 
-In addition, there is a stage-\bbone\ type $\fut A$, containing encapsulated
+The stage-\bbone\ type $\fut A$ contains encapsulated
 stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
-opaquely by stage-\bbone\ code, as they cannot be evaluated until stage \bbtwo.
+opaquely by stage-\bbone\ code, as (by the requirements of staging) they cannot be evaluated until stage \bbtwo.
 While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
@@
 definition above, it is in fact a staged program.
-The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns stage-\bbtwo\ result.
+The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns a stage-\bbtwo\ result.
 The {\tt if} predicates and exponent
@@
 \verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
+
 \item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
@@
 \verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
+
 \item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
@@
 
-We now consider how to use this formal system to encode inputs to partial evaluation and stage-splitting systems.
+The dynamics described in the previous section provide a formal description of partial evaluation.
 Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
-some labeled {\em dynamic}.  Once the static input is provided, we can partially evaluate the function to produce
+some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
 a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
@@
 Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is residual, also called ``f specialized to x."
-The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must evaluate it many times.
+
+% KAYVONF: good statement, but hold out for now
+%The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must %evaluate it many times.
 

### 2014-08-30 paper/techreport.tex
Minor changes on partial evaluation.

@@
 This document is a tech report on ongoing research into a program transformation
-technique which we call {\em stage splitting}.  The stage splitting operation statically transforms a program written in a modal language into multiple unstaged (monostage) terms.
+technique which we call {\em stage splitting}, which statically transforms a program written in a multi-stage (modal) language into multiple unstaged (monostage) terms.
 
@@
 \begin{itemize}
-\item \Cref{sec:semantics} describes \lang which unambiguously specifies the inputs to 
-our splitting transformation.  The type system of \lang\ is drawn from prior work, but its semantics are uniquely defined in anticipation of splitting.
+\item \Cref{sec:semantics} describes \lang\ which serves to unambiguously specify the input to splitting.
+The type system of \lang\ is drawn from prior work, but its semantics are uniquely defined in anticipation of splitting.
 

### 2014-09-01 paper/techreport.tex
Expanded the intro the encompass all of the interelated issues.

@@
 
-This document is a tech report on ongoing research into a program transformation
-technique which we call {\em stage splitting}, which statically transforms a program written in a multi-stage (modal) language into multiple unstaged (monostage) terms.
+It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).
+This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).
+Examples include,
+\begin{itemize}
+\item Computing $b^e$, where $b$ varies and $e$ is constant.
+\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
+\item Computing the shading of a surfacing in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
+\end{itemize}
 
-The idea of splitting has previously been explored as an optimization technique that improves performance in situations where multi-variate functions are invoked many times, with some inputs being invariant across invocations and others being dynamic (varying across invocations).
-Prior examples of splitting include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
-In this document we define the splitting transformation for the two-stage language \lang, a typed lambda calculus featuring products, sums, and isorecursive types.
+With these goals in mind, our initial concern is how to express multistage programs.
+To this end, we have slightly modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
+An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.
 
-\begin{itemize}
-\item \Cref{sec:semantics} describes \lang\ which serves to unambiguously specify the input to splitting.
-The type system of \lang\ is drawn from prior work, but its semantics are uniquely defined in anticipation of splitting.
+Once we have appropriately written {\em multi-stage} programs, we then must execute them in a way that respects the staging.
+In this paper, we explore two techniques for this: {\em partial evaluation} and {\em stage-splitting}.
 
-\item \Cref{sec:splitting} describes the splitting transformation on \lang.
-As stated above, splitting accepts terms in \lang\ as input, and converts them to terms in a standard unstaged language.
-This section also formally defines the correctness for the splitting transformation.
+The main idea of the former technique is to evaluate all of the constant portions of the program and embed the partial results into a {\em residual} program that depends only on the varying inputs.
+This is an old and well-explored idea, though the typical setup of partial evaluation implies a more powerful binding time analysis than is necessary for our language.
 
-\item \Cref{sec:examples} gives several examples of how splitting terms in \lang\ can be used
-to derive data structures that are communicated between the two unstaged terms, and discusses the effectiveness and limitations of this technique.
-Unlike prior work, where the data communicated between the resultant unstaged terms was a basic tuple of values, splitting \lang\ terms can generate non-trivial intermediate structures, such as trees.
-\end{itemize}
+The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way.
+This has previously been explored as an optimization technique for languages with fewer features or a more constrained staging model than \lang.
+Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+
+\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator.
+\Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
+Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.
+Notably, these examples make use of the stage-splitting features not available in prior work.
 

### 2014-09-02 paper/splitting.tex
Minor changes on partial evaluation.
Lots of improvements to splitting.

@@
 Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
-Despite this similarity in purpose, the form of stage-splitting is rather different. 
+However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
+before any arguments are available.
 Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
@@
 with the same $\llbracket \cdot \rrbracket$ notation as before.
-Importantly, the stage splitting operation does not depend on the stage-\bbone\ input, unlike partial evaluation.
+Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
+meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
+This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.
 
-As with the dynamics, our stage-splitting algorithm for expressions in \lang\ take 
+As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
 the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$. 
@@
 and $v_2$ contains only the information from $v$ that is relevant to the second stage.
-There are many reasonable ways to do this, 
-and we simply chose one that preserves as much structure as possible from $v$.
+We call this action {\em masking}.
+
+There are many reasonable ways to achieve this goal, 
+and we choose one that preserves as much structure as possible from $v$.
 Formally, partial value splitting comprises a single judgment, $v \vsplito [v_1,v_2]$, the rules for which are given in \cref{fig:valSplit}.
@@
 
-\subsection{Full Expression Splitting}
+\subsection{ Expression Splitting}
 
-We consider splitting of stage \bbtwo\ terms first, as it has a simpler form.  
-For example, consider \verb|hold{1+2} < 5|.
-We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
-which are called the {\em precomputation} and {\em residual}\footnote{We also used the term {\em residual} for the result of partial evaluation.
-Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
-and {\em evaluation-residual}.}, respectively.
+We now consider the problem of splitting expressions in \lang, starting with stage \bbtwo\ terms.  
+The goal here is to end up with two new terms: one that has all of the stage \bbone\ content of the original, and one with all of the stage \bbtwo\ content. 
+These are respectively called the {\em precomputation} and {\em residual}.
+\footnote{We also used the term {\em residual} for the result of partial evaluation.
+Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual} and {\em evaluation-residual}.}
 
-The form of splitting for stage \bbone\ terms is more complicated.
+For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
+We can split this into two terms, $p=``\verb|(1+2,3+4)|"$ and $r=``\verb|fn (L1,L2) => L1 < L2-5|"$, 
+with the as yet imprecise property that $r$ applied to $p$ yields the same final value as $e$.
+In general, all we're doing here is finding the contained nuggets of stage \bbone\ code and pulling them into a big tuple,
+an action we call {\em lifting}.
+
+Note that lifting and masking are different, despite both being within the purview of splitting.
+They particularly differ in how the resulting stage \bbone\ parts are used.
+The $v_1$ of partial value splitting is essentially the stage \bbone\ content of {\em result} of computation, 
+and it's accessible to whatever surrounding piece of code refers to the value being split.
+In contrast, the $p$ of stage \bbtwo\ expression splitting is only meant to support the stage \bbtwo\ part,
+and is not accessible externally.
+
+The distinction between lifting and masking is important because stage \bbone\ expression splitting does both.
 To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
-As before, the comparison operation is part of stage \bbtwo, 
-and the addition operation is a stage \bbone\ precomputation, the result of which will eventually become input to the residual.
-The multiplication operation, however, is neither part of the residual nor a precomputation to support it.
-Instead it is the {\em immediate result} of stage \bbone, 
-because it is available to the stage \bbone\ context around our original term.
-For example:
-\begin{lstlisting}
-let x = (next{hold{1+2} < 5},3*4) in
-next{ if prev{#1 x} then 7 else hold {#2 x} }
-\end{lstlisting}
-To support this, stage \bbone\ splitting produces two outputs: the {\em combined result} and the residual,
-where the combined result reduces to a tuple containing the immediate result and the precomputation.
+The stage \bbone\ result of masking is \verb|((),3*4)|, 
+and the precomputation (i.e. the stage \bbone\ result of lifting) is \verb|(1+2)|.
+To get the total stage \bbone\ result of splitting, we just take the tuple of those two parts: \verb|(((),3*4),1+2)|.  
+This is called the {\em combined result} for stage \bbone.
+\footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}
 
-We must also develop a notion of what it means for splitting to be correct.  
-Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
-In general, if a multi-stage program $P$ splits into stage-\bbone\ part $P_1$ and stage-\bbtwo\ part $P_2$, then
-\begin{enumerate}
-\item Partial evaluation of $P$ will terminate if and only if evaluation of $P_1$ terminates, and if so then
-\item the immediate results of each method will be equivalent, and
-\item the evaluation-residual and the splitting-residual ($P_2$ bound under the precomputation) are equivalent.
-\end{enumerate}
-
-This is straight-forward, given a careful definition of equivalence.
+The residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
+In this case, it is ``\verb|fn L => (L<5,())|".
 
@@
 
-\subsection {Speculation}
+\subsection {Properties}
+
+We'll now call out a few properties to notice in the splitting rules.
+
+\subsubsection {Speculation}
 
@@
 
-\subsection {Stage \bbone\ Divergence}
+\subsubsection {Stage \bbone\ Divergence}
 Consider the splitting rules for stage-\bbone\ {\tt if} and {\tt case} expressions.
@@
 
-\subsection {Stage \bbone\ Functions}
+\subsubsection {Stage \bbone\ Functions}
 
@@
 
-\subsection{Metatheory}
+\subsection{Correctness of Splitting}
 
-We start with two mutually dependent definitions of equivalence.  
+We can now develop an notion of what it means for splitting to be correct.  
+Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
+Thus, we start with two mutually dependent definitions of equivalence.  
 Both relate evaluation-residuals on the left with splitting residuals on the right,
-but $\equiv$ operates on terms, whereas $\sim$ on values.
+but $\equiv$ equates closed expressions, whereas $\sim$ equates values.
 
@@
 Substitution splitting works just like value splitting, which is why they use the same symbol.
-
-The jury is still out on how to make these strong enough to prove that the partitioning between stages is correct.
 

### 2014-09-02 paper/techreport.tex
Lots of improvements to splitting.

@@
 With these goals in mind, our initial concern is how to express multistage programs.
-To this end, we have slightly modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
+To this end, we have modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
 An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.
@@
 Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+Unlike partial evaluation, stage splitting is purely a static operation and is not dependent on the input data.
 

### 2014-09-03 paper/splitting.tex
Fix three minor typos.

@@
 doing little more than recuring into subvalues and duplicating the structure to both results.
-At base types, partial value splitting assign all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
+At base types, partial value splitting assigns all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
 Conversely at $\fut$ types, splitting assigns all content ({\em i.e.} the environment reference) to stage \bbtwo, and uses a nullary tuple for stage \bbone.

### 2014-09-03 paper/techreport.tex
Fix three minor typos.

@@
 \item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
-\item Computing the shading of a surfacing in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
+\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
 \end{itemize}

### 2014-09-03 paper/semantics.tex
Fix three minor typos.
Addressed Kayvon's comments in section 1-3

@@
 language called \langTwo. Then, we perform the remainder of the evaluation with 
-$\tworedsym$, an ordinary dynamic semantics for \langTwo.
+$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
+judgment are not shown, but they are standard).
 
@@
 The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
-C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
-enclosing $v$, via $\reify\xi{v}{f_a}$. By \ref{thm:reify-type}, the resulting
-residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
+C)$, which by canonical forms must have the form $v = \next~\hat y$. 
+Next, we reify this environment into a sequence of \verb|let|-bindings
+enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
+Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
 and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.

### 2014-09-03 paper/splitting.tex
Addressed Kayvon's comments in section 1-3

--- 2014-09-02 paper/splitting.tex
@@
 where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
-and $f_2$ uses that data structure and the stage-\bbtwo\ input to produce the final output.  
+and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
 More precisely, a stage splitter is any $s$ such that
@@
 As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
-the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$. 
+the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$, 
+which respectively split terms that type $@ \bbone$ and $@ \bbtwo$. 
 Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
@@
 
-The basic goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
-where $v_1$ contains only the information from $v$ that is relevant to the first stage, 
-and $v_2$ contains only the information from $v$ that is relevant to the second stage.
+Recall that, along with a future environment, partial values are the output of the partial evaluation of stage \bbone\ terms.
+In addition to base types, tuples, and injections, partial values may also contain references into the future environment.
+In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
+The goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
+where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 
+and $v_2$ contains only the information from $v$ that is likewise relevant to second stage code.
 We call this action {\em masking}.
@@
 \footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}
-
-The residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
-In this case, it is ``\verb|fn L => (L<5,())|".
+The splitting residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
+For our example, it is ``\verb|fn L => (L<5,())|".
 
@@
 We give the following end-to-end correctness lemmas for open terms. 
-It's a bit of a mess currently, but the $\Gamma$ is supposed to be all of the stage \bbone\ bindings, 
-whereas $\Gamma'$ is the stage \bbtwo\ bindings.
+In some of the theorems, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
+which respectively hold all of the stage \bbone\ and stage \bbtwo\ variable bindings.
 Substitution splitting works just like value splitting, which is why they use the same symbol.

### 2014-09-03 paper/techreport.tex
Addressed Kayvon's comments in section 1-3

--- 2014-09-01 paper/techreport.tex
@@
 With these goals in mind, our initial concern is how to express multistage programs.
-To this end, we have slightly modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
+To this end, we have modified prior work (\cite{davies96}) to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
 An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.
@@
 
-The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way.
-This has previously been explored as an optimization technique for languages with fewer features or a more constrained staging model than \lang.
+The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.
+This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang.
 Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.
 
-\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator.
+\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging.
 \Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.

### 2014-09-05 paper/semantics.tex
Minor grammatical improvements for clarity.

@@
 
-When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
+When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
 searches into the block's stage \bbtwo\ content to find any contained stage
@@
 Importantly, speculation will search into both branches of a stage-\bbtwo\ {\tt if}
-statement to evaluate the contained stage-\bbone\ code.  Thus the name.
+statement to evaluate all contained stage-\bbone\ code.  Thus the name.
 }.  The context is then reified (using the $\reifysym$
@@
 From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
-our semantics terminates strictly less often than that of \cite{davies96}.
+{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
 
@@
 where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
-Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is residual, also called ``f specialized to x."
+Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."
 

### 2014-09-05 paper/splitting.tex
Minor grammatical improvements for clarity.

@@
 However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
-before any arguments are available.
+before {\em any arguments} are available.
 Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
@@
 In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
-The goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
+The goal of partial value splitting is to take a partial value $v$ and from it create two single stage values $v_1$ and $v_2$,
 where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 

### 2014-09-05 paper/semantics.tex
Incorporated Kayvon's suggestions.

@@
 \noindent
-The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
-%\footnote{We render $\fut$ as {\tt \$} in code examples.}
+The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:
 
@@
 Achieving this behavior mechanically requires us to resolve a contradiction:
-we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
+we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
 Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.
@@
 \end{lstlisting}
-To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them. 
+To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
 As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
@@
 from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
-$v$) to residuals---in our example, we represented this environment with
-explicit substitutions. For all of the normal features of \lang
+$v$) to residuals---in our example above, we represented this environment with
+explicit substitutions. For all of the normal features of \lang\ 
 (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
-effect on values as standard evaluation, and the final future environment is
+effect on (partial) values as does standard evaluation, and the final future environment is
 gotten by merging the future environments of the subterms.
@@
 searches into the block's stage \bbtwo\ content to find any contained stage
-\bbone\ subexpressions and evaluate them in place.  We call this search process
-\emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes
+\bbone\ subexpressions and evaluate them in place.  This search process, called
+\emph{speculation}, is implemented by the $\redtwosym$ judgment, which takes
 a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
-speculated into a residual ($q$), we then return a fresh variable wrapped in a
+speculated into a residual ($q$), the output of $\redonesym$ is a fresh variable wrapped in a
 $\next$ block ($\next~\hat y$), along with a future environment which maps that
-variable to our residual ($\hat y \mapsto q$).
+variable to the residual ($\hat y \mapsto q$).
 
@@
 nothing but recursively speculate into every subexpression.  Once speculation
-finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which
+finds a $\prev$ block, it resumes stage \bbone\ evaluation of the contents, which
 produces a future context and (by canonical forms) a $\next$-wrapped variable
-($\next\{\hat y\}$)\footnote{
-Importantly, speculation will search into both branches of a stage-\bbtwo\ {\tt if}
-statement to evaluate all contained stage-\bbone\ code.  Thus the name.
-}.  The context is then reified (using the $\reifysym$
-judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of
-its $\next$---at the bottom.
+($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
+judgment) into a series of let bindings with $\mathtt{\hat y}$, stripped of
+its $\next$, at the bottom.
+
+Within speculation lies a subtle---if perhaps unintuitive---feature.  
+Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
+statement in its search for stage-\bbone\ code. 
+Thus the evaluation of that stage one code will occur {\em regardless of the eventual value of the predicate},
+and so a term like 
+\begin{lstlisting} 
+next{
+  if true 
+  then hold{1+2} 
+  else prev{spin() (* loops forever *)}
+}
+\end{lstlisting}
+will fail to evaluate at stage \bbone.
+This behavior is why the judgment is named ``speculation."
 
@@
 \end{mathpar}
-and treat $\next~q$ as a partial value, where $q$ is a residual,
-then we get precisely the semantics from \cite{davies96}. This essentially bypasses the
+and treat $\next~q$ as a partial value for any residual $q$,
+then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
 environment bookkeeping in $\redonesym$, by inlining residuals instead of
@@
 The dynamics described in the previous section provide a formal description of partial evaluation.
-Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
+Recall that in partial evaluation systems, we start with a multivariate function $f$ for which some input is labeled {\em static} and 
 some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces

### 2014-09-05 paper/splitting.tex
Incorporated Kayvon's suggestions.

@@
 
-\subsection{ Expression Splitting}
+\subsection{Expression Splitting}
 
@@
 
-The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$.
+The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
+which are in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
 For both, the input is a term in \lang, and the output is two terms in \langmono, an
@@
 We'll now call out a few properties to notice in the splitting rules.
+
+\subsubsection {Expression Splitting at Values}
+
+Note that general expression splitting always produces an effectively trivial precomputation when applied to partial values,
+where effectively trivial means that the precomputation is either a unit value, or a tuple or injection of an effectively trivial values.
+
+Partial value splitting is more than just a restriction of general expression splitting 
+precisely because it observes this property it at the structural level, doing away with the precomputation entirely.
+This makes sense intuitively, since a precomputation would imply that there is some computation left to do, 
+which is antithetical to the notion of values.
 
@@
 We give the following end-to-end correctness lemmas for open terms. 
-In some of the theorems, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
+Those readers not interested in the gritty details are advised to skip ahead to the theorems for closed terms, which give the important gist.
+
+In some of the lemmas, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
 which respectively hold all of the stage \bbone\ and stage \bbtwo\ variable bindings.

### 2014-09-08 paper/semantics.tex
commented out some duplicate stuff

@@
 
-\begin{remark}
-For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
-always compute an empty environment, and a partial value identical to the result
-of call-by-value evaluation of $e$.
-%derivationally equivalent to standard call-by-value evaluation.
-\end{remark}
+%\begin{remark}
+%For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
+%always compute an empty environment, and a partial value identical to the result
+%of call-by-value evaluation of $e$.
+%%derivationally equivalent to standard call-by-value evaluation.
+%\end{remark}
 
@@
 
-\TODO
-Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
-integer term is guaranteed not to depend on the table, although one might be
-produced. One may either discard the table, or evaluate everything in the table
-(and terminating with the partial value iff everything in the table terminates).
+%\TODO
+%Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
+%integer term is guaranteed not to depend on the table, although one might be
+%produced. One may either discard the table, or evaluate everything in the table
+%(and terminating with the partial value iff everything in the table terminates).

### 2014-09-23 paper/semantics.tex
implemented some suggested changes

@@
 there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
+That is, there is a one-way dependence between stages at the term level.
+
+Alternatively, the stages are mutually dependent at the term level.
 The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
@@
 
-The grammar and type system for \lang\ is given in
-\ref{fig:grammar,fig:statics}. (Typing judgments and context variables are annotated with stages.)
+The grammar and type system for \lang\ is given in \ref{fig:grammar,fig:statics}. 
+Typing judgments and context variables are annotated with stages after an $@$ symbol.
 Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
@@
 Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
-Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality\,\cite{cave14}.
+Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality,\cite{cave14}.
 The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
@@
 
-\subsection{Staged Evaluation}
-\label{sec:stagedsemantics}
-
-As a simple example of expressing staged programs in \lang, consider the following fast exponentiation algorithm, which 
-calculates $b^p$ in $\log p$ time:
-\[
-	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
-		1 &  p = 0 \\ 
-		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
-		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
-	\right .
-\]
-
-% Ignoring for the moment how to define recursive functions in \lang,
-\noindent
-The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:
-
+Recall the fast exponent function from above.
+We present a valid staging of it here:
 \begin{lstlisting} 
-let fexp (b : $int, e : int) : $int =
-	if e == 0 then
+let fexp (b : $int, p : int) : $int =
+	if p == 0 then
 		next{1}
-	else if (e mod 2) == 0 then
+	else if (p mod 2) == 0 then
 		next{let x = prev{fexp(b,p/2)} in x*x}
@@
 
-Although the code for {\tt fexp} appears very similar to the unstaged mathematical
-definition above, it is in fact a staged program.
+\TODO explain why we staged this the way we did
+
 The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns a stage-\bbtwo\ result. 
-The {\tt if} predicates and exponent
-decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
-%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
-%the decomposition of the exponent can be done at the stage \bbone\ without
-%requiring disruption of the elegant functional structure of the code.}
-%
-% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
-%
-%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs.
-We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
-%
-A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
-however in the case that stage \bbone\ expressions are contained inside
-stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.
+The {\tt if} predicates and exponent decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
+
+\subsection{Staged Evaluation}
+\label{sec:stagedsemantics}
+
+A key attribute of the fexp example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
+Ordinary term evaluation eliminates outermost redexes first, 
+however in the case that stage \bbone\ expressions are contained inside stage \bbtwo\ ones 
+(such as the recurive call to {\tt fexp} above), 
+this strategy conflicts with the precept of staged execution: 
+that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.
 
@@
 \begin{lstlisting}
-#2 (next {1+2}, 3+4)
+#2 (next {fib 20}, 2+3)
 \end{lstlisting}
 This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
-stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
-call-by-value evaluation strategy demands that we evaluate both components of
+stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. 
+In this example, ${\tt fib} : \rmint \to \rmint$ is a stage-\bbtwo\ reference to the Fibonacci function.
+A conventional call-by-value evaluation strategy demands that we evaluate both components of
 the pair to values before we project from the pair. The problem is that
-while \verb|next {1+2}| is not a value (in the sense that additional
-stage-\bbtwo\ computation steps are necessary to produce the result \verb|3|), evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
-Intuitively, the solution is to designate \verb|next {1+2}| as a value \emph{in
+while \verb|next {fib 20}| is not a value (in the sense that additional
+stage-\bbtwo\ computation steps are necessary to produce the result \verb|6765|), 
+evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
+Intuitively, the solution is to designate \verb|next {fib 20}| as a value \emph{in
 stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
@@
 \begin{lstlisting}
-#2 (next {1+2}, 7)
+#2 (next {fib 20}, 5)
 \end{lstlisting}
-then obtain \verb|7| as the result of projection.
+then obtain \verb|5| as the result of projection.
 
@@
 prev{
-  let x = (next {1+2}, 3+4) in
+  let x = (next {fib 20}, 3+4) in
   next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
@@
 Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
-depends on the value of \verb|1+2|, which is not reduced to \verb|3| until
+depends on the value of \verb|fib 20|, which is not reduced to \verb|3| until
 stage \bbtwo.
 
-If we simply treat \verb|(next {1+2}, 7)| as a value during stage~\bbone, and
+If we simply treat \verb|(next {fib 20}, 7)| as a value during stage~\bbone, and
 substitute it for the three occurrences of \verb|x| in the body of the
@@
 prev{
-  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
+  next{ 
+    prev{next {fib 20}} * prev{next {fib 20}} * 7 
+  }
 }
@@
 \begin{lstlisting} 
-(1+2) * (1+2) * 7
+(fib 20) * (fib 20) * 7
 \end{lstlisting}
-Stage-\bbtwo\ evaluation of this expression will compute \verb|1+2| twice. Of
-course, this is not particularly expensive, but in general, the duplicated
-computation could be anything, even a recursive call to the function! To avoid duplicating
-computations, we take a different approach.  Instead of
+Note how stage-\bbtwo\ evaluation of this expression will compute \verb|fib 20| twice.  
+To avoid duplicating computations, we take a different approach.  Instead of
 duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
@@
 \begin{lstlisting} 
-let y = 1+2 in y * y * 7
+let y = fib 20 in y * y * 7
 \end{lstlisting}
@@
 prev {
-[yhat|->1+2]
+[yhat|->fib 20]
   let x = (next{yhat}, 7) in
@@
 prev {
-[y|->1+2]
+[yhat|->fib 20]
   next{
@@
 prev {
-[yhat|->1+2]
+[yhat|->fib 20]
     next{ yhat * yhat * 7 }
@@
 prev {
-[yhat|->1+2]
+[yhat|->fib 20]
 [zhat|->yhat*yhat*7]
@@
 \begin{lstlisting} 
-let yhat = 1+2 in
+let yhat = fib 20 in
 let zhat = yhat * yhat * 7 in z
@@
 \input {figures/dynamics}
-
-%{{{ stuff nico commented out
-%Previous work (\cite{davies96}) focuses on the correspondence between the type system and existing temporal logics, whereas we care more about operational behavior and cost.
-In this section, we'll consider a few proposals for our language before settling on one we like.
-All of the proposals are call-by-value, differing primarily in how they handle values of $\fut$ type.
-
-%\subsection{The Erasure Semantics}
-%
-%We first consider the {\em erasure semantics}, so called because it corresponds to what one would get by interpreting \lamStaged as a single-stage language, ignoring all of the $\next$ and $\prev$ terms.
-This gives us two judgments, $\erasone$ and $\erastwo$, corresponding to {\em multistage evaluation} at \bbone and at \bbtwo.
-We call these judgments  ``multistage'' because they cause work to happen at both stages.
-%
-%Both judgments behave normally at non-staged features.  We cover their behavior at staged features below:
-%
-%\begin{mathpar}
-%\infer {\next~e \erasone \next~v} {e \erastwo v} \and
-%\infer {\prev~e \erastwo \next~v} {e \erasone v} \and
-%\infer {\pause~e \erasone \next~i} {e \erasone i}
-%\end{mathpar}
-%
-%Essentially, we immediately evaluate under the $\next$, yielding a value for $\fut$ types.
-The $\prev$ terms remove this wrapper.
-As expected, $\pause$ also gives us a way to inject into the wrapper at integers.
-%
-%The erasure semantics has some undesirable properties.
-By intention, it interleaves the execution of stage-\bbone~and stage-\bbtwo~code, so the evaluation can't really be said to be staged (i.e.
-stage-\bbone~work is done before stage-\bbtwo~work).
-Moreover, the erasure semantics cannot be equivalent to any semantics which does have this property!
-To see why, consider the following code, which types to int at \bbtwo:
-%
-%\begin{lstlisting} 
-%if 5*4*3*2 > 111 then
-%	hold{2+4}
-%else
-%	prev{ loopForever () (* does not terminate *) }
-%\end{lstlisting}
-%
-%Under the erasure semantics (using $\erastwo$), this code takes the top branch and evaluates to 6.
-But in order know that the {\tt loopForever} function need not be called, the predicate had to be evaluated prior.
-But the predicate is stage-\bbtwo, whereas {\tt loopForever} is stage-\bbone.
-To borrow terminology from \cite{cave14}, this violates causality.
-In order to avoid this problem, a valid semantics must {\em speculate} down the branches of any stage-\bbtwo~if or case statement (or similarly into the body of a stage-\bbtwo function) to find and evaluate all of the stage-\bbone~code.
-Both of the other semantics we will consider have this property.
-%
-%The benefit of the erasure semantics is that it's very natural, and has a familiar cost model.
-It also {\em obviously} produces the ``correct'' answer, so we can use the erasure semantics as a reference to prove the reasonableness of any other semantics.
-%
-%\subsection{Meta Semantics}
-%
-%A different semantics was provided in \cite{davies96}.  We briefly review a two-stage version of that semantics here.
-%
-%In the erasure semantics, $\next v$ is a value only if $v$ is fully reduced.
-But in the meta semantics, $\next e$ is a value only if $e$ has no $\prev$ terms; $v$ is allowed to have unreduced stage-2 computation.
-%
-%...
-%
-%The Davies semantics is comprised of two mutually recursive judgments: $\daviesz$ and $\davieso$.
-For some $\colone e A$, the $\daviesz$ judgment evaluates only the first-stage parts of $e$, leaving unevaluated second-stage code within $\next$s.
-(...
-This essentially gives us partial evaluation, and we're left to just evaluate the residual normally to get multi-stage evaluation...)
-%
-%The benefit of the meta semantics is that it gives us a very explicit notion of partial evaluation.
-This also, by construction, means that the meta semantics does all of the first stage-work {\em before} the second stage work begins.
-%
-%The $\next$-by-name behavior of the meta semantics of course means that it happily duplicates second-stage code, which could increase the cost.
-This makes reasoning about second-stage cost rather difficult.
-%
-%\begin{lstlisting} 
-%let x = next {4+5} in
-%next{prev{x} * prev{x}}
-%\end{lstlisting}
-%
-%\subsection{Our Semantics}
-%
-%We desire a semantics that meets the following goals:
-%
-%\begin{enumerate}
-%\item Modulo termination, it should be equivalent to the erasure semantics.
-%\item All of the first stage work should be completed before second stage work.  Ideally, it should just have a notion of partial evaluation, like the meta semantics.
-%\item Should be $\next$-by-value, rather than $\next$-by-name, like the erasure semantics.  
-%\end{enumerate}
-%
-%We meet all of these goals.
-%}}}
-%{{{ stuff carlo commented out
-%Abstractly, we can think of our evaluation as proceeding in the standard way for
-%stage-1 code. When the evaluator encounters a $\next \{e\}$ expression, it
-%places $e$ off to the side in a context and replaces the whole expression with a
-%reference to the context entry.  These references are then passed around as
-%stage-1 values for $\fut$ types.  But what if $e$ contains $\prev$ expressions?
-%To ensure that all stage-1 code is evaluated before any stage-2 code, we must
-%evaluate all of the 1-code contained in $e$ before inserting it into the table.
-%This entails searching $e$ for all contained $\prev$s and evaluating them in
-%place.
-
-%The evaluation judgment operates on stage-1 code, whereas the
-%speculation operates on stage-2 code.  Since $\next$ and $\prev$ are the
-%crossover points between 1-code and 2-code, they are correspondingly the only
-%places where the evaluation and speculation judgments depend on the other. 
-%}}}
 
@@
 {\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
+
+Returning to our {\tt fexp} example, we can speculate on the stage \bbtwo\ term
+$\verb|fn b => prev{fexp(next{b},10)}|$
+to get the residual
+\begin{lstlisting} 
+fn b : int =>
+  let x0 =
+    (let x1 = 
+      let x2 = 1 * b
+      in x2 * x2
+    in x1 * x1) * b
+  in x0 * x0
+\end{lstlisting}
+
+Note how the recursion causes duplicate code in the residual.
 

### 2014-09-23 paper/techreport.tex
implemented some suggested changes

@@
 With these goals in mind, our initial concern is how to express multistage programs.
-To this end, we have modified prior work (\cite{davies96}) to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
+To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
 An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.
@@
 Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.
+
+\TODO motivate this example
+\begin{lstlisting} 
+let fexp (b : int, p : int) : int =
+	if p == 0 then
+		1
+	else if (p mod 2) == 0 then
+		let x = fexp(b,p/2) in x*x
+	else
+		b * fexp(b,p-1)	
+\end{lstlisting}
 

### 2014-09-24 paper/semantics.tex
dumped some sentences for a new section 2

@@
-\section{The \lang\ Language}
+\section {The \lang\ Language}
+
+We now describe \lang, a simple two-stage language
+for which we will later define and analyze a stage splitter.
+
+For this purpose, we use {\em fast exponent} as a running example.
+It calculates $b^p$ in $O(\log p)$ time.  
+Using a standard unstaged ML-like syntax, it is
+\begin{lstlisting} 
+fun fexp (b : int, p : int) : int = 
+	if p == 0 then
+		1
+	else if (p mod 2) == 0 then
+		let x = fexp(b,p/2) in x*x
+	else
+		b * fexp(b,p-1)	
+\end{lstlisting}
+
+The main idea of \lang\ is that terms defined in it are 
+split across two {\em stages}, called \bbone\ and \bbtwo,
+such that information can flow from the stage \bbone\ portions of the term to the stage \bbtwo\ portions,
+but never from \bbtwo\ to \bbone.
+Our starting point for this is a typed lambda calculus with products, sums, and isorecursive types.
+Here is the fast exponent example in staged form:
+\begin{lstlisting} 
+let fexp (b : $int, p : int) : $int =
+	if p == 0 then
+		next{1}
+	else if (p mod 2) == 0 then
+		next{let x = prev{fexp(b,p/2)} in x*x}
+	else
+		next{prev{b} * prev{fexp(b,p-1)}}		
+\end{lstlisting}
+This staged version contains three new constructs: 
+$\next$ and $\prev$ at the term level, and $\fut$ at the type level.
+We can read $\next$ and $\prev$ as indications to 
+execute the code in the following block at stage \bbtwo\ and \bbone, respectively.
+The type constructor $\fut$ is a way for us to talk at stage \bbone\ about things that will not be available until stage \bbtwo.
+Thus, ${\tt fexp} : \fut\rmint \times \rmint \to \fut\rmint$ 
+can be read as a function that takes a base at stage \bbtwo\ and a power at stage \bbone, 
+and returns the exponent at stage \bbtwo.
+
+With this in mind, observe that the exponent decomposition and recursive calls in ${\tt fexp}$ 
+all occur in stage \bbone, whereas squaring and multiplication by the base occur at stage \bbtwo.
+
+This paper investigates a transformation called stage-splitting, 
+wherein a single multi-stage program is split into multiple single-stage programs.
+For ${\tt fexp}$, this is:
+
+\begin{lstlisting} 
+fun fexp1 (b, e) =
+  ((), roll (
+    if e == 0 then inj1 ()
+    else
+      inj2 (
+        if (e mod 2) == 0
+        then inj1 (#2 (fexp1 (b,e/2)))
+        else inj2 (#2 (fexp1 (b,e-1)))
+      )
+  ))
+
+fun fexp2 ((b, e), p) =
+	case unroll p of
+	  inj1 () => 1
+	| inj2 d =>
+		case d of
+		  inj1 r => let x = fexp2 ((b,()),r) in x*x
+		| inj2 r => b * fexp2 ((b,()),r)
+\end{lstlisting}
+
+\section{Statics and Dynamics}
 \label{sec:semantics}
@@
 \input {figures/statics}
-
-In this section, we describe \lang, a simple two-stage language
-for which we define and analyze stage splitting.
-
-% \lang\ is largely based off the
-% language \lamCircle\ by \cite{davies96}.  
-% which was used as a model for partial evaluation.  
-% Our version, despite a slightly different semantics, can be used for the same purpose.
-% The details of the partial evaluation view are shown in \cref{sec:partialeval}.
-
-\lang\ is essentially two copies of a typed lambda calculus with
-products, sums, and isorecursive types.%
+%
 \footnote{It is possible for different stages to have different sets of

### 2014-09-24 paper/semantics.tex
added figure and outlined a new example/intro (filling out at home)

--- 2014-09-23 paper/semantics.tex
@@
-\section{The \lang\ Language}
+\section {The \lang\ Language}
+
+\begin{figure*}
+\begin{subfigure}{0.5\textwidth}
+\begin{lstlisting} 
+datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) 
+  : (int*list*list) =
+  case unroll l of 
+    Empty => (0,Empty, Empty) 
+  | Cons (h,t) =>
+      let (s,left,right) = partition (p,t) in
+      if h<p 
+      then (s+1,Cons(h,left),right)
+      else (s,left,Cons(h,right))
+
+fun qs (l : list, i : int) : int = 
+  case l of
+    Empty => next {0}
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+        case compare i n of
+          LT => qs left i
+        | EQ => h
+        | GT => qs right (i-n-1)
+\end{lstlisting}
+\caption{Unstaged Code}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\begin{lstlisting} 
+datatype list = Empty | Cons of int * list
+
+fun partition (p : int, l : list) = ...
+fun qs (l : list, i: $int) : $int = 
+  case l of
+    Empty => next {0}
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      next{
+        let n = hold{n} in
+          case compare prev{i} n of
+            LT => prev {qs left i}
+          | EQ => hold {h}
+          | GT => prev {qs right next{prev{i}-n-1}}
+      }	
+\end{lstlisting}
+\caption{Staged Code}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\begin{lstlisting} 
+datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+datatype tree = Branch of int * int * tree * tree
+                | Leaf
+
+fun qs1 (l : list) : tree =
+  case l of
+    Empty => Leaf
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      Branch (n, h, qs1 left i, qs1 right i)
+
+fun qs2 (p : tree, i : int) : int = 
+  case unroll p of
+    Leaf => 0
+  | Branch (n,h,p1,p2) => 
+      case compare i n of
+        LT => qs2 (p1,i)
+      | EQ => h
+      | GT => qs2 (p2,i-n-1)
+\end{lstlisting}
+\caption{Split Code}
+\end{subfigure}
+\caption{Caption place holder}
+\end{figure*}
+
+Consider the quickselect algorithm in [the figure],
+which finds the $i$th largest element of a list.
+It operates by inspecting the head of the list, 
+partitioning the rest of the list by the head,
+comparing the size of the partition to the desired index,
+and the recursively selecting on the correct sublist.
+If the index was outside valid bounds, 0 is returned.
+[runtime]
+
+It has been observed that [dependency...].
+
+We can then leverage this dependency to split quickselect into two other functions [one precomputes all splits... build tree... other consumes].
+[We've shown this in figure...]
+[we like it as two functions because... multiple calls... efficient]
+
+The goal of our work is to do this transformation automatically.
+[first we need a language to ... unambiguously and clearly specify staging]
+[ideally that language looks similar to the original term]
+[The type system for this language should be enough to prove that splitting will work out.]
+
+[to this end, we present \lang]
+The main idea of \lang\ is that terms defined in it are 
+split across two {\em stages}, called \bbone\ and \bbtwo.
+If part of the term is in \bbone, then it will end up in the precomputation, else it will end up in the residual.
+[information can flow from the stage \bbone\ portions of the term to the stage \bbtwo\ portions,
+but never from \bbtwo\ to \bbone.  this property is required if we want to be able to split the term]
+
+[Check out the example, now staged all fancy.]
+
+This staged version contains three new constructs: 
+$\next$ and $\prev$ at the term level, and $\fut$ at the type level.
+We can read $\next$ and $\prev$ as indications to 
+execute the code in the following block at stage \bbtwo\ and \bbone, respectively.
+The type constructor $\fut$ is a way for us to talk at stage \bbone\ about things that will not be available until stage \bbtwo.
+Thus, ${\tt fexp} : \fut\rmint \times \rmint \to \fut\rmint$ 
+can be read as a function that takes a base at stage \bbtwo\ and a power at stage \bbone, 
+and returns the exponent at stage \bbtwo.
+
+[$\fut$ acts in a sanitary way]
+
+\section{Statics and Dynamics}
 \label{sec:semantics}

### 2014-09-25 paper/semantics.tex
beefed up example/intro

@@
 \begin{figure*}
+\label{fig:quickselect}
 \begin{subfigure}{0.5\textwidth}
@@
         case compare i n of
-          LT => qs left i
+          LT => qs (left, i)
         | EQ => h
-        | GT => qs right (i-n-1)
+        | GT => qs (right, i-n-1)
 \end{lstlisting}
@@
 datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
 
-fun partition (p : int, l : list) = ...
 fun qs (l : list, i: $int) : $int = 
@@
           case compare prev{i} n of
-            LT => prev {qs left i}
+            LT => prev {qs (left i)}
           | EQ => hold {h}
-          | GT => prev {qs right next{prev{i}-n-1}}
+          | GT => prev {qs (right, next{prev{i}-n-1)}}
       }	
@@
 
-Consider the quickselect algorithm in [the figure],
+Consider the quickselect algorithm in [figure part (a)],
 which finds the $i$th largest element of a list.
@@
 If the index was outside valid bounds, 0 is returned.
-[runtime]
+Assuming that the list is randomly sorted, 
+quickselect will take expected $O(n)$ time, where $n$ is the size of the list.
 
-It has been observed that [dependency...].
+Observe that within quickselect the partitioning of the list does not depend on $i$, 
+even through the recursive boundary.\footnote{There is a control-dependence on $i$, but we can ignore that for now.}
+We can leverage this dependency pattern to split quickselect into two other functions:
+one which precomputes all of the partitions of the input list to build a tree,
+and another which consumes this tree to perform an accelerated index lookup.
+This split version of quickselect is shown in [figure part (c)].
+One advantage of this form is that the tree can be built once per list
+and then used to accelerate many lookups.
 
-We can then leverage this dependency to split quickselect into two other functions [one precomputes all splits... build tree... other consumes].
-[We've shown this in figure...]
-[we like it as two functions because... multiple calls... efficient]
+The goal of our work is to automate this splitting transformation.
+Our first step is to define a language which clearly and unambiguously specifies how its terms should be split.
+Ideally, only minimal refactoring should be required to write terms within this language,
+and from there, the type system should be enough to prove that splitting can be carried out.
 
-The goal of our work is to do this transformation automatically.
-[first we need a language to ... unambiguously and clearly specify staging]
-[ideally that language looks similar to the original term]
-[The type system for this language should be enough to prove that splitting will work out.]
-
-[to this end, we present \lang]
-The main idea of \lang\ is that terms defined in it are 
-split across two {\em stages}, called \bbone\ and \bbtwo.
-If part of the term is in \bbone, then it will end up in the precomputation, else it will end up in the residual.
-[information can flow from the stage \bbone\ portions of the term to the stage \bbtwo\ portions,
-but never from \bbtwo\ to \bbone.  this property is required if we want to be able to split the term]
+To this end, we present \lang.
+The main idea of \lang\ is that all parts of terms defined in it can be identified with one of two {\em stages}, namely \bbone\ or \bbtwo.
+Those parts of the \lang\ term in \bbone\ will end up in the precomputation, 
+and those parts in \bbtwo\ will end up in the residual.
+Within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
+but never from \bbtwo\ to \bbone.  
+Indeed this is a necessary property if we want to be able to split the term.
 

### 2014-09-25 paper/semantics.tex
iteration on example/intro

@@
 
-[Check out the example, now staged all fancy.]
+See [figure part (b)] for a staged version of quickselect.
+This example contains all four of the special constructs of \lang: 
+$\fut$ at the type level, and $\next$, $\prev$, and $\pause$ at the term level.
+Notice that but for these new constructs, the staged version of quickselect is virtually identical to the unstaged version.
 
-This staged version contains three new constructs: 
-$\next$ and $\prev$ at the term level, and $\fut$ at the type level.
-We can read $\next$ and $\prev$ as indications to 
-execute the code in the following block at stage \bbtwo\ and \bbone, respectively.
 The type constructor $\fut$ is a way for us to talk at stage \bbone\ about things that will not be available until stage \bbtwo.
-Thus, ${\tt fexp} : \fut\rmint \times \rmint \to \fut\rmint$ 
-can be read as a function that takes a base at stage \bbtwo\ and a power at stage \bbone, 
-and returns the exponent at stage \bbtwo.
+In our example, the type signature ${\tt qs} : {\tt list} \times \fut \rmint \to \fut\rmint$ 
+indicates that ${\tt qs}$ is a function that takes a list ($l$) at stage \bbone\ and an integer ($k$) at stage \bbtwo, 
+and returns an int at stage \bbtwo\ (the $k$th largest element of $l$).
+Note that ${\tt qs}$ is itself bound at stage \bbone.
 
-[$\fut$ acts in a sanitary way]
+At the term level, $\pause$ is conceptually the simplest.
+It allows us to take an integer computed at stage one, and keep it around until stage \bbtwo.
+
+The staging of a term is achieved via a nesting of alternating $\next$ and $\prev$ blocks.  
+As $\next$, $\prev$, and $\pause$ are the only ways in \lang\ to alter the stage of a term,
+we surround their arguments with braces in \lang\ syntax.
+Specifically, $\next$ embeds a stage \bbtwo\ block within a stage \bbone\ environment,
+whereas $\prev$ operates vice versa.
+To ensure that information never flows backward in time, 
+the output of a $\next$ block is always a $\fut$,
+and the only construct that can eliminate $\fut$s is $\prev$.
+This means that any values created at stage \bbtwo\ are always handled in a sanitary way while being passed around in stage \bbone.
+
+[still get to WHY we staged like this, and talk about binding time analysis]
 

### 2014-09-26 paper/paper.tex
factored example section into its own file
reorganized splitting section
minor changes after meeting with nico
fix the issue with listings in footnotes
some text in section 3 about top-level operational semantics
adding colors to listings

--- 2014-06-19 paper/paper.tex
@@
         escapeinside={"*}{*"},
+        moredelim=*[is][\color{red}]{1`}{`},
+        moredelim=*[is][\color{blue}]{2`}{`},
         literate={|->}{{\ $\mapsto$\ }}1
@@
 \input{splitting}
-
+\input{algoexamples}
 \input{relatedwork}

### 2014-09-26 paper/semantics.tex
use minipage instead of deprecated subfigure (not in texlive 2014)
adding colors to listings

@@
 \label{fig:quickselect}
-\begin{subfigure}{0.5\textwidth}
+\begin{minipage}{0.5\textwidth}
 \begin{lstlisting} 
@@
 \caption{Unstaged Code}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \begin{lstlisting} 
@@
   case l of
-    Empty => next {0}
+    1`Empty` => next {2`0`}
   | Cons (h,t) => 
@@
 \caption{Staged Code}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \begin{lstlisting} 
@@
 \caption{Split Code}
-\end{subfigure}
+\end{minipage}
 \caption{Caption place holder}

### 2014-09-26 paper/semantics.tex
More work on section 2

--- 2014-09-25 paper/semantics.tex
@@
-\section {The \lang\ Language}
-
+\section {Staged Programming in \lang}
+\label{sec:staging}
 \begin{figure*}
 \label{fig:quickselect}
-\begin{subfigure}{0.5\textwidth}
+\begin{minipage}{0.5\textwidth}
 \begin{lstlisting} 
@@
 
-fun qs (l : list, i : int) : int = 
+fun qs (l : list, k : int) : int = 
   case l of
-    Empty => next {0}
+    Empty => 0
   | Cons (h,t) => 
       let (left,right,n) = partition h t in
-        case compare i n of
-          LT => qs (left, i)
+        case compare k n of
+          LT => qs (left, k)
         | EQ => h
-        | GT => qs (right, i-n-1)
+        | GT => qs (right, k-n-1)
 \end{lstlisting}
 \caption{Unstaged Code}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
 \begin{lstlisting} 
@@
 
-fun qs (l : list, i: $int) : $int = 
+fun qs (l : list, k : $int) : $int = 
   case l of
-    Empty => next {0}
+    1`Empty` => next {2`0`}
   | Cons (h,t) => 
@@
         let n = hold{n} in
-          case compare prev{i} n of
-            LT => prev {qs (left i)}
+          case compare prev{k} n of
+            LT => prev {qs (left k)}
           | EQ => hold {h}
-          | GT => prev {qs (right, next{prev{i}-n-1)}}
+          | GT => prev {qs (right, next{prev{k}-n-1)}}
       }	
@@
       let (left,right,n) = partition h t in
-      Branch (n, h, qs1 left i, qs1 right i)
+      Branch (n, h, qs1 left, qs1 right)
 
-fun qs2 (p : tree, i : int) : int = 
+fun qs2 (p : tree, k : int) : int = 
   case unroll p of
@@
   | Branch (n,h,p1,p2) => 
-      case compare i n of
-        LT => qs2 (p1,i)
+      case compare k n of
+        LT => qs2 (p1,k)
       | EQ => h
-      | GT => qs2 (p2,i-n-1)
+      | GT => qs2 (p2,k-n-1)
 \end{lstlisting}
 \caption{Split Code}
-\end{subfigure}
+\end{minipage}
 \caption{Caption place holder}
@@
 
-Consider the quickselect algorithm in [figure part (a)],
-which finds the $i$th largest element of a list.
+For an example of an algorithm amenable to staging techniques,
+consider the quickselect algorithm in [figure part (a)],
+which finds the $k$th largest element of a list.
 It operates by inspecting the head of the list, 
@@
 and the recursively selecting on the correct sublist.
-If the index was outside valid bounds, 0 is returned.
 Assuming that the list is randomly sorted, 
@@
 
-Observe that within quickselect the partitioning of the list does not depend on $i$, 
-even through the recursive boundary.\footnote{There is a control-dependence on $i$, but we can ignore that for now.}
+Importantly, quickselect has property that the partitioning of the list does not depend on $k$, 
+even through the recursive boundary.\footnote{There is a control-dependence on $k$, but we can ignore that for now.}
 We can leverage this dependency pattern to split quickselect into two other functions:
-one which precomputes all of the partitions of the input list to build a tree,
-and another which consumes this tree to perform an accelerated index lookup.
+one which precomputes all of the partitions of the input list to build a binary search tree,
+and another which uses this tree to perform an accelerated index lookup.
 This split version of quickselect is shown in [figure part (c)].
@@
 Ideally, only minimal refactoring should be required to write terms within this language,
-and from there, the type system should be enough to prove that splitting can be carried out.
+and from there, the type system should be enough to prove that a valid splitting exists.
 
@@
 The main idea of \lang\ is that all parts of terms defined in it can be identified with one of two {\em stages}, namely \bbone\ or \bbtwo.
-Those parts of the \lang\ term in \bbone\ will end up in the precomputation, 
+Intuitively, the stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
+subexpressions are evaluated before stage-\bbtwo\ ones.
+Correspondingly, after splitting those parts of the \lang\ term in \bbone\ will end up in the precomputation, 
 and those parts in \bbtwo\ will end up in the residual.
@@
 
-[Check out the example, now staged all fancy.]
+[Figure part (b)] shows a staged implementation of quickselect in \lang.
+As before, the inputs to the function are the list to select from and the index to select,
+but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
+The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
+whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
+Naturally the output type, representing the $k$th largest element of the list,
+is also $\fut\rmint$, since it cannot be known until the next stage.
 
-This staged version contains three new constructs: 
-$\next$ and $\prev$ at the term level, and $\fut$ at the type level.
-We can read $\next$ and $\prev$ as indications to 
-execute the code in the following block at stage \bbtwo\ and \bbone, respectively.
-The type constructor $\fut$ is a way for us to talk at stage \bbone\ about things that will not be available until stage \bbtwo.
-Thus, ${\tt fexp} : \fut\rmint \times \rmint \to \fut\rmint$ 
-can be read as a function that takes a base at stage \bbtwo\ and a power at stage \bbone, 
-and returns the exponent at stage \bbtwo.
+Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
+Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
+whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
+(We adapt the convention convention that the top-level context is stage \bbone.)
+The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
+For example, \verb|next{0}| from above has the type $\fut\rmint$.
+Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
+For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
+These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
+which is necessary to admit a properly staged implementation.
 
-[$\fut$ acts in a sanitary way]
+The code also contains two $\pause$ blocks.  
+This construct has same stage signature as $\prev$,
+but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage \bbone\ to stage \bbtwo.
+\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
+We instead provide it as a primitive to shorten examples.  
+Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
+This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
 
-\section{Statics and Dynamics}
+But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
+the staged version of quickselect is virtually identical to the unstaged version.
+The constructs that were added were placed in order to maximize the work done at stage \bbone\ while still conforming to the type signature.
+It would have also been valid to simply move the whole input list unchanged into stage \bbtwo\ at the very beginning, 
+but that would not be particularly interesting since it would result in an effectively trivial split 
+that's just the identity at stage \bbone\ and plus quickselect at stage \bbtwo.
+There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
+This process is known as {\em binding time analysis}, and we do not consider it here.
+Instead, we assume that all input programs are properly staged according to some programmer's intent.
+
+\section{\lang\ Statics and Dynamics}
 \label{sec:semantics}
@@
 \input {figures/statics}
-%
+
+As stated in \cref{sec:staging}, we operate on a language called \lang,
+which is a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
+
+\subsection{Statics}
+The grammar and type system of \lang\ are given in \ref{fig:grammar,fig:statics}. 
+They are a simple adaptation of those of \cite{davies96}, 
+restricted to two stages\footnote{This restriction is not a big deal...} and extended with general sums and recursion.
+
+Typing judgments and context variables are annotated with stages after an $@$ symbol.
+Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
+of a term or type.
+
 \footnote{It is possible for different stages to have different sets of
 features, but for simplicity we do not consider this.}
-Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
-expressing in which copy of the language the term resides. Intuitively, the
-stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.
+Every valid expression has both a type and a stage, either \bbone~or \bbtwo.
 
@@
 
-The grammar and type system for \lang\ is given in \ref{fig:grammar,fig:statics}. 
-Typing judgments and context variables are annotated with stages after an $@$ symbol.
-Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
-of a term or type.
 %We formulated our typing judgments in the style of \cite{davies96}, where the
@@
 
-\subsection{Staged Evaluation}
+\subsection{Dynamics}
 \label{sec:stagedsemantics}
 
-A key attribute of the fexp example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
+A key attribute of the quickselect example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
 Ordinary term evaluation eliminates outermost redexes first, 
@@
 stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
-monostaged
-\footnote{In general, \lang could have more than two stages, in which case the
-expression would be in a language with one fewer stage.}
-language called \langTwo. Then, we perform the remainder of the evaluation with 
+monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 
 $\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this

### 2014-09-28 paper/semantics.tex
Wrote a simple section 3.0 and 3.1

@@
 which is a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
+The grammar of \lang\ can be found in \cref{fig:grammar}.
 
 \subsection{Statics}
-The grammar and type system of \lang\ are given in \ref{fig:grammar,fig:statics}. 
-They are a simple adaptation of those of \cite{davies96}, 
-restricted to two stages\footnote{This restriction is not a big deal...} and extended with general sums and recursion.
+The grammar and type system of \lang\ are a simple adaptation of those of \cite{davies96}, 
+restricted to two stages\footnote{This restriction to two stages is made primarily for 
+simplicity of presentation.  All of these techniques could feasibly be extended to more
+complicated stage systems.} and extended with general sums and recursion.
 
-Typing judgments and context variables are annotated with stages after an $@$ symbol.
-Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
-of a term or type.
+As explained before, the $\fut$ constructor is a way to encapsulate a stage \bbtwo\ type at stage \bbone.  
+But because of the two-stage assumption, $\fut$ has no meaning when used at stage \bbtwo.
+This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
+In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
+This context is only augmented by recursive types.
 
-\footnote{It is possible for different stages to have different sets of
-features, but for simplicity we do not consider this.}
-Every valid expression has both a type and a stage, either \bbone~or \bbtwo.
-
-The stage-\bbone\ type $\fut A$ contains encapsulated
-stage-\bbtwo\ expressions of type $A$. Terms of type $\fut A$ are treated
-opaquely by stage-\bbone\ code, as (by the requirements of staging) they cannot be evaluated until stage \bbtwo.
-While $\fut$ allows us to embed stage-\bbtwo\ types within stage-\bbone\ types,
-there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.
-That is, there is a one-way dependence between stages at the term level.
-
-Alternatively, the stages are mutually dependent at the term level.
-The $\next$ constructor embeds stage-\bbtwo\ expressions into stage \bbone,
-while $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.  $\next$ and $\prev$ are the
-only ways in \lang\ to alter the stage of a term; we surround their arguments
-with braces in \lang\ syntax to clearly indicate stage boundaries within a
-program.
-
-%We formulated our typing judgments in the style of \cite{davies96}, where the
-%whole judgment is annotated with a stage.  
-%The grammar and type system for \lang\ is given in
-%\ref{fig:grammar,fig:statics}.
-% We annotate typing judgments and context variables with stages;
-%This is made manifest as rules which are entirely abstract over stage.
-%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
-Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
-%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
-Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
-Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality,\cite{cave14}.
-The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
-It is possible to implement $\pause$ from other \lang\ features, but
-we provide it as a core primitive to simplify our examples.
-
-Recall the fast exponent function from above.
-We present a valid staging of it here:
-\begin{lstlisting} 
-let fexp (b : $int, p : int) : $int =
-	if p == 0 then
-		next{1}
-	else if (p mod 2) == 0 then
-		next{let x = prev{fexp(b,p/2)} in x*x}
-	else
-		next{prev{b} * prev{fexp(b,p-1)}}		
-\end{lstlisting}
-
-\TODO explain why we staged this the way we did
-
-The function receives a stage-\bbone\ exponent {\tt e} and a stage-\bbtwo\ base {\tt b} and returns a stage-\bbtwo\ result. 
-The {\tt if} predicates and exponent decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
+The typing judgment, like type validity, is annotated with a stage.
+Thus we can read $\typeswor x A$ as saying that term $x$ has type $A$ at world $w$ under the context $\Gamma$.
+The typing rules are given in \cref{fig:statics}.
+The staging features ($\next$, $\prev$, and $\pause$) behave as described in the previous section,
+and all other features behavior in the standard way and are essentially agnostic to stage. 
+Variables in the context are annotated with the stage at which they were introduced 
+and can only be used at the same stage.
+As one would expect, the typing judgment only produces well-formed types 
+(by induction on the derivation of the typing judgment).
 

### 2014-09-28 paper/semantics.tex
Editing pass in Section 2 and Section 3.0.

@@
 \label{sec:staging}
+
 \begin{figure*}
@@
 \end{lstlisting}
-\caption{Unstaged Code}
+\caption{Unstaged implementation of quickselect.}
 \end{minipage}
@@
 \end{lstlisting}
-\caption{Staged Code}
+\caption{Staged implementation of quickselect in \lang.}
 \end{minipage}
@@
 \end{lstlisting}
-\caption{Split Code}
+\caption{Split (two-phase) implementation of quickselect.}
 \end{minipage}
@@
 
-For an example of an algorithm amenable to staging techniques,
-consider the quickselect algorithm in [figure part (a)],
-which finds the $k$th largest element of a list.
-It operates by inspecting the head of the list, 
-partitioning the rest of the list by the head,
-comparing the size of the partition to the desired index,
-and the recursively selecting on the correct sublist.
-Assuming that the list is randomly sorted, 
-quickselect will take expected $O(n)$ time, where $n$ is the size of the list.
+As an example of an algorithm amenable to staging techniques,
+consider the quickselect algorithm shown in Figure~2a.
+Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
+It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
+then recursing on the sublist containing the desired element.
 
-Importantly, quickselect has property that the partitioning of the list does not depend on $k$, 
-even through the recursive boundary.\footnote{There is a control-dependence on $k$, but we can ignore that for now.}
-We can leverage this dependency pattern to split quickselect into two other functions:
-one which precomputes all of the partitions of the input list to build a binary search tree,
-and another which uses this tree to perform an accelerated index lookup.
-This split version of quickselect is shown in [figure part (c)].
-One advantage of this form is that the tree can be built once per list
-and then used to accelerate many lookups.
+Importantly, one property of quickselect is that the partitioning of the list, even across recursive calls, depends only on the values of list elements, not on the value of $k$ 
+\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
+This independence allows the quickselect algorithm to be split into two functions:
+one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qs1} in Figure~2c),
+and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
+One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
+and then used to accelerate many future lookups for any value of $k$.
 
-The goal of our work is to automate this splitting transformation.
-Our first step is to define a language which clearly and unambiguously specifies how its terms should be split.
-Ideally, only minimal refactoring should be required to write terms within this language,
-and from there, the type system should be enough to prove that a valid splitting exists.
-
-To this end, we present \lang.
-The main idea of \lang\ is that all parts of terms defined in it can be identified with one of two {\em stages}, namely \bbone\ or \bbtwo.
-Intuitively, the stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.
-Correspondingly, after splitting those parts of the \lang\ term in \bbone\ will end up in the precomputation, 
-and those parts in \bbtwo\ will end up in the residual.
+This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
+\lang clearly and unambiguously specifies how its terms should be split
+% and features a type system that ensures valid splittings exist.
+% only minimal refactoring should be required to write terms within this language,
+%and from there, the type system should be enough to prove that a valid splitting exists.
+by identifying each term with one of two {\em stages}, namely \bbone\ or \bbtwo.
+Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
+subexpressions are evaluated before stage-\bbtwo\ ones.  After splitting, those parts of the \lang\ term in stage \bbone\ will end up in the precomputation, 
+and those parts in stage \bbtwo\ will end up in the residual.
 Within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
@@
 
-[Figure part (b)] shows a staged implementation of quickselect in \lang.
-As before, the inputs to the function are the list to select from and the index to select,
+Figure 2b shows a staged implementation of quickselect in \lang.
+As before, the function accepts as input the list to select from and a value of $k$,
 but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
@@
 Naturally the output type, representing the $k$th largest element of the list,
-is also $\fut\rmint$, since it cannot be known until the next stage.
+is also $\fut\rmint$, since it is not computed until stage \bbtwo.
 
@@
 whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
-(We adapt the convention convention that the top-level context is stage \bbone.)
+(We adapt the convention that the top-level context is stage \bbone.)
 The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
@@
 This construct has same stage signature as $\prev$,
-but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage \bbone\ to stage \bbtwo.
+but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage~\bbone\ to stage~\bbtwo.
 \footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
@@
 But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
-the staged version of quickselect is virtually identical to the unstaged version.
-The constructs that were added were placed in order to maximize the work done at stage \bbone\ while still conforming to the type signature.
-It would have also been valid to simply move the whole input list unchanged into stage \bbtwo\ at the very beginning, 
+the staged version of quickselect in Figure~2b is virtually identical to the unstaged version in Figure~2a.
+The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
+It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
 but that would not be particularly interesting since it would result in an effectively trivial split 
-that's just the identity at stage \bbone\ and plus quickselect at stage \bbtwo.
+that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
 There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
@@
 
-As stated in \cref{sec:staging}, we operate on a language called \lang,
-which is a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
-The grammar of \lang\ can be found in \cref{fig:grammar}.
+As stated in \cref{sec:staging}, in this paper we describe splitting in the context of \lang,
+a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
+The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
+restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
+simplicity of presentation.  All of these techniques could feasibly be extended to more
+complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.
 
 \subsection{Statics}
-The grammar and type system of \lang\ are a simple adaptation of those of \cite{davies96}, 
-restricted to two stages\footnote{This restriction to two stages is made primarily for 
-simplicity of presentation.  All of these techniques could feasibly be extended to more
-complicated stage systems.} and extended with general sums and recursion.
 
-As explained before, the $\fut$ constructor is a way to encapsulate a stage \bbtwo\ type at stage \bbone.  
-But because of the two-stage assumption, $\fut$ has no meaning when used at stage \bbtwo.
+In \lang, the $\fut$ constructor encapsulates a stage \bbtwo\ type at stage~\bbone.  
+But because of the two-stage assumption, $\fut$ has no meaning when used at stage~\bbtwo.
 This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
@@
 Ordinary term evaluation eliminates outermost redexes first, 
-however in the case that stage \bbone\ expressions are contained inside stage \bbtwo\ ones 
+however in the case that stage \bbone\ expressions are contained inside stage~\bbtwo\ ones 
 (such as the recurive call to {\tt fexp} above), 
@@
 Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
-subexpressions before any of its stage \bbtwo\ subexpressions. This results in a
-stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
+subexpressions before any of its stage~\bbtwo\ subexpressions. This results in a
+stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
 monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 

### 2014-09-28 paper/semantics.tex

@@
-\section {Staged Programming in \lang}
+\section {Example: Staged Quickselect in \lang}
 \label{sec:staging}
@@
 This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
-\lang clearly and unambiguously specifies how its terms should be split
+\lang unambiguously specifies how its terms should be split
 % and features a type system that ensures valid splittings exist.
@@
 %and from there, the type system should be enough to prove that a valid splitting exists.
-by identifying each term with one of two {\em stages}, namely \bbone\ or \bbtwo.
+by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
 Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
@@
 and those parts in stage \bbtwo\ will end up in the residual.
-Within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
+Typing rules for \lang ensure a valid splitting exists.  That is, 
+within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
 but never from \bbtwo\ to \bbone.
-Indeed this is a necessary property if we want to be able to split the term.
 
@@
 
-As stated in \cref{sec:staging}, in this paper we describe splitting in the context of \lang,
-a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
+As stated in \cref{sec:staging}, we use \lang,
+a typed two-stage lambda calculus featuring products, sums, and isorecursive types, as language to define and analyze the splitting transform.
 The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
@@
 
-\subsection{Non-Duplicating First-Stage Evaluation}
+\subsubsection{Non-Duplicating First-Stage Evaluation}
 
@@
 
-\subsection{Dynamics}
+\subsubsection{Dynamics}
 \label{ssec:dynamics}
@@
 
-\subsection {A Partial Evaluation System}
+\subsubsection {A Partial Evaluation System}
 \label{sec:partialeval}

### 2014-09-29 paper/semantics.tex
Small edits.

@@
 
-fun qs (l : list, k : int) : int = 
+fun qSelect (l : list, k : int) : int = 
   case l of
@@
         case compare k n of
-          LT => qs (left, k)
+          LT => qSelect (left, k)
         | EQ => h
-        | GT => qs (right, k-n-1)
+        | GT => qSelect (right, k-n-1)
 \end{lstlisting}
@@
 
-fun qs (l : list, k : $int) : $int = 
+fun qsStaged (l : list, k : $int) : $int = 
   case l of
@@
           case compare prev{k} n of
-            LT => prev {qs (left k)}
+            LT => prev {qsStaged (left k)}
           | EQ => hold {h}
-          | GT => prev {qs (right, next{prev{k}-n-1)}}
+          | GT => prev {qsStaged (right, 
+                             next{prev{k}-n-1)}}
       }	
@@
 
-fun qs1 (l : list) : tree =
+fun qSelect1 (l : list) : tree =
   case l of
@@
       let (left,right,n) = partition h t in
-      Branch (n, h, qs1 left, qs1 right)
+      Branch (n, h, qSelect1 left, qSelect1 right)
 
-fun qs2 (p : tree, k : int) : int = 
+fun qSelect2 (p : tree, k : int) : int = 
   case unroll p of
@@
       case compare k n of
-        LT => qs2 (p1,k)
+        LT => qSelect2 (p1,k)
       | EQ => h
-      | GT => qs2 (p2,k-n-1)
+      | GT => qSelect2 (p2,k-n-1)
 \end{lstlisting}
@@
 This independence allows the quickselect algorithm to be split into two functions:
-one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qs1} in Figure~2c),
+one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qSelect1} in Figure~2c),
 and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
@@
 Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.  After splitting, those parts of the \lang\ term in stage \bbone\ will end up in the precomputation, 
-and those parts in stage \bbtwo\ will end up in the residual.
+subexpressions are evaluated before stage-\bbtwo\ ones.  
+After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
+will end up in the stage \bbone\ function ({\tt qSelect1}), 
+and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
 Typing rules for \lang ensure a valid splitting exists.  That is, 
@@
 
-\subsubsection{Dynamics}
+\subsubsection{Evaluation and Speculation}
 \label{ssec:dynamics}
@@
 \input {figures/dynamics}
+
+The goal of the dynamics is to perform the above style of first-stage evaluation automatically.
+It comprises judgments: $\redonesym$ (evaluation),
+$\redtwosym$ (speculation), and $\reifysym$ (reification of substitutions into let bindings).
 
@@
 \end{itemize}
-
-% would like some kind of intro here that anticipates all the complexity.
-% The dynamics of \lang\ consists of three types of judgements: $\redonesym$ (stage-\bbone reduction), $\redtwosym$ (speculation), and $\reifysym$ (reification).
 
@@
 
-\subsubsection {A Partial Evaluation System}
+\subsubsection {Top-Level Evaluation}
 \label{sec:partialeval}

### 2014-09-29 paper/semantics.tex
Push more emphasis on the calling of quickselect many times.

@@
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
-This independence allows the quickselect algorithm to be split into two functions:
-one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qSelect1} in Figure~2c),
+Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qSelect1} in Figure~2c),
 and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).

### 2014-09-29 paper/example.tex
Split section 2 into it's own .tex file.

--- 2014-09-29 paper/semantics.tex
@@
 Instead, we assume that all input programs are properly staged according to some programmer's intent.
-
-\section{\lang\ Statics and Dynamics}
-\label{sec:semantics}
-
-\input {figures/grammars}
-\input {figures/statics}
-
-As stated in \cref{sec:staging}, we use \lang,
-a typed two-stage lambda calculus featuring products, sums, and isorecursive types, as language to define and analyze the splitting transform.
-The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
-restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
-simplicity of presentation.  All of these techniques could feasibly be extended to more
-complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.
-
-\subsection{Statics}
-
-In \lang, the $\fut$ constructor encapsulates a stage \bbtwo\ type at stage~\bbone.  
-But because of the two-stage assumption, $\fut$ has no meaning when used at stage~\bbtwo.
-This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
-In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
-This context is only augmented by recursive types.
-
-The typing judgment, like type validity, is annotated with a stage.
-Thus we can read $\typeswor x A$ as saying that term $x$ has type $A$ at world $w$ under the context $\Gamma$.
-The typing rules are given in \cref{fig:statics}.
-The staging features ($\next$, $\prev$, and $\pause$) behave as described in the previous section,
-and all other features behavior in the standard way and are essentially agnostic to stage. 
-Variables in the context are annotated with the stage at which they were introduced 
-and can only be used at the same stage.
-As one would expect, the typing judgment only produces well-formed types 
-(by induction on the derivation of the typing judgment).
-
-\subsection{Dynamics}
-\label{sec:stagedsemantics}
-
-A key attribute of the quickselect example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
-Ordinary term evaluation eliminates outermost redexes first, 
-however in the case that stage \bbone\ expressions are contained inside stage~\bbtwo\ ones 
-(such as the recurive call to {\tt fexp} above), 
-this strategy conflicts with the precept of staged execution: 
-that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.
-
-Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
-subexpressions before any of its stage~\bbtwo\ subexpressions. This results in a
-stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
-monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 
-$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
-judgment are not shown, but they are standard).
-
-\subsubsection{Non-Duplicating First-Stage Evaluation}
-
-To gain intuition about the challenges of implementing this staged dynamic
-semantics, consider the following example:
-\begin{lstlisting}
-#2 (next {fib 20}, 2+3)
-\end{lstlisting}
-This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
-stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. 
-In this example, ${\tt fib} : \rmint \to \rmint$ is a stage-\bbtwo\ reference to the Fibonacci function.
-A conventional call-by-value evaluation strategy demands that we evaluate both components of
-the pair to values before we project from the pair. The problem is that
-while \verb|next {fib 20}| is not a value (in the sense that additional
-stage-\bbtwo\ computation steps are necessary to produce the result \verb|6765|), 
-evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
-Intuitively, the solution is to designate \verb|next {fib 20}| as a value \emph{in
-stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
-Therefore, we evaluate the pair to
-\begin{lstlisting}
-#2 (next {fib 20}, 5)
-\end{lstlisting}
-then obtain \verb|5| as the result of projection.
-
-Now consider a more complex example where 
-stage-\bbone\ evaluation must substitute such a incompletely-evaluated
-expression. The following stage-\bbtwo\ term has type $\rmint$:
-\begin{lstlisting} 
-prev{
-  let x = (next {fib 20}, 3+4) in
-  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
-}
-\end{lstlisting}
-Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
-depends on the value of \verb|fib 20|, which is not reduced to \verb|3| until
-stage \bbtwo.
-
-If we simply treat \verb|(next {fib 20}, 7)| as a value during stage~\bbone, and
-substitute it for the three occurrences of \verb|x| in the body of the
-\verb|let| expression, the result of stage \bbone\ computation is
-\begin{lstlisting} 
-prev{
-  next{ 
-    prev{next {fib 20}} * prev{next {fib 20}} * 7 
-  }
-}
-\end{lstlisting}
-Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
-\begin{lstlisting} 
-(fib 20) * (fib 20) * 7
-\end{lstlisting}
-Note how stage-\bbtwo\ evaluation of this expression will compute \verb|fib 20| twice.  
-To avoid duplicating computations, we take a different approach.  Instead of
-duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
-a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
-This produces:
-\begin{lstlisting} 
-let y = fib 20 in y * y * 7
-\end{lstlisting}
-
-Achieving this behavior mechanically requires us to resolve a contradiction:
-we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
-Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.
-This substitution then floats up to the top of the containing $\prev$ block:
-\begin{lstlisting} 
-prev {
-[yhat|->fib 20]
-  let x = (next{yhat}, 7) in
-  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
-}
-\end{lstlisting}
-As a convention, we render the new variable with a %stylish and fashionable
-hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
-\begin{lstlisting} 
-prev {
-[yhat|->fib 20]
-  next{
-    prev{#1 (next {yhat}, 7)} * 
-    prev{#1 (next {yhat}, 7)} *
-    hold{#2 (next {yhat}, 7)}
-  }
-}
-\end{lstlisting}
-To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
-As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
-\begin{lstlisting} 
-prev {
-[yhat|->fib 20]
-    next{ yhat * yhat * 7 }
-}
-\end{lstlisting}
-Once again, we lift the contents of the $\next$ into a substitution:
-\begin{lstlisting} 
-prev {
-[yhat|->fib 20]
-[zhat|->yhat*yhat*7]
-    next{ zhat }
-}
-\end{lstlisting}
-Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
-\begin{lstlisting} 
-let yhat = fib 20 in
-let zhat = yhat * yhat * 7 in z
-\end{lstlisting}
-
-Thus we have evaluated all of the stage \bbone\ expressions of this program without duplicating the contents of $\next$ blocks.
-
-\subsubsection{Evaluation and Speculation}
-\label{ssec:dynamics}
-
-\input {figures/dynamics}
-
-The goal of the dynamics is to perform the above style of first-stage evaluation automatically.
-It comprises judgments: $\redonesym$ (evaluation),
-$\redtwosym$ (speculation), and $\reifysym$ (reification of substitutions into let bindings).
-
-The algorithm described above creates three different kinds of expressions which
-cannot be evaluated further at a particular stage:
-\begin{itemize}
-\item 
-Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
-but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
-In the example above, 
-\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
-
-\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
-stage \bbone\ subexpressions have all been fully evaluated. In the example
-above,
-\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
-
-\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
-are the results of a computation after both stages have been completed. In the
-example above, the term evaluates to \verb|63|.
-\end{itemize}
-
-The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
-environment} $\xi$ and a partial value $v$.  The future environment is a mapping
-from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
-$v$) to residuals---in our example above, we represented this environment with
-explicit substitutions. For all of the normal features of \lang\ 
-(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
-effect on (partial) values as does standard evaluation, and the final future environment is
-gotten by merging the future environments of the subterms.
-
-When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
-searches into the block's stage \bbtwo\ content to find any contained stage
-\bbone\ subexpressions and evaluate them in place.  This search process, called
-\emph{speculation}, is implemented by the $\redtwosym$ judgment, which takes
-a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
-speculated into a residual ($q$), the output of $\redonesym$ is a fresh variable wrapped in a
-$\next$ block ($\next~\hat y$), along with a future environment which maps that
-variable to the residual ($\hat y \mapsto q$).
-
-At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
-nothing but recursively speculate into every subexpression.  Once speculation
-finds a $\prev$ block, it resumes stage \bbone\ evaluation of the contents, which
-produces a future context and (by canonical forms) a $\next$-wrapped variable
-($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
-judgment) into a series of let bindings with $\mathtt{\hat y}$, stripped of
-its $\next$, at the bottom.
-
-Within speculation lies a subtle---if perhaps unintuitive---feature.  
-Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
-statement in its search for stage-\bbone\ code. 
-Thus the evaluation of that stage one code will occur {\em regardless of the eventual value of the predicate},
-and so a term like 
-\begin{lstlisting} 
-next{
-  if true 
-  then hold{1+2} 
-  else prev{spin() (* loops forever *)}
-}
-\end{lstlisting}
-will fail to evaluate at stage \bbone.
-This behavior is why the judgment is named ``speculation."
-
-The context ($\Gamma$) keeps track of stage \bbtwo\ variables in the input term. 
-These both appear in the original program at stage \bbtwo\ and are inserted by the semantics.
-
-As an optimization, we can include the special-case rule,
-\begin{mathpar}
-\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
-\end{mathpar}
-to avoid one-for-one variable bindings in the residual.
-We used this implicitly in the example in the previous section.
-
-If we change the $\next$ and $\prev$ rules to 
-\begin{mathpar}
-\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
-\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
-\end{mathpar}
-and treat $\next~q$ as a partial value for any residual $q$,
-then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
-environment bookkeeping in $\redonesym$, by inlining residuals instead of
-hoisting them in \verb|let|-bindings.
-From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
-
-Returning to our {\tt fexp} example, we can speculate on the stage \bbtwo\ term
-$\verb|fn b => prev{fexp(next{b},10)}|$
-to get the residual
-\begin{lstlisting} 
-fn b : int =>
-  let x0 =
-    (let x1 = 
-      let x2 = 1 * b
-      in x2 * x2
-    in x1 * x1) * b
-  in x0 * x0
-\end{lstlisting}
-
-Note how the recursion causes duplicate code in the residual.
-
-\subsubsection {Top-Level Evaluation}
-\label{sec:partialeval}
-
-The dynamics described in the previous section provide a formal description of partial evaluation.
-Recall that in partial evaluation systems, we start with a multivariate function $f$ for which some input is labeled {\em static} and 
-some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
-a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
-\[
-	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
-\]
-where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
-Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."
-
-% KAYVONF: good statement, but hold out for now
-%The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must %evaluate it many times.
-
-By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
-our dynamics can serve as a partial evaluator.   
-Specifically, we encode $f$ as a \lang\ expression with a function type of the form $A\to\fut(B\to C)$
-\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
-the following higher-order function which makes the adjustment:
-\begin{lstlisting} 
-let adjust (f : $int * int -> $int) =
-  fn (p : int) => 
-    next{
-      fn (b : int) => 
-        prev{f (next {b}, p)}
-    }
-\end{lstlisting}}.
-%
-Here $A$ represents the static input, $B$ represents the dynamic input, and $C$ represents the output.
-
-Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
-function:
-$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
-The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
-C)$, which by canonical forms must have the form $v = \next~\hat y$. 
-Next, we reify this environment into a sequence of \verb|let|-bindings
-enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
-Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
-and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
-
-That this sequence of evaluations is in fact staged follows from our
-characterizations of partial values, residuals, and values, that $\redonesym$
-outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
-
-%\begin{remark}
-%For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
-%always compute an empty environment, and a partial value identical to the result
-%of call-by-value evaluation of $e$.
-%%derivationally equivalent to standard call-by-value evaluation.
-%\end{remark}
-
-%\subsection{Metatheory}
-%
-%Recall that residuals live in \langTwo; we will indicate typing judgments in
-%\langTwo\ with $\vdash_\bbtwo$.
-%
-%%\begin{definition}
-%%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
-%%contains only stage-2 variables.
-%%\end{definition}
-%
-%\begin{definition}
-%An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
-%\begin{enumerate}
-%\item $\xi = \cdot$; or
-%\item $\xi = \xi',x:B\mapsto e$ where
-%$\Gamma\vdash\xi'\wf$ and
-%$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
-%%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
-%%$\Gamma,\dom{\xi'}\vdash e \res$.
-%\end{enumerate}
-%\end{definition}
-%
-%\begin{theorem}
-%If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
-%\end{theorem}
-%
-%\begin{theorem}
-%If $\diaonesub$ and $\typesone e A$ then
-%\begin{enumerate}
-%\item $\Gamma\vdash\xi\wf$;
-%\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
-%\item $\Gamma,\dom\xi\vdash v\pval$.
-%\end{enumerate}
-%\end{theorem}
-%
-%\begin{theorem}
-%If $\diatwosub$ and $\typestwo e A$ then
-%\begin{enumerate}
-%\item $\typeslangTwo q A$; and
-%\item $\Gamma\vdash_\bbtwo q\val$.
-%\end{enumerate}
-%\end{theorem}
-%
-%\begin{theorem}\label{thm:reify-type}
-%If $\Gamma\vdash\xi\wf$ and
-%$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
-%then 
-%$\reify{\xi}{\hat y}{q}$ and
-%$\typeslangTwo q A$.
-%\end{theorem}
-
-%\TODO
-%Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
-%integer term is guaranteed not to depend on the table, although one might be
-%produced. One may either discard the table, or evaluate everything in the table
-%(and terminating with the partial value iff everything in the table terminates).

### 2014-09-29 paper/semantics.tex
Split section 2 into it's own .tex file.

--- 2014-09-29 paper/semantics.tex
@@
-\section {Example: Staged Quickselect in \lang}
-\label{sec:staging}
-
-\begin{figure*}
-\label{fig:quickselect}
-\begin{minipage}{0.5\textwidth}
-\begin{lstlisting} 
-datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) 
-  : (int*list*list) =
-  case unroll l of 
-    Empty => (0,Empty, Empty) 
-  | Cons (h,t) =>
-      let (s,left,right) = partition (p,t) in
-      if h<p 
-      then (s+1,Cons(h,left),right)
-      else (s,left,Cons(h,right))
-
-fun qSelect (l : list, k : int) : int = 
-  case l of
-    Empty => 0
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-        case compare k n of
-          LT => qSelect (left, k)
-        | EQ => h
-        | GT => qSelect (right, k-n-1)
-\end{lstlisting}
-\caption{Unstaged implementation of quickselect.}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-\begin{lstlisting} 
-datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...
-
-fun qsStaged (l : list, k : $int) : $int = 
-  case l of
-    1`Empty` => next {2`0`}
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      next{
-        let n = hold{n} in
-          case compare prev{k} n of
-            LT => prev {qsStaged (left k)}
-          | EQ => hold {h}
-          | GT => prev {qsStaged (right, 
-                             next{prev{k}-n-1)}}
-      }	
-\end{lstlisting}
-\caption{Staged implementation of quickselect in \lang.}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-\begin{lstlisting}
-datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...
-
-datatype tree = Branch of int * int * tree * tree
-                | Leaf
-
-fun qSelect1 (l : list) : tree =
-  case l of
-    Empty => Leaf
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      Branch (n, h, qSelect1 left, qSelect1 right)
-
-fun qSelect2 (p : tree, k : int) : int = 
-  case unroll p of
-    Leaf => 0
-  | Branch (n,h,p1,p2) => 
-      case compare k n of
-        LT => qSelect2 (p1,k)
-      | EQ => h
-      | GT => qSelect2 (p2,k-n-1)
-\end{lstlisting}
-\caption{Split (two-phase) implementation of quickselect.}
-\end{minipage}
-\caption{Caption place holder}
-\end{figure*}
-
-As an example of an algorithm amenable to staging techniques,
-consider the quickselect algorithm shown in Figure~2a.
-Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
-It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
-then recursing on the sublist containing the desired element.
-
-Importantly, one property of quickselect is that the partitioning of the list, even across recursive calls, depends only on the values of list elements, not on the value of $k$ 
-\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
-This independence allows the quickselect algorithm to be split into two functions:
-one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qSelect1} in Figure~2c),
-and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
-One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
-and then used to accelerate many future lookups for any value of $k$.
-
-This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
-\lang unambiguously specifies how its terms should be split
-% and features a type system that ensures valid splittings exist.
-% only minimal refactoring should be required to write terms within this language,
-%and from there, the type system should be enough to prove that a valid splitting exists.
-by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
-Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.  
-After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
-will end up in the stage \bbone\ function ({\tt qSelect1}), 
-and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
-Typing rules for \lang ensure a valid splitting exists.  That is, 
-within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
-but never from \bbtwo\ to \bbone.
-
-Figure 2b shows a staged implementation of quickselect in \lang.
-As before, the function accepts as input the list to select from and a value of $k$,
-but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
-The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
-whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
-Naturally the output type, representing the $k$th largest element of the list,
-is also $\fut\rmint$, since it is not computed until stage \bbtwo.
-
-Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
-Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
-whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
-(We adapt the convention that the top-level context is stage \bbone.)
-The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
-For example, \verb|next{0}| from above has the type $\fut\rmint$.
-Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
-For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
-These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
-which is necessary to admit a properly staged implementation.
-
-The code also contains two $\pause$ blocks.  
-This construct has same stage signature as $\prev$,
-but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage~\bbone\ to stage~\bbtwo.
-\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
-We instead provide it as a primitive to shorten examples.  
-Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
-This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
-
-But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
-the staged version of quickselect in Figure~2b is virtually identical to the unstaged version in Figure~2a.
-The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
-It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
-but that would not be particularly interesting since it would result in an effectively trivial split 
-that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
-There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
-This process is known as {\em binding time analysis}, and we do not consider it here.
-Instead, we assume that all input programs are properly staged according to some programmer's intent.
-
 \section{\lang\ Statics and Dynamics}

### 2014-09-29 paper/example.tex
Further strengthening what is going on in the quickselect example.

@@
 As an example of an algorithm amenable to staging techniques,
-consider the quickselect algorithm shown in Figure~2a.
+consider the implementation of quickselect shown in Figure~2a.
 Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
@@
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
-Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes all partitions of the input list (to construct a binary search tree---see \texttt{qSelect1} in Figure~2c),
-and another which uses this tree to perform an accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
+Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in Figure~2c),
+and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
 One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once

### 2014-09-29 paper/example.tex
Minor edit.

@@
 
-Importantly, one property of quickselect is that the partitioning of the list, even across recursive calls, depends only on the values of list elements, not on the value of $k$ 
+Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.

### 2014-09-29 paper/example.tex
qs2 --> qSelect2

@@
 Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in Figure~2c),
-and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qs2}).
-One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
-and then used to accelerate many future lookups for any value of $k$.
+and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
+%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
+%and then used to accelerate many future lookups for any value of $k$.
 

### 2014-09-29 paper/intro.tex
moving around some stuff; merging techreport into paper

--- 2014-06-11 paper/intro.tex
@@
 
-We often find ourselves in situations where the input to a computation arrives in two parts, at different points in time.
+It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).
+This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).
+Examples include,
+\begin{itemize}
+\item Computing $b^e$, where $b$ varies and $e$ is constant.
+\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
+\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
+\end{itemize}
 
-The natural response is to try and do some part of the computation early, and the rest of the computation once the second input arrives.
+With these goals in mind, our initial concern is how to express multistage programs.
+To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
+An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.
 
-Since the computation runs in two parts, the total cost can be split as well, into $m+n$.
+Once we have appropriately written {\em multi-stage} programs, we then must execute them in a way that respects the staging.
+In this paper, we explore two techniques for this: {\em partial evaluation} and {\em stage-splitting}.
 
-Additionally, we can run the second stage multiple times.  Cost is now $m+bn$.
+The main idea of the former technique is to evaluate all of the constant portions of the program and embed the partial results into a {\em residual} program that depends only on the varying inputs.
+This is an old and well-explored idea, though the typical setup of partial evaluation implies a more powerful binding time analysis than is necessary for our language.
 
-Jorring et al.
-(\cite{jorring86}) identify three classes of staging techniques: meta-programming, partial evaluation, and stage-splitting.
-The first two of these have received significantly more attention than the third.
-Countless meta-programming systems exist (...twelve thousand citations...\cite{devito13}), and their background theory and type-systems are well understood (\cite{davies01}).
-Partial evaluation, too, is well-understood.
-Partial evaluation systems exist...
-.
-This paper explores both theory and applications for stage-splitting.
+The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.
+This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang.
+Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.
+
+\TODO motivate this example
+
+\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging.
+\Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
+Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.
+Notably, these examples make use of the stage-splitting features not available in prior work.

### 2014-09-29 paper/example.tex
a lot of work on section 3

@@
 It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
-then recursing on the sublist containing the desired element.
+then recurring on the sublist containing the desired element.
 
@@
 This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
-\lang unambiguously specifies how its terms should be split
+\lang\ unambiguously specifies how its terms should be split
 % and features a type system that ensures valid splittings exist.
@@
 and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
-Typing rules for \lang ensure a valid splitting exists.  That is, 
+Typing rules for \lang\ ensure a valid splitting exists.  That is, 
 within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,

### 2014-09-29 paper/intro.tex
a lot of work on section 3

@@
 This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).
-Examples include,
+Examples include:
 \begin{itemize}
@@
 This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang.
-Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
+Prior examples include the ``data specialization" work of Knoblock and Ruf\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
 Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

### 2014-09-29 paper/semantics.tex
a lot of work on section 3

@@
 
-As stated in \cref{sec:staging}, we use \lang,
-a typed two-stage lambda calculus featuring products, sums, and isorecursive types, as language to define and analyze the splitting transform.
-The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
-restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
-simplicity of presentation.  All of these techniques could feasibly be extended to more
-complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.
+\lang\ is a typed two-stage lambda calculus with products, sums, and isorecursive types.
+Its statics are adapted from the linear temporal logic given by \cite{davies96},
+restricted to two stages and extended with general sums and recursion.%
+\footnote{This restriction to two stages is made primarily for simplicity of
+presentation. All of these techniques could feasibly be extended to more
+complicated stage systems.}
+%The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of 
+%, as language to define and analyze the splitting transform.
+
+%The following subsections provide a brief overview of the type system for \lang,
+%and highlight important differences in its semantics from this prior work.
 
@@
 
-In \lang, the $\fut$ constructor encapsulates a stage \bbtwo\ type at stage~\bbone.  
-But because of the two-stage assumption, $\fut$ has no meaning when used at stage~\bbtwo.
-This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
-In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
-This context is only augmented by recursive types.
+In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
+stage~\bbone\ term. To ensure that the staging features are only applied to
+valid terms, we index our judgments not only by a type, but also a stage $w$.
 
-The typing judgment, like type validity, is annotated with a stage.
-Thus we can read $\typeswor x A$ as saying that term $x$ has type $A$ at world $w$ under the context $\Gamma$.
-The typing rules are given in \cref{fig:statics}.
-The staging features ($\next$, $\prev$, and $\pause$) behave as described in the previous section,
-and all other features behavior in the standard way and are essentially agnostic to stage. 
-Variables in the context are annotated with the stage at which they were introduced 
-and can only be used at the same stage.
-As one would expect, the typing judgment only produces well-formed types 
-(by induction on the derivation of the typing judgment).
+The type validity judgment $\Delta \vdash A \istypewor$, defined in
+\cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
+example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
+$\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
+$\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
+typing judgment only produces valid types.
+%(by induction on the derivation of the typing judgment).
+% under the assumptions in $\Delta$. This context is only augmented by recursive types.
+
+Except for the staging features $\next$, $\prev$, and $\pause$, the language
+features are essentially agnostic to stage. Variables in the context are
+annotated with the stage at which they were introduced and can only be used at
+the same stage.
+
+Note that there is no way to eliminate a term of type $\fut A$ into a stage
+\bbone\ term---the only elimination form is $\prev$, which results in a stage
+\bbtwo\ term. This is why well-typed terms cannot have any information flow
+from stage \bbtwo\ to stage \bbone, and what makes splitting possible.
 
@@
 
-A key attribute of the quickselect example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
-Ordinary term evaluation eliminates outermost redexes first, 
-however in the case that stage \bbone\ expressions are contained inside stage~\bbtwo\ ones 
-(such as the recurive call to {\tt fexp} above), 
-this strategy conflicts with the precept of staged execution: 
-that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.
+In the quickselect example, stage~\bbone\ terms are contained within
+stage~\bbtwo\ terms---for example, in the recursive call to {\tt qsStaged}. In
+this case, if we use an ordinary evaluation strategy reducing outermost redexes
+first, we would no longer have staged evaluation, as stage~\bbtwo\ code would
+be evaluated before stage~\bbone\ code.
 
 Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
-subexpressions before any of its stage~\bbtwo\ subexpressions. This results in a
-stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
-monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 
-$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
-judgment are not shown, but they are standard).
+subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
+a stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
+monostaged language called \langTwo. Then, at stage~\bbtwo, we complete
+evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
+rules for this judgment are not shown, but they are standard.)
 
-\subsubsection{Non-Duplicating First-Stage Evaluation}
+\subsubsection{Non-Duplicating Stage~\bbone\ Evaluation}
 
 To gain intuition about the challenges of implementing this staged dynamic
-semantics, consider the following example:
+semantics, consider the program:
 \begin{lstlisting}
-#2 (next {fib 20}, 2+3)
+#2 (next {f 20}, 2+3)
 \end{lstlisting}
-This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
-stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. 
-In this example, ${\tt fib} : \rmint \to \rmint$ is a stage-\bbtwo\ reference to the Fibonacci function.
-A conventional call-by-value evaluation strategy demands that we evaluate both components of
-the pair to values before we project from the pair. The problem is that
-while \verb|next {fib 20}| is not a value (in the sense that additional
-stage-\bbtwo\ computation steps are necessary to produce the result \verb|6765|), 
-evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
-Intuitively, the solution is to designate \verb|next {fib 20}| as a value \emph{in
-stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
-Therefore, we evaluate the pair to
+This is a stage~\bbone\ expression of type $\rmint$; the pair inside it is a
+stage~\bbone\ expression of type $(\fut\rmint)\times\rmint$; and ${\tt f} :
+\rmint \to \rmint$ is an expensive stage~\bbtwo\ function which we would like
+to avoid evaluating. A conventional call-by-value semantics would fully
+evaluate both components of the pair before projecting the second component.
+The problem is that, while \verb|next {f 20}| is not a value (in the sense that
+additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
+evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
+evaluation.
+
+Intuitively, the solution is to designate \verb|next {f 20}| as a value
+\emph{in stage \bbone}, even though it requires additional evaluation in stage
+\bbtwo. Therefore, the pair evaluates to
 \begin{lstlisting}
-#2 (next {fib 20}, 5)
+#2 (next {f 20}, 5)
 \end{lstlisting}
-then obtain \verb|5| as the result of projection.
+and the projection, in turn, immediately evaluates to \verb|5|.
 
-Now consider a more complex example where 
-stage-\bbone\ evaluation must substitute such a incompletely-evaluated
-expression. The following stage-\bbtwo\ term has type $\rmint$:
+Now consider a more complex example where stage~\bbone\ evaluation must
+substitute such a incompletely-evaluated expression. The following
+stage-\bbtwo\ term has type $\rmint$:
 \begin{lstlisting} 
 prev{
-  let x = (next {fib 20}, 3+4) in
+  let x = (next {f 20}, 3+4) in
   next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
@@
 \end{lstlisting}
-Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
-depends on the value of \verb|fib 20|, which is not reduced to \verb|3| until
-stage \bbtwo.
+Again, this term does not fully reduce at stage~\bbone, because it depends on
+the value of \verb|f 20|, which is not reduced until stage \bbtwo.
 
-If we simply treat \verb|(next {fib 20}, 7)| as a value during stage~\bbone, and
+If we simply treat \verb|(next {f 20}, 7)| as a value during stage~\bbone, and
 substitute it for the three occurrences of \verb|x| in the body of the
-\verb|let| expression, the result of stage \bbone\ computation is
+\verb|let| expression, the stage \bbone\ result would be
 \begin{lstlisting} 
@@
   next{ 
-    prev{next {fib 20}} * prev{next {fib 20}} * 7 
+    prev{next {f 20}} * prev{next {f 20}} * 7 
   }
@@
 \end{lstlisting}
-Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
+Finally, the $\prev$s and $\next$s cancel to yield the residual:
 \begin{lstlisting} 
-(fib 20) * (fib 20) * 7
+(f 20) * (f 20) * 7
 \end{lstlisting}
-Note how stage-\bbtwo\ evaluation of this expression will compute \verb|fib 20| twice.  
-To avoid duplicating computations, we take a different approach.  Instead of
-duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
-a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
-This produces:
-\begin{lstlisting} 
-let y = fib 20 in y * y * 7
-\end{lstlisting}
+But now stage \bbtwo\ must compute the expensive function call \verb|f 20|
+twice! To avoid this duplication, instead of treating \verb|next {f 20}| as a
+value and substituting it into the body of the \verb|let|, we instead bind
+\verb|f 20| to a stage~\bbtwo\ variable $\mathtt{\hat y}$, and substitute that
+variable.
 
-Achieving this behavior mechanically requires us to resolve a contradiction:
-we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
-Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.
-This substitution then floats up to the top of the containing $\prev$ block:
+More precisely, when evaluating the \verb|next|, we construct an explicit
+substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
+fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
+containing $\prev$ block:
 \begin{lstlisting} 
 prev {
-[yhat|->fib 20]
+[yhat|->f 20]
   let x = (next{yhat}, 7) in
@@
 \end{lstlisting}
-As a convention, we render the new variable with a %stylish and fashionable
-hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
+%As a convention, we render the new variable with a %stylish and fashionable
+%hat.  
+We proceed with stage~\bbone\ evaluation, which now duplicates the variable
+$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
 \begin{lstlisting} 
 prev {
-[yhat|->fib 20]
+[yhat|->f 20]
   next{
@@
 \end{lstlisting}
-To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
-As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
+To evaluate the large $\next$, we must first \emph{speculate} inside of it, to
+evaluate all the stage~\bbone\ terms contained within $\prev$s. These
+necessarily reduce to $\next$ expressions which cancel the $\prev$s, resulting
+in:
 \begin{lstlisting} 
 prev {
-[yhat|->fib 20]
+[yhat|->f 20]
     next{ yhat * yhat * 7 }
@@
 prev {
-[yhat|->fib 20]
+[yhat|->f 20]
 [zhat|->yhat*yhat*7]
@@
 \begin{lstlisting} 
-let yhat = fib 20 in
+let yhat = f 20 in
 let zhat = yhat * yhat * 7 in z
@@
 
-Thus we have evaluated all of the stage \bbone\ expressions of this program without duplicating the contents of $\next$ blocks.
+This concludes stage \bbone\ evaluation of the program---we have reduced all
+\bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
+evaluation, $\tworedsym$, then reduces this to a numeral.
 
@@
 
-\input {figures/dynamics}
+\input{figures/dynamics}
 
-The goal of the dynamics is to perform the above style of first-stage evaluation automatically.
-It comprises judgments: $\redonesym$ (evaluation),
-$\redtwosym$ (speculation), and $\reifysym$ (reification of substitutions into let bindings).
+Our dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
+described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
+(speculation), and $\reifysym$ (reification of substitutions into let
+bindings).
 
@@
 \begin{itemize}
-\item 
-Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
-but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
-In the example above, 
+\item Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully
+evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
+blocks.  In the example above, 
 \verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
@@
 
-\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
-are the results of a computation after both stages have been completed. In the
-example above, the term evaluates to \verb|63|.
+%\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
+%are the results of a computation after both stages have been completed. In the
+%example above, the term evaluates to \verb|63|.
 \end{itemize}
@@
 The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
-environment} $\xi$ and a partial value $v$.  The future environment is a mapping
-from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
-$v$) to residuals---in our example above, we represented this environment with
-explicit substitutions. For all of the normal features of \lang\ 
-(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
-effect on (partial) values as does standard evaluation, and the final future environment is
-gotten by merging the future environments of the subterms.
+environment} $\xi$ and a partial value $v$.  The future environment implements
+the explicit substitutions in our example---it maps fresh stage \bbtwo\
+variables (which may appear inside $\next$ blocks in $v$) to
+residuals. For non-staging features of \lang\ (\ref{fig:diaSemanticsCore}),
+$\redonesym$ is essentially standard call-by-value evaluation, and gathers the
+subterms' future environments into a single one.
 
-When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
-searches into the block's stage \bbtwo\ content to find any contained stage
-\bbone\ subexpressions and evaluate them in place.  This search process, called
-\emph{speculation}, is implemented by the $\redtwosym$ judgment, which takes
-a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
-speculated into a residual ($q$), the output of $\redonesym$ is a fresh variable wrapped in a
-$\next$ block ($\next~\hat y$), along with a future environment which maps that
-variable to the residual ($\hat y \mapsto q$).
+At $\next~e$, in contrast, $\redonesym$ must search through $e$ for any stage
+\bbone\ subexpressions, and evaluate them.  (See \ref{fig:diaSemanticsNP}.)
+This search process, called \emph{speculation}, is implemented by the
+$\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
 
-At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
-nothing but recursively speculate into every subexpression.  Once speculation
-finds a $\prev$ block, it resumes stage \bbone\ evaluation of the contents, which
-produces a future context and (by canonical forms) a $\next$-wrapped variable
-($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
-judgment) into a series of let bindings with $\mathtt{\hat y}$, stripped of
-its $\next$, at the bottom.
+For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
+all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
+which produces a future context $\xi$ and (by canonical forms) an expression
+$\next\{\hat y\}$. We then reify $\xi$ using $\reifysym$ into a series of let
+bindings enclosing $\mathtt{\hat y}$.
+
+Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
+fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
+environment which maps that variable to the residual ($\hat y \mapsto q$).
 
@@
 Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
-statement in its search for stage-\bbone\ code. 
-Thus the evaluation of that stage one code will occur {\em regardless of the eventual value of the predicate},
+statement in its search for stage~\bbone\ code. 
+Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
 and so a term like 
@@
 
-Returning to our {\tt fexp} example, we can speculate on the stage \bbtwo\ term
-$\verb|fn b => prev{fexp(next{b},10)}|$
-to get the residual
-\begin{lstlisting} 
-fn b : int =>
-  let x0 =
-    (let x1 = 
-      let x2 = 1 * b
-      in x2 * x2
-    in x1 * x1) * b
-  in x0 * x0
-\end{lstlisting}
-
-Note how the recursion causes duplicate code in the residual.
-
 \subsubsection {Top-Level Evaluation}
@@
 
-The dynamics described in the previous section provide a formal description of partial evaluation.
-Recall that in partial evaluation systems, we start with a multivariate function $f$ for which some input is labeled {\em static} and 
-some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
-a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
+These dynamics allow us to define a partial evaluator.
+Recall that partial evaluators operate on multivariate functions $f$, whose
+inputs are each labeled either {\em static} or {\em dynamic}. Once the static
+inputs are provided, partial evaluation of $f$ produces a residual depending on
+only the dynamic input. Mathematically, a partial evaluator is any $p$ such
+that
 \[
@@
 \]
-where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
-Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."
+where $\llbracket \cdot \rrbracket$ translates the text of a function to its
+mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
+input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
+specialized to $x$}.
 
@@
 
-By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
-our dynamics can serve as a partial evaluator.   
-Specifically, we encode $f$ as a \lang\ expression with a function type of the form $A\to\fut(B\to C)$
-\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
-the following higher-order function which makes the adjustment:
-\begin{lstlisting} 
-let adjust (f : $int * int -> $int) =
-  fn (p : int) => 
-    next{
-      fn (b : int) => 
-        prev{f (next {b}, p)}
-    }
-\end{lstlisting}}.
+We define a partial evaluator for \lang\ by identifying static with stage
+\bbone\ and dynamic with stage \bbtwo.  Specifically, we encode $f$ as a \lang\
+expression with a function type of the form $A\to\fut(B\to C)$.%
+%\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
+%the following higher-order function which makes the adjustment:
+%\begin{lstlisting} 
+%let adjust (f : $int * int -> $int) =
+%  fn (p : int) => 
+%    next{
+%      fn (b : int) => 
+%        prev{f (next {b}, p)}
+%    }
+%\end{lstlisting}}
 %
-Here $A$ represents the static input, $B$ represents the dynamic input, and $C$ represents the output.
+Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.
 
@@
 That this sequence of evaluations is in fact staged follows from our
-characterizations of partial values, residuals, and values, that $\redonesym$
+characterizations of partial values and residuals, that $\redonesym$
 outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
+
+For example, to specialize quickselect to a particular list, 
+\begin{lstlisting}
+qsStaged [5,2,7,4,1]
+\end{lstlisting}
+\TODO finish example
 

### 2014-09-29 paper/semantics.tex
fix unicode in pdf bookmarks
small fix

@@
-\section{\lang\ Statics and Dynamics}
+\section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
 \label{sec:semantics}
@@
 This concludes stage \bbone\ evaluation of the program---we have reduced all
-\bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
+stage \bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
 evaluation, $\tworedsym$, then reduces this to a numeral.
@@
 
-The algorithm described above creates three different kinds of expressions which
-cannot be evaluated further at a particular stage:
+The algorithm described above creates two different kinds of expressions which
+cannot be evaluated further at stage \bbone:
 \begin{itemize}

### 2014-09-30 paper/semantics.tex
Fixes from meeting with Kayvon.

@@
 complicated stage systems.} 
-%The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of 
-%, as language to define and analyze the splitting transform.
-
-%The following subsections provide a brief overview of the type system for \lang,
-%and highlight important differences in its semantics from this prior work.
+Although \cite{davies96} was mostly interested in describing partial evaluation,
+the system is equally apt as an input to splitting.
 
@@
 
-Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
+Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
 subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
-a stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
-monostaged language called \langTwo. Then, at stage~\bbtwo, we complete
+a stage \bbtwo\ term with no stage~\bbone\ content, 
+which can be described as a term in a monostaged language called \langTwo. 
+Then, at stage~\bbtwo, we complete
 evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
@@
 \end{lstlisting}
-But now stage \bbtwo\ must compute the expensive function call \verb|f 20|
-twice! To avoid this duplication, instead of treating \verb|next {f 20}| as a
+But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
+There are some systems, typically called {\em metaprogramming}, 
+where this duplication would be considered intended behavior.
+For our applications, we find the duplication undesirable and avoid it.
+Accordingly, instead of treating \verb|next {f 20}| as a
 value and substituting it into the body of the \verb|let|, we instead bind
@@
 
-Our dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
+The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
 described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
@@
 
-The algorithm described above creates two different kinds of expressions which
+\lang\ creates two different kinds of expressions which
 cannot be evaluated further at stage \bbone:
@@
 This behavior is why the judgment is named ``speculation."
+Although it may seem undesirable, this speculative behavior is a deliberate feature.
+Indeed it is critical to the quickselect example, 
+wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
 
@@
 We used this implicitly in the example in the previous section.
+
+\TODO [below here will get moved to related work]
 

### 2014-09-30 paper/example.tex
fix unicode in pdf bookmarks
made changes from meeting with Umut

@@
-\section {Example: Staged Quickselect in \lang}
+\section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
 \label{sec:staging}
@@
 consider the implementation of quickselect shown in Figure~2a.
-Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
-It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
+Quickselect is an algorithm for finding the $k$-th largest element in a list.
+It works by partitioning the input list into sublists containing elements less than and greater than the head element, 
 then recurring on the sublist containing the desired element.
 
-Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
+One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
@@
 and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
+
+The benefits are substantial: for a randomly ordered list of length $n$, 
+the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
+whereas \texttt{qSelect2} takes $O(\log n)$ time.
+Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
 %One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once

### 2014-09-30 paper/relatedwork.tex
made changes from meeting with Umut

@@
 \section{Related Work}
+
+\subsection{Davies}
+
+\cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
+and showed the equivalence between that type system and existing one for binding time analysis. 
+That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
+That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
+In fact, if we change the $\next$ and $\prev$ rules of ours to 
+\begin{mathpar}
+\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
+\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
+\end{mathpar}
+and treat $\next~q$ as a partial value for any residual $q$,
+then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
+environment bookkeeping in $\redonesym$, by inlining residuals instead of
+hoisting them in \verb|let|-bindings.
+From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
+{\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
+
+\subsection{Specialization}
 
@@
 
+\subsection{Graphics Systems}
+
 Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.
@@
 [RTSL and SH]
-
-[Discuss Yong's recent paper here.
-It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.
-They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.
-If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]
-
-Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis.
-That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language.
-Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.
-The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.
-They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.

### 2014-09-30 paper/semantics.tex
made changes from meeting with Umut

@@
 \begin{lstlisting}
-#2 (next {f 20}, 5)
+(next {f 20}, 5)
 \end{lstlisting}
@@
 \end{lstlisting}
-Again, this term does not fully reduce at stage~\bbone, because it depends on
-the value of \verb|f 20|, which is not reduced until stage \bbtwo.
+As with the simpler example, this term does not fully reduce at stage~\bbone, 
+because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.
 
-If we simply treat \verb|(next {f 20}, 7)| as a value during stage~\bbone, and
+Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
 substitute it for the three occurrences of \verb|x| in the body of the
-\verb|let| expression, the stage \bbone\ result would be
+\verb|let| expression, yielding
+\begin{lstlisting} 
+prev{
+  next{ prev{#1 (next {f 20}, 7)} 
+      * prev{#1 (next {f 20}, 7)} 
+      * hold{#2 (next {f 20}, 7)} }
+}
+\end{lstlisting}
+To then evaluate the large $\next$, we must first search it for stage~\bbone\ subterms to reduce in place.
+This search process is called {\em speculation}.
+In our example this means reducing all three projections to give,
 \begin{lstlisting} 
@@
 \end{lstlisting}
-Finally, the $\prev$s and $\next$s cancel to yield the residual:
+Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
 \begin{lstlisting} 
@@
 For our applications, we find the duplication undesirable and avoid it.
-Accordingly, instead of treating \verb|next {f 20}| as a
-value and substituting it into the body of the \verb|let|, we instead bind
-\verb|f 20| to a stage~\bbtwo\ variable $\mathtt{\hat y}$, and substitute that
-variable.
+
+To get this behavior, we do not treat \verb|next {f 20}| as a value to be substituted, 
+but instead bind \verb|f 20| to a fresh stage~\bbtwo\ variable $\mathtt{\hat y}$
+and then substitute that variable.
 
@@
 \end{lstlisting}
-To evaluate the large $\next$, we must first \emph{speculate} inside of it, to
-evaluate all the stage~\bbone\ terms contained within $\prev$s. These
-necessarily reduce to $\next$ expressions which cancel the $\prev$s, resulting
-in:
+We then speculate and reduce the projections to get:
 \begin{lstlisting} 
@@
 
-The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
-environment} $\xi$ and a partial value $v$.  The future environment implements
+The $\redonesym$ judgment takes an stage \bbone\ term\footnote{
+The input to $\redonesym$ may be open on stage \bbtwo\ variables,
+but the type system ensures that those must occur under a $\next$.
+Consequently $\redonesym$ will never directly encounter a variable.
+} to a {\em future environment} $\xi$ and a partial value $v$.  
+The future environment implements
 the explicit substitutions in our example---it maps fresh stage \bbtwo\
@@
 wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
+As always, it is the programmer's responsibility to ensure that her staging annotations 
+do not cause a program to loop forever.
 
@@
 to avoid one-for-one variable bindings in the residual.
-We used this implicitly in the example in the previous section.
-
-\TODO [below here will get moved to related work]
-
-If we change the $\next$ and $\prev$ rules to 
-\begin{mathpar}
-\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
-\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
-\end{mathpar}
-and treat $\next~q$ as a partial value for any residual $q$,
-then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
-environment bookkeeping in $\redonesym$, by inlining residuals instead of
-hoisting them in \verb|let|-bindings.
-From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
 

### 2014-09-30 paper/semantics.tex
made comments about prev reduction more explicit

@@
 \end{lstlisting}
-We then speculate and reduce the projections to get:
+We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
 \begin{lstlisting} 
@@
 \end{lstlisting}
-Once again, we lift the contents of the $\next$ into a substitution:
+Once again, we lift the results of speculation into a substitution:
 \begin{lstlisting} 
@@
 \end{lstlisting}
-Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
+Finally, when canceling the last $\next$ and $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
 \begin{lstlisting} 
 let yhat = f 20 in
-let zhat = yhat * yhat * 7 in z
+let zhat = yhat * yhat * 7 in zhat
 \end{lstlisting}

### 2014-10-01 paper/example.tex
fix subfigure references

--- 2014-09-29 paper/example.tex
@@
 \begin{figure*}
-\label{fig:quickselect}
-\begin{minipage}{0.5\textwidth}
+\begin{subfigure}{0.5\textwidth}
 \begin{lstlisting} 
@@
 \caption{Unstaged implementation of quickselect.}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
+\label{fig:quickselect}
+\end{subfigure}%
+\begin{subfigure}{0.5\textwidth}
 \begin{lstlisting} 
@@
 \caption{Staged implementation of quickselect in \lang.}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
+\label{fig:qsstaged}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
 \begin{lstlisting}
@@
 \caption{Split (two-phase) implementation of quickselect.}
-\end{minipage}
+\label{fig:qssplit}
+\end{subfigure}
 \caption{Caption place holder}
@@
 As an example of an algorithm amenable to staging techniques,
-consider the implementation of quickselect shown in Figure~2a.
+consider the implementation of quickselect shown in \ref{fig:quickselect}.
 Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
@@
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
-Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in Figure~2c),
+Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
 and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
@@
 
-Figure 2b shows a staged implementation of quickselect in \lang.
+\ref{fig:qsstaged} shows a staged implementation of quickselect in \lang.
 As before, the function accepts as input the list to select from and a value of $k$,
@@
 But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
-the staged version of quickselect in Figure~2b is virtually identical to the unstaged version in Figure~2a.
+the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
 The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.

### 2014-10-01 paper/example.tex

@@
 consider the implementation of quickselect shown in \ref{fig:quickselect}.
-Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
-It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
+Quickselect is an algorithm for finding the $k$-th largest element in a list.
+It works by partitioning the input list into sublists containing elements less than and greater than the head element, 
 then recurring on the sublist containing the desired element.
 
-Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
+One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.

### 2014-10-01 paper/paper.tex
updating to PLDI formatting
Add section 2 to paper.tex as well.
moving around some stuff; merging techreport into paper
fix unicode in pdf bookmarks
made changes from meeting with Umut
fix subfigure references
fix spacing in listings

@@
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
+\documentclass[pldi,preprint]{sigplanconf}
 
@@
 \usepackage{listings}
-\usepackage{hyperref}
+\usepackage{enumitem}
+\usepackage{url}
+\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
 \usepackage{cprotect}
+\usepackage{subcaption}
+\usepackage[T1]{fontenc}
+%\usepackage[utf8x]{inputenc}
+%\PrerenderUnicode{λ}
 
@@
 \crefname{theorem}{Theorem}{Theorems}
+
+\captionsetup[subfigure]{subrefformat=simple,labelformat=simple}
+\renewcommand\thesubfigure{(\alph{subfigure})}
 
@@
         columns=fullflexible,
+        keepspaces=true,
         basicstyle=\ttfamily,
@@
 \copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
+%\doi{nnnnnnn.nnnnnnn}
 
@@
 
+\input{example}
 \input{semantics}
 
-\vfill\eject\pagebreak
+%\vfill\eject\pagebreak
 \input{splitting}

### 2014-10-01 paper/example.tex
changes from Kayvon and David's comments; also some color

@@
 \begin{lstlisting} 
-datatype list = Empty | Cons of int * list
+1`datatype list = Empty | Cons of int * list
 fun partition (p : int, l : list) = ...
 
-fun qsStaged (l : list, k : $int) : $int = 
+fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
   case l of
-    1`Empty` => next {2`0`}
+    Empty => next {`2`0`1`}
   | Cons (h,t) => 
@@
       next{
-        let n = hold{n} in
-          case compare prev{k} n of
-            LT => prev {qsStaged (left k)}
-          | EQ => hold {h}
-          | GT => prev {qsStaged (right, 
-                             next{prev{k}-n-1)}}
-      }	
+        `2`let n = hold{`1`n`2`} in
+          case compare prev{`1`k`2`} n of
+            LT => prev {`1`qsStaged (left k)`2`}
+          | EQ => hold {`1`h`2`}
+          | GT => prev {`1`qsStaged (right, 
+                             next{prev{k}-n-1)}`2`}`1`
+      }`
 \end{lstlisting}
@@
 \begin{lstlisting}
-datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...`
 
@@
 
-fun qSelect1 (l : list) : tree =
+1`fun qSelect1 (l : list) : tree =
   case l of
@@
       let (left,right,n) = partition h t in
-      Branch (n, h, qSelect1 left, qSelect1 right)
+      Branch (n, h, qSelect1 left, qSelect1 right)`
 
-fun qSelect2 (p : tree, k : int) : int = 
+2`fun qSelect2 (p : tree, k : int) : int = 
   case unroll p of
@@
       | EQ => h
-      | GT => qSelect2 (p2,k-n-1)
+      | GT => qSelect2 (p2,k-n-1)`
 \end{lstlisting}
@@
 \footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
-Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
+Therefore, if quickselect were to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
 and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).  
+Equationally, we can represent this relationship as 
+\[
+\mathtt{qSelect(l,k)} \equiv \mathtt{qSelect2(qSelect1(l),k)}
+\]
 
-The benefits are substantial: for a randomly ordered list of length $n$, 
+The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
 the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
@@
 whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
-(We adapt the convention that the top-level context is stage \bbone.)
+(We adopt the convention that the top-level context is stage \bbone.)
 The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  

### 2014-10-01 paper/paper.tex
changes from Kayvon and David's comments; also some color

@@
 
-\lstset{language=ML,
+\lstset{%language=ML,
         columns=fullflexible,

### 2014-10-01 paper/semantics.tex
fix spacing in listings
changes from Kayvon and David's comments; also some color

@@
 
-\subsubsection{Non-Duplicating Stage~\bbone\ Evaluation}
+\subsubsection{Stage~\bbone\ Evaluation}
 
@@
 \begin{lstlisting}
-#2 (next {f 20}, 2+3)
+1`#2 (next {`2`f 20`1`}, 2+3)`
 \end{lstlisting}
-This is a stage~\bbone\ expression of type $\rmint$; the pair inside it is a
-stage~\bbone\ expression of type $(\fut\rmint)\times\rmint$; and ${\tt f} :
-\rmint \to \rmint$ is an expensive stage~\bbtwo\ function which we would like
+This is a stage~\bbone\ expression of type $\rmint~@~\bbone$; the pair inside it is a
+stage~\bbone\ expression of type $(\fut\rmint)\times\rmint~@~\bbone$; and ${\tt f} :
+\rmint \to \rmint~@~\bbtwo$ is an expensive stage~\bbtwo\ function which we would like
 to avoid evaluating. A conventional call-by-value semantics would fully
@@
 \begin{lstlisting}
-(next {f 20}, 5)
+1`(next {`2`f 20`1`}, 5)`
 \end{lstlisting}
@@
 Now consider a more complex example where stage~\bbone\ evaluation must
-substitute such a incompletely-evaluated expression. The following
+substitute such an incompletely-evaluated expression. The following
 stage-\bbtwo\ term has type $\rmint$:
 \begin{lstlisting} 
-prev{
-  let x = (next {f 20}, 3+4) in
-  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
-}
+2`prev{`
+  1`let x = (next {`2`f 20`1`}, 3+4) in
+  next{` 2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}` 1`}`
+2`}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev{
-  next{ prev{#1 (next {f 20}, 7)} 
-      * prev{#1 (next {f 20}, 7)} 
-      * hold{#2 (next {f 20}, 7)} }
-}
+2`prev{`
+  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
+      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
+      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
+}`
 \end{lstlisting}
-To then evaluate the large $\next$, we must first search it for stage~\bbone\ subterms to reduce in place.
+The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
+So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
 This search process is called {\em speculation}.
@@
   next{ 
-    prev{next {f 20}} * prev{next {f 20}} * 7 
+    prev{next {2`f 20`}} 2`*` prev{next {2`f 20`}} 2`* 7` 
   }
@@
 \begin{lstlisting} 
-(f 20) * (f 20) * 7
+2`(f 20) * (f 20) * 7`
 \end{lstlisting}
@@
 prev {
-[yhat|->f 20]
-  let x = (next{yhat}, 7) in
-  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
+[2`yhat`|->2`f 20`]
+  1`let x = (`next{2`yhat`}1`, 7) in`
+  next{prev{1`#1 x`} 2`*` prev{1`#1 x`} 2`*`  hold{1`#2 x`}}
 }
@@
 prev {
-[yhat|->f 20]
+[2`yhat`|->2`f 20`]
   next{
-    prev{#1 (next {yhat}, 7)} * 
-    prev{#1 (next {yhat}, 7)} *
-    hold{#2 (next {yhat}, 7)}
+    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*` 
+    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*`
+    hold{1`#2 (`next {2`yhat`}1`, 7)`}
   }
@@
 prev {
-[yhat|->f 20]
-    next{ yhat * yhat * 7 }
+[2`yhat`|->2`f 20`]
+    next{ 2`yhat * yhat * 7` }
 }
@@
 prev {
-[yhat|->f 20]
-[zhat|->yhat*yhat*7]
-    next{ zhat }
+[2`yhat`|->2`f 20`]
+[2`zhat`|->2`yhat * yhat * 7`]
+    next{ 2`zhat` }
 }
 \end{lstlisting}
-Finally, when canceling the last $\next$ and $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
+Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
 \begin{lstlisting} 
-let yhat = f 20 in
-let zhat = yhat * yhat * 7 in zhat
+2`let yhat = f 20 in
+let zhat = yhat * yhat * 7 in zhat`
 \end{lstlisting}
@@
 The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
-described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
-(speculation), and $\reifysym$ (reification of substitutions into let
-bindings).
+described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
+(speculation), and the auxiliary $\reifysym$ (reification). 
+Along with these are three judgments identifying forms of values:
+$\pvalsym$ (partial values), $\ressym$ (residuals), 
+and the auxiliary $\fcon$ (future context).
+The interrelation of these judgments can be summarized as follows:
+\begin{itemize}
+\item Evaluation sends a stage~\bbone\ term to a partial value.
+Because of the non-duplicating behavior, this sometimes produces explicit substitutions,
+which are aggregated into a single {\em future context} that is also the output of evaluation.
+\item Speculation sends a stage~\bbtwo\ term to a residual.
+\item Reification sends a residual and a future context to another residual.  
+\item Evaluation depends on speculation where stage~\bbone\ encapsulates stage~\bbtwo: $\next$. 
+\item Speculation depends on evaluation where stage~\bbtwo\ encapsulates stage~\bbone: $\prev$ and $\pause$.
+Reification is also used at these sites to handle the future context produced by evaluation.
+\end{itemize}
 
-\lang\ creates two different kinds of expressions which
-cannot be evaluated further at stage \bbone:
-\begin{itemize}
-\item Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully
+Partial values are stage \bbone\ terms that have been fully
 evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
 blocks.  In the example above, 
-\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
-
-\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
+\begin{lstlisting} 
+1`(next {`2`yhat`1`},7)`
+\end{lstlisting}
+is a partial value.
+Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
 stage \bbone\ subexpressions have all been fully evaluated. In the example
 above,
-\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
-
-%\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
-%are the results of a computation after both stages have been completed. In the
-%example above, the term evaluates to \verb|63|.
-\end{itemize}
+\begin{lstlisting} 
+2`let zhat = yhat*yhat*7 in zhat`
+\end{lstlisting}
+is a residual,
+These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
+They are different in that ``partial value" is a classification of stage~\bbone\ terms,
+while ``residual" is a classification of stage-\bbtwo\ terms.
 
@@
 
-At $\next~e$, in contrast, $\redonesym$ must search through $e$ for any stage
-\bbone\ subexpressions, and evaluate them.  (See \ref{fig:diaSemanticsNP}.)
-This search process, called \emph{speculation}, is implemented by the
-$\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
-
+At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
+subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
+This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
 For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
@@
 which produces a future context $\xi$ and (by canonical forms) an expression
-$\next\{\hat y\}$. We then reify $\xi$ using $\reifysym$ into a series of let
+$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
 bindings enclosing $\mathtt{\hat y}$.
@@
 next{
-  if true 
-  then hold{1+2} 
-  else prev{spin() (* loops forever *)}
+  2`if true 
+  then `hold{1`3+4`} 
+  2`else` prev{1`spin() (* loops forever *)`}
 }

### 2014-10-03 paper/relatedwork.tex
related work/meta-programming

@@
 {\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
+
+\subsection{Multi-stage (meta) programming}
+{\bf Umut: jotting down notes for now.}
+
+Multi-stage, a.k.a., meta programming or staged programming, gives the
+programmer linguistic facilities consisting of language constructs
+(usually also a type system, including a dynamic semantics) for
+expressing programs where inputs may come in stages.  Earlier work on
+staged programming was motivated by understanding the effects of
+binding-time analysis that partial evaluation relied on and allowing
+control over staging so that the programmer can control the
+termination and efficiency characteristics of staged programs.
+Several different approaches to staged programming has been proposed.
+
+Davies ...
+
+Taha's approach is similar to Davies, but extends Davies's approach
+with ``cross-stage persistency'' which allows a variable to be used in
+any later stage and a ``run'' construct that allows evaluating an
+expression.  Our language constructs therefore are also close to
+staged programming such as proposed by Taha.  In fact our static
+semantics is essentially identical to Taha's.  Though, as Davies, we
+don't include ``run'' in the language.  The dynamic semantics of
+staged languages... COMPLETE
+
+The key difference between our work and Taha's and similar work on
+staging is our splitting algorithm that can translate the staged code
+to a different code that can run significantly (sometimes
+asymptotically) faster. COMPARE DYNAMIC SEMANTICS
+
+ Unlike the evaluati
 

### 2014-10-06 paper/paper.tex
more code colors

--- 2014-10-01 paper/paper.tex
@@
 \usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
+\usepackage[usenames,dvipsnames]{color}
 \usepackage{cprotect}
@@
 
+\definecolor{stageOneColor}{RGB}{120,30,0}
+\definecolor{stageTwoColor}{RGB}{0,0,127}
 \begin{document}
 
-\lstset{language=ML,
+\lstset{%language=ML,
         columns=fullflexible,
@@
         escapeinside={"*}{*"},
-        moredelim=*[is][\color{red}]{1`}{`},
-        moredelim=*[is][\color{blue}]{2`}{`},
+        moredelim=*[is][\color{stageOneColor}]{1`}{`},
+        moredelim=*[is][\color{stageTwoColor}]{2`}{`},
         literate={|->}{{\ $\mapsto$\ }}1

### 2014-10-06 paper/semantics.tex
more code colors

@@
 \begin{lstlisting} 
-prev{
-  next{ 
-    prev{next {2`f 20`}} 2`*` prev{next {2`f 20`}} 2`* 7` 
-  }
-}
+2`prev{`1`
+  next{`2` 
+    prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
+  `1`}`2`
+}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev {
-[2`yhat`|->2`f 20`]
-  1`let x = (`next{2`yhat`}1`, 7) in`
-  next{prev{1`#1 x`} 2`*` prev{1`#1 x`} 2`*`  hold{1`#2 x`}}
-}
+2`prev {
+[yhat|->f 20]
+  `1`let x = (next{`2`yhat`1`}, 7) in
+  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
+}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev {
-[2`yhat`|->2`f 20`]
-  next{
-    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*` 
-    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*`
-    hold{1`#2 (`next {2`yhat`}1`, 7)`}
-  }
-}
+2`prev {
+[yhat|->f 20]
+  `1`next{`2`
+    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
+    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
+    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
+  `1`}`2`
+}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev {
-[2`yhat`|->2`f 20`]
-    next{ 2`yhat * yhat * 7` }
-}
+2`prev {
+[yhat|->f 20]
+    `1`next{`2`yhat * yhat * 7`1`}
+`2`}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-prev {
-[2`yhat`|->2`f 20`]
-[2`zhat`|->2`yhat * yhat * 7`]
-    next{ 2`zhat` }
-}
+2`prev {
+[yhat|->f 20]
+[zhat|->yhat * yhat * 7]
+    `1`next{`2`zhat`1`}`2`
+}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-next{
-  2`if true 
-  then `hold{1`3+4`} 
-  2`else` prev{1`spin() (* loops forever *)`}
-}
+1`next{
+  `2`if true 
+  then hold{`1`3+4`2`} 
+  else prev{`1`spin() (* loops forever *)`2`}`1`
+}`
 \end{lstlisting}

### 2014-10-06 paper/semantics.tex
added remarks

--- 2014-10-01 paper/semantics.tex
@@
 
+\ur{Partial values and residuals should be ``definitons'', i.e.,
+  use italics here.}
+
 Partial values are stage \bbone\ terms that have been fully
@@
 is a residual,
+\ur{Comma above! ``classification'' sounds odd to me.  just say they
+  ``are stage-1 terms.''}
+
 These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
@@
 while ``residual" is a classification of stage-\bbtwo\ terms.
+
+\ur{I would recommend using ``1-value'' and ``2-value'' instead of
+  partial values and residuals.  The former are more accurate, more
+  direct, more succinct, and does not lead to confusions with other
+  residuals.}
+
+\ur{Many incorrectly hyphenated ``partial values'' in this text.}
 
@@
 
+\ur{drop ``must''.}
 At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
@@
 
+\ur{This is really a continuation of the previous paragraph.}
 Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
@@
 will fail to evaluate at stage \bbone.
+\ur{This might be standard in the PE literature.  If that is the case
+  there is no need to elaborate as much.}
 This behavior is why the judgment is named ``speculation."
@@
 \label{sec:partialeval}
+
+\ur{The following several paragraphs should probably be somewhere
+  else, further ahead, or further back when we discuss the
+  relationship with PE.}
 

### 2014-10-06 paper/splitting.tex
added remarks

@@
 \label{sec:splitting}
+\ur{First para should probably be somewhere else concerning problem
+  definition and relationship to PE.  Based on this definition, it
+  might be good and correct to argue that this is more or less an
+  orthogonal problem.}
 
@@
 
+\ur{Reduce commentary.  Two sentences connected with an ``and.'' Split
+  sentences.}
+
 We now consider the problem of splitting stage-\bbone\ partial values.
@@
 
-There are many reasonable ways to achieve this goal, 
-and we choose one that preserves as much structure as possible from $v$.
-Formally, partial value splitting comprises a single judgment, $v \vsplito [v_1,v_2]$, the rules for which are given in \cref{fig:valSplit}.
-It essentially acts distributively at product, sum, and recursive types, 
-doing little more than recuring into subvalues and duplicating the structure to both results.
-At base types, partial value splitting assigns all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
-Conversely at $\fut$ types, splitting assigns all content ({\em i.e.} the environment reference) to stage \bbtwo, and uses a nullary tuple for stage \bbone.
+\ur{Next sentence is unnecessary commentary. It also sounds of
+  questinoable accuracy, not clear to me that there are many ways to
+  do this.}  There are many reasonable ways to achieve this goal, and
+we choose one that preserves as much structure as possible from $v$.
+Formally, partial value splitting comprises a single judgment, $v
+\vsplito [v_1,v_2]$, the rules for which are given in
+\cref{fig:valSplit}.  It essentially acts distributively at product,
+sum, and recursive types, doing little more than recuring into
+subvalues and duplicating the structure to both results.  At base
+types, partial value splitting assigns all content ({\em i.e.} the
+integer or boolean) to stage \bbone, and uses a nullary tuple
+placeholder for stage \bbtwo.  Conversely at $\fut$ types, splitting
+assigns all content ({\em i.e.} the environment reference) to stage
+\bbtwo, and uses a nullary tuple for stage \bbone.
 
-Functions are inherently more complicated. 
-Since the body of a function is an open term, the partial value splitting rule for functions must rely on general expression splitting, which is covered momentarily.
-The most we can say here is a function splits into two other functions, 
-whose bodies correspond to the stage \bbone\ and stage \bbtwo\ components of the original function's body.
+\ur{The case for ``fold'' seems missing in partial value splitting rules.}
+
+\ur{Drop, ``the most we can say.''  Be direct and explain.}
+Functions are inherently more complicated.  Since the body of a
+function is an open term, the partial value splitting rule for
+functions must rely on general expression splitting, which is covered
+momentarily.  The most we can say here is a function splits into two
+other functions, whose bodies correspond to the stage \bbone\ and
+stage \bbtwo\ components of the original function's body.
 
@@
 These are respectively called the {\em precomputation} and {\em residual}.
-\footnote{We also used the term {\em residual} for the result of partial evaluation.
-Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual} and {\em evaluation-residual}.}
+\footnote{We also used the term {\em residual} for the result of
+  partial evaluation.  Where necessary, we will distinguish between
+  these concepts by using the terms {\em splitting-residual} and {\em
+    evaluation-residual}.}  \ur{The reuse of residual is bothersome,
+  we should use different terminology, 1-value and 2-value perhaps.}
 
+\ur{We need a ``next'' I believe.  All our examples should start at
+  stage-1 for consistency.}
 For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
@@
 
+\ur{I have no idea what the next paragraph is trying to say.}
 Note that lifting and masking are different, despite both being within the purview of splitting.
@@
 
+\ur{``for both'' below is ambiguous.}
 The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
@@
 \footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
+\ur{Drop commentary about taste and such.  Explain that we could have
+  used lambda calculus as target but this is more direct.}
 
+\ur{Drop ``specifically''.}
 Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
@@
 For concision, we represent residuals as a term open on a single variable, rather than as functions.
-For both judgments, we use a context ($\Gamma$) to keep track of the open variables of $e$.
+For both judgments, we use a context ($\Gamma$) to keep track of the
+open variables of $e$.
+\ur{Case (pl./si.) mismatch in ``we represent ...'' sentence above.}
 
+\ur{Reformat rules to make them fit into the paper.}
+
+\ur{Explain the rules.  This is the most important part of the
+  paper. Work needed here.}
 The rules\footnote{The rules are written using patterns, including the open variable of the residual.} 

### 2014-10-06 paper/intro.tex
various edits with nico

@@
 
+\TODO integrate the following text into the new intro
+
+Recall that partial evaluators operate on multivariate functions $f$, whose
+inputs are each labeled either {\em static} or {\em dynamic}. Once the static
+inputs are provided, partial evaluation of $f$ produces a residual depending on
+only the dynamic input. Mathematically, a partial evaluator is any $p$ such
+that
+\[
+	\forall f,x. \exists f_x. \left[p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)\right]
+\]
+where $\llbracket \cdot \rrbracket$ translates the text of a function to its
+mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
+input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
+specialized to $x$}.
+
+\ur{  Based on this definition, it
+  might be good and correct to argue that this is more or less an
+  orthogonal problem.}
+
+Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
+However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
+before {\em any arguments} are available.
+Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
+where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
+and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
+More precisely, a stage splitter is any $s$ such that
+\[
+	\forall f. \exists f_1,f_2. 
+	\left[
+		\begin{array}{l}
+		s(f) = (f_1,f_2) \text{ and } \\
+		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
+		\end{array}
+	\right]
+\]
+with the same $\llbracket \cdot \rrbracket$ notation as before.
+Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
+meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
+This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.
+
 \Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging.

### 2014-10-06 paper/semantics.tex
various edits with nico

--- 2014-10-06 paper/semantics.tex
@@
 
-Except for the staging features $\next$, $\prev$, and $\pause$, the language
-features are essentially agnostic to stage. Variables in the context are
-annotated with the stage at which they were introduced and can only be used at
-the same stage.
+Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
+language features (i.e., the introductory and eliminatory forms for product,
+sum, and function types) preserve the stage of their subterms, and work at both
+stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
+\lang\ would simply consist of two non-interacting copies of a standard lambda
+calculus. Variables in the context are annotated with the stage at which they
+were introduced and can only be used at the same stage.
 
@@
 and the auxiliary $\fcon$ (future context).
-The interrelation of these judgments can be summarized as follows:
+These judgments are related as follows:
 \begin{itemize}
 \item Evaluation sends a stage~\bbone\ term to a partial value.
-Because of the non-duplicating behavior, this sometimes produces explicit substitutions,
-which are aggregated into a single {\em future context} that is also the output of evaluation.
+In order to avoid duplicating subterms, evaluation collects explicit
+substitutions into a single {\em future context} that is also the output of
+evaluation.
 \item Speculation sends a stage~\bbtwo\ term to a residual.
 \item Reification sends a residual and a future context to another residual.  
-\item Evaluation depends on speculation where stage~\bbone\ encapsulates stage~\bbtwo: $\next$. 
-\item Speculation depends on evaluation where stage~\bbtwo\ encapsulates stage~\bbone: $\prev$ and $\pause$.
-Reification is also used at these sites to handle the future context produced by evaluation.
+\item Evaluation depends on speculation for $\next$ terms, which have
+stage~\bbtwo\ subterms.
+\item Speculation depends on evaluation for $\prev$ and $\pause$, which have
+stage~\bbone\ subterms. The future context produced by evaluation is also
+reified here.
 \end{itemize}
 
-Partial values are stage \bbone\ terms that have been fully
-evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
+\emph{Partial values} are stage~\bbone\ terms that have been fully
+evaluated, but which may contain stage~\bbtwo\ variables wrapped in $\next$
 blocks.  In the example above, 
@@
 is a partial value.
-Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
+\emph{Residuals} ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
 stage \bbone\ subexpressions have all been fully evaluated. In the example
@@
 \end{lstlisting}
-is a residual,
-These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
-They are different in that ``partial value" is a classification of stage~\bbone\ terms,
-while ``residual" is a classification of stage-\bbtwo\ terms.
+is a residual.
+
+Note that partial values and residuals are both terms for which stage \bbone\
+computation has completed; however, partial values are stage~\bbone\ terms,
+while residuals are stage~\bbtwo\ terms.
+
+\ur{I would recommend using ``1-value'' and ``2-value'' instead of
+  partial values and residuals.  The former are more accurate, more
+  direct, more succinct, and does not lead to confusions with other
+  residuals.}
+
+\ur{Many incorrectly hyphenated ``partial values'' in this text.}
 
@@
 
-At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
+At $\next~e$, in contrast, $\redonesym$ speculates into $e$ for stage~\bbone\ 
 subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
@@
 bindings enclosing $\mathtt{\hat y}$.
-
 Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
@@
 
-These dynamics allow us to define a partial evaluator.
-Recall that partial evaluators operate on multivariate functions $f$, whose
-inputs are each labeled either {\em static} or {\em dynamic}. Once the static
-inputs are provided, partial evaluation of $f$ produces a residual depending on
-only the dynamic input. Mathematically, a partial evaluator is any $p$ such
-that
-\[
-	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
-\]
-where $\llbracket \cdot \rrbracket$ translates the text of a function to its
-mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
-input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
-specialized to $x$}.
-
 % KAYVONF: good statement, but hold out for now
@@
 
-We define a partial evaluator for \lang\ by identifying static with stage
-\bbone\ and dynamic with stage \bbtwo.  Specifically, we encode $f$ as a \lang\
-expression with a function type of the form $A\to\fut(B\to C)$.%
+These dynamics allow us to define a partial evaluator for \lang\ by identifying
+static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
+encode $f$ as a \lang\ expression with a function type of the form
+$A\to\fut(B\to C)$.%
 %\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply

### 2014-10-06 paper/splitting.tex
various edits with nico

@@
 \label{sec:splitting}
-\ur{First para should probably be somewhere else concerning problem
-  definition and relationship to PE.  Based on this definition, it
-  might be good and correct to argue that this is more or less an
-  orthogonal problem.}
-
-Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
-However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
-before {\em any arguments} are available.
-Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
-where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
-and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
-More precisely, a stage splitter is any $s$ such that
-\[
-	\forall f. \exists f_1,f_2. 
-	\left[
-		\begin{array}{l}
-		s(f) = (f_1,f_2) \text{ and } \\
-		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
-		\end{array}
-	\right]
-\]
-with the same $\llbracket \cdot \rrbracket$ notation as before.
-Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
-meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
-This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.
 
@@
 
-We now consider the problem of splitting stage-\bbone\ partial values.
+We now consider the problem of splitting stage~\bbone\ partial values.
 This serves as a simple didactic stepping-stone to the full formulation, 

### 2014-10-07 paper/splitting.tex
major work on 4.0 - 4.3

@@
 
-As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
-the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$, 
-which respectively split terms that type $@ \bbone$ and $@ \bbtwo$. 
-Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
-After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.
+%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
+%takes the form of two mutually-dependent judgments, $\splitonesym$ and
+%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
+%\bbtwo$.  Before diving into the full generality, 
+As discussed in the introduction, a stage splitter must take a term in \lang\
+and produce two
 
-\subsection{Partial Value Splitting}
+\TODO waiting on how the intro defines splitting
 
-\ur{Reduce commentary.  Two sentences connected with an ``and.'' Split
-  sentences.}
+In this section, we define splitting for partial values, then extend it to all
+terms, and prove correctness of this algorithm.
 
-We now consider the problem of splitting stage~\bbone\ partial values.
-This serves as a simple didactic stepping-stone to the full formulation, 
-and we will see later that it is a necessary component of the correctness theorem.
+\subsection{Partial Values}
 
-Recall that, along with a future environment, partial values are the output of the partial evaluation of stage \bbone\ terms.
-In addition to base types, tuples, and injections, partial values may also contain references into the future environment.
-In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
-The goal of partial value splitting is to take a partial value $v$ and from it create two single stage values $v_1$ and $v_2$,
-where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 
-and $v_2$ contains only the information from $v$ that is likewise relevant to second stage code.
-We call this action {\em masking}.
+Partial values, along with future environments, are produced by partially
+evaluating stage \bbone\ terms. Although partial values have no more stage
+\bbone\ computation remaining, they may contain variables bound by the future
+environment, and thus, may require stage \bbtwo\ computation.
 
-\ur{Next sentence is unnecessary commentary. It also sounds of
-  questinoable accuracy, not clear to me that there are many ways to
-  do this.}  There are many reasonable ways to achieve this goal, and
-we choose one that preserves as much structure as possible from $v$.
-Formally, partial value splitting comprises a single judgment, $v
-\vsplito [v_1,v_2]$, the rules for which are given in
-\cref{fig:valSplit}.  It essentially acts distributively at product,
-sum, and recursive types, doing little more than recuring into
-subvalues and duplicating the structure to both results.  At base
-types, partial value splitting assigns all content ({\em i.e.} the
-integer or boolean) to stage \bbone, and uses a nullary tuple
-placeholder for stage \bbtwo.  Conversely at $\fut$ types, splitting
-assigns all content ({\em i.e.} the environment reference) to stage
-\bbtwo, and uses a nullary tuple for stage \bbone.
+Splitting a partial value $v$ creates two single-stage values $v_1$ and $v_2$,
+where $v_w$ contains only the information required for stage $w$ evaluation of
+$v$. Splitting partial values is implemented by the judgment $v \vsplito
+[v_1,v_2]$ defined in \ref{fig:valSplit}. 
+At product, sum, and recursive types, splitting acts recursively, gathering all
+subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
+entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
+at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
+entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
+exposing only the relevant structure of a term is called \emph{masking}.
 
@@
 
-\ur{Drop, ``the most we can say.''  Be direct and explain.}
-Functions are inherently more complicated.  Since the body of a
-function is an open term, the partial value splitting rule for
-functions must rely on general expression splitting, which is covered
-momentarily.  The most we can say here is a function splits into two
-other functions, whose bodies correspond to the stage \bbone\ and
-stage \bbtwo\ components of the original function's body.
+Splitting a function is more complex, because function bodies are open terms,
+and so are split by the judgments described in the next subsection. Essentially,
+functions will split into two functions, whose bodies correspond to the stage
+\bbone\ and stage \bbtwo\ components of the original function body.
 
-\subsection{Expression Splitting}
+\subsection{General Splitting}
 
-We now consider the problem of splitting expressions in \lang, starting with stage \bbtwo\ terms.  
-The goal here is to end up with two new terms: one that has all of the stage \bbone\ content of the original, and one with all of the stage \bbtwo\ content. 
-These are respectively called the {\em precomputation} and {\em residual}.
+\ur{The reuse of residual is bothersome, we should use different terminology,
+1-value and 2-value perhaps.}
+
+Splitting arbitrary terms in \lang\ again results in two terms containing only
+the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
+called, respectively, the {\em precomputation} and {\em residual}.%
 \footnote{We also used the term {\em residual} for the result of
@@
 these concepts by using the terms {\em splitting-residual} and {\em
-    evaluation-residual}.}  \ur{The reuse of residual is bothersome,
-  we should use different terminology, 1-value and 2-value perhaps.}
+evaluation-residual}.}
+%
+For example, the stage \bbtwo\ term
+\begin{lstlisting}
+2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
+\end{lstlisting}
+splits into a precomputation \verb|(1+2,3+4)| and a residual
+\verb|fn (L1,L2) => L1 < L2-5|, with the property that the residual applied to
+the precomputation must agree with the original term. This process of hoisting
+all the stage \bbone\ subterms out of a term is called \emph{lifting}.
 
-\ur{We need a ``next'' I believe.  All our examples should start at
-  stage-1 for consistency.}
-For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
-We can split this into two terms, $p=``\verb|(1+2,3+4)|"$ and $r=``\verb|fn (L1,L2) => L1 < L2-5|"$, 
-with the as yet imprecise property that $r$ applied to $p$ yields the same final value as $e$.
-In general, all we're doing here is finding the contained nuggets of stage \bbone\ code and pulling them into a big tuple,
-an action we call {\em lifting}.
+Lifting and masking are two necessary ingredients for ensuring that splitting
+observes the stages of \lang. When splitting a partial value, the stage \bbone\
+result must \emph{mask} the stage \bbtwo\ computations in order to be
+single-staged. When splitting a stage \bbtwo\ term, the precomputation must
+\emph{lift} out all embedded stage \bbone\ computations, so they can be
+performed first, even though their results are only used in stage \bbtwo.
+%and it's accessible to whatever surrounding piece of code refers to the value being split.
+%is not accessible externally.
 
-\ur{I have no idea what the next paragraph is trying to say.}
-Note that lifting and masking are different, despite both being within the purview of splitting.
-They particularly differ in how the resulting stage \bbone\ parts are used.
-The $v_1$ of partial value splitting is essentially the stage \bbone\ content of {\em result} of computation, 
-and it's accessible to whatever surrounding piece of code refers to the value being split.
-In contrast, the $p$ of stage \bbtwo\ expression splitting is only meant to support the stage \bbtwo\ part,
-and is not accessible externally.
+Splitting a stage \bbone\ term, in contrast, requires both masking and splitting.
+Consider the term 
+\begin{lstlisting}
+1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
+\end{lstlisting}
+Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
+yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
+must occur at stage \bbone, the former is needed by stage \bbone, while the
+latter is needed by stage \bbtwo.
 
-The distinction between lifting and masking is important because stage \bbone\ expression splitting does both.
-To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
-The stage \bbone\ result of masking is \verb|((),3*4)|, 
-and the precomputation (i.e. the stage \bbone\ result of lifting) is \verb|(1+2)|.
-To get the total stage \bbone\ result of splitting, we just take the tuple of those two parts: \verb|(((),3*4),1+2)|.  
-This is called the {\em combined result} for stage \bbone.
-\footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}
-The splitting residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
-For our example, it is ``\verb|fn L => (L<5,())|".
+The precomputation for this term is a \emph{combined result} which is
+the pair of these components: \verb|(((),3*4),1+2)|. The splitting residual is
+likewise the pair of the stage \bbtwo\ results of lifting and masking:
+\verb|fn L => (L<5,())|.
+%\footnote{In this example, the combined result is a 2-tuple, but more generally
+%it only must be a term which reduces to a 2-tuple.}
 
@@
 
-\ur{``for both'' below is ambiguous.}
-The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
-which are in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
-For both, the input is a term in \lang, and the output is two terms in \langmono, an
-unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
-staging features (we say that implicitly it has only one stage), and it is untyped.  
-Modulo these differences, \langmono\ has all of the same features as \lang, 
-although this is a matter of taste
-\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
-\ur{Drop commentary about taste and such.  Explain that we could have
-  used lambda calculus as target but this is more direct.}
+%in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
+Splitting is implemented by the judgments $\splitonesym$ and $\splittwosym$,
+which take stage \bbone\ and stage \bbtwo\ terms respectively; these judgments
+are defined in \cref{fig:termSplitOne,fig:termSplitTwo} inductively on the
+structure of the term. Both judgments produce two terms in an unstaged language,
+\langmono, whose grammar is given in \ref{fig:monoGrammar}. This language is
+untyped and contains no staging features, but otherwise has the same term
+constructors as \lang.
+%\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
 
-\ur{Drop ``specifically''.}
-Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
-while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
-For concision, we represent residuals as a term open on a single variable, rather than as functions.
-For both judgments, we use a context ($\Gamma$) to keep track of the
-open variables of $e$.
-\ur{Case (pl./si.) mismatch in ``we represent ...'' sentence above.}
+The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
+$c$ and residual $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
+precomputation $p$ and a residual $l.r$. In both cases, the term $e$ is open on
+variables in the context $\Gamma$. We write the rules using pattern notation
+when appropriate.
+%For concision, we represent residuals as a term open on a single variable,
+%rather than as functions.
 
@@
   paper. Work needed here.}
-The rules\footnote{The rules are written using patterns, including the open variable of the residual.} 
-for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
+Every $\splittwosym$ rule works by bundling the precomputations of the constituent parts, and
+then unbundling them with a pattern.
+The rules for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
 The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
@@
 The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
-and projects out the integer result in the residual\footnote{The residual of an integer expression is usually trivial, 
+and projects out the integer result in the residual.%
+\footnote{The residual of an integer expression is usually trivial, 
 but we have to include it here for termination purposes.}
 
-The rules of stage \bbtwo\ splitting are given in \ref{fig:termSplitTwo}.  
-Every rule works by bundling the precomputations of the constituent parts, and then unbundling them with a pattern.
-The rules of stage \bbone\ splitting are given in \ref{fig:termSplitOne}.
+There are some particularly notable features of this definition of term
+splitting.
 
-\subsection {Properties}
+\paragraph {Expression Splitting at Values}
 
-We'll now call out a few properties to notice in the splitting rules.
+When applied to partial values, term splitting always produces a trivial
+precomputation, in the sense that the output is a nested sequence, possibly
+empty, of tuples and injections of \texttt{()}. The intuition is that a partial
+value has no remaining stage \bbone\ computation, and hence no precomputation.
 
-\subsubsection {Expression Splitting at Values}
+%Partial value splitting is more than just a restriction of general expression
+%splitting precisely because it observes this property it at the structural
+%level, doing away with the precomputation entirely.
 
-Note that general expression splitting always produces an effectively trivial precomputation when applied to partial values,
-where effectively trivial means that the precomputation is either a unit value, or a tuple or injection of an effectively trivial values.
+\paragraph {Speculation}
 
-Partial value splitting is more than just a restriction of general expression splitting 
-precisely because it observes this property it at the structural level, doing away with the precomputation entirely.
-This makes sense intuitively, since a precomputation would imply that there is some computation left to do, 
-which is antithetical to the notion of values.
+In the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the
+precomputation is lifted out from within branches. This is the manifestation of
+the speculation behavior from the semantics.
 
-\subsubsection {Speculation}
+\paragraph {Stage \bbone\ Divergence}
 
-Notice in the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the precomputation is lifted out from within branches.
-This is the manifestation of the speculation behavior from the semantics.
+In the splitting rules for stage \bbone\ {\tt if} and {\tt case} expressions, we
+have enough information at stage \bbone\ to know what branch to take, so there's
+no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
+taken branch only, and inject the precomputation into a sum type. In the
+residual, we case on that injection, and resume the stage \bbtwo\ portion of the
+correct branch.
 
-\subsubsection {Stage \bbone\ Divergence}
-Consider the splitting rules for stage-\bbone\ {\tt if} and {\tt case} expressions.
-In both cases we have enough information at stage \bbone\ to know what branch to take, so there's no need to speculate.
-Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject precomputation into a sum type.
-Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.
+\paragraph {Stage \bbone\ Functions}
 
-\subsubsection {Stage \bbone\ Functions}
-
-The trick when splitting stage \bbone\ functions is that the contents themselves may be multi-stage.
-We handle this by splitting them into two functions:
-one in the immediate result which handles all the stage \bbone\ content of the original,
-and one in the residual which handles all of the stage \bbtwo\ content.
-Note that stage \bbone\ $\lambda$-expressions themselves have only a trivial precomputation.
+The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
+handle this by splitting the function into two functions: one in the immediate
+result which handles all the stage \bbone\ content of the original, and one in
+the residual which handles all of the stage \bbtwo\ content. Note that stage
+\bbone\ lambdas are partial values, and so have only a trivial precomputation.
 

### 2014-10-07 paper/semantics.tex
tweaking the diagrams

@@
 
-The $\redonesym$ judgment takes an stage \bbone\ term\footnote{
-The input to $\redonesym$ may be open on stage \bbtwo\ variables,
-but the type system ensures that those must occur under a $\next$.
-Consequently $\redonesym$ will never directly encounter a variable.
-} to a {\em future environment} $\xi$ and a partial value $v$.  
-The future environment implements
-the explicit substitutions in our example---it maps fresh stage \bbtwo\
-variables (which may appear inside $\next$ blocks in $v$) to
-residuals. For non-staging features of \lang\ (\ref{fig:diaSemanticsCore}),
-$\redonesym$ is essentially standard call-by-value evaluation, and gathers the
-subterms' future environments into a single one.
+The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
+\bbone\ term%
+\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
+the type system ensures that those must occur under a $\next$. Consequently
+$\redonesym$ will never directly encounter a variable.}
+%
+to a {\em future environment} $\xi$ and a partial value $v$. The future
+environment implements the explicit substitutions in our example---it maps fresh
+stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
+residuals. For non-staging features of \lang, $\redonesym$ is essentially
+standard call-by-value evaluation, and gathers the subterms' future environments
+into a single one.
+
+\crem{Can we cut \ref{fig:diaSemanticsSpec} from the paper? Perhaps just put a
+few representative rules into \ref{fig:diaSemantics}?}
 
 At $\next~e$, in contrast, $\redonesym$ speculates into $e$ for stage~\bbone\ 
-subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
-This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
+subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
+which takes a stage \bbtwo\ term to a residual. 
 For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into

### 2014-10-08 paper/relatedwork.tex
wrote stuff in related work

@@
 
-\subsection{Davies}
+\subsection {Staged Computation}
+Partial evaluation is a well-studied technique for specializing programs to some of their input.  
+The usual phrasing of the partial evaluation problem implicitly includes binding time analysis,
+but many presentations start with a stage-annotated representation, as we have done.
+We've shown that our dynamic semantics can be used to perform partial evaluation 
+[need to do more reading to know exactly how it compares].
 
-\cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
-and showed the equivalence between that type system and existing one for binding time analysis. 
-That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
-That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
-In fact, if we change the $\next$ and $\prev$ rules of ours to 
-\begin{mathpar}
-\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
-\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
-\end{mathpar}
-and treat $\next~q$ as a partial value for any residual $q$,
-then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
-environment bookkeeping in $\redonesym$, by inlining residuals instead of
-hoisting them in \verb|let|-bindings.
-From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-{\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
-
-\subsection{Multi-stage (meta) programming}
-{\bf Umut: jotting down notes for now.}
-
-Multi-stage, a.k.a., meta programming or staged programming, gives the
-programmer linguistic facilities consisting of language constructs
-(usually also a type system, including a dynamic semantics) for
-expressing programs where inputs may come in stages.  Earlier work on
-staged programming was motivated by understanding the effects of
-binding-time analysis that partial evaluation relied on and allowing
-control over staging so that the programmer can control the
-termination and efficiency characteristics of staged programs.
-Several different approaches to staged programming has been proposed.
+One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
+and so they must necessarily unfold recursive calls.
+This potentially explodes the size of a residual, which can be a problem.
+Stage-splitting, in contrast, operates on terms which are open on first stage variables,
+and so this problem is avoided.
 
@@
 
-\subsection{Specialization}
+Multi-stage, a.k.a., meta programming or staged programming, gives the
+programmer linguistic facilities consisting of language constructs
+(usually also a type system, including a dynamic semantics) for
+expressing programs where inputs may come in stages.  Earlier work on
+staged programming was motivated by understanding the effects of
+binding-time analysis that partial evaluation relied on and allowing
+control over staging so that the programmer can control the
+termination and efficiency characteristics of staged programs.
+Several different approaches to staged programming has been proposed.
 
-Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.
-They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.
-They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.
-The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.
-Accordingly, their approach was entirely informal, with no implementation realized.
-Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."
+\subsection{Stage Splitting}
 
-Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.
-The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.
-Their shaders are written in a C-like language with basic arithmetic and if statements.
-Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.
+Stage-splitting itself has appeared in the literature---under 
+the names {\em pass separation} and {\em data specialization}--- for various simpler input languages.
 
-\subsection{Graphics Systems}
+Most often, splitting is considered a compiler optimization (\cite{jorring86},\cite{knoblock96})
+to minimize recomputation.
+In this use case, only the arguments to the program are annotated with stages, 
+and the splitter is left to perform an {\em ad hoc} binding time analysis to determine the 
+internal staging of the program.
+Because this occurs without programmer supervision, the binding time analyzers for these uses are 
+reluctant to place stage \bbone\ code within stage \bbtwo\ conditionals, 
+since that would result in speculative behavior that could increase the overall runtime.
+We don't worry about this since the annotations of \lang\ are assumed to represent the will of a programmer.
 
 Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  
-But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.
-Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.
-Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.
-Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].
+But instead of using a binding-time analysis to determine what parts of the program to put in what shader, 
+Spark gives the programmer manual control of term staging using a language that is in spirit similar to \lang.
 
-[RTSL and SH]
+We are also not the first to use stage splitting to derive algorithms.
+\cite{malmkjaer89}, expanding on the work of \cite{barzdins88}, defined a stage-splitter
+for a language of tail-recursive first-order equations with explicit staging annotations [double check that],
+and considered applications to string matching.
+
+Our work is distinct from all previous stage splitters in the following ways:
+\begin{itemize}
+\item Our input language, \lang, has a disciplined type theory.
+\item \lang\ has full first-class functions (and by extension, recursion), as well as sum types.
+\item Our stage splitting is proven correct.
+\end{itemize}
+
+\subsection{Davies}
+
+\cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
+and showed the equivalence between that type system and existing one for binding time analysis. 
+That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
+That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
+In fact, if we change the $\next$ and $\prev$ rules of ours to 
+\begin{mathpar}
+\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
+\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
+\end{mathpar}
+and treat $\next~q$ as a partial value for any residual $q$,
+then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
+environment bookkeeping in $\redonesym$, by inlining residuals instead of
+hoisting them in \verb|let|-bindings.
+From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
+{\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

### 2014-10-08 paper/relatedwork.tex
more notes on metaprogramming

@@
 
-Davies ...
-
-Taha's approach is similar to Davies, but extends Davies's approach
-with ``cross-stage persistency'' which allows a variable to be used in
-any later stage and a ``run'' construct that allows evaluating an
-expression.  Our language constructs therefore are also close to
-staged programming such as proposed by Taha.  In fact our static
-semantics is essentially identical to Taha's.  Though, as Davies, we
-don't include ``run'' in the language.  The dynamic semantics of
-staged languages... COMPLETE
-
-The key difference between our work and Taha's and similar work on
-staging is our splitting algorithm that can translate the staged code
-to a different code that can run significantly (sometimes
-asymptotically) faster. COMPARE DYNAMIC SEMANTICS
-
- Unlike the evaluati
-
-Multi-stage, a.k.a., meta programming or staged programming, gives the
-programmer linguistic facilities consisting of language constructs
-(usually also a type system, including a dynamic semantics) for
-expressing programs where inputs may come in stages.  Earlier work on
-staged programming was motivated by understanding the effects of
-binding-time analysis that partial evaluation relied on and allowing
-control over staging so that the programmer can control the
-termination and efficiency characteristics of staged programs.
-Several different approaches to staged programming has been proposed.
+There's another staging technique known as {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}).
+The differences between it and partial evaluation can be subtle.
+The core idea of metaprogramming is to interpret stage \bbtwo\ values as {\em code} in some object language
+which stage \bbone\ can manipulate.
+This has three implications:
+\begin{itemize}
+\item Stage annotations are now mandatory, rather than optional.
+\item The $\next$-contents-duplicating semantics is now intended behavior [this needs to be verified].
+This changes the cost of terms from what they would be in \lang's semantics, 
+but that's merely something for the programmer to keep in mind.
+\item In metaprogramming systems, it now makes sense to define an operator, \texttt{run}, 
+which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
+When you think about it, this is a form a backwards information flow.
+For this reason, the \texttt{run} feature is in direct conflict with stage-splitting,
+which by its definition cannot work for a language that allows backwards control flow.
+\end{itemize}
 

### 2014-10-09 paper/relatedwork.tex
related work/termination

@@
 {\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
+
+\ur{Umut: I am starting a paragraph on termination.  we might need to expand
+  on this.}
+\paragraph{Termination.}
+
+There are two reasons for non-termination: infinite loops and
+execution of errorneous code.  The splitting algorithm can cause the
+first.  But can it also cause the second? (If all computations really
+depend on stage 1 code, then would it be possible to for example see
+something like 1/0?  i think so but only if there is a meta dependency
+between the data in two stages?)
+
+Termination analysis in partial evaluation: see paper~\cite{AH96,GJ05}.

### 2014-10-09 paper/relatedwork.tex
an iteration on related work

@@
 
-\subsection {Staged Computation}
 Partial evaluation is a well-studied technique for specializing programs to some of their input.  
@@
 One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
-and so they must necessarily unfold recursive calls.
-This potentially explodes the size of a residual, which can be a problem.
-Stage-splitting, in contrast, operates on terms which are open on first stage variables,
-and so this problem is avoided.
-
-There's another staging technique known as {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}).
-The differences between it and partial evaluation can be subtle.
-The core idea of metaprogramming is to interpret stage \bbtwo\ values as {\em code} in some object language
-which stage \bbone\ can manipulate.
-This has three implications:
-\begin{itemize}
-\item Stage annotations are now mandatory, rather than optional.
-\item The $\next$-contents-duplicating semantics is now intended behavior [this needs to be verified].
-This changes the cost of terms from what they would be in \lang's semantics, 
-but that's merely something for the programmer to keep in mind.
-\item In metaprogramming systems, it now makes sense to define an operator, \texttt{run}, 
-which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
-When you think about it, this is a form a backwards information flow.
-For this reason, the \texttt{run} feature is in direct conflict with stage-splitting,
-which by its definition cannot work for a language that allows backwards control flow.
-\end{itemize}
-
-\subsection{Stage Splitting}
+and so they must necessarily unfold recursive calls and wait for all stage \bbone\ input to operate.
+There has been some work to avoid these issues by templating [need citations].
+Stage-splitting, in contrast, goes all the way to operating on terms which are open on first stage variables,
+and so these problems are avoided.
 
@@
 
-\subsection{Davies}
+The goal of {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}) is 
+to use code in a host language to write code in an object language.
+Like partial evaluation and stage-splitting, this can be modeled by a staged language with explicit annotations.
+Instead of identifying stages with points in time, stages are identified with host/object levels of the language,
+and so type safety ensures the proper separation of levels of the language and safety of the constructed object code.
+
+But the differences in goals of metaprogramming do cause a few differences in form.
+For example, it's generally preferred in the metaprogramming context to use 
+a dynamics that duplicates the contents of $\next$ blocks.
+Moreover, in metaprogramming systems it makes sense to define an operator, \texttt{run}, 
+which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
+This is essentially a form a backwards information flow.
+Thus, the \texttt{run} feature is directly incompatible with a temporal interpretation of stages,
+which stage-splitting requires.
 

### 2014-10-10 paper/relatedwork.tex
relatedwork

@@
 
+\subsection{Umut's notes}
+
+---- DO NOT DELETE ----
+
+Answer the following question for all of the related work:
+
+What happens when we try to write quickselect as a meta program?
+
+\paragraph{Partial Evaluation.}
+Online and offline.  Both approaches have their merits but online
+partial evaluation can be too slow and if stopped lead to an
+inefficient residual.  Offline partial evaluation is faster.  Also,
+full self-application has only been achieved by offline partial
+evaluation.
+
+Weaknesses of partial evaluation include~\cite{GJ05}:
+1) speedups linear in the subject programs run time (excuse moi?)
+
+2) good speedups require close knowledge of the program and some
+binding-time improvements.
+
+3) result of specialization can be hard to predict as it can lead to
+slowdown, code explosion, and non-termination at specialization time.
+
+\paragraph{Meta programming}
+
+What happens when we try to write quickselect as a meta program?
+
 \ur{Umut: I am starting a paragraph on termination.  we might need to expand
@@
 
+Is in possible for the residual program to not terminate (unexpectedly)?
+
 Termination analysis in partial evaluation: see paper~\cite{AH96,GJ05}.
+
+Based on Glenstrup and Jones (pp. 1154, 1155), a partial evaluator
+should be complete. That is every computation depending on the static
+input should be reduced to a value (during specialization).
+Unfortunately, completeness is difficult to achieve if we also require
+that the specialization is always terminating.
+
+One problem with their description is that it is not clear which
+termination problem they are talking about termination of the residual
+or the specializer.
+
+Nontermination during specialization: 
+consider code
+
+double x = doubleplus (x,0)
+
+doubleplus (u,v) = if u <= 0 then v else doubleplus (u-1,v+2)
+
+this can lead to non-termination if v is considered static.
+
+Is this a problem for us?
+
+\paragraph{Supercompilation.}
+~\cite{Turchin86}

### 2014-10-10 paper/relatedwork.tex
related work...

@@
 ~\cite{Turchin86}
+
+\paragraph{Intro.}
+
+A predominating theme in the design, implementatation, and
+optimization of algorithm and software is frequency reduction and
+precomputation.  The idea is to reduce the frequency of computations
+by reducing their frequency and ideally by precomputing when possible.
+
+The reason for why this theme is as predominant as it is that it is
+naturally common in many actual applications.  For example, we may
+have a collection of names or numbers that we frequently lookup, as
+for example part of a membership test as part of a larger system.  In
+some applications areas such as in graphics (the motivation for our
+work), this is so common that there is a relatively large literature
+on speeding up graphics computations by taking advantage of frequency
+reduction and precomputation.
+
+One way to express frequency of computations is to allow the
+programmer to stage computation explicitly by making explicit the
+stage of each computation and then by moving computations between
+stages to improve efficiency by performing work as early (to save time
+later when performance might be more critical) and  as infrequently
+as possible (to reduce overall time).
+
+For example, in meta-programming~\cite{}, the programmer can assign a
+program expression a stage (stage 1, 2, 3, ...), which indicates the
+stage at which that expression evaluates.  Combined with the ability
+for earlier stages to construct expressions of later stages,
+meta-programming techniques can be used to construct efficient
+programs that improve efficiency over their non-staged counterparts by
+reducing frequency reduction and precomputation.  For example, the
+programmer can write a program that takes a list of numbers $\ell$ as
+a first stage value and returns a second-stage function that performs
+a fast membership test on that list $\ell$ by inlining all the
+comparisons.  When given a key to lookup in the second stage, the
+membership test function would quickly return the result.

### 2014-10-11 paper/relatedwork.tex
added a quick preamble to the related work

@@
 \section{Related Work}
+There are several distinctive aspect of our work which we wish to compare and contrast.
+Firstly, \lang\ is a simply typed lambda calculus 
+with explicit stage annotations and a non-duplicating semantics.
+Its statics and dynamics enforce no backward information flow,
+which a is necessary property to define a splitting transformation
+that is equivalent to the dynamics.
+Finally, the splitting transformation itself operates on any well-typed open term in \lang.
 
 Partial evaluation is a well-studied technique for specializing programs to some of their input.  
-The usual phrasing of the partial evaluation problem implicitly includes binding time analysis,
-but many presentations start with a stage-annotated representation, as we have done.
+Often, the goal of partial evaluation systems is to go all the way from unstaged code to a specialized version.
+Although this problem definition implicitly includes binding time analysis,
+many presentations factor out that step and start with a stage-annotated representation, as we have done.
 We've shown that our dynamic semantics can be used to perform partial evaluation 

### 2014-10-13 paper/splitting.tex
quick incremental improvement to splitting rules descriptions

@@
 
-\ur{Explain the rules.  This is the most important part of the
-  paper. Work needed here.}
-Every $\splittwosym$ rule works by bundling the precomputations of the constituent parts, and
-then unbundling them with a pattern.
-The rules for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
+The $\splittwosym$ rules are conceptually simple and almost all very similar.  
+As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
+from contained stage \bbone\ code into a single expression $p$, 
+leaving a residual $l.r$ to interpret the results of that precomputation.
+Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
+that is, $p$ is a unit value and $l$ is never used in $r$.
+At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
+the precomputations of the constituent parts, and then unbundling them with a pattern,
+with the residual precisely resembling the original expression.
+
 The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
-while the rule for $\prev$ projects the combined result of its subexpression to just the precomputation.
+while the rule for $\prev$ projects the combined result of its subexpression to isolate the precomputation.
+It may seem like this $\prev$ rule is throwing information away, but we well show later that the first component 
+of the evaluated form of $c$ in the $\prev$ rule must always be unit.
 The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 

### 2014-10-14 paper/example.tex
first rough cut at a spelled-out splitting example

@@
 Instead, we assume that all input programs are properly staged according to some programmer's intent.
+
+\subsection{Splitting of Quickselect}
+
+Here, we will manually split apart the \texttt{qSelect} function in order to give some intuition about how the operation works.
+
+We start with the basic \texttt{qSelect}:
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list) (k : $`2`int`1`) : $`2`int`1` = 
+  case l of
+    Empty => next {`2`0`1`}
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      next{
+        `2`let n = hold{`1`n`2`} in
+        case compare prev{`1`k`2`} n of
+          LT => prev {`1`qsStaged left k`2`}
+        | EQ => hold {`1`h`2`}
+        | GT => prev {`1`qsStaged right 
+                             next{prev{k}-n-1)}`2`}`1`
+      }`
+\end{lstlisting}
+
+Again our goal is to split this into \texttt{qSelect1} and \texttt{qSelect2} such that \texttt{qStaged l i = qSelect2 (qSelect1 l) i}.  
+The first step in this process is to assume this conclusion at the recursive sites:
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list) (k : $`2`int`1`) : $`2`int`1` = 
+  case l of
+    Empty => next {`2`0`1`}
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      next{
+        `2`let n = hold{`1`n`2`} in
+        case compare prev{`1`k`2`} n of
+          LT => qSelect2 (`1`qSelect1 left`2`) k
+        | EQ => hold {`1`h`2`}
+        | GT => qSelect2 (`1`qSelect1 right`2`) 
+                                      (k-n-1)`1`
+      }`
+\end{lstlisting}
+
+We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list) (k : $`2`int`1`) : $`2`int`1` = 
+  case l of
+    Empty => next {`2`0`1`}
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      next{
+        `2`let n = hold{`1`n`2`} in
+        let p = `1`(h,qSelect1 left,qSelect1 right)`2`
+        case compare k n of
+          LT => qSelect2 (pi2 p) k
+        | EQ => pi1 p
+        | GT => qSelect2 (pi3 p) (k-n-1)`1`
+      }`
+\end{lstlisting}
+
+We continue to pull up the precomputation and include more in it:
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list) (k : $`2`int`1`) : $`2`int`1` = 
+  case l of
+    Empty => next {`2`0`1`}
+  | Cons (h,t) => 
+      let p =
+        let (left,right,n) = partition h t in
+        `1`(n,h,qSelect1 left,qSelect1 right)
+      in`2`
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+This completes the splitting of the cons branch.  
+The empty branch can also be split, though in a trivial way:
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list) (k : $`2`int`1`) : $`2`int`1` = 
+  case l of
+    Empty => let p = () in `2`0`1`
+  | Cons (h,t) => 
+      let p =
+        let (left,right,n) = partition h t in
+        `1`(n,h,qSelect1 left,qSelect1 right)
+      in`2`
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+To continue splitting the case statement, we need to us a sum type in the precomputation
+to remind the code what branch its in.
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list) (k : $`2`int`1`) : $`2`int`1` = 
+  let p = 
+    case l of
+      Empty => Top ()
+    | Cons (h,t) =>
+        let (left,right,n) = partition h t in`1`
+        Bot (n,h,qSelect1 left,qSelect1 right)
+  in`2`
+  case p of 
+    Top () => 0
+  | Bot p => 
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+We finish where we start, splitting \texttt{qsStaged} into
+\texttt{qSelect1} and \texttt{qSelect2}.  
+Omitting a few type annotations, that is:
+
+\begin{lstlisting} 
+1`fun qSelect1 (l : list) = 
+    case l of
+      Empty => Top ()
+    | Cons (h,t) =>
+        let (left,right,n) = partition h t in
+        Bot (n,h,qSelect1 left,qSelect1 right)
+`2`
+fun qSelect2 p (k : int) : int = 
+  case p of 
+    Top () => 0
+  | Bot p => 
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}

### 2014-10-14 paper/example.tex
removed staging annotations from the informal example

--- 2014-10-01 paper/example.tex
@@
 Instead, we assume that all input programs are properly staged according to some programmer's intent.
+
+\subsection{Splitting of Quickselect}
+
+Here, we will manually split apart the \texttt{qSelect} function in order to give some intuition about how the operation works.
+
+We start with the basic \texttt{qSelect}:
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ... `
+
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in`2`
+      case compare k n of
+        LT => `qSelect 1`left `2`k
+      | EQ => h
+      | GT => `qSelect 1`right `2`(k-n-1)`
+\end{lstlisting}
+
+Again our goal is to split this into \texttt{qSelect1} and \texttt{qSelect2} such that \texttt{qStaged l i = qSelect2 (qSelect1 l) i}.  
+The first step in this process is to assume this conclusion at the recursive sites:
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in`2`
+      case compare k n of
+        LT => qSelect2 (`1`qSelect1 left`2`) k
+      | EQ => h
+      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
+\end{lstlisting}
+
+We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
+      case compare k n of
+        LT => qSelect2 (pi2 p) k
+      | EQ => pi1 p
+      | GT => qSelect2 (pi3 p) (k-n-1)`
+\end{lstlisting}
+
+We continue to pull up the precomputation and include more in it:
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let p =
+        let (left,right,n) = partition h t in
+        `1`(n,h,qSelect1 left,qSelect1 right)
+      in`2`
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+This completes the splitting of the cons branch.  
+The empty branch can also be split, though in a trivial way:
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => let p = () in `2`0`1`
+  | Cons (h,t) => 
+      let p =
+        let (left,right,n) = partition h t in
+        `1`(n,h,qSelect1 left,qSelect1 right)
+      in`2`
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+To continue splitting the case statement, we need to us a sum type in the precomputation
+to remind the code what branch its in.
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  let p = 
+    case l of
+      Empty => Top ()
+    | Cons (h,t) =>
+        let (left,right,n) = partition h t in`1`
+        Bot (n,h,qSelect1 left,qSelect1 right)
+  in`2`
+  case p of 
+    Top () => 0
+  | Bot p => 
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+We finish where we start, splitting \texttt{qSelect} into
+\texttt{qSelect1} and \texttt{qSelect2}.  
+Omitting a few type annotations, that is:
+
+\begin{lstlisting} 
+1`fun qSelect1 (l : list) = 
+    case l of
+      Empty => Top ()
+    | Cons (h,t) =>
+        let (left,right,n) = partition h t in
+        Bot (n,h,qSelect1 left,qSelect1 right)
+`2`
+fun qSelect2 p (k : int) : int = 
+  case p of 
+    Top () => 0
+  | Bot p => 
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}

### 2014-10-14 paper/example.tex
worked the splitting example into the intro

@@
 
-One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
-\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.
-Therefore, if quickselect were to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in \ref{fig:qssplit}),
-and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).  
-Equationally, we can represent this relationship as 
-\[
-\mathtt{qSelect(l,k)} \equiv \mathtt{qSelect2(qSelect1(l),k)}
-\]
-
-The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
-the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
-whereas \texttt{qSelect2} takes $O(\log n)$ time.
-Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
-%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
-%and then used to accelerate many future lookups for any value of $k$.
-
-This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
-\lang\ unambiguously specifies how its terms should be split
-% and features a type system that ensures valid splittings exist.
-% only minimal refactoring should be required to write terms within this language,
-%and from there, the type system should be enough to prove that a valid splitting exists.
-by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
-Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.  
-After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
-will end up in the stage \bbone\ function ({\tt qSelect1}), 
-and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
-Typing rules for \lang\ ensure a valid splitting exists.  That is, 
-within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
-but never from \bbtwo\ to \bbone.
-
-\ref{fig:qsstaged} shows a staged implementation of quickselect in \lang.
-As before, the function accepts as input the list to select from and a value of $k$,
-but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
-The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
-whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
-Naturally the output type, representing the $k$th largest element of the list,
-is also $\fut\rmint$, since it is not computed until stage \bbtwo.
-
-Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
-Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
-whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
-(We adopt the convention that the top-level context is stage \bbone.)
-The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
-For example, \verb|next{0}| from above has the type $\fut\rmint$.
-Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
-For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
-These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
-which is necessary to admit a properly staged implementation.
-
-The code also contains two $\pause$ blocks.  
-This construct has same stage signature as $\prev$,
-but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage~\bbone\ to stage~\bbtwo.
-\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
-We instead provide it as a primitive to shorten examples.  
-Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
-This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
-
-But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
-the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
-The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
-It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
-but that would not be particularly interesting since it would result in an effectively trivial split 
-that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
-There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
-This process is known as {\em binding time analysis}, and we do not consider it here.
-Instead, we assume that all input programs are properly staged according to some programmer's intent.
-
-\subsection{Splitting of Quickselect}
-
-Here, we will manually split apart the \texttt{qSelect} function in order to give some intuition about how the operation works.
-
-We start with the basic \texttt{qSelect}:
+One notable property of quickselect is that the partitioning of the list, 
+even through its recursive calls, depends only on the values of list elements, and not on the value of $k$.
+To illustrate this, we present colorized version of \texttt{qSelect}:
 
@@
 
-Again our goal is to split this into \texttt{qSelect1} and \texttt{qSelect2} such that \texttt{qStaged l i = qSelect2 (qSelect1 l) i}.  
-The first step in this process is to assume this conclusion at the recursive sites:
+Here, the input $k$ and all parts of the computation that depend on it are colored blue,
+whereas all of the parts of the computation that do not depend on $k$ are colored red.
+The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.
 
+To be precise, our coloring strategy respects data dependence, but not control dependence---the 
+first arguments to the recursive calls to \texttt{qSelect} are colored red,
+despite appearing underneath a blue case statement.
+Because of this coloring strategy and the definition of data dependence, 
+information can flow from red code to blue code but not vice versa.
+
+Observe the that helper function \texttt{partition} is colored entirely red.
+This is where much of the work of the quickselect algorithm takes place, and it doesn't even depend on $k$!
+
+This suggests an optimization: we can split \texttt{qSelect} into two functions, 
+one for all of the red code and another for all of the blue code. 
+We'll call them \texttt{qSelect1} and \texttt{qSelect2}.
+Because of the dependence properties stated above, 
+\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
+whereas \texttt{qSelect1} will only depend on the input list $l$.
+Of course, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
+Equationally, we can represent this relationship as simply 
+\[
+\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
+\]
+
+With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from \texttt{qSelect}.
+Because the original function is recursive,
+the first step in this process is to assume this conclusion at the call sites.
+This is performed in place:
 \begin{lstlisting} 
@@
 \end{lstlisting}
-
 We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
-
 \begin{lstlisting} 
@@
 \end{lstlisting}
-
+This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
 We continue to pull up the precomputation and include more in it:
@@
 \end{lstlisting}
+
+The final version is shown in figure...
+Note that it's basically a tree algorithm...
+
+The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
+the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
+whereas \texttt{qSelect2} takes $O(\log n)$ time.
+Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
+%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
+%and then used to accelerate many future lookups for any value of $k$.
+
+This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
+\lang\ unambiguously specifies how its terms should be split
+% and features a type system that ensures valid splittings exist.
+% only minimal refactoring should be required to write terms within this language,
+%and from there, the type system should be enough to prove that a valid splitting exists.
+by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
+Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
+subexpressions are evaluated before stage-\bbtwo\ ones.  
+After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
+will end up in the stage \bbone\ function ({\tt qSelect1}), 
+and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
+Typing rules for \lang\ ensure a valid splitting exists.  That is, 
+within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
+but never from \bbtwo\ to \bbone.
+
+\ref{fig:qsstaged} shows a staged implementation of quickselect in \lang.
+As before, the function accepts as input the list to select from and a value of $k$,
+but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
+The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
+whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
+Naturally the output type, representing the $k$th largest element of the list,
+is also $\fut\rmint$, since it is not computed until stage \bbtwo.
+
+Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
+Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
+whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
+(We adopt the convention that the top-level context is stage \bbone.)
+The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
+For example, \verb|next{0}| from above has the type $\fut\rmint$.
+Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
+For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
+These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
+which is necessary to admit a properly staged implementation.
+
+The code also contains two $\pause$ blocks.  
+This construct has same stage signature as $\prev$,
+but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage~\bbone\ to stage~\bbtwo.
+\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
+We instead provide it as a primitive to shorten examples.  
+Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
+This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
+
+But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
+the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
+The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
+It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
+but that would not be particularly interesting since it would result in an effectively trivial split 
+that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
+There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
+This process is known as {\em binding time analysis}, and we do not consider it here.
+Instead, we assume that all input programs are properly staged according to some programmer's intent.

### 2014-10-15 paper/example.tex
added a few more notes about the example

@@
 
-To continue splitting the case statement, we need to us a sum type in the precomputation
+To continue splitting the case statement, we need to generate a sum type in the precomputation
 to remind the code what branch its in.
@@
 
-The final version is shown in figure...
-Note that it's basically a tree algorithm...
-
+The final version, cleaned up and given better datatype names, is shown in \ref{fig:qssplit}.
+Observe that \texttt{qSelect1} essentially builds a binary search tree from the input list, 
+and \texttt{qSelect2} searches that tree for the correct index.
 The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
@@
 Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
-%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
-%and then used to accelerate many future lookups for any value of $k$.
 
-This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.
+\nr{There's something else I need to say right here before the contribution statement below. 
+I want the reader to be thinking "yeah, this splitting business does seem cool."
+Bonus points if I also have them thinking "but it would be cooler if it were formal."}
+
+This paper formally describes and analyzes the splitting transformation exemplified above.
+(I also want to make it clear that recognizing \lang's appropriateness for this is itself is a contribution.)
+
+We start by defining a simple staged language, called \lang, to express the input to our splitting algorithm.  
 \lang\ unambiguously specifies how its terms should be split

### 2014-10-16 paper/example.tex
clarifications to section two, and a new word for splitting residuals

@@
 
-To be precise, our coloring strategy respects data dependence, but not control dependence---the 
+To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
 first arguments to the recursive calls to \texttt{qSelect} are colored red,
 despite appearing underneath a blue case statement.
-Because of this coloring strategy and the definition of data dependence, 
-information can flow from red code to blue code but not vice versa.
+This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.
 
-Observe the that helper function \texttt{partition} is colored entirely red.
-This is where much of the work of the quickselect algorithm takes place, and it doesn't even depend on $k$!
+Observe the that helper function \texttt{partition} is colored entirely red,
+which is valid because it is called with only red arguments.  
+This is an illustration of the property stated above, that 
+the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
+Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.
 
 This suggests an optimization: we can split \texttt{qSelect} into two functions, 
-one for all of the red code and another for all of the blue code. 
-We'll call them \texttt{qSelect1} and \texttt{qSelect2}.
+one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
 Because of the dependence properties stated above, 
@@
 whereas \texttt{qSelect1} will only depend on the input list $l$.
-Of course, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
+Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
 Equationally, we can represent this relationship as simply 
@@
 \]
+for some list \texttt{l} and index \texttt{k}.
 
-With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from \texttt{qSelect}.
+With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
 Because the original function is recursive,
-the first step in this process is to assume this conclusion at the call sites.
+the first step in this process is to assume the conclusion at the call sites.
 This is performed in place:

### 2014-10-16 paper/splitting.tex
clarifications to section two, and a new word for splitting residuals

@@
 
-\ur{The reuse of residual is bothersome, we should use different terminology,
-1-value and 2-value perhaps.}
-
 Splitting arbitrary terms in \lang\ again results in two terms containing only
 the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
-called, respectively, the {\em precomputation} and {\em residual}.%
-\footnote{We also used the term {\em residual} for the result of
-partial evaluation.  Where necessary, we will distinguish between
-these concepts by using the terms {\em splitting-residual} and {\em
-evaluation-residual}.}
+called, respectively, the {\em precomputation} and {\em resumer}.
 %
@@
 \end{lstlisting}
-splits into a precomputation \verb|(1+2,3+4)| and a residual
-\verb|fn (L1,L2) => L1 < L2-5|, with the property that the residual applied to
+splits into a precomputation \verb|(1+2,3+4)| and a resumer
+\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
 the precomputation must agree with the original term. This process of hoisting
@@
 The precomputation for this term is a \emph{combined result} which is
-the pair of these components: \verb|(((),3*4),1+2)|. The splitting residual is
+the pair of these components: \verb|(((),3*4),1+2)|. The resumer is
 likewise the pair of the stage \bbtwo\ results of lifting and masking:
@@
 The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
-$c$ and residual $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
-precomputation $p$ and a residual $l.r$. In both cases, the term $e$ is open on
+$c$ and resumer $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
+precomputation $p$ and a resumer $l.r$. In both cases, the term $e$ is open on
 variables in the context $\Gamma$. We write the rules using pattern notation
 when appropriate.
-%For concision, we represent residuals as a term open on a single variable,
+%For concision, we represent resumers as a term open on a single variable,
 %rather than as functions.
@@
 from contained stage \bbone\ code into a single expression $p$, 
-leaving a residual $l.r$ to interpret the results of that precomputation.
+leaving a resumer $l.r$ to interpret the results of that precomputation.
 Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
@@
 the precomputations of the constituent parts, and then unbundling them with a pattern,
-with the residual precisely resembling the original expression.
+with the resumer precisely resembling the original expression.
 
@@
 The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
-and projects out the integer result in the residual.%
-\footnote{The residual of an integer expression is usually trivial, 
+and projects out the integer result in the resumer.%
+\footnote{The resumer of an integer expression is usually trivial, 
 but we have to include it here for termination purposes.}
@@
 taken branch only, and inject the precomputation into a sum type. In the
-residual, we case on that injection, and resume the stage \bbtwo\ portion of the
+resumer, we case on that injection, and resume the stage \bbtwo\ portion of the
 correct branch.
@@
 result which handles all the stage \bbone\ content of the original, and one in
-the residual which handles all of the stage \bbtwo\ content. Note that stage
+the resumer which handles all of the stage \bbtwo\ content. Note that stage
 \bbone\ lambdas are partial values, and so have only a trivial precomputation.
@@
 Thus, we start with two mutually dependent definitions of equivalence.  
-Both relate evaluation-residuals on the left with splitting residuals on the right,
+Both relate residuals on the left with resumers on the right,
 but $\equiv$ equates closed expressions, whereas $\sim$ equates values.
@@
 \begin{definition}
-For evaluation-residual $q$ and splitting-residual $r$, define $q \equiv r : A$ to mean that 
+For residual $q$ and resumer $r$, define $q \equiv r : A$ to mean that 
 $q \tworedsym v_q$ iff $\reduce {r} {v_r}$ where $v_q \sim v_r : A$, 
@@
 \begin{definition}
-For evaluation-residual value $v_1$ and splitting-residual value $v_2$, define $v_1 \sim v_2 : A$ by the following cases:
+For residual value $v_1$ and resumer value $v_2$, define $v_1 \sim v_2 : A$ by the following cases:
 \begin{itemize}

### 2014-10-17 paper/umut.tex
taking notes

\section{Umut's notes}

...To be added to the paper...

Answer the following question for all of the related work:

What happens when we try to write quickselect as a meta program?

\newcommand{\drun}[2]{\lVert{#2}\rVert_{#1}}
\paragraph{Partial Evaluation.}

The following definitions are adapted from the partial evaluation
book.  For simplicity and uniformity, we assume that data and programs
are draws from the same set and that ill-behaved programs and
non-terminating programs all return $bottom$.

Definition[Residual].

Let $p \in P$ be a program in a language $P$, which takes two inputs
$d_s \in P$ (static data) and $d_d \in P$ (dynamic data).  The program
$r \in P$ is a {\em residual for $p$ with respect to $d_s$} if for all
$d_d$, residual behaves that same as the $p$ for static data $d_s$,
i.e., 
\[
\drun{P}{p} (d_s, d_d) = \drun{P}{r} (d_d).
\]

Definition[Partial Evaluator]
A {\em partial evaluator} or a {\em program specializer} is a program
$e \in P$ (for ``evaluate'') such that for every $p \in P$, and every
$d_s, d_d \in P$, 
\[
\drun{P}{p} (d_s, d_d) = \drun{P}{\drun{P}{e}(p, d_s)} (d_d).
\]

Foundations of partial evaluation go back to Kleene's s-n-m theorem
from 1952, which constructively proved the existence of partial
evaluators by using Turing machines~\cite{Kleene52}.  Kleene did not
intend to improve efficiency, however. The first use of the term
``partial evaluation'' appears in Lombardi and Raphael (1964), in the
context of a paper on incremental camputation. Futamura (1971)
considered self-application of partial evaluation, also using it for
the purpose of compiler generation.  Ershov enriched Futamura's work
by considering double self-application, which led to compiler
generator generators, calling also Futamura's equations between
partially evaluated programs as ``Futamura projections.''

Ershov also introduced a richer form of transformation, {\em mixed
  computation} that generalizes partial evaluation.  While in some
literature, mixed computation is characterized as similar to partial
evaluation, it is significantly more general, at least by definition.

Definition [Mixed computation]. A mixed-computation for a language $P$
consist of two algorithms $m_p, m_d \in P$ for specializing the data
and the program such that for any program $p \in P$ and any data $d
\in P$ the specialized program run with the specialized data behaves
the same as the original program run with the original data, i.e.,

\[
\drun{P}{p} (d) = \drun{P}{\drun{P}{m_p} (p,d)} (\drun{P}{m_d} (p,d)).
\]

Partial evaluation is trivially a special case of mixed computation,
because it allows a transformation of the program only on the static
data, and because it allows only a simple transformation on data (the
extraction of the dynamic data).

### 2014-10-17 paper/example.tex
remarks inserted

@@
 \label{sec:staging}
+
+\ur{Indentation seems broken. Be consistent.}
 
@@
 
+\ur{I find it hard to distinguish between blue and black on printed paper.}
+
 One notable property of quickselect is that the partitioning of the list, 
@@
 The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.
+
+\ur{I find control dependence issue very confusing.  Let's factor this
+out and talk about it in the end. }
 
@@
 
+\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}
+
 With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
@@
 the first step in this process is to assume the conclusion at the call sites.
+
+\ur{quickselect 1 and 2: how did we introduce them?}
+
 This is performed in place:
@@
 \end{lstlisting}
-We then pull the stage \bbone\ portion of each branch of the comparison up into a single tuple.
+We then pull the stage \bbone\ portion of each branch of the
+comparison up into a single tuple.
+\ur{Why did we include h}
+
 \begin{lstlisting} 
@@
 \end{lstlisting}
+
+\ur{Don't know what ``this last step'' refers to.}
 This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
@@
 
+\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
 \begin{lstlisting} 

### 2014-10-17 paper/abstract.tex
Wrote an abstract with Nico.

Many algorithms can perform useful work before receiving all of their inputs.
If we regard those inputs as arriving at different \emph{stages} of the
computation, then we can \emph{split} those algorithms into, for each stage, a
function performing all the work dependent only on the available inputs.

In this paper, we provide a theoretical understanding of this splitting process
by defining it as a program transformation on \lang, a typed lambda calculus
equipped with a modal staging operator. This approach extends splitting to
language features absent in prior work, including first-order functions and
disjoint unions.

These new features allow us to express some familiar algorithms which, when
split, yield asymptotic improvements; for example, quickselect of the $k$th
smallest element of a list $l$ splits into (1) sorting $l$ into a binary search
tree, then (2) finding the tree's $k$th leftmost element.

\TODO (Is this true?)
We also show that splitting a staged implementation of an interpreter yields a
partial evaluator.

%into one function per stage, performing all the work depending only on the
%available inputs.

%as much work as possible given the inputs available at each stage.

%In this way, splitting is essentially an algorithm generation technique.

% amortized

### 2014-10-17 paper/umut.tex

@@
 ...To be added to the paper...
+(TODO: make a pass to make consistent the terminology. it is a bit complex.)
 
@@
 \newcommand{\drun}[2]{\lVert{#2}\rVert_{#1}}
-\paragraph{Partial Evaluation.}
+
+\subsection{Mixed Computation, Partial Evaluation, and Data Specialization.}
 
 The following definitions are adapted from the partial evaluation
-book.  For simplicity and uniformity, we assume that data and programs
-are draws from the same set and that ill-behaved programs and
-non-terminating programs all return $bottom$.
+book, the cited papers, and Malmkjaer thesis.  For simplicity and
+uniformity, we assume that data and all programs are drawn from the
+same set and that ill-behaved programs and non-terminating programs
+all return $bottom$.
 
@@
 
-Ershov also introduced a richer form of transformation, {\em mixed
-  computation} that generalizes partial evaluation.  While in some
-literature, mixed computation is characterized as similar to partial
-evaluation, it is significantly more general, at least by definition.
+Data specialization (Barzdin and Bulyonkov 88) is a generalization of
+partial evaluation.
+
+Definition [Data Specialization].  A data specializer for a language
+$P$ consist of two algorithms $ds_p, ds_d \in P$ for specializing the
+data and the program such that for any program $p \in P$ and any
+static and dynamic data $d_s, d_d \in P$, the code-specialized program
+when run with static-data-specialized program behaves the same as the
+original program run with the original data, i.e.,
+
+\[
+\drun{P}{p} (d_s, d_d) = \drun{P}{\drun{P}{ds_p} (p)} (\drun{P}{ds_d}
+(p,d_s), d_d)).
+\]
+
+Seeing that partial evaluation is a special case of data
+specialization is not trivial, as it requires using an interpreter for
+the language $P$ (written in $P$).  Specifically, we can use a partial
+evaluator in place of data specializer $ds_d$ and an  interpreter for $P$ in
+place of the program specializer $ds_p$.
+
+Both partial evaluation and data specialization are special cases of
+mixed computation, which allows both the program and the data to be
+specialized with respect to each other.  Ershov introduced {\em mixed
+  computation}.
 
@@
 data, and because it allows only a simple transformation on data (the
-extraction of the dynamic data).
+extraction of the dynamic data).  Similarly data specialization is a
+special case of mixed computation because it allows the program to be
+specialized in the absence of data and only the static portion of the
+data to be specialized with respect to the program.
+
+\subsection{Techniques}
+
+Polyvariant.
+
+Call Unrolling.
+
+Elimination of duplicate work.

### 2014-10-20 paper/example.tex
working on example.tex

@@
 \label{sec:staging}
+
+\ur{I find it hard to distinguish between blue and black on printed paper.}
+
+\begin{figure*}
+\begin{subfigure}{0.5\textwidth}
+\begin{lstlisting} 
+datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) 
+  : (int*list*list) =
+  case unroll l of 
+    Empty => (0,Empty, Empty) 
+  | Cons (h,t) =>
+      let (s,left,right) = partition (p,t) in
+      if h<p 
+      then (s+1,Cons(h,left),right)
+      else (s,left,Cons(h,right))
+
+fun qSelect (l : list, k : int) : int = 
+  case l of
+    Empty => 0
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+        case compare k n of
+          LT => qSelect (left, k)
+        | EQ => h
+        | GT => qSelect (right, k-n-1)
+\end{lstlisting}
+\caption{Unstaged implementation of quickselect.}
+\label{fig:quickselect}
+\label{fig:qs-unstaged}
+\end{subfigure}%
+\begin{subfigure}{0.5\textwidth}
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...
+
+fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
+  case l of
+    Empty => next {`2`0`1`}
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      next{
+        `2`let n = hold{`1`n`2`} in
+          case compare prev{`1`k`2`} n of
+            LT => prev {`1`qsStaged (left k)`2`}
+          | EQ => hold {`1`h`2`}
+          | GT => prev {`1`qsStaged (right, 
+                             next{prev{k}-n-1)}`2`}`1`
+      }`
+\end{lstlisting}
+\caption{Staged implementation of quickselect in \lang.}
+\label{fig:qsstaged}
+\label{fig:qs-staged}
+\end{subfigure}
+\caption{Quickselect: traditional and staged.}
+\end{figure*}
+
+Staging is a technique for using language constructs and types to
+explicitly staged a computation according to the frequency or the
+arrival time of data.  As an example, suppose that we wish to issue
+order statistics queries on a collection of items represented as a
+list \texttt{l}.  For this tasks, we can use the expected linear time,
+quickselect algorithm~\cite{quickselect}, which takes a
+list~\texttt{l} and a rank~\texttt{k} and returns the element with
+rank~\texttt{k} in the list.  To this end, the algorithm, whose code
+is shown in ~\figref{qs-unstaged} partitions the list into two by
+using the first element as a pivot~\footnote{We assume that the list
+  is prepermuted to guarantee the expected linear time behavior.} and
+recurs down to one of the ``halves'' to find the desired element.  The
+``half'' to be considered recursively next is determined by the
+relationship of~\texttt{k} to the size of the first half~\texttt{n},
+which is returned by the function~\texttt{partition}.
+
+Continuing on our example, consider an application where we perform
+many order statistics queries using the same collection~\texttt{l}, but
+potentially with different ranks~\cite{k}.  To take advantage of this,
+we can {\em stage} quickselect so that the parts of the computation
+that depend on the collection~\texttt{l} is performed once, and the
+desired result is computed by using~\texttt{k} and the results from
+this prior stage.
+
+Prior work proposed staging for exactly these kinds of applications
+and proposed techniques for writing staged
+programs~\cite{staging}. Using techniques essentially the same as
+prior work, we can stage our quickselect code by using staged typing
+and inserting three keywords~\texttt{next, prev, hold} into the
+appropriate places in the code as shown in \figref{qs-staged}. To
+write the staged version, the programmer first observes that the
+rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
+rather than $\rmint$, indicating that it will be available in the
+second stage.  All types that are not of circle type are available in
+the first stage.  Naturally the output type, representing the $k$th
+largest element of the list, is also $\fut\rmint$, since it is not
+computed until the second stage. Having determined the type signature,
+the programmer can then insert $\next$ and $\prev$ constructs into the
+code. The construct $\next$ occurs in a stage \bbone\ context and
+indicates that the contents of its block are stage \bbtwo, whereas
+$\prev$ occurs in a stage \bbtwo\ context and indicates that the
+contents of its block is stage \bbone.  (We adopt the convention that
+the top-level context is stage \bbone.)  The output type of a $\next$
+block is the $\fut$'d version of the type of its stage
+\bbtwo\ contents.  For example, \verb|next{0}| from above has the type
+$\fut\rmint$.  Correspondingly, $\prev$ requires that its stage
+\bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
+For example, \verb|prev{k}| from above has type $\rmint$ at stage
+\bbtwo, since $k$ is bound to a $\fut\rmint$ at stage
+\bbone. Informally speaking, the
+$\next$ construct enables operating on second stage computations in
+the first stage essentially by treating them as boxes that can be
+moved around, stored in data structures, and passed to functions, and
+the $\prev$ construct enables unwrapping the contents of such a box to
+use in the second stage.
+
+\ur{$\pause$ should be called ``promote'' or something like that. I
+  believe this is what Taha calls cross stage persistence.  THat work should
+  be cited.}
+
+For convenience, the code also uses $\pause$ blocks.  This construct
+has same stage signature as $\prev$, but instead of ``unwrapping"
+$\fut$ types it simply promotes integers from stage~\bbone\ to
+stage~\bbtwo.
+\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
+We instead provide it as a primitive to shorten examples.  
+Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
+This is related to the notion of {\em mobility} in \cite{murphy05} and
+{\em stability} in \cite{krishnaswami13}.}
+
+We note that given the type signature of staged quickselect, there are
+several ways to insert the staging annotations.  The version we
+presented here maximizes the work done at stage~\bbone.
+
+\ur{I don't like fuzzy words like hygenic that dont mean anything.
+  What is the point of using them?  A paper ought to be written for
+  the reader not as a note to self.}
+
+%%%
+%% There has been the extensive research into the question of how to
+%% automatically add staging annotations to unstaged code.  This process
+%% is known as {\em binding time analysis}, and we do not consider it
+%% here.  Instead, we assume that all input programs are properly staged
+%% according to some programmer's intent.
+
+Prior work on meta programming allows us to evaluate such staged code
+in a staged fashion by first providing the first-stage arguments and
+generating a piece of code that is specialized for the arguments.  For
+example, we can evaluate the staged quickselect by providing a
+list~\texttt{l} to obtain a piece of code that then can be used with
+different ranks (\texttt{k} values) in the second stage.  Such a
+specialized code essentially performs all the first-stage computations
+first and uses their results to improve the efficiency in the second
+stage, where we issue many queries.  The type system guarantees that
+all first-stage computations depend only on the first stage values and
+can indeed be computed in the absence of second-stage values.
+
+This staged evaluation technique corresponds to {\em partial
+  evaluation} or {\em program specialization} and can bring out
+dramatic improvements in the efficiency of programs.  It has however
+one major limitation: since it requires the first stage arguments to
+be available, it cannot be performed at compile time, except of course
+when the first-stage values are available at compile time.  Since the
+evaluation in the first stage produces a piece of code that must also
+be compiled or interpreted to execute in the second stage, the
+approach can lead to code blowup and tends to be relatively expensive
+in practice.  Due to these reasons, the applications of partial
+evaluation are limited.  For example, in our quickselect example, it
+is unlikely to know the list~\texttt{l} that we will be operation on
+at compile time. Partial evaluation dynamically at run time is
+difficult because the first stage generates code that is linear in the
+size of the list, whose generation, dynamic compilation or
+interpretation can be excruciatingly slow.
+
+Knoblock and Ruf~\cite{KR} proposed  the {\em data specialization} technique
+
+limitations of
+
+Prior work on
+
+To illustrate this, we present colorized version of \texttt{qSelect}:
+
+\begin{lstlisting} 
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ... `
+
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in`2`
+      case compare k n of
+        LT => `qSelect 1`left `2`k
+      | EQ => h
+      | GT => `qSelect 1`right `2`(k-n-1)`
+\end{lstlisting}
+
+Here, the input $k$ and all parts of the computation that depend on it are colored blue,
+whereas all of the parts of the computation that do not depend on $k$ are colored red.
+The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.
+
+\ur{I find control dependence issue very confusing.  Let's factor this
+out and talk about it in the end. }
+
+To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
+first arguments to the recursive calls to \texttt{qSelect} are colored red,
+despite appearing underneath a blue case statement.
+This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.
+
+Observe the that helper function \texttt{partition} is colored entirely red,
+which is valid because it is called with only red arguments.  
+This is an illustration of the property stated above, that 
+the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
+Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.
+
+This suggests an optimization: we can split \texttt{qSelect} into two functions, 
+one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
+Because of the dependence properties stated above, 
+\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
+whereas \texttt{qSelect1} will only depend on the input list $l$.
+Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
+Equationally, we can represent this relationship as simply 
+\[
+\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
+\]
+for some list \texttt{l} and index \texttt{k}.
+
+\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}
+
+With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
+Because the original function is recursive,
+the first step in this process is to assume the conclusion at the call sites.
+
+\ur{quickselect 1 and 2: how did we introduce them?}
+
+This is performed in place:
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in`2`
+      case compare k n of
+        LT => qSelect2 (`1`qSelect1 left`2`) k
+      | EQ => h
+      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
+\end{lstlisting}
+We then pull the stage \bbone\ portion of each branch of the
+comparison up into a single tuple.
+\ur{Why did we include h}
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
+      case compare k n of
+        LT => qSelect2 (pi2 p) k
+      | EQ => pi1 p
+      | GT => qSelect2 (pi3 p) (k-n-1)`
+\end{lstlisting}
+
+\ur{Don't know what ``this last step'' refers to.}
+This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
+We continue to pull up the precomputation and include more in it:
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => `2`0`1`
+  | Cons (h,t) => 
+      let p =
+        let (left,right,n) = partition h t in
+        `1`(n,h,qSelect1 left,qSelect1 right)
+      in`2`
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+This completes the splitting of the cons branch.  
+The empty branch can also be split, though in a trivial way:
+
+\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  case l of
+    Empty => let p = () in `2`0`1`
+  | Cons (h,t) => 
+      let p =
+        let (left,right,n) = partition h t in
+        `1`(n,h,qSelect1 left,qSelect1 right)
+      in`2`
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+To continue splitting the case statement, we need to generate a sum type in the precomputation
+to remind the code what branch its in.
+
+\begin{lstlisting} 
+fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
+  let p = 
+    case l of
+      Empty => Top ()
+    | Cons (h,t) =>
+        let (left,right,n) = partition h t in`1`
+        Bot (n,h,qSelect1 left,qSelect1 right)
+  in`2`
+  case p of 
+    Top () => 0
+  | Bot p => 
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+We finish where we start, splitting \texttt{qSelect} into
+\texttt{qSelect1} and \texttt{qSelect2}.  
+Omitting a few type annotations, that is:
+
+\begin{lstlisting} 
+1`fun qSelect1 (l : list) = 
+    case l of
+      Empty => Top ()
+    | Cons (h,t) =>
+        let (left,right,n) = partition h t in
+        Bot (n,h,qSelect1 left,qSelect1 right)
+`2`
+fun qSelect2 p (k : int) : int = 
+  case p of 
+    Top () => 0
+  | Bot p => 
+      case compare k (pi1 p) of
+        LT => qSelect2 (pi3 p) k
+      | EQ => pi2 p
+      | GT => qSelect2 (pi4 p) (k-n-1)`
+\end{lstlisting}
+
+The final version, cleaned up and given better datatype names, is shown in \ref{fig:qssplit}.
+Observe that \texttt{qSelect1} essentially builds a binary search tree from the input list, 
+and \texttt{qSelect2} searches that tree for the correct index.
+The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
+the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
+whereas \texttt{qSelect2} takes $O(\log n)$ time.
+Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
+
+\nr{There's something else I need to say right here before the contribution statement below. 
+I want the reader to be thinking "yeah, this splitting business does seem cool."
+Bonus points if I also have them thinking "but it would be cooler if it were formal."}
+
+This paper formally describes and analyzes the splitting transformation exemplified above.
+(I also want to make it clear that recognizing \lang's appropriateness for this is itself is a contribution.)
+
+We start by defining a simple staged language, called \lang, to express the input to our splitting algorithm.  
+\lang\ unambiguously specifies how its terms should be split
+% and features a type system that ensures valid splittings exist.
+% only minimal refactoring should be required to write terms within this language,
+%and from there, the type system should be enough to prove that a valid splitting exists.
+by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
+Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
+subexpressions are evaluated before stage-\bbtwo\ ones.  
+After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
+will end up in the stage \bbone\ function ({\tt qSelect1}), 
+and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
+Typing rules for \lang\ ensure a valid splitting exists.  That is, 
+within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
+but never from \bbtwo\ to \bbone.
+
+\begin{figure}
+
+\begin{subfigure}{0.5\textwidth}
+\begin{lstlisting}
+1`datatype list = Empty | Cons of int * list
+fun partition (p : int, l : list) = ...`
+
+datatype tree = Branch of int * int * tree * tree
+                | Leaf
+
+1`fun qSelect1 (l : list) : tree =
+  case l of
+    Empty => Leaf
+  | Cons (h,t) => 
+      let (left,right,n) = partition h t in
+      Branch (n, h, qSelect1 left, qSelect1 right)`
+
+2`fun qSelect2 (p : tree, k : int) : int = 
+  case unroll p of
+    Leaf => 0
+  | Branch (n,h,p1,p2) => 
+      case compare k n of
+        LT => qSelect2 (p1,k)
+      | EQ => h
+      | GT => qSelect2 (p2,k-n-1)`
+\end{lstlisting}
+\caption{Split (two-phase) implementation of quickselect.}
+\label{fig:qssplit}
+\end{subfigure}
+\caption{Caption place holder}
+\end{figure}
+
+\begin{comment}
+
+%% Umut: this is the earlier version.  I am reorganizing this whole
+%% section.
 
@@
 Instead, we assume that all input programs are properly staged according to some programmer's intent.
+\end{comment}

### 2014-10-20 paper/example.tex
first pass on example.tex

@@
 
-\ur{I find it hard to distinguish between blue and black on printed paper.}
+\subsection{Comments, suggestions, and ``todo''}
+I rewrote this section.  There is more work to be done.
+
+\begin{itemize}
+\item 
+I find it hard to distinguish between blue and black on printed paper.
+
+\item
+$\pause$ should be called ``promote'' or something like that. I
+  believe this is what Taha calls cross stage persistence.  THat work should
+  be cited.
+
+\item
+I don't like fuzzy words like hygenic that dont mean anything.
+  What is the point of using them?  A paper ought to be written for
+  the reader not as a note to self.
+\end{itemize}
 
@@
 
-\ur{$\pause$ should be called ``promote'' or something like that. I
-  believe this is what Taha calls cross stage persistence.  THat work should
-  be cited.}
-
 For convenience, the code also uses $\pause$ blocks.  This construct
@@
 presented here maximizes the work done at stage~\bbone.
-
-\ur{I don't like fuzzy words like hygenic that dont mean anything.
-  What is the point of using them?  A paper ought to be written for
-  the reader not as a note to self.}
 
@@
 
-Knoblock and Ruf~\cite{KR} proposed  the {\em data specialization} technique
+Knoblock and Ruf~\cite{KR} proposed the {\em data specialization}
+technique to overcome some of these limitations of program
+specialization.  The idea behind this technique is to take a staged
+program and split it into a stage-1 and a stage-2 program without
+knowledge of any of the arguments.  Unlike program specialization,
+data specialization does not require knowledge of the first-stage
+arguments to produce fast code for the second stage.  Instead it yield
+a first stage program that maps the first stage arguments to an
+intermediate or {\em boundary data structure}, which is used by the
+second stage program along with the second stage arguments to produce
+the final results.  Knoblock and Ruf's observation is that the
+information required by the second stage can be represented solely as
+a data structure, making it unnecessary to generate and compile or
+interpret code dynamically at run time.  Data specialization can
+therefore be applied statically at compile time, leading to
+drastically improved run-time performance without any significant
+run-time overheads.
 
-limitations of
+By applying data specialization to our quickselect example, we would
+wish to derive two functions, \texttt{quickselect1} and
+\texttt{quickselect2}, that would be semantically equivalent to
+quickselect in the following sense:
+\\
+\texttt{quickselect l k = quickselect2 (quickselect1 l) k}.
+\\
+%
+Since \texttt{quickselect1} requires only on a first stage
+argument~\texttt{l}, we can run it once to produce a boundary data
+structure that can be used for all subsequent queries on~\texttt{l},
+e.g.,  the code consisting of $m$ calls 
+\begin{lstlisting}
+quickselect l 1
+quickselect l 2
+ @$\ldots$@ 
+quickselect l @$m$@
+\end{lstlisting}
+%
+can be replaced by 
+%
+\begin{lstlisting}
+b = quickselect1 l
+quickselect2 b 1
+quickselect b 2
+@$\ldots$@ 
+quickselect b @$m$@.
+\end{lstlisting}
 
-Prior work on
+The benefit of this ``optimization'' can be dramatic.  For example, in
+principle at least, the first stage function \texttt{quickselect1}
+could produce a balanced binary search tree holding the elements
+of~\texttt{l} and the second stage function \texttt{quickselect2}
+could use this balanced binary search tree and the given rank value to
+perform a logarithmic-time search in the tree to return the desired
+result.  Assuming that the list~\texttt{l} contains $n$ elements, this
+optimization would change the asymptotic complexity from expected
+(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} \cdot m
+\cdot\log{n})$, which for any $m \approx n$ reduces the complexity
+from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
+improvement.  Note that the desired optimized code shows above is
+intellectually more sophisticated than the code that we have started
+with: the optimized code is able to create a data structure, a
+balanced binary tree augmented with indexing information, and use a
+binary search technique over this tree to compute the result
+asymptotically more efficiently.  In fact, based our teaching
+experience, we can imagine this kind of problem to be a moderately
+difficult exam question in an undergraduate algorithms class, as it
+not only requires understanding of data structures such as binary
+search trees but also requires modifying them to augment with indexing
+information to support rank-based search.
+
+It is thus perhaps no surprise that the prior art in data
+specialization is not able to perform such ``optimizations''.
+Knoblock and Ruf's paper takes a very important step by formulating
+the problem and proposing a solution.  But their work applies to
+program fragments that don't involve sophisticated data types such as
+as sums and recursive types, and don't involve recursion. For example,
+their work would not apply to our quickselect example and similar ones
+involving recursion and/or  recursive types.
+
+In this paper, we show that it is possible to generalize data
+specialization to a fully general language with recursion and
+recursive types.  The key ideas behind our approach are: ...
+
+In what follows we present a relatively mechanical application of our
+techniques to our running example.
+
+\ur{I made up to here.  What follows should be shortened and made to
+  the point. Some time later...}
 

### 2014-10-20 paper/paper.tex
added remarks
first pass on example.tex

@@
 \usepackage{amsmath}
-%\usepackage{amssymb}
+\usepackage{amssymb}
 \usepackage{amsthm}
-\usepackage{bbm}
+%\usepackage{bbm}
 \usepackage{stmaryrd}
@@
 \usepackage{mathpartir}
-\usepackage{mathabx}
+%\usepackage{mathabx}
 \usepackage{listings}
@@
 %\PrerenderUnicode{λ}
+\usepackage{verbatim}
 
@@
 
-\input {macros}
+\input{macros}
+\input{macros-ref}
+\input{remarks}
 
@@
 \definecolor{stageTwoColor}{RGB}{0,0,127}
+
 \begin{document}
@@
         tabsize=4,
-        escapeinside={"*}{*"},
+%        escapeinside={"*}{*"},
+        escapeinside={@}{@},
         moredelim=*[is][\color{stageOneColor}]{1`}{`},
@@
 
-\input{intro}
+%% \input{intro}
 
@@
 
-%\vfill\eject\pagebreak
-\input{splitting}
+%% %\vfill\eject\pagebreak
+%%\input{splitting}
 \input{algoexamples}
 \input{relatedwork}
+\input{umut}
 

### 2014-10-20 paper/umut.tex
first pass on example.tex

@@
 
+\paragraph{Defitions and relationships.}
 The following definitions are adapted from the partial evaluation
@@
 
+THIS IS IMPORTANT. EXPAND ON THIS.  Knoblock and Ruf's data
+specialization is similar but statically splits the code. This is
+important because it is not feasible to perform splitting at run time
+as can be possible according to the definition of data splitting. This
+is similar to ours.  Their approach, however, seems to assume that the
+specialized code is a single non-recursive procedure.
+
 Both partial evaluation and data specialization are special cases of
@@
 
-\subsection{Techniques}
+\paragraph{Techniques}
 

### 2014-10-20 paper/example.tex
first pass on example.tex

@@
   the reader not as a note to self.
+
+\item the splitting example should shed more light to the techniques
+  and try to give more insight.
 \end{itemize}
@@
 many order statistics queries using the same collection~\texttt{l}, but
-potentially with different ranks~\cite{k}.  To take advantage of this,
-we can {\em stage} quickselect so that the parts of the computation
-that depend on the collection~\texttt{l} is performed once, and the
-desired result is computed by using~\texttt{k} and the results from
-this prior stage.
+ with $m$ different ranks~\texttt{k}:
+%
+\begin{lstlisting}
+quickselect l 1
+quickselect l 2
+ @$\ldots$@ 
+quickselect l @$m$@.
+\end{lstlisting}
+%
+Since \texttt{quickselect1} requires only on a first stage
+argument~\texttt{l}, we can run it once to produce a boundary data
+structure that can be used for all subsequent queries on~\texttt{l},
+e.g., the code consisting of $m$ calls To take advantage of this, we
+can {\em stage} quickselect so that the parts of the computation that
+depend on the collection~\texttt{l} is performed once, and the desired
+result is computed by using~\texttt{k} and the results from this prior
+stage.
 
@@
 largest element of the list, is also $\fut\rmint$, since it is not
-computed until the second stage. Having determined the type signature,
+computed until the second stage.
+
+Having determined the type signature,
 the programmer can then insert $\next$ and $\prev$ constructs into the
@@
 stage~\bbtwo.
-\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
-We instead provide it as a primitive to shorten examples.  
-Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
-This is related to the notion of {\em mobility} in \cite{murphy05} and
-{\em stability} in \cite{krishnaswami13}.}
-
-We note that given the type signature of staged quickselect, there are
-several ways to insert the staging annotations.  The version we
-presented here maximizes the work done at stage~\bbone.
+\footnote{It will turn out that $\pause$ is implementable---though it
+  takes some effort---from our other language features.  We instead
+  provide it as a primitive to shorten examples.  Furthermore, it
+  would be wise to extend $\pause$ to other base types, if we had
+  them, and to products and sums thereof.  This is related to the
+  notion of {\em mobility} in \cite{murphy05} and {\em stability} in
+  \cite{krishnaswami13}.} We note that given the type signature of
+staged quickselect, there are several ways to insert the staging
+annotations.  The version we presented here maximizes the work done at
+stage~\bbone.
 
@@
 optimization would change the asymptotic complexity from expected
-(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} \cdot m
-\cdot\log{n})$, which for any $m \approx n$ reduces the complexity
+(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the complexity
 from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time

### 2014-10-20 paper/umut.tex
working on the related work

@@
 partially evaluated programs as ``Futamura projections.''
+
+\paragraph{Staging and staging transformations.}
+
+To the best of our knowledge, Jorring and Sherlis are the
+first to define precisely the notion of {\em staging} by recognizing
+it to be a common optimization performed in software design, as the
+programmer tries to stage computations based on frequency of a
+computation and the availability of data~\cite{jorring86}.  They then
+define, staging as the problem of taking a function of the form
+$f(x,y)$ and deriving two functions $f_1$ and $f_2$ such that $f(x,y)
+= f_2(f_1(x),y)$.  As the authors point out, the key difference
+between staging and partial evaluation as that the latter performs no
+a priori transformation; instead when $x$ becomes available it
+specialized $f(x,\cdot)$ as much as possible.  Based on this notion of
+staging, Jorring and Sherlis also define the {\em meta programming}
+approach where $f(x, \cdot)$ is specialized by a meta program derived
+from the definition of $f(x,y)$.
+
+Having defined staging and meta programming problems, Jorring and
+Scherlis move on to describe techniques for them.  They present the
+technique called {\em pass separation} as a solution the problem of
+splitting a function into two such that the stage-1 function produces
+an intermediate data structure from the stage-1 data, which is then
+used by the stage-2 function along with stage-2 data to produce the
+final result.  Jorring and Scherlis then note describe how
+pass-separation can be applied to an interpreter to obtain a compiler
+from it by manually transforming the interpreter code into a compiler
+by analysis. They also describe how meta programming can be used to
+derive a specialized program not by creating a data structure as pass
+separation does by generating a more optimized code that ``freezes''
+certain actions to be performed at run time.
+
+Jorring and Scherlis highlight that their techniques are not meant to
+be applied automatically and that they describe what can be called a
+``vision.''  Indeed they describe their approach by way of examples as
+transformations that can be applied by the programmer manually and
+point out that they are powerful techniques that will require some time
+and further research to automate.
+
+Our work provides strong evidence that pass-separation transformations
+can indeed be performed on explicitly staged programs automatically.
+
+\ur{I didn't understand meta programming transformations that they do
+  and the relationship to followup work.}
 

### 2014-10-27 paper/splitting.tex
first pass on example.tex
added a few notes about mono

@@
 In this section, we define splitting for partial values, then extend it to all
-terms, and prove correctness of this algorithm.
+terms.
 
@@
 \bbone\ and stage \bbtwo\ components of the original function body.
+
+\ur{Unclear whether the term ``masking'' includes the case for
+  functions or not.  If it does, then it sounds rather
+  inaccurate.  If not, then it should be stated clearly that it does
+  not, but then what is the value of introducing one more term?}
 
@@
 
+\subsection{Cost}
+
+While these splitting rules are all correct in terms of producing the right values, applying them naively can result in terms with needlessly large cost.
+Consider the following example involving the factorial function:
+\begin{lstlisting}
+2`prev{
+  1`letfun fact (n : int) : int = 
+    if n <= 0 then 1 else fact(n-1)*n
+  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
+}`
+\end{lstlisting}
+
+The boundary type\footnote{The splitting results are untyped, but we add in type annotations to aid the reader.} of this is give by the following datatype declaration:
+\begin{lstlisting}
+datatype prec = L | R of prec
+\end{lstlisting}
+The example then splits into the precomputation,
+\begin{lstlisting}
+1`letfun fact (n : int) : int * prec = 
+  if n <= 0 then (1,L) 
+  else let (y,z) = fact(n-1) in (y*n,R z)
+in ((),fact 5)`
+\end{lstlisting}
+and the resumer
+\begin{lstlisting}
+2`l.
+letfun fact (n : unit, l0 : prec) : unit = 
+  case l0 of L => () | R l1 => fact ((),l0)
+in (fact (#2 l); #1 l)-100`
+\end{lstlisting}
+The problem here is that the naive system doesn't realize that the \texttt{fact} function has no stage \bbtwo\ content,
+and so it pessimistically retraces all of its steps in the resumer.  
+Fixing this requires some recursive reasoning.
+
+We instead tried to solve the issue by adding a single new staging annotation to \lang, called \texttt{mono}.  
+A \texttt{mono} block must appear in a stage \bbone\ context, and it indicates that its entire context is monostage.
+This can be encoded in our statics and dynamics by adding a new stage \bbmono\ and the judgements:
+\begin{mathpar}
+\infer{\typesone{\monoSt~e}A}{\typesmono e A} \and
+\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
+\end{mathpar}
+
+The splitting rule for \texttt{mono} is more difficult.
+
 \input{figures/splitting-misc}

### 2014-10-28 paper/splitting.tex
additions to splitting section

@@
 %\bbtwo$.  Before diving into the full generality,
-As discussed in the introduction, a stage splitter must take a term in \lang\
-and produce two
 
-\TODO waiting on how the intro defines splitting
+The goal of splitting is to take a multistage term and produce two new terms:
+one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
+Exactly how this goal manifests depends on the form of the input. 
+We consider four cases: stage \bbone\ values, stage \bbone\ terms, stage \bbtwo\ values,
+and stage \bbtwo\ terms.
 
-In this section, we define splitting for partial values, then extend it to all
-terms.
+We cover stage \bbtwo\ value splitting first, because it is trivial by construction.
+Since there is no stage \bbtwo\ expression that both is a value
+and contains \bbone\ subexpressions, 
+we say that stage \bbtwo\ values must contain no stage \bbone\ content.
+Accordingly, there is nothing really to split.
 
@@
 The splitting rule for \texttt{mono} is more difficult.
+...
+transition to dummy.
+
+The evaluation rules for dummy values are given by:
+\begin{mathpar}
+\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
+\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
+\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
+\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
+\end{mathpar}
+
+The basic idea here is that dummy will dynamically be whatever data structure you want it to be.
+
+This is safe because [nico is just throwing something out here] 
+it can only appear in the resumers of terms that have no circles in their \lang\ type.
+Since the $\prev$ rule requires a circle type, dummy values cannot leak from
+the resumers of stage \bbone\ terms into the resumers of stage \bbtwo\ terms.
 

### 2014-10-29 paper/splitting.tex
worked on global structure of splitting section

@@
 one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
-Exactly how this goal manifests depends on the form of the input. 
-We consider four cases: stage \bbone\ values, stage \bbone\ terms, stage \bbtwo\ values,
+Exactly what content is implied by this goal depends on the form of the input. 
+There are four cases: stage \bbone\ partial values, stage \bbone\ terms, stage \bbtwo\ values,
 and stage \bbtwo\ terms.
 
-We cover stage \bbtwo\ value splitting first, because it is trivial by construction.
-Since there is no stage \bbtwo\ expression that both is a value
-and contains \bbone\ subexpressions, 
+In all cases, the outputs of splitting are terms in an unstaged language, \langmono. 
+This language is untyped and contains no staging features, 
+but otherwise has the same term constructors as \lang. 
+It has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.
+
+\subsection {Stage \bbtwo\ Terms}
+
+We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
+In spirit, splitting takes a term $\coltwo e A$ and produces from it a first stage term
+$p : \tau$ (called the {\em precomputation}) and a second stage term $r : \tau \to A$ (called the {\em resumer}).
+Here, the responsibility of $p$ is to produce an intermediate data structure, 
+and the responsibility of $r$ is to consume that data structure and produce the same final value as $e$.
+
+Our actual presentation of stage splitting differs from this ideal in two ways.
+Firstly, the outputs are untyped, for reasons discussed later.
+Secondly, we do not actually represent the resumer as a function 
+but instead as a term open on a variable, typically $l.r$.
+This helps keep the splitting rules concise.
+
+ For example, the stage \bbtwo\ term
+\begin{lstlisting}
+2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
+\end{lstlisting}
+splits into a precomputation \verb|(1+2,3+4)| and a resumer
+\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
+the precomputation must agree with the original term. This process of hoisting
+all the stage \bbone\ subterms out of a term is called \emph{lifting}.
+
+Splitting is implemented by the judgment $\splittwosym$,
+the rules for which are given in \cref{fig:termSplitTwo}.  
+It operates inductively on the structure of the term.  
+As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
+from contained stage \bbone\ code into a single expression $p$, 
+leaving a resumer $l.r$ to interpret the results of that precomputation.
+Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
+that is, $p$ is a unit value and $l$ is never used in $r$.
+At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
+the precomputations of the constituent parts, and then unbundling them with a pattern,
+with the resumer precisely resembling the original expression.
+The $\prev$ rule depends on splitting stage \bbone\ terms, which is covered later.
+
+This is enough to define a correctness goal for splitting closed terms.
+
+\paragraph {Correctness of stage \bbtwo\ splitting.}
+If $e : A \splittwosym [p,l.r]$, then $e \equiv (\letin l p r)$
+
+\subsection {Stage \bbtwo\ Values}
+
+We next consider splitting for the special case of stage \bbtwo\ values.
+Observing that no stage \bbtwo\ expression both is a value and contains stage \bbone\ subexpressions, 
 we say that stage \bbtwo\ values must contain no stage \bbone\ content.
-Accordingly, there is nothing really to split.
+That is, stage \bbtwo\ value splitting is trivial by construction.
 
-\subsection{Partial Values}
+When applied to partial values, stage \bbtwo\ term splitting always produces a trivial
+precomputation, in the sense that the output is a nested sequence, possibly
+empty, of tuples and injections of \texttt{()}. The intuition is that a partial
+value has no remaining stage \bbone\ computation, and hence no precomputation.
 
-Partial values, along with future environments, are produced by partially
-evaluating stage \bbone\ terms. Although partial values have no more stage
-\bbone\ computation remaining, they may contain variables bound by the future
-environment, and thus, may require stage \bbtwo\ computation.
+%Partial value splitting is more than just a restriction of general expression
+%splitting precisely because it observes this property it at the structural
+%level, doing away with the precomputation entirely.
 
-Splitting a partial value $v$ creates two single-stage values $v_1$ and $v_2$,
+\subsection{Stage \bbone\ Partial Values}
+
+Partial values, along with future environments, are produced by partially evaluating stage \bbone\ terms. 
+They, unlike stage \bbtwo\ values, {\em can} have content at both stages.
+As discussed above, the {\tt quickselect} function itself is an example of a multi-stage values.
+
+Although partial values have no more stage
+\bbone\ computation remaining, they may contain $\next$ blocks
+(the contents of which must be variables bound by the future environment).
+The goal of splitting for stage \bbone\ partial values is therefore 
+to separate the contents of $\next$ blocks from the rest of the value, while maintaining some structure.
+
+Splitting a partial value $v$ creates two unstaged values $v_1$ and $v_2$,
 where $v_w$ contains only the information required for stage $w$ evaluation of
@@
 
-\ur{The case for ``fold'' seems missing in partial value splitting rules.}
+\ur{The case for ``roll'' seems missing in partial value splitting rules.}
 
@@
 
-\subsection{General Splitting}
-
-Splitting arbitrary terms in \lang\ again results in two terms containing only
-the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
-called, respectively, the {\em precomputation} and {\em resumer}.
-%
-For example, the stage \bbtwo\ term
-\begin{lstlisting}
-2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
-\end{lstlisting}
-splits into a precomputation \verb|(1+2,3+4)| and a resumer
-\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
-the precomputation must agree with the original term. This process of hoisting
-all the stage \bbone\ subterms out of a term is called \emph{lifting}.
+\subsection{Stage \bbone\ Term Splitting}
 
@@
 
-\subsection {Formal Setup}
-
-%in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
-Splitting is implemented by the judgments $\splitonesym$ and $\splittwosym$,
-which take stage \bbone\ and stage \bbtwo\ terms respectively; these judgments
-are defined in \cref{fig:termSplitOne,fig:termSplitTwo} inductively on the
-structure of the term. Both judgments produce two terms in an unstaged language,
-\langmono, whose grammar is given in \ref{fig:monoGrammar}. This language is
-untyped and contains no staging features, but otherwise has the same term
-constructors as \lang.
-%\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
-
+Splitting is implemented by the judgment $\splitonesym$,
+the rules for which are given in \cref{fig:termSplitOne}.  
+It operates inductively on the structure of the term.  
 The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
-$c$ and resumer $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
-precomputation $p$ and a resumer $l.r$. In both cases, the term $e$ is open on
-variables in the context $\Gamma$. We write the rules using pattern notation
-when appropriate.
-%For concision, we represent resumers as a term open on a single variable,
-%rather than as functions.
-
-\ur{Reformat rules to make them fit into the paper.}
-
-The $\splittwosym$ rules are conceptually simple and almost all very similar.  
-As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
-from contained stage \bbone\ code into a single expression $p$, 
-leaving a resumer $l.r$ to interpret the results of that precomputation.
-Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
-that is, $p$ is a unit value and $l$ is never used in $r$.
-At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
-the precomputations of the constituent parts, and then unbundling them with a pattern,
-with the resumer precisely resembling the original expression.
+$c$ and resumer $l.r$.
 
@@
 
+\subsection {Notable Properties}
+
+\nr{These points will probably get lifted into the preceding sections.}
+
 There are some particularly notable features of this definition of term
 splitting.
-
-\paragraph {Expression Splitting at Values}
-
-When applied to partial values, term splitting always produces a trivial
-precomputation, in the sense that the output is a nested sequence, possibly
-empty, of tuples and injections of \texttt{()}. The intuition is that a partial
-value has no remaining stage \bbone\ computation, and hence no precomputation.
-
-%Partial value splitting is more than just a restriction of general expression
-%splitting precisely because it observes this property it at the structural
-%level, doing away with the precomputation entirely.
 
@@
 
-%\subsection {Boundary Type Worst Case}
-%
-%\TODO add the example which is the worst case for figuring out boundary types
-%
+\subsection {Boundary Type Worst Case}
+
+\TODO add the example which shows that we need dynamic types
 

### 2014-10-29 paper/splitting.tex
more work on splitting

@@
 
-The goal of splitting is to take a multistage term and produce two new terms:
+The goal of splitting is to take a multistage term,
+which may feature interleaved stage \bbone\ and stage \bbtwo\ code,
+and untangle it to produce two separate terms:
 one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
-Exactly what content is implied by this goal depends on the form of the input. 
-There are four cases: stage \bbone\ partial values, stage \bbone\ terms, stage \bbtwo\ values,
-and stage \bbtwo\ terms.
+The splitting algorithm comes in several forms, depending on the stage of the input and whether it is a value.
+The stage \bbtwo\ versions will be addressed first,
+before the more complicated stage \bbone\ versions.
 
@@
 
-\subsection {Stage \bbtwo\ Terms}
+\subsection {Stage \bbtwo\ Splitting}
 
 We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
-In spirit, splitting takes a term $\coltwo e A$ and produces from it a first stage term
-$p : \tau$ (called the {\em precomputation}) and a second stage term $r : \tau \to A$ (called the {\em resumer}).
-Here, the responsibility of $p$ is to produce an intermediate data structure, 
+Splitting takes a term $\coltwo e A$ and produces from it a first stage term
+$p$ (called the {\em precomputation}) and a second stage term $l.r$ (called the {\em resumer})
+open on a single variable, such that $e$ is equivalent to $\letin l p r$.
+Intuitively, the responsibility of $p$ is to produce an intermediate data structure, 
 and the responsibility of $r$ is to consume that data structure and produce the same final value as $e$.
-
-Our actual presentation of stage splitting differs from this ideal in two ways.
-Firstly, the outputs are untyped, for reasons discussed later.
-Secondly, we do not actually represent the resumer as a function 
-but instead as a term open on a variable, typically $l.r$.
-This helps keep the splitting rules concise.
 
@@
 splits into a precomputation \verb|(1+2,3+4)| and a resumer
-\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
+\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to
 the precomputation must agree with the original term. This process of hoisting
-all the stage \bbone\ subterms out of a term is called \emph{lifting}.
+all the stage \bbone\ subterms out of a term is called \emph{precomputation lifting}.
 
@@
 
-This is enough to define a correctness goal for splitting closed terms.
+\subsection{(Partial) Value Splitting}
 
-\paragraph {Correctness of stage \bbtwo\ splitting.}
-If $e : A \splittwosym [p,l.r]$, then $e \equiv (\letin l p r)$
+Observe that when applied to a stage \bbtwo\ values, 
+splitting always produces an effectively trivial precomputation:
+either \texttt{()}, or nested tuples thereof.
+Intuitively, this makes sense---there is no way to embed stage \bbone\ code
+in the internals of a stage \bbtwo\ value, 
+thus they can have no stage \bbone\ content.
+Indeed it would be almost nonsensical to talk of an algorithm for splitting stage \bbtwo\ values,
+since the precondition, that the input is a stage \bbtwo\ value, is enough to imply the postcondition,
+that no stage interleaving remains.
 
-\subsection {Stage \bbtwo\ Values}
+The same is not true of partial values---partial values, a restriction on stage \bbone\ terms,
+{\em can} have stage \bbtwo\ content.
+Thus, it is meaningful to define some notion of splitting for them.
+This is useful to study, because the definition will later be used as a contract 
+with which general stage \bbone\ term splitting must comply.
 
-We next consider splitting for the special case of stage \bbtwo\ values.
-Observing that no stage \bbtwo\ expression both is a value and contains stage \bbone\ subexpressions, 
-we say that stage \bbtwo\ values must contain no stage \bbone\ content.
-That is, stage \bbtwo\ value splitting is trivial by construction.
-
-When applied to partial values, stage \bbtwo\ term splitting always produces a trivial
-precomputation, in the sense that the output is a nested sequence, possibly
-empty, of tuples and injections of \texttt{()}. The intuition is that a partial
-value has no remaining stage \bbone\ computation, and hence no precomputation.
-
-%Partial value splitting is more than just a restriction of general expression
-%splitting precisely because it observes this property it at the structural
-%level, doing away with the precomputation entirely.
-
-\subsection{Stage \bbone\ Partial Values}
-
-Partial values, along with future environments, are produced by partially evaluating stage \bbone\ terms. 
-They, unlike stage \bbtwo\ values, {\em can} have content at both stages.
-As discussed above, the {\tt quickselect} function itself is an example of a multi-stage values.
-
-Although partial values have no more stage
-\bbone\ computation remaining, they may contain $\next$ blocks
-(the contents of which must be variables bound by the future environment).
+Recall that a partial value is externally a stage \bbone\ thing; 
+it has a type $@~\bbone$, and along with a future context, it is the result of reducing a stage \bbone\ term.
+But internally, a partial value might contain stage \bbtwo\ subexpressions.
+There are two places where such embedding may occur: 
+$\next$ blocks, which must contain a variable reference to some future context; 
+and lambda values, which may contain a $\next$ in their body.
 The goal of splitting for stage \bbone\ partial values is therefore 
-to separate the contents of $\next$ blocks from the rest of the value, while maintaining some structure.
+to separate the contents of $\next$ blocks from the rest of the value, 
+while maintaining some structure.
 

### 2014-10-29 paper/splitting.tex
changes to intro of splitting

@@
 
-The goal of splitting is to take a multistage term,
+The goal of splitting is to take a multistage \lang\ term,
 which may feature interleaved stage \bbone\ and stage \bbtwo\ code,
 and untangle it to produce two separate terms:
-one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
-The splitting algorithm comes in several forms, depending on the stage of the input and whether it is a value.
-The stage \bbtwo\ versions will be addressed first,
-before the more complicated stage \bbone\ versions.
+one with all of the stage \bbone\ CONTENT, and one with all of
+the stage \bbtwo\ CONTENT.
 
-In all cases, the outputs of splitting are terms in an unstaged language, \langmono. 
-This language is untyped and contains no staging features, 
-but otherwise has the same term constructors as \lang. 
-It has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.
+Splitting emits terms in an unstaged language, \langmono, that is 
+untyped and contains no staging features, but otherwise has the same term constructors as \lang. 
+\langmono\ has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.
+
+Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than splitting stage 1 terms,
+we first describe the mechanics of splitting stage \bbtwo, before moving to the more complex
+case of stage \bbone\ term splitting.
+
+%The splitting algorithm comes in several forms, depending on the stage of the input and whether it is a value.
+%The stage \bbtwo\ versions will be addressed first,
+%before the more complicated stage \bbone\ versions.
 
@@
 
-We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
-Splitting takes a term $\coltwo e A$ and produces from it a first stage term
-$p$ (called the {\em precomputation}) and a second stage term $l.r$ (called the {\em resumer})
+% We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
+
+Splitting a stage-2 term $\coltwo e A$ produces a term
+$p$ (called the {\em precomputation}) and a term $l.r$ (called the {\em resumer})
 open on a single variable, such that $e$ is equivalent to $\letin l p r$.

### 2014-10-30 paper/obsolete/correctness.tex
minor tweaks

--- 2014-10-29 paper/splitting.tex
@@
-\section{Splitting Algorithm}
-\label{sec:splitting}
-
-%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
-%takes the form of two mutually-dependent judgments, $\splitonesym$ and
-%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
-%\bbtwo$.  Before diving into the full generality,
-
-The goal of splitting is to take a multistage term,
-which may feature interleaved stage \bbone\ and stage \bbtwo\ code,
-and untangle it to produce two separate terms:
-one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
-The splitting algorithm comes in several forms, depending on the stage of the input and whether it is a value.
-The stage \bbtwo\ versions will be addressed first,
-before the more complicated stage \bbone\ versions.
-
-In all cases, the outputs of splitting are terms in an unstaged language, \langmono. 
-This language is untyped and contains no staging features, 
-but otherwise has the same term constructors as \lang. 
-It has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.
-
-\subsection {Stage \bbtwo\ Splitting}
-
-We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
-Splitting takes a term $\coltwo e A$ and produces from it a first stage term
-$p$ (called the {\em precomputation}) and a second stage term $l.r$ (called the {\em resumer})
-open on a single variable, such that $e$ is equivalent to $\letin l p r$.
-Intuitively, the responsibility of $p$ is to produce an intermediate data structure, 
-and the responsibility of $r$ is to consume that data structure and produce the same final value as $e$.
-
- For example, the stage \bbtwo\ term
-\begin{lstlisting}
-2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
-\end{lstlisting}
-splits into a precomputation \verb|(1+2,3+4)| and a resumer
-\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to
-the precomputation must agree with the original term. This process of hoisting
-all the stage \bbone\ subterms out of a term is called \emph{precomputation lifting}.
-
-Splitting is implemented by the judgment $\splittwosym$,
-the rules for which are given in \cref{fig:termSplitTwo}.  
-It operates inductively on the structure of the term.  
-As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
-from contained stage \bbone\ code into a single expression $p$, 
-leaving a resumer $l.r$ to interpret the results of that precomputation.
-Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
-that is, $p$ is a unit value and $l$ is never used in $r$.
-At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
-the precomputations of the constituent parts, and then unbundling them with a pattern,
-with the resumer precisely resembling the original expression.
-The $\prev$ rule depends on splitting stage \bbone\ terms, which is covered later.
-
-\subsection{(Partial) Value Splitting}
-
-Observe that when applied to a stage \bbtwo\ values, 
-splitting always produces an effectively trivial precomputation:
-either \texttt{()}, or nested tuples thereof.
-Intuitively, this makes sense---there is no way to embed stage \bbone\ code
-in the internals of a stage \bbtwo\ value, 
-thus they can have no stage \bbone\ content.
-Indeed it would be almost nonsensical to talk of an algorithm for splitting stage \bbtwo\ values,
-since the precondition, that the input is a stage \bbtwo\ value, is enough to imply the postcondition,
-that no stage interleaving remains.
-
-The same is not true of partial values---partial values, a restriction on stage \bbone\ terms,
-{\em can} have stage \bbtwo\ content.
-Thus, it is meaningful to define some notion of splitting for them.
-This is useful to study, because the definition will later be used as a contract 
-with which general stage \bbone\ term splitting must comply.
-
-Recall that a partial value is externally a stage \bbone\ thing; 
-it has a type $@~\bbone$, and along with a future context, it is the result of reducing a stage \bbone\ term.
-But internally, a partial value might contain stage \bbtwo\ subexpressions.
-There are two places where such embedding may occur: 
-$\next$ blocks, which must contain a variable reference to some future context; 
-and lambda values, which may contain a $\next$ in their body.
-The goal of splitting for stage \bbone\ partial values is therefore 
-to separate the contents of $\next$ blocks from the rest of the value, 
-while maintaining some structure.
-
-Splitting a partial value $v$ creates two unstaged values $v_1$ and $v_2$,
-where $v_w$ contains only the information required for stage $w$ evaluation of
-$v$. Splitting partial values is implemented by the judgment $v \vsplito
-[v_1,v_2]$ defined in \ref{fig:valSplit}. 
-At product, sum, and recursive types, splitting acts recursively, gathering all
-subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
-entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
-at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
-entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
-exposing only the relevant structure of a term is called \emph{masking}.
-
-\ur{The case for ``roll'' seems missing in partial value splitting rules.}
-
-Splitting a function is more complex, because function bodies are open terms,
-and so are split by the judgments described in the next subsection. Essentially,
-functions will split into two functions, whose bodies correspond to the stage
-\bbone\ and stage \bbtwo\ components of the original function body.
-
-\ur{Unclear whether the term ``masking'' includes the case for
-  functions or not.  If it does, then it sounds rather
-  inaccurate.  If not, then it should be stated clearly that it does
-  not, but then what is the value of introducing one more term?}
-
-\subsection{Stage \bbone\ Term Splitting}
-
-Lifting and masking are two necessary ingredients for ensuring that splitting
-observes the stages of \lang. When splitting a partial value, the stage \bbone\
-result must \emph{mask} the stage \bbtwo\ computations in order to be
-single-staged. When splitting a stage \bbtwo\ term, the precomputation must
-\emph{lift} out all embedded stage \bbone\ computations, so they can be
-performed first, even though their results are only used in stage \bbtwo.
-%and it's accessible to whatever surrounding piece of code refers to the value being split.
-%is not accessible externally.
-
-Splitting a stage \bbone\ term, in contrast, requires both masking and splitting.
-Consider the term 
-\begin{lstlisting}
-1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
-\end{lstlisting}
-Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
-yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
-must occur at stage \bbone, the former is needed by stage \bbone, while the
-latter is needed by stage \bbtwo.
-
-The precomputation for this term is a \emph{combined result} which is
-the pair of these components: \verb|(((),3*4),1+2)|. The resumer is
-likewise the pair of the stage \bbtwo\ results of lifting and masking:
-\verb|fn L => (L<5,())|.
-%\footnote{In this example, the combined result is a 2-tuple, but more generally
-%it only must be a term which reduces to a 2-tuple.}
-
-Splitting is implemented by the judgment $\splitonesym$,
-the rules for which are given in \cref{fig:termSplitOne}.  
-It operates inductively on the structure of the term.  
-The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
-$c$ and resumer $l.r$.
-
-The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
-while the rule for $\prev$ projects the combined result of its subexpression to isolate the precomputation.
-It may seem like this $\prev$ rule is throwing information away, but we well show later that the first component 
-of the evaluated form of $c$ in the $\prev$ rule must always be unit.
-The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
-and projects out the integer result in the resumer.%
-\footnote{The resumer of an integer expression is usually trivial, 
-but we have to include it here for termination purposes.}
-
-\subsection {Notable Properties}
-
-\nr{These points will probably get lifted into the preceding sections.}
-
-There are some particularly notable features of this definition of term
-splitting.
-
-\paragraph {Speculation}
-
-In the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the
-precomputation is lifted out from within branches. This is the manifestation of
-the speculation behavior from the semantics.
-
-\paragraph {Stage \bbone\ Divergence}
-
-In the splitting rules for stage \bbone\ {\tt if} and {\tt case} expressions, we
-have enough information at stage \bbone\ to know what branch to take, so there's
-no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
-taken branch only, and inject the precomputation into a sum type. In the
-resumer, we case on that injection, and resume the stage \bbtwo\ portion of the
-correct branch.
-
-\paragraph {Stage \bbone\ Functions}
-
-The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
-handle this by splitting the function into two functions: one in the immediate
-result which handles all the stage \bbone\ content of the original, and one in
-the resumer which handles all of the stage \bbtwo\ content. Note that stage
-\bbone\ lambdas are partial values, and so have only a trivial precomputation.
-
-\subsection {Boundary Type Worst Case}
-
-\TODO add the example which shows that we need dynamic types
-
-\input{figures/splitting-term}
-
 \subsection{Correctness of Splitting}
@@
 \TODO write the theorem for $\vdash f:A\to\fut(B\to C)~@~\bbone$.
-
-%
-%\subsubsection{Simple Types}
-%
-%\begin{theorem}
-%If $\cdot\vdash e : A \splittwosym [p,l:\tau.r]$ then,
-%\begin{itemize}
-%\item $\cdot \vdash e : A~@~\bbtwo$ 
-%\item $\types [\cdot] p \tau$ and $\types [l:\tau] r A$ 
-%\item $\diatwo [\cdot] e q$ iff $\reduce p u$ and if so
-%\item $q \equiv (\letin{l}{u}{r})$
-%\end{itemize}
-%\end{theorem}
-%
-%\begin{theorem}
-%If $\cdot\vdash e : A \splitonesym [c,l:\tau.r]$ then,
-%\begin{itemize}
-%\item $\typesone [\cdot] e A$ 
-%\item $A \tsplito [A_1,A_2]$
-%\item $\types [\cdot] c {A_1 \times \tau}$ and $\types [l:\tau] r A_2$ 
-%\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ and if so
-%\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
-%\item $\reify \xi {v_2} q$
-%\item $q \equiv (\letin{l}{u}{r}) : A_2$
-%\end{itemize}
-%\end{theorem}
-%
-%\subsubsection{\bbone-Dependent Types}
-%\begin{theorem}
-%If $\cdot\vdash e : A \splitonesym [c,l:\tau.r]$ then,
-%\begin{itemize}
-%\item $\typesone [\cdot] e A$ 
-%\item $A \tsplito [A_1,a.A_2]$
-%\item $\types [\cdot] c {A_1 \times \tau}$ and $\types [l:\tau,a:A_1] r A_2$ 
-%\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ and if so
-%\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
-%\item $\reify \xi {v_2} q$
-%\item $q \equiv (\letin{l}{u}{r}) : A_2$
-%\end{itemize}
-%\end{theorem}
-
-\subsection{Cost}
-
-While these splitting rules are all correct in terms of producing the right values, applying them naively can result in terms with needlessly large cost.
-Consider the following example involving the factorial function:
-\begin{lstlisting}
-2`prev{
-  1`letfun fact (n : int) : int = 
-    if n <= 0 then 1 else fact(n-1)*n
-  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
-}`
-\end{lstlisting}
-
-The boundary type\footnote{The splitting results are untyped, but we add in type annotations to aid the reader.} of this is give by the following datatype declaration:
-\begin{lstlisting}
-datatype prec = L | R of prec
-\end{lstlisting}
-The example then splits into the precomputation,
-\begin{lstlisting}
-1`letfun fact (n : int) : int * prec = 
-  if n <= 0 then (1,L) 
-  else let (y,z) = fact(n-1) in (y*n,R z)
-in ((),fact 5)`
-\end{lstlisting}
-and the resumer
-\begin{lstlisting}
-2`l.
-letfun fact (n : unit, l0 : prec) : unit = 
-  case l0 of L => () | R l1 => fact ((),l0)
-in (fact (#2 l); #1 l)-100`
-\end{lstlisting}
-The problem here is that the naive system doesn't realize that the \texttt{fact} function has no stage \bbtwo\ content,
-and so it pessimistically retraces all of its steps in the resumer.  
-Fixing this requires some recursive reasoning.
-
-We instead tried to solve the issue by adding a single new staging annotation to \lang, called \texttt{mono}.  
-A \texttt{mono} block must appear in a stage \bbone\ context, and it indicates that its entire context is monostage.
-This can be encoded in our statics and dynamics by adding a new stage \bbmono\ and the judgements:
-\begin{mathpar}
-\infer{\typesone{\monoSt~e}A}{\typesmono e A} \and
-\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
-\end{mathpar}
-
-The splitting rule for \texttt{mono} is more difficult.
-...
-transition to dummy.
-
-The evaluation rules for dummy values are given by:
-\begin{mathpar}
-\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
-\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
-\end{mathpar}
-
-The basic idea here is that dummy will dynamically be whatever data structure you want it to be.
-
-This is safe because [nico is just throwing something out here] 
-it can only appear in the resumers of terms that have no circles in their \lang\ type.
-Since the $\prev$ rule requires a circle type, dummy values cannot leak from
-the resumers of stage \bbone\ terms into the resumers of stage \bbtwo\ terms.
-
-\input{figures/splitting-misc}

### 2014-10-30 paper/splitting.tex
minor tweaks

@@
 
-Observe that when applied to a stage \bbtwo\ values, 
+Observe that, when applied to stage \bbtwo\ values, 
 splitting always produces an effectively trivial precomputation:
@@
 
-\subsection {Boundary Type Worst Case}
+\paragraph {Typing the Boundary}
 
@@
 
-\subsection{Correctness of Splitting}
-
-We can now develop an notion of what it means for splitting to be correct.  
-Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
-Thus, we start with two mutually dependent definitions of equivalence.  
-Both relate residuals on the left with resumers on the right,
-but $\equiv$ equates closed expressions, whereas $\sim$ equates values.
-
-\begin{definition}
-For residual $q$ and resumer $r$, define $q \equiv r : A$ to mean that 
-$q \tworedsym v_q$ iff $\reduce {r} {v_r}$ where $v_q \sim v_r : A$, 
-\end{definition}
-
-\begin{definition}
-For residual value $v_1$ and resumer value $v_2$, define $v_1 \sim v_2 : A$ by the following cases:
-\begin{itemize}
-\item $i \sim i : \rmint$
-\item $(v_1,u_1) \sim (v_2,u_2) : A \times B$ where $v_1 \sim v_2 : A$ and $u_1 \sim u_2 : B$
-\item $\inl~v_1 \sim \inl~v_2 : A + B$ where $v_1 \sim v_2 : A$
-\item $\inr~v_1 \sim \inr~v_2 : A + B$ where $v_1 \sim v_2 : B$
-\item $\lam {x_1} A {e_1} \sim \lambda x_2.e_2 : A \to B$ where \\ $\forall (v_1 \sim v_2 : A). [v_1/x_1]e_1 \equiv [v_2/x_2]e_2 : B$
-\end{itemize}
-\end{definition}
-
-Essentially, we can read these as saying that two terms are equivalent if they evaluate to the same value,
-where "same" for functions means that those functions always evaluate to the same thing given equivalent inputs.
-
-We give the following end-to-end correctness lemmas for open terms. 
-Those readers not interested in the gritty details are advised to skip ahead to the theorems for closed terms, which give the important gist.
-
-In some of the lemmas, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
-which respectively hold all of the stage \bbone\ and stage \bbtwo\ variable bindings.
-Substitution splitting works just like value splitting, which is why they use the same symbol.
-
-\begin{lemma}
-If $\typesone e A$ then
-$\Gamma\vdash e : A \splitonesym [c,l.r]$.
-If $\typestwo e A$ then
-$\Gamma\vdash e : A \splittwosym [p,l.r]$.
-\end{lemma}
-
-\begin{lemma}
-If $\Gamma, \Gamma'\vdash e : A \splittwosym [p,l.r]$ then for all substitutions $\gamma : \Gamma$,
-\begin{itemize}
-\item $\gamma \vsplito [\gamma_1, \gamma_2]$
-\item $\diatwo [\Gamma'] {\gamma(e)} q$ iff $\reduce {\gamma_1(p)} u$ where
-\item $\Gamma' \vdash q \equiv (\letin{l}{u}{\gamma_2(r)})$
-\end{itemize}
-\end{lemma}
-
-\begin{lemma}
-If $\Gamma, \Gamma'\vdash e : A \splitonesym [c,l.r]$ then for all $\gamma : \Gamma$,
-\begin{itemize}
-\item $\gamma \vsplito [\gamma_1, \gamma_2]$
-\item $\diaone [\Gamma'] {\gamma(e)} {\xi;v}$ iff $\reduce {\gamma_1(c)} {(v_1,u)}$ where
-\item $\Gamma',\dom \xi \vdash v \vsplito [v_1,v_2]$
-\item $\reify \xi {v_2} q$
-\item $\Gamma'\vdash q \equiv (\letin{l}{u}{\gamma_2(r)})$
-\end{itemize}
-\end{lemma}
-
-%You should think of these theorems as saying that 
-%splitting commutes with evaluation.
-These lemmas are rather technical, but they ultimately imply that evaluating a
-closed term by splitting or by the dynamic semantics of \ref{ssec:dynamics} are
-equivalent. 
-We state this result for closed terms at each stage.
-
-\begin{theorem}[Correctness of splitting at $\bbone$]
-If $\vdash e:A~@~\bbone$, then (by splitting)
-\begin{itemize}
-\item $\vdash e : A \splitonesym [c,l.r]$
-\item $\reduce {c} {(v_1,u)}$
-\item $\reduce {(\letin{l}{u}{r})} v_S$
-\end{itemize}
-if and only if (by the staged dynamic semantics)
-\begin{itemize}
-\item $\diaone [] e {\xi;v}$
-\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
-\item $\reify \xi{v_2}q$
-\item $q \mathbin{\tworedsym} v_D$
-\end{itemize}
-and if so, then $v_D \sim v_S$.
-\end{theorem}
-
-\begin{theorem}[Correctness of splitting at $\bbtwo$]
-If $\vdash e:A~@~\bbtwo$, then (by splitting)
-\begin{itemize}
-\item $\vdash e : A \splittwosym [p,l.r]$
-\item $\reduce p u$
-\item $\reduce{(\letin{l}{u}{r})}{v_S}$
-\end{itemize}
-if and only if (by the staged dynamic semantics)
-\begin{itemize}
-\item $\diatwo [] e q$
-\item $q \mathbin{\tworedsym} v_D$
-\end{itemize}
-and if so, then $v_D \sim v_S$.
-\end{theorem}
-
-If we apply the former theorem to \verb|next{e}| of type $\fut A$, we
-essentially obtain the latter theorem at \verb|e|.
-
-The latter theorem implies that, given a multi-stage function $f:A\to\fut(B\to
-C)~@~\bbone$, the two methods of evaluating \verb|prev{f a} b| agree.
-However, we also expect that splitting $f$ directly will give us two functions,
-one which accepts an $A$ and outputs an intermediate value and boundary data,
-and another which takes in that boundary data and a $B$ and outputs a $C$.
-Moreover, the composition of these two functions should be extensionally equal
-to the staged dynamic semantics.
-
-\TODO write the theorem for $\vdash f:A\to\fut(B\to C)~@~\bbone$.
-
-%
-%\subsubsection{Simple Types}
-%
-%\begin{theorem}
-%If $\cdot\vdash e : A \splittwosym [p,l:\tau.r]$ then,
-%\begin{itemize}
-%\item $\cdot \vdash e : A~@~\bbtwo$ 
-%\item $\types [\cdot] p \tau$ and $\types [l:\tau] r A$ 
-%\item $\diatwo [\cdot] e q$ iff $\reduce p u$ and if so
-%\item $q \equiv (\letin{l}{u}{r})$
-%\end{itemize}
-%\end{theorem}
-%
-%\begin{theorem}
-%If $\cdot\vdash e : A \splitonesym [c,l:\tau.r]$ then,
-%\begin{itemize}
-%\item $\typesone [\cdot] e A$ 
-%\item $A \tsplito [A_1,A_2]$
-%\item $\types [\cdot] c {A_1 \times \tau}$ and $\types [l:\tau] r A_2$ 
-%\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ and if so
-%\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
-%\item $\reify \xi {v_2} q$
-%\item $q \equiv (\letin{l}{u}{r}) : A_2$
-%\end{itemize}
-%\end{theorem}
-%
-%\subsubsection{\bbone-Dependent Types}
-%\begin{theorem}
-%If $\cdot\vdash e : A \splitonesym [c,l:\tau.r]$ then,
-%\begin{itemize}
-%\item $\typesone [\cdot] e A$ 
-%\item $A \tsplito [A_1,a.A_2]$
-%\item $\types [\cdot] c {A_1 \times \tau}$ and $\types [l:\tau,a:A_1] r A_2$ 
-%\item $\diaone [\cdot] e {\xi;v}$ iff $\reduce c {(v_1,u)}$ and if so
-%\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
-%\item $\reify \xi {v_2} q$
-%\item $q \equiv (\letin{l}{u}{r}) : A_2$
-%\end{itemize}
-%\end{theorem}
+\TODO say something about correctness criteria for splitting? (see
+\texttt{obsolete/correctness.tex})
 
 \subsection{Cost}
+
+\TODO don't use that word
 

### 2014-10-30 paper/splitting.tex
few additions to splitting

@@
 
-\ur{The case for ``roll'' seems missing in partial value splitting rules.}
-
 Splitting a function is more complex, because function bodies are open terms,
@@
 \bbone\ and stage \bbtwo\ components of the original function body.
-
-\ur{Unclear whether the term ``masking'' includes the case for
-  functions or not.  If it does, then it sounds rather
-  inaccurate.  If not, then it should be stated clearly that it does
-  not, but then what is the value of introducing one more term?}
 
@@
 The splitting rule for \texttt{mono} is more difficult.
-...
-transition to dummy.
+The problem is that the rules needs to produce a resumer with the correct shape
+for whatever the type of the \texttt{mono} is.
+This is difficult in the presence of sums and general recursion,
+and so we introduce a new value in our target language called dummy.
+Dummy can only appear in resumers and dynamically takes the shape of whatever is required of it.
 
@@
 
-The basic idea here is that dummy will dynamically be whatever data structure you want it to be.
+Coming back to \texttt{mono}, the splitting rule is:
+\begin{mathpar}
+\infer{\splitone{\monoSt~e}A {e', {\tt dummy}} }{e \overset{C}{\rightarrow} e'} \and
+\infer{e_1~e_2 \overset{C}{\rightarrow} \pio~(e_1'~e_2')}{e_1 \overset{C}{\rightarrow} e_1' & e_2 \overset{C}{\rightarrow} e_2'} \and
+\infer{\lambda x.e \overset{C}{\rightarrow} \lambda x.(e',())}{e \overset{C}{\rightarrow} e'}
+\end{mathpar}
+And $\overset{C}{\rightarrow}$ operates like a map at all other terms.
 
-This is safe because [nico is just throwing something out here] 
+Our use of {\tt dummy} is safe because
 it can only appear in the resumers of terms that have no circles in their \lang\ type.

### 2014-10-30 paper/splitting.tex
heavy edits to 4.0-4.2

--- 2014-10-30 paper/splitting.tex
@@
+%\langmono\ has a evaluation judgment $\redsym$, which works in the expected way
+%to produce unstaged values.
+
 \section{Splitting Algorithm}
@@
 %$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
-%\bbtwo$.  Before diving into the full generality,
+%\bbtwo$.
 
-The goal of splitting is to take a multistage \lang\ term,
-which may feature interleaved stage \bbone\ and stage \bbtwo\ code,
-and untangle it to produce two separate terms:
-one with all of the stage \bbone\ CONTENT, and one with all of
-the stage \bbtwo\ CONTENT.
+Although terms in \lang\ clearly distinguish their stage \bbone\ and stage
+\bbtwo\ components, these components may be interleaved in complex ways. The
+goal of \emph{stage splitting} is to produce from a term in \lang\ a pair of
+terms in an unstaged language \langmono, which respectively perform all the
+stage \bbone\ and stage \bbtwo\ subcomputations of the original term.
 
-Splitting emits terms in an unstaged language, \langmono, that is 
-untyped and contains no staging features, but otherwise has the same term constructors as \lang. 
-\langmono\ has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.
+We begin by describing how to split stage \bbtwo\ terms.
 
-Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than splitting stage 1 terms,
-we first describe the mechanics of splitting stage \bbtwo, before moving to the more complex
-case of stage \bbone\ term splitting.
-
-%The splitting algorithm comes in several forms, depending on the stage of the input and whether it is a value.
-%The stage \bbtwo\ versions will be addressed first,
-%before the more complicated stage \bbone\ versions.
+%Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than
+%splitting stage 1 terms, we first describe the mechanics of splitting stage
+%\bbtwo, before moving to the more complex case of stage \bbone\ term splitting.
 
@@
 
-% We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
-
-Splitting a stage-2 term $\coltwo e A$ produces a term
-$p$ (called the {\em precomputation}) and a term $l.r$ (called the {\em resumer})
-open on a single variable, such that $e$ is equivalent to $\letin l p r$.
-Intuitively, the responsibility of $p$ is to produce an intermediate data structure, 
-and the responsibility of $r$ is to consume that data structure and produce the same final value as $e$.
+Splitting a stage-\bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
+term $p$ and a \emph{resumer} function $l.r$, such that $e$ is equivalent to
+$\letin l p r$. This is implemented by the judgment $\splittwosym$,
+(\cref{fig:termSplitTwo}), which operates inductively on the structure of the
+term.
 
@@
 splits into a precomputation \verb|(1+2,3+4)| and a resumer
-\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to
-the precomputation must agree with the original term. This process of hoisting
-all the stage \bbone\ subterms out of a term is called \emph{precomputation lifting}.
+\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to the
+precomputation must agree with the original term. This process of hoisting stage
+\bbone\ subterms out of $e$ is called \emph{precomputation lifting}, and
+intuitively, amounts to computing a data structure which is consumed by $r$.
 
-Splitting is implemented by the judgment $\splittwosym$,
-the rules for which are given in \cref{fig:termSplitTwo}.  
-It operates inductively on the structure of the term.  
-As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
-from contained stage \bbone\ code into a single expression $p$, 
-leaving a resumer $l.r$ to interpret the results of that precomputation.
-Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
-that is, $p$ is a unit value and $l$ is never used in $r$.
-At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
-the precomputations of the constituent parts, and then unbundling them with a pattern,
-with the resumer precisely resembling the original expression.
-The $\prev$ rule depends on splitting stage \bbone\ terms, which is covered later.
+Since base types and variables have no stage \bbone\ content, their
+precomputations are \texttt{()}, and are not used in the resumer. At every other
+expression except $\prev$, splitting bundles the precomputations of the
+constituent parts, and then unbundles them with a pattern, with the resumer
+precisely resembling the original expression. The $\prev$ rule depends on
+splitting stage \bbone\ terms, which is covered later.
+
+%As mentioned before, the only responsibility of $\splittwosym$ is to lift
+%out the precomputation from contained stage \bbone\ code into a single
+%expression $p$, leaving a resumer $l.r$ to interpret the results of that
+%precomputation.
 
@@
 
-Observe that, when applied to stage \bbtwo\ values, 
-splitting always produces an effectively trivial precomputation:
-either \texttt{()}, or nested tuples thereof.
-Intuitively, this makes sense---there is no way to embed stage \bbone\ code
-in the internals of a stage \bbtwo\ value, 
-thus they can have no stage \bbone\ content.
-Indeed it would be almost nonsensical to talk of an algorithm for splitting stage \bbtwo\ values,
-since the precondition, that the input is a stage \bbtwo\ value, is enough to imply the postcondition,
-that no stage interleaving remains.
+As described above, splitting a stage \bbtwo\ value always produces a trivial
+precomputation---either \texttt{()}, or nested tuples thereof. Intuitively,
+this is because stage-\bbtwo\ values contain no stage \bbone\ computations.
 
-The same is not true of partial values---partial values, a restriction on stage \bbone\ terms,
-{\em can} have stage \bbtwo\ content.
-Thus, it is meaningful to define some notion of splitting for them.
-This is useful to study, because the definition will later be used as a contract 
-with which general stage \bbone\ term splitting must comply.
+The same is not true of partial values (at stage \bbone), which \emph{can}
+contain stage \bbtwo\ computations, namely, inside $\next$ blocks, which must
+contain a variable reference to some future context, and inside lambdas, which
+may contain a $\next$ in their body.
 
-Recall that a partial value is externally a stage \bbone\ thing; 
-it has a type $@~\bbone$, and along with a future context, it is the result of reducing a stage \bbone\ term.
-But internally, a partial value might contain stage \bbtwo\ subexpressions.
-There are two places where such embedding may occur: 
-$\next$ blocks, which must contain a variable reference to some future context; 
-and lambda values, which may contain a $\next$ in their body.
-The goal of splitting for stage \bbone\ partial values is therefore 
-to separate the contents of $\next$ blocks from the rest of the value, 
-while maintaining some structure.
-
-Splitting a partial value $v$ creates two unstaged values $v_1$ and $v_2$,
-where $v_w$ contains only the information required for stage $w$ evaluation of
-$v$. Splitting partial values is implemented by the judgment $v \vsplito
-[v_1,v_2]$ defined in \ref{fig:valSplit}. 
+Thus, to split partial values, we must separate the contents of $\next$ blocks
+from the rest of the term, while maintaining the term's original structure. This
+is implemented by the judgment $v \vsplito [v_1,v_2]$ (\ref{fig:valSplit}).
 At product, sum, and recursive types, splitting acts recursively, gathering all
-subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
-entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
-at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
-entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
-exposing only the relevant structure of a term is called \emph{masking}.
+subterms' stage $w$ content into $v_w$.
+At base types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()}
+to stage \bbtwo);
+on the other hand, at $\fut$ types, splitting assigns $v$ (which must have the
+form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
+\bbone).
+This process of exposing only the relevant structure of a term at each stage is
+called \emph{masking}.
 
@@
   not, but then what is the value of introducing one more term?}
+
+%Indeed it would be almost nonsensical to talk of an algorithm for splitting
+%stage \bbtwo\ values, since the precondition, that the input is a stage \bbtwo\
+%value, is enough to imply the postcondition, that no stage interleaving remains
+
+%Thus, it is meaningful to define some notion of splitting for them.  This is
+%useful to study, because the definition will later be used as a contract with
+%which general stage \bbone\ term splitting must comply.
+
+%Recall that a partial value is externally a stage \bbone\ thing; it has a type
+%$@~\bbone$, and along with a future context, it is the result of reducing a
+%stage \bbone\ term. But internally, a partial value might contain stage \bbtwo\
+%subexpressions.
 

### 2014-10-30 paper/splitting.tex
small edits

@@
 
-Splitting a stage-\bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
+Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
 term $p$ and a \emph{resumer} function $l.r$, such that $e$ is equivalent to
-$\letin l p r$. This is implemented by the judgment $\splittwosym$,
+$\letin l p r$. This is implemented by the judgment $\splittwosym$ 
 (\cref{fig:termSplitTwo}), which operates inductively on the structure of the
@@
 precomputations are \texttt{()}, and are not used in the resumer. At every other
-expression except $\prev$, splitting bundles the precomputations of the
+expression except $\prev$ and $\pause$, splitting bundles the precomputations of the
 constituent parts, and then unbundles them with a pattern, with the resumer
-precisely resembling the original expression. The $\prev$ rule depends on
+precisely resembling the original expression. The $\prev$ and $\pause$ rules depends on
 splitting stage \bbone\ terms, which is covered later.
+
+Observe that in the rules for {\tt if}s, {\tt case}s, and functions, the
+precomputation is lifted out from within branches. This is the manifestation of
+the speculation behavior from the semantics.
 
@@
 
-The same is not true of partial values (at stage \bbone), which \emph{can}
+The same is not true of partial values (by definition, stage \bbone), which \emph{can}
 contain stage \bbtwo\ computations, namely, inside $\next$ blocks, which must
@@
 to stage \bbtwo);
-on the other hand, at $\fut$ types, splitting assigns $v$ (which must have the
+conversely, at $\fut$ types, splitting assigns $v$ (which must have the
 form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
@@
 \bbone\ and stage \bbtwo\ components of the original function body.
+
+\TODO talk about the changed function signature, which is necessitated by the precomputation;
+note that we have to assume {\em all} stage \bbone\ functions have meaningful precomputations
 
@@
 splitting.
-
-\paragraph {Speculation}
-
-In the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the
-precomputation is lifted out from within branches. This is the manifestation of
-the speculation behavior from the semantics.
 

### 2014-10-31 paper/splitting.tex
small changes; lots of commenting-out

@@
 \bbone\ and stage \bbtwo\ components of the original function body.
-
-\TODO talk about the changed function signature, which is necessitated by the precomputation;
-note that we have to assume {\em all} stage \bbone\ functions have meaningful precomputations
+Moreover, the body of the function may contain a precomputation,
+which should be saved and resumer appropriately wherever the function is applied.
 
@@
 
-\subsection {Notable Properties}
-
-\nr{These points will probably get lifted into the preceding sections.}
-
-There are some particularly notable features of this definition of term
-splitting.
-
-\paragraph {Stage \bbone\ Divergence}
-
-In the splitting rules for stage \bbone\ {\tt if} and {\tt case} expressions, we
+In the splitting rules for stage \bbone\ {\tt case} expressions, we
 have enough information at stage \bbone\ to know what branch to take, so there's
@@
 
-\paragraph {Stage \bbone\ Functions}
+%\subsection {Notable Properties}
+%
+%\nr{These points will probably get lifted into the preceding sections.}
+%
+%There are some particularly notable features of this definition of term
+%splitting.
 
-The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
-handle this by splitting the function into two functions: one in the immediate
-result which handles all the stage \bbone\ content of the original, and one in
-the resumer which handles all of the stage \bbtwo\ content. Note that stage
-\bbone\ lambdas are partial values, and so have only a trivial precomputation.
+%\paragraph {Stage \bbone\ Divergence}
+%
 
-\paragraph {Typing the Boundary}
-
-\TODO add the example which shows that we need dynamic types
+%\paragraph {Stage \bbone\ Functions}
+%
+%The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
+%handle this by splitting the function into two functions: one in the immediate
+%result which handles all the stage \bbone\ content of the original, and one in
+%the resumer which handles all of the stage \bbtwo\ content. Note that stage
+%\bbone\ lambdas are partial values, and so have only a trivial precomputation.
+%
+%\paragraph {Typing the Boundary}
+%
+%\TODO add the example which shows that we need dynamic types
 
@@
 
-\subsection{Cost}
+\subsection{Eliminating Unnecessary Code}
 

### 2014-10-31 paper/example.tex
rewrite 4.0 - 4.1 with nico

@@
 rank~\texttt{k} in the list.  To this end, the algorithm, whose code
-is shown in ~\figref{qs-unstaged} partitions the list into two by
-using the first element as a pivot~\footnote{We assume that the list
+is shown in ~\ref{fig:qs-unstaged} partitions the list into two by
+using the first element as a pivot\footnote{We assume that the list
   is prepermuted to guarantee the expected linear time behavior.} and
@@
 and inserting three keywords~\texttt{next, prev, hold} into the
-appropriate places in the code as shown in \figref{qs-staged}. To
+appropriate places in the code as shown in \ref{fig:qs-staged}. To
 write the staged version, the programmer first observes that the

### 2014-10-31 paper/splitting.tex
rewrite 4.0 - 4.1 with nico

@@
-%\langmono\ has a evaluation judgment $\redsym$, which works in the expected way
-%to produce unstaged values.
-
 \section{Splitting Algorithm}
 \label{sec:splitting}
-
-%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
-%takes the form of two mutually-dependent judgments, $\splitonesym$ and
-%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
-%\bbtwo$.
 
@@
 
-%Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than
-%splitting stage 1 terms, we first describe the mechanics of splitting stage
-%\bbtwo, before moving to the more complex case of stage \bbone\ term splitting.
-
 \subsection {Stage \bbtwo\ Splitting}
@@
 Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
-term $p$ and a \emph{resumer} function $l.r$, such that $e$ is equivalent to
-$\letin l p r$. This is implemented by the judgment $\splittwosym$ 
-(\cref{fig:termSplitTwo}), which operates inductively on the structure of the
-term.
+term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
+and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
+computations can depend on stage \bbone, $r$ can depend on $p$ via the variable
+$l$. The correctness of splitting thus requires that $e$ is equivalent to
+$\letin l p r$.
 
+Splitting, as implemented by the judgment $e \splittwosym [p,l.r]$
+(\cref{fig:termSplitTwo}), is guided by the structure of the term $e$. 
 For example, the stage \bbtwo\ term
 \begin{lstlisting}
-2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
+2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
 \end{lstlisting}
-splits into a precomputation \verb|(1+2,3+4)| and a resumer
-\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to the
-precomputation must agree with the original term. This process of hoisting stage
-\bbone\ subterms out of $e$ is called \emph{precomputation lifting}, and
-intuitively, amounts to computing a data structure which is consumed by $r$.
+has two subterms, the latter of which itself has two subterms. Two are $\pause$s
+containing stage \bbone\ computations, while the third has no stage \bbone\
+content. The precomputation \emph{lifts} these computations in the same tree
+structure: \verb|(1+2,(3+4,()))|.%
+\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
+The resumer is a function which computes the stage \bbtwo\ value of the original
+term by mapping the results of the precomputation in place of the $\pause$s:
+\verb|(L1,(L2,_)) => L1 < (L2 - 5)|.
 
-Since base types and variables have no stage \bbone\ content, their
-precomputations are \texttt{()}, and are not used in the resumer. At every other
-expression except $\prev$ and $\pause$, splitting bundles the precomputations of the
-constituent parts, and then unbundles them with a pattern, with the resumer
-precisely resembling the original expression. The $\prev$ and $\pause$ rules depends on
-splitting stage \bbone\ terms, which is covered later.
+In general, constants and variables have no stage \bbone\ content, so their
+precomputations are \texttt{()}, and their resumers return the original term's
+value. Non-staging constructs with $n$ subterms have precomputations which are
+$n$-tuples of their subterms' precomputations, and resumers which pass each
+component to the resumer of the respective subterm. (This behavior at
+\texttt{case} and functions gives rise to speculative execution of
+precomputations from underneath binders.) The $\prev$ and $\pause$ rules depends
+on splitting stage \bbone\ terms, which is covered later.
 
-Observe that in the rules for {\tt if}s, {\tt case}s, and functions, the
-precomputation is lifted out from within branches. This is the manifestation of
-the speculation behavior from the semantics.
+\subsection{Partial Value Splitting}
 
-%As mentioned before, the only responsibility of $\splittwosym$ is to lift
-%out the precomputation from contained stage \bbone\ code into a single
-%expression $p$, leaving a resumer $l.r$ to interpret the results of that
-%precomputation.
+%while maintaining the term's original structure.
 
-\subsection{(Partial) Value Splitting}
-
-As described above, splitting a stage \bbtwo\ value always produces a trivial
-precomputation---either \texttt{()}, or nested tuples thereof. Intuitively,
-this is because stage-\bbtwo\ values contain no stage \bbone\ computations.
-
-The same is not true of partial values (by definition, stage \bbone), which \emph{can}
-contain stage \bbtwo\ computations, namely, inside $\next$ blocks, which must
+Stage \bbtwo\ values can contain no stage \bbone\ subcomputations, but the
+reverse is not true of partial values (at stage \bbone). These \emph{can}
+contain stage \bbtwo\ computations, both inside $\next$ blocks, which must
 contain a variable reference to some future context, and inside lambdas, which
@@
 Thus, to split partial values, we must separate the contents of $\next$ blocks
-from the rest of the term, while maintaining the term's original structure. This
-is implemented by the judgment $v \vsplito [v_1,v_2]$ (\ref{fig:valSplit}).
-At product, sum, and recursive types, splitting acts recursively, gathering all
-subterms' stage $w$ content into $v_w$.
-At base types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()}
-to stage \bbtwo);
-conversely, at $\fut$ types, splitting assigns $v$ (which must have the
+from the rest of the term. This is implemented by the judgment $v \vsplito
+[v_1,v_2]$ (\ref{fig:valSplit}).
+
+No dependence here.
+
+For example, the stage \bbone\ term
+\begin{lstlisting}
+1`inl(3, next{`2`yhat`1`})`
+\end{lstlisting}
+splits into
+\verb|inl(3,())| and \texttt{((),$\mathtt{\hat{y}}$)}.
+
+sum types
+
+In general,
+at product and recursive types, splitting acts
+recursively, gathering all subterms' stage $w$ content into $v_w$. At base
+types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()} to stage
+\bbtwo); conversely, at $\fut$ types, splitting assigns $v$ (which must have the
 form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
 \bbone).
+
 This process of exposing only the relevant structure of a term at each stage is
@@
 \bbone\ and stage \bbtwo\ components of the original function body.
-Moreover, the body of the function may contain a precomputation,
-which should be saved and resumer appropriately wherever the function is applied.
+Because that body may contain a precomputation, the generated function signatures must properly emit and resume with that precomputation.
 
-%Indeed it would be almost nonsensical to talk of an algorithm for splitting
-%stage \bbtwo\ values, since the precondition, that the input is a stage \bbtwo\
-%value, is enough to imply the postcondition, that no stage interleaving remains
+%As described above, splitting a stage \bbtwo\ value always produces a trivial
+%precomputation---either \texttt{()}, or nested tuples thereof.
 
@@
 
-\subsection{Stage \bbone\ Term Splitting}
+\subsection{Stage \bbone\ Splitting}
+
+\crem{value splitting figure is missing injections}
+
+\TODO
+\begin{enumerate}
+\item Examples
+\item Lifting (and masking)
+\item Correctness? (Does this require partial value splitting?)
+\end{enumerate}
+
+\begin{theorem}[Correctness of splitting at $\bbone$]
+If $\vdash e:A~@~\bbone$, then (by splitting)
+\begin{itemize}
+\item $\vdash e : A \splitonesym [c,l.r]$
+\item $\reduce {c} {(v_1,u)}$
+\item $\reduce {(\letin{l}{u}{r})} v_S$
+\end{itemize}
+if and only if (by the staged dynamic semantics)
+\begin{itemize}
+\item $\diaone [] e {\xi;v}$
+\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
+\item $\reify \xi{v_2}q$
+\item $q \mathbin{\tworedsym} v_D$
+\end{itemize}
+and if so, then $v_D \sim v_S$.
+\end{theorem}
 
@@
 
-Splitting a stage \bbone\ term, in contrast, requires both masking and splitting.
+Splitting a stage \bbone\ term, in contrast, requires both masking and lifting.
 Consider the term 
@@
 
-%\subsection {Notable Properties}
-%
-%\nr{These points will probably get lifted into the preceding sections.}
-%
-%There are some particularly notable features of this definition of term
-%splitting.
-
-%\paragraph {Stage \bbone\ Divergence}
-%
-
-%\paragraph {Stage \bbone\ Functions}
-%
-%The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
-%handle this by splitting the function into two functions: one in the immediate
-%result which handles all the stage \bbone\ content of the original, and one in
-%the resumer which handles all of the stage \bbtwo\ content. Note that stage
-%\bbone\ lambdas are partial values, and so have only a trivial precomputation.
-%
-%\paragraph {Typing the Boundary}
-%
-%\TODO add the example which shows that we need dynamic types
-
 \input{figures/splitting-term}
-
-\TODO say something about correctness criteria for splitting? (see
-\texttt{obsolete/correctness.tex})
 
@@
 
-\TODO don't use that word
+\crem{don't say ``cost''}
 

### 2014-11-04 paper/splitting.tex
Comments from marathon meeting.

@@
 
-We begin by describing how to split stage \bbtwo\ terms.
+Since splitting stage \bbtwo\ terms is easier than splitting stage \bbone\ terms, 
+we begin by describing how to split stage \bbtwo\ terms.
 
@@
 and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
-computations can depend on stage \bbone, $r$ can depend on $p$ via the variable
+computations can depend on stage \bbone\ computations, $r$ can depend on $p$ via the variable
 $l$. The correctness of splitting thus requires that $e$ is equivalent to
@@
 
-Splitting, as implemented by the judgment $e \splittwosym [p,l.r]$
+% TODO this contract works because a stage-2 term reduces to a value
+% so we can simply compare (p,l.r) to that value
+
+Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym [p,l.r]$
 (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$. 
@@
 component to the resumer of the respective subterm. (This behavior at
-\texttt{case} and functions gives rise to speculative execution of
-precomputations from underneath binders.) The $\prev$ and $\pause$ rules depends
+\texttt{case} and functions causes precomputations to be lifted out from 
+underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends
 on splitting stage \bbone\ terms, which is covered later.
 
-\subsection{Partial Value Splitting}
+\subsection{Stage \bbone\ Splitting}
+
+\TODO
+The contract for splitting stage \bbone\ terms is more complex, because stage \bbone\ 
+terms reduce to a pair of a future context and a partial value, not simply to a value.
 

### 2014-11-04 paper/splitting.tex
rewrote 4.2 with nico

@@
 
-Since splitting stage \bbtwo\ terms is easier than splitting stage \bbone\ terms, 
-we begin by describing how to split stage \bbtwo\ terms.
+Since splitting stage \bbtwo\ terms is easier than splitting stage \bbone\
+terms, we begin by describing how to split stage \bbtwo\ terms.
 
@@
 and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
-computations can depend on stage \bbone\ computations, $r$ can depend on $p$ via the variable
-$l$. The correctness of splitting thus requires that $e$ is equivalent to
-$\letin l p r$.
+computations can depend on stage \bbone\ computations, $r$ can depend on $p$ via
+the variable $l$. The correctness of splitting thus requires that $e$ is
+equivalent to $\letin l p r$.
 
@@
 
-Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym [p,l.r]$
-(\cref{fig:termSplitTwo}), is guided by the structure of the term $e$. 
+Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym
+[p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.
 For example, the stage \bbtwo\ term
@@
 \texttt{case} and functions causes precomputations to be lifted out from
-underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends
-on splitting stage \bbone\ terms, which is covered later.
+underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends on
+splitting stage \bbone\ terms, which is covered later.
 
@@
 
-\TODO
-The contract for splitting stage \bbone\ terms is more complex, because stage \bbone\ 
-terms reduce to a pair of a future context and a partial value, not simply to a value.
+Splitting a stage \bbone\ term $\colone e A$ produces a term and a function,
+which respectively contain all the stage \bbone\ and stage \bbtwo\
+subcomputations of $e$. The situation is more complex than when splitting stage
+\bbtwo\ terms; to see why, consider the correctness criterion for splitting
+stage \bbone\ terms.
 
-%while maintaining the term's original structure.
+Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
+yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
+value $v$. It is tempting to compare the two results of evaluating $e$ to the
+two results of splitting it. To see where this goes wrong, consider the stage
+\bbone\ term
+\begin{lstlisting}
+1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+\end{lstlisting}
+which evaluates to
+\begin{lstlisting}
+2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
+\end{lstlisting}
+This partial value contains a $\next$, and so is itself multistaged:
+intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
+is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5}.
 
-Stage \bbtwo\ values can contain no stage \bbone\ subcomputations, but the
-reverse is not true of partial values (at stage \bbone). These \emph{can}
-contain stage \bbtwo\ computations, both inside $\next$ blocks, which must
-contain a variable reference to some future context, and inside lambdas, which
-may contain a $\next$ in their body.
+We formalize this notion via two \emph{masking} functions (\ref{fig:valSplit}),
+$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
+a partial value at each stage, essentially by replacing all stage \bbone\
+(resp., \bbtwo) components of a partial value by \texttt{()}. The first result
+of splitting must then compute the \bbone-mask of the partial value, and the
+second result the \bbtwo-mask.
 
-Thus, to split partial values, we must separate the contents of $\next$ blocks
-from the rest of the term. This is implemented by the judgment $v \vsplito
-[v_1,v_2]$ (\ref{fig:valSplit}).
+There is one final subtlety to splitting stage \bbone\ terms. In the example
+above, there are \emph{two} flavors of stage \bbone\ subcomputations:
+\texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
+a precomputation which is used by the computation of the \bbtwo-mask.
 
-No dependence here.
+Formally, splitting a stage \bbone\ term, as implemented by the judgment $e
+\splitonesym [c,l.r]$ (\ref{fig:termSplitOne}), yields a \emph{combined result}
+$c$ which computes a pair of the stage \bbone\ result and the precomputation,
+and a \emph{resumer} function $l.r$. The correctness of splitting requires that,
+for $e\mathbin{\redonesym}[\xi;v]$, we have that $\pio~c$ agrees with
+$\masko{v}$, and $\letin l{\pit~c}r$ agrees with $\maskt{\xi;v}$.
 
-For example, the stage \bbone\ term
-\begin{lstlisting}
-1`inl(3, next{`2`yhat`1`})`
-\end{lstlisting}
-splits into
-\verb|inl(3,())| and \texttt{((),$\mathtt{\hat{y}}$)}.
+The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
+resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
+which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
+\texttt{((),12)}, which agrees with the stage \bbtwo\ mask.
 
-sum types
+\paragraph{Stage \bbone\ masking}
 
-In general,
-at product and recursive types, splitting acts
-recursively, gathering all subterms' stage $w$ content into $v_w$. At base
-types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()} to stage
-\bbtwo); conversely, at $\fut$ types, splitting assigns $v$ (which must have the
-form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
-\bbone).
+The purpose of masking is to expose only the relevant structure of a partial
+value $v$ at each stage. For the most part, stage \bbone\ masking, $\masko{v}$,
+simply replaces all $\next$s in $v$ with \texttt{()}. The only exception is for
+functions $\lam xAe$; such a value represents a computation, so masking splits
+its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
+stage \bbone\ component, the combined result.
 
-This process of exposing only the relevant structure of a term at each stage is
-called \emph{masking}.
+\paragraph{Stage \bbtwo\ masking}
 
-Splitting a function is more complex, because function bodies are open terms,
-and so are split by the judgments described in the next subsection. Essentially,
-functions will split into two functions, whose bodies correspond to the stage
-\bbone\ and stage \bbtwo\ components of the original function body.
-Because that body may contain a precomputation, the generated function signatures must properly emit and resume with that precomputation.
+Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
+future context $\xi$, because the stage \bbtwo\ parts of $v$ contain variables
+in $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
+structure, and then reify the resulting term in the context $\xi$. This
+rewriting process replaces base type constants in $v$ with \texttt{()}, and
+$\next~\hat{y}$ with $\hat{y}$. Injection tags are eliminated, because the sum
+type is at stage \bbone, so the tag is not relevant at stage \bbtwo. For
+functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
+replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.
 
-%As described above, splitting a stage \bbtwo\ value always produces a trivial
-%precomputation---either \texttt{()}, or nested tuples thereof.
+\paragraph{Splitting}
 
-%Thus, it is meaningful to define some notion of splitting for them.  This is
-%useful to study, because the definition will later be used as a contract with
-%which general stage \bbone\ term splitting must comply.
+Splitting stage \bbone\ terms, as implemented by the judgment $e\splitonesym
+[c,l.r]$, is also guided by the structure of the term $e$.
 
-%Recall that a partial value is externally a stage \bbone\ thing; it has a type
-%$@~\bbone$, and along with a future context, it is the result of reducing a
-%stage \bbone\ term. But internally, a partial value might contain stage \bbtwo\
-%subexpressions.
-
-\subsection{Stage \bbone\ Splitting}
-
-\crem{value splitting figure is missing injections}
-
-\TODO
-\begin{enumerate}
-\item Examples
-\item Lifting (and masking)
-\item Correctness? (Does this require partial value splitting?)
-\end{enumerate}
-
-\begin{theorem}[Correctness of splitting at $\bbone$]
-If $\vdash e:A~@~\bbone$, then (by splitting)
-\begin{itemize}
-\item $\vdash e : A \splitonesym [c,l.r]$
-\item $\reduce {c} {(v_1,u)}$
-\item $\reduce {(\letin{l}{u}{r})} v_S$
-\end{itemize}
-if and only if (by the staged dynamic semantics)
-\begin{itemize}
-\item $\diaone [] e {\xi;v}$
-\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
-\item $\reify \xi{v_2}q$
-\item $q \mathbin{\tworedsym} v_D$
-\end{itemize}
-and if so, then $v_D \sim v_S$.
-\end{theorem}
-
-Lifting and masking are two necessary ingredients for ensuring that splitting
-observes the stages of \lang. When splitting a partial value, the stage \bbone\
-result must \emph{mask} the stage \bbtwo\ computations in order to be
-single-staged. When splitting a stage \bbtwo\ term, the precomputation must
-\emph{lift} out all embedded stage \bbone\ computations, so they can be
-performed first, even though their results are only used in stage \bbtwo.
-%and it's accessible to whatever surrounding piece of code refers to the value being split.
-%is not accessible externally.
-
-Splitting a stage \bbone\ term, in contrast, requires both masking and lifting.
-Consider the term 
-\begin{lstlisting}
-1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
-\end{lstlisting}
-Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
-yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
-must occur at stage \bbone, the former is needed by stage \bbone, while the
-latter is needed by stage \bbtwo.
-
-The precomputation for this term is a \emph{combined result} which is
-the pair of these components: \verb|(((),3*4),1+2)|. The resumer is
-likewise the pair of the stage \bbtwo\ results of lifting and masking:
-\verb|fn L => (L<5,())|.
-%\footnote{In this example, the combined result is a 2-tuple, but more generally
-%it only must be a term which reduces to a 2-tuple.}
-
-Splitting is implemented by the judgment $\splitonesym$,
-the rules for which are given in \cref{fig:termSplitOne}.  
-It operates inductively on the structure of the term.  
-The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
-$c$ and resumer $l.r$.
+\TODO polish the following explanation
 
@@
 correct branch.
+
+\crem{split value splitting into two judgments, and call it masking}
+
+\crem{masking figure is missing injections}
+
+\crem{rename future context to residual table}
 

### 2014-11-04 paper/splitting.tex
more work on 4.1 and 4.3

@@
 computations can depend on stage \bbone\ computations, $r$ can depend on $p$ via
-the variable $l$. The correctness of splitting thus requires that $e$ is
-equivalent to $\letin l p r$.
-
-% TODO this contract works because a stage-2 term reduces to a value
-% so we can simply compare (p,l.r) to that value
+the variable $l$. The correctness of splitting thus requires that evaluating
+$\letin l p r$ agrees with the value to which $e$ itself evaluates.
 
@@
 
-\subsection{Eliminating Unnecessary Code}
+\subsection{Optimizations}
 
-\crem{don't say ``cost''}
+The above splitting rules are correct, but can be optimized in many cases. One
+major optimization is to eliminate redundant instances of \texttt{()}---for
+example, if multiple subterms have a \texttt{()} precomputation, these can be
+combined.
 
-While these splitting rules are all correct in terms of producing the right values, applying them naively can result in terms with needlessly large cost.
-Consider the following example involving the factorial function:
+A similar issue can occur in the precomputations for recursive function bodies.
+Consider the stage \bbtwo\ term
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-The boundary type\footnote{The splitting results are untyped, but we add in type annotations to aid the reader.} of this is give by the following datatype declaration:
-\begin{lstlisting}
-datatype prec = L | R of prec
-\end{lstlisting}
-The example then splits into the precomputation,
+which splits into:
 \begin{lstlisting}
@@
   else let (y,z) = fact(n-1) in (y*n,R z)
-in ((),fact 5)`
-\end{lstlisting}
-and the resumer
-\begin{lstlisting}
-2`l.
+in fact 5`
+
+2`l => 
 letfun fact (n : unit, l0 : prec) : unit = 
   case l0 of L => () | R l1 => fact ((),l0)
-in (fact (#2 l); #1 l)-100`
+in (fact (pi2 l); pi1 l)-100`
 \end{lstlisting}
-The problem here is that the naive system doesn't realize that the \texttt{fact} function has no stage \bbtwo\ content,
-and so it pessimistically retraces all of its steps in the resumer.  
-Fixing this requires some recursive reasoning.
+\texttt{fact}, being a stage \bbone\ function, itself splits into two recursive
+functions, which compute a combined result (stage \bbone\ result and
+precomputation), and a stage \bbtwo\ result, respectively. This precomputation
+could be given the type
+\begin{lstlisting}
+datatype prec = L | R of prec
+\end{lstlisting}
+while the stage \bbtwo\ result is always \texttt{()}.
 
-We instead tried to solve the issue by adding a single new staging annotation to \lang, called \texttt{mono}.  
-A \texttt{mono} block must appear in a stage \bbone\ context, and it indicates that its entire context is monostage.
-This can be encoded in our statics and dynamics by adding a new stage \bbmono\ and the judgements:
+It is wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
+always returns \texttt{()}; worse yet, it has linear runtime!
+
+We solve this issue by adding a new staging annotation to \lang, called
+\texttt{mono}. $\monoSt~e$ is stage \bbone, and requires that $e$ contains
+no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
+rules to \lang:
 \begin{mathpar}
@@
 
-The splitting rule for \texttt{mono} is more difficult.
-The problem is that the rules needs to produce a resumer with the correct shape
-for whatever the type of the \texttt{mono} is.
-This is difficult in the presence of sums and general recursion,
-and so we introduce a new value in our target language called dummy.
-Dummy can only appear in resumers and dynamically takes the shape of whatever is required of it.
+\crem{Why is it difficult?}
 
-The evaluation rules for dummy values are given by:
+To split \texttt{mono} terms, we must produce a resumer with the correct shape
+for the enclosed term. This is difficult to produce statically, because the
+language has sums and general recursion. We instead produce it dynamically,
+using a new value in \langmono, \texttt{dummy}, which can be consumed as though
+it had any type:
 \begin{mathpar}
@@
 
-Coming back to \texttt{mono}, the splitting rule is:
+\crem{Explain $\overset{C}{\to}$.}
+
+The splitting rules are then
 \begin{mathpar}
@@
 
-Our use of {\tt dummy} is safe because
-it can only appear in the resumers of terms that have no circles in their \lang\ type.
-Since the $\prev$ rule requires a circle type, dummy values cannot leak from
-the resumers of stage \bbone\ terms into the resumers of stage \bbtwo\ terms.
+To see that this optimization is correct, notice that \texttt{dummy} can only
+appear in the resumers of terms that have no circles in their \lang\ type.
+Since the $\prev$ rule requires a circle type, dummy values cannot leak from the
+resumers of stage \bbone\ terms into the resumers of stage \bbtwo\ terms.
 
 \input{figures/splitting-misc}
+
+\subsection{Implementation}
+
+\crem{Say something about the implementation.}

### 2014-11-05 paper/splitting.tex
comments entered

@@
 
+\ur{Why is there a free variable ``Gamma'' in the figure for splitting
+  stage 2?}
+
 Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
@@
 
+\ur{I would not say that ``The correctness of splitting requires'',
+  because we don't know what splitting is.  Instead, it is better to
+  state this actively, e.g., say that we wish to split the code such
+  that the condition e = let l = p in r holds.  correctness will have
+  to go beyond this, by saying for example that all stage 1 parts are
+  in p etc.  Note that this correctness allows a trivial split
+   that shoves everything to r, e.g., p = () and r is ``e''
+  (might need to flatten).}
+
 Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym
 [p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.
+
+\ur{At this point, I would have expected a brief discussion of the
+  easy cases, like the paragraph that starts with ``In general'' below
+  does. You can move that paragraph here and edit to play this role.
+  You can then move onto the example below after stating that more
+  interesting things happen with prev and hold.}
+
 For example, the stage \bbtwo\ term
@@
 containing stage \bbone\ computations, while the third has no stage \bbone\
-content. The precomputation \emph{lifts} these computations in the same tree
+content.
+
+\ur{The discussion of the subterms is not helpful. what is the point?}
+
+The precomputation \emph{lifts} these computations in the same tree
 structure: \verb|(1+2,(3+4,()))|.%
+\ur{Not clear what a ``tree structure'' is. }
 \footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
@@
 
+\ur{The above example should refer to Figure 11 (not Figure 10).  But
+  this is not mentioned.}
+
+\ur{This paragraph should come first, followed by a paragraph that
+  talks about Figure 11.}
 In general, constants and variables have no stage \bbone\ content, so their
@@
 
+\ur{So what goes wrong?  I have no idea what you are talking about at
+  this point.  The thought seems unfinished.}
+
+\ur{The following paragraph makes no sense to me, especially because
+  it does not match Figure 11.  AS far as I can see Figure 11 does not
+talk about masking but value splitting.}
 We formalize this notion via two \emph{masking} functions (\ref{fig:valSplit}),
@@
 
+\ur{The following paragraph makes no sense to me, especially because
+  it does not match Figure 11.}
+
 There is one final subtlety to splitting stage \bbone\ terms. In the example
@@
 a precomputation which is used by the computation of the \bbtwo-mask.
+
+\ur{This paragraph is unintellegible to me.  What is the point?}
 

### 2014-11-05 paper/semantics.tex
few additions to splitting
lots of easy changes in 3 and 4

@@
 So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
-This search process is called {\em speculation}.
+This search process is called {\em specialization}.
 In our example this means reducing all three projections to give,
@@
 \end{lstlisting}
-We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
+We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
 \begin{lstlisting} 
@@
 \end{lstlisting}
-Once again, we lift the results of speculation into a substitution:
+Once again, we lift the results of specialization into a substitution:
 \begin{lstlisting} 
@@
 
-\subsubsection{Evaluation and Speculation}
+\subsubsection{Evaluation and Specialization}
 \label{ssec:dynamics}
@@
 described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
-(speculation), and the auxiliary $\reifysym$ (reification). 
+(specialization), and the auxiliary $\reifysym$ (reification). 
 Along with these are three judgments identifying forms of values:
 $\pvalsym$ (partial values), $\ressym$ (residuals), 
-and the auxiliary $\fcon$ (future context).
+and the auxiliary $\fcon$ (residual table).
 These judgments are related as follows:
@@
 In order to avoid duplicating subterms, evaluation collects explicit
-substitutions into a single {\em future context} that is also the output of
+substitutions into a single {\em residual table} that is also the output of
 evaluation.
-\item Speculation sends a stage~\bbtwo\ term to a residual.
-\item Reification sends a residual and a future context to another residual.  
-\item Evaluation depends on speculation for $\next$ terms, which have
+\item Specialization sends a stage~\bbtwo\ term to a residual.
+\item Reification sends a single residual and a residual table to another residual.  
+\item Evaluation depends on specialization for $\next$ terms, which have
 stage~\bbtwo\ subterms.
-\item Speculation depends on evaluation for $\prev$ and $\pause$, which have
-stage~\bbone\ subterms. The future context produced by evaluation is also
+\item Specialization depends on evaluation for $\prev$ and $\pause$, which have
+stage~\bbone\ subterms. The residual table produced by evaluation is also
 reified here.
@@
 
-\ur{I would recommend using ``1-value'' and ``2-value'' instead of
-  partial values and residuals.  The former are more accurate, more
-  direct, more succinct, and does not lead to confusions with other
-  residuals.}
-
-\ur{Many incorrectly hyphenated ``partial values'' in this text.}
-
 The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
@@
 %
-to a {\em future environment} $\xi$ and a partial value $v$. The future
-environment implements the explicit substitutions in our example---it maps fresh
+to a {\em residual table} $\xi$ and a partial value $v$. The residual
+table implements the explicit substitutions in our example---it maps fresh
 stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
 residuals. For non-staging features of \lang, $\redonesym$ is essentially
-standard call-by-value evaluation, and gathers the subterms' future environments
+standard call-by-value evaluation, and gathers the subterms' residual tables
 into a single one.
@@
 
-At $\next~e$, in contrast, $\redonesym$ speculates into $e$ for stage~\bbone\ 
+At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
 subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
 which takes a stage \bbtwo\ term to a residual. 
-For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
+For the most part (\ref{fig:diaSemanticsSpec}), specialization simply recurs into
 all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
-which produces a future context $\xi$ and (by canonical forms) an expression
+which produces a residual table $\xi$ and (by canonical forms) an expression
 $\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
 bindings enclosing $\mathtt{\hat y}$.
-Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
-fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
-environment which maps that variable to the residual ($\hat y \mapsto q$).
+Once specialization reduces $e$ to a residual $q$, the output of $\redonesym$ is a
+fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a residual
+table which maps that variable to the residual ($\hat y \mapsto q$).
 
-Within speculation lies a subtle---if perhaps unintuitive---feature.  
-Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
+Within specialization lies a subtle---if perhaps unintuitive---feature.  
+Observe that specialization will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
 statement in its search for stage~\bbone\ code. 
@@
 will fail to evaluate at stage \bbone.
-This behavior is why the judgment is named ``speculation."
-Although it may seem undesirable, this speculative behavior is a deliberate feature.
+Although it may seem undesirable, this behavior is a deliberate feature.
 Indeed it is critical to the quickselect example, 

### 2014-11-05 paper/splitting.tex
lots of easy changes in 3 and 4

@@
 \subsection {Stage \bbtwo\ Splitting}
-
-\ur{Why is there a free variable ``Gamma'' in the figure for splitting
-  stage 2?}
 
@@
 talk about masking but value splitting.}
-We formalize this notion via two \emph{masking} functions (\ref{fig:valSplit}),
+We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
 $\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
@@
 second result the \bbtwo-mask.
-
-\ur{The following paragraph makes no sense to me, especially because
-  it does not match Figure 11.}
 
@@
 Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
-future context $\xi$, because the stage \bbtwo\ parts of $v$ contain variables
-in $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
+residual table $\xi$, because the stage \bbtwo\ parts of $v$ contain references
+to $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
 structure, and then reify the resulting term in the context $\xi$. This
@@
 have enough information at stage \bbone\ to know what branch to take, so there's
-no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
+no need to evaluate both sides. Instead, we evaluate the stage \bbone\ portion of the
 taken branch only, and inject the precomputation into a sum type. In the
@@
 correct branch.
-
-\crem{split value splitting into two judgments, and call it masking}
-
-\crem{masking figure is missing injections}
-
-\crem{rename future context to residual table}
 

### 2014-11-05 paper/splitting.tex
One quick elaboration in section 4.

@@
 and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
-computations can depend on stage \bbone\ computations, $r$ can depend on $p$ via
-the variable $l$. The correctness of splitting thus requires that evaluating
+computations can depend on its stage \bbone\ computations within $e$, 
+$r$ must depend on $p$, which splitting accomplishes via the variable $l$. 
+Given this form, the goal for splitting is to ensure that evaluating
 $\letin l p r$ agrees with the value to which $e$ itself evaluates.
 
-\ur{I would not say that ``The correctness of splitting requires'',
-  because we don't know what splitting is.  Instead, it is better to
-  state this actively, e.g., say that we wish to split the code such
-  that the condition e = let l = p in r holds.  correctness will have
+\ur{Correctness will have
   to go beyond this, by saying for example that all stage 1 parts are

### 2014-11-05 paper/semantics.tex
minor changes to 4

@@
   `1`let x = (next{`2`yhat`1`}, 7) in
-  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
+  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}`2`
 }`
@@
 2`prev {
-[yhat|->f 20]
-[zhat|->yhat * yhat * 7]
+[yhat|->f 20,
+ zhat|->yhat * yhat * 7]
     `1`next{`2`zhat`1`}`2`

### 2014-11-05 paper/splitting.tex
minor changes to 4

@@
 term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
-and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
-computations can depend on its stage \bbone\ computations within $e$, 
-$r$ must depend on $p$, which splitting accomplishes via the variable $l$. 
+and \bbtwo\ subcomputations of $e$, respectively. As stage \bbtwo\
+computations within $e$ can depend on its stage \bbone\ computations, 
+$r$ must depend on the results of $p$, which splitting accomplishes via the variable $l$. 
 Given this form, the goal for splitting is to ensure that evaluating
@@
 [p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.
-
-\ur{At this point, I would have expected a brief discussion of the
-  easy cases, like the paragraph that starts with ``In general'' below
-  does. You can move that paragraph here and edit to play this role.
-  You can then move onto the example below after stating that more
-  interesting things happen with prev and hold.}
-
 For example, the stage \bbtwo\ term
@@
 
-\ur{The above example should refer to Figure 11 (not Figure 10).  But
-  this is not mentioned.}
+\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
 
-\ur{This paragraph should come first, followed by a paragraph that
-  talks about Figure 11.}
 In general, constants and variables have no stage \bbone\ content, so their
@@
 underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends on
-splitting stage \bbone\ terms, which is covered later.
+splitting stage \bbone\ terms, and so they are covered later.
 

### 2014-11-06 paper/splitting.tex
Editing cut at 4.0 and the intro paragraph of 4.1.

@@
 Although terms in \lang\ clearly distinguish their stage \bbone\ and stage
-\bbtwo\ components, these components may be interleaved in complex ways. The
-goal of \emph{stage splitting} is to produce from a term in \lang\ a pair of
-terms in an unstaged language \langmono, which respectively perform all the
-stage \bbone\ and stage \bbtwo\ subcomputations of the original term.
+\bbtwo\ components, these components may be interleaved in \lang\ code. The
+goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
+of a term $e$ in \lang\ into a pair of
+terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
+stage \bbone\ subcomputations in $e$ and $e_2$ performs all of the stage \bbtwo\ subcomputations.
+Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same value 
+as evaluating $e$ in \lang.
 
-Since splitting stage \bbtwo\ terms is easier than splitting stage \bbone\
-terms, we begin by describing how to split stage \bbtwo\ terms.
+Since the mechanics of splitting stage \bbtwo\ terms are conceptually simpler than those for stage \bbone\
+terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.
 
@@
 
-Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
-term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
-and \bbtwo\ subcomputations of $e$, respectively. As stage \bbtwo\
-computations within $e$ can depend on its stage \bbone\ computations, 
-$r$ must depend on the results of $p$, which splitting accomplishes via the variable $l$. 
-Given this form, the goal for splitting is to ensure that evaluating
-$\letin l p r$ agrees with the value to which $e$ itself evaluates.
+Splitting a stage \bbtwo\ term $\coltwo e A$ generates
+ a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
+of $e$ in place to create a residual, splitting statically isolates these computations into
+$p$.  The resumer accepts the value produced by the precomputation as input
+% (bound to the free variable $l$)
+and executes all stage 2 subcomputations.   
+Specifically, if $e$ in \lang\ reduces to the final value $v$, 
+then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
+(The resumer, bound to the output of $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})
 
-\ur{Correctness will have
-  to go beyond this, by saying for example that all stage 1 parts are
-  in p etc.  Note that this correctness allows a trivial split
-   that shoves everything to r, e.g., p = () and r is ``e''
-  (might need to flatten).}
+%	\ur{Correctness will have
+%	  to go beyond this, by saying for example that all stage 1 parts are
+%	  in p etc.  Note that this correctness allows a trivial split
+%	   that shoves everything to r, e.g., p = () and r is ``e''
+%	  (might need to flatten).}
 

### 2014-11-06 paper/splitting.tex
Completed pass through 4.1

@@
 
-Although terms in \lang\ clearly distinguish their stage \bbone\ and stage
+Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
 \bbtwo\ components, these components may be interleaved in \lang\ code. The
@@
 terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
-stage \bbone\ subcomputations in $e$ and $e_2$ performs all of the stage \bbtwo\ subcomputations.
+stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
 Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same value 
@@
 
-Since the mechanics of splitting stage \bbtwo\ terms are conceptually simpler than those for stage \bbone\
+Since the mechanics of splitting stage~\bbtwo\ terms are conceptually simpler than those for stage \bbone\
 terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.
@@
 
-Splitting a stage \bbtwo\ term $\coltwo e A$ generates
+Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
  a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
@@
 % (bound to the free variable $l$)
-and executes all stage 2 subcomputations.   
+and executes all stage~\bbtwo\ subcomputations.   
 Specifically, if $e$ in \lang\ reduces to the final value $v$, 
@@
 
-Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym
-[p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.
-For example, the stage \bbtwo\ term
-\begin{lstlisting}
-2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
-\end{lstlisting}
-has two subterms, the latter of which itself has two subterms. Two are $\pause$s
-containing stage \bbone\ computations, while the third has no stage \bbone\
-content.
+Splitting stage~\bbtwo\ terms, as implemented by the judgment $e \splittwosym
+[p,l.r]$ in \cref{fig:termSplitTwo} is defined recursively on the structure of $e$.
+In the terminal cases of constants and variables, splitting generates precomputations that are \texttt{()}
+(these terms do not involve any stage~\bbone\ computations) and resumers that produce the original term's value.
+For example, the integer constant \texttt{5} splits into the precomputation \texttt{()} and \texttt{l=>5}.
 
-\ur{The discussion of the subterms is not helpful. what is the point?}
+For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
+The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
+The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
+Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
+underneath stage \bbtwo\ binders. % should probably draw a parallel to the same behavior in dynamics
 
-The precomputation \emph{lifts} these computations in the same tree
-structure: \verb|(1+2,(3+4,()))|.%
-\ur{Not clear what a ``tree structure'' is. }
-\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
-The resumer is a function which computes the stage \bbtwo\ value of the original
-term by mapping the results of the precomputation in place of the $\pause$s:
-\verb|(L1,(L2,_)) => L1 < (L2 - 5)|.
+The $\prev$ and $\pause$ rules depends on
+splitting stage \bbone\ terms, and so they are covered later.
 
-\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
-
-In general, constants and variables have no stage \bbone\ content, so their
-precomputations are \texttt{()}, and their resumers return the original term's
-value. Non-staging constructs with $n$ subterms have precomputations which are
-$n$-tuples of their subterms' precomputations, and resumers which pass each
-component to the resumer of the respective subterm. (This behavior at
-\texttt{case} and functions causes precomputations to be lifted out from
-underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends on
-splitting stage \bbone\ terms, and so they are covered later.
+%For example, the stage \bbtwo\ term
+%\begin{lstlisting}
+%2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
+%\end{lstlisting}
+%has two subterms, the latter of which itself has two subterms. Two are $\pause$s
+%containing stage \bbone\ computations, while the third has no stage \bbone\
+%content. 
+%
+%\ur{The discussion of the subterms is not helpful. what is the point?}
+%
+%The precomputation \emph{lifts} these computations in the same tree
+%structure: \verb|(1+2,(3+4,()))|.%
+%\ur{Not clear what a ``tree structure'' is. }
+%\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
+%The resumer is a function which computes the stage \bbtwo\ value of the original
+%term by mapping the results of the precomputation in place of the $\pause$s:
+%\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
+%
+%\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
+%
+%Non-staging constructs with $n$ subterms have precomputations which are
+%$n$-tuples of their subterms' precomputations, and resumers which pass each
+%component to the resumer of the respective subterm.
 

### 2014-11-06 paper/splitting.tex
Setting up why the definition of Stage 1 splitting is what it is.

@@
 
-Splitting a stage \bbone\ term $\colone e A$ produces a term and a function,
-which respectively contain all the stage \bbone\ and stage \bbtwo\
-subcomputations of $e$. The situation is more complex than when splitting stage
-\bbtwo\ terms; to see why, consider the correctness criterion for splitting
-stage \bbone\ terms.
-
-Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
-yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
-value $v$. It is tempting to compare the two results of evaluating $e$ to the
-two results of splitting it. To see where this goes wrong, consider the stage
-\bbone\ term
+The goal of stage 2 splitting was to generate a precomputation and resumer such that the resumer and the \lang\ residual reduce to the same final value.
+However, stage~\bbone\ terms do not reduce to final values, rather they reduce to residual tables and partial values.
+For example, consider the stage 1 term:
 \begin{lstlisting}
@@
 \end{lstlisting}
-which evaluates to
+which evaluates (via $\mathbin{\redonesym}$), to the following residual table and a multi-stage tuple:
+
 \begin{lstlisting}
@@
 \end{lstlisting}
-This partial value contains a $\next$, and so is itself multistaged:
-intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
-is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5}.
 
-\ur{So what goes wrong?  I have no idea what you are talking about at
-  this point.  The thought seems unfinished.}
+%Like stage \bbtwo\ splitting, splitting a stage~\bbone\ term $\colone e A$ produces a term $c$ containing all stage~\bbone\ subcomputations in $e$ and a resumer function %$l.r$ containing all stage~\bbtwo\ subcomputations.
+%which respectively contain all the stage \bbone\ and stage \bbtwo\
+%subcomputations of $e$. 
+%But where the stage \bbtwo\ splitting goal was phrased terms ensure.
+
+%The situation is more complex than when splitting stage
+%\bbtwo\ terms; to see why, consider the correctness criterion for splitting
+%stage \bbone\ terms.
+
+%Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
+%yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
+%value $v$.
+
+Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
+
+%It is tempting to compare the two results of evaluating $e$ to the
+%two results of splitting it.
+
+%To see where this goes wrong, consider the stage
+%\bbone\ term
+%\begin{lstlisting}
+%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+%\end{lstlisting}
+%which evaluates to
+%\begin{lstlisting}
+%2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
+%\end{lstlisting}
+%This partial value contains a $\next$, and so is itself multistaged:
+%intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
+%is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5} in the table.
 

### 2014-11-06 paper/splitting.tex
setting up intro to masking definitions

@@
 
-Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
+Note that this result carries information at {\em both} stages: 
+at stage~\bbone\ it carries the value 3, and at stage~\bbtwo\ it carries the code $\letin \yhat {\mathtt{7+5}} \yhat$.
+
+But the whole point of splitting is that its results (and their evaluated forms) are monostage.
+Ergo, if we intend to compare the results of evaluation to the those of splitting, 
+then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.
+
+% Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
+To this end, we introduce two {\em masking} functions...
 
@@
 
-\ur{The following paragraph makes no sense to me, especially because
-  it does not match Figure 11.  AS far as I can see Figure 11 does not
-talk about masking but value splitting.}
 We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),

### 2014-11-06 paper/splitting.tex
Just made the transition to motivating the definition of splitting in 4.2.

@@
 stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
-Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same value 
+Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
 as evaluating $e$ in \lang.
@@
 then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
-(The resumer, bound to the output of $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})
+(Thus, the resumer, bound to the value output by $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})
 
@@
 Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
-underneath stage \bbtwo\ binders. % should probably draw a parallel to the same behavior in dynamics
+underneath stage \bbtwo\ binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
 
@@
 at stage~\bbone\ it carries the value 3, and at stage~\bbtwo\ it carries the code $\letin \yhat {\mathtt{7+5}} \yhat$. 
+Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms to create a resumer that reduces to the same value as the stage~\bbtwo\ carried by the partial value, and an $e1$ that not only performs the precomputation necessary for the resumer, but also generates values that match the stage~\bbone\ aspects of the partial value.
 
-But the whole point of splitting is that its results (and their evaluated forms) are monostage.
-Ergo, if we intend to compare the results of evaluation to the those of splitting, 
-then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.
+% Ergo, if we intend to compare the results of evaluation to the those of splitting, 
+% then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.
+
+% But the whole point of splitting is that its results (and their evaluated forms) are monostage.
 
 % Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
-To this end, we introduce two {\em masking} functions...
+
+To concretize what it means by ``match", we introduce two {\em masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the parts that are relevant at a particular stage.
+These functions are shown in \ref{fig:valMask}.
+They operate by replacing all stage \bbone\
+(resp., \bbtwo) components of a partial value by \texttt{()}.
 
@@
 
-We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
-$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
-a partial value at each stage, essentially by replacing all stage \bbone\
-(resp., \bbtwo) components of a partial value by \texttt{()}. The first result
+%We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
+%$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
+%a partial value at each stage, essentially by replacing all stage \bbone\
+%(resp., \bbtwo) components of a partial value by \texttt{()}.
+
+The first result
 of splitting must then compute the \bbone-mask of the partial value, and the
@@
 
-There is one final subtlety to splitting stage \bbone\ terms. In the example
-above, there are \emph{two} flavors of stage \bbone\ subcomputations:
-\texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
-a precomputation which is used by the computation of the \bbtwo-mask.
-
-\ur{This paragraph is unintellegible to me.  What is the point?}
+% There is one final subtlety to splitting stage \bbone\ terms. In the example
+% above, there are \emph{two} flavors of stage \bbone\ subcomputations:
+% \texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
+% a precomputation which is used by the computation of the \bbtwo-mask.
 

### 2014-11-06 paper/splitting.tex
First attempt at masking.

@@
 
-\subsection {Stage \bbtwo\ Splitting}
+\subsection {Splitting Stage \bbtwo\ Terms}
 
@@
 
-\subsection{Stage \bbone\ Splitting}
+\subsection{Splitting Stage \bbone\ Terms}
 
@@
 
-Note that this result carries information at {\em both} stages: 
-at stage~\bbone\ it carries the value 3, and at stage~\bbtwo\ it carries the code $\letin \yhat {\mathtt{7+5}} \yhat$. 
-Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms to create a resumer that reduces to the same value as the stage~\bbtwo\ carried by the partial value, and an $e1$ that not only performs the precomputation necessary for the resumer, but also generates values that match the stage~\bbone\ aspects of the partial value.
+Note that this result simultaneously represents information at {\em both} stages: 
+a stage~\bbone\ value 3, and the stage~\bbtwo\ computation $\letin \yhat {\mathtt{7+5}} \yhat$. 
+Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms in \lang\ into a resumer that reduces to the same value as the stage~\bbtwo\ computation, and an $e1$ that not only performs the precomputation expected by the resumer, but also computes values that match the stage~\bbone\ aspects of the partial value.
 
@@
 
-To concretize what it means by ``match", we introduce two {\em masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the parts that are relevant at a particular stage.
-These functions are shown in \ref{fig:valMask}.
-They operate by replacing all stage \bbone\
-(resp., \bbtwo) components of a partial value by \texttt{()}.
+\subsubsection{Partial Value Masking}
+
+To concretize the meaning of ``match", we introduce two {\em partial value masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the subterms that are relevant at a particular stage.
+The masking functions are defined in \ref{fig:valMask}.
+
+Since \texttt{next}'s only contain references to the residual table (computations that occur in stage~\bbtwo),
+the primary operation of $\masko{\xi;v}$ is to replace all $\next$s in $v$ with \texttt{()}.
+The result is a \langmono\ value with the same structure as $v$, but containing only its stage~\bbone\ information.
+Masking is more complex when $v$ is a function since the function's body itself may contain stage~\bbtwo\ subterms.
+As a result, masking of functions depends on the general behavior of stage~\bbone\ splitting, described shortly.
+
+While $\masko{\cdot}$ produces a value in \langmono,
+$\maskt{\xi;v}$ produces a \emph{term} in \langmono\ that corresponds to the stage~\bbtwo\ computation represented by $[\xi;v]$.
+Stage~\bbone\ information in $v$ is discarded by replacing base constants with \texttt{()}, replacing $\next~\hat{y}$ with $\hat{y}$, and eliminating all injection tags.
+(The sum type is at stage \bbone, so the tag is not relevant at stage \bbtwo.)
+Last, $\maskt{\cdot}$ reifies $\xi$ around the modified $v$ to produce a term in \langmono. 
+As with $\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on general term splitting.
+
+% For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
+% replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.
+
+%The only exception is for
+%functions $\lam xAe$; such a value represents a computation, so masking splits
+%its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
+%stage \bbone\ component, the combined result.
+
+%They operate by replacing all stage \bbone\
+%(resp., \bbtwo) components of a partial value by \texttt{()}.
 
@@
 %(resp., \bbtwo) components of a partial value by \texttt{()}.
+
+\subsubsection{Splitting}
 

### 2014-11-06 paper/splitting.tex
First cut at the start of Stage 2 splitting.

@@
 stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
-Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
-as evaluating $e$ in \lang.
+Informally, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
+as evaluating $e$ in \lang. (We make this intuition precise in the subsequent subsections.)
 
@@
 
-\subsubsection{Splitting}
+\subsubsection{Splitting Mechanics}
 
-The first result
-of splitting must then compute the \bbone-mask of the partial value, and the
-second result the \bbtwo-mask.
+Given the definition of masking, splitting a stage \bbone\ term $\colone e A$ 
+(where $e\mathbin{\redonesym}[\xi;v]$), yields a \emph{combined term}
+$c$ that performs all stage~\bbone\ computations in $e$ and a \emph{resumer} function $l.r$ that represents all stage~\bbtwo\ computations.
+Whereas in stage~\bbtwo\ term splitting,
+$p$ constitutes only the precomputation to the resumer, in stage~\bbone\ splitting we require $c$ to reduce (via~$\redsym$) to the tuple $(y,z)$, where $y$ is an \emph{immediate result} that is identical to $\masko{\xi;v}$, and $z$ is the required precomputation such that $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
 
-% There is one final subtlety to splitting stage \bbone\ terms. In the example
-% above, there are \emph{two} flavors of stage \bbone\ subcomputations:
-% \texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
-% a precomputation which is used by the computation of the \bbtwo-mask.
+Splitting stage~\bbone\ terms is implemented by the judgment $e \splitonesym [c,l.r]$ in \ref{fig:termSplitOne}.
 
-Formally, splitting a stage \bbone\ term, as implemented by the judgment $e
-\splitonesym [c,l.r]$ (\ref{fig:termSplitOne}), yields a \emph{combined result}
-$c$ which computes a pair of the stage \bbone\ result and the precomputation,
-and a \emph{resumer} function $l.r$. The correctness of splitting requires that,
-for $e\mathbin{\redonesym}[\xi;v]$, we have that $\pio~c$ agrees with
-$\masko{v}$, and $\letin l{\pit~c}r$ agrees with $\maskt{\xi;v}$.
+% The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
+% resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
+% which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
+% \texttt{((),12)}, which agrees with the stage \bbtwo\ mask.
 
-The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
-resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
-which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
-\texttt{((),12)}, which agrees with the stage \bbtwo\ mask.
-
-\paragraph{Stage \bbone\ masking}
-
-The purpose of masking is to expose only the relevant structure of a partial
-value $v$ at each stage. For the most part, stage \bbone\ masking, $\masko{v}$,
-simply replaces all $\next$s in $v$ with \texttt{()}. The only exception is for
-functions $\lam xAe$; such a value represents a computation, so masking splits
-its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
-stage \bbone\ component, the combined result.
-
-\paragraph{Stage \bbtwo\ masking}
-
-Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
-residual table $\xi$, because the stage \bbtwo\ parts of $v$ contain references
-to $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
-structure, and then reify the resulting term in the context $\xi$. This
-rewriting process replaces base type constants in $v$ with \texttt{()}, and
-$\next~\hat{y}$ with $\hat{y}$. Injection tags are eliminated, because the sum
-type is at stage \bbone, so the tag is not relevant at stage \bbtwo. For
-functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
-replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.
-
-\paragraph{Splitting}
-
-Splitting stage \bbone\ terms, as implemented by the judgment $e\splitonesym
-[c,l.r]$, is also guided by the structure of the term $e$.
-
-\TODO polish the following explanation
+\TODO REDO THE REST OF THE TEXT TO WALK THROUGH THE MAJOR RULES OF SPLITTING
 

### 2014-11-06 paper/splitting.tex
pass over splitting

@@
 
-Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
-\bbtwo\ components, these components may be interleaved in \lang\ code. The
-goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
-of a term $e$ in \lang\ into a pair of
-terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
-stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
-Informally, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
-as evaluating $e$ in \lang. (We make this intuition precise in the subsequent subsections.)
+Given a term in the source language \lang\ where terms of
+stage~\bbone\ and stage-\bbtwo\ may be interleaved, \emph{(stage)
+  splitting} separates the stage-\bbone\ and stage-\bbtwo\
+subcomputations into two separate terms, each performing the
+subcomputations corresponding to each stage.  More precisely, given a
+\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
+unstaged target language \langmono\ such that~$e_1$ performs all the
+stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
+stage~\bbtwo\ subcomputations.
 
-Since the mechanics of splitting stage~\bbtwo\ terms are conceptually simpler than those for stage \bbone\
-terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.
+\ur{Technically, target language is not lambda but lambda plus pairs.
+  Please use some other symbol than lambda, lambda-p lambda+ or
+  something like that, or simply:} \textsf{\bf \textsc{Mono}}.
 
-\subsection {Splitting Stage \bbtwo\ Terms}
+Splitting preserves correctness in the sense that evaluation of~$e$ in
+the staged language produces the same result as evaluating~$e_1$ and
+then~$e_2$ with the result from~$e_1$.  In this section, we present a
+splitting algorithm for the source language \lang\, specifying
+precisely the algorithm and the correctness condition.  We start by
+defining a masking relation that helps us to isolate stage-1 and
+stage-2 components of a partial value and then move on to the
+specification of the splitting algorithms.
 
-Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
- a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
-of $e$ in place to create a residual, splitting statically isolates these computations into
-$p$.  The resumer accepts the value produced by the precomputation as input
-% (bound to the free variable $l$)
-and executes all stage~\bbtwo\ subcomputations.   
-Specifically, if $e$ in \lang\ reduces to the final value $v$, 
-then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
-(Thus, the resumer, bound to the value output by $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})
+\subsection{Partial Value Masking}
+
+{\em Partial-value masking} or simply {\em masking} are a functions
+for each stage that project from the domain of a partial value and
+residual table to the a term in the target language \langmono.
+
+\ur{I am attempting to define this even though it is not clear from
+  the text what it is.  The definition should specify precisely the
+  domain and the range of the functions.  Does the domain include the
+  table?  The figure does not.  }
+
+\ref{fig:valMask} defines the functions, $\masko{\cdot}$ and
+$\maskt{\cdot}$. Since \texttt{next}'s only contain references to the
+residual table (computations that occur in stage~\bbtwo),
+$\masko{\xi;v}$ is to replaces all~$\next$s in~$v$ with~\texttt{()}.
+The result is a \langmono\ value with the same structure as~$v$, but
+containing only its stage~\bbone\ information. Masking is more complex
+when~$v$ is a function since the function's body itself may contain
+stage~\bbtwo\ subterms.  As a result, masking of functions depends on
+the general behavior of stage~\bbone\ splitting, described shortly.
+
+While $\masko{\cdot}$ produces a value in \langmono, $\maskt{\xi;v}$
+produces a term in \langmono\ that corresponds to the stage~\bbtwo\
+computation represented by~$[\xi;v]$.  Stage~\bbone\ information
+in~$v$ is discarded by replacing base constants with~\texttt{()},
+replacing $\next~\hat{y}$ with~$\hat{y}$, and eliminating all
+injection tags. (The sum type is at stage \bbone, so the tag is not
+relevant at stage \bbtwo.)  Last, $\maskt{\cdot}$ reifies~$\xi$ around
+the modified~$v$ to produce a term in \langmono.  As with
+$\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on
+general term splitting.
+
+\subsection{Splitting Algorithm}
+
+\paragraph{Splitting of stage-\bbone\ terms.}
+
+\ur{Fill this in.}
+
+OUTLINE:
+
+\begin{itemize}
+\item 
+The structure of the splitting judgment.
+
+\item
+The equivalence condition.
+
+\item
+Walk through the cases.
+\end{itemize}
+
+\paragraph{Splitting stage-\bbtwo\ terms.}
+
+Splitting a stage~\bbtwo\ term $\coltwo e A$ generates a
+\emph{precomputation}~$p$, and a~\emph{resumer} function $l.r$.  The
+resumer accepts the value produced by the precomputation as input
+($l$) and performs the stage~\bbtwo\ subcomputations of~$e$.
+Specifically, if~$e$ in~\lang\ reduces to the value~$v$, then after
+splitting the term $\letin l p r$ also reduces to~$v$ in \langmono.
+
+\ur{I don't quite understand the following.}
+(Thus, the resumer, bound to the value output by $p$, is equivalent to
+the residual term defined in~\ref{ssec:dynamics})
 
@@
 
-Splitting stage~\bbtwo\ terms, as implemented by the judgment $e \splittwosym
-[p,l.r]$ in \cref{fig:termSplitTwo} is defined recursively on the structure of $e$.
-In the terminal cases of constants and variables, splitting generates precomputations that are \texttt{()}
-(these terms do not involve any stage~\bbone\ computations) and resumers that produce the original term's value.
-For example, the integer constant \texttt{5} splits into the precomputation \texttt{()} and \texttt{l=>5}.
+The splitting algorithm for stage~\bbtwo\ terms, as specified by the
+judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
+recursively on the structure of~$e$.  In the terminal cases of
+constants and variables, splitting generates precomputations that are
+\texttt{()}, because these terms do not involve any stage~\bbone\
+computations, and resumers that produce the original term. For
+example, the integer constant \texttt{5} splits into the
+precomputation \texttt{()} and \texttt{l=>5}.
 
-For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
-The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
-The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
-Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
-underneath stage \bbtwo\ binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
+For all non-terminal stage~\bbtwo\ terms~$e$ in \lang\ (except $\prev$
+and $\pause$), splitting descends into~$e$, recursively splitting
+subterms and generating precomputations $p_1, \ldots, p_n$ and
+resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
+The precompuation of~$e$ is defined as the combined precomputations:
+$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
+argument $(l_1,\ldots,l_n)$ in a term that has the same structure
+of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
+Notice that splitting \texttt{case}s and functions causes the
+precomputation of subterms to be lifted out from underneath stage
+\bbtwo\
+binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
+\ur{This last point seems important.  It would be important to come
+  back to it.}
 
@@
 \subsection{Splitting Stage \bbone\ Terms}
+
+\ur{What follows is still a bit vague to me. Some
+  suggestions/questions: are ``term'' and ``computation'' being used
+  interchangeably? Not sure I understand this.  Isn't ``multi-stage''
+  tuple a partial-value?}
 
@@
 
-\subsubsection{Partial Value Masking}
-
-To concretize the meaning of ``match", we introduce two {\em partial value masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the subterms that are relevant at a particular stage.
-The masking functions are defined in \ref{fig:valMask}.
-
-Since \texttt{next}'s only contain references to the residual table (computations that occur in stage~\bbtwo),
-the primary operation of $\masko{\xi;v}$ is to replace all $\next$s in $v$ with \texttt{()}.
-The result is a \langmono\ value with the same structure as $v$, but containing only its stage~\bbone\ information.
-Masking is more complex when $v$ is a function since the function's body itself may contain stage~\bbtwo\ subterms.
-As a result, masking of functions depends on the general behavior of stage~\bbone\ splitting, described shortly.
-
-While $\masko{\cdot}$ produces a value in \langmono,
-$\maskt{\xi;v}$ produces a \emph{term} in \langmono\ that corresponds to the stage~\bbtwo\ computation represented by $[\xi;v]$.
-Stage~\bbone\ information in $v$ is discarded by replacing base constants with \texttt{()}, replacing $\next~\hat{y}$ with $\hat{y}$, and eliminating all injection tags.
-(The sum type is at stage \bbone, so the tag is not relevant at stage \bbtwo.)
-Last, $\maskt{\cdot}$ reifies $\xi$ around the modified $v$ to produce a term in \langmono. 
-As with $\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on general term splitting.
-
 % For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
@@
 
+% % Umut: I made a pass over this, see above.  I don't think I lost
+% % anything, so it should be fine to delete this.
+% % \section{Splitting Algorithm}
+% % \label{sec:splitting}
+% % Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
+% % \bbtwo\ components, these components may be interleaved in \lang\ code. The
+% % goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
+% % of a term $e$ in \lang\ into a pair of
+% % terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
+% % stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
+% % Informally, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
+% % as evaluating $e$ in \lang. (We make this intuition precise in the subsequent subsections.)
+
+% % Since the mechanics of splitting stage~\bbtwo\ terms are conceptually simpler than those for stage \bbone\
+% % terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.
+
+% % \subsection {Splitting Stage \bbtwo\ Terms}
+
+% % Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
+% %  a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
+% % of $e$ in place to create a residual, splitting statically isolates these computations into
+% % $p$.  The resumer accepts the value produced by the precomputation as input
+% % % (bound to the free variable $l$)
+% % and executes all stage~\bbtwo\ subcomputations.   
+% % Specifically, if $e$ in \lang\ reduces to the final value $v$, 
+% % then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
+% % (Thus, the resumer, bound to the value output by $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})
+
+% % %	\ur{Correctness will have
+% % %	  to go beyond this, by saying for example that all stage 1 parts are
+% % %	  in p etc.  Note that this correctness allows a trivial split
+% % %	   that shoves everything to r, e.g., p = () and r is ``e''
+% % %	  (might need to flatten).}
+
+% % Splitting stage~\bbtwo\ terms, as implemented by the judgment $e \splittwosym
+% % [p,l.r]$ in \cref{fig:termSplitTwo} is defined recursively on the structure of $e$.
+% % In the terminal cases of constants and variables, splitting generates precomputations that are \texttt{()}
+% % (these terms do not involve any stage~\bbone\ computations) and resumers that produce the original term's value.
+% % For example, the integer constant \texttt{5} splits into the precomputation \texttt{()} and \texttt{l=>5}.
+
+% % For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
+% % The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
+% % The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
+% % Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
+% % underneath stage \bbtwo\ binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
+
+% % The $\prev$ and $\pause$ rules depends on
+% % splitting stage \bbone\ terms, and so they are covered later.
+
+% % %For example, the stage \bbtwo\ term
+% % %\begin{lstlisting}
+% % %2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
+% % %\end{lstlisting}
+% % %has two subterms, the latter of which itself has two subterms. Two are $\pause$s
+% % %containing stage \bbone\ computations, while the third has no stage \bbone\
+% % %content. 
+% % %
+% % %\ur{The discussion of the subterms is not helpful. what is the point?}
+% % %
+% % %The precomputation \emph{lifts} these computations in the same tree
+% % %structure: \verb|(1+2,(3+4,()))|.%
+% % %\ur{Not clear what a ``tree structure'' is. }
+% % %\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
+% % %The resumer is a function which computes the stage \bbtwo\ value of the original
+% % %term by mapping the results of the precomputation in place of the $\pause$s:
+% % %\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
+% % %
+% % %\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
+% % %
+% % %Non-staging constructs with $n$ subterms have precomputations which are
+% % %$n$-tuples of their subterms' precomputations, and resumers which pass each
+% % %component to the resumer of the respective subterm.
+
+% % \subsection{Splitting Stage \bbone\ Terms}
+
+% % The goal of stage 2 splitting was to generate a precomputation and resumer such that the resumer and the \lang\ residual reduce to the same final value.
+However, stage~\bbone\ terms do not reduce to final values, rather they reduce to residual tables and partial values.
+For example, consider the stage 1 term:
+% % \begin{lstlisting}
+% % 1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+% % \end{lstlisting}
+% % which evaluates (via $\mathbin{\redonesym}$), to the following residual table and a multi-stage tuple:
+
+% % \begin{lstlisting}
+% % 2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
+% % \end{lstlisting}
+
+% % %Like stage \bbtwo\ splitting, splitting a stage~\bbone\ term $\colone e A$ produces a term $c$ containing all stage~\bbone\ subcomputations in $e$ and a resumer function %$l.r$ containing all stage~\bbtwo\ subcomputations.
+% % %which respectively contain all the stage \bbone\ and stage \bbtwo\
+% % %subcomputations of $e$. 
+% % %But where the stage \bbtwo\ splitting goal was phrased terms ensure.
+
+% % %The situation is more complex than when splitting stage
+% % %\bbtwo\ terms; to see why, consider the correctness criterion for splitting
+% % %stage \bbone\ terms.
+
+% % %Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
+% % %yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
+% % %value $v$.
+
+% % Note that this result simultaneously represents information at {\em both} stages: 
+% % a stage~\bbone\ value 3, and the stage~\bbtwo\ computation $\letin \yhat {\mathtt{7+5}} \yhat$. 
+% % Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms in \lang\ into a resumer that reduces to the same value as the stage~\bbtwo\ computation, and an $e1$ that not only performs the precomputation expected by the resumer, but also computes values that match the stage~\bbone\ aspects of the partial value.
+
+% % % Ergo, if we intend to compare the results of evaluation to the those of splitting, 
+% % % then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.
+
+% % % But the whole point of splitting is that its results (and their evaluated forms) are monostage.
+
+% % % Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
+
+% % \subsubsection{Partial Value Masking}
+
+% % To concretize the meaning of ``match", we introduce two {\em partial value masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the subterms that are relevant at a particular stage.
+The masking functions are defined in \ref{fig:valMask}.
+
+% % Since \texttt{next}'s only contain references to the residual table (computations that occur in stage~\bbtwo),
+% % the primary operation of $\masko{\xi;v}$ is to replace all $\next$s in $v$ with \texttt{()}.
+The result is a \langmono\ value with the same structure as $v$, but containing only its stage~\bbone\ information.
+Masking is more complex when $v$ is a function since the function's body itself may contain stage~\bbtwo\ subterms.
+As a result, masking of functions depends on the general behavior of stage~\bbone\ splitting, described shortly.
+
+% % While $\masko{\cdot}$ produces a value in \langmono,
+% % $\maskt{\xi;v}$ produces a \emph{term} in \langmono\ that corresponds to the stage~\bbtwo\ computation represented by $[\xi;v]$.
+% % Stage~\bbone\ information in $v$ is discarded by replacing base constants with \texttt{()}, replacing $\next~\hat{y}$ with $\hat{y}$, and eliminating all injection tags.
+(The sum type is at stage \bbone, so the tag is not relevant at stage \bbtwo.)
+% % Last, $\maskt{\cdot}$ reifies $\xi$ around the modified $v$ to produce a term in \langmono. 
+% % As with $\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on general term splitting.
+
+% % % For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
+% % % replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.
+
+% % %The only exception is for
+% % %functions $\lam xAe$; such a value represents a computation, so masking splits
+% % %its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
+% % %stage \bbone\ component, the combined result.
+
+% % %They operate by replacing all stage \bbone\
+% % %(resp., \bbtwo) components of a partial value by \texttt{()}.
+
+% % %It is tempting to compare the two results of evaluating $e$ to the
+% % %two results of splitting it.
+
+% % %To see where this goes wrong, consider the stage
+% % %\bbone\ term
+% % %\begin{lstlisting}
+% % %1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+% % %\end{lstlisting}
+% % %which evaluates to
+% % %\begin{lstlisting}
+% % %2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
+% % %\end{lstlisting}
+% % %This partial value contains a $\next$, and so is itself multistaged:
+% % %intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
+% % %is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5} in the table.
+
+% % %We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
+% % %$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
+% % %a partial value at each stage, essentially by replacing all stage \bbone\
+% % %(resp., \bbtwo) components of a partial value by \texttt{()}.
+
 \subsubsection{Splitting Mechanics}
+
+\ur{It starts getting rough here, mostly because combined term is not
+  defined.  Specifically, we should say what
+  the  form of the splitting judgment is. }
 

### 2014-11-06 paper/splitting.tex
pass over splitting
fixed figures
some work with nico

@@
 Given a term in the source language \lang\ where terms of
-stage~\bbone\ and stage-\bbtwo\ may be interleaved, \emph{(stage)
-  splitting} separates the stage-\bbone\ and stage-\bbtwo\
+stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
+  splitting} separates the stage~\bbone\ and stage~\bbtwo\
 subcomputations into two separate terms, each performing the
@@
 
-\ur{Technically, target language is not lambda but lambda plus pairs.
-  Please use some other symbol than lambda, lambda-p lambda+ or
-  something like that, or simply:} \textsf{\bf \textsc{Mono}}.
-
 Splitting preserves correctness in the sense that evaluation of~$e$ in
@@
 precisely the algorithm and the correctness condition.  We start by
-defining a masking relation that helps us to isolate stage-1 and
-stage-2 components of a partial value and then move on to the
+defining a masking relation that helps us to isolate stage~\bbone\ and
+stage~\bbtwo\ components of a partial value and then move on to the
 specification of the splitting algorithms.
@@
 
-{\em Partial-value masking} or simply {\em masking} are a functions
-for each stage that project from the domain of a partial value and
-residual table to the a term in the target language \langmono.
+Intuitively, we require that the terms obtained by splitting a term $e$, when
+evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
+the case of stage~\bbone\ terms, our dynamic semantics produce partial values
+containing information at \emph{both} stages, which goes against the idea that
+splitting separates the subcomputations belonging to each stage.
 
-\ur{I am attempting to define this even though it is not clear from
-  the text what it is.  The definition should specify precisely the
-  domain and the range of the functions.  Does the domain include the
-  table?  The figure does not.  }
+For example, consider the stage 1 term
+\begin{lstlisting}
+1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+\end{lstlisting}
+which evaluates via $\redonesym$ to a residual table and multi-stage tuple
+\begin{lstlisting}
+2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
+\end{lstlisting}
+This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
+computation $\letin \yhat {\mathtt{7+5}} \yhat$.
 
-\ref{fig:valMask} defines the functions, $\masko{\cdot}$ and
-$\maskt{\cdot}$. Since \texttt{next}'s only contain references to the
-residual table (computations that occur in stage~\bbtwo),
-$\masko{\xi;v}$ is to replaces all~$\next$s in~$v$ with~\texttt{()}.
-The result is a \langmono\ value with the same structure as~$v$, but
-containing only its stage~\bbone\ information. Masking is more complex
-when~$v$ is a function since the function's body itself may contain
-stage~\bbtwo\ subterms.  As a result, masking of functions depends on
-the general behavior of stage~\bbone\ splitting, described shortly.
+To clearly delineate these parts of the partial value, we define 
+\emph{partial-value masking}, or simply {\em masking}, functions at each stage,
+which compute from a residual table and partial value the content (expressed as
+a term in \langmono) belonging to each stage. \ref{fig:valMask} defines these
+masking functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$.
 
-While $\masko{\cdot}$ produces a value in \langmono, $\maskt{\xi;v}$
-produces a term in \langmono\ that corresponds to the stage~\bbtwo\
-computation represented by~$[\xi;v]$.  Stage~\bbone\ information
-in~$v$ is discarded by replacing base constants with~\texttt{()},
-replacing $\next~\hat{y}$ with~$\hat{y}$, and eliminating all
-injection tags. (The sum type is at stage \bbone, so the tag is not
-relevant at stage \bbtwo.)  Last, $\maskt{\cdot}$ reifies~$\xi$ around
-the modified~$v$ to produce a term in \langmono.  As with
-$\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on
-general term splitting.
+The goal of $\masko{v}$ is to produce a term in \langmono\ containing exactly
+the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
+content, we replace all $\next$ subterms of $v$ with \texttt{()}.  For lambdas,
+since these represent potentially-multistage computations, we split their bodies
+as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and return
+the stage~\bbone\ component. $\masko{v}$ is always a value in \langmono, and has
+the same structure as~$v$.
 
-\subsection{Splitting Algorithm}
+The goal of $\maskt{\xi;v}$ is to produce a term in \langmono\ containing
+exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike in stage \bbone\ masking,
+here we need access to the residual table $\xi$, because all stage~\bbtwo\
+content in $v$ has already been lifted out into $\xi$. To discard all
+stage~\bbone\ information in $v$, we replace base constants with~\texttt{()},
+replace $\next~\hat{y}$ with~$\hat{y}$, and replace all $n$-ary term
+constructors with an $n$-tuple of their subterms, themselves masked. As before,
+we mask functions by splitting their bodies as general stage~\bbone\ terms, and
+use the stage~\bbtwo\ component.
 
-\paragraph{Splitting of stage-\bbone\ terms.}
+After performing this transformation on $v$, we reify $\xi$ around the result,
+producing a term in \langmono.
+
+\subsection{Splitting Stage \bbone\ Terms}
+\label{sec:split-one}
 
@@
 
-OUTLINE:
+FOLLOW OUTLINE:
 
 \begin{itemize}
-\item 
-The structure of the splitting judgment.
-
-\item
-The equivalence condition.
-
-\item
-Walk through the cases.
+\item The structure of the splitting judgment.
+\item The equivalence condition.
+\item Walk through the cases.
 \end{itemize}
-
-\paragraph{Splitting stage-\bbtwo\ terms.}
-
-Splitting a stage~\bbtwo\ term $\coltwo e A$ generates a
-\emph{precomputation}~$p$, and a~\emph{resumer} function $l.r$.  The
-resumer accepts the value produced by the precomputation as input
-($l$) and performs the stage~\bbtwo\ subcomputations of~$e$.
-Specifically, if~$e$ in~\lang\ reduces to the value~$v$, then after
-splitting the term $\letin l p r$ also reduces to~$v$ in \langmono.
-
-\ur{I don't quite understand the following.}
-(Thus, the resumer, bound to the value output by $p$, is equivalent to
-the residual term defined in~\ref{ssec:dynamics})
-
-%	\ur{Correctness will have
-%	  to go beyond this, by saying for example that all stage 1 parts are
-%	  in p etc.  Note that this correctness allows a trivial split
-%	   that shoves everything to r, e.g., p = () and r is ``e''
-%	  (might need to flatten).}
-
-The splitting algorithm for stage~\bbtwo\ terms, as specified by the
-judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
-recursively on the structure of~$e$.  In the terminal cases of
-constants and variables, splitting generates precomputations that are
-\texttt{()}, because these terms do not involve any stage~\bbone\
-computations, and resumers that produce the original term. For
-example, the integer constant \texttt{5} splits into the
-precomputation \texttt{()} and \texttt{l=>5}.
-
-For all non-terminal stage~\bbtwo\ terms~$e$ in \lang\ (except $\prev$
-and $\pause$), splitting descends into~$e$, recursively splitting
-subterms and generating precomputations $p_1, \ldots, p_n$ and
-resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
-The precompuation of~$e$ is defined as the combined precomputations:
-$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
-argument $(l_1,\ldots,l_n)$ in a term that has the same structure
-of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
-Notice that splitting \texttt{case}s and functions causes the
-precomputation of subterms to be lifted out from underneath stage
-\bbtwo\
-binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
-\ur{This last point seems important.  It would be important to come
-  back to it.}
-
-The $\prev$ and $\pause$ rules depends on
-splitting stage \bbone\ terms, and so they are covered later.
-
-%For example, the stage \bbtwo\ term
-%\begin{lstlisting}
-%2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
-%\end{lstlisting}
-%has two subterms, the latter of which itself has two subterms. Two are $\pause$s
-%containing stage \bbone\ computations, while the third has no stage \bbone\
-%content. 
-%
-%\ur{The discussion of the subterms is not helpful. what is the point?}
-%
-%The precomputation \emph{lifts} these computations in the same tree
-%structure: \verb|(1+2,(3+4,()))|.%
-%\ur{Not clear what a ``tree structure'' is. }
-%\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
-%The resumer is a function which computes the stage \bbtwo\ value of the original
-%term by mapping the results of the precomputation in place of the $\pause$s:
-%\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
-%
-%\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
-%
-%Non-staging constructs with $n$ subterms have precomputations which are
-%$n$-tuples of their subterms' precomputations, and resumers which pass each
-%component to the resumer of the respective subterm.
-
-\subsection{Splitting Stage \bbone\ Terms}
 
@@
   tuple a partial-value?}
-
-The goal of stage 2 splitting was to generate a precomputation and resumer such that the resumer and the \lang\ residual reduce to the same final value.
-However, stage~\bbone\ terms do not reduce to final values, rather they reduce to residual tables and partial values.
-For example, consider the stage 1 term:
-\begin{lstlisting}
-1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
-\end{lstlisting}
-which evaluates (via $\mathbin{\redonesym}$), to the following residual table and a multi-stage tuple:
-
-\begin{lstlisting}
-2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
-\end{lstlisting}
 
@@
 %value $v$.
-
-Note that this result simultaneously represents information at {\em both} stages: 
-a stage~\bbone\ value 3, and the stage~\bbtwo\ computation $\letin \yhat {\mathtt{7+5}} \yhat$. 
-Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms in \lang\ into a resumer that reduces to the same value as the stage~\bbtwo\ computation, and an $e1$ that not only performs the precomputation expected by the resumer, but also computes values that match the stage~\bbone\ aspects of the partial value.
 
@@
 
-\subsubsection{Splitting Mechanics}
-
 \ur{It starts getting rough here, mostly because combined term is not
@@
 correct branch.
+
+The goal of stage 2 splitting was to generate a precomputation and resumer such
+that the resumer and the \lang\ residual reduce to the same final value.
+However, stage~\bbone\ terms do not reduce to final values, rather they reduce
+to residual tables and partial values.  
+Therefore, to make the results of splitting match the
+the partial value above, we choose to split \bbone\ terms in \lang\ into a
+resumer that reduces to the same value as the stage~\bbtwo\ computation, and an
+$e_1$ that not only performs the precomputation expected by the resumer, but
+also computes values that match the stage~\bbone\ aspects of the partial value.
+
+\subsection{Splitting Stage \bbtwo\ Terms}
+
+Splitting a stage~\bbtwo\ term $\coltwo e A$ generates a
+\emph{precomputation}~$p$, and a~\emph{resumer} function $l.r$.  The
+resumer accepts the value produced by the precomputation as input
+($l$) and performs the stage~\bbtwo\ subcomputations of~$e$.
+Specifically, if~$e$ in~\lang\ reduces to the value~$v$, then after
+splitting the term $\letin l p r$ also reduces to~$v$ in \langmono.
+
+\ur{I don't quite understand the following.}
+(Thus, the resumer, bound to the value output by $p$, is equivalent to
+the residual term defined in~\ref{ssec:dynamics})
+
+%	\ur{Correctness will have
+%	  to go beyond this, by saying for example that all stage 1 parts are
+%	  in p etc.  Note that this correctness allows a trivial split
+%	   that shoves everything to r, e.g., p = () and r is ``e''
+%	  (might need to flatten).}
+
+The splitting algorithm for stage~\bbtwo\ terms, as specified by the
+judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
+recursively on the structure of~$e$.  In the terminal cases of
+constants and variables, splitting generates precomputations that are
+\texttt{()}, because these terms do not involve any stage~\bbone\
+computations, and resumers that produce the original term. For
+example, the integer constant \texttt{5} splits into the
+precomputation \texttt{()} and \texttt{l=>5}.
+
+For all non-terminal stage~\bbtwo\ terms~$e$ in \lang\ (except $\prev$
+and $\pause$), splitting descends into~$e$, recursively splitting
+subterms and generating precomputations $p_1, \ldots, p_n$ and
+resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
+The precompuation of~$e$ is defined as the combined precomputations:
+$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
+argument $(l_1,\ldots,l_n)$ in a term that has the same structure
+of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
+Notice that splitting \texttt{case}s and functions causes the
+precomputation of subterms to be lifted out from underneath stage
+\bbtwo\
+binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
+\ur{This last point seems important.  It would be important to come
+  back to it.}
+
+The $\prev$ and $\pause$ rules depends on
+splitting stage \bbone\ terms, and so they are covered later.
+
+%For example, the stage \bbtwo\ term
+%\begin{lstlisting}
+%2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
+%\end{lstlisting}
+%has two subterms, the latter of which itself has two subterms. Two are $\pause$s
+%containing stage \bbone\ computations, while the third has no stage \bbone\
+%content. 
+%
+%\ur{The discussion of the subterms is not helpful. what is the point?}
+%
+%The precomputation \emph{lifts} these computations in the same tree
+%structure: \verb|(1+2,(3+4,()))|.%
+%\ur{Not clear what a ``tree structure'' is. }
+%\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
+%The resumer is a function which computes the stage \bbtwo\ value of the original
+%term by mapping the results of the precomputation in place of the $\pause$s:
+%\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
+%
+%\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
+%
+%Non-staging constructs with $n$ subterms have precomputations which are
+%$n$-tuples of their subterms' precomputations, and resumers which pass each
+%component to the resumer of the respective subterm.
 

### 2014-11-07 paper/splitting.tex
few comments about stage one splitting

@@
 
-The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
-while the rule for $\prev$ projects the combined result of its subexpression to isolate the precomputation.
-It may seem like this $\prev$ rule is throwing information away, but we well show later that the first component 
-of the evaluated form of $c$ in the $\prev$ rule must always be unit.
+In general, the
+
+The rule for $\next$ simply tuples up the precomputation of its subterm with a trivial immediate result,
+while the rule for $\prev$ projects the combined result of its subterm to isolate the precomputation.
+Since the argument to $\prev$ must have a $\fut$ type, its immediate result must reduce to \texttt{()},
+justifying why that part can be thrown away.
 The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
@@
 
-In the splitting rules for stage \bbone\ {\tt case} expressions, we
-have enough information at stage \bbone\ to know what branch to take, so there's
-no need to evaluate both sides. Instead, we evaluate the stage \bbone\ portion of the
-taken branch only, and inject the precomputation into a sum type. In the
-resumer, we case on that injection, and resume the stage \bbtwo\ portion of the
-correct branch.
+At stage \bbone\ {\tt case}s, the splitting rules adds injections to the precomputations of each branch. 
+This is then cased in the resumer to resume the stage \bbtwo\ portion of the correct branch.
+This rule and $\pause$ are the only two places where nontrivial information is added to the precomputation.
 

### 2014-11-07 paper/splitting.tex
Syncing up partial value masking with the splitting intro.

--- 2014-11-06 paper/splitting.tex
@@
 stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
-stage~\bbtwo\ subcomputations.
+stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
+the staged language produces the same result as evaluating~$e_1$ and 
+then~$e_2$ with the result from~$e_1$.
 
-Splitting preserves correctness in the sense that evaluation of~$e$ in
-the staged language produces the same result as evaluating~$e_1$ and
-then~$e_2$ with the result from~$e_1$.  In this section, we present a
-splitting algorithm for the source language \lang\, specifying
-precisely the algorithm and the correctness condition.  We start by
-defining a masking relation that helps us to isolate stage~\bbone\ and
-stage~\bbtwo\ components of a partial value and then move on to the
-specification of the splitting algorithms.
+In this section, we present a
+splitting algorithm for the source language \lang, specifying
+precisely the algorithm and its correctness condition.  
+We begin by defining a masking relation that serves to extract stage~\bbone\ and
+stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting must match. We then 
+move on to the specification of the splitting algorithm itself.
 
@@
 
-Intuitively, we require that the terms obtained by splitting a term $e$, when
-evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
-the case of stage~\bbone\ terms, our dynamic semantics produce partial values
-containing information at \emph{both} stages, which goes against the idea that
-splitting separates the subcomputations belonging to each stage.
+% Intuitively, we require that the terms obtained by splitting a term $e$, when
+% evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
+% the case of stage~\bbone\ terms, our dynamic semantics produce partial values
+% containing information at \emph{both} stages, which goes against the idea that
+% splitting separates the subcomputations belonging to each stage.
 
-For example, consider the stage 1 term
+%For example, the partial value below contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
+%computation $\letin \yhat {\mathtt{7+5}} \yhat$.
+
+% such as in the example below:
+
+%For example, consider the stage 1 term
+%\begin{lstlisting}
+%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+%\end{lstlisting}
+%which evaluates via $\redonesym$ to a 
+%
+%residual table and multi-stage tuple
+
+%This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
+%computation $\letin \yhat {\mathtt{7+5}} \yhat$.
+
+Recall that \lang\ partial values and residual tables together may
+contain content at \emph{both} stages. For example the term:
 \begin{lstlisting}
@@
 \end{lstlisting}
-which evaluates via $\redonesym$ to a residual table and multi-stage tuple
+evaluates via $\redonesym$ to:
 \begin{lstlisting}
@@
 \end{lstlisting}
-This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
+Intuitively, splitting this term should generate an $e1$ that reduces to the stage 1 content of the partial value (3), and an $e2$ corresponding to the stage~\bbtwo\
 computation $\letin \yhat {\mathtt{7+5}} \yhat$.
+To precisely delineate the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
+\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
+which take a residual table and partial value as input and emit a term in \langmono\ that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
+masking functions.
+% written $\masko{{-}}$ and $\maskt{{-};{-}}$.
 
-To clearly delineate these parts of the partial value, we define 
-\emph{partial-value masking}, or simply {\em masking}, functions at each stage,
-which compute from a residual table and partial value the content (expressed as
-a term in \langmono) belonging to each stage. \ref{fig:valMask} defines these
-masking functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$.
-
-The goal of $\masko{v}$ is to produce a term in \langmono\ containing exactly
+\emph{Stage~\bbone\ masking} of the partial value $v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
 the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
-content, we replace all $\next$ subterms of $v$ with \texttt{()}.  For lambdas,
-since these represent potentially-multistage computations, we split their bodies
-as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and return
-the stage~\bbone\ component. $\masko{v}$ is always a value in \langmono, and has
+content, $\masko{v}$ replaces all $\next$ subterms of $v$ with \texttt{()}.
+Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
+the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
 the same structure as~$v$.
 
-The goal of $\maskt{\xi;v}$ is to produce a term in \langmono\ containing
-exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike in stage \bbone\ masking,
-here we need access to the residual table $\xi$, because all stage~\bbtwo\
-content in $v$ has already been lifted out into $\xi$. To discard all
-stage~\bbone\ information in $v$, we replace base constants with~\texttt{()},
-replace $\next~\hat{y}$ with~$\hat{y}$, and replace all $n$-ary term
-constructors with an $n$-tuple of their subterms, themselves masked. As before,
-we mask functions by splitting their bodies as general stage~\bbone\ terms, and
-use the stage~\bbtwo\ component.
-
-After performing this transformation on $v$, we reify $\xi$ around the result,
+\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
+exactly the stage~\bbtwo\ content of $[\xi;v]$. Stage~\bbtwo\ masking 
+requires access to the residual table $\xi$ since all stage~\bbtwo\
+content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
+stage~\bbone\ content in $v$ by replacing base constants with~\texttt{()},
+replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
+constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ splitting, stage~\bbtwo\
+splititing masks functions by splitting their bodies as general stage~\bbone\ terms, and
+returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ reifies $\xi$ around the result,
 producing a term in \langmono.

### 2014-11-07 paper/splitting.tex
Small refinements to the text in partial value masking.  Verbosely using the terminology stage 1 masking and stage 2 masking.

@@
 stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
-the staged language produces the same result as evaluating~$e_1$ and 
-then~$e_2$ with the result from~$e_1$.
+the staged language produces the same result as evaluating~$e_1$, followed by evaluating 
+$e_2$ with the result from~$e_1$.
 
@@
 computation $\letin \yhat {\mathtt{7+5}} \yhat$.
-To precisely delineate the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
+To precisely define the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
 \emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
@@
 \emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
-exactly the stage~\bbtwo\ content of $[\xi;v]$. Stage~\bbtwo\ masking 
-requires access to the residual table $\xi$ since all stage~\bbtwo\
+exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
+requires access to the residual table $\xi$ because all stage~\bbtwo\
 content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
@@
 splititing masks functions by splitting their bodies as general stage~\bbone\ terms, and
-returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ reifies $\xi$ around the result,
+returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
 producing a term in \langmono.

### 2014-11-07 paper/splitting.tex
explaining split one rules

--- 2014-11-07 paper/splitting.tex
@@
 stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
-stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
+stage~\bbtwo\ subcomputations.
+
+Splitting preserves correctness in the sense that evaluation of~$e$ in
 the staged language produces the same result as evaluating~$e_1$ and
-then~$e_2$ with the result from~$e_1$.
-
-In this section, we present a
-splitting algorithm for the source language \lang, specifying
-precisely the algorithm and its correctness condition.  
-We begin by defining a masking relation that serves to extract stage~\bbone\ and
-stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting must match. We then 
-move on to the specification of the splitting algorithm itself.
+then~$e_2$ with the result from~$e_1$.  In this section, we present a
+splitting algorithm for the source language \lang\, specifying
+precisely the algorithm and the correctness condition.  We start by
+defining a masking relation that helps us to isolate stage~\bbone\ and
+stage~\bbtwo\ components of a partial value and then move on to the
+specification of the splitting algorithms.
 
@@
 
-% Intuitively, we require that the terms obtained by splitting a term $e$, when
-% evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
-% the case of stage~\bbone\ terms, our dynamic semantics produce partial values
-% containing information at \emph{both} stages, which goes against the idea that
-% splitting separates the subcomputations belonging to each stage.
+Intuitively, we require that the terms obtained by splitting a term $e$, when
+evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
+the case of stage~\bbone\ terms, our dynamic semantics produce partial values
+containing information at \emph{both} stages, which goes against the idea that
+splitting separates the subcomputations belonging to each stage.
 
-%For example, the partial value below contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
-%computation $\letin \yhat {\mathtt{7+5}} \yhat$.
-
-% such as in the example below:
-
-%For example, consider the stage 1 term
-%\begin{lstlisting}
-%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
-%\end{lstlisting}
-%which evaluates via $\redonesym$ to a 
-%
-%residual table and multi-stage tuple
-
-%This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
-%computation $\letin \yhat {\mathtt{7+5}} \yhat$.
-
-Recall that \lang\ partial values and residual tables together may
-contain content at \emph{both} stages. For example the term:
+For example, consider the stage 1 term
 \begin{lstlisting}
@@
 \end{lstlisting}
-evaluates via $\redonesym$ to:
+which evaluates via $\redonesym$ to a residual table and multi-stage tuple
 \begin{lstlisting}
@@
 \end{lstlisting}
-Intuitively, splitting this term should generate an $e1$ that reduces to the stage 1 content of the partial value (3), and an $e2$ corresponding to the stage~\bbtwo\
+This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
 computation $\letin \yhat {\mathtt{7+5}} \yhat$.
-To precisely delineate the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
-\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
-which take a residual table and partial value as input and emit a term in \langmono\ that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
-masking functions.
-% written $\masko{{-}}$ and $\maskt{{-};{-}}$.
 
-\emph{Stage~\bbone\ masking} of the partial value $v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
+To clearly delineate these parts of the partial value, we define 
+\emph{partial-value masking}, or simply {\em masking}, functions at each stage,
+which compute from a residual table and partial value the content (expressed as
+a term in \langmono) belonging to each stage. \ref{fig:valMask} defines these
+masking functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$.
+
+The goal of $\masko{v}$ is to produce a term in \langmono\ containing exactly
 the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
-content, $\masko{v}$ replaces all $\next$ subterms of $v$ with \texttt{()}.
-Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
-the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
+content, we replace all $\next$ subterms of $v$ with \texttt{()}.  For lambdas,
+since these represent potentially-multistage computations, we split their bodies
+as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and return
+the stage~\bbone\ component. $\masko{v}$ is always a value in \langmono, and has
 the same structure as~$v$.
 
-\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
-exactly the stage~\bbtwo\ content of $[\xi;v]$. Stage~\bbtwo\ masking 
-requires access to the residual table $\xi$ since all stage~\bbtwo\
-content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
-stage~\bbone\ content in $v$ by replacing base constants with~\texttt{()},
-replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
-constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ splitting, stage~\bbtwo\
-splititing masks functions by splitting their bodies as general stage~\bbone\ terms, and
-returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ reifies $\xi$ around the result,
+The goal of $\maskt{\xi;v}$ is to produce a term in \langmono\ containing
+exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike in stage \bbone\ masking,
+here we need access to the residual table $\xi$, because all stage~\bbtwo\
+content in $v$ has already been lifted out into $\xi$. To discard all
+stage~\bbone\ information in $v$, we replace base constants with~\texttt{()},
+replace $\next~\hat{y}$ with~$\hat{y}$, and replace all $n$-ary term
+constructors with an $n$-tuple of their subterms, themselves masked. As before,
+we mask functions by splitting their bodies as general stage~\bbone\ terms, and
+use the stage~\bbtwo\ component.
+
+After performing this transformation on $v$, we reify $\xi$ around the result,
 producing a term in \langmono.
@@
 
-In general, the
+When $e$ is a terminal (variable or base constant), the combined result
+is formed by tupling $e$ with a \texttt{()} precomputation, 
+and the resumer is just \texttt{()}.
+When $e$ is a non-terminal (except $\next$),
+splitting descends into $e$, recursively splitting subterms
+and generating combined terms $c_1,\ldots,c_n$.
+The new combined term of $e$ then works by calling $c_1,\ldots,c_n$,
+destucting them into variables $y_1,\ldots,y_n$ for immediate results 
+and $z_1,\ldots,z_n$ for precomputations.
+The immediate result of $e$ is then formed by replacing its subterms with $y_1,\ldots,y_n$, 
+and the precomputations is formed from the tuple of $z_1,\ldots,z_n$.
+
+In the These combined terms are each
+
+At stage \bbone\ {\tt case}s, the splitting rules adds injections to the precomputations of each branch. 
+This is then cased in the resumer to resume the stage \bbtwo\ portion of the correct branch.
+This rule and $\pause$ are the only two places where nontrivial information is added to the precomputation.
 
@@
 but we have to include it here for termination purposes.}
-
-At stage \bbone\ {\tt case}s, the splitting rules adds injections to the precomputations of each branch. 
-This is then cased in the resumer to resume the stage \bbtwo\ portion of the correct branch.
-This rule and $\pause$ are the only two places where nontrivial information is added to the precomputation.
 

### 2014-11-07 paper/splitting.tex
Big change to redo the transition from partial value masking to stage 1 splitting.

--- 2014-11-07 paper/splitting.tex
@@
 
-Given a term in the source language \lang\ where terms of
+Given a term $e$ in the source language \lang\ where terms of
 stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
   splitting} separates the stage~\bbone\ and stage~\bbtwo\
-subcomputations into two separate terms, each performing the
-subcomputations corresponding to each stage.  More precisely, given a
-\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
-unstaged target language \langmono\ such that~$e_1$ performs all the
+subcomputations into two terms, $e_1$ and $e_2$, in an unstaged target language \langmono\ 
+%each performing the
+%subcomputations corresponding to each stage.  More precisely, given a
+%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
+%unstaged target language \langmono\
+such that~$e_1$ performs all the
 stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
@@
 
-%For example, the partial value below contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
-%computation $\letin \yhat {\mathtt{7+5}} \yhat$.
-
-% such as in the example below:
-
-%For example, consider the stage 1 term
-%\begin{lstlisting}
-%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
-%\end{lstlisting}
-%which evaluates via $\redonesym$ to a 
-%
-%residual table and multi-stage tuple
-
-%This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
-%computation $\letin \yhat {\mathtt{7+5}} \yhat$.
-
 Recall that \lang\ partial values and residual tables together may
@@
 \end{lstlisting}
-Intuitively, splitting this term should generate an $e1$ that reduces to the stage 1 content of the partial value (3), and an $e2$ corresponding to the stage~\bbtwo\
+Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value (3), and an $e_2$ corresponding to the stage~\bbtwo\
 computation $\letin \yhat {\mathtt{7+5}} \yhat$.
@@
 \emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
-which take a residual table and partial value as input and emit a term in \langmono\ that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
+which take a residual table and partial value as input and emit a term in \langmono\
+that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
 masking functions.
-% written $\masko{{-}}$ and $\maskt{{-};{-}}$.
 
@@
 the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
-the same structure as~$v$.
+the same structure as~$v$.  For example, the stage~\bbone\ mask of the partial value given above is \texttt{(3,())}.
 
@@
 replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
-constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ splitting, stage~\bbtwo\
-splititing masks functions by splitting their bodies as general stage~\bbone\ terms, and
-returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
-producing a term in \langmono.
+constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ masking, stage~\bbtwo\
+masking splits the bodies of lambdas as general stage~\bbone\ terms, and
+returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
+producing a term in \langmono. The stage~\bbtwo\ mask of the partial value and residual table given above is:
+\begin{lstlisting}
+let yhat=7+5 in ((),yhat)
+\end{lstlisting}
 
@@
 \label{sec:split-one}
-
-\ur{Fill this in.}
-
-FOLLOW OUTLINE:
-
-\begin{itemize}
-\item The structure of the splitting judgment.
-\item The equivalence condition.
-\item Walk through the cases.
-\end{itemize}
-
-\ur{What follows is still a bit vague to me. Some
-  suggestions/questions: are ``term'' and ``computation'' being used
-  interchangeably? Not sure I understand this.  Isn't ``multi-stage''
-  tuple a partial-value?}
 
@@
 
-\ur{It starts getting rough here, mostly because combined term is not
-  defined.  Specifically, we should say what
-  the  form of the splitting judgment is. }
+Splitting a stage \bbone\ term $\colone e A$ 
+(where $e\mathbin{\redonesym}[\xi;v]$), yields a term 
+$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that represents all stage~\bbtwo\ subcomputations.
+For splitting to match the results of evaluating $e$ in \lang,
+we require that $e_1$ reduce to the tuple $(y,z)$,
+where $y$ is identical to $\masko{\xi;v}$, and 
+$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
+Intuitively, $e_1$ performs both the stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (\emph{precomputations}) required to produce the input for the resumer.
 
-Given the definition of masking, splitting a stage \bbone\ term $\colone e A$ 
-(where $e\mathbin{\redonesym}[\xi;v]$), yields a \emph{combined term}
-$c$ that performs all stage~\bbone\ computations in $e$ and a \emph{resumer} function $l.r$ that represents all stage~\bbtwo\ computations.
-Whereas in stage~\bbtwo\ term splitting,
-$p$ constitutes only the precomputation to the resumer, in stage~\bbone\ splitting we require $c$ to reduce (via~$\redsym$) to the tuple $(y,z)$, where $y$ is an \emph{immediate result} that is identical to $\masko{\xi;v}$, and $z$ is the required precomputation such that $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
-
-Splitting stage~\bbone\ terms is implemented by the judgment $e \splitonesym [c,l.r]$ in \ref{fig:termSplitOne}.
-
-% The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
-% resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
-% which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
-% \texttt{((),12)}, which agrees with the stage \bbtwo\ mask.
+The splitting algorithm for stage~\bbone\ terms, as specified by the
+judgment $e \splitonesym [e_1,l.r]$ in \cref{fig:termSplitOne}, proceeds
+recursively on the structure of~$e$.
 

### 2014-11-07 paper/splitting.tex
Minor grammar.
Paragraph break.
Typo.

@@
 splitting algorithm for the source language \lang, specifying
-precisely the algorithm and its correctness condition.  
+precisely the algorithm and its correctness conditions.  
 We begin by defining a masking relation that serves to extract stage~\bbone\ and
@@
 computation $\letin \yhat {\mathtt{7+5}} \yhat$.
+
 To precisely define the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
@@
 returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
-producing a term in \langmono. The stage~\bbtwo\ mask of the partial value and residual table given above is:
+producing a term in \langmono. For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
 \begin{lstlisting}

### 2014-11-07 paper/splitting.tex
more stage one splitting descriptions

--- 2014-11-07 paper/splitting.tex
@@
 
-\TODO REDO THE REST OF THE TEXT TO WALK THROUGH THE MAJOR RULES OF SPLITTING
+When $e$ is a terminal (variable or base constant), the combined result
+is formed by tupling $e$ with a \texttt{()} precomputation, 
+and the resumer is just constant \texttt{()}.  
+For example, \texttt{3} simply splits into combined result \texttt{((3,())} and \texttt{\_=>4}.
+When $e$ is a non-terminal (except $\next$),
+splitting descends into $e$, recursively splitting subterms
+and generating combined terms $c_1,\ldots,c_n$.
+The new combined term of $e$ then works by binding $c_1,\ldots,c_n$
+to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
+to isolate their immediate results and precomputations.
+The immediate result of $e$ is then formed by replacing its subterms with $y_1,\ldots,y_n$, 
+and the precomputation is formed from the tuple of $z_1,\ldots,z_n$.
 
-In general, the
-
-The rule for $\next$ simply tuples up the precomputation of its subterm with a trivial immediate result,
-while the rule for $\prev$ projects the combined result of its subterm to isolate the precomputation.
-Since the argument to $\prev$ must have a $\fut$ type, its immediate result must reduce to \texttt{()},
-justifying why that part can be thrown away.
-The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
-and projects out the integer result in the resumer.% 
-\footnote{The resumer of an integer expression is usually trivial, 
-but we have to include it here for termination purposes.}
-
-At stage \bbone\ {\tt case}s, the splitting rules adds injections to the precomputations of each branch. 
-This is then cased in the resumer to resume the stage \bbtwo\ portion of the correct branch.
+At {\tt case}s, only one of the branches' combined results is executed, and its precomputation injected and bundled with the precomputation of the predicate. 
+This precomputation is then cased in the resumer to resume the stage \bbtwo\ portion of the correct branch.
 This rule and $\pause$ are the only two places where nontrivial information is added to the precomputation.
 
-The goal of stage 2 splitting was to generate a precomputation and resumer such
-that the resumer and the \lang\ residual reduce to the same final value.
-However, stage~\bbone\ terms do not reduce to final values, rather they reduce
-to residual tables and partial values.  
-Therefore, to make the results of splitting match the
-the partial value above, we choose to split \bbone\ terms in \lang\ into a
-resumer that reduces to the same value as the stage~\bbtwo\ computation, and an
-$e_1$ that not only performs the precomputation expected by the resumer, but
-also computes values that match the stage~\bbone\ aspects of the partial value.
+[FUNCTIONS]
 
@@
 
-The $\prev$ and $\pause$ rules depends on
-splitting stage \bbone\ terms, and so they are covered later.
+\subsection{Interface Rules}
+
+The rule for $\next$ simply tuples up the precomputation of its subterm with a trivial immediate result,
+while the rule for $\prev$ projects the combined result of its subterm to isolate the precomputation.
+Since the argument to $\prev$ must have a $\fut$ type, its immediate result must reduce to \texttt{()},
+justifying why that part can be thrown away.
+The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
+and projects out the integer result in the resumer.% 
+\footnote{The resumer of an integer expression is usually trivial, 
+but we have to include it here for termination purposes.}
 

### 2014-11-07 paper/splitting.tex
These are minor grammar edits.

--- 2014-11-07 paper/splitting.tex
@@
 
-Given a term $e$ in the source language \lang\ where terms of
+Given a term $e$ in the source language \lang, where terms of
 stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
@@
 
-Splitting a stage \bbone\ term $\colone e A$ 
+Splitting a stage \bbone\ term in \lang\
+% $\colone e A$ 
 (where $e\mathbin{\redonesym}[\xi;v]$), yields a term 
-$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that represents all stage~\bbtwo\ subcomputations.
+$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
 For splitting to match the results of evaluating $e$ in \lang,
@@
 $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
-Intuitively, $e_1$ performs both the stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (\emph{precomputations}) required to produce the input for the resumer.
+Intuitively, $e_1$ performs both the stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (called \emph{precomputations}) required to produce the input required by the resumer.
 

### 2014-11-07 paper/splitting.tex
wrote about functions

@@
 
-[FUNCTIONS]
+Function introduction has a \texttt{()} precomputation, since functions are already fully reduced in our semantics.
+But since the body of a function may be multistage, splitting must continue into it.
+The immediate result is a new function formed out of the stage \bbone\ part of the original body,
+and the resumer is a new function formed out of the stage \bbtwo\ part of the original body.
+It is the responsibility of the application site to save the precomputation of the function body
+and then appropriately pass it to the resumer version of the function.
 

### 2014-11-07 paper/splitting.tex
Editing pass through stage 1 terminals.

--- 2014-11-07 paper/splitting.tex
@@
 $e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
-For splitting to match the results of evaluating $e$ in \lang,
+To ensure the results of splitting produce the same output as the evaluation $e$ in \lang,
 we require that $e_1$ reduce to the tuple $(y,z)$,
@@
 $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
-Intuitively, $e_1$ performs both the stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (called \emph{precomputations}) required to produce the input required by the resumer.
+Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (called \emph{precomputations}) required to produce the input required by the resumer, we refer to this term as the \emph{combined result}, and hereafter denote it as $c$.
 
 The splitting algorithm for stage~\bbone\ terms, as specified by the
-judgment $e \splitonesym [e_1,l.r]$ in \cref{fig:termSplitOne}, proceeds
+judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
 recursively on the structure of~$e$.
+When $e$ is a terminal (a variables or base constant)
+splitting yields a combined result formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
+(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into combined result \texttt{(3,())} and resumer \texttt{\_=>()}.
 
-When $e$ is a terminal (variable or base constant), the combined result
-is formed by tupling $e$ with a \texttt{()} precomputation, 
-and the resumer is just constant \texttt{()}.  
-For example, \texttt{3} simply splits into combined result \texttt{((3,())} and \texttt{\_=>4}.
+% Note the direct correspondence of these splitting outputs to the output of the stage~\bbone\ and~\bbtwo\ masking functions.
+
 When $e$ is a non-terminal (except $\next$),

### 2014-11-07 paper/splitting.tex
Rough editing pass through the stage 1 splitting description.

--- 2014-11-07 paper/splitting.tex
@@
 recursively on the structure of~$e$.
-When $e$ is a terminal (a variables or base constant)
+When $e$ is a terminal (a variable or base constant)
 splitting yields a combined result formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
-(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into combined result \texttt{(3,())} and resumer \texttt{\_=>()}.
+(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined result \texttt{(3,())} and resumer \texttt{\_=>()}.
 
@@
 
-When $e$ is a non-terminal (except $\next$),
-splitting descends into $e$, recursively splitting subterms
-and generating combined terms $c_1,\ldots,c_n$.
-The new combined term of $e$ then works by binding $c_1,\ldots,c_n$
+For all non-terminals (except $\next$),
+splitting descends into $e$, recursively splitting its $n$ subterms
+to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
+The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
 to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
-to isolate their immediate results and precomputations.
-The immediate result of $e$ is then formed by replacing its subterms with $y_1,\ldots,y_n$, 
-and the precomputation is formed from the tuple of $z_1,\ldots,z_n$.
+to isolate immediate results from the results of precomputations. Then,
+the immediate result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
+The resumer binds the precomputation results $z_1,\ldots,z_n$ to an
+argument $(l_1,\ldots,l_n)$ in a term that has the same structure
+of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
 
-At {\tt case}s, only one of the branches' combined results is executed, and its precomputation injected and bundled with the precomputation of the predicate. 
-This precomputation is then cased in the resumer to resume the stage \bbtwo\ portion of the correct branch.
-This rule and $\pause$ are the only two places where nontrivial information is added to the precomputation.
+The combined term of a {\tt case} executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
+The precomputation result $z_i$ for this branch is injected and bundled with that of the predicate ($z_1$).   
+$z_i$ is cased in the resumer to determine which branch's resumer should be executed.
+{\tt case} and $\pause$ are the only two rules where splitting adds non-trivial logic is added to the precomputation.
 
-[FUNCTIONS]
+% We call attention to the rule for {\tt case}s since it, and the rule for $\pause$, where splitting adds non-trivial logic to the precomputation.
+
+Function introduction has a \texttt{()} precomputation result,
+since functions are already fully reduced in our semantics.
+However, since the body of a function may itself be multi-stage, splitting must continue into it.
+The immediate result is a new function formed from the stage~\bbone\ part of the original body.
+The resumer is a new function formed out of the stage \bbtwo\ part of the original body.
+It is the responsibility of the application site to save the precomputation of the function body
+and pass it to the resumer version of the function.
 

### 2014-11-07 paper/splitting.tex
tightened up the optimization section

@@
 
-The above splitting rules are correct, but can be optimized in many cases. One
-major optimization is to eliminate redundant instances of \texttt{()}---for
-example, if multiple subterms have a \texttt{()} precomputation, these can be
-combined.
-
-A similar issue can occur in the precomputations for recursive function bodies.
-Consider the stage \bbtwo\ term
+The above splitting rules are correct, but can be optimized in many cases. 
+For instance, if multiple subterms have a \texttt{()} precomputation, these can be
+combined into a single \texttt{()}, with the residual changed accordingly.
+This becomes more difficult to fix when it occurs within a recursive function.
+For example, consider the stage \bbtwo\ term
 \begin{lstlisting}
@@
 \end{lstlisting}
-\texttt{fact}, being a stage \bbone\ function, itself splits into two recursive
+Here, the stage \bbone\ function \texttt{fact} splits into two recursive
 functions, which compute a combined result (stage \bbone\ result and
@@
 always returns \texttt{()}; worse yet, it has linear runtime!
-
-We solve this issue by adding a new staging annotation to \lang, called
-\texttt{mono}. $\monoSt~e$ is stage \bbone, and requires that $e$ contains
+But optimizing this away is in general a global operation,
+so we instead solve this issue by adding a new staging annotation to \lang, called
+\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
 no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
@@
 \begin{mathpar}
-\infer{\typesone{\monoSt~e}A}{\typesmono e A} \and
+\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
 \infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
 \end{mathpar}
-
-\crem{Why is it difficult?}
+Here, the ``safe" judgment requires that $A$ has no $\to$ types in negative position.
+Intuitively, this is to ensure that stage \bbtwo\ code cannot be substituted into a monostage region.
 
@@
 \end{mathpar}
-
-\crem{Explain $\overset{C}{\to}$.}
-
 The splitting rules are then
 \begin{mathpar}
-\infer{\splitone{\monoSt~e}A {e', {\tt dummy}} }{e \overset{C}{\rightarrow} e'} \and
-\infer{e_1~e_2 \overset{C}{\rightarrow} \pio~(e_1'~e_2')}{e_1 \overset{C}{\rightarrow} e_1' & e_2 \overset{C}{\rightarrow} e_2'} \and
-\infer{\lambda x.e \overset{C}{\rightarrow} \lambda x.(e',())}{e \overset{C}{\rightarrow} e'}
+\infer{\splitone{\monoSt~e}A {(e',()), \_.{\tt dummy}} }{e \overset{C}{\rightarrow} e'}
 \end{mathpar}
-And $\overset{C}{\rightarrow}$ operates like a map at all other terms.
-
-To see that this optimization is correct, notice that \texttt{dummy} can only
-appear in the resumers of terms that have no circles in their \lang\ type.
-Since the $\prev$ rule requires a circle type, dummy values cannot leak from the
-resumers of stage \bbone\ terms into the resumers of stage \bbtwo\ terms.
+Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
+by performing the two rewrites in place:
+\begin{align*}
+\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
+e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
+\end{align*}
+For $\splitone e A {c,l.r}$, \texttt{dummy} can only appear as the 
+final value of $r$ when $e$ has no $\fut$ in its \lang\ type.
+Since the $\prev$ rule requires a circle type, \texttt{dummy} values cannot leak 
+into the resumers of stage \bbtwo\ terms.
 

### 2014-11-07 paper/splitting.tex
Trying harder to get a transition to stage 2 splitting.

--- 2014-11-07 paper/splitting.tex
@@
 stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
-stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
-the staged language produces the same result as evaluating~$e_1$, followed by evaluating 
-$e_2$ with the result from~$e_1$.
+stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
+the staged language produces the same result as evaluating~$e_1$ in the target language, then further evaluating 
+$e_2$ in the target language with the result from~$e_1$.
 
@@
 
-%Like stage \bbtwo\ splitting, splitting a stage~\bbone\ term $\colone e A$ produces a term $c$ containing all stage~\bbone\ subcomputations in $e$ and a resumer function %$l.r$ containing all stage~\bbtwo\ subcomputations.
-%which respectively contain all the stage \bbone\ and stage \bbtwo\
-%subcomputations of $e$. 
-%But where the stage \bbtwo\ splitting goal was phrased terms ensure.
-
-%The situation is more complex than when splitting stage
-%\bbtwo\ terms; to see why, consider the correctness criterion for splitting
-%stage \bbone\ terms.
-
-%Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
-%yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
-%value $v$.
-
-% Ergo, if we intend to compare the results of evaluation to the those of splitting, 
-% then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.
-
-% But the whole point of splitting is that its results (and their evaluated forms) are monostage.
-
-% Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
-
-% For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
-% replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.
-
-%The only exception is for
-%functions $\lam xAe$; such a value represents a computation, so masking splits
-%its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
-%stage \bbone\ component, the combined result.
-
-%They operate by replacing all stage \bbone\
-%(resp., \bbtwo) components of a partial value by \texttt{()}.
-
-%It is tempting to compare the two results of evaluating $e$ to the
-%two results of splitting it.
-
-%To see where this goes wrong, consider the stage
-%\bbone\ term
-%\begin{lstlisting}
-%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
-%\end{lstlisting}
-%which evaluates to
-%\begin{lstlisting}
-%2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
-%\end{lstlisting}
-%This partial value contains a $\next$, and so is itself multistaged:
-%intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
-%is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5} in the table.
-
-%We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
-%$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
-%a partial value at each stage, essentially by replacing all stage \bbone\
-%(resp., \bbtwo) components of a partial value by \texttt{()}.
-
-% % Umut: I made a pass over this, see above.  I don't think I lost
-% % anything, so it should be fine to delete this.
-% % \section{Splitting Algorithm}
-% % \label{sec:splitting}
-% % Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
-% % \bbtwo\ components, these components may be interleaved in \lang\ code. The
-% % goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
-% % of a term $e$ in \lang\ into a pair of
-% % terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
-% % stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
-% % Informally, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
-% % as evaluating $e$ in \lang. (We make this intuition precise in the subsequent subsections.)
-
-% % Since the mechanics of splitting stage~\bbtwo\ terms are conceptually simpler than those for stage \bbone\
-% % terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.
-
-% % \subsection {Splitting Stage \bbtwo\ Terms}
-
-% % Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
-% %  a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
-% % of $e$ in place to create a residual, splitting statically isolates these computations into
-% % $p$.  The resumer accepts the value produced by the precomputation as input
-% % % (bound to the free variable $l$)
-% % and executes all stage~\bbtwo\ subcomputations.   
-% % Specifically, if $e$ in \lang\ reduces to the final value $v$, 
-% % then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
-% % (Thus, the resumer, bound to the value output by $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})
-
-% % %	\ur{Correctness will have
-% % %	  to go beyond this, by saying for example that all stage 1 parts are
-% % %	  in p etc.  Note that this correctness allows a trivial split
-% % %	   that shoves everything to r, e.g., p = () and r is ``e''
-% % %	  (might need to flatten).}
-
-% % Splitting stage~\bbtwo\ terms, as implemented by the judgment $e \splittwosym
-% % [p,l.r]$ in \cref{fig:termSplitTwo} is defined recursively on the structure of $e$.
-% % In the terminal cases of constants and variables, splitting generates precomputations that are \texttt{()}
-% % (these terms do not involve any stage~\bbone\ computations) and resumers that produce the original term's value.
-% % For example, the integer constant \texttt{5} splits into the precomputation \texttt{()} and \texttt{l=>5}.
-
-% % For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
-% % The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
-% % The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
-% % Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
-% % underneath stage \bbtwo\ binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
-
-% % The $\prev$ and $\pause$ rules depends on
-% % splitting stage \bbone\ terms, and so they are covered later.
-
-% % %For example, the stage \bbtwo\ term
-% % %\begin{lstlisting}
-% % %2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
-% % %\end{lstlisting}
-% % %has two subterms, the latter of which itself has two subterms. Two are $\pause$s
-% % %containing stage \bbone\ computations, while the third has no stage \bbone\
-% % %content. 
-% % %
-% % %\ur{The discussion of the subterms is not helpful. what is the point?}
-% % %
-% % %The precomputation \emph{lifts} these computations in the same tree
-% % %structure: \verb|(1+2,(3+4,()))|.%
-% % %\ur{Not clear what a ``tree structure'' is. }
-% % %\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
-% % %The resumer is a function which computes the stage \bbtwo\ value of the original
-% % %term by mapping the results of the precomputation in place of the $\pause$s:
-% % %\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
-% % %
-% % %\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
-% % %
-% % %Non-staging constructs with $n$ subterms have precomputations which are
-% % %$n$-tuples of their subterms' precomputations, and resumers which pass each
-% % %component to the resumer of the respective subterm.
-
-% % \subsection{Splitting Stage \bbone\ Terms}
-
-% % The goal of stage 2 splitting was to generate a precomputation and resumer such that the resumer and the \lang\ residual reduce to the same final value.
-However, stage~\bbone\ terms do not reduce to final values, rather they reduce to residual tables and partial values.
-For example, consider the stage 1 term:
-% % \begin{lstlisting}
-% % 1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
-% % \end{lstlisting}
-% % which evaluates (via $\mathbin{\redonesym}$), to the following residual table and a multi-stage tuple:
-
-% % \begin{lstlisting}
-% % 2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
-% % \end{lstlisting}
-
-% % %Like stage \bbtwo\ splitting, splitting a stage~\bbone\ term $\colone e A$ produces a term $c$ containing all stage~\bbone\ subcomputations in $e$ and a resumer function %$l.r$ containing all stage~\bbtwo\ subcomputations.
-% % %which respectively contain all the stage \bbone\ and stage \bbtwo\
-% % %subcomputations of $e$. 
-% % %But where the stage \bbtwo\ splitting goal was phrased terms ensure.
-
-% % %The situation is more complex than when splitting stage
-% % %\bbtwo\ terms; to see why, consider the correctness criterion for splitting
-% % %stage \bbone\ terms.
-
-% % %Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
-% % %yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
-% % %value $v$.
-
-% % Note that this result simultaneously represents information at {\em both} stages: 
-% % a stage~\bbone\ value 3, and the stage~\bbtwo\ computation $\letin \yhat {\mathtt{7+5}} \yhat$. 
-% % Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms in \lang\ into a resumer that reduces to the same value as the stage~\bbtwo\ computation, and an $e1$ that not only performs the precomputation expected by the resumer, but also computes values that match the stage~\bbone\ aspects of the partial value.
-
-% % % Ergo, if we intend to compare the results of evaluation to the those of splitting, 
-% % % then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.
-
-% % % But the whole point of splitting is that its results (and their evaluated forms) are monostage.
-
-% % % Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
-
-% % \subsubsection{Partial Value Masking}
-
-% % To concretize the meaning of ``match", we introduce two {\em partial value masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the subterms that are relevant at a particular stage.
-The masking functions are defined in \ref{fig:valMask}.
-
-% % Since \texttt{next}'s only contain references to the residual table (computations that occur in stage~\bbtwo),
-% % the primary operation of $\masko{\xi;v}$ is to replace all $\next$s in $v$ with \texttt{()}.
-The result is a \langmono\ value with the same structure as $v$, but containing only its stage~\bbone\ information.
-Masking is more complex when $v$ is a function since the function's body itself may contain stage~\bbtwo\ subterms.
-As a result, masking of functions depends on the general behavior of stage~\bbone\ splitting, described shortly.
-
-% % While $\masko{\cdot}$ produces a value in \langmono,
-% % $\maskt{\xi;v}$ produces a \emph{term} in \langmono\ that corresponds to the stage~\bbtwo\ computation represented by $[\xi;v]$.
-% % Stage~\bbone\ information in $v$ is discarded by replacing base constants with \texttt{()}, replacing $\next~\hat{y}$ with $\hat{y}$, and eliminating all injection tags.
-(The sum type is at stage \bbone, so the tag is not relevant at stage \bbtwo.)
-% % Last, $\maskt{\cdot}$ reifies $\xi$ around the modified $v$ to produce a term in \langmono. 
-% % As with $\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on general term splitting.
-
-% % % For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
-% % % replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.
-
-% % %The only exception is for
-% % %functions $\lam xAe$; such a value represents a computation, so masking splits
-% % %its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
-% % %stage \bbone\ component, the combined result.
-
-% % %They operate by replacing all stage \bbone\
-% % %(resp., \bbtwo) components of a partial value by \texttt{()}.
-
-% % %It is tempting to compare the two results of evaluating $e$ to the
-% % %two results of splitting it.
-
-% % %To see where this goes wrong, consider the stage
-% % %\bbone\ term
-% % %\begin{lstlisting}
-% % %1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
-% % %\end{lstlisting}
-% % %which evaluates to
-% % %\begin{lstlisting}
-% % %2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
-% % %\end{lstlisting}
-% % %This partial value contains a $\next$, and so is itself multistaged:
-% % %intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
-% % %is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5} in the table.
-
-% % %We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
-% % %$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
-% % %a partial value at each stage, essentially by replacing all stage \bbone\
-% % %(resp., \bbtwo) components of a partial value by \texttt{()}.
-
-Splitting a stage \bbone\ term in \lang\
+Splitting a stage \bbone\ term $e$ in \lang\
 % $\colone e A$ 
-(where $e\mathbin{\redonesym}[\xi;v]$), yields a term 
+yields a term 
 $e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
-To ensure the results of splitting produce the same output as the evaluation $e$ in \lang,
+To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$ in \lang,
 we require that $e_1$ reduce to the tuple $(y,z)$,
@@
 $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
-Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (called \emph{precomputations}) required to produce the input required by the resumer, we refer to this term as the \emph{combined result}, and hereafter denote it as $c$.
+Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (called the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.
 
@@
 When $e$ is a terminal (a variable or base constant)
-splitting yields a combined result formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
-(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined result \texttt{(3,())} and resumer \texttt{\_=>()}.
+splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
+(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
 
@@
 
-% We call attention to the rule for {\tt case}s since it, and the rule for $\pause$, where splitting adds non-trivial logic to the precomputation.
-
 Function introduction has a \texttt{()} precomputation result,
@@
 
-Splitting a stage~\bbtwo\ term $\coltwo e A$ generates a
-\emph{precomputation}~$p$, and a~\emph{resumer} function $l.r$.  The
-resumer accepts the value produced by the precomputation as input
-($l$) and performs the stage~\bbtwo\ subcomputations of~$e$.
-Specifically, if~$e$ in~\lang\ reduces to the value~$v$, then after
-splitting the term $\letin l p r$ also reduces to~$v$ in \langmono.
+Rather than match the
 
-\ur{I don't quite understand the following.}
-(Thus, the resumer, bound to the value output by $p$, is equivalent to
-the residual term defined in~\ref{ssec:dynamics})
+Splitting stage~\bbtwo\ terms in \lang\ takes on a simpler form than stage~\bbone\ terms since stage~\bbtwo\ terms because rather that the output of splitting
 
-%	\ur{Correctness will have
-%	  to go beyond this, by saying for example that all stage 1 parts are
-%	  in p etc.  Note that this correctness allows a trivial split
-%	   that shoves everything to r, e.g., p = () and r is ``e''
-%	  (might need to flatten).}
+Like stage one, we get two terms.
 
-The splitting algorithm for stage~\bbtwo\ terms, as specified by the
+But now the objective is to match the partial value, but the value after full evaluation.
+
+That means we need to match values.
+
+Splitting a stage~\bbtwo\ term $e$ in \lang\ generates a
+precomputation~$p$ that performs all stage~\bbone\ subcomputations in $e$,
+and a~resumer function $e_2=l.r$ that performs all stage~\bbtwo\ subcomputations.
+If full evaluation of $e$ in~\lang\ reduces to the value~$v$, then $\letin l p r$ must reduce to~$v$ in \langmono.
+
+% \ur{I don't quite understand the following.}
+% (Thus, the resumer, bound to the value output by $p$, is equivalent to
+% the residual term defined in~\ref{ssec:dynamics})
+
+As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms
+as specified by the
 judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
 recursively on the structure of~$e$.  In the terminal cases of
-constants and variables, splitting generates precomputations that are
-\texttt{()}, because these terms do not involve any stage~\bbone\
-computations, and resumers that produce the original term. For
+constants and variables, splitting generates \texttt{()} precomputations (they do not involve stage~\bbone\
+subcomputations), and resumers that produce the original term. For
 example, the integer constant \texttt{5} splits into the

### 2014-11-07 paper/splitting.tex
Edits through 4.3

@@
 
-\subsection{Splitting Stage \bbone\ Terms}
+\subsection{Stage \bbone\ Term Splitting}
 \label{sec:split-one}
@@
 splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
-(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
+(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
 
@@
 
-\subsection{Splitting Stage \bbtwo\ Terms}
+Since the results of splitting $\next$ terms depend on the output of splitting its stage~\bbtwo\ subterm,
+we defer description of $\next$ until after describing stage~\bbtwo\ term splitting. % in the subsequent section.
 
-Rather than match the
+\subsection{Stage \bbtwo\ Term Splitting}
 
-Splitting stage~\bbtwo\ terms in \lang\ takes on a simpler form than stage~\bbone\ terms since stage~\bbtwo\ terms because rather that the output of splitting
-
-Like stage one, we get two terms.
-
-But now the objective is to match the partial value, but the value after full evaluation.
-
-That means we need to match values.
-
-Splitting a stage~\bbtwo\ term $e$ in \lang\ generates a
-precomputation~$p$ that performs all stage~\bbone\ subcomputations in $e$,
-and a~resumer function $e_2=l.r$ that performs all stage~\bbtwo\ subcomputations.
-If full evaluation of $e$ in~\lang\ reduces to the value~$v$, then $\letin l p r$ must reduce to~$v$ in \langmono.
+Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
+splitting stage~\bbtwo\ terms in \lang\ assumes a simpler form than that of stage~\bbone\ term splitting. 
+Specifically, splitting a stage~\bbtwo\ term $e$ in \lang\ generates a precomputation term $p$
+(performing all stage~\bbone\ subcomputations in $e$) and a
+resumer function $e_2=l.r$ (performing all stage~\bbtwo\ subcomputations) such that when 
+$e$ reduces to the value~$v$, $\letin l p r$ reduces to an identical value in \langmono.
+Note that in contrast to the combined term produced during stage~\bbone\ splitting,
+the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
+to compute the input expected by the resumer.
 
@@
 
-As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms
-as specified by the
-judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
-recursively on the structure of~$e$.  In the terminal cases of
-constants and variables, splitting generates \texttt{()} precomputations (they do not involve stage~\bbone\
-subcomputations), and resumers that produce the original term. For
-example, the integer constant \texttt{5} splits into the
-precomputation \texttt{()} and \texttt{l=>5}.
+As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
+as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, is defined on the local structure of $e$.
+In the terminal cases of
+constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers that produce the original term. For
+example, the integer constant \texttt{3} splits into the
+precomputation \texttt{()} and resumer \texttt{\_=>3}.
 
-For all non-terminal stage~\bbtwo\ terms~$e$ in \lang\ (except $\prev$
-and $\pause$), splitting descends into~$e$, recursively splitting
-subterms and generating precomputations $p_1, \ldots, p_n$ and
-resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
-The precompuation of~$e$ is defined as the combined precomputations:
+For non-terminal terms~$e$ in \lang\ (except $\prev$
+and $\pause$)
+% recursively splitting
+% subterms and generating precomputations $p_1, \ldots, p_n$ and
+% resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
+the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
 $p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
 argument $(l_1,\ldots,l_n)$ in a term that has the same structure
-of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
-Notice that splitting \texttt{case}s and functions causes the
-precomputation of subterms to be lifted out from underneath stage
+of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.
+A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
+precomputation of subterms is lifted out from underneath stage
 \bbtwo\
@@
 
-\subsection{Interface Rules}
-
-The rule for $\next$ simply tuples up the precomputation of its subterm with a trivial immediate result,
-while the rule for $\prev$ projects the combined result of its subterm to isolate the precomputation.
-Since the argument to $\prev$ must have a $\fut$ type, its immediate result must reduce to \texttt{()},
-justifying why that part can be thrown away.
-The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
-and projects out the integer result in the resumer.% 
+Splitting $\prev$ generates a precomputation that projects the combined result of its stage~\bbone\ subterm.
+Since the argument to $\prev$ is of $\fut$ type, its immediate result reduces to \texttt{()}, justifying why it can be thrown away.
+$\pause$ treats the entire combined result of its stage~\bbone\ subexpression as a precomputation, 
+and projects out the integer result in the resumer. 
 \footnote{The resumer of an integer expression is usually trivial, 
 but we have to include it here for termination purposes.} 
-
-%For example, the stage \bbtwo\ term
-%\begin{lstlisting}
-%2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
-%\end{lstlisting}
-%has two subterms, the latter of which itself has two subterms. Two are $\pause$s
-%containing stage \bbone\ computations, while the third has no stage \bbone\
-%content. 
-%
-%\ur{The discussion of the subterms is not helpful. what is the point?}
-%
-%The precomputation \emph{lifts} these computations in the same tree
-%structure: \verb|(1+2,(3+4,()))|.%
-%\ur{Not clear what a ``tree structure'' is. }
-%\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
-%The resumer is a function which computes the stage \bbtwo\ value of the original
-%term by mapping the results of the precomputation in place of the $\pause$s:
-%\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
-%
-%\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
-%
-%Non-staging constructs with $n$ subterms have precomputations which are
-%$n$-tuples of their subterms' precomputations, and resumers which pass each
-%component to the resumer of the respective subterm.
+Finally, splitting $\next$ simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial immediate result \texttt{()}.
 
@@
 
-\subsection{Optimizations}
+\subsection{Optimization: Monostage Functions}
+
+\textbf{KF: the missing point here is that the splitting algorithm was designed based on the local structure of \lang\ terms.
+To make the algorithm derivation case compelling,
+we need to optimize away some of that structure.  But this is hard to do through functions.}
 
@@
 
-\subsection{Implementation}
-
-\crem{Say something about the implementation.}
+% \subsection{Implementation}
+%
+% \crem{Say something about the implementation.}

### 2014-11-07 paper/splitting.tex
Minor wordsmithing.

@@
 We begin by defining a masking relation that serves to extract stage~\bbone\ and
-stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting must match. We then 
+stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting are designed to match. We then 
 move on to the specification of the splitting algorithm itself.
@@
 \end{lstlisting}
-Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value (3), and an $e_2$ corresponding to the stage~\bbtwo\
-computation $\letin \yhat {\mathtt{7+5}} \yhat$.
+Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value \texttt{3}, and an $e_2$ corresponding to the stage~\bbtwo\ computation: \\
+$\letin \yhat {\mathtt{7+5}} \yhat$.
 
-To precisely define the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
+To precisely specify the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
 \emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
@@
 $\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
-Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (called the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.
+Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (a value we call the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.
 
@@
 splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
-(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
+(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
 
@@
 
-The combined term of a {\tt case} executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
+Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
 The precomputation result $z_i$ for this branch is injected and bundled with that of the predicate ($z_1$).   
@@
 In the terminal cases of
-constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers that produce the original term. For
-example, the integer constant \texttt{3} splits into the
+constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers consisting of the original term.
+For example, the integer constant \texttt{3} splits into the
 precomputation \texttt{()} and resumer \texttt{\_=>3}.
 
-For non-terminal terms~$e$ in \lang\ (except $\prev$
+For all non-terminal terms~$e$ in \lang\ (except $\prev$
 and $\pause$)
@@
 of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.
+
 A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the

### 2014-11-07 paper/figures/splitting-overview.tex
added splitting overview skeleton

\begin{figure*}

**Evaluation:

If  
$e: A @ 1$
then
$e$ $red_1$ residual table, partial value

$red_1$: 1-eval

If 
$e: A @ 2$ 
then
$e$ $red_2$ $q$, where $q$ is a residual in an unstaged language.

$red_2$: 2-eval,  which we think of as being identical to specialization.

** Splitting

$e:A@1$ splits into $c$, which represents all stage-1 subcomputations
where $c$ reduces $(y,b)$, where $y$ is the 1-result and $b$ is the
boundary value.

** Correctness

$y$ is equal to one-mask of evaluating e with 1-eval.

let l = b in r $red$ $v$ (vanilla reduction), where $v$ is identical
to the reduced value (via $red$) of two-mask of  e with 1-eval.

\caption{Stage \bbtwo\ Splitting}
\label{fig:termSplitTwo}
\end{figure*}

### 2014-11-07 paper/figures/splitting-overview.tex
Check terminology of splitting. Update summary figure.

@@
 
-**Evaluation:
+\textbf{\lang\ Evaluation:}
 
-If  
-$e: A @ 1$
-then
-$e$ $red_1$ residual table, partial value
+\hspace{2em}\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.
 
-$red_1$: 1-eval
+\hspace{2em}\bbtwo-Evaluation: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo
 
-If 
-$e: A @ 2$ 
-then
-$e$ $red_2$ $q$, where $q$ is a residual in an unstaged language.
+% $red_2$: 2-eval,  which we think of as being identical to specialization.
 
-$red_2$: 2-eval,  which we think of as being identical to specialization.
+\vspace{.75em}
+\textbf{\lang\ Splitting:}
 
-** Splitting
+\hspace{2em}\bbone-Splitting Structure: $e \splitonesym [c,l.r]$, where:
 
-$e:A@1$ splits into $c$, which represents all stage-1 subcomputations
-where $c$ reduces $(y,b)$, where $y$ is the 1-result and $b$ is the
-boundary value.
+\hspace{4em}$c$ is the \emph{combined term} (representing all stage~\bbone\ subcomputations in $e$)
 
-** Correctness
+\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputatiosn in $e$)
 
-$y$ is equal to one-mask of evaluating e with 1-eval.
+\hspace{4em}$c~\redsym~(y,b)$, where $y$ is the \emph{\bbone-result} of $e$, and $b$ is the \emph{boundary value} of $e$
 
-let l = b in r $red$ $v$ (vanilla reduction), where $v$ is identical
-to the reduced value (via $red$) of two-mask of  e with 1-eval.
+\hspace{2em}\bbone-Splitting Correctness: If $e\mathbin{\redonesym}[\xi;v]$, then:
 
-\caption{Stage \bbtwo\ Splitting}
+\hspace{4em}$y$ is identical to $\masko{\xi;v}$
+
+\hspace{4em}$\maskt{\xi;v}$ and \texttt{(let l=b in r)} reduce (via $\redsym$) to identical values.
+
+\hspace{2em}\bbtwo-Splitting Structure: $e \splittwosym [p,l.r]$, where:
+
+\hspace{4em}$p$ is the \emph{precomputation} (representing all stage~\bbone\ subcomputations in $e$)
+
+\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputations in $e$)
+
+\hspace{2em}\bbtwo-Splitting Correctness: If $e~\redtwosym~q$, and $q~\redsym~v$, then \texttt{(let l=b in r)}~$\redsym~v$
+
+\caption{Summary of \lang\ evaluation and splitting.}
 \label{fig:termSplitTwo}

### 2014-11-07 paper/splitting.tex
added splitting overview skeleton
Check terminology of splitting. Update summary figure.

@@
 \label{sec:splitting}
+
+\input{figures/splitting-overview}
 
@@
   splitting} separates the stage~\bbone\ and stage~\bbtwo\
-subcomputations into two terms, $e_1$ and $e_2$, in an unstaged target language \langmono\ 
+subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
 %each performing the
@@
 such that~$e_1$ performs all the
-stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
+stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
 stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
-the staged language produces the same result as evaluating~$e_1$ in the target language, then further evaluating 
-$e_2$ in the target language with the result from~$e_1$.
+the staged language produces the same result as evaluating~$e_1$ in the target language,
+then further evaluating~$e_2$ in the target language with the result from~$e_1$.
 
@@
 
-\subsection{Partial Value Masking}
+\subsection{Masking}
 
@@
 \end{lstlisting}
-Intuitively, splitting this term should generate an $e_1$ that reduces to the stage 1 content of the partial value \texttt{3}, and an $e_2$ corresponding to the stage~\bbtwo\ computation: \\
+Intuitively, splitting this term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value \texttt{3},
+and an~$e_2$ corresponding to the stage~\bbtwo\ computation: \\
 $\letin \yhat {\mathtt{7+5}} \yhat$.
@@
 To precisely specify the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
-\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
+\emph{masking} functions, written $\masko{{-};{-}}$ and $\maskt{{-};{-}}$, 
 which take a residual table and partial value as input and emit a term in \langmono\
@@
 
-\emph{Stage~\bbone\ masking} of the partial value $v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
-the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
-content, $\masko{v}$ replaces all $\next$ subterms of $v$ with \texttt{()}.
+\emph{Stage~\bbone\ masking} of the partial value~$v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
+the stage~\bbone\ content of~$v$. Since \next\ terms only contain stage~\bbtwo\
+content, $\masko{v}$ replaces all $\next$ subterms of~$v$ with \texttt{()}.
 Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
@@
 
-\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
+\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context~$\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
 exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
 requires access to the residual table $\xi$ because all stage~\bbtwo\
-content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
-stage~\bbone\ content in $v$ by replacing base constants with~\texttt{()},
+content in~$v$ has been lifted out into~$\xi$. Stage~\bbtwo\ masking discards
+stage~\bbone\ content in~$v$ by replacing base constants with~\texttt{()},
 replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
@@
 masking splits the bodies of lambdas as general stage~\bbone\ terms, and
-returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from $v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
+returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from~$v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
 producing a term in \langmono. For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
@@
 
-Splitting a stage \bbone\ term $e$ in \lang\
-% $\colone e A$ 
-yields a term 
+Splitting a stage~\bbone\ term $e$ in \lang\ yields a term 
 $e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
-To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$ in \lang,
-we require that $e_1$ reduce to the tuple $(y,z)$,
+To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$,
+we require that $e_1$ reduce to the tuple $(y,b)$,
 where $y$ is identical to $\masko{\xi;v}$, and 
-$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$.
-Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (a value we call the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.
+$\letin l b r$ reduces to the same value as $\maskt{\xi;v}$.
+Since~$e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (a value we call the \emph{stage~\bbone\ result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the \emph{boundary value} $b$ required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.
 
@@
 The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
-to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
-to isolate immediate results from the results of precomputations. Then,
-the immediate result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
-The resumer binds the precomputation results $z_1,\ldots,z_n$ to an
+to the patterns $(y_1,b_1),\ldots,(y_n,b_n)$
+to isolate \bbone-results from boundary values. Then,
+the \bbone-result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
+The resumer binds the boundary values $b_1,\ldots,b_n$ to an
 argument $(l_1,\ldots,l_n)$ in a term that has the same structure
@@
 
-Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
-The precomputation result $z_i$ for this branch is injected and bundled with that of the predicate ($z_1$).   
-$z_i$ is cased in the resumer to determine which branch's resumer should be executed.
+Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the \bbone-result $y_1$ of the predicate.
+The boundary value $b_i$ for this branch is injected and bundled with that of the predicate ($b_1$).   
+$b_i$ is cased in the resumer to determine which branch's resumer should be executed.
 {\tt case} and $\pause$ are the only two rules where splitting adds non-trivial logic is added to the precomputation.
 
-Function introduction has a \texttt{()} precomputation result,
+Function introduction has a \texttt{()} boundary value,
 since functions are already fully reduced in our semantics.
 However, since the body of a function may itself be multi-stage, splitting must continue into it.
-The immediate result is a new function formed from the stage~\bbone\ part of the original body.
-The resumer is a new function formed out of the stage \bbtwo\ part of the original body.
+The \bbone-result is a new function formed from the stage~\bbone\ part of the original body.
+The resumer is a new function formed out of the stage~\bbtwo\ part of the original body.
 It is the responsibility of the application site to save the precomputation of the function body
@@
 Since the results of splitting $\next$ terms depend on the output of splitting its stage~\bbtwo\ subterm,
-we defer description of $\next$ until after describing stage~\bbtwo\ term splitting. % in the subsequent section.
+we defer description of $\next$ until after describing stage~\bbtwo\ term splitting.
 
@@
 the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
-$p=(p_1,\ldots,p_n)$.  The resumer binds each precomputation to an
+$p=(p_1,\ldots,p_n)$.  The resumer binds each boundary value to an
 argument $(l_1,\ldots,l_n)$ in a term that has the same structure
@@
 
-Splitting $\prev$ generates a precomputation that projects the combined result of its stage~\bbone\ subterm.
-Since the argument to $\prev$ is of $\fut$ type, its immediate result reduces to \texttt{()}, justifying why it can be thrown away.
-$\pause$ treats the entire combined result of its stage~\bbone\ subexpression as a precomputation, 
+Splitting $\prev$ generates a precomputation that projects the \bbone-result of its stage~\bbone\ subterm.
+Since the argument to $\prev$ is of $\fut$ type, its \bbone-result reduces to \texttt{()}, justifying why it can be thrown away.
+$\pause$ treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
 and projects out the integer result in the resumer. 
@@
 but we have to include it here for termination purposes.} 
-Finally, splitting $\next$ simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial immediate result \texttt{()}.
+Finally, splitting $\next$ simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial \bbone-result \texttt{()}.
 
@@
 This becomes more difficult to fix when it occurs within a recursive function.
-For example, consider the stage \bbtwo\ term
+For example, consider the stage~\bbtwo\ term
 \begin{lstlisting}

### 2014-11-07 paper/splitting.tex
added implementation and optimization section

--- 2014-11-07 paper/splitting.tex
@@
 \input{figures/splitting-term}
-
-\subsection{Optimization: Monostage Functions}
-
-\textbf{KF: the missing point here is that the splitting algorithm was designed based on the local structure of \lang\ terms.
-To make the algorithm derivation case compelling,
-we need to optimize away some of that structure.  But this is hard to do through functions.}
-
-The above splitting rules are correct, but can be optimized in many cases. 
-For instance, if multiple subterms have a \texttt{()} precomputation, these can be
-combined into a single \texttt{()}, with the residual changed accordingly.
-This becomes more difficult to fix when it occurs within a recursive function.
-For example, consider the stage \bbtwo\ term
-\begin{lstlisting}
-2`prev{
-  1`letfun fact (n : int) : int = 
-    if n <= 0 then 1 else fact(n-1)*n
-  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
-}`
-\end{lstlisting}
-which splits into:
-\begin{lstlisting}
-1`letfun fact (n : int) : int * prec = 
-  if n <= 0 then (1,L) 
-  else let (y,z) = fact(n-1) in (y*n,R z)
-in fact 5`
-
-2`l => 
-letfun fact (n : unit, l0 : prec) : unit = 
-  case l0 of L => () | R l1 => fact ((),l0)
-in (fact (pi2 l); pi1 l)-100`
-\end{lstlisting}
-Here, the stage \bbone\ function \texttt{fact} splits into two recursive
-functions, which compute a combined result (stage \bbone\ result and
-precomputation), and a stage \bbtwo\ result, respectively. This precomputation
-could be given the type
-\begin{lstlisting}
-datatype prec = L | R of prec
-\end{lstlisting}
-while the stage \bbtwo\ result is always \texttt{()}.
-
-It is wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
-always returns \texttt{()}; worse yet, it has linear runtime!
-But optimizing this away is in general a global operation,
-so we instead solve this issue by adding a new staging annotation to \lang, called
-\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
-no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
-rules to \lang:
-\begin{mathpar}
-\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
-\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
-\end{mathpar}
-Here, the ``safe" judgment requires that $A$ has no $\to$ types in negative position.
-Intuitively, this is to ensure that stage \bbtwo\ code cannot be substituted into a monostage region.
-
-To split \texttt{mono} terms, we must produce a resumer with the correct shape
-for the enclosed term. This is difficult to produce statically, because the
-language has sums and general recursion. We instead produce it dynamically,
-using a new value in \langmono, \texttt{dummy}, which can be consumed as though
-it had any type:
-\begin{mathpar}
-\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
-\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
-\end{mathpar}
-The splitting rules are then
-\begin{mathpar}
-\infer{\splitone{\monoSt~e}A {(e',()), \_.{\tt dummy}} }{e \overset{C}{\rightarrow} e'}
-\end{mathpar}
-Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
-by performing the two rewrites in place:
-\begin{align*}
-\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
-e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
-\end{align*}
-For $\splitone e A {c,l.r}$, \texttt{dummy} can only appear as the 
-final value of $r$ when $e$ has no $\fut$ in its \lang\ type.
-Since the $\prev$ rule requires a circle type, \texttt{dummy} values cannot leak 
-into the resumers of stage \bbtwo\ terms.
-
 \input{figures/splitting-misc}
-
-% \subsection{Implementation}
-%
-% \crem{Say something about the implementation.}

### 2014-11-07 paper/splitting.tex

--- 2014-11-07 paper/splitting.tex
@@
 \input{figures/splitting-term}
-
-\subsection{Optimization: Monostage Functions}
-
-\textbf{KF: the missing point here is that the splitting algorithm was designed based on the local structure of \lang\ terms.
-To make the algorithm derivation case compelling,
-we need to optimize away some of that structure.  But this is hard to do through functions.}
-
-The above splitting rules are correct, but can be optimized in many cases. 
-For instance, if multiple subterms have a \texttt{()} precomputation, these can be
-combined into a single \texttt{()}, with the residual changed accordingly.
-This becomes more difficult to fix when it occurs within a recursive function.
-For example, consider the stage~\bbtwo\ term
-\begin{lstlisting}
-2`prev{
-  1`letfun fact (n : int) : int = 
-    if n <= 0 then 1 else fact(n-1)*n
-  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
-}`
-\end{lstlisting}
-which splits into:
-\begin{lstlisting}
-1`letfun fact (n : int) : int * prec = 
-  if n <= 0 then (1,L) 
-  else let (y,z) = fact(n-1) in (y*n,R z)
-in fact 5`
-
-2`l => 
-letfun fact (n : unit, l0 : prec) : unit = 
-  case l0 of L => () | R l1 => fact ((),l0)
-in (fact (pi2 l); pi1 l)-100`
-\end{lstlisting}
-Here, the stage \bbone\ function \texttt{fact} splits into two recursive
-functions, which compute a combined result (stage \bbone\ result and
-precomputation), and a stage \bbtwo\ result, respectively. This precomputation
-could be given the type
-\begin{lstlisting}
-datatype prec = L | R of prec
-\end{lstlisting}
-while the stage \bbtwo\ result is always \texttt{()}.
-
-It is wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
-always returns \texttt{()}; worse yet, it has linear runtime!
-But optimizing this away is in general a global operation,
-so we instead solve this issue by adding a new staging annotation to \lang, called
-\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
-no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
-rules to \lang:
-\begin{mathpar}
-\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
-\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
-\end{mathpar}
-Here, the ``safe" judgment requires that $A$ has no $\to$ types in negative position.
-Intuitively, this is to ensure that stage \bbtwo\ code cannot be substituted into a monostage region.
-
-To split \texttt{mono} terms, we must produce a resumer with the correct shape
-for the enclosed term. This is difficult to produce statically, because the
-language has sums and general recursion. We instead produce it dynamically,
-using a new value in \langmono, \texttt{dummy}, which can be consumed as though
-it had any type:
-\begin{mathpar}
-\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
-\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
-\end{mathpar}
-The splitting rules are then
-\begin{mathpar}
-\infer{\splitone{\monoSt~e}A {(e',()), \_.{\tt dummy}} }{e \overset{C}{\rightarrow} e'}
-\end{mathpar}
-Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
-by performing the two rewrites in place:
-\begin{align*}
-\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
-e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
-\end{align*}
-For $\splitone e A {c,l.r}$, \texttt{dummy} can only appear as the 
-final value of $r$ when $e$ has no $\fut$ in its \lang\ type.
-Since the $\prev$ rule requires a circle type, \texttt{dummy} values cannot leak 
-into the resumers of stage \bbtwo\ terms.
-
 \input{figures/splitting-misc}
-
-% \subsection{Implementation}
-%
-% \crem{Say something about the implementation.}

### 2014-11-08 paper/implementation.tex

\section {Implementation and Optimization}

We have implemented the semantics and splitting algorithm of \lang\ in Standard ML.
In addition to the features shown here, our implementation includes more base types, 
$n$-ary products and sums, if and let statements, and a basic elaboration step.

The rules of \cref{sec:splitting} can be optimized in many cases.
For instance, multiple trivial subterm precomputations can be merged:
\[
\infer {\splittwo {(e_1,e_2)} A {p,l.(r_1,r_2)}}{\splittwo {e_1} A {(),\_.r_1} & \splittwo {e_2} A {p,l.r_2}}
\]
This sort of optimization is easy, because it is local on the structure of $e$,
but the problem becomes more difficult when trying to optimize across functions, including recursive calls.
For example, consider the stage \bbtwo\ term,
\begin{lstlisting}
2`prev{
  1`letfun fact (n : int) : int = 
    if n <= 0 then 1 else fact(n-1)*n
  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
}`
\end{lstlisting}
Noting that the precomputation of the body of \texttt{fact} could be given the type,
\begin{lstlisting}
datatype prec = L | R of prec
\end{lstlisting}
the whole expression splits into:
\begin{lstlisting}
1`letfun fact (n : int) : int * prec = 
  if n <= 0 then (1,L) 
  else let (y,z) = fact(n-1) in (y*n,R z)
in fact 5`

2`l => 
letfun fact (n : unit, l0 : prec) : unit = 
  case l0 of L => () | R l1 => fact ((),l0)
in (fact (pi2 l); pi1 l)-100`
\end{lstlisting}

Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
always returns \texttt{()}; worse yet, it has linear runtime!
But recognizing and optimizing this away is in general a global operation,
so we instead solve this issue by adding a new staging annotation to \lang, called
\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
rules to \lang:
\begin{mathpar}
\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
\end{mathpar}
Here, the ``safe" judgment requires that $A$ has no $\to$ types in negative position.
Intuitively, this is to ensure that stage \bbtwo\ code cannot be substituted into a monostage region.

To split \texttt{mono} terms, we must produce a resumer with the correct shape
for the enclosed term. This is difficult to produce statically, because the
language has sums and general recursion. We instead produce it dynamically,
using a new value in \langmono, \texttt{dummy}, which can be consumed as though
it had any type:
\begin{mathpar}
\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
\end{mathpar}
The splitting rules are then
\begin{mathpar}
\infer{\splitone{\monoSt~e}A {(e',()), \_.{\tt dummy}} }{e \overset{C}{\rightarrow} e'}
\end{mathpar}
Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
by performing the two rewrites in place:
\begin{align*}
\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
\end{align*}
For $\splitone e A {c,l.r}$, \texttt{dummy} can only appear as the 
final value of $r$ when $e$ has no $\fut$ in its \lang\ type.
Since the $\prev$ rule requires a circle type, \texttt{dummy} values cannot leak 
into the resumers of stage \bbtwo\ terms.

### 2014-11-08 paper/splitting.tex
cleaning up latex warnings

@@
 The splitting algorithm for stage~\bbone\ terms, as specified by the
-judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
+judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplit}, proceeds
 recursively on the structure of~$e$.
@@
 As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
-as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, is defined on the local structure of $e$.
+as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplit}, is defined on the local structure of $e$.
 In the terminal cases of

### 2014-11-08 paper/implementation.tex
small edit to intro of implementation

@@
 
-The rules of \cref{sec:splitting} can be optimized in many cases.
-For instance, multiple trivial subterm precomputations can be merged:
+Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
+In many cases, the results can be improved by considering special cases in individual rules.
+For instance, trivial subterm precomputations needn't be saved at tuples:
 \[
@@
 \]
-This sort of optimization is easy, because it is local on the structure of $e$,
-but the problem becomes more difficult when trying to optimize across functions, including recursive calls.
+The problem becomes more difficult when trying to optimize across functions, including recursive calls.
 For example, consider the stage \bbtwo\ term,

### 2014-11-08 paper/semantics.tex
working on the semantics

--- 2014-11-05 paper/semantics.tex
@@
+\section{Terminology and Overview}
+
+\input{figures/splitting-overview}
+
+We present an overview of the formal development along with our
+terminology, as summarized in \figref{terminology}.
+
+\figref{terminology}
+
+Our starting point is a two-staged lambda calculus, whose syntax and
+types are very similar to the multi-stage languages considered in the
+literature on partial evaluation and meta programming.  Specifically,
+we consider a two-stage, which we call \lang\, which is variant of
+Davies`s multi-stage language.  Apart from the restriction to two
+stages instead of multiple stages, there are no significant
+differences between our syntax and the static semantics and those of
+Davies'. We restrict our attention to two stages, because it
+simplifies the (many) details of the splitting problem, without
+altering the essence of the problem.  We expect that the techniques
+presented here can be extended to multi-stage languages.
+
+The dynamic semantics of \lang\ differs from Davies's approach and
+similar approaches in meta programming literature, because it avoids
+duplication of staged work. Duplication takes place in meta
+programming when a second-stage computation is silently duplicated by
+the semantics.  In addition to leading to increases in run time, a
+duplicating semantics makes it harder to reason about the splitting
+algorithm, whose development and exposition is our main goal.  Apart
+from the non-duplication, the dynamic semantics of \lang\ is
+essentially identical to  Davies'.
+
+We define the dynamics semantics of \lang\ not as an explicit
+evaluation relation but as a relation that produces a term in an
+unstaged, completely standard language, which we refer to as
+\langmono\, because it is essentially lambda calculus extended with
+product types. Given a \lang\ term, the dynamic semantics can be used
+to reduce the term to a \langmono\ program, called {\em residual} by
+using one application of stage-1 evaluation, one application of
+reification, followed by one application of stage-2 evaluation.
+
+\ur{Splitting...}
+
 \section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
@@
 \input {figures/statics}
-
-\lang\ is a typed two-stage lambda calculus with products, sums, and isorecursive types.
-Its statics are adapted from the linear temporal logic given by \cite{davies96},
-restricted to two stages and extended with general sums and recursion.%
-\footnote{This restriction to two stages is made primarily for simplicity of
-presentation. All of these techniques could feasibly be extended to more
-complicated stage systems.} 
-Although \cite{davies96} was mostly interested in describing partial evaluation,
-the system is equally apt as an input to splitting.
 
@@
 
-In the quickselect example, stage~\bbone\ terms are contained within
-stage~\bbtwo\ terms---for example, in the recursive call to {\tt qsStaged}. In
-this case, if we use an ordinary evaluation strategy reducing outermost redexes
-first, we would no longer have staged evaluation, as stage~\bbtwo\ code would
-be evaluated before stage~\bbone\ code.
+As is the case in essentially all staged language, the dynamic
+semantics evaluates a term to a residual (a term in the target
+language \langmono) in two stages. In the first stage, all stage-1
+terms are evaluated; in the second stage all stage-2 terms are
+evaluated.  To eliminate duplication, our semantics lifts out each
+stage-2 term nested inside of a stage-1 expressions, into table which
+makes it possible for the stage-2 term, actually the corresponding
+residual obtained by reducing all stage-1 subterms, to be referred to
+by use a dynamically generated variable (a label).
 
-Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
-subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
-a stage \bbtwo\ term with no stage~\bbone\ content, 
-which can be described as a term in a monostaged language called \langTwo. 
-Then, at stage~\bbtwo, we complete
-evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
-rules for this judgment are not shown, but they are standard.)
+Since stage-1 evaluation does not evaluate stage-2 terms and since
+stage-2 terms may be nested inside stage-1 terms, stage-1 evaluation
+cannot reduce \lang\ terms to values but only to {\em partial values},
+which are stage~\bbone\ terms that have been evaluated except for
+their stage-2 subterms, which can only be~\bbtwo\ variables wrapped in
+$\next$ blocks.  \figref{validPvalues} defines partial values.
 
-\subsubsection{Stage~\bbone\ Evaluation}
-
-To gain intuition about the challenges of implementing this staged dynamic
-semantics, consider the program:
-\begin{lstlisting}
-1`#2 (next {`2`f 20`1`}, 2+3)`
-\end{lstlisting}
-This is a stage~\bbone\ expression of type $\rmint~@~\bbone$; the pair inside it is a
-stage~\bbone\ expression of type $(\fut\rmint)\times\rmint~@~\bbone$; and ${\tt f} :
-\rmint \to \rmint~@~\bbtwo$ is an expensive stage~\bbtwo\ function which we would like
-to avoid evaluating. A conventional call-by-value semantics would fully
-evaluate both components of the pair before projecting the second component.
-The problem is that, while \verb|next {f 20}| is not a value (in the sense that
-additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
-evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
-evaluation.
-
-Intuitively, the solution is to designate \verb|next {f 20}| as a value
-\emph{in stage \bbone}, even though it requires additional evaluation in stage
-\bbtwo. Therefore, the pair evaluates to
-\begin{lstlisting}
-1`(next {`2`f 20`1`}, 5)`
-\end{lstlisting}
-and the projection, in turn, immediately evaluates to \verb|5|.
-
-Now consider a more complex example where stage~\bbone\ evaluation must
-substitute such an incompletely-evaluated expression. The following
-stage-\bbtwo\ term has type $\rmint$:
-\begin{lstlisting} 
-2`prev{`
-  1`let x = (next {`2`f 20`1`}, 3+4) in
-  next{` 2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}` 1`}`
-2`}`
-\end{lstlisting}
-As with the simpler example, this term does not fully reduce at stage~\bbone, 
-because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.
-
-Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
-substitute it for the three occurrences of \verb|x| in the body of the
-\verb|let| expression, yielding
-\begin{lstlisting} 
-2`prev{`
-  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
-      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
-      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
-}`
-\end{lstlisting}
-The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
-So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
-This search process is called {\em specialization}.
-In our example this means reducing all three projections to give,
-\begin{lstlisting} 
-2`prev{`1`
-  next{`2` 
-    prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
-  `1`}`2`
-}`
-\end{lstlisting}
-Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
-\begin{lstlisting} 
-2`(f 20) * (f 20) * 7`
-\end{lstlisting}
-But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
-There are some systems, typically called {\em metaprogramming}, 
-where this duplication would be considered intended behavior.
-For our applications, we find the duplication undesirable and avoid it.
-
-To get this behavior, we do not treat \verb|next {f 20}| as a value to be substituted, 
-but instead bind \verb|f 20| to a fresh stage~\bbtwo\ variable $\mathtt{\hat y}$
-and then substitute that variable.
-
-More precisely, when evaluating the \verb|next|, we construct an explicit
-substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
-fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
-containing $\prev$ block:
-\begin{lstlisting} 
-2`prev {
-[yhat|->f 20]
-  `1`let x = (next{`2`yhat`1`}, 7) in
-  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
-}`
-\end{lstlisting}
-%As a convention, we render the new variable with a %stylish and fashionable
-%hat.  
-We proceed with stage~\bbone\ evaluation, which now duplicates the variable
-$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
-\begin{lstlisting} 
-2`prev {
-[yhat|->f 20]
-  `1`next{`2`
-    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
-    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
-    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
-  `1`}`2`
-}`
-\end{lstlisting}
-We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
-\begin{lstlisting} 
-2`prev {
-[yhat|->f 20]
-    `1`next{`2`yhat * yhat * 7`1`}
-`2`}`
-\end{lstlisting}
-Once again, we lift the results of specialization into a substitution:
-\begin{lstlisting} 
-2`prev {
-[yhat|->f 20]
-[zhat|->yhat * yhat * 7]
-    `1`next{`2`zhat`1`}`2`
-}`
-\end{lstlisting}
-Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
-\begin{lstlisting} 
-2`let yhat = f 20 in
-let zhat = yhat * yhat * 7 in zhat`
-\end{lstlisting}
-
-This concludes stage \bbone\ evaluation of the program---we have reduced all
-stage \bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
-evaluation, $\tworedsym$, then reduces this to a numeral.
-
-\subsubsection{Evaluation and Specialization}
-\label{ssec:dynamics}
+The dynamic semantics consist of two evaluation relations $\redonesym$
+and $\redtwosym$, called {\em 1-evaluation} and {\em 2-evaluation},
+that respectively evaluates stage-1 and stage-2 expressions.
 
 \input{figures/dynamics}
+
+The relation 1-evaluation maps a stage-1 term to a partial value and a
+residual table.
+
+ In the example above, 
+\begin{lstlisting} 
+1`(next {`2`yhat`1`},7)`
+\end{lstlisting}
+is a partial value.
+
+In the example
+above,
+\begin{lstlisting} 
+2`let zhat = yhat*yhat*7 in zhat`
+\end{lstlisting}
+is a residual.
 
@@
 
-\emph{Partial values} are stage~\bbone\ terms that have been fully
-evaluated, but which may contain stage~\bbtwo\ variables wrapped in $\next$
-blocks.  In the example above, 
-\begin{lstlisting} 
-1`(next {`2`yhat`1`},7)`
-\end{lstlisting}
-is a partial value.
-\emph{Residuals} ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
-stage \bbone\ subexpressions have all been fully evaluated. In the example
-above,
-\begin{lstlisting} 
-2`let zhat = yhat*yhat*7 in zhat`
-\end{lstlisting}
-is a residual.
-
 Note that partial values and residuals are both terms for which stage \bbone\
@@
 into a single one.
+
+The relation 2-evaluation takes a stage-2 term and maps to a residual.
+
+To generate the residual table, the dynamics semantics relies on a
+{\em reification} relation, $\reifysym$, that maps a table to a
+residual, essentially generating a target expression representing the
+computations in the table.
+
+We can evaluate a \lang\ term as follows: we first apply 1-evaluation
+to obtain a partial value and a residual table, we then construct a
+2-term from the partial value and the residual table by using
+reification, and then reduce this term to a residual via 2-evaluation.
+
+Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
+subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
+a stage \bbtwo\ term with no stage~\bbone\ content, 
+which can be described as a term in a monostaged language called \langTwo. 
+Then, at stage~\bbtwo, we complete
+evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
+rules for this judgment are not shown, but they are standard.)
 
@@
 
+%% Umut: superseded by what is above.
+%% Version Nov 7, 2014
+%% \section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
+%% \label{sec:semantics}
+
+%% \input {figures/grammars}
+%% \input {figures/statics}
+
+%% \lang\ is a typed two-stage lambda calculus with products, sums, and
+%% isorecursive types.  Its statics are adapted from the linear temporal
+%% logic given by \cite{davies96}, restricted to two stages and extended
+%% with general sums and recursion.%
+%% \footnote{This restriction to two stages is made primarily for simplicity of
+%% presentation. All of these techniques could feasibly be extended to more
+%% complicated stage systems.} 
+%% Although \cite{davies96} was mostly interested in describing partial evaluation,
+%% the system is equally apt as an input to splitting.
+
+%% \subsection{Statics}
+
+%% In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
+%% stage~\bbone\ term. To ensure that the staging features are only applied to
+%% valid terms, we index our judgments not only by a type, but also a stage $w$.
+
+%% The type validity judgment $\Delta \vdash A \istypewor$, defined in
+%% \cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
+%% example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
+%% $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
+%% $\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
+%% typing judgment only produces valid types.
+%% %(by induction on the derivation of the typing judgment).
+%% % under the assumptions in $\Delta$. This context is only augmented by recursive types.
+
+%% Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
+%% language features (i.e., the introductory and eliminatory forms for product,
+%% sum, and function types) preserve the stage of their subterms, and work at both
+%% stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
+%% \lang\ would simply consist of two non-interacting copies of a standard lambda
+%% calculus. Variables in the context are annotated with the stage at which they
+%% were introduced and can only be used at the same stage.
+
+%% Note that there is no way to eliminate a term of type $\fut A$ into a stage
+%% \bbone\ term---the only elimination form is $\prev$, which results in a stage
+%% \bbtwo\ term. This is why well-typed terms cannot have any information flow
+%% from stage \bbtwo\ to stage \bbone, and what makes splitting possible.
+
+%% \subsection{Dynamics}
+%% \label{sec:stagedsemantics}
+
+%% Specifically, the dynamic semantics consist of two evaluation
+%% relations $\redonesym$ and $\redtwosym$, called {\em 1-evaluation} and
+%% {\em 2-evaluation}, that respectively evaluates stage-1 and stage-2
+%% expressions. The relation 1-evaluation maps a stage-1 term to a
+%% partial value and a residual table. A {\em partial value}, can be
+%% thought as a value except that it might include references to stage-2
+%% code stored in the {\em residual table}, which maps variables to
+%% residuals.  The relation 2-evaluation takes a stage-2 term and maps to
+%% a residual.  To generate the residual table, the dynamics semantics
+%% relies on a {\em reification} relation, $\reifysym$, that maps a table
+%% to a residual, essentially generating a target expression representing
+%% the computations in the table.  We can evaluate a \lang\ term as
+%% follows: we first apply 1-evaluation to obtain a partial value and a
+%% residual table, we then construct a 2-term from the partial value and
+%% the residual table by using reification, and then reduce this term to
+%% a residual via 2-evaluation.
+
+%% In the quickselect example, stage~\bbone\ terms are contained within
+%% stage~\bbtwo\ terms---for example, in the recursive call to {\tt qsStaged}. In
+%% this case, if we use an ordinary evaluation strategy reducing outermost redexes
+%% first, we would no longer have staged evaluation, as stage~\bbtwo\ code would
+%% be evaluated before stage~\bbone\ code.
+
+%% Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
+%% subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
+%% a stage \bbtwo\ term with no stage~\bbone\ content, 
+%% which can be described as a term in a monostaged language called \langTwo. 
+%% Then, at stage~\bbtwo, we complete
+%% evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
+%% rules for this judgment are not shown, but they are standard.)
+
+%% \subsubsection{Stage~\bbone\ Evaluation}
+
+%% To gain intuition about the challenges of implementing this staged dynamic
+%% semantics, consider the program:
+%% \begin{lstlisting}
+%% 1`#2 (next {`2`f 20`1`}, 2+3)`
+%% \end{lstlisting}
+%% This is a stage~\bbone\ expression of type $\rmint~@~\bbone$; the pair inside it is a
+%% stage~\bbone\ expression of type $(\fut\rmint)\times\rmint~@~\bbone$; and ${\tt f} :
+%% \rmint \to \rmint~@~\bbtwo$ is an expensive stage~\bbtwo\ function which we would like
+%% to avoid evaluating. A conventional call-by-value semantics would fully
+%% evaluate both components of the pair before projecting the second component.
+%% The problem is that, while \verb|next {f 20}| is not a value (in the sense that
+%% additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
+%% evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
+%% evaluation.
+
+%% Intuitively, the solution is to designate \verb|next {f 20}| as a value
+%% \emph{in stage \bbone}, even though it requires additional evaluation in stage
+%% \bbtwo. Therefore, the pair evaluates to
+%% \begin{lstlisting}
+%% 1`(next {`2`f 20`1`}, 5)`
+%% \end{lstlisting}
+%% and the projection, in turn, immediately evaluates to \verb|5|.
+
+%% Now consider a more complex example where stage~\bbone\ evaluation must
+%% substitute such an incompletely-evaluated expression. The following
+%% stage-\bbtwo\ term has type $\rmint$:
+%% \begin{lstlisting} 
+%% 2`prev{`
+%%   1`let x = (next {`2`f 20`1`}, 3+4) in
+%%   next{` 2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}` 1`}`
+%% 2`}`
+%% \end{lstlisting}
+%% As with the simpler example, this term does not fully reduce at stage~\bbone, 
+%% because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.
+
+%% Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
+%% substitute it for the three occurrences of \verb|x| in the body of the
+%% \verb|let| expression, yielding
+%% \begin{lstlisting} 
+%% 2`prev{`
+%%   1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
+%%       * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
+%%       * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
+%% }`
+%% \end{lstlisting}
+%% The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
+%% So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
+%% This search process is called {\em specialization}.
+%% In our example this means reducing all three projections to give,
+%% \begin{lstlisting} 
+%% 2`prev{`1`
+%%   next{`2` 
+%%     prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
+%%   `1`}`2`
+%% }`
+%% \end{lstlisting}
+%% Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
+%% \begin{lstlisting} 
+%% 2`(f 20) * (f 20) * 7`
+%% \end{lstlisting}
+%% But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
+%% There are some systems, typically called {\em metaprogramming}, 
+%% where this duplication would be considered intended behavior.
+%% For our applications, we find the duplication undesirable and avoid it.
+
+%% To get this behavior, we do not treat \verb|next {f 20}| as a value to be substituted, 
+%% but instead bind \verb|f 20| to a fresh stage~\bbtwo\ variable $\mathtt{\hat y}$
+%% and then substitute that variable.
+
+%% More precisely, when evaluating the \verb|next|, we construct an explicit
+%% substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
+%% fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
+%% containing $\prev$ block:
+%% \begin{lstlisting} 
+%% 2`prev {
+%% [yhat|->f 20]
+%%   `1`let x = (next{`2`yhat`1`}, 7) in
+%%   next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}`2`
+%% }`
+%% \end{lstlisting}
+%% %As a convention, we render the new variable with a %stylish and fashionable
+%% %hat.  
+%% We proceed with stage~\bbone\ evaluation, which now duplicates the variable
+%% $\mathtt{\hat y}$ rather than the expression \verb|f 20|.
+%% \begin{lstlisting} 
+%% 2`prev {
+%% [yhat|->f 20]
+%%   `1`next{`2`
+%%     prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
+%%     prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
+%%     hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
+%%   `1`}`2`
+%% }`
+%% \end{lstlisting}
+%% We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
+%% \begin{lstlisting} 
+%% 2`prev {
+%% [yhat|->f 20]
+%%     `1`next{`2`yhat * yhat * 7`1`}
+%% `2`}`
+%% \end{lstlisting}
+%% Once again, we lift the results of specialization into a substitution:
+%% \begin{lstlisting} 
+%% 2`prev {
+%% [yhat|->f 20,
+%%  zhat|->yhat * yhat * 7]
+%%     `1`next{`2`zhat`1`}`2`
+%% }`
+%% \end{lstlisting}
+%% Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
+%% \begin{lstlisting} 
+%% 2`let yhat = f 20 in
+%% let zhat = yhat * yhat * 7 in zhat`
+%% \end{lstlisting}
+
+%% This concludes stage \bbone\ evaluation of the program---we have reduced all
+%% stage \bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
+%% evaluation, $\tworedsym$, then reduces this to a numeral.
+
+%% \subsubsection{Evaluation and Specialization}
+%% \label{ssec:dynamics}
+
+%% \input{figures/dynamics}
+
+%% The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
+%% described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
+%% (specialization), and the auxiliary $\reifysym$ (reification). 
+%% Along with these are three judgments identifying forms of values:
+%% $\pvalsym$ (partial values), $\ressym$ (residuals), 
+%% and the auxiliary $\fcon$ (residual table).
+%% These judgments are related as follows:
+%% \begin{itemize}
+%% \item Evaluation sends a stage~\bbone\ term to a partial value.
+%% In order to avoid duplicating subterms, evaluation collects explicit
+%% substitutions into a single {\em residual table} that is also the output of
+%% evaluation.
+%% \item Specialization sends a stage~\bbtwo\ term to a residual.
+%% \item Reification sends a single residual and a residual table to another residual.  
+%% \item Evaluation depends on specialization for $\next$ terms, which have
+%% stage~\bbtwo\ subterms.
+%% \item Specialization depends on evaluation for $\prev$ and $\pause$, which have
+%% stage~\bbone\ subterms. The residual table produced by evaluation is also
+%% reified here.
+%% \end{itemize}
+
+%% \emph{Partial values} are stage~\bbone\ terms that have been fully
+%% evaluated, but which may contain stage~\bbtwo\ variables wrapped in $\next$
+%% blocks.  In the example above, 
+%% \begin{lstlisting} 
+%% 1`(next {`2`yhat`1`},7)`
+%% \end{lstlisting}
+%% is a partial value.
+%% \emph{Residuals} ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
+%% stage \bbone\ subexpressions have all been fully evaluated. In the example
+%% above,
+%% \begin{lstlisting} 
+%% 2`let zhat = yhat*yhat*7 in zhat`
+%% \end{lstlisting}
+%% is a residual.
+
+%% Note that partial values and residuals are both terms for which stage \bbone\
+%% computation has completed; however, partial values are stage~\bbone\ terms,
+%% while residuals are stage~\bbtwo\ terms.
+
+%% The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
+%% \bbone\ term%
+%% \footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
+%% the type system ensures that those must occur under a $\next$. Consequently
+%% $\redonesym$ will never directly encounter a variable.}
+%% %
+%% to a {\em residual table} $\xi$ and a partial value $v$. The residual
+%% table implements the explicit substitutions in our example---it maps fresh
+%% stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
+%% residuals. For non-staging features of \lang, $\redonesym$ is essentially
+%% standard call-by-value evaluation, and gathers the subterms' residual tables
+%% into a single one.
+
+%% \crem{Can we cut \ref{fig:diaSemanticsSpec} from the paper? Perhaps just put a
+%% few representative rules into \ref{fig:diaSemantics}?}
+
+%% At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
+%% subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
+%% which takes a stage \bbtwo\ term to a residual. 
+%% For the most part (\ref{fig:diaSemanticsSpec}), specialization simply recurs into
+%% all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
+%% which produces a residual table $\xi$ and (by canonical forms) an expression
+%% $\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
+%% bindings enclosing $\mathtt{\hat y}$.
+%% Once specialization reduces $e$ to a residual $q$, the output of $\redonesym$ is a
+%% fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a residual
+%% table which maps that variable to the residual ($\hat y \mapsto q$).
+
+%% Within specialization lies a subtle---if perhaps unintuitive---feature.  
+%% Observe that specialization will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
+%% statement in its search for stage~\bbone\ code. 
+%% Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
+%% and so a term like 
+%% \begin{lstlisting} 
+%% 1`next{
+%%   `2`if true 
+%%   then hold{`1`3+4`2`} 
+%%   else prev{`1`spin() (* loops forever *)`2`}`1`
+%% }`
+%% \end{lstlisting}
+%% will fail to evaluate at stage \bbone.
+%% Although it may seem undesirable, this behavior is a deliberate feature.
+%% Indeed it is critical to the quickselect example, 
+%% wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
+%% As always, it is the programmer's responsibility to ensure that her staging annotations 
+%% do not cause a program to loop forever.
+
+%% The context ($\Gamma$) keeps track of stage \bbtwo\ variables in the input term. 
+%% These both appear in the original program at stage \bbtwo\ and are inserted by the semantics.
+
+%% As an optimization, we can include the special-case rule,
+%% \begin{mathpar}
+%% \inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
+%% \end{mathpar}
+%% to avoid one-for-one variable bindings in the residual.
+
+%% \subsubsection {Top-Level Evaluation}
+%% \label{sec:partialeval}
+
+%% % KAYVONF: good statement, but hold out for now
+%% %The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must %evaluate it many times.
+
+%% These dynamics allow us to define a partial evaluator for \lang\ by identifying
+%% static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
+%% encode $f$ as a \lang\ expression with a function type of the form
+%% $A\to\fut(B\to C)$.%
+%% %\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
+%% %the following higher-order function which makes the adjustment:
+%% %\begin{lstlisting} 
+%% %let adjust (f : $int * int -> $int) =
+%% %  fn (p : int) => 
+%% %    next{
+%% %      fn (b : int) => 
+%% %        prev{f (next {b}, p)}
+%% %    }
+%% %\end{lstlisting}}
+%% %
+%% Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.
+
+%% Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
+%% function:
+%% $\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
+%% The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
+%% C)$, which by canonical forms must have the form $v = \next~\hat y$. 
+%% Next, we reify this environment into a sequence of \verb|let|-bindings
+%% enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
+%% Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
+%% and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
+
+%% That this sequence of evaluations is in fact staged follows from our
+%% characterizations of partial values and residuals, that $\redonesym$
+%% outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
+
+%% For example, to specialize quickselect to a particular list, 
+%% \begin{lstlisting}
+%% qsStaged [5,2,7,4,1]
+%% \end{lstlisting}
+%% \TODO finish example
+
 %\begin{remark}

### 2014-11-08 paper/splitting.tex
Making text consistent with new terminology.

@@
 
-% Intuitively, we require that the terms obtained by splitting a term $e$, when
-% evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
-% the case of stage~\bbone\ terms, our dynamic semantics produce partial values
-% containing information at \emph{both} stages, which goes against the idea that
-% splitting separates the subcomputations belonging to each stage.
-
-Recall that \lang\ partial values and residual tables together may
+Recall that partial values and residual tables produced by \bbone-evaluation may
 contain content at \emph{both} stages. For example the term:
@@
 \end{lstlisting}
-Intuitively, splitting this term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value \texttt{3},
+To match the results of \bbone-evaluation,
+splitting the term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value, \texttt{3},
 and an~$e_2$ corresponding to the stage~\bbtwo\ computation: \\
@@
 
-To precisely specify the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
+to precisely specific the stage~\bbone\ and stage~\bbtwo\ components of \bbone-evaluation results, we define 
 \emph{masking} functions, written $\masko{{-};{-}}$ and $\maskt{{-};{-}}$, 
 which take a residual table and partial value as input and emit a term in \langmono\
-that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
-masking functions.
+that corresponds to only the content belonging to a single stage.
+\ref{fig:valMask} defines these masking functions.
 
@@
 content, $\masko{v}$ replaces all $\next$ subterms of~$v$ with \texttt{()}.
-Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
+Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits the body of lambdas as a general stage~\bbone\ term (as described in \ref{sec:split-one}), and returns
 the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
@@
 exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
-requires access to the residual table $\xi$ because all stage~\bbtwo\
-content in~$v$ has been lifted out into~$\xi$. Stage~\bbtwo\ masking discards
+requires access to the residual table $\xi$ because after \bbone-evaluation, all stage~\bbtwo\
+has been lifted out of~$v$. Stage~\bbtwo\ masking discards
 stage~\bbone\ content in~$v$ by replacing base constants with~\texttt{()},
@@
 returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from~$v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
-producing a term in \langmono. For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
+producing a term in \langmono. This term reduces to a value that with the same structure as~$v$.
+For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
 \begin{lstlisting}

### 2014-11-09 paper/figures/splitting-overview.tex
minor tweaks to figures
working on the semantics

@@
-\begin{figure*}
+\begin{figure}
+\textbf{Languages:}
+\begin{itemize}
+\item \lang: two-staged lambda calculus with product, sum, and
+  recursive types
+\item \langmono: an unstaged lambda calculus with products.
+\end{itemize}
 
-\textbf{\lang\ Evaluation:}
+\textbf{\lang\ Evaluation Relations:}
+\begin{itemize}
+\item 
+\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.
 
-\hspace{2em}\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.
+\item
+\bbtwo-Evaluation: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo
 
-\hspace{2em}\bbtwo-Evaluation: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo
+\item 
+Table reification: \ur{Fill in...}
+\end{itemize}
 
@@
 
-\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputatiosn in $e$)
+\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputations in $e$)
 
@@
 \caption{Summary of \lang\ evaluation and splitting.}
-\label{fig:termSplitTwo}
-\end{figure*}
+\label{fig:terminology}
+\label{fig:termSplitSummary}
+\end{figure}

### 2014-11-09 paper/semantics.tex
working on the semantics

@@
 language \langmono) in two stages. In the first stage, all stage-1
-terms are evaluated; in the second stage all stage-2 terms are
-evaluated.  To eliminate duplication, our semantics lifts out each
-stage-2 term nested inside of a stage-1 expressions, into table which
-makes it possible for the stage-2 term, actually the corresponding
-residual obtained by reducing all stage-1 subterms, to be referred to
-by use a dynamically generated variable (a label).
+terms are evaluated by a relation, called {\em 1-evaluation}; in the
+second stage all stage-2 terms are evaluated by a relation
+$\redtwosym$, called {\em 2-evaluation}.  To eliminate duplication,
+our semantics lifts out each stage-2 term nested inside of a stage-1
+expressions, into a table which makes it possible for the stage-2
+term---more precisely the corresponding residual obtained by reducing
+all stage-1 subterms---to be referred to by use a dynamically
+generated variable (a label).
 
 Since stage-1 evaluation does not evaluate stage-2 terms and since
-stage-2 terms may be nested inside stage-1 terms, stage-1 evaluation
-cannot reduce \lang\ terms to values but only to {\em partial values},
-which are stage~\bbone\ terms that have been evaluated except for
-their stage-2 subterms, which can only be~\bbtwo\ variables wrapped in
-$\next$ blocks.  \figref{validPvalues} defines partial values.
+stage-2 terms may be nested inside stage-1 terms, 1-evaluation cannot
+reduce \lang\ terms to values.  Instead 1-evaluation yields a {\em
+  partial value}, which which is a stage~\bbone\ term that have been
+evaluated except for its stage-2 subterms, which can only
+be~\bbtwo\ variables wrapped in $\next$ blocks.  \figref{validPvalues}
+defines partial values.
 
-The dynamic semantics consist of two evaluation relations $\redonesym$
-and $\redtwosym$, called {\em 1-evaluation} and {\em 2-evaluation},
-that respectively evaluates stage-1 and stage-2 expressions.
+In order to avoid duplication of stage-2 computations, 1-evaluation
+labels stage-2 computations by generating variable symbols for them as
+needed and stores their residuals in a {\em residual table} keyed by
+the variable.
 
-\input{figures/dynamics}
+The relation 1-evaluation, defined by $\redonesym$ judgment as shown
+in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
+\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
+the type system ensures that those must occur under a $\next$. Consequently
+$\redonesym$ will never directly encounter a variable.}
+%
+to a residual table $\xi$ and a partial value $v$.  For non-staging
+features of \lang, $\redonesym$ is essentially standard call-by-value
+evaluation but in addition gathers the subterms' residual tables into
+a single table. \ur{The syntax of residual tables should be defined in
+  the terminology figure} Residual table entries are created by the
+evaluation of \next: 1-evaluation reduces a \next expression by
+reducing the body of the \next with 2-evaluation to obtain a
+residual $q$ and returns a residual table that maps a freshly
+generated variable to the residual.  \ur{Variable should be fresh in
+  the rule.}
 
-The relation 1-evaluation maps a stage-1 term to a partial value and a
-residual table.
+The relation 2-evaluation, defined by $\redtwosym$ judgment, as shows
+in \figref{diaSemantics}, takes a stage \bbtwo\ term and reduces it to
+a residual.  Apart from the evaluation of \prev and \pause, the
+evaluation is straightforward. To reduce a \prev term, 2-evaluation
+reduces the body by 1-evaluation to obtain a residual table and a
+partial value.  It then reifies the residual table and the partial
+value into an residual term, by creating for each variable-residual
+pair in the table a corresponding variable binding in the residual,
+finally ending with the partial value.
+\ur{There seems to be errors in the semantics in the handling of
+  recursive types.  I don't see any reason to do any reductions here.
+If you want to, the description above needs to be updated.}
 
- In the example above, 
-\begin{lstlisting} 
-1`(next {`2`yhat`1`},7)`
-\end{lstlisting}
-is a partial value.
+...UMUT: UP TO HERE...
 
-In the example
-above,
-\begin{lstlisting} 
-2`let zhat = yhat*yhat*7 in zhat`
-\end{lstlisting}
-is a residual.
+ In addition, 1-evaluation {\em reifies} the contents of
+the residual table into a residual when changing from stage-2 to stage
 
-The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
-described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
-(specialization), and the auxiliary $\reifysym$ (reification). 
+%% Note that partial values and residuals are both terms for which stage \bbone\
+%% computation has completed; however, partial values are stage~\bbone\ terms,
+%% while residuals are stage~\bbtwo\ terms.
+
+ The residual
+table implements the explicit substitutions in our example---it maps fresh
+stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
+residuals.
+
+In order to avoid duplicating subterms, evaluation collects explicit
+substitutions into a single {\em residual table} that is also the output of
+evaluation.
+
+ the auxiliary $\reifysym$ (reification). 
 Along with these are three judgments identifying forms of values:
@@
 \item Evaluation sends a stage~\bbone\ term to a partial value.
-In order to avoid duplicating subterms, evaluation collects explicit
-substitutions into a single {\em residual table} that is also the output of
-evaluation.
+
 \item Specialization sends a stage~\bbtwo\ term to a residual.
+
 \item Reification sends a single residual and a residual table to another residual.
+
 \item Evaluation depends on specialization for $\next$ terms, which have
@@
 \end{itemize}
-
-Note that partial values and residuals are both terms for which stage \bbone\
-computation has completed; however, partial values are stage~\bbone\ terms,
-while residuals are stage~\bbtwo\ terms.
-
-The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
-\bbone\ term%
-\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
-the type system ensures that those must occur under a $\next$. Consequently
-$\redonesym$ will never directly encounter a variable.}
-%
-to a {\em residual table} $\xi$ and a partial value $v$. The residual
-table implements the explicit substitutions in our example---it maps fresh
-stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
-residuals. For non-staging features of \lang, $\redonesym$ is essentially
-standard call-by-value evaluation, and gathers the subterms' residual tables
-into a single one.
 

### 2014-11-09 paper/figures/splitting-overview.tex
BUGFIX: fixed error in resumer in Stage-2 splitting.
Small cleanup in 4.2 and 4.3 to be consistent with terminology.

--- 2014-11-07 paper/figures/splitting-overview.tex
@@
 
-\hspace{2em}\bbtwo-Splitting Correctness: If $e~\redtwosym~q$, and $q~\redsym~v$, then \texttt{(let l=b in r)}~$\redsym~v$
+\hspace{2em}\bbtwo-Splitting Correctness: If $e~\redtwosym~q$, and $q~\redsym~v$, then \texttt{(let l=p in r)}~$\redsym~v$
 
 \caption{Summary of \lang\ evaluation and splitting.}
-\label{fig:termSplitTwo}
+\label{fig:termSplitSummary}
 \end{figure*}

### 2014-11-09 paper/splitting.tex
BUGFIX: fixed error in resumer in Stage-2 splitting.
Small cleanup in 4.2 and 4.3 to be consistent with terminology.

@@
 Splitting a stage~\bbone\ term $e$ in \lang\ yields a term 
-$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
-To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$,
+$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer term} $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
+To ensure the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$,
 we require that $e_1$ reduce to the tuple $(y,b)$,
-where $y$ is identical to $\masko{\xi;v}$, and 
+where $y$, called the \emph{\bbone-result} of $e$, is identical to $\masko{\xi;v}$, and that
 $\letin l b r$ reduces to the same value as $\maskt{\xi;v}$.
-Since~$e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (a value we call the \emph{stage~\bbone\ result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the \emph{boundary value} $b$ required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.
+(We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.) Since~$e_1$ performs both stage~\bbone\ subcomputations needed to produce the \bbone-result and subcomputations (refered to as \emph{precomputations}) that produce the boundary value, we refer to this term as the \emph{combined term}, and hereafter denote it as~$c$.
 
@@
 (performing all stage~\bbone\ subcomputations in $e$) and a
-resumer function $e_2=l.r$ (performing all stage~\bbtwo\ subcomputations) such that when 
+resumer term $e_2=l.r$ (performing all stage~\bbtwo\ subcomputations) such that when 
 $e$ reduces to the value~$v$, $\letin l p r$ reduces to an identical value in \langmono.
@@
 the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
-to compute the input expected by the resumer.
+to compute the boundary value expected by the resumer.
 

### 2014-11-09 paper/implementation.tex
talked about new optimization, added iota example

@@
 \]
+This sort of optimization does not affect the contract of splitting at all,
+because the contract only requires that there be a precomputation that the residual can accept,
+with no restrictions on the form of that precomputation.
+Further gains can be realized by changing the contract of splitting itself.
+
+Let a type be {\em present} if it has no $\to$s or $\fut$s in its type.
+Observe that for any $v$ with a present type, $\maskt{v}$ will always be a nested bunch of \texttt{()}s,
+which is to say that it carries no information.
+
+\nr{We probably need an example of this.}
+
+Of course, \texttt{()} is a more compact representation that carries no information.
+Thus, we can change the definition of masking so that $\maskt{v}$ is always \texttt{()},
+and change the splitting rules accordingly.
+
+Back to the boundary type problem.
 The problem becomes more difficult when trying to optimize across functions, including recursive calls.
@@
 \infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
-\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
-\end{mathpar}
-Here, the ``safe" judgment requires that $A$ has no $\to$ types in negative position.
-Intuitively, this is to ensure that stage \bbtwo\ code cannot be substituted into a monostage region.
-
-To split \texttt{mono} terms, we must produce a resumer with the correct shape
-for the enclosed term. This is difficult to produce statically, because the
-language has sums and general recursion. We instead produce it dynamically,
-using a new value in \langmono, \texttt{dummy}, which can be consumed as though
-it had any type:
-\begin{mathpar}
-\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
-\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
-\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
-\end{mathpar}
-The splitting rules are then
-\begin{mathpar}
-\infer{\splitone{\monoSt~e}A {(e',()), \_.{\tt dummy}} }{e \overset{C}{\rightarrow} e'}
+\infer{A\to B~\mathrm{safe}}{A~\mathrm{present} & B~\mathrm{safe}} \and
+\infer{A~\mathrm{safe}}{A~\mathrm{present}} \and
+\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}} \and
+\infer{\splitone{\monoSt~e}A {(e',()), \_.r} }{e \overset{C}{\rightarrow} e' & A \overset{G}{\rightarrow} r} \and
+\infer{A \overset{G}{\rightarrow} ()}{A ~\mathrm{present}} \and
+\infer{(A \to B)\overset{G}{\rightarrow} r}{B \overset{G}{\rightarrow} \lambda \_.r }
 \end{mathpar}
@@
 \end{align*}
-For $\splitone e A {c,l.r}$, \texttt{dummy} can only appear as the 
-final value of $r$ when $e$ has no $\fut$ in its \lang\ type.
-Since the $\prev$ rule requires a circle type, \texttt{dummy} values cannot leak 
-into the resumers of stage \bbtwo\ terms.
+And then $\overset{G}{\rightarrow}$ generates a resumer that corresponds to the type $A$.
+It operates inductively on the safety judgement of $A$.
+At the safe $A$ base case, we make use of the prior optimization that says all safe $A$s have a resumer that evaluates to \texttt{()}.

### 2014-11-09 paper/semantics.tex
complete rewrite of section on the language and semantics

@@
 
+\input{figures/dynamics}
+
 Since stage-1 evaluation does not evaluate stage-2 terms and since
@@
 residual $q$ and returns a residual table that maps a freshly
-generated variable to the residual.  \ur{Variable should be fresh in
-  the rule.}
+generated variable to the residual.
 
@@
 in \figref{diaSemantics}, takes a stage \bbtwo\ term and reduces it to
-a residual.  Apart from the evaluation of \prev and \pause, the
+a residual.  Apart from the evaluation of \prev\ and \pause, the
 evaluation is straightforward. To reduce a \prev term, 2-evaluation
@@
 finally ending with the partial value.
-\ur{There seems to be errors in the semantics in the handling of
-  recursive types.  I don't see any reason to do any reductions here.
-If you want to, the description above needs to be updated.}
 
-...UMUT: UP TO HERE...
-
- In addition, 1-evaluation {\em reifies} the contents of
-the residual table into a residual when changing from stage-2 to stage
-
-%% Note that partial values and residuals are both terms for which stage \bbone\
-%% computation has completed; however, partial values are stage~\bbone\ terms,
-%% while residuals are stage~\bbtwo\ terms.
-
- The residual
-table implements the explicit substitutions in our example---it maps fresh
-stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
-residuals.
-
-In order to avoid duplicating subterms, evaluation collects explicit
-substitutions into a single {\em residual table} that is also the output of
-evaluation.
-
- the auxiliary $\reifysym$ (reification). 
-Along with these are three judgments identifying forms of values:
-$\pvalsym$ (partial values), $\ressym$ (residuals), 
-and the auxiliary $\fcon$ (residual table).
-These judgments are related as follows:
-\begin{itemize}
-\item Evaluation sends a stage~\bbone\ term to a partial value.
-
-\item Specialization sends a stage~\bbtwo\ term to a residual.
-
-\item Reification sends a single residual and a residual table to another residual.
-
-\item Evaluation depends on specialization for $\next$ terms, which have
-stage~\bbtwo\ subterms.
-\item Specialization depends on evaluation for $\prev$ and $\pause$, which have
-stage~\bbone\ subterms. The residual table produced by evaluation is also
-reified here.
-\end{itemize}
-
-The relation 2-evaluation takes a stage-2 term and maps to a residual.
-
-To generate the residual table, the dynamics semantics relies on a
-{\em reification} relation, $\reifysym$, that maps a table to a
-residual, essentially generating a target expression representing the
-computations in the table.
-
-We can evaluate a \lang\ term as follows: we first apply 1-evaluation
-to obtain a partial value and a residual table, we then construct a
-2-term from the partial value and the residual table by using
-reification, and then reduce this term to a residual via 2-evaluation.
-
-Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
-subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
-a stage \bbtwo\ term with no stage~\bbone\ content, 
-which can be described as a term in a monostaged language called \langTwo. 
-Then, at stage~\bbtwo, we complete
-evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
-rules for this judgment are not shown, but they are standard.)
-
-\crem{Can we cut \ref{fig:diaSemanticsSpec} from the paper? Perhaps just put a
-few representative rules into \ref{fig:diaSemantics}?}
-
-At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
-subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
-which takes a stage \bbtwo\ term to a residual. 
-For the most part (\ref{fig:diaSemanticsSpec}), specialization simply recurs into
-all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
-which produces a residual table $\xi$ and (by canonical forms) an expression
-$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
-bindings enclosing $\mathtt{\hat y}$.
-Once specialization reduces $e$ to a residual $q$, the output of $\redonesym$ is a
-fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a residual
-table which maps that variable to the residual ($\hat y \mapsto q$).
-
-Within specialization lies a subtle---if perhaps unintuitive---feature.  
-Observe that specialization will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
-statement in its search for stage~\bbone\ code. 
-Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
-and so a term like 
-\begin{lstlisting} 
-1`next{
-  `2`if true 
-  then hold{`1`3+4`2`} 
-  else prev{`1`spin() (* loops forever *)`2`}`1`
-}`
-\end{lstlisting}
-will fail to evaluate at stage \bbone.
-Although it may seem undesirable, this behavior is a deliberate feature.
-Indeed it is critical to the quickselect example, 
-wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
-As always, it is the programmer's responsibility to ensure that her staging annotations 
-do not cause a program to loop forever.
-
-The context ($\Gamma$) keeps track of stage \bbtwo\ variables in the input term. 
-These both appear in the original program at stage \bbtwo\ and are inserted by the semantics.
-
-As an optimization, we can include the special-case rule,
-\begin{mathpar}
-\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
-\end{mathpar}
-to avoid one-for-one variable bindings in the residual.
-
-\subsubsection {Top-Level Evaluation}
-\label{sec:partialeval}
-
-% KAYVONF: good statement, but hold out for now
-%The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must %evaluate it many times.
-
-These dynamics allow us to define a partial evaluator for \lang\ by identifying
-static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
-encode $f$ as a \lang\ expression with a function type of the form
-$A\to\fut(B\to C)$.%
-%\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
-%the following higher-order function which makes the adjustment:
-%\begin{lstlisting} 
-%let adjust (f : $int * int -> $int) =
-%  fn (p : int) => 
-%    next{
-%      fn (b : int) => 
-%        prev{f (next {b}, p)}
-%    }
-%\end{lstlisting}}
-%
-Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.
-
-Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
-function:
-$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
-The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
-C)$, which by canonical forms must have the form $v = \next~\hat y$. 
-Next, we reify this environment into a sequence of \verb|let|-bindings
-enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
-Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
-and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
-
-That this sequence of evaluations is in fact staged follows from our
-characterizations of partial values and residuals, that $\redonesym$
-outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
-
-For example, to specialize quickselect to a particular list, 
-\begin{lstlisting}
-qsStaged [5,2,7,4,1]
-\end{lstlisting}
-\TODO finish example
+We can use the evaluation relations in the dynamic semantics to
+evaluate a term to evaluate a \lang\ term as follows. We first apply
+1-evaluation to obtain a partial value and a residual table. We then
+construct a 2-term from the partial value and the residual table by
+using reification, and then reduce this term to a residual via
+2-evaluation.  Finally, we evaluate the residual, which is a term in a
+our monostaged target language \langmono, to a value by using  a
+standard evaluation relation for \langmono.
 

### 2014-11-10 paper/semantics.tex
more minor fixes

@@
 We present an overview of the formal development along with our
-terminology, as summarized in \figref{terminology}.
+terminology, as summarized in \ref{fig:termSplitSummary}.
 
-\figref{terminology}
+\ref{fig:termSplitSummary}
 
@@
 evaluated except for its stage-2 subterms, which can only
-be~\bbtwo\ variables wrapped in $\next$ blocks.  \figref{validPvalues}
+be~\bbtwo\ variables wrapped in $\next$ blocks.  \ref{fig:validPvalues}
 defines partial values.
@@
 The relation 2-evaluation, defined by $\redtwosym$ judgment, as shows
-in \figref{diaSemantics}, takes a stage \bbtwo\ term and reduces it to
+in \ref{fig:diaSemantics}, takes a stage \bbtwo\ term and reduces it to
 a residual.  Apart from the evaluation of \prev\ and \pause, the

### 2014-11-10 paper/semantics.tex
fold the type validity figure into the syntax figure
more minor fixes

@@
 
+\crem{Cut references to type validity.}
 The type validity judgment $\Delta \vdash A \istypewor$, defined in
@@
   the terminology figure} Residual table entries are created by the
-evaluation of \next: 1-evaluation reduces a \next expression by
-reducing the body of the \next with 2-evaluation to obtain a
+evaluation of \next: 1-evaluation reduces a \next\ expression by
+reducing the body of the \next\ with 2-evaluation to obtain a
 residual $q$ and returns a residual table that maps a freshly
@@
 a residual.  Apart from the evaluation of \prev\ and \pause, the
-evaluation is straightforward. To reduce a \prev term, 2-evaluation
+evaluation is straightforward. To reduce a \prev\ term, 2-evaluation
 reduces the body by 1-evaluation to obtain a residual table and a

### 2014-11-11 paper/figures/splitting-overview.tex
more minor fixes
a bit of work on terminology section

--- 2014-11-09 paper/figures/splitting-overview.tex
@@
-\begin{figure}
+\begin{figure*}
 \textbf{Languages:}
 \begin{itemize}
-\item \lang: two-staged lambda calculus with product, sum, and
+\item \lang: two-staged typed lambda calculus with product, sum, and
   recursive types
-\item \langmono: an unstaged lambda calculus with products.
+\item \langTwo: monostaged typed lambda calculus
+\item \langmono: an unstaged lambda calculus
 \end{itemize}
 
-\textbf{\lang\ Evaluation Relations:}
+\textbf{\lang\ Dynamic Semantics:}
 \begin{itemize}
@@
 \item 
-Table reification: \ur{Fill in...}
+Table reification: $[\xi;q]\mathbin{\reifysym} q'$, where $q,q'$ are residuals,
+$\xi$ is a residual table.
+
+\item
+Residual evaluation: $q\mathbin{\redsym} v$, where $q$ is a residual, $v$ is a
+value.
 \end{itemize}
 
-% $red_2$: 2-eval,  which we think of as being identical to specialization.
-
-\vspace{.75em}
 \textbf{\lang\ Splitting:}
@@
 
-\hspace{4em}$c$ is the \emph{combined term} (representing all stage~\bbone\ subcomputations in $e$)
+\hspace{4em}$c$ is the \emph{combined term}
+(representing all stage~\bbone\ subcomputations in $e$)
+in \langmono
 
-\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputations in $e$)
+\hspace{4em}$l.r$ is the \emph{resumer}
+(representing all stage~\bbtwo\ subcomputations in $e$)
+in \langmono
 
-\hspace{4em}$c~\redsym~(y,b)$, where $y$ is the \emph{\bbone-result} of $e$, and $b$ is the \emph{boundary value} of $e$
+\hspace{4em}$c~\redsym~(y,b)$, where 
+$y$ is the \emph{\bbone-result} of $e$, and
+$b$ is the \emph{boundary value} of $e$
 
@@
 
-\hspace{4em}$\maskt{\xi;v}$ and \texttt{(let l=b in r)} reduce (via $\redsym$) to identical values.
+\hspace{4em}$\maskt{\xi;v}$ and \texttt{(let l=b in r)} 
+reduce (via $\redsym$) to equivalent values.
 
@@
 
-\hspace{4em}$p$ is the \emph{precomputation} (representing all stage~\bbone\ subcomputations in $e$)
+\hspace{4em}$p$ is the \emph{precomputation}
+(representing all stage~\bbone\ subcomputations in $e$)
+in \langmono
 
-\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputations in $e$)
+\hspace{4em}$l.r$ is the \emph{resumer}
+(representing all stage~\bbtwo\ subcomputations in $e$)
+in \langmono
 
-\hspace{2em}\bbtwo-Splitting Correctness: If $e~\redtwosym~q$, and $q~\redsym~v$, then \texttt{(let l=b in r)}~$\redsym~v$
+\hspace{2em}\bbtwo-Splitting Correctness:
+If $e~\redtwosym~q$, and $q~\redsym~v$, then
+\texttt{(let l=p in r)}~$\redsym~v'$ and $v,v'$ are equivalent.
+% TODO: FIX THIS
+% if e reduces to q, then p reduces to u
+% and q == [u/l]r
+% that is, stage 1 reduces, even if stage 2 doesn't
 
 \caption{Summary of \lang\ evaluation and splitting.}
-\label{fig:terminology}
 \label{fig:termSplitSummary}
-\end{figure}
+\end{figure*}

### 2014-11-11 paper/obsolete/correctness.tex
a bit of work on terminology section

@@
+\subsection{Metatheory}
+
+Recall that residuals live in \langTwo; we will indicate typing judgments in
+\langTwo\ with $\vdash_\bbtwo$.
+
+%\begin{definition}
+%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
+%contains only stage-2 variables.
+%\end{definition}
+
+\begin{definition}
+An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
+\begin{enumerate}
+\item $\xi = \cdot$; or
+\item $\xi = \xi',x:B\mapsto e$ where
+$\Gamma\vdash\xi'\wf$ and
+$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
+%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
+%$\Gamma,\dom{\xi'}\vdash e \res$.
+\end{enumerate}
+\end{definition}
+
+\begin{theorem}
+If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
+\end{theorem}
+
+\begin{theorem}
+If $\diaonesub$ and $\typesone e A$ then
+\begin{enumerate}
+\item $\Gamma\vdash\xi\wf$;
+\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
+\item $\Gamma,\dom\xi\vdash v\pval$.
+\end{enumerate}
+\end{theorem}
+
+\begin{theorem}
+If $\diatwosub$ and $\typestwo e A$ then
+\begin{enumerate}
+\item $\typeslangTwo q A$; and
+\item $\Gamma\vdash_\bbtwo q\val$.
+\end{enumerate}
+\end{theorem}
+
+\begin{theorem}\label{thm:reify-type}
+If $\Gamma\vdash\xi\wf$ and
+$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
+then 
+$\reify{\xi}{\hat y}{q}$ and
+$\typeslangTwo q A$.
+\end{theorem}
+
+\TODO
+Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
+integer term is guaranteed not to depend on the table, although one might be
+produced. One may either discard the table, or evaluate everything in the table
+(and terminating with the partial value iff everything in the table terminates).
+
 \subsection{Correctness of Splitting}

### 2014-11-11 paper/semantics.tex
improve figures, yet again
a bit of work on terminology section

@@
 
-We present an overview of the formal development along with our
-terminology, as summarized in \ref{fig:termSplitSummary}.
+%We present an overview of the formal development along with our
+%terminology, as summarized in \ref{fig:termSplitSummary}.
 
-\ref{fig:termSplitSummary}
-
-Our starting point is a two-staged lambda calculus, whose syntax and
-types are very similar to the multi-stage languages considered in the
-literature on partial evaluation and meta programming.  Specifically,
-we consider a two-stage, which we call \lang\, which is variant of
-Davies`s multi-stage language.  Apart from the restriction to two
-stages instead of multiple stages, there are no significant
-differences between our syntax and the static semantics and those of
-Davies'. We restrict our attention to two stages, because it
-simplifies the (many) details of the splitting problem, without
-altering the essence of the problem.  We expect that the techniques
-presented here can be extended to multi-stage languages.
+Our starting point is \lang, a typed, two-stage lambda calculus with products,
+sums, and isorecursive types. This language is adapted from \cite{davies96}, and
+is very similar to the staged languages considered in the partial evaluation and
+metaprogramming literature. In this paper, we restrict our attention to a
+two-stage language to simplify our presentation of the splitting algorithm, but
+we expect that the techniques presented here extend to more stages.
 
 The dynamic semantics of \lang\ differs from Davies's approach and
-similar approaches in meta programming literature, because it avoids
+similar approaches in metaprogramming literature, because it avoids
 duplication of staged work. Duplication takes place in meta
@@
 
+\crem{Fold previous section in here.}
+
 \input {figures/grammars}
-\input {figures/statics}
 
 \subsection{Statics}
+
+\input {figures/statics}
 
@@
 
-\crem{Cut references to type validity.}
-The type validity judgment $\Delta \vdash A \istypewor$, defined in
-\cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
-example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
-$\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
-$\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
-typing judgment only produces valid types.
-%(by induction on the derivation of the typing judgment).
-% under the assumptions in $\Delta$. This context is only augmented by recursive types.
+The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in
+the context $\Gamma$, and is defined in \cref{fig:statics}.
+
+\crem{Notice that certain types are impossible.}
 
@@
 
-As is the case in essentially all staged language, the dynamic
-semantics evaluates a term to a residual (a term in the target
-language \langmono) in two stages. In the first stage, all stage-1
+As is the case in essentially all staged languages, the dynamic
+semantics evaluate a term to a residual (a term in the target
+language \langTwo) in two stages. In the first stage, all stage-1
 terms are evaluated by a relation, called {\em 1-evaluation}; in the
@@
 evaluated except for its stage-2 subterms, which can only
-be~\bbtwo\ variables wrapped in $\next$ blocks.  \ref{fig:validPvalues}
+be~\bbtwo\ variables wrapped in $\next$ blocks.  \ref{fig:grammar}
 defines partial values.
@@
 evaluation but in addition gathers the subterms' residual tables into
-a single table. \ur{The syntax of residual tables should be defined in
-  the terminology figure} Residual table entries are created by the
+a single table. Residual table entries are created by the
 evaluation of \next: 1-evaluation reduces a \next\ expression by
@@
 %% Version Nov 7, 2014
-%% \section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
-%% \label{sec:semantics}
-
-%% \input {figures/grammars}
-%% \input {figures/statics}
-
-%% \lang\ is a typed two-stage lambda calculus with products, sums, and
-%% isorecursive types.  Its statics are adapted from the linear temporal
-%% logic given by \cite{davies96}, restricted to two stages and extended
-%% with general sums and recursion.%
-%% \footnote{This restriction to two stages is made primarily for simplicity of
-%% presentation. All of these techniques could feasibly be extended to more
-%% complicated stage systems.} 
-%% Although \cite{davies96} was mostly interested in describing partial evaluation,
-%% the system is equally apt as an input to splitting.
-
-%% \subsection{Statics}
-
-%% In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
-%% stage~\bbone\ term. To ensure that the staging features are only applied to
-%% valid terms, we index our judgments not only by a type, but also a stage $w$.
-
-%% The type validity judgment $\Delta \vdash A \istypewor$, defined in
-%% \cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
-%% example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
-%% $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
-%% $\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
-%% typing judgment only produces valid types.
-%% %(by induction on the derivation of the typing judgment).
-%% % under the assumptions in $\Delta$. This context is only augmented by recursive types.
-
-%% Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
-%% language features (i.e., the introductory and eliminatory forms for product,
-%% sum, and function types) preserve the stage of their subterms, and work at both
-%% stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
-%% \lang\ would simply consist of two non-interacting copies of a standard lambda
-%% calculus. Variables in the context are annotated with the stage at which they
-%% were introduced and can only be used at the same stage.
-
-%% Note that there is no way to eliminate a term of type $\fut A$ into a stage
-%% \bbone\ term---the only elimination form is $\prev$, which results in a stage
-%% \bbtwo\ term. This is why well-typed terms cannot have any information flow
-%% from stage \bbtwo\ to stage \bbone, and what makes splitting possible.
 
@@
 
-%% \crem{Can we cut \ref{fig:diaSemanticsSpec} from the paper? Perhaps just put a
-%% few representative rules into \ref{fig:diaSemantics}?}
-
 %% At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
@@
 %% \TODO finish example
-
-%\begin{remark}
-%For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
-%always compute an empty environment, and a partial value identical to the result
-%of call-by-value evaluation of $e$.
-%%derivationally equivalent to standard call-by-value evaluation.
-%\end{remark}
-
-%\subsection{Metatheory}
-%
-%Recall that residuals live in \langTwo; we will indicate typing judgments in
-%\langTwo\ with $\vdash_\bbtwo$.
-%
-%%\begin{definition}
-%%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
-%%contains only stage-2 variables.
-%%\end{definition}
-%
-%\begin{definition}
-%An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
-%\begin{enumerate}
-%\item $\xi = \cdot$; or
-%\item $\xi = \xi',x:B\mapsto e$ where
-%$\Gamma\vdash\xi'\wf$ and
-%$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
-%%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
-%%$\Gamma,\dom{\xi'}\vdash e \res$.
-%\end{enumerate}
-%\end{definition}
-%
-%\begin{theorem}
-%If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
-%\end{theorem}
-%
-%\begin{theorem}
-%If $\diaonesub$ and $\typesone e A$ then
-%\begin{enumerate}
-%\item $\Gamma\vdash\xi\wf$;
-%\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
-%\item $\Gamma,\dom\xi\vdash v\pval$.
-%\end{enumerate}
-%\end{theorem}
-%
-%\begin{theorem}
-%If $\diatwosub$ and $\typestwo e A$ then
-%\begin{enumerate}
-%\item $\typeslangTwo q A$; and
-%\item $\Gamma\vdash_\bbtwo q\val$.
-%\end{enumerate}
-%\end{theorem}
-%
-%\begin{theorem}\label{thm:reify-type}
-%If $\Gamma\vdash\xi\wf$ and
-%$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
-%then 
-%$\reify{\xi}{\hat y}{q}$ and
-%$\typeslangTwo q A$.
-%\end{theorem}
-
-%\TODO
-%Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
-%integer term is guaranteed not to depend on the table, although one might be
-%produced. One may either discard the table, or evaluate everything in the table
-%(and terminating with the partial value iff everything in the table terminates).

### 2014-11-11 paper/example.tex
cleaning up latex warnings
major reorganization of section 2 (in progress) - moving lots of text around!

@@
 
-\subsection{Comments, suggestions, and ``todo''}
-I rewrote this section.  There is more work to be done.
-
+\ur{
 \begin{itemize}
@@
 I find it hard to distinguish between blue and black on printed paper.
-
 \item
@@
   be cited.
-
-\item
-I don't like fuzzy words like hygenic that dont mean anything.
-  What is the point of using them?  A paper ought to be written for
-  the reader not as a note to self.
-
 \item the splitting example should shed more light to the techniques
@@
 \end{itemize}
+}
 
-\begin{figure*}
-\begin{subfigure}{0.5\textwidth}
-\begin{lstlisting} 
-datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) 
-  : (int*list*list) =
-  case unroll l of 
-    Empty => (0,Empty, Empty) 
-  | Cons (h,t) =>
-      let (s,left,right) = partition (p,t) in
-      if h<p 
-      then (s+1,Cons(h,left),right)
-      else (s,left,Cons(h,right))
+\crem{
+Outline:
+\begin{enumerate}
+\item Say we want to run a bunch of quickselect queries on a single list.
+\item We observe that some of the code depends only on the list.
+\item We can write this code in \lang\ to make that dependency clear.
+\item We can intuitively split into two functions.
+\item This is asymptotically better.
+\end{enumerate}
+}
 
-fun qSelect (l : list, k : int) : int = 
-  case l of
-    Empty => 0
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-        case compare k n of
-          LT => qSelect (left, k)
-        | EQ => h
-        | GT => qSelect (right, k-n-1)
-\end{lstlisting}
-\caption{Unstaged implementation of quickselect.}
-\label{fig:quickselect}
-\label{fig:qs-unstaged}
-\end{subfigure}%
-\begin{subfigure}{0.5\textwidth}
-\begin{lstlisting} 
-1`datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...
-
-fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
-  case l of
-    Empty => next {`2`0`1`}
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      next{
-        `2`let n = hold{`1`n`2`} in
-          case compare prev{`1`k`2`} n of
-            LT => prev {`1`qsStaged (left k)`2`}
-          | EQ => hold {`1`h`2`}
-          | GT => prev {`1`qsStaged (right, 
-                             next{prev{k}-n-1)}`2`}`1`
-      }`
-\end{lstlisting}
-\caption{Staged implementation of quickselect in \lang.}
-\label{fig:qsstaged}
-\label{fig:qs-staged}
-\end{subfigure}
-\caption{Quickselect: traditional and staged.}
-\end{figure*}
+\input{figures/quickselect}
 
 Staging is a technique for using language constructs and types to
-explicitly staged a computation according to the frequency or the
-arrival time of data.  As an example, suppose that we wish to issue
-order statistics queries on a collection of items represented as a
-list \texttt{l}.  For this tasks, we can use the expected linear time,
-quickselect algorithm~\cite{quickselect}, which takes a
-list~\texttt{l} and a rank~\texttt{k} and returns the element with
-rank~\texttt{k} in the list.  To this end, the algorithm, whose code
-is shown in ~\ref{fig:qs-unstaged} partitions the list into two by
-using the first element as a pivot\footnote{We assume that the list
-  is prepermuted to guarantee the expected linear time behavior.} and
-recurs down to one of the ``halves'' to find the desired element.  The
-``half'' to be considered recursively next is determined by the
-relationship of~\texttt{k} to the size of the first half~\texttt{n},
-which is returned by the function~\texttt{partition}.
+explicitly divide a computation according to the frequency or the
+arrival time of data.
 
-Continuing on our example, consider an application where we perform
-many order statistics queries using the same collection~\texttt{l}, but
- with $m$ different ranks~\texttt{k}:
+Suppose that we wish to issue order statistics queries on a collection of items
+represented as a list \texttt{l}. We can use quickselect \cite{quickselect}, an
+expected linear time algorithm which takes a list \texttt{l} and a rank
+\texttt{k} and returns the element with rank \texttt{k} in \texttt{l}.
+
+Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
+by using the first element as a pivot%
+\footnote{We assume that the list is prepermuted to guarantee the expected
+linear time behavior.}
+and then recurs on one of the two resulting sides to find the desired element. The
+side chosen is determined by the relationship of \texttt{k} to the size of the
+first half \texttt{n}, which is returned by the \texttt{partition} function
+along with the two sides.
+
+Now, consider an application where we perform many order statistics queries
+on the same collection \texttt{l}, but with $m$ different ranks \texttt{k}:
 %
@@
 %
-Since \texttt{quickselect1} requires only on a first stage
-argument~\texttt{l}, we can run it once to produce a boundary data
-structure that can be used for all subsequent queries on~\texttt{l},
-e.g., the code consisting of $m$ calls To take advantage of this, we
-can {\em stage} quickselect so that the parts of the computation that
-depend on the collection~\texttt{l} is performed once, and the desired
-result is computed by using~\texttt{k} and the results from this prior
-stage.
+To take advantage of this, we can \emph{stage} quickselect so that the parts of
+the computation that depend on the collection~\texttt{l} is performed once, and
+the desired result is computed by using~\texttt{k} and the results from this
+prior stage.
 
-Prior work proposed staging for exactly these kinds of applications
-and proposed techniques for writing staged
-programs~\cite{staging}. Using techniques essentially the same as
+%% There has been the extensive research into the question of how to
+%% automatically add staging annotations to unstaged code.  This process
+%% is known as {\em binding time analysis}, and we do not consider it
+%% here.  Instead, we assume that all input programs are properly staged
+%% according to some programmer's intent.
+
+\crem{missing text}
+
+Using techniques essentially the same as
 prior work, we can stage our quickselect code by using staged typing
@@
 
-Having determined the type signature,
-the programmer can then insert $\next$ and $\prev$ constructs into the
-code. The construct $\next$ occurs in a stage \bbone\ context and
-indicates that the contents of its block are stage \bbtwo, whereas
-$\prev$ occurs in a stage \bbtwo\ context and indicates that the
-contents of its block is stage \bbone.  (We adopt the convention that
-the top-level context is stage \bbone.)  The output type of a $\next$
-block is the $\fut$'d version of the type of its stage
-\bbtwo\ contents.  For example, \verb|next{0}| from above has the type
-$\fut\rmint$.  Correspondingly, $\prev$ requires that its stage
-\bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
-For example, \verb|prev{k}| from above has type $\rmint$ at stage
-\bbtwo, since $k$ is bound to a $\fut\rmint$ at stage
-\bbone. Informally speaking, the
-$\next$ construct enables operating on second stage computations in
-the first stage essentially by treating them as boxes that can be
-moved around, stored in data structures, and passed to functions, and
-the $\prev$ construct enables unwrapping the contents of such a box to
-use in the second stage.
+\crem{explain circle types}
+
+The construct $\next$ occurs in a stage \bbone\ context and indicates that the
+contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
+\bbtwo\ context and indicates that the contents of its block is stage \bbone.
+(We adopt the convention that the top-level context is stage \bbone.)
+
+Informally speaking, the $\next$ construct enables operating on second stage
+computations in the first stage essentially by treating them as boxes that can
+be moved around, stored in data structures, and passed to functions, and the
+$\prev$ construct enables unwrapping the contents of such a box to use in the
+second stage.
 
 For convenience, the code also uses $\pause$ blocks.  This construct
-has same stage signature as $\prev$, but instead of ``unwrapping"
+has same stage signature as $\prev$, but instead of ``unwrapping''
 $\fut$ types it simply promotes integers from stage~\bbone\ to
-stage~\bbtwo.
-\footnote{It will turn out that $\pause$ is implementable---though it
-  takes some effort---from our other language features.  We instead
-  provide it as a primitive to shorten examples.  Furthermore, it
-  would be wise to extend $\pause$ to other base types, if we had
-  them, and to products and sums thereof.  This is related to the
-  notion of {\em mobility} in \cite{murphy05} and {\em stability} in
-  \cite{krishnaswami13}.} We note that given the type signature of
-staged quickselect, there are several ways to insert the staging
-annotations.  The version we presented here maximizes the work done at
-stage~\bbone.
+stage~\bbtwo.%
+\footnote{$\pause$ can be implemented from our other language features, with
+some effort. We provide it as a primitive to shorten examples. Furthermore, it
+would be wise to extend $\pause$ to other base types, if we had them, and to
+products and sums thereof.  This is related to the notion of {\em mobility} in
+\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
 
-%%%
-%% There has been the extensive research into the question of how to
-%% automatically add staging annotations to unstaged code.  This process
-%% is known as {\em binding time analysis}, and we do not consider it
-%% here.  Instead, we assume that all input programs are properly staged
-%% according to some programmer's intent.
+We note that given the type signature of staged quickselect, there are several
+ways to insert the staging annotations.  The version we presented here maximizes
+the work done at stage~\bbone.
 
-Prior work on meta programming allows us to evaluate such staged code
-in a staged fashion by first providing the first-stage arguments and
-generating a piece of code that is specialized for the arguments.  For
-example, we can evaluate the staged quickselect by providing a
-list~\texttt{l} to obtain a piece of code that then can be used with
-different ranks (\texttt{k} values) in the second stage.  Such a
-specialized code essentially performs all the first-stage computations
-first and uses their results to improve the efficiency in the second
-stage, where we issue many queries.  The type system guarantees that
-all first-stage computations depend only on the first stage values and
-can indeed be computed in the absence of second-stage values.
+The type system guarantees that all first-stage computations depend only on the
+first stage values and can indeed be computed in the absence of second-stage
+values.
 
-This staged evaluation technique corresponds to {\em partial
-  evaluation} or {\em program specialization} and can bring out
-dramatic improvements in the efficiency of programs.  It has however
-one major limitation: since it requires the first stage arguments to
-be available, it cannot be performed at compile time, except of course
-when the first-stage values are available at compile time.  Since the
-evaluation in the first stage produces a piece of code that must also
-be compiled or interpreted to execute in the second stage, the
-approach can lead to code blowup and tends to be relatively expensive
-in practice.  Due to these reasons, the applications of partial
-evaluation are limited.  For example, in our quickselect example, it
-is unlikely to know the list~\texttt{l} that we will be operation on
-at compile time. Partial evaluation dynamically at run time is
-difficult because the first stage generates code that is linear in the
-size of the list, whose generation, dynamic compilation or
-interpretation can be excruciatingly slow.
+\crem{partial evaluation requires compile-time access to first stage, whereas we
+don't}
 
-Knoblock and Ruf~\cite{KR} proposed the {\em data specialization}
-technique to overcome some of these limitations of program
-specialization.  The idea behind this technique is to take a staged
-program and split it into a stage-1 and a stage-2 program without
-knowledge of any of the arguments.  Unlike program specialization,
-data specialization does not require knowledge of the first-stage
-arguments to produce fast code for the second stage.  Instead it yield
-a first stage program that maps the first stage arguments to an
-intermediate or {\em boundary data structure}, which is used by the
-second stage program along with the second stage arguments to produce
-the final results.  Knoblock and Ruf's observation is that the
-information required by the second stage can be represented solely as
-a data structure, making it unnecessary to generate and compile or
-interpret code dynamically at run time.  Data specialization can
-therefore be applied statically at compile time, leading to
-drastically improved run-time performance without any significant
-run-time overheads.
+\crem{introduce boundary terminology}
 
-By applying data specialization to our quickselect example, we would
-wish to derive two functions, \texttt{quickselect1} and
-\texttt{quickselect2}, that would be semantically equivalent to
-quickselect in the following sense:
+Here, we wish to derive two functions, \texttt{quickselect1} and
+\texttt{quickselect2}, that are semantically equivalent to quickselect in
+the sense that:
 \\
@@
 \\
-%
+\crem{and moreover they correspond to the stages}
+
 Since \texttt{quickselect1} requires only on a first stage
@@
 
-The benefit of this ``optimization'' can be dramatic.  For example, in
-principle at least, the first stage function \texttt{quickselect1}
-could produce a balanced binary search tree holding the elements
-of~\texttt{l} and the second stage function \texttt{quickselect2}
-could use this balanced binary search tree and the given rank value to
-perform a logarithmic-time search in the tree to return the desired
-result.  Assuming that the list~\texttt{l} contains $n$ elements, this
-optimization would change the asymptotic complexity from expected
-(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the complexity
-from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
-improvement.  Note that the desired optimized code shows above is
-intellectually more sophisticated than the code that we have started
-with: the optimized code is able to create a data structure, a
-balanced binary tree augmented with indexing information, and use a
-binary search technique over this tree to compute the result
-asymptotically more efficiently.  In fact, based our teaching
-experience, we can imagine this kind of problem to be a moderately
-difficult exam question in an undergraduate algorithms class, as it
-not only requires understanding of data structures such as binary
-search trees but also requires modifying them to augment with indexing
-information to support rank-based search.
-
-It is thus perhaps no surprise that the prior art in data
-specialization is not able to perform such ``optimizations''.
-Knoblock and Ruf's paper takes a very important step by formulating
-the problem and proposing a solution.  But their work applies to
-program fragments that don't involve sophisticated data types such as
-as sums and recursive types, and don't involve recursion. For example,
-their work would not apply to our quickselect example and similar ones
-involving recursion and/or  recursive types.
-
 In this paper, we show that it is possible to generalize data
@@
 
-In what follows we present a relatively mechanical application of our
-techniques to our running example.
+\crem{walk through manual example here}
 
-\ur{I made up to here.  What follows should be shortened and made to
-  the point. Some time later...}
+The final version, cleaned up and given better datatype names, is shown in
+\ref{fig:qssplit}.  Observe that \texttt{qSelect1} essentially builds a binary
+search tree from the input list, and \texttt{qSelect2} searches that tree for
+the correct index.
 
-To illustrate this, we present colorized version of \texttt{qSelect}:
+Assuming that the list~\texttt{l} contains $n$ elements, this optimization
+changes the asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$
+to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
+complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
+improvement.
 
-\begin{lstlisting} 
-1`datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ... `
+Note that the desired optimized code shows above is intellectually more
+sophisticated than the code that we have started with: the optimized code is
+able to create a data structure, a balanced binary tree augmented with indexing
+information, and use a binary search technique over this tree to compute the
+result asymptotically more efficiently.
+%In fact, based our teaching experience, we can imagine this kind of problem to
+%be a moderately difficult exam question in an undergraduate algorithms class, as
+%it not only requires understanding of data structures such as binary search
+%trees but also requires modifying them to augment with indexing information to
+%support rank-based search.
 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in`2`
-      case compare k n of
-        LT => `qSelect 1`left `2`k
-      | EQ => h
-      | GT => `qSelect 1`right `2`(k-n-1)`
-\end{lstlisting}
+This paper formally describes and analyzes the splitting transformation
+exemplified above.  (I also want to make it clear that recognizing \lang's
+appropriateness for this is itself is a contribution.)
 
-Here, the input $k$ and all parts of the computation that depend on it are colored blue,
-whereas all of the parts of the computation that do not depend on $k$ are colored red.
-The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.
-
-\ur{I find control dependence issue very confusing.  Let's factor this
-out and talk about it in the end. }
-
-To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
-first arguments to the recursive calls to \texttt{qSelect} are colored red,
-despite appearing underneath a blue case statement.
-This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.
-
-Observe the that helper function \texttt{partition} is colored entirely red,
-which is valid because it is called with only red arguments.  
-This is an illustration of the property stated above, that 
-the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
-Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.
-
-This suggests an optimization: we can split \texttt{qSelect} into two functions, 
-one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
-Because of the dependence properties stated above, 
-\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
-whereas \texttt{qSelect1} will only depend on the input list $l$.
-Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
-Equationally, we can represent this relationship as simply 
-\[
-\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
-\]
-for some list \texttt{l} and index \texttt{k}.
-
-\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}
-
-With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
-Because the original function is recursive,
-the first step in this process is to assume the conclusion at the call sites.
-
-\ur{quickselect 1 and 2: how did we introduce them?}
-
-This is performed in place:
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in`2`
-      case compare k n of
-        LT => qSelect2 (`1`qSelect1 left`2`) k
-      | EQ => h
-      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
-\end{lstlisting}
-We then pull the stage \bbone\ portion of each branch of the
-comparison up into a single tuple.
-\ur{Why did we include h}
-
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
-      case compare k n of
-        LT => qSelect2 (pi2 p) k
-      | EQ => pi1 p
-      | GT => qSelect2 (pi3 p) (k-n-1)`
-\end{lstlisting}
-
-\ur{Don't know what ``this last step'' refers to.}
-This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
-We continue to pull up the precomputation and include more in it:
-
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let p =
-        let (left,right,n) = partition h t in
-        `1`(n,h,qSelect1 left,qSelect1 right)
-      in`2`
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-This completes the splitting of the cons branch.  
-The empty branch can also be split, though in a trivial way:
-
-\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => let p = () in `2`0`1`
-  | Cons (h,t) => 
-      let p =
-        let (left,right,n) = partition h t in
-        `1`(n,h,qSelect1 left,qSelect1 right)
-      in`2`
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-To continue splitting the case statement, we need to generate a sum type in the precomputation
-to remind the code what branch its in.
-
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  let p = 
-    case l of
-      Empty => Top ()
-    | Cons (h,t) =>
-        let (left,right,n) = partition h t in`1`
-        Bot (n,h,qSelect1 left,qSelect1 right)
-  in`2`
-  case p of 
-    Top () => 0
-  | Bot p => 
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-We finish where we start, splitting \texttt{qSelect} into
-\texttt{qSelect1} and \texttt{qSelect2}.  
-Omitting a few type annotations, that is:
-
-\begin{lstlisting} 
-1`fun qSelect1 (l : list) = 
-    case l of
-      Empty => Top ()
-    | Cons (h,t) =>
-        let (left,right,n) = partition h t in
-        Bot (n,h,qSelect1 left,qSelect1 right)
-`2`
-fun qSelect2 p (k : int) : int = 
-  case p of 
-    Top () => 0
-  | Bot p => 
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-The final version, cleaned up and given better datatype names, is shown in \ref{fig:qssplit}.
-Observe that \texttt{qSelect1} essentially builds a binary search tree from the input list, 
-and \texttt{qSelect2} searches that tree for the correct index.
-The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
-the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
-whereas \texttt{qSelect2} takes $O(\log n)$ time.
-Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
-
-\nr{There's something else I need to say right here before the contribution statement below. 
-I want the reader to be thinking "yeah, this splitting business does seem cool."
-Bonus points if I also have them thinking "but it would be cooler if it were formal."}
-
-This paper formally describes and analyzes the splitting transformation exemplified above.
-(I also want to make it clear that recognizing \lang's appropriateness for this is itself is a contribution.)
-
-We start by defining a simple staged language, called \lang, to express the input to our splitting algorithm.  
-\lang\ unambiguously specifies how its terms should be split
-% and features a type system that ensures valid splittings exist.
-% only minimal refactoring should be required to write terms within this language,
-%and from there, the type system should be enough to prove that a valid splitting exists.
-by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
-Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.  
-After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
-will end up in the stage \bbone\ function ({\tt qSelect1}), 
-and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
-Typing rules for \lang\ ensure a valid splitting exists.  That is, 
-within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
-but never from \bbtwo\ to \bbone.
-
-\begin{figure}
-
-\begin{subfigure}{0.5\textwidth}
-\begin{lstlisting}
-1`datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...`
-
-datatype tree = Branch of int * int * tree * tree
-                | Leaf
-
-1`fun qSelect1 (l : list) : tree =
-  case l of
-    Empty => Leaf
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      Branch (n, h, qSelect1 left, qSelect1 right)`
-
-2`fun qSelect2 (p : tree, k : int) : int = 
-  case unroll p of
-    Leaf => 0
-  | Branch (n,h,p1,p2) => 
-      case compare k n of
-        LT => qSelect2 (p1,k)
-      | EQ => h
-      | GT => qSelect2 (p2,k-n-1)`
-\end{lstlisting}
-\caption{Split (two-phase) implementation of quickselect.}
-\label{fig:qssplit}
-\end{subfigure}
-\caption{Caption place holder}
-\end{figure}
-
-\begin{comment}
-
-%% Umut: this is the earlier version.  I am reorganizing this whole
-%% section.
-
-\ur{Indentation seems broken. Be consistent.}
-
-\begin{figure*}
-\begin{subfigure}{0.5\textwidth}
-\begin{lstlisting} 
-datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) 
-  : (int*list*list) =
-  case unroll l of 
-    Empty => (0,Empty, Empty) 
-  | Cons (h,t) =>
-      let (s,left,right) = partition (p,t) in
-      if h<p 
-      then (s+1,Cons(h,left),right)
-      else (s,left,Cons(h,right))
-
-fun qSelect (l : list, k : int) : int = 
-  case l of
-    Empty => 0
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-        case compare k n of
-          LT => qSelect (left, k)
-        | EQ => h
-        | GT => qSelect (right, k-n-1)
-\end{lstlisting}
-\caption{Unstaged implementation of quickselect.}
-\label{fig:quickselect}
-\end{subfigure}%
-\begin{subfigure}{0.5\textwidth}
-\begin{lstlisting} 
-1`datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...
-
-fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
-  case l of
-    Empty => next {`2`0`1`}
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      next{
-        `2`let n = hold{`1`n`2`} in
-          case compare prev{`1`k`2`} n of
-            LT => prev {`1`qsStaged (left k)`2`}
-          | EQ => hold {`1`h`2`}
-          | GT => prev {`1`qsStaged (right, 
-                             next{prev{k}-n-1)}`2`}`1`
-      }`
-\end{lstlisting}
-\caption{Staged implementation of quickselect in \lang.}
-\label{fig:qsstaged}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\begin{lstlisting}
-1`datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ...`
-
-datatype tree = Branch of int * int * tree * tree
-                | Leaf
-
-1`fun qSelect1 (l : list) : tree =
-  case l of
-    Empty => Leaf
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      Branch (n, h, qSelect1 left, qSelect1 right)`
-
-2`fun qSelect2 (p : tree, k : int) : int = 
-  case unroll p of
-    Leaf => 0
-  | Branch (n,h,p1,p2) => 
-      case compare k n of
-        LT => qSelect2 (p1,k)
-      | EQ => h
-      | GT => qSelect2 (p2,k-n-1)`
-\end{lstlisting}
-\caption{Split (two-phase) implementation of quickselect.}
-\label{fig:qssplit}
-\end{subfigure}
-\caption{Caption place holder}
-\end{figure*}
-
-As an example of an algorithm amenable to staging techniques,
-consider the implementation of quickselect shown in \ref{fig:quickselect}.
-Quickselect is an algorithm for finding the $k$-th largest element in a list.
-It works by partitioning the input list into sublists containing elements less than and greater than the head element, 
-then recurring on the sublist containing the desired element.
-
-\ur{I find it hard to distinguish between blue and black on printed paper.}
-
-One notable property of quickselect is that the partitioning of the list, 
-even through its recursive calls, depends only on the values of list elements, and not on the value of $k$.
-To illustrate this, we present colorized version of \texttt{qSelect}:
-
-\begin{lstlisting} 
-1`datatype list = Empty | Cons of int * list
-fun partition (p : int, l : list) = ... `
-
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in`2`
-      case compare k n of
-        LT => `qSelect 1`left `2`k
-      | EQ => h
-      | GT => `qSelect 1`right `2`(k-n-1)`
-\end{lstlisting}
-
-Here, the input $k$ and all parts of the computation that depend on it are colored blue,
-whereas all of the parts of the computation that do not depend on $k$ are colored red.
-The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.
-
-\ur{I find control dependence issue very confusing.  Let's factor this
-out and talk about it in the end. }
-
-To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
-first arguments to the recursive calls to \texttt{qSelect} are colored red,
-despite appearing underneath a blue case statement.
-This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.
-
-Observe the that helper function \texttt{partition} is colored entirely red,
-which is valid because it is called with only red arguments.  
-This is an illustration of the property stated above, that 
-the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
-Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.
-
-This suggests an optimization: we can split \texttt{qSelect} into two functions, 
-one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
-Because of the dependence properties stated above, 
-\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
-whereas \texttt{qSelect1} will only depend on the input list $l$.
-Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
-Equationally, we can represent this relationship as simply 
-\[
-\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
-\]
-for some list \texttt{l} and index \texttt{k}.
-
-\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}
-
-With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
-Because the original function is recursive,
-the first step in this process is to assume the conclusion at the call sites.
-
-\ur{quickselect 1 and 2: how did we introduce them?}
-
-This is performed in place:
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in`2`
-      case compare k n of
-        LT => qSelect2 (`1`qSelect1 left`2`) k
-      | EQ => h
-      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
-\end{lstlisting}
-We then pull the stage \bbone\ portion of each branch of the
-comparison up into a single tuple.
-\ur{Why did we include h}
-
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let (left,right,n) = partition h t in
-      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
-      case compare k n of
-        LT => qSelect2 (pi2 p) k
-      | EQ => pi1 p
-      | GT => qSelect2 (pi3 p) (k-n-1)`
-\end{lstlisting}
-
-\ur{Don't know what ``this last step'' refers to.}
-This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
-We continue to pull up the precomputation and include more in it:
-
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => `2`0`1`
-  | Cons (h,t) => 
-      let p =
-        let (left,right,n) = partition h t in
-        `1`(n,h,qSelect1 left,qSelect1 right)
-      in`2`
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-This completes the splitting of the cons branch.  
-The empty branch can also be split, though in a trivial way:
-
-\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  case l of
-    Empty => let p = () in `2`0`1`
-  | Cons (h,t) => 
-      let p =
-        let (left,right,n) = partition h t in
-        `1`(n,h,qSelect1 left,qSelect1 right)
-      in`2`
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-To continue splitting the case statement, we need to generate a sum type in the precomputation
-to remind the code what branch its in.
-
-\begin{lstlisting} 
-fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
-  let p = 
-    case l of
-      Empty => Top ()
-    | Cons (h,t) =>
-        let (left,right,n) = partition h t in`1`
-        Bot (n,h,qSelect1 left,qSelect1 right)
-  in`2`
-  case p of 
-    Top () => 0
-  | Bot p => 
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-We finish where we start, splitting \texttt{qSelect} into
-\texttt{qSelect1} and \texttt{qSelect2}.  
-Omitting a few type annotations, that is:
-
-\begin{lstlisting} 
-1`fun qSelect1 (l : list) = 
-    case l of
-      Empty => Top ()
-    | Cons (h,t) =>
-        let (left,right,n) = partition h t in
-        Bot (n,h,qSelect1 left,qSelect1 right)
-`2`
-fun qSelect2 p (k : int) : int = 
-  case p of 
-    Top () => 0
-  | Bot p => 
-      case compare k (pi1 p) of
-        LT => qSelect2 (pi3 p) k
-      | EQ => pi2 p
-      | GT => qSelect2 (pi4 p) (k-n-1)`
-\end{lstlisting}
-
-The final version, cleaned up and given better datatype names, is shown in \ref{fig:qssplit}.
-Observe that \texttt{qSelect1} essentially builds a binary search tree from the input list, 
-and \texttt{qSelect2} searches that tree for the correct index.
-The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
-the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
-whereas \texttt{qSelect2} takes $O(\log n)$ time.
-Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
-
-\nr{There's something else I need to say right here before the contribution statement below. 
-I want the reader to be thinking "yeah, this splitting business does seem cool."
-Bonus points if I also have them thinking "but it would be cooler if it were formal."}
-
-This paper formally describes and analyzes the splitting transformation exemplified above.
-(I also want to make it clear that recognizing \lang's appropriateness for this is itself is a contribution.)
-
-We start by defining a simple staged language, called \lang, to express the input to our splitting algorithm.  
-\lang\ unambiguously specifies how its terms should be split
-% and features a type system that ensures valid splittings exist.
-% only minimal refactoring should be required to write terms within this language,
-%and from there, the type system should be enough to prove that a valid splitting exists.
-by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
-Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
-subexpressions are evaluated before stage-\bbtwo\ ones.  
-After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
-will end up in the stage \bbone\ function ({\tt qSelect1}), 
-and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
-Typing rules for \lang\ ensure a valid splitting exists.  That is, 
-within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
-but never from \bbtwo\ to \bbone.
-
-\ref{fig:qsstaged} shows a staged implementation of quickselect in \lang.
-As before, the function accepts as input the list to select from and a value of $k$,
-but now the latter is represented with the type $\fut\rmint$ rather than $\rmint$.
-The difference here is that an $\rmint$ is an integer available at the current stage (stage \bbone), 
-whereas a $\fut\rmint$ is an integer available only at the next stage (stage \bbtwo).
-Naturally the output type, representing the $k$th largest element of the list,
-is also $\fut\rmint$, since it is not computed until stage \bbtwo.
-
-Each construct in the body of quickselect is now associated with a stage via an interleaving of $\next$ and $\prev$ blocks.  
-Specifically, $\next$ occurs in a stage \bbone\ context and indicates that the contents of its block are stage \bbtwo, 
-whereas $\prev$ occurs in a stage \bbtwo\ context and indicates that the contents of its block is stage \bbone.
-(We adopt the convention that the top-level context is stage \bbone.)
-The output type of a $\next$ block is the $\fut$'d version of the type of its stage \bbtwo\ contents.  
-For example, \verb|next{0}| from above has the type $\fut\rmint$.
-Correspondingly, $\prev$ requires that its stage \bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
-For example, \verb|prev{k}| from above has type $\rmint$ at stage \bbtwo, since $k$ is bound to a $\fut\rmint$ at stage \bbone.
-These type restrictions essentially enforce that ``later stage content" is always treated hygienically at stage \bbone,
-which is necessary to admit a properly staged implementation.
-
-The code also contains two $\pause$ blocks.  
-This construct has same stage signature as $\prev$,
-but instead of ``unwrapping" $\fut$ types it simply promotes integers from stage~\bbone\ to stage~\bbtwo.
-\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
-We instead provide it as a primitive to shorten examples.  
-Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
-This is related to the notion of {\em mobility} in \cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
-
-But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
-the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
-The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
-It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
-but that would not be particularly interesting since it would result in an effectively trivial split 
-that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
-There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
-This process is known as {\em binding time analysis}, and we do not consider it here.
-Instead, we assume that all input programs are properly staged according to some programmer's intent.
-\end{comment}
+\input{figures/quickselect-split}

### 2014-11-11 paper/intro.tex
major reorganization of section 2 (in progress) - moving lots of text around!

@@
 \section{Introduction}
+
+\crem{This text was lifted from section 2.}
+
+Prior work on metaprogramming allows us to evaluate such staged code
+in a staged fashion by first providing the first-stage arguments and
+generating a piece of code that is specialized for the arguments.  For
+example, we can evaluate the staged quickselect by providing a
+list~\texttt{l} to obtain a piece of code that then can be used with
+different ranks (\texttt{k} values) in the second stage.  Such a
+specialized code essentially performs all the first-stage computations
+first and uses their results to improve the efficiency in the second
+stage, where we issue many queries.
+
+This staged evaluation technique corresponds to {\em partial
+  evaluation} or {\em program specialization} and can bring out
+dramatic improvements in the efficiency of programs.  It has however
+one major limitation: since it requires the first stage arguments to
+be available, it cannot be performed at compile time, except of course
+when the first-stage values are available at compile time.  Since the
+evaluation in the first stage produces a piece of code that must also
+be compiled or interpreted to execute in the second stage, the
+approach can lead to code blowup and tends to be relatively expensive
+in practice.  Due to these reasons, the applications of partial
+evaluation are limited.  For example, in our quickselect example, it
+is unlikely to know the list~\texttt{l} that we will be operation on
+at compile time. Partial evaluation dynamically at run time is
+difficult because the first stage generates code that is linear in the
+size of the list, whose generation, dynamic compilation or
+interpretation can be excruciatingly slow.
+
+\cite{knoblock96} proposed the {\em data specialization}
+technique to overcome some of these limitations of program
+specialization.  The idea behind this technique is to take a staged
+program and split it into a stage-1 and a stage-2 program without
+knowledge of any of the arguments.  Unlike program specialization,
+data specialization does not require knowledge of the first-stage
+arguments to produce fast code for the second stage.  Instead it yield
+a first stage program that maps the first stage arguments to an
+intermediate or {\em boundary data structure}, which is used by the
+second stage program along with the second stage arguments to produce
+the final results.  Knoblock and Ruf's observation is that the
+information required by the second stage can be represented solely as
+a data structure, making it unnecessary to generate and compile or
+interpret code dynamically at run time.  Data specialization can
+therefore be applied statically at compile time, leading to
+drastically improved run-time performance without any significant
+run-time overheads.
+
+It is thus perhaps no surprise that the prior art in data
+specialization is not able to perform such ``optimizations''.
+Knoblock and Ruf's paper takes a very important step by formulating
+the problem and proposing a solution.  But their work applies to
+program fragments that don't involve sophisticated data types such as
+as sums and recursive types, and don't involve recursion. For example,
+their work would not apply to our quickselect example and similar ones
+involving recursion and/or  recursive types.
+
+\subsection{Old intro}
 

### 2014-11-11 paper/obsolete/qselect-example.tex
major reorganization of section 2 (in progress) - moving lots of text around!

To illustrate this, we present colorized version of \texttt{qSelect}:

\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ... `

fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => `qSelect 1`left `2`k
      | EQ => h
      | GT => `qSelect 1`right `2`(k-n-1)`
\end{lstlisting}

Here, the input $k$ and all parts of the computation that depend on it are colored blue,
whereas all of the parts of the computation that do not depend on $k$ are colored red.
The \texttt{qSelect} variable itself, due to its status as a function with a mixed-color body, is left black.

\ur{I find control dependence issue very confusing.  Let's factor this
out and talk about it in the end. }

To be precise, we use "depend on" to mean data dependence, but not control dependence---the 
first arguments to the recursive calls to \texttt{qSelect} are colored red,
despite appearing underneath a blue case statement.
This coloring rule is enough to ensure that information can flow from red code to blue code but not vice versa.

Observe the that helper function \texttt{partition} is colored entirely red,
which is valid because it is called with only red arguments.  
This is an illustration of the property stated above, that 
the partitioning of the list, even through recursion, depends only on the list itself and not on $k$.
Moreover, the property is notable because much of the work of the quickselect algorithm takes place in \texttt{partition}.

This suggests an optimization: we can split \texttt{qSelect} into two functions, 
one for all of the red code (\texttt{qSelect1}) and another for all of the blue code (\texttt{qSelect2}). 
Because of the dependence properties stated above, 
\texttt{qSelect2} will need only depend on the results of \texttt{qSelect1} and the input index $k$,
whereas \texttt{qSelect1} will only depend on the input list $l$.
Additionally, the meaning of \texttt{qSelect1} appropriately composed with \texttt{qSelect2} should be the same as \texttt{qSelect}, too.
Equationally, we can represent this relationship as simply 
\[
\mathtt{qSelect~l~k} \equiv \mathtt{qSelect2~(qSelect1~l)~k}
\]
for some list \texttt{l} and index \texttt{k}.

\ur{I don't follow the explanation below, e.g., ``assuming conclusion''}

With this goal in mind, we now set about deriving \texttt{qSelect1} and \texttt{qSelect2} from our colorized \texttt{qSelect}.
Because the original function is recursive,
the first step in this process is to assume the conclusion at the call sites.

\ur{quickselect 1 and 2: how did we introduce them?}

This is performed in place:
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in`2`
      case compare k n of
        LT => qSelect2 (`1`qSelect1 left`2`) k
      | EQ => h
      | GT => qSelect2 (`1`qSelect1 right`2`) (k-n-1)`
\end{lstlisting}
We then pull the stage \bbone\ portion of each branch of the
comparison up into a single tuple.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      let p = `1`(h,qSelect1 left,qSelect1 right) in`2`
      case compare k n of
        LT => qSelect2 (pi2 p) k
      | EQ => pi1 p
      | GT => qSelect2 (pi3 p) (k-n-1)`
\end{lstlisting}

\ur{Don't know what ``this last step'' refers to.}
This last step is necessary, but we can perform it only because we're willing to ignore the control dependence.
We continue to pull up the precomputation and include more in it:

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

This completes the splitting of the cons branch.  
The empty branch can also be split, though in a trivial way:

\ur{Not clear, ``can be split'' split.  Does or doesn't (split) should be the verb?}
\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  case l of
    Empty => let p = () in `2`0`1`
  | Cons (h,t) => 
      let p =
        let (left,right,n) = partition h t in
        `1`(n,h,qSelect1 left,qSelect1 right)
      in`2`
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

To continue splitting the case statement, we need to generate a sum type in the precomputation
to remind the code what branch its in.

\begin{lstlisting} 
fun qSelect (1`l : list`) (2`k : int`) : 2`int` = 1`
  let p = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in`1`
        Bot (n,h,qSelect1 left,qSelect1 right)
  in`2`
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

We finish where we start, splitting \texttt{qSelect} into
\texttt{qSelect1} and \texttt{qSelect2}.  
Omitting a few type annotations, that is:

\begin{lstlisting} 
1`fun qSelect1 (l : list) = 
    case l of
      Empty => Top ()
    | Cons (h,t) =>
        let (left,right,n) = partition h t in
        Bot (n,h,qSelect1 left,qSelect1 right)
`2`
fun qSelect2 p (k : int) : int = 
  case p of 
    Top () => 0
  | Bot p => 
      case compare k (pi1 p) of
        LT => qSelect2 (pi3 p) k
      | EQ => pi2 p
      | GT => qSelect2 (pi4 p) (k-n-1)`
\end{lstlisting}

### 2014-11-11 paper/semantics.tex
major reorganization of section 2 (in progress) - moving lots of text around!

@@
 %terminology, as summarized in \ref{fig:termSplitSummary}.
+
+\crem{Fold in the next paragraph:}
+
+We start by defining a simple staged language, called \lang, to express the
+input to our splitting algorithm.  \lang\ unambiguously specifies how its terms
+should be split
+% and features a type system that ensures valid splittings exist.
+% only minimal refactoring should be required to write terms within this language,
+%and from there, the type system should be enough to prove that a valid splitting exists.
+by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
+Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate
+it---all stage-\bbone\ subexpressions are evaluated before stage-\bbtwo\ ones.
+After splitting, those parts of the \lang\ term that were identified with stage
+\bbone\ will end up in the stage \bbone\ function ({\tt qSelect1}), and those
+parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function
+({\tt qSelect2}).  Typing rules for \lang\ ensure a valid splitting exists.
+That is, within well-typed terms, information can flow from stage \bbone\ to
+stage \bbtwo\ portions, but never from \bbtwo\ to \bbone.
 

### 2014-11-11 paper/splitting.tex
various fixes
added residual++ splitting subection

@@
 \begin{lstlisting}
-1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
+1`(1+2, next{`2`7 + 5`1`})`
 \end{lstlisting}
@@
 \input{figures/splitting-misc}
+
+\subsection{Improved Definition}
+\label{sec:resumers}
+
+The definitions of masking and splitting can be augmented to better handle special cases.
+One such special case is the behavior of stage \bbone\ splitting at stage \bbone\ types that contain no $\fut$s or $\to$s.
+We call such types {\em monostage} (judgmentally, $A~mono$), because a partial value $v$ with a monostage type can contain no $\next$s.
+
+Intuitively, for $[\xi;v] : A$ and $A~mono$, we would expect that $\maskt{\xi;v}$ only reduce to something ``trivial."
+There are two reasonable definitions of trivial here: 
+\begin{enumerate}
+\item a value is trivial if it is exactly \texttt{()},
+\item a value is trivial if it is in the closure of \texttt{()} and \texttt{(-,-)}
+\end{enumerate}
+Indeed, as $\maskt{-;-}$ is currently defined, this property holds for the latter definition of trivial,
+but we can make it hold for the former definition by adjusting just the tuple rule:
+\[
+\maskt{\cdot;(v_1,v_2)} = \left\{ \begin{array}{ll} 
+\mathtt{()} &\text{if } (v_1,v_2):A~mono\\ 
+(\maskt{\cdot;v_1},\maskt{\cdot;v_2}) &\text{otherwise} \end{array} \right.
+\]
+From here, the splitting rules can be made to comport with this new 
+definition adding special-case tuple, projection, and application 
+rules\footnote{The old tuple, projection, and application rules 
+must also be augmented to apply only when the $mono$ premises don't hold.
+Also, all splitting rules need to be augmented to carry types, though
+we omit the contexts here.}:
+\begin{mathpar}
+\infer
+	{(e_1,e_2):A\times B\splitonesym [\ldots,(l_1,l_2).(r_1;r_2;())]}
+	{e_1 : A \splitonesym [c_1,l_1.r_1] 
+	& e_2 : B \splitonesym [c_2,l_2.r_2] 
+	& A \times B~mono} 
+\and
+\infer
+	{\pio~e:A\splitonesym [\ldots,l.(r;())]}
+	{e : A\times B \splitonesym [c,l.r] 
+	& A \times B~mono} 
+\and
+\infer
+	{e_1~e_2: B\splitonesym [\ldots,(l_1,l_2,l_3).(r_1 (r_2,l_3); ())]}
+	{e_1 : A \to B \splitonesym [c_1,l_1.r_1] 
+	& e_2 : A \splitonesym [c_2,l_2.r_2] 
+	& B~mono} 
+\end{mathpar}

### 2014-11-11 paper/implementation.tex
partway through editing section 2
added half example

@@
 \]
+
 This sort of optimization does not affect the contract of splitting at all,

### 2014-11-11 paper/example.tex

@@
   be cited.
-\item the splitting example should shed more light to the techniques
-  and try to give more insight.
 \end{itemize}
@@
 
-\crem{
-Outline:
-\begin{enumerate}
-\item Say we want to run a bunch of quickselect queries on a single list.
-\item We observe that some of the code depends only on the list.
-\item We can write this code in \lang\ to make that dependency clear.
-\item We can intuitively split into two functions.
-\item This is asymptotically better.
-\end{enumerate}
-}
-
 \input{figures/quickselect}
-
-Staging is a technique for using language constructs and types to
-explicitly divide a computation according to the frequency or the
-arrival time of data.
 
@@
 linear time behavior.}
-and then recurs on one of the two resulting sides to find the desired element. The
-side chosen is determined by the relationship of \texttt{k} to the size of the
-first half \texttt{n}, which is returned by the \texttt{partition} function
-along with the two sides.
+and then recurs on one of the two resulting sides to find the desired element.
+The side chosen is determined by the relationship of \texttt{k} to the size of
+the first half \texttt{n}, which is returned by the \texttt{partition} function
+along with the two sides themselves.
 
-Now, consider an application where we perform many order statistics queries
-on the same collection \texttt{l}, but with $m$ different ranks \texttt{k}:
+Now, consider an application where we perform many order statistics queries on
+the same collection \texttt{l}, but with $m$ different ranks \texttt{k}.
+Certainly, it is possible to implement this with $m$ calls to
+\texttt{quickselect}:
 %
 \begin{lstlisting}
-quickselect l 1
-quickselect l 2
- @$\ldots$@ 
-quickselect l @$m$@.
+quickselect l @$k_1$@
+quickselect l @$k_2$@
+ @$\vdots$@ 
+quickselect l @$k_m$@.
 \end{lstlisting}
 %
-To take advantage of this, we can \emph{stage} quickselect so that the parts of
-the computation that depend on the collection~\texttt{l} is performed once, and
-the desired result is computed by using~\texttt{k} and the results from this
-prior stage.
+Can we do better?
 
-%% There has been the extensive research into the question of how to
-%% automatically add staging annotations to unstaged code.  This process
-%% is known as {\em binding time analysis}, and we do not consider it
-%% here.  Instead, we assume that all input programs are properly staged
-%% according to some programmer's intent.
+\subsection{Staging}
 
-\crem{missing text}
+An astute programmer might notice that much of the code in \texttt{quickselect}
+does \emph{not} depend on the rank \texttt{k}---for example, the list is
+partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
+to \texttt{qSelect} are guarded by a comparison to \texttt{k}, the lists passed
+in to those recursive calls also don't depend on \texttt{k}. Thus, if we are
+willing to change the evaluation order of the language (and evaluate under the
+\texttt{case} statement), it should be possible to set things up so that
+\emph{all} the computation involving \texttt{l} is performed before \emph{any}
+of the computation involving \texttt{k}.
 
-Using techniques essentially the same as
-prior work, we can stage our quickselect code by using staged typing
-and inserting three keywords~\texttt{next, prev, hold} into the
-appropriate places in the code as shown in \ref{fig:qs-staged}. To
-write the staged version, the programmer first observes that the
-rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
-rather than $\rmint$, indicating that it will be available in the
-second stage.  All types that are not of circle type are available in
-the first stage.  Naturally the output type, representing the $k$th
-largest element of the list, is also $\fut\rmint$, since it is not
-computed until the second stage.
+We can formalize this intuition by writing quickselect in \lang\
+(\ref{sec:semantics}), a \emph{staged}, typed lambda calculus. In \lang, we can
+directly express the idea that the argument \texttt{l} is known at the
+\emph{first} stage of the computation, while the argument \texttt{k} is only
+known at the \emph{second}, and all purely-stage-\bbone\ computation occurs
+before any stage \bbtwo\ computation.
 
-\crem{explain circle types}
+A staged version of quickselect is written in \ref{fig:qs-staged}, where stage
+\bbone\ code is red, and stage \bbtwo\ code is blue. Whereas before
+\texttt{qSelect} took a \textrm{list} and \rmint\ and returned an \rmint, now we
+must be more careful---at stage \bbone, \texttt{qsStaged} takes a \textrm{list}
+and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a
+$\fut\rmint$.
 
-The construct $\next$ occurs in a stage \bbone\ context and indicates that the
-contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
-\bbtwo\ context and indicates that the contents of its block is stage \bbone.
-(We adopt the convention that the top-level context is stage \bbone.)
+At stage \bbone, all terms of non-circle type are available for
+immediate use. Indeed, we are able to case on \texttt{l} as usual, and in the
+\texttt{Cons} branch, pass it to the stage \bbone\ function \texttt{partition}.
 
-Informally speaking, the $\next$ construct enables operating on second stage
-computations in the first stage essentially by treating them as boxes that can
-be moved around, stored in data structures, and passed to functions, and the
-$\prev$ construct enables unwrapping the contents of such a box to use in the
-second stage.
+\crem{Very rough text for the remainder of the section. Be a bit more explicit.}
 
-For convenience, the code also uses $\pause$ blocks.  This construct
-has same stage signature as $\prev$, but instead of ``unwrapping''
-$\fut$ types it simply promotes integers from stage~\bbone\ to
-stage~\bbtwo.%
+However, \texttt{k} has type $\fut\rmint$, so we cannot use it as part of a
+stage \bbone\ computation. Luckily, we are trying to produce a $\fut\rmint$. To
+create one of these, we can wrap $\next$ around a stage \bbtwo\ computation of
+an integer; in that computation, we can wrap $\prev$ around the $\fut\rmint$
+\texttt{k} in order to use it as a stage \bbtwo\ integer.
+
+(There's a subtle difference here between an \emph{int later} at the first
+stage, and an int at the second stage.)
+
+Essentially, $\next$ allows us to pass around opaque stage \bbtwo\ computations
+at stage \bbone, and the $\prev$ construct lets us unwrap those computations for
+use at stage \bbtwo.
+
+For convenience, the code also uses $\pause$ blocks. This construct has same
+stage signature as $\prev$, but instead of ``unwrapping'' $\fut$ types it simply
+promotes integers from stage \bbone\ to stage \bbtwo.%
 \footnote{$\pause$ can be implemented from our other language features, with
@@
 
-We note that given the type signature of staged quickselect, there are several
-ways to insert the staging annotations.  The version we presented here maximizes
-the work done at stage~\bbone.
+The process of adding staging annotations to unstaged code has been the subject
+of extensive research under the name of \emph{binding time analysis}. In this
+paper, we assume that these annotations have been provided by the programmer (or
+perhaps a binding time analysis tool), and do not consider the problem of
+generating such annotations. There are many ways to annotate any program,
+including \texttt{qselect}; the version presented here matches the intuition for
+the maximal amount of work that can be performed at stage \bbone.
 
-The type system guarantees that all first-stage computations depend only on the
-first stage values and can indeed be computed in the absence of second-stage
-values.
+The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
+typechecks, it is a valid way of dividing the program into two stages, where the
+stage \bbone\ computations can all be performed prior to the stage \bbtwo\
+computations. In \ref{sec:semantics} we provide an evaluation strategy for
+\lang\ which validates that claim.
 
-\crem{partial evaluation requires compile-time access to first stage, whereas we
-don't}
+%The construct $\next$ occurs in a stage \bbone\ context and indicates that the
+%contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
+%\bbtwo\ context and indicates that the contents of its block is stage \bbone.
+
+%(We adopt the convention that the top-level context is stage \bbone.)
+
+\subsection{Splitting Staged Programs}
+
+\input{figures/quickselect-split}
+
+Because we have found a valid staging \texttt{qsStaged}, one might ask what
+happens if we split this function into \emph{two} functions, one of which
+performs all the stage \bbone\ computations given \texttt{l}, and the second
+which performs all the stage \bbtwo\ computations given \texttt{k} and the
+results of the previous.
+
+Intuitively, what should this be? All calls to \texttt{partition} are in stage
+\bbone, so this means we build a binary search tree from \texttt{l}. All
+decisions on which side to recur down happen in stage \bbtwo, so this means we
+traverse the tree once we have the rank \texttt{k}.
+
+(There are two kinds of things we must pass to the second function---which stage
+\bbtwo\ computations to use, and also anything that was held.)
 
@@
 
-Here, we wish to derive two functions, \texttt{quickselect1} and
-\texttt{quickselect2}, that are semantically equivalent to quickselect in
-the sense that:
+Formally, we wish to derive two functions, \texttt{qSelect1} and
+\texttt{qSelect2}, which respectively contain all the stage \bbone\ and stage
+\bbtwo\ content of \texttt{qSelect}, and are moreover semantically equivalent to
+\texttt{qSelect} in the sense that:
 \\
-\texttt{quickselect l k = quickselect2 (quickselect1 l) k}.
+\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
 \\
-\crem{and moreover they correspond to the stages}
 
 Since \texttt{quickselect1} requires only on a first stage
-argument~\texttt{l}, we can run it once to produce a boundary data
-structure that can be used for all subsequent queries on~\texttt{l},
-e.g.,  the code consisting of $m$ calls 
-\begin{lstlisting}
-quickselect l 1
-quickselect l 2
- @$\ldots$@ 
-quickselect l @$m$@
-\end{lstlisting}
-%
-can be replaced by 
+argument \texttt{l}, we can run it once to produce a boundary data
+structure that can be used for all subsequent queries on \texttt{l},
+e.g., we can perform $m$ queries by simply running:
 %
 \begin{lstlisting}
-b = quickselect1 l
-quickselect2 b 1
-quickselect b 2
-@$\ldots$@ 
-quickselect b @$m$@.
+let b = qSelect1 l in
+  qSelect2 b @$k_1$@
+  qSelect2 b @$k_2$@
+   @$\vdots$@ 
+  qSelect2 b @$k_m$@.
 \end{lstlisting}
 
-In this paper, we show that it is possible to generalize data
-specialization to a fully general language with recursion and
-recursive types.  The key ideas behind our approach are: ...
-
-\crem{walk through manual example here}
-
-The final version, cleaned up and given better datatype names, is shown in
-\ref{fig:qssplit}.  Observe that \texttt{qSelect1} essentially builds a binary
-search tree from the input list, and \texttt{qSelect2} searches that tree for
-the correct index.
+Indeed, in \ref{sec:splitting,sec:implementation} we describe an algorithm for
+splitting staged terms, which on \texttt{qsStaged} actually emits the algorithm
+in \ref{fig:qssplit}.
 
@@
 
-Note that the desired optimized code shows above is intellectually more
-sophisticated than the code that we have started with: the optimized code is
-able to create a data structure, a balanced binary tree augmented with indexing
-information, and use a binary search technique over this tree to compute the
-result asymptotically more efficiently.
+%Note that the desired optimized code shows above is intellectually more
+%sophisticated than the code that we have started with: the optimized code is
+%able to create a data structure, a balanced binary tree augmented with indexing
+%information, and use a binary search technique over this tree to compute the
+%result asymptotically more efficiently.
 %In fact, based our teaching experience, we can imagine this kind of problem to
@@
 
-This paper formally describes and analyzes the splitting transformation
-exemplified above.  (I also want to make it clear that recognizing \lang's
-appropriateness for this is itself is a contribution.)
-
-\input{figures/quickselect-split}
+%(I also want to make it clear that recognizing \lang's appropriateness for this
+%is itself is a contribution.)

### 2014-11-12 paper/example.tex
finished (a draft of) section 2

@@
 Now, consider an application where we perform many order statistics queries on
-the same collection \texttt{l}, but with $m$ different ranks \texttt{k}.
+the same collection \texttt{l}, but with $m$ different ranks
+$\mathtt{k_1},\dots,\mathtt{k_m}$.
 Certainly, it is possible to implement this with $m$ calls to
@@
 \begin{lstlisting}
-quickselect l @$k_1$@
-quickselect l @$k_2$@
+quickselect l @$\mathtt{k_1}$@
+quickselect l @$\mathtt{k_2}$@
  @$\vdots$@ 
-quickselect l @$k_m$@.
+quickselect l @$\mathtt{k_m}$@.
 \end{lstlisting}
@@
 An astute programmer might notice that much of the code in \texttt{quickselect}
-does \emph{not} depend on the rank \texttt{k}---for example, the list is
+does \emph{not} depend on the rank \texttt{k}; for example, the list is
 partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
-to \texttt{qSelect} are guarded by a comparison to \texttt{k}, the lists passed
-in to those recursive calls also don't depend on \texttt{k}. Thus, if we are
-willing to change the evaluation order of the language (and evaluate under the
-\texttt{case} statement), it should be possible to set things up so that
-\emph{all} the computation involving \texttt{l} is performed before \emph{any}
-of the computation involving \texttt{k}.
+to \texttt{qSelect} are guarded by a comparison to \texttt{k}, \texttt{k} does
+not determine the arguments to those recursive calls---it only determines which
+call is made. Thus, if we are willing to change the evaluation order of the
+language (and evaluate under the \texttt{case} statement), it should be possible
+to set things up so that \emph{all} the computation involving \texttt{l} is
+performed before \emph{any} of the computation involving \texttt{k}.
 
-We can formalize this intuition by writing quickselect in \lang\
-(\ref{sec:semantics}), a \emph{staged}, typed lambda calculus. In \lang, we can
-directly express the idea that the argument \texttt{l} is known at the
-\emph{first} stage of the computation, while the argument \texttt{k} is only
-known at the \emph{second}, and all purely-stage-\bbone\ computation occurs
-before any stage \bbtwo\ computation.
+We can formalize this intuition by writing quickselect in a \emph{staged}
+language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
+lambda calculus. This allows us to directly express the idea that the argument
+\texttt{l} is known at the \emph{first} stage of the computation, while the
+argument \texttt{k} is only known at the \emph{second}, and all
+purely-stage-\bbone\ computation occurs before any stage \bbtwo\ computation.
 
-A staged version of quickselect is written in \ref{fig:qs-staged}, where stage
-\bbone\ code is red, and stage \bbtwo\ code is blue. Whereas before
-\texttt{qSelect} took a \textrm{list} and \rmint\ and returned an \rmint, now we
-must be more careful---at stage \bbone, \texttt{qsStaged} takes a \textrm{list}
-and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a
-$\fut\rmint$.
+We define a staged version of quickselect in \ref{fig:qs-staged}, writing stage
+\bbone\ code in red, and stage \bbtwo\ code in blue. Whereas \texttt{qSelect}
+sends a \textrm{list} and \rmint\ to an \rmint, \texttt{qsStaged} is more
+precise---it is a stage \bbone\ function which takes a \textrm{list} and a
+$\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
 
-At stage \bbone, all terms of non-circle type are available for
-immediate use. Indeed, we are able to case on \texttt{l} as usual, and in the
-\texttt{Cons} branch, pass it to the stage \bbone\ function \texttt{partition}.
+The stage \bbone\ portions of \texttt{qsStaged} look like \texttt{qSelect}: any
+term of non-circle type is available for immediate use. Indeed, we case on
+\texttt{l} as usual, and in the \texttt{Cons} branch, \texttt{partition} it.
 
-\crem{Very rough text for the remainder of the section. Be a bit more explicit.}
+The next part of the original \texttt{qSelect} function depends on \texttt{k},
+which has type $\fut\rmint$, indicating that it is only available to stage
+\bbtwo\ computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$,
+the remainder of the function is stage \bbtwo\ code.
 
-However, \texttt{k} has type $\fut\rmint$, so we cannot use it as part of a
-stage \bbone\ computation. Luckily, we are trying to produce a $\fut\rmint$. To
-create one of these, we can wrap $\next$ around a stage \bbtwo\ computation of
-an integer; in that computation, we can wrap $\prev$ around the $\fut\rmint$
-\texttt{k} in order to use it as a stage \bbtwo\ integer.
-
-(There's a subtle difference here between an \emph{int later} at the first
-stage, and an int at the second stage.)
-
-Essentially, $\next$ allows us to pass around opaque stage \bbtwo\ computations
-at stage \bbone, and the $\prev$ construct lets us unwrap those computations for
-use at stage \bbtwo.
-
-For convenience, the code also uses $\pause$ blocks. This construct has same
-stage signature as $\prev$, but instead of ``unwrapping'' $\fut$ types it simply
-promotes integers from stage \bbone\ to stage \bbtwo.%
+The $\next$ wraps a stage \bbtwo\ expression of type $\rmint$ (the \texttt{case}
+expression) as a stage \bbone\ expression of type $\fut\rmint$ (the result type
+of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
+$\prev$ unwraps stage \bbone\ $\fut\rmint$s (the results of the recursive calls
+to \texttt{qsStaged}) as stage \bbtwo\ $\rmint$s. In the \texttt{EQ} branch,
+$\pause$ promotes an integer at stage \bbone\ (the head of \texttt{l}) directly
+to an integer at stage \bbtwo.%
 \footnote{$\pause$ can be implemented from our other language features, with
@@
 
-The process of adding staging annotations to unstaged code has been the subject
-of extensive research under the name of \emph{binding time analysis}. In this
-paper, we assume that these annotations have been provided by the programmer (or
-perhaps a binding time analysis tool), and do not consider the problem of
-generating such annotations. There are many ways to annotate any program,
-including \texttt{qselect}; the version presented here matches the intuition for
-the maximal amount of work that can be performed at stage \bbone.
+The process of adding staging annotations ($\fut$ types, $\next$, $\prev$, and
+$\pause$) to unstaged code has been the subject of extensive research under the
+name of \emph{binding time analysis}. In this paper, we assume that these
+annotations have been provided by the programmer (or perhaps a binding time
+analysis tool), and do not consider the problem of generating such annotations.
+There are many ways to annotate any program, including \texttt{qSelect}; we
+chose annotations which maximize the work performed at stage \bbone.
 
-The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
-typechecks, it is a valid way of dividing the program into two stages, where the
-stage \bbone\ computations can all be performed prior to the stage \bbtwo\
-computations. In \ref{sec:semantics} we provide an evaluation strategy for
-\lang\ which validates that claim.
-
-%The construct $\next$ occurs in a stage \bbone\ context and indicates that the
-%contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
-%\bbtwo\ context and indicates that the contents of its block is stage \bbone.
+%The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
+%typechecks, it is a valid way of dividing the program into two stages, where the
+%stage \bbone\ computations can all be performed prior to the stage \bbtwo\
+%computations. In \ref{sec:semantics} we provide an evaluation strategy for
+%\lang\ which validates that claim.
 
 %(We adopt the convention that the top-level context is stage \bbone.)
+
+%pass around opaque stage \bbtwo\ computations at stage \bbone
 
@@
 
-Because we have found a valid staging \texttt{qsStaged}, one might ask what
-happens if we split this function into \emph{two} functions, one of which
-performs all the stage \bbone\ computations given \texttt{l}, and the second
-which performs all the stage \bbtwo\ computations given \texttt{k} and the
-results of the previous.
+An astute programmer, having noticed that \texttt{quickselect} can be staged in
+this fashion, might try to split it into a pair of functions, one which performs
+all the work depending only on \texttt{l} (stage \bbone), and one which uses
+\texttt{k} and the result of that work on \texttt{l} to compute the element with
+rank \texttt{k} in \texttt{l}.
 
-Intuitively, what should this be? All calls to \texttt{partition} are in stage
-\bbone, so this means we build a binary search tree from \texttt{l}. All
-decisions on which side to recur down happen in stage \bbtwo, so this means we
-traverse the tree once we have the rank \texttt{k}.
+Intuitively, \texttt{l} determines the result of all calls to
+\texttt{partition}, and \texttt{k} only determines which calls are made. So we
+can preprocess \texttt{l} by recursively dividing it into halves smaller and
+greater than the pivot---that is, building a binary search tree. Then, once we
+have \texttt{k}, we can recur on this tree, choosing whichever branch has the
+\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. Crucially,
+because \texttt{partition} has no stage \bbtwo\ content, we need not perform any
+of its work at stage \bbtwo!
 
-(There are two kinds of things we must pass to the second function---which stage
-\bbtwo\ computations to use, and also anything that was held.)
-
-\crem{introduce boundary terminology}
-
-Formally, we wish to derive two functions, \texttt{qSelect1} and
-\texttt{qSelect2}, which respectively contain all the stage \bbone\ and stage
-\bbtwo\ content of \texttt{qSelect}, and are moreover semantically equivalent to
-\texttt{qSelect} in the sense that:
-\\
-\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
-\\
-
-Since \texttt{quickselect1} requires only on a first stage
-argument \texttt{l}, we can run it once to produce a boundary data
-structure that can be used for all subsequent queries on \texttt{l},
-e.g., we can perform $m$ queries by simply running:
+We have implemented this splitting of quickselect in \ref{fig:qs-split}.
+\texttt{qSelect1} builds a binary search tree from the list \texttt{l}, and
+\texttt{qSelect2} takes such a tree and a rank \texttt{k} and computes the
+answer. This allows us to efficiently perform many order statistics queries on
+\texttt{l} by caching the tree and reusing it for many different ranks
+$\mathtt{k_1},\dots,\mathtt{k_m}$:
 %
@@
 
-Indeed, in \ref{sec:splitting,sec:implementation} we describe an algorithm for
-splitting staged terms, which on \texttt{qsStaged} actually emits the algorithm
-in \ref{fig:qssplit}.
-
-Assuming that the list~\texttt{l} contains $n$ elements, this optimization
-changes the asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$
-to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
+Assuming \texttt{l} contains $n$ elements, this optimization changes the
+asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$ to
+$\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
 complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
 improvement.
+
+In this paper, we develop a splitting algorithm
+(\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
+produces a pair of terms $e_1,e_2$ containing respectively all the stage \bbone\
+and stage \bbtwo\ content of $e$, such that $e_2~e_1$ is equivalent to $e$. In
+the case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
+described above.
+
+%\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
+
+%Because the tree passes information across the stage boundary, we call it the
+%\emph{boundary data structure}.
 
@@
 %result asymptotically more efficiently.
+
 %In fact, based our teaching experience, we can imagine this kind of problem to

### 2014-11-12 paper/semantics.tex
migrate overview section's text into semantics section

--- 2014-11-11 paper/semantics.tex
@@
-\section{Terminology and Overview}
+\section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
+\label{sec:semantics}
 
@@
 
-%We present an overview of the formal development along with our
-%terminology, as summarized in \ref{fig:termSplitSummary}.
+\input {figures/grammars}
 
-Our starting point is \lang, a typed, two-stage lambda calculus with products,
-sums, and isorecursive types. This language is adapted from \cite{davies96}, and
-is very similar to the staged languages considered in the partial evaluation and
-metaprogramming literature. In this paper, we restrict our attention to a
-two-stage language to simplify our presentation of the splitting algorithm, but
-we expect that the techniques presented here extend to more stages.
+The input to our splitting algorithm is given in \lang, a typed, two-stage
+lambda calculus with products, sums, and isorecursive types. This language is
+adapted from \cite{davies96}, and is very similar to the staged languages
+considered in the partial evaluation and metaprogramming literature. In this
+paper, we restrict our attention to a two-stage language to simplify our
+presentation of the splitting algorithm, but we expect that the techniques
+presented here extend to more stages.
 
-The dynamic semantics of \lang\ differs from Davies's approach and
-similar approaches in metaprogramming literature, because it avoids
-duplication of staged work. Duplication takes place in meta
-programming when a second-stage computation is silently duplicated by
-the semantics.  In addition to leading to increases in run time, a
-duplicating semantics makes it harder to reason about the splitting
-algorithm, whose development and exposition is our main goal.  Apart
-from the non-duplication, the dynamic semantics of \lang\ is
-essentially identical to  Davies'.
+Our dynamic semantics of \lang\ differ from those given in \cite{davies96} and
+the metaprogramming literature, because we avoid duplicating stage \bbtwo\
+computations. In addition to being more efficient, our non-duplicating semantics
+are easier to compare with our splitting algorithm, whose development and
+exposition is our main goal.
 
-We define the dynamics semantics of \lang\ not as an explicit
-evaluation relation but as a relation that produces a term in an
-unstaged, completely standard language, which we refer to as
-\langmono\, because it is essentially lambda calculus extended with
-product types. Given a \lang\ term, the dynamic semantics can be used
-to reduce the term to a \langmono\ program, called {\em residual} by
-using one application of stage-1 evaluation, one application of
-reification, followed by one application of stage-2 evaluation.
-
-\ur{Splitting...}
-
-\section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
-\label{sec:semantics}
-
-\crem{Fold previous section in here.}
-
-\input {figures/grammars}
+%Apart from the non-duplication, the dynamic semantics of \lang\ is essentially
+%identical to Davies'.
 
@@
 \label{sec:stagedsemantics}
+
+\crem{fold in the next paragraph}
+
+Given a \lang\ term, the dynamic semantics can be used
+to reduce the term to a \langmono\ program, called {\em residual} by
+using one application of stage-1 evaluation, one application of
+reification, followed by one application of stage-2 evaluation.
 

### 2014-11-12 paper/semantics.tex
2-evaluation -> residualization

@@
 second stage all stage-2 terms are evaluated by a relation
-$\redtwosym$, called {\em 2-evaluation}.  To eliminate duplication,
+$\redtwosym$, called {\em residualization}.  To eliminate duplication,
 our semantics lifts out each stage-2 term nested inside of a stage-1
@@
 evaluation of \next: 1-evaluation reduces a \next\ expression by
-reducing the body of the \next\ with 2-evaluation to obtain a
+reducing the body of the \next\ with residualization to obtain a
 residual $q$ and returns a residual table that maps a freshly
@@
 
-The relation 2-evaluation, defined by $\redtwosym$ judgment, as shows
+The relation residualization, defined by $\redtwosym$ judgment, as shows
 in \ref{fig:diaSemantics}, takes a stage \bbtwo\ term and reduces it to
 a residual.  Apart from the evaluation of \prev\ and \pause, the
-evaluation is straightforward. To reduce a \prev\ term, 2-evaluation
+evaluation is straightforward. To reduce a \prev\ term, residualization
 reduces the body by 1-evaluation to obtain a residual table and a
@@
 using reification, and then reduce this term to a residual via
-2-evaluation.  Finally, we evaluate the residual, which is a term in a
+residualization.  Finally, we evaluate the residual, which is a term in a
 our monostaged target language \langmono, to a value by using  a
@@
 %% relations $\redonesym$ and $\redtwosym$, called {\em 1-evaluation} and
-%% {\em 2-evaluation}, that respectively evaluates stage-1 and stage-2
+%% {\em residualization}, that respectively evaluates stage-1 and stage-2
 %% expressions. The relation 1-evaluation maps a stage-1 term to a
@@
 %% code stored in the {\em residual table}, which maps variables to
-%% residuals.  The relation 2-evaluation takes a stage-2 term and maps to
+%% residuals.  The relation residualization takes a stage-2 term and maps to
 %% a residual.  To generate the residual table, the dynamics semantics
@@
 %% the residual table by using reification, and then reduce this term to
-%% a residual via 2-evaluation.
+%% a residual via residualization.
 

### 2014-11-13 paper/splitting.tex
expansion of 5.4

--- 2014-11-09 paper/splitting.tex
@@
 \label{sec:splitting}
-
-\input{figures/splitting-overview}
 
@@
 \input{figures/splitting-misc}
+
+\subsection{Improved Definition}
+\label{sec:resumers}
+
+The definitions of masking and splitting can be augmented to better handle special cases,
+particularly ones that depend on the type of the term we're splitting.
+One such special case is the behavior of stage \bbone\ splitting at stage \bbone\ types that contain no $\fut$s or $\to$s.
+We call such types {\em monostage} (judgmentally, $A~mono$), 
+because a partial value $v$ with a monostage type can contain no $\next$s,
+and so must comprise only stage \bbone\ information.
+
+Intuitively, for $[\xi;v] : A$ and $A~mono$, 
+we would expect that $\maskt{\xi;v}$ only reduce to something that carries no information.
+This indeed happens: for example, $\maskt {\cdot;\mathtt{(inl~3,5)}}$ reduces to \texttt{((),())}.
+But while \texttt{((),())} technically ``carries no information," it still has non-trivial structure.
+This is not only unintuitive, but potentially expensive.
+Instead, we desire a stronger property, that $\maskt{\xi;v}$ only reduce to exactly \texttt{()}. 
+We can achieve this by adjusting just the tuple rule:
+\[
+\maskt{\cdot;(v_1,v_2)} = \left\{ \begin{array}{ll} 
+\mathtt{()} &\text{if } (v_1,v_2):A~mono\\ 
+(\maskt{\cdot;v_1},\maskt{\cdot;v_2}) &\text{otherwise} \end{array} \right.
+\]
+From here, the splitting rules can be made to comport with this new 
+definition adding special-case tuple, projection, and application 
+rules:
+\begin{mathpar}
+\infer
+	{(e_1,e_2):A\times B\splitonesym [\ldots,(l_1,l_2).(r_1;r_2;())]}
+	{e_1 : A \splitonesym [c_1,l_1.r_1] 
+	& e_2 : B \splitonesym [c_2,l_2.r_2] 
+	& A \times B~mono} 
+\and
+\infer
+	{\pio~e:A\splitonesym [\ldots,l.(r;())]}
+	{e : A\times B \splitonesym [c,l.r] 
+	& A \times B~mono} 
+\and
+\infer
+	{e_1~e_2: B\splitonesym [\ldots,(l_1,l_2,l_3).(r_1 (r_2,l_3); ())]}
+	{e_1 : A \to B \splitonesym [c_1,l_1.r_1] 
+	& e_2 : A \splitonesym [c_2,l_2.r_2] 
+	& B~mono} 
+\end{mathpar}
+The old tuple, projection, and application rules 
+must also be augmented to apply only when the $mono$ premises don't hold.
+Also, all splitting rules need to be augmented to carry types, though
+we omit the contexts here.

### 2014-11-26 paper/implementation.tex
some work on the mono section

@@
 \section {Implementation and Optimization}
+\label{sec:implementation}
 
@@
 with no restrictions on the form of that precomputation.
-Further gains can be realized by changing the contract of splitting itself.
 
-Let a type be {\em present} if it has no $\to$s or $\fut$s in its type.
-Observe that for any $v$ with a present type, $\maskt{v}$ will always be a nested bunch of \texttt{()}s,
-which is to say that it carries no information.
-
-\nr{We probably need an example of this.}
-
-Of course, \texttt{()} is a more compact representation that carries no information.
-Thus, we can change the definition of masking so that $\maskt{v}$ is always \texttt{()},
-and change the splitting rules accordingly.
-
-Back to the boundary type problem.
-The problem becomes more difficult when trying to optimize across functions, including recursive calls.
+The problem becomes more difficult when trying to optimize across function boundaries.
 For example, consider the stage \bbtwo\ term,
@@
 2`prev{
-  1`letfun fact (n : int) : int = 
-    if n <= 0 then 1 else fact(n-1)*n
-  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
+  1`letfun double (n : int) = n + n
+  in next{`2`hold {`1`double 5`2`}-7`1`}`2`
 }`
 \end{lstlisting}
-Noting that the precomputation of the body of \texttt{fact} could be given the type,
+Using the rules above, this term splits into:
 \begin{lstlisting}
-datatype prec = L | R of prec
-\end{lstlisting}
-the whole expression splits into:
-\begin{lstlisting}
-1`letfun fact (n : int) : int * prec = 
-  if n <= 0 then (1,L) 
-  else let (y,z) = fact(n-1) in (y*n,R z)
-in fact 5`
+1`letfun double (n : int) = (n+n, ())
+in double 5`
 
 2`l => 
-letfun fact (n : unit, l0 : prec) : unit = 
-  case l0 of L => () | R l1 => fact ((),l0)
-in (fact (pi2 l); pi1 l)-100`
+letfun double (n : unit, l0 : unit) = ()
+in (double (pi2 l); pi1 l)-100`
 \end{lstlisting}
 
-Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
-always returns \texttt{()}; worse yet, it has linear runtime!
-But recognizing and optimizing this away is in general a global operation,
-so we instead solve this issue by adding a new staging annotation to \lang, called
-\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
-no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
-rules to \lang:
-\begin{mathpar}
-\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
-\infer{A\to B~\mathrm{safe}}{A~\mathrm{present} & B~\mathrm{safe}} \and
-\infer{A~\mathrm{safe}}{A~\mathrm{present}} \and
-\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}} \and
-\infer{\splitone{\monoSt~e}A {(e',()), \_.r} }{e \overset{C}{\rightarrow} e' & A \overset{G}{\rightarrow} r} \and
-\infer{A \overset{G}{\rightarrow} ()}{A ~\mathrm{present}} \and
-\infer{(A \to B)\overset{G}{\rightarrow} r}{B \overset{G}{\rightarrow} \lambda \_.r }
-\end{mathpar}
-Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
-by performing the two rewrites in place:
-\begin{align*}
-\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
-e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
-\end{align*}
-And then $\overset{G}{\rightarrow}$ generates a resumer that corresponds to the type $A$.
-It operates inductively on the safety judgement of $A$.
-At the safe $A$ base case, we make use of the prior optimization that says all safe $A$s have a resumer that evaluates to \texttt{()}.
+Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{double}, since it always returns \texttt{()}. 
+In this case, the wasted work is small, but if \texttt{double} were a recursive function,
+then it's runtime may be unbounded.
+
+There are two ways plausible ways to solve this problem:
+either we could define an optimization step that comes 
+after splitting and show that it tends to fix the inefficiencies that splitting creates; 
+or we could give more structure to the input language so that splitting can generate more efficient code directly.
+
+We choose to investigate the latter approach.
+Specifically, we add new staging features to \lang\ which allow us to indicate that a piece of code is {\em strictly} 
+stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
+At the type level, this manifests as a constructor called $\monoType$.
+So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
+which might fail to terminate at stage \bbtwo.  
+Alternatively $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
+which {\em cannot} have any content at stage \bbtwo.
+Naturally, the well-formedness of $\monoType A$ requires that $A$ contain no $\fut$s.
+
+At the term level, we add the $\monoTerm$, $\lifttag$, \texttt{let mono} constructs to move information
+in and out of $\monoType$.
+
+\input{figures/mono.tex}

### 2014-12-01 paper/implementation.tex
added a more about mono

@@
 letfun double (n : unit, l0 : unit) = ()
-in (double (pi2 l); pi1 l)-100`
+in (double (pi2 l); pi1 l)-7`
 \end{lstlisting}
@@
 So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
-which might fail to terminate at stage \bbtwo.  
-Alternatively $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
+which might fail to terminate at stage \bbtwo,  
+but $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
 which {\em cannot} have any content at stage \bbtwo.
@@
 
-At the term level, we add the $\monoTerm$, $\lifttag$, \texttt{let mono} constructs to move information
-in and out of $\monoType$.
+At the term level, we add the 
+$\monoTerm$, \texttt{letmono}, $\lifttag$, and $\liftint$ 
+constructs to move information in and out of $\monoType$.  
+The statics, dynamics, and splitting rules for these constructs are given in \cref{fig:monoRules}.
+
+The $\monoTerm$ construct allows us to declare a block of code as entirely stage \bbone,
+equivalent to saying that it has no $\next$ blocks.
+This is enforced in the type system by checking at the new world $\bbmono$ within the $\monoTerm$ block.
+Accordingly, $\monoTerm$ blocks are the intro form for the $\monoType$ modality.
+
+The \texttt{letmono} construct allows us to form let bindings of monostage code without going into a $\monoTerm$ block.
+This construct, along with $\monoTerm$ blocks, is enough to allow us to form terms of the following types:
+\begin{itemize}
+\item $\monoType(A \times B) \to \monoType A \times \monoType B$ 
+\item $\monoType A \times \monoType B \to \monoType(A \times B)$
+\item $\monoType A + \monoType B \to \monoType(A + B)$
+\item $\monoType(A \to B) \to (\monoType A \to \monoType B)$
+\end{itemize}
+There are two notable types absent in this list, namely the converses of the latter two entries.
+Lets double-check to make sure that this is the behavior we'd want.
+
+First consider $(\monoType A \to \monoType B) \to \monoType(A \to B)$.
+Were this type inhabited, it would suggest that any function which might not terminate at stage $\bbtwo$
+can be transmuted into a function that does all the same work but definately terminates at stage $\bbtwo$.
+That seems definitely wrong, so it makes sense that this type is uninhabited.
+
+Next, consider $\monoType(A + B) \to \monoType A + \monoType B$.
+That would suggest that if there's a sum tag computed in a monostage way,
+then that sum tag can be lifted out from under the $\monoType$.
+This feature {\em does} sound reasonable, and so we add it explicitly
+with the $\lifttag$ construct. 
+Similarly, we add $\liftint$ to move integers out from under $\monoType$.
+\footnote{As with $\pause$, $\liftint$ can be defined in the language with
+some effort, but we provide it as a convenience.}
+
+With these new features in place, we can now define the \texttt{double}
+function so that it splits reasonably.
+\begin{lstlisting}
+2`prev{
+  1`let mono{`3`double`1`} = mono{`3`fn n : int => n+n`1`}
+  in next{`2`hold {`1`lift(mono{`3`double 5`1`})`2`}-7`1`}`2`
+}`
+\end{lstlisting}
+This now splits into:
+\begin{lstlisting}
+1`let double = (fn n : int => n+n)
+in (double 5,())`
+
+2`l => in (pi1 l)-7`
+\end{lstlisting}
 

### 2014-12-01 paper/paper.tex
uncomment the splitting section
worked on global structure of splitting section
minor tweaks
small changes; lots of commenting-out
rewrite 4.0 - 4.1 with nico
comments entered
minor
migrate overview section's text into semantics section
some work on the mono section
added a more about mono

@@
 \usepackage{amsthm}
-%\usepackage{bbm}
 \usepackage{stmaryrd}
@@
 \usepackage{mathpartir}
+\usepackage{bbm}
 %\usepackage{mathabx}
@@
 \AtBeginDocument{\renewcommand\ref\cref}
+\crefname{section}{Section}{Sections}
 \crefname{figure}{Figure}{Figures}
@@
 \definecolor{stageTwoColor}{RGB}{0,0,127}
+\definecolor{stageMonoColor}{RGB}{0,120,0}
 
@@
         moredelim=*[is][\color{stageTwoColor}]{2`}{`},
+        moredelim=*[is][\color{stageMonoColor}]{3`}{`},
         literate={|->}{{\ $\mapsto$\ }}1
@@
 
-\title{Stage-Splitting a Modal Language}
+\title{Data Specialization for a Staged Lambda Calculus}
 
@@
 
+\section{Intro}
+(placeholder in order to keep section numbers stable)
+
+In \ref{fig:termSplitSummary}, we summarize the terminology of our
+formal development.
+
 %% \input{intro}
@@
 %% %\vfill\eject\pagebreak
-%%\input{splitting}
+\input{splitting}
+\input{implementation}
 \input{algoexamples}

### 2014-12-18 sematics/boundary/boundary.tex
working on boundary type document

@@
 \usepackage{amsmath}
-%\usepackage{amssymb}
+\usepackage{amssymb}
 \usepackage{amsthm}
@@
 \usepackage{graphicx}
+\usepackage[usenames,dvipsnames]{color}
 
@@
 \newcommand {\next}{asdlfkj}
-\input {../../paper/macros}
+\input{../../paper/macros}
 
@@
 
+\newcommand {\ibksym} {\rotatebox[origin=c]{270}{$\therefore$}}
+\newcommand {\btsym} {\rotatebox[origin=c]{270}{$\triangle$}}
 \newcommand {\ktype} {\rm Type}
-\newcommand {\ibk} [2] {{#1}~\rotatebox[origin=c]{270}{$\therefore$}~{#2}}
-\newcommand {\btcont} [3] {{#1}~\triangleright~[{#2},{#3}]}
-\newcommand {\bty} [4] [\Gamma] {{#1}\vdash {#2}\triangleright[{#3},{#4}]}
+\newcommand {\ibk} [2] {{#1}~\ibksym~{#2}}
+\newcommand {\btj} [4] {{#1}:{#2}~@~{#3}~\btsym~{#4}}
+\newcommand {\btyo} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbone{#4}}
+\newcommand {\btyt} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbtwo{#4}}
+\newcommand {\btyw} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}w{#4}}
 \newcommand {\btysub} [2] [] {\bty{e_{#1}:{#2}}{\tau_{#1}}{\sigma_{#1}}}
 
-Consider a language with products and functions.  The internal boundary kind is given by:
+\section{Goals}
+
+The purpose of this document is to explain how to type the boundary produced by our splitting algorithm.
+We'll start with a very simple input language, the boundary of which can be typed in an easy way.
+We'll then start adding features to the language, thereby inducing complexity in the boundary type system.
+
+For the most part, we'll do all of our work with a single new judgement $\btj e A w \tau$, 
+which indicates that the expression $e$ (which has type $A$ at world $w$) has the boundary type $\tau$.
+Ideally, the splitting rules will agree with the boundary type given to us by this judgment.
+That is, if $\btj e A \bbtwo \tau$ and $\splitone{e}{A}{p,l.r}$, 
+then $p:\tau$ and $l:\tau \vdash r:A$ (and a similar relationship at stage \bbone, covered later).
+Mind you, we don't actually bother checking that property in this document.
+
+\section{Straight-line Code}
+
+We first consider an input language which has no functions or divergence.
+At either stage, the boundary type of any expression is 
+just the product of the boundary types of its subexpressions
+(unit in the nullary case).
+The only exception to this pattern is the $\pause$ rule,
+which also multiplies an int into the boundary type.
 
 \begin{mathpar}
-\infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
-\infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
-\infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
-\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
+\infer {\btyw i {\rm int}\rmunit} {\cdot} \and
+\infer {\btyw b {\rm bool}\rmunit} {\cdot} \and
+\infer {\btyw {(e_1,e_2)} {A \times B} {\tau_1 \times \tau_2}} 
+	{\btyw {e_1} A {\tau_1} 
+	&\btyw {e_2} B {\tau_2}} \and
+\infer {\btyw {\pio~e} A \tau} 
+	{\btyw {e} {A \times B} {\tau}} \and
+\infer {\btyw {\pit~e} B \tau} 
+	{\btyw {e} {A \times B} {\tau}} \and
+\infer {\btyo {\next~e} B \tau} 
+	{\btyt {e} {A \times B} {\tau}} \and
+\infer {\btyt {\prev~e} B \tau} 
+	{\btyo {e} {A \times B} {\tau}} \and
+\infer {\btyt {\pause~e} \rmint \tau} 
+	{\btyo e \rmint {\tau \times \rmint}}
 \end{mathpar}
 
-For a statement $e$, the judgment $\btcont {e:A} \tau \sigma$ means that $e$ has the internal boundary type $\tau$ and the external boundary type-constructor $\sigma$.
-Invariant: for $\ibk A \kappa$, if $\btcont {e:A} \tau \sigma$ then $\sigma : \kappa$.
+Finally, we look at the let and var rules.
 
 \begin{mathpar}
-\infer {\bty {i:\rm int}\rmunit {()}} {\cdot} \and
-\infer {\bty {b:\rm bool}\rmunit {()}} {\cdot} \and
-\infer {\bty {\pause~e:\fut\rm int}{\tau\times{\rm int}} {()}} {\bty {e:\rm int}\tau {()}} \and
-\infer {\bty [\Gamma,\btcont {x:A} \rmunit \sigma] {x:A}\rmunit\sigma} {\cdot} \and
-\infer 
-	{\bty {(e_1,e_2):A\times B} {\tau_1 \times \tau_2} {(\sigma_1, \sigma_2)}} 
-	{\btysub [1] A & \btysub [2] B} 
-	\\
-\infer 
-	{\bty {\pio~e:A} \tau {(\pi_1\sigma)}} 
-	{\btysub {A\times B}} \and
-\infer 
-	{\bty {\pit~e:B} \tau {(\pi_2\sigma)}} 
-	{\btysub {A\times B}} 
-	\\
-\infer 
-	{\bty {\lam x A e : A \to B} \rmunit {(\lambda \alpha:\kappa.(\tau,\sigma))}} 
-	{\ibk A \kappa & \bty [\Gamma, \btcont {x:A} \rmunit \alpha] {e:B} \tau \sigma} 
-	\and
-\infer 
-	{\bty {e_1~e_2 : B}
-		{\tau_1 \times \tau_2 \times \pi_1 (\sigma_1~\sigma_2)} 
-		{(\pi_2(\sigma_1~\sigma_2))} }
-	{\btysub [1] {A \to B} & \btysub [2] A } 
+\infer [let] {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
+	{\btyw {e_1} A {\tau_1} 
+	&\btyw [\Gamma,\btj x A w \rmunit] {e_2} B {\tau_2}} \and
+\infer [var] {\btyw [\Gamma,\btj x A w \rmunit] {x} A \rmunit} {\cdot}
+\end{mathpar}
+Notice that we've defined these two rules in a slightly different way than usual.
+Typically, we'd say that whenever some variable $x$ is in the context 
+(because it was put there at a binder site),
+we can form the judgment that $x$ reference sites have a unit boundary.
+Instead what we've done is put the whole boundary type judgment in context at the binding site,
+and then we pull it out of the context at the reference site.
+Strictly speaking, the var rule is merely a special case of the 
+general structural rule stating ``you can use what you assume,'' so we'll omit it from here on out.
+
+This is a pretty common technique  called {\em hypothetical judgments}.
+For now, the use of hypothetical judgments is merely a stylistic difference,
+but it will become important when things get more complicated.
+
+\section{Adding Functions and Fixed Points}
+
+Lets turn this language to 11.  
+In addition to straight-line code, we'll add \texttt{if}s, functions, and fixed points,
+which the astute reader will notice is enough to make the language Turing-complete.
+Before getting into the rules, consider the following example, 
+where $E$ is some purely stage \bbone\ expression:
+
+\[
+\left(
+\tallif {E}
+	{(\lam x \rmint {\next\{\pause\{x\}+1\})}}
+	{(\lam x \rmint {\next\{\pause\{x\}+\pause\{x*x\}\}})}
+\right ) 5
+\]
+
+What do we expect the boundary type to be for this whole piece of code?
+Well, it depends. 
+If $E$ evaluates to true, then we'd expect the boundary type to be $\rmint$,
+but if $E$ evaluates to false, we'd expect the boundary type to be $\rmint \times \rmint$.
+Ergo, figuring out the boundary type of this piece of code is at least as hard 
+as figuring out what any given expression evaluates to.
+Of course, that's not decidable. Yikes.
+
+Fortunately for us, that's not really the end of it.  
+Just because we can't know the boundary type of a piece of code doesn't mean we can't say anything.  
+Maybe instead of giving the {\em exact} answer, we can give a list of possibilities.
+
+\section{Adding Just \texttt{if}-Expressions}
+
+So to explore what this means, lets dial back the language a bit.  
+In addition to straight-line from before, lets just add \texttt{if} expressions.
+It turns out this is easy to integrate into the existing framework:
+
+\begin{mathpar}
+\infer {\btyo {\ifthen {e_1}{e_2}{e_3}} {A} {\tau_1 \times (\tau_2 + \tau_3)}} 
+	{\btyo {e_1} \rmbool {\tau_1} 
+	&\btyo {e_2} A {\tau_2}
+	&\btyo {e_3} A {\tau_3}} \and
+\infer {\btyt {\ifthen {e_1}{e_2}{e_3}} {A} {\tau_1 \times \tau_2 \times \tau_3}} 
+	{\btyt {e_1} \rmbool {\tau_1} 
+	&\btyt {e_2} A {\tau_2}
+	&\btyt {e_3} A {\tau_3}}
 \end{mathpar}
 
+The only particular care we took here was to observe speculation in the stage \bbtwo\ case,
+which induces a product (whereas stage \bbone\ had a sum).
+We can rewrite the problematic example in our restricted language 
+by distributing the application into the if and applying the lambda in both branches:
+\[
+\tallif {E}
+	{(\next\{\pause\{5\}+1\})}
+	{(\next\{\pause\{5\}+\pause\{5*5\}\})}
+\]
+We our new rules, we can say that the boundary type of this expression is
+$\rmunit \times (\rmint + \rmint \times \rmint)$.
+Sure, this isn't the {\em exact} boundary type, 
+but it's as much as we can know without an oracle.
+
+\section{Adding Functions (But Not Ifs)}
+
+% Consider a language with products and functions.  The internal boundary kind is given by:
+
+% \begin{mathpar}
+% \infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
+% \infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
+% \infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
+% %\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
+% \end{mathpar}
+
 \end{document}

### 2014-12-19 sematics/boundary/boundary.tex
updates from meeting

@@
 	{\btyw {e} {A \times B} {\tau}} \and
-\infer {\btyo {\next~e} B \tau} 
-	{\btyt {e} {A \times B} {\tau}} \and
-\infer {\btyt {\prev~e} B \tau} 
-	{\btyo {e} {A \times B} {\tau}} \and
-\infer {\btyt {\pause~e} \rmint \tau} 
-	{\btyo e \rmint {\tau \times \rmint}}
+\infer {\btyo {\next~e} {\fut A} \tau} 
+	{\btyt {e} A {\tau}} \and
+\infer {\btyt {\prev~e} A \tau} 
+	{\btyo {e} {\fut A} {\tau}} \and
+\infer {\btyt {\pause~e} \rmint {\tau \times \rmint}} 
+	{\btyo e \rmint \tau}
 \end{mathpar}
@@
 
-\section{Adding Functions (But Not Ifs)}
+\section{Adding Just Functions}
+
+Lets now take a different stab at this, by adding functions to our straight-line code,
+but leaving \texttt{if}s out for now.
+Again, lets consider an example:
+\[
+L1 := (\lam x \rmint {\next\{5\}})
+\]
+The boundary type of $L1$ is obviously unit, right?
+After all, lambdas are values, and values have unit boundary.
+So then what about this example:
+\[
+L2 := (\lam x \rmint {\next\{\pause\{x\}+1\}})
+\]
+I'd like to think it's the same story.
+But what if we apply a value to these functions?
+Consider
+\[
+(\lam x \rmint {\next\{5\}})~3
+\]
+versus
+\[
+(\lam x \rmint {\next\{\pause\{x\}+1\}})~3
+\]
+The boundary type of the first expression should still be unit 
+(even after substitution it has no $\pause$s),
+but the bound type of the second expression should definitely be int.
+
+There's a problem here: $L1$ and $L2$ had the same boundary type before application, but different ones after.
+This suggests that if we want the boundary type to be a compositional property (rather than global),
+then we'll need to keep track of more information.
+
+We do this by making a distinction between the {\em internal} boundary of an expression,
+and its {\em external} boundary.  
+$L1$ and $L2$ have the same external boundary, namely unit,
+but they have different internal boundaries.  
+In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
+whereas the internal boundary of $L2$ should be unit.
 

### 2015-01-06 sematics/mono/mono.tex
worked on mono report

\documentclass[]{article}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{color}

\title{Boundary Types}
\author{Nico}

\newcommand {\next}{asdlfkj}
\input{../../paper/macros}

\begin{document}
\maketitle

\section {Goals}

The goal of this document is to motivate a three-world interpretation of two-staged languages.
I'll be using some labels in a way that is completely opposite from previous formulations,
so reader beware.

\section {Two Independent Languages}

As our baseline, we start with two independent languages, called L\bbone\ and L\bbtwo.
These languages can in general have any desired set of constructs.
In this document, we'll assume that both languages have products, sums, functions, and some base types.

The valid types of L\bbone\ and L\bbtwo\ are given by the judgements 
$A \istypeone$ and $A \istypetwo$, respectively.
Likewise, the typing judgements are given by
$\colone e A$ and $\coltwo e A$.
As it stands, these languages are completely independent.
That is, the rules for the $@~\bbone$ judgements and $@~\bbtwo$ judgements never depend on each other.

We call the thing after the $@$ sign a {\em world}.
So far, this has only been either \bbone\ or \bbtwo.
We can save space in our presentation by abstracting judgements over worlds,
conventionally using the metavariable $w$.
Thus, the valid type judgement becomes $A \istypewor$ 
and the typing judgement becomes $\colwor e A$.  
We can also define the rules parametrically over world, where appropriate.

\section{Bridging the Languages: A Roadmap}

Our goal in this project is to define some sort of linguistic superstructure that bridges L\bbone\ and L\bbtwo.
In particular, we'll be looking for a way to do this that admits a {\em temporal interpretation},
wherein the language L\bbone\ is thought to operate at one point in time, 
the language L\bbtwo\ is thought to operate at a later point in time, 
and information cannot pass from L\bbtwo\ to L\bbone, 
lest there be a violation of causality.
Strictly speaking, we should probably call this {\em directionality},
but we go with a temporal metaphor, because time is the most accessable example of a directional process.

We'll start by adding a new world \bbem\ (for {\em mixed}) which encapsulates and coordiantes the other two languages.
We'll then proceed by adding features, one at a time, to the \bbem-level language,
starting with structure constructs, then moving on to products, functions, and sums.

All along the way, we'll be keeping track of how well the language admits temporal interpretation.

\section{Structural Constructs}

We start with a few structural constructs to get us off the ground,
the most important being {\em encapsulation} rules, at the type and term level.
These rules allow us to encapsulate computations of world \bbone\ or \bbtwo,
and pass around their result at the mixed level.

\begin{mathpar}
\infer {\typesm e {\curr A}}{\typesone e A} \and
\infer {\typesm e {\fut A}}{\typestwo e A}
\end{mathpar}

\end{document}

### 2015-01-06 sematics/mono/mono.tex
more on a three-world system

@@
 \usepackage{graphicx}
+\usepackage{bbm}
 \usepackage[usenames,dvipsnames]{color}
 
-\title{Boundary Types}
+\title{Mixed Worlds}
 \author{Nico}
@@
 
-All along the way, we'll be keeping track of how well the language admits temporal interpretation.
+All along the way, we'll be keeping track of how well the language admits a temporal interpretation.
+In particular, we do this by defining a partial evaluation semantics and a stage-splitter [pretend I explain what these mean].
 
@@
 We start with a few structural constructs to get us off the ground,
-the most important being {\em encapsulation} rules, at the type and term level.
+the most important being {\em encapsulation} rules at the type and term level.
 These rules allow us to encapsulate computations of world \bbone\ or \bbtwo,
@@
 \begin{mathpar}
-\infer {\typesm e {\curr A}}{\typesone e A} \and
-\infer {\typesm e {\fut A}}{\typestwo e A}
+\infer {\curr A \istypem}{A \istypeone} \and
+\infer {\fut A \istypem}{A \istypetwo} \\
+\infer {\typesm {\monoTerm~e} {\curr A}}{\typesone e A} \and
+\infer {\typesm {\next~e} {\fut A}}{\typestwo e A} 
 \end{mathpar}
 
+These are essentially the only base types at \bbem.
+We'll also add a general way to form let bindings at stage two:
+
+\begin{mathpar}
+\infer {\typesm {\letin x {e_1} {e_2}} B} {\typesm {e_1} A & \typesm [\Gamma,\colm x A] {e_2} B} 
+\end{mathpar}
+
+And some simple ways to eliminate $\curr$ and $\fut$:
+\begin{mathpar}
+\infer {\typesm {\letin {\monoTerm \{x\}} {e_1} {e_2}} B} {\typesm {e_1} {\curr A} & \typesm [\Gamma,\colone x A] {e_2} B} \and
+\infer {\typesm {\letin {\next \{x\}} {e_1} {e_2}} B} {\typesm {e_1} {\fut A} & \typesm [\Gamma,\coltwo x A] {e_2} B} 
+\end{mathpar}
+
+Finally, we need a way to move base types from \bbone\ to \bbtwo:
+\begin{mathpar}
+\infer {\typesm {\pause~e} {\fut \rmint}} {\typesm e {\curr \rmint}} 
+\end{mathpar}
+
+So far we have nothing that breaks a temporal interpretation, and our splitting rules can be given by:
+\begin{mathpar}
+\infer {\splitone {\pause~e} {\fut \rmint} {((),c),(y,l).(r;y)}} {\splitonesub {} A} \and
+\infer {\splitone {\monoTerm~e} {\curr A}  {(e,()),\_.()}}		{\cdot} \and
+\infer {\splitone {\next~e} {\curr A} {((),()),\_.e}}			{\cdot} \and
+\infer {\splitone {\letin x {e_1} {e_2}} B 
+			{\left(
+				\talllet{(x,z_1)}{c_1}{
+				\talllet{(y,z_2)}{c_2}{
+				(y, (z_1, z_2))}}
+			\right),
+			(l_1,l_2).\letin{x}{r_1}{r_2}}}
+		{\splitonesub 1 A & \splitonesub 2 A} \and
+\infer {\splitone {\letin {\monoTerm \{x\}} {e_1} {e_2}} B 
+			{\left(
+				\talllet{(x,z_1)}{c_1}{
+				\talllet{(y,z_2)}{c_2}{
+				(y, (z_1, z_2))}}
+			\right),
+			(l_1,l_2).(r_1;r_2)}}
+		{\splitonesub 1 A & \splitonesub 2 A} \and
+\infer {\splitone {\letin {\next \{x\}} {e_1} {e_2}} B 
+			{\left(
+				\talllet{(y,z)}{c_2}{
+				(y, (\pit~c_1, z_2))}
+			\right),
+			(l_1,l_2).\letin{x}{r_1}{r_2}}}
+		{\splitonesub 1 A & \splitonesub 2 A}
+\end{mathpar}
+
+\section{Adding Product and Functions}
+
+Adding mixed products and functions is straight-forward.
+
+\begin{mathpar}
+\infer {A \times B \istypem}{A \istypem & B \istypem} \and
+\infer {A \to B \istypem}{A \istypem & B \istypem} \and
+\infer [\times I]	{\typesm {(e_1,e_2)}{A\times B}}				{\typesm {e_1} A & \typesm {e_2} B} 														\and
+\infer [\times E_1]	{\typesm {\pio~e} A}							{\typesm e {A\times B}} 																\and
+\infer [\to I]		{\typesm {\lam{x}{A}{e}} {A \to B}}				{A \istypem & \typesm [\Gamma,\colm x A] e B} 										\and
+\infer [\to E]		{\typesm {e_1~e_2} {B}}							{\typesm {e_1} {A \to B} & \typesm {e_2} A} 											
+\end{mathpar}
+And the splitting rules:
+\begin{mathpar}
+\infersplitone [\to I]	{\spl {\lam{x}{A}{e}}{A \to B}{(\lambda x.c,()),\_.(\lambda (x,l).r)}}					{\sub [\Gamma,\col x A] {} B} 									\and
+\infersplitone [\to E]	{\spl {e_1~e_2}{B}{\left(
+									\talllet{(y_1,z_1)}{c_1}{
+									\talllet{(y_2,z_2)}{c_2}{
+									\talllet{(y_3,z_3)}{y_1~y_2}{(y_3,(z_1,z_2,z_3))}}}\right),
+							\begin{aligned}
+									(l_1,&l_2,l_3).\\[-\baselineskip] % terrible hax
+									&(r_1~(r_2,l_3))\\
+							\end{aligned}
+							}}																									{\sub 1 {A \to B} & \sub 2 A}							\and
+\infersplitone [\times I] 	{\spl {(e_1,e_2)}{A\times B}
+									{\left(
+										\talllet{(y_1,z_1)}{c_1}{
+										\talllet{(y_2,z_2)}{c_2}{
+										((y_1, y_2),(z_1, z_2))}}
+									\right),
+									(l_1,l_2).(r_1,r_2)
+							}}																				{\sub 1 A & \sub 2 B} \and
+\infersplitone [\times E_1]	{\spl {\pio~e}{A}{\talllet{(y,z)}{c}{(\pio~y,z)},l.\pio~r}}								{\sub {} {A\times B}} \and
+\end{mathpar}
+
+\section{Adding Sums}
+
+So adding mixed sums is actually a bit more difficult.  It forces us to answer the question: when is the tag known?
+
+\section{Adding \prev}
+
+We can also add \prev, which is more expressive than the current $\fut$ elim form, but this requires changing all our rules.
+
 \end{document}

### 2015-01-21 paper/example.tex
migrate overview section's text into semantics section

@@
   be cited.
+\end{itemize}
+}
+
+\crem{
+Notes from 1/16:
+\begin{itemize}
+\item We are modeling computations where some inputs arrive at stage 1 and
+others arrive at stage 2. As a result, some subcomputations can be performed at
+stage 1, and others at stage 2.
+\item \emph{Statics}: M12 is a language with three worlds: 1, 2, and M.
+We model the computations described above as functions in world M.
+\item \emph{Dynamics}: We describe how to perform all of the stage 1
+subcomputations in an M12 program. Our semantics are mutually defined in terms
+of how to reduce terms in each of the three worlds.
 \end{itemize}
@@
 produces a pair of terms $e_1,e_2$ containing respectively all the stage \bbone\
-and stage \bbtwo\ content of $e$, such that $e_2~e_1$ is equivalent to $e$. In
-the case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
+and stage \bbtwo\ content of $e$, such that $e_2~e_1$ is equivalent to $e$.
+(Such a splitting is always possible because $e$ is well-typed in \lang.) In the
+case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
 described above.

### 2015-01-22 paper/splitting.tex
introduced a simple staged language

@@
+%!TEX root = paper.tex
+
 \section{Splitting Algorithm}
@@
 
-Given a term $e$ in the source language \lang, where terms of
-stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
-  splitting} separates the stage~\bbone\ and stage~\bbtwo\
-subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
-%each performing the
-%subcomputations corresponding to each stage.  More precisely, given a
-%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
-%unstaged target language \langmono\
-such that~$e_1$ performs all the
-stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
-stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
-the staged language produces the same result as evaluating~$e_1$ in the target language,
-then further evaluating~$e_2$ in the target language with the result from~$e_1$.
+The primary strength of \lang\ is the expressive power afforded by
+its ability to syntactically inteleave work intended for stage \bbone\
+and work intended for stage \bbtwo, as well as it ability to form
+mixed-stage abstractions.
 
-In this section, we present a
-splitting algorithm for the source language \lang, specifying
-precisely the algorithm and its correctness conditions.  
-We begin by defining a masking relation that serves to extract stage~\bbone\ and
-stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting are designed to match. We then 
-move on to the specification of the splitting algorithm itself.
+But one could imagine a simpler multistage language, \langS,
+wherein the stages are syntactically separated {\em in the source}.
+That is, there is only one multistage construct, and it has the form:
+\[
+\dothen {e_1} x {e_2}
+\]
+where $e_1$ and $e_2$ are entirely monostage terms that
+encode the stage \bbone\ work and stage \bbtwo\ work, respectively,
+of the whole program.
+The type system for this language has only a single multistage type,
+$A \& B$ which means the program produces a value of type $A$ 
+as the output of stage \bbone\ and a value of type $B$ as the
+output of stage \bbtwo.  
+There is also only one typing rule:
+\[
+\infer {\dothen {e_1} x {e_2} : A \& B} {\types[\emptyC]{e_1}{A\times \tau} & \types[x : \tau]{e_2}{B}}
+\]
+Phase \bbone\ evaluation is given by
+\[
+\infer {\dothen {e_1} x {e_2} \Downarrow_\bbone^{S} [v_1; [u/x]e_2] } {\reduce{e_1}{(v_1,u)}}
+\]
+and phase \bbtwo\ evaluation is given by
+\[
+\infer {[v_1; e_2] \Downarrow_\bbtwo^{S} [v_1; v_2] } {\reduce{e_2}{v_2}}
+\]
+
+[Need to talk about the other flavor in \langS, which produces no first-stage output.]
+
+% Given a term $e$ in the source language \lang, where terms of
+% stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
+%   splitting} separates the stage~\bbone\ and stage~\bbtwo\
+% subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
+% %each performing the
+% %subcomputations corresponding to each stage.  More precisely, given a
+% %\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
+% %unstaged target language \langmono\
+% such that~$e_1$ performs all the
+% stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
+% stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
+% the staged language produces the same result as evaluating~$e_1$ in the target language,
+% then further evaluating~$e_2$ in the target language with the result from~$e_1$.
+
+The goal of stage splitting is to translate terms of \lang\ into equivalent terms of \langS.
+[Specifically, there are versions for each flavor.]
+
+We begin by defining {\em masking} operations that relate the full and partial values of \lang\ and \langS.
+With our end goal defined, we move on to the specification of the term splitting algorithm itself.
 

### 2015-01-22 paper/example.tex
small changes to section 2

@@
 argument \texttt{k} is only known at the \emph{second}, and all
-purely-stage-\bbone\ computation occurs before any stage \bbtwo\ computation.
+first-stage computation occurs before any second-stage computation.
 
-We define a staged version of quickselect in \ref{fig:qs-staged}, writing stage
-\bbone\ code in red, and stage \bbtwo\ code in blue. Whereas \texttt{qSelect}
-sends a \textrm{list} and \rmint\ to an \rmint, \texttt{qsStaged} is more
-precise---it is a stage \bbone\ function which takes a \textrm{list} and a
-$\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
+We define a staged version of quickselect in \ref{fig:qs-staged}, writing
+first-stage computations in red, and second-stage computations in blue. While
+\texttt{qSelect} sends a \textrm{list} and \rmint\ to an \rmint,
+\texttt{qsStaged} has a more precise type---it is a first-stage function which
+takes a \textrm{list} and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and
+returns a $\fut\rmint$.
+
+\crem{Next three paragraphs are waiting on exact code for \ref{fig:qs-staged}.
+Use three colors?}
 
@@
 There are many ways to annotate any program, including \texttt{qSelect}; we
-chose annotations which maximize the work performed at stage \bbone.
+chose annotations which maximize the work performed in the first stage.
 
@@
 this fashion, might try to split it into a pair of functions, one which performs
-all the work depending only on \texttt{l} (stage \bbone), and one which uses
+all the work depending only on \texttt{l} (the first stage), and one which uses
 \texttt{k} and the result of that work on \texttt{l} to compute the element with
@@
 of its work at stage \bbtwo!
+
+\crem{Use $\bbonep$ to explain the above.}
 
@@
 (\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
-produces a pair of terms $e_1,e_2$ containing respectively all the stage \bbone\
-and stage \bbtwo\ content of $e$, such that $e_2~e_1$ is equivalent to $e$.
+produces a pair of terms $e_1,e_2$ containing respectively all the first- and
+second-stage content of $e$, such that $e_2~e_1$ is equivalent to $e$.
 (Such a splitting is always possible because $e$ is well-typed in \lang.) In the

### 2015-01-27 paper/splitting.tex
edits all over (syntax) and masking section

--- 2014-11-13 paper/splitting.tex
@@
 
-Given a term $e$ in the source language \lang, where terms of
-stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
-  splitting} separates the stage~\bbone\ and stage~\bbtwo\
-subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
-%each performing the
-%subcomputations corresponding to each stage.  More precisely, given a
-%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
-%unstaged target language \langmono\
-such that~$e_1$ performs all the
-stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
-stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
-the staged language produces the same result as evaluating~$e_1$ in the target language,
-then further evaluating~$e_2$ in the target language with the result from~$e_1$.
+\begin{concretesyntax}
 
-In this section, we present a
-splitting algorithm for the source language \lang, specifying
-precisely the algorithm and its correctness conditions.  
-We begin by defining a masking relation that serves to extract stage~\bbone\ and
-stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting are designed to match. We then 
-move on to the specification of the splitting algorithm itself.
+The primary strength of \lang\ is the expressive power afforded by
+its ability to syntactically interleave work intended for stage \bbone\
+and work intended for stage \bbtwo, as well as it ability to form
+mixed-stage abstractions.
+[Sentence about how we actual want this in separated monostage form.]
+In this section, we introduce a {\em splitting transformation} whose job 
+is to translate \lang\ terms into a form where the stages are 
+seperated {\em at the top level}.
+
+Recall that $\colmix {e_1}{A_1}$ represents multistage computations with multistage results,
+whereas $\coltwo {e_2}{A_2}$ represents multistage computions with stage \bbtwo\ results.
+Similarly, our splitting algorithm has two forms, 
+given by the judgements $\splitonesym$ and $\splittwosym$, 
+which respectively take these types of \lang\ terms as input.
+For the outputs, we introduce new constructs 
+to encode multistage values and computations that are syntactically separated at the top level.
 
@@
 
-Recall that partial values and residual tables produced by \bbone-evaluation may
-contain content at \emph{both} stages. For example the term:
+Recall that partial values (including residual tables) produced by \bbone-evaluation may
+contain content at \emph{both} stages. 
+In particular, partial values have the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.
+For example, the term
 \begin{lstlisting}
@@
 \end{lstlisting}
-To match the results of \bbone-evaluation,
-splitting the term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value, \texttt{3},
-and an~$e_2$ corresponding to the stage~\bbtwo\ computation: \\
-$\letin \yhat {\mathtt{7+5}} \yhat$.
 
-to precisely specific the stage~\bbone\ and stage~\bbtwo\ components of \bbone-evaluation results, we define 
-\emph{masking} functions, written $\masko{{-};{-}}$ and $\maskt{{-};{-}}$, 
-which take a residual table and partial value as input and emit a term in \langmono\
-that corresponds to only the content belonging to a single stage.
-\ref{fig:valMask} defines these masking functions.
+What we desire is a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
+from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
+allowing us to later define a term translation that respects this value translation.
 
-\emph{Stage~\bbone\ masking} of the partial value~$v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
-the stage~\bbone\ content of~$v$. Since \next\ terms only contain stage~\bbtwo\
-content, $\masko{v}$ replaces all $\next$ subterms of~$v$ with \texttt{()}.
-Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits the body of lambdas as a general stage~\bbone\ term (as described in \ref{sec:split-one}), and returns
-the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
-the same structure as~$v$.  For example, the stage~\bbone\ mask of the partial value given above is \texttt{(3,())}.
+For this purpose, we define a relation called {\em masking}, written $\vsplito$, 
+which sends partial values to {\em masked values}.
+Masked values have only one form, written $\mval {i} {r}$, 
+where $i$ is a monostage value (the {\em immediate value}) and $r$ is a monostage term (the {\em resumer}).
+The masking relation is defined in \ref{fig:valMask}.
 
-\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context~$\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
-exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
-requires access to the residual table $\xi$ because after \bbone-evaluation, all stage~\bbtwo\
-has been lifted out of~$v$. Stage~\bbtwo\ masking discards
-stage~\bbone\ content in~$v$ by replacing base constants with~\texttt{()},
-replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
-constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ masking, stage~\bbtwo\
-masking splits the bodies of lambdas as general stage~\bbone\ terms, and
-returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from~$v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
-producing a term in \langmono. This term reduces to a value that with the same structure as~$v$.
-For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
+Masking operates by first inducting on the entries of the residual table.  
+Being purely stage~\bbtwo\ content, these are reified into let statements at the top of the resumer.
+Once the table is empty, masking inducts on value itself.
+
+Masking assigns purely stage~\bbone\ values to the immediate value
+and likewise assign references into the residual table to the resumer.
+In both cases, the offside component is assigned to \texttt{()}, to represent trivial information.
+
+Masking distributes into tuples, injections, and rolls, since their subvalues may have content at both stages.
+However, the tags of injections and rolls are replicated only in the immediate value, 
+since they represent stage~\bbone\ information.
+
+Since lambdas may represent multi-stage computations, 
+masking splits the body of lambdas as a general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
+and packages the resulting terms as functions.
+
+The exmaple above is masked into,
 \begin{lstlisting}
-let yhat=7+5 in ((),yhat)
+mv((3,()), let yhat=7+5 in ((),yhat))
 \end{lstlisting}
@@
 
-\input{figures/splitting-term}
-\input{figures/splitting-misc}
-
 \subsection{Improved Definition}
@@
 we omit the contexts here.
+
+\end{concretesyntax}
+
+\input{figures/splitting-term}
+\input{figures/splitting-misc}

### 2015-01-27 paper/splitting.tex
some work on stage one splitting and general syntax stuff

@@
 
-\begin{concretesyntax}
+\begin{abstrsyn}
 
@@
 
-Splitting a stage~\bbone\ term $e$ in \lang\ yields a term 
-$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer term} $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
-To ensure the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$,
-we require that $e_1$ reduce to the tuple $(y,b)$,
-where $y$, called the \emph{\bbone-result} of $e$, is identical to $\masko{\xi;v}$, and that
-$\letin l b r$ reduces to the same value as $\maskt{\xi;v}$.
-(We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.) Since~$e_1$ performs both stage~\bbone\ subcomputations needed to produce the \bbone-result and subcomputations (refered to as \emph{precomputations}) that produce the boundary value, we refer to this term as the \emph{combined term}, and hereafter denote it as~$c$.
+Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
+To represent this in a separated form, we use the construct $\pipeM c l r$, 
+where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
+This ``multi-output pipeline'' can be partially evaluated with the following rule:
+\[
+\infer {\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
+\]
+We require that $c$ (the \emph{combined term}) reduce to the tuple $(i,b)$,
+where $i$ becomes the immediate result of the masked value,
+and where $b$ is plugged into the open variable of the resumer.
+We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.
 
 The splitting algorithm for stage~\bbone\ terms, as specified by the
-judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplit}, proceeds
+judgment $e \splitonesym \pipeM c l r$ in \cref{fig:termSplit}, proceeds
 recursively on the structure of~$e$.
+Our goal for the transformation is that 
+if $\diaone e {\xi,v}$ and $[\xi,v] \vsplito \mval i r$, 
+then $\splitone e A {\pipeM c l {r'}}$ and $\sepredone {\pipeM c l {r'}} {\mval i {r''}}$,
+where $r \equiv r''$ in some standard equivalence.
+Intuitively, translation and evaluation should commute.
+
 When $e$ is a terminal (a variable or base constant)
@@
 (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
-
-% Note the direct correspondence of these splitting outputs to the output of the stage~\bbone\ and~\bbtwo\ masking functions.
 
@@
 The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
-to the patterns $(y_1,b_1),\ldots,(y_n,b_n)$
+to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
 to isolate \bbone-results from boundary values. Then,
@@
 
-\subsection{Improved Definition}
-\label{sec:resumers}
-
-The definitions of masking and splitting can be augmented to better handle special cases,
-particularly ones that depend on the type of the term we're splitting.
-One such special case is the behavior of stage \bbone\ splitting at stage \bbone\ types that contain no $\fut$s or $\to$s.
-We call such types {\em monostage} (judgmentally, $A~mono$), 
-because a partial value $v$ with a monostage type can contain no $\next$s,
-and so must comprise only stage \bbone\ information.
-
-Intuitively, for $[\xi;v] : A$ and $A~mono$, 
-we would expect that $\maskt{\xi;v}$ only reduce to something that carries no information.
-This indeed happens: for example, $\maskt {\cdot;\mathtt{(inl~3,5)}}$ reduces to \texttt{((),())}.
-But while \texttt{((),())} technically ``carries no information," it still has non-trivial structure.
-This is not only unintuitive, but potentially expensive.
-Instead, we desire a stronger property, that $\maskt{\xi;v}$ only reduce to exactly \texttt{()}. 
-We can achieve this by adjusting just the tuple rule:
-\[
-\maskt{\cdot;(v_1,v_2)} = \left\{ \begin{array}{ll} 
-\mathtt{()} &\text{if } (v_1,v_2):A~mono\\ 
-(\maskt{\cdot;v_1},\maskt{\cdot;v_2}) &\text{otherwise} \end{array} \right.
-\]
-From here, the splitting rules can be made to comport with this new 
-definition adding special-case tuple, projection, and application 
-rules:
-\begin{mathpar}
-\infer
-	{(e_1,e_2):A\times B\splitonesym [\ldots,(l_1,l_2).(r_1;r_2;())]}
-	{e_1 : A \splitonesym [c_1,l_1.r_1] 
-	& e_2 : B \splitonesym [c_2,l_2.r_2] 
-	& A \times B~mono} 
-\and
-\infer
-	{\pio~e:A\splitonesym [\ldots,l.(r;())]}
-	{e : A\times B \splitonesym [c,l.r] 
-	& A \times B~mono} 
-\and
-\infer
-	{e_1~e_2: B\splitonesym [\ldots,(l_1,l_2,l_3).(r_1 (r_2,l_3); ())]}
-	{e_1 : A \to B \splitonesym [c_1,l_1.r_1] 
-	& e_2 : A \splitonesym [c_2,l_2.r_2] 
-	& B~mono} 
-\end{mathpar}
-The old tuple, projection, and application rules 
-must also be augmented to apply only when the $mono$ premises don't hold.
-Also, all splitting rules need to be augmented to carry types, though
-we omit the contexts here.
-
-\end{concretesyntax}
+\end{abstrsyn}
 

### 2015-01-30 paper/paper.tex
changes to figures
switch to icfp formatting
edits all over (syntax) and masking section
changed staged quickselect code

@@
-\documentclass[pldi,preprint]{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 
@@
 \usepackage{bbm}
-%\usepackage{mathabx}
+\usepackage{mathabx}
 \usepackage{listings}
@@
 \usepackage[T1]{fontenc}
-%\usepackage[utf8x]{inputenc}
-%\PrerenderUnicode{λ}
+\usepackage[utf8x]{inputenc}
+\PrerenderUnicode{λ}
 \usepackage{verbatim}
@@
                  {beta}{{$\beta$}}1
+                 {atsign}{{@}}1
+                 {^}{{$\curr$}}1
                  {$}{{$\fut$}}1}
@@
 \section{Intro}
+
 (placeholder in order to keep section numbers stable)

### 2015-01-30 paper/implementation.tex
edits all over (syntax) and masking section
some work on stage one splitting and general syntax stuff
various edits and renames

--- 2014-11-26 paper/implementation.tex
@@
+%!TEX root = paper.tex
+
 \section {Implementation and Optimization}
@@
 
+\begin{concretesyntax}
 We have implemented the semantics and splitting algorithm of \lang\ in Standard ML.
@@
 stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
-At the type level, this manifests as a constructor called $\monoType$.
+At the type level, this manifests as a constructor called $\curr$.
 So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
-which might fail to terminate at stage \bbtwo.  
-Alternatively $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
+which might fail to terminate at stage \bbtwo, 
+but $\curr (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
 which {\em cannot} have any content at stage \bbtwo.
-Naturally, the well-formedness of $\monoType A$ requires that $A$ contain no $\fut$s.
+Naturally, the well-formedness of $\curr A$ requires that $A$ contain no $\fut$s.
 
-At the term level, we add the $\monoTerm$, $\lifttag$, \texttt{let mono} constructs to move information
-in and out of $\monoType$.
+With these new features in place, we can now define the \texttt{double}
+function so that it splits reasonably.
+\begin{lstlisting}
+2`prev{
+  1`let mono{`3`double`1`} = mono{`3`fn n : int => n+n`1`}
+  in next{`2`hold {`1`lift(mono{`3`double 5`1`})`2`}-7`1`}`2`
+}`
+\end{lstlisting}
+This now splits into:
+\begin{lstlisting}
+1`let double = (fn n : int => n+n)
+in (double 5,())`
 
-\input{figures/mono.tex}
+2`l => in (pi1 l)-7`
+\end{lstlisting}
+\end{concretesyntax}

### 2015-02-01 paper/splitting.tex
various edits to spliting section

@@
 mixed-stage abstractions.
-[Sentence about how we actual want this in separated monostage form.]
-In this section, we introduce a {\em splitting transformation} whose job 
-is to translate \lang\ terms into a form where the stages are 
-seperated {\em at the top level}.
+To achieve this, \lang\ has three classes of multistage construct:
+\begin{itemize}
+\item terms $\colmix {e_1}{A_1}$ to represent multistage computations with multistage results,
+\item terms $\coltwo {e_2}{A_2}$ to represent multistage computions with purely stage \bbtwo\ results, and
+\item patial values $[\xi; v]$ to represent computations which have produced stage \bbone\ results,
+but for which some stage \bbtwo\ computation remains.
+\end{itemize}
 
-Recall that $\colmix {e_1}{A_1}$ represents multistage computations with multistage results,
-whereas $\coltwo {e_2}{A_2}$ represents multistage computions with stage \bbtwo\ results.
-Similarly, our splitting algorithm has two forms, 
-given by the judgements $\splitonesym$ and $\splittwosym$, 
-which respectively take these types of \lang\ terms as input.
-For the outputs, we introduce new constructs 
-to encode multistage values and computations that are syntactically separated at the top level.
+Our goal in this section will be to, for each class above, 
+show how to translate members of that class into
+an equivalent form where the stages are separated at the top level.
+
+More formally, we introduce three new constructs 
+(\texttt{mv}, \texttt{pipeM}, \texttt{pipeS}) 
+to represent the separated form of each class, 
+and three judgements ($\vsplito$,$\splitonesym$,$\splittwosym$)
+mapping interleaved forms to separated forms.
+In summary:
+
+\begin{center}
+\begin{tabular} {|c|c|c|c|}
+\hline
+Class
+	& \begin{tabular}[x]{@{}c@{}}Interleaved\\Form\end{tabular} 
+	& \begin{tabular}[x]{@{}c@{}}Separated\\Form\end{tabular} 
+	& \begin{tabular}[x]{@{}c@{}}Splitting\\Relation\end{tabular} 
+	\\ \hline
+Partial Value
+	& $[\xi; v]$
+	& $\mval i r$
+	& $\vsplito$
+	\\ \hline
+\begin{tabular}[x]{@{}c@{}}Multistage Term\\(Outputs at \bbone,\bbtwo)\end{tabular} 
+	& $\colmix e A$
+	& $\pipeM c l r$
+	& $\splitonesym$
+	\\ \hline
+\begin{tabular}[x]{@{}c@{}}Multistage Term\\(Output at \bbtwo)\end{tabular} 
+	& $\coltwo e A$
+	& $\pipeS p l r$
+	& $\splittwosym$
+	\\ \hline
+\end{tabular}
+\end{center}
 
@@
 
-Recall that partial values (including residual tables) produced by \bbone-evaluation may
-contain content at \emph{both} stages. 
-In particular, partial values have the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.
+Recall that partial values are multistage terms with the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.
+
 For example, the term
@@
 
-The exmaple above is masked into,
+The example above is masked into,
 \begin{lstlisting}
@@
 Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
-To represent this in a separated form, we use the construct $\pipeM c l r$, 
+To represent this in a separated form, we use the ``multi-output pipeline'' construct $\pipeM c l r$, 
 where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
-This ``multi-output pipeline'' can be partially evaluated with the following rule:
+This can be partially evaluated (into a masked value) with the following rule:
 \[
@@
 Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
-splitting stage~\bbtwo\ terms in \lang\ assumes a simpler form than that of stage~\bbone\ term splitting. 
-Specifically, splitting a stage~\bbtwo\ term $e$ in \lang\ generates a precomputation term $p$
-(performing all stage~\bbone\ subcomputations in $e$) and a
-resumer term $e_2=l.r$ (performing all stage~\bbtwo\ subcomputations) such that when 
-$e$ reduces to the value~$v$, $\letin l p r$ reduces to an identical value in \langmono.
+stage~\bbtwo\ term splitting assumes a simpler form than does stage~\bbone\ term splitting. 
+Specifically, splitting a term $\coltwo e A$ in \lang\ generates a ``single-output pipeline''
+term $\pipeS p l r$ where $p$ is a monostage {\em precomputation} term 
+that performs all of the stage~\bbone\ subcomputations in $e$, and $l.r$ 
+is a resumer which that performs all of the stage~\bbtwo\ subcomputations in $e$.
+This can be partially evaluated into a residual via the rule,
+\[
+\infer {\sepredone {\pipeS p l r} {[b/l]r}} {\reduce p b} 
+\]
 Note that in contrast to the combined term produced during stage~\bbone\ splitting,
@@
 binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
-\ur{This last point seems important.  It would be important to come
-  back to it.}
 

### 2015-02-02 paper/example.tex
edits all over (syntax) and masking section
small edits to the front

@@
+%!TEX root = paper.tex
+
 \section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
 \label{sec:staging}
+
+\begin{concretesyntax}
 
@@
 \texttt{qsStaged} has a more precise type---it is a first-stage function which
-takes a \textrm{list} and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and
-returns a $\fut\rmint$.
-
-\crem{Next three paragraphs are waiting on exact code for \ref{fig:qs-staged}.
-Use three colors?}
+takes a \textrm{list} (ignore the $\curr$ for now) and a 
+$\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
 
@@
 %pass around opaque stage \bbtwo\ computations at stage \bbone
+
+The quickselect example also contains some {\em purity annotations}, 
+$\curr$ and \texttt{@pure}, which indicate regions of code that are 
+contain no stage \bbtwo\ code (i.e. they are purely stage \bbone),
+which allow more efficient code generation down the line.
 
@@
 (\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
-produces a pair of terms $e_1,e_2$ containing respectively all the first- and
-second-stage content of $e$, such that $e_2~e_1$ is equivalent to $e$.
+produces an equivalent term where the stages of computation 
+are syntactically seperated at the top level.
 (Such a splitting is always possible because $e$ is well-typed in \lang.) In the
@@
 %is itself is a contribution.)
+
+\end{concretesyntax}

### 2015-02-02 paper/semantics.tex
small fixes and edits

@@
+%!TEX root = paper.tex
+
 \section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
@@
 
-The input to our splitting algorithm is given in \lang, a typed, two-stage
+The input to our splitting algorithm is given in \lang, a typed, modal
 lambda calculus with products, sums, and isorecursive types. This language is
 adapted from \cite{davies96}, and is very similar to the staged languages
-considered in the partial evaluation and metaprogramming literature. In this
-paper, we restrict our attention to a two-stage language to simplify our
-presentation of the splitting algorithm, but we expect that the techniques
+considered in the partial evaluation and metaprogramming literature,
+though it differs in three key ways:
+
+\begin{enumerate}
+
+\item We restrict our attention to a two-stage language to simplify our
+presentation of the splitting algorithm, though we expect that the techniques
 presented here extend to more stages.
 
-Our dynamic semantics of \lang\ differ from those given in \cite{davies96} and
+\item Although our language describes computations with two {\em stages}
+(\bbone\ and \bbtwo), its type system makes use of three {\em worlds}
+(\bbonem, \bbonep, \bbtwo).  For reasons discussed later, we require
+that stage \bbone\ computations be classified as being either pure
+(occuring at \bbonep) or mixed (occuring at \bbonem). 
+Only mixed code may transition to stage/world \bbtwo.
+
+\item Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
 the metaprogramming literature, because we avoid duplicating stage \bbtwo\
@@
 exposition is our main goal.
+
+\end{enumerate}
 
@@
 
-In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
-stage~\bbone\ term. To ensure that the staging features are only applied to
-valid terms, we index our judgments not only by a type, but also a stage $w$.
-
+To ensure that the staging features are only applied in a valid way, 
+we index our judgments not only by a type, but also a stage $w$.
 The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in

### 2015-02-04 paper/figures/splitting-overview.tex
major reorganization of section 2 (in progress) - moving lots of text around!
2-evaluation -> residualization
elaborated on the splitting overview figures

@@
+%!TEX root = ../paper.tex
+
 \begin{figure*}
@@
 
-\textbf{\lang\ Dynamic Semantics:}
+\textbf{\lang\ Dynamic Semantics (\ref{sec:semantics})}
 \begin{itemize}
@@
 \item
-\bbtwo-Evaluation: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo
+Residualization: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo
 
@@
 
-\textbf{\lang\ Splitting:}
-
-\hspace{2em}\bbone-Splitting Structure: $e \splitonesym [c,l.r]$, where:
-
-\hspace{4em}$c$ is the \emph{combined term}
-(representing all stage~\bbone\ subcomputations in $e$)
-in \langmono
-
-\hspace{4em}$l.r$ is the \emph{resumer}
-(representing all stage~\bbtwo\ subcomputations in $e$)
-in \langmono
-
-\hspace{4em}$c~\redsym~(y,b)$, where 
-$y$ is the \emph{\bbone-result} of $e$, and
-$b$ is the \emph{boundary value} of $e$
-
-\hspace{2em}\bbone-Splitting Correctness: If $e\mathbin{\redonesym}[\xi;v]$, then:
-
-\hspace{4em}$y$ is identical to $\masko{\xi;v}$
-
-\hspace{4em}$\maskt{\xi;v}$ and \texttt{(let l=b in r)} 
-reduce (via $\redsym$) to equivalent values.
-
-\hspace{2em}\bbtwo-Splitting Structure: $e \splittwosym [p,l.r]$, where:
-
-\hspace{4em}$p$ is the \emph{precomputation}
-(representing all stage~\bbone\ subcomputations in $e$)
-in \langmono
-
-\hspace{4em}$l.r$ is the \emph{resumer}
-(representing all stage~\bbtwo\ subcomputations in $e$)
-in \langmono
-
-\hspace{2em}\bbtwo-Splitting Correctness:
-If $e~\redtwosym~q$, and $q~\redsym~v$, then
-\texttt{(let l=p in r)}~$\redsym~v'$ and $v,v'$ are equivalent.
-% TODO: FIX THIS
-% if e reduces to q, then p reduces to u
-% and q == [u/l]r
-% that is, stage 1 reduces, even if stage 2 doesn't
-
-\caption{Summary of \lang\ evaluation and splitting.}
+\caption{Summary of \lang\ evaluation.}
 \label{fig:termSplitSummary}
 \end{figure*}
+
+\begin{figure*}
+\begin{abstrsyn}
+
+\textbf{Single-Output Pipelines:}
+
+\hspace{2em} Construction: $\pipeS p l r$, where
+
+	\hspace{4em} $p$ is the {\em precomputation}, a monostage term representing the stage \bbone\ part of the program
+
+	\hspace{4em} $l.r$ is the {\em resumer}, a monostage term representing the stage \bbtwo\ part of the program
+
+\hspace{2em} Evaluation Form: $\sepredtwo {\pipeS p l r} q$, where
+
+	\hspace{4em} $\pipeS p l r$ is the single-output pipeline to be evaluated
+
+	\hspace{4em} $q$ is the resulting residual, a monostage term representing remaining stage \bbtwo\ work
+
+\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeS p l r} {[b/l]r}$, where $b$ (called the {\em boundary value}) is generated by $\reduce p b$
+
+\textbf{Mutiple-Output Pipelines:}
+
+\hspace{2em} Construction: $\pipeM c l r$, where
+
+	\hspace{4em} $c$ is the {\em combined result}, a monostage term representing the stage \bbone\ part of the program
+
+	\hspace{4em} $l.r$ is the {\em resumer}, a monostage term representing the stage \bbtwo\ part of the program
+
+\hspace{2em} Evaluation Form: $\sepredone {\pipeM c l r} {\mval i q}$, where
+
+	\hspace{4em} $\pipeM c l r$ is the multi-output pipeline to be evaluated
+
+	\hspace{4em} $\mval i q$ is the resulting masked value (defined below)
+
+\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also the {\em boundary value}) is generated by $\reduce c (i,b)$
+
+\textbf{Masked Values:}
+
+\hspace{2em} Construction: $\mval i q$, where
+
+	\hspace{4em} $i$ is the {\em immediate result}, a monostage value representing the stage \bbone\ result
+
+	\hspace{4em} $q$ is the {\em residual}, a monostage term representing stage \bbtwo\ work left to do
+
+\end{abstrsyn}
+\caption{Overview of Output Language.}
+\label{fig:outputTermSummary}
+\end{figure*}
+
+\setlength{\unitlength}{2.8pt}
+\begin{figure*}
+\begin{abstrsyn}
+\centering
+\begin{tabular}{c|c}
+
+{\Large \bf Splitting at World \bbonem} & 
+{\Large \bf Splitting at World \bbtwo} \\ \hline
+
+\begin{minipage}[t]{0.45\textwidth}
+Splitting Terms: $(\colmix e A) \splitonesym \pipeM p l r$, where
+\par \hspace{2em} $\colmix e A$ is the \lang\ input term to be split 
+\par \hspace{2em} $\pipeM p l r$ is the resulting multi-output pipeline
+\\
+Masking Partial Values: $[\xi;v] \vsplito \mval i q$, where
+\par \hspace{2em} $[\xi;v]$ is a partial value in \lang, resulting from $\redonesym$
+\par \hspace{2em} $\mval i q$ is the resulting masked value
+\end{minipage}
+&
+
+\begin{minipage}[t]{0.45\textwidth}
+Splitting Terms: $(\coltwo e A) \splittwosym \pipeS p l r$, where 	
+\par \hspace{2em} $\coltwo e A$ is the \lang\ input term to be split
+\par \hspace{2em} $\pipeS p l r$ is the resulting single-output pipeline
+\end{minipage}
+
+\\
+
+\begin{picture} (80,54) (10,30)
+
+\thicklines
+\put(25,37){\oval(22,10)}
+\put(25,75){\oval(22,10)}
+\put(75,37){\oval(22,10)}
+\put(75,75){\oval(22,10)}
+\put(36,37){\vector(1,0){28}}
+\put(36,75){\vector(1,0){28}}
+\put(25,70){\vector(0,-1){28}}
+\put(75,70){\vector(0,-1){28}}
+
+\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colone e A$}}}
+\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$[\xi,v]$}}}
+\put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
+\put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\mval i q$}}}
+
+\put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splitonesym\right)$}}
+\put(25,32){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
+\put(76,55){\parbox[l]{20 \unitlength}{Separated-Form \\ Evaluation $\left(\sepredonesym\right)$}}
+\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation $\left(\redonesym\right)$}}
+
+\put(28,66){\parbox[t]{44 \unitlength}{ 
+	\textbf{Correctness Condition:} \\ 
+	Splitting followed by evaluation (top and right arrows)
+	should produce the same result as 
+	evaluation followed by masking (left and bottom arrows),
+	up to a term equivalence.
+}}
+\end{picture}
+
+&
+\begin{picture} (80,54) (10,30)
+
+\thicklines
+\put(25,75){\oval(22,10)}
+\put(75,37){\oval(22,10)}
+\put(75,75){\oval(22,10)}
+\put(25,37){\vector(1,0){39}}
+\put(36,75){\vector(1,0){28}}
+\put(25,70){\line(0,-1){33}}
+\put(75,70){\vector(0,-1){28}}
+
+\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\coltwo e A$}}}
+\put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeS p l r$}}}
+\put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$q$}}}
+
+\put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splittwosym\right)$}}
+\put(76,55){\parbox[l]{20 \unitlength}{Separated-Form \\ Evaluation $\left(\sepredtwosym\right)$}}
+\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation $\left(\redtwosym\right)$}}
+
+\put(28,66){\parbox[t]{44 \unitlength}{ 
+	\textbf{Correctness Condition:} \\ 
+	Splitting followed by evaluation (top and right arrows)
+	should produce the same result as 
+	residualization (left and bottom arrow),
+	up to a term equivalence.
+}}
+\end{picture}
+
+\\
+
+\end{tabular}
+
+\end{abstrsyn}
+\caption{Overview of Splitting.}
+\label{fig:outputTermSummary}
+\end{figure*}

### 2015-02-04 paper/splitting.tex
wrote a new intro to splitting section

@@
 mixed-stage abstractions.
-To achieve this, \lang\ has three classes of multistage construct:
+In this section, we provide an algorithm to translate the multistage terms and values of \lang\
+into equivalent but simpler forms where the stages are separated at the top syntactic level.
+
+Since \lang\ itself is not sufficient to express the syntactically-seperated forms 
+that we desire as the output of splitting,
+we introduce three new constructs for just that purpose:
 \begin{itemize}
-\item terms $\colmix {e_1}{A_1}$ to represent multistage computations with multistage results,
-\item terms $\coltwo {e_2}{A_2}$ to represent multistage computions with purely stage \bbtwo\ results, and
-\item patial values $[\xi; v]$ to represent computations which have produced stage \bbone\ results,
-but for which some stage \bbtwo\ computation remains.
+\item {\em Single-output pipelines}, \texttt{pipeS}, analogous to \lang\ terms at \bbtwo.
+\item {\em Multi-output pipelines}, \texttt{pipeM}, analogous to \lang\ terms at \bbonem. 
+\item {\em Masked values}, \texttt{mv}, analogous to \lang\ partial values.
 \end{itemize}
 
-Our goal in this section will be to, for each class above, 
-show how to translate members of that class into
-an equivalent form where the stages are separated at the top level.
+We also define analogs to evaluation and residualization for these new terms.
+So, whereas residualization ($\redtwosym$) sends multistage terms at world \bbtwo\ to residuals, 
+separated-form residualization ($\sepredtwosym$) sends \texttt{pipeS} terms to residuals.
+And whereas evaluation ($\redonesym$) sends multistage terms at world \bbonem\ to partial values,
+seperated-form evaluation ($\sepredonesym$) sends a \texttt{pipeM} terms to \texttt{mv} terms.
+An overview of these features is provided in \ref{fig:outputTermSummary}.
+That figure also provides the actual rules for $\sepredonesym$ and $\sepredtwosym$.
+\footnote{Note how there's one rule for each, and they're rather simple.
+Indeed, the simplicity of these reduction rules is one reason why we 
+would want to translate into this separated form, rather than use the semantics of \ref{sec:semantics}.}
 
-More formally, we introduce three new constructs 
-(\texttt{mv}, \texttt{pipeM}, \texttt{pipeS}) 
-to represent the separated form of each class, 
-and three judgements ($\vsplito$,$\splitonesym$,$\splittwosym$)
-mapping interleaved forms to separated forms.
-In summary:
+With these output terms defined, we can state the goal of our splitting algorithm:
+to translate terms in \lang\ to a syntactically-separted form, 
+such that the semantics is preserved.
 
-\begin{center}
-\begin{tabular} {|c|c|c|c|}
-\hline
-Class
-	& \begin{tabular}[x]{@{}c@{}}Interleaved\\Form\end{tabular} 
-	& \begin{tabular}[x]{@{}c@{}}Separated\\Form\end{tabular} 
-	& \begin{tabular}[x]{@{}c@{}}Splitting\\Relation\end{tabular} 
-	\\ \hline
-Partial Value
-	& $[\xi; v]$
-	& $\mval i r$
-	& $\vsplito$
-	\\ \hline
-\begin{tabular}[x]{@{}c@{}}Multistage Term\\(Outputs at \bbone,\bbtwo)\end{tabular} 
-	& $\colmix e A$
-	& $\pipeM c l r$
-	& $\splitonesym$
-	\\ \hline
-\begin{tabular}[x]{@{}c@{}}Multistage Term\\(Output at \bbtwo)\end{tabular} 
-	& $\coltwo e A$
-	& $\pipeS p l r$
-	& $\splittwosym$
-	\\ \hline
-\end{tabular}
-\end{center}
+Exactly what is meant by ``such that the semantics is preserved'' depends on whether we're translating
+terms at \bbonem\ or \bbtwo.  
+In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splitonesym$ relation,
+this constraint is straightforward: $e$ and $\pipeS p l r$ can both be reduced to residuals
+via $\redtwosym$ and $\sepredtwosym$, so we only need to require that those
+residuals be equivalent for the translation to be correct.
+In the former case, where an $\colone e A$ is translated to a $\pipeM c l r$ with the $\splittwosym$ relation, 
+things are more complicated: 
+the output of $\redonesym$ is a partial value and the output of $\sepredonesym$ is a masked value,
+so we first need a way to relate those forms if we want to talk about correctness of the term translation.
+That is the purpose of of the {\em masking} relation, written $\vsplito$.
+
+This section proceeds by first defining masking ($\vsplito$), 
+then using it to motivate the definition of \bbonem-translation ($\splitonesym$),
+and finally coming back to define \bbtwo-tranlastion ($\splittwosym$).
+A summary of these operations and the relationship between them is provided in \ref{fig:splittingSummary}.
 

### 2015-02-05 paper/splitting.tex
added to intro to splitting section

@@
 mixed-stage abstractions.
+
 In this section, we provide an algorithm to translate the multistage terms and values of \lang\
-into equivalent but simpler forms where the stages are separated at the top syntactic level.
-
-Since \lang\ itself is not sufficient to express the syntactically-seperated forms 
-that we desire as the output of splitting,
-we introduce three new constructs for just that purpose:
+into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
+of the computation are {\em syntactically seperated} from each other.
+Specifically, we desire a outputs of the form $\mathcal{C}(e_1;e_2)$ where  
+$e_1$ is a monostage term that encodes all stage \bbone\ computation, 
+and $e_2$ is a monostage term that encodes all stage \bbtwo\ computation.
+To represent outputs in this way, we introduce three new constructs:
 \begin{itemize}
-\item {\em Single-output pipelines}, \texttt{pipeS}, analogous to \lang\ terms at \bbtwo.
-\item {\em Multi-output pipelines}, \texttt{pipeM}, analogous to \lang\ terms at \bbonem. 
-\item {\em Masked values}, \texttt{mv}, analogous to \lang\ partial values.
+\item {\em Single-output pipelines}, \texttt{pipeS}, for the output of splitting \lang\ terms at world \bbtwo.
+\item {\em Multi-output pipelines}, \texttt{pipeM}, for the output of splitting \lang\ terms at world \bbonem. 
+\item {\em Masked values}, \texttt{mv}, which will serve a purpose analogous to \lang\ partial values.
 \end{itemize}
 
-We also define analogs to evaluation and residualization for these new terms.
+We also define analogs to evaluation and residualization for these new terms, 
+called $\sepredonesym$ and $\sepredtwosym$.
 So, whereas residualization ($\redtwosym$) sends multistage terms at world \bbtwo\ to residuals, 
@@
 That figure also provides the actual rules for $\sepredonesym$ and $\sepredtwosym$.
-\footnote{Note how there's one rule for each, and they're rather simple.
+Note how there's one rule for each, and they're rather simple.
 Indeed, the simplicity of these reduction rules is one reason why we 
-would want to translate into this separated form, rather than use the semantics of \ref{sec:semantics}.}
+would want to translate into this separated form, rather than use the semantics of \ref{sec:semantics}.
 

### 2015-02-05 paper/splitting.tex
few small edits

@@
 of the computation are {\em syntactically seperated} from each other.
-Specifically, we desire a outputs of the form $\mathcal{C}(e_1;e_2)$ where  
+That is, we desire a outputs of the form $\mathcal{C}(e_1;e_2)$ where  
 $e_1$ is a monostage term that encodes all stage \bbone\ computation, 
@@
 We also define analogs to evaluation and residualization for these new terms, 
-called $\sepredonesym$ and $\sepredtwosym$.
-So, whereas residualization ($\redtwosym$) sends multistage terms at world \bbtwo\ to residuals, 
-separated-form residualization ($\sepredtwosym$) sends \texttt{pipeS} terms to residuals.
-And whereas evaluation ($\redonesym$) sends multistage terms at world \bbonem\ to partial values,
-seperated-form evaluation ($\sepredonesym$) sends a \texttt{pipeM} terms to \texttt{mv} terms.
+called $\sepredonesym$ and $\sepredtwosym$, 
+which respectively send \texttt{pipeM} terms to \texttt{mv} terms
+and \texttt{pipeS} terms to residuals.
 An overview of these features is provided in \ref{fig:outputTermSummary}.
@@
 terms at \bbonem\ or \bbtwo.  
-In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splitonesym$ relation,
+In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splittwosym$ relation,
 this constraint is straightforward: $e$ and $\pipeS p l r$ can both be reduced to residuals
@@
 residuals be equivalent for the translation to be correct.
-In the former case, where an $\colone e A$ is translated to a $\pipeM c l r$ with the $\splittwosym$ relation, 
+In the former case, where an $\colone e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
 things are more complicated: 

### 2015-02-05 paper/splitting.tex
tiny edits to section 4

@@
 its ability to syntactically interleave work intended for stage \bbone\
-and work intended for stage \bbtwo, as well as it ability to form
+and work intended for stage \bbtwo, as well as its ability to form
 mixed-stage abstractions.
@@
 into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
-of the computation are {\em syntactically seperated} from each other.
-That is, we desire a outputs of the form $\mathcal{C}(e_1;e_2)$ where  
-$e_1$ is a monostage term that encodes all stage \bbone\ computation, 
-and $e_2$ is a monostage term that encodes all stage \bbtwo\ computation.
+of the computation are {\em syntactically separated} from each other.
+That is, we desire outputs of the form $\mathcal{C}(e_1;e_2)$ where  
+$e_1$ is a monostage term encoding all stage \bbone\ computation, 
+and $e_2$ is a monostage term encoding all stage \bbtwo\ computation.
 To represent outputs in this way, we introduce three new constructs:
@@
 With these output terms defined, we can state the goal of our splitting algorithm:
-to translate terms in \lang\ to a syntactically-separted form, 
+to translate terms in \lang\ to a syntactically-separated form, 
 such that the semantics is preserved.
@@
 so we first need a way to relate those forms if we want to talk about correctness of the term translation.
-That is the purpose of of the {\em masking} relation, written $\vsplito$.
+That is the purpose of the {\em masking} relation, written $\vsplito$.
 
-This section proceeds by first defining masking ($\vsplito$), 
-then using it to motivate the definition of \bbonem-translation ($\splitonesym$),
-and finally coming back to define \bbtwo-tranlastion ($\splittwosym$).
+This section proceeds by defining masking ($\vsplito$), 
+then using masking to motivate the definition of \bbonem-translation ($\splitonesym$),
+and finally coming back to define \bbtwo-translation ($\splittwosym$).
 A summary of these operations and the relationship between them is provided in \ref{fig:splittingSummary}.
@@
 
-What we desire is a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
+We want a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
 from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
@@
 Since lambdas may represent multi-stage computations, 
-masking splits the body of lambdas as a general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
+masking splits the body of lambdas as general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
 and packages the resulting terms as functions.

### 2015-02-05 paper/splitting.tex
edits to splitting section

--- 2015-02-05 paper/splitting.tex
@@
 
-Recall that partial values are multistage terms with the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.
-
-For example, the term
-\begin{lstlisting}
-1`(1+2, next{`2`7 + 5`1`})`
-\end{lstlisting}
-evaluates via $\redonesym$ to:
-\begin{lstlisting}
-2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
-\end{lstlisting}
-
-What we desire is a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
-from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
-allowing us to later define a term translation that respects this value translation.
-
-For this purpose, we define a relation called {\em masking}, written $\vsplito$, 
-which sends partial values to {\em masked values}.
-Masked values have only one form, written $\mval {i} {r}$, 
-where $i$ is a monostage value (the {\em immediate value}) and $r$ is a monostage term (the {\em resumer}).
-The masking relation is defined in \ref{fig:valMask}.
+The point of the masking operation is simply to convert a partial value into a masked value $\mval i q$
+by assigning all of the stage \bbone\ content to $i$ and all of the stage \bbtwo\ content to $q$.
+The rules of the masking relation are given in \ref{fig:valMask}.
 
@@
 
-The example above is masked into,
-\begin{lstlisting}
-mv((3,()), let yhat=7+5 in ((),yhat))
-\end{lstlisting}
-
-\subsection{Stage \bbone\ Term Splitting}
+\subsection{Term Splitting at \bbonem}
 \label{sec:split-one}
 
-Recall that a term $\colone e A$ represents a multistage computation with a multistage result.
-To represent this in a separated form, we use the ``multi-output pipeline'' construct $\pipeM c l r$, 
-where $c$ is a monostage term and $r$ is a monostage term open on the variable $l$.
-This can be partially evaluated (into a masked value) with the following rule:
-\[
-\infer {\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
-\]
-We require that $c$ (the \emph{combined term}) reduce to the tuple $(i,b)$,
-where $i$ becomes the immediate result of the masked value,
-and where $b$ is plugged into the open variable of the resumer.
-We refer to $b$ as the \emph{boundary value} of the split since it carries information between stage~\bbone\ and stage~\bbtwo\ subcomputations.
-
-The splitting algorithm for stage~\bbone\ terms, as specified by the
-judgment $e \splitonesym \pipeM c l r$ in \cref{fig:termSplit}, proceeds
-recursively on the structure of~$e$.
-Our goal for the transformation is that 
-if $\diaone e {\xi,v}$ and $[\xi,v] \vsplito \mval i r$, 
-then $\splitone e A {\pipeM c l {r'}}$ and $\sepredone {\pipeM c l {r'}} {\mval i {r''}}$,
-where $r \equiv r''$ in some standard equivalence.
-Intuitively, translation and evaluation should commute.
+We now show how to translate terms $\colone e A$ into the form $\pipeM c l r$,
+pursuant to the correctness condition given in \ref{fig:splittingSummary}.
+The algorithm is specified by the $\splitonesym$ relation (\cref{fig:termSplit}), 
+which proceeds recursively on the structure of~$e$.
 
 When $e$ is a terminal (a variable or base constant)
-splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}.
-(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
+splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, 
+and the trivial resumer \texttt{()}. 
+(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  
+For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
 
-For all non-terminals (except $\next$),
+For all non-terminals (except \texttt{next}),
 splitting descends into $e$, recursively splitting its $n$ subterms
@@
 $b_i$ is cased in the resumer to determine which branch's resumer should be executed.
-{\tt case} and $\pause$ are the only two rules where splitting adds non-trivial logic is added to the precomputation.
+{\tt case} and \texttt{hold} are the only two rules where splitting adds non-trivial logic is added to the precomputation.
 
@@
 
-Since the results of splitting $\next$ terms depend on the output of splitting its stage~\bbtwo\ subterm,
-we defer description of $\next$ until after describing stage~\bbtwo\ term splitting.
+Since the results of splitting \texttt{next} terms depend on the output of splitting its stage~\bbtwo\ subterm,
+we defer description of \texttt{next} until after describing stage~\bbtwo\ term splitting.
 
-\subsection{Stage \bbtwo\ Term Splitting}
+\subsection{Term Splitting at \bbtwo}
 
-Because stage~\bbtwo\ terms in \lang\ fully reduce to values (as opposed to partial values),
-stage~\bbtwo\ term splitting assumes a simpler form than does stage~\bbone\ term splitting. 
-Specifically, splitting a term $\coltwo e A$ in \lang\ generates a ``single-output pipeline''
-term $\pipeS p l r$ where $p$ is a monostage {\em precomputation} term 
-that performs all of the stage~\bbone\ subcomputations in $e$, and $l.r$ 
-is a resumer which that performs all of the stage~\bbtwo\ subcomputations in $e$.
-This can be partially evaluated into a residual via the rule,
-\[
-\infer {\sepredone {\pipeS p l r} {[b/l]r}} {\reduce p b} 
-\]
-Note that in contrast to the combined term produced during stage~\bbone\ splitting,
-the precomputation $p$ generated by stage~\bbtwo\ splitting exists entirely
-to compute the boundary value expected by the resumer.
+Because stage~\bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
+term splitting at \bbtwo\ assumes a simpler form than the version at \bbonem\ does. 
+The algorithm is specified by the $\splittwosym$ relation in \cref{fig:termSplit}.
 
-% \ur{I don't quite understand the following.}
-% (Thus, the resumer, bound to the value output by $p$, is equivalent to
-% the residual term defined in~\ref{ssec:dynamics})
-
-As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms,
-as specified by the judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplit}, is defined on the local structure of $e$.
 In the terminal cases of
@@
 
-For all non-terminal terms~$e$ in \lang\ (except $\prev$
-and $\pause$)
-% recursively splitting
-% subterms and generating precomputations $p_1, \ldots, p_n$ and
-% resumers $r_1, \ldots, r_n$ for a term with~$n$ immediate subterms.
-the precomputation of~$e$ is defined as the combined precomputations of $e$'s $n$ subterms:
+More generally, for all (except \texttt{prev} and \texttt{hold}) 
+$n$-ary terms $e = \mathcal{C}\ttlpar e_1 \ttsemi \ldots \ttsemi e_n \ttrpar$ 
+the precomputation is the tupled precomputations of $e$'s $n$ subterms:
 $p=(p_1,\ldots,p_n)$.  The resumer binds each boundary value to an
 argument $(l_1,\ldots,l_n)$ in a term that has the same structure
-of~$e$ but where each subterm is replaced by its corresponding resumer $r_i$.
+of~$e$ but where each subterm is replaced by its corresponding resumer:
+$r = \mathcal{C}\ttlpar r_1 \ttsemi \ldots \ttsemi r_n \ttrpar$ .
+Notably, at \texttt{case}s and functions the
+precomputation of subterms is lifted out from underneath stage \bbtwo\ binders.  
+% TODO: should probably draw a parallel to the same behavior in dynamics
 
-A notable properly of splitting stage~\bbtwo\ \texttt{case}s and functions is that the
-precomputation of subterms is lifted out from underneath stage
-\bbtwo\
-binders.  % TODO: should probably draw a parallel to the same behavior in dynamics
-
-Splitting $\prev$ generates a precomputation that projects the \bbone-result of its stage~\bbone\ subterm.
-Since the argument to $\prev$ is of $\fut$ type, its \bbone-result reduces to \texttt{()}, justifying why it can be thrown away.
-$\pause$ treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
+Splitting \texttt{prev} generates a precomputation that projects the \bbone-result of its stage~\bbone\ subterm.
+Since the argument to \texttt{prev} is of $\fut$ type, its \bbone-result reduces to \texttt{()}, justifying why it can be thrown away.
+\texttt{hold} treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
 and projects out the integer result in the resumer. 
@@
 but we have to include it here for termination purposes.} 
-Finally, splitting $\next$ simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial \bbone-result \texttt{()}.
+Finally, splitting \texttt{next} simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial \bbone-result \texttt{()}.
 

### 2015-02-05 paper/figures/splitting-overview.tex
wrote a new intro to splitting section
pared down implementation section

@@
 \caption{Summary of \lang\ evaluation.}
-\label{fig:termSplitSummary}
+\label{fig:dynamicsSummary}
 \end{figure*}
@@
 \end{abstrsyn}
-\caption{Overview of Output Language.}
+\caption{Overview of the output terms of splitting.}
 \label{fig:outputTermSummary}
@@
 	Splitting followed by evaluation (top and right arrows)
-	should produce the same result as 
+	produce the same result as 
 	evaluation followed by masking (left and bottom arrows),
@@
 	Splitting followed by evaluation (top and right arrows)
-	should produce the same result as 
+	produce the same result as 
 	residualization (left and bottom arrow),
@@
 \caption{Overview of Splitting.}
-\label{fig:outputTermSummary}
+\label{fig:splittingSummary}
 \end{figure*}

### 2015-02-05 paper/implementation.tex
pared down implementation section

@@
 \]
+Our implementation uses many optimizations like these.
 
-This sort of optimization does not affect the contract of splitting at all,
-because the contract only requires that there be a precomputation that the residual can accept,
-with no restrictions on the form of that precomputation.
-
-The problem becomes more difficult when trying to optimize across function boundaries.
-For example, consider the stage \bbtwo\ term,
-\begin{lstlisting}
-2`prev{
-  1`letfun double (n : int) = n + n
-  in next{`2`hold {`1`double 5`2`}-7`1`}`2`
-}`
-\end{lstlisting}
-Using the rules above, this term splits into:
-\begin{lstlisting}
-1`letfun double (n : int) = (n+n, ())
-in double 5`
-
-2`l => 
-letfun double (n : unit, l0 : unit) = ()
-in (double (pi2 l); pi1 l)-7`
-\end{lstlisting}
-
-Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{double}, since it always returns \texttt{()}. 
-In this case, the wasted work is small, but if \texttt{double} were a recursive function,
-then it's runtime may be unbounded.
-
-There are two ways plausible ways to solve this problem:
-either we could define an optimization step that comes 
-after splitting and show that it tends to fix the inefficiencies that splitting creates; 
-or we could give more structure to the input language so that splitting can generate more efficient code directly.
-
-We choose to investigate the latter approach.
-Specifically, we add new staging features to \lang\ which allow us to indicate that a piece of code is {\em strictly} 
-stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
-At the type level, this manifests as a constructor called $\curr$.
-So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
-which might fail to terminate at stage \bbtwo, 
-but $\curr (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
-which {\em cannot} have any content at stage \bbtwo.
-Naturally, the well-formedness of $\curr A$ requires that $A$ contain no $\fut$s.
-
-With these new features in place, we can now define the \texttt{double}
-function so that it splits reasonably.
-\begin{lstlisting}
-2`prev{
-  1`let mono{`3`double`1`} = mono{`3`fn n : int => n+n`1`}
-  in next{`2`hold {`1`lift(mono{`3`double 5`1`})`2`}-7`1`}`2`
-}`
-\end{lstlisting}
-This now splits into:
-\begin{lstlisting}
-1`let double = (fn n : int => n+n)
-in (double 5,())`
-
-2`l => in (pi1 l)-7`
-\end{lstlisting}
 \end{concretesyntax}

### 2015-02-09 paper/figures/splitting-overview.tex
implemented some kayvon comments

@@
 
-\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also the {\em boundary value}) is generated by $\reduce c (i,b)$
+\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also called the {\em boundary value}) is generated by $\reduce c (i,b)$
 

### 2015-02-10 paper/splitting.tex
fixes to figures

@@
 residuals be equivalent for the translation to be correct.
-In the former case, where an $\colone e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
+In the former case, where an $\colmix e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
 things are more complicated: 
@@
 
-We now show how to translate terms $\colone e A$ into the form $\pipeM c l r$,
+We now show how to translate terms $\colmix e A$ into the form $\pipeM c l r$,
 pursuant to the correctness condition given in \ref{fig:splittingSummary}.
@@
 
+\subsection {Three World System?}
+
+Why do we need a three world system?
+
 \end{abstrsyn}

### 2015-02-10 paper/figures/splitting-overview.tex
fixes to figures
fixed figures

--- 2015-02-05 paper/figures/splitting-overview.tex
@@
 
-\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also the {\em boundary value}) is generated by $\reduce c (i,b)$
+\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also called the {\em boundary value}) is generated by $\reduce c {\tup{i,b}}$
 
@@
 
-\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colone e A$}}}
+\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
 \put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$[\xi,v]$}}}

### 2015-02-11 paper/example.tex
fixed figures
edited section 2.

--- 2015-01-22 paper/example.tex
@@
+%!TEX root = paper.tex
+
 \section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
 \label{sec:staging}
+
+\begin{abstrsyn}
 
@@
 stage 1, and others at stage 2.
-\item \emph{Statics}: M12 is a language with three worlds: 1, 2, and M.
-We model the computations described above as functions in world M.
+\item \emph{Statics}: M12 is a language with three worlds: 1M, 1P, and 2.
+We model the computations described above as functions in world M, and their
+results as values in 2.
 \item \emph{Dynamics}: We describe how to perform all of the stage 1
@@
 The side chosen is determined by the relationship of \texttt{k} to the size of
-the first half \texttt{n}, which is returned by the \texttt{partition} function
+the first half \texttt{n}, which is returned by the \texttt{part}ition function
 along with the two sides themselves.
@@
 language (and evaluate under the \texttt{case} statement), it should be possible
-to set things up so that \emph{all} the computation involving \texttt{l} is
-performed before \emph{any} of the computation involving \texttt{k}.
+to force \emph{all} the computations involving \texttt{l} to be performed before
+\emph{any} of the computations involving \texttt{k}.
 
@@
 argument \texttt{k} is only known at the \emph{second}, and all
-first-stage computation occurs before any second-stage computation.
+first-stage computations occur before any second-stage computations.
 
@@
 first-stage computations in red, and second-stage computations in blue. While
-\texttt{qSelect} sends a \textrm{list} and \rmint\ to an \rmint,
+\texttt{qSelect} sends a \textrm{list} and $\rmint$ to an $\rmint$,
 \texttt{qsStaged} has a more precise type---it is a first-stage function which
-takes a \textrm{list} and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and
-returns a $\fut\rmint$.
+takes a \textrm{list} (ignoring the $\curr$ for the moment) and a $\fut\rmint$
+(an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
 
-\crem{Next three paragraphs are waiting on exact code for \ref{fig:qs-staged}.
-Use three colors?}
+Ignoring for the moment all \texttt{pure} annotations, the first-stage code in
+\texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
+non-circle type are available for immediate use. Indeed, we case on the list as
+usual, and in the \texttt{Cons} branch, \texttt{part}ition it.
 
-The stage \bbone\ portions of \texttt{qsStaged} look like \texttt{qSelect}: any
-term of non-circle type is available for immediate use. Indeed, we case on
-\texttt{l} as usual, and in the \texttt{Cons} branch, \texttt{partition} it.
+The \texttt{case} expression in \texttt{qSelect} depends on \texttt{k}, whose
+type $\fut\rmint$ indicates that it is only available to second-stage
+computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$, the
+remainder of the function is second-stage code.
 
-The next part of the original \texttt{qSelect} function depends on \texttt{k},
-which has type $\fut\rmint$, indicating that it is only available to stage
-\bbtwo\ computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$,
-the remainder of the function is stage \bbtwo\ code.
-
-The $\next$ wraps a stage \bbtwo\ expression of type $\rmint$ (the \texttt{case}
-expression) as a stage \bbone\ expression of type $\fut\rmint$ (the result type
+The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
+expression) as a first-stage expression of type $\fut\rmint$ (the result type
 of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
-$\prev$ unwraps stage \bbone\ $\fut\rmint$s (the results of the recursive calls
-to \texttt{qsStaged}) as stage \bbtwo\ $\rmint$s. In the \texttt{EQ} branch,
-$\pause$ promotes an integer at stage \bbone\ (the head of \texttt{l}) directly
-to an integer at stage \bbtwo.%
+$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
+to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
+$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
+second-stage $\rmint$.%
 \footnote{$\pause$ can be implemented from our other language features, with
@@
 
-%The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
-%typechecks, it is a valid way of dividing the program into two stages, where the
-%stage \bbone\ computations can all be performed prior to the stage \bbtwo\
-%computations. In \ref{sec:semantics} we provide an evaluation strategy for
-%\lang\ which validates that claim.
-
-%(We adopt the convention that the top-level context is stage \bbone.)
-
-%pass around opaque stage \bbtwo\ computations at stage \bbone
+Our type system, discussed in \ref{sec:semantics}, ensures that the staging
+annotations are consistent, in the sense that computations marked as
+first-stage cannot depend on ones marked as second-stage. The $\curr$ and
+\texttt{pure} annotations further distinguish those first-stage computations
+which do not contain any embedded second-stage code. (Code \emph{inside}
+\texttt{pure}, like the \texttt{part}ition function, is guaranteed not to
+contain second-stage code; other code, like \texttt{qsStaged}, may.) We will
+discuss the importance of this additional distinction in \ref{sec:splitting}.
 
@@
 Intuitively, \texttt{l} determines the result of all calls to
-\texttt{partition}, and \texttt{k} only determines which calls are made. So we
+\texttt{part}ition, and \texttt{k} only determines which calls are made. So we
 can preprocess \texttt{l} by recursively dividing it into halves smaller and
@@
 have \texttt{k}, we can recur on this tree, choosing whichever branch has the
-\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. Crucially,
-because \texttt{partition} has no stage \bbtwo\ content, we need not perform any
-of its work at stage \bbtwo!
-
-\crem{Use $\bbonep$ to explain the above.}
+\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. And because
+\texttt{part}ition contains no second-stage code, we can run it entirely in the
+first stage!
 
@@
 In this paper, we develop a splitting algorithm
-(\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
-produces a pair of terms $e_1,e_2$ containing respectively all the first- and
-second-stage content of $e$, such that $e_2~e_1$ is equivalent to $e$.
-(Such a splitting is always possible because $e$ is well-typed in \lang.) In the
-case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
-described above.
+(\ref{sec:splitting,sec:implementation}) which, given a program $e$ in \lang,
+produces an equivalent pair of programs which correspond precisely to the two
+stages of computation in $e$.
+(Splitting is always possible because the staging annotations in $e$ are
+consistent, because $e$ is well-typed in \lang.) In the case of
+\texttt{qsStaged}, our splitting algorithm produces the algorithm described
+above.
 
@@
 %is itself is a contribution.)
+
+\end{abstrsyn}

### 2015-02-11 paper/figures/splitting-overview.tex
Fixed up the evaluation summary figure

@@
 
-\begin{figure*}
-\textbf{Languages:}
-\begin{itemize}
-\item \lang: two-staged typed lambda calculus with product, sum, and
-  recursive types
-\item \langTwo: monostaged typed lambda calculus
-\item \langmono: an unstaged lambda calculus
-\end{itemize}
+\begin{figure}
+\begin{abstrsyn}
 
-\textbf{\lang\ Dynamic Semantics (\ref{sec:semantics})}
-\begin{itemize}
-\item 
-\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.
+\textbf{Monostage Evaluation (also used for \bbone-Evaluation at \bbonep):}
 
-\item
-Residualization: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo
+\hspace{2em} Evaluation Form: $\reduce e v$, where
 
-\item 
-Table reification: $[\xi;q]\mathbin{\reifysym} q'$, where $q,q'$ are residuals,
-$\xi$ is a residual table.
+	\hspace{4em} $e$ is the monostage term to be evaluated
 
-\item
-Residual evaluation: $q\mathbin{\redsym} v$, where $q$ is a residual, $v$ is a
-value.
-\end{itemize}
+	\hspace{4em} $v$ is the resulting value
 
+\textbf{\bbone-Evaluation at \bbonem:}
+
+\hspace{2em} Evaluation Form (Evaluation): $\diaone e {\xi;v}$, where
+
+	\hspace{4em} $e$ is the term at world \bbonem\ to be evaluated
+
+	\hspace{4em} $\xi$ is the resulting residual table
+
+	\hspace{4em} $v$ is the resulting partial value
+
+\textbf{\bbone-Evaluation at \bbtwo:}
+
+\hspace{2em} Evaluation Form (Residualization): $\diatwo e q$, where
+
+	\hspace{4em} $e$ is the term at world \bbtwo\ to be evaluated
+
+	\hspace{4em} $q$ is the resulting residual
+
+\hspace{2em} Auxiliary Judgement (Reification): $\reify \xi q q'$, where
+
+	\hspace{4em} $\xi$ is the residual table to be reified into \texttt{let} terms
+
+	\hspace{4em} $q$ is a residual to be placed at the bottom of the \texttt{let}s
+
+	\hspace{4em} $q'$ is the resulting residual
+
+\end{abstrsyn}
 \caption{Summary of \lang\ evaluation.}
 \label{fig:dynamicsSummary}
-\end{figure*}
+\end{figure}
 

### 2015-02-12 paper/semantics.tex
worked on 3.0-3.1

@@
 
-The input to our splitting algorithm is given in \lang, a typed, modal
-lambda calculus with products, sums, and isorecursive types. This language is
-adapted from \cite{davies96}, and is very similar to the staged languages
-considered in the partial evaluation and metaprogramming literature,
-though it differs in three key ways:
+We express staged programs as terms in \lang, a typed, modal lambda calculus
+with products, sums, and isorecursive types. This language is adapted from
+\cite{davies96}, and is very similar to other staged languages considered in the
+partial evaluation and metaprogramming literature, though it differs in three
+key ways:
 
@@
 
-\item We restrict our attention to a two-stage language to simplify our
-presentation of the splitting algorithm, though we expect that the techniques
+\item We restrict \lang\ to exactly two stages, to simplify our
+presentation of our stage-splitting algorithm. We expect that the techniques
 presented here extend to more stages.
 
-\item Although our language describes computations with two {\em stages}
-(\bbone\ and \bbtwo), its type system makes use of three {\em worlds}
-(\bbonem, \bbonep, \bbtwo).  For reasons discussed later, we require
-that stage \bbone\ computations be classified as being either pure
-(occuring at \bbonep) or mixed (occuring at \bbonem). 
-Only mixed code may transition to stage/world \bbtwo.
+\item Although \lang\ describes two-stage computations, its modal type system
+has three distinct \emph{worlds} \bbonem, \bbonep, and \bbtwo. Essentially, this
+further distinguishes those first-stage computations which are \emph{purely
+first-stage} (\bbonep) or contain a \emph{mix} of code from each stage
+(\bbonem). This distinction proves useful when defining our splitting algorithm.
+%Only mixed code may transition to stage/world \bbtwo.
 
@@
 
-%Apart from the non-duplication, the dynamic semantics of \lang\ is essentially
-%identical to Davies'.
+\crem{Do we actually want to make this ``more efficient'' claim?}
 
@@
 
-To ensure that the staging features are only applied in a valid way, 
-we index our judgments not only by a type, but also a stage $w$.
-The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in
-the context $\Gamma$, and is defined in \cref{fig:statics}.
+\begin{abstrsyn}
 
-\crem{Notice that certain types are impossible.}
+The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
+$e$ has type $A$ at world $w$, in the context $\Gamma$. (Variables in the
+context are also annotated with the world at which they live.)
 
-Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
-language features (i.e., the introductory and eliminatory forms for product,
-sum, and function types) preserve the stage of their subterms, and work at both
-stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
-\lang\ would simply consist of two non-interacting copies of a standard lambda
-calculus. Variables in the context are annotated with the stage at which they
-were introduced and can only be used at the same stage.
+All three worlds (\bbonem, \bbonep, and \bbtwo) contain product, sum, function,
+and recursive types defined in the usual fashion. For example, given
+two terms $\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form
+the pair $\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world
+\bbonem.
 
-Note that there is no way to eliminate a term of type $\fut A$ into a stage
-\bbone\ term---the only elimination form is $\prev$, which results in a stage
-\bbtwo\ term. This is why well-typed terms cannot have any information flow
-from stage \bbtwo\ to stage \bbone, and what makes splitting possible.
+Differing worlds (and hence, differing stages of computation) interact by means
+of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
+which classifies second-stage computations of type $A$. Given a world \bbtwo\
+term $e$ of type $A$, $\next{e}$ has type $\fut A$ at world
+\bbonem. This essentially encapsulates $e$ as a computation to be evaluated in
+the future---first-stage computations can shuffle it around, but not use its
+result. The only way to use a $\fut A$ is to wrap it with $\prev$, yielding an
+$A$ at world \bbtwo.
+
+%This is why well-typed terms cannot have any information flow from stage
+%\bbtwo\ to stage \bbone, and what makes splitting possible.
+
+$\curr A$ is a type in world \bbonem\ which classifies purely-first-stage
+computations of type $A$. Given a world \bbonep\ term $e$ of type $A$,
+$\pure{e}$ has type $\curr A$ at world \bbonem. ($e$ is guaranteed not to contain
+second-stage computations because $\fut$ types are not available in world
+\bbonep.)
+
+\crem{Need to explain why we chose letp and caseP as the abstractions.}
+
+\crem{Need to explain why \bbonem\ doesn't have ints.}
+
+\end{abstrsyn}
 

### 2015-02-13 paper/figures/splitting-overview.tex

@@
 
-\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also called the {\em boundary value}) is generated by $\reduce c {\tup{i,b}}$
+\hspace{2em} Evaluation Rule: $\sepredone {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also called the {\em boundary value}) is generated by $\reduce c {\tup{i,b}}$
 

### 2015-02-13 paper/splitting.tex

@@
 
-In this section, we provide an algorithm to translate the multistage terms and values of \lang\
+In this section, we provide a {\em splitting algorithm} to translate the multistage terms of \lang\,
 into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
 of the computation are {\em syntactically separated} from each other.
-That is, we desire outputs of the form $\mathcal{C}(e_1;e_2)$ where  
-$e_1$ is a monostage term encoding all stage \bbone\ computation, 
-and $e_2$ is a monostage term encoding all stage \bbtwo\ computation.
-To represent outputs in this way, we introduce three new constructs:
-\begin{itemize}
-\item {\em Single-output pipelines}, \texttt{pipeS}, for the output of splitting \lang\ terms at world \bbtwo.
-\item {\em Multi-output pipelines}, \texttt{pipeM}, for the output of splitting \lang\ terms at world \bbonem. 
-\item {\em Masked values}, \texttt{mv}, which will serve a purpose analogous to \lang\ partial values.
-\end{itemize}
 
-We also define analogs to evaluation and residualization for these new terms, 
-called $\sepredonesym$ and $\sepredtwosym$, 
-which respectively send \texttt{pipeM} terms to \texttt{mv} terms
-and \texttt{pipeS} terms to residuals.
-An overview of these features is provided in \ref{fig:outputTermSummary}.
-That figure also provides the actual rules for $\sepredonesym$ and $\sepredtwosym$.
-Note how there's one rule for each, and they're rather simple.
-Indeed, the simplicity of these reduction rules is one reason why we 
-would want to translate into this separated form, rather than use the semantics of \ref{sec:semantics}.
+More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
+we can translate $e$ into some syntactically separated program $\mathcal{P}$ 
+(via the new relation $e \splittwosym \mathcal{P}$), which
+can also be reduced to the residual $q$, up to an equivalence.
+We enforce the ``syntactically separated'' condition structurally by saying that 
+$\mathcal{P}$ must have the form $\pipeS p l r$ where  
+$p$ (called the {\em precomputation}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
+and $l.r$ (called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.  
+We can then residualize $\mathcal{P}$ with the $\sepredtwosym$ judgement, which is defined by the single rule
+which evaluates the precomputation and plugs it in for $l$:
+\[
+\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
+\]
+A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.
 
-With these output terms defined, we can state the goal of our splitting algorithm:
-to translate terms in \lang\ to a syntactically-separated form, 
-such that the semantics is preserved.
+Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
+we also provide a way to split those as well.
+This is given by the new relation $e \splitonesym \mathcal{P}$,
+where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
+$c$ (called the {\em combined term}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
+and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.
 
-Exactly what is meant by ``such that the semantics is preserved'' depends on whether we're translating
-terms at \bbonem\ or \bbtwo.  
-In the latter case, where an $\coltwo e A$ is translated to a $\pipeS p l r$ with the $\splittwosym$ relation,
-this constraint is straightforward: $e$ and $\pipeS p l r$ can both be reduced to residuals
-via $\redtwosym$ and $\sepredtwosym$, so we only need to require that those
-residuals be equivalent for the translation to be correct.
-In the former case, where an $\colmix e A$ is translated to a $\pipeM c l r$ with the $\splitonesym$ relation, 
-things are more complicated: 
-the output of $\redonesym$ is a partial value and the output of $\sepredonesym$ is a masked value,
-so we first need a way to relate those forms if we want to talk about correctness of the term translation.
-That is the purpose of the {\em masking} relation, written $\vsplito$.
+Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{pipeM} terms,
+with one complication arising from the fact that \lang\ terms at world \bbonem\ produces multistage output.
+For example, consider the term
+\begin{lstlisting}
+(next {1+2}, pure{3+4})
+\end{lstlisting}
+which reduces (via $\redonesym$) to the residual table and partial value
+\begin{lstlisting}
+[yhat |-> 1+2] (next {yhat}, pure{7})
+\end{lstlisting}
+Since this output has interleaved stages, so theres no way a syntactically seperated term could directly reduce to it.
+
+We resolve this mismatch by introducing one more translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
+which converts residual tables and partial values like the those above into a syntactically seperated version $\mathcal{V}$.
+As before, we enforce the ``syntactically seperated'' property structurally,
+saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
+$i$ is a monostage value respresenting the stage \bbone\ part of $v$,
+and $q$ is a monostage term representing all of the stage \bbtwo\ computation in $\xi$ and $v$.
+
+With this machinery in place, we can define $\sepredtwosym$ as just creating masked values with the single rule:
+\[
+\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
+\]
+Here we see the primary difference between terms at world \bbtwo\ and world \bbone.
+Since $\coltwo e A$ reduces to an entirely stage \bbtwo\ result, 
+its stage \bbone\ subcomputations only exist to internally pass their value to stage \bbtwo;
+but since $\colmix e A$ reduces to a multistage result, 
+its stage \bbone\ subcomputations exist {\em both} to pass their value to stage \bbtwo\ {\em and} to be present in the result.
+Correspondingly, in the $\sepredtwosym$ rule, $p$ reduces only to a boundary value which is passed to the resumer;
+but in the $\sepredonesym$ rule, $c$ reduces to a tuple containing both the immediate result and the boundary value,
+with only the latter being passed to the resumer.
+
+A summary of these operations and the relationship between them is provided on the left side of \ref{fig:splittingSummary}.
 
@@
 and finally coming back to define \bbtwo-translation ($\splittwosym$).
-A summary of these operations and the relationship between them is provided in \ref{fig:splittingSummary}.
 

### 2015-02-13 paper/splitting.tex
work on section 4.0

@@
 
-In this section, we provide a {\em splitting algorithm} to translate the multistage terms of \lang\,
-into equivalent but simpler forms where the stage \bbone\ and stage \bbtwo\ portions 
-of the computation are {\em syntactically separated} from each other.
+In this section, we provide a {\em splitting algorithm} to translate the multistage terms of \lang\
+into a simpler but equivalent form.
+The output of splitting is simpler because it is {\em syntactically separated}; that is, 
+stage \bbone\ and stage \bbtwo\ are represented by distinct subterms and are only connected at the top syntactic level.
+The output of splitting is equivalent (to its input) because it can be evaluated to the same residual 
+that would be produced by the semantics of \ref{sec:semantics}.
 
 More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
-we can translate $e$ into some syntactically separated program $\mathcal{P}$ 
+splitting translates $e$ into some syntactically separated program $\mathcal{P}$ 
 (via the new relation $e \splittwosym \mathcal{P}$), which
-can also be reduced to the residual $q$, up to an equivalence.
-We enforce the ``syntactically separated'' condition structurally by saying that 
-$\mathcal{P}$ must have the form $\pipeS p l r$ where  
+can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
+We enforce the ``syntactically separated'' condition by saying that $\mathcal{P}$ 
+can only have one form, namely $\pipeS p l r$, where  
 $p$ (called the {\em precomputation}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
 and $l.r$ (called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.  
-We can then residualize $\mathcal{P}$ with the $\sepredtwosym$ judgement, which is defined by the single rule
-which evaluates the precomputation and plugs it in for $l$:
+Likewise, $\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and plugs it in for $l$:
 \[
@@
 \]
+Here we call $b$ the {\em boundary value}, as it represents the communication at the boundary between stages.
 A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

### 2015-02-14 paper/splitting.tex
small fixes, and removed ints from theory

@@
 its stage \bbone\ subcomputations exist {\em both} to pass their value to stage \bbtwo\ {\em and} to be present in the result.
-Correspondingly, in the $\sepredtwosym$ rule, $p$ reduces only to a boundary value which is passed to the resumer;
-but in the $\sepredonesym$ rule, $c$ reduces to a tuple containing both the immediate result and the boundary value,
+Correspondingly, in the $\sepredtwosym$ rule, the precomptuation $p$ reduces only to a boundary value which is passed to the resumer;
+but in the $\sepredonesym$ rule, the combined term $c$ reduces to a tuple containing both the immediate result and the boundary value,
 with only the latter being passed to the resumer.
@@
 
-When $e$ is a terminal (a variable or base constant)
+When $e$ is a terminal (a variable or unit)
 splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, 
@@
 to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
-to isolate \bbone-results from boundary values. Then,
-the \bbone-result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
+to isolate immediate results from boundary values. Then,
+the immediate result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
 The resumer binds the boundary values $b_1,\ldots,b_n$ to an
@@
 
-Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the \bbone-result $y_1$ of the predicate.
+Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
 The boundary value $b_i$ for this branch is injected and bundled with that of the predicate ($b_1$).   
@@
 However, since the body of a function may itself be multi-stage, splitting must continue into it.
-The \bbone-result is a new function formed from the stage~\bbone\ part of the original body.
+The immediate result is a new function formed from the stage~\bbone\ part of the original body.
 The resumer is a new function formed out of the stage~\bbtwo\ part of the original body.
@@
 
-Splitting \texttt{prev} generates a precomputation that projects the \bbone-result of its stage~\bbone\ subterm.
-Since the argument to \texttt{prev} is of $\fut$ type, its \bbone-result reduces to \texttt{()}, justifying why it can be thrown away.
+Splitting \texttt{prev} generates a precomputation that projects the immediate result of its stage~\bbone\ subterm.
+Since the argument to \texttt{prev} is of $\fut$ type, its immediate result reduces to $\tup{}$, justifying why it can be thrown away.
 \texttt{hold} treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
@@
 but we have to include it here for termination purposes.} 
-Finally, splitting \texttt{next} simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial \bbone-result \texttt{()}.
+Finally, splitting \texttt{next} simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial immediate result $\tup{}$.
 

### 2015-02-14 paper/figures/splitting-overview.tex
changed name of pipes

@@
 
-\textbf{Single-Output Pipelines:}
+\textbf{Syntactically Seperated Term with Stage \bbtwo\ Output:}
 
@@
 
-	\hspace{4em} $\pipeS p l r$ is the single-output pipeline to be evaluated
+	\hspace{4em} $\pipeS p l r$ is the syntactically seperated term to be evaluated
 
@@
 
-\textbf{Mutiple-Output Pipelines:}
+\textbf{Syntactically Seperated Term with Multistage Output:}
 
@@
 
-	\hspace{4em} $\pipeM c l r$ is the multi-output pipeline to be evaluated
+	\hspace{4em} $\pipeM c l r$ is the syntactically seperated term to be evaluated
 
@@
 \par \hspace{2em} $\colmix e A$ is the \lang\ input term to be split 
-\par \hspace{2em} $\pipeM p l r$ is the resulting multi-output pipeline
+\par \hspace{2em} $\pipeM p l r$ is the syntactically seperated result
 \\
@@
 \par \hspace{2em} $\coltwo e A$ is the \lang\ input term to be split
-\par \hspace{2em} $\pipeS p l r$ is the resulting single-output pipeline
+\par \hspace{2em} $\pipeS p l r$ is the syntactically seperated result
 \end{minipage}

### 2015-02-14 paper/splitting.tex
changed name of pipes

@@
 
-Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{pipeM} terms,
+Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{sepM} terms,
 with one complication arising from the fact that \lang\ terms at world \bbonem\ produces multistage output.

### 2015-02-16 paper/splitting.tex
various edits to splitting section

@@
 \end{lstlisting}
-Since this output has interleaved stages, so theres no way a syntactically seperated term could directly reduce to it.
+Since this output has interleaved stages, theres no way a syntactically seperated term could directly reduce to it.
 
@@
 and $q$ is a monostage term representing all of the stage \bbtwo\ computation in $\xi$ and $v$.
+Note that masking exists only for the purposes of stating correctness; it is not part of the splitting algorithm itself.
 
@@
 Masking assigns purely stage~\bbone\ values to the immediate value
-and likewise assign references into the residual table to the resumer.
-In both cases, the offside component is assigned to \texttt{()}, to represent trivial information.
+and likewise assigns references into the residual table to the resumer.
+In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
 

### 2015-02-16 paper/relatedwork.tex
first pass on example.tex
minor
finished a pass through 3.0-3.1

@@
 \section{Related Work}
+
+\crem{\emph{This text came from section 3:}
+\lang\ is adapted from \cite{davies96}, and is very similar to other staged
+languages considered in the partial evaluation and metaprogramming literature.
+Unlike \cite{davies96}, we restrict \lang\ to exactly two stages, to simplify
+our presentation of our stage-splitting algorithm. We expect that the techniques
+presented here extend to more stages.
+Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
+the metaprogramming literature, because we avoid duplicating stage \bbtwo\
+computations. In addition to being more efficient, our non-duplicating semantics
+are easier to compare with our splitting algorithm, whose development and
+exposition is our main goal.
+}
+
 There are several distinctive aspect of our work which we wish to compare and contrast.
@@
 
-\subsection{Umut's notes}
-
----- DO NOT DELETE ----
-
-Answer the following question for all of the related work:
-
-What happens when we try to write quickselect as a meta program?
-
-\paragraph{Partial Evaluation.}
 Online and offline.  Both approaches have their merits but online

### 2015-02-16 paper/semantics.tex
note
finished a pass through 3.0-3.1

@@
 
-We express staged programs as terms in \lang, a typed, modal lambda calculus
-with products, sums, and isorecursive types. This language is adapted from
-\cite{davies96}, and is very similar to other staged languages considered in the
-partial evaluation and metaprogramming literature, though it differs in three
-key ways:
-
-\begin{enumerate}
-
-\item We restrict \lang\ to exactly two stages, to simplify our
-presentation of our stage-splitting algorithm. We expect that the techniques
-presented here extend to more stages.
-
-\item Although \lang\ describes two-stage computations, its modal type system
-has three distinct \emph{worlds} \bbonem, \bbonep, and \bbtwo. Essentially, this
-further distinguishes those first-stage computations which are \emph{purely
-first-stage} (\bbonep) or contain a \emph{mix} of code from each stage
-(\bbonem). This distinction proves useful when defining our splitting algorithm.
-%Only mixed code may transition to stage/world \bbtwo.
-
-\item Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
-the metaprogramming literature, because we avoid duplicating stage \bbtwo\
-computations. In addition to being more efficient, our non-duplicating semantics
-are easier to compare with our splitting algorithm, whose development and
-exposition is our main goal.
-
-\end{enumerate}
-
-\crem{Do we actually want to make this ``more efficient'' claim?}
+We express two-stage programs as terms in \lang, a typed, modal lambda calculus
+with products, sums, and isorecursive types. Its type system has three distinct
+\emph{worlds} \bbonep, \bbonem, and \bbtwo, which respectively classify
+\emph{purely first-stage} computations (\bbonep); 
+first-stage computations containing a \emph{mix} of code from each stage
+(\bbonem); 
+and second-stage computations (\bbtwo).
 
@@
 
-All three worlds (\bbonem, \bbonep, and \bbtwo) contain product, sum, function,
-and recursive types defined in the usual fashion. For example, given
-two terms $\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form
-the pair $\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world
-\bbonem.
+All three worlds contain unit, product, sum, function, and recursive types
+defined in the usual fashion. For example, given two terms $\typesone{e_1}A$ and
+$\typesone{e_2}B$ at world \bbonem, one can form the pair
+$\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world \bbonem.
 
@@
 of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
-which classifies second-stage computations of type $A$. Given a world \bbtwo\
-term $e$ of type $A$, $\next{e}$ has type $\fut A$ at world
-\bbonem. This essentially encapsulates $e$ as a computation to be evaluated in
-the future---first-stage computations can shuffle it around, but not use its
-result. The only way to use a $\fut A$ is to wrap it with $\prev$, yielding an
-$A$ at world \bbtwo.
+which classifies second-stage computations of type $A$. Given a term $e$ of type
+$A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. This essentially
+encapsulates $e$ as a computation to be evaluated in the future---first-stage
+computations can shuffle it around, but not use its result.  The only way to use
+a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.
 
@@
 %\bbtwo\ to stage \bbone, and what makes splitting possible.
+
+Although our products and functions are restricted to types at the same world,
+$\fut$ essentially allows us to build ``mixed-stage'' products and functions.
+For example, quickselect is a function at world \bbonem\ which takes a
+$\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
+second-stage computation of an integer) to a $\fut\rmint$ (a second-stage
+computation of an integer).
 
@@
 
-\crem{Need to explain why we chose letp and caseP as the abstractions.}
+An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
+using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
+$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
+p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
+projection, of type $\curr A$.
 
-\crem{Need to explain why \bbonem\ doesn't have ints.}
+Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
+$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
+open on $\curr{A}$ and $\curr{B}$ respectively.
+
+The example code in this paper makes liberal use of $\rmint$s and various
+functions on these, as well as a function $\pause$ which takes a $\curr\rmint$
+to a $\fut\rmint$.
 

### 2015-02-17 paper/example.tex
finished a pass through 3.0-3.1
really simple pure2grnd pass

@@
 }
+
+\crem{Now that hold is not officially in the language, I think we have to put a
+function prototype in the quickselect code figure. (And move the footnote about
+it to the end of 3.1.}
 
@@
 
-Ignoring for the moment all \texttt{pure} annotations, the first-stage code in
+Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
 \texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
@@
 first-stage cannot depend on ones marked as second-stage. The $\curr$ and
-\texttt{pure} annotations further distinguish those first-stage computations
+\texttt{grnd} annotations further distinguish those first-stage computations
 which do not contain any embedded second-stage code. (Code \emph{inside}
-\texttt{pure}, like the \texttt{part}ition function, is guaranteed not to
+\texttt{grnd}, like the \texttt{part}ition function, is guaranteed not to
 contain second-stage code; other code, like \texttt{qsStaged}, may.) We will

### 2015-02-17 paper/semantics.tex
really simple pure2grnd pass

@@
 \emph{worlds} \bbonep, \bbonem, and \bbtwo, which respectively classify
-\emph{purely first-stage} computations (\bbonep); 
+\emph{ground first-stage} computations (\bbonep); 
 first-stage computations containing a \emph{mix} of code from each stage

### 2015-02-18 paper/semantics.tex
finished a pass through 3.2

@@
 
-\input {figures/grammars}
+\input{figures/grammars}
 
@@
 
-\input {figures/statics}
+\input{figures/statics}
 
@@
 
-\crem{fold in the next paragraph}
-
-Given a \lang\ term, the dynamic semantics can be used
-to reduce the term to a \langmono\ program, called {\em residual} by
-using one application of stage-1 evaluation, one application of
-reification, followed by one application of stage-2 evaluation.
-
-As is the case in essentially all staged languages, the dynamic
-semantics evaluate a term to a residual (a term in the target
-language \langTwo) in two stages. In the first stage, all stage-1
-terms are evaluated by a relation, called {\em 1-evaluation}; in the
-second stage all stage-2 terms are evaluated by a relation
-$\redtwosym$, called {\em residualization}.  To eliminate duplication,
-our semantics lifts out each stage-2 term nested inside of a stage-1
-expressions, into a table which makes it possible for the stage-2
-term---more precisely the corresponding residual obtained by reducing
-all stage-1 subterms---to be referred to by use a dynamically
-generated variable (a label).
-
 \input{figures/dynamics}
 
-Since stage-1 evaluation does not evaluate stage-2 terms and since
-stage-2 terms may be nested inside stage-1 terms, 1-evaluation cannot
-reduce \lang\ terms to values.  Instead 1-evaluation yields a {\em
-  partial value}, which which is a stage~\bbone\ term that have been
-evaluated except for its stage-2 subterms, which can only
-be~\bbtwo\ variables wrapped in $\next$ blocks.  \ref{fig:grammar}
-defines partial values.
+\begin{abstrsyn}
 
-In order to avoid duplication of stage-2 computations, 1-evaluation
-labels stage-2 computations by generating variable symbols for them as
-needed and stores their residuals in a {\em residual table} keyed by
-the variable.
+We evaluate \lang\ terms in two stages. The first stage of evaluation:
+\begin{enumerate}
+\item fully reduces terms in \bbonep, as they are purely first-stage;
+\item reduces terms in \bbonem, skipping over any world \bbtwo\ terms they may
+contain (inside $\next$);
+\item does not reduce terms in \bbtwo, but scans them for any \bbonem\ terms
+they may contain (inside $\prev$) and reduces those.
+\end{enumerate}
+These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
+big-step judgments, respectively (all defined in \ref{fig:diaSemantics}).
 
-The relation 1-evaluation, defined by $\redonesym$ judgment as shown
-in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
-\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
-the type system ensures that those must occur under a $\next$. Consequently
-$\redonesym$ will never directly encounter a variable.}
-%
-to a residual table $\xi$ and a partial value $v$.  For non-staging
-features of \lang, $\redonesym$ is essentially standard call-by-value
-evaluation but in addition gathers the subterms' residual tables into
-a single table. Residual table entries are created by the
-evaluation of \next: 1-evaluation reduces a \next\ expression by
-reducing the body of the \next\ with residualization to obtain a
-residual $q$ and returns a residual table that maps a freshly
-generated variable to the residual.
+This process ultimately produces a monostaged term, called a \emph{residual},
+all of whose redexes are second-stage computations. The second stage of
+evaluation is therefore just ordinary (unstaged) evaluation of this residual,
+which we notate $\redsym$. (We assume there is a single notion of unstaged
+evaluation applicable to both \bbonep\ terms and residuals. For space reasons,
+we omit its definition.)
 
-The relation residualization, defined by $\redtwosym$ judgment, as shows
-in \ref{fig:diaSemantics}, takes a stage \bbtwo\ term and reduces it to
-a residual.  Apart from the evaluation of \prev\ and \pause, the
-evaluation is straightforward. To reduce a \prev\ term, residualization
-reduces the body by 1-evaluation to obtain a residual table and a
-partial value.  It then reifies the residual table and the partial
-value into an residual term, by creating for each variable-residual
-pair in the table a corresponding variable binding in the residual,
-finally ending with the partial value.
+Notice that our type system makes it possible to evaluate terms in a staged
+fashion, but doing so still requires some machinery, since the stages are
+syntactically interleaved within an individual term. For example,
+\texttt{qsStaged} contains a first-stage recursive call inside a second-stage
+\texttt{case}; thus, an ordinary outermost-first evaluation strategy would not
+be properly staged.
+
+\paragraph{First-stage evaluation at \bbonem\ ($\redonesym$)}
+The difficulty of evaluating terms at \bbonem\ is that they contain both terms
+at \bbonep\ (via $\curr$) and at \bbtwo\ (via $\fut$); the other evaluation
+rules are ordinary call-by-value rules. When we encounter $\pure{e}$, we
+evaluate $e$ as a monostaged term; \texttt{letg} and \texttt{caseg} simply reach
+inside $\pure$ and evaluate like \texttt{let} and \texttt{case}.
+
+$\next{e}$ is trickier, because we should not evaluate $e$ (at \bbtwo) yet.
+There are two difficulties---$e$ may contain other code at \bbonem\ (which we
+handle with $\redtwosym$, discussed below); and $\next{e}$ may be part of a
+first-stage redex, which we \emph{do} need to reduce. For example, in
+\begin{lstlisting} 
+1`(fn x : $int => e') (next{`2`e`1`})`
+\end{lstlisting}
+we must reduce this first-stage function application, but cannot reduce $e$ yet.
+We could reduce this to $[\next{e}/x]e'$, but this may potentially duplicate an
+expensive computation $e$. Instead, we choose to hoist $e$ outside, binding it
+to a temporary variable $\hat{y}$, and substituting that variable instead:
+\begin{lstlisting} 
+2`let yhat = e in `1`[next{`2`yhat`1`}/x]e'`
+\end{lstlisting}
+then proceed by evaluating $[\next{\hat{y}}/x]e'$.
+
+We achieve this by threading through a \emph{residual table} $\xi$, which
+functions as a top-level sequence of \texttt{let}-bindings of residuals $e$ to
+variables $\hat{y}$ in an evaluated \bbonem\ term. Thus, $\redonesym$ evaluation
+yields a pair $[\xi;v]$ of a residual table and a \emph{partial value} $v$
+containing variables bound in that table. The grammar of residual tables and
+partial values (terms fully evaluated at the first but not second stage) is
+given in \ref{fig:values}.
+
+%The input to $\redonesym$ may be open on stage \bbtwo\ variables, but the type
+%system ensures that those must occur under a $\next$. Consequently $\redonesym$
+%will never directly encounter a variable.
+
+\paragraph{First-stage evaluation at \bbtwo\ ($\redtwosym$)}
+Since terms at world \bbtwo\ represent second-stage computations, we should not
+evaluate them yet. However, we do need to evaluate any first-stage computations
+they contain (inside $\prev$), even if those computations are underneath
+binders.
+
+For all constructs other than $\prev{e}$, $\redtwosym$ evaluation proceeds
+recursively on all subterms, yielding a monostaged residual $q$. At $\prev{e}$,
+we evaluate $e$ using $\redonesym$ to $[\xi;v]$. Because $\colmix e{\fut A}$,
+$v$ must be $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
+$\xi$ into a sequence of second-stage \texttt{let} bindings around $\hat{y}$,
+using the auxiliary judgment $\reify\xi q{q'}$ defined in
+\ref{fig:diaSemantics}, then return the resulting residual $q'$.
+
+\end{abstrsyn}
+
+%Note that partial values and residuals are both terms for which stage \bbone\
+%computation has completed; however, partial values are stage~\bbone\ terms,
+%while residuals are stage~\bbtwo\ terms.
+
+%As an optimization, we can include the special-case rule,
+%\begin{mathpar}
+%\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
+%\end{mathpar}
+%to avoid one-for-one variable bindings in the residual.
+
+\subsection {Top-Level Evaluation}
+
+\crem{TODO below here}
 
@@
 
-%% Umut: superseded by what is above.
-%% Version Nov 7, 2014
+We can evaluate a \lang\ term as
+follows: we first apply 1-evaluation to obtain a partial value and a
+residual table, we then construct a 2-term from the partial value and
+the residual table by using reification, and then reduce this term to
+a residual via residualization.
 
-%% \subsection{Dynamics}
-%% \label{sec:stagedsemantics}
+These dynamics allow us to define a partial evaluator for \lang\ by identifying
+static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
+encode $f$ as a \lang\ expression with a function type of the form
+$A\to\fut(B\to C)$.%
 
-%% Specifically, the dynamic semantics consist of two evaluation
-%% relations $\redonesym$ and $\redtwosym$, called {\em 1-evaluation} and
-%% {\em residualization}, that respectively evaluates stage-1 and stage-2
-%% expressions. The relation 1-evaluation maps a stage-1 term to a
-%% partial value and a residual table. A {\em partial value}, can be
-%% thought as a value except that it might include references to stage-2
-%% code stored in the {\em residual table}, which maps variables to
-%% residuals.  The relation residualization takes a stage-2 term and maps to
-%% a residual.  To generate the residual table, the dynamics semantics
-%% relies on a {\em reification} relation, $\reifysym$, that maps a table
-%% to a residual, essentially generating a target expression representing
-%% the computations in the table.  We can evaluate a \lang\ term as
-%% follows: we first apply 1-evaluation to obtain a partial value and a
-%% residual table, we then construct a 2-term from the partial value and
-%% the residual table by using reification, and then reduce this term to
-%% a residual via residualization.
+Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.
 
-%% In the quickselect example, stage~\bbone\ terms are contained within
-%% stage~\bbtwo\ terms---for example, in the recursive call to {\tt qsStaged}. In
-%% this case, if we use an ordinary evaluation strategy reducing outermost redexes
-%% first, we would no longer have staged evaluation, as stage~\bbtwo\ code would
-%% be evaluated before stage~\bbone\ code.
+Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
+function:
+$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
+The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
+C)$, which by canonical forms must have the form $v = \next~\hat y$. 
+Next, we reify this environment into a sequence of \verb|let|-bindings
+enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
+Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
+and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
 
-%% Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
-%% subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
-%% a stage \bbtwo\ term with no stage~\bbone\ content, 
-%% which can be described as a term in a monostaged language called \langTwo. 
-%% Then, at stage~\bbtwo, we complete
-%% evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
-%% rules for this judgment are not shown, but they are standard.)
+That this sequence of evaluations is in fact staged follows from our
+characterizations of partial values and residuals, that $\redonesym$
+outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
 
-%% \subsubsection{Stage~\bbone\ Evaluation}
+For example, to specialize quickselect to a particular list, 
+\begin{lstlisting}
+qsStaged [5,2,7,4,1]
+\end{lstlisting}
+\TODO finish example
 
-%% To gain intuition about the challenges of implementing this staged dynamic
-%% semantics, consider the program:
-%% \begin{lstlisting}
-%% 1`#2 (next {`2`f 20`1`}, 2+3)`
-%% \end{lstlisting}
-%% This is a stage~\bbone\ expression of type $\rmint~@~\bbone$; the pair inside it is a
-%% stage~\bbone\ expression of type $(\fut\rmint)\times\rmint~@~\bbone$; and ${\tt f} :
-%% \rmint \to \rmint~@~\bbtwo$ is an expensive stage~\bbtwo\ function which we would like
-%% to avoid evaluating. A conventional call-by-value semantics would fully
-%% evaluate both components of the pair before projecting the second component.
-%% The problem is that, while \verb|next {f 20}| is not a value (in the sense that
-%% additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
-%% evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
-%% evaluation.
-
-%% Intuitively, the solution is to designate \verb|next {f 20}| as a value
-%% \emph{in stage \bbone}, even though it requires additional evaluation in stage
-%% \bbtwo. Therefore, the pair evaluates to
-%% \begin{lstlisting}
-%% 1`(next {`2`f 20`1`}, 5)`
-%% \end{lstlisting}
-%% and the projection, in turn, immediately evaluates to \verb|5|.
-
-%% Now consider a more complex example where stage~\bbone\ evaluation must
-%% substitute such an incompletely-evaluated expression. The following
-%% stage-\bbtwo\ term has type $\rmint$:
-%% \begin{lstlisting} 
-%% 2`prev{`
-%%   1`let x = (next {`2`f 20`1`}, 3+4) in
-%%   next{` 2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}` 1`}`
-%% 2`}`
-%% \end{lstlisting}
-%% As with the simpler example, this term does not fully reduce at stage~\bbone, 
-%% because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.
-
-%% Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
-%% substitute it for the three occurrences of \verb|x| in the body of the
-%% \verb|let| expression, yielding
-%% \begin{lstlisting} 
-%% 2`prev{`
-%%   1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
-%%       * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
-%%       * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
-%% }`
-%% \end{lstlisting}
-%% The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
-%% So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
-%% This search process is called {\em specialization}.
-%% In our example this means reducing all three projections to give,
-%% \begin{lstlisting} 
-%% 2`prev{`1`
-%%   next{`2` 
-%%     prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
-%%   `1`}`2`
-%% }`
-%% \end{lstlisting}
-%% Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
-%% \begin{lstlisting} 
-%% 2`(f 20) * (f 20) * 7`
-%% \end{lstlisting}
-%% But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
-%% There are some systems, typically called {\em metaprogramming}, 
-%% where this duplication would be considered intended behavior.
-%% For our applications, we find the duplication undesirable and avoid it.
-
-%% To get this behavior, we do not treat \verb|next {f 20}| as a value to be substituted, 
-%% but instead bind \verb|f 20| to a fresh stage~\bbtwo\ variable $\mathtt{\hat y}$
-%% and then substitute that variable.
-
-%% More precisely, when evaluating the \verb|next|, we construct an explicit
-%% substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
-%% fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
-%% containing $\prev$ block:
-%% \begin{lstlisting} 
-%% 2`prev {
-%% [yhat|->f 20]
-%%   `1`let x = (next{`2`yhat`1`}, 7) in
-%%   next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}`2`
-%% }`
-%% \end{lstlisting}
-%% %As a convention, we render the new variable with a %stylish and fashionable
-%% %hat.  
-%% We proceed with stage~\bbone\ evaluation, which now duplicates the variable
-%% $\mathtt{\hat y}$ rather than the expression \verb|f 20|.
-%% \begin{lstlisting} 
-%% 2`prev {
-%% [yhat|->f 20]
-%%   `1`next{`2`
-%%     prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
-%%     prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
-%%     hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
-%%   `1`}`2`
-%% }`
-%% \end{lstlisting}
-%% We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
-%% \begin{lstlisting} 
-%% 2`prev {
-%% [yhat|->f 20]
-%%     `1`next{`2`yhat * yhat * 7`1`}
-%% `2`}`
-%% \end{lstlisting}
-%% Once again, we lift the results of specialization into a substitution:
-%% \begin{lstlisting} 
-%% 2`prev {
-%% [yhat|->f 20,
-%%  zhat|->yhat * yhat * 7]
-%%     `1`next{`2`zhat`1`}`2`
-%% }`
-%% \end{lstlisting}
-%% Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
-%% \begin{lstlisting} 
-%% 2`let yhat = f 20 in
-%% let zhat = yhat * yhat * 7 in zhat`
-%% \end{lstlisting}
-
-%% This concludes stage \bbone\ evaluation of the program---we have reduced all
-%% stage \bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
-%% evaluation, $\tworedsym$, then reduces this to a numeral.
-
-%% \subsubsection{Evaluation and Specialization}
-%% \label{ssec:dynamics}
-
-%% \input{figures/dynamics}
-
-%% The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
-%% described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
-%% (specialization), and the auxiliary $\reifysym$ (reification). 
-%% Along with these are three judgments identifying forms of values:
-%% $\pvalsym$ (partial values), $\ressym$ (residuals), 
-%% and the auxiliary $\fcon$ (residual table).
-%% These judgments are related as follows:
-%% \begin{itemize}
-%% \item Evaluation sends a stage~\bbone\ term to a partial value.
-%% In order to avoid duplicating subterms, evaluation collects explicit
-%% substitutions into a single {\em residual table} that is also the output of
-%% evaluation.
-%% \item Specialization sends a stage~\bbtwo\ term to a residual.
-%% \item Reification sends a single residual and a residual table to another residual.  
-%% \item Evaluation depends on specialization for $\next$ terms, which have
-%% stage~\bbtwo\ subterms.
-%% \item Specialization depends on evaluation for $\prev$ and $\pause$, which have
-%% stage~\bbone\ subterms. The residual table produced by evaluation is also
-%% reified here.
-%% \end{itemize}
-
-%% \emph{Partial values} are stage~\bbone\ terms that have been fully
-%% evaluated, but which may contain stage~\bbtwo\ variables wrapped in $\next$
-%% blocks.  In the example above, 
-%% \begin{lstlisting} 
-%% 1`(next {`2`yhat`1`},7)`
-%% \end{lstlisting}
-%% is a partial value.
-%% \emph{Residuals} ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
-%% stage \bbone\ subexpressions have all been fully evaluated. In the example
-%% above,
-%% \begin{lstlisting} 
-%% 2`let zhat = yhat*yhat*7 in zhat`
-%% \end{lstlisting}
-%% is a residual.
-
-%% Note that partial values and residuals are both terms for which stage \bbone\
-%% computation has completed; however, partial values are stage~\bbone\ terms,
-%% while residuals are stage~\bbtwo\ terms.
-
-%% The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
-%% \bbone\ term%
-%% \footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
-%% the type system ensures that those must occur under a $\next$. Consequently
-%% $\redonesym$ will never directly encounter a variable.}
-%% %
-%% to a {\em residual table} $\xi$ and a partial value $v$. The residual
-%% table implements the explicit substitutions in our example---it maps fresh
-%% stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
-%% residuals. For non-staging features of \lang, $\redonesym$ is essentially
-%% standard call-by-value evaluation, and gathers the subterms' residual tables
-%% into a single one.
-
-%% At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
-%% subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
-%% which takes a stage \bbtwo\ term to a residual. 
-%% For the most part (\ref{fig:diaSemanticsSpec}), specialization simply recurs into
-%% all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
-%% which produces a residual table $\xi$ and (by canonical forms) an expression
-%% $\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
-%% bindings enclosing $\mathtt{\hat y}$.
-%% Once specialization reduces $e$ to a residual $q$, the output of $\redonesym$ is a
-%% fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a residual
-%% table which maps that variable to the residual ($\hat y \mapsto q$).
-
-%% Within specialization lies a subtle---if perhaps unintuitive---feature.  
-%% Observe that specialization will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
-%% statement in its search for stage~\bbone\ code. 
-%% Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
-%% and so a term like 
-%% \begin{lstlisting} 
-%% 1`next{
-%%   `2`if true 
-%%   then hold{`1`3+4`2`} 
-%%   else prev{`1`spin() (* loops forever *)`2`}`1`
-%% }`
-%% \end{lstlisting}
-%% will fail to evaluate at stage \bbone.
-%% Although it may seem undesirable, this behavior is a deliberate feature.
-%% Indeed it is critical to the quickselect example, 
-%% wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
-%% As always, it is the programmer's responsibility to ensure that her staging annotations 
-%% do not cause a program to loop forever.
-
-%% The context ($\Gamma$) keeps track of stage \bbtwo\ variables in the input term. 
-%% These both appear in the original program at stage \bbtwo\ and are inserted by the semantics.
-
-%% As an optimization, we can include the special-case rule,
-%% \begin{mathpar}
-%% \inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
-%% \end{mathpar}
-%% to avoid one-for-one variable bindings in the residual.
-
-%% \subsubsection {Top-Level Evaluation}
-%% \label{sec:partialeval}
-
-%% % KAYVONF: good statement, but hold out for now
-%% %The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must %evaluate it many times.
-
-%% These dynamics allow us to define a partial evaluator for \lang\ by identifying
-%% static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
-%% encode $f$ as a \lang\ expression with a function type of the form
-%% $A\to\fut(B\to C)$.%
-%% %\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
-%% %the following higher-order function which makes the adjustment:
-%% %\begin{lstlisting} 
-%% %let adjust (f : $int * int -> $int) =
-%% %  fn (p : int) => 
-%% %    next{
-%% %      fn (b : int) => 
-%% %        prev{f (next {b}, p)}
-%% %    }
-%% %\end{lstlisting}}
-%% %
-%% Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.
-
-%% Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
-%% function:
-%% $\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
-%% The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
-%% C)$, which by canonical forms must have the form $v = \next~\hat y$. 
-%% Next, we reify this environment into a sequence of \verb|let|-bindings
-%% enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
-%% Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
-%% and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
-
-%% That this sequence of evaluations is in fact staged follows from our
-%% characterizations of partial values and residuals, that $\redonesym$
-%% outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
-
-%% For example, to specialize quickselect to a particular list, 
-%% \begin{lstlisting}
-%% qsStaged [5,2,7,4,1]
-%% \end{lstlisting}
-%% \TODO finish example
+%We can rewrite \texttt{fexp} in this form, or simply apply
+%the following higher-order function which makes the adjustment:
+%\begin{lstlisting} 
+%let adjust (f : $int * int -> $int) =
+%  fn (p : int) => 
+%    next{
+%      fn (b : int) => 
+%        prev{f (next {b}, p)}
+%    }
+%\end{lstlisting}

### 2015-02-18 paper/figures/splitting-overview.tex
really simple pure2grnd pass
cleaning up figure captions

--- 2015-02-13 paper/figures/splitting-overview.tex
@@
 
-\textbf{Single-Output Pipelines:}
+\textbf{Syntactically Separated Term with Stage \bbtwo\ Output:}
 
@@
 
-	\hspace{4em} $\pipeS p l r$ is the single-output pipeline to be evaluated
+	\hspace{4em} $\pipeS p l r$ is the syntactically separated term to be evaluated
 
@@
 
-\textbf{Mutiple-Output Pipelines:}
+\textbf{Syntactically Separated Term with Multistage Output:}
 
@@
 
-	\hspace{4em} $\pipeM c l r$ is the multi-output pipeline to be evaluated
+	\hspace{4em} $\pipeM c l r$ is the syntactically separated term to be evaluated
 
@@
 \par \hspace{2em} $\colmix e A$ is the \lang\ input term to be split 
-\par \hspace{2em} $\pipeM p l r$ is the resulting multi-output pipeline
+\par \hspace{2em} $\pipeM p l r$ is the syntactically separated result
 \\
@@
 \par \hspace{2em} $\coltwo e A$ is the \lang\ input term to be split
-\par \hspace{2em} $\pipeS p l r$ is the resulting single-output pipeline
+\par \hspace{2em} $\pipeS p l r$ is the syntactically separated result
 \end{minipage}
@@
 \put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
-\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$[\xi,v]$}}}
+\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$[\xi;v]$}}}
 \put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
@@
 \end{abstrsyn}
-\caption{Overview of Splitting.}
+\caption{Overview of splitting.}
 \label{fig:splittingSummary}

### 2015-02-18 paper/figures/splitting-overview.tex
almost done messing with figures...

@@
 %!TEX root = ../paper.tex
-
-\begin{figure}
-\begin{abstrsyn}
-
-\textbf{Monostage Evaluation (also used for \bbone-Evaluation at \bbonep):}
-
-\hspace{2em} Evaluation Form: $\reduce e v$, where
-
-	\hspace{4em} $e$ is the monostage term to be evaluated
-
-	\hspace{4em} $v$ is the resulting value
-
-\textbf{\bbone-Evaluation at \bbonem:}
-
-\hspace{2em} Evaluation Form (Evaluation): $\diaone e {\xi;v}$, where
-
-	\hspace{4em} $e$ is the term at world \bbonem\ to be evaluated
-
-	\hspace{4em} $\xi$ is the resulting residual table
-
-	\hspace{4em} $v$ is the resulting partial value
-
-\textbf{\bbone-Evaluation at \bbtwo:}
-
-\hspace{2em} Evaluation Form (Residualization): $\diatwo e q$, where
-
-	\hspace{4em} $e$ is the term at world \bbtwo\ to be evaluated
-
-	\hspace{4em} $q$ is the resulting residual
-
-\hspace{2em} Auxiliary Judgement (Reification): $\reify \xi q q'$, where
-
-	\hspace{4em} $\xi$ is the residual table to be reified into \texttt{let} terms
-
-	\hspace{4em} $q$ is a residual to be placed at the bottom of the \texttt{let}s
-
-	\hspace{4em} $q'$ is the resulting residual
-
-\end{abstrsyn}
-\caption{Summary of \lang\ evaluation.}
-\label{fig:dynamicsSummary}
-\end{figure}
 

### 2015-02-18 paper/semantics.tex
one figure per file
almost done messing with figures...
finished a pass through 3.3

@@
 
-\input{figures/splitting-overview}
+\input{figures/values}
 
@@
 
+\input{figures/dynamics-overview}
+
 \input{figures/dynamics}
@@
 \begin{lstlisting} 
-2`let yhat = e in `1`[next{`2`yhat`1`}/x]e'`
+2`let val yhat = e in `1`[next{`2`yhat`1`}/x]e'`
 \end{lstlisting}
@@
 
-\subsection {Top-Level Evaluation}
+\subsection{Evaluating Staged Programs}
 
-\crem{TODO below here}
+We consider a staged program to be a closed term $e$ at world \bbtwo. The first
+stage of evaluation is $e\mathbin{\redtwosym} q$, which reduces all code at
+\bbonem\ and \bbonep\ inside $e$. The second stage of evaluation is $q
+\mathbin{\redsym} u$, or monostaged evaluation of the residual $q$, which yields
+a value $u$.
 
-We can use the evaluation relations in the dynamic semantics to
-evaluate a term to evaluate a \lang\ term as follows. We first apply
-1-evaluation to obtain a partial value and a residual table. We then
-construct a 2-term from the partial value and the residual table by
-using reification, and then reduce this term to a residual via
-residualization.  Finally, we evaluate the residual, which is a term in a
-our monostaged target language \langmono, to a value by using  a
-standard evaluation relation for \langmono.
+A multistage function, such as \texttt{qsStaged} from \ref{sec:staging}, has
+type $A\to \fut (B \to C)$ at \bbonem.%
+\footnote{We can rewrite \texttt{qsStaged} in this curried form, or apply a
+higher-order currying function.}
+To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
+evaluate the program:
+\begin{lstlisting}
+2`prev{`1`f a`2`} b`
+\end{lstlisting}
 
-We can evaluate a \lang\ term as
-follows: we first apply 1-evaluation to obtain a partial value and a
-residual table, we then construct a 2-term from the partial value and
-the residual table by using reification, and then reduce this term to
-a residual via residualization.
-
-These dynamics allow us to define a partial evaluator for \lang\ by identifying
-static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
-encode $f$ as a \lang\ expression with a function type of the form
-$A\to\fut(B\to C)$.%
-
-Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.
-
-Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
-function:
-$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
-The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
-C)$, which by canonical forms must have the form $v = \next~\hat y$. 
-Next, we reify this environment into a sequence of \verb|let|-bindings
-enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
-Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
-and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.
-
-That this sequence of evaluations is in fact staged follows from our
-characterizations of partial values and residuals, that $\redonesym$
-outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.
-
-For example, to specialize quickselect to a particular list, 
+We can also directly observe the reuse of first-stage computations across
+multiple second-stage computations. To run many order statistics queries
+$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
 \begin{lstlisting}
-qsStaged [5,2,7,4,1]
+2`let val qs = prev{`1`qsStaged `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
+2`in (qs k1,...,qs km)`
 \end{lstlisting}
-\TODO finish example
-
-%We can rewrite \texttt{fexp} in this form, or simply apply
-%the following higher-order function which makes the adjustment:
-%\begin{lstlisting} 
-%let adjust (f : $int * int -> $int) =
-%  fn (p : int) => 
-%    next{
-%      fn (b : int) => 
-%        prev{f (next {b}, p)}
-%    }
-%\end{lstlisting}
+To see that the partial application of \texttt{qsStaged} is only evaluated once,
+observe that after the first stage of evaluation, the residual has the form:
+\begin{lstlisting}
+2`let val yhat = ...`
+2`    val qs = yhat`
+2`in (qs k1,...,qs km)`
+\end{lstlisting}
+where $\hat{y}$ is bound to a function containing no first-stage redexes.

### 2015-02-18 paper/splitting.tex
really simple pure2grnd pass
finished a pass through 3.3
small fixes to splitting section

@@
 Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{sepM} terms,
-with one complication arising from the fact that \lang\ terms at world \bbonem\ produces multistage output.
+with one complication arising from the fact that \lang\ terms at world \bbonem\ produce multistage output.
 For example, consider the term
 \begin{lstlisting}
-(next {1+2}, pure{3+4})
+(next {1+2}, grnd{3+4})
 \end{lstlisting}
@@
 \begin{lstlisting}
-[yhat |-> 1+2] (next {yhat}, pure{7})
+[yhat |-> 1+2] (next {yhat}, grnd{7})
 \end{lstlisting}
@@
 We resolve this mismatch by introducing one more translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
-which converts residual tables and partial values like the those above into a syntactically seperated version $\mathcal{V}$.
+which is just a technical device to convert residual tables and partial values---like the those above---into a syntactically seperated version $\mathcal{V}$.
 As before, we enforce the ``syntactically seperated'' property structurally,
@@
 
-With this machinery in place, we can define $\sepredtwosym$ as just creating masked values with the single rule:
+With this machinery in place, we can define $\sepredonesym$ as just creating masked values with the single rule:
 \[
@@
 
-Masking assigns purely stage~\bbone\ values to the immediate value
+Masking assigns ground stage~\bbone\ values to the immediate value
 and likewise assigns references into the residual table to the resumer.
@@
 
-When $e$ is a terminal (a variable or unit)
-splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, 
-and the trivial resumer \texttt{()}. 
-(Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ subcomputations.)  
-For example, the integer constant \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.
+When $e$ is unit or a variable,
+splitting yields a combined term formed by tupling $e$ with a $\tup{}$ precomputation.
+These expressions, as they contain no stage~\bbtwo\ subcomputations, have a trivial resumer, $\tup{}$.
+Similarly, the contents of a \texttt{grnd} block are assigned entirely to the immediate value,
+with trivial precomputation and resumer.
 
@@
 $b_i$ is cased in the resumer to determine which branch's resumer should be executed.
-{\tt case} and \texttt{hold} are the only two rules where splitting adds non-trivial logic is added to the precomputation.
+{\tt case} is the only rule where splitting adds non-trivial logic is added to the precomputation.
 
-Function introduction has a \texttt{()} boundary value,
+Function introduction has a $\tup{}$ boundary value,
 since functions are already fully reduced in our semantics.
@@
 
-More generally, for all (except \texttt{prev} and \texttt{hold}) 
+More generally, for all (except \texttt{prev}) 
 $n$-ary terms $e = \mathcal{C}\ttlpar e_1 \ttsemi \ldots \ttsemi e_n \ttrpar$ 
@@
 Since the argument to \texttt{prev} is of $\fut$ type, its immediate result reduces to $\tup{}$, justifying why it can be thrown away.
-\texttt{hold} treats the entire combined term of its stage~\bbone\ subexpression as a precomputation, 
-and projects out the integer result in the resumer. 
-\footnote{The resumer of an integer expression is usually trivial, 
-but we have to include it here for termination purposes.} 
 Finally, splitting \texttt{next} simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial immediate result $\tup{}$.

### 2015-02-18 paper/implementation.tex
fixed figures
various edits to splitting section
a little work on examples and implementation

@@
 
-\section {Implementation and Optimization}
+\section {Implementation}
 \label{sec:implementation}
 
-\begin{concretesyntax}
-We have implemented the semantics and splitting algorithm of \lang\ in Standard ML.
-In addition to the features shown here, our implementation includes more base types, 
-$n$-ary products and sums, if and let statements, and a basic elaboration step.
+\begin{abstrsyn}
+We have implemented the semantics and splitting algorithm of \lang\ in Standard ML,
+as well as a semantics for the output language.
+In addition to the features shown here, our implementation includes base types, 
+$n$-ary products and sums, patterns, if and let statements, and a basic elaboration step.
+We cover some of the details here.
 
+\paragraph {Splitting Optimizations} 
 Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
@@
 \[
-\infer {\splittwo {(e_1,e_2)} A {p,l.(r_1,r_2)}}{\splittwo {e_1} A {(),\_.r_1} & \splittwo {e_2} A {p,l.r_2}}
+\infer {\splittwo {\tup{e_1,e_2}} A p l {\tup{r_1,r_2}}} {\splittwo {e_1} A {\tup{}} {\_}{r_1} & \splittwo {e_2} A  p l {r_2}}
 \]
@@
 
-\end{concretesyntax}
+\paragraph {Patterns}
+
+Our implemention includes patterns to eliminate products, $\fut$, and $\curr$.
+We split patterns directly (in a way similar to the standard elim forms), 
+rather than translate them away before splitting.
+This, along with the optimized splitting algorithm,
+goes a long way toward producing readable outputs.
+
+\paragraph {Syntax}
+
+Our concrete syntax is used in examples throughout the paper.
+Its main notable property is that world boundaries are always noted by braces.
+
+Functions and datatypes declarations are elaborated in the standard way.
+We also implement declaration-level staging features.  For example,
+\begin{lstlisting}
+1`atsigngrnd{
+  datatype list = Empty | Cons of int * list
+  fun part (...) = ...
+}
+fun qsStaged (...) = ...`
+\end{lstlisting}
+indicates that the \texttt{list} datatype and \texttt{part} function
+are declared and defined at world \bbonep, though this is accomplished by 
+translating the above into,
+\begin{lstlisting}
+1`val grnd{Empty} = grnd{roll (inj ...)}
+val grnd{Cons}  = grnd{fn (...) => roll (inj ...)}
+val grnd{part}  = grnd{fn (...) => ...}
+val qsStaged    = fn (...) => ...`
+\end{lstlisting}
+
+\end{abstrsyn}

### 2015-02-19 paper/splitting.tex
simple pass through section 4 to remove "stage n" terminology.

@@
 
+\input{figures/splitting-overview}
+
 \begin{abstrsyn}
@@
 The primary strength of \lang\ is the expressive power afforded by
-its ability to syntactically interleave work intended for stage \bbone\
-and work intended for stage \bbtwo, as well as its ability to form
-mixed-stage abstractions.
+its ability to syntactically interleave first- and second-stage computations,
+as well as its ability to form mixed-stage abstractions.
 
@@
 The output of splitting is simpler because it is {\em syntactically separated}; that is, 
-stage \bbone\ and stage \bbtwo\ are represented by distinct subterms and are only connected at the top syntactic level.
+the first and second stages are represented by distinct subterms and are only connected at the top syntactic level.
 The output of splitting is equivalent (to its input) because it can be evaluated to the same residual 
@@
 can only have one form, namely $\pipeS p l r$, where  
-$p$ (called the {\em precomputation}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
-and $l.r$ (called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.  
+$p$ (called the {\em precomputation}) is a monostage term encoding all the
+first-stage computation that was in $e$, 
+and $l.r$ (called the {\em resumer}) is a monostage term encoding all the
+second-stage computation that was in $e$.  
 Likewise, $\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and plugs it in for $l$:
@@
 where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
-$c$ (called the {\em combined term}) is a monostage term encoding all the stage \bbone\ computation that was in $e$, 
-and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the stage \bbtwo\ computation that was in $e$.
+$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computation that was in $e$, 
+and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
+second-stage computation that was in $e$.
 
@@
 \end{lstlisting}
-Since this output has interleaved stages, theres no way a syntactically seperated term could directly reduce to it.
+Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
 
 We resolve this mismatch by introducing one more translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
-which is just a technical device to convert residual tables and partial values---like the those above---into a syntactically seperated version $\mathcal{V}$.
-As before, we enforce the ``syntactically seperated'' property structurally,
+which is just a technical device to convert residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
+As before, we enforce the ``syntactically separated'' property structurally,
 saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
-$i$ is a monostage value respresenting the stage \bbone\ part of $v$,
-and $q$ is a monostage term representing all of the stage \bbtwo\ computation in $\xi$ and $v$.
+$i$ is a monostage value representing the first-stage part of $v$,
+and $q$ is a monostage term representing all of the second-stage computation in $\xi$ and $v$.
 Note that masking exists only for the purposes of stating correctness; it is not part of the splitting algorithm itself.
@@
 \]
-Here we see the primary difference between terms at world \bbtwo\ and world \bbone.
-Since $\coltwo e A$ reduces to an entirely stage \bbtwo\ result, 
-its stage \bbone\ subcomputations only exist to internally pass their value to stage \bbtwo;
+Here we see the primary difference between world \bbtwo\ terms and world \bbonem\ terms.
+Since $\coltwo e A$ reduces to an entirely second-stage result, 
+its first-stage subcomputations only exist to internally pass their value to the
+second stage;
 but since $\colmix e A$ reduces to a multistage result, 
-its stage \bbone\ subcomputations exist {\em both} to pass their value to stage \bbtwo\ {\em and} to be present in the result.
+its first-stage subcomputations exist {\em both} to pass their value to the
+second stage {\em and} to be present in the result.
 Correspondingly, in the $\sepredtwosym$ rule, the precomptuation $p$ reduces only to a boundary value which is passed to the resumer;
@@
 The point of the masking operation is simply to convert a partial value into a masked value $\mval i q$
-by assigning all of the stage \bbone\ content to $i$ and all of the stage \bbtwo\ content to $q$.
+by assigning all of the first-stage content to $i$ and all of the second-stage content to $q$.
 The rules of the masking relation are given in \ref{fig:valMask}.
@@
 Masking operates by first inducting on the entries of the residual table.  
-Being purely stage~\bbtwo\ content, these are reified into let statements at the top of the resumer.
+Being purely second-stage content, these are reified into let statements at the top of the resumer.
 Once the table is empty, masking inducts on value itself.
 
-Masking assigns ground stage~\bbone\ values to the immediate value
+Masking assigns values in \bbonep\ to the immediate value
 and likewise assigns references into the residual table to the resumer.
@@
 However, the tags of injections and rolls are replicated only in the immediate value, 
-since they represent stage~\bbone\ information.
+since they represent first-stage information.
 
 Since lambdas may represent multi-stage computations, 
-masking splits the body of lambdas as general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
+masking splits the body of lambdas as general world \bbonem\ terms (as described in \ref{sec:split-one}), 
 and packages the resulting terms as functions.
@@
 splitting yields a combined term formed by tupling $e$ with a $\tup{}$ precomputation.
-These expressions, as they contain no stage~\bbtwo\ subcomputations, have a trivial resumer, $\tup{}$.
+These expressions, as they contain no second-stage subcomputations, have a trivial resumer, $\tup{}$.
 Similarly, the contents of a \texttt{grnd} block are assigned entirely to the immediate value,
@@
 However, since the body of a function may itself be multi-stage, splitting must continue into it.
-The immediate result is a new function formed from the stage~\bbone\ part of the original body.
-The resumer is a new function formed out of the stage~\bbtwo\ part of the original body.
+The immediate result is a new function formed from the first-stage part of the original body.
+The resumer is a new function formed out of the second-stage part of the original body.
 It is the responsibility of the application site to save the precomputation of the function body
@@
 
-Since the results of splitting \texttt{next} terms depend on the output of splitting its stage~\bbtwo\ subterm,
-we defer description of \texttt{next} until after describing stage~\bbtwo\ term splitting.
+Since the results of splitting \texttt{next} terms depend on the output of
+splitting its world \bbtwo\ subterm,
+we defer description of \texttt{next} until after describing world \bbtwo\ term splitting.
 
@@
 
-Because stage~\bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
-term splitting at \bbtwo\ assumes a simpler form than the version at \bbonem\ does. 
+Because world \bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
+term splitting at world \bbtwo\ assumes a simpler form than the version at \bbonem\ does. 
 The algorithm is specified by the $\splittwosym$ relation in \cref{fig:termSplit}.
@@
 Notably, at \texttt{case}s and functions the
-precomputation of subterms is lifted out from underneath stage \bbtwo\ binders.  
+precomputation of subterms is lifted out from underneath world \bbtwo\ binders.  
 % TODO: should probably draw a parallel to the same behavior in dynamics
 
-Splitting \texttt{prev} generates a precomputation that projects the immediate result of its stage~\bbone\ subterm.
+Splitting \texttt{prev} generates a precomputation that projects the immediate
+result of its world \bbone\ subterm.
 Since the argument to \texttt{prev} is of $\fut$ type, its immediate result reduces to $\tup{}$, justifying why it can be thrown away.
-Finally, splitting \texttt{next} simply tuples up the precomputation of its stage~\bbtwo\ subterm with a trivial immediate result $\tup{}$.
+Finally, splitting \texttt{next} simply tuples up the precomputation of its
+world \bbtwo\ subterm with a trivial immediate result $\tup{}$.
 

### 2015-02-20 paper/semantics.tex
section 3 - reword the remark about unstaged evaluation

@@
 These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
-big-step judgments, respectively (all defined in \ref{fig:diaSemantics}).
+big-step judgments, respectively (defined in \ref{fig:diaSemantics}).
 
-This process ultimately produces a monostaged term, called a \emph{residual},
-all of whose redexes are second-stage computations. The second stage of
-evaluation is therefore just ordinary (unstaged) evaluation of this residual,
-which we notate $\redsym$. (We assume there is a single notion of unstaged
-evaluation applicable to both \bbonep\ terms and residuals. For space reasons,
-we omit its definition.)
+This process ultimately produces a monostaged term, called a \emph{residual}
+(\ref{fig:values}), all of whose redexes are second-stage computations. The
+second stage of evaluation is therefore just ordinary (unstaged) evaluation of
+this residual, which we notate $\redsym$. (For simplicity, we treat residuals
+and \bbonep\ terms as living in a single language equipped with a standard
+evaluation $\redsym$, whose definition we omit for space reasons.)
 

### 2015-02-20 paper/example.tex

@@
 }
-
-\crem{
-Notes from 1/16:
-\begin{itemize}
-\item We are modeling computations where some inputs arrive at stage 1 and
-others arrive at stage 2. As a result, some subcomputations can be performed at
-stage 1, and others at stage 2.
-\item \emph{Statics}: M12 is a language with three worlds: 1M, 1P, and 2.
-We model the computations described above as functions in world M, and their
-results as values in 2.
-\item \emph{Dynamics}: We describe how to perform all of the stage 1
-subcomputations in an M12 program. Our semantics are mutually defined in terms
-of how to reduce terms in each of the three worlds.
-\end{itemize}
-}
-
-\crem{Now that hold is not officially in the language, I think we have to put a
-function prototype in the quickselect code figure. (And move the footnote about
-it to the end of 3.1.}
 
@@
 The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
-expression) as a first-stage expression of type $\fut\rmint$ (the result type
+expression) as a first-stage expression of type $\fut\rmint$ (the result 
 of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
@@
 $\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
-second-stage $\rmint$.%
-\footnote{$\pause$ can be implemented from our other language features, with
-some effort. We provide it as a primitive to shorten examples. Furthermore, it
-would be wise to extend $\pause$ to other base types, if we had them, and to
-products and sums thereof.  This is related to the notion of {\em mobility} in
-\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}
+second-stage $\rmint$.
 
-The process of adding staging annotations ($\fut$ types, $\next$, $\prev$, and
-$\pause$) to unstaged code has been the subject of extensive research under the
+The process of adding staging annotations ($\fut$ types, $\next$, and $\prev$)
+to unstaged code has been the subject of extensive research under the
 name of \emph{binding time analysis}. In this paper, we assume that these

### 2015-02-20 paper/semantics.tex

@@
 functions on these, as well as a function $\pause$ which takes a $\curr\rmint$
-to a $\fut\rmint$.
+to a $\fut\rmint$.%
+\footnote{$\pause$ is definable in \lang\ given an inductive definition of
+$\rmint$s. In practice, we provide both $\rmint$s and $\pause$ as primitives.
+It is sensible to extend $\pause$ to all base types and to products and sums
+thereof. This is related to the notion of {\em mobility} in \cite{murphy05} and
+{\em stability} in \cite{krishnaswami13}.}
 

### 2015-02-20 paper/splitting.tex
mention fixpoints; change first paragraph of section 4

@@
 
-The primary strength of \lang\ is the expressive power afforded by
-its ability to syntactically interleave first- and second-stage computations,
-as well as its ability to form mixed-stage abstractions.
+Although the type system of \lang\ ensures that we can perform all first-stage
+computations before any second-stage ones, actually evaluating terms in this
+fashion is somewhat intricate, because stages are syntactically interleaved
+within terms. In this section, we present a \emph{splitting algorithm} which
+statically converts a term in \lang\ into a pair of 
+syntactically % I don't like this word but temporarily here for continuity
+separate programs corresponding to the two stages of computation in that term.
 
-In this section, we provide a {\em splitting algorithm} to translate the multistage terms of \lang\
-into a simpler but equivalent form.
-The output of splitting is simpler because it is {\em syntactically separated}; that is, 
-the first and second stages are represented by distinct subterms and are only connected at the top syntactic level.
-The output of splitting is equivalent (to its input) because it can be evaluated to the same residual 
-that would be produced by the semantics of \ref{sec:semantics}.
+%The output of splitting is simpler because it is {\em syntactically separated};
+%The output of splitting is equivalent (to its input) because it can be
+%evaluated to the same residual that would be produced by the semantics of
+%\ref{sec:semantics}.
 
 More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
-splitting translates $e$ into some syntactically separated program $\mathcal{P}$ 
+$e$ splits into a syntactically separated program $\mathcal{P}$ 
 (via the new relation $e \splittwosym \mathcal{P}$), which

### 2015-02-20 paper/implementation.tex

@@
 \begin{abstrsyn}
-We have implemented the semantics and splitting algorithm of \lang\ in Standard ML,
-as well as a semantics for the output language.
+We have implemented the semantics and splitting algorithm of \lang\ in Standard ML.
+Specifically, our implemention includes parsers, printers, and interpretters for both \lang\ 
+and the output language, as well as a splitting algorithm which mediates between the two.
 In addition to the features shown here, our implementation includes base types, 
@@
 
-\paragraph {Splitting Optimizations} 
-Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
-In many cases, the results can be improved by considering special cases in individual rules.
-For instance, trivial subterm precomputations needn't be saved at tuples:
-\[
-\infer {\splittwo {\tup{e_1,e_2}} A p l {\tup{r_1,r_2}}} {\splittwo {e_1} A {\tup{}} {\_}{r_1} & \splittwo {e_2} A  p l {r_2}}
-\]
-Our implementation uses many optimizations like these.
-
-\paragraph {Patterns}
-
-Our implemention includes patterns to eliminate products, $\fut$, and $\curr$.
-We split patterns directly (in a way similar to the standard elim forms), 
-rather than translate them away before splitting.
-This, along with the optimized splitting algorithm,
-goes a long way toward producing readable outputs.
-
 \paragraph {Syntax}
-
 Our concrete syntax is used in examples throughout the paper.
@@
 
+\paragraph {Patterns}
+Our implemention includes patterns in the input language to eliminate products, $\fut$, and $\curr$.
+As implied by our splitting rules, we also include product patterns in the output language.
+For the sake of readability, we split input patterns directly (in a way similar to the standard elim forms), 
+rather than translate them away before splitting.
+
+\paragraph {Specialized Splitting} 
+Our implemented splitting algorithm also includes many special cases to create more readable outputs.
+Most important among these are cases for eliminating trivial precomputations.
+These special cases all fit within the presented framework of our correctness condition.
+
 \end{abstrsyn}

### 2015-02-21 paper/implementation.tex
rework section 5

--- 2015-02-05 paper/implementation.tex
@@
 
-\section {Implementation and Optimization}
+\section {Implementation}
 \label{sec:implementation}
 
-\begin{concretesyntax}
-We have implemented the semantics and splitting algorithm of \lang\ in Standard ML.
-In addition to the features shown here, our implementation includes more base types, 
-$n$-ary products and sums, if and let statements, and a basic elaboration step.
+\begin{abstrsyn}
+We have a Standard ML implementation of the \lang\ language, a staged
+interpreter, the splitting algorithm, and an interpreter for split programs.
+This implementation extends the language described in \ref{sec:semantics} with 
+\rmint{}s,
+\rmbool{}s,
+\texttt{let} statements,
+$n$-ary sums and products,
+datatype and function declarations, and
+deep pattern matching (including $\next{}$ and $\pure{}$ patterns).
+The code snippets in this paper are written in our concrete syntax, using 
+these additional features when convenient.
 
-Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
-In many cases, the results can be improved by considering special cases in individual rules.
-For instance, trivial subterm precomputations needn't be saved at tuples:
-\[
-\infer {\splittwo {(e_1,e_2)} A {p,l.(r_1,r_2)}}{\splittwo {e_1} A {(),\_.r_1} & \splittwo {e_2} A {p,l.r_2}}
-\]
-Our implementation uses many optimizations like these.
+We allow staging annotations around declarations. For example,
+\begin{lstlisting}
+1`atsigngrnd{
+  datatype list = Empty | Cons of int * list
+  fun part (...) = ...
+}
+fun qsStaged (...) = ...`
+\end{lstlisting}
+declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
+by elaborating into:
+\begin{lstlisting}
+1`val grnd{Empty} = grnd{roll (inj ...)}
+val grnd{Cons}  = grnd{fn (...) => roll (inj ...)}
+val grnd{part}  = grnd{fn (...) => ...}
+val qsStaged    = fn (...) => ...`
+\end{lstlisting}
 
-\end{concretesyntax}
+We also have a number of optimizations which drastically improve the readability
+of split programs. We split patterns directly, rather than translating them into
+lower-level constructs. Finally, we perform many local simplifications during
+the splitting algorithm, most notably, not pairing precomputations when one is
+known to be $\tup{}$.
+
+%Specifically, our implemention includes parsers, printers, and interpreters
+
+%Its main notable property is that world boundaries are always noted by braces.
+
+%We also emit tuple patterns liberally in split programs.
+
+\end{abstrsyn}

### 2015-02-21 paper/intro.tex
partway through editing section 2
culling some old intro text

--- 2014-10-06 paper/intro.tex
@@
 
-It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).
-This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).
-Examples include:
+\crem{Notes from 2/21:
+\begin{enumerate}
+\item see K/R intro.
+\item function specialization is an important optimization technique which can
+be performed whether or not the function's arguments are known.
+\item data specialization statically specializes a multi-argument program
+\emph{before} having any of its arguments, unlike partial evaluation.
+\item we are doing data specialization, but on lambda calculus (cf C-like
+languages, as in graphics papers).
+\item this additional expressive power gives us some interesting
+results---asymptotic speedup on some algorithms (and generation of recursive
+types from recursive functions), and cross-cutting (?) modularity (see graphics
+applications).
+\end{enumerate}
+}
+
+\subsection{Old intro}
+
+partial evaluation requires compile-time access to first stage, whereas we don't
+
+Prior work on metaprogramming allows us to evaluate such staged code
+in a staged fashion by first providing the first-stage arguments and
+generating a piece of code that is specialized for the arguments.  For
+example, we can evaluate the staged quickselect by providing a
+list~\texttt{l} to obtain a piece of code that then can be used with
+different ranks (\texttt{k} values) in the second stage.
+Such a specialized code essentially performs all the first-stage computations
+first and uses their results to improve the efficiency in the second stage,
+where we issue many queries.
+
+For example, in our quickselect example, it is unlikely to know the
+list~\texttt{l} that we will be operation on at compile time.
+
+\cite{knoblock96} proposed the {\em data specialization}
+technique to overcome some of these limitations of program
+specialization.  
+Unlike program specialization,
+data specialization does not require knowledge of the first-stage
+arguments to produce fast code for the second stage.  
+Instead it yield
+a first stage program that maps the first stage arguments to an
+intermediate or {\em boundary data structure}, which is used by the
+second stage program along with the second stage arguments to produce
+the final results.
+
+But their work applies to program fragments that don't involve sophisticated
+data types such as as sums and recursive types, and don't involve recursion.
+
+\subsection{Oldest intro}
+
+This most often arises when
+we want to evaluate a multivariate function several times, with some inputs
+dynamic (varying across invocations) and some static (constant across
+invocations).
 \begin{itemize}
 \item Computing $b^e$, where $b$ varies and $e$ is constant.
-\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
-\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
+\item Finding the $i$th largest element in a list, where the list is constant
+and $i$ varies.
+\item Computing the shading of a surface in a 3D rendering, where some
+contributors to the final pixel color can be approximated as constant over
+pixel blocks.  
 \end{itemize}
-
-With these goals in mind, our initial concern is how to express multistage programs.
-To this end, we have created \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types.
-An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.
-
-Once we have appropriately written {\em multi-stage} programs, we then must execute them in a way that respects the staging.
-In this paper, we explore two techniques for this: {\em partial evaluation} and {\em stage-splitting}.
-
-The main idea of the former technique is to evaluate all of the constant portions of the program and embed the partial results into a {\em residual} program that depends only on the varying inputs.
-This is an old and well-explored idea, though the typical setup of partial evaluation implies a more powerful binding time analysis than is necessary for our language.
-
-The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.
-This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang.
-Prior examples include the ``data specialization" work of Knoblock and Ruf\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.
-Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.
-
-\TODO motivate this example
-
-\TODO integrate the following text into the new intro
+Prior examples
+include the ``data specialization" work of Knoblock and Ruf~\cite{knoblock96}
+as well as the compilation of high-level shading languages for computer graphics
+such as RTSL~\cite{Proudfoot:2001} and Spark~\cite{Foley:2011}.
 
@@
 inputs are provided, partial evaluation of $f$ produces a residual depending on
-only the dynamic input. Mathematically, a partial evaluator is any $p$ such
-that
+only the dynamic input. Mathematically, a partial evaluator is any $p$ such that
 \[
@@
 
-\ur{  Based on this definition, it
-  might be good and correct to argue that this is more or less an
-  orthogonal problem.}
-
-Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
-However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
-before {\em any arguments} are available.
-Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
-where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
-and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
-More precisely, a stage splitter is any $s$ such that
+Taking a multivariate function $f$ as input, stage-splitting produces two
+functions $f_1$ and $f_2$, where $f_1$ uses the stage-\bbone\ input to produce a
+data structure, and $f_2$ consumes that data structure and the stage-\bbtwo\
+input to produce the final output.  More precisely, a stage splitter is any $s$
+such that
 \[
@@
 with the same $\llbracket \cdot \rrbracket$ notation as before.
-Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
-meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
+The stage splitting operation does not depend on the stage-\bbone\ input.
 This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.
 
-\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging.
-\Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
-Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.
-Notably, these examples make use of the stage-splitting features not available in prior work.
+\Cref{sec:semantics} describes the grammar and type system of \lang, as well as
+a reference semantics which is essentially a partial evaluator that respects
+staging. \Cref{sec:splitting} describes a stage-splitting transformation for
+\lang\ and proves its equivalence to the reference semantics.  Finally,
+\Cref{sec:examples} gives several examples of how stage-splitting terms in
+\lang\ can be used to derive classic data structures, and discusses the
+effectiveness and limitations of this technique.  Notably, these examples make
+use of the stage-splitting features not available in prior work.

### 2015-02-22 paper/intro.tex

@@
 
-\crem{Notes from 2/21:
-\begin{enumerate}
-\item see K/R intro.
-\item function specialization is an important optimization technique which can
-be performed whether or not the function's arguments are known.
-\item data specialization statically specializes a multi-argument program
-\emph{before} having any of its arguments, unlike partial evaluation.
-\item we are doing data specialization, but on lambda calculus (cf C-like
-languages, as in graphics papers).
-\item this additional expressive power gives us some interesting
-results---asymptotic speedup on some algorithms (and generation of recursive
-types from recursive functions), and cross-cutting (?) modularity (see graphics
-applications).
-\end{enumerate}
-}
+Multi-argument functions can frequently perform useful work before receiving all
+of their inputs, or are often called numerous times with one argument fixed. An
+important program optimization is therefore to \emph{specialize} such a function
+$f$ to its fixed argument $a$, by executing those computations in $f$ which
+depend only on $a$. This ensures that calls to the specialized function require
+only computations which depend on its varying argument.
 
-\subsection{Old intro}
+\emph{Program specialization}, or partial evaluation, is a well-known
+specialization technique which, given $a$, transforms $f$ into a new function
+$f_a(-)$ which computes $f(a,-)$. This transformation essentially substitutes
+$a$ for the first argument of $f$, then evaluates in place any subexpressions of
+$f$ depending only on that argument.
 
-partial evaluation requires compile-time access to first stage, whereas we don't
+\emph{Data specialization} \cite{knoblock96} is a different technique which
+specializes $f$ \emph{without} the fixed argument $a$. Instead, it transforms
+$f$ into a pair of functions $f_1$ and $f_2$. $f_1(a)$ produces a data structure
+populated with the results of the subexpressions depending only on $a$. $f_2$
+takes this data structure and the varying argument, and computes the result;
+that is, $f_2(f_1(a),-)$ is the same as $f(a,-)$. Importantly, unlike in partial
+evaluation, neither $f_1$ nor $f_2$ depend on $a$!
 
-Prior work on metaprogramming allows us to evaluate such staged code
-in a staged fashion by first providing the first-stage arguments and
-generating a piece of code that is specialized for the arguments.  For
-example, we can evaluate the staged quickselect by providing a
-list~\texttt{l} to obtain a piece of code that then can be used with
-different ranks (\texttt{k} values) in the second stage.
-Such a specialized code essentially performs all the first-stage computations
-first and uses their results to improve the efficiency in the second stage,
-where we issue many queries.
+In this paper, we extend the technique of data specialization to a typed lambda
+calculus with recursive and higher-order functions, and recursive datatypes.
+Previous work has been limited to C-like languages (see graphics papers) and
+doesn't handle certain stuff.
 
-For example, in our quickselect example, it is unlikely to know the
-list~\texttt{l} that we will be operation on at compile time.
+This additional expressive power allows us to perform data specialization on new
+kinds of functions, like quickselect. We get some interesting results from
+this---including algorithm synthesis with asymptotic speedup. Higher-order
+functions also allow us to write combinators which cut across both stages (see
+graphics stuff).
 
-\cite{knoblock96} proposed the {\em data specialization}
-technique to overcome some of these limitations of program
-specialization.  
-Unlike program specialization,
-data specialization does not require knowledge of the first-stage
-arguments to produce fast code for the second stage.  
-Instead it yield
-a first stage program that maps the first stage arguments to an
-intermediate or {\em boundary data structure}, which is used by the
-second stage program along with the second stage arguments to produce
-the final results.
+We start by walking through a quickselect example in \ref{sec:example}.
+In \ref{sec:semantics}, we describe our staged language \lang, including its
+type system and operational semantics.
+In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
+algorithm which performs data specialization.
+Finally, in \ref{sec:examples}, we walk through what our algorithm does to
+various programs, like deriving classic algorithms.
 
-But their work applies to program fragments that don't involve sophisticated
-data types such as as sums and recursive types, and don't involve recursion.
+%It has potentially both control-flow and data dependencies on $a$.
 
-\subsection{Oldest intro}
+%Many programs have arguments coming at different frequencies. It can be
+%important to specialize these functions on some of their arguments, whether or
+%not we know those arguments, in order to perform invariant computations only
+%once.
 
-This most often arises when
-we want to evaluate a multivariate function several times, with some inputs
-dynamic (varying across invocations) and some static (constant across
-invocations).
-\begin{itemize}
-\item Computing $b^e$, where $b$ varies and $e$ is constant.
-\item Finding the $i$th largest element in a list, where the list is constant
-and $i$ varies.
-\item Computing the shading of a surface in a 3D rendering, where some
-contributors to the final pixel color can be approximated as constant over
-pixel blocks.  
-\end{itemize}
-Prior examples
-include the ``data specialization" work of Knoblock and Ruf~\cite{knoblock96}
-as well as the compilation of high-level shading languages for computer graphics
-such as RTSL~\cite{Proudfoot:2001} and Spark~\cite{Foley:2011}.
+%For example, quickselect. Or, the compilation of high-level shading languages
+%for computer graphics such as RTSL~\cite{Proudfoot:2001} and
+%Spark~\cite{Foley:2011}.
 
-Recall that partial evaluators operate on multivariate functions $f$, whose
-inputs are each labeled either {\em static} or {\em dynamic}. Once the static
-inputs are provided, partial evaluation of $f$ produces a residual depending on
-only the dynamic input. Mathematically, a partial evaluator is any $p$ such that
-\[
-	\forall f,x. \exists f_x. \left[p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)\right]
-\]
-where $\llbracket \cdot \rrbracket$ translates the text of a function to its
-mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
-input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
-specialized to $x$}.
-
-Taking a multivariate function $f$ as input, stage-splitting produces two
-functions $f_1$ and $f_2$, where $f_1$ uses the stage-\bbone\ input to produce a
-data structure, and $f_2$ consumes that data structure and the stage-\bbtwo\
-input to produce the final output.  More precisely, a stage splitter is any $s$
-such that
-\[
-	\forall f. \exists f_1,f_2. 
-	\left[
-		\begin{array}{l}
-		s(f) = (f_1,f_2) \text{ and } \\
-		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
-		\end{array}
-	\right]
-\]
-with the same $\llbracket \cdot \rrbracket$ notation as before.
-The stage splitting operation does not depend on the stage-\bbone\ input.
-This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.
-
-\Cref{sec:semantics} describes the grammar and type system of \lang, as well as
-a reference semantics which is essentially a partial evaluator that respects
-staging. \Cref{sec:splitting} describes a stage-splitting transformation for
-\lang\ and proves its equivalence to the reference semantics.  Finally,
-\Cref{sec:examples} gives several examples of how stage-splitting terms in
-\lang\ can be used to derive classic data structures, and discusses the
-effectiveness and limitations of this technique.  Notably, these examples make
-use of the stage-splitting features not available in prior work.
+%\begin{itemize}
+%\item Computing $b^e$, where $b$ varies and $e$ is constant.
+%\item Finding the $i$th largest element in a list, where the list is constant
+%and $i$ varies.
+%\item Computing the shading of a surface in a 3D rendering, where some
+%contributors to the final pixel color can be approximated as constant over
+%pixel blocks.  
+%\end{itemize}

### 2015-02-22 paper/semantics.tex
mention fixpoints; change first paragraph of section 4

@@
 
-All three worlds contain unit, product, sum, function, and recursive types
-defined in the usual fashion. For example, given two terms $\typesone{e_1}A$ and
-$\typesone{e_2}B$ at world \bbonem, one can form the pair
+All three worlds contain fixpoints and unit, product, sum, function, and
+recursive types defined in the usual fashion. For example, given two terms
+$\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form the pair
 $\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world \bbonem.
@@
 
-A multistage function, such as \texttt{qsStaged} from \ref{sec:staging}, has
+A multistage function, such as \texttt{qsStaged} from \ref{sec:example}, has
 type $A\to \fut (B \to C)$ at \bbonem.%

### 2015-02-22 paper/intro.tex
finished a pass through the intro

@@
 
-\emph{Data specialization} \cite{knoblock96} is a different technique which
-specializes $f$ \emph{without} the fixed argument $a$. Instead, it transforms
-$f$ into a pair of functions $f_1$ and $f_2$. $f_1(a)$ produces a data structure
-populated with the results of the subexpressions depending only on $a$. $f_2$
-takes this data structure and the varying argument, and computes the result;
-that is, $f_2(f_1(a),-)$ is the same as $f(a,-)$. Importantly, unlike in partial
-evaluation, neither $f_1$ nor $f_2$ depend on $a$!
+\emph{Data specialization} \cite{knoblock96} is a technique for specializing $f$
+\emph{without} the fixed argument $a$, instead splitting $f$ into a pair of
+functions $f_1$ and $f_2$. $f_1(a)$ produces a data structure containing the
+results of the computations which depend only on $a$; $f_2$ then completes the
+computation, given this data structure and the varying argument; that is,
+$f_2(f_1(a),-)$ computes $f(a,-)$.
 
-In this paper, we extend the technique of data specialization to a typed lambda
-calculus with recursive and higher-order functions, and recursive datatypes.
-Previous work has been limited to C-like languages (see graphics papers) and
-doesn't handle certain stuff.
+Previous work on data specialization has been limited to simple, imperative
+languages. In this paper, we extend data specialization to a typed lambda
+calculus, allowing us to specialize more interesting programs.
 
-This additional expressive power allows us to perform data specialization on new
-kinds of functions, like quickselect. We get some interesting results from
-this---including algorithm synthesis with asymptotic speedup. Higher-order
-functions also allow us to write combinators which cut across both stages (see
-graphics stuff).
+When splitting certain recursive functions, like \texttt{quickselect}, our
+algorithm synthesizes recursive data structures and traversal algorithms which
+yield asymptotic speedups over the original function. Splitting higher-order
+combinators, like \texttt{map}, provides compositional reasoning at the source
+level while cross-cutting fixed runtime stages, as in graphical shading
+languages like Spark~\cite{Foley:2011}.
 
-We start by walking through a quickselect example in \ref{sec:example}.
+We start in \ref{sec:example} with an extended example of splitting
+\texttt{quickselect}.
 In \ref{sec:semantics}, we describe our staged language \lang, including its
@@
 In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
-algorithm which performs data specialization.
-Finally, in \ref{sec:examples}, we walk through what our algorithm does to
-various programs, like deriving classic algorithms.
-
-%It has potentially both control-flow and data dependencies on $a$.
-
-%Many programs have arguments coming at different frequencies. It can be
-%important to specialize these functions on some of their arguments, whether or
-%not we know those arguments, in order to perform invariant computations only
-%once.
-
-%For example, quickselect. Or, the compilation of high-level shading languages
-%for computer graphics such as RTSL~\cite{Proudfoot:2001} and
-%Spark~\cite{Foley:2011}.
-
-%\begin{itemize}
-%\item Computing $b^e$, where $b$ varies and $e$ is constant.
-%\item Finding the $i$th largest element in a list, where the list is constant
-%and $i$ varies.
-%\item Computing the shading of a surface in a 3D rendering, where some
-%contributors to the final pixel color can be approximated as constant over
-%pixel blocks.  
-%\end{itemize}
+algorithm for performing data specialization.
+Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
+transforms a variety of other programs.

### 2015-02-22 paper/abstract.tex
various fixes
tweak abstract to line up with intro

@@
 
-In this paper, we provide a theoretical understanding of this splitting process
-by defining it as a program transformation on \lang, a typed lambda calculus
-equipped with a modal staging operator. This approach extends splitting to
-language features absent in prior work, including first-order functions and
-disjoint unions.
+Prior work on splitting has been limited to simple, imperative languages; in
+this paper, we extend splitting to \lang, a typed lambda calculus equipped with
+a modal staging operator, allowing us to split recursive and higher-order
+functions.
 
@@
 
-\TODO (Is this true?)
-We also show that splitting a staged implementation of an interpreter yields a
-partial evaluator.
-
-%into one function per stage, performing all the work depending only on the
-%available inputs.
-
-%as much work as possible given the inputs available at each stage.
-
-%In this way, splitting is essentially an algorithm generation technique.
-
-% amortized
+% something about compositional, cross-cutting code

### 2015-02-22 paper/splitting.tex
add some citations to intro

@@
 
-Since lambdas may represent multi-stage computations, 
+Since lambdas may represent multistage computations, 
 masking splits the body of lambdas as general world \bbonem\ terms (as described in \ref{sec:split-one}), 
@@
 since functions are already fully reduced in our semantics.
-However, since the body of a function may itself be multi-stage, splitting must continue into it.
+However, since the body of a function may itself be multistage, splitting must continue into it.
 The immediate result is a new function formed from the first-stage part of the original body.

### 2015-02-22 paper/splitting.tex
worked on examples and ground explanation

@@
 
-\subsection {Three World System?}
+\subsection {Necessity of \bbonep}
 
-Why do we need a three world system?
+We have not yet given a justification for why first stage code must be
+partitioned between the worlds \bbonem\ and \bbonep.
+Certainly one could imagine a simpler system without all the \texttt{grnd} annotations.  
+Why wouldn't this work?
+
+To see why, consider a function with a monostage body. 
+For example, one which simply doubles its argument:
+\begin{lstlisting}
+2`prev{
+  1`let fun double (n : int) = n + n
+  in next{`2`hold {`1`double 5`2`}-7`1`}`2`
+}`
+\end{lstlisting}
+Using the rule for splitting at \bbonem, this term splits into:
+\begin{lstlisting}
+1`let fun double (n : int) = (n+n, ())
+in double 5`
+
+2`l => 
+let fun double (n : unit, l0 : unit) = ()
+in (double (pi2 l); pi1 l)-7`
+\end{lstlisting}
+
+Observe that it's wasteful to run the stage \bbtwo\ version of \texttt{double}, since it always returns \texttt{()}. 
+In this case, the wasted work is small, but if \texttt{double} were a recursive function,
+then it's runtime may be unbounded.
+
+Optimizing such an inefficiency away after-the-fact is in general a global program analysis,
+since functions like \texttt{double} could be passed around as values.  
+So instead of relying on hefty analysis, 
+we take the approach of adding enough structure to the input language's type system to
+allow the input code to {\em prove} itself to be---in some parts---monostage.
+
+We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
+Yet even if a language designer desired a different implicit/explicit trade-off,
+then the three-world \lang\ would still be useful as a typed intermediate representation.
 

### 2015-02-22 paper/relatedwork.tex
add quickselect citation
related work

--- 2014-10-10 paper/relatedwork.tex
@@
 
-Partial evaluation is a well-studied technique for specializing programs to some of their input.  
-The usual phrasing of the partial evaluation problem implicitly includes binding time analysis,
-but many presentations start with a stage-annotated representation, as we have done.
-We've shown that our dynamic semantics can be used to perform partial evaluation 
-[need to do more reading to know exactly how it compares].
+As Jorring and Scherlis argue frequency reduction and precomputation
+are common techniques for both designing algorithms and performing
+compiler optimizations~\cite{JS86-staging}.  The idea behind
+frequency reduction is to identify computations that are performed
+multiple times and pull them ahead so that they can be performed once
+and used as needed later.  The idea behind precomputation is to
+identify computations that can be performed earlier, for example at
+compile time if their inputs are available statically and perform them
+at that earlier time. It is not difficult to see that many interesting
+algorithm design techniques and compiler optimizations do exactly
+frequency reduction and precomputation.  For example, dynamic
+programming and loop hoisting are a frequency reduction techniques,
+whereas dynamic data structures and incremental computation are
+frequency reduction techniques.
 
-One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
-and so they must necessarily unfold recursive calls and wait for all stage \bbone\ input to operate.
-There has been some work to avoid these issues by templating [need citations].
-Stage-splitting, in contrast, goes all the way to operating on terms which are open on first stage variables,
-and so these problems are avoided.
+Perhaps one of the most well studied example of precomputation is
+partial evaluation.  Partial evaluation distinguishes between a static
+(compile-time) and dynamic (run-time) stage. Given a program and the
+values of its static inputs, which are the subset of the inputs that
+become available at compilation time, partial evaluation generates a
+specialized program by performing the computations that depend on the
+statics inputs~\cite{jones96}.  The roots of partial evaluation go
+back to Kleene, who in 1952 proved that a multivariate program
+(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
+$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
+that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
+y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
+however.  Lombardi appears to be the first person to use the term in
+his work on incremental computation~\cine{Lombardi67}.  Since then,
+there has a lot of work on partial evaluation, the only broad-strokes
+of which we can mention in this here.  We refer the interested reader
+to the excellent book by Jones, Gommard, and Sestoft for a
+comprehesive discussion of work until early 90's~\cite{JGS93}.
 
-Stage-splitting itself has appeared in the literature---under 
-the names {\em pass separation} and {\em data specialization}--- for various simpler input languages.
+At a high level, earlier approaches to partial evaluation can be
+viewed as operating in two stages: binding time analysis and program
+specialization.  For a multivariate program, e.g., $f(x_s,x_d)$, with
+clearly marked static and dynamic arguments, binding-time analysis
+identifies all the parts of the program that can be computed by the
+knowledge of static arguments. Using this information and the static
+arguments themselves, program specialization specializes the program
+to a partially evaluated program that can be used with many different
+dynamic arguments.  This approach has been applied to construct
+partial evaluators for a number of languages such as 
+Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
+part first-order subsets.
 
-Most often, splitting is considered a compiler optimization (\cite{jorring86},\cite{knoblock96})
-to minimize recomputation.
-In this use case, only the arguments to the program are annotated with stages, 
-and the splitter is left to perform an {\em ad hoc} binding time analysis to determine the 
-internal staging of the program.
-Because this occurs without programmer supervision, the binding time analyzers for these uses are 
-reluctant to place stage \bbone\ code within stage \bbtwo\ conditionals, 
-since that would result in speculative behavior that could increase the overall runtime.
-We don't worry about this since the annotations of \lang\ are assumed to represent the will of a programmer.
+Subsequence experience with binding time analysis based approach
+showed that it can be difficult to control, leading to specialized
+programs whose performance can be difficult to predict. This led to
+investigations based on type systems for making explicit in the
+program the stage of each
+computation~\cite{GJ91-lambda,NN92-twolevel}.  Davies~\cite{Davies96}
+presented a logical construction of binding-time type systems by
+deriving a type system via the Curry-Howard isomorphism applied to
+temporal logic.  In subsequence work, Davies and Pfenning proposed a
+new type system for staged computation based on a particular fragment
+of modal logic~\cite{DP01-modal}.  These later approaches can be
+viewed more generally meta programming, where evaluating a program at
+a certain stage with its inputs at that stage gives another program to
+be executed at the next stage (e.g.,~\cite{JS86-staging}). The work on
+MetaML scaled type-based ideas to a full-scale language by developing
+statically typed programming language based on ML that enables the
+programmer to express programs with multiple
+stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
+to Davies~\cite{Davies96} but extends it in several important ways.
+Nanevksi and Pfenning further extended the these techniques by
+allowing free variable to occur within staged
+computations~\cite{NP05-nn}.
 
-Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  
-But instead of using a binding-time analysis to determine what parts of the program to put in what shader, 
-Spark gives the programmer manual control of term staging using a language that is in spirit similar to \lang.
+The type-system component of our work is closely related to this later
+line of work on metaprogramming and staged computation.  The specific
+extension to the typed lambda calculus that we use here is based on
+the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
+small ways to account for the specific evaluation model that we
+propose.  The key difference between our work and this prior work is
+the staging or the splitting algorithm, which given a two-staged
+program splits it into two stages that can be executed sequentially,
+passing information from the earlier to the latter stage by
+construction. 
+%
+\ur{Do prior work evaluate first-stage expressions inside second stage
+expressions?}
+%
+Unlike in partial evaluation, our splitting algorithm requires no
+knowledge of the arguments themselves. The only information that is
+required is their staging. Using just the staging information, the
+splitting algorithm can generate programs that can perform non-trivial
+transformations on the code automatically.  As our examples illustrate
+such staging transformations can improve the efficiency of the source
+program, sometimes in asymptotically significant ways.  
+We are not aware of prior work in the context of higher-order
+functional languages that presents such a splitting algorithm.
 
-We are also not the first to use stage splitting to derive algorithms.
-\cite{malmkjaer89}, expanding on the work of \cite{barzdins88}, defined a stage-splitter
-for a language of tail-recursive first-order equations with explicit staging annotations [double check that],
-and considered applications to string matching.
+Perhaps the closest work to ours is the work on ``data
+specialization'' by Knoblock and Ruf~\cite{knoblock96}.  As our work,
+Knoblock and Ruf are concerned about the problem of splitting a
+program into two stages.  They only consider, however, a very simple
+first-order language and straightline programs.
 
-Our work is distinct from all previous stage splitters in the following ways:
-\begin{itemize}
-\item Our input language, \lang, has a disciplined type theory.
-\item \lang\ has full first-class functions (and by extension, recursion), as well as sum types.
-\item Our stage splitting is proven correct.
-\end{itemize}
+%% \section{Related Work}
 
-The goal of {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}) is 
-to use code in a host language to write code in an object language.
-Like partial evaluation and stage-splitting, this can be modeled by a staged language with explicit annotations.
-Instead of identifying stages with points in time, stages are identified with host/object levels of the language,
-and so type safety ensures the proper separation of levels of the language and safety of the constructed object code.
+%% \ur{
+%% \begin{itemize}
+%% \item 
+%% I find it hard to distinguish between blue and black on printed paper.
+%% \item
+%% $\pause$ should be called ``promote'' or something like that. I
+%%   believe this is what Taha calls cross stage persistence.  THat work should
+%%   be cited.
+%% \end{itemize}
+%% }
 
-But the differences in goals of metaprogramming do cause a few differences in form.
-For example, it's generally preferred in the metaprogramming context to use 
-a dynamics that duplicates the contents of $\next$ blocks.
-Moreover, in metaprogramming systems it makes sense to define an operator, \texttt{run}, 
-which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
-This is essentially a form a backwards information flow.
-Thus, the \texttt{run} feature is directly incompatible with a temporal interpretation of stages,
-which stage-splitting requires.
+%% \crem{\emph{This text came from section 3:}
+%% \lang\ is adapted from \cite{davies96}, and is very similar to other staged
+%% languages considered in the partial evaluation and metaprogramming literature.
+%% Unlike \cite{davies96}, we restrict \lang\ to exactly two stages, to simplify
+%% our presentation of our stage-splitting algorithm. We expect that the techniques
+%% presented here extend to more stages.
+%% Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
+%% the metaprogramming literature, because we avoid duplicating stage \bbtwo\
+%% computations. In addition to being more efficient, our non-duplicating semantics
+%% are easier to compare with our splitting algorithm, whose development and
+%% exposition is our main goal.
+%% }
 
-\cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
-and showed the equivalence between that type system and existing one for binding time analysis. 
-That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
-That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
-In fact, if we change the $\next$ and $\prev$ rules of ours to 
-\begin{mathpar}
-\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
-\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
-\end{mathpar}
-and treat $\next~q$ as a partial value for any residual $q$,
-then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
-environment bookkeeping in $\redonesym$, by inlining residuals instead of
-hoisting them in \verb|let|-bindings.
-From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-{\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
+%% There are several distinctive aspect of our work which we wish to compare and contrast.
+%% Firstly, \lang\ is a simply typed lambda calculus 
+%% with explicit stage annotations and a non-duplicating semantics.
+%% Its statics and dynamics enforce no backward information flow,
+%% which a is necessary property to define a splitting transformation
+%% that is equivalent to the dynamics.
+%% Finally, the splitting transformation itself operates on any well-typed open term in \lang.
 
-\subsection{Umut's notes}
+%% Partial evaluation is a well-studied technique for specializing programs to some of their input.  
+%% Often, the goal of partial evaluation systems is to go all the way from unstaged code to a specialized version.
+%% Although this problem definition implicitly includes binding time analysis,
+%% many presentations factor out that step and start with a stage-annotated representation, as we have done.
+%% We've shown that our dynamic semantics can be used to perform partial evaluation 
+%% [need to do more reading to know exactly how it compares].
 
----- DO NOT DELETE ----
+%% One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
+%% and so they must necessarily unfold recursive calls and wait for all stage \bbone\ input to operate.
+%% There has been some work to avoid these issues by templating [need citations].
+%% Stage-splitting, in contrast, goes all the way to operating on terms which are open on first stage variables,
+%% and so these problems are avoided.
 
-Answer the following question for all of the related work:
+%% Stage-splitting itself has appeared in the literature---under 
+%% the names {\em pass separation} and {\em data specialization}--- for various simpler input languages.
 
-What happens when we try to write quickselect as a meta program?
+%% Most often, splitting is considered a compiler optimization (\cite{jorring86},\cite{knoblock96})
+%% to minimize recomputation.
+%% In this use case, only the arguments to the program are annotated with stages, 
+%% and the splitter is left to perform an {\em ad hoc} binding time analysis to determine the 
+%% internal staging of the program.
+%% Because this occurs without programmer supervision, the binding time analyzers for these uses are 
+%% reluctant to place stage \bbone\ code within stage \bbtwo\ conditionals, 
+%% since that would result in speculative behavior that could increase the overall runtime.
+%% We don't worry about this since the annotations of \lang\ are assumed to represent the will of a programmer.
 
-\paragraph{Partial Evaluation.}
-Online and offline.  Both approaches have their merits but online
-partial evaluation can be too slow and if stopped lead to an
-inefficient residual.  Offline partial evaluation is faster.  Also,
-full self-application has only been achieved by offline partial
-evaluation.
+%% Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  
+%% But instead of using a binding-time analysis to determine what parts of the program to put in what shader, 
+%% Spark gives the programmer manual control of term staging using a language that is in spirit similar to \lang.
 
-Weaknesses of partial evaluation include~\cite{GJ05}:
-1) speedups linear in the subject programs run time (excuse moi?)
+%% We are also not the first to use stage splitting to derive algorithms.
+%% \cite{malmkjaer89}, expanding on the work of \cite{barzdins88}, defined a stage-splitter
+%% for a language of tail-recursive first-order equations with explicit staging annotations [double check that],
+%% and considered applications to string matching.
 
-2) good speedups require close knowledge of the program and some
-binding-time improvements.
+%% Our work is distinct from all previous stage splitters in the following ways:
+%% \begin{itemize}
+%% \item Our input language, \lang, has a disciplined type theory.
+%% \item \lang\ has full first-class functions (and by extension, recursion), as well as sum types.
+%% \item Our stage splitting is proven correct.
+%% \end{itemize}
 
-3) result of specialization can be hard to predict as it can lead to
-slowdown, code explosion, and non-termination at specialization time.
+%% The goal of {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}) is 
+%% to use code in a host language to write code in an object language.
+%% Like partial evaluation and stage-splitting, this can be modeled by a staged language with explicit annotations.
+%% Instead of identifying stages with points in time, stages are identified with host/object levels of the language,
+%% and so type safety ensures the proper separation of levels of the language and safety of the constructed object code.
 
-\paragraph{Meta programming}
+%% But the differences in goals of metaprogramming do cause a few differences in form.
+%% For example, it's generally preferred in the metaprogramming context to use 
+%% a dynamics that duplicates the contents of $\next$ blocks.
+%% Moreover, in metaprogramming systems it makes sense to define an operator, \texttt{run}, 
+%% which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
+%% This is essentially a form a backwards information flow.
+%% Thus, the \texttt{run} feature is directly incompatible with a temporal interpretation of stages,
+%% which stage-splitting requires.
 
-What happens when we try to write quickselect as a meta program?
+%% \cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
+%% and showed the equivalence between that type system and existing one for binding time analysis. 
+%% That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
+%% That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
+%% In fact, if we change the $\next$ and $\prev$ rules of ours to 
+%% \begin{mathpar}
+%% \infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
+%% \infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
+%% \end{mathpar}
+%% and treat $\next~q$ as a partial value for any residual $q$,
+%% then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
+%% environment bookkeeping in $\redonesym$, by inlining residuals instead of
+%% hoisting them in \verb|let|-bindings.
+%% From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
+%% {\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
 
-\ur{Umut: I am starting a paragraph on termination.  we might need to expand
-  on this.}
-\paragraph{Termination.}
+%% Online and offline.  Both approaches have their merits but online
+%% partial evaluation can be too slow and if stopped lead to an
+%% inefficient residual.  Offline partial evaluation is faster.  Also,
+%% full self-application has only been achieved by offline partial
+%% evaluation.
 
-There are two reasons for non-termination: infinite loops and
-execution of errorneous code.  The splitting algorithm can cause the
-first.  But can it also cause the second? (If all computations really
-depend on stage 1 code, then would it be possible to for example see
-something like 1/0?  i think so but only if there is a meta dependency
-between the data in two stages?)
+%% Weaknesses of partial evaluation include~\cite{GJ05}:
+%% 1) speedups linear in the subject programs run time (excuse moi?)
 
-Is in possible for the residual program to not terminate (unexpectedly)?
+%% 2) good speedups require close knowledge of the program and some
+%% binding-time improvements.
 
-Termination analysis in partial evaluation: see paper~\cite{AH96,GJ05}.
+%% 3) result of specialization can be hard to predict as it can lead to
+%% slowdown, code explosion, and non-termination at specialization time.
 
-Based on Glenstrup and Jones (pp. 1154, 1155), a partial evaluator
-should be complete. That is every computation depending on the static
-input should be reduced to a value (during specialization).
-Unfortunately, completeness is difficult to achieve if we also require
-that the specialization is always terminating.
+%% \paragraph{Meta programming}
 
-One problem with their description is that it is not clear which
-termination problem they are talking about termination of the residual
-or the specializer.
+%% What happens when we try to write quickselect as a meta program?
 
-Nontermination during specialization: 
-consider code
+%% \ur{Umut: I am starting a paragraph on termination.  we might need to expand
+%%   on this.}
+%% \paragraph{Termination.}
 
-double x = doubleplus (x,0)
+%% There are two reasons for non-termination: infinite loops and
+%% execution of errorneous code.  The splitting algorithm can cause the
+%% first.  But can it also cause the second? (If all computations really
+%% depend on stage 1 code, then would it be possible to for example see
+%% something like 1/0?  i think so but only if there is a meta dependency
+%% between the data in two stages?)
 
-doubleplus (u,v) = if u <= 0 then v else doubleplus (u-1,v+2)
+%% Is in possible for the residual program to not terminate (unexpectedly)?
 
-this can lead to non-termination if v is considered static.
+%% Termination analysis in partial evaluation: see paper~\cite{AH96,GJ05}.
 
-Is this a problem for us?
+%% Based on Glenstrup and Jones (pp. 1154, 1155), a partial evaluator
+%% should be complete. That is every computation depending on the static
+%% input should be reduced to a value (during specialization).
+%% Unfortunately, completeness is difficult to achieve if we also require
+%% that the specialization is always terminating.
 
-\paragraph{Supercompilation.}
-~\cite{Turchin86}
+%% One problem with their description is that it is not clear which
+%% termination problem they are talking about termination of the residual
+%% or the specializer.
 
-\paragraph{Intro.}
+%% Nontermination during specialization: 
+%% consider code
 
-A predominating theme in the design, implementatation, and
-optimization of algorithm and software is frequency reduction and
-precomputation.  The idea is to reduce the frequency of computations
-by reducing their frequency and ideally by precomputing when possible.
+%% double x = doubleplus (x,0)
 
-The reason for why this theme is as predominant as it is that it is
-naturally common in many actual applications.  For example, we may
-have a collection of names or numbers that we frequently lookup, as
-for example part of a membership test as part of a larger system.  In
-some applications areas such as in graphics (the motivation for our
-work), this is so common that there is a relatively large literature
-on speeding up graphics computations by taking advantage of frequency
-reduction and precomputation.
+%% doubleplus (u,v) = if u <= 0 then v else doubleplus (u-1,v+2)
 
-One way to express frequency of computations is to allow the
-programmer to stage computation explicitly by making explicit the
-stage of each computation and then by moving computations between
-stages to improve efficiency by performing work as early (to save time
-later when performance might be more critical) and  as infrequently
-as possible (to reduce overall time).
+%% this can lead to non-termination if v is considered static.
 
-For example, in meta-programming~\cite{}, the programmer can assign a
-program expression a stage (stage 1, 2, 3, ...), which indicates the
-stage at which that expression evaluates.  Combined with the ability
-for earlier stages to construct expressions of later stages,
-meta-programming techniques can be used to construct efficient
-programs that improve efficiency over their non-staged counterparts by
-reducing frequency reduction and precomputation.  For example, the
-programmer can write a program that takes a list of numbers $\ell$ as
-a first stage value and returns a second-stage function that performs
-a fast membership test on that list $\ell$ by inlining all the
-comparisons.  When given a key to lookup in the second stage, the
-membership test function would quickly return the result.
+%% Is this a problem for us?
+
+%% \paragraph{Supercompilation.}
+%% ~\cite{Turchin86}
+
+%% \paragraph{Intro.}
+
+%% A predominating theme in the design, implementatation, and
+%% optimization of algorithm and software is frequency reduction and
+%% precomputation.  The idea is to reduce the frequency of computations
+%% by reducing their frequency and ideally by precomputing when possible.
+
+%% The reason for why this theme is as predominant as it is that it is
+%% naturally common in many actual applications.  For example, we may
+%% have a collection of names or numbers that we frequently lookup, as
+%% for example part of a membership test as part of a larger system.  In
+%% some applications areas such as in graphics (the motivation for our
+%% work), this is so common that there is a relatively large literature
+%% on speeding up graphics computations by taking advantage of frequency
+%% reduction and precomputation.
+
+%% One way to express frequency of computations is to allow the
+%% programmer to stage computation explicitly by making explicit the
+%% stage of each computation and then by moving computations between
+%% stages to improve efficiency by performing work as early (to save time
+%% later when performance might be more critical) and  as infrequently
+%% as possible (to reduce overall time).
+
+%% For example, in meta-programming~\cite{}, the programmer can assign a
+%% program expression a stage (stage 1, 2, 3, ...), which indicates the
+%% stage at which that expression evaluates.  Combined with the ability
+%% for earlier stages to construct expressions of later stages,
+%% meta-programming techniques can be used to construct efficient
+%% programs that improve efficiency over their non-staged counterparts by
+%% reducing frequency reduction and precomputation.  For example, the
+%% programmer can write a program that takes a list of numbers $\ell$ as
+%% a first stage value and returns a second-stage function that performs
+%% a fast membership test on that list $\ell$ by inlining all the
+%% comparisons.  When given a key to lookup in the second stage, the
+%% membership test function would quickly return the result.

### 2015-02-22 paper/umut.tex
related work

@@
+\section{Related Work}
+
+As Jorring and Scherlis argue frequency reduction and precomputation
+are common techniques for both designing algorithms and performing
+compiler optimizations~\cite{JS86-staging}.  The idea behind
+frequency reduction is to identify computations that are performed
+multiple times and pull them ahead so that they can be performed once
+and used as needed later.  The idea behind precomputation is to
+identify computations that can be performed earlier, for example at
+compile time if their inputs are available statically and perform them
+at that earlier time. It is not difficult to see that many interesting
+algorithm design techniques and compiler optimizations do exactly
+frequency reduction and precomputation.  For example, dynamic
+programming and loop hoisting are a frequency reduction techniques,
+whereas dynamic data structures and incremental computation are
+frequency reduction techniques.
+
+Perhaps one of the most well studied example of precomputation is
+partial evaluation.  Partial evaluation distinguishes between a static
+(compile-time) and dynamic (run-time) stage. Given a program and the
+values of its static inputs, which are the subset of the inputs that
+become available at compilation time, partial evaluation generates a
+specialized program by performing the computations that depend on the
+statics inputs~\cite{jones96}.  The roots of partial evaluation go
+back to Kleene, who in 1952 proved that a multivariate program
+(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
+$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
+that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
+y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
+however.  Lombardi appears to be the first person to use the term in
+his work on incremental computation~\cine{Lombardi67}.  Since then,
+there has a lot of work on partial evaluation, the only broad-strokes
+of which we can mention in this here.  We refer the interested reader
+to the excellent book by Jones, Gommard, and Sestoft for a
+comprehesive discussion of work until early 90's~\cite{JGS93}.
+
+At a high level, earlier approaches to partial evaluation can be
+viewed as operating in two stages: binding time analysis and program
+specialization.  For a multivariate program, e.g., $f(x_s,x_d)$, with
+clearly marked static and dynamic arguments, binding-time analysis
+identifies all the parts of the program that can be computed by the
+knowledge of static arguments. Using this information and the static
+arguments themselves, program specialization specializes the program
+to a partially evaluated program that can be used with many different
+dynamic arguments.  This approach has been applied to construct
+partial evaluators for a number of languages such as 
+Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
+part first-order subsets.
+
+Subsequence experience with binding time analysis based approach
+showed that it can be difficult to control, leading to specialized
+programs whose performance can be difficult to predict. This led to
+investigations based on type systems for making explicit in the
+program the stage of each
+computation~\cite{GJ91-lambda,NN92-twolevel}.  Davies~\cite{Davies96}
+presented a logical construction of binding-time type systems by
+deriving a type system via the Curry-Howard isomorphism applied to
+temporal logic.  In subsequence work, Davies and Pfenning proposed a
+new type system for staged computation based on a particular fragment
+of modal logic~\cite{DP01-modal}.  These later approaches can be
+viewed more generally meta programming, where evaluating a program at
+a certain stage with its inputs at that stage gives another program to
+be executed at the next stage (e.g.,~\cite{JS86-staging}). The work on
+MetaML scaled type-based ideas to a full-scale language by developing
+statically typed programming language based on ML that enables the
+programmer to express programs with multiple
+stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
+to Davies~\cite{Davies96} but extends it in several important ways.
+Nanevksi and Pfenning further extended the these techniques by
+allowing free variable to occur within staged
+computations~\cite{NP05-nn}.
+
+The type-system component of our work is closely related to this later
+line of work on metaprogramming and staged computation.  The specific
+extension to the typed lambda calculus that we use here is based on
+the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
+small ways to account for the specific evaluation model that we
+propose.  The key difference between our work and this prior work is
+the staging or the splitting algorithm, which given a two-staged
+program splits it into two stages that can be executed sequentially,
+passing information from the earlier to the latter stage by
+construction. 
+%
+\ur{Do prior work evaluate first-stage expressions inside second stage
+expressions?}
+%
+Unlike in partial evaluation, our splitting algorithm requires no
+knowledge of the arguments themselves. The only information that is
+required is their staging. Using just the staging information, the
+splitting algorithm can generate programs that can perform non-trivial
+transformations on the code automatically.  As our examples illustrate
+such staging transformations can improve the efficiency of the source
+program, sometimes in asymptotically significant ways.  
+We are not aware of prior work in the context of higher-order
+functional languages that presents such a splitting algorithm.
+
+Perhaps the closest work to ours is the work on ``data
+specialization'' by Knoblock and Ruf~\cite{knoblock96}.  As our work,
+Knoblock and Ruf are concerned about the problem of splitting a
+program into two stages.  They only consider, however, a very simple
+first-order language and straightline programs.
+
+\begin{comment}
+
 \section{Umut's notes}
@@
 Elimination of duplicate work.
+\end{comment}

### 2015-02-22 paper/example.tex

@@
 \section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
-\label{sec:staging}
+\label{sec:example}
 
 \begin{abstrsyn}
-
-\ur{
-\begin{itemize}
-\item 
-I find it hard to distinguish between blue and black on printed paper.
-\item
-$\pause$ should be called ``promote'' or something like that. I
-  believe this is what Taha calls cross stage persistence.  THat work should
-  be cited.
-\end{itemize}
-}
 
@@
 Suppose that we wish to issue order statistics queries on a collection of items
-represented as a list \texttt{l}. We can use quickselect \cite{quickselect}, an
+represented as a list \texttt{l}. We can use quickselect \cite{Hoare:1961}, an
 expected linear time algorithm which takes a list \texttt{l} and a rank

### 2015-02-22 paper/implementation.tex

@@
 
-We also have a number of optimizations which drastically improve the readability
-of split programs. We split patterns directly, rather than translating them into
-lower-level constructs. Finally, we perform many local simplifications during
-the splitting algorithm, most notably, not pairing precomputations when one is
-known to be $\tup{}$.
+In the splitting algorithm, we perform a number of optimizations which
+drastically improve the readability of split programs. For example, we split
+patterns directly, instead of first translating them into lower-level
+constructs. We also take advantage of many local simplifications, most notably,
+not pairing precomputations when one is known to be $\tup{}$.
 

### 2015-02-22 paper/intro.tex

@@
 
-\emph{Program specialization}, or partial evaluation, is a well-known
-specialization technique which, given $a$, transforms $f$ into a new function
-$f_a(-)$ which computes $f(a,-)$. This transformation essentially substitutes
-$a$ for the first argument of $f$, then evaluates in place any subexpressions of
-$f$ depending only on that argument.
+\emph{Program specialization}, or partial evaluation \cite{futamura71,jones96},
+is a well-known specialization technique which, given $a$, transforms $f$ into a
+new function $f_a(-)$ which computes $f(a,-)$. This transformation essentially
+substitutes $a$ for the first argument of $f$, then evaluates in place any
+subexpressions of $f$ depending only on that argument.
 
-\emph{Data specialization} \cite{knoblock96} is a technique for specializing $f$
-\emph{without} the fixed argument $a$, instead splitting $f$ into a pair of
-functions $f_1$ and $f_2$. $f_1(a)$ produces a data structure containing the
-results of the computations which depend only on $a$; $f_2$ then completes the
-computation, given this data structure and the varying argument; that is,
-$f_2(f_1(a),-)$ computes $f(a,-)$.
+\emph{Data specialization} \cite{knoblock96,jorring86} 
+is a technique for specializing $f$ \emph{without} the fixed argument $a$,
+instead splitting $f$ into a pair of functions $f_1$ and $f_2$. $f_1(a)$
+produces a data structure containing the results of the computations which
+depend only on $a$; $f_2$ then completes the computation, given this data
+structure and the varying argument; that is, $f_2(f_1(a),-)$ computes $f(a,-)$.
 

### 2015-02-23 paper/relatedwork.tex
related work

@@
 are common techniques for both designing algorithms and performing
-compiler optimizations~\cite{JS86-staging}.  The idea behind
-frequency reduction is to identify computations that are performed
-multiple times and pull them ahead so that they can be performed once
-and used as needed later.  The idea behind precomputation is to
-identify computations that can be performed earlier, for example at
-compile time if their inputs are available statically and perform them
-at that earlier time. It is not difficult to see that many interesting
-algorithm design techniques and compiler optimizations do exactly
-frequency reduction and precomputation.  For example, dynamic
-programming and loop hoisting are a frequency reduction techniques,
-whereas dynamic data structures and incremental computation are
-frequency reduction techniques.
+compiler optimizations~\cite{JS86-staging}.  The idea behind frequency
+reduction is to identify computations that are performed multiple
+times and pull them ahead so that they can be performed once and used
+later as needed.  The idea behind precomputation is to identify
+computations that can be performed earlier, for example at compile
+time if their inputs are available statically and perform them at that
+earlier time.  Many interesting algorithm design techniques and
+compiler optimizations perform frequency reduction and precomputation.
+For example, dynamic programming and loop hoisting are a frequency
+reduction techniques, whereas dynamic data structures and incremental
+computation are precomputation techniques.
 
@@
 however.  Lombardi appears to be the first person to use the term in
-his work on incremental computation~\cine{Lombardi67}.  Since then,
-there has a lot of work on partial evaluation, the only broad-strokes
-of which we can mention in this here.  We refer the interested reader
-to the excellent book by Jones, Gommard, and Sestoft for a
-comprehesive discussion of work until early 90's~\cite{JGS93}.
+his work on incremental computation~\cite{Lombardi67}.  Since then,
+there has a lot of work on partial evaluation, only broad-strokes of
+which we can mention in this here.  We refer the interested reader to
+the excellent book by Jones, Gommard, and Sestoft for a comprehesive
+discussion of work until early 90's~\cite{JGS93}.
 
@@
 viewed as operating in two stages: binding time analysis and program
-specialization.  For a multivariate program, e.g., $f(x_s,x_d)$, with
-clearly marked static and dynamic arguments, binding-time analysis
-identifies all the parts of the program that can be computed by the
-knowledge of static arguments. Using this information and the static
-arguments themselves, program specialization specializes the program
-to a partially evaluated program that can be used with many different
-dynamic arguments.  This approach has been applied to construct
-partial evaluators for a number of languages such as 
+specialization.  For a multivariate program with clearly marked static
+and dynamic arguments, binding-time analysis identifies all the parts
+of the program that can be computed by the knowledge of static
+arguments. Using this information and the static arguments themselves,
+program specialization specializes the program to a partially
+evaluated program that can be used with many different dynamic
+arguments.  This approach has been applied to construct partial
+evaluators for a number of languages such as
 Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
@@
 
-Subsequence experience with binding time analysis based approach
-showed that it can be difficult to control, leading to specialized
-programs whose performance can be difficult to predict. This led to
-investigations based on type systems for making explicit in the
-program the stage of each
-computation~\cite{GJ91-lambda,NN92-twolevel}.  Davies~\cite{Davies96}
-presented a logical construction of binding-time type systems by
-deriving a type system via the Curry-Howard isomorphism applied to
-temporal logic.  In subsequence work, Davies and Pfenning proposed a
-new type system for staged computation based on a particular fragment
-of modal logic~\cite{DP01-modal}.  These later approaches can be
-viewed more generally meta programming, where evaluating a program at
-a certain stage with its inputs at that stage gives another program to
-be executed at the next stage (e.g.,~\cite{JS86-staging}). The work on
-MetaML scaled type-based ideas to a full-scale language by developing
-statically typed programming language based on ML that enables the
-programmer to express programs with multiple
+Experience with binding time analysis showed that it can be difficult
+to control, leading to programs whose performance can be difficult to
+predict. This led to investigations based on type systems for making
+explicit in the program the stage of each
+computation~\cite{GJ91-lambda,NN92-twolevel} and writing ``meta
+programs'' that when evaluated at a stage, give another program to be
+evaluated at the next stage. Jorring and Scherlis describe this idea
+meta programming in their paper (~\cite{JS86-staging}) as one of the
+three possible approaches to taking advantage of
+staging. Davies~\cite{Davies96} presented a logical construction of
+binding-time type systems by deriving a type system via the
+Curry-Howard isomorphism applied to temporal logic.  Davies and
+Pfenning proposed a new type system for staged computation based on a
+particular fragment of modal logic~\cite{DP01-modal}. The work on
+MetaML extended type-based techniques to a full-scale language by
+developing statically typed programming language based on ML that
+enables the programmer to express programs with multiple
 stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar

### 2015-02-23 paper/intro.tex

@@
 
-\emph{Data specialization} \cite{knoblock96,jorring86} 
+\emph{Data specialization} \cite{knoblock96,JS86-staging} 
 is a technique for specializing $f$ \emph{without} the fixed argument $a$,
@@
 structure and the varying argument; that is, $f_2(f_1(a),-)$ computes $f(a,-)$.
+Crucially, none of the generated code ($f_1$ or $f_2$) depends on $a$!
 

### 2015-02-24 paper/example.tex
small edits all over

@@
 \texttt{qsStaged} has a more precise type---it is a first-stage function which
-takes a \textrm{list} (ignoring the $\curr$ for the moment) and a $\fut\rmint$
+takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$
 (an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
@@
 
-Our type system, discussed in \ref{sec:semantics}, ensures that the staging
+\lang's type system, discussed in \ref{sec:semantics}, ensures that the staging
 annotations are consistent, in the sense that computations marked as
 first-stage cannot depend on ones marked as second-stage. The $\curr$ and
-\texttt{grnd} annotations further distinguish those first-stage computations
+\texttt{grnd} (``ground'') annotations further distinguish those first-stage computations
 which do not contain any embedded second-stage code. (Code \emph{inside}
 \texttt{grnd}, like the \texttt{part}ition function, is guaranteed not to
-contain second-stage code; other code, like \texttt{qsStaged}, may.) We will
+contain second-stage code; first stage code outside a \texttt{grnd}, 
+like \texttt{qsStaged}, may.) We will
 discuss the importance of this additional distinction in \ref{sec:splitting}.
@@
 all the work depending only on \texttt{l} (the first stage), and one which uses
-\texttt{k} and the result of that work on \texttt{l} to compute the element with
+that partial result and \texttt{k} to compute the element with
 rank \texttt{k} in \texttt{l}.
@@
 \texttt{part}ition contains no second-stage code, we can run it entirely in the
-first stage!
+first stage.
 
@@
 $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
-complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
-improvement.
+complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a significant improvement.
 

### 2015-02-24 paper/semantics.tex
small edits all over

@@
 
+\input{figures/grammars}
+
 \input{figures/values}
-
-\input{figures/grammars}
 
@@
 
-Although our products and functions are restricted to types at the same world,
-$\fut$ essentially allows us to build ``mixed-stage'' products and functions.
+Although products and functions are restricted to types at the same world,
+$\fut$ allows costruction of ``mixed-stage'' products and functions.
 For example, quickselect is a function at world \bbonem\ which takes a
 $\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
-second-stage computation of an integer) to a $\fut\rmint$ (a second-stage
+second-stage integer) to a $\fut\rmint$ (a second-stage
 computation of an integer).
@@
 
-We evaluate \lang\ terms in two stages. The first stage of evaluation:
+We evaluate \lang\ terms in two stages. The first stage of evaluation has 
+three interdependent modes of operation, depending on the world of the term it is evaluating.
+Specifically, first stage evaluation:
 \begin{enumerate}
 \item fully reduces terms in \bbonep, as they are purely first-stage;
-\item reduces terms in \bbonem, skipping over any world \bbtwo\ terms they may
-contain (inside $\next$);
-\item does not reduce terms in \bbtwo, but scans them for any \bbonem\ terms
-they may contain (inside $\prev$) and reduces those.
+\item reduces terms in \bbonem, save for any world \bbtwo\ terms they may
+contain (inside \texttt{next}s);
+\item scans terms in \bbtwo\ for any world \bbonem\ subterms (inside \texttt{prev}s), 
+and reduces those in place.
 \end{enumerate}
@@
 
-This process ultimately produces a monostaged term, called a \emph{residual}
+First-stage evaluation ultimately produces a monostaged term, called a \emph{residual}
 (\ref{fig:values}), all of whose redexes are second-stage computations. The
@@
 we evaluate $e$ using $\redonesym$ to $[\xi;v]$. Because $\colmix e{\fut A}$,
-$v$ must be $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
+$v$ must have the canonical form $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
 $\xi$ into a sequence of second-stage \texttt{let} bindings around $\hat{y}$,

### 2015-02-24 paper/splitting.tex
small edits all over

--- 2015-02-22 paper/splitting.tex
@@
 computations before any second-stage ones, actually evaluating terms in this
-fashion is somewhat intricate, because stages are syntactically interleaved
+fashion is rather involved, because stages are syntactically interleaved
 within terms. In this section, we present a \emph{splitting algorithm} which
@@
 Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
-we also provide a way to split those as well.
+we must also provide a way to split those as well.
 This is given by the new relation $e \splitonesym \mathcal{P}$,
@@
 
-Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{sepM} terms,
-with one complication arising from the fact that \lang\ terms at world \bbonem\ produce multistage output.
+Similar to above, we also define a $\sepredtwosym$ relation 
+which can be used to evaluate these seperated world \bbonep terms.
+One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
 For example, consider the term
@@
 
-We resolve this mismatch by introducing one more translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
-which is just a technical device to convert residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
+We resolve this mismatch by introducing a translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
+which converts residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
 As before, we enforce the ``syntactically separated'' property structurally,
@@
 and $q$ is a monostage term representing all of the second-stage computation in $\xi$ and $v$.
-Note that masking exists only for the purposes of stating correctness; it is not part of the splitting algorithm itself.
+Note that masking is a technical device for the purpose of stating correctness of splitting at \bbonem; 
+it is not part of the splitting algorithm itself.
 
@@
 
-\subsection {Three World System?}
+\subsection {Necessity of \bbonep}
 
-Why do we need a three world system?
+We have not yet given a justification for why first stage code must be
+partitioned between the worlds \bbonem\ and \bbonep.
+Certainly one could imagine a simpler system without all the \texttt{grnd} annotations.  
+Why wouldn't this work?
+
+In order to be correct, any code at \bbonem\ must be split with the 
+pessimistic assumption that it may result in work at the second stage.
+When this assumption turns out not to be true---that is, purely monostage code---
+then splitting may produce second stage code which is needlessly costly 
+(an example of this will occur in the next section).
+Detecting and optimizing this case is in general a global program analysis,
+since functions can be passed around as values.  
+So instead of relying on hefty analysis, 
+we take the approach of adding enough structure to the input language's type system to
+allow the input code to {\em prove} itself to be---in some parts---monostage.
+
+We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
+Yet even if a language designer desired a different implicit/explicit trade-off,
+then the three-world \lang\ would still be useful as a typed intermediate representation.
 

### 2015-02-24 paper/figures/splitting-overview.tex
redo syntactically-separated-term overview figure to look better

@@
 
-\textbf{Syntactically Separated Term with Stage \bbtwo\ Output:}
+\textbf{Syntactically Separated Term with Stage \bbtwo\ Output ---
+$\pipeS p l r$}
+\begin{enumerate}
+\item[] Precomputation $p$ is a monostaged term containing the program's
+first-stage computations.
+\item[] Resumer $l.r$ is a monostaged term containing the program's 
+second-stage computations.
+\item[] Evaluates via $\sepredtwo {\pipeS p l r} q$, where the resulting
+residual $q$ is a monostaged term containing the remaining second-stage
+computations.
+\item[] $\sepredtwo {\pipeS p l r} {[b/l]r}$, where $b$ (the boundary value) is
+generated by $\reduce p b$.
+\end{enumerate}
 
-\hspace{2em} Construction: $\pipeS p l r$, where
+\textbf{Syntactically Separated Term with Multistage Output ---
+$\pipeM c l r$}
+\begin{enumerate}
+\item[] Combined result $c$ is a monostage term containing the program's
+first-stage computations.
+\item[] Resumer $l.r$ is a monostage term containing the program's 
+second-stage computations.
+\item[] Evaluates via $\sepredone {\pipeM c l r} {\mval i q}$, where 
+$\mval i q$ is the resulting masked value (defined below).
+\item[] $\sepredone {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (the boundary
+value) is generated by $\reduce c {\tup{i,b}}$.
+\end{enumerate}
 
-	\hspace{4em} $p$ is the {\em precomputation}, a monostage term representing the stage \bbone\ part of the program
-
-	\hspace{4em} $l.r$ is the {\em resumer}, a monostage term representing the stage \bbtwo\ part of the program
-
-\hspace{2em} Evaluation Form: $\sepredtwo {\pipeS p l r} q$, where
-
-	\hspace{4em} $\pipeS p l r$ is the syntactically separated term to be evaluated
-
-	\hspace{4em} $q$ is the resulting residual, a monostage term representing remaining stage \bbtwo\ work
-
-\hspace{2em} Evaluation Rule: $\sepredtwo {\pipeS p l r} {[b/l]r}$, where $b$ (called the {\em boundary value}) is generated by $\reduce p b$
-
-\textbf{Syntactically Separated Term with Multistage Output:}
-
-\hspace{2em} Construction: $\pipeM c l r$, where
-
-	\hspace{4em} $c$ is the {\em combined result}, a monostage term representing the stage \bbone\ part of the program
-
-	\hspace{4em} $l.r$ is the {\em resumer}, a monostage term representing the stage \bbtwo\ part of the program
-
-\hspace{2em} Evaluation Form: $\sepredone {\pipeM c l r} {\mval i q}$, where
-
-	\hspace{4em} $\pipeM c l r$ is the syntactically separated term to be evaluated
-
-	\hspace{4em} $\mval i q$ is the resulting masked value (defined below)
-
-\hspace{2em} Evaluation Rule: $\sepredone {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (also called the {\em boundary value}) is generated by $\reduce c {\tup{i,b}}$
-
-\textbf{Masked Values:}
-
-\hspace{2em} Construction: $\mval i q$, where
-
-	\hspace{4em} $i$ is the {\em immediate result}, a monostage value representing the stage \bbone\ result
-
-	\hspace{4em} $q$ is the {\em residual}, a monostage term representing stage \bbtwo\ work left to do
+\textbf{Masked Values ---
+$\mval i q$}
+\begin{enumerate}
+\item[] Immediate result $i$ is a monostaged value representing the first-stage
+result.
+\item[] Residual $q$ is a monostaged term representing the second-stage work
+left to do.
+\end{enumerate}
 

### 2015-02-24 paper/intro.tex
small edits all over
working on the intro

@@
 \section{Introduction}
+
+Consider the following function that takes as arguments an $x$ and $m$
+$y$'s and invokes a function $f(\cdot, \cdot)$ on consecutive
+values of $y$'s.
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+  @$(f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@.
+\end{lstlisting}
+Let's assume for concreteness that $f(x,y)$ takes time to the size of
+the first argument.  The function $F(x_0, y_0 \ldots y_{m-1}$ then
+takes time $O(n \cdot m)$ where $n$ is the size of the input $x_0$.
+
+Such computations, where we apply a multivariate function to arguments
+where only one of the arguments change over time, are common.  In
+fact, many research problems revolve around improving performance of
+such functions.  For example, algorithm design techniques such as
+dynamic programming take advantage of the fact that the function $f$
+above is being passed the same argument many times, and thus optimize
+it so that it can re-use results related to $f$.  Similarly, 
+crucial compiler optimization techniques such as loop hoisting and
+common subexpression evaluation take identify common common
+computations across all invocations of $f$ and perform them once and
+for all at the beginning of the computation.
+
+The common theme in all such techniques is to achieve what Jorring and
+Scherlis back in 1980's called: {\em frequency reduction} and {\em
+  precomputation}~\cite{JS86-staging}.  The idea behind frequency
+reduction is to identify computations that are performed multiple
+times and pull them ahead so that they can be performed once and used
+later as needed.  The idea behind precomputation is to identify
+computations that can be performed earlier, for example at compile
+time if their inputs are available statically and perform them at that
+earlier time.  Jorring and Scherlis discuss three possible approaches
+to achieving frequency reduction and procomputation by performing {\em
+  stage transformations} that stages the computation into distinct
+stages where the earlier stages host computations that can be
+performed earlier or that can might be used frequently.  The three
+approaches are
+\begin{itemize}
+\item 
+partial evaluation,
+
+\item
+meta programming, and
+
+\item 
+pass separation.
+
+\end{itemize}
+
+The idea behind partial evaluation is to take advantage of the fact
+that certain data becomes available earlier than others.
+Specifically, partial evaluation distinguishes between a {\em static}
+and {\em dynamic} stage and performs all static computations first,
+followed be the dynamic stage.  For example, suppose that we wish to
+compute $F(x_0, y_0 \ldots y_{m-1})$ and that $x_0$ is a static
+variable becoming available at an earlier stage.  We can take
+advantage of this by specializing $f$ for $x_0$ such that $f(x,y) =
+f_{x0}(y)$, where $f_{x0}$ is the function specialized for $x0$.  We
+can then re-write our function $F$ as follows.
+
+\begin{lstlisting}
+fun @$F_{x0}(y_0, y_1 \ldots y_{m-1}$@ = 
+  @$(f_{x0}(y_0);  f_{x0}(y_1)  \ldots f_{x0}(ym)$@.
+\end{lstlisting}
+
+There has been much research on partial evaluation (\secref{related}).
+Broadly speaking, earlier (pre 90's) research used binding-time
+analysis to determine the static and dynamic parts of a computation
+and specialized the program based on that.  Binding time analysis can
+be fully automated but can be difficult to predict.  Later research
+tends to use {\em meta programming}, where the programmer writes a
+program that given the static values generates a specialized program
+for those values.  Meta programming gives the programmer relatively
+fine-grained control over partial evaluation but comes at the cost of
+more complex reasoning and a loss of automation.
+
+The third and final approach, called pass separation, assumes much
+less information to perform a stage transformations.  Instead of
+requiring the static data, it requires just the knowledge of which
+computations are static and dynamic and uses this knowledge to
+generate a stage one (static) and stage two (dynamic) computations.
+Applied to our example, pass separation would {\em split} function
+into $f1$ and $f2$ such that $f(x,y) = f2(f1(x),y)$.  We can then
+rewrite our example as
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+@$(f2(f1(x), y_0);  f2(f1(x), y_1)  \ldots f2(f1(x), y_{m-1})$@.
+\end{lstlisting}
+%
+A compiler can then rewrite this as 
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+let @$fx = f1(x)$@
+in @$(f2(fx, y_0);  f2(fx, y_1)  \ldots f2(fx, y_{m-1})$@.
+\end{lstlisting}
+We refer to such a program as {\em untangled}.
+
+%
+The key difference with partial evaluation is that we have not
+specialized the program to any specific static value but simply
+translated it into an equivalent program where the first stage
+function $f1$ only consumes first-stage data.  We can evaluate such an
+expression in two stages: first, we evaluate all $f1(x)$'s and reduce
+them to a final value that can be used directly in the second stage
+(we can be smart about this evaluation and evaluate $f1(x)$ just
+once). Second, we evaluate all instances of $f2$'s and compute the
+final value.
+
+Since pass separation requires strictly less information than partial
+evaluation and meta programming, it can be a powerful technique that
+can be applied with far less information.  For example, the code that
+we obtained in our example, can be instantiated with a value of $x$
+that becomes available at run time and still obtain significant
+benefits.  As we show in this paper, these benefits can be
+asymptotically significant.  In other words, we can obtain some
+benefit of partial evaluation at run time in cases when we don't know
+the values of arguments ahead of time.  If, however, we do know the
+values ahead of time, then we can also apply partial evaluation to the
+untangled program, deriving again all the benefits of partial
+evaluation.
+
+As the example illustrates, untangling the program into two stages
+seems to have no cost.  So why don't we apply this kind of untangling
+all time time?
+
+\smallskip
+
+Untangling a program into stages to can be done manually.  In fact
+Jorring and Sherlis describe how to do this for some examples.
+Automation, however, has been difficult.  According to Jorring and
+Sherlis ``... partial evaluation technique is readily mechanized,
+while staging transformation approach will elude full automation for
+some time.''
+
+\smallskip
+
+That time has come.
+
+In this paper...
+
+\paragraph{Old intro below.}
 
@@
 languages. In this paper, we extend data specialization to a typed lambda
-calculus, allowing us to specialize more interesting programs.
+calculus, allowing us to specialize a broader class of programs.
 

### 2015-02-25 paper/paper.tex
wrote a new intro to splitting section
expanded figures
minor
a little work on examples and implementation
culling some old intro text
related work
title
small edits all over
a bit more work

@@
                  {atsign}{{@}}1
+                 {literalone}{1}1 % hax
+                 {literaltwo}{2}1 % hax
                  {^}{{$\curr$}}1
@@
 
+%% Vote on your favorite title! Lower is better!
 \title{Data Specialization for a Staged Lambda Calculus}
+\title{Algorithmically Staged Lambda Calculus}
+\title{Algorithmic Staging}
+\title{Automatically Splitting a Staged Lambda Calculus}
 
@@
 
-\section{Intro}
-
-(placeholder in order to keep section numbers stable)
-
-In \ref{fig:termSplitSummary}, we summarize the terminology of our
-formal development.
-
-%% \input{intro}
+\input{intro}
 
 \input{example}
+
 \input{semantics}
 
-%% %\vfill\eject\pagebreak
 \input{splitting}
+
 \input{implementation}
+
 \input{algoexamples}
 \input{relatedwork}
-\input{umut}
+%\input{umut}
 
@@
 
+%% %\vfill\eject\pagebreak
+
 \bibliography{paper}

### 2015-02-25 paper/splitting.tex
various symbol changes

@@
 
-We resolve this mismatch by introducing a translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
+We resolve this mismatch by introducing a translation, called {\em masking} and written $\rtab \xi v \vsplito \mathcal{V}$,
 which converts residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.

### 2015-02-25 paper/relatedwork.tex

@@
 
-Perhaps the closest work to ours is the work on ``data
-specialization'' by Knoblock and Ruf~\cite{knoblock96}.  As our work,
-Knoblock and Ruf are concerned about the problem of splitting a
-program into two stages.  They only consider, however, a very simple
-first-order language and straightline programs.
+Perhaps the closest work to ours is algorithm for ``data
+specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  As our
+work, Knoblock and Ruf are concerned about the problem of splitting an
+explicitly staged program into two stages.  However, they only
+consider, simple first-order language and straight-line
+programs. Automatic pass seperation has also been a topic of interest
+in programming systems for computer graphics, where, to achieve
+high-performance, programs are typically manually separated into
+components by frequency of execution (Seperation into a pre-determined
+set of per-frequency modules is \emph{required} by the execution model
+of modern graphics architectures.)  The software engineering
+challenges modifying multiple per-stage programs have led to
+suggestions of writing graphics programs in an explicitly staged
+programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
+deferring the task of pass seperation to the compiler. However, all
+prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
+limited to simple, imperitive languages.
 

### 2015-02-25 paper/splitting.tex
tweaked examples

@@
 \subsection {Necessity of \bbonep}
+\label{sec:needGround}
 
@@
 pessimistic assumption that it may result in work at the second stage.
-When this assumption turns out not to be true---that is, purely monostage code---
-then splitting may produce second stage code which is needlessly costly 
-(an example of this will occur in the next section).
+When this assumption turns out not to be true---that is, purely monostage code---then 
+splitting may produce second stage code which is needlessly costly 
+(an example of this will occur in \ref{sec:exampleQS}).
 Detecting and optimizing this case is in general a global program analysis,

### 2015-02-25 paper/intro.tex
intro pass complete

--- 2015-02-23 paper/intro.tex
@@
+\newcommand{\langname}{LAMDA12}
 \section{Introduction}
+
+Consider a function that takes as arguments an $x$ and $y_1 \ldots
+y_{m-1}$ and invokes a function $f(\cdot, \cdot)$ on consecutive
+values of $y$'s.  Using ``;'' for  sequential composition, we can
+write such a function as
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@.
+\end{lstlisting}
+%
+%% Let's assume for concreteness that $f(x,y)$ takes time to the size of
+%% the first argument.  The function $F(x_0, y_0 \ldots y_{m-1}$ then
+%% takes time $O(n \cdot m)$ where $n$ is the size of the input $x_0$.
+
+Such computations, where we apply a multivariate function to arguments
+where one of the argument remains the same are common.  In fact, many
+research problems aim at improving efficiency of such functions.  For
+example, algorithm design techniques such as dynamic programming
+identify the computation in $f$ that depend on $x$ try to ensure that
+such computations are performed no more than once.  Similarly, crucial
+compiler optimization techniques such as loop hoisting and common
+subexpression elimination take identify common computations across all
+invocations of $f$ and perform them once and for all at the beginning
+of the computation. The whole area of incremental computation aims at
+adapting a computation to small changes, which may be applicable to
+this problem depending on the closeness of $y_i$'s.
+
+The common theme in all such design and optimization techniques is to
+achieve what Jorring and Scherlis back in 1980's called: {\em
+  frequency reduction} or {\em precomputation}~\cite{JS86-staging}.
+The idea behind frequency reduction is to identify computations that
+are performed multiple times and pull them ahead so that they can be
+performed once and used later as needed.  The idea behind
+precomputation is to identify computations that can be performed
+earlier, for example at compile time if their inputs are available
+statically and perform them at that earlier time.
+
+There has been much work on language-based techniques for performing
+frequency reduction and precomputation.  Broadly speaking, these
+techniques can be categorized into
+\begin{itemize}
+\item 
+partial evaluation,
+
+\item
+meta programming, and
+
+\item 
+pass separation or data specialization.
+
+\end{itemize}
+
+Partial evaluation performs precomputation by distinguishing between a
+{\em static} and {\em dynamic} stage and performs all static
+computations first, followed be the dynamic stage.  For example,
+suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
+statically known $x = 0$.  We can derive a specialization of $f$ at
+$0$, written $f_0$, such that $f(x,y) = f_{0}(y)$ and specialize $F$
+for $x=0$ as
+\begin{lstlisting}
+fun @$F_{0}(y_0, y_1 \ldots y_{m-1}$@ = 
+  @$(f_{0}(y_0);  f_{0}(y_1)  \ldots f_{0}(ym)$@.
+\end{lstlisting}
+%
+Such a partial evaluation typically proceed by performing a
+binding-time analysis to determine the static and dynamic parts of $f$
+and $F$ and specializes them based on the results of the binding-time
+analysis.
+
+A closely related technique, {\em meta programming} enables the
+programmer to write a program that takes the static values and
+generates a specialized program for those values.  Meta programming
+gives the programmer relatively fine-grained control over partial
+evaluation by requiring staging annotations in the program.  Such
+staging annotations make explicit the stage of each computation and
+can overcome some of the intricacies of binding-time analysis but this
+comes at the cost of more complex reasoning and some loss of
+automation (partial evaluation can in principle be applied fully
+automatically without relying on any annotations).
+
+The third and final approach, called {\em pass separation} or {\em
+  data specialization}, assumes much less information: instead of
+requiring the static data, it requires just the knowledge of which
+computations are static and dynamic and uses this knowledge to
+generate two stage computations. 
+%
+Applied to our example, pass separation would {\em split} the function
+$f$ into $f_s$ (static) and $f_d$ (dynamic) such that $f(x,y) =
+f_d(f_s(x),y)$.  We can then rewrite our example as
+%
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+let @$fsx = f_s(x)$@
+in @$(f_d(fsx, y_0);  f_d(fsx, y_1)  \ldots f_d(fsx, y_{m-1})$@.
+\end{lstlisting}
+We refer to such a program as {\em stratified} program.
+
+The key difference with partial evaluated program and the stratified
+program is that the latter has lost to generality: it work for any
+arguments, whereas the partially evaluated program is specialized for
+a particular value of $x$ ($x = 0$ in the example).  Thus, the pass
+separation can be applied without leading to any loss of generality.
+Going back to the terminology of Jorring and Scherlis, pass separation
+is a frequency reduction, whereas as partial evaluation and
+meta-programming are precomputation techniques.  Specifically, if
+desired partial evaluation can be applied further to a stratified
+program as an orthogonal improvement.
+
+But what is the benefit or pass separation? Let's assume for
+concreteness that $f(x,y)$ takes time in the size of the first
+argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes time $O(n
+\cdot m)$ where $n$ is the size of $x$. Assume further that in the
+stratified code the static function $f_s(x)$ takes $O(n)$ time and the
+$f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With these run
+times, the stratified code has complexity $O(n + m\log{n})$ instead of
+$O(n \cdot m)$, a significant asymptotic improvement.  (In this paper,
+we will achieve such a reduction in complexity automatically.)
+
+In summary pass separation causes no loss of generalization and can
+improve efficiency, so why is it not applied liberally?
+
+Indeed, in fact it is: in computer graphics, program are typically
+manually stratified for frequency reduction.  In fact, stratification
+into a pre-determined set ``modules'' is required by the execution
+model of modern graphics architectures.  Such manual stratification is
+a massive challenge: it breaks composition because the invariants are
+scattered over the code written over multiple strata and because
+modification to any one stratum requires modifying all others to
+match.  Ideally, we wish to write just one program and stratify it
+into multiple strata automatically.
+
+Such automatic stratification turns out to be challenging: the first
+successful approach that we know of is by Knoblock and
+Ruf\,\cite{knoblock96}, which like our work was inspired by computer
+graphics applications.  While practically effective, Knoblock and Ruf
+consider only a very simple language and are not able to stratify
+recursive programs.  Recent
+efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} .... 
+but still cannot stratify recursive programs.
+%
+In fact back in 80,s Jorring and Sherlis was able to stratify
+reasonably complex programs but only manually and note that
+``... partial evaluation technique is readily mechanized, while
+staging transformation approach will elude full automation for some
+time.''
+
+In this paper we make significant progress on the problem of automatic
+stratification of higher-order programs.  As our starting point, we
+take the modal language \langname with explicit staging in the style
+of Davies~\cite{Davies}, where the ``circle'' modality denotes
+computation in a later stage, and present a staged operational
+semantics similar for \langname (\secref{language}).  We then present
+a splitting algorithm that stratifies a \langname program into a two
+mono-staged programs, one for the first stage and for the second stage
+(\secref{splitting}).  Since each program is at a mono-stage, they are
+expressed in conventional function languages.
+
+The crux of the splitting algorithm is splitting recursive mixed stage
+functions that involve both a first stage and second stage
+computations.  When split, such a function turns into a first-stage
+function that yields a {\em boundary value} that encodes the
+``first-stage'' part of the computation.  If the function recurs on a
+first-stage argument, then the boundary value is typically a recursive
+data structure.  In other words, the splitting algorithm maps
+computational recursion into a recursive data type. If the boundary
+structure is generated as a result of a conditional, it is typically
+encoded as a (recursive) sum type, such as a tree, where the kind of
+the node indicates the recursion status of the first-stage evaluation.
+The ``second stage'' part of the computation takes the boundary value
+as an argument as well as the second stage argument and uses them to
+complete the computation, often by traversing the now precomputed
+boundary in light of the now available second-stage argument.
+
+Since it deals with unrestricted recursion and a rich language
+consisting of first-class functions, recursion, and sum and recursive
+types, the splitting algorithm is naturally complex. With some care,
+the algorithm nevertheless can be specified reasonably succinctly.  In
+fact, we have implemented the algorithm (\secref{implementation}) and
+applied to several examples (\secref{examples}).  As an interesting
+example, we show that a function 
+\begin{lstlisting}
+fun @$F(x: int list, y_0:int, y_1:int \ldots y_{m-1}:int$@ = 
+  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
+\end{lstlisting}
+where $f(x,y)$ selects the element of the list $x$ with rank $y$, when
+splits yields an asymptotically more efficient function that runs in
+expected $O(n\log{n} + m\log{n})$ time instead of expected $O(n \cdot
+m)$ time, a near linear time improvement.  Interestingly the code
+output by our splitting algorithm auto-generates the ``quicksort''
+algorithm, which it uses to build a binary search tree as a boundary
+value in the first stage, and in the second stage, generated a
+``binary search'' algorithm that takes the tree and performs a binary
+search on the tree as guided by the given rank.  To the best of our
+knowledge, no prior approaches to this problem can perform such
+complex transformations on code that can yield asymptotic
+improvements in run time.
+
+\paragraph{Old intro below.}
 

### 2015-02-25 paper/intro.tex
minor edits

@@
-\newcommand{\langname}{LAMDA12}
+\newcommand{\langname}{LAMDA12\xspace}
 \section{Introduction}
@@
 suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
-statically known $x = 0$.  We can derive a specialization of $f$ at
-$0$, written $f_0$, such that $f(x,y) = f_{0}(y)$ and specialize $F$
-for $x=0$ as
+statically known $x = [~]$.  We can derive a specialization of $f$ at
+$[~]$, written $f_{[~]}$, such that $f(x,y) = f_{[~]}(y)$ and specialize $F$
+for $x=[~]$ as
 \begin{lstlisting}
-fun @$F_{0}(y_0, y_1 \ldots y_{m-1}$@ = 
-  @$(f_{0}(y_0);  f_{0}(y_1)  \ldots f_{0}(ym)$@.
+fun @$F_{[~]}(y_0, y_1 \ldots y_{m-1})$@ = 
+  @$(f_{[~]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
 \end{lstlisting}
@@
 \begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
-let @$fsx = f_s(x)$@
-in @$(f_d(fsx, y_0);  f_d(fsx, y_1)  \ldots f_d(fsx, y_{m-1})$@.
+fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
+let @$fx = f_s(x)$@
+in @$(f_d(fx, y_0);  f_d(fx, y_1)  \ldots f_d(fx, y_{m-1}))$@.
 \end{lstlisting}
@@
 The key difference with partial evaluated program and the stratified
-program is that the latter has lost to generality: it work for any
+program is that the latter has lost no generality: it works with all
 arguments, whereas the partially evaluated program is specialized for
-a particular value of $x$ ($x = 0$ in the example).  Thus, the pass
+a particular value of $x$ ($x = [~]$ in the example).  Thus, the pass
 separation can be applied without leading to any loss of generality.
-Going back to the terminology of Jorring and Scherlis, pass separation
-is a frequency reduction, whereas as partial evaluation and
-meta-programming are precomputation techniques.  Specifically, if
-desired partial evaluation can be applied further to a stratified
-program as an orthogonal improvement.
+In fact, this is quite natural: going back to the terminology of
+Jorring and Scherlis, pass separation is a frequency reduction,
+whereas as partial evaluation and meta-programming are precomputation
+techniques. Consequently, if desired, partial evaluation can be
+orthogonally applied to a stratified program.
 
-But what is the benefit or pass separation? Let's assume for
-concreteness that $f(x,y)$ takes time in the size of the first
-argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes time $O(n
-\cdot m)$ where $n$ is the size of $x$. Assume further that in the
-stratified code the static function $f_s(x)$ takes $O(n)$ time and the
-$f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With these run
-times, the stratified code has complexity $O(n + m\log{n})$ instead of
-$O(n \cdot m)$, a significant asymptotic improvement.  (In this paper,
-we will achieve such a reduction in complexity automatically.)
+% Let's assume for concreteness that $f(x,y)$ takes time in the size of
+% the first argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes
+% time $O(n \cdot m)$ where $n$ is the size of $x$. Assume further that
+% in the stratified code the static function $f_s(x)$ takes $O(n)$ time
+% and the $f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With
+% these run times, the stratified code has complexity $O(n + m\log{n})$
+% instead of $O(n \cdot m)$, a significant asymptotic improvement.  (In
+% this paper, we will achieve such a reduction in complexity
+% automatically.)
 
-In summary pass separation causes no loss of generalization and can
-improve efficiency, so why is it not applied liberally?
+But what is the benefit or pass separation? The big benefit is
+efficiency. By stratifying the program, pass separation factors out
+frequently used the parts of computation, computes them first, and
+then reuse them later instead of computing them again.  Since it can
+be applied statically with no loss of generality, pass separation can
+significantly improve performance.  In fact, in computer graphics,
+program are typically manually stratified both for improved
+efficiency. Such stratification is in fact required by the execution
+model of some modern graphics architectures.
 
-Indeed, in fact it is: in computer graphics, program are typically
-manually stratified for frequency reduction.  In fact, stratification
-into a pre-determined set ``modules'' is required by the execution
-model of modern graphics architectures.  Such manual stratification is
-a massive challenge: it breaks composition because the invariants are
-scattered over the code written over multiple strata and because
-modification to any one stratum requires modifying all others to
-match.  Ideally, we wish to write just one program and stratify it
-into multiple strata automatically.
-
-Such automatic stratification turns out to be challenging: the first
+While highly desirable, manual stratification is difficult and highly
+nonmodular: program invariants are scattered over the code written in
+multiple strata and because modification to any one stratum requires
+modifying all others to match.  Ideally, we wish to write just one
+program and stratify it into multiple strata automatically.  But such
+automatic stratification turns out to be challenging: the first
 successful approach that we know of is by Knoblock and
@@
 recursive programs.  Recent
-efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} .... 
-but still cannot stratify recursive programs.
+efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
+cannot stratify recursive programs.
 %
@@
 In this paper we make significant progress on the problem of automatic
-stratification of higher-order programs.  As our starting point, we
-take the modal language \langname with explicit staging in the style
-of Davies~\cite{Davies}, where the ``circle'' modality denotes
-computation in a later stage, and present a staged operational
+stratification of higher-order functional programs.  As our starting
+point, we take the modal language \langname with explicit staging in
+the style of Davies~\cite{Davies}, where the ``circle'' modality
+denotes computation in a later stage, and present a staged operational
 semantics similar for \langname (\secref{language}).  We then present
@@
 functions that involve both a first stage and second stage
-computations.  When split, such a function turns into a first-stage
+computations.  When split, such a function turns into a stratum-1
 function that yields a {\em boundary value} that encodes the
@@
 the node indicates the recursion status of the first-stage evaluation.
-The ``second stage'' part of the computation takes the boundary value
-as an argument as well as the second stage argument and uses them to
-complete the computation, often by traversing the now precomputed
-boundary in light of the now available second-stage argument.
+The stratum-2 part of the computation, which corresponds to the
+``second stage'' takes the boundary value as an argument as well as
+the second-stage argument and uses them to complete the computation,
+often by traversing the now precomputed boundary in light of the now
+available second-stage argument.
 
@@
 applied to several examples (\secref{examples}).  As an interesting
-example, we show that a function 
+example, we consider a function
 \begin{lstlisting}
-fun @$F(x: int list, y_0:int, y_1:int \ldots y_{m-1}:int$@ = 
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
   @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
 \end{lstlisting}
-where $f(x,y)$ selects the element of the list $x$ with rank $y$, when
-splits yields an asymptotically more efficient function that runs in
-expected $O(n\log{n} + m\log{n})$ time instead of expected $O(n \cdot
-m)$ time, a near linear time improvement.  Interestingly the code
-output by our splitting algorithm auto-generates the ``quicksort''
-algorithm, which it uses to build a binary search tree as a boundary
-value in the first stage, and in the second stage, generated a
-``binary search'' algorithm that takes the tree and performs a binary
-search on the tree as guided by the given rank.  To the best of our
-knowledge, no prior approaches to this problem can perform such
-complex transformations on code that can yield asymptotic
-improvements in run time.
+where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
+using the ``quickselect'' algorithm.  When applied to this function,
+our split algorithm yields an asymptotically more efficient function
+that runs in expected $O(n\log{n} + m\log{n})$ time instead of
+expected $O(n \cdot m)$ time---a near linear time improvement.
+Interestingly the code output by our splitting algorithm
+auto-generates the ``quicksort'' algorithm, which it uses to build a
+binary search tree as a boundary value in the first stage.  For the
+second stage, the splitting algorithm generates a ``binary search''
+that takes the tree and performs a binary search on the tree as guided
+by the given rank.  To the best of our knowledge, no prior approaches
+can perform such complex transformations on higher-order code, nor can
+they yield such asymptotic improvements in run time.
 

### 2015-02-26 paper/semantics.tex
edits to masking and top of dynamics

@@
 We evaluate \lang\ terms in two stages. The first stage of evaluation has 
-three interdependent modes of operation, depending on the world of the term it is evaluating.
+three mutually recursive modes of operation, depending on the world of the term it is evaluating.
 Specifically, first stage evaluation:
-\begin{enumerate}
+\begin{itemize}
 \item fully reduces terms in \bbonep, as they are purely first-stage;
@@
 and reduces those in place.
-\end{enumerate}
+\end{itemize}
 These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
@@
 variables $\hat{y}$ in an evaluated \bbonem\ term. Thus, $\redonesym$ evaluation
-yields a pair $[\xi;v]$ of a residual table and a \emph{partial value} $v$
+yields a pair $\rtab \xi v$ of a residual table and a \emph{partial value} $v$
 containing variables bound in that table. The grammar of residual tables and
@@
 recursively on all subterms, yielding a monostaged residual $q$. At $\prev{e}$,
-we evaluate $e$ using $\redonesym$ to $[\xi;v]$. Because $\colmix e{\fut A}$,
+we evaluate $e$ using $\redonesym$ to $\rtab \xi v$. Because $\colmix e{\fut A}$,
 $v$ must have the canonical form $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}

### 2015-02-26 paper/splitting.tex
changed masking figure to use inference rules
edits to masking and top of dynamics

@@
 
-The point of the masking operation is simply to convert a partial value into a masked value $\mval i q$
-by assigning all of the first-stage content to $i$ and all of the second-stage content to $q$.
+The masking operation converts a partial value into a masked value $\mval i q$,
+where $i$ is a monostage value encoding all of the first-stage information of the original partial value,
+and $q$ is a monostage term encoding all of the second-stage information of the original partial value.
+Remember that masking will not play a role in the final splitting algorithm, 
+but instead exists to specify the contract to which splitting at \bbonem\ must adhere.
+Accordingly, masking should be thought of as a definition, and as such it is in some parts arbitrary.
+
 The rules of the masking relation are given in \ref{fig:valMask}.
-
 Masking operates by first inducting on the entries of the residual table.  
@@
 
-Masking assigns values in \bbonep\ to the immediate value
+Masking assigns ground values to the immediate value
 and likewise assigns references into the residual table to the resumer.
 In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
+Note that the \texttt{grnd} and \texttt{next} annotations are erased.
 
@@
 
+\input{figures/splitting-misc}
 \input{figures/splitting-term}
-\input{figures/splitting-misc}

### 2015-02-26 paper/splitting.tex
Beginning pass through spitting section.

@@
 
-Although the type system of \lang\ ensures that we can perform all first-stage
-computations before any second-stage ones, actually evaluating terms in this
-fashion is rather involved, because stages are syntactically interleaved
-within terms. In this section, we present a \emph{splitting algorithm} which
+%Although the type system of \lang\ ensures that we can perform all first-stage
+%computations before any second-stage ones, actually evaluating terms in this
+%fashion is rather involved, because stages are syntactically interleaved
+%within terms.
+
+In this section, we present a \emph{splitting algorithm} which
 statically converts a term in \lang\ into a pair of 
-syntactically % I don't like this word but temporarily here for continuity
-separate programs corresponding to the two stages of computation in that term.
+%syntactically % I don't like this word but temporarily here for continuity
+separate, monostaged programs that correspond to the first and second stage computations in that term.
 
@@
 $e$ splits into a syntactically separated program $\mathcal{P}$ 
-(via the new relation $e \splittwosym \mathcal{P}$), which
+(via the new relation $e \splittwosym \mathcal{P}$) which
 can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
-We enforce the ``syntactically separated'' condition by saying that $\mathcal{P}$ 
-can only have one form, namely $\pipeS p l r$, where  
-$p$ (called the {\em precomputation}) is a monostage term encoding all the
-first-stage computation that was in $e$, 
-and $l.r$ (called the {\em resumer}) is a monostage term encoding all the
-second-stage computation that was in $e$.  
-Likewise, $\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and plugs it in for $l$:
+By ``syntactically separated'' we mean that $\mathcal{P}$ 
+has the form $\pipeS p l r$, where  
+$p$ (called the {\em precomputation}) is a monostage term representing all the
+first-stage computations in $e$, 
+and $l.r$ (called the {\em resumer}) is a monostage term representing all the
+second-stage computations in $e$.  
+$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
 \[
@@
 \]
-Here we call $b$ the {\em boundary value}, as it represents the communication at the boundary between stages.
+We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
 A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.
 
-Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
-we must also provide a way to split those as well.
+Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
+\bbonem\ terms. %we must also provide a way to split those as well.
 This is given by the new relation $e \splitonesym \mathcal{P}$,
 where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
-$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computation that was in $e$, 
+$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computations in $e$, 
 and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
-second-stage computation that was in $e$.
+second-stage computations in $e$.
 
 Similar to above, we also define a $\sepredtwosym$ relation 
-which can be used to evaluate these seperated world \bbonep terms.
+which can be used to evaluate these seperated world \bbonep\ terms.
 One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
@@
 We resolve this mismatch by introducing a translation, called {\em masking} and written $\rtab \xi v \vsplito \mathcal{V}$,
-which converts residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
-As before, we enforce the ``syntactically separated'' property structurally,
-saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
-$i$ is a monostage value representing the first-stage part of $v$,
-and $q$ is a monostage term representing all of the second-stage computation in $\xi$ and $v$.
-Note that masking is a technical device for the purpose of stating correctness of splitting at \bbonem; 
+which converts residual tables and partial values---like those in the example above---into a syntactically separated version $\mathcal{V}$.
+As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value}) where 
+$i$ is a monostage value representing the first stage part of $v$,
+and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.
+Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
 it is not part of the splitting algorithm itself.
 
-With this machinery in place, we can define $\sepredonesym$ as just creating masked values with the single rule:
+Given this definition of masking, we define $\sepredonesym$ by a single rule that creates masked values:
 \[
@@
 second stage {\em and} to be present in the result.
-Correspondingly, in the $\sepredtwosym$ rule, the precomptuation $p$ reduces only to a boundary value which is passed to the resumer;
+Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$ reduces only to a boundary value which is passed to the resumer;
 but in the $\sepredonesym$ rule, the combined term $c$ reduces to a tuple containing both the immediate result and the boundary value,
@@
 This section proceeds by defining masking ($\vsplito$), 
-then using masking to motivate the definition of \bbonem-translation ($\splitonesym$),
-and finally coming back to define \bbtwo-translation ($\splittwosym$).
+then using masking to motivate the definition of \bbonem-term splitting ($\splitonesym$),
+and finally coming back to define \bbtwo-term splitting ($\splittwosym$).
 

### 2015-02-26 paper/intro.tex
started editing intro

--- 2015-02-25 paper/intro.tex
@@
 where one of the argument remains the same are common. In fact, many
-research problems aim at improving efficiency of such functions.  For
+research problems aim to improve the efficiency of such functions.  For
 example, algorithm design techniques such as dynamic programming
-identify the computation in $f$ that depend on $x$ try to ensure that
+identify the computation in $f$ that depend on $x$, and try to ensure that
 such computations are performed no more than once.  Similarly, crucial
 compiler optimization techniques such as loop hoisting and common
-subexpression elimination take identify common computations across all
+subexpression elimination identify common computations across all
 invocations of $f$ and perform them once and for all at the beginning
@@
 adapting a computation to small changes, which may be applicable to
-this problem depending on the closeness of $y_i$'s.
+this problem if the $y_i$'s are sufficiently similar.
 
 The common theme in all such design and optimization techniques is to
-achieve what Jorring and Scherlis back in 1980's called: {\em
+achieve what in the 1980s J{\o}rring and Scherlis called {\em
 frequency reduction} or {\em precomputation}~\cite{JS86-staging}.
 The idea behind frequency reduction is to identify computations that
-are performed multiple times and pull them ahead so that they can be
+are performed multiple times and hoist them so that they can be
 performed once and used later as needed.  The idea behind
 precomputation is to identify computations that can be performed
-earlier, for example at compile time if their inputs are available
-statically and perform them at that earlier time.
+earlier---for example, at compile time if their inputs are known
+statically---and perform them at that earlier time.
 
@@
 \begin{itemize}
-\item 
-partial evaluation,
-
-\item
-meta programming, and
-
-\item 
-pass separation or data specialization.
-
+\item partial evaluation,
+\item metaprogramming, and
+\item pass separation or data specialization.
 \end{itemize}
 
-Partial evaluation performs precomputation by distinguishing between a
-{\em static} and {\em dynamic} stage and performs all static
-computations first, followed be the dynamic stage.  For example,
+Partial evaluation achieves precomputation by distinguishing between 
+{\em static} and {\em dynamic} stages, and performing all static
+computations first, followed by dynamic ones.  For example,
 suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
@@
 
-A closely related technique, {\em meta programming} enables the
+A closely related technique, {\em metaprogramming}, enables the
 programmer to write a program that takes the static values and
-generates a specialized program for those values.  Meta programming
+generates a specialized program for those values.  Metaprogramming
 gives the programmer relatively fine-grained control over partial
@@
 \end{lstlisting}
-We refer to such a program as {\em stratified} program.
+We call such a program {\em stratified}.
 
-The key difference with partial evaluated program and the stratified
-program is that the latter has lost to generality: it work for any
-arguments, whereas the partially evaluated program is specialized for
-a particular value of $x$ ($x = 0$ in the example).  Thus, the pass
-separation can be applied without leading to any loss of generality.
-Going back to the terminology of Jorring and Scherlis, pass separation
+The key difference between partially-evaluated and stratified
+programs is that the latter is more general: it works for any
+arguments, whereas the partially-evaluated program is specialized for
+a particular value of $x$ ($x = 0$ in the example).  
+In the terminology of J{\o}rring and Scherlis, pass separation
 is a frequency reduction, whereas as partial evaluation and
-meta-programming are precomputation techniques.  Specifically, if
-desired partial evaluation can be applied further to a stratified
+metaprogramming are precomputation techniques.  Specifically, if
+desired, partial evaluation can be applied further to a stratified
 program as an orthogonal improvement.
 
-But what is the benefit or pass separation? Let's assume for
+But what is the benefit of pass separation? Let's assume for
 concreteness that $f(x,y)$ takes time in the size of the first
@@
 
-In summary pass separation causes no loss of generalization and can
+In summary, pass separation causes no loss of generality and can
 improve efficiency, so why is it not applied liberally?
 
-Indeed, in fact it is: in computer graphics, program are typically
+Indeed, in fact it is: in computer graphics, programs are typically
 manually stratified for frequency reduction.  In fact, stratification
@@
 successful approach that we know of is by Knoblock and
-Ruf\,\cite{knoblock96}, which like our work was inspired by computer
+Ruf\,\cite{knoblock96}, which, like our work, was inspired by computer
 graphics applications.  While practically effective, Knoblock and Ruf
@@
 stratification of higher-order programs.  As our starting point, we
-take the modal language \langname with explicit staging in the style
-of Davies~\cite{Davies}, where the ``circle'' modality denotes
+take the modal language \lang\ with explicit staging in the style
+of Davies~\cite{davies96}, where the ``circle'' modality denotes
 computation in a later stage, and present a staged operational
-semantics similar for \langname (\secref{language}).  We then present
-a splitting algorithm that stratifies a \langname program into a two
+semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then present
+a splitting algorithm that stratifies a \lang\ program into a two
 mono-staged programs, one for the first stage and for the second stage
-(\secref{splitting}).  Since each program is at a mono-stage, they are
-expressed in conventional function languages.
+(\ref{sec:splitting}).  Since each program is monostaged, they are
+expressed in conventional functional languages.
 
@@
 the algorithm nevertheless can be specified reasonably succinctly.  In
-fact, we have implemented the algorithm (\secref{implementation}) and
-applied to several examples (\secref{examples}).  As an interesting
+fact, we have implemented the algorithm (\ref{sec:implementation}) and
+applied to several examples (\ref{sec:examples}).  As an interesting
 example, we show that a function 
@@
 complex transformations on code that can yield asymptotic
-improvements in run time.
+improvements in runtime.
 

### 2015-02-26 paper/relatedwork.tex
minor edits

@@
 three possible approaches to taking advantage of
-staging. Davies~\cite{Davies96} presented a logical construction of
+staging. Davies~\cite{davies96} presented a logical construction of
 binding-time type systems by deriving a type system via the
@@
 stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
-to Davies~\cite{Davies96} but extends it in several important ways.
+to Davies~\cite{davies96} but extends it in several important ways.
 Nanevksi and Pfenning further extended the these techniques by

### 2015-02-26 paper/splitting.tex
Hacking on 4.0.  Trying to define the requirements of splitting more intuitively.

@@
 $e$ splits into a syntactically separated program $\mathcal{P}$ 
-(via the new relation $e \splittwosym \mathcal{P}$) which
-can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
+(via the new relation $e \splittwosym \mathcal{P}$) which is
+reduced to a residual $q'$ (via the new relation $\mathcal{P} \sepredtwosym q'$), which is the same as $q$ up to an equivalence.
 By ``syntactically separated'' we mean that $\mathcal{P}$ 
@@
 Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
-\bbonem\ terms. %we must also provide a way to split those as well.
-This is given by the new relation $e \splitonesym \mathcal{P}$,
-where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
-$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computations in $e$, 
-and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
-second-stage computations in $e$.
+\bbonem\ terms.
+However, defining correctness requirements for splitting \bbonem\ terms is
+challenging because unlike \bbtwo\ terms (which reduce to monostage
+residuals), \bbonem\ terms reduce to partial values and
+residual tables that represent computations at both stages.
+%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
+For example, the \bbonem\ term:
+\begin{lstlisting}
+(next {1+2}, grnd{3+4})
+\end{lstlisting}
+reduces (via $\redonesym$) to the residual table and partial value:
+\begin{lstlisting}
+[yhat |-> 1+2] (next {yhat}, grnd{7})
+\end{lstlisting}
+
+To resolve the mismatch between the multi-stage computations
+represented by residual tables and partial values and the desired monostaged
+outputs of splitting,
+% Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
+we define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mathcal{V}$,
+which converts residual tables and partial values into a syntactically separated {\em masked value} of the form $\mval i q$.
+%version $\mathcal{V}$.
+%As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
+The first component, $i$ of the masked value holds the result of all
+first stage computations in $v$.  The second component, $q$ is a
+residual representing all second stage computations in $\rtab \xi v$.
+
+%where 
+%$i$ is a monostage value representing the first stage part of $v$,
+%and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.
+
+%Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
+%it is not part of the splitting algorithm itself.
+
+Given this definition of masking, the requirements for splitting
+\bbonem\ terms are as follows. For any $\colmix e A$ which reduces to
+$\rtab \xi v$ (via $\redonesym$), $\rtab \xi v$ masks to $\mval i q$
+(via $\vsplito$), $e$ splits into a syntactically seperated program
+$\mathcal{P}$ of the form $\pipeM c l r$ (via the new relation $e
+\splitonesym \mathcal{P}$) which is reduced to a masked value $\mval i
+{q'}$ (via the new relation $\sepredtwosym$) where $q'$ is the same as
+$q$ up to an equivalence.
+
+%Given the definition of masking, \bbonem\ terms split via the new
+%relation $e \splitonesym \mathcal{P}$, where $\mathcal{P}$ must have
+%the form $\pipeM c l r$
+where $c$ (called the {\em combined term}) is
+a monostage term encoding all the first-stage computations in $e$, and
+$l.r$ (again called the {\em resumer}) is a monostage term encoding
+all the second-stage computations in $e$.
 
@@
 which can be used to evaluate these seperated world \bbonep\ terms.
-One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
-For example, consider the term
-\begin{lstlisting}
-(next {1+2}, grnd{3+4})
-\end{lstlisting}
-which reduces (via $\redonesym$) to the residual table and partial value
-\begin{lstlisting}
-[yhat |-> 1+2] (next {yhat}, grnd{7})
-\end{lstlisting}
-Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
-
-We resolve this mismatch by introducing a translation, called {\em masking} and written $\rtab \xi v \vsplito \mathcal{V}$,
-which converts residual tables and partial values---like those in the example above---into a syntactically separated version $\mathcal{V}$.
-As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value}) where 
-$i$ is a monostage value representing the first stage part of $v$,
-and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.
-Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
-it is not part of the splitting algorithm itself.
 

### 2015-02-26 paper/splitting.tex
Finish hacking up 4.0.  The section still uses the "seperated form" indirections, but I think it now flows nicely since the similarity between 1M splitting and 2 splitting in now more clear.

@@
 
-Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
-\bbonem\ terms.
-However, defining correctness requirements for splitting \bbonem\ terms is
+Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
+\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
+defining correctness requirements for splitting \bbonem\ terms is
 challenging because unlike \bbtwo\ terms (which reduce to monostage
-residuals), \bbonem\ terms reduce to partial values and
-residual tables that represent computations at both stages.
+residuals that can be directly compared to the monostage result of
+$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
+tables that represent computations at both stages.
 %One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
@@
 % Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
-we define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mathcal{V}$,
-which converts residual tables and partial values into a syntactically separated {\em masked value} of the form $\mval i q$.
+we choose to define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mval i q$,
+which converts residual tables and partial values into a syntactically separated {\em masked value}.
 %version $\mathcal{V}$.
 %As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
-The first component, $i$ of the masked value holds the result of all
-first stage computations in $v$.  The second component, $q$ is a
-residual representing all second stage computations in $\rtab \xi v$.
+The first component of the masked value, $i$, is a value holding the
+result of all first stage computations in $v$.  The second component,
+$q$, is a residual representing all second stage computations in $\rtab
+\xi v$.
 
@@
 
-Given this definition of masking, the requirements for splitting
-\bbonem\ terms are as follows. For any $\colmix e A$ which reduces to
-$\rtab \xi v$ (via $\redonesym$), $\rtab \xi v$ masks to $\mval i q$
-(via $\vsplito$), $e$ splits into a syntactically seperated program
-$\mathcal{P}$ of the form $\pipeM c l r$ (via the new relation $e
-\splitonesym \mathcal{P}$) which is reduced to a masked value $\mval i
-{q'}$ (via the new relation $\sepredtwosym$) where $q'$ is the same as
-$q$ up to an equivalence.
+We define the requirements for splitting \bbonem\ terms using the
+results of masking. Specifically, for any $\colmix e A$ which reduces
+to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
+$\mval i q$ (via $\vsplito$).  Then $e$ splits into a syntactically
+seperated program $\mathcal{P}$ (via the new relation $e \splitonesym
+\mathcal{P}$) which is reduced to a masked value $\mval i {q'}$ (via
+the new relation $\sepredtwosym$) where $q'$ is the same as $q$ up to
+an equivalence.
 
@@
 %the form $\pipeM c l r$
-where $c$ (called the {\em combined term}) is
-a monostage term encoding all the first-stage computations in $e$, and
+Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form
+$\pipeM c l r$ where $c$ (called the {\em combined term}) is a
+monostage term encoding all the first-stage computations in $e$, and
 $l.r$ (again called the {\em resumer}) is a monostage term encoding
-all the second-stage computations in $e$.
-
-Similar to above, we also define a $\sepredtwosym$ relation 
-which can be used to evaluate these seperated world \bbonep\ terms.
-
-Given this definition of masking, we define $\sepredonesym$ by a single rule that creates masked values:
+all the second-stage computations in $e$.  $\sepredonesym$ is defined
+by a single rule that yields masked values:
 \[
@@
 \]
-Here we see the primary difference between world \bbtwo\ terms and world \bbonem\ terms.
-Since $\coltwo e A$ reduces to an entirely second-stage result, 
-its first-stage subcomputations only exist to internally pass their value to the
-second stage;
-but since $\colmix e A$ reduces to a multistage result, 
-its first-stage subcomputations exist {\em both} to pass their value to the
-second stage {\em and} to be present in the result.
-Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$ reduces only to a boundary value which is passed to the resumer;
-but in the $\sepredonesym$ rule, the combined term $c$ reduces to a tuple containing both the immediate result and the boundary value,
-with only the latter being passed to the resumer.
 
-A summary of these operations and the relationship between them is provided on the left side of \ref{fig:splittingSummary}.
+Note that since $\coltwo e A$ reduces to an entirely second-stage
+residual, its first-stage computations only exist to generate input
+for second stage computations.  In contrast, since $\colmix e A$
+reduces to a multi-stage residual table and partial value, its first
+stage computations serve to both generate the value used as input for
+the second stage {\em and also} produce a first-stage value.
+Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
+generated via splitting reduces only to a boundary value which is
+passed to the resumer; but in the $\sepredonesym$ rule, the combined
+term $c$ reduces to a tuple containing both the first stage result $i$
+and the boundary value $b$ provided as input to the resumer.  A
+summary of the relationship between the results of directly evaluating
+\lang\ and the results of splitting is provided at left in
+\ref{fig:splittingSummary}.
 
-This section proceeds by defining masking ($\vsplito$), 
-then using masking to motivate the definition of \bbonem-term splitting ($\splitonesym$),
-and finally coming back to define \bbtwo-term splitting ($\splittwosym$).
+In the following subsections we formally define the masking
+transformation ($\vsplito$), then describe the algorithm for splitting
+both \bbonem\ terms ($\splitonesym$) and \bbtwo\ terms
+($\splittwosym$).
 

### 2015-02-26 paper/splitting.tex
More preample to 4.0.

@@
 In this section, we present a \emph{splitting algorithm} which
-statically converts a term in \lang\ into a pair of 
-%syntactically % I don't like this word but temporarily here for continuity
-separate, monostaged programs that correspond to the first and second stage computations in that term.
+statically converts a term in \lang\ into a pair of separate,
+monostaged terms that individually correspond to the first and second
+stage computations of the \lang\ term. Due to important differences in
+structure, we describe both the correctness condition and algorithm
+for splitting of \bbonem\ terms separately from that of \bbtwo\ terms.
+By construction, \bbonep\ terms are monostage and do not require
+splitting.
 
@@
 
-More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
+%More precisely
+For any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
 $e$ splits into a syntactically separated program $\mathcal{P}$ 
@@
 
-Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
-\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
-defining correctness requirements for splitting \bbonem\ terms is
+%Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
+%\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
+Defining the correctness requirements for splitting \bbonem\ terms is
 challenging because unlike \bbtwo\ terms (which reduce to monostage
@@
 Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
-generated via splitting reduces only to a boundary value which is
+generated via splitting reduces only to a boundary value $b$ which is
 passed to the resumer; but in the $\sepredonesym$ rule, the combined
 term $c$ reduces to a tuple containing both the first stage result $i$
-and the boundary value $b$ provided as input to the resumer.  A
-summary of the relationship between the results of directly evaluating
-\lang\ and the results of splitting is provided at left in
-\ref{fig:splittingSummary}.
+and the boundary value $b$. A summary of the relationship between the
+results of directly evaluating \lang\ and the results of splitting is
+provided at left in \ref{fig:splittingSummary}.
 
-In the following subsections we formally define the masking
-transformation ($\vsplito$), then describe the algorithm for splitting
-both \bbonem\ terms ($\splitonesym$) and \bbtwo\ terms
-($\splittwosym$).
+The following subsections formally define the masking transformation
+($\vsplito$), then describe the algorithm for splitting both
+\bbonem\ terms ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
 

### 2015-02-26 paper/paper.tex
minor edits
dangerous and experimental new masking section

@@
         moredelim=*[is][\color{stageMonoColor}]{3`}{`},
+        moredelim=*[is][\colorbox{black}]{4`}{`},
         literate={|->}{{\ $\mapsto$\ }}1

### 2015-02-26 paper/splitting.tex
dangerous and experimental new masking section

--- 2015-02-26 paper/splitting.tex
@@
 
-The rules of the masking relation are given in \ref{fig:valMask}.
-Masking operates by first inducting on the entries of the residual table.  
-Being purely second-stage content, these are reified into let statements at the top of the resumer.
-Once the table is empty, masking inducts on value itself.
+So to motivate the definition we chose, consider the following residual table and partial value,
+colored as before:
+\begin{lstlisting}
+2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (grnd{7}, next{`2`yhat`1`}))`
+\end{lstlisting}
+To reduce this term to just its first-stage information ($i$ in the output), we start by redacting all of the second stage (blue) parts,
+along with the surrounding \texttt{next} annotations:
+\begin{lstlisting}
+4`##########`1` (`4`######`1`, injL (grnd{7}, `4`######`1`))`
+\end{lstlisting}
+The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
+\begin{lstlisting}
+1`((), injL (grnd{7},()))`
+\end{lstlisting}
+Finally, we drop \texttt{grnd} annotations:
+\begin{lstlisting}
+1`((), injL (7,()))`
+\end{lstlisting}
+To reduce the original partial value to just its second-stage information ($q$ in the output), 
+we start by redacting all \texttt{grnd} blocks, \texttt{next} annotations, and injection tags
+(essentially, the tuple structure is kept, but all other structure is discarded):
+\begin{lstlisting}
+2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
+\end{lstlisting}
+Where the \texttt{grnd} blocks left a hole, we put in a unit values:
+\begin{lstlisting}
+2`[yhat |-> 1+2] (yhat, ((),yhat))`
+\end{lstlisting}
+Finally, we reify the residual table into \texttt{let} bindings:
+\begin{lstlisting}
+2`let yhat = 1+2 in (yhat, ((),yhat))`
+\end{lstlisting}
 
-Masking assigns ground values to the immediate value
-and likewise assigns references into the residual table to the resumer.
-In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
-Note that the \texttt{grnd} and \texttt{next} annotations are erased.
+The formal implementation of the masking relation is given in \ref{fig:valMask}.
+Note that since lambdas may represent multistage computations, 
+masking splits the body of lambdas as general world \bbonem\ terms (as described in \ref{sec:split-one}), 
+and packages the resulting terms as functions in both the first and second stage.
+% Masking operates by first inducting on the entries of the residual table.  
+% Being purely second-stage content, these are reified into let statements at the top of the resumer.
+% Once the table is empty, masking inducts on value itself.
 
-Masking distributes into tuples, injections, and rolls, since their subvalues may have content at both stages.
-However, the tags of injections and rolls are replicated only in the immediate value, 
-since they represent first-stage information.
+% Masking assigns ground values to the immediate value
+% and likewise assigns references into the residual table to the resumer.
+% In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
+% Note that the \texttt{grnd} and \texttt{next} annotations are erased.
 
-Since lambdas may represent multistage computations, 
-masking splits the body of lambdas as general world \bbonem\ terms (as described in \ref{sec:split-one}), 
-and packages the resulting terms as functions.
+% Masking distributes into tuples, injections, and rolls, since their subvalues may have content at both stages.
+% However, the tags of injections and rolls are replicated only in the immediate value, 
+% since they represent first-stage information.
 

### 2015-02-26 paper/splitting.tex
Half-baked. Just checking in to merge.

--- 2015-02-26 paper/splitting.tex
@@
 stage computations of the \lang\ term. Due to important differences in
-structure, we describe both the correctness condition and algorithm
-for splitting of \bbonem\ terms separately from that of \bbtwo\ terms.
-By construction, \bbonep\ terms are monostage and do not require
-splitting.
+structure, we describe both the splitting correctness condition and
+splitting algorithm for \bbonem\ terms separately from that of
+\bbtwo\ terms.  By construction, \bbonep\ terms are monostage and do
+not require splitting.
 
@@
 %More precisely
+
+As described in \ref{sec:example},
+
 For any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),

### 2015-02-26 paper/splitting.tex
Attempting to add more "ease in" to the world 2 splitting correctness definition in 4.0.

@@
 
-%The output of splitting is simpler because it is {\em syntactically separated};
-%The output of splitting is equivalent (to its input) because it can be
-%evaluated to the same residual that would be produced by the semantics of
-%\ref{sec:semantics}.
-
-%More precisely
-
-As described in \ref{sec:example},
-
-For any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
-$e$ splits into a syntactically separated program $\mathcal{P}$ 
-(via the new relation $e \splittwosym \mathcal{P}$) which is
-reduced to a residual $q'$ (via the new relation $\mathcal{P} \sepredtwosym q'$), which is the same as $q$ up to an equivalence.
-By ``syntactically separated'' we mean that $\mathcal{P}$ 
-has the form $\pipeS p l r$, where  
-$p$ (called the {\em precomputation}) is a monostage term representing all the
-first-stage computations in $e$, 
-and $l.r$ (called the {\em resumer}) is a monostage term representing all the
-second-stage computations in $e$.  
-$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
+\textbf{Correctness condition for \bbtwo\ term splitting.}
+Intuitively, as suggested in \ref{sec:example}, splitting a
+\lang\ program (a \bbtwo\ term) should yield two monostage terms where
+evaluating the second term (corresponding to second stage
+computations) on the output of the first (first stage computations),
+yields the same result as direct evaluation of the \lang\ program.
+More precisely, for any $\coltwo e A$ which reduces to a residual $q$
+(via $\diatwo e q$), $e$ splits into a syntactically separated program
+$\mathcal{P}$ (via the new relation $e \splittwosym \mathcal{P}$)
+which is reduced to a residual $q'$ (via the new relation $\mathcal{P}
+\sepredtwosym q'$), which is the same as $q$ up to an equivalence.  By
+``syntactically separated'' we mean that $\mathcal{P}$ has the form
+$\pipeS p l r$, where $p$ (called the {\em precomputation}) is a
+monostage term representing all the first-stage computations in $e$,
+and $l.r$ (called the {\em resumer}) is a monostage term representing
+all the second-stage computations in $e$.  $\sepredtwosym$ is defined
+by a single rule, which evaluates the precomputation and substitutes
+the resulting value in for $l$ in the resumer:
 \[
@@
 
-%Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
-%\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
-Defining the correctness requirements for splitting \bbonem\ terms is
+\textbf{Correctness condition for \bbonem\ term splitting.} Defining
+the correctness requirements for splitting \bbonem\ terms is
 challenging because unlike \bbtwo\ terms (which reduce to monostage
@@
 %One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
-For example, the \bbonem\ term:
-\begin{lstlisting}
-(next {1+2}, grnd{3+4})
-\end{lstlisting}
-reduces (via $\redonesym$) to the residual table and partial value:
-\begin{lstlisting}
-[yhat |-> 1+2] (next {yhat}, grnd{7})
-\end{lstlisting}
+%For example, the \bbonem\ term:
+%\begin{lstlisting}
+%(next {1+2}, grnd{3+4})
+%\end{lstlisting}
+%reduces (via $\redonesym$) to the residual table and partial value:
+%\begin{lstlisting}
+%[yhat |-> 1+2] (next {yhat}, grnd{7})
+%\end{lstlisting}
 
 To resolve the mismatch between the multi-stage computations
-represented by residual tables and partial values and the desired monostaged
-outputs of splitting,
-% Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
-we choose to define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mval i q$,
-which converts residual tables and partial values into a syntactically separated {\em masked value}.
-%version $\mathcal{V}$.
-%As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
-The first component of the masked value, $i$, is a value holding the
-result of all first stage computations in $v$.  The second component,
-$q$, is a residual representing all second stage computations in $\rtab
-\xi v$.
-
-%where 
-%$i$ is a monostage value representing the first stage part of $v$,
-%and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.
-
-%Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
-%it is not part of the splitting algorithm itself.
+represented by residual tables and partial values and the desired
+monostaged outputs of splitting, we choose to define a translation,
+called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
+converts residual tables and partial values into a syntactically
+separated {\em masked value}.  The first component of the masked
+value, $i$, is a value holding the result of all first stage
+computations in $v$.  The second component, $q$, is a residual
+representing all second stage computations in $\rtab \xi v$.
 
@@
 
-%Given the definition of masking, \bbonem\ terms split via the new
-%relation $e \splitonesym \mathcal{P}$, where $\mathcal{P}$ must have
-%the form $\pipeM c l r$
 Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form

### 2015-02-26 paper/splitting.tex
Move 4.0 over to the new notation.

@@
 
-\textbf{Correctness condition for \bbtwo\ term splitting.}
-Intuitively, as suggested in \ref{sec:example}, splitting a
-\lang\ program (a \bbtwo\ term) should yield two monostage terms where
-evaluating the second term (corresponding to second stage
-computations) on the output of the first (first stage computations),
-yields the same result as direct evaluation of the \lang\ program.
-More precisely, for any $\coltwo e A$ which reduces to a residual $q$
-(via $\diatwo e q$), $e$ splits into a syntactically separated program
-$\mathcal{P}$ (via the new relation $e \splittwosym \mathcal{P}$)
-which is reduced to a residual $q'$ (via the new relation $\mathcal{P}
-\sepredtwosym q'$), which is the same as $q$ up to an equivalence.  By
-``syntactically separated'' we mean that $\mathcal{P}$ has the form
-$\pipeS p l r$, where $p$ (called the {\em precomputation}) is a
-monostage term representing all the first-stage computations in $e$,
-and $l.r$ (called the {\em resumer}) is a monostage term representing
-all the second-stage computations in $e$.  $\sepredtwosym$ is defined
-by a single rule, which evaluates the precomputation and substitutes
-the resulting value in for $l$ in the resumer:
-\[
-\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
-\]
-We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
-A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.
+\textbf{Correctness condition for \bbtwo\ term splitting.}  As
+suggested informally in \ref{sec:example}, splitting a \lang\ program
+(a \bbtwo\ term) should yield two monostage terms where evaluating the
+second term (corresponding to second stage computations) on the output
+of the first term (first stage computations), yields the same result
+as direct evaluation of the \lang\ program.  More precisely, for any
+$\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$), $e$
+splits (via the new relation~$\splittwosym$) into a syntactically
+separated program $\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is
+the same as $q$ up to an equivalence. The term $p$ generated by the
+split (called the {\em precomputation}) is a monostage term
+representing all the first-stage computations in $e$, and $l.r$
+(called the {\em resumer}) is a monostage term representing all the
+second-stage computations in $e$.  We refer to the result of the
+precomputation ($b$) the {\em boundary value} of the split since it
+represents the data communicated between the first and second stage
+computations.  The mechanics of splitting \bbtwo\ terms and their
+relationship to the outputs of evaluation is provided at right in
+\ref{fig:splittingSummary}.
+
+% $\sepredtwosym$ is defined
+% by a single rule, which evaluates the precomputation and substitutes
+% the resulting value in for $l$ in the resumer:
+%\[
+%\infer{{[b/l]r}\equiv q} {\reduce p b}
+%\]
 
@@
 challenging because unlike \bbtwo\ terms (which reduce to monostage
-residuals that can be directly compared to the monostage result of
-$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
-tables that represent computations at both stages.
+residuals that can be directly compared to the resumer),
+\bbonem\ terms reduce to partial values and residual tables that
+represent computations at both stages.
 %One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
@@
 to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
-$\mval i q$ (via $\vsplito$).  Then $e$ splits into a syntactically
-seperated program $\mathcal{P}$ (via the new relation $e \splitonesym
-\mathcal{P}$) which is reduced to a masked value $\mval i {q'}$ (via
-the new relation $\sepredtwosym$) where $q'$ is the same as $q$ up to
-an equivalence.
+$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
+relation~$\splitonesym$) into a syntactically separated program
+$\pipeM c l r$ where $\reduce c (i,b)$, and $[b/l]r$ is the same as
+$q$ up to an equivalence.  The term $c$ generated by splitting (called
+the {\em combined term}) is a monostage term encoding all the
+first-stage computations in $e$, and $l.r$ (again called the {\em
+  resumer}) is a monostage term encoding all the second-stage
+computations in $e$.  A summary of splitting \bbonem\ terms is
+provided at left in \ref{fig:splittingSummary}.
 
-Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form
-$\pipeM c l r$ where $c$ (called the {\em combined term}) is a
-monostage term encoding all the first-stage computations in $e$, and
-$l.r$ (again called the {\em resumer}) is a monostage term encoding
-all the second-stage computations in $e$.  $\sepredonesym$ is defined
-by a single rule that yields masked values:
-\[
-\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
-\]
+%$\sepredonesym$ is defined by a single rule that yields masked values:
+%\[
+%\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
+%\]
 
@@
 the second stage {\em and also} produce a first-stage value.
-Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
-generated via splitting reduces only to a boundary value $b$ which is
-passed to the resumer; but in the $\sepredonesym$ rule, the combined
-term $c$ reduces to a tuple containing both the first stage result $i$
-and the boundary value $b$. A summary of the relationship between the
-results of directly evaluating \lang\ and the results of splitting is
-provided at left in \ref{fig:splittingSummary}.
+Correspondingly, the precomputation $p$ generated when splitting
+\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
+the resumer; but the combined term $c$ resulting from \bbtwo\ term
+splitting reduces to a tuple containing both the first stage result
+$i$ and the boundary value $b$ (motivating the name {\em combined
+term}).
 
@@
 
-The masking operation converts a partial value into a masked value $\mval i q$,
+Masking converts a partial value and associated residual table into a masked value $\mval i q$,
 where $i$ is a monostage value encoding all of the first-stage information of the original partial value,

### 2015-02-26 paper/figures/splitting-overview.tex
various symbol changes
changed figure 8 to be smaller and not use sepred

@@
 
-\textbf{Syntactically Separated Term with Stage \bbtwo\ Output ---
-$\pipeS p l r$}
+\textbf{Masking Partial Values ---
+$\rtab \xi v \vsplito \mval i q$}
 \begin{enumerate}
-\item[] Precomputation $p$ is a monostaged term containing the program's
-first-stage computations.
-\item[] Resumer $l.r$ is a monostaged term containing the program's 
-second-stage computations.
-\item[] Evaluates via $\sepredtwo {\pipeS p l r} q$, where the resulting
-residual $q$ is a monostaged term containing the remaining second-stage
-computations.
-\item[] $\sepredtwo {\pipeS p l r} {[b/l]r}$, where $b$ (the boundary value) is
-generated by $\reduce p b$.
-\end{enumerate}
-
-\textbf{Syntactically Separated Term with Multistage Output ---
-$\pipeM c l r$}
-\begin{enumerate}
-\item[] Combined result $c$ is a monostage term containing the program's
-first-stage computations.
-\item[] Resumer $l.r$ is a monostage term containing the program's 
-second-stage computations.
-\item[] Evaluates via $\sepredone {\pipeM c l r} {\mval i q}$, where 
-$\mval i q$ is the resulting masked value (defined below).
-\item[] $\sepredone {\pipeM c l r} {\mval i {[b/l]r}}$, where $b$ (the boundary
-value) is generated by $\reduce c {\tup{i,b}}$.
-\end{enumerate}
-
-\textbf{Masked Values ---
-$\mval i q$}
-\begin{enumerate}
+\item[] Input $\rtab \xi v$ is the residual table and partial value to be masked.
 \item[] Immediate result $i$ is a monostaged value representing the first-stage
@@
 
-\end{abstrsyn}
-\caption{Overview of the output terms of splitting.}
-\label{fig:outputTermSummary}
-\end{figure*}
+\textbf{Splitting Terms at World \bbtwo\ ---
+$\splittwo e A p l r$}
+\begin{enumerate}
+\item[] Input $\coltwo e A$ is the \lang\ program to be split.
+\item[] Precomputation $p$ is a monostaged term containing the program's first-stage computations.
+\item[] Resumer $l.r$ is a monostaged term containing the program's second-stage computations.
+\end{enumerate}
 
+\textbf{Splitting Terms at World \bbonem\ ---
+$\splitone e A c {l.r}$}
+\begin{enumerate}
+\item[] Input $\colmix e A$ is the \lang\ program to be split.
+\item[] Combined result $c$ is a monostage term containing the program's first-stage computations.
+\item[] Resumer $l.r$ is a monostage term containing the program's second-stage computations.
+\end{enumerate}
+
+\hrule
+\vspace{1em}
+
+\centering
 \setlength{\unitlength}{2.8pt}
-\begin{figure*}
-\begin{abstrsyn}
-\centering
 \begin{tabular}{c|c}
-
 {\Large \bf Splitting at World \bbonem} & 
 {\Large \bf Splitting at World \bbtwo} \\ \hline
-
-\begin{minipage}[t]{0.45\textwidth}
-Splitting Terms: $(\colmix e A) \splitonesym \pipeM p l r$, where
-\par \hspace{2em} $\colmix e A$ is the \lang\ input term to be split 
-\par \hspace{2em} $\pipeM p l r$ is the syntactically separated result
-\\
-Masking Partial Values: $[\xi;v] \vsplito \mval i q$, where
-\par \hspace{2em} $[\xi;v]$ is a partial value in \lang, resulting from $\redonesym$
-\par \hspace{2em} $\mval i q$ is the resulting masked value
-\end{minipage}
-&
-
-\begin{minipage}[t]{0.45\textwidth}
-Splitting Terms: $(\coltwo e A) \splittwosym \pipeS p l r$, where 	
-\par \hspace{2em} $\coltwo e A$ is the \lang\ input term to be split
-\par \hspace{2em} $\pipeS p l r$ is the syntactically separated result
-\end{minipage}
-
-\\
 
@@
 \put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
-\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$[\xi;v]$}}}
+\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\rtab \xi v$}}}
 \put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
@@
 \put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splitonesym\right)$}}
-\put(25,32){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
-\put(76,55){\parbox[l]{20 \unitlength}{Separated-Form \\ Evaluation $\left(\sepredonesym\right)$}}
-\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation $\left(\redonesym\right)$}}
+\put(25,33){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
+\put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\$\reduce c {\tup{i,b}}$\\$q\equiv [b/l]r$ }}
+\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation: \\$\diaone e \xi v$}}
 
@@
 \put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splittwosym\right)$}}
-\put(76,55){\parbox[l]{20 \unitlength}{Separated-Form \\ Evaluation $\left(\sepredtwosym\right)$}}
-\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation $\left(\redtwosym\right)$}}
+\put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\ $\reduce p b$\\$q\equiv [b/l]r$}}
+\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation: \\ $\diatwo e q$}}
 
@@
 	produce the same result as 
-	residualization (left and bottom arrow),
+	direct evaluation (left and bottom arrow),
 	up to a term equivalence.

### 2015-02-26 paper/figures/splitting-overview.tex
moved masking to last
figure tweaks

@@
 
-\textbf{Masking Partial Values ---
+\textbf{Splitting Terms at World \bbtwo\ ---
+$\splittwo {\coltwo e A} A p l r$}
+\begin{enumerate}
+\item[] Precomputation $p$ is a monostaged term containing the first-stage computations in $e$.
+\item[] Resumer $l.r$ is a monostaged term containing the second-stage computations in $e$.
+\end{enumerate}
+
+\textbf{Splitting Terms at World \bbonem\ ---
+$\splitone {\colmix e A} A c {l.r}$}
+\begin{enumerate}
+\item[] Combined result $c$ is a monostage term containing the first-stage computations in $e$.
+\item[] Resumer $l.r$ is a monostage term containing the second-stage computations in $e$.
+\end{enumerate}
+
+\textbf{Masking ---
 $\rtab \xi v \vsplito \mval i q$}
@@
 \item[] Input $\rtab \xi v$ is the residual table and partial value to be masked.
-\item[] Immediate result $i$ is a monostaged value representing the first-stage
-result.
-\item[] Residual $q$ is a monostaged term representing the second-stage work
-left to do.
-\end{enumerate}
-
-\textbf{Splitting Terms at World \bbtwo\ ---
-$\splittwo e A p l r$}
-\begin{enumerate}
-\item[] Input $\coltwo e A$ is the \lang\ program to be split.
-\item[] Precomputation $p$ is a monostaged term containing the program's first-stage computations.
-\item[] Resumer $l.r$ is a monostaged term containing the program's second-stage computations.
-\end{enumerate}
-
-\textbf{Splitting Terms at World \bbonem\ ---
-$\splitone e A c {l.r}$}
-\begin{enumerate}
-\item[] Input $\colmix e A$ is the \lang\ program to be split.
-\item[] Combined result $c$ is a monostage term containing the program's first-stage computations.
-\item[] Resumer $l.r$ is a monostage term containing the program's second-stage computations.
+\item[] Immediate value $i$ is a monostage value representing the first-stage components of $v$.
+\item[] Residual $q$ is a monostaged term representing the second-stage computations in $\xi$ and $v$.
 \end{enumerate}
@@
 \begin{tabular}{c|c}
-{\Large \bf Splitting at World \bbonem} & 
-{\Large \bf Splitting at World \bbtwo} \\ \hline
+{\large \bf Splitting at World \bbonem} & 
+{\large \bf Splitting at World \bbtwo} \\
 
@@
 \put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\$\reduce c {\tup{i,b}}$\\$q\equiv [b/l]r$ }}
-\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation: \\$\diaone e \xi v$}}
+\put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\$\diaone e \xi v$}}
 
@@
 \put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\ $\reduce p b$\\$q\equiv [b/l]r$}}
-\put(9,55){\parbox[r]{20 \unitlength}{Interleaved \\ Evaluation: \\ $\diatwo e q$}}
+\put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\ $\diatwo e q$}}
 

### 2015-02-26 paper/splitting.tex
Checkin so I can go home. Make masking text similar to Figure 8.

@@
 of the first term (first stage computations), yields the same result
-as direct evaluation of the \lang\ program.  More precisely, for any
-$\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$), $e$
-splits (via the new relation~$\splittwosym$) into a syntactically
-separated program $\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is
-the same as $q$ up to an equivalence. The term $p$ generated by the
-split (called the {\em precomputation}) is a monostage term
-representing all the first-stage computations in $e$, and $l.r$
-(called the {\em resumer}) is a monostage term representing all the
-second-stage computations in $e$.  We refer to the result of the
-precomputation ($b$) the {\em boundary value} of the split since it
-represents the data communicated between the first and second stage
-computations.  The mechanics of splitting \bbtwo\ terms and their
-relationship to the outputs of evaluation is provided at right in
-\ref{fig:splittingSummary}.
+as direct evaluation of the original \lang\ program.  More precisely,
+for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e
+q$), $e$ splits (via the new relation~$\splittwosym$) into a
+syntactically separated program $\pipeS p l r$ where $\reduce p b$,
+and $[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
+generated by the split (called the {\em precomputation}, due to its
+role producing the inputs for subsequent second stage computation) is
+a monostage term representing all the first-stage computations in
+$e$. The term $l.r$ (called the {\em resumer}) is a monostage term
+representing all the second-stage computations in $e$.  We refer to
+the result of the precomputation ($b$) the {\em boundary value} of the
+split since it represents the data communicated between the first and
+second stage computations.  The mechanics of splitting \bbtwo\ terms
+and their relationship to the outputs of direct \lang\ evaluation is
+provided at right in \ref{fig:splittingSummary}.
 
@@
 challenging because unlike \bbtwo\ terms (which reduce to monostage
-residuals that can be directly compared to the resumer),
-\bbonem\ terms reduce to partial values and residual tables that
-represent computations at both stages.
+residuals that can be directly compared to resumers), \bbonem\ terms
+reduce to residual tables and partial values that represent
+computations at both stages.
 %One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
@@
 converts residual tables and partial values into a syntactically
-separated {\em masked value}.  The first component of the masked
-value, $i$, is a value holding the result of all first stage
-computations in $v$.  The second component, $q$, is a residual
-representing all second stage computations in $\rtab \xi v$.
+separated {\em masked value}.  The first part of the masked value,
+$i$, is a value holding the result of all first stage components of
+$v$.  The second component, $q$, is a residual representing all second
+stage computations in $\xi$ and $v$.
 
@@
 \bbtwo\ terms reduces only to a boundary value $b$ which is passed to
-the resumer; but the combined term $c$ resulting from \bbtwo\ term
+the resumer; but the combined term $c$ resulting from \bbonem\ term
 splitting reduces to a tuple containing both the first stage result
@@
 
-Masking converts a partial value and associated residual table into a masked value $\mval i q$,
-where $i$ is a monostage value encoding all of the first-stage information of the original partial value,
-and $q$ is a monostage term encoding all of the second-stage information of the original partial value.
-Remember that masking will not play a role in the final splitting algorithm, 
-but instead exists to specify the contract to which splitting at \bbonem\ must adhere.
-Accordingly, masking should be thought of as a definition, and as such it is in some parts arbitrary.
+Masking converts a partial value and associated residual table into a
+masked value $\mval i q$, where $i$ is a monostage value encoding all
+first stage components of the partial value, and $q$ is a monostage
+term encoding all of the second-stage computations of the partial
+value.  Remember that masking will not play a role in the final
+splitting algorithm, but instead exists to specify the contract to
+which splitting at \bbonem\ must adhere.  Accordingly, masking should
+be thought of as a definition, and as such it is in some parts
+arbitrary.
 

### 2015-02-26 paper/splitting.tex
A bunch of minor editing operations in 4.0

@@
 
-In this section, we present a \emph{splitting algorithm} which
-statically converts a term in \lang\ into a pair of separate,
-monostaged terms that individually correspond to the first and second
-stage computations of the \lang\ term. Due to important differences in
-structure, we describe both the splitting correctness condition and
+In this section, we present a {\em splitting algorithm} that
+statically converts a term in \lang\ into a pair of monostaged terms
+that individually correspond to the first and second stage
+computations of the original \lang\ term. Due to important differences
+in structure, we describe both the splitting correctness condition and
 splitting algorithm for \bbonem\ terms separately from that of
@@
 for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e
-q$), $e$ splits (via the new relation~$\splittwosym$) into a
-syntactically separated program $\pipeS p l r$ where $\reduce p b$,
-and $[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
+q$), $e$ splits (via the new relation~$\splittwosym$) into the
+syntactically separated terms $\pipeS p l r$ where $\reduce p b$, and
+$[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
 generated by the split (called the {\em precomputation}, due to its
@@
 split since it represents the data communicated between the first and
-second stage computations.  The mechanics of splitting \bbtwo\ terms
-and their relationship to the outputs of direct \lang\ evaluation is
-provided at right in \ref{fig:splittingSummary}.
+second stage computations.  The process of splitting \bbtwo\ terms
+(and the relationship of these operations to the outputs of direct
+\lang\ evaluation) is provided at right in \ref{fig:splittingSummary}.
+% and described in detail in \ref{sec:split-two}.
 
@@
 separated {\em masked value}.  The first part of the masked value,
-$i$, is a value holding the result of all first stage components of
-$v$.  The second component, $q$, is a residual representing all second
-stage computations in $\xi$ and $v$.
+$i$, is a monostage value holding the result of all first stage
+components of $v$.  The second component, $q$, is a residual
+representing all second stage computations in $\xi$ and $v$. (A precise definition of masking is given in \ref{sec:masking}.)
 
@@
 $\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
-relation~$\splitonesym$) into a syntactically separated program
-$\pipeM c l r$ where $\reduce c (i,b)$, and $[b/l]r$ is the same as
-$q$ up to an equivalence.  The term $c$ generated by splitting (called
-the {\em combined term}) is a monostage term encoding all the
-first-stage computations in $e$, and $l.r$ (again called the {\em
-  resumer}) is a monostage term encoding all the second-stage
-computations in $e$.  A summary of splitting \bbonem\ terms is
-provided at left in \ref{fig:splittingSummary}.
+relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
+$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
+equivalence.  The term $c$ generated by splitting (called the {\em
+  combined term}) is a monostage term encoding all the first-stage
+computations in $e$, and $l.r$ (again called the {\em resumer}) is a
+monostage term encoding all the second-stage computations in $e$.  The
+process of splitting \bbonem\ terms is summarized at left in
+\ref{fig:splittingSummary}.
+% and described in detail in \ref{sec:split-one}.
 
@@
 the second stage {\em and also} produce a first-stage value.
-Correspondingly, the precomputation $p$ generated when splitting
+Correspondingly, the precomputation $p$ resulting from splitting
 \bbtwo\ terms reduces only to a boundary value $b$ which is passed to
-the resumer; but the combined term $c$ resulting from \bbonem\ term
-splitting reduces to a tuple containing both the first stage result
-$i$ and the boundary value $b$ (motivating the name {\em combined
-term}).
+the resumer; but the combined term $c$ resulting from splitting
+\bbonem\ terms reduces to a tuple containing both the first stage {\em
+  immediate result} $i$ and the boundary value $b$ (motivating the
+name {\em combined terxm}).
 
@@
 \subsection{Masking}
+\label{sec:masking}
 

### 2015-02-26 paper/splitting.tex
Pass through 4.1
Decided to remove "holes" and just keep text saying they were removed.

@@
 
-  Masking converts a partial value and associated residual table into a
-masked value $\mval i q$, where $i$ is a monostage value encoding all
-first stage components of the partial value, and $q$ is a monostage
-term encoding all of the second-stage computations of the partial
-value.  Remember that masking will not play a role in the final
-splitting algorithm, but instead exists to specify the contract to
-which splitting at \bbonem\ must adhere.  Accordingly, masking should
-be thought of as a definition, and as such it is in some parts
-arbitrary.
-
-So to motivate the definition we chose, consider the following residual table and partial value,
-colored as before:
+Masking converts a residual table and associated partial value into a
+masked value $\mval i q$.
+%that explicitly separates the first stage components of 
+%result from the residual.
+%where $i$ is a monostage value encoding all
+%first stage components of the partial value, and $q$ is a monostage
+%term encoding all of the second-stage computations of the partial
+%value.
+Masking is not a transformation performed as part of the splitting
+algorithm.  Instead, masking defines an intuitive notion of the first
+stage and second stage ``components'' of residual tables and partial
+values, and it is used to define the requirements of what output
+splitting of \bbonem\ terms must produce.  A precise definition of the
+masking relation is given in \ref{fig:valMask}, but to gain intuition
+about the intent of masking operations, consider the following
+residual table and partial value, colored as before:
 \begin{lstlisting}
@@
 \end{lstlisting}
-To reduce this term to just its first-stage information ($i$ in the output), we start by redacting all of the second stage (blue) parts,
-along with the surrounding \texttt{next} annotations:
-\begin{lstlisting}
-4`##########`1` (`4`######`1`, injL (grnd{7}, `4`######`1`))`
-\end{lstlisting}
-The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
+To construct a value representing its first-stage components ($i$ in
+the resulting masked value), masking first redacts all second stage
+(blue) parts, along with the surrounding \texttt{next}
+annotations. (This redaction removes the entire residual table since
+it only represents stage two computations.) The resulting ``holes'' in
+the term are replaced with unit values.
+%\begin{lstlisting}
+%4`##########`1` (`4`######`1`, injL (grnd{7}, `4`######`1`))`
+%\end{lstlisting}
+%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Finally, we drop \texttt{grnd} annotations:
+Finally, masking dropping \texttt{grnd} annotations yielding a value.
 \begin{lstlisting}
@@
 \end{lstlisting}
-To reduce the original partial value to just its second-stage information ($q$ in the output), 
-we start by redacting all \texttt{grnd} blocks, \texttt{next} annotations, and injection tags
-(essentially, the tuple structure is kept, but all other structure is discarded):
-\begin{lstlisting}
-2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
-\end{lstlisting}
-Where the \texttt{grnd} blocks left a hole, we put in a unit values:
+To construct a residual corresponding to the stage two computations
+represented by the residual table and partial value ($q$ in resulting
+the masked value), masking first redacts all \texttt{grnd} blocks
+(replaced with unit), \texttt{next} annotations, and injection tags
+(essentially, the tuple structure is kept, but all other structure is
+discarded):
+%\begin{lstlisting}
+%2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
+%\end{lstlisting}
+%Where the \texttt{grnd} blocks left a hole, we put in a unit values:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Finally, we reify the residual table into \texttt{let} bindings:
+Then masking reifies the residual table into \texttt{let} bindings,
+yielding a residual:
 \begin{lstlisting}
@@
 
-The formal implementation of the masking relation is given in \ref{fig:valMask}.
-Note that since lambdas may represent multistage computations, 
-masking splits the body of lambdas as general world \bbonem\ terms (as described in \ref{sec:split-one}), 
-and packages the resulting terms as functions in both the first and second stage.
+Note in \ref{fig:valMask} that since lambdas may represent multistage
+computations, masking splits the body of lambdas as general world
+\bbonem\ terms (as described in \ref{sec:split-one}), and packages the
+resulting terms as functions in both the first and second stage parts
+of the masked value.
 % Masking operates by first inducting on the entries of the residual table.  

### 2015-02-26 paper/intro.tex

--- 2015-02-25 paper/intro.tex
@@
 automatic stratification turns out to be challenging: the first
+
+Such automatic stratification turns out to be challenging: the first
 successful approach that we know of is by Knoblock and
-Ruf\,\cite{knoblock96}, which like our work was inspired by computer
+Ruf\,\cite{knoblock96}, which, like our work, was inspired by computer
 graphics applications.  While practically effective, Knoblock and Ruf
@@
 point, we take the modal language \langname with explicit staging in
-the style of Davies~\cite{Davies}, where the ``circle'' modality
+the style of Davies~\cite{davies96}, where the ``circle'' modality
 denotes computation in a later stage, and present a staged operational
-semantics similar for \langname (\secref{language}).  We then present
-a splitting algorithm that stratifies a \langname program into a two
-mono-staged programs, one for the first stage and for the second stage
-(\secref{splitting}).  Since each program is at a mono-stage, they are
-expressed in conventional function languages.
+semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
+present a splitting algorithm that stratifies a \langname program into
+a two mono-staged programs, one for the first stage and for the second
+stage (\ref{sec:splitting}).  Since each program is monostaged, they
+are expressed in conventional functional languages.
 
@@
 the algorithm nevertheless can be specified reasonably succinctly.  In
-fact, we have implemented the algorithm (\secref{implementation}) and
-applied to several examples (\secref{examples}).  As an interesting
-example, we consider a function
+fact, we have implemented the algorithm (\ref{sec:implementation}) and
+applied to several examples (\ref{sec:examples}).  As an interesting
+example, we show that a function
+
 \begin{lstlisting}

### 2015-02-26 paper/intro.tex
minor edits to intro

@@
-\newcommand{\langname}{LAMDA12\xspace}
 \section{Introduction}
@@
 separation can be applied without leading to any loss of generality.
-In fact, this is quite natural: going back to the terminology of
-Jorring and Scherlis, pass separation is a frequency reduction,
-whereas as partial evaluation and meta-programming are precomputation
+In fact, this is quite natural---in the terminology of
+J{\o}rring and Scherlis, pass separation is a frequency reduction,
+whereas as partial evaluation and metaprogramming are precomputation
 techniques. Consequently, if desired, partial evaluation can be
@@
 
-But what is the benefit or pass separation? The big benefit is
+But what is the benefit of pass separation? The big benefit is
 efficiency. By stratifying the program, pass separation factors out
@@
 stratification of higher-order functional programs.  As our starting
-point, we take the modal language \langname with explicit staging in
+point, we take the modal language \lang\ with explicit staging in
 the style of Davies~\cite{davies96}, where the ``circle'' modality
@@
 semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
-present a splitting algorithm that stratifies a \langname program into
+present a splitting algorithm that stratifies a \lang\ program into
 a two mono-staged programs, one for the first stage and for the second

### 2015-02-26 paper/splitting.tex
Remove references to "correctness" and "syntatically separated".

@@
 computations of the original \lang\ term. Due to important differences
-in structure, we describe both the splitting correctness condition and
-splitting algorithm for \bbonem\ terms separately from that of
-\bbtwo\ terms.  By construction, \bbonep\ terms are monostage and do
-not require splitting.
+in structure, we describe both the invariants and algorithm for
+splitting \bbonem\ terms separately from that of \bbtwo\ terms.  By
+construction, \bbonep\ terms are monostage and do not require
+splitting.
 
-\textbf{Correctness condition for \bbtwo\ term splitting.}  As
-suggested informally in \ref{sec:example}, splitting a \lang\ program
-(a \bbtwo\ term) should yield two monostage terms where evaluating the
-second term (corresponding to second stage computations) on the output
-of the first term (first stage computations), yields the same result
-as direct evaluation of the original \lang\ program.  More precisely,
-for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e
-q$), $e$ splits (via the new relation~$\splittwosym$) into the
-syntactically separated terms $\pipeS p l r$ where $\reduce p b$, and
-$[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
-generated by the split (called the {\em precomputation}, due to its
-role producing the inputs for subsequent second stage computation) is
-a monostage term representing all the first-stage computations in
-$e$. The term $l.r$ (called the {\em resumer}) is a monostage term
-representing all the second-stage computations in $e$.  We refer to
-the result of the precomputation ($b$) the {\em boundary value} of the
-split since it represents the data communicated between the first and
-second stage computations.  The process of splitting \bbtwo\ terms
-(and the relationship of these operations to the outputs of direct
-\lang\ evaluation) is provided at right in \ref{fig:splittingSummary}.
+\textbf{\bbtwo\ term splitting.}  As suggested informally in
+\ref{sec:example}, splitting a \lang\ program (a \bbtwo\ term) should
+yield two monostage terms where evaluating the second term
+(corresponding to second stage computations) on the output of the
+first term (first stage computations), yields the same result as
+direct evaluation of the original \lang\ program.  More precisely, for
+any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
+$e$ splits (via the new relation~$\splittwosym$) into a pair of terms
+$\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is the same as $q$ up
+to an equivalence. The term $p$ generated by the split (called the
+{\em precomputation}, due to its role producing the inputs for
+subsequent second stage computation) is a monostage term representing
+all the first-stage computations in $e$. The term $l.r$ (called the
+{\em resumer}) is a monostage term representing all the second-stage
+computations in $e$.  We refer to the result of the precomputation
+($b$) the {\em boundary value} of the split since it represents the
+data communicated between the first and second stage computations.
+The process of splitting \bbtwo\ terms (and the relationship of these
+operations to the outputs of direct \lang\ evaluation) is provided at
+right in \ref{fig:splittingSummary}.
 % and described in detail in \ref{sec:split-two}.
@@
 
-\textbf{Correctness condition for \bbonem\ term splitting.} Defining
-the correctness requirements for splitting \bbonem\ terms is
-challenging because unlike \bbtwo\ terms (which reduce to monostage
-residuals that can be directly compared to resumers), \bbonem\ terms
-reduce to residual tables and partial values that represent
-computations at both stages.
+\textbf{\bbonem\ term splitting.} Defining the requirements for
+splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
+(which reduce to monostage residuals that can be directly compared to
+resumers), \bbonem\ terms reduce to residual tables and partial values
+that represent computations at both stages.
 %One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
@@
 called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
-converts residual tables and partial values into a syntactically
-separated {\em masked value}.  The first part of the masked value,
-$i$, is a monostage value holding the result of all first stage
-components of $v$.  The second component, $q$, is a residual
-representing all second stage computations in $\xi$ and $v$. (A precise definition of masking is given in \ref{sec:masking}.)
+converts residual tables and partial values into a {\em masked value}.
+The first part of the masked value, $i$, is a monostage value holding
+the result of all first stage components of $v$.  The second
+component, $q$, is a residual representing all second stage
+computations in $\xi$ and $v$. (A precise definition of masking is
+given in \ref{sec:masking}.)
 
@@
   immediate result} $i$ and the boundary value $b$ (motivating the
-name {\em combined terxm}).
+name {\em combined term}).
 

### 2015-02-26 paper/intro.tex
minor

@@
 Such computations, where we apply a multivariate function to arguments
-where one of the argument remains the same are common. In fact, many
-research problems aim to improve the efficiency of such functions.  For
-example, algorithm design techniques such as dynamic programming
-identify the computation in $f$ that depend on $x$, and try to ensure that
-such computations are performed no more than once.  Similarly, crucial
-compiler optimization techniques such as loop hoisting and common
-subexpression elimination identify common computations across all
-invocations of $f$ and perform them once and for all at the beginning
-of the computation. The whole area of incremental computation aims at
-adapting a computation to small changes, which may be applicable to
-this problem if the $y_i$'s are sufficiently similar.
+where one of the argument remains the same are common.  For example,
+many algorithm-design techniques revolve around the idea of finding
+the computation in $f$ that depends only on $x$, and performing them
+once and for all to use.  Similarly, crucial compiler optimization
+techniques such as loop hoisting and common subexpression elimination
+identify common computations across all invocations of $f$ and perform
+them once and for all at the beginning of the computation.  A
+particular application area where such algorithm-design and code
+optimization techniques are widely applied include real-time computer
+graphics.  There, to achieve high-performance, modern graphics
+architectures requires writing such programs.  Programmers therefore
+manually separate code into distinct {\em passes} based on their
+required frequency of execution.  For example, graphics computations
+are expressed as a sequence of passes where each pass may perform a
+computation at increasingly finer grain, e.g., per object, per region
+of the screen, and per pixel, where later computations use the results
+from the prior computations.
 
-The common theme in all such design and optimization techniques is to
-achieve what in the 1980s J{\o}rring and Scherlis called {\em
-frequency reduction} or {\em precomputation}~\cite{JS86-staging}.
-The idea behind frequency reduction is to identify computations that
-are performed multiple times and hoist them so that they can be
-performed once and used later as needed.  The idea behind
-precomputation is to identify computations that can be performed
-earlier---for example, at compile time if their inputs are known
-statically---and perform them at that earlier time.
+%
+While performant, programming with manually separated passes results
+in complex code where invariants must hold across different passes,
+and local changes for one pass may require changes to other passes to
+``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
+%
+In other words, manual pass separation breaks key programming
+abstractions such as composition and modularity.
+%
+Graphics researchers therefore have suggested using explicitly staged
+programming languages\,\cite{Proudfoot:2001,Foley:2011,He:2014},
+deferring to the compiler the tedious task of separating the code into
+passes. However, all prior efforts have been limited to simple
+languages that do not support recursion or first-class functions.
+
+Programming-languages researchers have studied similar problems. In
+the 1980s, J{\o}rring and Scherlis identified {\em frequency
+  reduction} or {\em precomputation}~\cite{JS86-staging} as the common
+mechanisms for evaluating efficiently staged computations.  The idea
+behind frequency reduction is to identify computations that are
+performed multiple times and hoist them so that they can be performed
+once and used later as needed.  The idea behind precomputation is to
+identify computations that can be performed earlier---for example, at
+compile time if their inputs are known statically---and perform them
+at that earlier time.
+
+There has been much work on language-based techniques for performing
+precomputation. Partial evaluation~\cite{} achieves precomputation by
+specializing functions for specific values.  Going back to our
+example, we can specialize $f$ at $[1,2, \ldots, n]$, written
+$f_{[1,2, \ldots, n]}$, such that $f(x,y) = f_{[1,2, \ldots, n]}(y)$
+and specialize $F$ for the same input as
+\begin{lstlisting}
+fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
+  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
+\end{lstlisting}
+%
+Such a partial evaluation typically proceed by performing a
+binding-time analysis to determine the static and dynamic parts of $f$
+and $F$ and specializes them based on the results of the binding-time
+analysis. A closely related technique, {\em metaprogramming}, enables
+the programmer to write programs that generate specialized code.
+Metaprogramming enables fine-grained control over specialization by
+requiring staging annotations that make explicit the stage of each
+computation.
+
+While previous work shows that precomputation can be performed
+essentially automatically (via partial evaluation or meta
+programming), frequency reduction has been less explored.  In their
+original paper, J{\o}rring and Scherlis proposed {\em pass separation}
+as a manual method for {\em splitting} a program into multiple passes.
+%
+Applied to our example, pass separation would split the function $f$
+into $f_1$ and $f_2$ such that $f(x,y) = f_2(f_1(x),y)$.  We can then
+rewrite our example as a multi-pass program:
+%
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
+  let @$fx = f_1(x)$@
+  in @$(f_2(fx, y_0);  f_2(fx, y_1)  \ldots f_2(fx, y_{m-1}))$@.
+\end{lstlisting}
+%
+The key difference between partial evaluation (meta programming) and
+pass separation is that the latter does not require the first
+argument~$x$: it works with all arguments, whereas the partially
+evaluated program is specialized for a particular value of $x$ ($x =
+[1,2, \ldots, n]$ in the example). Specifically, if desired, partial
+evaluation can be applied subsequently in our example, by partially
+evaluating function~$f_2$ on $f_1(x)$.
+
+However, unlike for precomputation, there is no available technique
+for pass separation except for very simple languages
+\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.  Notably omitted
+features include recursion and first-class functions.  We are thus
+interested in the problem of splitting automatically a functional
+language.
+%
+
+As our starting point, we take the modal language \lang\ with explicit
+staging in the style of Davies~\cite{davies96}, where the ``circle''
+modality denotes computation in a later stage, and present a staged
+operational semantics similar for \lang\ (\ref{sec:stagedsemantics}).
+We then present a splitting algorithm that stratifies a \lang\ program
+into a two mono-staged programs, one for the first stage and for the
+second stage (\ref{sec:splitting}).  Since each program is monostaged,
+they are expressed in conventional functional languages.
+
+The crux of the splitting algorithm is splitting recursive mixed stage
+functions that involve both a first stage and second stage
+computations.  When split, such a function turns into a stratum-1
+function that yields a {\em boundary value} that encodes the
+``first-stage'' part of the computation.  If the function recurs on a
+first-stage argument, then the boundary value is typically a recursive
+data structure.  In other words, the splitting algorithm maps
+computational recursion into a recursive data type. If the boundary
+structure is generated as a result of a conditional, it is typically
+encoded as a (recursive) sum type, such as a tree, where the kind of
+the node indicates the recursion status of the first-stage evaluation.
+The stratum-2 part of the computation, which corresponds to the
+``second stage'' takes the boundary value as an argument as well as
+the second-stage argument and uses them to complete the computation,
+often by traversing the now precomputed boundary in light of the now
+available second-stage argument.
+
+Since it deals with unrestricted recursion and a rich language
+consisting of first-class functions, recursion, and sum and recursive
+types, the splitting algorithm is naturally complex. With some care,
+the algorithm nevertheless can be specified reasonably succinctly.  In
+fact, we have implemented the algorithm (\ref{sec:implementation}) and
+applied to several examples (\ref{sec:examples}).  As an interesting
+example, we show that a function
+
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
+\end{lstlisting}
+where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
+using the ``quickselect'' algorithm.  When applied to this function,
+our split algorithm yields an asymptotically more efficient function
+that runs in expected $O(n\log{n} + m\log{n})$ time instead of
+expected $O(n \cdot m)$ time---a near linear time improvement.
+Interestingly the code output by our splitting algorithm
+auto-generates the ``quicksort'' algorithm, which it uses to build a
+binary search tree as a boundary value in the first stage.  For the
+second stage, the splitting algorithm generates a ``binary search''
+that takes the tree and performs a binary search on the tree as guided
+by the given rank.  To the best of our knowledge, no prior approaches
+can perform such complex transformations on higher-order code, nor can
+they yield such asymptotic improvements in run time.
+
+\begin{comment}
+\section{Introduction}
+
+Consider a function that takes as arguments an $x$ and $y_1 \ldots
+y_{m-1}$ and invokes a function $f(\cdot, \cdot)$ on consecutive
+values of $y$'s.  Using ``;'' for  sequential composition, we can
+write such a function as
+\begin{lstlisting}
+fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
+  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@.
+\end{lstlisting}
+%
+%% Let's assume for concreteness that $f(x,y)$ takes time to the size of
+%% the first argument.  The function $F(x_0, y_0 \ldots y_{m-1}$ then
+%% takes time $O(n \cdot m)$ where $n$ is the size of the input $x_0$.
+
+Such computations, where we apply a multivariate function to arguments
+where one of the argument remains the same are common.  For example,
+many algorithm-design techniques revolve around the idea of finding
+the computation in $f$ that depends only on $x$, and performing them
+once and for all to use.  Similarly, crucial compiler optimization
+techniques such as loop hoisting and common subexpression elimination
+identify common computations across all invocations of $f$ and perform
+them once and for all at the beginning of the computation.  A
+particular application area where such algorithm-design and code
+optimization techniques are widely applied include real-time computer
+graphics.  There, to achieve high-performance, programmers manually
+separate code into distinct {\em passes} based on their required
+frequency of execution.  For example, graphics computations are often
+expressed as a sequence of passes where each pass may perform a
+computation in increasingly finer grain, e.g., per object, per region
+of the screen, and per pixel, where later computations use the results
+from the prior computations.
+
+%
+While performant, this programming style of programming with manually
+separated passes results in complex code where invariants must hold
+across different passes, and local changes for one pass may require
+changes to other passes to ``plumb'' inputs from all other
+frequencies\,\cite{Foley:2011}.
+%
+In other words, manual pass separation breaks key programming
+abstractions such as composition and modularity that we hold dear.
+%
+Graphics researchers therefore have suggested writing graphics
+programs in explicitly staged programming
+languages\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring the
+tedious task of separating the code into passes as needed for
+performance to the compiler. However, all prior efforts have been
+limited to simple languages that do not support recursion or
+first-class functions.
+
+Motivated by the crucial need for improved abstractions for performant
+multi-stage programs, we investigate in this paper, the problem of
+automatically splitting a higher-order program into a collection of
+programs each of which gets executed in a single pass.
+%
+This is not a new problem.
+%
+Common theme in this problem domain is what in the 1980s J{\o}rring
+and Scherlis called {\em frequency reduction} or {\em
+  precomputation}~\cite{JS86-staging}.  The idea behind frequency
+reduction is to identify computations that are performed multiple
+times and hoist them so that they can be performed once and used later
+as needed.  The idea behind precomputation is to identify computations
+that can be performed earlier---for example, at compile time if their
+inputs are known statically---and perform them at that earlier time.
 
@@
 {\em static} and {\em dynamic} stages, and performing all static
-computations first, followed by dynamic ones.  For example,
-suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
-statically known $x = [~]$.  We can derive a specialization of $f$ at
-$[~]$, written $f_{[~]}$, such that $f(x,y) = f_{[~]}(y)$ and specialize $F$
-for $x=[~]$ as
+computations first, followed by dynamic ones.  For example, suppose
+that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a statically
+known $x = [1,2, \ldots, n]$.  We can derive a specialization of $f$
+at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots, n]}$, such that
+$f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for $x=[~]$ as
 \begin{lstlisting}
-fun @$F_{[~]}(y_0, y_1 \ldots y_{m-1})$@ = 
-  @$(f_{[~]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
+fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
+  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
 \end{lstlisting}
@@
 Applied to our example, pass separation would {\em split} the function
-$f$ into $f_s$ (static) and $f_d$ (dynamic) such that $f(x,y) =
-f_d(f_s(x),y)$.  We can then rewrite our example as
+$f$ into $f_s$ and $f_d$, corresponding to the static and dynamic
+passes, such that $f(x,y) = f_d(f_s(x),y)$.  We can then rewrite our
+example as
 %
@@
 \end{lstlisting}
-We refer to such a program as {\em stratified} program.
+We refer to such a program as a {\em multi-pass} program.
 
@@
 arguments, whereas the partially evaluated program is specialized for
-a particular value of $x$ ($x = [~]$ in the example).  Thus, the pass
-separation can be applied without leading to any loss of generality.
-In fact, this is quite natural---in the terminology of
+a particular value of $x$ ($x = [1,2, \ldots, n]$ in the example).
+Thus, the pass separation can be applied without leading to any loss
+of generality.  In fact, this is quite natural---in the terminology of
 J{\o}rring and Scherlis, pass separation is a frequency reduction,
@@
 techniques. Consequently, if desired, partial evaluation can be
-orthogonally applied to a stratified program.
+orthogonally applied to a multi-pass program.
+%
+The crucial benefit of pass-separation is efficiency: separating the
+passes in the program factors out frequently used the parts of
+computation, computes them first, and then reuse them later instead of
+computing them again. This is precisely the reason for why it is
+employed broadly in graphics software even though it results in
+code that is difficult to design, implement, and maintain.
 
-% Let's assume for concreteness that $f(x,y)$ takes time in the size of
-% the first argument.  The function $F(x, y_0 \ldots y_{m-1}$ then takes
-% time $O(n \cdot m)$ where $n$ is the size of $x$. Assume further that
-% in the stratified code the static function $f_s(x)$ takes $O(n)$ time
-% and the $f_d(fsx, \cdot)$ function takes $O(\log{n})$ time.  With
-% these run times, the stratified code has complexity $O(n + m\log{n})$
-% instead of $O(n \cdot m)$, a significant asymptotic improvement.  (In
-% this paper, we will achieve such a reduction in complexity
-% automatically.)
-
-But what is the benefit of pass separation? The big benefit is
-efficiency. By stratifying the program, pass separation factors out
-frequently used the parts of computation, computes them first, and
-then reuse them later instead of computing them again.  Since it can
-be applied statically with no loss of generality, pass separation can
-significantly improve performance.  In fact, in computer graphics,
-program are typically manually stratified both for improved
-efficiency. Such stratification is in fact required by the execution
-model of some modern graphics architectures.
-
-While highly desirable, manual stratification is difficult and highly
-nonmodular: program invariants are scattered over the code written in
-multiple strata and because modification to any one stratum requires
-modifying all others to match.  Ideally, we wish to write just one
-program and stratify it into multiple strata automatically.  But such
-automatic stratification turns out to be challenging: the first
-
-Such automatic stratification turns out to be challenging: the first
-successful approach that we know of is by Knoblock and
-Ruf\,\cite{knoblock96}, which, like our work, was inspired by computer
-graphics applications.  While practically effective, Knoblock and Ruf
-consider only a very simple language and are not able to stratify
-recursive programs.  Recent
+Thus, we are interested in the question of whether it would be
+possible to obtain the best of the both worlds, by automatically
+pass-separating programs.  Such automatic stratification has been
+shown to be challenging. The first successful approach that we know of
+is by Knoblock and Ruf\,\cite{knoblock96}, which, like our work, was
+inspired by computer graphics applications.  While practically
+effective, Knoblock and Ruf consider only a very simple language and
+are not able to stratify recursive programs.  Recent
 efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
 cannot stratify recursive programs.
-%
-In fact back in 80,s Jorring and Sherlis was able to stratify
-reasonably complex programs but only manually and note that
-``... partial evaluation technique is readily mechanized, while
-staging transformation approach will elude full automation for some
-time.''
 
 In this paper we make significant progress on the problem of automatic
-stratification of higher-order functional programs.  As our starting
-point, we take the modal language \lang\ with explicit staging in
-the style of Davies~\cite{davies96}, where the ``circle'' modality
-denotes computation in a later stage, and present a staged operational
-semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
-present a splitting algorithm that stratifies a \lang\ program into
-a two mono-staged programs, one for the first stage and for the second
-stage (\ref{sec:splitting}).  Since each program is monostaged, they
-are expressed in conventional functional languages.
+stratification of programs expressed in a rich higher-order functional
+language.  As our starting point, we take the modal language \lang\
+with explicit staging in the style of Davies~\cite{davies96}, where
+the ``circle'' modality denotes computation in a later stage, and
+present a staged operational semantics similar for \lang\
+(\ref{sec:stagedsemantics}).  We then present a splitting algorithm
+that stratifies a \lang\ program into a two mono-staged programs, one
+for the first stage and for the second stage (\ref{sec:splitting}).
+Since each program is monostaged, they are expressed in conventional
+functional languages.
 
@@
 transforms a variety of other programs.
+\end{comment}

### 2015-02-26 paper/intro.tex
uncomment splitting section

@@
 
-\begin{comment}
-\section{Introduction}
-
-Consider a function that takes as arguments an $x$ and $y_1 \ldots
-y_{m-1}$ and invokes a function $f(\cdot, \cdot)$ on consecutive
-values of $y$'s.  Using ``;'' for  sequential composition, we can
-write such a function as
-\begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
-  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@.
-\end{lstlisting}
-%
-%% Let's assume for concreteness that $f(x,y)$ takes time to the size of
-%% the first argument.  The function $F(x_0, y_0 \ldots y_{m-1}$ then
-%% takes time $O(n \cdot m)$ where $n$ is the size of the input $x_0$.
-
-Such computations, where we apply a multivariate function to arguments
-where one of the argument remains the same are common.  For example,
-many algorithm-design techniques revolve around the idea of finding
-the computation in $f$ that depends only on $x$, and performing them
-once and for all to use.  Similarly, crucial compiler optimization
-techniques such as loop hoisting and common subexpression elimination
-identify common computations across all invocations of $f$ and perform
-them once and for all at the beginning of the computation.  A
-particular application area where such algorithm-design and code
-optimization techniques are widely applied include real-time computer
-graphics.  There, to achieve high-performance, programmers manually
-separate code into distinct {\em passes} based on their required
-frequency of execution.  For example, graphics computations are often
-expressed as a sequence of passes where each pass may perform a
-computation in increasingly finer grain, e.g., per object, per region
-of the screen, and per pixel, where later computations use the results
-from the prior computations.
-
-%
-While performant, this programming style of programming with manually
-separated passes results in complex code where invariants must hold
-across different passes, and local changes for one pass may require
-changes to other passes to ``plumb'' inputs from all other
-frequencies\,\cite{Foley:2011}.
-%
-In other words, manual pass separation breaks key programming
-abstractions such as composition and modularity that we hold dear.
-%
-Graphics researchers therefore have suggested writing graphics
-programs in explicitly staged programming
-languages\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring the
-tedious task of separating the code into passes as needed for
-performance to the compiler. However, all prior efforts have been
-limited to simple languages that do not support recursion or
-first-class functions.
-
-Motivated by the crucial need for improved abstractions for performant
-multi-stage programs, we investigate in this paper, the problem of
-automatically splitting a higher-order program into a collection of
-programs each of which gets executed in a single pass.
-%
-This is not a new problem.
-%
-Common theme in this problem domain is what in the 1980s J{\o}rring
-and Scherlis called {\em frequency reduction} or {\em
-  precomputation}~\cite{JS86-staging}.  The idea behind frequency
-reduction is to identify computations that are performed multiple
-times and hoist them so that they can be performed once and used later
-as needed.  The idea behind precomputation is to identify computations
-that can be performed earlier---for example, at compile time if their
-inputs are known statically---and perform them at that earlier time.
-
-There has been much work on language-based techniques for performing
-frequency reduction and precomputation.  Broadly speaking, these
-techniques can be categorized into
-\begin{itemize}
-\item partial evaluation,
-\item metaprogramming, and
-\item pass separation or data specialization.
-\end{itemize}
-
-Partial evaluation achieves precomputation by distinguishing between
-{\em static} and {\em dynamic} stages, and performing all static
-computations first, followed by dynamic ones.  For example, suppose
-that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a statically
-known $x = [1,2, \ldots, n]$.  We can derive a specialization of $f$
-at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots, n]}$, such that
-$f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for $x=[~]$ as
-\begin{lstlisting}
-fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
-  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
-\end{lstlisting}
-%
-Such a partial evaluation typically proceed by performing a
-binding-time analysis to determine the static and dynamic parts of $f$
-and $F$ and specializes them based on the results of the binding-time
-analysis.
-
-A closely related technique, {\em metaprogramming}, enables the
-programmer to write a program that takes the static values and
-generates a specialized program for those values.  Metaprogramming
-gives the programmer relatively fine-grained control over partial
-evaluation by requiring staging annotations in the program.  Such
-staging annotations make explicit the stage of each computation and
-can overcome some of the intricacies of binding-time analysis but this
-comes at the cost of more complex reasoning and some loss of
-automation (partial evaluation can in principle be applied fully
-automatically without relying on any annotations).
-
-The third and final approach, called {\em pass separation} or {\em
-  data specialization}, assumes much less information: instead of
-requiring the static data, it requires just the knowledge of which
-computations are static and dynamic and uses this knowledge to
-generate two stage computations. 
-%
-Applied to our example, pass separation would {\em split} the function
-$f$ into $f_s$ and $f_d$, corresponding to the static and dynamic
-passes, such that $f(x,y) = f_d(f_s(x),y)$.  We can then rewrite our
-example as
-%
-\begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
-let @$fx = f_s(x)$@
-in @$(f_d(fx, y_0);  f_d(fx, y_1)  \ldots f_d(fx, y_{m-1}))$@.
-\end{lstlisting}
-We refer to such a program as a {\em multi-pass} program.
-
-The key difference with partial evaluated program and the stratified
-program is that the latter has lost no generality: it works with all
-arguments, whereas the partially evaluated program is specialized for
-a particular value of $x$ ($x = [1,2, \ldots, n]$ in the example).
-Thus, the pass separation can be applied without leading to any loss
-of generality.  In fact, this is quite natural---in the terminology of
-J{\o}rring and Scherlis, pass separation is a frequency reduction,
-whereas as partial evaluation and metaprogramming are precomputation
-techniques. Consequently, if desired, partial evaluation can be
-orthogonally applied to a multi-pass program.
-%
-The crucial benefit of pass-separation is efficiency: separating the
-passes in the program factors out frequently used the parts of
-computation, computes them first, and then reuse them later instead of
-computing them again. This is precisely the reason for why it is
-employed broadly in graphics software even though it results in
-code that is difficult to design, implement, and maintain.
-
-Thus, we are interested in the question of whether it would be
-possible to obtain the best of the both worlds, by automatically
-pass-separating programs.  Such automatic stratification has been
-shown to be challenging. The first successful approach that we know of
-is by Knoblock and Ruf\,\cite{knoblock96}, which, like our work, was
-inspired by computer graphics applications.  While practically
-effective, Knoblock and Ruf consider only a very simple language and
-are not able to stratify recursive programs.  Recent
-efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
-cannot stratify recursive programs.
-
-In this paper we make significant progress on the problem of automatic
-stratification of programs expressed in a rich higher-order functional
-language.  As our starting point, we take the modal language \lang\
-with explicit staging in the style of Davies~\cite{davies96}, where
-the ``circle'' modality denotes computation in a later stage, and
-present a staged operational semantics similar for \lang\
-(\ref{sec:stagedsemantics}).  We then present a splitting algorithm
-that stratifies a \lang\ program into a two mono-staged programs, one
-for the first stage and for the second stage (\ref{sec:splitting}).
-Since each program is monostaged, they are expressed in conventional
-functional languages.
-
-The crux of the splitting algorithm is splitting recursive mixed stage
-functions that involve both a first stage and second stage
-computations.  When split, such a function turns into a stratum-1
-function that yields a {\em boundary value} that encodes the
-``first-stage'' part of the computation.  If the function recurs on a
-first-stage argument, then the boundary value is typically a recursive
-data structure.  In other words, the splitting algorithm maps
-computational recursion into a recursive data type. If the boundary
-structure is generated as a result of a conditional, it is typically
-encoded as a (recursive) sum type, such as a tree, where the kind of
-the node indicates the recursion status of the first-stage evaluation.
-The stratum-2 part of the computation, which corresponds to the
-``second stage'' takes the boundary value as an argument as well as
-the second-stage argument and uses them to complete the computation,
-often by traversing the now precomputed boundary in light of the now
-available second-stage argument.
-
-Since it deals with unrestricted recursion and a rich language
-consisting of first-class functions, recursion, and sum and recursive
-types, the splitting algorithm is naturally complex. With some care,
-the algorithm nevertheless can be specified reasonably succinctly.  In
-fact, we have implemented the algorithm (\ref{sec:implementation}) and
-applied to several examples (\ref{sec:examples}).  As an interesting
-example, we show that a function
-
-\begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
-  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
-\end{lstlisting}
-where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
-using the ``quickselect'' algorithm.  When applied to this function,
-our split algorithm yields an asymptotically more efficient function
-that runs in expected $O(n\log{n} + m\log{n})$ time instead of
-expected $O(n \cdot m)$ time---a near linear time improvement.
-Interestingly the code output by our splitting algorithm
-auto-generates the ``quicksort'' algorithm, which it uses to build a
-binary search tree as a boundary value in the first stage.  For the
-second stage, the splitting algorithm generates a ``binary search''
-that takes the tree and performs a binary search on the tree as guided
-by the given rank.  To the best of our knowledge, no prior approaches
-can perform such complex transformations on higher-order code, nor can
-they yield such asymptotic improvements in run time.
-
 \paragraph{Old intro below.}
@@
 transforms a variety of other programs.
-\end{comment}

### 2015-02-26 paper/splitting.tex
Check in of preliminary 2-term intro to splitting.  Now working on 1M

@@
 
-\input{figures/splitting-overview}
+%\input{figures/splitting-overview}
 
@@
 
-%Although the type system of \lang\ ensures that we can perform all first-stage
-%computations before any second-stage ones, actually evaluating terms in this
-%fashion is rather involved, because stages are syntactically interleaved
-%within terms.
-
-In this section, we present a {\em splitting algorithm} that
-statically converts a term in \lang\ into a pair of monostaged terms
-that individually correspond to the first and second stage
-computations of the original \lang\ term. Due to important differences
-in structure, we describe both the invariants and algorithm for
-splitting \bbonem\ terms separately from that of \bbtwo\ terms.  By
+Splitting statically transforms terms in \lang\ into a pair of
+monostaged terms that correspond to the first stage and second stage
+computations in the original term. The idea is that evaluating the
+first monostaged term in a first pass, and then evaluating the second
+monostaged term on the results of the first pass, yields the same
+result as fully evaluating the original \lang\ program.  Complexity
+arises because \lang\ evaluation of \bbonem\ terms may also generate
+results in the first stage. (For example, a \bbonem\ term of type
+$\curr \rmint$ results in an \rmint\ in the first stage.) Handling the
+asymmetry between the results of evaluating \bbonem\ terms (which may
+produce results at both the first and second stage) and \bbtwo\ terms
+(which only produce second stage results) is fundamental to the
+structure of the splitting algorithm presented in this section.  By
 construction, \bbonep\ terms are monostage and do not require
@@
 
-\textbf{\bbtwo\ term splitting.}  As suggested informally in
-\ref{sec:example}, splitting a \lang\ program (a \bbtwo\ term) should
-yield two monostage terms where evaluating the second term
-(corresponding to second stage computations) on the output of the
-first term (first stage computations), yields the same result as
-direct evaluation of the original \lang\ program.  More precisely, for
-any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
-$e$ splits (via the new relation~$\splittwosym$) into a pair of terms
-$\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is the same as $q$ up
-to an equivalence. The term $p$ generated by the split (called the
-{\em precomputation}, due to its role producing the inputs for
-subsequent second stage computation) is a monostage term representing
-all the first-stage computations in $e$. The term $l.r$ (called the
-{\em resumer}) is a monostage term representing all the second-stage
-computations in $e$.  We refer to the result of the precomputation
-($b$) the {\em boundary value} of the split since it represents the
-data communicated between the first and second stage computations.
-The process of splitting \bbtwo\ terms (and the relationship of these
-operations to the outputs of direct \lang\ evaluation) is provided at
-right in \ref{fig:splittingSummary}.
-% and described in detail in \ref{sec:split-two}.
+\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
+to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
+$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
+yields a pair of monostaged terms $p$ (containing first stage
+computations in $e$) and $l.r$ (containing second stage computations
+in $e$). Executing $p$ in a first pass produces $b$ (via $\reduce p
+b$). Then, in a second pass, $\reduce {[b/l]r} {v'} = v$.  Since
+execution of the first pass serves to generate input for the second
+pass, we say $p$ is a {\em precomputation} that produces a {\em
+  boundary value} ($b$) that is communicated to the {\em resumer}
+($l.r$).
+
+The process of splitting \bbtwo\ terms is given by via the new
+relation~$\splittwosym$.
+
+% (and the relationship of these
+% operations to the outputs of direct \lang\ evaluation) is provided at
+% right in \ref{fig:splittingSummary}.
 
@@
 
-\textbf{\bbonem\ term splitting.} Defining the requirements for
-splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
-(which reduce to monostage residuals that can be directly compared to
-resumers), \bbonem\ terms reduce to residual tables and partial values
-that represent computations at both stages.
-%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
-%For example, the \bbonem\ term:
-%\begin{lstlisting}
-%(next {1+2}, grnd{3+4})
-%\end{lstlisting}
-%reduces (via $\redonesym$) to the residual table and partial value:
-%\begin{lstlisting}
-%[yhat |-> 1+2] (next {yhat}, grnd{7})
-%\end{lstlisting}
+\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
+to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
+$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
+relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
+$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
+equivalence.  The term $c$ generated by splitting (called the {\em
+  combined term}) is a monostage term encoding all the first-stage
+computations in $e$, and $l.r$ (again called the {\em resumer}) is a
+monostage term encoding all the second-stage computations in $e$.  The
+process of splitting \bbonem\ terms is summarized at left in
+\ref{fig:splittingSummary}.
+
+%Defining the requirements for
+%splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
+%(which reduce to monostage residuals that can be directly compared to
+%resumers), \bbonem\ terms reduce to residual tables and partial values
+%that represent computations at both stages.
 
@@
 
-We define the requirements for splitting \bbonem\ terms using the
-results of masking. Specifically, for any $\colmix e A$ which reduces
-to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
-$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
-relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
-$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
-equivalence.  The term $c$ generated by splitting (called the {\em
-  combined term}) is a monostage term encoding all the first-stage
-computations in $e$, and $l.r$ (again called the {\em resumer}) is a
-monostage term encoding all the second-stage computations in $e$.  The
-process of splitting \bbonem\ terms is summarized at left in
-\ref{fig:splittingSummary}.
-% and described in detail in \ref{sec:split-one}.
-
-%$\sepredonesym$ is defined by a single rule that yields masked values:
-%\[
-%\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
-%\]
+% We define the requirements for splitting \bbonem\ terms using the
+% results of masking.
 

### 2015-02-26 paper/splitting.tex
More tweaks to 2-splitting.  Now actually moving on to 1M

@@
 $\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
-yields a pair of monostaged terms $p$ (containing first stage
-computations in $e$) and $l.r$ (containing second stage computations
-in $e$). Executing $p$ in a first pass produces $b$ (via $\reduce p
-b$). Then, in a second pass, $\reduce {[b/l]r} {v'} = v$.  Since
-execution of the first pass serves to generate input for the second
-pass, we say $p$ is a {\em precomputation} that produces a {\em
-  boundary value} ($b$) that is communicated to the {\em resumer}
-($l.r$).
+yields a pair of monostaged terms $p$ (containing all first stage
+computations in $e$) and $l.r$ (containing all second stage
+computations in $e$). Executing $p$ in a first pass produces $b$ (via
+$\reduce p b$). Then, in the second pass, $\reduce {[b/l]r} {v'} = v$.
+Since execution of the first pass serves to generate input for the
+second pass, we say $p$ is a {\em precomputation} that produces a {\em
+  boundary value} ($b$) for the {\em resumer} ($l.r$).
 
-The process of splitting \bbtwo\ terms is given by via the new
-relation~$\splittwosym$.
+% The process of splitting \bbtwo\ terms is given by via the new
+% relation~$\splittwosym$.
 

### 2015-02-26 paper/splitting.tex
rewriting splitting rules explanation

--- 2015-02-26 paper/splitting.tex
@@
 
-When $e$ is unit or a variable,
-splitting yields a combined term formed by tupling $e$ with a $\tup{}$ precomputation.
-These expressions, as they contain no second-stage subcomputations, have a trivial resumer, $\tup{}$.
-Similarly, the contents of a \texttt{grnd} block are assigned entirely to the immediate value,
-with trivial precomputation and resumer.
+We start with splitting the unit value, $\tup{}$.
+Masking tells us that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
+Plugging this in to the contract for splitting at \bbonem\ gives:
+\[
+	\text{if } \splitone {\tup {}} A c {l.r} \text{ then } \reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
+\]
+Although the only hard requirement for $b$, $c$, and $l.r$ is to satisfy these relations,
+our actual choices are designed to do the least work possible.
+In particular, we choose $b = \tup{}$; $c=\tup{\tup{},\tup{}}$; and $l.r = \_.\tup{}$.
+That is, we choose a trivial boundary value.
+All together, this yield the rule in \cref{fig:termSplit}.
+This exercise proceeds much the same way for \texttt{grnd} blocks, 
+with the purely first stage contents of the block standing in for one of the $\tup{}$s.
 
-For all non-terminals (except \texttt{next}),
-splitting descends into $e$, recursively splitting its $n$ subterms
-to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
-The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
-to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
-to isolate immediate results from boundary values. Then,
-the immediate result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
-The resumer binds the boundary values $b_1,\ldots,b_n$ to an
-argument $(l_1,\ldots,l_n)$ in a term that has the same structure
-of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
+We next consider splitting a non-terminal, namely $\inl {e}$.
+From the semantics, we already know that 
+for some $\xi$ and $v$, $\diaone {e} \xi v$ and $\diaone {\inl e} \xi {\inl v}$.
+Masking these outputs gives $\rtab \xi v \vsplito \mval i q$ 
+and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$.
+Since splitting works from the inside out, we already know that
+\[
+	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
+\]
+and what we need to do is find some $b'$, $c'$, and $l'.r'$ satisfying
+\[
+	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l']r' \equiv q
+\]
+Once again, we take the simplest solution, which in this case is $b' = b$; $l'.r' = l.r$;
+and $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$.  
+This sort of technique generalize to all constructs which don't involve control flow.
+
+\nr{working from here down}
 

### 2015-02-27 paper/appendix.tex

\section{Additional examples}

Here we include some additional examples of our splitting algorithm.

\subsection{Trie}

\crem{pending Nico}

\subsection {Mixed Map Combinator}

In \ref{sec:exampleQS}, we suggested that staged quickselect could be applied to one list and many queries.
In this example, we implement that behavior by passing \texttt{qs} to a higher order function.
Such a combinator, which operates over both stages at once, is possible because of the generality of \lang's abstractions.

Note that we have to define a datatype for integer lists at the second stage, which is given by \texttt{list2}.
\begin{lstlisting} 
1`atsignnext{`2`
  datatype list2 = Empty2 | Cons2 of int * list2
`1`}

type qsType = ^list*$`2`int`1`->$`2`int`1`

// map : qsType -> ^list * $`2`list2`1` -> $`2`list2`1`
fun map f (l, q) = 
next {`2`
  let 
  fun m Empty2 = Empty2
    | m (Cons2(h,t)) = Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
  in m prev{`1`q`2`}
`1`}

val qsMany = map qs`
\end{lstlisting}
The \texttt{map} function splits into the following two functions:
\begin{lstlisting} 
2`datatype list2 = Empty2 | Cons2 of int * list2

`1`fun map1 f = (fn (l,()) => ((), #2 (f (l, ()))), ())
`2`fun map2 (f,()) ((l,q), p : tree) =
  let 
  fun m Empty2 = Empty2
    | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
  in m q`
\end{lstlisting}

As desired, \texttt{qs1} (which is passed to \texttt{map1} via variable \texttt{f}) is executed only once per invocation of \texttt{map},
whereas \texttt{qs2} (which is passed to \texttt{map2} via variable \texttt{f}) is executed once per query.

### 2015-02-27 paper/intro.tex

@@
 
-Consider a function that takes as arguments an $x$ and $y_1 \ldots
-y_{m-1}$ and invokes a function $f(\cdot, \cdot)$ on consecutive
-values of $y$'s.  Using ``;'' for  sequential composition, we can
-write such a function as
+Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, in order to
+compute $f(x,y_i)$ for each $y_i$:
 \begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
-  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@.
+fun F(x, @$y_0$@, @$y_1$@, @$\ldots$@ @$y_{m-1}$@) = 
+  f(x, @$y_0$@);  f(x, @$y_1$@); @$\ldots$@ f(x, @$y_{m-1}$@)
 \end{lstlisting}
 %
-%% Let's assume for concreteness that $f(x,y)$ takes time to the size of
-%% the first argument.  The function $F(x_0, y_0 \ldots y_{m-1}$ then
-%% takes time $O(n \cdot m)$ where $n$ is the size of the input $x_0$.
+Such computations, in which a multivariate function is applied many times with
+one argument fixed, are common. Many algorithm design techniques revolve around
+the idea of finding the computations in $f$ that depend only on $x$, and
+performing them only once. Similarly, compiler optimization techniques like loop
+hoisting and common subexpression elimination serve to identify the computations
+common to all invocations of $f$, and perform them a single time at the
+beginning of the computation.
 
-Such computations, where we apply a multivariate function to arguments
-where one of the argument remains the same are common.  For example,
-many algorithm-design techniques revolve around the idea of finding
-the computation in $f$ that depends only on $x$, and performing them
-once and for all to use.  Similarly, crucial compiler optimization
-techniques such as loop hoisting and common subexpression elimination
-identify common computations across all invocations of $f$ and perform
-them once and for all at the beginning of the computation.  A
-particular application area where such algorithm-design and code
-optimization techniques are widely applied include real-time computer
-graphics.  There, to achieve high-performance, modern graphics
-architectures requires writing such programs.  Programmers therefore
-manually separate code into distinct {\em passes} based on their
-required frequency of execution.  For example, graphics computations
-are expressed as a sequence of passes where each pass may perform a
-computation at increasingly finer grain, e.g., per object, per region
-of the screen, and per pixel, where later computations use the results
-from the prior computations.
+These techniques are particularly essential in real-time computer graphics; in
+fact, modern graphics architectures \emph{require} that code be separated into
+distinct {\em passes} based on their required frequency of execution. For
+example, graphics computations are expressed as a sequence of passes where each
+pass may perform a computation at increasingly finer grain, e.g., per object,
+per region of the screen, and per pixel, where later computations use the
+results from the prior computations.
 
-%
 While performant, programming with manually separated passes results
@@
 the 1980s, J{\o}rring and Scherlis identified {\em frequency
-  reduction} or {\em precomputation}~\cite{JS86-staging} as the common
+reduction} and {\em precomputation}~\cite{JS86-staging} as common
 mechanisms for evaluating efficiently staged computations.  The idea
@@
 There has been much work on language-based techniques for performing
-precomputation. Partial evaluation~\cite{} achieves precomputation by
-specializing functions for specific values.  Going back to our
-example, we can specialize $f$ at $[1,2, \ldots, n]$, written
-$f_{[1,2, \ldots, n]}$, such that $f(x,y) = f_{[1,2, \ldots, n]}(y)$
-and specialize $F$ for the same input as
+precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
+precomputation by specializing functions for specific values. Going back to our
+example, we can specialize $f$ at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots,
+n]}$, such that $f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for the
+same input as
 \begin{lstlisting}
@@
 %
-Such a partial evaluation typically proceed by performing a
+Partial evaluation typically proceeds by performing a
 binding-time analysis to determine the static and dynamic parts of $f$
@@
 
-As our starting point, we take the modal language \lang\ with explicit
-staging in the style of Davies~\cite{davies96}, where the ``circle''
+As our starting point, we consider a modal, typed language \lang\ with explicit
+staging in the style of Davies~\cite{davies96}, where the $\fut$
 modality denotes computation in a later stage, and present a staged
 operational semantics similar for \lang\ (\ref{sec:stagedsemantics}).
-We then present a splitting algorithm that stratifies a \lang\ program
-into a two mono-staged programs, one for the first stage and for the
+We then present a splitting algorithm that pass-separates a \lang\ program
+into two monostaged programs, one for the first stage and for the
 second stage (\ref{sec:splitting}).  Since each program is monostaged,
@@
 functions that involve both a first stage and second stage
-computations.  When split, such a function turns into a stratum-1
+computations.  When split, such a function turns into a first-pass
 function that yields a {\em boundary value} that encodes the
@@
 the node indicates the recursion status of the first-stage evaluation.
-The stratum-2 part of the computation, which corresponds to the
+The second-pass part of the computation, which corresponds to the
 ``second stage'' takes the boundary value as an argument as well as
@@
 fact, we have implemented the algorithm (\ref{sec:implementation}) and
-applied to several examples (\ref{sec:examples}).  As an interesting
-example, we show that a function
-
-\begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
-  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
-\end{lstlisting}
-where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
-using the ``quickselect'' algorithm.  When applied to this function,
-our split algorithm yields an asymptotically more efficient function
-that runs in expected $O(n\log{n} + m\log{n})$ time instead of
-expected $O(n \cdot m)$ time---a near linear time improvement.
-Interestingly the code output by our splitting algorithm
-auto-generates the ``quicksort'' algorithm, which it uses to build a
-binary search tree as a boundary value in the first stage.  For the
-second stage, the splitting algorithm generates a ``binary search''
-that takes the tree and performs a binary search on the tree as guided
-by the given rank.  To the best of our knowledge, no prior approaches
-can perform such complex transformations on higher-order code, nor can
-they yield such asymptotic improvements in run time.
-
-\paragraph{Old intro below.}
-
-Multi-argument functions can frequently perform useful work before receiving all
-of their inputs, or are often called numerous times with one argument fixed. An
-important program optimization is therefore to \emph{specialize} such a function
-$f$ to its fixed argument $a$, by executing those computations in $f$ which
-depend only on $a$. This ensures that calls to the specialized function require
-only computations which depend on its varying argument.
-
-\emph{Program specialization}, or partial evaluation \cite{futamura71,jones96},
-is a well-known specialization technique which, given $a$, transforms $f$ into a
-new function $f_a(-)$ which computes $f(a,-)$. This transformation essentially
-substitutes $a$ for the first argument of $f$, then evaluates in place any
-subexpressions of $f$ depending only on that argument.
-
-\emph{Data specialization} \cite{knoblock96,JS86-staging} 
-is a technique for specializing $f$ \emph{without} the fixed argument $a$,
-instead splitting $f$ into a pair of functions $f_1$ and $f_2$. $f_1(a)$
-produces a data structure containing the results of the computations which
-depend only on $a$; $f_2$ then completes the computation, given this data
-structure and the varying argument; that is, $f_2(f_1(a),-)$ computes $f(a,-)$.
-Crucially, none of the generated code ($f_1$ or $f_2$) depends on $a$!
-
-Previous work on data specialization has been limited to simple, imperative
-languages. In this paper, we extend data specialization to a typed lambda
-calculus, allowing us to specialize a broader class of programs.
-
-When splitting certain recursive functions, like \texttt{quickselect}, our
-algorithm synthesizes recursive data structures and traversal algorithms which
-yield asymptotic speedups over the original function. Splitting higher-order
-combinators, like \texttt{map}, provides compositional reasoning at the source
-level while cross-cutting fixed runtime stages, as in graphical shading
-languages like Spark~\cite{Foley:2011}.
-
-We start in \ref{sec:example} with an extended example of splitting
-\texttt{quickselect}.
-In \ref{sec:semantics}, we describe our staged language \lang, including its
-type system and operational semantics.
-In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
-algorithm for performing data specialization.
-Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
-transforms a variety of other programs.
+applied it to a number of staged programs (\ref{sec:examples}) ranging
+from straight-line arithmetic operations to recursive and higher-order
+functions. In the case of \texttt{quickselect}, which we discuss next, 
+the split code is asymptotically faster over many $y_i$.

### 2015-02-27 paper/splitting.tex
fix to splitting rule
Complete pass through 4.0

--- 2015-02-26 paper/splitting.tex
@@
 $\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
-yields a pair of monostaged terms $p$ (containing all first stage
-computations in $e$) and $l.r$ (containing all second stage
-computations in $e$). Executing $p$ in a first pass produces $b$ (via
-$\reduce p b$). Then, in the second pass, $\reduce {[b/l]r} {v'} = v$.
-Since execution of the first pass serves to generate input for the
-second pass, we say $p$ is a {\em precomputation} that produces a {\em
-  boundary value} ($b$) for the {\em resumer} ($l.r$).
+yields a pair of monostaged terms $p$ and $l.r$. ($p$ and $l.r$
+contain all first and second stage computations in $e$ respectively.)
+Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
+in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
+$v$.  Since execution of the first pass serves to generate input for
+the second pass, we say $p$ is a {\em precomputation} that produces a
+{\em boundary value} ($b$) for the {\em resumer} ($l.r$).
 
@@
 
-% (and the relationship of these
-% operations to the outputs of direct \lang\ evaluation) is provided at
-% right in \ref{fig:splittingSummary}.
-
-% $\sepredtwosym$ is defined
-% by a single rule, which evaluates the precomputation and substitutes
-% the resulting value in for $l$ in the resumer:
-%\[
-%\infer{{[b/l]r}\equiv q} {\reduce p b}
-%\]
+%masking yields the masked value $\mval i q$ (via~$\vsplito$)
 
 \textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
-to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
-$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
-relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
-$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
-equivalence.  The term $c$ generated by splitting (called the {\em
-  combined term}) is a monostage term encoding all the first-stage
-computations in $e$, and $l.r$ (again called the {\em resumer}) is a
-monostage term encoding all the second-stage computations in $e$.  The
-process of splitting \bbonem\ terms is summarized at left in
-\ref{fig:splittingSummary}.
+to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
+relation~$\splitonesym$) again yields the pair of monostage terms $c$
+and $l.r$
+%(corresponding to the first stage and second stage
+%computations in $e$).
+Evaluating $c$ in the first pass produces the pair $(i,b)$, where the
+first element $i$ matches the first stage results contained in $v$.
+For example, in the simple case where $\rtab \xi v$ contains entirely
+first stage computation, such as:
+\begin{lstlisting}
+2`[]`1` grnd{1+2}`
+\end{lstlisting}
+the first stage result is the value 3.  The second element of the pair
+is used as input to the resumer in second pass evaluation as $\reduce
+{[b/l]r} {v'}$, where $v'$ is the same as result produced by second
+stage computations in $e$.  For example, in the case where $\rtab \xi
+v$ contains entirely second stage computation, such as:
+\begin{lstlisting}
+2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
+\end{lstlisting}
+the second stage result is the value 5. Since in general residual
+tables and partial values describe a mixture of both first and second
+stage computations, in the subsequent subsection we define an
+operation, called {\em masking}, that separates residual tables and
+partial values into stage one and stage two results. We then describe
+the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
+\bbtwo\ terms ($\splittwosym$).
 
-%Defining the requirements for
-%splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
-%(which reduce to monostage residuals that can be directly compared to
-%resumers), \bbonem\ terms reduce to residual tables and partial values
-%that represent computations at both stages.
+%Since $c$ ensapulates computations that produce both $e$'s first-stage
+%result and also the boundary value required by the resumer, we refer
+%to it as a {\em combined term}.
+
+%Note that since $\coltwo e A$ reduces to an entirely second-stage
+%residual, its first-stage computations only exist to generate input
+%for second stage computations.  In contrast, since $\colmix e A$
+%reduces to a multi-stage residual table and partial value, its first
+%stage computations serve to both generate the value used as input for
+%the second stage {\em and also} produce a first-stage value.
+%Correspondingly, the precomputation $p$ resulting from splitting
+%\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
+%the resumer; but the combined term $c$ resulting from splitting
+%\bbonem\ terms reduces to a tuple containing both the first stage {\em
+%  immediate result} $i$ and the boundary value $b$ (motivating the
+%name {\em combined term}).
+
+\subsection{Masking}
+\label{sec:masking}
 
@@
 given in \ref{sec:masking}.)
-
-% We define the requirements for splitting \bbonem\ terms using the
-% results of masking.
-
-Note that since $\coltwo e A$ reduces to an entirely second-stage
-residual, its first-stage computations only exist to generate input
-for second stage computations.  In contrast, since $\colmix e A$
-reduces to a multi-stage residual table and partial value, its first
-stage computations serve to both generate the value used as input for
-the second stage {\em and also} produce a first-stage value.
-Correspondingly, the precomputation $p$ resulting from splitting
-\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
-the resumer; but the combined term $c$ resulting from splitting
-\bbonem\ terms reduces to a tuple containing both the first stage {\em
-  immediate result} $i$ and the boundary value $b$ (motivating the
-name {\em combined term}).
-
-The following subsections formally define the masking transformation
-($\vsplito$), then describe the algorithm for splitting both
-\bbonem\ terms ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
-
-\subsection{Masking}
-\label{sec:masking}
 

### 2015-02-27 paper/splitting.tex

@@
 
-When $e$ is unit or a variable,
-splitting yields a combined term formed by tupling $e$ with a $\tup{}$ precomputation.
-These expressions, as they contain no second-stage subcomputations, have a trivial resumer, $\tup{}$.
-Similarly, the contents of a \texttt{grnd} block are assigned entirely to the immediate value,
-with trivial precomputation and resumer.
+We start with splitting the unit value, $\tup{}$.
+Masking tells us that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
+Plugging this in to the contract for splitting at \bbonem\ gives:
+\[
+	\text{if } \splitone {\tup {}} A c {l.r} \text{ then } \reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
+\]
+Although the only hard requirement for $b$, $c$, and $l.r$ is to satisfy these relations,
+our actual choices are designed to do the least work possible.
+In particular, we choose $b = \tup{}$; $c=\tup{\tup{},\tup{}}$; and $l.r = \_.\tup{}$.
+That is, we choose a trivial boundary value.
+All together, this yield the rule in \cref{fig:termSplit}.
+This exercise proceeds much the same way for \texttt{grnd} blocks, 
+with the purely first stage contents of the block standing in for one of the $\tup{}$s.
 
-For all non-terminals (except \texttt{next}),
-splitting descends into $e$, recursively splitting its $n$ subterms
-to produce their respective combined terms $c_1,\ldots,c_n$ and resumers $r_1, \ldots, r_n$.
-The combined term of $e$ is formed by binding $c_1,\ldots,c_n$
-to the patterns $(y_1,z_1),\ldots,(y_n,z_n)$
-to isolate immediate results from boundary values. Then,
-the immediate result of $e$ is formed by replacing $e$'s subterms with $y_1,\ldots,y_n$.
-The resumer binds the boundary values $b_1,\ldots,b_n$ to an
-argument $(l_1,\ldots,l_n)$ in a term that has the same structure
-of~$e$ but where each subterm is replaced by its resumer ($r_i$'s).
+We next consider splitting a non-terminal, namely $\inl {e}$.
+From the semantics, we already know that 
+for some $\xi$ and $v$, $\diaone {e} \xi v$ and $\diaone {\inl e} \xi {\inl v}$.
+Masking these outputs then gives $\rtab \xi v \vsplito \mval i q$ 
+and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$.
+Since splitting works from the inside out, we already know that
+\[
+	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
+\]
+and what we need to do is find some $b'$, $c'$, and $l'.r'$ satisfying
+\[
+	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l']r' \equiv q
+\]
+Once again, we take the simplest solution, which in this case is $b' = b$; $l'.r' = l.r$;
+and $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$.  
+This sort of technique generalize to all constructs which don't involve control flow.
+
+\nr{working from here down}
 

### 2015-02-27 paper/figures/splitting-overview.tex

@@
 	Splitting followed by evaluation (top and right arrows)
-	produce the same result as 
-	evaluation followed by masking (left and bottom arrows),
+	produces the same result as 
+	\lang\ evaluation followed by masking (left and bottom arrows),
 	up to a term equivalence.
@@
 	Splitting followed by evaluation (top and right arrows)
-	produce the same result as 
-	direct evaluation (left and bottom arrow),
+	produces the same result as 
+	\lang\ evaluation (left and bottom arrow),
 	up to a term equivalence.

### 2015-02-27 paper/obsolete/intro-newer.tex

% first cut at intro in feb 2015

Multi-argument functions can frequently perform useful work before receiving all
of their inputs, or are often called numerous times with one argument fixed. An
important program optimization is therefore to \emph{specialize} such a function
$f$ to its fixed argument $a$, by executing those computations in $f$ which
depend only on $a$. This ensures that calls to the specialized function require
only computations which depend on its varying argument.

\emph{Program specialization}, or partial evaluation \cite{futamura71,jones96},
is a well-known specialization technique which, given $a$, transforms $f$ into a
new function $f_a(-)$ which computes $f(a,-)$. This transformation essentially
substitutes $a$ for the first argument of $f$, then evaluates in place any
subexpressions of $f$ depending only on that argument.

\emph{Data specialization} \cite{knoblock96,JS86-staging} 
is a technique for specializing $f$ \emph{without} the fixed argument $a$,
instead splitting $f$ into a pair of functions $f_1$ and $f_2$. $f_1(a)$
produces a data structure containing the results of the computations which
depend only on $a$; $f_2$ then completes the computation, given this data
structure and the varying argument; that is, $f_2(f_1(a),-)$ computes $f(a,-)$.
Crucially, none of the generated code ($f_1$ or $f_2$) depends on $a$!

Previous work on data specialization has been limited to simple, imperative
languages. In this paper, we extend data specialization to a typed lambda
calculus, allowing us to specialize a broader class of programs.

When splitting certain recursive functions, like \texttt{quickselect}, our
algorithm synthesizes recursive data structures and traversal algorithms which
yield asymptotic speedups over the original function. Splitting higher-order
combinators, like \texttt{map}, provides compositional reasoning at the source
level while cross-cutting fixed runtime stages, as in graphical shading
languages like Spark~\cite{Foley:2011}.

We start in \ref{sec:example} with an extended example of splitting
\texttt{quickselect}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm for performing data specialization.
Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
transforms a variety of other programs.

### 2015-02-27 paper/semantics.tex

@@
 Although products and functions are restricted to types at the same world,
-$\fut$ allows costruction of ``mixed-stage'' products and functions.
+$\fut$ allows construction of ``mixed-stage'' products and functions.
 For example, quickselect is a function at world \bbonem\ which takes a
@@
 
-We evaluate \lang\ terms in two stages. The first stage of evaluation has 
-three mutually recursive modes of operation, depending on the world of the term it is evaluating.
-Specifically, first stage evaluation:
+Although the type system of \lang\ (specifically, the opaque nature of $\fut$
+types) ensures that we can perform all first-stage computations before any
+second-stage ones, the fact that first- and second-stage computations are
+interleaved in \lang\ terms means that realizing this notion of staged
+evaluation requires some care.
+
+We evaluate \lang\ terms in two stages. The first stage of evaluation has three
+mutually recursive modes of operation, depending on the world of the term it is
+evaluating. Specifically, first stage evaluation:
 \begin{itemize}

### 2015-02-27 paper/splitting.tex

@@
 monostaged term on the results of the first pass, yields the same
-result as fully evaluating the original \lang\ program.  Complexity
-arises because \lang\ evaluation of \bbonem\ terms may also generate
-results in the first stage. (For example, a \bbonem\ term of type
-$\curr \rmint$ results in an \rmint\ in the first stage.) Handling the
-asymmetry between the results of evaluating \bbonem\ terms (which may
-produce results at both the first and second stage) and \bbtwo\ terms
-(which only produce second stage results) is fundamental to the
-structure of the splitting algorithm presented in this section.  By
-construction, \bbonep\ terms are monostage and do not require
-splitting.
+result as fully evaluating the original \lang\ program (performing
+both first stage and second stage evaluation).  While world
+\bbtwo\ terms fully evaluate to values in stage two, complexity arises
+because evaluation of \bbonem\ terms may also generate results in the
+first stage. (For example, a \bbonem\ term of type $\curr \rmint$
+results in an \rmint\ in the first stage.) Handling this asymmetry
+between the results of evaluating \bbonem\ terms (which may produce
+results at both the first and second stage) and \bbtwo\ terms (which
+only produce second stage results) is fundamental to the structure of
+the splitting algorithm presented in this section.  By construction,
+\bbonep\ terms are monostage and do not require splitting.
 
@@
 relation~$\splitonesym$) again yields the pair of monostage terms $c$
-and $l.r$
-%(corresponding to the first stage and second stage
-%computations in $e$).
-Evaluating $c$ in the first pass produces the pair $(i,b)$, where the
-first element $i$ matches the first stage results contained in $v$.
-For example, in the simple case where $\rtab \xi v$ contains entirely
-first stage computation, such as:
+and $l.r$. Evaluating $c$ in the first pass produces the pair $(i,b)$,
+where the first element $i$ matches the first stage results contained
+in $v$.  For example, in the simple case where $\rtab \xi v$ contains
+entirely first stage computation, such as:
 \begin{lstlisting}

### 2015-02-27 paper/splitting.tex

@@
 Splitting statically transforms terms in \lang\ into a pair of
-monostaged terms that correspond to the first stage and second stage
-computations in the original term. The idea is that evaluating the
-first monostaged term in a first pass, and then evaluating the second
-monostaged term on the results of the first pass, yields the same
-result as fully evaluating the original \lang\ program (performing
-both first stage and second stage evaluation).  While world
-\bbtwo\ terms fully evaluate to values in stage two, complexity arises
-because evaluation of \bbonem\ terms may also generate results in the
-first stage. (For example, a \bbonem\ term of type $\curr \rmint$
-results in an \rmint\ in the first stage.) Handling this asymmetry
-between the results of evaluating \bbonem\ terms (which may produce
-results at both the first and second stage) and \bbtwo\ terms (which
-only produce second stage results) is fundamental to the structure of
-the splitting algorithm presented in this section.  By construction,
-\bbonep\ terms are monostage and do not require splitting.
+monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first
+stage and second stage computations in the original term. Since full
+evaluation of world \bbtwo\ terms (both first stage and second stage
+evaluation) only produces values in stage two, the idea of splitting
+is to generate terms where evaluating $f_1$ in a first pass, and then
+evaluating $f_2$ on the result of the first pass, yields the same
+value as full evaluation of the original \lang\ program.
+
+Similarly, when splitting \bbonem\ terms, first pass evaluation of
+$f_1$ should produce the same value as first stage evaluation of the
+original \lang\ term.  However, complexity arises because
+\bbonem\ terms may also contain computations to be evaluated in the
+second stage. (For example, a \bbonem\ term of type $\fut \rmint$
+results in an \rmint\ value in the second stage.) Thus we also want
+the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
+the same value as fully evaluating all second stage computations in
+the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
+which only produce values in the first stage, we only need to compare
+the result of $f_1$ with that of the \lang\ term.
+
+Handling the asymmetry between the results of evaluating
+\bbonem\ terms (which may produce results at both the first and second
+stage) and \bbtwo\ terms (which only produce second stage results) is
+fundamental to the structure of the splitting algorithm presented in
+this section.
 

### 2015-02-27 paper/splitting.tex
Update masking to be consistent with the new 4.0.

@@
 
+\newpage
+
 \subsection{Masking}
@@
 
-To resolve the mismatch between the multi-stage computations
-represented by residual tables and partial values and the desired
-monostaged outputs of splitting, we choose to define a translation,
-called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
-converts residual tables and partial values into a {\em masked value}.
-The first part of the masked value, $i$, is a monostage value holding
-the result of all first stage components of $v$.  The second
-component, $q$, is a residual representing all second stage
-computations in $\xi$ and $v$. (A precise definition of masking is
-given in \ref{sec:masking}.)
+%Masking separates residual tables and partial values into first stage
+%results and second stage residuals.
 
-Masking converts a residual table and associated partial value into a
-masked value $\mval i q$.
+Masking (given by the relation $\vsplito$) converts a residual table $\xi$
+and associated partial value $v$ into a masked value $\mval i q$.
 %that explicitly separates the first stage components of 
@@
 %value.
-Masking is not a transformation performed as part of the splitting
-algorithm.  Instead, masking defines an intuitive notion of the first
-stage and second stage ``components'' of residual tables and partial
-values, and it is used to define the requirements of what output
-splitting of \bbonem\ terms must produce.  A precise definition of the
-masking relation is given in \ref{fig:valMask}, but to gain intuition
-about the intent of masking operations, consider the following
-residual table and partial value, colored as before:
+The first part of the masked value, $i$, is a monostage value holding
+the result of all first stage components of $v$.  The second
+component, $q$, is a residual representing all second stage
+computations in $\xi$ and $v$.
+
+%(A precise definition of masking is
+%given in \ref{sec:masking}.)
+
+%Masking is not a transformation performed as part of the splitting
+%algorithm.  Instead, masking defines an intuitive notion of the first
+%stage and second stage ``components'' of residual tables and partial
+%values, and it is used to define the requirements of what output
+%splitting of \bbonem\ terms must produce.
+To provide intuition about the intent of masking, consider the
+following residual table and partial value:
 \begin{lstlisting}
@@
 \end{lstlisting}
-To construct a value representing its first-stage components ($i$ in
-the resulting masked value), masking first redacts all second stage
-(blue) parts, along with the surrounding \texttt{next}
-annotations. (This redaction removes the entire residual table since
-it only represents stage two computations.) The resulting ``holes'' in
-the term are replaced with unit values.
+To construct the value $i$ representing its first-stage components,
+masking first redacts all second stage (blue) parts, along with the
+surrounding \texttt{next} annotations. (This redaction removes the
+entire residual table since it only represents second stage
+computations.) The resulting ``holes'' in the term are replaced with
+unit values.
 %\begin{lstlisting}
@@
 \end{lstlisting}
-Finally, masking dropping \texttt{grnd} annotations yielding a value.
+Finally, masking drops \texttt{grnd} annotations yielding:
 \begin{lstlisting}
@@
 \end{lstlisting}
-To construct a residual corresponding to the stage two computations
-represented by the residual table and partial value ($q$ in resulting
-the masked value), masking first redacts all \texttt{grnd} blocks
-(replaced with unit), \texttt{next} annotations, and injection tags
-(essentially, the tuple structure is kept, but all other structure is
-discarded):
+To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{grnd} blocks
+(replacing them with unit), \texttt{next} annotations, and injection tags:
 %\begin{lstlisting}
@@
 \end{lstlisting}
-Then masking reifies the residual table into \texttt{let} bindings,
-yielding a residual:
+Then the residual table is reified into \texttt{let} bindings,
+yielding:
 \begin{lstlisting}
@@
 \end{lstlisting}
+A precise definition of the masking relation is given in
+\ref{fig:valMask}.
 
-Note in \ref{fig:valMask} that since lambdas may represent multistage
-computations, masking splits the body of lambdas as general world
-\bbonem\ terms (as described in \ref{sec:split-one}), and packages the
-resulting terms as functions in both the first and second stage parts
-of the masked value.
+%Since lambdas may represent multi-stage computations, masking splits
+%the body of lambdas as general world \bbonem\ terms (as described in
+%\ref{sec:split-one}), and packages the resulting terms as functions in
+%both the first and second stage parts of the masked value.
+
 % Masking operates by first inducting on the entries of the residual table.  

### 2015-02-27 paper/splitting.tex
Fix stray page break.
halfway through new 4.2

--- 2015-02-27 paper/splitting.tex
@@
 We now show how to translate terms $\colmix e A$ into the form $\pipeM c l r$,
-pursuant to the correctness condition given in \ref{fig:splittingSummary}.
+pursuant to our splitting invariant.
 The algorithm is specified by the $\splitonesym$ relation (\cref{fig:termSplit}), 
-which proceeds recursively on the structure of~$e$.
+which proceeds recursively on the structure of~$e$. We derive a few cases here.
 
-We start with splitting the unit value, $\tup{}$.
-Masking tells us that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
-Plugging this in to the contract for splitting at \bbonem\ gives:
+\paragraph {Unit} 
+Consider splitting the unit value, $\tup{}$.
+Masking says that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
+So in order for the split $\splitone {\tup {}} A c {l.r}$ to be valid, 
+we need a $c$ and $l.r$ satisfying,
 \[
-	\text{if } \splitone {\tup {}} A c {l.r} \text{ then } \reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
+	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
 \]
-Although the only hard requirement for $b$, $c$, and $l.r$ is to satisfy these relations,
-our actual choices are designed to do the least work possible.
-In particular, we choose $b = \tup{}$; $c=\tup{\tup{},\tup{}}$; and $l.r = \_.\tup{}$.
-That is, we choose a trivial boundary value.
-All together, this yield the rule in \cref{fig:termSplit}.
-This exercise proceeds much the same way for \texttt{grnd} blocks, 
-with the purely first stage contents of the block standing in for one of the $\tup{}$s.
+Our algorithm chooses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
+Splitting $\pure{e}$ follows in similar fashion, 
+where the immediate result is instead the result of evaluating $e$.
 
-We next consider splitting a non-terminal, namely $\inl {e}$.
-From the semantics, we already know that 
-for some $\xi$ and $v$, $\diaone {e} \xi v$ and $\diaone {\inl e} \xi {\inl v}$.
-Masking these outputs then gives $\rtab \xi v \vsplito \mval i q$ 
-and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$.
-Since splitting works from the inside out, we already know that
+\paragraph {Injections} 
+Now consider splitting a non-terminal, $\inl {e}$,
+where $\diaone {e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, 
+and by the splitting invariant for $e$,
 \[
@@
 \]
-and what we need to do is find some $b'$, $c'$, and $l'.r'$ satisfying
+Since, $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
+the validity of the split $\splitone {\inl e} A {c'} {l.r'}$ requires,
 \[
-	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l']r' \equiv q
+	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
 \]
-Once again, we take the simplest solution, which in this case is $b' = b$; $l'.r' = l.r$;
-and $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$.  
-This sort of technique generalize to all constructs which don't involve control flow.
+Our algorithm chooses $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$ and $l.r' = l.r$.
 
-\nr{working from here down}
+\paragraph{Tuple} 
+For an example with more than one argument, consider a tuple $\tup{e_1,e_2}$,
+where $\diaone {e_1} \xi {v_1} \vsplito \mval {i_1} {q_1}$
+and $\diaone {e_2} \xi {v_2} \vsplito \mval {i_2} {q_2}$.
+The splitting invariants for $e_1$ and $e_2$ give
+\[
+	\splitone {e_i} A {c_i} {l_i.r_i} \text{ and } \reduce {c_i} {\tup{i_i,b_i}} \text{ and } [b_i/l_i]r_i \equiv q_i
+\]
+for $i=1,2$.
+
+\paragraph {Case}
+The most complicated rule is that of $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms, because of their divergent control flow.
+We derive the correct splitting rule by considering the two possible execution paths seperately.
+If the first branch was taken, then the predicate evaluated to $\diaone {e_1} {\xi_1} {\inl {v_1}}$ 
+and the first branch evaluated to $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$.
+Similarly these ouputs mask as $\rtab {\xi_1,\xi _2} {v_2} \vsplito \mval {i_2} {q_2}$ and $\rtab {\xi_1,\xi _2} {v_2} \vsplito \mval {i_2} {q_2}$.
+Again because splitting works from the inside out, we already know that that the predicate obeys
+\[
+	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
+\]
+and the first branch obeys
+\[
+	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} \text{ and } [b_2/l_2]r_2 \equiv q_2
+\]
 

### 2015-02-27 paper/example.tex

@@
 
+It is well-known that for any given input list with $n$ elements, the
+quick-select algorithm requires $\Theta(n)$ expected time (for any
+second argument) if we select keys for the partition function
+uniformly randomly or identically if we permute the input. Under this
+assumption, our example program requires expected $\Theta(nm)$ time.
+
+Suppose now that $m$ is large (close to $n$ for example).  In this
+case, our program is essentially a quadratic-time algorithm, which can
+be very slow even for modest values of $m$ and $n$. In this case, to
+improve the run-time it appears that we have no choice but rewrite our
+program to use a different algorithm.  One way to do this would be to
+pre-sort the input list $l$ and then lookup the element with the
+desired rank (as specified by the second argument).  While this
+approach might sound sensible, attaining the desired performance
+requires answering one more important question: what should the data
+structure for storing the sorted input be?  Unfortunately using a list
+does not work, because each look up in a list requires linear time on
+average. One option is to use arrays but in this case, we would need
+to implement a binary search to perform the lookup.  Another option is
+to use a balanced binary search tree data structure, which supports
+lookups in logarithmic time.  Using either one of these techniques, we
+can reduce the run-time to $O(n\log{n} + m\log{n})$, which is
+significantly better that $O(n^2)$.
+
+In summary, to achieve the desired improvement, we wrote a two-pass
+program, where the first pass sorts the input and the second pass
+performs a binary search either on an array or a balanced tree. As is
+probably apparent to the reader, the resulting program is both
+algorithmically and structurally more complex: it requires conversions
+between different data structures (list to array or binary search
+tree) and requires implementing additional algorithms such as sorting
+and binary search, also possibly binary search trees.  As suggested by
+this relatively small example and experienced by the graphics
+community, the design, implementation, and maintenance of such
+multi-pass programs can be difficult due to the increased complexity.
+
+Instead of relying an writing multipass programs manually, in this
+paper, we present techniques for automatically splitting staged
+programs into multi-pass programs.  The key innovation behind our
+approach is a {\em splitting algorithm} that splits a multistage
+program into separate passes where each pass is a conventional program
+expressed in a conventional functional language.  Our splitting
+algorithm makes no restrictions on the input program: essentially any
+functional program, including higher order programs, can be split.
+Furthermore, the splitting algorithm is able to perform highly
+non-trivial transformations on the program that can improve efficiency
+significantly.
+
+When applied to the quick-select example as described, our splitting
+algorithm yields a two-pass program that uses the binary-search-tree
+based implementation described above.  Specifically, in the first
+pass, the program takes the input list and constructs a
+probabilistically balanced binary search tree, which is isomorphic to
+a treap data structure~\cite{treaps}.  In the second pass, the program
+performs a binary search tree lookup, by walking the tree to find the
+element with the desired rank.  The resulting program takes the
+$O(n\log{n} + m\log{n})$.
+
+We note that the splitting algorithm operates by composing local
+transformations on the subterms of the input program.  In particular,
+the algorithm has no special knowledge of the quick-select algorithm,
+binary search trees, or how to perform lookups on binary search trees,
+but it is able to derive all of these from the input program.  But
+how?
+
+The technical detail of the algorithm are relatively involved but in
+the hope that it aids understanding we present a brief informal
+overview of how the algorithm operates.  This exposition is
+necessarily imprecise but the algorithm is made precise in
+\ref{sec:splitting} and implemented
+(\ref{sec:implementation}{sec:examples}).
+
+At a high level, the splitting algorithm operates on our example by
+using the code for quickselect and the knowledge that the list (the
+first argument) is a first-stage argumint and the rank (second
+argument) is a second stage argumunt, the algorithm then derives two
+functions: \texttt{qsStage1} and \texttt{qsStage2} as follows.
+
+\paragraph{Pass-1 code.}
+The splitting algorithm ``imitates'' an execution of quick-select with
+just the first-stage input by evaluating as much code as possible with
+the available information.  Specifically the algorithm generates a
+function \texttt{qsStage1} to perform all the recursive calls of
+\texttt{qs} and to evaluate all instances of the partition function,
+which depend only on the input list.  The function \texttt{qsStage1}
+produces a trace of the execution of \texttt{qs} on its first input by
+generating a tree that collects the results from all recursive calls
+along with a tag that indicates the control branches taken.  The trace
+ends up being represented as a tree because \texttt{qs} has a binary
+control structure casing on the list. In fact, function
+\texttt{qsStage1} constructs a binary search tree, keyed by the
+``pivot'' used by the partition, because the ``pivot'' is needed in
+the second stage to construct the result.  We call the result data
+structure of \texttt{qsStace1} the {\em boundary}, as it passes
+information from the first pass to the second.
+
+\paragraph{Pass-2 code}
+As the splitting algorithm ``scans'' \texttt{qs} for computations that
+can be performed in the first stage and collects them into
+\texttt{qsStage1}, it also collects computations that must be left to
+the second stage in a separate function \texttt{qsStage2}.  This
+function, which is executed in the second pass, takes as argument the
+boundary and the second-stage argument (the rank) and performs a
+lookup in the boundary data structure.  But how does \texttt{qsStage2}
+knows to traverse the tree? It does not. Since the splitting algorithm
+recorded the control flow of the first stage execution in the boundary
+data structure, \texttt{qsStage2} simply follows this control
+structure and performs at each point the parts of the computation from
+\texttt{qs} that can now be performed in the presence of the
+second-stage argument (the rank).
+
+\begin{comment}
+
 An astute programmer, having noticed that \texttt{quickselect} can be staged in
@@
 above.
-
+\end{comment}
 %\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
@@
 
+\subsection{Quickselect in more detail}
+
+Carlo writes this.
+
 \end{abstrsyn}

### 2015-02-27 paper/splitting.tex
few more lines of splitting

@@
 %name {\em combined term}).
-
-\newpage
 
@@
 For an example with more than one argument, consider a tuple $\tup{e_1,e_2}$,
-where $\diaone {e_1} \xi {v_1} \vsplito \mval {i_1} {q_1}$
-and $\diaone {e_2} \xi {v_2} \vsplito \mval {i_2} {q_2}$.
-The splitting invariants for $e_1$ and $e_2$ give
+where $\diaone {e_k} {\xi_k} {v_k}$ and $\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$
+for $k = 1,2$.
+Likewise, the splitting invariants for $e_1$ and $e_2$ give
 \[
-	\splitone {e_i} A {c_i} {l_i.r_i} \text{ and } \reduce {c_i} {\tup{i_i,b_i}} \text{ and } [b_i/l_i]r_i \equiv q_i
+	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
 \]
-for $i=1,2$.
+Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
+and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
+the validity of the split $\splitone {\tup{e_1,e_2}} A {c'} {l.r'}$ requires,
+\[
+	\reduce {c'} {\tup{\tup{i_1,i_2},b'}} \text{ and } [b'/l]r' \equiv q
+\]
 

### 2015-02-27 paper/relatedwork.tex
fix merge conflicts (correctly, i hope) and references in related work

@@
 prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
-limited to simple, imperitive languages.
+limited to simple, imperative languages.
+
+\paragraph{Memoization.} Leaves code structure unchanged.
+
+\paragraph{Incremental computation.} Usually not applicable within a
+computation, require ``mutator'' ``core'' behavior.
 

### 2015-02-27 paper/splitting.tex
Move to {p|l.r} syntax in 4.0 to be consistent with rest of document.

@@
 $\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
-yields a pair of monostaged terms $p$ and $l.r$. ($p$ and $l.r$
+yields a pair of monostaged terms {$p$|$l.r$}. ($p$ and $l.r$
 contain all first and second stage computations in $e$ respectively.)
@@
 to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
-relation~$\splitonesym$) again yields the pair of monostage terms $c$
-and $l.r$. Evaluating $c$ in the first pass produces the pair $(i,b)$,
-where the first element $i$ matches the first stage results contained
-in $v$.  For example, in the simple case where $\rtab \xi v$ contains
-entirely first stage computation, such as:
+relation~$\splitonesym$) again yields the pair of monostage terms
+{$c$|$l.r$}. Evaluating $c$ in the first pass produces the pair
+$(i,b)$, where the first element $i$ matches the first stage results
+contained in $v$.  For example, in the simple case where $\rtab \xi v$
+contains entirely first stage computation, such as:
 \begin{lstlisting}
@@
 the first stage result is the value 3.  The second element of the pair
-is used as input to the resumer in second pass evaluation as $\reduce
-{[b/l]r} {v'}$, where $v'$ is the same as result produced by second
-stage computations in $e$.  For example, in the case where $\rtab \xi
-v$ contains entirely second stage computation, such as:
+is the boundary value used as input to the resumer in second pass
+evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
+result produced by second stage computations in $e$.  For example, in
+the case where $\rtab \xi v$ contains entirely second stage
+computation, such as:
 \begin{lstlisting}
@@
 
-\subsection {Necessity of \bbonep}
+\subsection {Role of World \bbonep}
 \label{sec:needGround}
 
-We have not yet given a justification for why first stage code must be
-partitioned between the worlds \bbonem\ and \bbonep.
-Certainly one could imagine a simpler system without all the \texttt{grnd} annotations.  
-Why wouldn't this work?
+The splitting algorithm described in the previous subsections is
+defined based on the local structure of \lang\ terms.  One artifact of
+this design is that splitting generates
+
+For example, the rule for splitting \bbonem\ case terms
+
+% splitting rules are local term driven
+
+% generate a lot of cruft
+
+% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content
+
+% recursive case would be recursive calls
+
+% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff
+
+% skip splitting.
+
+% example is partition in 6.3
 
@@
 
-We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
-Yet even if a language designer desired a different implicit/explicit trade-off,
-then the three-world \lang\ would still be useful as a typed intermediate representation.
+%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
+%Yet even if a language designer desired a different implicit/explicit trade-off,
+%then the three-world \lang\ would still be useful as a typed intermediate representation.
 

### 2015-02-27 paper/splitting.tex
Trying again with the "pair" syntax.

@@
 $\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
-yields a pair of monostaged terms {$p$|$l.r$}. ($p$ and $l.r$
+yields a pair of monostaged terms $\pipeS p l r$. ($p$ and $l.r$
 contain all first and second stage computations in $e$ respectively.)
@@
 relation~$\splitonesym$) again yields the pair of monostage terms
-{$c$|$l.r$}. Evaluating $c$ in the first pass produces the pair
+$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
 $(i,b)$, where the first element $i$ matches the first stage results

### 2015-02-27 paper/intro.tex
intro, work in progress

@@
 
-Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, in order to
-compute $f(x,y_i)$ for each $y_i$:
+Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, and computes
+$f(x,y_i)$ for each $y_i$:
 \begin{lstlisting}
-fun F(x, @$y_0$@, @$y_1$@, @$\ldots$@ @$y_{m-1}$@) = 
-  f(x, @$y_0$@);  f(x, @$y_1$@); @$\ldots$@ f(x, @$y_{m-1}$@)
+fun @$F(x, y_0, y_1, \ldots, y_{m-1})$@ = 
+  @$f(x, y_0);  f(x, y_1); \ldots; f(x, y_{m-1})$@
 \end{lstlisting}
 %
-Such computations, in which a multivariate function is applied many times with
-one argument fixed, are common. Many algorithm design techniques revolve around
-the idea of finding the computations in $f$ that depend only on $x$, and
-performing them only once. Similarly, compiler optimization techniques like loop
-hoisting and common subexpression elimination serve to identify the computations
-common to all invocations of $f$, and perform them a single time at the
-beginning of the computation.
+This scenario, in which a multivariate function is applied many times with one
+argument fixed, arises frequently. 
+\crem{?? Many algorithm design techniques revolve around the idea of finding the
+computations in $f$ that depend only on $x$, and
+performing them only once.
+\\
+?? It's advantageous to find the computations in f that depend on x and perform
+them only once.}
+This is the intent of compiler optimization techniques like loop hoisting and
+common subexpression elimination, which identify the computations common to all
+invocations of $f$, and perform them once at the beginning of the computation.
 
-These techniques are particularly essential in real-time computer graphics; in
-fact, modern graphics architectures \emph{require} that code be separated into
-distinct {\em passes} based on their required frequency of execution. For
-example, graphics computations are expressed as a sequence of passes where each
-pass may perform a computation at increasingly finer grain, e.g., per object,
-per region of the screen, and per pixel, where later computations use the
-results from the prior computations.
+Programs structured in this way are particularly important in real-time computer
+graphics; in fact, modern graphics architectures \emph{require} that graphics
+computations be organized into distinct {\em passes} which perform increasingly
+fine-grained computations (for example, per object, per region of the screen,
+and per pixel, where later computations use the results from the prior
+computations).
 
-While performant, programming with manually separated passes results
-in complex code where invariants must hold across different passes,
-and local changes for one pass may require changes to other passes to
+While performant, this paradigm of programming with explicit collections of
+passes results in complex code where invariants must hold across different
+passes, and local changes for one pass may require changes to other passes to
 ``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
 %
-In other words, manual pass separation breaks key programming
+In other words, requiring manual separation of passes breaks key programming
 abstractions such as composition and modularity.
 %
-Graphics researchers therefore have suggested using explicitly staged
-programming languages\,\cite{Proudfoot:2001,Foley:2011,He:2014},
-deferring to the compiler the tedious task of separating the code into
-passes. However, all prior efforts have been limited to simple
-languages that do not support recursion or first-class functions.
+Graphics researchers therefore have suggested using explicitly staged programs
+\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
+mechanical task of separating such programs into passes. However, prior work in
+automatic pass separation has been limited to simple languages that do not
+support recursion or first-class functions.
 
-Programming-languages researchers have studied similar problems. In
-the 1980s, J{\o}rring and Scherlis identified {\em frequency
-reduction} and {\em precomputation}~\cite{JS86-staging} as common
-mechanisms for evaluating efficiently staged computations.  The idea
-behind frequency reduction is to identify computations that are
-performed multiple times and hoist them so that they can be performed
-once and used later as needed.  The idea behind precomputation is to
-identify computations that can be performed earlier---for example, at
-compile time if their inputs are known statically---and perform them
-at that earlier time.
+Programming languages researchers have studied similar problems. In the 1980s,
+J{\o}rring and Scherlis identified {\em frequency reduction} and {\em
+precomputation}~\cite{JS86-staging} as mechanisms for efficiently evaluating
+staged computations. To perform frequency reduction, one identifies and hoists
+computations that are performed multiple times, in order to compute them only
+once. To perform precomputation, one identifies computations that can be
+performed in advance, and does so---for example, at compile time if the relevant
+inputs are statically known.
 
@@
 precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
-precomputation by specializing functions for specific values. Going back to our
-example, we can specialize $f$ at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots,
-n]}$, such that $f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for the
-same input as
+precomputation by specializing functions to specific arguments. Going back to
+our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
+$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
 \begin{lstlisting}
-fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
-  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
+fun @$F_v(y_0, y_1, \ldots, y_{m-1})$@ = 
+  @$f_v(y_0);  f_v(y_1); \ldots; f_v(y_{m-1})$@.
 \end{lstlisting}
+eliminating the need to recompute
+\crem{WIP especially below here}
+TODO.
 %
-Partial evaluation typically proceeds by performing a
-binding-time analysis to determine the static and dynamic parts of $f$
-and $F$ and specializes them based on the results of the binding-time
-analysis. A closely related technique, {\em metaprogramming}, enables
-the programmer to write programs that generate specialized code.
-Metaprogramming enables fine-grained control over specialization by
-requiring staging annotations that make explicit the stage of each
-computation.
+A closely related technique, {\em metaprogramming}, allows writing programs that
+generate specialized code. Metaprogramming enables fine-grained control over
+specialization by requiring staging annotations that make explicit the stage of
+each computation. (In partial evaluation, we typically assume that a
+\emph{binding-time analysis} has automatically determined which parts of $f$ can
+be specialized.)
 

### 2015-02-27 paper/splitting.tex
explaining stage one spliting

--- 2015-02-27 paper/splitting.tex
@@
 \]
-Our algorithm chooses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
+Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
 Splitting $\pure{e}$ follows in similar fashion, 
@@
 \]
-Our algorithm chooses $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$ and $l.r' = l.r$.
+To satisfy this, our algorithm uses the same boundary value as the subexpression,
+and so $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$ and $l.r' = l.r$.
 
@@
 and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
-the validity of the split $\splitone {\tup{e_1,e_2}} A {c'} {l.r'}$ requires,
+the validity of the split $\splitone {\tup{e_1,e_2}} A {c} {l.r}$ requires,
 \[
-	\reduce {c'} {\tup{\tup{i_1,i_2},b'}} \text{ and } [b'/l]r' \equiv q
+	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
+\]
+To satisfy this, our algorithm uses a boundary value which is the tuple of the boundary values of the subexpressions,
+$b = \tup{b_1,b_2}$. This is deconstructed with a tuple pattern, $l.r = \tup{l_1,l_2}.\tup{r_1,r_2}$,
+and constructed using,
+\[
+c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
 \]
@@
 \paragraph {Case}
-The most complicated rule is that of $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms, because of their divergent control flow.
-We derive the correct splitting rule by considering the two possible execution paths seperately.
-If the first branch was taken, then the predicate evaluated to $\diaone {e_1} {\xi_1} {\inl {v_1}}$ 
-and the first branch evaluated to $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$.
-Similarly these ouputs mask as $\rtab {\xi_1,\xi _2} {v_2} \vsplito \mval {i_2} {q_2}$ and $\rtab {\xi_1,\xi _2} {v_2} \vsplito \mval {i_2} {q_2}$.
-Again because splitting works from the inside out, we already know that that the predicate obeys
-\[
-	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
-\]
-and the first branch obeys
-\[
-	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} \text{ and } [b_2/l_2]r_2 \equiv q_2
-\]
+Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
+Consider that when $\reduce {e_1} {\inl {v_1}}$, both the first and second stage component of $e_2$ will be executed, 
+but $e_3$ will be ignored entirely. The reverse occurs when $\reduce {e_1} {\inr {v_1}}$.
+This implies that both the first and second stage results of splitting will need to have \texttt{case} statements.
 
-Splitting {\tt case} yields a combined term that executes one of the branches' combined terms based on the immediate result $y_1$ of the predicate.
-The boundary value $b_i$ for this branch is injected and bundled with that of the predicate ($b_1$). 
-$b_i$ is cased in the resumer to determine which branch's resumer should be executed.
-{\tt case} is the only rule where splitting adds non-trivial logic is added to the precomputation.
+Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
+$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.
+Thus the algorithm cases on $c_1$ to run the correct branch of first stage code.
+The boundary value from this correct branch, $b_2$ or $b_3$, is packaged as either $\tup {b_1, \inl {b_2}}$ or $\tup {b_1, \inr {b_3}}$.
 
+The resumer for \texttt{case} operates by untupling the boundary value to get $b_1$ and either $\inl {b_2}$ or $\inr {b_3}$.
+The resumer of the predicate is run using $b_1$, and the remaining injection is inspected to run the correct branch resumer.
+
+\paragraph {Function} 
 Function introduction has a $\tup{}$ boundary value,
@@
 and pass it to the resumer version of the function.
-
-Since the results of splitting \texttt{next} terms depend on the output of
-splitting its world \bbtwo\ subterm,
-we defer description of \texttt{next} until after describing world \bbtwo\ term splitting.
 

### 2015-02-27 paper/splitting.tex

@@
 $\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
-yields a pair of monostaged terms $p$ and $l.r$. ($p$ and $l.r$
+yields a pair of monostaged terms $\pipeS p l r$. ($p$ and $l.r$
 contain all first and second stage computations in $e$ respectively.)
@@
 to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
-relation~$\splitonesym$) again yields the pair of monostage terms $c$
-and $l.r$. Evaluating $c$ in the first pass produces the pair $(i,b)$,
-where the first element $i$ matches the first stage results contained
-in $v$.  For example, in the simple case where $\rtab \xi v$ contains
-entirely first stage computation, such as:
+relation~$\splitonesym$) again yields the pair of monostage terms
+$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
+$(i,b)$, where the first element $i$ matches the first stage results
+contained in $v$.  For example, in the simple case where $\rtab \xi v$
+contains entirely first stage computation, such as:
 \begin{lstlisting}

### 2015-02-27 paper/splitting.tex
Quick first pass at 4.4.

--- 2015-02-27 paper/splitting.tex
@@
 operation, called {\em masking}, that separates residual tables and
-partial values into stage one and stage two results. We then describe
-the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
-\bbtwo\ terms ($\splittwosym$).
+partial values into their first and second stage results. We then
+describe the algorithm for splitting both \bbonem\ terms
+($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
 
@@
 
-The splitting algorithm described in the previous subsections is
-defined based on the local structure of \lang\ terms.  One artifact of
-this design is that splitting generates
+The splitting algorithm described in the previous subsections operates
+on the local structure of \lang\ terms.  One artifact of this design
+is that splitting \bbonem\ terms may generate resumers containing
+unnecessary logic.  For example, the rule for splitting
+\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
+argument into the boundary value, then decodes this tag in the
+resumer.  This logic occurs regardless of whether the terms forming
+the branches of the \texttt{case} contain second stage computations.
+Worse, if this \texttt{case} appeared in the body of a recursive
+function with no other second stage computations, splitting would
+generate a resumer with (useless) recursive calls.
 
-For example, the rule for splitting \bbonem\ case terms
+Rather than attempt global optimization of the outputs of splitting,
+we instead introduce world \bbonep\ and use the type system to
+statically prove terms contain no second stage computations.  In these
+situations, splitting is trivial---splitting $\pure e$ terms yields a
+unit resumer.
 
-% splitting rules are local term driven
-
-% generate a lot of cruft
-
-% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content
-
-% recursive case would be recursive calls
-
-% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff
-
-% skip splitting.
-
-% example is partition in 6.3
-
-In order to be correct, any code at \bbonem\ must be split with the 
-pessimistic assumption that it may result in work at the second stage.
-When this assumption turns out not to be true---that is, purely monostage code---then 
-splitting may produce second stage code which is needlessly costly 
-(an example of this will occur in \ref{sec:exampleQS}).
-Detecting and optimizing this case is in general a global program analysis,
-since functions can be passed around as values.  
-So instead of relying on hefty analysis, 
-we take the approach of adding enough structure to the input language's type system to
-allow the input code to {\em prove} itself to be---in some parts---monostage.
-
-%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
-%Yet even if a language designer desired a different implicit/explicit trade-off,
-%then the three-world \lang\ would still be useful as a typed intermediate representation.
+The \texttt{partition} function in the \texttt{quickselect} example in
+\ref{sec:exampleQS} is an example of such a function that contains no
+second stage computations, but, if not defined within a $\pure$
+annotation, would cause splitting to generate a recursive resumer.
 

### 2015-02-27 paper/splitting.tex
Fix typo.

@@
 argument into the boundary value, then decodes this tag in the
-resumer.  This logic occurs regardless of whether the terms forming
-the branches of the \texttt{case} contain second stage computations.
+resumer. This logic occurs regardless of whether the terms forming the
+branches of the \texttt{case} contain second stage computations.
 Worse, if this \texttt{case} appeared in the body of a recursive
@@
 we instead introduce world \bbonep\ and use the type system to
-statically prove terms contain no second stage computations.  In these
-situations, splitting is trivial---splitting $\pure e$ terms yields a
-unit resumer.
+statically prove certain terms contain no second stage computations.
+In these situations, splitting is trivial---splitting $\pure e$ terms
+yields a unit resumer.
 

### 2015-02-27 paper/intro.tex
finished a major pass through the intro (pending the first paragraph)

@@
 \,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
-mechanical task of separating such programs into passes. However, prior work in
-automatic pass separation has been limited to simple languages that do not
-support recursion or first-class functions.
+mechanical task of separating such programs into passes.
 
@@
 \end{lstlisting}
-eliminating the need to recompute
-\crem{WIP especially below here}
-TODO.
-%
+eliminating the need to recompute the parts of $f(v,-)$ which do not depend on
+the second argument.
+
 A closely related technique, {\em metaprogramming}, allows writing programs that
@@
 
-While previous work shows that precomputation can be performed
-essentially automatically (via partial evaluation or meta
-programming), frequency reduction has been less explored.  In their
-original paper, J{\o}rring and Scherlis proposed {\em pass separation}
-as a manual method for {\em splitting} a program into multiple passes.
+While partial evaluation and metaprogramming have been studied extensively,
+frequency reduction techniques have not seen the same attention. In
+\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
+manual method for {\em splitting} a program into multiple passes.
 %
-Applied to our example, pass separation would split the function $f$
-into $f_1$ and $f_2$ such that $f(x,y) = f_2(f_1(x),y)$.  We can then
-rewrite our example as a multi-pass program:
+In our example, pass separation would split the function $f$ into two passes
+$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
+evaluating the first pass on $x$, and then the second pass on each $y_i$.
 %
 \begin{lstlisting}
-fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
+fun @$F_\textrm{staged}(x, y_0, y_1, \ldots, y_{m-1})$@ = 
   let @$fx = f_1(x)$@
@@
 %
-The key difference between partial evaluation (meta programming) and
-pass separation is that the latter does not require the first
-argument~$x$: it works with all arguments, whereas the partially
-evaluated program is specialized for a particular value of $x$ ($x =
-[1,2, \ldots, n]$ in the example). Specifically, if desired, partial
-evaluation can be applied subsequently in our example, by partially
-evaluating function~$f_2$ on $f_1(x)$.
+The key difference between partial evaluation (or metaprogramming) and pass
+separation is that the latter can be performed without access to the first
+argument $x$: $F_\textrm{staged}$ works for any $x$, while $F_v$ is defined
+only for $x=v$. (Of course, when a particular $v$ \emph{is} known, one may
+always use partial evaluation to specialize $F_\textrm{staged}$ to that $v$.)
 
-However, unlike for precomputation, there is no available technique
-for pass separation except for very simple languages
-\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.  Notably omitted
-features include recursion and first-class functions.  We are thus
-interested in the problem of splitting automatically a functional
-language.
+Prior work on partial evaluation and metaprogramming has shown that
+precomputation can largely be automated for functional languages. In contrast,
+automatic pass separation has been limited to simple languages that do not
+support recursion or first-class
+functions~\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In this paper, we
+extend this work by presenting a pass separation algorithm for a typed lambda
+calculus.
+
+As our starting point, we consider an explicitly-two-staged, typed lambda
+calculus \lang\ in the style of Davies~\cite{davies96}, with a $\fut$ modality
+denoting computation in a later stage (\ref{sec:semantics}). We then present a
+splitting algorithm to pass-separate a \lang\ program into two passes
+corresponding to the stages of computation in that program
+(\ref{sec:splitting}). These passes are monostaged programs, expressed in a
+conventional functional language.
+
+The most interesting case involves splitting a mixed-stage function (like $f$ in
+the example) which is moreover recursive. This yields a first pass function
+$f_1$ which computes from its first-stage argument $x$ a \emph{boundary} data
+structure encoding the interstage communication in $f(x,-)$. When $f$ recurs on
+$x$, this boundary structure itself has a recursive type.
 %
+The second pass is a function $f_2$ which takes the boundary data structure
+$f_1(x)$ and a second-stage argument $y_i$, and completes the computation
+$f(x,y_i)$, traversing the boundary structure in light of the particular
+argument $y_i$.
 
-As our starting point, we consider a modal, typed language \lang\ with explicit
-staging in the style of Davies~\cite{davies96}, where the $\fut$
-modality denotes computation in a later stage, and present a staged
-operational semantics similar for \lang\ (\ref{sec:stagedsemantics}).
-We then present a splitting algorithm that pass-separates a \lang\ program
-into two monostaged programs, one for the first stage and for the
-second stage (\ref{sec:splitting}).  Since each program is monostaged,
-they are expressed in conventional functional languages.
+We have implemented this splitting algorithm (\ref{sec:implementation}) and
+applied it to a number of staged programs (\ref{sec:examples}) ranging from
+straight-line arithmetic operations to recursive and higher-order functions. In
+the case of \texttt{quickselect}, which we discuss next, the split code is in
+fact asymptotically faster over many $y_i$.
 
-The crux of the splitting algorithm is splitting recursive mixed stage
-functions that involve both a first stage and second stage
-computations.  When split, such a function turns into a first-pass
-function that yields a {\em boundary value} that encodes the
-``first-stage'' part of the computation.  If the function recurs on a
-first-stage argument, then the boundary value is typically a recursive
-data structure.  In other words, the splitting algorithm maps
-computational recursion into a recursive data type. If the boundary
-structure is generated as a result of a conditional, it is typically
-encoded as a (recursive) sum type, such as a tree, where the kind of
-the node indicates the recursion status of the first-stage evaluation.
-The second-pass part of the computation, which corresponds to the
-``second stage'' takes the boundary value as an argument as well as
-the second-stage argument and uses them to complete the computation,
-often by traversing the now precomputed boundary in light of the now
-available second-stage argument.
-
-Since it deals with unrestricted recursion and a rich language
-consisting of first-class functions, recursion, and sum and recursive
-types, the splitting algorithm is naturally complex. With some care,
-the algorithm nevertheless can be specified reasonably succinctly. In
-fact, we have implemented the algorithm (\ref{sec:implementation}) and
-applied it to a number of staged programs (\ref{sec:examples}) ranging
-from straight-line arithmetic operations to recursive and higher-order
-functions. In the case of \texttt{quickselect}, which we discuss next, 
-the split code is asymptotically faster over many $y_i$.
+%The crux of the splitting algorithm is splitting recursive mixed stage
+%functions that involve both a first stage and second stage computations.  When
+%split, such a function turns into a first-pass function that yields a {\em
+%boundary value} that encodes the ``first-stage'' part of the computation.  If
+%the function recurs on a first-stage argument, then the boundary value is
+%typically a recursive data structure.  In other words, the splitting algorithm
+%maps computational recursion into a recursive data type. If the boundary
+%structure is generated as a result of a conditional, it is typically encoded as
+%a (recursive) sum type, such as a tree, where the kind of the node indicates
+%the recursion status of the first-stage evaluation.  The second-pass part of
+%the computation, which corresponds to the ``second stage'' takes the boundary
+%value as an argument as well as the second-stage argument and uses them to
+%complete the computation, often by traversing the now precomputed boundary in
+%light of the now available second-stage argument.

### 2015-02-27 paper/intro.tex
complete intro

@@
 %
-This scenario, in which a multivariate function is applied many times with one
-argument fixed, arises frequently. 
-\crem{?? Many algorithm design techniques revolve around the idea of finding the
-computations in $f$ that depend only on $x$, and
-performing them only once.
-\\
-?? It's advantageous to find the computations in f that depend on x and perform
-them only once.}
-This is the intent of compiler optimization techniques like loop hoisting and
-common subexpression elimination, which identify the computations common to all
-invocations of $f$, and perform them once at the beginning of the computation.
+In this common scenario, a multivariate function is applied many times with one
+argument fixed, so it is advantageous to find the computations in $f$ that
+depend on $x$ and perform them only once. This is the intent of compiler
+optimization techniques like loop hoisting and common subexpression elimination,
+which identify the computations common to all invocations of $f$, and perform
+them once at the beginning of the computation.
 
-Programs structured in this way are particularly important in real-time computer
-graphics; in fact, modern graphics architectures \emph{require} that graphics
-computations be organized into distinct {\em passes} which perform increasingly
-fine-grained computations (for example, per object, per region of the screen,
-and per pixel, where later computations use the results from the prior
-computations).
+Programs featuring this structure are particularly important in real-time
+computer graphics; in fact, modern graphics architectures \emph{require} that
+graphics computations be organized into distinct {\em passes} which perform
+increasingly fine-grained computations (for example, per object, per region of
+the screen, and per pixel, where many computations in later passes use the
+results from a single computation in an earlier pass).
 
-While performant, this paradigm of programming with explicit collections of
-passes results in complex code where invariants must hold across different
-passes, and local changes for one pass may require changes to other passes to
-``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
+While performant, this paradigm of programming with explicit passes results in
+complex code where invariants must hold across different passes, and local
+changes for one pass may require changes to other passes to ``plumb'' inputs
+from all other frequencies\,\cite{Foley:2011}.
 %

### 2015-02-27 paper/splitting.tex
Incorporate Carlo's comments into 4.4.

--- 2015-02-27 paper/splitting.tex
@@
 Rather than attempt global optimization of the outputs of splitting,
-we instead introduce world \bbonep\ and use the type system to
-statically prove terms contain no second stage computations.  In these
-situations, splitting is trivial---splitting $\pure e$ terms yields a
-unit resumer.
+we instead leverage the type system to indicate when a term contains
+no second-stage computations by adding a third world \bbonep\ whose
+terms are purely first stage.  While not necessary for ordinary staged
+term evaluation (\ref{sec:stagedsemantics}), this annotation is
+particularly useful for splitting. Specifically, splitting $\pure e$
+terms trivially yields a unit resumer.
 

### 2015-02-27 paper/splitting.tex
worked on splitting rules exposition

--- 2015-02-27 paper/splitting.tex
@@
 operation, called {\em masking}, that separates residual tables and
-partial values into their first and second stage results. We then
-describe the algorithm for splitting both \bbonem\ terms
-($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
+partial values into stage one and stage two results. We then describe
+the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
+\bbtwo\ terms ($\splittwosym$).
 
@@
 
-Since the results of splitting \texttt{next} terms depend on the output of
-splitting its world \bbtwo\ subterm,
-we defer description of \texttt{next} until after describing world \bbtwo\ term splitting.
-
 \subsection{Term Splitting at \bbtwo}
@@
 Because world \bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
-term splitting at world \bbtwo\ assumes a simpler form than the version at \bbonem\ does. 
+term splitting at world \bbtwo\ is simpler and more uniform than the version at \bbonem. 
 The algorithm is specified by the $\splittwosym$ relation in \cref{fig:termSplit}.
 
-In the terminal cases of
-constants and variables, splitting generates trivial precomputations that are \texttt{()}, and resumers consisting of the original term.
-For example, the integer constant \texttt{3} splits into the
-precomputation \texttt{()} and resumer \texttt{\_=>3}.
-
-More generally, for all (except \texttt{prev}) 
-$n$-ary terms $e = \mathcal{C}\ttlpar e_1 \ttsemi \ldots \ttsemi e_n \ttrpar$ 
-the precomputation is the tupled precomputations of $e$'s $n$ subterms:
-$p=(p_1,\ldots,p_n)$.  The resumer binds each boundary value to an
-argument $(l_1,\ldots,l_n)$ in a term that has the same structure
-of~$e$ but where each subterm is replaced by its corresponding resumer:
-$r = \mathcal{C}\ttlpar r_1 \ttsemi \ldots \ttsemi r_n \ttrpar$ .
-Notably, at \texttt{case}s and functions the
-precomputation of subterms is lifted out from underneath world \bbtwo\ binders.  
-% TODO: should probably draw a parallel to the same behavior in dynamics
+Consider any (except \texttt{prev}) 
+$n$-ary term $e = \scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$.
+First note that for any $\diatwo e q$, $q$ must have the form $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$,
+even under binders and for nullary terms.
+For each $k \in \{1,\ldots,n\}$, the invariant for stage \bbtwo\ splitting of $e_k$ is,
+\[
+	\splittwo {e_k} A {p_k} {l_k} {r_k} \text{ and } \reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
+\]
+To match the invariant for a split $\splittwo {\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}} A p l r$, we require,
+\[
+	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
+\]
+Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
+$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
+Again, this works even for nullary terms and under binders.
 
@@
 
-The splitting algorithm described in the previous subsections operates
-on the local structure of \lang\ terms.  One artifact of this design
-is that splitting \bbonem\ terms may generate resumers containing
-unnecessary logic.  For example, the rule for splitting
-\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
-argument into the boundary value, then decodes this tag in the
-resumer.  This logic occurs regardless of whether the terms forming
-the branches of the \texttt{case} contain second stage computations.
-Worse, if this \texttt{case} appeared in the body of a recursive
-function with no other second stage computations, splitting would
-generate a resumer with (useless) recursive calls.
+The splitting algorithm described in the previous subsections is
+defined based on the local structure of \lang\ terms.  One artifact of
+this design is that splitting generates
 
-Rather than attempt global optimization of the outputs of splitting,
-we instead introduce world \bbonep\ and use the type system to
-statically prove terms contain no second stage computations.  In these
-situations, splitting is trivial---splitting $\pure e$ terms yields a
-unit resumer.
+For example, the rule for splitting \bbonem\ case terms
 
-The \texttt{partition} function in the \texttt{quickselect} example in
-\ref{sec:exampleQS} is an example of such a function that contains no
-second stage computations, but, if not defined within a $\pure$
-annotation, would cause splitting to generate a recursive resumer.
+% splitting rules are local term driven
+
+% generate a lot of cruft
+
+% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content
+
+% recursive case would be recursive calls
+
+% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff
+
+% skip splitting.
+
+% example is partition in 6.3
+
+In order to be correct, any code at \bbonem\ must be split with the 
+pessimistic assumption that it may result in work at the second stage.
+When this assumption turns out not to be true---that is, purely monostage code---then 
+splitting may produce second stage code which is needlessly costly 
+(an example of this will occur in \ref{sec:exampleQS}).
+Detecting and optimizing this case is in general a global program analysis,
+since functions can be passed around as values.  
+So instead of relying on hefty analysis, 
+we take the approach of adding enough structure to the input language's type system to
+allow the input code to {\em prove} itself to be---in some parts---monostage.
+
+%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
+%Yet even if a language designer desired a different implicit/explicit trade-off,
+%then the three-world \lang\ would still be useful as a typed intermediate representation.
 

### 2015-02-27 paper/splitting.tex

@@
 
-The splitting algorithm described in the previous subsections is
-defined based on the local structure of \lang\ terms.  One artifact of
-this design is that splitting generates
+The splitting algorithm described in the previous subsections operates
+on the local structure of \lang\ terms.  One artifact of this design
+is that splitting \bbonem\ terms may generate resumers containing
+unnecessary logic.  For example, the rule for splitting
+\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
+argument into the boundary value, then decodes this tag in the
+resumer. This logic occurs regardless of whether the terms forming the
+branches of the \texttt{case} contain second stage computations.
+Worse, if this \texttt{case} appeared in the body of a recursive
+function with no other second stage computations, splitting would
+generate a resumer with (useless) recursive calls.
 
-For example, the rule for splitting \bbonem\ case terms
+Rather than attempt global optimization of the outputs of splitting,
+we instead leverage the type system to indicate when a term contains
+no second-stage computations by adding a third world \bbonep\ whose
+terms are purely first stage.  While not necessary for ordinary staged
+term evaluation (\ref{sec:stagedsemantics}), this annotation is
+particularly useful for splitting. Specifically, splitting $\pure e$
+terms trivially yields a unit resumer.
 
-% splitting rules are local term driven
-
-% generate a lot of cruft
-
-% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content
-
-% recursive case would be recursive calls
-
-% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff
-
-% skip splitting.
-
-% example is partition in 6.3
-
-In order to be correct, any code at \bbonem\ must be split with the 
-pessimistic assumption that it may result in work at the second stage.
-When this assumption turns out not to be true---that is, purely monostage code---then 
-splitting may produce second stage code which is needlessly costly 
-(an example of this will occur in \ref{sec:exampleQS}).
-Detecting and optimizing this case is in general a global program analysis,
-since functions can be passed around as values.  
-So instead of relying on hefty analysis, 
-we take the approach of adding enough structure to the input language's type system to
-allow the input code to {\em prove} itself to be---in some parts---monostage.
-
-%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
-%Yet even if a language designer desired a different implicit/explicit trade-off,
-%then the three-world \lang\ would still be useful as a typed intermediate representation.
+The \texttt{partition} function in the \texttt{quickselect} example in
+\ref{sec:exampleQS} is an example of such a function that contains no
+second stage computations, but, if not defined within a $\pure$
+annotation, would cause splitting to generate a recursive resumer.
 

### 2015-02-27 paper/splitting.tex
Add purely.  Carlo has a more ideas as well.
Wordsmith unit in 4.2.

@@
 
-We now show how to translate terms $\colmix e A$ into the form $\pipeM c l r$,
-pursuant to our splitting invariant.
-The algorithm is specified by the $\splitonesym$ relation (\cref{fig:termSplit}), 
-which proceeds recursively on the structure of~$e$. We derive a few cases here.
+We now describe how to split \bbonem\ terms. The algorithm, given in
+\cref{fig:termSplit}, proceeds recursively on the structure of~$e$. We
+derive a few cases here.
 
-\paragraph {Unit} 
-Consider splitting the unit value, $\tup{}$.
-Masking says that units, being trivial, only split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
-So in order for the split $\splitone {\tup {}} A c {l.r}$ to be valid, 
-we need a $c$ and $l.r$ satisfying,
+\paragraph {Unit.} 
+As a simple first case, consider splitting the unit value, $\tup{}$
+which masks to $\mval {\tup{}} {\tup{}}$.
+%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
+Splitting must produce a $c$ and $l.r$ satisfying,
 \[
@@
 Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
-Splitting $\pure{e}$ follows in similar fashion, 
-where the immediate result is instead the result of evaluating $e$.
+Splitting $\pure{e}$ follows in similar fashion, except the immediate
+result is instead the result of evaluating the \bbonep\ term $e$.
 
-\paragraph {Injections} 
-Now consider splitting a non-terminal, $\inl {e}$,
-where $\diaone {e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, 
-and by the splitting invariant for $e$,
+\paragraph {Injections.}
+Now consider splitting the non-terminal, $\inl {e}$, where $\diaone
+{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$. Therefore:
 \[
@@
 
-\paragraph{Tuple} 
+\paragraph{Tuple.} 
 For an example with more than one argument, consider a tuple $\tup{e_1,e_2}$,
@@
 The splitting algorithm described in the previous subsections operates
-on the local structure of \lang\ terms.  One artifact of this design
+purely on the local structure of \lang\ terms.  One artifact of this design
 is that splitting \bbonem\ terms may generate resumers containing

### 2015-02-27 paper/splitting.tex
4.2 through injections.

@@
 \]
-Since, $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
-the validity of the split $\splitone {\inl e} A {c'} {l.r'}$ requires,
+Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
+splitting must produce a $c'$ and $l.r'$ satisfying,
 \[
@@
 \]
-To satisfy this, our algorithm uses the same boundary value as the subexpression,
-and so $c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}$ and $l.r' = l.r$.
-
-\paragraph{Tuple.} 
-For an example with more than one argument, consider a tuple $\tup{e_1,e_2}$,
-where $\diaone {e_k} {\xi_k} {v_k}$ and $\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$
-for $k = 1,2$.
-Likewise, the splitting invariants for $e_1$ and $e_2$ give
+Our algorithm simply emits the resumer of the subexpression ($l.r' =
+l.r$), and generates a first-pass term $c'$ that produces the same
+boundary value ($b$) as the subexpression. The entire first-pass term
+is:
+\[
+c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
+\]
+\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
+where $\diaone {e_k} {\xi_k} {v_k}$ and $\rtab {\xi_k} {v_k} \vsplito
+\mval {i_k} {q_k}$ for $k = 1,2$.  Inductively:
 \[

### 2015-02-27 paper/splitting.tex
More edits on injections.

@@
 
-We now describe how to split \bbonem\ terms. The algorithm, given in
-\cref{fig:termSplit}, proceeds recursively on the structure of~$e$. We
-derive a few cases here.
+Given the definition of masking, we now describe how to split
+\bbonem\ terms. The algorithm, given in \cref{fig:termSplit}, proceeds
+recursively on the structure of~$e$. We derive a few cases here.
 
@@
 Splitting $\pure{e}$ follows in similar fashion, except the immediate
-result is instead the result of evaluating the \bbonep\ term $e$.
+result now becomes the result of evaluating the \bbonep\ term $e$.
 
@@
 \]
-Our algorithm simply emits the resumer of the subexpression ($l.r' =
-l.r$), and generates a first-pass term $c'$ that produces the same
-boundary value ($b$) as the subexpression. The entire first-pass term
-is:
+Our algorithm generates a first-pass term $c'$ that produces the same
+boundary value ($b$) as the split of the subexpression $e$:
 \[
@@
 \]
+Splitting also generates the same resumer as the split of the
+subexpression ($l.r' = l.r$)
+
 \paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,

### 2015-02-27 paper/splitting.tex
Splitting text tightening.

@@
 and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
-the validity of the split $\splitone {\tup{e_1,e_2}} A {c} {l.r}$ requires,
+splitting generates:
 \[
@@
 \]
-To satisfy this, our algorithm uses a boundary value which is the tuple of the boundary values of the subexpressions,
-$b = \tup{b_1,b_2}$. This is deconstructed with a tuple pattern, $l.r = \tup{l_1,l_2}.\tup{r_1,r_2}$,
-and constructed using,
+The resulting boundary value is the tuple of the boundary values of
+the subexpressions, $b = \tup{b_1,b_2}$.  The value is 
+constructed using,
 \[
@@
 \]
+And deconstructed with a tuple pattern, $l.r = \tup{l_1,l_2}.\tup{r_1,r_2}$,
 

### 2015-02-27 paper/abstract.tex
working on the abstract

@@
+A staged program is a program where expressions are assigned to a
+stage corresponding to their order of evaluation.  For example, in a
+two-staged program, some of the inputs become available in stage-1 and
+others in stage-2.  Evaluation of such a program proceeds by
+evaluating all the stage-1 expressions to construct a stage-2
+expression, which is then evaluated in stage-2.  Programming
+languages, evaluation strategies, and type systems for staged programs
+have been studied extensively by the research on partial evaluation
+and meta programming.
+
+In this paper, we consider a relatively less considered evaluation
+strategy for staged programs by splitting a staged program into a
+multiple passes, where each pass corresponds to a stage of evaluation.
+For example, a 2-stage program would be split into two programs pass-1
+and pass-2, one for each phase, and evaluation proceeds by evaluating
+proceeds by evaluating pass-
+
 Many algorithms can perform useful work before receiving all of their inputs.

### 2015-02-27 paper/abstract.tex
pass on the abstract

--- 2015-02-22 paper/abstract.tex
@@
-Many algorithms can perform useful work before receiving all of their inputs.
-If we regard those inputs as arriving at different \emph{stages} of the
-computation, then we can \emph{split} those algorithms into, for each stage, a
-function performing all the work dependent only on the available inputs.
+A staged program is a program where expressions are assigned to a
+stage corresponding to their order of evaluation.  For example, in a
+two-staged program, some of the inputs become available in stage-1 and
+others in stage-2.  Evaluation of such a program proceeds by
+evaluating all the stage-1 expressions to construct a stage-2
+expression, which is then evaluated in stage-2.  Programming
+languages, evaluation strategies, and type systems for staged programs
+have been studied extensively in the areas of partial evaluation and
+meta programming.
 
-Prior work on splitting has been limited to simple, imperative languages; in
-this paper, we extend splitting to \lang, a typed lambda calculus equipped with
-a modal staging operator, allowing us to split recursive and higher-order
-functions.
+In this paper, we present techniques for splitting staged programs
+into a multi-pass program, where each pass consists of an un-staged,
+conventional program.  To evaluate a multi-pass program, we evaluate
+each pass, passing the output to next pass; the final pass produces
+the result.  This technique, called {\em pass separation} was
+conceived in the 80`s but J{\o}rring and Scherlis and has been studied
+but previously proposed technique all apply only to simple programs
+that lack important features such as recursion and first-class
+functions.  Our techniques, formalized by considering a staged lambda
+calculus, make no restrictions on the kinds of programs that can be
+written.  We also present an implementation.
 
-These new features allow us to express some familiar algorithms which, when
-split, yield asymptotic improvements; for example, quickselect of the $k$th
-smallest element of a list $l$ splits into (1) sorting $l$ into a binary search
-tree, then (2) finding the tree's $k$th leftmost element.
+The motivations for pass separation are 1) improvements in efficiency
+by assigning frequent computations to earlier passes so that they can
+be re-used (cheaply) later and, 2) performing such scheduling
+statically at compile time without loss of any generality.  To assess
+the effectiveness of our techniques for pass separation, we consider
+several examples and demonstrate that our implementation is able to
+perform pass separation statically as expected.  In a particularly
+interesting example, involving recursion and recursive data, we show
+that our techniques perform highly non-trivial algorithmic
+transformations, improving efficiency asymptotically by nearly a
+linear factor.
 
-% something about compositional, cross-cutting code
+% Many algorithms can perform useful work before receiving all of their inputs.
+% If we regard those inputs as arriving at different \emph{stages} of the
+% computation, then we can \emph{split} those algorithms into, for each stage, a
+% function performing all the work dependent only on the available inputs.
+
+% Prior work on splitting has been limited to simple, imperative languages; in
+% this paper, we extend splitting to \lang, a typed lambda calculus equipped with
+% a modal staging operator, allowing us to split recursive and higher-order
+% functions.
+
+% These new features allow us to express some familiar algorithms which, when
+% split, yield asymptotic improvements; for example, quickselect of the $k$th
+% smallest element of a list $l$ splits into (1) sorting $l$ into a binary search
+% tree, then (2) finding the tree's $k$th leftmost element.
+
+% % something about compositional, cross-cutting code

### 2015-02-27 paper/intro.tex
pass on the abstract

@@
 \begin{lstlisting}
-fun @$F_\textrm{staged}(x, y_0, y_1, \ldots, y_{m-1})$@ = 
+fun @$F_\textrm{multipass}(x, y_0, y_1, \ldots, y_{m-1})$@ = 
   let @$fx = f_1(x)$@
@@
 %
-The key difference between partial evaluation (or metaprogramming) and pass
-separation is that the latter can be performed without access to the first
-argument $x$: $F_\textrm{staged}$ works for any $x$, while $F_v$ is defined
-only for $x=v$. (Of course, when a particular $v$ \emph{is} known, one may
-always use partial evaluation to specialize $F_\textrm{staged}$ to that $v$.)
+The key difference between partial evaluation (or metaprogramming) and
+pass separation is that the latter can be performed without access to
+the first argument $x$: $F_\textrm{multipass}$ works for any $x$,
+while $F_v$ is defined only for $x=v$.  Note that unlike
+precomputation, frequency reduction does not require knowledge of
+arguments; if a particular $v$ is known, we can use partial evaluation
+to specialize $F_\textrm{multipass}$ to that $v$.
 

### 2015-02-27 paper/overview.tex
...

--- 2015-02-27 paper/example.tex
@@
 
-\section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
-\label{sec:example}
+\section{Overview}
+\label{sec:examplea}
+
+We present an overview of our approach by considering an example.  To
+highlight the ideas, we abstract over some technical details, which
+are fully formalized in the rest of the paper.  The example itself is
+also implemented in our implementation.
+
+\paragraph{Computing order statistics.}
 
@@
 
-\input{figures/quickselect}
+\input{figures/quickselect-aligned}
 
-Suppose that we wish to issue order statistics queries on a collection of items
-represented as a list \texttt{l}. We can use quickselect \cite{Hoare:1961}, an
-expected linear time algorithm which takes a list \texttt{l} and a rank
-\texttt{k} and returns the element with rank \texttt{k} in \texttt{l}.
-
-Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
-by using the first element as a pivot%
-\footnote{We assume that the list is prepermuted to guarantee the expected
-linear time behavior.}
-and then recurs on one of the two resulting sides to find the desired element.
-The side chosen is determined by the relationship of \texttt{k} to the size of
-the first half \texttt{n}, which is returned by the \texttt{part}ition function
-along with the two sides themselves.
-
-Now, consider an application where we perform many order statistics queries on
-the same collection \texttt{l}, but with $m$ different ranks
-$\mathtt{k_1},\dots,\mathtt{k_m}$.
-Certainly, it is possible to implement this with $m$ calls to
-\texttt{quickselect}:
+Suppose that we wish to compute a series of order statistics queries
+on a list \texttt{l}. To this end, we can use
+quickselect~\cite{Hoare:1961}, which give a list \texttt{l} and a rank
+\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
+As implemented in an ML-like language in \ref{fig:qs-unstaged},
+function \texttt{qs} partitions the list by using the first element as
+a pivot and then recurs on one of the two resulting sides to find the
+desired element based on the relationship of \texttt{k} to the size of
+the first half \texttt{i}.  Assuming that the input is uniformly
+randomly ordered (which can be achieved by pre-permutation of the
+input), function \texttt{qs} runs in expected linear time.
+%
+\footnote{We assume that the list is prepermuted to guarantee the
+  expected linear time behavior.} 
+%
+Using function \texttt{qs}, we can perform many order statistics
+queries, for example but with $m$ different ranks
+$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
 %
 \begin{lstlisting}
-quickselect l @$\mathtt{k_1}$@
-quickselect l @$\mathtt{k_2}$@
- @$\vdots$@ 
-quickselect l @$\mathtt{k_m}$@.
+qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
 \end{lstlisting}
 %
-Can we do better?
+Unfortunately, this approach requires $O(n \cdot m)$ time, making it
+expensive to perform order statistics on more than a small number of
+times. This is unfortunate, because for example, because we wish to be
+able to map \texttt{qs} to over another list and do so
+efficiently.
+
+One way to regain efficiency is to change the algorithm that we use to
+perform the computation in multiple passes.  One option would be to
+pre-sort the input list \texttt{l} into a sorted list \texttt{s} in
+the first pass, and then perform lookups in \texttt{s} in the second
+pass.  Unfortunately, since \texttt{s} is a list, a lookup would
+require linear time (also on average), leading to no improvement in
+efficiency.  We can resolve this issue by using a data structure that
+can provide faster access.  For example, after sorting \texttt{s} we
+can copy in into an array and then perform binary searches to find the
+element with the desired rank (copying the input to an array and
+sorting would not work well in a functional language).  As another
+option, we can represent construct a binary search tree from the input
+and perform size-based searches on the binary search tree to find the
+elements with the desired rank.  Such an approach would fit nicely
+into the functional paradigm of computing but requires additional
+algorithmic ingenuity: to ensure efficiency we would need to store at
+each note the size of its subtree (these are sometimes called
+``augmented trees'').  In summary, we are able to improve efficiency
+by replacing our simple solution with a multi-pass algorithm that
+first preprocesses the input to generate a lookup data structure and
+then in the second pass performs fast lookups.  As outlined above, in
+addition, this transformation is highly non-trivial as it involves
+reasoning about intricate algorithmic concerns, and implementing more
+complex algorithms.
+
+Such transformations, called pass-separation by J{\o}rring and
+Scherlis are commonly employed.  For example, as briefly mentioned in
+the introduction, modern graphics software is written exactly in this
+way.  Unfortunately, as the example illustrates, they lead to complex
+software, leading to the question of whether it would be possible to
+perform such transformations automatically.  We answer this question
+affirmatively.  Our approach is to start with a staged language and
+then use a splitting algorithm to generate multi-pass programs from
+staged programs.
 
@@
 
-An astute programmer might notice that much of the code in \texttt{quickselect}
-does \emph{not} depend on the rank \texttt{k}; for example, the list is
-partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
-to \texttt{qSelect} are guarded by a comparison to \texttt{k}, \texttt{k} does
-not determine the arguments to those recursive calls---it only determines which
-call is made. Thus, if we are willing to change the evaluation order of the
-language (and evaluate under the \texttt{case} statement), it should be possible
-to force \emph{all} the computations involving \texttt{l} to be performed before
-\emph{any} of the computations involving \texttt{k}.
+The idea behind staged programming is to use staging annotations or
+types so indicate the stage of each expression.  To apply this
+technique to our example, suppose that we have a language with two
+stages.  We can now write a staged version of \texttt{qs}, called
+\texttt{qss} in \ref{fig:qs-staged}, making the input list a stage-1
+and the rank a stage-2, more precisely the input list has type
+$\curr\mathrm{list}$ (an integer list ``now''), the rank has type
+$\fut\rmint$ (an integer in the ``future''), and return has type
+$\fut\rmint$.  We can then annotate the rest of the body of
+\texttt{qs}, by using the staging annotations \texttt{prev} and
+\texttt{next}, which respectively include a stage-1 expressions in
+stage-2, and a stage-2 expression in stage-1.  To help readability, we
+write the first-stage computations in red, and the second-stage
+computations in blue.
+
+the
+function \texttt{qs} does \emph{not} depend on the rank \texttt{k};
+for example, the list is partitioned before \texttt{k} is ever
+used. Moreover, while the recursive calls to \texttt{qSelect} are
+guarded by a comparison to \texttt{k}, \texttt{k} does not determine
+the arguments to those recursive calls---it only determines which call
+is made. Thus, if we are willing to change the evaluation order of the
+language (and evaluate under the \texttt{case} statement), it should
+be possible to force \emph{all} the computations involving \texttt{l}
+to be performed before \emph{any} of the computations involving
+\texttt{k}.
 
@@
 to implement a binary search to perform the lookup.  Another option is
-to use a balanced binary search tree data structure, which supports
-lookups in logarithmic time.  Using either one of these techniques, we
-can reduce the run-time to $O(n\log{n} + m\log{n})$, which is
-significantly better that $O(n^2)$.
+to implement a balanced binary search tree data structure for storing
+the elements in the list and then use balanced binary-search-tree
+lookups, which are logarithmic time.  Using either one of these
+techniques, we can reduce the run-time to $\Theta(n\log{n} + m\log{n})$,
+which is significantly better that $\Theta(n^2)$.
 
@@
 tree) and requires implementing additional algorithms such as sorting
-and binary search, also possibly binary search trees.  As suggested by
-this relatively small example and experienced by the graphics
-community, the design, implementation, and maintenance of such
-multi-pass programs can be difficult due to the increased complexity.
+and binary search or possibly binary search trees.
+
+%% As suggested by this relatively small example and experienced for
+%% example by the graphics community, the design, implementation, and
+%% maintenance of such multi-pass programs can be difficult due to the
+%% increased complexity.
 
@@
 algorithm yields a two-pass program that uses the binary-search-tree
-based implementation described above.  Specifically, in the first
-pass, the program takes the input list and constructs a
-probabilistically balanced binary search tree, which is isomorphic to
-a treap data structure~\cite{treaps}.  In the second pass, the program
-performs a binary search tree lookup, by walking the tree to find the
-element with the desired rank.  The resulting program takes the
-$O(n\log{n} + m\log{n})$.
+based implementation outlined above.  Specifically, in the first pass,
+the program takes the input list and constructs a probabilistically
+balanced binary search tree, which is isomorphic to a treap data
+structure~\cite{treaps}.  In the second pass, the program performs,
+for each rank, a binary search tree lookup, by walking the tree to
+find the element with the desired rank.  The resulting program, shown
+in \ref{fig:????}  takes the $O(n\log{n} + m\log{n})$.
 
-We note that the splitting algorithm operates by composing local
-transformations on the subterms of the input program.  In particular,
-the algorithm has no special knowledge of the quick-select algorithm,
-binary search trees, or how to perform lookups on binary search trees,
-but it is able to derive all of these from the input program.  But
-how?
-
-The technical detail of the algorithm are relatively involved but in
-the hope that it aids understanding we present a brief informal
-overview of how the algorithm operates.  This exposition is
-necessarily imprecise but the algorithm is made precise in
-\ref{sec:splitting} and implemented
-(\ref{sec:implementation}{sec:examples}).
+To create the multi-pass algorithm, the splitting algorithm operates
+by composing local transformations on the subterms of the input
+program.  In particular, the algorithm has no special knowledge of the
+quick-select algorithm, binary search trees, or how to perform lookups
+on binary search trees, but it is able to derive all of these from the
+input program.  But how?
+%
+The details of the algorithm are relatively involved but in the hope
+that it aids understanding we present a brief informal overview of how
+the algorithm operates.  
+%
+%% This exposition is necessarily imprecise but the algorithm is made
+%% precise in \ref{sec:splitting} and implemented
+%% (\ref{sec:implementation}{sec:examples}).
+%
 
 At a high level, the splitting algorithm operates on our example by
-using the code for quickselect and the knowledge that the list (the
-first argument) is a first-stage argumint and the rank (second
-argument) is a second stage argumunt, the algorithm then derives two
-functions: \texttt{qsStage1} and \texttt{qsStage2} as follows.
+using the program code and the knowledge that the list (the first
+argument) is first-stage and the rank ( the second argument) is second
+stage, the algorithm then derives two functions: \texttt{qsStage1} and
+\texttt{qsStage2} as follows.
 
 \paragraph{Pass-1 code.}
-The splitting algorithm ``imitates'' an execution of quick-select with
-just the first-stage input by evaluating as much code as possible with
-the available information.  Specifically the algorithm generates a
-function \texttt{qsStage1} to perform all the recursive calls of
-\texttt{qs} and to evaluate all instances of the partition function,
-which depend only on the input list.  The function \texttt{qsStage1}
-produces a trace of the execution of \texttt{qs} on its first input by
-generating a tree that collects the results from all recursive calls
-along with a tag that indicates the control branches taken.  The trace
-ends up being represented as a tree because \texttt{qs} has a binary
-control structure casing on the list. In fact, function
-\texttt{qsStage1} constructs a binary search tree, keyed by the
-``pivot'' used by the partition, because the ``pivot'' is needed in
-the second stage to construct the result.  We call the result data
-structure of \texttt{qsStace1} the {\em boundary}, as it passes
+The splitting algorithm scans the program and imitates an execution of
+\texttt{qs} with just the first-stage input by evaluating as much code
+as possible with the available first-stage values.  Specifically the
+algorithm generates a function \texttt{qsStage1} to perform all the
+recursive calls of \texttt{qs} and to evaluate all instances of the
+partition function, which depend only on the input list.  The function
+\texttt{qsStage1} produces a trace of the execution of \texttt{qs} on
+its first input by generating a tree that collects the results from
+all recursive calls along with a tag that indicates the control
+branches taken.  The trace is represented as a tree because
+\texttt{qs} has a binary control structure (casing on the list), which
+leads to a binary tree data structure.  To ensure that the trace
+contains the necessary results to complete the execution in the second
+pass, where second-stage values may be used, the function
+\texttt{qsStage1} includes in the nodes of the tree information such
+as the key used for splitting the list at that recursive call.  The
+resulting tree is thus a binary search tree, keyed by the ``pivot''
+used by the partition.  In general we refer to the result data
+structure of \texttt{qsStace1} as the {\em boundary}, as it passes
 information from the first pass to the second.
@@
 \paragraph{Pass-2 code}
-As the splitting algorithm ``scans'' \texttt{qs} for computations that
-can be performed in the first stage and collects them into
+As the splitting algorithm scans \texttt{qs} for computations that can
+be performed in the first stage and collects them into
 \texttt{qsStage1}, it also collects computations that must be left to
@@
 \texttt{qs} that can now be performed in the presence of the
-second-stage argument (the rank).
+second-stage argument (the rank).  In the context of our example, this
+performs a lookup on the boundary data structure by using the supplied
+rank.
+
+\paragraph{Summary.}
+In summary, the splitting algorithm scans the program code for
+first-stage computations (which depend only on first-stage values) and
+separates them into function of the first pass. This function performs
+the stage-1 computations and places the results into a boundary data
+structure that both records the control flow and the results from the
+first stage computation at each control-flow point.  As the splitting
+function generates the code for the first pass, it also generates the
+code for the second pass as a function that walks both the boundary
+data (which essentially imitates execution) and at each control point
+performs the part of the computation that depends on second-stage
+values. We evaluate the resulting program by first evaluating the
+function for the first pass and passing its return value (the
+boundary) to the second pass, which then returns the desired result.
 

### 2015-02-27 paper/conclusion.tex
conclusion

\section{Conclusion}

This paper presents techniques to translate staged lambda calculus
programs to multi-pass programs, which evaluate in a sequence of
passes, each of which is a conventional program.  We present a
formalization of the techniques in a way that can be implemented in
practice.  While we have also provided an implementation that
successfully splits the examples considered, we have stopped short of
offering a full proof of correctness, which we leave for future work.
We are also interested in providing a more complete implementation and
investigating how these techniques can be integrated into existing
practical systems such as existing staged programming systems for
computer graphics.  On the more conceptual side, we plan to extend the
techniques to richer source languages, for example those that include
parallelism.

### 2015-02-27 paper/conclusion.tex
Add conclusion to paper.
One editing pass over conclusion.

@@
 
-This paper presents techniques to translate staged lambda calculus
-programs to multi-pass programs, which evaluate in a sequence of
-passes, each of which is a conventional program.  We present a
+This paper presents an algorithm for translating programs written in a
+staged lambda calculus into multi-pass programs, which evaluate as two
+monostage programs in a sequence of two dependent passes. We present a
 formalization of the techniques in a way that can be implemented in
-practice.  While we have also provided an implementation that
-successfully splits the examples considered, we have stopped short of
-offering a full proof of correctness, which we leave for future work.
-We are also interested in providing a more complete implementation and
-investigating how these techniques can be integrated into existing
-practical systems such as existing staged programming systems for
-computer graphics.  On the more conceptual side, we plan to extend the
-techniques to richer source languages, for example those that include
-parallelism.
+practice.  While we have provided an implementation of this algorithm
+that successfully splits the examples considered in this paper, we
+have stopped short of offering a full proof of correctness, which we
+leave for future work.
+
+In addition, we are interested in providing a more complete system
+implementation as well as investigating how ideas from \lang\ and the
+its splitting algorithm can be integrated into existing practical
+systems such as staged programming systems for computer graphics
+pipelines.  On the more conceptual side, it would be interesting to
+consider the extension of our splitting techniques to richer source
+languages, for example, those that include parallelism.

### 2015-02-27 paper/abstract.tex
pass on abstract

--- 2015-02-22 paper/abstract.tex
@@
-Many algorithms can perform useful work before receiving all of their inputs.
-If we regard those inputs as arriving at different \emph{stages} of the
-computation, then we can \emph{split} those algorithms into, for each stage, a
-function performing all the work dependent only on the available inputs.
+Staged program languages assign a stage to each expression and
+evaluate each expression in its assigned stage.  For example, in a
+two-staged program, some expressions are evaluated in the first stage
+and others in the second stage.  Programming languages, evaluation
+strategies, and type systems for staged programs where stages roughly
+correspond to the arrival order of inputs have been studied
+extensively in the areas of partial evaluation and meta programming.
 
-Prior work on splitting has been limited to simple, imperative languages; in
-this paper, we extend splitting to \lang, a typed lambda calculus equipped with
-a modal staging operator, allowing us to split recursive and higher-order
+In this paper, we consider staged programs written in a typed lambda
+calculus with a (``circle'') modality denoting computation in a later
+stages and present techniques for splitting them into multi-pass
+programs. Such multi-pass programs evaluate in a sequence of
+evaluations of un-staged, conventional programs each of which
+corresponds to a pass.  Conceived in the 80`s but J{\o}rring and
+Scherlis, this {\em pass-separation} technique has been studied but
+existing techniques are either manual or apply only to simple programs
+that lack important features such as recursion and first-class
 functions.
 
-These new features allow us to express some familiar algorithms which, when
-split, yield asymptotic improvements; for example, quickselect of the $k$th
-smallest element of a list $l$ splits into (1) sorting $l$ into a binary search
-tree, then (2) finding the tree's $k$th leftmost element.
+% We make no restrictions on the kinds of programs that can
+% be written.  We also present an implementation.
+% %
 
-% something about compositional, cross-cutting code
+Since we build on a lambda calculus, our techniques enable the
+programmer to write staged programs by using powerful programming
+abstractions in a composable and modular fashion. Our splitting
+transformation maps such programs into multi-pass programs that can
+run significantly more efficiently by assigning frequent computations
+to earlier passes so that they can be re-used (cheaply) later.  Our
+experiments based on a prototype implementation show that our
+splitting algorithm is able perform highly non-trivial algorithmic
+transformations, sometimes automatically improving efficiency
+asymptotically by nearly a linear factor.  Since splitting
+transformation takes place at compile time, these benefits are
+achieved without significant run-time penalty.
+
+% The motivations for pass separation are 1) improvements in efficiency
+% by assigning frequent computations to earlier passes so that they can
+% be re-used (cheaply) later and, 2) performing such scheduling
+% statically at compile time without loss of any generality.  To assess
+% the effectiveness of our techniques for pass separation, we consider
+% several examples and demonstrate that our implementation is able to
+% perform pass separation statically as expected.  In a particularly
+% interesting example, involving recursion and recursive data, we show
+% that our techniques perform highly non-trivial algorithmic
+% transformations, improving efficiency asymptotically by nearly a
+% linear factor.
+
+% Many algorithms can perform useful work before receiving all of their inputs.
+% If we regard those inputs as arriving at different \emph{stages} of the
+% computation, then we can \emph{split} those algorithms into, for each stage, a
+% function performing all the work dependent only on the available inputs.
+
+% Prior work on splitting has been limited to simple, imperative languages; in
+% this paper, we extend splitting to \lang, a typed lambda calculus equipped with
+% a modal staging operator, allowing us to split recursive and higher-order
+% functions.
+
+% These new features allow us to express some familiar algorithms which, when
+% split, yield asymptotic improvements; for example, quickselect of the $k$th
+% smallest element of a list $l$ splits into (1) sorting $l$ into a binary search
+% tree, then (2) finding the tree's $k$th leftmost element.
+
+% % something about compositional, cross-cutting code

### 2015-02-27 paper/intro.tex
GPU pipeline architecture was described in enough detail to require a citation.

@@
 the screen, and per pixel, where many computations in later passes use the
-results from a single computation in an earlier pass).
+results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
 

### 2015-02-27 paper/overview.tex
working on the overview

@@
 The idea behind staged programming is to use staging annotations or
-types so indicate the stage of each expression.  To apply this
+types to indicate the stage of each expression.  To apply this
 technique to our example, suppose that we have a language with two
-stages.  We can now write a staged version of \texttt{qs}, called
-\texttt{qss} in \ref{fig:qs-staged}, making the input list a stage-1
-and the rank a stage-2, more precisely the input list has type
-$\curr\mathrm{list}$ (an integer list ``now''), the rank has type
-$\fut\rmint$ (an integer in the ``future''), and return has type
-$\fut\rmint$.  We can then annotate the rest of the body of
-\texttt{qs}, by using the staging annotations \texttt{prev} and
-\texttt{next}, which respectively include a stage-1 expressions in
-stage-2, and a stage-2 expression in stage-1.  To help readability, we
-write the first-stage computations in red, and the second-stage
-computations in blue.
+worlds representing the two stages and a third, {\em ground} world
+that contains terms with no staging.
+
+We can write a staged version of \texttt{qs}, called
+\texttt{qss} in \ref{fig:qs-staged}, by making the input list a
+stage-1 and the rank a stage-2 argument. 
+%
+More precisely the input list
+has type $\curr\mathrm{list}$ (an integer list ``now''), the rank has
+type $\fut\rmint$ (an integer in the ``future''), and return has type
+$\fut\rmint$.  
+%
+We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
+with the staging annotations \texttt{prev} and \texttt{next}, which
+transition between a stage-1 and stage-2 code, and $\texttt{g}$, which
+marks ground terms.  
+%
+We also use a third function
+\texttt{hold}, which can be implemented with \texttt{prev} and
+\texttt{next}, for promoting a stage-1 integer to a stage-2 integer;
+the signature of \texttt{hold} is
+% 
+\lstinline{1`hold : ^int -> $`2`int`}.
+
+To help readability, we write the first-stage
+computations in red, and the second-stage computations in blue.
+
+Our type system ensures that the staging annotations are consistent,
+in the sense that computations marked as first-stage cannot depend on
+ones marked as second-stage.
+%
+The process of adding staging annotations to unstaged code has been
+the subject of extensive research (\secref{related}). We assume that
+these annotations have been provided; we do not consider the problem
+of generating them. In this example, there are other ways to annotate
+\texttt{qs}, but we chose annotations that maximize the work performed
+in the first stage
+
+% The $\curr$ and \texttt{grnd} (``ground'') annotations
+% further distinguish those first-stage computations which do not
+% contain any embedded second-stage code. (Code \emph{inside}
+% \texttt{grnd}, like the \texttt{part}ition function, is guaranteed not
+% to contain second-stage code; first stage code outside a
+% \texttt{grnd}, like \texttt{qsStaged}, may.) We will discuss the
+% importance of this additional distinction in \ref{sec:splitting}.
+
+% Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
+% \texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
+% non-circle type are available for immediate use. Indeed, we case on the list as
+% usual, and in the \texttt{Cons} branch, \texttt{part}ition it.
 
@@
 
-We can formalize this intuition by writing quickselect in a \emph{staged}
-language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
-lambda calculus. This allows us to directly express the idea that the argument
-\texttt{l} is known at the \emph{first} stage of the computation, while the
-argument \texttt{k} is only known at the \emph{second}, and all
-first-stage computations occur before any second-stage computations.
-
-We define a staged version of quickselect in \ref{fig:qs-staged}, writing
-first-stage computations in red, and second-stage computations in blue. While
-\texttt{qSelect} sends a \textrm{list} and $\rmint$ to an $\rmint$,
-\texttt{qsStaged} has a more precise type---it is a first-stage function which
-takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$
-(an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
-
-Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
-\texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
-non-circle type are available for immediate use. Indeed, we case on the list as
-usual, and in the \texttt{Cons} branch, \texttt{part}ition it.
-
 The \texttt{case} expression in \texttt{qSelect} depends on \texttt{k}, whose
@@
 second-stage $\rmint$.
-
-The process of adding staging annotations ($\fut$ types, $\next$, and $\prev$)
-to unstaged code has been the subject of extensive research under the
-name of \emph{binding time analysis}. In this paper, we assume that these
-annotations have been provided by the programmer (or perhaps a binding time
-analysis tool), and do not consider the problem of generating such annotations.
-There are many ways to annotate any program, including \texttt{qSelect}; we
-chose annotations which maximize the work performed in the first stage.
-
-\lang's type system, discussed in \ref{sec:semantics}, ensures that the staging
-annotations are consistent, in the sense that computations marked as
-first-stage cannot depend on ones marked as second-stage. The $\curr$ and
-\texttt{grnd} (``ground'') annotations further distinguish those first-stage computations
-which do not contain any embedded second-stage code. (Code \emph{inside}
-\texttt{grnd}, like the \texttt{part}ition function, is guaranteed not to
-contain second-stage code; first stage code outside a \texttt{grnd}, 
-like \texttt{qsStaged}, may.) We will
-discuss the importance of this additional distinction in \ref{sec:splitting}.
 

### 2015-02-27 paper/splitting.tex
described next and prev splitting more technically

@@
 $l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
-Again, this works even for nullary terms and under binders.
 
-Splitting \texttt{prev} generates a precomputation that projects the immediate
-result of its world \bbone\ subterm.
-Since the argument to \texttt{prev} is of $\fut$ type, its immediate result reduces to $\tup{}$, justifying why it can be thrown away.
-Finally, splitting \texttt{next} simply tuples up the precomputation of its
-world \bbtwo\ subterm with a trivial immediate result $\tup{}$.
+Again, this holds even for nullary terms and under binders.
+In particular, \texttt{case} splits according to the rule:
+\[
+\infersplittwo	{	\begin{array} {l}
+						\caseof{e_1}{x_2.e_2}{x_3.e_3} \splittwosym \\
+						\hspace{5em} \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}		
+					\end{array}}	
+				{\sub 1 ? & \sub 2 ? & \sub 3 ?} 								
+\]
+Note how the precomputations are pulled out from underneath the binder into a tuple that is always executed in the first stage,
+regardless of what branch is eventually executed in the second stage.
+
+\paragraph{Next and Prev}
+Since \texttt{next} and \texttt{prev} do nothing more than convert between 
+second stage terms at \bbtwo\ and second stage terms encapsulated within $\fut$s at \bbonem,
+their splitting rules are relatively simple.
+
+Given an $\splittwo e ? p l r$ and $\diatwo e q$ for which the splitting invariant holds, we know
+\[
+	\reduce p b \text{ and } [b/l]r \equiv q
+\]
+The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$ 
+and $\diaone {\next {e'}} {\yhat \mapsto q} {\next \yhat}$
+and $\rtab {\yhat \mapsto q} {\next \yhat} \vsplito \mval {\tup{}} {\letin \yhat q \yhat}$
+is
+\[
+	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin \yhat q \yhat
+\]
+which can be fulfilled using $c' = \tup{\tup{},p}$ and $l.r = l'.r'$.
+Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
 

### 2015-02-27 paper/splitting.tex

--- 2015-02-27 paper/splitting.tex
@@
 $l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
-Again, this works even for nullary terms and under binders.
+
+Again, this holds even for nullary terms and under binders.
+In particular, \texttt{case} splits according to the rule:
+\[
+\infersplittwo	{	\begin{array} {l}
+						\caseof{e_1}{x_2.e_2}{x_3.e_3} \splittwosym \\
+						\hspace{5em} \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}		
+					\end{array}}	
+				{\sub 1 ? & \sub 2 ? & \sub 3 ?} 								
+\]
+Note how the precomputations are pulled out from underneath the binder into a tuple that is always executed in the first stage,
+regardless of what branch is eventually executed in the second stage.
 

### 2015-02-27 paper/splitting.tex

--- 2015-02-27 paper/splitting.tex
@@
 Our algorithm generates a first-pass term $c'$ that produces the same
-boundary value ($b$) as the split of the subexpression $e$:
+boundary value ($b$) as the split of the subterm $e$:
 \[
@@
 Splitting also generates the same resumer as the split of the
-subexpression ($l.r' = l.r$)
+subterm ($l.r' = l.r$)
 
@@
 The resulting boundary value is the tuple of the boundary values of
-the subexpressions, $b = \tup{b_1,b_2}$.  The value is 
+the subterms, $b = \tup{b_1,b_2}$.  The value is 
 constructed using,

### 2015-02-27 paper/overview.tex

@@
 The process of adding staging annotations to unstaged code has been
-the subject of extensive research (\secref{related}). We assume that
+the subject of extensive research (\ref{sec:related}). We assume that
 these annotations have been provided; we do not consider the problem
@@
 find the element with the desired rank.  The resulting program, shown
-in \ref{fig:????}  takes the $O(n\log{n} + m\log{n})$.
+in \ref{fig:qs-split}  takes the $O(n\log{n} + m\log{n})$.
 

### 2015-02-27 paper/implementation.tex
more small cleanups

@@
 \begin{lstlisting}
-1`atsigngrnd{
+3`atsigngrnd{`1`
   datatype list = Empty | Cons of int * list
   fun part (...) = ...
-}
+`3`}`1`
 fun qsStaged (...) = ...`
@@
 \begin{lstlisting}
-1`val grnd{Empty} = grnd{roll (inj ...)}
-val grnd{Cons}  = grnd{fn (...) => roll (inj ...)}
-val grnd{part}  = grnd{fn (...) => ...}
+1`val 3`grnd{`Empty3`}` = 3`grnd{`roll (inj ...)3`}`
+val 3`grnd{`Cons3`}`  = 3`grnd{`fn (...) => roll (inj ...)3`}`
+val 3`grnd{`part3`}`  = 3`grnd{`fn (...) => ...3`}`
 val qsStaged    = fn (...) => ...`

### 2015-02-27 paper/semantics.tex
fixing various broken things
more small cleanups

@@
 
-\paragraph{First-stage evaluation at \bbonem\ ($\redonesym$)}
+\paragraph{First-stage evaluation at \bbonem\ ($\redonesym$).}
 The difficulty of evaluating terms at \bbonem\ is that they contain both terms
@@
 
-\paragraph{First-stage evaluation at \bbtwo\ ($\redtwosym$)}
+\paragraph{First-stage evaluation at \bbtwo\ ($\redtwosym$).}
 Since terms at world \bbtwo\ represent second-stage computations, we should not
@@
 
-A multistage function, such as \texttt{qsStaged} from \ref{sec:example}, has
+A multistage function, such as \texttt{qsStaged} from \ref{sec:examplea}, has
 type $A\to \fut (B \to C)$ at \bbonem.%

### 2015-02-27 paper/overview.tex
pass on overviet

--- 2015-02-27 paper/overview.tex
@@
 
-\input{figures/quickselect-split}
+\input{figures/quickselect-split-aligned}
 
-It is well-known that for any given input list with $n$ elements, the
-quick-select algorithm requires $\Theta(n)$ expected time (for any
-second argument) if we select keys for the partition function
-uniformly randomly or identically if we permute the input. Under this
-assumption, our example program requires expected $\Theta(nm)$ time.
+The staged quickselect code shown in \ref{fig:qs-staged} makes
+explicit the staging of all terms, making it natural to ask, whether
+we can transform this code into manually implemented multi-pass code
+described above and match the efficiency improvements of the manual
+implementation. Considering the algorithmic knowledge and the
+considerations needed, this may seem like a tall order.
+Interestingly, our splitting algorithm achieves exactly this.
 
-Suppose now that $m$ is large (close to $n$ for example).  In this
-case, our program is essentially a quadratic-time algorithm, which can
-be very slow even for modest values of $m$ and $n$. In this case, to
-improve the run-time it appears that we have no choice but rewrite our
-program to use a different algorithm.  One way to do this would be to
-pre-sort the input list $l$ and then lookup the element with the
-desired rank (as specified by the second argument).  While this
-approach might sound sensible, attaining the desired performance
-requires answering one more important question: what should the data
-structure for storing the sorted input be?  Unfortunately using a list
-does not work, because each look up in a list requires linear time on
-average. One option is to use arrays but in this case, we would need
-to implement a binary search to perform the lookup.  Another option is
-to implement a balanced binary search tree data structure for storing
-the elements in the list and then use balanced binary-search-tree
-lookups, which are logarithmic time.  Using either one of these
-techniques, we can reduce the run-time to $\Theta(n\log{n} + m\log{n})$,
-which is significantly better that $\Theta(n^2)$.
-
-In summary, to achieve the desired improvement, we wrote a two-pass
-program, where the first pass sorts the input and the second pass
-performs a binary search either on an array or a balanced tree. As is
-probably apparent to the reader, the resulting program is both
-algorithmically and structurally more complex: it requires conversions
-between different data structures (list to array or binary search
-tree) and requires implementing additional algorithms such as sorting
-and binary search or possibly binary search trees.
-
-%% As suggested by this relatively small example and experienced for
-%% example by the graphics community, the design, implementation, and
-%% maintenance of such multi-pass programs can be difficult due to the
-%% increased complexity.
-
-Instead of relying an writing multipass programs manually, in this
-paper, we present techniques for automatically splitting staged
-programs into multi-pass programs.  The key innovation behind our
-approach is a {\em splitting algorithm} that splits a multistage
-program into separate passes where each pass is a conventional program
-expressed in a conventional functional language.  Our splitting
-algorithm makes no restrictions on the input program: essentially any
-functional program, including higher order programs, can be split.
-Furthermore, the splitting algorithm is able to perform highly
-non-trivial transformations on the program that can improve efficiency
-significantly.
-
-When applied to the quick-select example as described, our splitting
+In the rest of this section, we present a brief, high-level overview
+of the main ideas behind this algorithm using the quickselect example.
+%
+When applied to the staged code in \ref{fig:qs-staged}, our splitting
 algorithm yields a two-pass program that uses the binary-search-tree
-based implementation outlined above.  Specifically, in the first pass,
-the program takes the input list and constructs a probabilistically
+based implementation.  Specifically, in its first pass, the two-pass
+program takes the input list and constructs a probabilistically
 balanced binary search tree, which is isomorphic to a treap data
@@
 for each rank, a binary search tree lookup, by walking the tree to
-find the element with the desired rank.  The resulting program, shown
-in \ref{fig:????}  takes the $O(n\log{n} + m\log{n})$.
+find the element with the desired rank.  
+%
+\ref{fig:qs-staged} illustrates the code for the first and second
+passes \texttt{qs1} and \texttt{qs2} of the staged function
+\texttt{qss}.
 
@@
 program.  In particular, the algorithm has no special knowledge of the
-quick-select algorithm, binary search trees, or how to perform lookups
+quickselect algorithm, binary search trees, or how to perform lookups
 on binary search trees, but it is able to derive all of these from the
-input program.  But how?
-%
-The details of the algorithm are relatively involved but in the hope
-that it aids understanding we present a brief informal overview of how
-the algorithm operates.  
-%
-%% This exposition is necessarily imprecise but the algorithm is made
-%% precise in \ref{sec:splitting} and implemented
-%% (\ref{sec:implementation}{sec:examples}).
+input program.
 %
 
-At a high level, the splitting algorithm operates on our example by
-using the program code and the knowledge that the list (the first
-argument) is first-stage and the rank ( the second argument) is second
-stage, the algorithm then derives two functions: \texttt{qsStage1} and
-\texttt{qsStage2} as follows.
+The splitting algorithm scans the program code for first-stage
+computations (which depend only on first-stage values) and separates
+them into the function, \texttt{qs1} of the first pass. This function
+performs the stage-1 computations and places the results into a
+boundary data structure that both records the control flow and the
+results from the first stage computation at each control-flow point.
+The function \texttt{qs1} performs all the recursive calls and
+evaluate all instances of the partition function, which depend only on
+the input list.  The function produces a boundary data structure that
+collects the results from all recursive calls along with a tag that
+indicates the control branches taken.  Since \texttt{qss} has a binary
+control structure (casing on the list), the boundary is a binary tree.
+To ensure that the trace contains the necessary results to complete
+the execution in the second pass, where second-stage values may be
+used, the splitting algorithm includes in the nodes of the tree
+information such as the "pivot" used for splitting and the size of the
+subtree at that note.  The resulting tree is thus not just a tree but
+a binary search tree, keyed by the pivot and augmented with "size"
+information.
 
-\paragraph{Pass-1 code.}
-The splitting algorithm scans the program and imitates an execution of
-\texttt{qs} with just the first-stage input by evaluating as much code
-as possible with the available first-stage values.  Specifically the
-algorithm generates a function \texttt{qsStage1} to perform all the
-recursive calls of \texttt{qs} and to evaluate all instances of the
-partition function, which depend only on the input list.  The function
-\texttt{qsStage1} produces a trace of the execution of \texttt{qs} on
-its first input by generating a tree that collects the results from
-all recursive calls along with a tag that indicates the control
-branches taken.  The trace is represented as a tree because
-\texttt{qs} has a binary control structure (casing on the list), which
-leads to a binary tree data structure.  To ensure that the trace
-contains the necessary results to complete the execution in the second
-pass, where second-stage values may be used, the function
-\texttt{qsStage1} includes in the nodes of the tree information such
-as the key used for splitting the list at that recursive call.  The
-resulting tree is thus a binary search tree, keyed by the ``pivot''
-used by the partition.  In general we refer to the result data
-structure of \texttt{qsStace1} as the {\em boundary}, as it passes
-information from the first pass to the second.
+As the splitting algorithm scans \texttt{qss} for computations that
+can be performed in the first stage, it also collects computations
+that must be left to the second stage in a separate function
+\texttt{qs2}.  This function, which is executed in the second pass,
+takes as argument the boundary and the second-stage argument (the
+rank) and performs a lookup in the boundary data structure.  But how
+does \texttt{qsStage2} knows to traverse the tree? It does not. Since
+the splitting algorithm recorded the control flow of the first stage
+execution in the boundary data structure, \texttt{qs2} simply follows
+this control structure and performs at each point the parts of the
+computation from \texttt{qss} that can now be performed in the
+presence of the second-stage argument (the rank).  In the context of
+our example, this performs a lookup on the boundary data structure by
+using the supplied rank.
 
-\paragraph{Pass-2 code}
-As the splitting algorithm scans \texttt{qs} for computations that can
-be performed in the first stage and collects them into
-\texttt{qsStage1}, it also collects computations that must be left to
-the second stage in a separate function \texttt{qsStage2}.  This
-function, which is executed in the second pass, takes as argument the
-boundary and the second-stage argument (the rank) and performs a
-lookup in the boundary data structure.  But how does \texttt{qsStage2}
-knows to traverse the tree? It does not. Since the splitting algorithm
-recorded the control flow of the first stage execution in the boundary
-data structure, \texttt{qsStage2} simply follows this control
-structure and performs at each point the parts of the computation from
-\texttt{qs} that can now be performed in the presence of the
-second-stage argument (the rank).  In the context of our example, this
-performs a lookup on the boundary data structure by using the supplied
-rank.
-
-\paragraph{Summary.}
-In summary, the splitting algorithm scans the program code for
-first-stage computations (which depend only on first-stage values) and
-separates them into function of the first pass. This function performs
-the stage-1 computations and places the results into a boundary data
-structure that both records the control flow and the results from the
-first stage computation at each control-flow point.  As the splitting
-function generates the code for the first pass, it also generates the
-code for the second pass as a function that walks both the boundary
-data (which essentially imitates execution) and at each control point
-performs the part of the computation that depends on second-stage
-values. We evaluate the resulting program by first evaluating the
-function for the first pass and passing its return value (the
-boundary) to the second pass, which then returns the desired result.
+\begin{theorem}
+  Consider an execution of \texttt{qs1} with a randomly permuted input
+  of $n$ keys and performing $m$ executions of \texttt{qs2} with the
+  result of \texttt{qs1}.  The total run-time for this computation is
+  $\Theta(n\log{n} + m\log{n})$ in expectation.
+\end{theorem}
 

### 2015-02-27 paper/figures/quickselect-split-aligned.tex
minor

\begin{figure}
\begin{lstlisting}
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...`

datatype tree = Leaf
              | Branch of int * int * tree * tree

1`fun qs1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons ht => 
      let (left,right,n) = partition ht in
      Branch (n, #1 ht, qs1 left, qSelect1 right)`

2`fun qs2 (p : tree, k : int) : int = 
  case p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qs2 (p1,k)
      | EQ => h
      | GT => qs2 (p2,k-n-1)`
\end{lstlisting}
\caption{Two-pass implementation of quickselect.}
\label{fig:qs-split}
\end{figure}

### 2015-02-27 paper/overview.tex
pass over the overview

@@
 % usual, and in the \texttt{Cons} branch, \texttt{part}ition it.
-
-the
-function \texttt{qs} does \emph{not} depend on the rank \texttt{k};
-for example, the list is partitioned before \texttt{k} is ever
-used. Moreover, while the recursive calls to \texttt{qSelect} are
-guarded by a comparison to \texttt{k}, \texttt{k} does not determine
-the arguments to those recursive calls---it only determines which call
-is made. Thus, if we are willing to change the evaluation order of the
-language (and evaluate under the \texttt{case} statement), it should
-be possible to force \emph{all} the computations involving \texttt{l}
-to be performed before \emph{any} of the computations involving
-\texttt{k}.
-
-The \texttt{case} expression in \texttt{qSelect} depends on \texttt{k}, whose
-type $\fut\rmint$ indicates that it is only available to second-stage
-computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$, the
-remainder of the function is second-stage code.
-
-The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
-expression) as a first-stage expression of type $\fut\rmint$ (the result 
-of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
-$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
-to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
-$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
-second-stage $\rmint$.
 

### 2015-02-27 paper/semantics.tex
pass over the overview

@@
 \subsection{Statics}
+
+\fbox{\parbox{5in}{
+\paragraph{Blend.}
+the
+function \texttt{qs} does \emph{not} depend on the rank \texttt{k};
+for example, the list is partitioned before \texttt{k} is ever
+used. Moreover, while the recursive calls to \texttt{qSelect} are
+guarded by a comparison to \texttt{k}, \texttt{k} does not determine
+the arguments to those recursive calls---it only determines which call
+is made. Thus, if we are willing to change the evaluation order of the
+language (and evaluate under the \texttt{case} statement), it should
+be possible to force \emph{all} the computations involving \texttt{l}
+to be performed before \emph{any} of the computations involving
+\texttt{k}.
+
+The \texttt{case} expression in \texttt{qSelect} depends on \texttt{k}, whose
+type $\fut\rmint$ indicates that it is only available to second-stage
+computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$, the
+remainder of the function is second-stage code.
+
+The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
+expression) as a first-stage expression of type $\fut\rmint$ (the result 
+of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
+$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
+to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
+$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
+second-stage $\rmint$.
+
+}}
 

### 2015-02-27 paper/semantics.tex

@@
 \subsection{Statics}
-
-\fbox{\parbox{5in}{
-\paragraph{Blend.}
-the
-function \texttt{qs} does \emph{not} depend on the rank \texttt{k};
-for example, the list is partitioned before \texttt{k} is ever
-used. Moreover, while the recursive calls to \texttt{qSelect} are
-guarded by a comparison to \texttt{k}, \texttt{k} does not determine
-the arguments to those recursive calls---it only determines which call
-is made. Thus, if we are willing to change the evaluation order of the
-language (and evaluate under the \texttt{case} statement), it should
-be possible to force \emph{all} the computations involving \texttt{l}
-to be performed before \emph{any} of the computations involving
-\texttt{k}.
-
-The \texttt{case} expression in \texttt{qSelect} depends on \texttt{k}, whose
-type $\fut\rmint$ indicates that it is only available to second-stage
-computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$, the
-remainder of the function is second-stage code.
-
-The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
-expression) as a first-stage expression of type $\fut\rmint$ (the result 
-of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
-$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
-to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
-$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
-second-stage $\rmint$.
-
-}}
 
@@
 {\em stability} in \cite{krishnaswami13}.}
+
+\crem{blend in the rest of the text in this section:}
+%
+\texttt{qsStaged} has a more precise type---it is a first-stage function which
+takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$ (an
+$\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
+
+Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
+\texttt{part}ition it. The \texttt{case} expression in \texttt{qSelect} depends
+on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
+second-stage computations. Since \texttt{qsStaged} itself produces a
+$\fut\rmint$, the remainder of the function is second-stage code.
+
+The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
+expression) as a first-stage expression of type $\fut\rmint$ (the result 
+of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
+$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
+to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
+$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
+second-stage $\rmint$.
+
+\crem{does this remark go anywhere?}
+
+Thus, if we are willing to change the evaluation order of the
+language (and evaluate under the \texttt{case} statement), it should
+be possible to force \emph{all} the computations involving \texttt{l}
+to be performed before \emph{any} of the computations involving
+\texttt{k}.
 

### 2015-02-27 paper/figures/quickselect-split-aligned.tex
pass on the overview

@@
       let (left,right,n) = partition ht in
-      Branch (n, #1 ht, qs1 left, qSelect1 right)`
+      Branch (i, #1 ht, qs1 left, qSelect1 right)`
 
@@
     Leaf => 0
-  | Branch (n,h,p1,p2) => 
-      case compare k n of
+  | Branch (i,h,p1,p2) => 
+      case compare k i of
         LT => qs2 (p1,k)
       | EQ => h
-      | GT => qs2 (p2,k-n-1)`
+      | GT => qs2 (p2,k-i-1)`
 \end{lstlisting}

### 2015-02-27 paper/overview.tex
pass on the overview

@@
 %
-\footnote{We assume that the list is prepermuted to guarantee the
-  expected linear time behavior.} 
-%
 Using function \texttt{qs}, we can perform many order statistics
-queries, for example but with $m$ different ranks
+queries, for example with $m$ different ranks
 $\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
@@
 expensive to perform order statistics on more than a small number of
-times. This is unfortunate, because for example, because we wish to be
-able to map \texttt{qs} to over another list and do so
-efficiently.
+times. This is unfortunate: we wish to be able to map \texttt{qs} to
+over another list and do so efficiently.
 
-One way to regain efficiency is to change the algorithm that we use to
-perform the computation in multiple passes.  One option would be to
-pre-sort the input list \texttt{l} into a sorted list \texttt{s} in
-the first pass, and then perform lookups in \texttt{s} in the second
-pass.  Unfortunately, since \texttt{s} is a list, a lookup would
-require linear time (also on average), leading to no improvement in
-efficiency.  We can resolve this issue by using a data structure that
-can provide faster access.  For example, after sorting \texttt{s} we
-can copy in into an array and then perform binary searches to find the
-element with the desired rank (copying the input to an array and
-sorting would not work well in a functional language).  As another
-option, we can represent construct a binary search tree from the input
-and perform size-based searches on the binary search tree to find the
-elements with the desired rank.  Such an approach would fit nicely
-into the functional paradigm of computing but requires additional
-algorithmic ingenuity: to ensure efficiency we would need to store at
-each note the size of its subtree (these are sometimes called
-``augmented trees'').  In summary, we are able to improve efficiency
-by replacing our simple solution with a multi-pass algorithm that
-first preprocesses the input to generate a lookup data structure and
-then in the second pass performs fast lookups.  As outlined above, in
-addition, this transformation is highly non-trivial as it involves
-reasoning about intricate algorithmic concerns, and implementing more
-complex algorithms.
+One way to regain efficiency is to use a multi-pass algorithm that
+performs frequently performed computations first so that they can be
+reused.  It is nontrivial what such computations are in this example
+but after some reflection, we can realize that they are the
+comparisons performed by the partition function.  To minimize them, we
+can pre-sort the input list \texttt{l} into a sorted list \texttt{s}
+in the first pass, and then perform lookups in \texttt{s} in the
+second pass.  Unfortunately, since \texttt{s} is a list, a lookup
+would require linear time (also on average), leading to no improvement
+in efficiency.  But this is not too hard: after sorting \texttt{s} we
+can copy in into an array and use binary search to find the element
+with the desired rank.  As another option, we can construct a binary
+search tree from the input and perform size-based searches on the
+binary search tree to find the elements with the desired rank.  Such
+an approach would fit nicely into the functional paradigm but requires
+one more additional algorithmic ingenuity: to ensure efficiency we
+would need to store at each note the size of its subtree (these are
+sometimes called ``augmented trees'').  In summary, we are able to
+improve efficiency by replacing our simple solution with a multi-pass
+algorithm that first preprocesses the input to generate a lookup data
+structure and then in the second pass performs fast lookups.  As
+outlined above, in addition, this transformation is highly non-trivial
+as it involves reasoning about intricate algorithmic concerns, and
+implementing more complex algorithms and data structures.  Both of
+these approaches however improve the run-time to (expected)
+$\Theta(n\log{n} + m\log{n})$, a near linear-time improvement in $m$.
 
@@
 Scherlis are commonly employed.  For example, as briefly mentioned in
-the introduction, modern graphics software is written exactly in this
-way.  Unfortunately, as the example illustrates, they lead to complex
-software, leading to the question of whether it would be possible to
+\secref{intro}, modern graphics software is written exactly in this
+way.  Unfortunately, as the example illustrates, multipass programs
+are naturally complex.  We thus ask: is it would be possible to
 perform such transformations automatically.  We answer this question
@@
 
-We can write a staged version of \texttt{qs}, called
-\texttt{qss} in \ref{fig:qs-staged}, by making the input list a
+We can write a staged version of \texttt{qs}, called \texttt{qss},
+whose code is shown in \ref{fig:qs-staged}, where the first-stage code
+are typeset in red, and the second-stage code are typeset in blue.  To
+obtain \texttt{qss} from \texttt{qs}, we first make the input list a
 stage-1 and the rank a stage-2 argument.
 %
-More precisely the input list
-has type $\curr\mathrm{list}$ (an integer list ``now''), the rank has
-type $\fut\rmint$ (an integer in the ``future''), and return has type
-$\fut\rmint$.  
+More precisely the input list has type $\curr\mathrm{list}$ (an
+integer list ``now''), the rank has type $\fut\rmint$ (an integer in
+the ``future''), and the return has type $\fut\rmint$.
 %
@@
 with the staging annotations \texttt{prev} and \texttt{next}, which
-transition between a stage-1 and stage-2 code, and $\texttt{g}$, which
+transition between a stage-1 and stage-2 code, and~$\texttt{g}$, which
 marks ground terms.  
@@
 \lstinline{1`hold : ^int -> $`2`int`}.
-
-To help readability, we write the first-stage
-computations in red, and the second-stage computations in blue.
+%
 
@@
 Interestingly, our splitting algorithm achieves exactly this.
-
 In the rest of this section, we present a brief, high-level overview
@@
 %
+
 When applied to the staged code in \ref{fig:qs-staged}, our splitting
@@
 quickselect algorithm, binary search trees, or how to perform lookups
-on binary search trees, but it is able to derive all of these from the
-input program.
+on binary search trees, but it is able to derive them from the input
+program.
 %
@@
 boundary data structure that both records the control flow and the
-results from the first stage computation at each control-flow point.
-The function \texttt{qs1} performs all the recursive calls and
-evaluate all instances of the partition function, which depend only on
-the input list.  The function produces a boundary data structure that
-collects the results from all recursive calls along with a tag that
-indicates the control branches taken.  Since \texttt{qss} has a binary
-control structure (casing on the list), the boundary is a binary tree.
-To ensure that the trace contains the necessary results to complete
-the execution in the second pass, where second-stage values may be
-used, the splitting algorithm includes in the nodes of the tree
-information such as the "pivot" used for splitting and the size of the
-subtree at that note.  The resulting tree is thus not just a tree but
-a binary search tree, keyed by the pivot and augmented with "size"
-information.
+results from the first stage at relevant control-flow points.  More
+specifically, the function \texttt{qs1} performs all the recursive
+calls and evaluates all instances of the partition function, which
+depend only on the input list.  The function produces a boundary data
+structure that collects the results from all recursive calls along
+with a tag that indicates the control branches taken.  Since
+\texttt{qss} has a binary control structure (casing on the list), the
+boundary is a binary tree.  To ensure that the trace contains the
+necessary results to complete the execution in the second pass, where
+second-stage values may be used, the splitting algorithm includes in
+the nodes of the tree information such as the "pivot" used for
+splitting and the size of the subtree at that note.  The resulting
+tree is thus not just a tree but a binary search tree, keyed by the
+pivot and augmented with "size" information.
 
@@
 rank) and performs a lookup in the boundary data structure.  But how
-does \texttt{qsStage2} knows to traverse the tree? It does not. Since
+does \texttt{qs2} knows to traverse the tree? It does not. Since
 the splitting algorithm recorded the control flow of the first stage
@@
 
-\subsection{Quickselect in more detail}
-
-Carlo writes this.
-
 \end{abstrsyn}

### 2015-02-27 paper/overview.tex
proof entered

@@
 \end{theorem}
+\begin{proof}
+  We use a few standard facts.  We note first that \texttt{qs1} is
+  isomorphic to a quick-sort algorithm, which builds a binary search
+  tree instead of combining the returned results in a list.  The
+  run-time for \texttt{qs1} is thus (expected) $O(n\log{n})$. 
+%
+  The depth of the tree generated by \texttt{qs1} is isomorphic to the
+    depth of a run of the quickselect algorithm or equivalently to the
+    span of a parallel implementation of the quicksort algorithm, both
+    of which are known to be (expected) $\Theta(\log{n})$.
+%
+    By inspection of \texttt{qs2}, we can see that the algorithm takes
+    time proportional to the depth of the binary search tree, which is
+    $\Theta(\log{n})$. Consider now a random variable for each of the
+    $m$ invocations of \texttt{qs2}, each of which takes expected
+    $\Theta(\log{n})$ time.  Since the total time is the sum of these
+    random variables and since the expectation of the sum is sum of
+    the expectations of the random variables, the bound follows.
+\end{proof}
 

### 2015-02-27 paper/semantics.tex
blending 3.1

@@
 
-\crem{does this remark go anywhere?}
-
-Thus, if we are willing to change the evaluation order of the
-language (and evaluate under the \texttt{case} statement), it should
-be possible to force \emph{all} the computations involving \texttt{l}
-to be performed before \emph{any} of the computations involving
-\texttt{k}.
-
 \end{abstrsyn}
@@
 evaluate them yet. However, we do need to evaluate any first-stage computations
-they contain (inside $\prev$), even if those computations are underneath
-binders.
+they contain (inside $\prev$), \emph{even if} those computations are guarded or
+underneath binders.
 

### 2015-02-27 paper/splitting.tex
fixed yhat

@@
 \paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
-where $\diaone {e_k} {\xi_k} {v_k}$ and $\rtab {\xi_k} {v_k} \vsplito
-\mval {i_k} {q_k}$ for $k = 1,2$.  Inductively:
+where $\diaone {e_k} {\xi_k} {v_k}$ and 
+$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
+Inductively:
 \[
@@
 Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
-and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
+and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
 splitting generates:
@@
 Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
+
+We consider the two possible control paths seperately.  In the first case, 
+$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
+Inductively,
+\[
+	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
+\]
+Additionally, the branch first evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$ and $\rtab {\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
+Inductively,
+\[
+	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]e_2} {\tup{i_2,b_2}} \text{ and } [b_2/l_2](\letin x {q_1} {r_2}) \equiv q_2
+\]
+
+So we need to satisfy.....
+
 Consider that when $\reduce {e_1} {\inl {v_1}}$, both the first and second stage component of $e_2$ will be executed, 
@@
 their splitting rules are relatively simple.
-
-Given an $\splittwo e ? p l r$ and $\diatwo e q$ for which the splitting invariant holds, we know
+Given an $\splittwo e ? p l r$ and $\diatwo e q$, the splitting invariant says,
 \[
@@
 \]
-The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$ 
-and $\diaone {\next {e'}} {\yhat \mapsto q} {\next \yhat}$
-and $\rtab {\yhat \mapsto q} {\next \yhat} \vsplito \mval {\tup{}} {\letin \yhat q \yhat}$
-is
+The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$,
+where $\diaone {\next {e'}} {{\hat y} \mapsto q} {\next {\hat y}}$
+and $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$
+then requires
 \[
-	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin \yhat q \yhat
+	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
 \]
-which can be fulfilled using $c' = \tup{\tup{},p}$ and $l.r = l'.r'$.
+which can be fulfilled using $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
 Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.

### 2015-02-27 paper/overview.tex
related work

@@
 balanced binary search tree, which is isomorphic to a treap data
-structure~\cite{treaps}.  In the second pass, the program performs,
+structure~\cite{SeidelAr96}.  In the second pass, the program performs,
 for each rank, a binary search tree lookup, by walking the tree to

### 2015-02-27 paper/relatedwork.tex
fixing various broken things
related work

@@
 \section{Related Work}
+\label{sec:related}
 
@@
 times and pull them ahead so that they can be performed once and used
-later as needed.  The idea behind precomputation is to identify
-computations that can be performed earlier, for example at compile
-time if their inputs are available statically and perform them at that
-earlier time.  Many interesting algorithm design techniques and
-compiler optimizations perform frequency reduction and precomputation.
-For example, dynamic programming and loop hoisting are a frequency
-reduction techniques, whereas dynamic data structures and incremental
-computation are precomputation techniques.
+later as needed.  For example, dynamic programming and loop hoisting,
+as well as the work presented here are a frequency reduction
+techniques.
+%
+The idea behind precomputation is to identify computations that can be
+performed earlier, for example at compile time if their inputs are
+available statically and perform them at that earlier time.  For
+example, dynamic algorithms, partial evaluation, and incremental
+computation are all precomputation techniques.
 
+\paragraph{Precomputation techniques.}
 Perhaps one of the most well studied example of precomputation is
@@
 %
-\ur{Do prior work evaluate first-stage expressions inside second stage
-expressions?}
-%
 Unlike in partial evaluation, our splitting algorithm requires no
@@
 
+An important class of pre-computation techniques include incremental
+computation, where a program can respond to small changes to its input
+efficiently by recomputing only the parts of the computation affected
+by the
+changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
+As other pre-computation techniques, incremental computation can be
+thought as a staged model, where changes (inputs) may come over a
+period of time in stages. 
+%
+It does not, however, fix any of the inputs or anticipate the future
+inputs, allowing in the most general case for all data to change. As a
+result, its benefits are dependent on the data instance (nature of the
+inputs). In addition, incremental computation techniques typically
+utilize significant run-time (dynamic) information, because no
+assumptions about input instances are made.  For example, we can apply
+incremental computation to our quickselect example
+(ref{sec:overview}), but we would not be able to achieve the bounds
+that we did because incremental computation techniques would unfold
+the quickselect function based on the demanded ranks, which can
+require linear time (as opposed to logarithmic) at each stage.  In
+addition, such an implementation would maintain rather sophisticated
+data structures dynamically at run time, the benefits of which depend
+highly on the forms of changes applied.
+
+\paragraph{Frequency-reduction techniques.}
 Perhaps the closest work to ours is algorithm for ``data
@@
 limited to simple, imperative languages.
-
-\paragraph{Memoization.} Leaves code structure unchanged.
-
-\paragraph{Incremental computation.} Usually not applicable within a
-computation, require ``mutator'' ``core'' behavior.
 

### 2015-02-27 paper/overview.tex
minor

@@
   We use a few standard facts.  We note first that \texttt{qs1} is
-  isomorphic to a quick-sort algorithm, which builds a binary search
-  tree instead of combining the returned results in a list.  The
-  run-time for \texttt{qs1} is thus (expected) $O(n\log{n})$. 
+  isomorphic to a quick-sort algorithm but builds a binary search tree
+  instead of combining the returned results in a list.  The run-time
+  for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
 %

### 2015-02-27 paper/splitting.tex
finished a rough draft of the case case

@@
 Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
+Here we consider only the case where
+$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
+(The \texttt{inr} case is analogous.)
 
-We consider the two possible control paths seperately.  In the first case, 
-$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
 Inductively,
@@
 \]
-Additionally, the branch first evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$ and $\rtab {\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
+Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
+and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$,
+and so the entire case evaluates like:
+$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$.
+
 Inductively,
 \[
-	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]e_2} {\tup{i_2,b_2}} \text{ and } [b_2/l_2](\letin x {q_1} {r_2}) \equiv q_2
+	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} 
 \]
+\[
+\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
+\]
+(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
+When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
+of $v_1$ into $c_2$ and $l_2.r_2$.)
 
-So we need to satisfy.....
+The invariant, applied to the entire \texttt{case}, requires some $c'$ and
+$l'.r'$ such that 
+\[
+  \reduce {c'} {\tup{i_2,b'}} \text{ and }
+  [b'/l']r' \equiv q_2
+\]
+We choose
+\[
+c' =
+{\left(
+\talllet{\tup{y_1,z_1}}{c_1}{
+\tallcase{y_1}
+{x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
+{x_3.\cdots\ttrpar}
+}\right)}
+\]
+\[
+l'.r' =
+\tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
+\]
+If you work it out you will see that $c'$ reduces to
+$\tup{i_2,\tup{b_1,\inl{b_2}}}$, meaning that 
+$b' = \tup{b_1,\inl{b_2}}$, and
+$[b'/l']r'$ essentially reduces to
+$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, which is exactly the 
+$[q_1/x_2][b_2/l_2]r_2$ we expected.
 
-Consider that when $\reduce {e_1} {\inl {v_1}}$, both the first and second stage component of $e_2$ will be executed, 
-but $e_3$ will be ignored entirely. The reverse occurs when $\reduce {e_1} {\inr {v_1}}$.
-This implies that both the first and second stage results of splitting will need to have \texttt{case} statements.
+In other words, $c'$ cases on $i_1$ to run the first branch of first stage
+code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
+{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
+the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
+similarly constrained by the $\texttt{inr}$ case.
 
-Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
-$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.
-Thus the algorithm cases on $c_1$ to run the correct branch of first stage code.
-The boundary value from this correct branch, $b_2$ or $b_3$, is packaged as either $\tup {b_1, \inl {b_2}}$ or $\tup {b_1, \inr {b_3}}$.
-
-The resumer for \texttt{case} operates by untupling the boundary value to get $b_1$ and either $\inl {b_2}$ or $\inr {b_3}$.
-The resumer of the predicate is run using $b_1$, and the remaining injection is inspected to run the correct branch resumer.
+%Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
+%$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.
 

### 2015-02-27 paper/abstract.tex

@@
-Staged program languages assign a stage to each expression and
+Staged programming languages assign a stage to each expression and
 evaluate each expression in its assigned stage.  For example, in a
@@
 evaluations of un-staged, conventional programs each of which
-corresponds to a pass.  Conceived in the 80`s but J{\o}rring and
+corresponds to a pass.  Conceived in the 80's but J{\o}rring and
 Scherlis, this {\em pass-separation} technique has been studied but

### 2015-02-27 paper/intro.tex

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 
-A closely related technique, {\em metaprogramming}, allows writing programs that
-generate specialized code. Metaprogramming enables fine-grained control over
-specialization by requiring staging annotations that make explicit the stage of
-each computation. (In partial evaluation, we typically assume that a
-\emph{binding-time analysis} has automatically determined which parts of $f$ can
-be specialized.)
+A closely related technique, {\em metaprogramming}, allows writing
+programs that generate specialized code
+(e.g.,~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
+enables fine-grained control over specialization by requiring staging
+annotations that make explicit the stage of each computation. (In
+partial evaluation, we typically assume that a \emph{binding-time
+  analysis} has automatically determined which parts of $f$ can be
+specialized.)
 

### 2015-02-27 paper/relatedwork.tex
Quick pass.

@@
 
-As Jorring and Scherlis argue frequency reduction and precomputation
-are common techniques for both designing algorithms and performing
-compiler optimizations~\cite{JS86-staging}.  The idea behind frequency
+Frequency reduction and precomputation are common techniques for both
+designing algorithms and performing compiler
+optimizations~\cite{JS86-staging}.  The idea behind frequency
 reduction is to identify computations that are performed multiple
 times and pull them ahead so that they can be performed once and used
-later as needed.  For example, dynamic programming and loop hoisting,
-as well as the work presented here are a frequency reduction
+later as needed.  Dynamic programming, loop hoisting, and pass
+separation (presented here) are all examples of frequency reduction
 techniques.
@@
 The idea behind precomputation is to identify computations that can be
-performed earlier, for example at compile time if their inputs are
-available statically and perform them at that earlier time.  For
-example, dynamic algorithms, partial evaluation, and incremental
-computation are all precomputation techniques.
+performed earlier (e.g., at compile time) if their inputs are
+available statically and perform them at that earlier time. Dynamic
+algorithms, partial evaluation, and incremental computation are all
+examples of precomputation techniques.
 
@@
 (compile-time) and dynamic (run-time) stage. Given a program and the
-values of its static inputs, which are the subset of the inputs that
-become available at compilation time, partial evaluation generates a
-specialized program by performing the computations that depend on the
-statics inputs~\cite{jones96}.  The roots of partial evaluation go
-back to Kleene, who in 1952 proved that a multivariate program
-(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
-$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
-that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
-y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
-however.  Lombardi appears to be the first person to use the term in
-his work on incremental computation~\cite{Lombardi67}.  Since then,
-there has a lot of work on partial evaluation, only broad-strokes of
-which we can mention in this here.  We refer the interested reader to
-the excellent book by Jones, Gommard, and Sestoft for a comprehesive
-discussion of work until early 90's~\cite{JGS93}.
+values of static inputs (inputs that are available at compilation
+time), partial evaluation generates a specialized program by
+performing the computations that depend on the statics
+inputs~\cite{jones96}.  The roots of partial evaluation go back to
+Kleene, who in 1952 proved that a multivariate program (function)
+$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
+a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
+\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
+Kleene did not intend to improve efficiency, however.  Lombardi
+appears to be the first person to use the term in his work on
+incremental computation~\cite{Lombardi67}.  Since then, there has a
+lot of work on partial evaluation, only broad-strokes of which we can
+mention in this here.  We refer the interested reader to the excellent
+book by Jones, Gommard, and Sestoft for a comprehesive discussion of
+work until early 90's~\cite{JGS93}.
 
-At a high level, earlier approaches to partial evaluation can be
-viewed as operating in two stages: binding time analysis and program
-specialization.  For a multivariate program with clearly marked static
-and dynamic arguments, binding-time analysis identifies all the parts
-of the program that can be computed by the knowledge of static
-arguments. Using this information and the static arguments themselves,
-program specialization specializes the program to a partially
-evaluated program that can be used with many different dynamic
-arguments.  This approach has been applied to construct partial
-evaluators for a number of languages such as
-Scheme~\cite{OB91-Similix,Consel88-Schism}, considering for the most
-part first-order subsets.
+Earlier approaches to partial evaluation can be viewed as operating in
+two stages: binding time analysis and program specialization.  For a
+multivariate program with clearly marked static and dynamic arguments,
+binding-time analysis identifies all the parts of the program that can
+be computed by the knowledge of static arguments. Using this
+information and the static arguments themselves, program
+specialization specializes the program to a partially evaluated
+program that can be used with many different dynamic arguments.  This
+approach has been applied to construct partial evaluators for a number
+of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism},
+considering for the most part first-order subsets.
 
@@
 
-The type-system component of our work is closely related to this later
-line of work on metaprogramming and staged computation.  The specific
+The type-system of \lang\ is closely related to this later line of
+work on metaprogramming and staged computation.  The specific
 extension to the typed lambda calculus that we use here is based on
@@
 propose.  The key difference between our work and this prior work is
-the staging or the splitting algorithm, which given a two-staged
-program splits it into two stages that can be executed sequentially,
-passing information from the earlier to the latter stage by
-construction. 
+our focus on the problem of pass separation: we developed an algorithm
+for splitting terms in the proposed staged language.
+%the staging or the splitting algorithm, which given a two-staged
+%program splits it into two stages that can be executed sequentially,
+%passing information from the earlier to the latter stage by
+%construction.
 %
-Unlike in partial evaluation, our splitting algorithm requires no
-knowledge of the arguments themselves. The only information that is
-required is their staging. Using just the staging information, the
-splitting algorithm can generate programs that can perform non-trivial
-transformations on the code automatically.  As our examples illustrate
-such staging transformations can improve the efficiency of the source
-program, sometimes in asymptotically significant ways.  
-We are not aware of prior work in the context of higher-order
-functional languages that presents such a splitting algorithm.
+Unlike partial evaluation, splitting requires no knowledge of program
+argument values, only their staging.
+%Using just the staging
+%information, the splitting algorithm can generate programs that can
+%perform non-trivial transformations on the code automatically.  As our
+%examples illustrate such staging transformations can improve the
+%efficiency of the source program, sometimes in asymptotically
+%significant ways.
+We are not aware of prior work in the context of
+higher-order functional languages that presents such a splitting
+algorithm.
 
@@
 changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
-As other pre-computation techniques, incremental computation can be
-thought as a staged model, where changes (inputs) may come over a
+Like other pre-computation techniques, incremental computation can be
+thought of as a staged model, where changes (in inputs) come over a
 period of time in stages.
 %
-It does not, however, fix any of the inputs or anticipate the future
+However, incremental computation does not fix any of its inputs or anticipate the future
 inputs, allowing in the most general case for all data to change. As a
@@
 
-\paragraph{Frequency-reduction techniques.}
-Perhaps the closest work to ours is algorithm for ``data
-specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  As our
-work, Knoblock and Ruf are concerned about the problem of splitting an
-explicitly staged program into two stages.  However, they only
-consider, simple first-order language and straight-line
-programs. Automatic pass seperation has also been a topic of interest
-in programming systems for computer graphics, where, to achieve
+\paragraph{Frequency-reduction via pass separation.}
+Perhaps the closest work to ours is the algorithm for ``data
+specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
+our work, Knoblock and Ruf are concerned about the problem of
+statically splitting an explicitly staged program into two stages.
+However, they only consider simple first-order language and
+straight-line programs and also treat all computations guarded by
+second stage conditionals as second stage computations.
+
+Automatic pass seperation has also been a topic of interest in
+programming systems for computer graphics, where, to achieve
 high-performance, programs are typically manually separated into

### 2015-02-27 paper/examples.tex
algoexamples renamed examples

%!TEX root = paper.tex
 \section{Examples of Splitting}
\label{sec:examples}

Now we investigate the behavior of our splitting algorithm on several examples.
The split code that follows is the output of our splitting implementation; for
clarity, we have performed some minor optimizations and manually added type
annotations (including datatype declarations and constructor names), as our
algorithm does not type its output.

\subsection{Dot Product}

Our first example, \texttt{dot}, appears in \cite{knoblock96}. \texttt{dot} is a
first-order, non-recursive function---precisely the sort of code studied in
prior work on pass separation for imperative languages. \texttt{dot} takes the
dot product of two three-dimensional integer vectors, where the first two
coordinates are first-stage, and the last coordinate is second-stage. 
%
\begin{lstlisting} 
1`type vec = ^int * ^int * $`2`int`

1`//dot : vec * vec -> $int
fun dot (
  (3`grnd{`1`xliteralone`3`}`1`,`3`grnd{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
  (3`grnd{`1`xliteraltwo`3`}`1`,`3`grnd{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
next{
  2`prev{`1`hold `3`grnd{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
1`}`
\end{lstlisting}
%
The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

Our algorithm splits \texttt{dot} into the two functions:
%
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()),(x2,y2,())) 
  = ((), (x1*x2)+(y1*y2))`

2`fun dot2 ((((),(),z1),((),(),z2)),l) 
  = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, except
that they write the precomputation into a mutable cache.

\subsection{Exponentiation by Squaring}

Our next example, \texttt{exp}, is a mainstay of the partial evaluation
literature (for example, in \cite{jones96}). \texttt{exp} recursively computes
$b^e$ using exponentiation by squaring, where $e$ is known at the first stage,
and $b$ is known at the second stage. 
%
\begin{lstlisting} 
1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`grnd{`1`e`3`}`1` : ^int) = 
  if `3`grnd{`1`e == 0`3`}`1` then 
    next{`2`literalone`1`}
  else if `3`grnd {`1`(e mod 2) == 0`3`}`1` then
    exp(next{`2`b*b`1`},`3`grnd{`1`e/literaltwo`3`}`1`)
  else 
    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`grnd{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
\end{lstlisting}
%
Because \texttt{exp} is a recursive function whose conditionals test the parity
of the exponent argument, the sequence of branches taken corresponds exactly to
the binary representation of $e$. Partially evaluating \texttt{exp} with $e$
eliminates all of the conditionals, selecting and expanding the appropriate
branch in each case.

Our algorithm, on the other hand, produces:
%
\begin{lstlisting} 
datatype binnat = Zero | Even of nat | Odd of nat

1`fun exp1 (b : unit, e : int) =
  if e==0 then
    ((), Zero)
  else if (e mod 2)==0 then 
    ((), Even (#2 (exp1 ((), e/2))))
  else 
    ((), Odd (#2 (exp1 ((), (e-1)/2))))`

2`fun exp2 ((b : int, e : unit), l : binnat) =
    case l of
      Zero => 1
    | Even n => exp2 ((b*b, ()), n)
    | Odd n => b * exp2 ((b*b, ()), n)`
\end{lstlisting}
%
\texttt{exp1} recursively performs parity tests on $e$, but unlike \texttt{exp},
it simply computes a data structure (a \texttt{binnat}) recording which branches
were taken. \texttt{exp2} takes $b$ and a \texttt{binnat} $l$, and uses $l$ to
determine how to compute with $b$.

Of course, the \texttt{binnat} computed by \texttt{exp1} is precisely the binary
representation of $e$! While partial evaluation realizes \texttt{exp}'s
control-flow dependency on a fixed $e$ by recursively expanding its branches in
place, we explicitly record this dependency generically over all $e$ by creating
a boundary data structure. This occurs in the $\splitonesym$ rule for
\texttt{case}, which emits a tag corresponding to the taken branch in the
precomputation, and \texttt{case}s on it (as $l_b$) in the residual.

Because splitting \texttt{exp} does not eliminate its conditionals, partial
evaluation is more useful in this case. (Notice, however, that partially
evaluating \texttt{exp2} on a \texttt{binnat} is essentially the same as as
partially evaluating \texttt{exp} on the corresponding \rmint.) Nevertheless,
splitting \texttt{exp} still demonstrates how our algorithm finds interesting
data structures latent in the structure of recursive functions.

\subsection{Quickselect}
\label {sec:exampleQS}

Let us return to the quickselect algorithm, which we discussed at
length in \ref{sec:overview}. (The code is in \ref{fig:qs-staged}.)
\texttt{qss} finds the $k$th largest element of a list $l$ by
recursively \texttt{partition}ing the list by its first element, then recurring
on the side containing the $k$th largest element. $l$ is first-stage and $k$ is
second-stage.

Stage-splitting \texttt{qss} produces:
%
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
datatype tree = Leaf
              | Branch of int * tree * int * tree

1`fun partition ((p,l):int*list) : (int*list*list) = 
  case l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) => 
      let val (s,left,right) = partition (p,t) in 
      if h<p 
      then (s+1,Cons(h,left),right) 
      else (s,left,Cons(h,right))

fun qs1 (l : list, k : unit) = ((), 
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      Branch (
        let (left,right,n) = partition (h,t) in
        (n, #2 (qs1 left k), h, #2 (qs1 right k))
      )
  )`

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare (k, n) of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except that the mixed-stage argument to both passes is
is explicitly masked, and we do not suppress the trivial immediate result of
\texttt{qs1}.

\texttt{qs1} \texttt{partition}s $l$, but since the comparison with $k$ (to
determine which half of $l$ to recur on) is at the second stage, it simply
recurs on \emph{both} halves, pairing up the results along with $h$ (the head of
$l$) and $n$ (the size of the left half). \texttt{qs2} takes $k$ and this tree
$p$, and uses $k$ to determine how to traverse $p$.

How does our splitting algorithm generate binary search trees and a traversal
algorithm? The $\splittwosym$ rule for \texttt{case} tuples up the
precomputations for its branches, and in the residual, selects the residual
corresponding to the appropriate branch. The tree is implicit in the structure
of the code; ordinarily, the quickselect algorithm only explores a single
branch, but the staging annotations force the entire tree to be built.

This is an instance where stage-splitting is more practical than partial
evaluation---if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup, by not expanding some calls to
\texttt{partition}, would result in duplicating first-stage computations.)

Thus, \texttt{qs1} performs $\Theta(n \log n)$ work per call, whereas
\texttt{qs2} performs expected $\Theta(\log n)$ work.  This results in a net
speedup over standard quickselect if we perform $\Omega(n / \log n)$ queries on
the same list.

Note that the recursive \texttt{partition} function is defined within a \texttt{grnd}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{partition}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{grnd} annotations, however, this can be prevented.

To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
the second pass, evaluate \texttt{qs2} on that tree and each rank.
\begin{lstlisting}
1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
\end{lstlisting}

\subsection {Composing Graphics Pipeline Programs}

A major motivation for staged programming in computer graphics is to express
logic related to a single application concern conveniently as a single staged
function (referred to as a pipeline program)\,\cite{Foley:2011}, rather than
as a collection of functions distributed across different GPU pipeline stages.

Pipeline programs can be directly expressed as staged functions in \lang\, with splitting 
used to generate monostaged code required for each GPU pipeline stage
(as well as the plumbing to communicate values between the stages).
Composition of such functions in \lang\ allows more complex pipelines to be expressed in a modular programming style
that is becoming increasingly desirable in graphics shading languages\,\cite{Foley:2011,He:2014}. A simple example of composition of staged functions (representing GPU pipeline programs) is given below.  The code defines a pipeline as a staged function that takes an object definition in first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel. The code defines a pipeline combinator \texttt{shade} which modulates the results of the object's reflectance at a given pixel with the results of a function that computes the object's albedo (surface color).  
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline,
           albedo : pipeline) : pipeline =
  (fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl   (obj,next{2`xy`})`2`}
     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
\end{lstlisting}
%
Most pipeline programming models either force the user to manually split their programs (which harms modularity).
Others, such as \cite{Foley:2011}, offer only a few multistage types (namely, first order functions) and give 
combinators like \texttt{par} as language features.
In contrast, \lang\ is powerful enough to define those combinators in-language.

### 2015-02-27 paper/overview.tex
overview section is now sec:overview
removing obsolete text from comments

@@
 \section{Overview}
-\label{sec:examplea}
+\label{sec:overview}
 
@@
 
-\begin{comment}
-
-An astute programmer, having noticed that \texttt{quickselect} can be staged in
-this fashion, might try to split it into a pair of functions, one which performs
-all the work depending only on \texttt{l} (the first stage), and one which uses
-that partial result and \texttt{k} to compute the element with
-rank \texttt{k} in \texttt{l}.
-
-Intuitively, \texttt{l} determines the result of all calls to
-\texttt{part}ition, and \texttt{k} only determines which calls are made. So we
-can preprocess \texttt{l} by recursively dividing it into halves smaller and
-greater than the pivot---that is, building a binary search tree. Then, once we
-have \texttt{k}, we can recur on this tree, choosing whichever branch has the
-\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. And because
-\texttt{part}ition contains no second-stage code, we can run it entirely in the
-first stage.
-
-We have implemented this splitting of quickselect in \ref{fig:qs-split}.
-\texttt{qSelect1} builds a binary search tree from the list \texttt{l}, and
-\texttt{qSelect2} takes such a tree and a rank \texttt{k} and computes the
-answer. This allows us to efficiently perform many order statistics queries on
-\texttt{l} by caching the tree and reusing it for many different ranks
-$\mathtt{k_1},\dots,\mathtt{k_m}$:
-%
-\begin{lstlisting}
-let b = qSelect1 l in
-  qSelect2 b @$k_1$@
-  qSelect2 b @$k_2$@
-   @$\vdots$@ 
-  qSelect2 b @$k_m$@.
-\end{lstlisting}
-
-Assuming \texttt{l} contains $n$ elements, this optimization changes the
-asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$ to
-$\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
-complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a significant improvement.
-
-In this paper, we develop a splitting algorithm
-(\ref{sec:splitting,sec:implementation}) which, given a program $e$ in \lang,
-produces an equivalent pair of programs which correspond precisely to the two
-stages of computation in $e$.
-(Splitting is always possible because the staging annotations in $e$ are
-consistent, because $e$ is well-typed in \lang.) In the case of
-\texttt{qsStaged}, our splitting algorithm produces the algorithm described
-above.
-\end{comment}
-%\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
-
-%Because the tree passes information across the stage boundary, we call it the
-%\emph{boundary data structure}.
-
-%Note that the desired optimized code shows above is intellectually more
-%sophisticated than the code that we have started with: the optimized code is
-%able to create a data structure, a balanced binary tree augmented with indexing
-%information, and use a binary search technique over this tree to compute the
-%result asymptotically more efficiently.
-
-%In fact, based our teaching experience, we can imagine this kind of problem to
-%be a moderately difficult exam question in an undergraduate algorithms class, as
-%it not only requires understanding of data structures such as binary search
-%trees but also requires modifying them to augment with indexing information to
-%support rank-based search.
-
-%(I also want to make it clear that recognizing \lang's appropriateness for this
-%is itself is a contribution.)
-
 \end{abstrsyn}

### 2015-02-27 paper/implementation.tex
eliminate inconsistent uses of 'part'ition

@@
   datatype list = Empty | Cons of int * list
-  fun part (...) = ...
+  fun partition (...) = ...
 `3`}`1`
@@
 \end{lstlisting}
-declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
+declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
 by elaborating into:
@@
 val 3`grnd{`Cons3`}`  = 3`grnd{`fn (...) => roll (inj ...)3`}`
-val 3`grnd{`part3`}`  = 3`grnd{`fn (...) => ...3`}`
+val 3`grnd{`partition3`}`  = 3`grnd{`fn (...) => ...3`}`
 val qsStaged    = fn (...) => ...`

### 2015-02-27 paper/implementation.tex
renaming qsStaged

@@
 `3`}`1`
-fun qsStaged (...) = ...`
+fun qss (...) = ...`
 \end{lstlisting}
@@
 val 3`grnd{`partition3`}` = 3`grnd{`fn (...) => ...3`}`
-val qsStaged    = fn (...) => ...`
+val qss = fn (...) => ...`
 \end{lstlisting}

### 2015-02-27 paper/semantics.tex
overview section is now sec:overview
eliminate inconsistent uses of 'part'ition
renaming qsStaged

@@
 %
-\texttt{qsStaged} has a more precise type---it is a first-stage function which
+\texttt{qss} has a more precise type---it is a first-stage function which
 takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$ (an
@@
 Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
-\texttt{part}ition it. The \texttt{case} expression in \texttt{qSelect} depends
+\texttt{partition} it. The \texttt{case} expression in \texttt{qSelect} depends
 on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
-second-stage computations. Since \texttt{qsStaged} itself produces a
+second-stage computations. Since \texttt{qss} itself produces a
 $\fut\rmint$, the remainder of the function is second-stage code.
@@
 expression) as a first-stage expression of type $\fut\rmint$ (the result 
-of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
+of \texttt{qss}). Inside the \texttt{LT} and \texttt{GT} branches, the
 $\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
-to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
+to \texttt{qss}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
 $\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
@@
 syntactically interleaved within an individual term. For example,
-\texttt{qsStaged} contains a first-stage recursive call inside a second-stage
+\texttt{qss} contains a first-stage recursive call inside a second-stage
 \texttt{case}; thus, an ordinary outermost-first evaluation strategy would not
@@
 
-A multistage function, such as \texttt{qsStaged} from \ref{sec:examplea}, has
+A multistage function, such as \texttt{qss} from \ref{sec:overview}, has
 type $A\to \fut (B \to C)$ at \bbonem.%
-\footnote{We can rewrite \texttt{qsStaged} in this curried form, or apply a
+\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
 higher-order currying function.}
@@
 \begin{lstlisting}
-2`let val qs = prev{`1`qsStaged `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
+2`let val qs = prev{`1`qss `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
 2`in (qs k1,...,qs km)`
 \end{lstlisting}
-To see that the partial application of \texttt{qsStaged} is only evaluated once,
+To see that the partial application of \texttt{qss} is only evaluated once,
 observe that after the first stage of evaluation, the residual has the form:

### 2015-02-28 paper/abstract.tex
large pass for consistency/small bugfixes

@@
 In this paper, we consider staged programs written in a typed lambda
-calculus with a (``circle'') modality denoting computation in a later
+calculus with a $\fut$ modality denoting computation in a later
 stages and present techniques for splitting them into multi-pass
 programs. Such multi-pass programs evaluate in a sequence of
-evaluations of un-staged, conventional programs each of which
-corresponds to a pass.  Conceived in the 80's but J{\o}rring and
-Scherlis, this {\em pass-separation} technique has been studied but
-existing techniques are either manual or apply only to simple programs
+evaluations of unstaged, conventional programs each of which
+corresponds to a pass.  Conceived in the 80's by J{\o}rring and
+Scherlis, this {\em pass-separation} technique has been studied, but
+existing techniques are either manual or applicable only to simple programs
 that lack important features such as recursion and first-class
@@
 run significantly more efficiently by assigning frequent computations
-to earlier passes so that they can be re-used (cheaply) later.  Our
+to earlier passes so that they can be reused (cheaply) later.  Our
 experiments based on a prototype implementation show that our

### 2015-02-28 paper/examples.tex
large pass for consistency/small bugfixes

@@
 fun dot (
-  (3`grnd{`1`xliteralone`3`}`1`,`3`grnd{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
-  (3`grnd{`1`xliteraltwo`3`}`1`,`3`grnd{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
+  (3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
+  (3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
 next{
-  2`prev{`1`hold `3`grnd{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
+  2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
 1`}`
@@
 \begin{lstlisting} 
-1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`grnd{`1`e`3`}`1` : ^int) = 
-  if `3`grnd{`1`e == 0`3`}`1` then 
+1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`gr{`1`e`3`}`1` : ^int) = 
+  if `3`gr{`1`e == 0`3`}`1` then 
     next{`2`literalone`1`}
-  else if `3`grnd {`1`(e mod 2) == 0`3`}`1` then
-    exp(next{`2`b*b`1`},`3`grnd{`1`e/literaltwo`3`}`1`)
+  else if `3`gr{`1`(e mod 2) == 0`3`}`1` then
+    exp(next{`2`b*b`1`},`3`gr{`1`e/literaltwo`3`}`1`)
   else 
-    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`grnd{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
+    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`gr{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
 \end{lstlisting}
@@
 
-Note that the recursive \texttt{partition} function is defined within a \texttt{grnd}
+Note that the recursive \texttt{partition} function is defined within a \texttt{gr}
 annotation.  As explained in \ref{sec:needGround}, defining \texttt{partition}
@@
 enough to overpower the asymptotic speedup gained elsewhere.
-With \texttt{grnd} annotations, however, this can be prevented.
+With \texttt{gr} annotations, however, this can be prevented.
 

### 2015-02-28 paper/implementation.tex
large pass for consistency/small bugfixes

@@
 \begin{lstlisting}
-3`atsigngrnd{`1`
+3`atsigngr{`1`
   datatype list = Empty | Cons of int * list
@@
 \begin{lstlisting}
-1`val 3`grnd{`Empty3`}` = 3`grnd{`roll (inj ...)3`}`
-val 3`grnd{`Cons3`}` = 3`grnd{`fn (...) => roll (inj ...)3`}`
-val 3`grnd{`partition3`}` = 3`grnd{`fn (...) => ...3`}`
+1`val 3`gr{`Empty3`}` = 3`gr{`roll (inj ...)3`}`
+val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
+val 3`gr{`partition3`}` = 3`gr{`fn (...) => ...3`}`
 val qss = fn (...) => ...`

### 2015-02-28 paper/intro.tex
large pass for consistency/small bugfixes

@@
 
-Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, and computes
+Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
 $f(x,y_i)$ for each $y_i$:
 \begin{lstlisting}
-fun @$F(x, y_0, y_1, \ldots, y_{m-1})$@ = 
-  @$f(x, y_0);  f(x, y_1); \ldots; f(x, y_{m-1})$@
+fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
+  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
 \end{lstlisting}
@@
 \begin{lstlisting}
-fun @$F_v(y_0, y_1, \ldots, y_{m-1})$@ = 
-  @$f_v(y_0);  f_v(y_1); \ldots; f_v(y_{m-1})$@.
+fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
+  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
 \end{lstlisting}
-eliminating the need to recompute the parts of $f(v,-)$ which do not depend on
-the second argument.
+eliminating the need to compute $m$ times those parts of $f(v,-)$ which do not
+depend on the second argument.
 
 A closely related technique, {\em metaprogramming}, allows writing
-programs that generate specialized code
-(e.g.,~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
-enables fine-grained control over specialization by requiring staging
-annotations that make explicit the stage of each computation. (In
+programs that generate specialized
+code~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
+enables fine-grained control over specialization by requiring \emph{staging
+annotations} that explicitly mark the stage of each computation. (In
 partial evaluation, we typically assume that a \emph{binding-time
@@
 \begin{lstlisting}
-fun @$F_\textrm{multipass}(x, y_0, y_1, \ldots, y_{m-1})$@ = 
-  let @$fx = f_1(x)$@
-  in @$(f_2(fx, y_0);  f_2(fx, y_1)  \ldots f_2(fx, y_{m-1}))$@.
+fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
+  let @$z = f_1(x)$@
+  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
 \end{lstlisting}
 %
-The key difference between partial evaluation (or metaprogramming) and
-pass separation is that the latter can be performed without access to
-the first argument $x$: $F_\textrm{multipass}$ works for any $x$,
+The key difference between pass separation and partial evaluation (or
+metaprogramming) is that the former can be performed without access to
+the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
 while $F_v$ is defined only for $x=v$.  Note that unlike
 precomputation, frequency reduction does not require knowledge of
-arguments; if a particular $v$ is known, we can use partial evaluation
+arguments; however, if a particular $v$ is known, we can use partial evaluation
 to specialize $F_\textrm{multipass}$ to that $v$.

### 2015-02-28 paper/overview.tex
delete obsolete quickselect figure
continuing a pass through the figures
eliminate inconsistent uses of 'part'ition
large pass for consistency/small bugfixes

@@
 
-\input{figures/quickselect-aligned}
+\input{figures/quickselect}
 
@@
 on a list \texttt{l}. To this end, we can use
-quickselect~\cite{Hoare:1961}, which give a list \texttt{l} and a rank
+quickselect~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
 \texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
@@
 \lstinline{1`hold : ^int -> $`2`int`}.
-%
 
@@
 \texttt{qs}, but we chose annotations that maximize the work performed
-in the first stage
-
-% The $\curr$ and \texttt{grnd} (``ground'') annotations
-% further distinguish those first-stage computations which do not
-% contain any embedded second-stage code. (Code \emph{inside}
-% \texttt{grnd}, like the \texttt{part}ition function, is guaranteed not
-% to contain second-stage code; first stage code outside a
-% \texttt{grnd}, like \texttt{qsStaged}, may.) We will discuss the
-% importance of this additional distinction in \ref{sec:splitting}.
-
-% Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
-% \texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
-% non-circle type are available for immediate use. Indeed, we case on the list as
-% usual, and in the \texttt{Cons} branch, \texttt{part}ition it.
+in the first stage.
 
@@
 
-\input{figures/quickselect-split-aligned}
+\input{figures/quickselect-split}
 

### 2015-02-28 paper/splitting.tex
large pass for consistency/small bugfixes

@@
 \begin{lstlisting}
-2`[]`1` grnd{1+2}`
+2`[]`1` 3`gr{`1`literalone+literaltwo`3`}`
 \end{lstlisting}
@@
 \begin{lstlisting}
-2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (grnd{7}, next{`2`yhat`1`}))`
+2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (3`gr{`1`7`3`}`, next{`2`yhat`1`}))`
 \end{lstlisting}
@@
 %\begin{lstlisting}
-%4`##########`1` (`4`######`1`, injL (grnd{7}, `4`######`1`))`
+%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
 %\end{lstlisting}
@@
 \begin{lstlisting}
-1`((), injL (grnd{7},()))`
+1`((), injL (3`gr{`1`7`3`}`,()))`
 \end{lstlisting}
-Finally, masking drops \texttt{grnd} annotations yielding:
+Finally, masking drops \texttt{gr} annotations yielding:
 \begin{lstlisting}
@@
 \end{lstlisting}
-To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{grnd} blocks
+To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{gr} blocks
 (replacing them with unit), \texttt{next} annotations, and injection tags:
@@
 %\end{lstlisting}
-%Where the \texttt{grnd} blocks left a hole, we put in a unit values:
+%Where the \texttt{gr} blocks left a hole, we put in a unit values:
 \begin{lstlisting}
@@
 % In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
-% Note that the \texttt{grnd} and \texttt{next} annotations are erased.
+% Note that the \texttt{gr} and \texttt{next} annotations are erased.
 

### 2015-02-28 paper/examples.tex
hyphenating "first-stage" and "second-stage" when appropriate

@@
 A simple example of composition of staged functions (representing GPU pipeline programs) is given below.
-The code defines a pipeline as a staged function that takes an object definition in first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel.
+The code defines a pipeline as a staged function that takes an object definition in the first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel.
 The code defines a pipeline combinator \texttt{shade} which modulates the results of the object's reflectance at a given pixel with the results of a function that computes the object's albedo (surface color).

### 2015-02-28 paper/overview.tex
hyphenating "first-stage" and "second-stage" when appropriate

@@
 does \texttt{qs2} knows to traverse the tree? It does not. Since
-the splitting algorithm recorded the control flow of the first stage
+the splitting algorithm recorded the control flow of the first-stage
 execution in the boundary data structure, \texttt{qs2} simply follows

### 2015-02-28 paper/relatedwork.tex
hyphenating "first-stage" and "second-stage" when appropriate

@@
 straight-line programs and also treat all computations guarded by
-second stage conditionals as second stage computations.
+second-stage conditionals as second-stage computations.
 

### 2015-02-28 paper/splitting.tex
hyphenating "first-stage" and "second-stage" when appropriate

@@
 Splitting statically transforms terms in \lang\ into a pair of
-monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first
-stage and second stage computations in the original term. Since full
-evaluation of world \bbtwo\ terms (both first stage and second stage
+monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
+and second-stage computations in the original term. Since full
+evaluation of world \bbtwo\ terms (both first-stage and second-stage
 evaluation) only produces values in stage two, the idea of splitting
@@
 Similarly, when splitting \bbonem\ terms, first pass evaluation of
-$f_1$ should produce the same value as first stage evaluation of the
+$f_1$ should produce the same value as first-stage evaluation of the
 original \lang\ term.  However, complexity arises because
@@
 the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
-the same value as fully evaluating all second stage computations in
+the same value as fully evaluating all second-stage computations in
 the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
@@
 \bbonem\ terms (which may produce results at both the first and second
-stage) and \bbtwo\ terms (which only produce second stage results) is
+stage) and \bbtwo\ terms (which only produce second-stage results) is
 fundamental to the structure of the splitting algorithm presented in
@@
 yields a pair of monostaged terms $\pipeS p l r$. ($p$ and $l.r$
-contain all first and second stage computations in $e$ respectively.)
+contain all first- and second-stage computations in $e$ respectively.)
 Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
@@
 $\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
-$(i,b)$, where the first element $i$ matches the first stage results
+$(i,b)$, where the first element $i$ matches the first-stage results
 contained in $v$.  For example, in the simple case where $\rtab \xi v$
-contains entirely first stage computation, such as:
+contains entirely first-stage computation, such as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-the first stage result is the value 3.  The second element of the pair
+the first-stage result is the value 3.  The second element of the pair
 is the boundary value used as input to the resumer in second pass
 evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
-result produced by second stage computations in $e$.  For example, in
-the case where $\rtab \xi v$ contains entirely second stage
+result produced by second-stage computations in $e$.  For example, in
+the case where $\rtab \xi v$ contains entirely second-stage
 computation, such as:
@@
 \end{lstlisting}
-the second stage result is the value 5. Since in general residual
+the second-stage result is the value 5. Since in general residual
 tables and partial values describe a mixture of both first and second
@@
 operation, called {\em masking}, that separates residual tables and
-partial values into their first and second stage results. We then
+partial values into their first- and second-stage results. We then
 describe the algorithm for splitting both \bbonem\ terms
@@
 %residual, its first-stage computations only exist to generate input
-%for second stage computations.  In contrast, since $\colmix e A$
+%for second-stage computations.  In contrast, since $\colmix e A$
 %reduces to a multi-stage residual table and partial value, its first
@@
 
-%Masking separates residual tables and partial values into first stage
+%Masking separates residual tables and partial values into first-stage
 %results and second stage residuals.
@@
 %where $i$ is a monostage value encoding all
-%first stage components of the partial value, and $q$ is a monostage
+%first-stage components of the partial value, and $q$ is a monostage
 %term encoding all of the second-stage computations of the partial
@@
 The first part of the masked value, $i$, is a monostage value holding
-the result of all first stage components of $v$.  The second
-component, $q$, is a residual representing all second stage
+the result of all first-stage components of $v$.  The second
+component, $q$, is a residual representing all second-stage
 computations in $\xi$ and $v$.
@@
 %Masking is not a transformation performed as part of the splitting
-%algorithm.  Instead, masking defines an intuitive notion of the first
-%stage and second stage ``components'' of residual tables and partial
+%algorithm.  Instead, masking defines an intuitive notion of the first-stage 
+%and second-stage ``components'' of residual tables and partial
 %values, and it is used to define the requirements of what output
@@
 To construct the value $i$ representing its first-stage components,
-masking first redacts all second stage (blue) parts, along with the
+masking first redacts all second-stage (blue) parts, along with the
 surrounding \texttt{next} annotations. (This redaction removes the
-entire residual table since it only represents second stage
+entire residual table since it only represents second-stage
 computations.) The resulting ``holes'' in the term are replaced with
@@
 \end{lstlisting}
-To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{gr} blocks
+To construct the residual $q$ (corresponding to second-stage computations) masking first redacts all \texttt{gr} blocks
 (replacing them with unit), \texttt{next} annotations, and injection tags:
@@
 %\ref{sec:split-one}), and packages the resulting terms as functions in
-%both the first and second stage parts of the masked value.
+%both the first- and second-stage parts of the masked value.
 
@@
 
-In other words, $c'$ cases on $i_1$ to run the first branch of first stage
+In other words, $c'$ cases on $i_1$ to run the first branch of first-stage
 code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
@@
 Since \texttt{next} and \texttt{prev} do nothing more than convert between 
-second stage terms at \bbtwo\ and second stage terms encapsulated within $\fut$s at \bbonem,
+second-stage terms at \bbtwo\ and second-stage terms encapsulated within $\fut$s at \bbonem,
 their splitting rules are relatively simple.
@@
 resumer. This logic occurs regardless of whether the terms forming the
-branches of the \texttt{case} contain second stage computations.
+branches of the \texttt{case} contain second-stage computations.
 Worse, if this \texttt{case} appeared in the body of a recursive
-function with no other second stage computations, splitting would
+function with no other second-stage computations, splitting would
 generate a resumer with (useless) recursive calls.
@@
 no second-stage computations by adding a third world \bbonep\ whose
-terms are purely first stage.  While not necessary for ordinary staged
+terms are purely first-stage.  While not necessary for ordinary staged
 term evaluation (\ref{sec:stagedsemantics}), this annotation is
@@
 \ref{sec:exampleQS} is an example of such a function that contains no
-second stage computations, but, if not defined within a $\pure$
+second-stage computations, but, if not defined within a $\pure$
 annotation, would cause splitting to generate a recursive resumer.

### 2015-02-28 paper/examples.tex
more terminology shifts

@@
 
-fun qs1 (l : list, k : unit) = ((), 
-  case l of
+fun qs1 (l : list, k : unit) = 
+  ((), case l of
     Empty => Leaf
@@
         let (left,right,n) = partition (h,t) in
-        (n, #2 (qs1 left k), h, #2 (qs1 right k))
-      )
-  )`
+        (n, #2 (qs1 left k), h, #2 (qs1 right k))))`
 
@@
 \ref{fig:qs-split}, except that the mixed-stage argument to both passes is
-is explicitly masked, and we do not suppress the trivial immediate result of
-\texttt{qs1}.
+is explicitly masked, and we do not suppress the trivial first-stage result
+returned by \texttt{qs1}.
 

### 2015-02-28 paper/overview.tex
more terminology shifts

@@
 randomly ordered (which can be achieved by pre-permutation of the
-input), function \texttt{qs} runs in expected linear time.
+input), function \texttt{qs} runs in expected $O(n)$ time.
 %
@@
 Scherlis are commonly employed.  For example, as briefly mentioned in
-\secref{intro}, modern graphics software is written exactly in this
+\ref{sec:intro}, modern graphics software is written exactly in this
 way.  Unfortunately, as the example illustrates, multipass programs
-are naturally complex.  We thus ask: is it would be possible to
+are naturally complex.  We thus ask: is it possible to
 perform such transformations automatically.  We answer this question
@@
 The idea behind staged programming is to use staging annotations or
-types to indicate the stage of each expression.  To apply this
+types to indicate the stage of each subterm.  To apply this
 technique to our example, suppose that we have a language with two

### 2015-02-28 paper/semantics.tex
large pass for consistency/small bugfixes
hyphenating "first-stage" and "second-stage" when appropriate
more terminology shifts

@@
 Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
-\texttt{partition} it. The \texttt{case} expression in \texttt{qSelect} depends
+\texttt{partition} it. The \texttt{case} term in \texttt{qSelect} depends
 on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
@@
 
-The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
-expression) as a first-stage expression of type $\fut\rmint$ (the result 
+The $\next$ wraps a second-stage term of type $\rmint$ (the \texttt{case}
+term) as a first-stage term of type $\fut\rmint$ (the result 
 of \texttt{qss}). Inside the \texttt{LT} and \texttt{GT} branches, the
@@
 mutually recursive modes of operation, depending on the world of the term it is
-evaluating. Specifically, first stage evaluation:
+evaluating. Specifically, first-stage evaluation:
 \begin{itemize}
@@
 \begin{lstlisting}
-2`let val qs = prev{`1`qss `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
+2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
 2`in (qs k1,...,qs km)`

### 2015-02-28 paper/splitting.tex
more terminology shifts

@@
 Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
-Splitting $\pure{e}$ follows in similar fashion, except the immediate
-result now becomes the result of evaluating the \bbonep\ term $e$.
+Splitting $\pure{e}$ follows in similar fashion, except the first 
+component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.
 
@@
 However, since the body of a function may itself be multistage, splitting must continue into it.
-The immediate result is a new function formed from the first-stage part of the original body.
+The $i$ is a new function formed from the first-stage part of the original body.
 The resumer is a new function formed out of the second-stage part of the original body.

### 2015-02-28 paper/overview.tex
stage-n => nth-stage

@@
 are typeset in red, and the second-stage code are typeset in blue.  To
-obtain \texttt{qss} from \texttt{qs}, we first make the input list a
-stage-1 and the rank a stage-2 argument.
+obtain \texttt{qss} from \texttt{qs}, we first make the input list 
+first-stage and the rank second-stage.
 %
@@
 with the staging annotations \texttt{prev} and \texttt{next}, which
-transition between a stage-1 and stage-2 code, and~$\texttt{g}$, which
-marks ground terms.  
+transition between first-stage and second-stage code, and~$\texttt{g}$, 
+which marks ground terms.  
 %
@@
 \texttt{hold}, which can be implemented with \texttt{prev} and
-\texttt{next}, for promoting a stage-1 integer to a stage-2 integer;
+\texttt{next}, for promoting a first-stage integer to a second-stage integer;
 the signature of \texttt{hold} is
@@
 them into the function, \texttt{qs1} of the first pass. This function
-performs the stage-1 computations and places the results into a
+performs the first-stage computations and places the results into a
 boundary data structure that both records the control flow and the

### 2015-02-28 paper/semantics.tex
get rid of the text walking through qss -- doesn't really fit anymore

@@
 $\fut$ allows construction of ``mixed-stage'' products and functions.
-For example, quickselect is a function at world \bbonem\ which takes a
+For example, \texttt{qss} is a function at world \bbonem\ which takes a
 $\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
@@
 {\em stability} in \cite{krishnaswami13}.}
-
-\crem{blend in the rest of the text in this section:}
-%
-\texttt{qss} has a more precise type---it is a first-stage function which
-takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$ (an
-$\rmint$ \emph{in the future}), and returns a $\fut\rmint$.
-
-Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
-\texttt{partition} it. The \texttt{case} term in \texttt{qSelect} depends
-on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
-second-stage computations. Since \texttt{qss} itself produces a
-$\fut\rmint$, the remainder of the function is second-stage code.
-
-The $\next$ wraps a second-stage term of type $\rmint$ (the \texttt{case}
-term) as a first-stage term of type $\fut\rmint$ (the result 
-of \texttt{qss}). Inside the \texttt{LT} and \texttt{GT} branches, the
-$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
-to \texttt{qss}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
-$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
-second-stage $\rmint$.
 

### 2015-02-28 paper/overview.tex
tightened up 2.0

@@
 
-We present an overview of our approach by considering an example.  To
-highlight the ideas, we abstract over some technical details, which
-are fully formalized in the rest of the paper.  The example itself is
-also implemented in our implementation.
-
-\paragraph{Computing order statistics.}
-
 \begin{abstrsyn}
@@
 
-Suppose that we wish to compute a series of order statistics queries
-on a list \texttt{l}. To this end, we can use
-quickselect~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
+Suppose that we wish to perform a series of order statistics queries
+on a list \texttt{l}. To this end, we can use the quickselect
+algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
 \texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
 As implemented in an ML-like language in \ref{fig:qs-unstaged},
-function \texttt{qs} partitions the list by using the first element as
-a pivot and then recurs on one of the two resulting sides to find the
-desired element based on the relationship of \texttt{k} to the size of
-the first half \texttt{i}.  Assuming that the input is uniformly
-randomly ordered (which can be achieved by pre-permutation of the
-input), function \texttt{qs} runs in expected $O(n)$ time.
+\texttt{qs} partitions \texttt{l} using the first element as
+a pivot, and then recurs on one of the two resulting sides, depending on
+the relationship of \texttt{k} to the size \texttt{i} of the first half, in
+order to find the desired element.  Assuming that the input is uniformly
+randomly ordered (which can be achieved by prepermuting the input), \texttt{qs}
+runs in expected $O(n)$ time.
 %
-Using function \texttt{qs}, we can perform many order statistics
-queries, for example with $m$ different ranks
-$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
+Using \texttt{qs}, we can perform $m$ different order statistics queries with
+ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
 %
 \begin{lstlisting}
-qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
+(qs l k1, qs l k2, ..., qs l km)
 \end{lstlisting}
 %
-Unfortunately, this approach requires $O(n \cdot m)$ time, making it
-expensive to perform order statistics on more than a small number of
-times. This is unfortunate: we wish to be able to map \texttt{qs} to
-over another list and do so efficiently.
+Unfortunately, this approach requires $O(n \cdot m)$ time.
 
-One way to regain efficiency is to use a multi-pass algorithm that
-performs frequently performed computations first so that they can be
-reused.  It is nontrivial what such computations are in this example
-but after some reflection, we can realize that they are the
-comparisons performed by the partition function.  To minimize them, we
-can pre-sort the input list \texttt{l} into a sorted list \texttt{s}
-in the first pass, and then perform lookups in \texttt{s} in the
-second pass.  Unfortunately, since \texttt{s} is a list, a lookup
-would require linear time (also on average), leading to no improvement
-in efficiency.  But this is not too hard: after sorting \texttt{s} we
-can copy in into an array and use binary search to find the element
-with the desired rank.  As another option, we can construct a binary
-search tree from the input and perform size-based searches on the
-binary search tree to find the elements with the desired rank.  Such
-an approach would fit nicely into the functional paradigm but requires
-one more additional algorithmic ingenuity: to ensure efficiency we
-would need to store at each note the size of its subtree (these are
-sometimes called ``augmented trees'').  In summary, we are able to
-improve efficiency by replacing our simple solution with a multi-pass
-algorithm that first preprocesses the input to generate a lookup data
-structure and then in the second pass performs fast lookups.  As
-outlined above, in addition, this transformation is highly non-trivial
-as it involves reasoning about intricate algorithmic concerns, and
-implementing more complex algorithms and data structures.  Both of
-these approaches however improve the run-time to (expected)
-$\Theta(n\log{n} + m\log{n})$, a near linear-time improvement in $m$.
+\pagebreak
+We can attempt to improve on this algorithm by factoring out computations shared
+between these calls to \texttt{qs}. In particular, the comparisons performed by
+\texttt{partition} depend only on \texttt{l}, although different calls to
+\texttt{partition} are made by each call to \texttt{qs}.
 
-Such transformations, called pass-separation by J{\o}rring and
-Scherlis are commonly employed.  For example, as briefly mentioned in
-\ref{sec:intro}, modern graphics software is written exactly in this
-way.  Unfortunately, as the example illustrates, multipass programs
-are naturally complex.  We thus ask: is it possible to
-perform such transformations automatically.  We answer this question
-affirmatively.  Our approach is to start with a staged language and
-then use a splitting algorithm to generate multi-pass programs from
-staged programs.
+A first cut at a solution might be to, in a first pass, sort \texttt{l} into
+\texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
+Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since since
+\texttt{s} is a list, leading to no improvement in efficiency.
+
+A more sophisticated solution is to, in a first pass, construct a binary search
+tree out of \texttt{l}, and in a second pass, simply look for the \texttt{k}th
+leftmost element of that tree. To do this efficiently requires one more
+innovation---storing at each node the size of its left subtree (i.e., using
+\emph{augmented trees}). This approach has (expected) runtime $\Theta(n\log{n} +
+m\log{n})$, a nearly linear-time improvement in $m$.
+
+In both cases, we replace our simple solution with a two-pass algorithm which
+first processes the input into a lookup data structure, and then subsequently
+performs fast lookups. Improving algorithms in this way is highly non-trivial,
+as it requires reasoning about intricate algorithmic concerns, and implementing
+more complex algorithms and data structures.
+
+In this paper, we present a splitting algorithm which transforms staged programs
+into multi-pass programs. In certain cases, this algorithm \emph{automatically}
+improves algorithms in the way described above.
+
+%Such transformations, called pass-separation by J{\o}rring and
+%Scherlis are commonly employed.  For example, as briefly mentioned in
+%\ref{sec:intro}, modern graphics software is written exactly in this
+%way.  Unfortunately, as the example illustrates, multipass programs
+%are naturally complex.
 
 \subsection{Staging}
+
+%The example itself is also implemented in our implementation.
 

### 2015-02-28 paper/abstract.tex
Editing pass on abstract (to tighten a bit)

@@
 Staged programming languages assign a stage to each expression and
-evaluate each expression in its assigned stage.  For example, in a
-two-staged program, some expressions are evaluated in the first stage
-and others in the second stage.  Programming languages, evaluation
-strategies, and type systems for staged programs where stages roughly
-correspond to the arrival order of inputs have been studied
-extensively in the areas of partial evaluation and meta programming.
-
-In this paper, we consider staged programs written in a typed lambda
-calculus with a $\fut$ modality denoting computation in a later
-stages and present techniques for splitting them into multi-pass
-programs. Such multi-pass programs evaluate in a sequence of
-evaluations of unstaged, conventional programs each of which
-corresponds to a pass.  Conceived in the 80's by J{\o}rring and
-Scherlis, this {\em pass-separation} technique has been studied, but
-existing techniques are either manual or applicable only to simple programs
+evaluate each expression in its assigned stage.
+%For example, in a
+%two-staged program, some expressions are evaluated in the first stage
+%and others in the second stage.
+%Programming languages, evaluation
+%strategies, and type systems for staged programs where stages roughly
+%correspond to the arrival order of inputs have been studied
+%extensively in the areas of partial evaluation and meta programming.
+While staged programming has been studied extensively in the areas of
+partial evaluation and meta programming, in this paper we present an
+algorithm for statically {\em splitting} programs witten in a
+two-stage typed lambda calculus (with a $\fut$ modality denoting
+computation in a later stage) into multi-pass programs that are
+structured as two dependent, unstaged programs and are evaluated using
+two distinct execution passes.  While splitting, also called pass
+separation or data specialization, has been studied in the past, prior
+techniques were either manual or applicable only to simple programs
 that lack important features such as recursion and first-class
@@
 
-% We make no restrictions on the kinds of programs that can
-% be written.  We also present an implementation.
-% %
-
-Since we build on a lambda calculus, our techniques enable the
-programmer to write staged programs by using powerful programming
-abstractions in a composable and modular fashion. Our splitting
-transformation maps such programs into multi-pass programs that can
-run significantly more efficiently by assigning frequent computations
-to earlier passes so that they can be reused (cheaply) later.  Our
-experiments based on a prototype implementation show that our
-splitting algorithm is able perform highly non-trivial algorithmic
+Since our work builds on a lambda calculus, our splitting algorithm
+enables the programmer to write staged programs by using powerful
+programming abstractions in a composable and modular fashion. Our
+splitting transformation maps such programs into multi-pass programs
+that can run significantly more efficiently by assigning frequent
+computations to earlier passes so that they can be reused (cheaply)
+later.  Our experiments based on a prototype implementation show that
+our splitting algorithm is able perform highly non-trivial algorithmic
 transformations, sometimes automatically improving efficiency

### 2015-02-28 paper/overview.tex
remove accidental page break
tightening 2.1

@@
 
-\pagebreak
 We can attempt to improve on this algorithm by factoring out computations shared
@@
 
-The idea behind staged programming is to use staging annotations or
-types to indicate the stage of each subterm.  To apply this
-technique to our example, suppose that we have a language with two
-worlds representing the two stages and a third, {\em ground} world
-that contains terms with no staging.
+The idea behind staged programming is to use staging annotations---in our case,
+guided by types---to indicate the stage of each subterm. We consider a
+language \lang\ with two modal worlds representing the two stages, and a third
+{\em ground} world of unstaged terms.
 
-We can write a staged version of \texttt{qs}, called \texttt{qss},
-whose code is shown in \ref{fig:qs-staged}, where the first-stage code
-are typeset in red, and the second-stage code are typeset in blue.  To
-obtain \texttt{qss} from \texttt{qs}, we first make the input list 
-first-stage and the rank second-stage.
+In \ref{fig:qs-staged} we show a staged version of \texttt{qs}, called
+\texttt{qss}, where first-stage code is colored red, and second-stage blue. In
+\texttt{qss}, we regard the input list as arriving in the first stage
+(with type $\curr\mathrm{list}$, a list ``now''), the input rank as arriving in
+the second stage (with type $\fut\rmint$, an integer in the ``future''), and the
+result as being produced in the second stage (with type $\fut\rmint$).
+
+\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
+\texttt{prev} and \texttt{next}, signaling transitions between first- and
+second-stage code, and $\texttt{g}$, signaling ground terms.  
 %
-More precisely the input list has type $\curr\mathrm{list}$ (an
-integer list ``now''), the rank has type $\fut\rmint$ (an integer in
-the ``future''), and the return has type $\fut\rmint$.
+We also use a function \texttt{hold},
+\lstinline{1`hold : ^int -> $`2`int`},
+to promote first-stage integers to second-stage integers.
 %
-We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
-with the staging annotations \texttt{prev} and \texttt{next}, which
-transition between first-stage and second-stage code, and~$\texttt{g}$, 
-which marks ground terms.  
-%
-We also use a third function
-\texttt{hold}, which can be implemented with \texttt{prev} and
-\texttt{next}, for promoting a first-stage integer to a second-stage integer;
-the signature of \texttt{hold} is
-% 
-\lstinline{1`hold : ^int -> $`2`int`}.
+Our type system ensures that the staging annotations in \texttt{qss} are
+consistent, in the sense that computations marked as first-stage cannot depend
+on ones marked as second-stage.
 
-Our type system ensures that the staging annotations are consistent,
-in the sense that computations marked as first-stage cannot depend on
-ones marked as second-stage.
-%
 The process of adding staging annotations to unstaged code has been
-the subject of extensive research (\ref{sec:related}). We assume that
-these annotations have been provided; we do not consider the problem
-of generating them. In this example, there are other ways to annotate
-\texttt{qs}, but we chose annotations that maximize the work performed
-in the first stage.
+the subject of extensive research (\ref{sec:related}). In this paper, we do not
+consider this problem, instead assuming that the annotations have been provided.
+In the case of \texttt{qss}, we have specifically chosen annotations which
+maximize the work performed in the first stage.
 

### 2015-02-28 paper/relatedwork.tex
Finished pass through related work.

@@
 Perhaps one of the most well studied example of precomputation is
-partial evaluation.  Partial evaluation distinguishes between a static
-(compile-time) and dynamic (run-time) stage. Given a program and the
-values of static inputs (inputs that are available at compilation
+partial evaluation.  Partial evaluation distinguishes between static
+(compile-time) and dynamic (run-time) stages. Given a program and values 
+for all static inputs (inputs available at compilation
 time), partial evaluation generates a specialized program by
-performing the computations that depend on the statics
+performing computations that depend only on the static
 inputs~\cite{jones96}.  The roots of partial evaluation go back to
@@
 appears to be the first person to use the term in his work on
-incremental computation~\cite{Lombardi67}.  Since then, there has a
-lot of work on partial evaluation, only broad-strokes of which we can
-mention in this here.  We refer the interested reader to the excellent
-book by Jones, Gommard, and Sestoft for a comprehesive discussion of
-work until early 90's~\cite{JGS93}.
+incremental computation~\cite{Lombardi67}.  There has since been 
+an enormous body of work on partial evaluation, and we refer the reader to the
+book by Jones, Gommard, and Sestoft for a comprehensive discussion of
+work until the early 90's~\cite{JGS93}.
 
-Earlier approaches to partial evaluation can be viewed as operating in
+Early approaches to partial evaluation can be viewed as operating in
 two stages: binding time analysis and program specialization.  For a
@@
 be computed by the knowledge of static arguments. Using this
-information and the static arguments themselves, program
-specialization specializes the program to a partially evaluated
-program that can be used with many different dynamic arguments.  This
+information and the values of static arguments, program
+specialization specializes the original program to a partially evaluated
+one that operates on many different dynamic arguments.  This
 approach has been applied to construct partial evaluators for a number
@@
 Experience with binding time analysis showed that it can be difficult
-to control, leading to programs whose performance can be difficult to
+to control, leading to programs whose performance was difficult to
 predict. This led to investigations based on type systems for making
-explicit in the program the stage of each
-computation~\cite{GJ91-lambda,NN92-twolevel} and writing ``meta
-programs'' that when evaluated at a stage, give another program to be
-evaluated at the next stage. Jorring and Scherlis describe this idea
-meta programming in their paper (~\cite{JS86-staging}) as one of the
-three possible approaches to taking advantage of
-staging. Davies~\cite{davies96} presented a logical construction of
-binding-time type systems by deriving a type system via the
-Curry-Howard isomorphism applied to temporal logic.  Davies and
-Pfenning proposed a new type system for staged computation based on a
-particular fragment of modal logic~\cite{DP01-modal}. The work on
-MetaML extended type-based techniques to a full-scale language by
-developing statically typed programming language based on ML that
+the stage of computations explicit in programs~\cite{GJ91-lambda,NN92-twolevel}
+and writing ``meta programs'' that, when evaluated at a stage, yield
+a new program to be evaluated at the next stage.
+%J{\o}rring and
+%Scherlis describe the idea of meta programming as one of the three possible approaches to
+%taking advantage of staging~\cite{JS86-staging}.
+Davies~\cite{davies96} presented a
+logical construction of binding-time type systems by deriving a type
+system via the Curry-Howard isomorphism applied to temporal logic.
+Davies and Pfenning proposed a new type system for staged computation
+based on a particular fragment of modal logic~\cite{DP01-modal}. The
+work on MetaML extended type-based techniques to a full-scale language
+by developing statically typed programming language based on ML that
 enables the programmer to express programs with multiple
@@
 Nanevksi and Pfenning further extended the these techniques by
-allowing free variable to occur within staged
+allowing free variables to occur within staged
 computations~\cite{NP05-nn}.
@@
 small ways to account for the specific evaluation model that we
-propose.  The key difference between our work and this prior work is
-our focus on the problem of pass separation: we developed an algorithm
-for splitting terms in the proposed staged language.
+propose, however the key difference between our work and this prior work is
+that we instead focus on the problem of pass separation.
+%we developed an algorithm
+%for splitting terms in the proposed staged language.
 %the staging or the splitting algorithm, which given a two-staged
@@
 %
-Unlike partial evaluation, splitting requires no knowledge of program
-argument values, only their staging.
+%Unlike partial evaluation, splitting requires no knowledge of program
+%argument values, only their staging.
 %Using just the staging
@@
 %significant ways.
-We are not aware of prior work in the context of
-higher-order functional languages that presents such a splitting
-algorithm.
 
-An important class of pre-computation techniques include incremental
-computation, where a program can respond to small changes to its input
-efficiently by recomputing only the parts of the computation affected
+Another class of pre-computation techniques is incremental
+computation, where a program can efficiently respond to small changes
+to its input by only recomputing parts of the computation affected
 by the
@@
 Like other pre-computation techniques, incremental computation can be
-thought of as a staged model, where changes (in inputs) come over a
+thought of as a staged model, where changes in inputs come over a
 period of time in stages.
 %
-However, incremental computation does not fix any of its inputs or anticipate the future
-inputs, allowing in the most general case for all data to change. As a
-result, its benefits are dependent on the data instance (nature of the
-inputs). In addition, incremental computation techniques typically
-utilize significant run-time (dynamic) information, because no
-assumptions about input instances are made.  For example, we can apply
-incremental computation to our quickselect example
-(ref{sec:overview}), but we would not be able to achieve the bounds
-that we did because incremental computation techniques would unfold
-the quickselect function based on the demanded ranks, which can
-require linear time (as opposed to logarithmic) at each stage.  In
-addition, such an implementation would maintain rather sophisticated
+However, incremental computation does not requiring fixing any of its
+inputs and allows, in the general case, for all inputs to
+change. Thus, its benefits are dependent the input changes made and
+techniques typically utilize significant run-time (dynamic)
+information.  For example, while we can apply incremental computation
+to our quickselect example (\ref{sec:overview}), doing so would not
+realize the same bounds as splitting since incremental computation
+techniques would unfold the quickselect function based on the demanded
+ranks, which can require linear time (as opposed to logarithmic) at
+each stage.  Such an implementation would also maintain sophisticated
 data structures dynamically at run time, the benefits of which depend
-highly on the forms of changes applied.
+highly on input changes applied.
 
@@
 statically splitting an explicitly staged program into two stages.
-However, they only consider simple first-order language and
-straight-line programs and also treat all computations guarded by
-second-stage conditionals as second-stage computations.
+However, they only consider a simple first-order language and
+straight-line programs. Their work also treats all computations
+guarded by second-stage conditionals as second-stage computations,
+which would prevent optimization of programs such as quickselect.
 
-Automatic pass seperation has also been a topic of interest in
+Automatic pass separation has also been a topic of interest in
 programming systems for computer graphics, where, to achieve
 high-performance, programs are typically manually separated into
-components by frequency of execution (Seperation into a pre-determined
+components by frequency of execution (Separation into a pre-determined
 set of per-frequency modules is \emph{required} by the execution model
@@
 programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
-deferring the task of pass seperation to the compiler. However, all
+deferring the task of pass separation to the compiler. However, all
 prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
-limited to simple, imperative languages.
+limited to simple, imperative languages. We are not aware of prior
+work the presents a splitting algorithm in the context of higher-order
+functional languages.
 

### 2015-02-28 paper/conclusion.tex
Editing pass on conclusion.

@@
 
-This paper presents an algorithm for translating programs written in a
-staged lambda calculus into multi-pass programs, which evaluate as two
-monostage programs in a sequence of two dependent passes. We present a
-formalization of the techniques in a way that can be implemented in
-practice.  While we have provided an implementation of this algorithm
-that successfully splits the examples considered in this paper, we
-have stopped short of offering a full proof of correctness, which we
-leave for future work.
+This paper provides a formal presentation of an algorithm for
+statically translating programs written in a two-stage lambda calculus
+into multi-pass programs, which evaluate as two monostage programs in
+two dependent passes.  While we have provided an implementation of
+this algorithm that successfully splits the examples considered in
+this paper, we have stopped short of offering a full proof of
+correctness, which we leave for future work.
 
-In addition, we are interested in providing a more complete system
-implementation as well as investigating how ideas from \lang\ and the
-its splitting algorithm can be integrated into existing practical
-systems such as staged programming systems for computer graphics
-pipelines.  On the more conceptual side, it would be interesting to
-consider the extension of our splitting techniques to richer source
-languages, for example, those that include parallelism.
+In addition, we are interested in exploring the behavior of splitting
+for a broader set of examples as well as investigating how ideas from
+\lang\ and its splitting algorithm can be integrated into existing
+practical systems such as staged programming systems for computer
+graphics pipelines.  On the more conceptual side, it would be
+interesting to consider the extension of our splitting techniques to
+richer source languages that include mutation, parallel constructs,
+and more than two stages.

### 2015-02-28 paper/intro.tex
more terminology shifts
Editing pass on abstract (to tighten a bit)
tightened up 2.2

@@
 results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
-
 While performant, this paradigm of programming with explicit passes results in
@@
 precomputation by specializing functions to specific arguments. Going back to
-our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
-$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
+our example, if $f$ specializes to a particular $v$, written $f_v$, such that
+$f(v,y) = f_v(y)$, then $F$ can be specialized to $v$ as
 \begin{lstlisting}
@@
 straight-line arithmetic operations to recursive and higher-order functions. In
-the case of \texttt{quickselect}, which we discuss next, the split code is in
-fact asymptotically faster over many $y_i$.
+the case of the quickselect algorithm, which we discuss next, the split code is
+in fact asymptotically faster over many $y_i$.
 

### 2015-02-28 paper/overview.tex
tightened up 2.2

@@
 leftmost element of that tree. To do this efficiently requires one more
-innovation---storing at each node the size of its left subtree (i.e., using
-\emph{augmented trees}). This approach has (expected) runtime $\Theta(n\log{n} +
-m\log{n})$, a nearly linear-time improvement in $m$.
+innovation---augmenting the tree by storing at each node the size of its left
+subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
+nearly linear-time improvement in $m$.
 
@@
 
-In this paper, we present a splitting algorithm which transforms staged programs
-into multi-pass programs. In certain cases, this algorithm \emph{automatically}
-improves algorithms in the way described above.
+In this paper, we present a splitting algorithm which automatically transforms
+staged programs into multi-pass programs. In certain cases, this algorithm even
+improves input algorithms in the way described above!
 
@@
 \texttt{prev} and \texttt{next}, signaling transitions between first- and
-second-stage code, and $\texttt{g}$, signaling ground terms.  
+second-stage code, and $\texttt{gr}$, signaling ground terms.  
 %
-We also use a function \texttt{hold},
-\lstinline{1`hold : ^int -> $`2`int`},
+We also use a function
+%\lstinline{1`hold : ^int -> $`2`int`},
 to promote first-stage integers to second-stage integers.
@@
 
-The staged quickselect code shown in \ref{fig:qs-staged} makes
-explicit the staging of all terms, making it natural to ask, whether
-we can transform this code into manually implemented multi-pass code
-described above and match the efficiency improvements of the manual
-implementation. Considering the algorithmic knowledge and the
-considerations needed, this may seem like a tall order.
-Interestingly, our splitting algorithm achieves exactly this.
-In the rest of this section, we present a brief, high-level overview
-of the main ideas behind this algorithm using the quickselect example.
+In the rest of this section, we present a high-level overview of the main ideas
+behind our splitting algorithm, applied to \texttt{qss}.
 %
+Splitting \texttt{qss} yields a two-pass program that creates a
+probabilistically balanced, augmented binary search tree (which is isomorphic to
+a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
+particular, its first pass (\texttt{qs1}) constructs such a binary search tree,
+and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
+information to find the element of the desired rank.
+%
+The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-staged}.
 
-When applied to the staged code in \ref{fig:qs-staged}, our splitting
-algorithm yields a two-pass program that uses the binary-search-tree
-based implementation.  Specifically, in its first pass, the two-pass
-program takes the input list and constructs a probabilistically
-balanced binary search tree, which is isomorphic to a treap data
-structure~\cite{SeidelAr96}.  In the second pass, the program performs,
-for each rank, a binary search tree lookup, by walking the tree to
-find the element with the desired rank.
+%Our splitting algorithm is guided by the local structure of the input program.
+%In particular, the algorithm has no special knowledge of the quickselect
+%algorithm, binary search trees, or how to perform lookups on binary search
+%trees, but it is able to derive them from the input program.
+
+Our splitting algorithm scans \texttt{qss} for first-stage computations,
+gathering them into \texttt{qs1}. Given \texttt{l}, this pass performs these
+computations and places the information needed by the subsequent pass into a
+boundary data structure.
 %
-\ref{fig:qs-staged} illustrates the code for the first and second
-passes \texttt{qs1} and \texttt{qs2} of the staged function
-\texttt{qss}.
+In particular, \texttt{qs1} performs all recursive calls and evaluates all
+instances of \texttt{partition} (since it depends only on \texttt{l}). It
+produces a boundary data structure that collects the results from these
+recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
+in which that call occurred. Since the recursive calls occur in two different
+branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
+%
+Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
+(the pivot/head of the list) in the boundary structure, because those
+computations are \emph{held} for use in the second stage.
+%
+The final result is a binary search tree augmented with size information, and
+whose keys are the pivots.
 
-To create the multi-pass algorithm, the splitting algorithm operates
-by composing local transformations on the subterms of the input
-program.  In particular, the algorithm has no special knowledge of the
-quickselect algorithm, binary search trees, or how to perform lookups
-on binary search trees, but it is able to derive them from the input
-program.
-%
-
-The splitting algorithm scans the program code for first-stage
-computations (which depend only on first-stage values) and separates
-them into the function, \texttt{qs1} of the first pass. This function
-performs the first-stage computations and places the results into a
-boundary data structure that both records the control flow and the
-results from the first stage at relevant control-flow points.  More
-specifically, the function \texttt{qs1} performs all the recursive
-calls and evaluates all instances of the partition function, which
-depend only on the input list.  The function produces a boundary data
-structure that collects the results from all recursive calls along
-with a tag that indicates the control branches taken.  Since
-\texttt{qss} has a binary control structure (casing on the list), the
-boundary is a binary tree.  To ensure that the trace contains the
-necessary results to complete the execution in the second pass, where
-second-stage values may be used, the splitting algorithm includes in
-the nodes of the tree information such as the "pivot" used for
-splitting and the size of the subtree at that note.  The resulting
-tree is thus not just a tree but a binary search tree, keyed by the
-pivot and augmented with "size" information.
-
-As the splitting algorithm scans \texttt{qss} for computations that
-can be performed in the first stage, it also collects computations
-that must be left to the second stage in a separate function
-\texttt{qs2}.  This function, which is executed in the second pass,
-takes as argument the boundary and the second-stage argument (the
-rank) and performs a lookup in the boundary data structure.  But how
-does \texttt{qs2} knows to traverse the tree? It does not. Since
-the splitting algorithm recorded the control flow of the first-stage
-execution in the boundary data structure, \texttt{qs2} simply follows
-this control structure and performs at each point the parts of the
-computation from \texttt{qss} that can now be performed in the
-presence of the second-stage argument (the rank).  In the context of
-our example, this performs a lookup on the boundary data structure by
-using the supplied rank.
+Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
+computations, gathering them into \texttt{qs2}. This pass is given the boundary
+data structure and the rank \texttt{k}, and finishes the computation. Now that
+\texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,
+choosing which recursive call of \texttt{qss} is actually relevant for this
+\texttt{k}. Since the boundary data structure contains the first-stage data for
+\emph{all} of the recursive calls, performing these comparisons essentially
+walks the tree, using the rank along with the size data \texttt{i} in order to
+look up the \texttt{k}th leftmost node in the tree.
 
@@
   We use a few standard facts.  We note first that \texttt{qs1} is
-  isomorphic to a quick-sort algorithm but builds a binary search tree
+  isomorphic to a quicksort algorithm but builds a binary search tree
   instead of combining the returned results in a list.  The run-time
@@
   The depth of the tree generated by \texttt{qs1} is isomorphic to the
-    depth of a run of the quickselect algorithm or equivalently to the
+    depth of a run of the quickselect algorithm, or equivalently to the
     span of a parallel implementation of the quicksort algorithm, both
@@
     $\Theta(\log{n})$ time.  Since the total time is the sum of these
-    random variables and since the expectation of the sum is sum of
+    random variables, and the expectation of the sum is the sum of
     the expectations of the random variables, the bound follows.

### 2015-02-28 paper/relatedwork.tex

@@
 Perhaps the closest work to ours is the algorithm for ``data
-specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
-our work, Knoblock and Ruf are concerned about the problem of
-statically splitting an explicitly staged program into two stages.
-However, they only consider a simple first-order language and
+specialization'' given by Knoblock and Ruf~\cite{knoblock96}, which
+also seeks to statically split an explicitly staged program into two
+stages.  However, they only consider a simple first-order language and
 straight-line programs. Their work also treats all computations
 guarded by second-stage conditionals as second-stage computations,
-which would prevent optimization of programs such as quickselect.
+which would prevent optimization (via splitting) of programs such as
+quickselect.
 
@@
 deferring the task of pass separation to the compiler. However, all
-prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
-limited to simple, imperative languages. We are not aware of prior
-work the presents a splitting algorithm in the context of higher-order
-functional languages.
+prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
+been limited to simple, imperative languages. We are not aware of
+prior work the presents a splitting algorithm in the context of
+higher-order functional languages.
 

### 2015-02-28 paper/overview.tex
minor fixes

@@
 We also use a function
-%\lstinline{1`hold : ^int -> $`2`int`},
+\lstinline{1`hold : ^int -> $`2`int`},
 to promote first-stage integers to second-stage integers.
@@
 
-\begin{theorem}
+\begin{theorem*}
   Consider an execution of \texttt{qs1} with a randomly permuted input
@@
   $\Theta(n\log{n} + m\log{n})$ in expectation.
-\end{theorem}
+\end{theorem*}
 \begin{proof}

### 2015-02-28 paper/overview.tex
wordsmithing

@@
 \texttt{qss}, where first-stage code is colored red, and second-stage blue. In
-\texttt{qss}, we regard the input list as arriving in the first stage
-(with type $\curr\mathrm{list}$, a list ``now''), the input rank as arriving in
-the second stage (with type $\fut\rmint$, an integer in the ``future''), and the
-result as being produced in the second stage (with type $\fut\rmint$).
+\texttt{qss}, we regard the input list \texttt{l} as arriving in the first stage
+(with type $\curr\mathrm{list}$, a list ``now''), the input rank \texttt{k} as
+arriving in the second stage (with type $\fut\rmint$, an integer in the
+``future''), and the result as being produced in the second stage (with type
+$\fut\rmint$).
 

### 2015-02-28 paper/splitting.tex
4.3 all the way to next/prev

@@
 Our algorithm generates a first-pass term $c'$ that produces the same
-boundary value ($b$) as the split of the subterm $e$:
+boundary value as splitting $e$:
 \[
@@
 \]
-Splitting also generates the same resumer as the split of the
-subterm ($l.r' = l.r$)
+Splitting the injection also generates the same resumer as splitting
+the subterm $e$: ($l.r' = l.r$).
 
@@
 \]
-And deconstructed with a tuple pattern, $l.r = \tup{l_1,l_2}.\tup{r_1,r_2}$,
-
-\paragraph {Case}
+And deconstructed in the resumer with the tuple pattern,
+\[
+l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
+\]
+\paragraph {Case.}
 Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
 Here we consider only the case where
-$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
+$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
 (The \texttt{inr} case is analogous.)
 
+\noindent
 Inductively,
@@
 Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
-and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$,
-and so the entire case evaluates like:
-$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$.
-
-Inductively,
+and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
+Therefore:
 \[
@@
 
-The invariant, applied to the entire \texttt{case}, requires some $c'$ and
+The entire case evaluates as
+$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
+some $c'$ and
 $l'.r'$ such that 
@@
 \]
-If you work it out you will see that $c'$ reduces to
-$\tup{i_2,\tup{b_1,\inl{b_2}}}$, meaning that 
-$b' = \tup{b_1,\inl{b_2}}$, and
-$[b'/l']r'$ essentially reduces to
-$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, which is exactly the 
-$[q_1/x_2][b_2/l_2]r_2$ we expected.
+The term $c'$ reduces to
+$\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
+$b' = \tup{b_1,\inl{b_2}}$. Therefore
+$[b'/l']r'$ reduces to
+$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
+$[q_1/x_2][b_2/l_2]r_2$ required.
 
-In other words, $c'$ cases on $i_1$ to run the first branch of first-stage
+More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
 code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
@@
 
-\paragraph {Function} 
-Function introduction has a $\tup{}$ boundary value,
-since functions are already fully reduced in our semantics.
-However, since the body of a function may itself be multistage, splitting must continue into it.
-The $i$ is a new function formed from the first-stage part of the original body.
-The resumer is a new function formed out of the second-stage part of the original body.
-It is the responsibility of the application site to save the precomputation of the function body
-and pass it to the resumer version of the function.
+\paragraph {Function.} 
+Function introduction has a $\tup{}$ boundary value, since functions
+are already fully reduced in our semantics.  However, since the body
+of a function may itself be multistage, splitting must continue into
+it.  The first-pass term generate by split produces an $i$ that is a
+new function formed from the first-stage part of the original body.
+The resumer is a new function formed out of the second-stage part of
+the original body.  It is the responsibility of the application site
+to save the precomputation of the function body and pass it to the
+resumer version of the function.
 
@@
 
-Because world \bbtwo\ terms in \lang\ reduce to monostage residuals (as opposed to partial values),
-term splitting at world \bbtwo\ is simpler and more uniform than the version at \bbonem. 
-The algorithm is specified by the $\splittwosym$ relation in \cref{fig:termSplit}.
+Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
+(as opposed to partial values), term splitting at world
+\bbtwo\ (specified by the $\splittwosym$ relation in
+\cref{fig:termSplit}) is simpler than that of \bbonem.
 
-Consider any (except \texttt{prev}) 
-$n$-ary term $e = \scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$.
-First note that for any $\diatwo e q$, $q$ must have the form $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$,
-even under binders and for nullary terms.
-For each $k \in \{1,\ldots,n\}$, the invariant for stage \bbtwo\ splitting of $e_k$ is,
+Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
+  \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
+Note that for any $\diatwo e q$, the residual $q$ must have the form
+$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even under binders and
+for nullary terms.  For each $k \in \{1,\ldots,n\}$ splitting the
+subterm $e_k$ yields $p_k$ and $l_k.r_k$ where:
 \[
-	\splittwo {e_k} A {p_k} {l_k} {r_k} \text{ and } \reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
+	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
 \]
-To match the invariant for a split $\splittwo {\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}} A p l r$, we require,
+Correspondingly, splitting the $n$-ary $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where:
 \[
@@
 \]
-Note how the precomputations are pulled out from underneath the binder into a tuple that is always executed in the first stage,
+Note how the precomputations are pulled out from underneath the binder
+into a tuple that is always executed in the first stage. This occurs
 regardless of what branch is eventually executed in the second stage.
 
-\paragraph{Next and Prev}
+\paragraph{Next and Prev.}
 Since \texttt{next} and \texttt{prev} do nothing more than convert between 

### 2015-02-28 paper/abstract.tex

@@
 partial evaluation and meta programming, in this paper we present an
-algorithm for statically {\em splitting} programs witten in a
+algorithm for statically {\em splitting} programs written in a
 two-stage typed lambda calculus (with a $\fut$ modality denoting
@@
 structured as two dependent, unstaged programs and are evaluated using
-two distinct execution passes.  While splitting, also called pass
-separation or data specialization, has been studied in the past, prior
-techniques were either manual or applicable only to simple programs
+two distinct execution passes.  While splitting has been studied in the past
+under the names \emph{pass separation} and \emph{data specialization},
+such techniques were either manual or applicable only to simple programs
 that lack important features such as recursion and first-class
@@
 later.  Our experiments based on a prototype implementation show that
-our splitting algorithm is able perform highly non-trivial algorithmic
+our splitting algorithm is able to perform highly non-trivial algorithmic
 transformations, sometimes automatically improving efficiency
-asymptotically by nearly a linear factor.  Since splitting
+asymptotically by nearly a linear factor.  Since the splitting
 transformation takes place at compile time, these benefits are

### 2015-02-28 paper/relatedwork.tex

@@
 designing algorithms and performing compiler
-optimizations~\cite{JS86-staging}.  The idea behind frequency
-reduction is to identify computations that are performed multiple
-times and pull them ahead so that they can be performed once and used
-later as needed.  Dynamic programming, loop hoisting, and pass
-separation (presented here) are all examples of frequency reduction
-techniques.
+optimizations~\cite{JS86-staging}.
 %
@@
 examples of precomputation techniques.
+%
+The idea behind frequency
+reduction is to identify computations that are performed multiple
+times and pull them ahead so that they can be performed once and used
+later as needed.  Dynamic programming, loop hoisting, and pass
+separation (presented here) are all examples of frequency reduction
+techniques.
 
 \paragraph{Precomputation techniques.}
-Perhaps one of the most well studied example of precomputation is
+Perhaps one of the most studied examples of precomputation is
 partial evaluation.  Partial evaluation distinguishes between static
-(compile-time) and dynamic (run-time) stages. Given a program and values 
-for all static inputs (inputs available at compilation
-time), partial evaluation generates a specialized program by
+(compile-time) and dynamic (runtime) stages. Given a program and values 
+for all static inputs, partial evaluation generates a specialized program by
 performing computations that depend only on the static
@@
 information and the values of static arguments, program
-specialization specializes the original program to a partially evaluated
+specialization specializes the original program to a partially-evaluated
 one that operates on many different dynamic arguments.  This
@@
 of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism}, 
-considering for the most part first-order subsets.
+although for the most part restricted to first-order subsets.
 
@@
 the stage of computations explicit in programs~\cite{GJ91-lambda,NN92-twolevel}
-and writing ``meta programs'' that, when evaluated at a stage, yield
+and writing ``metaprograms'' that, when evaluated at a stage, yield
 a new program to be evaluated at the next stage.
@@
 logical construction of binding-time type systems by deriving a type
-system via the Curry-Howard isomorphism applied to temporal logic.
+system via the Curry-Howard correspondence applied to temporal logic.
 Davies and Pfenning proposed a new type system for staged computation
@@
 work on MetaML extended type-based techniques to a full-scale language
-by developing statically typed programming language based on ML that
+by developing a statically typed programming language based on ML that
 enables the programmer to express programs with multiple
@@
 extension to the typed lambda calculus that we use here is based on
-the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
+the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ in
 small ways to account for the specific evaluation model that we
-propose, however the key difference between our work and this prior work is
+propose; however, the key difference between our work and this prior work is
 that we instead focus on the problem of pass separation.
@@
 
-Another class of pre-computation techniques is incremental
+Another class of precomputation techniques is incremental
 computation, where a program can efficiently respond to small changes
@@
 changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
-Like other pre-computation techniques, incremental computation can be
+Like other precomputation techniques, incremental computation can be
 thought of as a staged model, where changes in inputs come over a
@@
 inputs and allows, in the general case, for all inputs to
-change. Thus, its benefits are dependent the input changes made and
+change. Thus, its benefits are dependent on the input changes made and
 techniques typically utilize significant run-time (dynamic)
@@
 \paragraph{Frequency-reduction via pass separation.}
-Perhaps the closest work to ours is the algorithm for ``data
-specialization'' given by Knoblock and Ruf~\cite{knoblock96}, which
+Perhaps the closest work to ours is the algorithm for data
+specialization given by Knoblock and Ruf~\cite{knoblock96}, which
 also seeks to statically split an explicitly staged program into two
@@
 programming systems for computer graphics, where, to achieve
-high-performance, programs are typically manually separated into
-components by frequency of execution (Separation into a pre-determined
+high performance, programs are typically manually separated into
+components by frequency of execution (Indeed, separation into a pre-determined
 set of per-frequency modules is \emph{required} by the execution model
 of modern graphics architectures.)  The software engineering
-challenges modifying multiple per-stage programs have led to
-suggestions of writing graphics programs in an explicitly staged
+challenges of modifying multiple per-stage programs have led to
+suggestions of writing graphics programs in an explicitly-staged
 programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
@@
 been limited to simple, imperative languages. We are not aware of
-prior work the presents a splitting algorithm in the context of
+prior work that presents a splitting algorithm in the context of
 higher-order functional languages.

### 2015-02-28 paper/splitting.tex
Spellcheck.

@@
 Again, this holds even for nullary terms and under binders.
-In particular, \texttt{case} splits according to the rule:
+In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
 \[
-\infersplittwo	{	\begin{array} {l}
-						\caseof{e_1}{x_2.e_2}{x_3.e_3} \splittwosym \\
-						\hspace{5em} \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}		
-					\end{array}}	
-				{\sub 1 ? & \sub 2 ? & \sub 3 ?} 								
+\pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
 \]
-Note how the precomputations are pulled out from underneath the binder
-into a tuple that is always executed in the first stage. This occurs
-regardless of what branch is eventually executed in the second stage.
+Note how the precomputations of $e_2$ and $e_3$ are pulled out from
+underneath the binder into a tuple that is always executed in the
+first stage. This occurs regardless of what branch is eventually
+executed in the second stage.
 
@@
 Since \texttt{next} and \texttt{prev} do nothing more than convert between 
-second-stage terms at \bbtwo\ and second-stage terms encapsulated within $\fut$s at \bbonem,
+second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
 their splitting rules are relatively simple.
-Given an $\splittwo e ? p l r$ and $\diatwo e q$, the splitting invariant says,
+Given $\diatwo e q$, splitting yields $p$ and $l.r$ where:
 \[
@@
 \]
-The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$,
-where $\diaone {\next {e'}} {{\hat y} \mapsto q} {\next {\hat y}}$
-and $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$
-then requires
+Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
+and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where:
 \[
@@
 \]
-which can be fulfilled using $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
+Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
 Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
@@
 
-The \texttt{partition} function in the \texttt{quickselect} example in
+The \texttt{partition} function in the quickselect example in
 \ref{sec:exampleQS} is an example of such a function that contains no

### 2015-02-28 paper/appendix.tex
added trie, which is all lies

@@
 whereas \texttt{qs2} (which is passed to \texttt{map2} via variable \texttt{f}) is executed once per query.
+
+\subsection{Trie}
+
+Our last example is one which generates a trie intermediate data structure.
+
+The function {\tt exists} below determines whether a query string (s) occurs in a given set of strings (from the alphabet $\{A, B,C\}$).
+It solves this problem by partitioning the provided set of strings into three subsets,
+one which contains all words that start with $A$,
+one which contains all words that start with $B$, and one for $C$.
+When partitioning, we also throw away the first letter, since we know it's uniform for the whole subset.
+Also while doing this partitioning, we also keep track of whether the empty word occurred at all.
+We then look at the query string.  If it's empty, we return whether the empty word was in the set.
+Otherwise, we look at the first letter, pick the appropriate subset, 
+and recursively search for the rest of our query string in that.
+
+We stage this algorithm by saying that the word set comes at stage \bbone\
+and the query string comes at stage \bbtwo.
+
+\begin{lstlisting} 
+datatype letter = A | B | C
+datatype string = EmptyS | ConsS of letter * string
+datatype list = EmptyL | ConsL of string * list
+/* helper function */
+fun partition (l : list) : (bool * list * list * list) =
+ case l of
+   EmptyL => (false,EmptyL,EmptyL,EmptyL)
+ | ConsL (s,ss) =>
+     let (anyEmpty,a,b,c) = partition ss in
+     case unroll s of
+       EmptyS => (true,a,b,c)
+     | ConsS (z,zs) =>
+         case z of 
+           A => (anyEmpty, ConsL(zs,a), b, c) 
+         | B => (anyEmpty, a, ConsL(zs,b), c) 
+         | C => (anyEmpty, a, b, ConsL(zs, c))
+fun exists ((l,s) : list * $string) : $bool =
+ case l of 
+   EmptyL => next{false} 
+ | ConsL _ => 
+   let (anyEmpty, a, b, c) = partition l in
+   next {
+     case prev{s} of
+       EmptyS => prev{
+         if anyEmpty then next{true} else next{false}
+         }
+     | ConsS (z,zs) =>
+         case z of
+           A => prev{exists (a,next{zs})}
+         | B => prev{exists (b,next{zs})} 
+         | C => prev{exists (c,next{zs})}
+   }
+\end{lstlisting}
+
+\noindent
+Running the splitting algorithm on this code produces:
+
+\begin{lstlisting} 
+datatype _ = EmptyS | ConsS of _
+datatype _ = EmptyL | ConsL of _
+fun partition l =
+ case l of
+   EmptyL => (false,EmptyL,EmptyL,EmptyL)
+ | ConsL (s,ss) =>
+     let (anyEmpty,a,b,c) = partition ss in
+     case unroll s of
+       EmptyS => (true,a,b,c)
+     | ConsS (z,zs) =>
+         case z of 
+           A => (anyEmpty, ConsL(zs,a), b, c) 
+         | B => (anyEmpty, a, ConsL(zs,b), c) 
+         | C => (anyEmpty, a, b, ConsL(zs, c))
+fun exists1 (l,s) = roll ((), 
+ case l of 
+   EmptyL => inj1 ()
+ | ConsL _ => 
+   let (anyEmpty, a, b, c) = partition l in
+   inj2 (
+   	if anyEmpty then inj1 () else inj2 (), 
+   	#2 (exists1 (a,())), 
+   	#2 (exists1 (b,())), 
+   	#2 (exists1 (c,()))
+   )
+ )
+
+datatype _ = EmptyS | ConsS of _
+fun exists2 ((l,s), p) =
+ case unroll p of 
+   inj1 () => false
+ | inj2 (anyEmpty,p_a,p_b,p_c) => 
+     case s of
+       EmptyS => 
+         case anyEmpty of 
+           inj1 () => true 
+         | inj2 () => false
+     | ConsS (z,zs) =>
+         case z of
+           A => exists2 (((),zs),p_a)
+         | B => exists2 (((),zs),p_b)
+         | C => exists2 (((),zs),p_c)
+\end{lstlisting}
+
+As promised, this split version creates a trie in the first stage, and consumes it in the second stage.

### 2015-02-28 paper/appendix.tex
Wordsmithing text in Appendix.

@@
 
-Here we include some additional examples of our splitting algorithm.
-
-\subsection{Trie}
-
-\crem{pending Nico}
+This appendix includes additional examples of the behavior of our
+splitting algorithm.
 
@@
 
-In \ref{sec:exampleQS}, we suggested that staged quickselect could be applied to one list and many queries.
-In this example, we implement that behavior by passing \texttt{qs} to a higher order function.
-Such a combinator, which operates over both stages at once, is possible because of the generality of \lang's abstractions.
-
-Note that we have to define a datatype for integer lists at the second stage, which is given by \texttt{list2}.
+In \ref{sec:exampleQS}, we showed the output of splitting quickselect
+applied to one list and $m$ queries.  Here, we generalize to mapping
+quickselect onto a list of queries. Note that we have to define a
+datatype for integer lists at the second stage, which is given by
+\texttt{list2}.
 \begin{lstlisting} 

### 2015-02-28 paper/splitting.tex

@@
 \]
-Our algorithm generates a first-pass term $c'$ that produces the same
+Our algorithm generates a first pass $c'$ that produces the same
 boundary value as splitting $e$:
@@
 of a function may itself be multistage, splitting must continue into
-it.  The first-pass term generate by split produces an $i$ that is a
-new function formed from the first-stage part of the original body.
-The resumer is a new function formed out of the second-stage part of
-the original body.  It is the responsibility of the application site
-to save the precomputation of the function body and pass it to the
+it.  The first pass generated by split produces an $i$ that is a new
+function formed from the first-stage part of the original body.  The
+resumer is a new function formed out of the second-stage part of the
+original body.  It is the responsibility of the application site to
+save the precomputation of the function body and pass it to the
 resumer version of the function.

### 2015-02-28 paper/appendix.tex
working on trie
trie input code

@@
+%!TEX root = paper.tex
+
 \section{Additional examples}
@@
 \begin{lstlisting} 
+atsignmono { 
 datatype letter = A | B | C 
@@
 datatype list   = EmptyL | ConsL of string * list
-/* helper function */
-fun partition (l : list) : (bool * list * list * list) =
- case l of
-   EmptyL => (false,EmptyL,EmptyL,EmptyL)
- | ConsL (s,ss) =>
-     let (anyEmpty,a,b,c) = partition ss in
-     case unroll s of
+
+fun partition (EmptyL) = (false,EmptyL,EmptyL,EmptyL) 
+| partition (ConsL (s,ss)) = 
+  let val (anyEmpty,a,b,c) = partition ss in 
+  case s of 
     EmptyS => (true,a,b,c) 
-     | ConsS (z,zs) =>
-         case z of 
-           A => (anyEmpty, ConsL(zs,a), b, c) 
-         | B => (anyEmpty, a, ConsL(zs,b), c) 
-         | C => (anyEmpty, a, b, ConsL(zs, c))
-fun exists ((l,s) : list * $string) : $bool =
- case l of 
-   EmptyL => next{false} 
- | ConsL _ => 
-   let (anyEmpty, a, b, c) = partition l in
-   next {
-     case prev{s} of
-       EmptyS => prev{
-         if anyEmpty then next{true} else next{false}
+  | ConsS (A,z) => (anyEmpty,ConsL(z,a),b,c) 
+  | ConsS (B,z) => (anyEmpty,a,ConsL(z,b),c) 
+  | ConsS (C,z) => (anyEmpty,a,b,ConsL(z,c)) 
 }
-     | ConsS (z,zs) =>
-         case z of
-           A => prev{exists (a,next{zs})}
-         | B => prev{exists (b,next{zs})} 
-         | C => prev{exists (c,next{zs})}
+atsignnext { 
+datatype letter2 = A2 | B2 | C2
+datatype string2 = EmptyS2 | ConsS2 of letter2 * string2
+}
+
+fun exists (mono{EmptyL},next{s}) = next{false} 
+  | exists (mono{ConsL ht},next{s}) = 
+	let atsignmono{
+	  val (anyEmpty,a,b,c) = partition (ConsL ht)
+	} in next { 
+	case s of 
+	  EmptyS2 => prev{hold mono{anyEmpty}} 
+	| ConsS2 (A2,z) => prev{exists(mono{a},next{z})} 
+	| ConsS2 (B2,z) => prev{exists(mono{b},next{z})}
+	| ConsS2 (C2,z) => prev{exists(mono{c},next{z})}  
 	} 

### 2015-02-28 paper/appendix.tex
put in the output of trie

@@
 \begin{lstlisting}
-datatype _ = EmptyS | ConsS of _
-datatype _ = EmptyL | ConsL of _
-fun partition l =
- case l of
-   EmptyL => (false,EmptyL,EmptyL,EmptyL)
- | ConsL (s,ss) =>
-     let (anyEmpty,a,b,c) = partition ss in
-     case unroll s of
+
+datatype trie = Leaf 
+			  | Branch of bool * trie * trie * trie
+
+1`datatype letter = A | B | C 
+datatype string = EmptyS | ConsS of letter * string
+datatype list   = EmptyL | ConsL of string * list
+fun partition (EmptyL) = (false,EmptyL,EmptyL,EmptyL) 
+  | partition (ConsL (s,ss)) = 
+  let val (anyEmpty,a,b,c) = partition ss in 
+  case s of 
     EmptyS => (true,a,b,c) 
-     | ConsS (z,zs) =>
-         case z of 
-           A => (anyEmpty, ConsL(zs,a), b, c) 
-         | B => (anyEmpty, a, ConsL(zs,b), c) 
-         | C => (anyEmpty, a, b, ConsL(zs, c))
-fun exists1 (l,s) = roll ((), 
- case l of 
-   EmptyL => inj1 ()
- | ConsL _ => 
-   let (anyEmpty, a, b, c) = partition l in
-   inj2 (
-   	if anyEmpty then inj1 () else inj2 (), 
-   	#2 (exists1 (a,())), 
-   	#2 (exists1 (b,())), 
-   	#2 (exists1 (c,()))
-   )
- )
+  | ConsS (A,z) => (anyEmpty,ConsL(z,a),b,c) 
+  | ConsS (B,z) => (anyEmpty,a,ConsL(z,b),c) 
+  | ConsS (C,z) => (anyEmpty,a,b,ConsL(z,c)) 
+in
+fun exists (EmptyL,()) = ((), Leaf)
+  | exists (ConsL ht,()) =
+	let 
+	  val (anyEmpty,a,b,c) = partition (ConsL ht) 
+	in ((), 
+		Branch (anyEmpty, 
+			#2 (exists (a, ())), 
+			#2 (exists (b, ())), 
+			#2 (exists (c, ()))))
 
-datatype _ = EmptyS | ConsS of _
-fun exists2 ((l,s), p) =
- case unroll p of 
-   inj1 () => false
- | inj2 (anyEmpty,p_a,p_b,p_c) => 
+`2`datatype letter2 = A2 | B2 | C2
+datatype string2 = EmptyS2 | ConsS2 of letter2 * string2
+fun exists (((),s), Leaf) = false
+  | exists (((),s), Branch (e,a,b,c)) = 
     case s of
-       EmptyS => 
-         case anyEmpty of 
-           inj1 () => true 
-         | inj2 () => false
-     | ConsS (z,zs) =>
-         case z of
-           A => exists2 (((),zs),p_a)
-         | B => exists2 (((),zs),p_b)
-         | C => exists2 (((),zs),p_c)
+	  EmptyS2 => e
+	| ConsS2 (A,z) => exists (((), z), a)
+	| ConsS2 (B,z) => exists (((), z), b)
+	| ConsS2 (C,z) => exists (((), z), c)`
+
 \end{lstlisting}

### 2015-02-28 paper/appendix.tex
popped in qsmany

@@
     | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
-  in m q`
+  in m q
+
+`1`val (qsMany1,()) = map1 qs
+`2`val qsMany2 = map2 (qs, ()) `
 \end{lstlisting}

### 2015-02-28 paper/splitting.tex

@@
 %Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
-Splitting must produce a $c$ and $l.r$ satisfying,
+Splitting must produce a $c$ and $l.r$ satisfying
 \[
@@
 \paragraph {Injections.}
-Now consider splitting the non-terminal, $\inl {e}$, where $\diaone
-{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$. Therefore:
+Now consider splitting the non-terminal $\inl {e}$, where $\diaone
+{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
 \[
@@
 Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
-splitting must produce a $c'$ and $l.r'$ satisfying,
+splitting must produce a $c'$ and $l.r'$ satisfying
 \[
@@
 Splitting the injection also generates the same resumer as splitting
-the subterm $e$: ($l.r' = l.r$).
+the subterm $e$ (that is, $l.r' = l.r$).
 
@@
 the subterms, $b = \tup{b_1,b_2}$.  The value is 
-constructed using,
+constructed using
 \[
@@
 \]
-And deconstructed in the resumer with the tuple pattern,
+and deconstructed in the resumer with a tuple pattern:
 \[
@@
 Therefore:
-\[
-	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} 
-\]
-\[
+\begin{gather*}
+	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
+  {\tup{i_2,b_2}} \\
 \text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
-\]
+\end{gather*}
 (Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
@@
 are already fully reduced in our semantics.  However, since the body
-of a function may itself be multistage, splitting must continue into
-it.  The first pass generated by split produces an $i$ that is a new
+of a function may itself be multistage, splitting must descend into
+it.  The first pass generated by splitting produces an $i$ that is a new
 function formed from the first-stage part of the original body.  The
@@
 Note that for any $\diatwo e q$, the residual $q$ must have the form
-$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even under binders and
-for nullary terms.  For each $k \in \{1,\ldots,n\}$ splitting the
-subterm $e_k$ yields $p_k$ and $l_k.r_k$ where:
+$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
+nullary or the $q_i$ have binders.
+For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
 \[
@@
 \]
-Correspondingly, splitting the $n$-ary $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where:
+Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
 \[
@@
 
-Again, this holds even for nullary terms and under binders.
+%Again, this holds even for nullary terms and under binders.
 In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
@@
 
-\paragraph{Next and Prev.}
+\paragraph{Staging constructs.}
 Since \texttt{next} and \texttt{prev} do nothing more than convert between 
@@
 their splitting rules are relatively simple.
-Given $\diatwo e q$, splitting yields $p$ and $l.r$ where:
+Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
 \[
@@
 Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
-and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where:
+and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
 \[

### 2015-02-28 paper/examples.tex
minor syntax bug

@@
 
-1`//dot : vec * vec -> $int
+1`dot : vec * vec -> $int
 fun dot (
@@
       Branch (
-        let (left,right,n) = partition (h,t) in
+        let val (left,right,n) = partition (h,t) in
         (n, #2 (qs1 left k), h, #2 (qs1 right k))))`

### 2015-02-28 paper/appendix.tex

@@
 
-Our last example is one which generates a trie intermediate data structure.
+Our last example generates a trie intermediate data structure.
 
@@
 and the query string comes at stage \bbtwo.
-
 \begin{lstlisting} 
@@
 \end{lstlisting}
-
-\noindent
 Running the splitting algorithm on this code produces:
-
 \begin{lstlisting} 
-
 datatype trie = Leaf 

### 2015-02-28 paper/appendix.tex
clean up trie exposition

@@
 
-Our last example generates a trie intermediate data structure.
+The function {\tt exists} determines whether a query string \texttt{s} occurs in
+a given list of strings (over the alphabet $\{A,B,C\}$). It starts by
+partitioning the provided set of strings into three subsets, corresponding to
+the set of strings starting with $A$ (resp., $B$ and $C$), and stripping that
+first letter off of each of those strings. While processing the input list, it
+also keeps track of whether the empty string has been encountered.
+%
+We then look at the query string \texttt{s}. If it's empty, we return whether
+the empty string was encountered. Otherwise, we choose the partition which
+agrees with the first letter of \texttt{s}, and recursively searches that list
+of strings for the remainder of \texttt{s}.
 
-The function {\tt exists} below determines whether a query string (s) occurs in a given set of strings (from the alphabet $\{A, B,C\}$).
-It solves this problem by partitioning the provided set of strings into three subsets,
-one which contains all words that start with $A$,
-one which contains all words that start with $B$, and one for $C$.
-When partitioning, we also throw away the first letter, since we know it's uniform for the whole subset.
-Also while doing this partitioning, we also keep track of whether the empty word occurred at all.
-We then look at the query string.  If it's empty, we return whether the empty word was in the set.
-Otherwise, we look at the first letter, pick the appropriate subset, 
-and recursively search for the rest of our query string in that.
-
-We stage this algorithm by saying that the word set comes at stage \bbone\
-and the query string comes at stage \bbtwo.
+Our staging annotations indicate that the list of strings are first-stage, while
+the query string is second-stage.
 \begin{lstlisting} 
-atsignmono { 
+atsigngr { 
 datatype letter = A | B | C 
@@
 
-fun exists (mono{EmptyL},next{s}) = next{false} 
-  | exists (mono{ConsL ht},next{s}) = 
-	let atsignmono{
+fun exists (gr{EmptyL},next{s}) = next{false} 
+  | exists (gr{ConsL ht},next{s}) = 
+	let atsigngr{
 	  val (anyEmpty,a,b,c) = partition (ConsL ht)
@@
 	case s of 
-	  EmptyS2 => prev{hold mono{anyEmpty}} 
-	| ConsS2 (A2,z) => prev{exists(mono{a},next{z})} 
-	| ConsS2 (B2,z) => prev{exists(mono{b},next{z})}
-	| ConsS2 (C2,z) => prev{exists(mono{c},next{z})}  
+	  EmptyS2 => prev{hold gr{anyEmpty}} 
+	| ConsS2 (A2,z) => prev{exists(gr{a},next{z})} 
+	| ConsS2 (B2,z) => prev{exists(gr{b},next{z})}
+	| ConsS2 (C2,z) => prev{exists(gr{c},next{z})}  
 	} 
@@
 
-As promised, this split version creates a trie in the first stage, and consumes it in the second stage.
+\texttt{exists1} creates a trie (or prefix tree) as the boundary data structure,
+which is traversed by \texttt{exists2}.

### 2015-02-28 paper/appendix.tex
added color annotations

@@
 \begin{lstlisting} 
-atsigngr { 
+3`atsigngr { `1`
 datatype letter = A | B | C 
@@
   | ConsS (C,z) => (anyEmpty,a,b,ConsL(z,c)) 
-}
-atsignnext { 
+`3`}`1`
+atsignnext { `2`
 datatype letter2 = A2 | B2 | C2
 datatype string2 = EmptyS2 | ConsS2 of letter2 * string2
-}
+`1`}
 
-fun exists (gr{EmptyL},next{s}) = next{false} 
-  | exists (gr{ConsL ht},next{s}) = 
-	let atsigngr{
+fun exists (`3`gr{`1`EmptyL`3`}`1`,next{`2`s`1`}) = next{`2`false`1`} 
+  | exists (`3`gr{`1`ConsL ht`3`}`1`,next{`2`s`1`}) = 
+	let `3`atsigngr{`1`
 	  val (anyEmpty,a,b,c) = partition (ConsL ht)
-	} in next { 
+	`3`}`1` in next { `2`
 	case s of 
-	  EmptyS2 => prev{hold gr{anyEmpty}} 
-	| ConsS2 (A2,z) => prev{exists(gr{a},next{z})} 
-	| ConsS2 (B2,z) => prev{exists(gr{b},next{z})}
-	| ConsS2 (C2,z) => prev{exists(gr{c},next{z})}  
-	} 
+	  EmptyS2 => prev{`1`hold `3`gr{`1`anyEmpty`3`}`2`} 
+	| ConsS2 (A2,z) => prev{`1`exists(`3`gr{`1`a`3`}`1`,next{`2`z`1`})`2`} 
+	| ConsS2 (B2,z) => prev{`1`exists(`3`gr{`1`b`3`}`1`,next{`2`z`1`})`2`}
+	| ConsS2 (C2,z) => prev{`1`exists(`3`gr{`1`c`3`}`1`,next{`2`z`1`})`2`}  
+	`1`}` 
 \end{lstlisting}
@@
 in
-fun exists (EmptyL,()) = ((), Leaf)
-  | exists (ConsL ht,()) =
+fun exists1 (EmptyL,()) = ((), Leaf)
+  | exists1 (ConsL ht,()) =
 	let 
@@
 		Branch (anyEmpty, 
-			#2 (exists (a, ())), 
-			#2 (exists (b, ())), 
-			#2 (exists (c, ()))))
+			#2 (exists1 (a, ())), 
+			#2 (exists1 (b, ())), 
+			#2 (exists1 (c, ()))))
 
 `2`datatype letter2 = A2 | B2 | C2
-datatype string2 = EmptyS2 | ConsS2 of letter2 * string2
-fun exists (((),s), Leaf) = false
-  | exists (((),s), Branch (e,a,b,c)) = 
+datatype string2 = EmptyS2 
+                 | ConsS2 of letter2 * string2
+fun exists2 (((),s), Leaf) = false
+  | exists2 (((),s), Branch (e,a,b,c)) = 
     case s of
 	  EmptyS2 => e
-	| ConsS2 (A,z) => exists (((), z), a)
-	| ConsS2 (B,z) => exists (((), z), b)
-	| ConsS2 (C,z) => exists (((), z), c)`
+	| ConsS2 (A,z) => exists2 (((), z), a)
+	| ConsS2 (B,z) => exists2 (((), z), b)
+	| ConsS2 (C,z) => exists2 (((), z), c)`
 

### 2015-02-28 paper/appendix.tex
edits on A.1

@@
 
-type qsType = ^list*$`2`int`1`->$`2`int`1`
-
-// map : qsType -> ^list * $`2`list2`1` -> $`2`list2`1`
-fun map f (l, q) = 
-next {`2`
+map : (^list * $`2`int`1` -> $`2`int`1`)
+      -> ^list * $`2`list2`1` -> $`2`list2`1`
+fun map f (l, q) = next {`2`
   let 
   fun m Empty2 = Empty2
-    | m (Cons2(h,t)) = Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
-  in m prev{`1`q`2`}
-`1`}
+    | m (Cons2(h,t)) =
+                    Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
+  in m prev{`1`q`2`}`1`}
 
-val qsMany = map qs`
+val mapqss = map qss`
 \end{lstlisting}
@@
 
-`1`fun map1 f = (fn (l,()) => ((), #2 (f (l, ()))), ())
+`1`fun map1 f = (fn (l,()) => ((),#2 (f (l,()))),())
 `2`fun map2 (f,()) ((l,q), p : tree) =
@@
 
-`1`val (qsMany1,()) = map1 qs
-`2`val qsMany2 = map2 (qs, ()) `
+`1`val (mapqss1,()) = map1 qs1
+`2`val mapqss2 = map2 (qs2, ())`
 \end{lstlisting}
-
-As desired, \texttt{qs1} (which is passed to \texttt{map1} via variable \texttt{f}) is executed only once per invocation of \texttt{map},
-whereas \texttt{qs2} (which is passed to \texttt{map2} via variable \texttt{f}) is executed once per query.
+%
+Indeed, \texttt{qs1}, which is passed to \texttt{map1}, is evaluated only in the
+first pass, whereas \texttt{qs2}, which is passed to \texttt{map2}, is evaluated
+in the second pass, once per query \texttt{k}.
 

### 2015-03-13 sematics/smallstep/makepix.py
added more rules and made a build system

@@ -1,0 +1,16 @@
+import re
+from subprocess import call
+
+f = open('bits.tex','r')
+s = f.read()
+f.close()
+
+p = re.compile (r"\\bit{(.*?)}\s*%beginbit(.*?)%endbit", re.DOTALL)
+k = 0
+for (name,bit) in p.findall (s):
+	bitfile = open ('outputs/bit.tex','w')
+	bitfile.write(bit)
+	bitfile.close()
+	call (['latex', 'presentationbits.tex'])
+	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-o', 'temp.png', 'presentationbits.dvi'])
+	call (['mv', 'temp.png', 'outputs/'+name+'.png'])

### 2015-03-14 sematics/smallstep/makepix.py
changed build system to use single file

@@ -1,16 +1,29 @@
 import re
+import bisect
 from subprocess import call
 
-f = open('bits.tex','r')
+f = open('presentationbits.tex','r')
 s = f.read()
 f.close()
 
-p = re.compile (r"\\bit{(.*?)}\s*%beginbit(.*?)%endbit", re.DOTALL)
+lits = re.finditer (r"%beginlit(.*?)%endlit", s, re.DOTALL)
+bits = re.finditer (r"%beginbit(.*?)%endbit", s, re.DOTALL)
+
+litdict = {l.start() : l.group(1) for l in lits}
+
 k = 0
-for (name,bit) in p.findall (s):
-	bitfile = open ('outputs/bit.tex','w')
-	bitfile.write(bit)
+for bit in bits:
+	name = str(k)
+	k = k + 1
+	fullfile = litdict.copy()
+	fullfile[bit.start()] = bit.group(1)
+
+
+	bitfile = open ('temp.tex','w')
+	for (loc,text) in iter(sorted(fullfile.items())):
+		bitfile.write(text)	
 	bitfile.close()
-	call (['latex', 'presentationbits.tex'])
-	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-o', 'temp.png', 'presentationbits.dvi'])
+	
+	call (['latex', 'temp.tex'])
+	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-o', 'temp.png', 'temp.dvi'])
 	call (['mv', 'temp.png', 'outputs/'+name+'.png'])

### 2015-03-14 sematics/smallstep/makepix.py
build system somewhat cleans up after itself

@@ -18,12 +18,17 @@ for bit in bits:
 	fullfile = litdict.copy()
 	fullfile[bit.start()] = bit.group(1)
 
+	tempname = 'temp'
+	temptexfile = tempname+'.tex'
+	tempdvifile = tempname+'.dvi'
+	temppngfile = tempname+'.png'
 
-	bitfile = open ('temp.tex','w')
+	bitfile = open (temptexfile,'w')
 	for (loc,text) in iter(sorted(fullfile.items())):
 		bitfile.write(text)	
 	bitfile.close()
 	
-	call (['latex', 'temp.tex'])
-	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-o', 'temp.png', 'temp.dvi'])
-	call (['mv', 'temp.png', 'outputs/'+name+'.png'])
+	call (['latex', temptexfile])
+	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-o', temppngfile, tempdvifile])
+	call (['mv', temppngfile, 'outputs/'+name+'.png'])
+	call (['rm', tempdvifile, temptexfile])

### 2015-03-16 sematics/smallstep/makepix.py
outputs now have transparent background

@@ -29,6 +29,6 @@ for bit in bits:
 	bitfile.close()
 	
 	call (['latex', temptexfile])
-	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-o', temppngfile, tempdvifile])
+	call (['dvipng', '-T', 'tight', '-D', '480', '-Q', '7', '-bg', 'Transparent', '-o', temppngfile, tempdvifile])
 	call (['mv', temppngfile, 'outputs/'+name+'.png'])
 	call (['rm', tempdvifile, temptexfile])

### 2015-05-16 paper/semantics.tex
added figure with a smallstep semantics
adding in stuff about new semantics

@@
 
-\subsection{Dynamics}
+\subsection{(Small Step) Dynamics}
+We provide a small-step semantics which evaluates \lang\ terms in two passes,
+first reducing only redexes in the first stage portions of code,
+then reducing those in second stage portions of code.
+Because first- and second-stage code are
+interleaved in \lang\ terms, realizing this notion of staged
+evaluation requires some care.
+
+The first pass of evaluation comprises three judgements: 
+steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
+where both terms are at world $w$;
+doneness $e \done$, which asserts that a term $e$ at world $w$ has no more first stage redeces;
+and the auxiliary hoisting $\lift e {e'}$, which will be covered later.
+
+One important feature of our semantics is that it reduces {\em all} first stage redexes
+in the first pass, even those that occur underneath second stage binders.
+This does not cause evaluation to ``get stuck'' however, 
+because first stage code cannot depend on the results of second stage code,
+as enforced by our type system.
+
+More formally, the type safety of our dynamics can be stated
+with the usual progress and preservation lemmas, annotated with a world:
+\\{\bf Progress.~} If $\typeswor e A$, then either $\stepwor e e'$ or $e \done$. 
+\\{\bf Preservation.~} If $\typeswor e A$ and $\stepwor e e'$, then $\typeswor e A$. 
+Here, $\Gamma$ may be any list of variable bindings at world \bbtwo,
+for those second stage variables under which we are evaluating.
+Also bear in mind that these lemmas will require minor amendments once hoisting is considered.
+
+\paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
+In the $w = \bbtwo$ case, this means that $e$ contains no first stage code at all; it is a residual.
+In the $w = \bbonep$ case, this means a $e$ comprises only tuples, injections, and base value; that is, it is a value.
+
+This second pass is in fact identical to standard monostage lambda calculus evaluation, 
+so we will not cover it here.
+
+We show that the type system of \lang\ (specifically, the opaque nature of $\fut$
+types) is sufficient to guarantee that our semantics can proceed.
+
+\subsection{(Big Step) Dynamics}
 \label{sec:stagedsemantics}

### 2015-05-16 paper/semantics.tex
work on the semantics

--- 2015-02-28 paper/semantics.tex
@@
 
-\subsection{Dynamics}
+\subsection{(Small Step) Dynamics}
+We provide a small-step semantics which evaluates \lang\ terms in two passes,
+first reducing only redexes in the first stage portions of code,
+then reducing those in second stage portions of code.
+Because first- and second-stage code are
+interleaved in \lang\ terms, realizing this notion of staged
+evaluation requires some care.
+
+[Example without hoisting.]
+[Paragraph about why we want hoisting.]
+[Gist of hoisting solution.]
+
+The first pass of evaluation comprises three judgements: 
+steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
+where both terms are at world $w$;
+doneness $e \done$, which asserts that a term $e$ at world $w$ has no more first stage redexes;
+and hoisting steps $\lift e {e'}$, which are available only at world \bbonem.
+
+One important feature of our semantics is that it reduces {\em all} first stage redexes
+in the first pass, even if they occur underneath second stage binders.
+This does not cause evaluation to ``get stuck'' however, 
+because first stage code cannot depend on the results of second stage code,
+as enforced by our type system.
+
+More formally, the type safety of our dynamics can be stated
+with the usual progress and preservation lemmas, annotated with a world:
+\\{\bf Progress.~} If $\typeswor e A$, then either 
+$\stepwor e {e'}$ or $e \done$ or $\lift e {e'}$ (hoisting available only at $w = \bbonem$).
+\\{\bf Preservation.~} For normal steps, if $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$. 
+For hoisting steps, if $\typesone e A$ and $\lift e {e'}$, then $\typesone {e'} A$.
+
+In the typing rules above, $\Gamma$ may be any list of variable bindings at world \bbtwo,
+representing the second stage variables under which we are evaluating.
+
+\paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
+In this way, it generalizes over the notions of values and residuals.
+
+\paragraph{Standard Steps.} The step rules are split into reduction rules
+which encode the elimination of redexes, 
+and substep rules which define where such a reduction can occur within a larger term.
+In our semantics, the reduction rules are restricted to worlds \bbonep\ and \bbonem,
+so that only first stage redexes are eliminated.
+Substep rules are provided for all worlds, as well as for $\next$ and $\prev$.
+Also note the extra substep rules at world \bbtwo, which allow steps under binders.
+
+\paragraph{Hoisting Steps.} 
+Once the contents of a next block have been reduced to a residual,
+they are replaced by a fresh variable and then bound to that
+variable in a let term just above the innermost containing prev.
+This is acomplished by a hoisting step,
+most of whose rules are substep rules which ``bubble up'' the
+terms to be bound.
+
+\subsection{(Big Step) Dynamics}
 \label{sec:stagedsemantics}

### 2015-05-16 paper/semantics.tex
more work on small-step semantics

@@
 \subsection{(Small Step) Dynamics}
-We provide a small-step semantics which evaluates \lang\ terms in two passes,
-first reducing only redexes in the first stage portions of code,
-then reducing those in second stage portions of code.
+We provide a small-step semantics which,
+given a \lang\ term $e$, evaluates its first stage subterms
+either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
+of until all of $e$'s first stage subterms have been evaluated away
+(when $e$ is at world \bbtwo).
 Because first- and second-stage code are
-interleaved in \lang\ terms, realizing this notion of staged
+interleaved in \lang\ terms, this first pass of
 evaluation requires some care.
 
-[Example without hoisting.]
-[Paragraph about why we want hoisting.]
-[Gist of hoisting solution.]
+Moreover, our staged semantics contains one innovation over those of 
+\cite{davies96} and \cite{taha-thesis-99}, which is the addition
+of hoisting steps.  
+To illustrate, consider the example
+\begin{lstlisting}
+prev{let x = next{2*3} in next{prev{x}+4}}
+\end{lstlisting}
+A non-staged semantics which ignores $\next$s and $\prev$s
+would proceed on this term by reducing \verb|2+3| to \verb|6| in place.
+A stage semantics does not, because it only reduces first stage redexes, and
+\verb|2+3| is at the second stage.  
+The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
+substituting \verb|next{2+3}| in for \verb|x| to yield
+\begin{lstlisting}
+prev{next{prev{next{2*3}}+4}}
+\end{lstlisting}
+This takes one step to 
+\begin{lstlisting}
+prev{next{2*3+4}}
+\end{lstlisting}
+and then one more step to
+\begin{lstlisting}
+2*3+4
+\end{lstlisting}
+which is the final entirely-second-stage result, called a {\em residual}.
+Our semantics differs from this formulation in one way.
+Consider a slightly modified example where \verb|x| appears twice:
+\begin{lstlisting}
+prev{let x = next{2*3} in next{prev{x}+prev{x}}}
+\end{lstlisting}
+A standard stage semantics would, in four steps, evaluate this to 
+\begin{lstlisting}
+2*3+2*3
+\end{lstlisting}
+Note how the multiplication is duplicated in the residual.
+We find this behavior potentially undesirable for our applications,
+so our semantics instead {\em hoists} the contents of a next statement 
+to the outermost containing $\prev$, turning
+\begin{lstlisting}
+prev{let x = next{2*3} in next{prev{x}+prev{x}}}
+\end{lstlisting}
+into
+\begin{lstlisting}
+let y = 2*3 in
+prev{let x = next{y} in next{prev{x}+prev{x}}}
+\end{lstlisting}
+This then evaluates to 
+\begin{lstlisting}
+let y = 2*3 in y+y
+\end{lstlisting}
+in four steps.  
+This hoisting step occurs as the last step in the evaluation of a $\next$.
 
-The first pass of evaluation comprises three judgements: 
+Formally, the dynamic semantics comprises three judgements: 
 steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
@@
 \paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
-In this way, it generalizes over the notions of values and residuals.
+It generalizes over the notions of values and residuals.
 

### 2015-05-16 paper/semantics.tex
more stuff in semantics

@@
 either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
-of until all of $e$'s first stage subterms have been evaluated away
+or until all of $e$'s first stage subterms have been evaluated away
 (when $e$ is at world \bbtwo).
@@
 \cite{davies96} and \cite{taha-thesis-99}, which is the addition
-of hoisting steps.  
+of {\em hoisting steps}.  
 To illustrate, consider the example
@@
 A non-staged semantics which ignores $\next$s and $\prev$s
-would proceed on this term by reducing \verb|2+3| to \verb|6| in place.
-A stage semantics does not, because it only reduces first stage redexes, and
-\verb|2+3| is at the second stage.  
+would proceed on this term by reducing \verb|2*3| to \verb|6| in place.
+A staged semantics does not, because it only reduces first stage redexes, and
+\verb|2*3| is at the second stage.  
 The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
-substituting \verb|next{2+3}| in for \verb|x| to yield
+substituting \verb|next{2*3}| in for \verb|x| to yield
 \begin{lstlisting}

### 2015-06-23 paper/semantics.tex
editted statics

@@
 
-We express two-stage programs as terms in \lang, a typed, modal lambda calculus
-with products, sums, and isorecursive types. Its type system has three distinct
-\emph{worlds} \bbonep, \bbonem, and \bbtwo, which respectively classify
-\emph{ground first-stage} computations (\bbonep); 
-first-stage computations containing a \emph{mix} of code from each stage
-(\bbonem); 
-and second-stage computations (\bbtwo).
+We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
+The type system of \lang\ (\ref{sec:statics}) classifies both the stage and type of a 
+computation, and is sufficient to show that the staged dynamics of \lang\ (\ref{sec:dynamics})
+do not get stuck.
 
 \subsection{Statics}
+\label{sec:statics}
 
@@
 The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
-$e$ has type $A$ at world $w$, in the context $\Gamma$. (Variables in the
-context are also annotated with the world at which they live.)
+$e$ has type $A$ at world $w$, in the context $\Gamma$. 
+Here, a \emph{world} is a slightly finer classification than a stage.
+Whereas there is only one world, \bbtwo, for second-stage computations,
+there are two worlds corresponding to the first stage:
+\bbonep\ for \emph{ground} first-stage computations, and 
+\bbonem\ for first-stage computations containing a \emph{mix} of code from each stage. 
+The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
+and will be discussed in \ref{sec:needGround}.
 
-All three worlds contain fixpoints and unit, product, sum, function, and
-recursive types defined in the usual fashion. For example, given two terms
-$\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form the pair
-$\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world \bbonem.
+All three worlds contain unit, product, and function types defined in the usual fashion;
+sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
+These ``standard'' features can only be constructed from subterms of the same world,
+and variables can only be used at the same world where they were introduced.
 
 Differing worlds (and hence, differing stages of computation) interact by means
-of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
-which classifies second-stage computations of type $A$. Given a term $e$ of type
-$A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. This essentially
+of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
+of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonep.
+For example, given a term $e$ of type $A$ at world \bbtwo, 
+$\next{e}$ has type $\fut A$ at \bbonem. This essentially
 encapsulates $e$ as a computation to be evaluated in the future---first-stage
@@
 a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.
-
-%This is why well-typed terms cannot have any information flow from stage
-%\bbtwo\ to stage \bbone, and what makes splitting possible.
 
@@
 \subsection{(Small Step) Dynamics}
+\label{sec:dynamics}
+
 We provide a small-step semantics which,

### 2015-06-24 paper/semantics.tex

@@
 We express two-stage programs as terms in \lang, a typed, modal lambda calculus.
-The type system of \lang\ (\ref{sec:statics}) classifies both the stage and type of a 
-computation, and is sufficient to show that the staged dynamics of \lang\ (\ref{sec:dynamics})
-do not get stuck.
+
+One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
+which is a slightly finer classification than a stage.
+Whereas there is only one world, \bbtwo, for second-stage computations,
+there are two worlds corresponding to the first stage:
+\bbonem\ for \emph{mixed} first-stage computations that may contain second stage subterms with $\next$ blocks,
+and \bbonep\ for \emph{ground} first-stage computations that may not.
+The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
+and will be discussed in \ref{sec:needGround}.
+
+The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
+Note that some production rules are parameterized by world.
+Our syntax also distinguishes values from general terms [need ``series expression'' citation!], 
+as this will be useful in defining our splitting algorithm.
+There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
+All represent computations which have no remaining work in the first stage.
 
@@
 $e$ has type $A$ at world $w$, in the context $\Gamma$.
-Here, a \emph{world} is a slightly finer classification than a stage.
-Whereas there is only one world, \bbtwo, for second-stage computations,
-there are two worlds corresponding to the first stage:
-\bbonep\ for \emph{ground} first-stage computations, and 
-\bbonem\ for first-stage computations containing a \emph{mix} of code from each stage. 
-The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
-and will be discussed in \ref{sec:needGround}.
 
@@
 
-\subsection{(Small Step) Dynamics}
+\subsection{Dynamics}
 \label{sec:dynamics}
 
-We provide a small-step semantics which,
-given a \lang\ term $e$, evaluates its first stage subterms
-either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
-or until all of $e$'s first stage subterms have been evaluated away
-(when $e$ is at world \bbtwo).
-Because first- and second-stage code are
-interleaved in \lang\ terms, this first pass of
-evaluation requires some care.
+We provide a small-step semantics which evaluates the first stage subterms
+of any \lang\ term.  
+This requires notions of values and steps that are specialized to each world.
 
-Moreover, our staged semantics contains one innovation over those of 
-\cite{davies96} and \cite{taha-thesis-99}, which is the addition
-of {\em hoisting steps}.  
-To illustrate, consider the example
+\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
+Since no reduction of stage two terms is to take place, it does nothing but search for \prev\ blocks within $e$.
+A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
+To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
+This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.
+
+\paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
+Thus, $\steppure e {e'}$ searches for redexes and reduces them in the standard way, and a ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.
+
+\paragraph{World \bbonem.}
+Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
+It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
+The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
+This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable,
+as they would be in the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
+For example, consider:
 \begin{lstlisting} 
-prev{let x = next{2*3} in next{prev{x}+4}}
+2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
 \end{lstlisting}
-A non-staged semantics which ignores $\next$s and $\prev$s
-would proceed on this term by reducing \verb|2*3| to \verb|6| in place.
-A staged semantics does not, because it only reduces first stage redexes, and
-\verb|2*3| is at the second stage.  
-The staged semantics of \cite{davies96} and \cite{taha-thesis-99} instead proceed by
-substituting \verb|next{2*3}| in for \verb|x| to yield
+We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
+expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
+to a temporary variable $\hat{y}$, and substituting that variable instead:
 \begin{lstlisting} 
-prev{next{prev{next{2*3}}+4}}
+2`let val yhat = e in prev{`1`[next{`2`yhat`1`}/x]e'`2`}`
 \end{lstlisting}
-This takes one step to 
-\begin{lstlisting}
-prev{next{2*3+4}}
-\end{lstlisting}
-and then one more step to
-\begin{lstlisting}
-2*3+4
-\end{lstlisting}
-which is the final entirely-second-stage result, called a {\em residual}.
-Our semantics differs from this formulation in one way.
-Consider a slightly modified example where \verb|x| appears twice:
-\begin{lstlisting}
-prev{let x = next{2*3} in next{prev{x}+prev{x}}}
-\end{lstlisting}
-A standard stage semantics would, in four steps, evaluate this to 
-\begin{lstlisting}
-2*3+2*3
-\end{lstlisting}
-Note how the multiplication is duplicated in the residual.
-We find this behavior potentially undesirable for our applications,
-so our semantics instead {\em hoists} the contents of a next statement 
-to the outermost containing $\prev$, turning
-\begin{lstlisting}
-prev{let x = next{2*3} in next{prev{x}+prev{x}}}
-\end{lstlisting}
-into
-\begin{lstlisting}
-let y = 2*3 in
-prev{let x = next{y} in next{prev{x}+prev{x}}}
-\end{lstlisting}
-This then evaluates to 
-\begin{lstlisting}
-let y = 2*3 in y+y
-\end{lstlisting}
-in four steps.  
-This hoisting step occurs as the last step in the evaluation of a $\next$.
+This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
+They steps are introduced when a \next\ block has contents that are a residual but not already a variable.
+In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,
+where it is reified into a {\tt let} statement.
 
-Formally, the dynamic semantics comprises three judgements: 
-steps $\stepwor e e'$, which asserts that a term $e$ performs some internal reduction to produce $e'$,
-where both terms are at world $w$;
-doneness $e \done$, which asserts that a term $e$ at world $w$ has no more first stage redexes;
-and hoisting steps $\lift e {e'}$, which are available only at world \bbonem.
+\subsection{Type Safety}
 
-One important feature of our semantics is that it reduces {\em all} first stage redexes
-in the first pass, even if they occur underneath second stage binders.
-This does not cause evaluation to ``get stuck'' however, 
-because first stage code cannot depend on the results of second stage code,
-as enforced by our type system.
-
-More formally, the type safety of our dynamics can be stated
-with the usual progress and preservation lemmas, annotated with a world:
+The statics and dynamics of \lang\ are related by a type safety theorem
+comprising the usual progress and preservation lemmas, again annotated by world:
 \\{\bf Progress.~} If $\typeswor e A$, then either 
@@
 representing the second stage variables under which we are evaluating.
-
-\paragraph{Done.} The judgement $e \done$ indicates that a term $e$ contains no first stage redexes.
-It generalizes over the notions of values and residuals.
-
-\paragraph{Standard Steps.} The step rules are split into reduction rules
-which encode the elimination of redexes, 
-and substep rules which define where such a reduction can occur within a larger term.
-In our semantics, the reduction rules are restricted to worlds \bbonep\ and \bbonem,
-so that only first stage redexes are eliminated.
-Substep rules are provided for all worlds, as well as for $\next$ and $\prev$.
-Also note the extra substep rules at world \bbtwo, which allow steps under binders.
-
-\paragraph{Hoisting Steps.} 
-Once the contents of a next block have been reduced to a residual,
-they are replaced by a fresh variable and then bound to that
-variable in a let term just above the innermost containing prev.
-This is acomplished by a hoisting step,
-most of whose rules are substep rules which ``bubble up'' the
-terms to be bound.
-
-\subsection{(Big Step) Dynamics}
-\label{sec:stagedsemantics}
-
-\input{figures/dynamics-overview}
-
-\input{figures/dynamics}
-
-\begin{abstrsyn}
-
-Although the type system of \lang\ (specifically, the opaque nature of $\fut$
-types) ensures that we can perform all first-stage computations before any
-second-stage ones, the fact that first- and second-stage computations are
-interleaved in \lang\ terms means that realizing this notion of staged
-evaluation requires some care.
-
-We evaluate \lang\ terms in two stages. The first stage of evaluation has three
-mutually recursive modes of operation, depending on the world of the term it is
-evaluating. Specifically, first-stage evaluation:
-\begin{itemize}
-\item fully reduces terms in \bbonep, as they are purely first-stage;
-\item reduces terms in \bbonem, save for any world \bbtwo\ terms they may
-contain (inside \texttt{next}s);
-\item scans terms in \bbtwo\ for any world \bbonem\ subterms (inside \texttt{prev}s), 
-and reduces those in place.
-\end{itemize}
-These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
-big-step judgments, respectively (defined in \ref{fig:diaSemantics}).
-
-First-stage evaluation ultimately produces a monostaged term, called a \emph{residual}
-(\ref{fig:values}), all of whose redexes are second-stage computations. The
-second stage of evaluation is therefore just ordinary (unstaged) evaluation of
-this residual, which we notate $\redsym$. (For simplicity, we treat residuals
-and \bbonep\ terms as living in a single language equipped with a standard
-evaluation $\redsym$, whose definition we omit for space reasons.)
-
-Notice that our type system makes it possible to evaluate terms in a staged
-fashion, but doing so still requires some machinery, since the stages are
-syntactically interleaved within an individual term. For example,
-\texttt{qss} contains a first-stage recursive call inside a second-stage
-\texttt{case}; thus, an ordinary outermost-first evaluation strategy would not
-be properly staged.
-
-\paragraph{First-stage evaluation at \bbonem\ ($\redonesym$).}
-The difficulty of evaluating terms at \bbonem\ is that they contain both terms
-at \bbonep\ (via $\curr$) and at \bbtwo\ (via $\fut$); the other evaluation
-rules are ordinary call-by-value rules. When we encounter $\pure{e}$, we
-evaluate $e$ as a monostaged term; \texttt{letg} and \texttt{caseg} simply reach
-inside $\pure$ and evaluate like \texttt{let} and \texttt{case}.
-
-$\next{e}$ is trickier, because we should not evaluate $e$ (at \bbtwo) yet.
-There are two difficulties---$e$ may contain other code at \bbonem\ (which we
-handle with $\redtwosym$, discussed below); and $\next{e}$ may be part of a
-first-stage redex, which we \emph{do} need to reduce. For example, in
-\begin{lstlisting} 
-1`(fn x : $int => e') (next{`2`e`1`})`
-\end{lstlisting}
-we must reduce this first-stage function application, but cannot reduce $e$ yet.
-We could reduce this to $[\next{e}/x]e'$, but this may potentially duplicate an
-expensive computation $e$. Instead, we choose to hoist $e$ outside, binding it
-to a temporary variable $\hat{y}$, and substituting that variable instead:
-\begin{lstlisting} 
-2`let val yhat = e in `1`[next{`2`yhat`1`}/x]e'`
-\end{lstlisting}
-then proceed by evaluating $[\next{\hat{y}}/x]e'$.
-
-We achieve this by threading through a \emph{residual table} $\xi$, which
-functions as a top-level sequence of \texttt{let}-bindings of residuals $e$ to
-variables $\hat{y}$ in an evaluated \bbonem\ term. Thus, $\redonesym$ evaluation
-yields a pair $\rtab \xi v$ of a residual table and a \emph{partial value} $v$
-containing variables bound in that table. The grammar of residual tables and
-partial values (terms fully evaluated at the first but not second stage) is
-given in \ref{fig:values}.
-
-%The input to $\redonesym$ may be open on stage \bbtwo\ variables, but the type
-%system ensures that those must occur under a $\next$. Consequently $\redonesym$
-%will never directly encounter a variable.
-
-\paragraph{First-stage evaluation at \bbtwo\ ($\redtwosym$).}
-Since terms at world \bbtwo\ represent second-stage computations, we should not
-evaluate them yet. However, we do need to evaluate any first-stage computations
-they contain (inside $\prev$), \emph{even if} those computations are guarded or
-underneath binders.
-
-For all constructs other than $\prev{e}$, $\redtwosym$ evaluation proceeds
-recursively on all subterms, yielding a monostaged residual $q$. At $\prev{e}$,
-we evaluate $e$ using $\redonesym$ to $\rtab \xi v$. Because $\colmix e{\fut A}$,
-$v$ must have the canonical form $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
-$\xi$ into a sequence of second-stage \texttt{let} bindings around $\hat{y}$,
-using the auxiliary judgment $\reify\xi q{q'}$ defined in
-\ref{fig:diaSemantics}, then return the resulting residual $q'$.
-
-\end{abstrsyn}
-
-%Note that partial values and residuals are both terms for which stage \bbone\
-%computation has completed; however, partial values are stage~\bbone\ terms,
-%while residuals are stage~\bbtwo\ terms.
-
-%As an optimization, we can include the special-case rule,
-%\begin{mathpar}
-%\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
-%\end{mathpar}
-%to avoid one-for-one variable bindings in the residual.
 

### 2015-06-24 paper/semantics.tex
working on semantics section

@@
 there are two worlds corresponding to the first stage:
-\bbonem\ for \emph{mixed} first-stage computations that may contain second stage subterms with $\next$ blocks,
-and \bbonep\ for \emph{ground} first-stage computations that may not.
-The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
+\bbonem\ for \emph{mixed} first-stage computations, which may contain second stage subterms with $\next$ blocks,
+and \bbonep\ for \emph{ground} first-stage computations, which that may not.
+The distinction between these two first-stage worlds is necessitated by our splitting algorithm
 and will be discussed in \ref{sec:needGround}.
@@
 
+\input{figures/dynamics}
+
 We provide a small-step semantics which evaluates the first stage subterms
@@
 \paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
-Since no reduction of stage two terms is to take place, it does nothing but search for \prev\ blocks within $e$.
+Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
 A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
@@
 This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
-They steps are introduced when a \next\ block has contents that are a residual but not already a variable.
+These steps are introduced when a \next\ block has contents that are a residual but not already a variable.
 In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,

### 2015-06-27 paper/semantics.tex
changes to semantics

@@
 
-\input{figures/values}
-
 We express two-stage programs as terms in \lang, a typed, modal lambda calculus.
@@
 One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
-which is a slightly finer classification than a stage.
+which includes and subsumes the concept of a stage.
+Our system has three worlds: \bbonem, \bbonep, and \bbtwo.
 Whereas there is only one world, \bbtwo, for second-stage computations,
@@
 Note that some production rules are parameterized by world.
-Our syntax also distinguishes values from general terms [need ``series expression'' citation!], 
-as this will be useful in defining our splitting algorithm.
+Our syntax also distinguishes values from general terms, 
+with an underline (``$\exv v$'') noting the explicit inclusion of values into general terms.
 There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
@@
 
-We provide a small-step semantics which evaluates the first stage subterms
-of any \lang\ term.  
-This requires notions of values and steps that are specialized to each world.
+The central tenet of a staged language
+is that first stage code should be evaluated entirely before second stage code.
+Accordingly, our dynamics operates in two passes.
+The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
+$\exv q$, and the second pass further reduces this residual to a value.
+Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
+Accordingly, this paper concerns itself only with the first pass of evaluation.
+
+Since $\coltwo e A$ may be constructed out of terms at other world, 
+our dynamics requires notions of values and steps that are specialized to each world.
 
@@
 The statics and dynamics of \lang\ are related by a type safety theorem
-comprising the usual progress and preservation lemmas, again annotated by world:
-\\{\bf Progress.~} If $\typeswor e A$, then either 
-$\stepwor e {e'}$ or $e \done$ or $\lift e {e'}$ (hoisting available only at $w = \bbonem$).
-\\{\bf Preservation.~} For normal steps, if $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$. 
-For hoisting steps, if $\typesone e A$ and $\lift e {e'}$, then $\typesone {e'} A$.
-
-In the typing rules above, $\Gamma$ may be any list of variable bindings at world \bbtwo,
-representing the second stage variables under which we are evaluating.
+comprising the usual progress and preservation lemmas, again annotated by world.
+In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
+representing the second stage binders under which we are evaluating.
+\\{\bf Progress.~} 
+\begin{itemize}
+\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
+\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
+\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
+\end{itemize}
+{\bf Preservation.~} 
+\begin{itemize}
+\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
+\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
+\end{itemize}
 
 \subsection{Evaluating Staged Programs}
-
-We consider a staged program to be a closed term $e$ at world \bbtwo. The first
-stage of evaluation is $e\mathbin{\redtwosym} q$, which reduces all code at
-\bbonem\ and \bbonep\ inside $e$. The second stage of evaluation is $q
-\mathbin{\redsym} u$, or monostaged evaluation of the residual $q$, which yields
-a value $u$.
 

### 2015-06-28 paper/splitting.tex
working on splitting

@@
 
-\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
-to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
-$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
-yields a pair of monostaged terms $\pipeS p l r$. ($p$ and $l.r$
-contain all first- and second-stage computations in $e$ respectively.)
-Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
-in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
-$v$.  Since execution of the first pass serves to generate input for
+\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$
+which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
+splitting $e$ (via the new relation~$\splittwosym$)
+yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
+contain all first- and second-stage computations in $e$ respectively.
+
+First pass evaluation reduces $p$ to $b$,
+which is plugged in for $l$ to produce $[b/l]r$.
+Correctness of splitting states that this should be equivalent 
+to the residual that would have been produced by direct evaluation.
+That is, $[b/l]r \equiv q$.
+
+Since execution of the first pass serves to generate input for
 the second pass, we say $p$ is a {\em precomputation} that produces a
@@
 \textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
-to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
-relation~$\splitonesym$) again yields the pair of monostage terms
-$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
+to $\exv v$ (via repeated $\stepsym \bbonem$), 
+splitting $e$ (via the new relation~$\splitonesym$) again 
+yields the pair of monostage terms $\pipeM c l r$. 
+Evaluating $c$ in the first pass produces the pair
 $(i,b)$, where the first element $i$ matches the first-stage results
-contained in $v$.  For example, in the simple case where $\rtab \xi v$
+contained in $v$.  For example, in the simple case where $e$
 contains entirely first-stage computation, such as:
 \begin{lstlisting}
-2`[]`1` 3`gr{`1`literalone+literaltwo`3`}`
+1` 3`gr{`1`literalone+literaltwo`3`}`
 \end{lstlisting}
-the first-stage result is the value 3.  The second element of the pair
+the first-stage result is the value 3.  The second element of the pair, $b$,
 is the boundary value used as input to the resumer in second pass
-evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
+evaluation as $[b/l]r \equiv q$, where $q$ is the same as
 result produced by second-stage computations in $e$.  For example, in

### 2015-06-30 paper/splitting.tex
working on splitting

@@
 Correctness of splitting states that this should be equivalent 
-to the residual that would have been produced by direct evaluation.
-That is, $[b/l]r \equiv q$.
-
+to the residual produced by direct evaluation, $q$.
 Since execution of the first pass serves to generate input for
@@
 {\em boundary value} ($b$) for the {\em resumer} ($l.r$).
+More precisely, we can state correctness as \ref{thm:allCorrect}:
+\begin{theorem}
+\label{thm:allCorrect}
+If $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$ and $\splitone e A p {l.r}$,
+then $\reduce p b$ and $[b/l]r \equiv q$.
+\end{theorem}
+where $\reduce e v$ indicates standard monostage reduction of the term $e$ to the value $v$,
+and $e_1 \equiv e_2$ indicates a monostage equivalence (defined later).
 
-% The process of splitting \bbtwo\ terms is given by via the new
-% relation~$\splittwosym$.
-
-%masking yields the masked value $\mval i q$ (via~$\vsplito$)
+We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
+In the base case, where $e$ is already a residual of the form $\exv q$, we know
+$\splitone {\exv q} A {\exv{\tup{}}} {\_.q}$, and so by uniqueness of splitting, 
+$p = b = \tup{}$ and $r = q$.
+It directly follows that $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
 

### 2015-07-01 paper/splitting.tex
working on splitting section

@@
 
-\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$
+Our splitting algorithm is given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$,
+which apply to \bbtwo\ terms, \bbonem\ terms, and partial values, respectively.
+Before discussing the meaning of these outputs, 
+we state a few theorems regarding the totality and uniqueness of splitting:
+
+\begin{theorem} \leavevmode
+\label{thm:total} 
+\begin{itemize} 
+\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
+\item For term $e$, if $\typesone e A$, then $\splitone e A p {l.r}$.
+\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
+\end{itemize}
+\end{theorem}
+
+\begin{theorem} \leavevmode
+\label{thm:unique}
+\begin{itemize} 
+\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, $l = l'$, and $r = r'$.
+\item If $\splitone e A p {l.r}$ and $\splitone e A {p'} {l'.r'}$, then $p = p'$, $l = l'$, and $r = r'$.
+\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
+\end{itemize}
+\end{theorem}
+
+The proofs are by induction on the typing derivation
+and simultaneous induction on the splitting derivations, respectively.
+
+\subsection{Correctness}
+
+For any $\coltwo e A$
 which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
@@
 \label{thm:allCorrect}
-If $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$ and $\splitone e A p {l.r}$,
+If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
 then $\reduce p b$ and $[b/l]r \equiv q$.
 \end{theorem}
-where $\reduce e v$ indicates standard monostage reduction of the term $e$ to the value $v$,
-and $e_1 \equiv e_2$ indicates a monostage equivalence (defined later).
+In this theorem, $\reduce e v$ indicates standard monostage reduction of the term $e$ to the value $v$,
+and $\equiv$ indicates a monostage equivalence, which we cover later.
 
@@
 In the base case, where $e$ is already a residual of the form $\exv q$, we know
-$\splitone {\exv q} A {\exv{\tup{}}} {\_.q}$, and so by uniqueness of splitting, 
+$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
 $p = b = \tup{}$ and $r = q$.
 It directly follows that $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
+In the recursive case, we have 
+$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
+To help this case, we introduce to a new lemma, whose proof will concern the rest of this section:
+\begin{lemma}
+\label{lem:twoCorrect}
+If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
+then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
+\end{lemma}
+By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
+The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
+From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
+and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed
 
-\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
-to $\exv v$ (via repeated $\stepsym \bbonem$), 
+All that remains is to prove \ref{lem:twoCorrect}.  
+We will do this by induction on the derivation of the step.
+However, since second-stage steps can depend on first-stage steps,
+we first cover the definition of splitting terms at \bbonem.
+
+For any $\colmix e A$, 
 splitting $e$ (via the new relation~$\splitonesym$) again 
@@
 Evaluating $c$ in the first pass produces the pair
-$(i,b)$, where the first element $i$ matches the first-stage results
-contained in $v$.  For example, in the simple case where $e$
+$\tup{i,b}$, where the first element $i$ represents the first stage part of the results of $e$,
+and $b$ is the boundary value.
+We can now define an analogue of \ref{lem:twoCorrect} for steps at \bbonem:
+\begin{lemma}
+\label{lem:oneCorrect}
+If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
+then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
+\end{lemma}
+Like with the \bbtwo\ version, this requires that the boundary-value-applied resumer be equivalent both sides of the step.
+But additionally, it also requires that the $i$ be exactly identical on both sides of the evaluation.
+
+...
+
+For example, in the simple case where $e$
 contains entirely first-stage computation, such as:
@@
 terms are purely first-stage.  While not necessary for ordinary staged
-term evaluation (\ref{sec:stagedsemantics}), this annotation is
+term evaluation (\ref{sec:semantics}), this annotation is
 particularly useful for splitting. Specifically, splitting $\pure e$

### 2015-07-01 paper/splitting.tex
working on splitting section

@@
 \begin{itemize} 
-\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, $l = l'$, and $r = r'$.
-\item If $\splitone e A p {l.r}$ and $\splitone e A {p'} {l'.r'}$, then $p = p'$, $l = l'$, and $r = r'$.
+\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
+\item If $\splitone e A p {l.r}$ and $\splitone e A {p'} {l'.r'}$, then $p = p'$, and $l.r = l'.r'$.
 \item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
@@
 
-\subsection{Correctness}
+\subsection{Partial Value Splitting}
+\label{sec:masking}
+
+Partial value splitting (given by the relation $\vsplito$) converts a 
+partial value $v$ into a split value $\mval i q$.
+The first part of the masked value, $i$, is a monostage value holding
+the result of all first-stage components of $v$.  The second
+component, $q$, is a residual representing all second-stage
+variable references of $v$.
+
+To provide intuition about the intent of partial value splitting, consider the
+following partial value:
+\begin{lstlisting}
+1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
+\end{lstlisting}
+To construct the value $i$ representing its first-stage components,
+splitting first redacts all second-stage (blue) parts, along with the
+surrounding \texttt{next} annotations. 
+The resulting ``holes'' in the term are replaced with
+unit values.
+%\begin{lstlisting}
+%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
+%\end{lstlisting}
+%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
+\begin{lstlisting}
+1`((), (3`gr{`1`injL 7`3`}`,()))`
+\end{lstlisting}
+Finally, partial value splitting drops \texttt{gr} annotations yielding:
+\begin{lstlisting}
+1`((), (injL 7,()))`
+\end{lstlisting}
+To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
+(replacing them with unit) and \texttt{next} annotations:
+\begin{lstlisting}
+2`(yhat, ((),yhat))`
+\end{lstlisting}
+A precise definition of the partial value splitting relation is given in
+\ref{fig:valMask}.
+
+%Since lambdas may represent multi-stage computations, masking splits
+%the body of lambdas as general world \bbonem\ terms (as described in
+%\ref{sec:split-one}), and packages the resulting terms as functions in
+%both the first- and second-stage parts of the masked value.
+
+% Masking operates by first inducting on the entries of the residual table.  
+% Being purely second-stage content, these are reified into let statements at the top of the resumer.
+% Once the table is empty, masking inducts on value itself.
+
+% Masking assigns ground values to the immediate value
+% and likewise assigns references into the residual table to the resumer.
+% In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
+% Note that the \texttt{gr} and \texttt{next} annotations are erased.
+
+% Masking distributes into tuples, injections, and rolls, since their subvalues may have content at both stages.
+% However, the tags of injections and rolls are replicated only in the immediate value, 
+% since they represent first-stage information.
+
+\subsection{Term Splitting}
 
@@
 
-First pass evaluation reduces $p$ to $b$,
-which is plugged in for $l$ to produce $[b/l]r$.
-Correctness of splitting states that this should be equivalent 
-to the residual produced by direct evaluation, $q$.
+This output is a multistage program in its own right.
+It evaluates by reducing $p$ to $b$ in the first pass,
+then plugging that result in for $l$ to produce $[b/l]r$,
+which is evaluated in the second pass.
 Since execution of the first pass serves to generate input for
@@
 {\em boundary value} ($b$) for the {\em resumer} ($l.r$).
-More precisely, we can state correctness as \ref{thm:allCorrect}:
+
+We define the correctness of splitting 
+by saying that $[b/l]r$, the \emph{applied resumer}, should be equivalent 
+to the residual produced by direct evaluation, $q$.
+More precisely:
 \begin{theorem}
@@
 \end{theorem}
-In this theorem, $\reduce e v$ indicates standard monostage reduction of the term $e$ to the value $v$,
-and $\equiv$ indicates a monostage equivalence, which we cover later.
+In this theorem, ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
+and ``$\equiv$'' indicates a monostage equivalence, which we cover later.
 
@@
 \end{lemma}
-Like with the \bbtwo\ version, this requires that the boundary-value-applied resumer be equivalent both sides of the step.
-But additionally, it also requires that the $i$ be exactly identical on both sides of the evaluation.
+Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
+But additionally, it also requires that first stages results be exactly identical.
 
+\subsection{writing from here down...}
 ...
@@
 %name {\em combined term}).
-
-\subsection{Masking}
-\label{sec:masking}
-
-%Masking separates residual tables and partial values into first-stage
-%results and second stage residuals.
-
-Masking (given by the relation $\vsplito$) converts a residual table $\xi$
-and associated partial value $v$ into a masked value $\mval i q$.
-%that explicitly separates the first stage components of 
-%result from the residual.
-%where $i$ is a monostage value encoding all
-%first-stage components of the partial value, and $q$ is a monostage
-%term encoding all of the second-stage computations of the partial
-%value.
-The first part of the masked value, $i$, is a monostage value holding
-the result of all first-stage components of $v$.  The second
-component, $q$, is a residual representing all second-stage
-computations in $\xi$ and $v$.
-
-%(A precise definition of masking is
-%given in \ref{sec:masking}.)
-
-%Masking is not a transformation performed as part of the splitting
-%algorithm.  Instead, masking defines an intuitive notion of the first-stage 
-%and second-stage ``components'' of residual tables and partial
-%values, and it is used to define the requirements of what output
-%splitting of \bbonem\ terms must produce.
-To provide intuition about the intent of masking, consider the
-following residual table and partial value:
-\begin{lstlisting}
-2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (3`gr{`1`7`3`}`, next{`2`yhat`1`}))`
-\end{lstlisting}
-To construct the value $i$ representing its first-stage components,
-masking first redacts all second-stage (blue) parts, along with the
-surrounding \texttt{next} annotations. (This redaction removes the
-entire residual table since it only represents second-stage
-computations.) The resulting ``holes'' in the term are replaced with
-unit values.
-%\begin{lstlisting}
-%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
-%\end{lstlisting}
-%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
-\begin{lstlisting}
-1`((), injL (3`gr{`1`7`3`}`,()))`
-\end{lstlisting}
-Finally, masking drops \texttt{gr} annotations yielding:
-\begin{lstlisting}
-1`((), injL (7,()))`
-\end{lstlisting}
-To construct the residual $q$ (corresponding to second-stage computations) masking first redacts all \texttt{gr} blocks
-(replacing them with unit), \texttt{next} annotations, and injection tags:
-%\begin{lstlisting}
-%2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
-%\end{lstlisting}
-%Where the \texttt{gr} blocks left a hole, we put in a unit values:
-\begin{lstlisting}
-2`[yhat |-> 1+2] (yhat, ((),yhat))`
-\end{lstlisting}
-Then the residual table is reified into \texttt{let} bindings,
-yielding:
-\begin{lstlisting}
-2`let yhat = 1+2 in (yhat, ((),yhat))`
-\end{lstlisting}
-A precise definition of the masking relation is given in
-\ref{fig:valMask}.
-
-%Since lambdas may represent multi-stage computations, masking splits
-%the body of lambdas as general world \bbonem\ terms (as described in
-%\ref{sec:split-one}), and packages the resulting terms as functions in
-%both the first- and second-stage parts of the masked value.
-
-% Masking operates by first inducting on the entries of the residual table.  
-% Being purely second-stage content, these are reified into let statements at the top of the resumer.
-% Once the table is empty, masking inducts on value itself.
-
-% Masking assigns ground values to the immediate value
-% and likewise assigns references into the residual table to the resumer.
-% In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
-% Note that the \texttt{gr} and \texttt{next} annotations are erased.
-
-% Masking distributes into tuples, injections, and rolls, since their subvalues may have content at both stages.
-% However, the tags of injections and rolls are replicated only in the immediate value, 
-% since they represent first-stage information.
 

### 2015-07-03 paper/figures/splitting-overview.tex
work on the splitting and semantics sections

@@
 
-\textbf{Splitting Terms at World \bbtwo\ ---
-$\splittwo {\coltwo e A} A p l r$}
-\begin{enumerate}
-\item[] Precomputation $p$ is a monostaged term containing the first-stage computations in $e$.
-\item[] Resumer $l.r$ is a monostaged term containing the second-stage computations in $e$.
-\end{enumerate}
+\begin{theorem} \leavevmode
+\label{thm:total} 
+\begin{itemize} 
+\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
+\item For term $e$, if $\typesone e A$, then $\splitone e A c {l.r}$.
+\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
+\end{itemize}
+\end{theorem}
 
-\textbf{Splitting Terms at World \bbonem\ ---
-$\splitone {\colmix e A} A c {l.r}$}
-\begin{enumerate}
-\item[] Combined result $c$ is a monostage term containing the first-stage computations in $e$.
-\item[] Resumer $l.r$ is a monostage term containing the second-stage computations in $e$.
-\end{enumerate}
+\begin{theorem} \leavevmode
+\label{thm:unique}
+\begin{itemize} 
+\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
+\item If $\splitone e A p {l.r}$ and $\splitone e A {c'} {l'.r'}$, then $c = c'$, and $l.r = l'.r'$.
+\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
+\end{itemize}
+\end{theorem}
 
-\textbf{Masking ---
-$\rtab \xi v \vsplito \mval i q$}
-\begin{enumerate}
-\item[] Input $\rtab \xi v$ is the residual table and partial value to be masked.
-\item[] Immediate value $i$ is a monostage value representing the first-stage components of $v$.
-\item[] Residual $q$ is a monostaged term representing the second-stage computations in $\xi$ and $v$.
-\end{enumerate}
+\begin{theorem}
+\label{thm:allCorrect}
+If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
+then $\reduce p b$ and $[b/l]r \equiv q$.
+\end{theorem}
 
-\hrule
-\vspace{1em}
+\begin{lemma}
+\label{lem:twoCorrect}
+If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
+then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
+\end{lemma}
 
-\centering
-\setlength{\unitlength}{2.8pt}
-\begin{tabular}{c|c}
-{\large \bf Splitting at World \bbonem} & 
-{\large \bf Splitting at World \bbtwo} \\
+\begin{lemma}
+\label{lem:oneCorrect}
+If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
+then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
+\end{lemma}
 
-\begin{picture} (80,54) (10,30)
+\begin{lemma}
+\label{lem:oneCorrect}
+If $e \stepsym \bbonep e'$ and $\reduce {e'} i$, then $\reduce e i$.
+\end{lemma}
 
-\thicklines
-\put(25,37){\oval(22,10)}
-\put(25,75){\oval(22,10)}
-\put(75,37){\oval(22,10)}
-\put(75,75){\oval(22,10)}
-\put(36,37){\vector(1,0){28}}
-\put(36,75){\vector(1,0){28}}
-\put(25,70){\vector(0,-1){28}}
-\put(75,70){\vector(0,-1){28}}
+% \textbf{Splitting Terms at World \bbtwo\ ---
+% $\splittwo {\coltwo e A} A p l r$}
+% \begin{enumerate}
+% \item[] Precomputation $p$ is a monostaged term containing the first-stage computations in $e$.
+% \item[] Resumer $l.r$ is a monostaged term containing the second-stage computations in $e$.
+% \end{enumerate}
 
-\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
-\put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\rtab \xi v$}}}
-\put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
-\put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\mval i q$}}}
+% \textbf{Splitting Terms at World \bbonem\ ---
+% $\splitone {\colmix e A} A c {l.r}$}
+% \begin{enumerate}
+% \item[] Combined result $c$ is a monostage term containing the first-stage computations in $e$.
+% \item[] Resumer $l.r$ is a monostage term containing the second-stage computations in $e$.
+% \end{enumerate}
 
-\put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splitonesym\right)$}}
-\put(25,33){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
-\put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\$\reduce c {\tup{i,b}}$\\$q\equiv [b/l]r$ }}
-\put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\$\diaone e \xi v$}}
+% \textbf{Masking ---
+% $\rtab \xi v \vsplito \mval i q$}
+% \begin{enumerate}
+% \item[] Input $\rtab \xi v$ is the residual table and partial value to be masked.
+% \item[] Immediate value $i$ is a monostage value representing the first-stage components of $v$.
+% \item[] Residual $q$ is a monostaged term representing the second-stage computations in $\xi$ and $v$.
+% \end{enumerate}
 
-\put(28,66){\parbox[t]{44 \unitlength}{ 
-	\textbf{Correctness Condition:} \\ 
-	Splitting followed by evaluation (top and right arrows)
-	produces the same result as 
-	\lang\ evaluation followed by masking (left and bottom arrows),
-	up to a term equivalence.
-}}
-\end{picture}
+% \hrule
+% \vspace{1em}
 
-&
-\begin{picture} (80,54) (10,30)
+% \centering
+% \setlength{\unitlength}{2.8pt}
+% \begin{tabular}{c|c}
+% {\large \bf Splitting at World \bbonem} & 
+% {\large \bf Splitting at World \bbtwo} \\
 
-\thicklines
-\put(25,75){\oval(22,10)}
-\put(75,37){\oval(22,10)}
-\put(75,75){\oval(22,10)}
-\put(25,37){\vector(1,0){39}}
-\put(36,75){\vector(1,0){28}}
-\put(25,70){\line(0,-1){33}}
-\put(75,70){\vector(0,-1){28}}
+% \begin{picture} (80,54) (10,30)
 
-\put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\coltwo e A$}}}
-\put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeS p l r$}}}
-\put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$q$}}}
+% \thicklines
+% \put(25,37){\oval(22,10)}
+% \put(25,75){\oval(22,10)}
+% \put(75,37){\oval(22,10)}
+% \put(75,75){\oval(22,10)}
+% \put(36,37){\vector(1,0){28}}
+% \put(36,75){\vector(1,0){28}}
+% \put(25,70){\vector(0,-1){28}}
+% \put(75,70){\vector(0,-1){28}}
 
-\put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splittwosym\right)$}}
-\put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\ $\reduce p b$\\$q\equiv [b/l]r$}}
-\put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\ $\diatwo e q$}}
+% \put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\colmix e A$}}}
+% \put(14,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\rtab \xi v$}}}
+% \put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeM c l r$}}}
+% \put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\mval i q$}}}
 
-\put(28,66){\parbox[t]{44 \unitlength}{ 
-	\textbf{Correctness Condition:} \\ 
-	Splitting followed by evaluation (top and right arrows)
-	produces the same result as 
-	\lang\ evaluation (left and bottom arrow),
-	up to a term equivalence.
-}}
-\end{picture}
+% \put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splitonesym\right)$}}
+% \put(25,33){\makebox[50 \unitlength]{Masking $\left(\vsplito\right)$}}
+% \put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\$\reduce c {\tup{i,b}}$\\$q\equiv [b/l]r$ }}
+% \put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\$\diaone e \xi v$}}
 
-\\
+% \put(28,66){\parbox[t]{44 \unitlength}{ 
+% 	\textbf{Correctness Condition:} \\ 
+% 	Splitting followed by evaluation (top and right arrows)
+% 	produces the same result as 
+% 	\lang\ evaluation followed by masking (left and bottom arrows),
+% 	up to a term equivalence.
+% }}
+% \end{picture}
 
-\end{tabular}
+% &
+% \begin{picture} (80,54) (10,30)
+
+% \thicklines
+% \put(25,75){\oval(22,10)}
+% \put(75,37){\oval(22,10)}
+% \put(75,75){\oval(22,10)}
+% \put(25,37){\vector(1,0){39}}
+% \put(36,75){\vector(1,0){28}}
+% \put(25,70){\line(0,-1){33}}
+% \put(75,70){\vector(0,-1){28}}
+
+% \put(14,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\coltwo e A$}}}
+% \put(64,75){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$\pipeS p l r$}}}
+% \put(64,37){\raisebox{-0.5ex}{\makebox[22 \unitlength]{$q$}}}
+
+% \put(25,78){\makebox[50 \unitlength]{Splitting $\left(\splittwosym\right)$}}
+% \put(76,55){\parbox[l]{20 \unitlength}{Separated \\ Evaluation: \\ $\reduce p b$\\$q\equiv [b/l]r$}}
+% \put(9,55){\parbox[r]{20 \unitlength}{First-Stage \\ \lang\ Eval.: \\ $\diatwo e q$}}
+
+% \put(28,66){\parbox[t]{44 \unitlength}{ 
+% 	\textbf{Correctness Condition:} \\ 
+% 	Splitting followed by evaluation (top and right arrows)
+% 	produces the same result as 
+% 	\lang\ evaluation (left and bottom arrow),
+% 	up to a term equivalence.
+% }}
+% \end{picture}
+
+% \\
+
+% \end{tabular}
 
 \end{abstrsyn}
-\caption{Overview of splitting.}
+\caption{\TODO put a real caption here and format the figure nicely}
 \label{fig:splittingSummary}

### 2015-07-03 paper/semantics.tex
work on the splitting and semantics sections

@@
 there are two worlds corresponding to the first stage:
-\bbonem\ for \emph{mixed} first-stage computations, which may contain second stage subterms with $\next$ blocks,
-and \bbonep\ for \emph{ground} first-stage computations, which that may not.
+\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
+and \bbonep\ for \emph{ground} first-stage computations, which may not.
 The distinction between these two first-stage worlds is necessitated by our splitting algorithm
@@
 Our syntax also distinguishes values from general terms, 
-with an underline (``$\exv v$'') noting the explicit inclusion of values into general terms.
+with underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
 There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
@@
 
+$\curr A$ is a type in world \bbonem\ which classifies purely-first-stage
+computations of type $A$. Given a world \bbonep\ term $e$ of type $A$,
+$\pure{e}$ has type $\curr A$ at world \bbonem. ($e$ is guaranteed not to contain
+second-stage computations because $\fut$ types are not available in world
+\bbonep.)
+
 Although products and functions are restricted to types at the same world,
@@
 computation of an integer).
-
-$\curr A$ is a type in world \bbonem\ which classifies purely-first-stage
-computations of type $A$. Given a world \bbonep\ term $e$ of type $A$,
-$\pure{e}$ has type $\curr A$ at world \bbonem. ($e$ is guaranteed not to contain
-second-stage computations because $\fut$ types are not available in world
-\bbonep.)
 
@@
 The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
-$\exv q$, and the second pass further reduces this residual to a value.
+$\exv q$, and the second pass further reduces this residual.
 Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
-Accordingly, this paper concerns itself only with the first pass of evaluation.
+Consequently, this paper is concerned only with the first pass of evaluation.
 
-Since $\coltwo e A$ may be constructed out of terms at other world, 
+Since $\coltwo e A$ may be constructed out of terms at other worlds, 
 our dynamics requires notions of values and steps that are specialized to each world.
@@
 The statics and dynamics of \lang\ are related by a type safety theorem
-comprising the usual progress and preservation lemmas, again annotated by world.
+comprising progress and preservation, again annotated by world.
 In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
 representing the second stage binders under which we are evaluating.
-\\{\bf Progress.~} 
+\begin{theorem}\leavevmode
+\label{thm:progress} 
 \begin{itemize} 
@@
 \end{itemize}
-{\bf Preservation.~} 
+\end{theorem}
+\begin{theorem}\leavevmode
+\label{thm:preservation} 
 \begin{itemize} 
@@
 \end{itemize}
+\end{theorem}
 
@@
 
-A multistage function, such as \texttt{qss} from \ref{sec:overview}, has
+The multistage function \texttt{qss} from \ref{sec:overview} has
 type $A\to \fut (B \to C)$ at \bbonem.%

### 2015-07-03 paper/splitting.tex
work on the splitting and semantics sections

@@
 
-%\input{figures/splitting-overview}
+\input{figures/splitting-overview}
 
@@
 
-Splitting statically transforms terms in \lang\ into a pair of
-monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
-and second-stage computations in the original term. Since full
-evaluation of world \bbtwo\ terms (both first-stage and second-stage
-evaluation) only produces values in stage two, the idea of splitting
-is to generate terms where evaluating $f_1$ in a first pass, and then
-evaluating $f_2$ on the result of the first pass, yields the same
-value as full evaluation of the original \lang\ program.
+% Splitting statically transforms terms in \lang\ into a pair of
+% monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
+% and second-stage computations in the original term.
 
-Similarly, when splitting \bbonem\ terms, first pass evaluation of
-$f_1$ should produce the same value as first-stage evaluation of the
-original \lang\ term.  However, complexity arises because
-\bbonem\ terms may also contain computations to be evaluated in the
-second stage. (For example, a \bbonem\ term of type $\fut \rmint$
-results in an \rmint\ value in the second stage.) Thus we also want
-the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
-the same value as fully evaluating all second-stage computations in
-the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
-which only produce values in the first stage, we only need to compare
-the result of $f_1$ with that of the \lang\ term.
+% Recall that full evaluation (both the first and second pass) of a term $\coltwo e A$ 
+% only produces its ``final result'' at the second stage.
+% In contrast, \bbonem\ terms may have final results at both stages. 
+% For example, consider the term:
+% \begin{lstlisting}
+% 1`(`3`gr{`1`4+5`3`}`1`, next{`2`6+7`1`})`
+% \end{lstlisting}
 
-Handling the asymmetry between the results of evaluating
-\bbonem\ terms (which may produce results at both the first and second
-stage) and \bbtwo\ terms (which only produce second-stage results) is
-fundamental to the structure of the splitting algorithm presented in
-this section.
+% Thus, the idea of splitting for \bbtwo\ terms
+% is to generate terms where evaluating $f_1$ in a first pass, and then
+% evaluating $f_2$ on the result of the first pass, yields the same
+% value as full evaluation of the original \lang\ program.
+
+% first pass evaluation of
+% $f_1$ should produce the same value as first-stage evaluation of the
+% original \lang\ term.  However, complexity arises because
+% \bbonem\ terms Thus we also want
+% the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
+% the same value as fully evaluating all second-stage computations in
+% the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
+% which only produce values in the first stage, we only need to compare
+% the result of $f_1$ with that of the \lang\ term.
+
+% Handling the asymmetry between the results of evaluating
+% \bbonem\ terms (which may produce results at both the first and second
+% stage) and \bbtwo\ terms (which only produce second-stage results) is
+% fundamental to the structure of the splitting algorithm presented in
+% this section.
 
@@
 Before discussing the meaning of these outputs, 
-we state a few theorems regarding the totality and uniqueness of splitting:
-
-\begin{theorem} \leavevmode
-\label{thm:total} 
-\begin{itemize} 
-\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
-\item For term $e$, if $\typesone e A$, then $\splitone e A p {l.r}$.
-\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
-\end{itemize}
-\end{theorem}
-
-\begin{theorem} \leavevmode
-\label{thm:unique}
-\begin{itemize} 
-\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
-\item If $\splitone e A p {l.r}$ and $\splitone e A {p'} {l'.r'}$, then $p = p'$, and $l.r = l'.r'$.
-\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
-\end{itemize}
-\end{theorem}
-
-The proofs are by induction on the typing derivation
+we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
+That is, each splitting relation is a total function.
+We prove these theorems by straightforward induction on the typing derivation
 and simultaneous induction on the splitting derivations, respectively.
@@
 
-\subsection{Term Splitting}
+\subsection{World \bbtwo\ Term Splitting}
 
@@
 which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
-splitting $e$ (via the new relation~$\splittwosym$)
+splitting $e$ (via $\splittwosym$)
 yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
@@
 to the residual produced by direct evaluation, $q$.
-More precisely:
-\begin{theorem}
-\label{thm:allCorrect}
-If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
-then $\reduce p b$ and $[b/l]r \equiv q$.
-\end{theorem}
-In this theorem, ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
+We state this more precisely as \ref{thm:allCorrect},
+where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
 and ``$\equiv$'' indicates a monostage equivalence, which we cover later.
@@
 $\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
-$p = b = \tup{}$ and $r = q$.
-It directly follows that $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
+$p = \tup{}$ and $r = q$.
+From here, we can directly derive $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
+
 In the recursive case, we have 
 $\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
-To help this case, we introduce to a new lemma, whose proof will concern the rest of this section:
-\begin{lemma}
-\label{lem:twoCorrect}
-If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
-then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
-\end{lemma}
+To help this case, we introduce \ref{lem:twoCorrect}, whose proof will concern the rest of this section.
 By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
@@
 
+\subsection{World \bbonem\ Term Splitting}
+
+\TODO why the added complexity for \bbonem\ terms?
+
 For any $\colmix e A$, 
-splitting $e$ (via the new relation~$\splitonesym$) again 
+splitting $e$ (via $\splitonesym$) again 
 yields the pair of monostage terms $\pipeM c l r$. 
@@
 and $b$ is the boundary value.
-We can now define an analogue of \ref{lem:twoCorrect} for steps at \bbonem:
-\begin{lemma}
-\label{lem:oneCorrect}
-If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
-then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
-\end{lemma}
+Since execution of the first pass both serves to generate input for
+the second pass, we say $c$ is a {\em combined term} that produces a
+\emph{immediate result} ($i$) and a boundary value ($b$) for the resumer ($l.r$).
+
+The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ is given in \ref{lem:oneCorrect}.
 Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
@@
 
-\subsection{writing from here down...}
-...
+\TODO talk about lifting step correctness
 
-For example, in the simple case where $e$
-contains entirely first-stage computation, such as:
-\begin{lstlisting}
-1` 3`gr{`1`literalone+literaltwo`3`}`
-\end{lstlisting}
-the first-stage result is the value 3.  The second element of the pair, $b$,
-is the boundary value used as input to the resumer in second pass
-evaluation as $[b/l]r \equiv q$, where $q$ is the same as
-result produced by second-stage computations in $e$.  For example, in
-the case where $\rtab \xi v$ contains entirely second-stage
-computation, such as:
-\begin{lstlisting}
-2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
-\end{lstlisting}
-the second-stage result is the value 5. Since in general residual
-tables and partial values describe a mixture of both first and second
-stage computations, in the subsequent subsection we define an
-operation, called {\em masking}, that separates residual tables and
-partial values into their first- and second-stage results. We then
-describe the algorithm for splitting both \bbonem\ terms
-($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
+\TODO talk about ground case
 
-%Since $c$ ensapulates computations that produce both $e$'s first-stage
-%result and also the boundary value required by the resumer, we refer
-%to it as a {\em combined term}.
+\subsection{Example cases.}
 
-%Note that since $\coltwo e A$ reduces to an entirely second-stage
-%residual, its first-stage computations only exist to generate input
-%for second-stage computations.  In contrast, since $\colmix e A$
-%reduces to a multi-stage residual table and partial value, its first
-%stage computations serve to both generate the value used as input for
-%the second stage {\em and also} produce a first-stage value.
-%Correspondingly, the precomputation $p$ resulting from splitting
-%\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
-%the resumer; but the combined term $c$ resulting from splitting
-%\bbonem\ terms reduces to a tuple containing both the first stage {\em
-%  immediate result} $i$ and the boundary value $b$ (motivating the
-%name {\em combined term}).
+\TODO do some cases of the proof, and use the opportunity to discuss why splitting does what it does
 
-\subsection{Term Splitting at \bbonem}
-\label{sec:split-one}
+% For example, in the simple case where $e$
+% contains entirely first-stage computation, such as:
+% \begin{lstlisting}
+% 1` 3`gr{`1`literalone+literaltwo`3`}`
+% \end{lstlisting}
+% the first-stage result is the value 3.  The second element of the pair, $b$,
+% is the boundary value used as input to the resumer in second pass
+% evaluation as $[b/l]r \equiv q$, where $q$ is the same as
+% result produced by second-stage computations in $e$.  For example, in
+% the case where $\rtab \xi v$ contains entirely second-stage
+% computation, such as:
+% \begin{lstlisting}
+% 2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
+% \end{lstlisting}
+% the second-stage result is the value 5. Since in general residual
+% tables and partial values describe a mixture of both first and second
+% stage computations, in the subsequent subsection we define an
+% operation, called {\em masking}, that separates residual tables and
+% partial values into their first- and second-stage results. We then
+% describe the algorithm for splitting both \bbonem\ terms
+% ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
 
-Given the definition of masking, we now describe how to split
-\bbonem\ terms. The algorithm, given in \cref{fig:termSplit}, proceeds
-recursively on the structure of~$e$. We derive a few cases here.
+% %Since $c$ ensapulates computations that produce both $e$'s first-stage
+% %result and also the boundary value required by the resumer, we refer
+% %to it as a {\em combined term}.
 
-\paragraph {Unit.} 
-As a simple first case, consider splitting the unit value, $\tup{}$
-which masks to $\mval {\tup{}} {\tup{}}$.
-%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
-Splitting must produce a $c$ and $l.r$ satisfying
-\[
-	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
-\]
-Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
-Splitting $\pure{e}$ follows in similar fashion, except the first 
-component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.
+% %Note that since $\coltwo e A$ reduces to an entirely second-stage
+% %residual, its first-stage computations only exist to generate input
+% %for second-stage computations.  In contrast, since $\colmix e A$
+% %reduces to a multi-stage residual table and partial value, its first
+% %stage computations serve to both generate the value used as input for
+% %the second stage {\em and also} produce a first-stage value.
+% %Correspondingly, the precomputation $p$ resulting from splitting
+% %\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
+% %the resumer; but the combined term $c$ resulting from splitting
+% %\bbonem\ terms reduces to a tuple containing both the first stage {\em
+% %  immediate result} $i$ and the boundary value $b$ (motivating the
+% %name {\em combined term}).
 
-\paragraph {Injections.}
-Now consider splitting the non-terminal $\inl {e}$, where $\diaone
-{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
-\[
-	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
-\]
-Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
-splitting must produce a $c'$ and $l.r'$ satisfying
-\[
-	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
-\]
-Our algorithm generates a first pass $c'$ that produces the same
-boundary value as splitting $e$:
-\[
-c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
-\]
-Splitting the injection also generates the same resumer as splitting
-the subterm $e$ (that is, $l.r' = l.r$).
+% \subsection{Term Splitting at \bbonem}
+% \label{sec:split-one}
 
-\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
-where $\diaone {e_k} {\xi_k} {v_k}$ and 
-$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
-Inductively:
-\[
-	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
-\]
-Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
-and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
-splitting generates:
-\[
-	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
-\]
-The resulting boundary value is the tuple of the boundary values of
-the subterms, $b = \tup{b_1,b_2}$.  The value is 
-constructed using
-\[
-c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
-\]
-and deconstructed in the resumer with a tuple pattern:
-\[
-l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
-\]
-\paragraph {Case.}
-Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
-Here we consider only the case where
-$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
-(The \texttt{inr} case is analogous.)
+% Given the definition of masking, we now describe how to split
+% \bbonem\ terms. The algorithm, given in \cref{fig:termSplit}, proceeds
+% recursively on the structure of~$e$. We derive a few cases here.
 
-\noindent
-Inductively,
-\[
-	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
-\]
-Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
-and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
-Therefore:
-\begin{gather*}
-	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
-  {\tup{i_2,b_2}} \\
-\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
-\end{gather*}
-(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
-When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
-of $v_1$ into $c_2$ and $l_2.r_2$.)
+% \paragraph {Unit.} 
+% As a simple first case, consider splitting the unit value, $\tup{}$
+% which masks to $\mval {\tup{}} {\tup{}}$.
+% %Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
+% Splitting must produce a $c$ and $l.r$ satisfying
+% \[
+% 	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
+% \]
+% Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
+% Splitting $\pure{e}$ follows in similar fashion, except the first 
+% component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.
 
-The entire case evaluates as
-$\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
-some $c'$ and
-$l'.r'$ such that 
-\[
-  \reduce {c'} {\tup{i_2,b'}} \text{ and }
-  [b'/l']r' \equiv q_2
-\]
-We choose
-\[
-c' =
-{\left(
-\talllet{\tup{y_1,z_1}}{c_1}{
-\tallcase{y_1}
-{x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
-{x_3.\cdots\ttrpar}
-}\right)}
-\]
-\[
-l'.r' =
-\tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
-\]
-The term $c'$ reduces to
-$\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
-$b' = \tup{b_1,\inl{b_2}}$. Therefore
-$[b'/l']r'$ reduces to
-$[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
-$[q_1/x_2][b_2/l_2]r_2$ required.
+% \paragraph {Injections.}
+% Now consider splitting the non-terminal $\inl {e}$, where $\diaone
+% {e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
+% \[
+% 	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
+% \]
+% Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
+% splitting must produce a $c'$ and $l.r'$ satisfying
+% \[
+% 	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
+% \]
+% Our algorithm generates a first pass $c'$ that produces the same
+% boundary value as splitting $e$:
+% \[
+% c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
+% \]
+% Splitting the injection also generates the same resumer as splitting
+% the subterm $e$ (that is, $l.r' = l.r$).
 
-More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
-code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
-{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
-the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
-similarly constrained by the $\texttt{inr}$ case.
+% \paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
+% where $\diaone {e_k} {\xi_k} {v_k}$ and 
+% $\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
+% Inductively:
+% \[
+% 	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
+% \]
+% Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
+% and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
+% splitting generates:
+% \[
+% 	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
+% \]
+% The resulting boundary value is the tuple of the boundary values of
+% the subterms, $b = \tup{b_1,b_2}$.  The value is 
+% constructed using
+% \[
+% c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
+% \]
+% and deconstructed in the resumer with a tuple pattern:
+% \[
+% l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
+% \]
+% \paragraph {Case.}
+% Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
+% Here we consider only the case where
+% $\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
+% (The \texttt{inr} case is analogous.)
 
-%Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
-%$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.
+% \noindent
+% Inductively,
+% \[
+% 	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
+% \]
+% Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
+% and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
+% Therefore:
+% \begin{gather*}
+% 	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
+%   {\tup{i_2,b_2}} \\
+% \text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
+% \end{gather*}
+% (Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
+% When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
+% of $v_1$ into $c_2$ and $l_2.r_2$.)
 
-\paragraph {Function.} 
-Function introduction has a $\tup{}$ boundary value, since functions
-are already fully reduced in our semantics.  However, since the body
-of a function may itself be multistage, splitting must descend into
-it.  The first pass generated by splitting produces an $i$ that is a new
-function formed from the first-stage part of the original body.  The
-resumer is a new function formed out of the second-stage part of the
-original body.  It is the responsibility of the application site to
-save the precomputation of the function body and pass it to the
-resumer version of the function.
+% The entire case evaluates as
+% $\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
+% some $c'$ and
+% $l'.r'$ such that 
+% \[
+%   \reduce {c'} {\tup{i_2,b'}} \text{ and }
+%   [b'/l']r' \equiv q_2
+% \]
+% We choose
+% \[
+% c' =
+% {\left(
+% \talllet{\tup{y_1,z_1}}{c_1}{
+% \tallcase{y_1}
+% {x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
+% {x_3.\cdots\ttrpar}
+% }\right)}
+% \]
+% \[
+% l'.r' =
+% \tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
+% \]
+% The term $c'$ reduces to
+% $\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
+% $b' = \tup{b_1,\inl{b_2}}$. Therefore
+% $[b'/l']r'$ reduces to
+% $[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
+% $[q_1/x_2][b_2/l_2]r_2$ required.
 
-\subsection{Term Splitting at \bbtwo}
+% More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
+% code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
+% {b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
+% the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
+% similarly constrained by the $\texttt{inr}$ case.
 
-Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
-(as opposed to partial values), term splitting at world
-\bbtwo\ (specified by the $\splittwosym$ relation in
-\cref{fig:termSplit}) is simpler than that of \bbonem.
+% %Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
+% %$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.
 
-Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
-  \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
-Note that for any $\diatwo e q$, the residual $q$ must have the form
-$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
-nullary or the $q_i$ have binders.
-For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
-\[
-	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
-\]
-Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
-\[
-	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
-\]
-Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
-$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
+% \paragraph {Function.} 
+% Function introduction has a $\tup{}$ boundary value, since functions
+% are already fully reduced in our semantics.  However, since the body
+% of a function may itself be multistage, splitting must descend into
+% it.  The first pass generated by splitting produces an $i$ that is a new
+% function formed from the first-stage part of the original body.  The
+% resumer is a new function formed out of the second-stage part of the
+% original body.  It is the responsibility of the application site to
+% save the precomputation of the function body and pass it to the
+% resumer version of the function.
 
-%Again, this holds even for nullary terms and under binders.
-In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
-\[
-\pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
-\]
-Note how the precomputations of $e_2$ and $e_3$ are pulled out from
-underneath the binder into a tuple that is always executed in the
-first stage. This occurs regardless of what branch is eventually
-executed in the second stage.
+% \subsection{Term Splitting at \bbtwo}
 
-\paragraph{Staging constructs.}
-Since \texttt{next} and \texttt{prev} do nothing more than convert between 
-second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
-their splitting rules are relatively simple.
-Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
-\[
-	\reduce p b \text{ and } [b/l]r \equiv q
-\]
-Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
-and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
-\[
-	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
-\]
-Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
-Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
+% Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
+% (as opposed to partial values), term splitting at world
+% \bbtwo\ (specified by the $\splittwosym$ relation in
+% \cref{fig:termSplit}) is simpler than that of \bbonem.
+
+% Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
+%   \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
+% Note that for any $\diatwo e q$, the residual $q$ must have the form
+% $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
+% nullary or the $q_i$ have binders.
+% For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
+% \[
+% 	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
+% \]
+% Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
+% \[
+% 	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
+% \]
+% Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
+% $l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
+
+% %Again, this holds even for nullary terms and under binders.
+% In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
+% \[
+% \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
+% \]
+% Note how the precomputations of $e_2$ and $e_3$ are pulled out from
+% underneath the binder into a tuple that is always executed in the
+% first stage. This occurs regardless of what branch is eventually
+% executed in the second stage.
+
+% \paragraph{Staging constructs.}
+% Since \texttt{next} and \texttt{prev} do nothing more than convert between 
+% second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
+% their splitting rules are relatively simple.
+% Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
+% \[
+% 	\reduce p b \text{ and } [b/l]r \equiv q
+% \]
+% Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
+% and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
+% \[
+% 	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
+% \]
+% Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
+% Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
 

### 2015-07-03 paper/semantics.tex
fixed up semantics section

@@
 of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
-of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonep.
+of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.
 For example, given a term $e$ of type $A$ at world \bbtwo, 
-$\next{e}$ has type $\fut A$ at \bbonem. This essentially
-encapsulates $e$ as a computation to be evaluated in the future---first-stage
-computations can shuffle it around, but not use its result.  The only way to use
-a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.
+$\next{e}$ has type $\fut A$ at \bbonem. 
+This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
+but it provides a handle (of type $\fut A$) to that eventual value which can be moved around now.
+Computations at \bbonem\ can shuffle this handle around, but not access the future result it refers to.  
+The only way to eliminate a $\fut$ wrapper is to do so from the future by using a {\tt prev}, yielding an $A$ at \bbtwo.
 
@@
 Accordingly, our dynamics operates in two passes.
-The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
-$\exv q$, and the second pass further reduces this residual.
-Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
+The first pass takes an input program $\coltwo e A$
+and reduces all of its first-stage subterms in place,
+eventually resulting in a residual $\exv q$.
+The second pass further reduces this residual.
+Since a residual $q$ is a monostage object, this second pass is just standard monostage evaluation.
 Consequently, this paper is concerned only with the first pass of evaluation.
@@
 
-\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
+\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
 Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
@@
 The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
-This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable,
-as they would be in the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
+This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
+This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
+Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future''
+ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
+We find this interpretation to be more natural for our applications.
 For example, consider:
@@
 expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
-to a temporary variable $\hat{y}$, and substituting that variable instead:
+to a temporary variable $y$, and substituting that variable instead:
 \begin{lstlisting} 
-2`let val yhat = e in prev{`1`[next{`2`yhat`1`}/x]e'`2`}`
+2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
 \end{lstlisting}
-This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
-These steps are introduced when a \next\ block has contents that are a residual but not already a variable.
-In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,
+This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
+These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
+In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
 where it is reified into a {\tt let} statement.
@@
 \subsection{Type Safety}
+
+\TODO pull out the theorems into a figure
 
@@
 The multistage function \texttt{qss} from \ref{sec:overview} has
-type $A\to \fut (B \to C)$ at \bbonem.%
+type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
 \footnote{We can rewrite \texttt{qss} in this curried form, or apply a
@@
 \begin{lstlisting}
-2`let val yhat = ...`
-2`    val qs = yhat`
+2`let val y = ...`
+2`    val qs = y`
 2`in (qs k1,...,qs km)`

### 2015-07-05 paper/figures/splitting-overview.tex
filling out explanations

@@
 
-\begin{theorem} \leavevmode
+\begin{theorem} [Splitting Totality] \leavevmode
 \label{thm:total} 
@@
 
-\begin{theorem} \leavevmode
+\begin{theorem} [Splitting Uniqueness] \leavevmode
 \label{thm:unique}
@@
 
-\begin{theorem}
+\begin{theorem} [Program Correctness]
 \label{thm:allCorrect}
@@
 
-\begin{lemma}
+\begin{lemma} [\bbtwo\ Step Correctness]
 \label{lem:twoCorrect}
@@
 
-\begin{lemma}
+\begin{lemma} [\bbonem\ Step Correctness]
 \label{lem:oneCorrect}
-If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
+If $e \stepsym \bbonem e'$, $\splitone e A c {l.r}$, $\splitone {e'} A {c'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
 then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
@@
 
-\begin{lemma}
-\label{lem:oneCorrect}
-If $e \stepsym \bbonep e'$ and $\reduce {e'} i$, then $\reduce e i$.
+\begin{lemma} [Hoisting Step Correctness]
+\label{lem:hoistCorrect}
+If $\lift e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, then $c = c'$, and $l.r \equiv l'.r'$.
 \end{lemma}
@@
 \end{abstrsyn}
-\caption{\TODO put a real caption here and format the figure nicely}
+\caption{Splitting theorems.}
 \label{fig:splittingSummary}

### 2015-07-05 paper/semantics.tex
filling out explanations

@@
 \input{figures/dynamics}
+\input{figures/dynamics-overview}
 
@@
 
-\TODO pull out the theorems into a figure
-
-The statics and dynamics of \lang\ are related by a type safety theorem
-comprising progress and preservation, again annotated by world.
-In all cases below, $\Gamma$ may be any list of variable bindings at world \bbtwo,
+The statics and dynamics of \lang\ are related by a type safety theorems
+in figure \ref{fig:typeSafety}, again annotated by world.
+In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
 representing the second stage binders under which we are evaluating.
-\begin{theorem}\leavevmode
-\label{thm:progress} 
-\begin{itemize} 
-\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
-\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
-\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
-\end{itemize}
-\end{theorem}
-\begin{theorem}\leavevmode
-\label{thm:preservation} 
-\begin{itemize} 
-\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
-\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
-\end{itemize}
-\end{theorem}
 

### 2015-07-05 paper/splitting.tex
filling out explanations

@@
 
-% Splitting statically transforms terms in \lang\ into a pair of
-% monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
-% and second-stage computations in the original term.
+The goal of a stage splitting translation is to send programs in a multistage language
+to a form where the stages are seperated at the top level.
+To be valid, such a translation must meet two requirements:
+at values it must not lose information,
+and at terms it must respect the semantics.
 
-% Recall that full evaluation (both the first and second pass) of a term $\coltwo e A$ 
-% only produces its ``final result'' at the second stage.
-% In contrast, \bbonem\ terms may have final results at both stages. 
-% For example, consider the term:
-% \begin{lstlisting}
-% 1`(`3`gr{`1`4+5`3`}`1`, next{`2`6+7`1`})`
-% \end{lstlisting}
+For a two-stage language like \lang, this means translating a multistage
+object into a pair of monostage objects
+which respectively embody the first-stage and second-stage content of the original.
+Since \lang\ has three classes of multistage object, 
+our splitting algorithm has three forms: 
+\bbtwo-term splitting, \bbonem-term splitting, and partial value splitting.
+These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
+The rules of all three judgements are given in \ref{fig:termSplit}.
 
-% Thus, the idea of splitting for \bbtwo\ terms
-% is to generate terms where evaluating $f_1$ in a first pass, and then
-% evaluating $f_2$ on the result of the first pass, yields the same
-% value as full evaluation of the original \lang\ program.
-
-% first pass evaluation of
-% $f_1$ should produce the same value as first-stage evaluation of the
-% original \lang\ term.  However, complexity arises because
-% \bbonem\ terms Thus we also want
-% the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
-% the same value as fully evaluating all second-stage computations in
-% the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
-% which only produce values in the first stage, we only need to compare
-% the result of $f_1$ with that of the \lang\ term.
-
-% Handling the asymmetry between the results of evaluating
-% \bbonem\ terms (which may produce results at both the first and second
-% stage) and \bbtwo\ terms (which only produce second-stage results) is
-% fundamental to the structure of the splitting algorithm presented in
-% this section.
-
-Our splitting algorithm is given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$,
-which apply to \bbtwo\ terms, \bbonem\ terms, and partial values, respectively.
-Before discussing the meaning of these outputs, 
+Before discussing the validity of these translations, 
 we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
@@
 partial value $v$ into a split value $\mval i q$.
-The first part of the masked value, $i$, is a monostage value holding
+The first component of the split value, the \emph{immediate value} $i$, is a monostage value holding
 the result of all first-stage components of $v$.  The second
@@
 \ref{fig:valMask}.
+Observe that partial value splitting does not lose any meaningful information (namely injection tags),
+which is the validity requirement for a translation of value forms.
 
@@
 
-\TODO why the added complexity for \bbonem\ terms?
-
 For any $\colmix e A$, 
@@
 yields the pair of monostage terms $\pipeM c l r$.
-Evaluating $c$ in the first pass produces the pair
-$\tup{i,b}$, where the first element $i$ represents the first stage part of the results of $e$,
-and $b$ is the boundary value.
-Since execution of the first pass both serves to generate input for
-the second pass, we say $c$ is a {\em combined term} that produces a
-\emph{immediate result} ($i$) and a boundary value ($b$) for the resumer ($l.r$).
 
-The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ is given in \ref{lem:oneCorrect}.
+The responsibilities of this first stage component, $c$, can be seen as a combination of the
+those of world \bbtwo\ term splitting and partial value splitting.
+Since $e$ is a term, the first stage computations within it may yet still
+affect its second stage computations,
+and so $c$ needs to produces a boundary value $b$ to be passed to the resumer ($l.r$).
+But since $e$ is a world \bbonem\ construct, it has an eventual result at the 
+first stage as well as the second, and so it must produce an immediate value $i$.
+The term $c$ meets both of these responsibilities by reducing to the tuple $\tup{i,b}$,
+and so we call it a {\em combined term}.
+
+The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
 Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
-But additionally, it also requires that first stages results be exactly identical.
+But additionally, it also requires that first stage results be exactly identical.
 
-\TODO talk about lifting step correctness
-
-\TODO talk about ground case
+The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
+Because hoisting steps are nothing but rearrangement of second stage code,
+this lemma can use the strong requirement of idnetical combined terms.
 
@@
 
-% For example, in the simple case where $e$
-% contains entirely first-stage computation, such as:
-% \begin{lstlisting}
-% 1` 3`gr{`1`literalone+literaltwo`3`}`
-% \end{lstlisting}
-% the first-stage result is the value 3.  The second element of the pair, $b$,
-% is the boundary value used as input to the resumer in second pass
-% evaluation as $[b/l]r \equiv q$, where $q$ is the same as
-% result produced by second-stage computations in $e$.  For example, in
-% the case where $\rtab \xi v$ contains entirely second-stage
-% computation, such as:
-% \begin{lstlisting}
-% 2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
-% \end{lstlisting}
-% the second-stage result is the value 5. Since in general residual
-% tables and partial values describe a mixture of both first and second
-% stage computations, in the subsequent subsection we define an
-% operation, called {\em masking}, that separates residual tables and
-% partial values into their first- and second-stage results. We then
-% describe the algorithm for splitting both \bbonem\ terms
-% ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).
-
-% %Since $c$ ensapulates computations that produce both $e$'s first-stage
-% %result and also the boundary value required by the resumer, we refer
-% %to it as a {\em combined term}.
-
-% %Note that since $\coltwo e A$ reduces to an entirely second-stage
-% %residual, its first-stage computations only exist to generate input
-% %for second-stage computations.  In contrast, since $\colmix e A$
-% %reduces to a multi-stage residual table and partial value, its first
-% %stage computations serve to both generate the value used as input for
-% %the second stage {\em and also} produce a first-stage value.
-% %Correspondingly, the precomputation $p$ resulting from splitting
-% %\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
-% %the resumer; but the combined term $c$ resulting from splitting
-% %\bbonem\ terms reduces to a tuple containing both the first stage {\em
-% %  immediate result} $i$ and the boundary value $b$ (motivating the
-% %name {\em combined term}).
+\paragraph{\bbtwo\ Compatility.}
+\paragraph{\bbone\ Compatility.}
+\paragraph{Functions.}
+\paragraph{Next and prev.}
+\paragraph{Ground and Caseg.}
 

### 2015-07-06 paper/splitting.tex
added proof cases

@@
 partial value $v$ into a split value $\mval i q$.
-The first component of the split value, the \emph{immediate value} $i$, is a monostage value holding
+The first component, the \emph{immediate value} $i$, is a monostage value holding
 the result of all first-stage components of $v$.  The second
@@
 
-\TODO do some cases of the proof, and use the opportunity to discuss why splitting does what it does
+In this section, we prove the \ref{lem:oneCorrect} and \ref{lem:twoCorrect} lemmas
+by considering a few exemplar cases.
 
-\paragraph{\bbtwo\ Compatility.}
-\paragraph{\bbone\ Compatility.}
-\paragraph{Functions.}
-\paragraph{Next and prev.}
-\paragraph{Ground and Caseg.}
+\paragraph{Reduction of {\tt pi1}.}
+Define $E = \pio {\exv{\tup {v_1,v_2}}}$ and  
+$C = \letin{\tup{y,z}}{\exv{\tup{\tup {i_1,i_2},\tup{}}}}{\tup{\pio {\exv y},\exv z}}$.
+We are given $\stepmix  E {\exv {v_1}}$,
+$\splitone E A C {l.\pio {\exv {\tup{q_1,q_2}}}}$, and
+$\splitone {\exv {v_1}} A {\exv {\tup{i_1,\tup{}}}} {\_.\exv{q_1}}$,
+and we need to show $\reduce C {\tup {i_1,\tup{}}}$
+and $\pio {\exv {\tup{q_1,q_2}}} \equiv q_1$.
+Both of these can be derived directly.
+This pattern, where the outputs can be directly derived, extends to the {\tt pi2} and {\tt prev} reduction rules and all value marking rules.
+
+\paragraph{Reduction of Application.} Let $E = \app {\exv{\fix f x e}} {\exv v}$,
+$E' = [\fix fxe, v/f,x] e$, \\ 
+$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$, \\
+$Q = \fix f {\tup{x,\roll l}} r$, and \\
+$C = \left[
+\talllet{\tup{y_1,z_1}}{\exv {\tup {I,\tup{}}}}{
+\talllet{\tup{y_2,z_2}} {\exv {\tup {i_1,\tup{}}}}{
+\talllet{\tup{y_3,z_3}}{\app{\exv {y_1}}{\exv {y_2}}}{\exv{\tup{y_3,\tup{z_1,z_2,z_3}}}\ttrpar\ttrpar\ttrpar}}}\right]$.
+\\
+We are given $\stepmix E {E'}$,
+$\splitone E A C {\tup{\_,\_,l}.\app{\exv Q}{\exv{\tup{q_1,l}}}}$,
+$\splitone {E'} A {[I,i_1/f,x]c} {l.[Q,q_1/f,x]r}$,
+and $\reduce {[I,i_1/f,x]c} {\tup{i,b}}$.
+From this, we can directly derive $\reduce C {\tup{i,\tup {\tup{},\tup{},\roll b}}}$
+and $\app{\exv Q}{\exv{\tup{q_1,\roll b}}} \equiv [Q,q_1,b/f,x,l] r$.
+This pattern applies to all of the other reduction rules involving substitution,
+namely those for {\tt caseg}, and {\tt letg}. 
+Observe that, in our formal version of the language, 
+{\tt caseg} is the only construct where information is added to the boundary value.
+
+\paragraph{Compatility of {\tt pi1} at \bbtwo.}
+We are given 
+$\steptwo {\pio e} {\pio {e'}}$, 
+$\splittwo {\pio e} A p l {\pio r}$, 
+$\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
+and $\reduce {p'} {b'}$.
+Inversion of the first three yields
+$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
+Using \ref{lem:twoCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
+From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
+This pattern generalizes to all world \bbtwo\ compatibility rules.
+
+\paragraph{Compatility of {\tt pi1} at \bbonem.}
+We are given
+$\stepmix {\pio e} {\pio {e'}}$, 
+$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
+\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
+and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
+Inversion of the first three yields 
+$\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
+and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
+Using \ref{lem:oneCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
+From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 
+$[b/l]\pio r \equiv [b'/l']\pio {r}$ directly.
+This pattern generalizes to all world \bbonem\ compatibility rules.
+
+\paragraph{Compatibility of {\tt gr}.}
+We are given 
+$\stepmix {\pure e} {\pure {e'}}$, \\
+$\splitone {\pure e} A {\tup {e,\exv {\tup{}}}} {\_.\exv{\tup{}}}$, 
+$\splitone {\pure {e'}} A {\tup {e',\exv {\tup{}}}} {\_.\exv{\tup{}}}$,
+and $\reduce {\tup {e',\exv {\tup{}}}} {\tup{i,\tup{}}}$.
+Inversion of the step and reduction yield \mbox{$\steppure e {e'}$} and $\reduce {e'} i$.
+As a simple property of monostage reduction (which $\stepsym \bbonep$ is), we know $\reduce e i$.
+From here, we can derive $\reduce {\tup {e,\exv {\tup{}}}} {\tup{i,\tup{}}}$ and $\exv{\tup {}} \equiv \exv{\tup {}}$ directly.
 
@@
 
-\input{figures/splitting-misc}
 \input{figures/splitting-term}

### 2015-07-06 paper/semantics.tex
various small edits

@@
 
-One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
-which includes and subsumes the concept of a stage.
-Our system has three worlds: \bbonem, \bbonep, and \bbtwo.
+Although \lang\ describes computations that occur in two stages,
+we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
+A world is essentially a slightly finer classification than a stage.
 Whereas there is only one world, \bbtwo, for second-stage computations,
@@
 and \bbonep\ for \emph{ground} first-stage computations, which may not.
-The distinction between these two first-stage worlds is necessitated by our splitting algorithm
+The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
 and will be discussed in \ref{sec:needGround}.
@@
 Our syntax also distinguishes values from general terms, 
-with underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
+with the underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
 There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
@@
 Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
-It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
+It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
 The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
@@
 This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
-Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future''
+Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
 ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
-We find this interpretation to be more natural for our applications.
-For example, consider:
+This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
+One consequence of the stronger requirement on partial values is that we will need a new kind of step to put terms into that form.
+To illustrate, consider:
 \begin{lstlisting}
@@
 
-The multistage function \texttt{qss} from \ref{sec:overview} has
+The multistage function \texttt{qss} from \ref{sec:overview} has a
 type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
@@
 \end{lstlisting}
-where $\hat{y}$ is bound to a function containing no first-stage redexes.
+where $y$ is bound to a function containing no first-stage redexes.

### 2015-07-06 paper/splitting.tex
various small edits

@@
 For a two-stage language like \lang, this means translating a multistage
-object into a pair of monostage objects
+term into a pair of monostage terms
 which respectively embody the first-stage and second-stage content of the original.
-Since \lang\ has three classes of multistage object, 
+Since \lang\ has three classes of multistage term, 
 our splitting algorithm has three forms: 
@@
 These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
-The rules of all three judgements are given in \ref{fig:termSplit}.
+The rules of all three judgements are given in \ref{fig:termSplit}, 
+and they operate locally on the structure of the term.
+Note that there's no need to provide forms of splitting for ground terms or residuals,
+since those classes of term are already monostage by construction.
 
@@
 A precise definition of the partial value splitting relation is given in
-\ref{fig:valMask}.
+\ref{fig:termSplit}.
 Observe that partial value splitting does not lose any meaningful information (namely injection tags),
@@
 This output is a multistage program in its own right.
-It evaluates by reducing $p$ to $b$ in the first pass,
+It evaluates by reducing $p$ to the value $b$ in the first pass,
 then plugging that result in for $l$ to produce $[b/l]r$,
@@
 
-We define the correctness of splitting 
-by saying that $[b/l]r$, the \emph{applied resumer}, should be equivalent 
+The validity of the splitting translation requires that this simple 
+evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
+That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
 to the residual produced by direct evaluation, $q$.
-We state this more precisely as \ref{thm:allCorrect},
+This condition, called \emph{correctness},
+is stated more precisely as \ref{thm:allCorrect},
 where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
@@
 
-All that remains is to prove \ref{lem:twoCorrect}.  
-We will do this by induction on the derivation of the step.
+We prove \ref{lem:twoCorrect} by induction on the derivation of the step.
 However, since second-stage steps can depend on first-stage steps,
@@
 The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
-Like with the \bbtwo\ version, this requires that the applied resumer be equivalent both sides of the step.
+As in the world \bbtwo\ version, correctness requires that the applied resumer from either side of the step be equivalent.
 But additionally, it also requires that first stage results be exactly identical.
@@
 Because hoisting steps are nothing but rearrangement of second stage code,
-this lemma can use the strong requirement of idnetical combined terms.
+this lemma can use the strong requirement of identical combined terms.
 
@@
 
-In this section, we prove the \ref{lem:oneCorrect} and \ref{lem:twoCorrect} lemmas
-by considering a few exemplar cases.
+In this section, we consider a few exemplar cases 
+from the proofs of \ref{lem:oneCorrect} and \ref{lem:twoCorrect}.
 
@@
 
-\paragraph{Compatility of {\tt pi1} at \bbtwo.}
+\paragraph{Compatibility of {\tt pi1} at \bbtwo.}
 We are given 
@@
 
-\paragraph{Compatility of {\tt pi1} at \bbonem.}
+\paragraph{Compatibility of {\tt pi1} at \bbonem.}
 We are given
@@
 unnecessary logic.  For example, the rule for splitting
-\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
+\bbonem\ \texttt{caseg} terms inserts the tag from the \texttt{caseg}
 argument into the boundary value, then decodes this tag in the
 resumer. This logic occurs regardless of whether the terms forming the
-branches of the \texttt{case} contain second-stage computations.
-Worse, if this \texttt{case} appeared in the body of a recursive
+branches of the \texttt{caseg} contain second-stage computations.
+Worse, if this \texttt{caseg} appeared in the body of a recursive
 function with no other second-stage computations, splitting would
@@
 term evaluation (\ref{sec:semantics}), this annotation is
-particularly useful for splitting. Specifically, splitting $\pure e$
+particularly useful for splitting, since splitting $\pure e$
 terms trivially yields a unit resumer.
@@
 
+\TODO consider expanding, based on comments from reviews
+
 \end{abstrsyn}

### 2015-07-06 paper/semantics.tex
a few more edits to the semantics

@@
 All three worlds contain unit, product, and function types defined in the usual fashion;
-sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
+sum and recursive types are provided for worlds \bbonep\ and \bbtwo\ 
+(we omit them from \bbonem, because they are not needed for our examples).
 These ``standard'' features can only be constructed from subterms of the same world,
 and variables can only be used at the same world where they were introduced.
-
-Differing worlds (and hence, differing stages of computation) interact by means
+Thus differing worlds (and hence, differing stages of computation) only interact by means
 of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
 of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.
-For example, given a term $e$ of type $A$ at world \bbtwo, 
-$\next{e}$ has type $\fut A$ at \bbonem. 
+
+At the term level, {\tt next} blocks can be used to form future computations:
+given a term $e$ of type $A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. 
 This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
-but it provides a handle (of type $\fut A$) to that eventual value which can be moved around now.
-Computations at \bbonem\ can shuffle this handle around, but not access the future result it refers to.  
-The only way to eliminate a $\fut$ wrapper is to do so from the future by using a {\tt prev}, yielding an $A$ at \bbtwo.
+and it provides a handle (of type $\fut A$) now to that eventual value.
+Computations at \bbonem\ can shuffle this handle around as a value, but the future result it refers to cannot be accessed.
+This is because the only way to eliminate a $\fut$ wrapper is by using a {\tt prev}, which yields an $A$ at \bbtwo.
+This feature was adapted from linear temporal logic, via \cite{davies96}.
 
@@
 
+An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
+using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
+$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
+p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
+projection, of type $\curr A$.
+This elimination form, in contrast to that of $\fut$, does not permit world \bbonem\ subterms
+within any world \bbonep\ term.
+
+Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
+$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
+open on $\curr{A}$ and $\curr{B}$ respectively.
+Without this, there would be no way to get information out from under a $\curr$ wrapper.
+
 Although products and functions are restricted to types at the same world,
@@
 computation of an integer).
-
-An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
-using the $\letp xe{e'}$ construct, which binds $\colpure xA$ in a \bbonem\ term
-$e'$. This allows us to compute under $\curr$---for example, given a $\colmix
-p{\curr(A\times B)}$, the term $\letp{x}{p}{\pure{\pio x}}$ computes its first
-projection, of type $\curr A$.
-
-Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
-$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
-open on $\curr{A}$ and $\curr{B}$ respectively.
 
@@
 The first pass takes an input program $\coltwo e A$
-and reduces all of its first-stage subterms in place,
+and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
 eventually resulting in a residual $\exv q$.
 The second pass further reduces this residual.
-Since a residual $q$ is a monostage object, this second pass is just standard monostage evaluation.
-Consequently, this paper is concerned only with the first pass of evaluation.
+Since a residual $q$ is monostage by definition, this second pass is just standard unstaged evaluation.
+Consequently, this paper does not describe in any detail the second pass of evaluation.
 
@@
 \paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
-Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
+Since first pass evaluation should not reduce stage two terms, this judgement does nothing but search for \prev\ blocks within $e$.
 A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 

### 2015-07-07 paper/semantics.tex
adding more structure to splitting section

@@
 Consequently, this paper does not describe in any detail the second pass of evaluation.
+Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.
 
@@
 \paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
-Since first pass evaluation should not reduce stage two terms, this judgement does nothing but search for \prev\ blocks within $e$.
+Since first pass evaluation should not reduce stage two terms, this judgement does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
 A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
@@
 \paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
-Thus, $\steppure e {e'}$ searches for redexes and reduces them in the standard way, and a ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.
+Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code (compatibility rules) 
+where it either performs a beta reduction (reduction rules) or marks the term as being evaluated (value marking rules). 
+A ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.
 
 \paragraph{World \bbonem.}
-Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
+Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
 It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
@@
 where it is reified into a {\tt let} statement.
+\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
+Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}
 

### 2015-07-07 paper/splitting.tex
adding more structure to splitting section

@@
 
-The goal of a stage splitting translation is to send programs in a multistage language
-to a form where the stages are seperated at the top level.
-To be valid, such a translation must meet two requirements:
-at values it must not lose information,
-and at terms it must respect the semantics.
+The goal of a stage splitting translation 
+is to a send program $\mathcal P$ in a multistage language
+to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
+By equivalent, we mean that $\mathcal P$ and $\mathcal P'$ should produce 
+the same answer under their respective semantics.
 
-For a two-stage language like \lang, this means translating a multistage
-term into a pair of monostage terms
-which respectively embody the first-stage and second-stage content of the original.
 Since \lang\ has three classes of multistage term, 
 our splitting algorithm has three forms: 
-\bbtwo-term splitting, \bbonem-term splitting, and partial value splitting.
-These are given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$.
-The rules of all three judgements are given in \ref{fig:termSplit}, 
-and they operate locally on the structure of the term.
+$\splittwo e A p l r$ for \bbtwo-terms, 
+$\splitone e A c {l.r}$ for \bbonem-terms, and 
+$v \vsplito \mval i q$ for partial values.
+In each form the output has two parts, corresponding to the
+first-stage ($p$, $c$, and $i$) and second-stage ($l.r$, and $q$) content of the input.
 Note that there's no need to provide forms of splitting for ground terms or residuals,
@@
 
-Before discussing the validity of these translations, 
+The rules of the three splitting judgements are given in \ref{fig:termSplit}.
+They are simply recursive on structure of the term, 
+implying that the splitting algorithm runs in linear time.
+
+Before discussing the correctness of these translations, 
 we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
-That is, each splitting relation is a total function.
+That is, each splitting judgement defines a total function.
 We prove these theorems by straightforward induction on the typing derivation
 and simultaneous induction on the splitting derivations, respectively.
+
+\subsection{World \bbtwo\ Term Splitting}
+
+Top level programs in \lang\ are given by a term $\coltwo e A$.
+Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
+We write this as:
+$
+e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q
+$.
+
+Splitting $e$ (via $\splittwosym$)
+yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
+contain all first- and second-stage computations in $e$ respectively.
+This output is a multistage program in its own right.
+It evaluates by reducing $p$ to the value $b$ in the first pass,
+then plugging that result in for $l$ to produce $[b/l]r$,
+which is evaluated in the second pass.
+Since execution of the first pass serves to generate input for
+the second pass, we say $p$ is a {\em precomputation} that produces a
+{\em boundary value} ($b$) for the {\em resumer} ($l.r$).
+
+The validity of the splitting translation requires that this simple 
+evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
+That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
+to the residual produced by direct evaluation, $q$.
+This condition, called \emph{correctness},
+is stated more precisely as \ref{thm:allCorrect},
+where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
+and ``$\equiv$'' indicates a monostage equivalence, which we cover later.
+
+We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
+In the base case, where $e$ is already a residual of the form $\exv q$, we know
+$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
+$p = \tup{}$ and $r = q$.
+From here, we can directly derive $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
+
+In the recursive case, we have 
+$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
+To help this case, we introduce \ref{lem:twoCorrect}, whose proof will concern the rest of this section.
+By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
+The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
+From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
+and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed
+
+We prove \ref{lem:twoCorrect} by induction on the derivation of the step.
+However, since second-stage steps can depend on first-stage steps,
+we first cover the definition of splitting terms at \bbonem.
 
@@
 % since they represent first-stage information.
-
-\subsection{World \bbtwo\ Term Splitting}
-
-For any $\coltwo e A$
-which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
-splitting $e$ (via $\splittwosym$)
-yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
-contain all first- and second-stage computations in $e$ respectively.
-
-This output is a multistage program in its own right.
-It evaluates by reducing $p$ to the value $b$ in the first pass,
-then plugging that result in for $l$ to produce $[b/l]r$,
-which is evaluated in the second pass.
-Since execution of the first pass serves to generate input for
-the second pass, we say $p$ is a {\em precomputation} that produces a
-{\em boundary value} ($b$) for the {\em resumer} ($l.r$).
-
-The validity of the splitting translation requires that this simple 
-evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
-That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
-to the residual produced by direct evaluation, $q$.
-This condition, called \emph{correctness},
-is stated more precisely as \ref{thm:allCorrect},
-where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
-and ``$\equiv$'' indicates a monostage equivalence, which we cover later.
-
-We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
-In the base case, where $e$ is already a residual of the form $\exv q$, we know
-$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
-$p = \tup{}$ and $r = q$.
-From here, we can directly derive $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
-
-In the recursive case, we have 
-$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
-To help this case, we introduce \ref{lem:twoCorrect}, whose proof will concern the rest of this section.
-By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
-The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
-From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
-and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed
-
-We prove \ref{lem:twoCorrect} by induction on the derivation of the step.
-However, since second-stage steps can depend on first-stage steps,
-we first cover the definition of splitting terms at \bbonem.
 

### 2015-07-08 paper/splitting.tex
fleshed out splitting section

@@
 Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
-We write this as:
-$
-e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q
-$.
+We write this as $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$.
 
-Splitting $e$ (via $\splittwosym$)
-yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
-contain all first- and second-stage computations in $e$ respectively.
+Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
 This output is a multistage program in its own right.
-It evaluates by reducing $p$ to the value $b$ in the first pass,
+It is evaluated by reducing $p$ to the value $b$ in the first pass,
 then plugging that result in for $l$ to produce $[b/l]r$,
@@
 
-The validity of the splitting translation requires that this simple 
+The correctness of the splitting translation requires that this simple 
 evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
@@
 to the residual produced by direct evaluation, $q$.
-This condition, called \emph{correctness},
-is stated more precisely as \ref{thm:allCorrect},
+This condition is stated more precisely as \ref{thm:allCorrect},
 where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
@@
 
-In the recursive case, we have 
-$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
-To help this case, we introduce \ref{lem:twoCorrect}, whose proof will concern the rest of this section.
+In the recursive case, where the evaluation takes at least one step, we have 
+$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
+as well as \mbox{$\typestwo [\cdot] e A$} 
+and \mbox{$\splittwo e A p l r$}.
+To help this case, we introduce \ref{lem:twoCorrect}, 
+which essentially states that any single step is correct,
+and whose proof will concern the rest of this section.
 By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
@@
 From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
-and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed
+and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
 
-We prove \ref{lem:twoCorrect} by induction on the derivation of the step.
-However, since second-stage steps can depend on first-stage steps,
-we first cover the definition of splitting terms at \bbonem.
+This concludes the proof of program correctness, modulo \ref{lem:twoCorrect}.
+However in order to prove that lemma, 
+we will need to state analogous version for steps at \bbonem,
+since the various kinds multistage term in \lang\ are mutually dependent.
+Thus, this section proceeds by covering the definition of splitting at \bbonem, 
+starting with the value form at that world.
 
@@
 partial value $v$ into a split value $\mval i q$.
-The first component, the \emph{immediate value} $i$, is a monostage value holding
-the result of all first-stage components of $v$.  The second
-component, $q$, is a residual representing all second-stage
-variable references of $v$.
+Since partial values, by definition, have no remaining work in the first pass
+and since transfer of information between the stages occurs in the first pass of evalution,
+we know that the second stage components of $v$ can no longer depend on it's first stage components.
+Analogously, this must hold of the output of splitting, 
+which is why $q$---unlike the resumers of world \bbtwo\ term splitting---is not open on a variable.
 
-To provide intuition about the intent of partial value splitting, consider the
+Thus instead of a pipeline relationship, $i$ and $q$ bear more of a parallel relationship.
+That is, they embody the complementary portions of $v$ that are relevant to each stage.
+For this reason we call $i$ the \emph{immediate value} and $q$ the residual.
+
+To provide intuition about the behavior of partial value splitting, consider the
 following partial value:
 \begin{lstlisting}
-1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
+1` (next{`2`y`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`y`1`}))`
 \end{lstlisting}
@@
 \begin{lstlisting}
-2`(yhat, ((),yhat))`
+2`(y, ((),y))`
 \end{lstlisting}
@@
 \ref{fig:termSplit}.
-Observe that partial value splitting does not lose any meaningful information (namely injection tags),
-which is the validity requirement for a translation of value forms.
+
+In some regards, the exact formulation of partial value splitting is arbitrary.
+For instance, we chose to replace ``holes'' with unit values,
+but in fact we could have used any value there and it would make no difference in the end.
+There are however, at least some parts of the definition that are not arbitrary.
+Importantly, partial value splitting must not lose any meaningful information,
+such as injection tags.
 

### 2015-07-08 paper/paper.tex
minor
uncomment splitting section
adding appendix, and moving map there
...
algoexamples renamed examples
minor fixes
I can't remember what this change was. It's late.
supplimentified
updated format stuff

@@
 \usepackage{verbatim}
+\usepackage{xspace}
 
@@
 \newtheorem{theorem}{Theorem}[section]
+\newtheorem*{theorem*}{Theorem}
 \newtheorem{lemma}[theorem]{Lemma}
@@
 
-%% Vote on your favorite title! Lower is better!
-\title{Data Specialization for a Staged Lambda Calculus}
-\title{Algorithmically Staged Lambda Calculus}
-\title{Algorithmic Staging}
 \title{Automatically Splitting a Staged Lambda Calculus}
+\subtitle{(Supplementary Material in Appendix)}
 
-\authorinfo{Nicolas Feltman \and Carlo Angiuli \and Umut Acar \and Kayvon Fatahalian}
-           {Carnegie Mellon University}
-           {\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}
+\authorinfo{Anon A. Anon}%Nicolas Feltman \and Carlo Angiuli \and Umut A. Acar \and Kayvon Fatahalian}
+           {Anonymous Institute}%Carnegie Mellon University}
+           {anon@anon.com}%\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}
 
@@
 
-\category{CR-number}{subcategory}{third-level}
+%\category{CR-number}{subcategory}{third-level}
 
@@
 
-\input{example}
+\input{overview}
 
@@
 
-\input{algoexamples}
+\input{examples}
 \input{relatedwork}
+\input{conclusion}
+
 %\input{umut}
-
-%\appendix
-%\section{Appendix Title}
-%This is the text of the appendix, if you need one.
 
@@
 
-% We recommend abbrvnat bibliography style.
+\bibliography{paper}
+\bibliographystyle{abbrvnat}
 
@@
 
-\bibliography{paper}
-\bibliographystyle{abbrvnat}
+\newpage\appendix
+\input{appendix}
 

### 2015-07-08 paper/relatedwork.tex
small changes to related work

@@
+%!TEX root = paper.tex
+
 \section{Related Work}
@@
 
-\paragraph{Frequency-reduction via pass separation.}
-Perhaps the closest work to ours is the algorithm for data
-specialization given by Knoblock and Ruf~\cite{knoblock96}, which
-also seeks to statically split an explicitly staged program into two
-stages.  However, they only consider a simple first-order language and
-straight-line programs. Their work also treats all computations
-guarded by second-stage conditionals as second-stage computations,
-which would prevent optimization (via splitting) of programs such as
-quickselect.
+\paragraph{Stage Splitting.}
+
+Algorithms similar to ours have appeared in the partial evaluation community,
+where they are used as a preprocessing step to partial evaluation. 
+For instance, \cite{Mogensen89a} and \cite{DeNiel91} defined splitting algorithms
+for programs consistening of a set of first-order recursive functions.
+In addition to the lack of first-class functions, 
+the binding time specificiations used by these systems are more complicated and less expressive than the type system of \lang.
+The question of binding time analysis for a higher-order language was visited in \cite{Mogensen89b},
+however the implications for splitting were not explicitly considered.
 
 Automatic pass separation has also been a topic of interest in
-programming systems for computer graphics, where, to achieve
+programming systems for computer graphics, where to achieve
 high performance, programs are typically manually separated into
-components by frequency of execution (Indeed, separation into a pre-determined
+components by frequency of execution. (Indeed, separation into a pre-determined
 set of per-frequency modules is \emph{required} by the execution model
-of modern graphics architectures.)  The software engineering
-challenges of modifying multiple per-stage programs have led to
-suggestions of writing graphics programs in an explicitly-staged
-programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
-deferring the task of pass separation to the compiler. However, all
-prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
-been limited to simple, imperative languages. We are not aware of
-prior work that presents a splitting algorithm in the context of
-higher-order functional languages.
+of modern graphics architectures.)  
+The software engineering challenges of modifying programs in the split form (see \cite{Foley:2011}) have led to
+suggestions of differing the task of splitting to the compiler.
+Early work (\cite{knoblock96}) sought to determine the staging of programs automatically through use of binding time analysis,
+however later work (\cite{Proudfoot:2001,Foley:2011,He:2014}) has focused on giving programmers explicit control of the staging through 
+programming language features. 
+All of these efforts in graphics have been limited to simple, imperative languages.
 

### 2015-07-09 paper/relatedwork.tex
in the middle of fixing related work

@@
 
-Algorithms similar to ours have appeared in the partial evaluation community,
-where they are used as a preprocessing step to partial evaluation. 
+Algorithms for stage splitting have appeared in the literature under the names 
+{\em pass separation}, {\em binding time separation}, {\em data specialization}, 
+{\em program bifurcation}, and as various unnamed compiler passes.
+
+In the partial evaluation community, stage splitting has been investigated 
+mostly as a preprocessing step for self-applied partial evaluators. 
 For instance, \cite{Mogensen89a} and \cite{DeNiel91} defined splitting algorithms
 for programs consistening of a set of first-order recursive functions.
+Those works used a grammar-based binding-time specification that's 
+relatively complicated compared to the type system of \lang.
+This complication mostly stems from the decision to start with an untyped language;
+.
+
 In addition to the lack of first-class functions, 
 the binding time specificiations used by these systems are more complicated and less expressive than the type system of \lang.
+
 The question of binding time analysis for a higher-order language was visited in \cite{Mogensen89b},

### 2015-07-09 paper/relatedwork.tex
more in related work

@@
 for programs consistening of a set of first-order recursive functions.
-Those works used a grammar-based binding-time specification that's 
+Those works use a grammar-based binding-time specification that's 
 relatively complicated compared to the type system of \lang.
-This complication mostly stems from the decision to start with an untyped language;
-.
+This complication mostly arises from the decision to start with an untyped language:
+one must first find a way to say ``argument $x$ will be a tuple'' before one can say
+``the first component of $x$ is available at the second stage,'' 
+and so much of their binding-time specification just amounts to a type system.
+Formulations more like \lang\ (namely the two-level system of \cite{NN92-twolevel}) were known at the time,
+but \cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
+Davies essentially solved this limitation in \cite{davies96} (pending our further tweak with $\curr$),
+and the question of stage splitting as we have presented it has been ripe ever since.
+Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b},
+but the implications for splitting were not explicitly considered.
 
-In addition to the lack of first-class functions, 
-the binding time specificiations used by these systems are more complicated and less expressive than the type system of \lang.
-
-The question of binding time analysis for a higher-order language was visited in \cite{Mogensen89b},
-however the implications for splitting were not explicitly considered.
-
-Automatic pass separation has also been a topic of interest in
+Splitting-like algorithms have also been a topic of interest in
 programming systems for computer graphics, where to achieve
@@
 The software engineering challenges of modifying programs in the split form (see \cite{Foley:2011}) have led to
-suggestions of differing the task of splitting to the compiler.
-Early work (\cite{knoblock96}) sought to determine the staging of programs automatically through use of binding time analysis,
-however later work (\cite{Proudfoot:2001,Foley:2011,He:2014}) has focused on giving programmers explicit control of the staging through 
+suggestions of defering the task of splitting to the compiler.
+Early work \cite{knoblock96} sought to determine the staging of programs automatically through use of binding time analysis,
+however later work \cite{Proudfoot:2001,Foley:2011,He:2014} has focused on giving programmers explicit control of the staging through 
 programming language features. 

### 2015-07-09 paper/examples.tex
add a remark pointing to the supplemental material
various fixes

@@
 evaluation is more useful in this case. (Notice, however, that partially
-evaluating \texttt{exp2} on a \texttt{binnat} is essentially the same as as
+evaluating \texttt{exp2} on a \texttt{binnat} is essentially the same as 
 partially evaluating \texttt{exp} on the corresponding \rmint.) Nevertheless,
@@
 \end{lstlisting}
+%
+In the appendix provided as supplemental material, we extend this example by
+splitting \texttt{map (qss l)} over an arbitrary list of ranks \texttt{ks}. We also
+provide an example of a program whose boundary data structure is a trie.
 

### 2015-07-09 paper/overview.tex
various fixes

@@
 order to find the desired element.  Assuming that the input is uniformly
-randomly ordered (which can be achieved by prepermuting the input), \texttt{qs}
+randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
 runs in expected $O(n)$ time.
@@
 \texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
-Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since since
+Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
 \texttt{s} is a list, leading to no improvement in efficiency.

### 2015-07-09 paper/splitting.tex
various fixes

@@
 The rules of the three splitting judgements are given in \ref{fig:termSplit}.
-They are simply recursive on structure of the term, 
+They are simply recursive on the structure of the term, 
 implying that the splitting algorithm runs in linear time.
@@
 and since transfer of information between the stages occurs in the first pass of evalution,
-we know that the second stage components of $v$ can no longer depend on it's first stage components.
+we know that the second stage components of $v$ can no longer depend on its first stage components.
 Analogously, this must hold of the output of splitting, 

### 2015-07-09 paper/examples.tex
various edits

@@
 This is nearly identical to the cleaned-up code we presented in
-\ref{fig:qs-split}, except that the mixed-stage argument to both passes is
+\ref{fig:qs-split}, except that the mixed-stage argument to both passes 
 is explicitly masked, and we do not suppress the trivial first-stage result

### 2015-07-09 paper/semantics.tex
various edits

@@
 We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
-
 Although \lang\ describes computations that occur in two stages,
@@
 This is because the only way to eliminate a $\fut$ wrapper is by using a {\tt prev}, which yields an $A$ at \bbtwo.
-This feature was adapted from linear temporal logic, via \cite{davies96}.
+This feature was adapted from linear temporal logic, via \cite{davies96},
+and ensures that there can be no flow of information from the second stage to the first.
 
@@
 \bbonep.)
-
 An $e$ of type $\curr A$ at \bbonem\ can be unwrapped as an $A$ at \bbonep\
@@
 
-Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we need the
+Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we introduce the
 $\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
 open on $\curr{A}$ and $\curr{B}$ respectively.
-Without this, there would be no way to get information out from under a $\curr$ wrapper.
+Importantly, this feature allows information to flow from \bbonep\ to \bbonem,
+which would not be possible otherwise.
 
@@
 Accordingly, our dynamics operates in two passes.
-The first pass takes an input program $\coltwo e A$
+The first pass takes an input top-level program $\coltwo e A$
 and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
@@
 The second pass further reduces this residual.
-Since a residual $q$ is monostage by definition, this second pass is just standard unstaged evaluation.
-Consequently, this paper does not describe in any detail the second pass of evaluation.
+Since a residual $q$ is monostage by definition, this second pass is standard unstaged evaluation
+and is not described in further detail in this paper.
 Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.
@@
 This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
-One consequence of the stronger requirement on partial values is that we will need a new kind of step to put terms into that form.
+\TODO Actually explain what MP and PE want this for.
+One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
 To illustrate, consider:
@@
 \end{lstlisting}
+\TODO what?
 To see that the partial application of \texttt{qss} is only evaluated once,

### 2015-07-09 paper/semantics.tex
said what metaprogrammings goal was

@@
 ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
-This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
-\TODO Actually explain what MP and PE want this for.
+This contrast stems from differing goals: metaprogramming explicitly intends to model code writing code.
 One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.

### 2015-07-09 paper/relatedwork.tex
Clean up the graphics related work section to be as factual as possible.

@@
 mostly as a preprocessing step for self-applied partial evaluators. 
-For instance, \cite{Mogensen89a} and \cite{DeNiel91} defined splitting algorithms
+For instance,~\cite{Mogensen89a} and~\cite{DeNiel91} defined splitting algorithms
 for programs consistening of a set of first-order recursive functions.
@@
 and so much of their binding-time specification just amounts to a type system.
-Formulations more like \lang\ (namely the two-level system of \cite{NN92-twolevel}) were known at the time,
-but \cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
-Davies essentially solved this limitation in \cite{davies96} (pending our further tweak with $\curr$),
+Formulations more like \lang\ (namely the two-level system of~\cite{NN92-twolevel}) were known at the time,
+but~\cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
+Davies essentially solved this limitation in~\cite{davies96} (pending our further tweak with $\curr$),
 and the question of stage splitting as we have presented it has been ripe ever since.
@@
 
-Splitting-like algorithms have also been a topic of interest in
-programming systems for computer graphics, where to achieve
-high performance, programs are typically manually separated into
-components by frequency of execution. (Indeed, separation into a pre-determined
-set of per-frequency modules is \emph{required} by the execution model
-of modern graphics architectures.)  
-The software engineering challenges of modifying programs in the split form (see \cite{Foley:2011}) have led to
-suggestions of defering the task of splitting to the compiler.
-Early work \cite{knoblock96} sought to determine the staging of programs automatically through use of binding time analysis,
-however later work \cite{Proudfoot:2001,Foley:2011,He:2014} has focused on giving programmers explicit control of the staging through 
-programming language features. 
-All of these efforts in graphics have been limited to simple, imperative languages.
-
-%% \section{Related Work}
-
-%% \ur{
-%% \begin{itemize}
-%% \item 
-%% I find it hard to distinguish between blue and black on printed paper.
-%% \item
-%% $\pause$ should be called ``promote'' or something like that. I
-%%   believe this is what Taha calls cross stage persistence.  THat work should
-%%   be cited.
-%% \end{itemize}
-%% }
-
-%% \crem{\emph{This text came from section 3:}
-%% \lang\ is adapted from \cite{davies96}, and is very similar to other staged
-%% languages considered in the partial evaluation and metaprogramming literature.
-%% Unlike \cite{davies96}, we restrict \lang\ to exactly two stages, to simplify
-%% our presentation of our stage-splitting algorithm. We expect that the techniques
-%% presented here extend to more stages.
-%% Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
-%% the metaprogramming literature, because we avoid duplicating stage \bbtwo\
-%% computations. In addition to being more efficient, our non-duplicating semantics
-%% are easier to compare with our splitting algorithm, whose development and
-%% exposition is our main goal.
-%% }
-
-%% There are several distinctive aspect of our work which we wish to compare and contrast.
-%% Firstly, \lang\ is a simply typed lambda calculus 
-%% with explicit stage annotations and a non-duplicating semantics.
-%% Its statics and dynamics enforce no backward information flow,
-%% which a is necessary property to define a splitting transformation
-%% that is equivalent to the dynamics.
-%% Finally, the splitting transformation itself operates on any well-typed open term in \lang.
-
-%% Partial evaluation is a well-studied technique for specializing programs to some of their input.  
-%% Often, the goal of partial evaluation systems is to go all the way from unstaged code to a specialized version.
-%% Although this problem definition implicitly includes binding time analysis,
-%% many presentations factor out that step and start with a stage-annotated representation, as we have done.
-%% We've shown that our dynamic semantics can be used to perform partial evaluation 
-%% [need to do more reading to know exactly how it compares].
-
-%% One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
-%% and so they must necessarily unfold recursive calls and wait for all stage \bbone\ input to operate.
-%% There has been some work to avoid these issues by templating [need citations].
-%% Stage-splitting, in contrast, goes all the way to operating on terms which are open on first stage variables,
-%% and so these problems are avoided.
-
-%% Stage-splitting itself has appeared in the literature---under 
-%% the names {\em pass separation} and {\em data specialization}--- for various simpler input languages.
-
-%% Most often, splitting is considered a compiler optimization (\cite{jorring86},\cite{knoblock96})
-%% to minimize recomputation.
-%% In this use case, only the arguments to the program are annotated with stages, 
-%% and the splitter is left to perform an {\em ad hoc} binding time analysis to determine the 
-%% internal staging of the program.
-%% Because this occurs without programmer supervision, the binding time analyzers for these uses are 
-%% reluctant to place stage \bbone\ code within stage \bbtwo\ conditionals, 
-%% since that would result in speculative behavior that could increase the overall runtime.
-%% We don't worry about this since the annotations of \lang\ are assumed to represent the will of a programmer.
-
-%% Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  
-%% But instead of using a binding-time analysis to determine what parts of the program to put in what shader, 
-%% Spark gives the programmer manual control of term staging using a language that is in spirit similar to \lang.
-
-%% We are also not the first to use stage splitting to derive algorithms.
-%% \cite{malmkjaer89}, expanding on the work of \cite{barzdins88}, defined a stage-splitter
-%% for a language of tail-recursive first-order equations with explicit staging annotations [double check that],
-%% and considered applications to string matching.
-
-%% Our work is distinct from all previous stage splitters in the following ways:
-%% \begin{itemize}
-%% \item Our input language, \lang, has a disciplined type theory.
-%% \item \lang\ has full first-class functions (and by extension, recursion), as well as sum types.
-%% \item Our stage splitting is proven correct.
-%% \end{itemize}
-
-%% The goal of {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}) is 
-%% to use code in a host language to write code in an object language.
-%% Like partial evaluation and stage-splitting, this can be modeled by a staged language with explicit annotations.
-%% Instead of identifying stages with points in time, stages are identified with host/object levels of the language,
-%% and so type safety ensures the proper separation of levels of the language and safety of the constructed object code.
-
-%% But the differences in goals of metaprogramming do cause a few differences in form.
-%% For example, it's generally preferred in the metaprogramming context to use 
-%% a dynamics that duplicates the contents of $\next$ blocks.
-%% Moreover, in metaprogramming systems it makes sense to define an operator, \texttt{run}, 
-%% which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
-%% This is essentially a form a backwards information flow.
-%% Thus, the \texttt{run} feature is directly incompatible with a temporal interpretation of stages,
-%% which stage-splitting requires.
-
-%% \cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
-%% and showed the equivalence between that type system and existing one for binding time analysis. 
-%% That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
-%% That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
-%% In fact, if we change the $\next$ and $\prev$ rules of ours to 
-%% \begin{mathpar}
-%% \infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
-%% \infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
-%% \end{mathpar}
-%% and treat $\next~q$ as a partial value for any residual $q$,
-%% then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
-%% environment bookkeeping in $\redonesym$, by inlining residuals instead of
-%% hoisting them in \verb|let|-bindings.
-%% From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
-%% {\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.
+Splitting algorithms have also been a topic of interest in programming
+systems for computer graphics, where to achieve high performance,
+programs are typically manually separated into components by frequency
+of execution. (Indeed, separation into a pre-determined set of
+per-frequency modules: e.g., per frame, per pixel, per triangle, is
+\emph{required} by the execution model of modern graphics
+architectures.)  The software engineering challenges of modifying
+graphics programs written in a split form have led to explorations of
+deferring the task of splitting to a compiler\,\cite{Foley:2011}.
+While early work sought to determine the staging of programs
+automatically via binding time analysis\,\cite{knoblock96}, more
+recent designs, like \lang, require explicit staging by the
+programmer\,\cite{Proudfoot:2001,Foley:2011,He:2014}.  However,
+splitting algorithms developed in all of these efforts have been
+limited to simple, imperative graphics shading languages that lack
+simple features such as functions.
 
@@
 %% Is this a problem for us?
-
-%% \paragraph{Supercompilation.}
-%% ~\cite{Turchin86}
-
-%% \paragraph{Intro.}
-
-%% A predominating theme in the design, implementatation, and
-%% optimization of algorithm and software is frequency reduction and
-%% precomputation.  The idea is to reduce the frequency of computations
-%% by reducing their frequency and ideally by precomputing when possible.
-
-%% The reason for why this theme is as predominant as it is that it is
-%% naturally common in many actual applications.  For example, we may
-%% have a collection of names or numbers that we frequently lookup, as
-%% for example part of a membership test as part of a larger system.  In
-%% some applications areas such as in graphics (the motivation for our
-%% work), this is so common that there is a relatively large literature
-%% on speeding up graphics computations by taking advantage of frequency
-%% reduction and precomputation.
-
-%% One way to express frequency of computations is to allow the
-%% programmer to stage computation explicitly by making explicit the
-%% stage of each computation and then by moving computations between
-%% stages to improve efficiency by performing work as early (to save time
-%% later when performance might be more critical) and  as infrequently
-%% as possible (to reduce overall time).
-
-%% For example, in meta-programming~\cite{}, the programmer can assign a
-%% program expression a stage (stage 1, 2, 3, ...), which indicates the
-%% stage at which that expression evaluates.  Combined with the ability
-%% for earlier stages to construct expressions of later stages,
-%% meta-programming techniques can be used to construct efficient
-%% programs that improve efficiency over their non-staged counterparts by
-%% reducing frequency reduction and precomputation.  For example, the
-%% programmer can write a program that takes a list of numbers $\ell$ as
-%% a first stage value and returns a second-stage function that performs
-%% a fast membership test on that list $\ell$ by inlining all the
-%% comparisons.  When given a key to lookup in the second stage, the
-%% membership test function would quickly return the result.

### 2015-07-09 paper/semantics.tex

@@
 
-Lastly, to case on a term $e$ of type $\curr(A+B)$ at world \bbonem, we introduce the
-$\caseP{e}{x.e_1}{x.e_2}$ construct, whose branches are world \bbonem\ terms
-open on $\curr{A}$ and $\curr{B}$ respectively.
-Importantly, this feature allows information to flow from \bbonep\ to \bbonem,
-which would not be possible otherwise.
+These features are sufficient to ensure that mixed code does not leak into ground code,
+however they also prevent information from ever escaping a $\curr$ wrapper.
+So to allow the latter behavior but not the former, we introduce the $\caseP{e}{x.e_1}{x.e_2}$ construct, 
+whose predicate is of type $\curr(A+B)$ and
+whose branches are world \bbonem\ terms open on $\curr{A}$ and $\curr{B}$ respectively. 
+This essentially allows code at \bbonem\ to inspect an injection tag within a $\curr$.
 
@@
 ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
-This contrast stems from differing goals: metaprogramming explicitly intends to model code writing code.
+This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
+whereas our applications feel more natural with a eager interpretation of {\tt next}.
 One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.

### 2015-07-10 paper/splitting.tex
added equivalence figure

@@
 \input{figures/splitting-overview}
+
+\TODO add related work
 
@@
 The rules of the three splitting judgements are given in \ref{fig:termSplit}.
-They are simply recursive on the structure of the term, 
-implying that the splitting algorithm runs in linear time.
-
-Before discussing the correctness of these translations, 
-we note that splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
+Since the rules are simply recursive on the structure of the term, 
+the splitting algorithm runs in linear time on the size of the input program.
+Splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
 That is, each splitting judgement defines a total function.
@@
 Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
-This output is a multistage program in its own right.
-It is evaluated by reducing $p$ to the value $b$ in the first pass,
+This output is evaluated in two passes by reducing $p$ to the value $b$ in the first pass,
 then plugging that result in for $l$ to produce $[b/l]r$,
@@
 
+\TODO ADD EQUIVALENCE FIGURE
+
 We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
 In the base case, where $e$ is already a residual of the form $\exv q$, we know
-$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
+$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, so by uniqueness of splitting, 
 $p = \tup{}$ and $r = q$.
@@
 and \mbox{$\splittwo e A p l r$}.
+By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
+The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
+
+\TODO defin the gap, then close it with the lemma
 To help this case, we introduce \ref{lem:twoCorrect}, 
@@
 and whose proof will concern the rest of this section.
-By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
-The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
+
 From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
@@
 
-This concludes the proof of program correctness, modulo \ref{lem:twoCorrect}.
-However in order to prove that lemma, 
+The above proof of \ref{thm:allCorrect} invokes \ref{lem:twoCorrect},
+In order to prove that lemma, 
 we will need to state analogous version for steps at \bbonem,
@@
 
-Thus instead of a pipeline relationship, $i$ and $q$ bear more of a parallel relationship.
-That is, they embody the complementary portions of $v$ that are relevant to each stage.
+Thus $i$ and $q$ embody the complementary portions of $v$ that are relevant to each stage.
 For this reason we call $i$ the \emph{immediate value} and $q$ the residual.
@@
 \input{figures/splitting-term}
+\input{figures/splitting-misc}

### 2015-07-10 paper/relatedwork.tex

@@
 For instance,~\cite{Mogensen89a} and~\cite{DeNiel91} defined splitting algorithms
-for programs consistening of a set of first-order recursive functions.
+for programs consisting of a set of first-order recursive functions.
 Those works use a grammar-based binding-time specification that's 

### 2015-07-10 paper/splitting.tex
expand on the importance of ground in 4.5

@@
 
-The \texttt{partition} function in the quickselect example in
-\ref{sec:exampleQS} is an example of such a function that contains no
-second-stage computations, but, if not defined within a $\pure$
-annotation, would cause splitting to generate a recursive resumer.
+An illustrative example is the \texttt{partition} function in the quickselect
+example of \ref{sec:exampleQS}. If \texttt{partition} were defined at \bbonem\
+instead of \bbonep, then (like \texttt{qs}) it would split into two functions
+\texttt{partition1} and \texttt{partition2}, the latter of which recursively
+computes the (trivial) second-stage component of \texttt{partition}'s result.
+Moreover, \texttt{qs2} would call \texttt{partition2}, just as \texttt{qs1}
+calls \texttt{partition1}:
+\begin{lstlisting}
+2`fun qs2 (p : tree, k : int) : int =
+... | Branch (i,h,p1,p2) =>
+  let val () = partition2 ((),()) in
+  case compare k i of ...
+\end{lstlisting}
+By explicitly marking that \texttt{partition} has no second-stage computations,
+we avoid both generating the resumer \texttt{partition2}, and calling it in
+\texttt{qs2}.
 
-\TODO consider expanding, based on comments from reviews
+\TODO somebody should read the new text about partition
 

### 2015-07-10 paper/overview.tex
Made a few tiny edits while reading.

@@
 \begin{proof}
-  We use a few standard facts.  We note first that \texttt{qs1} is
+  We note first that \texttt{qs1} is
   isomorphic to a quicksort algorithm but builds a binary search tree

### 2015-07-10 paper/splitting.tex
attempting to lift some more discussion into 4.0

--- 2015-07-10 paper/splitting.tex
@@
 
+Recall that a partial value $v$, by definition, has no remaining work in the first pass.
+Since transfer of information between the stages occurs during the first pass, 
+there can be no remaining internal communication between the stages of $v$.
+In contrast, a term $e$ (at either \bbonem\ or \bbtwo) could still evaluate further in the first pass,
+and so there may remain internal communication to happen between the stages of $e$.
+
+This difference is why the second-stage output of either form of term splitting ($l.r$) is open on a variable,
+while that of partial value splitting ($q$) is not.
+Substitution for $l$ in $r$ is how communication between stages is represented in these outputs.
+
 \subsection{World \bbtwo\ Term Splitting}
@@
 and ``$\equiv$'' indicates a monostage equivalence, which we cover later.
-
-\TODO ADD EQUIVALENCE FIGURE
 
@@
 from the proofs of \ref{lem:oneCorrect} and \ref{lem:twoCorrect}.
+The equivalence relation ``$\equiv$'' is defined in \ref{fig:equiv}.
 

### 2015-07-10 paper/relatedwork.tex

@@
 \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
-Kleene did not intend to improve efficiency, however.  Lombardi
-appears to be the first person to use the term in his work on
-incremental computation~\cite{Lombardi67}.  There has since been 
-an enormous body of work on partial evaluation, and we refer the reader to the
-book by Jones, Gommard, and Sestoft for a comprehensive discussion of
-work until the early 90's~\cite{JGS93}.
+%Kleene did not intend to improve efficiency, however.  Lombardi
+%appears to be the first person to use the term in his work on
+%incremental computation~\cite{Lombardi67}.
+There has since been an enormous body of work on partial evaluation,
+and we refer the reader to the book by Jones, Gommard, and Sestoft for
+a comprehensive discussion of work until the early 90's~\cite{JGS93}.
 
@@
 be computed by the knowledge of static arguments. Using this
-information and the values of static arguments, program
-specialization specializes the original program to a partially-evaluated
-one that operates on many different dynamic arguments.  This
-approach has been applied to construct partial evaluators for a number
-of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism}, 
-although for the most part restricted to first-order subsets.
+information and the values of static arguments, program specialization
+specializes the original program to a partially-evaluated one that
+operates on many different dynamic arguments.  This approach has been
+applied to construct partial evaluators for a number of languages such
+as Scheme~\cite{OB91-Similix,Consel88-Schism}, although for the most
+part restricted to first-order subsets.
 
@@
 predict. This led to investigations based on type systems for making
-the stage of computations explicit in programs~\cite{GJ91-lambda,NN92-twolevel}
-and writing ``metaprograms'' that, when evaluated at a stage, yield
-a new program to be evaluated at the next stage.
+the stage of computations explicit in
+programs~\cite{GJ91-lambda,NN92-twolevel} and writing ``metaprograms''
+that, when evaluated at a stage, yield a new program to be evaluated
+at the next stage.
 %J{\o}rring and
@@
 small ways to account for the specific evaluation model that we
-propose; however, the key difference between our work and this prior work is
-that we instead focus on the problem of pass separation.
+propose; however, the key difference between our work and this prior
+work is that we instead focus on the problem of splitting.
 %we developed an algorithm
@@
 
-Another class of precomputation techniques is incremental
-computation, where a program can efficiently respond to small changes
-to its input by only recomputing parts of the computation affected
-by the
+Another class of precomputation techniques is incremental computation,
+where a program can efficiently respond to small changes in its inputs
+by only recomputing parts of the computation affected by the
 changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
-Like other precomputation techniques, incremental computation can be
-thought of as a staged model, where changes in inputs come over a
-period of time in stages.
+Thus, incremental computation can be thought of as a staged model,
+where changes in inputs come over a period of time in stages.
 %
-However, incremental computation does not requiring fixing any of its
-inputs and allows, in the general case, for all inputs to
-change. Thus, its benefits are dependent on the input changes made and
-techniques typically utilize significant run-time (dynamic)
-information.  For example, while we can apply incremental computation
-to our quickselect example (\ref{sec:overview}), doing so would not
-realize the same bounds as splitting since incremental computation
-techniques would unfold the quickselect function based on the demanded
-ranks, which can require linear time (as opposed to logarithmic) at
-each stage.  Such an implementation would also maintain sophisticated
-data structures dynamically at run time, the benefits of which depend
-highly on input changes applied.
+However, unlike splitting, incremental computation does not require
+fixing any of its inputs and, in the general case, allows for all
+program inputs to change. Thus, the benefits of incremental
+computation depends on what changes to inputs are made. For example,
+while it is possible to apply incremental computation to our
+quickselect example (\ref{sec:overview}), techniques would unfold the
+quickselect function based on the demanded ranks, potentially
+incurring linear time cost at each step of the algorithm (as opposed
+to logarithmic result produced by splitting). Unlike the results of
+splitting, such an implementation would also maintain sophisticated
+data structures dynamically at run time to track what computations
+must be performed.
 
@@
 architectures.)  The software engineering challenges of modifying
-graphics programs written in a split form have led to explorations of
+graphics programs written in a split form have spurred interest in
 deferring the task of splitting to a compiler\,\cite{Foley:2011}.

### 2015-07-10 paper/splitting.tex
splitting reorganization

--- 2015-07-10 paper/splitting.tex
@@
 
-\subsection{World \bbtwo\ Term Splitting}
+\subsection{Outputs of Splitting}
 
-Top level programs in \lang\ are given by a term $\coltwo e A$.
-Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
-We write this as $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$.
-
-Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
-This output is evaluated in two passes by reducing $p$ to the value $b$ in the first pass,
-then plugging that result in for $l$ to produce $[b/l]r$,
-which is evaluated in the second pass.
+Splitting a top level program $\coltwo e A$, via $\splittwosym$, yields $\pipeS p l r$.
+Like $e$, this output is evaluated in two passes.
+The first pass reduces $p$ to the value $b$ and plugs this result in for $l$ to produce $[b/l]r$;
+the second pass evaluates $[b/l]r$.
+The relationship between the two stages in this case is thus like a pipeline,
+which is why we write them with a `\mytt{|}' in between.
 Since execution of the first pass serves to generate input for
@@
 
-The correctness of the splitting translation requires that this simple 
-evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
+\TODO prior work statement
+
+Splitting a partial value $v$, via $\vsplito$, yields $\mval i q$.
+Since partial values, by definition, have no remaining work in the first pass
+and since transfer of information between the stages occurs in the first pass of evalution,
+we know that the second stage components of $v$ can no longer depend on its first stage components.
+Analogously, this must also hold for the output of splitting, 
+which is why $q$---unlike the resumer of world \bbtwo\ term splitting---is not open on a variable.
+Thus, $i$ and $q$ are operationally independent, but they represent the complementary portions of $v$ that are relevant to each stage.
+We call $i$ the \emph{immediate value} and $q$ the residual.
+
+For any $\colmix e A$, splitting $e$ via $\splitonesym$ yields the pair of monostage terms $\pipeM c l r$. 
+This output form is essentially a hybrid of the previous two.
+Because $e$ is a term, $c$ needs to produces a boundary value $b$ to be passed to the resumer ($l.r$)
+And since $e$ types at world \bbonem, it has an eventual result at the 
+first stage as well as the second, and so it must produce an immediate value $i$.
+The term $c$ meets both of these responsibilities by reducing to the tuple $\tup{i,b}$,
+and so we call it a \emph{combined term}.
+
+\subsection{World \bbtwo\ Term Splitting}
+
+The correctness of the splitting translation requires that the simple 
+evaluate-and-plug semantics on the output produces the same answer as the staged semantics of the previous section.
 That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
@@
 \label{sec:masking}
-
-Partial value splitting (given by the relation $\vsplito$) converts a 
-partial value $v$ into a split value $\mval i q$.
-Since partial values, by definition, have no remaining work in the first pass
-and since transfer of information between the stages occurs in the first pass of evalution,
-we know that the second stage components of $v$ can no longer depend on its first stage components.
-Analogously, this must hold of the output of splitting, 
-which is why $q$---unlike the resumers of world \bbtwo\ term splitting---is not open on a variable.
-
-Thus $i$ and $q$ embody the complementary portions of $v$ that are relevant to each stage.
-For this reason we call $i$ the \emph{immediate value} and $q$ the residual.
 
@@
 
-For any $\colmix e A$, 
-splitting $e$ (via $\splitonesym$) again 
-yields the pair of monostage terms $\pipeM c l r$.
-
-The responsibilities of this first stage component, $c$, can be seen as a combination of the
-those of world \bbtwo\ term splitting and partial value splitting.
-Since $e$ is a term, the first stage computations within it may yet still
-affect its second stage computations,
-and so $c$ needs to produces a boundary value $b$ to be passed to the resumer ($l.r$).
-But since $e$ is a world \bbonem\ construct, it has an eventual result at the 
-first stage as well as the second, and so it must produce an immediate value $i$.
-The term $c$ meets both of these responsibilities by reducing to the tuple $\tup{i,b}$,
-and so we call it a {\em combined term}.
-
 The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.

### 2015-07-10 paper/abstract.tex
various small edits
reviewer comments

@@
 splitting transformation maps such programs into multi-pass programs
-that can run significantly more efficiently by assigning frequent
+that can run asymptotically more efficiently by assigning frequent
 computations to earlier passes so that they can be reused (cheaply)
@@
 achieved without significant run-time penalty.
+
+\TODO We also proved some stuff.
 

### 2015-07-10 paper/examples.tex
reviewer comments

@@
 \begin{lstlisting} 
-datatype binnat = Zero | Even of nat | Odd of nat
+datatype binnat = Zero
+                | Even of binnat 
+                | Odd of binnat
 
@@
 \texttt{qss} finds the $k$th largest element of a list $l$ by
-recursively \texttt{partition}ing the list by its first element, then recurring
+recursively partitioning the list by its first element, then recurring
 on the side containing the $k$th largest element. $l$ is first-stage and $k$ is
@@
   | Branch (n,left,h,right) =>
-      case compare (k, n) of 
+      case compare k n of 
         LT => qs2 (((), k), left) 
@@
 This is nearly identical to the cleaned-up code we presented in
-\ref{fig:qs-split}, except that the mixed-stage argument to both passes 
-is explicitly masked, and we do not suppress the trivial first-stage result
-returned by \texttt{qs1}.
+\ref{fig:qs-split}, except we do not suppress the trivial inputs and outputs of 
+\texttt{qs1} and \texttt{qs2}.
 
-\texttt{qs1} \texttt{partition}s $l$, but since the comparison with $k$ (to
+The function \texttt{qs1} partitions $l$, but since the comparison with $k$ (to
 determine which half of $l$ to recur on) is at the second stage, it simply

### 2015-07-10 paper/intro.tex
minor fixes
reviewer comments

@@
+%!TEX root = paper.tex
+
 \section{Introduction}
@@
 changes for one pass may require changes to other passes to ``plumb'' inputs
-from all other frequencies\,\cite{Foley:2011}.
+from all other frequencies (\TODO explain what ``frequency'' means here)\,\cite{Foley:2011}.
 %
@@
 $f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
-evaluating the first pass on $x$, and then the second pass on each $y_i$.
+evaluating the first pass on $x$, and using the result $z$ to evaluate the
+second pass on each $y_i$.
 %

### 2015-07-10 paper/overview.tex
reviewer comments

@@
 the relationship of \texttt{k} to the size \texttt{i} of the first half, in
-order to find the desired element.  Assuming that the input is uniformly
+order to find the desired element.  
+If the index is out of range, a default value of \texttt{0} is returned.
+Assuming that the input is uniformly
 randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
@@
 
-The process of adding staging annotations to unstaged code has been
+The process of automatically adding staging annotations to unstaged code has been
 the subject of extensive research (\ref{sec:related}). In this paper, we do not
-consider this problem, instead assuming that the annotations have been provided.
+consider this problem, instead assuming that the annotations already exist.
 In the case of \texttt{qss}, we have specifically chosen annotations which
@@
 %
-The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-staged}.
+The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-split}.
 

### 2015-07-10 paper/relatedwork.tex
reviewer comments

@@
 applied to construct partial evaluators for a number of languages such
-as Scheme~\cite{OB91-Similix,Consel88-Schism}, although for the most
-part restricted to first-order subsets.
+as Scheme~\cite{OB91-Similix,Consel88-Schism}.
 

### 2015-07-10 paper/semantics.tex
reviewer comments

@@
 All three worlds contain unit, product, and function types defined in the usual fashion;
-sum and recursive types are provided for worlds \bbonep\ and \bbtwo\ 
-(we omit them from \bbonem, because they are not needed for our examples).
+sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
+\footnote{We omit sums and recursive types from \bbonem\ to save space, though we do not believe they would present any unique difficulty.}
 These ``standard'' features can only be constructed from subterms of the same world,

### 2015-07-10 paper/splitting.tex
some iteration on splitting

@@
 \input{figures/splitting-overview}
-
-\TODO add related work
 
@@
 
-\TODO prior work statement
-
 Splitting a partial value $v$, via $\vsplito$, yields $\mval i q$.
@@
 
-The correctness of the splitting translation requires that the simple 
+The dynamic correctness of the splitting translation requires that the simple 
 evaluate-and-plug semantics on the output produces the same answer as the staged semantics of the previous section.
 That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
-to the residual produced by direct evaluation, $q$.
+to the residual $q$ produced by direct evaluation, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$.
 This condition is stated more precisely as \ref{thm:allCorrect},
@@
 By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
-The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
-
-\TODO defin the gap, then close it with the lemma
-To help this case, we introduce \ref{lem:twoCorrect}, 
+From here, the inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
+All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
+To close this gap we introduce \ref{lem:twoCorrect}, 
 which essentially states that any single step is correct,
 and whose proof will concern the rest of this section.
+After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
 
-From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
-and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
-
-The above proof of \ref{thm:allCorrect} invokes \ref{lem:twoCorrect},
-In order to prove that lemma, 
+In order to prove that \ref{lem:twoCorrect},
 we will need to state analogous version for steps at \bbonem,
@@
 
-The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
-As in the world \bbtwo\ version, correctness requires that the applied resumer from either side of the step be equivalent.
-But additionally, it also requires that first stage results be exactly identical.
+The correctness of \bbonem\ term splitting with respect to a $\stepsym \bbonem$ step is given in \ref{lem:oneCorrect}.
+It's very similar to the world \bbtwo\ version, 
+in that the reduction of the first stage part of $e'$ should imply reduction of the first stage part of $e$,
+and that the resulting applied resumers should be equivalent.
+But split forms at \bbonem\ have one more piece of output than those at \bbtwo,
+namely the immediate value $i$.
+\ref{lem:oneCorrect} accounts for this by saying that 
+that immediate value must be exactly identical on both sides of the step.
 
@@
 
-Rather than attempt global optimization of the outputs of splitting,
-we instead leverage the type system to indicate when a term contains
-no second-stage computations by adding a third world \bbonep\ whose
-terms are purely first-stage.  While not necessary for ordinary staged
-term evaluation (\ref{sec:semantics}), this annotation is
-particularly useful for splitting, since splitting $\pure e$
-terms trivially yields a unit resumer.
-
 An illustrative example is the \texttt{partition} function in the quickselect
 example of \ref{sec:exampleQS}. If \texttt{partition} were defined at \bbonem\
-instead of \bbonep, then (like \texttt{qs}) it would split into two functions
+then (like \texttt{qs}) it would split into two functions
 \texttt{partition1} and \texttt{partition2}, the latter of which recursively
@@
 \end{lstlisting}
-By explicitly marking that \texttt{partition} has no second-stage computations,
-we avoid both generating the resumer \texttt{partition2}, and calling it in
-\texttt{qs2}.
 
-\TODO somebody should read the new text about partition
+Rather than attempt global optimization of the outputs of splitting,
+we instead leverage the type system to indicate when a term contains
+no second-stage computations by adding a third world \bbonep\ whose
+terms are purely first-stage.  
+Defining \texttt{partition} in this world is 
+tantamount to proving it has no second-stage computations,
+allowing splitting to avoid generating the resumer \texttt{partition2}
+and calling it from \texttt{qs2}.
 
+Recall that this annotation is not especially useful 
+for ordinary staged term evaluation (\ref{sec:semantics}).
+We suspect the key factor is that splitting is a static transformation,
+since other static trasofrmations \cite{Mogensen89a,DeNiel91} 
+also benefited from a clear distinction between ground and mixed terms.
+Ground is also similar to the validity mechanism in ML5 \cite{murphy05}.
 \end{abstrsyn}

### 2015-07-10 paper/relatedwork.tex
Minor editing on some of the work I did last night.

@@
 work is that we instead focus on the problem of splitting.
-%we developed an algorithm
-%for splitting terms in the proposed staged language.
-%the staging or the splitting algorithm, which given a two-staged
-%program splits it into two stages that can be executed sequentially,
-%passing information from the earlier to the latter stage by
-%construction.
-%
-%Unlike partial evaluation, splitting requires no knowledge of program
-%argument values, only their staging.
-%Using just the staging
-%information, the splitting algorithm can generate programs that can
-%perform non-trivial transformations on the code automatically.  As our
-%examples illustrate such staging transformations can improve the
-%efficiency of the source program, sometimes in asymptotically
-%significant ways.
 
@@
 where changes in inputs come over a period of time in stages.
-%
 However, unlike splitting, incremental computation does not require
@@
 program inputs to change. Thus, the benefits of incremental
-computation depends on what changes to inputs are made. For example,
-while it is possible to apply incremental computation to our
-quickselect example (\ref{sec:overview}), techniques would unfold the
+computation depend on what changes to inputs are made. For example,
+while it is possible to apply incremental computation to the
+quickselect example in \ref{sec:overview}, techniques would unfold the
 quickselect function based on the demanded ranks, potentially
 incurring linear time cost at each step of the algorithm (as opposed
-to logarithmic result produced by splitting). Unlike the results of
-splitting, such an implementation would also maintain sophisticated
+to the logarithmic result produced by splitting).  Unlike splitting,
+incremental computation techniques must also maintain sophisticated
 data structures dynamically at run time to track what computations
@@
 limited to simple, imperative graphics shading languages that lack
-simple features such as functions.
+even simple features such as functions.
 

### 2015-07-10 paper/abstract.tex
abstract draft

--- 2015-02-28 paper/abstract.tex
@@
+%!TEX root = paper.tex
+
 Staged programming languages assign a stage to each expression and
-evaluate each expression in its assigned stage.
-%For example, in a
-%two-staged program, some expressions are evaluated in the first stage
-%and others in the second stage.
-%Programming languages, evaluation
-%strategies, and type systems for staged programs where stages roughly
-%correspond to the arrival order of inputs have been studied
-%extensively in the areas of partial evaluation and meta programming.
-While staged programming has been studied extensively in the areas of
-partial evaluation and meta programming, in this paper we present an
-algorithm for statically {\em splitting} programs witten in a
-two-stage typed lambda calculus (with a $\fut$ modality denoting
-computation in a later stage) into multi-pass programs that are
-structured as two dependent, unstaged programs and are evaluated using
-two distinct execution passes.  While splitting, also called pass
-separation or data specialization, has been studied in the past, prior
-techniques were either manual or applicable only to simple programs
-that lack important features such as recursion and first-class
-functions.
+evaluate each expression in its assigned stage,
+and can be used to describe programs where inputs arrive at different times or rates.
 
-Since our work builds on a lambda calculus, our splitting algorithm
-enables the programmer to write staged programs by using powerful
-programming abstractions in a composable and modular fashion. Our
-splitting transformation maps such programs into multi-pass programs
-that can run significantly more efficiently by assigning frequent
-computations to earlier passes so that they can be reused (cheaply)
-later.  Our experiments based on a prototype implementation show that
-our splitting algorithm is able perform highly non-trivial algorithmic
-transformations, sometimes automatically improving efficiency
-asymptotically by nearly a linear factor.  Since splitting
-transformation takes place at compile time, these benefits are
-achieved without significant run-time penalty.
+In this paper we present an 
+algorithm for statically {\em splitting} these mixed-staged programs 
+into two dependent, unstaged programs where the outputs of the first part can be efficiently reused accorss multiple invocations of the second program.
 
-% The motivations for pass separation are 1) improvements in efficiency
-% by assigning frequent computations to earlier passes so that they can
-% be re-used (cheaply) later and, 2) performing such scheduling
-% statically at compile time without loss of any generality.  To assess
-% the effectiveness of our techniques for pass separation, we consider
-% several examples and demonstrate that our implementation is able to
-% perform pass separation statically as expected.  In a particularly
-% interesting example, involving recursion and recursive data, we show
-% that our techniques perform highly non-trivial algorithmic
-% transformations, improving efficiency asymptotically by nearly a
-% linear factor.
+While previous algorithms for performing this transformation 
+(also called \emph{pass separation} and \emph{data specialization}) operate on simpler imperative languages,
+out algorithm operates on an explicitly-two-staged, typed lambda calculus \lang, 
+with a $\fut$ modality denoting computation at a later stage,
+and a $\curr$ modality noting purely first stage code.
+Notably, our algorithm is the first to with support for mixed-stage recursive and higher-order functions.
 
-% Many algorithms can perform useful work before receiving all of their inputs.
-% If we regard those inputs as arriving at different \emph{stages} of the
-% computation, then we can \emph{split} those algorithms into, for each stage, a
-% function performing all the work dependent only on the available inputs.
+We have proven the dynamic correctness of our splitting algorithm with respect
+to a partial-evaluation semantics, and mechanized this proof in Twelf.
 
-% Prior work on splitting has been limited to simple, imperative languages; in
-% this paper, we extend splitting to \lang, a typed lambda calculus equipped with
-% a modal staging operator, allowing us to split recursive and higher-order
-% functions.
-
-% These new features allow us to express some familiar algorithms which, when
-% split, yield asymptotic improvements; for example, quickselect of the $k$th
-% smallest element of a list $l$ splits into (1) sorting $l$ into a binary search
-% tree, then (2) finding the tree's $k$th leftmost element.
-
-% % something about compositional, cross-cutting code
+We have implemented the algorithm in a prototype compiler,
+and we show that the ability to split a language with recursion and higher-order features
+can enabled non-trivial algorithmic transformations that asymptotically improve code efficiency by nearly a linear factor,
+as well as modular expression of staged programming,

### 2015-07-10 paper/splitting.tex
abstract draft

@@
 
-Recall that this annotation is not especially useful 
-for ordinary staged term evaluation (\ref{sec:semantics}).
-We suspect the key factor is that splitting is a static transformation,
-since other static trasofrmations \cite{Mogensen89a,DeNiel91} 
-also benefited from a clear distinction between ground and mixed terms.
-Ground is also similar to the validity mechanism in ML5 \cite{murphy05}.
+Since direct staged term evaluation (\ref{sec:semantics})
+reduces all first stage terms to value forms without any remaining stage two work,
+the distinction between \bbonem\ and \bbonep\ is unnecessary.
+In contrast, when performing program transformations before the first stage inputs are known,
+it is valuable to form a clear distinction between ground and mixed terms.
+This was similarly observed in prior work seeking to implement self-applicable partial evaluators\,\cite{Mogensen89a,Mogensen89b}.
+While this paper assumes that ground annotations already exist in the input,
+it may be possible to use binding-time analysis techniques to automatically insert them.
+Ground is also similar to the validity mechanism in ML5\,\cite{murphy05}.
 \end{abstrsyn}

### 2015-07-10 paper/relatedwork.tex
related work

@@
 
+\TODO move this
+
+In the partial evaluation community, metastatic staging transforms (called \emph{program bifurcation} and \emph{binding time separation}) 
+similar to splitting have been investigated 
+as a preprocessing step necessary for simplifying the specialization phase of self-applied partial evaluators. 
+For instance, Mogensen\,\cite{Mogensen89a} and Niel et al.\,\cite{DeNiel91} defined transformations for a set of first-order recursive functions.
+
+Those works operate on the results of binding time analysis
+
+and so much of their binding-time specification just amounts to a type system.
+Formulations like \lang\ (namely the two-level system of~\cite{NN92-twolevel}) were known at the time,
+but~\cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
+
+Davies essentially solved this limitation in~\cite{davies96}
+
+Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b}.
+
 \paragraph{Stage Splitting.}
@@
 Algorithms for stage splitting have appeared in the literature under the names 
-{\em pass separation}, {\em binding time separation}, {\em data specialization}, 
-{\em program bifurcation}, and as various unnamed compiler passes.
-
-In the partial evaluation community, stage splitting has been investigated 
-mostly as a preprocessing step for self-applied partial evaluators. 
-For instance,~\cite{Mogensen89a} and~\cite{DeNiel91} defined splitting algorithms
-for programs consisting of a set of first-order recursive functions.
-Those works use a grammar-based binding-time specification that's 
-relatively complicated compared to the type system of \lang.
-This complication mostly arises from the decision to start with an untyped language:
-one must first find a way to say ``argument $x$ will be a tuple'' before one can say
-``the first component of $x$ is available at the second stage,'' 
-and so much of their binding-time specification just amounts to a type system.
-Formulations more like \lang\ (namely the two-level system of~\cite{NN92-twolevel}) were known at the time,
-but~\cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
-Davies essentially solved this limitation in~\cite{davies96} (pending our further tweak with $\curr$),
-and the question of stage splitting as we have presented it has been ripe ever since.
-Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b},
-but the implications for splitting were not explicitly considered.
+{\em pass separation}\,\cite{JS86-staging} and {\em data specialization}\,\cite{knoblock96}.
 

### 2015-07-10 paper/abstract.tex
Make some notes on missing points.

@@
 
-Staged programming languages assign a stage to each expression and
-evaluate each expression in its assigned stage,
-and can be used to describe programs where inputs arrive at different times or rates.
+Staged programming languages assign a stage to each program expression
+and evaluate each expression in its assigned stage. A common use of
+staged languages is to describe programs where inputs arrive at
+different times or rates. In this paper we present an algorithm for
+statically {\em splitting} these mixed-staged programs into two
+unstaged, but dependent, programs where the outputs of the first
+program can be efficiently reused across multiple invocations of the
+second. While previous algorithms for performing this transformation
+(also called \emph{pass separation} and \emph{data specialization})
+were limited to operate on simpler, imperative languages, we define a
+splitting algorithm for an explicitly-two-stage, typed lambda
+calculus \lang\ with a $\fut$ modality denoting computation at a later
+stage, and a $\curr$ modality noting purely first stage code.  Most
+notably, the algorithm splits mixed-stage recursive and higher-order
+functions.  We have proven the dynamic correctness of our splitting
+algorithm with respect to a partial-evaluation semantics, and
+mechanized this proof in Twelf. We have also implemented the algorithm
+in a prototype compiler, and demonstrate that the ability to split
+programs in a language featuring recursion and higher-order features
+enables non-trivial algorithmic transformations that asymptotically
+improve code efficiency by nearly a linear factor, and also
+facilitates modular expression of staged programs.
 
-In this paper we present an 
-algorithm for statically {\em splitting} these mixed-staged programs 
-into two dependent, unstaged programs where the outputs of the first part can be efficiently reused accorss multiple invocations of the second program.
-
-While previous algorithms for performing this transformation 
-(also called \emph{pass separation} and \emph{data specialization}) operate on simpler imperative languages,
-out algorithm operates on an explicitly-two-staged, typed lambda calculus \lang, 
-with a $\fut$ modality denoting computation at a later stage,
-and a $\curr$ modality noting purely first stage code.
-Notably, our algorithm is the first to with support for mixed-stage recursive and higher-order functions.
-
-We have proven the dynamic correctness of our splitting algorithm with respect
-to a partial-evaluation semantics, and mechanized this proof in Twelf.
-
-We have implemented the algorithm in a prototype compiler,
-and we show that the ability to split a language with recursion and higher-order features
-can enabled non-trivial algorithmic transformations that asymptotically improve code efficiency by nearly a linear factor,
-as well as modular expression of staged programming,
+% todo: static transformation carried out prior to any of the programs
+% inputs being available, and thus has low runtime overhead (and,
+% unline partial evaluation, is applicable when dynamic code
+% generation is not possible)

### 2015-07-10 paper/relatedwork.tex
background

@@
 inputs~\cite{jones96}.  The roots of partial evaluation go back to
-Kleene, who in 1952 proved that a multivariate program (function)
-$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
-a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
-\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
+Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52}.
 %Kleene did not intend to improve efficiency, however.  Lombardi
@@
 as Scheme~\cite{OB91-Similix,Consel88-Schism}.
+
+\emph{Program bifurcation}\,\cite{DeNiel91} (also called \emph{binding time separation}\,\cite{Mogensen89a})
+is a metastatic staging transformations
+used as a preprocessing step to simplify the specialization phase of self-applied partial evaluators. 
+Bifurcation is similar to stage splitting in that it factors the input program
+into two versions based on the stages of the input.
+However, in bifurcation there is no building of data structures to be use in the resumer;
+instead, the second stage version of the code also has 
+access to the first stage inputs and just recomputes whatever it needs.
+That difference aside, the grammar-based binding-time specifications used in bifurcation
+were capable of descring data structures with purely-static, purely-dynamic, and mixed stage content,
+much like the type system of \lang.
 
@@
 
-\TODO move this
-
-In the partial evaluation community, metastatic staging transforms (called \emph{program bifurcation} and \emph{binding time separation}) 
-similar to splitting have been investigated 
-as a preprocessing step necessary for simplifying the specialization phase of self-applied partial evaluators. 
-For instance, Mogensen\,\cite{Mogensen89a} and Niel et al.\,\cite{DeNiel91} defined transformations for a set of first-order recursive functions.
-
-Those works operate on the results of binding time analysis
-
-and so much of their binding-time specification just amounts to a type system.
-Formulations like \lang\ (namely the two-level system of~\cite{NN92-twolevel}) were known at the time,
-but~\cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
-
-Davies essentially solved this limitation in~\cite{davies96}
-
-Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b}.
-
 \paragraph{Stage Splitting.}
@@
 {\em pass separation}\,\cite{JS86-staging} and {\em data specialization}\,\cite{knoblock96}.
+Perhaps the closest work to ours is the algorithm for data
+specialization given by Knoblock and Ruf~\cite{knoblock96}, which
+also seeks to statically split an explicitly staged program into two
+stages.  However, they only consider a simple first-order language and
+straight-line programs. Their work also treats all computations
+guarded by second-stage conditionals as second-stage computations,
+which would prevent optimization (via splitting) of programs such as
+quickselect.
 
-Splitting algorithms have also been a topic of interest in programming
-systems for computer graphics, where to achieve high performance,
-programs are typically manually separated into components by frequency
-of execution. (Indeed, separation into a pre-determined set of
-per-frequency modules: e.g., per frame, per pixel, per triangle, is
-\emph{required} by the execution model of modern graphics
-architectures.)  The software engineering challenges of modifying
-graphics programs written in a split form have spurred interest in
-deferring the task of splitting to a compiler\,\cite{Foley:2011}.
-While early work sought to determine the staging of programs
-automatically via binding time analysis\,\cite{knoblock96}, more
-recent designs, like \lang, require explicit staging by the
-programmer\,\cite{Proudfoot:2001,Foley:2011,He:2014}.  However,
-splitting algorithms developed in all of these efforts have been
-limited to simple, imperative graphics shading languages that lack
-even simple features such as functions.
+Splitting algorithms have also been a topic of interest in
+programming systems for computer graphics, where, to achieve
+high performance, programs are typically manually separated into
+components by frequency of execution (Indeed, separation into a pre-determined
+set of per-frequency modules is \emph{required} by the execution model
+of modern graphics architectures.)  The software engineering
+challenges of modifying multiple per-stage programs have led to
+suggestions of writing graphics programs in an explicitly-staged
+programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
+deferring the task of pass separation to the compiler. However, all
+prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
+been limited to simple, imperative languages.
 

### 2015-07-10 paper/examples.tex
Remove graphics from intro. Move to examples.

@@
 
-A major motivation for staged programming in computer graphics is to express
-logic related to a single application concern conveniently as a single staged
-function (referred to as a pipeline program)\,\cite{Foley:2011}, rather than
-as a collection of functions distributed across different GPU pipeline stages.
+Programs featuring staged structure are particularly important in real-time
+computer graphics since modern graphics architectures \emph{require} that
+graphics computations be organized into distinct {\em passes} which perform
+increasingly fine-grained computations (e.g., per object, per screen region,
+per pixel) where many computations in later passes use the
+results from a single computation in an earlier pass\,\cite{OpenGL4Spec}.
 
-Pipeline programs can be directly expressed as staged functions in \lang\, with splitting 
-used to generate monostaged code required for each GPU pipeline stage
-(as well as the plumbing to communicate values between the stages).
-Composition of such functions in \lang\ allows more complex pipelines to be expressed in a modular programming style
-that is becoming increasingly desirable in graphics shading languages\,\cite{Foley:2011,He:2014}.
-A simple example of composition of staged functions (representing GPU pipeline programs) is given below.
-The code defines a pipeline as a staged function that takes an object definition in the first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel.
-The code defines a pipeline combinator \texttt{shade} which modulates the results of the object's reflectance at a given pixel with the results of a function that computes the object's albedo (surface color).
+While performant, programming with explicit passes results in complex
+code where invariants must hold across different passes, and local
+changes for one pass may require changes to other passes to ``plumb''
+its inputs through all upstream passes. Thus, requiring manual
+separation of passes breaks key programming abstractions such as
+composition and modularity.  Graphics researchers therefore have
+suggested using explicitly staged programs
+\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express logic related to
+a single application concern conveniently as a single staged function
+(referred to as a pipeline program)\,\cite{Foley:2011}.
+
+Pipeline programs can be directly expressed as staged functions in
+\lang\ with splitting used to generate monostaged code required for
+each GPU pipeline stage. Composition of such functions in
+\lang\ allows more complex pipelines to be expressed in a modular
+programming style that is becoming increasingly desirable as graphics
+programs increase in complexity\,\cite{Foley:2011,He:2014}.
+
+A simple example of composition of staged functions (representing GPU
+pipeline programs) is given below.  The code defines a pipeline as a
+staged function that takes an object definition in the first stage
+(\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second
+stage and emits the color of the object at the specified pixel. The
+code defines a pipeline combinator \texttt{shade} which modulates the
+results of the object's reflectance at a given pixel with the results
+of a function that computes the object's albedo (surface color).
 %
@@
 %
-Most pipeline programming models either force the user to manually split their programs (which harms modularity).
-Others, such as \cite{Foley:2011}, offer only a few multistage types (namely, first order functions) and give 
-combinators like \texttt{par} as language features.
-In contrast, \lang\ is powerful enough to define those combinators in-language.
+Whereas prior systems provides combinators like \texttt{par} as
+language features\,\cite{Foley:2011} to support composition, \lang\ is
+powerful enough to define those combinators in-language.

### 2015-07-10 paper/intro.tex
Remove graphics from intro. Move to examples.

@@
 %
-In this common scenario, a multivariate function is applied many times with one
-argument fixed, so it is advantageous to find the computations in $f$ that
-depend on $x$ and perform them only once. This is the intent of compiler
-optimization techniques like loop hoisting and common subexpression elimination,
-which identify the computations common to all invocations of $f$, and perform
-them once at the beginning of the computation.
+In this scenario, a multivariate function is applied many times with
+one argument fixed, so it is advantageous to find the computations in
+$f$ that depend on $x$ and then stage program execution so they are
+performed only once at the beginning of execution.
 
-Programs featuring this structure are particularly important in real-time
-computer graphics; in fact, modern graphics architectures \emph{require} that
-graphics computations be organized into distinct {\em passes} which perform
-increasingly fine-grained computations (for example, per object, per region of
-the screen, and per pixel, where many computations in later passes use the
-results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
-While performant, this paradigm of programming with explicit passes results in
-complex code where invariants must hold across different passes, and local
-changes for one pass may require changes to other passes to ``plumb'' inputs
-from all other frequencies (\TODO explain what ``frequency'' means here)\,\cite{Foley:2011}.
-%
-In other words, requiring manual separation of passes breaks key programming
-abstractions such as composition and modularity.
-%
-Graphics researchers therefore have suggested using explicitly staged programs
-\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
-mechanical task of separating such programs into passes.
+%Programs featuring this structure are particularly important in real-time
+%computer graphics; in fact, modern graphics architectures \emph{require} that
+%graphics computations be organized into distinct {\em passes} which perform
+%increasingly fine-grained computations (for example, per object, per region of
+%the screen, and per pixel, where many computations in later passes use the
+%results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
+%While performant, this paradigm of programming with explicit passes results in
+%complex code where invariants must hold across different passes, and local
+%changes for one pass may require changes to other passes to ``plumb'' inputs
+%from all other frequencies (\TODO explain what ``frequency'' means here)\,\cite{Foley:2011}.
+%In other words, requiring manual separation of passes breaks key programming
+%abstractions such as composition and modularity.
+%Graphics researchers therefore have suggested using explicitly staged programs
+%\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
+%mechanical task of separating such programs into passes.
 
-Programming languages researchers have studied similar problems. In the 1980s,
-J{\o}rring and Scherlis identified {\em frequency reduction} and {\em
-precomputation}~\cite{JS86-staging} as mechanisms for efficiently evaluating
-staged computations. To perform frequency reduction, one identifies and hoists
-computations that are performed multiple times, in order to compute them only
-once. To perform precomputation, one identifies computations that can be
-performed in advance, and does so---for example, at compile time if the relevant
+%Programming languages researchers have studied similar problems, with
+%J{\o}rring and Scherlis identifying {\em frequency reduction} and {\em
+% precomputation} as mechanisms for efficiently evaluating staged
+%computations\,\cite{JS86-staging}.
+
+J{\o}rring and Scherlis classify automatic program staging
+transformations, such as the one described above, as forms of {\em
+  frequency reduction} or {\em
+  precomputation}\,\cite{JS86-staging}. To perform frequency
+reduction, one identifies and hoists computations that are performed
+multiple times, in order to compute them only once. To perform
+precomputation, one identifies computations that can be performed in
+advance, and does so---for example, at compile time if the relevant
 inputs are statically known.
 
-There has been much work on language-based techniques for performing
-precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
-precomputation by specializing functions to specific arguments. Going back to
-our example, if $f$ specializes to a particular $v$, written $f_v$, such that
-$f(v,y) = f_v(y)$, then $F$ can be specialized to $v$ as
+One common precomputation technique is partial
+evaluation\,\cite{futamura71,jones96}, which relies on dynamic
+compilation to specialize functions to known argument values. Going
+back to our example, if $f$ specializes to a particular $v$, written
+$f_v$, such that $f(v,y) = f_v(y)$, then $F$ can be specialized to $v$
+as
 \begin{lstlisting}
@@
 \end{lstlisting}
-eliminating the need to compute $m$ times those parts of $f(v,-)$ which do not
-depend on the second argument.
+eliminating the need to compute $m$ times those parts of $f(v,-)$
+which do not depend on the second argument.
 
-A closely related technique, {\em metaprogramming}, allows writing
-programs that generate specialized
-code~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
-enables fine-grained control over specialization by requiring \emph{staging
-annotations} that explicitly mark the stage of each computation. (In
-partial evaluation, we typically assume that a \emph{binding-time
-  analysis} has automatically determined which parts of $f$ can be
-specialized.)
+Closely related to partial evaluation is {\em metaprogramming}, where
+known values represent program code to be executed in a later stage
+\,\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming enables
+fine-grained control over specialization by requiring explicit
+\emph{staging annotations} that mark the stage of each expression.
 
-While partial evaluation and metaprogramming have been studied extensively,
-frequency reduction techniques have not seen the same attention. In
-\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
-manual method for {\em splitting} a program into multiple passes.
-%
-In our example, pass separation would split the function $f$ into two passes
-$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
-evaluating the first pass on $x$, and using the result $z$ to evaluate the
-second pass on each $y_i$.
+%(In partial evaluation, we typically assume that a
+%\emph{binding-time analysis} has automatically determined which parts
+%of $f$ can be specialized.)
+
+%This is the intent of compiler optimization techniques like loop
+%hoisting and common subexpression elimination, which identify the
+%computations common to all invocations of $f$, and perform them once
+%at the beginning of the computation.
+
+%While partial evaluation and metaprogramming have been studied
+%extensively, frequency reduction techniques have not seen the same
+%attention.
+
+While simple forms of frequency reduction include standard compiler
+optimizations such as loop hoisting and common subexpression
+elimination, J{\o}rring and Scherlis proposed the more general
+transformation of {\em splitting} a program into multiple passes
+(called pass separation in \,\cite{JS86-staging}). In our example,
+splitting transforms the function $f$ into two passes $f_1$ and $f_2$
+such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
+evaluating the first pass on $x$, and using the result $z$ to evaluate
+the second pass on each $y_i$.
 %
@@
 %
-The key difference between pass separation and partial evaluation (or
+The key difference between splitting and partial evaluation (or
 metaprogramming) is that the former can be performed without access to
 the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
-while $F_v$ is defined only for $x=v$.  Note that unlike
-precomputation, frequency reduction does not require knowledge of
-arguments; however, if a particular $v$ is known, we can use partial evaluation
-to specialize $F_\textrm{multipass}$ to that $v$.
+while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
+evaluation, splitting is a static program transformation
+(``metastatic'' in parial evaluation terms) and does not require
+dynamic code generation.
 
-Prior work on partial evaluation and metaprogramming has shown that
-precomputation can largely be automated for functional languages. In contrast,
-automatic pass separation has been limited to simple languages that do not
-support recursion or first-class
-functions~\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In this paper, we
-extend this work by presenting a pass separation algorithm for a typed lambda
-calculus.
+% Note that unlike precomputation, frequency reduction does not require
+%knowledge of arguments; however, if a particular $v$ is known, we can
+%use partial evaluation to specialize $F_\textrm{multipass}$ to that
+%$v$.
 
-As our starting point, we consider an explicitly-two-staged, typed lambda
-calculus \lang\ in the style of Davies~\cite{davies96}, with a $\fut$ modality
-denoting computation in a later stage (\ref{sec:semantics}). We then present a
-splitting algorithm to pass-separate a \lang\ program into two passes
-corresponding to the stages of computation in that program
-(\ref{sec:splitting}). These passes are monostaged programs, expressed in a
-conventional functional language.
+Prior work on partial evaluation and metaprogramming has demonstrated
+automatic application of these techniques on higher order functional
+languages. In contrast, automatic splitting transforms have been
+limited to simple languages that do not support recursion or
+first-class
+functions\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In
+this paper, we extend this work by presenting a splitting algorithm
+for a typed lambda calculus.
 
-The most interesting case involves splitting a mixed-stage function (like $f$ in
-the example) which is moreover recursive. This yields a first pass function
-$f_1$ which computes from its first-stage argument $x$ a \emph{boundary} data
-structure encoding the interstage communication in $f(x,-)$. When $f$ recurs on
-$x$, this boundary structure itself has a recursive type.
-%
-The second pass is a function $f_2$ which takes the boundary data structure
-$f_1(x)$ and a second-stage argument $y_i$, and completes the computation
-$f(x,y_i)$, traversing the boundary structure in light of the particular
-argument $y_i$.
+As our starting point, we consider an explicitly-two-staged, typed
+lambda calculus \lang\ in the style of Davies\,\cite{davies96}, with a
+$\fut$ modality denoting computation in a later stage
+(\ref{sec:semantics}). We then present a splitting algorithm that
+transforms \lang\ program into two passes corresponding to the stages
+of computation in that program (\ref{sec:splitting}). These passes are
+monostaged programs, expressed in a conventional functional language.
 
-We have implemented this splitting algorithm (\ref{sec:implementation}) and
-applied it to a number of staged programs (\ref{sec:examples}) ranging from
-straight-line arithmetic operations to recursive and higher-order functions. In
-the case of the quickselect algorithm, which we discuss next, the split code is
-in fact asymptotically faster over many $y_i$.
+The most interesting case involves splitting a recursive mixed-stage
+function (like $f$ in the example). This yields a first pass function
+$f_1$ which computes from its first-stage argument $x$ a
+\emph{boundary} data structure encoding the interstage communication
+in $f(x,-)$. When $f$ recurs on $x$, this boundary structure itself
+has a recursive type.  The second pass is a function $f_2$ which takes
+the boundary data structure $f_1(x)$ and a second-stage argument
+$y_i$, and completes the computation $f(x,y_i)$, traversing the
+boundary structure in light of the particular argument $y_i$.
 
-%The crux of the splitting algorithm is splitting recursive mixed stage
-%functions that involve both a first stage and second stage computations.  When
-%split, such a function turns into a first-pass function that yields a {\em
-%boundary value} that encodes the ``first-stage'' part of the computation.  If
-%the function recurs on a first-stage argument, then the boundary value is
-%typically a recursive data structure.  In other words, the splitting algorithm
-%maps computational recursion into a recursive data type. If the boundary
-%structure is generated as a result of a conditional, it is typically encoded as
-%a (recursive) sum type, such as a tree, where the kind of the node indicates
-%the recursion status of the first-stage evaluation.  The second-pass part of
-%the computation, which corresponds to the ``second stage'' takes the boundary
-%value as an argument as well as the second-stage argument and uses them to
-%complete the computation, often by traversing the now precomputed boundary in
-%light of the now available second-stage argument.
+We have implemented this splitting algorithm
+(\ref{sec:implementation}) and applied it to a number of staged
+programs (\ref{sec:examples}) ranging from straight-line arithmetic
+operations to recursive and higher-order functions. In the case of the
+quickselect algorithm, which we discuss next, the split code is in
+fact asymptotically faster over many $y_i$.

### 2015-07-10 paper/relatedwork.tex
Remove graphics from intro. Move to examples.

@@
 That difference aside, the grammar-based binding-time specifications used in bifurcation
-were capable of descring data structures with purely-static, purely-dynamic, and mixed stage content,
+were capable of describing data structures with purely-static, purely-dynamic, and mixed stage content,
 much like the type system of \lang.
@@
 
-Algorithms for stage splitting have appeared in the literature under the names 
-{\em pass separation}\,\cite{JS86-staging} and {\em data specialization}\,\cite{knoblock96}.
-Perhaps the closest work to ours is the algorithm for data
-specialization given by Knoblock and Ruf~\cite{knoblock96}, which
-also seeks to statically split an explicitly staged program into two
-stages.  However, they only consider a simple first-order language and
-straight-line programs. Their work also treats all computations
-guarded by second-stage conditionals as second-stage computations,
-which would prevent optimization (via splitting) of programs such as
-quickselect.
+Algorithms for stage splitting have appeared in the literature under
+the names {\em pass separation}\,\cite{JS86-staging} and {\em data
+  specialization}\,\cite{knoblock96}.  Perhaps the closest work to
+ours is the algorithm for data specialization given by Knoblock and
+Ruf~\cite{knoblock96}, which also seeks to statically split an
+explicitly staged program into two stages.  However, they only
+consider a simple first-order language and straight-line
+programs. Their work also treats all computations guarded by
+second-stage conditionals as second-stage computations, which would
+prevent optimization (via splitting) of programs such as quickselect.
 
-Splitting algorithms have also been a topic of interest in
-programming systems for computer graphics, where, to achieve
-high performance, programs are typically manually separated into
-components by frequency of execution (Indeed, separation into a pre-determined
-set of per-frequency modules is \emph{required} by the execution model
-of modern graphics architectures.)  The software engineering
-challenges of modifying multiple per-stage programs have led to
-suggestions of writing graphics programs in an explicitly-staged
-programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
-deferring the task of pass separation to the compiler. However, all
-prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
-been limited to simple, imperative languages.
+Splitting algorithms have also been a topic of interest in programming
+systems for computer graphics, where, to achieve high performance,
+programs are typically manually separated into components by frequency
+of execution corresponding to support graphics hardware pipeline
+stages. The software engineering challenges of modifying multiple
+per-stage programs have led to suggestions of writing graphics
+programs in an explicitly-staged programming
+language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and deferring the
+task of pass separation to the compiler. However, all prior efforts,
+like that of Knoblock and Ruf\,\cite{knoblock96} have been limited to
+simple, imperative languages.
 

### 2015-07-10 paper/examples.tex
Fix a bunch of references to "passes" in the graphics section.

@@
 
-Programs featuring staged structure are particularly important in real-time
-computer graphics since modern graphics architectures \emph{require} that
-graphics computations be organized into distinct {\em passes} which perform
-increasingly fine-grained computations (e.g., per object, per screen region,
-per pixel) where many computations in later passes use the
-results from a single computation in an earlier pass\,\cite{OpenGL4Spec}.
+Programs featuring staged structure are particularly important in
+real-time computer graphics since modern graphics architectures
+\emph{require} that graphics computations be organized into a pipeline
+of stages which perform increasingly fine-grained computations (e.g.,
+per object, per screen region, per pixel). Many computations in later
+stages use the results from a single computation in an earlier
+stage\,\cite{OpenGL4Spec}.
 
-While performant, programming with explicit passes results in complex
-code where invariants must hold across different passes, and local
-changes for one pass may require changes to other passes to ``plumb''
-its inputs through all upstream passes. Thus, requiring manual
-separation of passes breaks key programming abstractions such as
-composition and modularity.  Graphics researchers therefore have
-suggested using explicitly staged programs
+While performant, manually partitioning programs by stage results in
+complex code where invariants must hold across different stages, and
+local changes to the logic of one stage may require changes to that of
+upstream stages other stages to ``plumb'' its inputs through the
+pipeline.  Thus, manual separation of programs by stage breaks key
+programming abstractions such as composition and modularity.  Graphics
+researchers therefore have suggested using mixed-staged programs
 \,\cite{Proudfoot:2001,Foley:2011,He:2014} to express logic related to

### 2015-07-11 paper/relatedwork.tex
Pound on program bifurcation really hard in the related work.

@@
 %
-The idea behind frequency
-reduction is to identify computations that are performed multiple
-times and pull them ahead so that they can be performed once and used
-later as needed.  Dynamic programming, loop hoisting, and pass
-separation (presented here) are all examples of frequency reduction
-techniques.
+The idea behind frequency reduction is to identify computations that
+are performed multiple times and pull them ahead so that they can be
+performed once and used later as needed.  Dynamic programming, loop
+hoisting, and splitting (presented here) are all examples of frequency
+reduction techniques.
 
 \paragraph{Precomputation techniques.}
-Perhaps one of the most studied examples of precomputation is
-partial evaluation.  Partial evaluation distinguishes between static
-(compile-time) and dynamic (runtime) stages. Given a program and values 
-for all static inputs, partial evaluation generates a specialized program by
+Perhaps one of the most studied examples of precomputation is partial
+evaluation, which distinguishes between static (compile-time) and
+dynamic (runtime) stages. Given a program and values for all static
+inputs, partial evaluation generates a specialized program by
 performing computations that depend only on the static
 inputs~\cite{jones96}.  The roots of partial evaluation go back to
-Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52}.
+Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52},
 %Kleene did not intend to improve efficiency, however.  Lombardi
@@
 %incremental computation~\cite{Lombardi67}.
-There has since been an enormous body of work on partial evaluation,
+%There has since been an enormous body of work on partial evaluation,
 and we refer the reader to the book by Jones, Gommard, and Sestoft for
-a comprehensive discussion of work until the early 90's~\cite{JGS93}.
+a comprehensive discussion of partial evaluation work until the early
+90's~\cite{JGS93}.
 
@@
 
-\emph{Program bifurcation}\,\cite{DeNiel91} (also called \emph{binding time separation}\,\cite{Mogensen89a})
-is a metastatic staging transformations
-used as a preprocessing step to simplify the specialization phase of self-applied partial evaluators. 
-Bifurcation is similar to stage splitting in that it factors the input program
-into two versions based on the stages of the input.
-However, in bifurcation there is no building of data structures to be use in the resumer;
-instead, the second stage version of the code also has 
-access to the first stage inputs and just recomputes whatever it needs.
-That difference aside, the grammar-based binding-time specifications used in bifurcation
-were capable of describing data structures with purely-static, purely-dynamic, and mixed stage content,
-much like the type system of \lang.
+In an attempt to reduce the complexity of algorithms for
+specialization used in self-applicable partial evaluators, researchers
+explored the use of staging transformations that, similar to
+splitting, partition an input program into two programs: one that
+produces static outputs from static inputs, and a second that produces
+dynamic outputs from both static and dynamic inputs.  These metastatic
+transformations, referred to as \emph{program
+  bifurcation}\,\cite{DeNiel91} or as \emph{binding time
+  separation}\,\cite{Mogensen89a}), were successfully applied to
+functional languages that supported
+recursion\,\cite{DeNiel91}. However, unlike splitting, where the goal
+is to emit code where first stage results are computed once and then
+reused in multiple invocations of second stage execution, program
+bifurcation only serves as a preprocessing step for the subsequent
+specialization phase of partial evaluation. The second function
+emitted by program bifurcation has access to the first stage inputs
+and recomputes first stage computations it needs. (As noted by
+Knoblock and Ruf\,\cite{knoblock96}, it may be possible to modify the
+program bifurcation algorithm to cache and reuse the intermediate
+results, but this was never attempted.) The grammar-based binding-time
+specifications used in program bifurcation were capable of describing
+data structures with purely-static, purely-dynamic, and mixed stage
+content, much like the type system of \lang.
 
@@
 that, when evaluated at a stage, yield a new program to be evaluated
-at the next stage.
-%J{\o}rring and
-%Scherlis describe the idea of meta programming as one of the three possible approaches to
-%taking advantage of staging~\cite{JS86-staging}.
-Davies~\cite{davies96} presented a
-logical construction of binding-time type systems by deriving a type
-system via the Curry-Howard correspondence applied to temporal logic.
-Davies and Pfenning proposed a new type system for staged computation
-based on a particular fragment of modal logic~\cite{DP01-modal}. The
-work on MetaML extended type-based techniques to a full-scale language
-by developing a statically typed programming language based on ML that
+at the next stage.  Davies~\cite{davies96} presented a logical
+construction of binding-time type systems by deriving a type system
+via the Curry-Howard correspondence applied to temporal logic.  Davies
+and Pfenning proposed a new type system for staged computation based
+on a particular fragment of modal logic~\cite{DP01-modal}. The work on
+MetaML extended type-based techniques to a full-scale language by
+developing a statically typed programming language based on ML that
 enables the programmer to express programs with multiple
@@
 
-Splitting algorithms have also been a topic of interest in programming
-systems for computer graphics, where, to achieve high performance,
-programs are typically manually separated into components by frequency
-of execution corresponding to support graphics hardware pipeline
-stages. The software engineering challenges of modifying multiple
-per-stage programs have led to suggestions of writing graphics
-programs in an explicitly-staged programming
-language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and deferring the
-task of pass separation to the compiler. However, all prior efforts,
-like that of Knoblock and Ruf\,\cite{knoblock96} have been limited to
-simple, imperative languages.
-
-%% Online and offline.  Both approaches have their merits but online
-%% partial evaluation can be too slow and if stopped lead to an
-%% inefficient residual.  Offline partial evaluation is faster.  Also,
-%% full self-application has only been achieved by offline partial
-%% evaluation.
-
-%% Weaknesses of partial evaluation include~\cite{GJ05}:
-%% 1) speedups linear in the subject programs run time (excuse moi?)
-
-%% 2) good speedups require close knowledge of the program and some
-%% binding-time improvements.
-
-%% 3) result of specialization can be hard to predict as it can lead to
-%% slowdown, code explosion, and non-termination at specialization time.
-
-%% \paragraph{Meta programming}
-
-%% What happens when we try to write quickselect as a meta program?
-
-%% \ur{Umut: I am starting a paragraph on termination.  we might need to expand
-%%   on this.}
-%% \paragraph{Termination.}
-
-%% There are two reasons for non-termination: infinite loops and
-%% execution of errorneous code.  The splitting algorithm can cause the
-%% first.  But can it also cause the second? (If all computations really
-%% depend on stage 1 code, then would it be possible to for example see
-%% something like 1/0?  i think so but only if there is a meta dependency
-%% between the data in two stages?)
-
-%% Is in possible for the residual program to not terminate (unexpectedly)?
-
-%% Termination analysis in partial evaluation: see paper~\cite{AH96,GJ05}.
-
-%% Based on Glenstrup and Jones (pp. 1154, 1155), a partial evaluator
-%% should be complete. That is every computation depending on the static
-%% input should be reduced to a value (during specialization).
-%% Unfortunately, completeness is difficult to achieve if we also require
-%% that the specialization is always terminating.
-
-%% One problem with their description is that it is not clear which
-%% termination problem they are talking about termination of the residual
-%% or the specializer.
-
-%% Nontermination during specialization: 
-%% consider code
-
-%% double x = doubleplus (x,0)
-
-%% doubleplus (u,v) = if u <= 0 then v else doubleplus (u-1,v+2)
-
-%% this can lead to non-termination if v is considered static.
-
-%% Is this a problem for us?
+As noted in \ref{sec:graphics}, splitting algorithms have also been a
+topic of interest in programming systems for computer graphics, where,
+to achieve high performance, programs are manually separated into
+components by frequency of execution corresponding to graphics
+hardware pipeline stages. The software engineering challenges of
+modifying multiple per-stage programs have led to suggestions of
+writing graphics programs in an explicitly-staged programming
+language\,\cite{Proudfoot:2001,Foley:2011,He:2014} and deferring the
+task of pass separation to the compiler. However, all prior splitting
+efforts in computer graphics, like that of Knoblock and
+Ruf\,\cite{knoblock96} have been limited to simple, imperative
+languages.

### 2015-07-11 paper/relatedwork.tex
Clarity by placing two sentances next to each otehr to contrast splitting and prog bifur.

@@
 
-In an attempt to reduce the complexity of algorithms for
-specialization used in self-applicable partial evaluators, researchers
-explored the use of staging transformations that, similar to
-splitting, partition an input program into two programs: one that
-produces static outputs from static inputs, and a second that produces
-dynamic outputs from both static and dynamic inputs.  These metastatic
-transformations, referred to as \emph{program
-  bifurcation}\,\cite{DeNiel91} or as \emph{binding time
-  separation}\,\cite{Mogensen89a}), were successfully applied to
-functional languages that supported
-recursion\,\cite{DeNiel91}. However, unlike splitting, where the goal
-is to emit code where first stage results are computed once and then
-reused in multiple invocations of second stage execution, program
-bifurcation only serves as a preprocessing step for the subsequent
-specialization phase of partial evaluation. The second function
-emitted by program bifurcation has access to the first stage inputs
-and recomputes first stage computations it needs. (As noted by
-Knoblock and Ruf\,\cite{knoblock96}, it may be possible to modify the
-program bifurcation algorithm to cache and reuse the intermediate
-results, but this was never attempted.) The grammar-based binding-time
+In an attempt to reduce the complexity of the specialization phase of
+self-applicable partial evaluators, researchers explored the use of
+staging transformations that, similar to splitting, partition an input
+program into two programs: one that produces static outputs from
+static inputs, and a second that produces dynamic outputs from both
+static and dynamic inputs.  These metastatic transformations, referred
+to as \emph{program bifurcation}\,\cite{DeNiel91} or as \emph{binding
+  time separation}\,\cite{Mogensen89a}), serve as a preprocessing step
+to specialization and were successfully applied to functional
+languages that supported recursion\,\cite{DeNiel91}. However, unlike
+splitting, where the goal is to emit code where first stage results
+are computed once and then reused in multiple invocations of second
+stage execution, the second function produced by program bifurcation
+does not use the results of the first; Instead,
+%only serves as a preprocessing step for the subsequent
+%specialization phase of partial evaluation.
+it has access to the first stage inputs and recomputes all required
+first stage computations. (As noted by Knoblock and
+Ruf\,\cite{knoblock96}, it may be possible to modify the program
+bifurcation algorithm to cache and reuse the intermediate results, but
+this was never attempted.) The grammar-based binding-time
 specifications used in program bifurcation were capable of describing

### 2015-07-11 paper/relatedwork.tex
similar to --> much like

@@
 self-applicable partial evaluators, researchers explored the use of
-staging transformations that, similar to splitting, partition an input
+staging transformations that, much like splitting, partition an input
 program into two programs: one that produces static outputs from

### 2015-07-11 paper/intro.tex
fiddling with figures

@@
 elimination, J{\o}rring and Scherlis proposed the more general
-transformation of {\em splitting} a program into multiple passes
+transformation of {\em splitting} a program into multiple subfunctions
 (called pass separation in \,\cite{JS86-staging}). In our example,
-splitting transforms the function $f$ into two passes $f_1$ and $f_2$
+splitting transforms the function $f$ into two others $f_1$ and $f_2$
 such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
-evaluating the first pass on $x$, and using the result $z$ to evaluate
-the second pass on each $y_i$.
+evaluating the first function on $x$, and using the result $z$ to evaluate
+the second function on each $y_i$.
 %
@@
 evaluation, splitting is a static program transformation
-(``metastatic'' in parial evaluation terms) and does not require
+(``metastatic'' in partial evaluation terms) and does not require
 dynamic code generation.
@@
 languages. In contrast, automatic splitting transforms have been
-limited to simple languages that do not support recursion or
-first-class
-functions\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In
-this paper, we extend this work by presenting a splitting algorithm
-for a typed lambda calculus.
+limited to simple languages\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.
 
-As our starting point, we consider an explicitly-two-staged, typed
-lambda calculus \lang\ in the style of Davies\,\cite{davies96}, with a
-$\fut$ modality denoting computation in a later stage
-(\ref{sec:semantics}). We then present a splitting algorithm that
-transforms \lang\ program into two passes corresponding to the stages
-of computation in that program (\ref{sec:splitting}). These passes are
-monostaged programs, expressed in a conventional functional language.
+In this paper, we present a splitting algorithm for \lang, 
+a two-staged typed lambda calculus in the style of Davies\,\cite{davies96},
+with support for recursion and first-class functions.
+Like Davies, \lang\ uses a $\fut$ modality to denote computation in the second stage,
+but to aid splitting we also add a $\curr$ modality to denote purely first-stage computations. 
+The dynamic semantics (\ref{sec:semantics}) of \lang\ are that of Davies, 
+modified to provide an eager behavior between stages, which is more intuitive in the context of splitting.
+We then prove the correctness of our splitting algorithm (\ref{sec:splitting}) for \lang\ with respect to the semantics.
+Finally, we discuss our implementation of this splitting algorithm
+(\ref{sec:implementation}) and demonstrate its power and behavior
+for a number of staged programs ranging from straight-line arithmetic
+operations to recursive and higher-order functions (\ref{sec:examples}).
 
-The most interesting case involves splitting a recursive mixed-stage
-function (like $f$ in the example). This yields a first pass function
-$f_1$ which computes from its first-stage argument $x$ a
-\emph{boundary} data structure encoding the interstage communication
-in $f(x,-)$. When $f$ recurs on $x$, this boundary structure itself
-has a recursive type.  The second pass is a function $f_2$ which takes
-the boundary data structure $f_1(x)$ and a second-stage argument
-$y_i$, and completes the computation $f(x,y_i)$, traversing the
-boundary structure in light of the particular argument $y_i$.
-
-We have implemented this splitting algorithm
-(\ref{sec:implementation}) and applied it to a number of staged
-programs (\ref{sec:examples}) ranging from straight-line arithmetic
-operations to recursive and higher-order functions. In the case of the
-quickselect algorithm, which we discuss next, the split code is in
-fact asymptotically faster over many $y_i$.
+We demonstrate that splitting a recursive mixed-stage $f$
+yields an $f_1$ which computes a recursive data structure
+and an $f_2$ which traverses that structure
+in light of information available at the second stage.
+In the case of the quickselect algorithm, which we discuss next, 
+the split code executes asymptotically faster than a unstaged evaluation of $f$.

### 2015-07-11 paper/paper.tex
various edits
attempting to lift some more discussion into 4.0
related work
fiddling with figures

@@
 \crefname{theorem}{Theorem}{Theorems}
+\crefname{lemma}{Lemma}{Lemmas}
 
@@
 
-\title{Automatically Splitting a Staged Lambda Calculus}
+\title{Automatically Splitting a Two-Stage Lambda Calculus}
 \subtitle{(Supplementary Material in Appendix)}
 
-\authorinfo{Anon A. Anon}%Nicolas Feltman \and Carlo Angiuli \and Umut A. Acar \and Kayvon Fatahalian}
-           {Anonymous Institute}%Carnegie Mellon University}
-           {anon@anon.com}%\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}
+\authorinfo{~}%Nicolas Feltman \and Carlo Angiuli \and Umut A. Acar \and Kayvon Fatahalian}
+           {~}%Carnegie Mellon University}
+           {~}%\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}
 

### 2015-07-11 paper/semantics.tex
fiddling with figures

@@
 The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
-This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
+This strong requirement ensures that second-stage computations are not duplicated 
+when partial values are substituted for a variable.
 This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
@@
 These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
-In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
+In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
 where it is reified into a {\tt let} statement.
@@
 
-The statics and dynamics of \lang\ are related by a type safety theorems
-in figure \ref{fig:typeSafety}, again annotated by world.
+The statics and dynamics of \lang\ are related by the type safety theorems
+in \ref{fig:typeSafety}, again annotated by world.
 In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,

### 2015-07-11 paper/splitting.tex
fiddling with figures

@@
 The goal of a stage splitting translation 
-is to a send program $\mathcal P$ in a multistage language
+is to send a program $\mathcal P$ in a multistage language
 to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
@@
 
-In order to prove that \ref{lem:twoCorrect},
+In order to prove \ref{lem:twoCorrect},
 we will need to state analogous version for steps at \bbonem,
-since the various kinds multistage term in \lang\ are mutually dependent.
+since the various kinds of multistage term in \lang\ are mutually dependent.
 Thus, this section proceeds by covering the definition of splitting at \bbonem, 
@@
 \end{lstlisting}
-Finally, partial value splitting drops \texttt{gr} annotations yielding:
+Finally, partial value splitting drops \texttt{gr} annotations, yielding:
 \begin{lstlisting}

### 2015-07-11 paper/overview.tex
fixed a TODO

@@
 \texttt{qs} partitions \texttt{l} using the first element as
-a pivot, and then recurs on one of the two resulting sides, depending on
+a pivot and then recurs on one of the two resulting sides, depending on
 the relationship of \texttt{k} to the size \texttt{i} of the first half, in
@@
 A first cut at a solution might be to, in a first pass, sort \texttt{l} into
-\texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
+\texttt{s} and then in a second pass, perform lookups in \texttt{s}.
 Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
@@
 A more sophisticated solution is to, in a first pass, construct a binary search
-tree out of \texttt{l}, and in a second pass, simply look for the \texttt{k}th
+tree out of \texttt{l} and in a second pass, simply look for the \texttt{k}th
 leftmost element of that tree. To do this efficiently requires one more
@@
 a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
-particular, its first pass (\texttt{qs1}) constructs such a binary search tree,
+particular, its first pass (\texttt{qs1}) constructs such a binary search tree
 and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
@@
 computations, gathering them into \texttt{qs2}. This pass is given the boundary
-data structure and the rank \texttt{k}, and finishes the computation. Now that
+data structure and the rank \texttt{k}, and it finishes the computation. Now that
 \texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,

### 2015-07-11 paper/semantics.tex
fixed a TODO

@@
 This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
-whereas our applications feel more natural with a eager interpretation of {\tt next}.
+whereas our applications feel more natural with an eager interpretation of {\tt next}.
 One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
@@
 
-The multistage function \texttt{qss} from \ref{sec:overview} has a
-type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
+Multistage functions, such as \texttt{qss} from \ref{sec:overview},
+can be represented as terms with a type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
 \footnote{We can rewrite \texttt{qss} in this curried form, or apply a
 higher-order currying function.} 
-To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
+To apply such a function $f$ to arguments $\colmix a A$ and $\coltwo b B$, simply
 evaluate the program:
@@
 \begin{lstlisting}
-2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
+2`let fun qs k = 
+      prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
 2`in (qs k1,...,qs km)`
 \end{lstlisting}
-\TODO what?
 To see that the partial application of \texttt{qss} is only evaluated once,
@@
 \begin{lstlisting}
-2`let val y = ...`
-2`    val qs = y`
+2`let fun qs k = 
+	  let val y = ... in y
 2`in (qs k1,...,qs km)`
 \end{lstlisting}
-where $y$ is bound to a function containing no first-stage redexes.
+where $y$ is bound to a large nest of case statements open on $k$.

### 2015-07-11 paper/conclusion.tex
reviewer comments
compressed figures

--- 2015-02-27 paper/conclusion.tex
@@
+%!TEX root = paper.tex
+
 \section{Conclusion}
 
-This paper presents techniques to translate staged lambda calculus
-programs to multi-pass programs, which evaluate in a sequence of
-passes, each of which is a conventional program.  We present a
-formalization of the techniques in a way that can be implemented in
-practice.  While we have also provided an implementation that
-successfully splits the examples considered, we have stopped short of
-offering a full proof of correctness, which we leave for future work.
-We are also interested in providing a more complete implementation and
-investigating how these techniques can be integrated into existing
-practical systems such as existing staged programming systems for
-computer graphics.  On the more conceptual side, we plan to extend the
-techniques to richer source languages, for example those that include
-parallelism.
+This paper presents a splitting algorithm for \lang, 
+a two-staged typed lambda calculus with support for recursion and first-class functions.
+The type system of \lang\ uses three worlds \bbonem, \bbonep, and \bbtwo\ to classify code by its stage,
+with the modalities $\fut$ and $\curr$ providing internalizations 
+of second-stage code and ground first-stage code to the mixed world.
+We present a dynamic semantics that evaluates terms in two passes
+and provides an eager interpretation of {\tt next} via hoisting steps.
+We prove the correctness of our splitting algorithm against the dynamic semantics,
+and implement this proof in Twelf.
+Finally, we discuss our implementation of \lang\ and its splitting algorithm
+and analyze their behavior for a number of staged programs,
+including those with higher-order and recursive functions.
+
+Looking forward, we are interested in investigating the behavior of splitting 
+in the presence of richer language features such as mutation, polymorphism,
+parallel constructs, and more than two stages.
+Also while we have proven the dynamic correctness of our splitting algorithm,
+we do not yet have a characterization of the types of the output
+or of the cost behavior of output terms.

### 2015-07-11 paper/abstract.tex
various small edits

@@
 notably, the algorithm splits mixed-stage recursive and higher-order
-functions.  We have proven the dynamic correctness of our splitting
+functions.  We prove the dynamic correctness of our splitting
 algorithm with respect to a partial-evaluation semantics, and
-mechanized this proof in Twelf. We have also implemented the algorithm
+mechanize this proof in Twelf. We also implement the algorithm
 in a prototype compiler, and demonstrate that the ability to split
@@
 enables non-trivial algorithmic transformations that asymptotically
-improve code efficiency by nearly a linear factor, and also
+improve code efficiency by nearly a linear factor and also
 facilitates modular expression of staged programs.

### 2015-07-11 paper/intro.tex
figure fiddling
various small edits

@@
 
+\input{figures/quickselect}
 Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
@@
 precomputation, one identifies computations that can be performed in
-advance, and does so---for example, at compile time if the relevant
+advance and does so---for example, at compile time if the relevant
 inputs are statically known.
@@
 such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
-evaluating the first function on $x$, and using the result $z$ to evaluate
+evaluating $f_1$ on $x$, and using the result $z$ to evaluate
 the second function on each $y_i$.
@@
 automatic application of these techniques on higher order functional
-languages. In contrast, automatic splitting transforms have been
-limited to simple languages\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.
+languages. In contrast, automatic splitting transformations have been
+limited to simpler languages\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.
 
@@
 
-We demonstrate that splitting a recursive mixed-stage $f$
+We also demonstrate that splitting a recursive mixed-stage $f$
 yields an $f_1$ which computes a recursive data structure
@@
 In the case of the quickselect algorithm, which we discuss next, 
-the split code executes asymptotically faster than a unstaged evaluation of $f$.
+the split code executes asymptotically faster than an unstaged evaluation of $f$.

### 2015-07-11 paper/overview.tex
figure fiddling
various small edits

@@
 \begin{abstrsyn}
-
-\input{figures/quickselect}
 
@@
 
-A first cut at a solution might be to, in a first pass, sort \texttt{l} into
-\texttt{s} and then in a second pass, perform lookups in \texttt{s}.
+A first cut at a solution might be to first sort \texttt{l} into
+\texttt{s} and then perform lookups in \texttt{s}.
 Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
@@
 
-A more sophisticated solution is to, in a first pass, construct a binary search
-tree out of \texttt{l} and in a second pass, simply look for the \texttt{k}th
+A more sophisticated solution is to construct a binary search
+tree out of \texttt{l} and then simply look for the \texttt{k}th
 leftmost element of that tree. To do this efficiently requires one more
@@
 
-In both cases, we replace our simple solution with a two-pass algorithm which
+In both cases, we replace our simple solution with a two-part algorithm which
 first processes the input into a lookup data structure, and then subsequently
@@
 In this paper, we present a splitting algorithm which automatically transforms
-staged programs into multi-pass programs. In certain cases, this algorithm even
+staged programs into multi-part programs. In certain cases, this algorithm even
 improves input algorithms in the way described above!
-
-%Such transformations, called pass-separation by J{\o}rring and
-%Scherlis are commonly employed.  For example, as briefly mentioned in
-%\ref{sec:intro}, modern graphics software is written exactly in this
-%way.  Unfortunately, as the example illustrates, multipass programs
-%are naturally complex.
 
@@
 The idea behind staged programming is to use staging annotations---in our case,
-guided by types---to indicate the stage of each subterm. We consider a
-language \lang\ with two modal worlds representing the two stages, and a third
-{\em ground} world of unstaged terms.
+guided by types---to indicate the stage of each subterm.
 
@@
 
-The process of automatically adding staging annotations to unstaged code has been
+The process of automatically adding staging annotations to unstaged code,
+called \emph{binding time analysis}, has been
 the subject of extensive research (\ref{sec:related}). In this paper, we do not
@@
 %
-Splitting \texttt{qss} yields a two-pass program that creates a
+Splitting \texttt{qss} yields a two-part program that creates a
 probabilistically balanced, augmented binary search tree (which is isomorphic to
 a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
-particular, its first pass (\texttt{qs1}) constructs such a binary search tree
-and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
+particular, its first part (\texttt{qs1}) constructs such a binary search tree
+and its second part (\texttt{qs2}) traverses the tree, using the embedded size
 information to find the element of the desired rank.
@@
 Our splitting algorithm scans \texttt{qss} for first-stage computations,
-gathering them into \texttt{qs1}. Given \texttt{l}, this pass performs these
-computations and places the information needed by the subsequent pass into a
+gathering them into \texttt{qs1}. Given \texttt{l}, this function performs these
+computations and places the information needed by the subsequent function into a
 boundary data structure.
@@
 Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
-computations, gathering them into \texttt{qs2}. This pass is given the boundary
+computations, gathering them into \texttt{qs2}. This function is given the boundary
 data structure and the rank \texttt{k}, and it finishes the computation. Now that

### 2015-07-11 paper/semantics.tex
various small edits

@@
 The central tenet of a staged language
-is that first stage code should be evaluated entirely before second stage code.
+is that first-stage code should be evaluated entirely before second-stage code.
 Accordingly, our dynamics operates in two passes.
@@
 The second pass further reduces this residual.
-Since a residual $q$ is monostage by definition, this second pass is standard unstaged evaluation
+Since $q$ is monostage by definition, this second pass is standard unstaged evaluation
 and is not described in further detail in this paper.
@@
 our dynamics requires notions of values and steps that are specialized to each world.
+The rules for all parts of first-pass evaluation are given in \ref{fig:diaSemantics}.
+In this and later figures, we extensively use an $\mathcal{S}[\text{-}]$ construction
+to indicate a \emph{shallow} evaluation context which looks a single level deep.
 
-\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
-Since first pass evaluation should not reduce stage two terms, this judgement does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
+\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgment \mbox{$\steptwo e {e'}$}.
+Since first pass evaluation should not reduce stage two terms, this judgment does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
 A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
@@
 \paragraph{World \bbonem.}
-Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
+Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
 It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
@@
 \end{lstlisting}
-This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
+This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
 These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
@@
 representing the second stage binders under which we are evaluating.
+Note how the progress theorem for world \bbonem\ states that every well-typed term must take 
+either a standard step or a hoisting step.
 

### 2015-07-30 sematics/boundary/boundary.tex
working on doc about typing boundary

@@
 \usepackage{mathpartir}
+\usepackage{bbm}
 \usepackage{mathabx}
@@
 
-\title{Boundary Types}
+\title{Typing the Outputs}
 \author{Nico}
@@
 \maketitle
+
+\begin{abstrsyn}
 
@@
 
+\section {Intro}
+
+The purpose of this document is to explain an approach to typing the outputs produced by our splitting algorithm.
+As we will see, this will require a dependent type system.
+The dependence comes in two flavors: dependence on the boundary, and dependence in partial value splitting.
+This first flavor has its origin in first-class functions, and the second flavor comes from mixed-stage sums.
+We will focus mostly on dependence on the boundary, and consider mixed-stage sums at the end.
+
 \section{Goals}
 
-The purpose of this document is to explain how to type the boundary produced by our splitting algorithm.
 We'll start with a very simple input language, the boundary of which can be typed in an easy way.
@@
 Ideally, the splitting rules will agree with the boundary type given to us by this judgment.
-That is, if $\btj e A \bbtwo \tau$ and $\splitone{e}{A}{p,l.r}$, 
+That is, if $\btj e A \bbtwo \tau $ and $\splittwo{e}{A}{p}{l}{r}$, 
 then $p:\tau$ and $l:\tau \vdash r:A$ (and a similar relationship at stage \bbone, covered later).
@@
 	&\btyw {e_2} B {\tau_2}} \and
-\infer {\btyw {\pio~e} A \tau} 
+\infer {\btyw {\pio e} A \tau} 
 	{\btyw {e} {A \times B} {\tau}} \and
-\infer {\btyw {\pit~e} B \tau} 
+\infer {\btyw {\pit e} B \tau} 
 	{\btyw {e} {A \times B} {\tau}} \and
-\infer {\btyo {\next~e} {\fut A} \tau} 
+\infer {\btyo {\next e} {\fut A} \tau} 
 	{\btyt {e} A {\tau}} \and
-\infer {\btyt {\prev~e} A \tau} 
+\infer {\btyt {\prev e} A \tau} 
 	{\btyo {e} {\fut A} {\tau}} \and
-\infer {\btyt {\pause~e} \rmint {\tau \times \rmint}} 
+\infer {\btyt {\pause e} \rmint {\tau \times \rmint}} 
 	{\btyo e \rmint \tau}
@@
 
+\end{abstrsyn}
 \end{document}

### 2015-08-01 sematics/boundary/boundary.tex
working on boundary type stuffs

@@
 \usepackage[usenames,dvipsnames]{color}
+\usepackage{framed}
 
@@
 
+\newcommand {\colfun} [3] {{#1}:>{#2}~\myatsign~{#3}}
+
 \newcommand {\ibksym} {\rotatebox[origin=c]{270}{$\therefore$}}
@@
 \newcommand {\ibk} [2] {{#1}~\ibksym~{#2}}
-\newcommand {\btj} [4] {{#1}:{#2}~@~{#3}~\btsym~{#4}}
+\newcommand {\btj} [4] {{#1}:{#2}~\myatsign~{#3}~\btsym~{#4}}
 \newcommand {\btyo} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbone{#4}}
@@
 The purpose of this document is to explain an approach to typing the outputs produced by our splitting algorithm.
-As we will see, this will require a dependent type system.
-The dependence comes in two flavors: dependence on the boundary, and dependence in partial value splitting.
-This first flavor has its origin in first-class functions, and the second flavor comes from mixed-stage sums.
-We will focus mostly on dependence on the boundary, and consider mixed-stage sums at the end.
+As we will see, this will become somewhat convoluted, eventually requiring a dependent type system.
+In general, the problem of typing the output has two subparts:
+\begin{enumerate}
+\item \textbf{Typing the boundary.}  The most interesting and important contribution of this document is characterizing the 
+\emph{boundary type}, which describes the communication between the two stages.  
+Figuring out the boundary type is a self-contained problem, 
+in that the boundary type of a term depends only on the boundary types of its subterms.
 
-\section{Goals}
+\item \textbf{Typing the rest.}  With a good characterization of the boundary type,
+we can then consider the broader question of assigning types to the outputs of splitting.
+\end{enumerate}
 
-We'll start with a very simple input language, the boundary of which can be typed in an easy way.
-We'll then start adding features to the language, thereby inducing complexity in the boundary type system.
+For the most part, all of the complication lies in typing the boundary.
+Relative to that, the second part is straight-forward 
+(with the exception of mixed-stage sums, which add some details that are independent of the boundary type).
 
-For the most part, we'll do all of our work with a single new judgement $\btj e A w \tau$, 
+Most of the complication in boundary typing arises from two sources: divergence and higher-order functions,
+with the combination of these features introducing even further complexities.
+Accordingly, this document proceeds by first considering a fragment of the language with
+divergence but only first-order recursive functions (Section \ref{sec:firstorder}),
+before considering a language with order functions (Section \ref{sec:higherorder}),
+and the finally putting it all together.
+
+\section{Input Language Fragments}
+
+This section defines a few input language fragments for later reference.
+Those fragments include ``core'' features (\emph{i.e.} simple ones that present no difficulty for any typing system), 
+if expressions, first-order recursive functions, higher-order functions, and finally sum/recursive types.
+The two function fragments will be used exclusively.
+
+\begin{framed}
+\noindent\textbf{Core Features (Base Types, Products, Circles, Let):}
+\begin{mathpar}
+\infertypeswor [\rmunit] 	{\ty {\tup{}}\rmunit}								{\cdot} 																	\and
+\infertypeswor [int]		{\ty {i} \rmint}									{\cdot}																		\and
+\infertypeswor [hold]		{\typestwo {\pause e} \rmint}						{\typesone e \rmint}				 										\and
+\infertypeswor [int]		{\ty {b} \rmbool}									{\cdot}																		\and
+\infertypeswor [hold]		{\typestwo {\pause e} \rmbool}						{\typesone e \rmbool}				 										\and
+\infertypeswor [\times I]	{\ty {\tup{e_1,e_2}}{A\times B}}					{\ty {e_1} A & \ty {e_2} B} 												\and
+\infertypeswor [\times E_1]	{\ty {\pio e} A}									{\ty e {A\times B}} 														\and
+\infertypeswor [\times E_2]	{\ty {\pit e} B}									{\ty e {A\times B}} 														\and
+\infertypeswor [\fut I]		{\typesone {\next e}{\fut A}}						{\typestwo e A} 															\and
+\infertypeswor [\fut E]		{\typestwo {\prev e} A}								{\typesone e {\fut A}} 														\and
+\infertypeswor [\fut E]		{\ty {\letin x {e_1} {e_2}} B}						{\ty {e_1} A & \ty [,\col x A] {e_2} B } 									\and
+\end{mathpar}
+\end{framed}
+
+\begin{framed}
+\noindent\textbf{If Expressions:}
+\begin{mathpar}
+\infertypeswor [+ E]		{\ty {\ifthen{e_1}{e_2}{e_3}} A}					{\ty {e_1}\rmbool & \ty{e_2} A & \ty{e_3} A}
+\end{mathpar}
+\end{framed}
+
+We implement first order recursive functions by introducing a new typing judgement \mbox{$\colfun f {A\to B} w$},
+which says that $f$ is a function from $A$ to $B$ at world $w$.  
+We use a different typing judgement here in order to keep functions and other values distinct.
+We also allow let statements to work for second stage functions 
+(for reasons explained later, first stage function lets can be safely eliminated \emph{a priori} by substitution).
+Note how the only way to derive a function type is to directly declare a function, or to assume it.
+\begin{framed}
+\noindent\textbf{First-Order Recursive Functions:}
+\begin{mathpar}
+\infertypeswor [\to I]		{\Gamma \vdash \colfun {\fix fx{e_1}} {A \to B} w}	{\ty [,\colfun f {A \to B} w,\col x A] {e_1} B} 							\and
+%\infertypeswor [fix]		{\ty {\fix fxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
+\infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\Gamma \vdash \colfun {e_1} {A \to B} w & \ty {e_2} A} 					\and
+\infertypeswor [\fut E]		{\typestwo {\letin x {e_1} {e_2}} B}				{\Gamma \vdash \colfun {e_1} {A \to B} \bbtwo 
+																				&\typestwo [\Gamma,\colfun x {A \to B} \bbtwo] {e_2} B } 					\and
+\end{mathpar}
+\end{framed}
+
+\begin{framed}
+\noindent\textbf{Higher-Order Functions:}
+\begin{mathpar}
+\infertypeswor [\to I]		{\ty {\lam {x}{e}} {A \to B}}						{\ty [,\col x A] e B} 														\and
+%\infertypeswor [fix]		{\ty {\fix fxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
+\infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 										\and
+\end{mathpar}
+\end{framed}
+
+\begin{framed}
+\noindent\textbf{Sums and Recursion:}
+\begin{mathpar}
+\infertypeswor [+ I_1]		{\ty {\inl e} {A + B}}								{\ty e A} 																	\and
+\infertypeswor [+ I_2]		{\ty {\inr e} {A + B}}								{\ty e B} 																	\and
+\infertypeswor [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}			{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C}	\and
+\infertypeswor [\mu I]		{\ty {\roll e} {\mu \alpha.\tau}}					{\ty e {[\mualphatau / \alpha]\tau}} 										\and
+\infertypeswor [\mu E]		{\ty {\unroll e} {[\mualphatau / \alpha]\tau}}		{\ty e \mualphatau} 														\and
+\end{mathpar}
+\end{framed}
+
+\section{Values and Computations}
+
+...
+
+\section{First Order Recursive Functions}
+\label{sec:firstorder}
+
+For the most part, we'll do all of our work with a single new judgment \mbox{$\btj e A w \tau$}, 
 which indicates that the expression $e$ (which has type $A$ at world $w$) has the boundary type $\tau$.
-Ideally, the splitting rules will agree with the boundary type given to us by this judgment.
-That is, if $\btj e A \bbtwo \tau $ and $\splittwo{e}{A}{p}{l}{r}$, 
+The goal is for the splitting rule to agree with the boundary type given to us by this judgment.
+That is, if $\btj e A \bbtwo \tau$ and $\splittwo{e}{A}{p}{l}{r}$, 
 then $p:\tau$ and $l:\tau \vdash r:A$ (and a similar relationship at stage \bbone, covered later).
-Mind you, we don't actually bother checking that property in this document.
 
-\section{Straight-line Code}
+\subsection{Straight-line Code}
 
-We first consider an input language which has no functions or divergence.
+Consider an input language which has no functions or divergence.
 At either stage, the boundary type of any expression is 
@@
 The only exception to this pattern is the $\pause$ rule,
-which also multiplies an int into the boundary type.
+which also multiplies an integer into the boundary type.
 
@@
 \infer {\btyw b {\rm bool}\rmunit} {\cdot} \and
-\infer {\btyw {(e_1,e_2)} {A \times B} {\tau_1 \times \tau_2}} 
+\infer {\btyw {\tup{e_1,e_2}} {A \times B} {\tau_1 \times \tau_2}} 
 	{\btyw {e_1} A {\tau_1} 
@@
 
-Finally, we look at the let and var rules.
+Finally, we look at the let rule, which introduces a variable into the context.\footnote{
+Note the use of {\em hypothetical judgments}, 
+wherein the judgements for a variable are given as assumptions at a declaration site rather than a rule for the reference site.
+For now, the use of hypothetical judgments is merely a stylistic difference,
+but it will become important when things get more complicated.}
 
 \begin{mathpar}
-\infer [let] {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
+\infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
 	{\btyw {e_1} A {\tau_1} 
-	&\btyw [\Gamma,\btj x A w \rmunit] {e_2} B {\tau_2}} \and
-\infer [var] {\btyw [\Gamma,\btj x A w \rmunit] {x} A \rmunit} {\cdot}
+	&\btyw [\Gamma,\btj x A w \rmunit] {e_2} B {\tau_2}}
 \end{mathpar}
-Notice that we've defined these two rules in a slightly different way than usual.
-Typically, we'd say that whenever some variable $x$ is in the context 
-(because it was put there at a binder site),
-we can form the judgment that $x$ reference sites have a unit boundary.
-Instead what we've done is put the whole boundary type judgment in context at the binding site,
-and then we pull it out of the context at the reference site.
-Strictly speaking, the var rule is merely a special case of the 
-general structural rule stating ``you can use what you assume,'' so we'll omit it from here on out.
 
-This is a pretty common technique  called {\em hypothetical judgments}.
-For now, the use of hypothetical judgments is merely a stylistic difference,
-but it will become important when things get more complicated.
+Notice how variable is introduced with a unit boundary value. 
+This is because.
 
-\section{Adding Functions and Fixed Points}
-
-Lets turn this language to 11.  
-In addition to straight-line code, we'll add \texttt{if}s, functions, and fixed points,
-which the astute reader will notice is enough to make the language Turing-complete.
-Before getting into the rules, consider the following example, 
-where $E$ is some purely stage \bbone\ expression:
-
-\[
-\left(
-\tallif {E}
-	{(\lam x \rmint {\next\{\pause\{x\}+1\})}}
-	{(\lam x \rmint {\next\{\pause\{x\}+\pause\{x*x\}\}})}
-\right ) 5
-\]
-
-What do we expect the boundary type to be for this whole piece of code?
-Well, it depends. 
-If $E$ evaluates to true, then we'd expect the boundary type to be $\rmint$,
-but if $E$ evaluates to false, we'd expect the boundary type to be $\rmint \times \rmint$.
-Ergo, figuring out the boundary type of this piece of code is at least as hard 
-as figuring out what any given expression evaluates to.
-Of course, that's not decidable. Yikes.
-
-Fortunately for us, that's not really the end of it.  
-Just because we can't know the boundary type of a piece of code doesn't mean we can't say anything.  
-Maybe instead of giving the {\em exact} answer, we can give a list of possibilities.
-
-\section{Adding Just \texttt{if}-Expressions}
+\section{First-Order Recursive Functions}
 
@@
 
-The only particular care we took here was to observe speculation in the stage \bbtwo\ case,
+The only particular care we took here was to observe speculative behavior in the stage \bbtwo\ case,
 which induces a product (whereas stage \bbone\ had a sum).
@@
 \tallif {E}
-	{(\next\{\pause\{5\}+1\})}
-	{(\next\{\pause\{5\}+\pause\{5*5\}\})}
+	{\next{\pause{5}+1}}
+	{\next{\pause{5}+\pause{5*5}}}
 \]
@@
 $\rmunit \times (\rmint + \rmint \times \rmint)$.
-Sure, this isn't the {\em exact} boundary type, 
-but it's as much as we can know without an oracle.
 
-\section{Adding Just Functions}
+\section{Adding Functions and Fixed Points}
+
+To get a taste of why the problem is hard,
+we consider adding lots of features, namely \texttt{if}s, functions, and fixed points.
+The astute reader will notice that these are enough to make the language Turing-complete.
+Before getting into the rules, consider the following example, 
+where $E$ is some purely stage \bbone\ expression:
+
+\[
+\left(
+\tallif {E}
+	{\lam x {\next{\pause{x}+1}}}
+	{\lam x {\next{\pause{x}+\pause{x*x}}}}
+\right ) 5
+\]
+
+What do we expect the eventual boundary value for this whole piece of code to be?
+Well, it depends. 
+If $E$ evaluates to true, then we'd expect the boundary value to be an integer (that is, have type $\rmint$),
+but if $E$ evaluates to false, we'd expect the boundary value to be the tuple of two integers (that is, have type $\rmint \times \rmint$).
+Ergo, the only way to have a boundary type for this whole piece of code is to either,
+\begin{itemize}
+\item figure out what $E$ evaluates, or
+\item use a description for boundary types that mentions both $\rmint$ and $\rmint \times \rmint$, 
+without committing to one.
+\end{itemize}
+
+Of course, since our language is Turing complete, figuring out what $E$ evaluates to is not decidable,
+so we'll go with the second option.
+
+\section{Higher-Order Functions}
+\label{sec:higherorder}
 
@@
 \[
-L1 := (\lam x \rmint {\next\{5\}})
+L1 := \lam x {\next{5}}
 \]
@@
 \[
-L2 := (\lam x \rmint {\next\{\pause\{x\}+1\}})
+L2 := \lam x {\next{\pause{x}+1}}
 \]
-I'd like to think it's the same story.
+I'd like to think it's the same story: unit boundary.
 But what if we apply a value to these functions?
@@
 \[
-(\lam x \rmint {\next\{5\}})~3
+\app {\lam x {\next{5}}} 3
 \]
@@
 \[
-(\lam x \rmint {\next\{\pause\{x\}+1\}})~3
+\app {\lam x {\next{\pause{x}+1}}} 3
 \]
 The boundary type of the first expression should still be unit 
-(even after substitution it has no $\pause$s),
+(even after substitution it has no {\tt hold}s),
 but the bound type of the second expression should definitely be int.
@@
 There's a problem here: $L1$ and $L2$ had the same boundary type before application, but different ones after.
-This suggests that if we want the boundary type to be a compositional property (rather than global),
-then we'll need to keep track of more information.
+This suggests that if we want the boundary type to be a compositional property,
+then we'll need to keep track of more information than we do above.
 
-We do this by making a distinction between the {\em internal} boundary of an expression,
-and its {\em external} boundary.  
-$L1$ and $L2$ have the same external boundary, namely unit,
-but they have different internal boundaries.  
-In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
-whereas the internal boundary of $L2$ should be unit.
+\subsection{Adding Boundary Annotations}
+
+Our first attempt at a solution will be to add more information to function types in our input language.
+This new information simply tells us what the boundary type will be yielded when the function is applied.
+
+[solution where we write the boundary type above the arrows]
+
+\subsection{Separating Internal and External Boundaries}
+
+There's something unsatisfying about this solution. 
+The type system of the input language was perfectly functional before all this mess about boundary types.
+
+[show how to separate the internal from external boundaries]
+
+\subsection{Higher order funcitons}
+
+[Take the system from the previous section and consider higher-order functions.]
+
+% We do this by making a distinction between the {\em internal} boundary of an expression,
+% and its {\em external} boundary.  
+% $L1$ and $L2$ have the same external boundary, namely unit,
+% but they have different internal boundaries.  
+% In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
+% whereas the internal boundary of $L2$ should be unit.
 

### 2015-08-13 sematics/boundary/boundary.tex
working on values and computations section

@@
 We use a different typing judgement here in order to keep functions and other values distinct.
+That is, the $\to$ connective only ever appears to the right of ``$:>$'', and never to the right of a ``$:$".
 We also allow let statements to work for second stage functions 
@@
 \infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\Gamma \vdash \colfun {e_1} {A \to B} w & \ty {e_2} A} 					\and
-\infertypeswor [\fut E]		{\typestwo {\letin x {e_1} {e_2}} B}				{\Gamma \vdash \colfun {e_1} {A \to B} \bbtwo 
-																				&\typestwo [\Gamma,\colfun x {A \to B} \bbtwo] {e_2} B } 					\and
+\infertypeswor [\fut E]		{\typestwo {\letin f {e_1} {e_2}} B}				{\Gamma \vdash \colfun {e_1} {A \to B} \bbtwo 
+																				&\typestwo [\Gamma,\colfun f {A \to B} \bbtwo] {e_2} B } 					\and
 \end{mathpar}
@@
 
-...
+It turns out that, for the purposes of splitting, it will be handy to differentiate terms into two classes: 
+\emph{values} which are fully reduced, and \emph{computations} which may have some reduction remaining.
+Importantly, we want to \emph{explicitly note} when subterms of a computation are actually values,
+and so we introduce a new piece of syntax to mark just that inclusion: the underline constructor, written $\exv v$
+where $v$ is a term (more precisely, a value).  
+Note that the type of the underline is just the type of it's contents, which avoids interaction with the type system.
+
+Because staging and the value/computation distinction are both related to reduction, they will interact non-trivially.
+For this reason, we'll first devise a  system for a monostage language, before then considering a staged language.
+
+\subsection{Monostage System}
+
+Priori work has observed the difference between values and computations
+via a syntactic distinction (i.e. saying there are two kinds of terms, value-terms and computation-terms).
+We instead observe the difference judgementally: there's only one kind of term, 
+but there are two judgements ($\valsym$ and $\compsym$) which classify terms as one or the other.
+
+For instance, the aforementioned underline constructor always yields a computation,
+but its contents are a value:
+
+\begin{framed}
+\noindent\textbf{Underlines:}
+\begin{mathpar}
+\infer {\exv v \comp} {v \val} \and
+\infer {\exv v : A} {v : A} 
+\end{mathpar}
+\end{framed}
+
+Int, bool, and unit base terms are always values.  
+Intro forms, (tuples, injections, and rolls) are values so long as all their subterms are.
+
+\begin{framed}
+\noindent\textbf{Standard Values:}
+\begin{mathpar}
+\infer {\Gamma \vdash \tup{} \val}						{\cdot} 											\and
+\infer {\Gamma \vdash i \val}							{\cdot}												\and
+\infer {\Gamma \vdash b \val}							{\cdot}												\and
+\infer {\Gamma \vdash \tup{e_1,e_2} \val}				{\Gamma \vdash e_1 \val & \Gamma \vdash e_2 \val} 	\and
+\infer {\Gamma \vdash \inl e \val}						{\Gamma \vdash e \val} 								\and
+\infer {\Gamma \vdash \inr e \val}						{\Gamma \vdash e \val} 								\and
+\infer {\Gamma \vdash \roll e \val}						{\Gamma \vdash e \val} 										
+\end{mathpar}
+\end{framed}
+
+Inversely, intro forms are also computations whenever all of the subterms are.
+The elim forms, lets, and ifs are always computations.
+Note that because we're studying a call-by-value language, all variables are introduced as values.
+
+\begin{framed}
+\noindent\textbf{Standard Computations:}
+\begin{mathpar}
+\infer {\Gamma \vdash \tup{e_1,e_2} \comp}				{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp} 	\and
+\infer {\Gamma \vdash \inl e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \inr e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \roll e \comp}					{\Gamma \vdash e \comp} 										
+\end{mathpar}
+\end{framed}
 

### 2015-08-20 sematics/boundary/boundary.tex
various additions

@@
 
-\newcommand {\colfun} [3] {{#1}:>{#2}~\myatsign~{#3}}
+\newcommand {\colfunsym} {:>}
+\newcommand {\colfun} [3] {{#1}\colfunsym{#2}~\myatsign~{#3}}
 
@@
 
+\section{Various Monostage Formalisms}
+
+In this section, we'll cover a few background formalisms in the context of a monostage language.
+They will be extended to a staged language in the next section.
+This section is primarily intended to be a reference.
+
+\subsection{First-Order Recursive Functions}
+
+We implement first order recursive functions by introducing a new typing judgement \mbox{$f \colfunsym A \to B$},
+which says that $f$ is a function from $A$ to $B$.  
+We use a different typing judgement here in order to keep functions and other values distinct.
+That is, the $\to$ connective only ever appears to the right of ``$:>$'', and never to the right of a ``$:$".
+We also allow let statements to work for functions.
+Note how the only way to derive a function type is to directly declare a function, or to assume it as a variable.
+\begin{framed}
+\noindent\textbf{First-Order Recursive Functions:}
+\begin{mathpar}
+\infertypeswor [\to I]		{\Gamma \vdash {\fix fx{e_1}} \colfunsym {A \to B}}	{\types [\Gamma, f \colfunsym {A \to B},\col x A] {e_1} B} 							\and
+\infertypeswor [\to E]		{\types {\app {e_1}{e_2}} {B}}							{\Gamma \vdash {e_1} \colfunsym {A \to B} & \ty {e_2} A} 					\and
+\infertypeswor [\fut E]		{\types {\letin f {e_1} {e_2}} B}				{\Gamma \vdash {e_1} \colfunsym {A \to B} 
+																			&\types [\Gamma, f \colfunsym {A \to B}] {e_2} B } 					\and
+\end{mathpar}
+\end{framed}
+
+\subsection{Values and Computations}
+
+It turns out that, for the purposes of splitting, it will be handy to differentiate terms into two classes: 
+\emph{values} which are fully reduced, and \emph{computations} which may have some reduction remaining.
+Importantly, we want to \emph{explicitly note} when subterms of a computation are actually values,
+and so we introduce a new piece of syntax to mark just that inclusion: the underline constructor, written $\exv v$
+where $v$ is a term (more precisely, a value).  
+The type of the underline is just the type of it's contents, which avoids interaction with the type system.
+
+Because staging and the value/computation distinction are both related to reduction, they will interact non-trivially.
+For this reason, we'll first devise a value system for a monostage language.
+
+Prior work has observed the difference between values and computations
+via a syntactic distinction (i.e. saying there are two kinds of terms, value-terms and computation-terms).
+We instead observe the difference judgementally: there's only one kind of term, 
+but there are two judgements ($\valsym$ and $\compsym$) which classify terms as one or the other.
+
+For instance, the aforementioned underline constructor always yields a computation,
+but its contents are a value.
+
+\begin{framed}
+\noindent\textbf{Underlines:}
+\begin{mathpar}
+\infer {\Gamma \vdash \exv v \comp} {\Gamma \vdash v \val} \and
+\infer {\Gamma \vdash \exv v : A} {\Gamma \vdash v : A} 
+\end{mathpar}
+\end{framed}
+
+Int, bool, and unit base terms are always values.  
+Intro forms, (tuples, injections, and rolls) are values so long as all their subterms are.
+
+\begin{framed}
+\noindent\textbf{Standard Values:}
+\begin{mathpar}
+\infer {\Gamma \vdash \tup{} \val}						{\cdot} 											\and
+\infer {\Gamma \vdash i \val}							{\cdot}												\and
+\infer {\Gamma \vdash b \val}							{\cdot}												\and
+\infer {\Gamma \vdash \tup{e_1,e_2} \val}				{\Gamma \vdash e_1 \val & \Gamma \vdash e_2 \val} 	\and
+\infer {\Gamma \vdash \inl e \val}						{\Gamma \vdash e \val} 								\and
+\infer {\Gamma \vdash \inr e \val}						{\Gamma \vdash e \val} 								\and
+\infer {\Gamma \vdash \roll e \val}						{\Gamma \vdash e \val} 										
+\end{mathpar}
+\end{framed}
+
+Inversely, intro forms are also computations whenever all of the subterms are.
+The elim forms, {\tt let}s, and {\tt if}s are always computations.
+Note that because we're studying a call-by-value language, all variables are introduced as values.
+
+\begin{framed}
+\noindent\textbf{Standard Computations:}
+\begin{mathpar}
+\infer {\Gamma \vdash \tup{e_1,e_2} \comp}				{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp} 	\and
+\infer {\Gamma \vdash \inl e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \inr e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \roll e \comp}					{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \pio e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \pit e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \unroll e \comp}					{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \comp}		{\Gamma \vdash e_1 \comp 
+														&\Gamma \vdash e_2 \comp 
+														&\Gamma \vdash e_3 \comp}	\and
+\infer {\Gamma \vdash \letin x {e_1} {e_2} \comp}		{\Gamma \vdash e_1 \comp 
+														& \Gamma, x \val \vdash e_2 \comp } 				
+\end{mathpar}
+\end{framed}
+
+Higher-order functions are the only constructs where value terms have computation subterms.
+
+\begin{framed}
+\noindent\textbf{Higher-Order Functions:}
+\begin{mathpar}			
+\infer {\Gamma \vdash \lam x e \val}		{\Gamma, x \val \vdash e \comp} \and
+\infer {\Gamma \vdash \app{e_1}{e_2} \comp}	{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp}
+\end{mathpar}
+\end{framed}
+
+A term with a first-order function type, $e \colfunsym A \to B$, can only be a lambda itself or a variable declaration.
+Ergo, they are always values and never computations.
+For this reason, first-order functions don't really participate in the value/computation system.
+
+\begin{framed}
+\noindent\textbf{First-Order Recursive Functions:}
+\begin{mathpar}
+\infer {\Gamma \vdash \app{e_1}{e_2} \comp}	{\Gamma \vdash e_2 \comp}
+\end{mathpar}
+\end{framed}
+
 \section{Input Language Fragments}
@@
 This section defines a few input language fragments for later reference.
-Those fragments include ``core'' features (\emph{i.e.} simple ones that present no difficulty for any typing system), 
+Those fragments include ``core'' features (base types, products, $\fut$ types, and lets, which present no difficulty for output typing), 
 if expressions, first-order recursive functions, higher-order functions, and finally sum/recursive types.
@@
 \begin{framed}
-\noindent\textbf{Core Features (Base Types, Products, Circles, Let):}
+\noindent\textbf{Core Feature Typing:}
 \begin{mathpar}
+\infer {\Gamma \vdash \exv e : A} {\Gamma \vdash e : A} 																									\and
 \infertypeswor [\rmunit] 	{\ty {\tup{}}\rmunit}								{\cdot} 																	\and
@@
 \begin{framed}
+\noindent\textbf{Core Feature Values/Computations:}
+\begin{mathpar}
+\infer {\Gamma \vdash \exv e \compo} 					{\Gamma \vdash e \valo} 								\and
+\infer {\Gamma \vdash \tup{} \valo}						{\cdot} 												\and
+\infer {\Gamma \vdash i \valo}							{\cdot}													\and
+\infer {\Gamma \vdash b \valo}							{\cdot}													\and
+\infer {\Gamma \vdash \tup{e_1,e_2} \valo}				{\Gamma \vdash e_1 \valo & \Gamma \vdash e_2 \valo} 	\and
+\infer {\Gamma \vdash \tup{e_1,e_2} \compo}				{\Gamma \vdash e_1 \compo & \Gamma \vdash e_2 \compo} 	\and
+\infer {\Gamma \vdash \pio e \compo}					{\Gamma \vdash e \compo} 								\and
+\infer {\Gamma \vdash \pit e \compo}					{\Gamma \vdash e \compo} 								\and
+\end{mathpar}
+\end{framed}
+
+\begin{framed}
 \noindent\textbf{If Expressions:}
 \begin{mathpar}
-\infertypeswor [+ E]		{\ty {\ifthen{e_1}{e_2}{e_3}} A}					{\ty {e_1}\rmbool & \ty{e_2} A & \ty{e_3} A}
+\infertypeswor [+ E]		{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1}\rmbool & \ty{e_2} A & \ty{e_3} A} \and
+\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \compo}			{\Gamma \vdash e_1 \compo 
+																&\Gamma \vdash e_2 \compo 
+																&\Gamma \vdash e_3 \compo}
 \end{mathpar}
@@
 \infertypeswor [\mu E]		{\ty {\unroll e} {[\mualphatau / \alpha]\tau}}		{\ty e \mualphatau} 														\and
-\end{mathpar}
-\end{framed}
-
-\section{Values and Computations}
-
-It turns out that, for the purposes of splitting, it will be handy to differentiate terms into two classes: 
-\emph{values} which are fully reduced, and \emph{computations} which may have some reduction remaining.
-Importantly, we want to \emph{explicitly note} when subterms of a computation are actually values,
-and so we introduce a new piece of syntax to mark just that inclusion: the underline constructor, written $\exv v$
-where $v$ is a term (more precisely, a value).  
-Note that the type of the underline is just the type of it's contents, which avoids interaction with the type system.
-
-Because staging and the value/computation distinction are both related to reduction, they will interact non-trivially.
-For this reason, we'll first devise a  system for a monostage language, before then considering a staged language.
-
-\subsection{Monostage System}
-
-Priori work has observed the difference between values and computations
-via a syntactic distinction (i.e. saying there are two kinds of terms, value-terms and computation-terms).
-We instead observe the difference judgementally: there's only one kind of term, 
-but there are two judgements ($\valsym$ and $\compsym$) which classify terms as one or the other.
-
-For instance, the aforementioned underline constructor always yields a computation,
-but its contents are a value:
-
-\begin{framed}
-\noindent\textbf{Underlines:}
-\begin{mathpar}
-\infer {\exv v \comp} {v \val} \and
-\infer {\exv v : A} {v : A} 
-\end{mathpar}
-\end{framed}
-
-Int, bool, and unit base terms are always values.  
-Intro forms, (tuples, injections, and rolls) are values so long as all their subterms are.
-
-\begin{framed}
-\noindent\textbf{Standard Values:}
-\begin{mathpar}
-\infer {\Gamma \vdash \tup{} \val}						{\cdot} 											\and
-\infer {\Gamma \vdash i \val}							{\cdot}												\and
-\infer {\Gamma \vdash b \val}							{\cdot}												\and
-\infer {\Gamma \vdash \tup{e_1,e_2} \val}				{\Gamma \vdash e_1 \val & \Gamma \vdash e_2 \val} 	\and
-\infer {\Gamma \vdash \inl e \val}						{\Gamma \vdash e \val} 								\and
-\infer {\Gamma \vdash \inr e \val}						{\Gamma \vdash e \val} 								\and
-\infer {\Gamma \vdash \roll e \val}						{\Gamma \vdash e \val} 										
-\end{mathpar}
-\end{framed}
-
-Inversely, intro forms are also computations whenever all of the subterms are.
-The elim forms, lets, and ifs are always computations.
-Note that because we're studying a call-by-value language, all variables are introduced as values.
-
-\begin{framed}
-\noindent\textbf{Standard Computations:}
-\begin{mathpar}
-\infer {\Gamma \vdash \tup{e_1,e_2} \comp}				{\Gamma \vdash e_1 \comp & \Gamma \vdash e_2 \comp} 	\and
-\infer {\Gamma \vdash \inl e \comp}						{\Gamma \vdash e \comp} 								\and
-\infer {\Gamma \vdash \inr e \comp}						{\Gamma \vdash e \comp} 								\and
-\infer {\Gamma \vdash \roll e \comp}					{\Gamma \vdash e \comp} 										
+\infer {\Gamma \vdash \inl e \valo}						{\Gamma \vdash e \valo} 								\and
+\infer {\Gamma \vdash \inl e \compo}					{\Gamma \vdash e \compo} 								\and
+\infer {\Gamma \vdash \inr e \valo}						{\Gamma \vdash e \valo} 								\and
+\infer {\Gamma \vdash \inr e \compo}					{\Gamma \vdash e \compo} 								\and
+\infer {\Gamma \vdash \roll e \valo}					{\Gamma \vdash e \valo} 								\and
+\infer {\Gamma \vdash \roll e \compo}					{\Gamma \vdash e \compo} 								\and
+\infer {\Gamma \vdash \unroll e \compo}					{\Gamma \vdash e \compo} 								\and
+\infer {\Gamma \vdash \letin x {e_1} {e_2} \compo}		{\Gamma \vdash e_1 \compo 
+														& \Gamma, x \valo \vdash e_2 \compo } 	
 \end{mathpar}

### 2015-08-21 sematics/boundary/boundary.tex
adding boundary type stuff

@@
 \newcommand {\btsym} {\rotatebox[origin=c]{270}{$\triangle$}}
+\newcommand {\ibtsym} {\mathrel{\ooalign{$\btsym$\cr \hidewidth\hbox{$\bullet\mkern3mu$}\cr}}}
 \newcommand {\ktype} {\rm Type}
@@
 \newcommand {\btj} [4] {{#1}:{#2}~\myatsign~{#3}~\btsym~{#4}}
+\newcommand {\ibtj} [3] {{#1}\colfunsym{#2}~\myatsign~\bbone~\ibtsym~{#3}}
 \newcommand {\btyo} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbone{#4}}
@@
 \newcommand {\btysub} [2] [] {\bty{e_{#1}:{#2}}{\tau_{#1}}{\sigma_{#1}}}
+\newcommand {\ibtyo} [4] [\Gamma] {{#1}\vdash \ibtj{#2}{#3}{#4}}
+\newcommand {\ibtyt} [4] [\Gamma] {{#1}\vdash \ibtj{#2}{#3}\bbtwo{#4}}
 
@@
 
+In this section, we develop a boundary-type system for first-order recursive functions
+with {\tt if}s, but no sum types.
+
+\subsection {Core Features and {\tt if}s}
+
 For the most part, we'll do all of our work with a single new judgment \mbox{$\btj e A w \tau$}, 
-which indicates that the expression $e$ (which has type $A$ at world $w$) has the boundary type $\tau$.
+which indicates that the term $e$ (where $\colwor e A$ and $e \comp$) has the boundary type $\tau$.
 The goal is for the splitting rule to agree with the boundary type given to us by this judgment.
@@
 
-\subsection{Straight-line Code}
-
-Consider an input language which has no functions or divergence.
-At either stage, the boundary type of any expression is 
-just the product of the boundary types of its subexpressions
-(unit in the nullary case).
+For core features at either stage, the boundary type of any term is 
+just the product of the boundary types of its computation subterms.
 The only exception to this pattern is the $\pause$ rule,
@@
 
+\begin{framed}
+\noindent\textbf{Core Feature Boundary Types:}
 \begin{mathpar}
-\infer {\btyw i {\rm int}\rmunit} {\cdot} \and
-\infer {\btyw b {\rm bool}\rmunit} {\cdot} \and
+\infer {\btyw {\exv v} A\rmunit} {\cdot} \and
 \infer {\btyw {\tup{e_1,e_2}} {A \times B} {\tau_1 \times \tau_2}} 
@@
 \infer {\btyt {\pause e} \rmint {\tau \times \rmint}} 
-	{\btyo e \rmint \tau}
-\end{mathpar}
-
-Finally, we look at the let rule, which introduces a variable into the context.\footnote{
-Note the use of {\em hypothetical judgments}, 
-wherein the judgements for a variable are given as assumptions at a declaration site rather than a rule for the reference site.
-For now, the use of hypothetical judgments is merely a stylistic difference,
-but it will become important when things get more complicated.}
-
-\begin{mathpar}
+	{\btyo e \rmint \tau} \and
 \infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
 	{\btyw {e_1} A {\tau_1} 
-	&\btyw [\Gamma,\btj x A w \rmunit] {e_2} B {\tau_2}}
+	&\btyw {e_2} B {\tau_2}}
 \end{mathpar}
+\end{framed}
 
-Notice how variable is introduced with a unit boundary value. 
-This is because.
+Note that in the {\tt let} rule, we needn't declare a boundary type for the $x$ variable,
+since variables are values, and only computations have a boundary type.
+Rules for \texttt{if} terms are similarly easy:
 
-\section{First-Order Recursive Functions}
-
-So to explore what this means, lets dial back the language a bit.  
-In addition to straight-line from before, lets just add \texttt{if} expressions. 
-It turns out this is easy to integrate into the existing framework:
-
+\begin{framed}
+\noindent\textbf{If Expression Boundary Types:}
 \begin{mathpar}
@@
 \end{mathpar}
+\end{framed}
 
@@
 which induces a product (whereas stage \bbone\ had a sum).
-We can rewrite the problematic example in our restricted language 
-by distributing the application into the if and applying the lambda in both branches:
+
+\subsection {Functions}
+
+Consider the following example, which is open on the variable \mbox{$\colfun g {\rmint \to \rmint} \bbone$}.
 \[
-\tallif {E}
-	{\next{\pause{5}+1}}
-	{\next{\pause{5}+\pause{5*5}}}
+\app g {\app g 5}
 \]
-We our new rules, we can say that the boundary type of this expression is
-$\rmunit \times (\rmint + \rmint \times \rmint)$.
+Clearly, the boundary type of this computation depends on the particular function $g$ is bound to.
+Thus, in order for a boundary type system to be complete,
+the variable $g$ must be associated with information about the boundary type that is produced where $g$ is applied.
+This is called the \emph{internal boundary type} of $g$.
 
-\section{Adding Functions and Fixed Points}
+Our system works by associating every function derivation \mbox{$\colfun g {A \to B} \bbone$} with an internal boundary type $\sigma$
+via the judgement $\ibtj g {A \to B} \sigma$.  
+Functions delcared at world $\bbtwo$ do not need internal boundar types.
 
-To get a taste of why the problem is hard,
-we consider adding lots of features, namely \texttt{if}s, functions, and fixed points.
-The astute reader will notice that these are enough to make the language Turing-complete.
-Before getting into the rules, consider the following example, 
-where $E$ is some purely stage \bbone\ expression:
+\begin{framed}
+\noindent\textbf{First-Order Recursive Function Boundary Types:}
+\begin{mathpar}
+\infer {\btyo {\app {f}{e}} {A} {\tau \times \sigma}} 
+	{\ibtyo {f} {A \to B} {\sigma} 
+	&\btyo {e} A {\tau}} \and
+% \infer {\btyt {\fix fxe} {A} {\tau}} 
+% 	{\btyt {e} A {\tau}} \and
+% \infer {\btyt {\app {f}{e}} {A} {\tau}} 
+% 	{\btyt {e} A {\tau}} \and
+% \infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
+% 	{\btyw {e_1} A {\tau_1} 
+% 	&\btyw {e_2} B {\tau_2}}
+\end{mathpar}
+\end{framed}
 
-\[
-\left(
-\tallif {E}
-	{\lam x {\next{\pause{x}+1}}}
-	{\lam x {\next{\pause{x}+\pause{x*x}}}}
-\right ) 5
-\]
+% \section{Adding Functions and Fixed Points}
 
-What do we expect the eventual boundary value for this whole piece of code to be?
-Well, it depends. 
-If $E$ evaluates to true, then we'd expect the boundary value to be an integer (that is, have type $\rmint$),
-but if $E$ evaluates to false, we'd expect the boundary value to be the tuple of two integers (that is, have type $\rmint \times \rmint$).
-Ergo, the only way to have a boundary type for this whole piece of code is to either,
-\begin{itemize}
-\item figure out what $E$ evaluates, or
-\item use a description for boundary types that mentions both $\rmint$ and $\rmint \times \rmint$, 
-without committing to one.
-\end{itemize}
+% To get a taste of why the problem is hard,
+% we consider adding lots of features, namely \texttt{if}s, functions, and fixed points.
+% The astute reader will notice that these are enough to make the language Turing-complete.
+% Before getting into the rules, consider the following example, 
+% where $E$ is some purely stage \bbone\ expression:
 
-Of course, since our language is Turing complete, figuring out what $E$ evaluates to is not decidable,
-so we'll go with the second option.
+% \[
+% \left(
+% \tallif {E}
+% 	{\lam x {\next{\pause{x}+1}}}
+% 	{\lam x {\next{\pause{x}+\pause{x*x}}}}
+% \right ) 5
+% \]
 
-\section{Higher-Order Functions}
-\label{sec:higherorder}
+% What do we expect the eventual boundary value for this whole piece of code to be?
+% Well, it depends. 
+% If $E$ evaluates to true, then we'd expect the boundary value to be an integer (that is, have type $\rmint$),
+% but if $E$ evaluates to false, we'd expect the boundary value to be the tuple of two integers (that is, have type $\rmint \times \rmint$).
+% Ergo, the only way to have a boundary type for this whole piece of code is to either,
+% \begin{itemize}
+% \item figure out what $E$ evaluates, or
+% \item use a description for boundary types that mentions both $\rmint$ and $\rmint \times \rmint$, 
+% without committing to one.
+% \end{itemize}
 
-Lets now take a different stab at this, by adding functions to our straight-line code,
-but leaving \texttt{if}s out for now.
-Again, lets consider an example:
-\[
-L1 := \lam x {\next{5}}
-\]
-The boundary type of $L1$ is obviously unit, right?
-After all, lambdas are values, and values have unit boundary.
-So then what about this example:
-\[
-L2 := \lam x {\next{\pause{x}+1}}
-\]
-I'd like to think it's the same story: unit boundary.
-But what if we apply a value to these functions?
-Consider
-\[
-\app {\lam x {\next{5}}} 3
-\]
-versus
-\[
-\app {\lam x {\next{\pause{x}+1}}} 3
-\]
-The boundary type of the first expression should still be unit 
-(even after substitution it has no {\tt hold}s),
-but the bound type of the second expression should definitely be int.
+% Of course, since our language is Turing complete, figuring out what $E$ evaluates to is not decidable,
+% so we'll go with the second option.
 
-There's a problem here: $L1$ and $L2$ had the same boundary type before application, but different ones after.
-This suggests that if we want the boundary type to be a compositional property,
-then we'll need to keep track of more information than we do above.
+% \section{Higher-Order Functions}
+% \label{sec:higherorder}
 
-\subsection{Adding Boundary Annotations}
+% Lets now take a different stab at this, by adding functions to our straight-line code,
+% but leaving \texttt{if}s out for now.
+% Again, lets consider an example:
+% \[
+% L1 := \lam x {\next{5}}
+% \]
+% The boundary type of $L1$ is obviously unit, right?
+% After all, lambdas are values, and values have unit boundary.
+% So then what about this example:
+% \[
+% L2 := \lam x {\next{\pause{x}+1}}
+% \]
+% I'd like to think it's the same story: unit boundary.
+% But what if we apply a value to these functions?
+% Consider
+% \[
+% \app {\lam x {\next{5}}} 3
+% \]
+% versus
+% \[
+% \app {\lam x {\next{\pause{x}+1}}} 3
+% \]
+% The boundary type of the first expression should still be unit 
+% (even after substitution it has no {\tt hold}s),
+% but the bound type of the second expression should definitely be int.
 
-Our first attempt at a solution will be to add more information to function types in our input language.
-This new information simply tells us what the boundary type will be yielded when the function is applied.
+% There's a problem here: $L1$ and $L2$ had the same boundary type before application, but different ones after.
+% This suggests that if we want the boundary type to be a compositional property,
+% then we'll need to keep track of more information than we do above.
 
-[solution where we write the boundary type above the arrows]
+% \subsection{Adding Boundary Annotations}
 
-\subsection{Separating Internal and External Boundaries}
+% Our first attempt at a solution will be to add more information to function types in our input language.
+% This new information simply tells us what the boundary type will be yielded when the function is applied.
 
-There's something unsatisfying about this solution. 
-The type system of the input language was perfectly functional before all this mess about boundary types.
+% [solution where we write the boundary type above the arrows]
 
-[show how to separate the internal from external boundaries]
+% \subsection{Separating Internal and External Boundaries}
 
-\subsection{Higher order funcitons}
+% There's something unsatisfying about this solution. 
+% The type system of the input language was perfectly functional before all this mess about boundary types.
 
-[Take the system from the previous section and consider higher-order functions.]
+% [show how to separate the internal from external boundaries]
 
-% We do this by making a distinction between the {\em internal} boundary of an expression,
-% and its {\em external} boundary.  
-% $L1$ and $L2$ have the same external boundary, namely unit,
-% but they have different internal boundaries.  
-% In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
-% whereas the internal boundary of $L2$ should be unit.
+% \subsection{Higher order funcitons}
 
-% Consider a language with products and functions.  The internal boundary kind is given by:
+% [Take the system from the previous section and consider higher-order functions.]
 
-% \begin{mathpar}
-% \infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
-% \infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
-% \infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
-% %\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
-% \end{mathpar}
+% % We do this by making a distinction between the {\em internal} boundary of an expression,
+% % and its {\em external} boundary.  
+% % $L1$ and $L2$ have the same external boundary, namely unit,
+% % but they have different internal boundaries.  
+% % In particular, the interal boundary of $L1$ should be int (or isomorphic to it), 
+% % whereas the internal boundary of $L2$ should be unit.
+
+% % Consider a language with products and functions.  The internal boundary kind is given by:
+
+% % \begin{mathpar}
+% % \infer {\ibk {\rm int}{\rm Unit}} {\cdot} \and 
+% % \infer {\ibk {\rm bool}{\rm Unit}} {\cdot} \and 
+% % \infer {\ibk {A \times B}{\kappa_A \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} \and 
+% % %\infer {\ibk {A \to B}{\kappa_A \to \ktype \times \kappa_B}} {\ibk A {\kappa_A} & \ibk B {\kappa_B}} 
+% % \end{mathpar}
 

### 2015-08-21 sematics/boundary/boundary.tex
working on boundary typing document

@@
 \newcommand {\btj} [4] {{#1}:{#2}~\myatsign~{#3}~\btsym~{#4}}
-\newcommand {\ibtj} [3] {{#1}\colfunsym{#2}~\myatsign~\bbone~\ibtsym~{#3}}
+\newcommand {\ibtj} [3] {{#1}\colfunsym{#2}~\myatsign~\bbone~\hspace{-0.4em}\ibtsym\hspace{-0.2em}~{#3}}
 \newcommand {\btyo} [4] [\Gamma] {{#1}\vdash \btj{#2}{#3}\bbone{#4}}
@@
 We implement first order recursive functions by introducing a new typing judgement \mbox{$f \colfunsym A \to B$},
-which says that $f$ is a function from $A$ to $B$.  
-We use a different typing judgement here in order to keep functions and other values distinct.
-That is, the $\to$ connective only ever appears to the right of ``$:>$'', and never to the right of a ``$:$".
-We also allow let statements to work for functions.
-Note how the only way to derive a function type is to directly declare a function, or to assume it as a variable.
+which says that $f$ is a function from $A$ to $B$.\footnote{We use $f$ for derived functions, and $u$ for variables refering to functions.}  
+Using a different judgement keeps functions and other values distinct.
+That is, the $\to$ connective only ever appears immediately to the right of ``$:>$'', and never to the right of a ``$:$".
+We also add a special {\tt letf} statement for reusing functions within terms.
+
 \begin{framed}
@@
 \begin{mathpar}
-\infertypeswor [\to I]		{\Gamma \vdash {\fix fx{e_1}} \colfunsym {A \to B}}	{\types [\Gamma, f \colfunsym {A \to B},\col x A] {e_1} B} 							\and
-\infertypeswor [\to E]		{\types {\app {e_1}{e_2}} {B}}							{\Gamma \vdash {e_1} \colfunsym {A \to B} & \ty {e_2} A} 					\and
-\infertypeswor [\fut E]		{\types {\letin f {e_1} {e_2}} B}				{\Gamma \vdash {e_1} \colfunsym {A \to B} 
-																			&\types [\Gamma, f \colfunsym {A \to B}] {e_2} B } 					\and
+\infertypeswor [\to I]		{\Gamma \vdash {\fix ux{e}} \colfunsym {A \to B}}	{\types [\Gamma, u \colfunsym {A \to B},\col x A] {e} B} 							\and
+\infertypeswor [\to E]		{\types {\app {f}{e}} {B}}							{\Gamma \vdash {f} \colfunsym {A \to B} & \ty {e} A} 					\and
+\infertypeswor [\fut E]		{\types {\letfin u {f} {e}} C}						{\Gamma \vdash {f} \colfunsym {A \to B} 
+																				&\types [\Gamma, u \colfunsym {A \to B}] e C } 					\and
 \end{mathpar}
 \end{framed}
+The important thing to note here is that functions have only two forms: direct declaration or reference to an assumed variable.
 
@@
 \infer {\Gamma \vdash \unroll e \comp}						{\Gamma \vdash e \comp} 								\and
+\infer {\Gamma \vdash \letin x {e_1} {e_2} \comp}			{\Gamma \vdash e_1 \comp 
+															& \Gamma, x \val \vdash e_2 \comp } 					\and
 \infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \comp}			{\Gamma \vdash e_1 \comp 
@@
 															&\Gamma \vdash e_3 \comp}								\and
-\infer {\Gamma \vdash \letin x {e_1} {e_2} \comp}		{\Gamma \vdash e_1 \comp 
-														& \Gamma, x \val \vdash e_2 \comp } 				
+\infer {\Gamma\vdash\caseof{e_1}{x_2.e_2}{x_3.e_3}\comp}	{\Gamma \vdash e_1 \comp 
+															& \Gamma, x_2 \val \vdash e_2 \comp 
+															& \Gamma, x_3 \val \vdash e_3 \comp } 				
 \end{mathpar}
@@
 \begin{mathpar}
-\infer {\Gamma \vdash \app{e_1}{e_2} \comp}	{\Gamma \vdash e_2 \comp}
+\infer {\Gamma \vdash \app{f}{e} \comp}	{\Gamma \vdash e \comp} \and
+\infer {\Gamma \vdash \letfin{u}f{e} \comp}	{\Gamma \vdash e \comp}
 \end{mathpar}
@@
 
+The important thing to note when extending the value/computation system to stages is that
+we care about values \emph{with respect to the first pass of evaluation}.
+Thus $e \valt$ essentially means that $e$ is a residual.
+
 \begin{framed}
@@
 \begin{mathpar}
-\infer {\Gamma \vdash \exv e \compo} 					{\Gamma \vdash e \valo} 								\and
-\infer {\Gamma \vdash \tup{} \valo}						{\cdot} 												\and
-\infer {\Gamma \vdash i \valo}							{\cdot}													\and
-\infer {\Gamma \vdash b \valo}							{\cdot}													\and
-\infer {\Gamma \vdash \tup{e_1,e_2} \valo}				{\Gamma \vdash e_1 \valo & \Gamma \vdash e_2 \valo} 	\and
-\infer {\Gamma \vdash \tup{e_1,e_2} \compo}				{\Gamma \vdash e_1 \compo & \Gamma \vdash e_2 \compo} 	\and
-\infer {\Gamma \vdash \pio e \compo}					{\Gamma \vdash e \compo} 								\and
-\infer {\Gamma \vdash \pit e \compo}					{\Gamma \vdash e \compo} 								\and
+\infer {\Gamma \vdash \exv e \compw} 					{\Gamma \vdash e \valw} 								\and
+\infer {\Gamma \vdash \tup{} \valw}						{\cdot} 												\and
+\infer {\Gamma \vdash i \valw}							{\cdot}													\and
+\infer {\Gamma \vdash b \valw}							{\cdot}													\and
+\infer {\Gamma \vdash \tup{e_1,e_2} \valw}				{\Gamma \vdash e_1 \valw & \Gamma \vdash e_2 \valw} 	\and
+\infer {\Gamma \vdash \tup{e_1,e_2} \compw}				{\Gamma \vdash e_1 \compw & \Gamma \vdash e_2 \compw} 	\and
+\infer {\Gamma \vdash \pio e \compw}					{\Gamma \vdash e \compw} 								\and
+\infer {\Gamma \vdash \pit e \compw}					{\Gamma \vdash e \compw} 								\and
+\infer {\Gamma \vdash \letin x{e_1}{e_2} \compw}		{\Gamma \vdash e_1 \compw & \Gamma, x \valw \vdash e_2 \compw} 						\\
+\infer {\Gamma \vdash \next e \valo}					{\Gamma \vdash e \valt} 								\and
+\infer {\Gamma \vdash \next e \compo}					{\Gamma \vdash e \compt} 								\and
+\infer {\Gamma \vdash \prev e \compt}					{\Gamma \vdash e \compo} 								\\
+\infer {\Gamma \vdash \pio e \valt}						{\Gamma \vdash e \valt} 								\and
+\infer {\Gamma \vdash \pit e \valt}						{\Gamma \vdash e \valt} 								\and
+\infer {\Gamma \vdash \letin x{e_1}{e_2} \valt}			{\Gamma \vdash e_1 \valt & \Gamma, x \valt \vdash e_2 \valt} 						
 \end{mathpar}
@@
 \infertypeswor [+ E]		{\ty {\ifthen{e_1}{e_2}{e_3}} A}	{\ty {e_1}\rmbool & \ty{e_2} A & \ty{e_3} A} \and
-\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \compo}			{\Gamma \vdash e_1 \compo 
-																&\Gamma \vdash e_2 \compo 
-																&\Gamma \vdash e_3 \compo}
+\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \compw}			{\Gamma \vdash e_1 \compw 
+																&\Gamma \vdash e_2 \compw 
+																&\Gamma \vdash e_3 \compw} \and
+\infer {\Gamma \vdash \ifthen{e_1}{e_2}{e_3} \valt}				{\Gamma \vdash e_1 \valt 
+																&\Gamma \vdash e_2 \valt 
+																&\Gamma \vdash e_3 \valt}
 \end{mathpar}
@@
 
-We implement first order recursive functions by introducing a new typing judgement \mbox{$\colfun f {A\to B} w$},
-which says that $f$ is a function from $A$ to $B$ at world $w$.  
-We use a different typing judgement here in order to keep functions and other values distinct.
-That is, the $\to$ connective only ever appears to the right of ``$:>$'', and never to the right of a ``$:$".
-We also allow let statements to work for second stage functions 
-(for reasons explained later, first stage function lets can be safely eliminated \emph{a priori} by substitution).
-Note how the only way to derive a function type is to directly declare a function, or to assume it.
+As in the monostage system, first-stage first-order functions don't participate in the value/computation system.
+Second-stage functions do, however, since we need to mark whether they contain first stage code.
 \begin{framed}
@@
 \begin{mathpar}
-\infertypeswor [\to I]		{\Gamma \vdash \colfun {\fix fx{e_1}} {A \to B} w}	{\ty [,\colfun f {A \to B} w,\col x A] {e_1} B} 							\and
-%\infertypeswor [fix]		{\ty {\fix fxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
-\infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\Gamma \vdash \colfun {e_1} {A \to B} w & \ty {e_2} A} 					\and
-\infertypeswor [\fut E]		{\typestwo {\letin f {e_1} {e_2}} B}				{\Gamma \vdash \colfun {e_1} {A \to B} \bbtwo 
-																				&\typestwo [\Gamma,\colfun f {A \to B} \bbtwo] {e_2} B } 					\and
+\infertypeswor [\to I]		{\Gamma \vdash \colfun {\fix ux{e}} {A \to B} w}	{\ty [,\colfun u {A \to B} w,\col x A] {e} B} 							\and
+%\infertypeswor [fix]		{\ty {\fix uxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
+\infertypeswor [\to E]		{\ty {\app {f}{e}} {B}}								{\Gamma \vdash \colfun f {A \to B} w & \ty e A} 					\and
+\infertypeswor [\fut E]		{\ty {\letfin u f e} B}								{\Gamma \vdash \colfun f {A \to B} w 
+																				&\ty [,\colfun u {A \to B} w] e B } 					\and
+\infer {\Gamma \vdash \fix uxe \valw}											{\Gamma, u \valw, x \valw \vdash e \valw}				\and
+\infer {\Gamma \vdash \fix uxe \compt}											{\Gamma, u \valt, x \valt \vdash e \compt}				\and
+\infer {\Gamma \vdash \app{f}{e} \compw}										{\Gamma \vdash e \compw}								\and
+\infer {\Gamma \vdash \app{f}{e} \valt}											{\Gamma \vdash e \valt}									\and
+\infer {\Gamma \vdash \letfin{u}f{e} \compo}									{\Gamma \vdash e \compo}								\and
+\infer {\Gamma \vdash \letfin{u}f{e} \compt}									{\Gamma, u \compt \vdash e \compt}
 \end{mathpar}
@@
 \infertypeswor [\to I]		{\ty {\lam {x}{e}} {A \to B}}						{\ty [,\col x A] e B} 														\and
-%\infertypeswor [fix]		{\ty {\fix fxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
+%\infertypeswor [fix]		{\ty {\fix uxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
 \infertypeswor [\to E]		{\ty {\app {e_1}{e_2}} {B}}							{\ty {e_1} {A \to B} & \ty {e_2} A} 										\and
+...
 \end{mathpar}
@@
 \begin{mathpar}
-\infertypeswor [+ I_1]		{\ty {\inl e} {A + B}}								{\ty e A} 																	\and
-\infertypeswor [+ I_2]		{\ty {\inr e} {A + B}}								{\ty e B} 																	\and
-\infertypeswor [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}			{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C}	\and
-\infertypeswor [\mu I]		{\ty {\roll e} {\mu \alpha.\tau}}					{\ty e {[\mualphatau / \alpha]\tau}} 										\and
-\infertypeswor [\mu E]		{\ty {\unroll e} {[\mualphatau / \alpha]\tau}}		{\ty e \mualphatau} 														\and
-\infer {\Gamma \vdash \inl e \valo}						{\Gamma \vdash e \valo} 								\and
-\infer {\Gamma \vdash \inl e \compo}					{\Gamma \vdash e \compo} 								\and
-\infer {\Gamma \vdash \inr e \valo}						{\Gamma \vdash e \valo} 								\and
-\infer {\Gamma \vdash \inr e \compo}					{\Gamma \vdash e \compo} 								\and
-\infer {\Gamma \vdash \roll e \valo}					{\Gamma \vdash e \valo} 								\and
-\infer {\Gamma \vdash \roll e \compo}					{\Gamma \vdash e \compo} 								\and
-\infer {\Gamma \vdash \unroll e \compo}					{\Gamma \vdash e \compo} 								\and
-\infer {\Gamma \vdash \letin x {e_1} {e_2} \compo}		{\Gamma \vdash e_1 \compo 
-														& \Gamma, x \valo \vdash e_2 \compo } 	
+...
+% \infertypeswor [+ I_1]		{\ty {\inl e} {A + B}}								{\ty e A} 																	\and
+% \infertypeswor [+ I_2]		{\ty {\inr e} {A + B}}								{\ty e B} 																	\and
+% \infertypeswor [+ E]		{\ty {\caseof{e_1}{x_2.e_2}{x_3.e_3}} C}			{\ty {e_1}{A+B} & \ty[,\col {x_2} A]{e_2} C & \ty[,\col {x_3} B]{e_3} C}	\and
+% \infertypeswor [\mu I]		{\ty {\roll e} {\mu \alpha.\tau}}					{\ty e {[\mualphatau / \alpha]\tau}} 										\and
+% \infertypeswor [\mu E]		{\ty {\unroll e} {[\mualphatau / \alpha]\tau}}		{\ty e \mualphatau} 														\and
+% \infer {\Gamma \vdash \inl e \valo}						{\Gamma \vdash e \valo} 								\and
+% \infer {\Gamma \vdash \inl e \compo}					{\Gamma \vdash e \compo} 								\and
+% \infer {\Gamma \vdash \inr e \valo}						{\Gamma \vdash e \valo} 								\and
+% \infer {\Gamma \vdash \inr e \compo}					{\Gamma \vdash e \compo} 								\and
+% \infer {\Gamma \vdash \roll e \valo}					{\Gamma \vdash e \valo} 								\and
+% \infer {\Gamma \vdash \roll e \compo}					{\Gamma \vdash e \compo} 								\and
+% \infer {\Gamma \vdash \unroll e \compo}					{\Gamma \vdash e \compo} 								\and
+% \infer {\Gamma \vdash \letin x {e_1} {e_2} \compo}		{\Gamma \vdash e_1 \compo 
+% 														& \Gamma, x \valo \vdash e_2 \compo } 	
 \end{mathpar}
@@
 via the judgement $\ibtj g {A \to B} \sigma$.  
-Functions delcared at world $\bbtwo$ do not need internal boundar types.
+Functions declared at world $\bbtwo$ do not need internal boundary types.
 
@@
 \begin{mathpar}
-\infer {\btyo {\app {f}{e}} {A} {\tau \times \sigma}} 
+\infer {\ibtyo {\fix uxe} {A \to B} {\mu \alpha.\tau}} 
+	{\btyo [\Gamma, \alpha \wf, \ibtj u {A \to B} {\alpha}] e B {\tau}} \and
+\infer {\btyo {\app {f}{e}} {B} {\tau \times \sigma}} 
 	{\ibtyo {f} {A \to B} {\sigma} 
 	&\btyo {e} A {\tau}} \and
-% \infer {\btyt {\fix fxe} {A} {\tau}} 
+\infer {\btyo {\letin u f e} C \tau}
+	{\ibtyo {f} {A \to B} {\sigma} 
+	&\btyo [\Gamma, \ibtj u {A \to B} \sigma] {e} C {\tau}}
+% \infer {\btyt {\fix uxe} {A} {\tau}} 
 % 	{\btyt {e} A {\tau}} \and
@@
 \end{framed}
+
+The interesting case here is function. 
+Essentially, this works by inventing a type variable $\alpha$ and using that as the internal boundary type of the recursive function variable $u$.
+This yields a boundary type $\tau$ for the function body which is open on $\alpha$.
+This is then wrapped with a least-fixed-point to yield the internal boundary type for the whole function.
 

### 2015-10-09 paper/overview.tex
changed paper style

@@
 
-\begin{theorem*}
+\begin{theorem}
   Consider an execution of \texttt{qs1} with a randomly permuted input
@@
   $\Theta(n\log{n} + m\log{n})$ in expectation.
-\end{theorem*}
+\end{theorem}
 \begin{proof}

### 2015-10-09 sematics/boundary/boundary.tex
some old changes to boundary type stuff

@@
 \newcommand {\colfun} [3] {{#1}\colfunsym{#2}~\myatsign~{#3}}
+\newcommand {\unfunwrap}[1] {\lfloor{#1}\rfloor}
+\newcommand {\funwrap}[1] {\lceil{#1}\rceil}
 
@@
 \begin{mathpar}
-\infertypeswor [\to I]		{\Gamma \vdash {\fix ux{e}} \colfunsym {A \to B}}	{\types [\Gamma, u \colfunsym {A \to B},\col x A] {e} B} 							\and
-\infertypeswor [\to E]		{\types {\app {f}{e}} {B}}							{\Gamma \vdash {f} \colfunsym {A \to B} & \ty {e} A} 					\and
+\infertypeswor [\to I]		{\Gamma \vdash {\fix ux{e}} \colfunsym {A \to B}}	{\types [\Gamma, u \colfunsym {A \to B}, x : A] e B} 							\and
+\infertypeswor [\to E]		{\types {\app {f}{e}} {B}}							{\Gamma \vdash {f} \colfunsym {A \to B} & \types e A} 					\and
 \infertypeswor [\fut E]		{\types {\letfin u {f} {e}} C}						{\Gamma \vdash {f} \colfunsym {A \to B} 
@@
 The important thing to note here is that functions have only two forms: direct declaration or reference to an assumed variable.
+
+\subsection{Higher-Order Second-Class Functions}
+
+We can extend this formalism to include functions which are higher-order but still second-class.
+
+\begin{framed}
+\noindent\textbf{Higher-Order Second-Class Functions:}
+\begin{mathpar}
+\infertypeswor [\to I]		{\Gamma \vdash \funwrap e \colfunsym \overline A}	{\types e A} 													\and
+\infertypeswor [\to I]		{\types {\unfunwrap f} A}							{\Gamma \vdash f \colfunsym \overline A} 						\and
+\infertypeswor [\to I]		{\Gamma \vdash {\lam uf} \colfunsym {A \to B}}		{\Gamma, u \colfunsym A \vdash f \colfunsym B} 					\and
+\infertypeswor [\to E]		{\types {\app {f_1}{f_2}} {B}}						{\Gamma \vdash {f_1} \colfunsym {A \to B} 
+																				&\Gamma \vdash {f_2} \colfunsym A} 								\and
+\infertypeswor [\fut E]		{\types {\letfin u {f} {e}} C}						{\Gamma \vdash {f} \colfunsym {A \to B} 
+																				&\types [\Gamma, u \colfunsym {A \to B}] e C } 					\and
+\end{mathpar}
+\end{framed}
 
@@
 \infertypeswor [\to E]		{\ty {\app {f}{e}} {B}}								{\Gamma \vdash \colfun f {A \to B} w & \ty e A} 					\and
-\infertypeswor [\fut E]		{\ty {\letfin u f e} B}								{\Gamma \vdash \colfun f {A \to B} w 
-																				&\ty [,\colfun u {A \to B} w] e B } 					\and
+\infertypeswor [\fut E]		{\ty {\letfin u f e} C}								{\Gamma \vdash \colfun f {A \to B} w 
+																				&\ty [,\colfun u {A \to B} w] e C } 					\and
 \infer {\Gamma \vdash \fix uxe \valw}											{\Gamma, u \valw, x \valw \vdash e \valw}				\and
@@
 \infer {\Gamma \vdash \letfin{u}f{e} \compt}									{\Gamma, u \compt \vdash e \compt}
+\end{mathpar}
+\end{framed}
+
+\begin{framed}
+\noindent\textbf{Higher-Order Second-Class Functions:}
+\begin{mathpar}
+\infertypeswor [\to I]		{\Gamma\vdash \colfun{\funwrap e}{\overline A} w}	{\ty e A} 													\and
+\infertypeswor [\to I]		{\ty {\unfunwrap f} A}								{\Gamma\vdash \colfun f{\overline A} w} 						\\
+\infertypeswor [\to I]		{\Gamma \vdash \colfun {\lam xe} {A \to B} w}		{\Gamma, \colfun x A w \vdash \colfun e B w} 			\and
+%\infertypeswor [fix]		{\ty {\fix uxe} {A \to B}}							{\ty [,\col f {A \to B},\col x A] e B} 										\and
+\infertypeswor [\to E]		{\Gamma \vdash \colfun {\app {f_1}{f_2}} {B} w}		{\Gamma \vdash \colfun {f_1} {A \to B} w 
+																				&\Gamma \vdash \colfun {f_2} A w} 						\and
+\infertypeswor [\fut E]		{\ty {\letfin u f e} B}								{\Gamma \vdash \colfun f A w 
+																				&\ty [,\colfun u A w] e B} 								\\
+\infer {\Gamma \vdash \funwrap e \valo}											{\Gamma \vdash e \valo}									\and
+\infer {\Gamma \vdash \unfunwrap f \valo}										{\Gamma \vdash f \valo}									\\
+\infer {\Gamma \vdash \lam xe \valo}											{\Gamma, x \valo \vdash e \compo}						\and
+\infer {\Gamma \vdash \lam xe \valt}											{\Gamma, x \valt \vdash e \valt}						\and
+\infer {\Gamma \vdash \lam xe \compt}											{\Gamma, x \valt \vdash e \compt}						\and
+\infer {\Gamma \vdash \app{f}{e} \compw}										{\Gamma \vdash e \compw}								\and
+\infer {\Gamma \vdash \app{f}{e} \valt}											{\Gamma \vdash e \valt}									\and
+\infer {\Gamma \vdash \letfin{u}f{e} \compo}									{\Gamma \vdash e \compo}								\and
+\infer {\Gamma \vdash \letfin{u}f{e} \compt}									{\Gamma, u \valt \vdash e \compt}
 \end{mathpar}
@@
 
-\section{First Order Recursive Functions}
+\section{Second-Class Functions}
 \label{sec:firstorder}
@@
 
-\subsection {Functions}
+\subsection {First-Order Functions}
 
@@
 This is then wrapped with a least-fixed-point to yield the internal boundary type for the whole function.
+
+\subsection {Higher-Order Functions}
+
+Now we consider an extension to higher-order, but still second-class, functions.
+The main difference is that now, after the $\ibtsym$ we can have a type constructor, 
+which is either an actual type, a nullary-tuple constructor, or a function from one type constructor to another.
+
+\begin{framed}
+\noindent\textbf{Higher-Order Function Boundary Types:}
+\begin{mathpar}
+\infer {\ibtyo {\funwrap e} {\overline A} {()}}   {\cdot} \and
+\infer {\btyo {\unfunwrap e} {\overline A} \rmunit}   {\cdot} \and
+\infer {\ibtyo {\lam uf} {A \to B} {\lambda \alpha.\sigma}}
+	{\ibtyo [\Gamma, \alpha \wf, \ibtj u A {\alpha}] f B {\sigma}} \and
+\infer {\ibtyo {\app {f_1}{f_2}} {B} {(\sigma_1~\sigma_2)}} 
+	{\ibtyo {f_1} {A \to B} {\sigma_1} 
+	&\ibtyo {f_2} A {\sigma_2}} \and
+\infer {\btyo {\letin u f e} C \tau}
+	{\ibtyo {f} A {\sigma} 
+	&\btyo [\Gamma, \ibtj u A \sigma] {e} C {\tau}}
+% \infer {\btyt {\fix uxe} {A} {\tau}} 
+% 	{\btyt {e} A {\tau}} \and
+% \infer {\btyt {\app {f}{e}} {A} {\tau}} 
+% 	{\btyt {e} A {\tau}} \and
+% \infer {\btyw {\letin x {e_1} {e_2}} {B} {\tau_1 \times \tau_2}} 
+% 	{\btyw {e_1} A {\tau_1} 
+% 	&\btyw {e_2} B {\tau_2}}
+\end{mathpar}
+\end{framed}
 

### 2015-10-12 paper/examples.tex
Pound on program bifurcation really hard in the related work.
format fiddling
lots of cleaning and compression

@@
 1`dot : vec * vec -> $int
-fun dot (
-  (3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
-  (3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
-next{
-  2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
-1`}`
+fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
+  next{ 2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)` 1`}`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-1`fun dot1 ((x1,y1,()),(x2,y2,())) 
-  = ((), (x1*x2)+(y1*y2))`
-
-2`fun dot2 ((((),(),z1),((),(),z2)),l) 
-  = l+(z1*z2)`
+1`fun dot1 ((x1,y1,()), (x2,y2,()))    = ((), (x1*x2)+(y1*y2))`
+2`fun dot2 ((((),(),z1),((),(),z2)),l) = l+(z1*z2)`
 \end{lstlisting}
@@
 \begin{lstlisting} 
-datatype list = Empty | Cons of int * list
 datatype tree = Leaf
               | Branch of int * tree * int * tree
+1`datatype list = Empty | Cons of int * list
 
-1`fun partition ((p,l):int*list) : (int*list*list) = 
+fun part ((p,l):int*list) : (int*list*list) = 
   case l of 
@@
   | Cons (h,t) => 
-      let val (s,left,right) = partition (p,t) in 
+      let val (n,left,right) = part (p,t) in 
       if h<p 
-      then (s+1,Cons(h,left),right) 
-      else (s,left,Cons(h,right))
+      then (n+1,Cons(h,left),right) 
+      else (n,left,Cons(h,right))
 
@@
       Branch (
-        let val (left,right,n) = partition (h,t) in
+        let val (n,left,right) = part (h,t) in
         (n, #2 (qs1 left k), h, #2 (qs1 right k))))`
@@
 %
-In the appendix provided as supplemental material, we extend this example by
-splitting \texttt{map (qss l)} over an arbitrary list of ranks \texttt{ks}. We also
-provide an example of a program whose boundary data structure is a trie.
 
-\subsection {Composing Graphics Pipeline Programs}
+\subsection{Composing Graphics Pipeline Programs}
+\label{sec:graphics}
 

### 2015-10-12 paper/overview.tex
lots of cleaning and compression

@@
 
-\begin{theorem}
-  Consider an execution of \texttt{qs1} with a randomly permuted input
-  of $n$ keys and performing $m$ executions of \texttt{qs2} with the
-  result of \texttt{qs1}.  The total run-time for this computation is
-  $\Theta(n\log{n} + m\log{n})$ in expectation.
-\end{theorem}
-\begin{proof}
-  We note first that \texttt{qs1} is
-  isomorphic to a quicksort algorithm but builds a binary search tree
-  instead of combining the returned results in a list.  The run-time
-  for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
-%
-  The depth of the tree generated by \texttt{qs1} is isomorphic to the
-    depth of a run of the quickselect algorithm, or equivalently to the
-    span of a parallel implementation of the quicksort algorithm, both
-    of which are known to be (expected) $\Theta(\log{n})$.
-%
-    By inspection of \texttt{qs2}, we can see that the algorithm takes
-    time proportional to the depth of the binary search tree, which is
-    $\Theta(\log{n})$. Consider now a random variable for each of the
-    $m$ invocations of \texttt{qs2}, each of which takes expected
-    $\Theta(\log{n})$ time.  Since the total time is the sum of these
-    random variables, and the expectation of the sum is the sum of
-    the expectations of the random variables, the bound follows.
-\end{proof}
+% \begin{theorem}
+%   Consider an execution of \texttt{qs1} with a randomly permuted input
+%   of $n$ keys and performing $m$ executions of \texttt{qs2} with the
+%   result of \texttt{qs1}.  The total run-time for this computation is
+%   $\Theta(n\log{n} + m\log{n})$ in expectation.
+% \end{theorem}
+% \begin{proof}
+%   We note first that \texttt{qs1} is
+%   isomorphic to a quicksort algorithm but builds a binary search tree
+%   instead of combining the returned results in a list.  The run-time
+%   for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
+% %
+%   The depth of the tree generated by \texttt{qs1} is isomorphic to the
+%     depth of a run of the quickselect algorithm, or equivalently to the
+%     span of a parallel implementation of the quicksort algorithm, both
+%     of which are known to be (expected) $\Theta(\log{n})$.
+% %
+%     By inspection of \texttt{qs2}, we can see that the algorithm takes
+%     time proportional to the depth of the binary search tree, which is
+%     $\Theta(\log{n})$. Consider now a random variable for each of the
+%     $m$ invocations of \texttt{qs2}, each of which takes expected
+%     $\Theta(\log{n})$ time.  Since the total time is the sum of these
+%     random variables, and the expectation of the sum is the sum of
+%     the expectations of the random variables, the bound follows.
+% \end{proof}
 

### 2015-10-12 paper/semantics.tex
lots of cleaning and compression

@@
 
-All three worlds contain unit, product, and function types defined in the usual fashion;
-sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
-\footnote{We omit sums and recursive types from \bbonem\ to save space, though we do not believe they would present any unique difficulty.}
+All three worlds contain unit, product, function, sum and recursive types defined in the usual fashion.
 These ``standard'' features can only be constructed from subterms of the same world,
@@
 \input{figures/dynamics}
-\input{figures/dynamics-overview}
 
@@
 
+\begin{abstrsyn}
+\begin{theorem} [Progress] \leavevmode
+\label{thm:progress} 
+\begin{itemize} 
+\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
+\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
+\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
+\end{itemize}
+\end{theorem}
+\begin{theorem} [Preservation] \leavevmode
+\label{thm:preservation} 
+\begin{itemize} 
+\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
+\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
+\end{itemize}
+\end{theorem}
+\end{abstrsyn}
+
 \subsection{Evaluating Staged Programs}

### 2015-10-12 paper/splitting.tex
lots of cleaning and compression

@@
 to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
-By equivalent, we mean that $\mathcal P$ and $\mathcal P'$ should produce 
+More specifically, $\mathcal P$ and $\mathcal P'$ should produce 
 the same answer under their respective semantics.
@@
 the splitting algorithm runs in linear time on the size of the input program.
-Splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
+Splitting is defined for all well-typed inputs (\ref{thm:total}), and it produces unique results (\ref{thm:unique}).
 That is, each splitting judgement defines a total function.
+
+\begin{theorem} [Splitting Totality] \leavevmode
+\label{thm:total} 
+\begin{itemize} 
+\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
+\item For term $e$, if $\typesone e A$, then $\splitone e A c {l.r}$.
+\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
+\end{itemize}
+\end{theorem}
+
+\begin{theorem} [Splitting Uniqueness] \leavevmode
+\label{thm:unique}
+\begin{itemize} 
+\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
+\item If $\splitone e A p {l.r}$ and $\splitone e A {c'} {l'.r'}$, then $c = c'$, and $l.r = l'.r'$.
+\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
+\end{itemize}
+\end{theorem}
+
 We prove these theorems by straightforward induction on the typing derivation
@@
 
+\begin{theorem} [End-to-End Correctness]
+\label{thm:allCorrect}
+If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
+then $\reduce p b$ and $[b/l]r \equiv q$.
+\end{theorem}
+
 We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
@@
 and whose proof will concern the rest of this section.
+
+\begin{lemma} [Single Step Correctness]
+\label{lem:twoCorrect}
+\begin{itemize} 
+\item If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
+then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
+\item If $e \stepsym \bbonem e'$, $\splitone e A c {l.r}$, $\splitone {e'} A {c'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
+then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
+\end{itemize}
+\end{lemma}
+
 After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
-
 In order to prove \ref{lem:twoCorrect},
@@
 
+\begin{lemma} [Hoisting Step Correctness]
+\label{lem:hoistCorrect}
+If $\lift e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, then $c = c'$, and $l.r \equiv l'.r'$.
+\end{lemma}
+
 \subsection{Example cases.}
@@
 $E' = [\fix fxe, v/f,x] e$, \\ 
-$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$, \\
+$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$,
 $Q = \fix f {\tup{x,\roll l}} r$, and \\

### 2015-10-14 paper/relatedwork.tex
changes all over

@@
 
-In an attempt to reduce the complexity of the specialization phase of
-self-applicable partial evaluators, researchers explored the use of
-staging transformations that, much like splitting, partition an input
-program into two programs: one that produces static outputs from
-static inputs, and a second that produces dynamic outputs from both
-static and dynamic inputs.  These metastatic transformations, referred
-to as \emph{program bifurcation}\,\cite{DeNiel91} or as \emph{binding
-  time separation}\,\cite{Mogensen89a}), serve as a preprocessing step
-to specialization and were successfully applied to functional
-languages that supported recursion\,\cite{DeNiel91}. However, unlike
-splitting, where the goal is to emit code where first stage results
+Researchers have explored other staging transformations that, 
+like splitting, partition an input program into static and dynamic components.  
+In particular, \emph{binding time separation}\,\cite{Mogensen89a} 
+(also called \emph{program bifurcation}\,\cite{DeNiel91})
+has been used as a preprocessor step in partial evaluators,
+allowing more efficient specialization of programs with 
+mixed-stage data structures without changes to the specializer itself.
+Notably, the grammar-based binding-time specifications used in binding 
+time separation were capable of describing
+data structures with purely-static, purely-dynamic, and mixed stage
+content, much like the type system of \lang\
+(though this correspondence is less clear without our addition of 
+\bbonep\ and $\curr$).
+
+That said, unlike splitting, where the goal is to emit code where first stage results
 are computed once and then reused in multiple invocations of second
-stage execution, the second function produced by program bifurcation
-does not use the results of the first; Instead,
+stage execution, the second (dynamic) function produced by binding time separation
+does not use the results of the first; instead,
 %only serves as a preprocessing step for the subsequent
@@
 it has access to the first stage inputs and recomputes all required
-first stage computations. (As noted by Knoblock and
+first stage computations. As noted by Knoblock and
 Ruf\,\cite{knoblock96}, it may be possible to modify the program
 bifurcation algorithm to cache and reuse the intermediate results, but
-this was never attempted.) The grammar-based binding-time
-specifications used in program bifurcation were capable of describing
-data structures with purely-static, purely-dynamic, and mixed stage
-content, much like the type system of \lang.
+this was never attempted.
 

### 2015-10-14 paper/splitting.tex
changes all over

@@
 All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
-To close this gap we introduce \ref{lem:twoCorrect}, 
+To close this gap we introduce \ref{lem:stepCorrect}, 
 which essentially states that any single step is correct,
@@
 \begin{lemma} [Single Step Correctness]
-\label{lem:twoCorrect}
+\label{lem:stepCorrect}
 \begin{itemize} 
@@
 After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
-In order to prove \ref{lem:twoCorrect},
+In order to prove \ref{lem:stepCorrect},
 we will need to state analogous version for steps at \bbonem,
@@
 
-The correctness of \bbonem\ term splitting with respect to a $\stepsym \bbonem$ step is given in \ref{lem:oneCorrect}.
+The correctness of \bbonem\ term splitting with respect to a $\stepsym \bbonem$ step is given in \ref{lem:stepCorrect}.
 It's very similar to the world \bbtwo\ version, 
@@
 namely the immediate value $i$.
-\ref{lem:oneCorrect} accounts for this by saying that 
+\ref{lem:stepCorrect} accounts for this by saying that 
 that immediate value must be exactly identical on both sides of the step.
@@
 In this section, we consider a few exemplar cases 
-from the proofs of \ref{lem:oneCorrect} and \ref{lem:twoCorrect}.
+from the proof of \ref{lem:stepCorrect}.
 The equivalence relation ``$\equiv$'' is defined in \ref{fig:equiv}.
@@
 $\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
-Using \ref{lem:twoCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
+Using \ref{lem:stepCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
 From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
@@
 and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
-Using \ref{lem:oneCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
+Using \ref{lem:stepCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
 From this, we can derive $\reduce {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b}}$ and 

### 2015-10-14 paper/relatedwork.tex
small clarity fix re bifurcation

@@
 Researchers have explored other staging transformations that, 
-like splitting, partition an input program into static and dynamic components.  
+like splitting, partition an input two-stage program into two components,
+on corresponding to each stage.  
 In particular, \emph{binding time separation}\,\cite{Mogensen89a} 
@@
 has been used as a preprocessor step in partial evaluators,
-allowing more efficient specialization of programs with 
+allowing efficient specialization of programs with 
 mixed-stage data structures without changes to the specializer itself.
@@
 
-That said, unlike splitting, where the goal is to emit code where first stage results
+However unlike splitting, where the goal is to emit code where first stage results
 are computed once and then reused in multiple invocations of second

### 2015-10-15 paper/relatedwork.tex
stuff on values

@@
 Notably, the grammar-based binding-time specifications used in binding 
-time separation were capable of describing
-data structures with purely-static, purely-dynamic, and mixed stage
+time separation are capable of describing
+data structures with purely-static, purely-dynamic, and mixed-stage
 content, much like the type system of \lang\

### 2015-10-15 paper/semantics.tex
changes all over
stuff on values

@@
 The abstract syntax of \lang\ is presented as a grammar in \ref{fig:grammar}.  
-Note that some production rules are parameterized by world.
-Our syntax also distinguishes values from general terms, 
-with the underline term constructor (``$\exv v$'') explicitly noting the inclusion of values into general terms.
-There are three forms of value: \emph{partial values} for \bbonem, \emph{ground values} for \bbonep, and \emph{residuals} for \bbtwo.
-All represent computations which have no remaining work in the first stage.
+To simplify the upcoming translation in \ref{sec:splitting}, 
+we have chosen to statically distinguish between values and general computations,
+using an underline constructor (``$\exv v$'') which explicitly note the parts of a computation that have been reduced to a value.
+\footnote{In this presentation, the value/computation distinction is encoded intrinsically at the syntactic level.
+However, we found it more convenient in the formal Twelf implementation to maintain the distinction with an extrinsic judgement.}
+Moreover, the value/computation distinction interacts non-trivially with the three possible world classifications.
+As a result, we end up with six classes of term in the grammar.
+
+The key feature of all three forms of value is that they have no remaining work in the first stage.
+That is, all of the first-stage portions of a value are fully reduced (except the bodies of first-stage functions).
+In the case of values at \bbonep, which we call \emph{ground values}, this collapses to just the standard notion of values in a monostage language.
+In the case of values at \bbtwo, which we call \emph{residuals}, this collapses to just standard monostage terms.
+Lastly, values at \bbonem\ end up with more of a mixed character, so we call them \emph{partial values}.
 
@@
 
-The statics and dynamics of \lang\ are related by the type safety theorems
-in \ref{fig:typeSafety}, again annotated by world.
+The statics and dynamics of \lang\ are related by the type safety theorems below, again annotated by world.
 In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,

### 2015-10-16 paper/examples.tex
various edits

@@
 runtime generation of a huge amount of code simultaneously encoding the tree and
-traversal algorithm. (Avoiding the code blowup, by not expanding some calls to
-\texttt{partition}, would result in duplicating first-stage computations.)
+traversal algorithm. (Avoiding the code blowup by not expanding some calls to
+\texttt{part}, would result in duplicating first-stage computations.)
 
@@
 
-Note that the recursive \texttt{partition} function is defined within a \texttt{gr}
-annotation.  As explained in \ref{sec:needGround}, defining \texttt{partition}
+Note that the recursive \texttt{part} function is defined within a \texttt{gr}
+annotation.  As explained in \ref{sec:needGround}, defining \texttt{part}
 at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
@@
 \lang\ with splitting used to generate monostaged code required for
-each GPU pipeline stage. Composition of such functions in
+each graphics pipeline stage. Composition of such functions in
 \lang\ allows more complex pipelines to be expressed in a modular
@@
 
-A simple example of composition of staged functions (representing GPU
+A simple example of composition of staged functions (representing graphics
 pipeline programs) is given below.  The code defines a pipeline as a
@@
 %
-Whereas prior systems provides combinators like \texttt{par} as
+Whereas prior systems provide combinators like \texttt{par} as
 language features\,\cite{Foley:2011} to support composition, \lang\ is

### 2015-10-16 paper/implementation.tex
various small edits
various edits

@@
 \label{sec:implementation}
+
+\subsection {Proof Implementation}
+
+We have encoded the type system, dynamic semantics, splitting algorithm, and output semantics in Twelf,
+as well as all of the theorems of \ref{sec:semantics,sec:splitting}, and their proofs. 
+Details can be found in the supplementary materials.
+
+\subsection {Example Implementation}
 
@@
   datatype list = Empty | Cons of int * list
-  fun partition (...) = ...
+  fun part (...) = ...
 `3`}`1`
@@
 \end{lstlisting}
-declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
+declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
 by elaborating into:
@@
 val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
-val 3`gr{`partition3`}` = 3`gr{`fn (...) => ...3`}`
+val 3`gr{`part3`}` = 3`gr{`fn (...) => ...3`}`
 val qss = fn (...) => ...`

### 2015-10-16 paper/intro.tex
various edits

@@
 The dynamic semantics (\ref{sec:semantics}) of \lang\ are that of Davies, 
-modified to provide an eager behavior between stages, which is more intuitive in the context of splitting.
+modified to provide an eager behavior between stages, which we believe is more intuitive in the context of splitting.
 We then prove the correctness of our splitting algorithm (\ref{sec:splitting}) for \lang\ with respect to the semantics.

### 2015-10-16 paper/overview.tex
various edits

@@
 Suppose that we wish to perform a series of order statistics queries
-on a list \texttt{l}. To this end, we can use the quickselect
-algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
+on a list~\texttt{l}. To this end, we can use the quickselect
+algorithm~\cite{Hoare:1961}, which given a list~\texttt{l} and a rank
 \texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
@@
 between these calls to \texttt{qs}. In particular, the comparisons performed by
-\texttt{partition} depend only on \texttt{l}, although different calls to
-\texttt{partition} are made by each call to \texttt{qs}.
+\texttt{part} depend only on \texttt{l}, although different calls to
+\texttt{part} are made by each call to \texttt{qs}.
 
@@
 \texttt{qss}, where first-stage code is colored red, and second-stage blue. In
-\texttt{qss}, we regard the input list \texttt{l} as arriving in the first stage
+\texttt{qss}, we regard the input list~\texttt{l} as arriving in the first stage
 (with type $\curr\mathrm{list}$, a list ``now''), the input rank \texttt{k} as
@@
 \texttt{prev} and \texttt{next}, signaling transitions between first- and
-second-stage code, and $\texttt{gr}$, signaling ground terms.  
+second-stage code. Additionally, $\texttt{gr}$ (\emph{ground}) annotations 
+mark certain first stage components as being purely first-stage, rather than mixed-stage.
 %
@@
 \lstinline{1`hold : ^int -> $`2`int`}, 
+\begin{comment}$ this comment block is here for syntax highlighting reasons\end{comment}
 to promote first-stage integers to second-stage integers.
@@
 In particular, \texttt{qs1} performs all recursive calls and evaluates all
-instances of \texttt{partition} (since it depends only on \texttt{l}). It
+instances of \texttt{part} (since it depends only on \texttt{l}). It
 produces a boundary data structure that collects the results from these

### 2015-10-16 paper/relatedwork.tex
various edits

@@
 like splitting, partition an input two-stage program into two components,
-on corresponding to each stage.  
+one corresponding to each stage.  
 In particular, \emph{binding time separation}\,\cite{Mogensen89a} 

### 2015-10-16 paper/semantics.tex
various edits

@@
 \input{figures/dynamics}
-
+\begin{abstrsyn}
 The central tenet of a staged language
@@
 \end{lstlisting}
-We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
-expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
+We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
+expensive computation $e$ depending on how many times $x$ appears in $e$. 
+Instead, we choose to \emph{hoist} $e$ outside, binding it
 to a temporary variable $y$, and substituting that variable instead:
@@
 2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
-\end{lstlisting}
+\end{lstlisting}\end{abstrsyn}
 This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
+We read this as saying somewhere within $e$ there was the subterm $q$ which needs to be hoisted out,
+yielding the new term $e'$ which has a new variable $y$ where the $q$ used to be.
 These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.

### 2015-10-16 paper/splitting.tex
various edits

@@
 $\splitone e A c {l.r}$ for \bbonem-terms, and 
-$v \vsplito \mval i q$ for partial values.
+\mbox{$v \vsplito \mval i q$} for partial values.
 In each form the output has two parts, corresponding to the
@@
 
-An illustrative example is the \texttt{partition} function in the quickselect
-example of \ref{sec:exampleQS}. If \texttt{partition} were defined at \bbonem\
+An illustrative example is the \texttt{part} function in the quickselect
+example of \ref{sec:exampleQS}. If \texttt{part} were defined at \bbonem\
 then (like \texttt{qs}) it would split into two functions
-\texttt{partition1} and \texttt{partition2}, the latter of which recursively
-computes the (trivial) second-stage component of \texttt{partition}'s result.
-Moreover, \texttt{qs2} would call \texttt{partition2}, just as \texttt{qs1}
-calls \texttt{partition1}:
+\texttt{part1} and \texttt{part2}, the latter of which recursively
+computes the (trivial) second-stage component of \texttt{part}'s result.
+Moreover, \texttt{qs2} would call \texttt{part2}, just as \texttt{qs1}
+calls \texttt{part1}:
 \begin{lstlisting}
@@
 ... | Branch (i,h,p1,p2) =>
-  let val () = partition2 ((),()) in
+  let val () = part2 ((),()) in
   case compare k i of ...
@@
 terms are purely first-stage.  
-Defining \texttt{partition} in this world is 
+Defining \texttt{part} in this world is 
 tantamount to proving it has no second-stage computations,
-allowing splitting to avoid generating the resumer \texttt{partition2}
+allowing splitting to avoid generating the resumer \texttt{part2}
 and calling it from \texttt{qs2}.

### 2015-10-17 paper/examples.tex
various revisions

@@
 
-1`fun shade (refl : pipeline,
-           albedo : pipeline) : pipeline =
-  (fn (obj : object, next{2`xy`} : $2`coord`) =>
-       2`prev{`1`refl   (obj,next{2`xy`})`2`}
-     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
+1`fun shade (refl : pipeline, albedo : pipeline) : pipeline =
+  fn (obj : object, next{2`xy`} : $2`coord`) =>
+       2`prev{`1`refl (obj,next{2`xy`})`2`} * prev{`1`albedo (obj,next{2`xy`})`2`}`
 \end{lstlisting}

### 2015-10-17 paper/relatedwork.tex
various revisions

@@
 this was never attempted.
+Alternatively, our algorithm could potentially be used as the basis of a general
+bifurcation algorithm in a partial evaluator.
 
@@
 extension to the typed lambda calculus that we use here is based on
-the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ in
-small ways to account for the specific evaluation model that we
-propose; however, the key difference between our work and this prior
+the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ 
+in the restriction to two stages and the addition of $\curr$; 
+however, the key difference between our work and this prior
 work is that we instead focus on the problem of splitting.
@@
 efforts in computer graphics, like that of Knoblock and
-Ruf\,\cite{knoblock96} have been limited to simple, imperative
+Ruf\,\cite{knoblock96}, have been limited to simple, imperative
 languages.
+
+\paragraph{Defunctionalization.}
+Defunctionalization \cite{Reynolds72} is a program transformation
+that eliminates high-order functions and replaces them with lower
+order functions that manipulate a data structure encoding the original control flow.
+This operation has similarity to our splitting transformation, 
+which eliminates staging in a program by encoding control flow in a data structure
+passed between the stage one and stage two outputs.
+It would interesting to explore the possibility of a semantics-preserving transformation
+to convert a staged program into a higher-order form,
+and then applying defunctionalization to obtain our results.

### 2015-10-17 paper/semantics.tex
various revisions

@@
 there are two worlds corresponding to the first stage:
-\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
+\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within {\tt next} blocks,
 and \bbonep\ for \emph{ground} first-stage computations, which may not.
@@
 \paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
-Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code (compatibility rules) 
-where it either performs a beta reduction (reduction rules) or marks the term as being evaluated (value marking rules). 
+Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code where it performs a reduction. 
 A ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.
@@
 \paragraph{World \bbonem.}
-Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
-It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
-The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
+Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a reduction.
+It also descends into {\tt gr} and {\tt next} blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
+The value form for \bbonem, called a partial value, comprises units, tuples, functions, {\tt gr} blocks of ground values, and {\tt next} blocks containing \emph{only a stage two variable}.
 This strong requirement ensures that second-stage computations are not duplicated 
@@
 We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
-expensive computation $e$ depending on how many times $x$ appears in $e$. 
+expensive computation $e$ depending on how many times $x$ appears in $e'$. 
 Instead, we choose to \emph{hoist} $e$ outside, binding it
@@
 This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
-We read this as saying somewhere within $e$ there was the subterm $q$ which needs to be hoisted out,
-yielding the new term $e'$ which has a new variable $y$ where the $q$ used to be.
+We read this as saying that somewhere within $e$ there was a subterm $q$ which needs to be hoisted out,
+yielding the new term $e'$ which has a new variable $y$ where $q$ used to be.
 These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
@@
 where it is reified into a {\tt let} statement.
-\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
+\footnote{Because a program is a term at \bbtwo, this {\tt prev} always exists.
 Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}
+
+Consider the following example, where {\tt P} has type $\curr(\rmunit + \rmunit)$,\begin{lstlisting}
+           1`caseg P of _ => next{`2`0`1`} | _ => next{`2`literalone`1`}`
+\end{lstlisting}
+Depending on what {\tt P} evaluates to in the first stage, whole term will step to either {\tt next\{0\}} or {\tt next\{1\}}.
+In this sense, we can see {\tt case} (at world \bbonem) and {\tt caseg} as the constructs that facilitate
+all cross-stage communication.
 
@@
 either a standard step or a hoisting step.
-
 \begin{abstrsyn}
@@
 \begin{itemize} 
-\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
+\item If $\typesone e A$ and $\lift e {e'}$, \\
+	then $\typestwo q B$ and \mbox{$\typesone [\Gamma,\coltwo y B] {e'} A$}.
 \item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
@@
 \begin{lstlisting}
-2`let fun qs k = 
-      prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
+2`let fun qs k = prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
 2`in (qs k1,...,qs km)`
@@
 \begin{lstlisting}
-2`let fun qs k = 
-	  let val y = ... in y
+2`let fun qs k = (let val y = ... in y)
 2`in (qs k1,...,qs km)`

### 2015-10-17 paper/splitting.tex
various revisions

@@
 Ground is also similar to the validity mechanism in ML5\,\cite{murphy05}.
+
+\subsection{Typing the Boundary Data Structure}
+
+One of the central features of our splitting algorithm is that
+it encodes the control flow behavior of the original 
+staged program into the boundary data structure.
+For instance, the {\tt case} and {\tt caseg} splitting rules 
+put injection tags on the boundary based on which branch was taken,
+and the {\tt fn} rule adds a roll tag to the boundary.
+As a result, the boundary value passed between the two output programs has a (potentially recursive) structure like a tree or list.
+
+This structure can be described with a type; for instance, the staged quickselect yields a binary search tree.
+Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
+However, we do not yet have a formal characterization of these boundary types that is total for all \lang\ programs.
+\footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
+replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.
+We chose the tags, however, in order to keep the typed interpretation more natural, even in absence of a formal result.
+}
 \end{abstrsyn}

### 2015-10-17 paper/semantics.tex
mostly formatting

@@
 \paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgment \mbox{$\steptwo e {e'}$}.
-Since first pass evaluation should not reduce stage two terms, this judgment does nothing but traverse $e$ to find \prev\ blocks, under which it performs in-place reductions.
+Since first pass evaluation should not reduce stage two terms, this judgment does nothing but traverse $e$ to find {\tt prev} blocks, under which it performs in-place reductions.
 A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
-To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
+To be a residual, a term must have no first stage subterms (equivalently, no {\tt prev}s), even within the body of a function or branches of a case.
 This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.
@@
 \begin{abstrsyn}
-\begin{theorem} [Progress] \leavevmode
+\begin{theorem} [Progress]
 \label{thm:progress} 
+\vspace{-0.75em}
 \begin{itemize} 
@@
 \end{theorem}
-\begin{theorem} [Preservation] \leavevmode
+\begin{theorem} [Preservation]
 \label{thm:preservation} 
+\vspace{-0.75em}
 \begin{itemize} 

### 2015-10-17 paper/splitting.tex
mostly formatting

@@
 \label{thm:total} 
+\vspace{-0.75em}
 \begin{itemize} 
@@
 \label{thm:unique}
+\vspace{-0.75em}
 \begin{itemize} 
@@
 \label{lem:stepCorrect}
+\vspace{-0.75em}
 \begin{itemize} 
@@
 
-\subsection{Example cases.}
+\subsection{Example Cases}
 
@@
 \paragraph{Reduction of {\tt pi1}.}
-Define $E = \pio {\exv{\tup {v_1,v_2}}}$ and  
-$C = \letin{\tup{y,z}}{\exv{\tup{\tup {i_1,i_2},\tup{}}}}{\tup{\pio {\exv y},\exv z}}$.
+Define 
+$C = \letin{\tup{y,z}}{\exv{\tup{\tup {i_1,i_2},\tup{}}}}{\tup{\pio {\exv y},\exv z}}$ and \\
+\mbox{$E = \pio {\exv{\tup {v_1,v_2}}}$}.
 We are given $\stepmix  E {\exv {v_1}}$,
@@
 $\splitone {\exv {v_1}} A {\exv {\tup{i_1,\tup{}}}} {\_.\exv{q_1}}$,
-and we need to show $\reduce C {\tup {i_1,\tup{}}}$
-and $\pio {\exv {\tup{q_1,q_2}}} \equiv q_1$.
+and we need to show \mbox{$\reduce C {\tup {i_1,\tup{}}}$}
+and \mbox{$\pio {\exv {\tup{q_1,q_2}}} \equiv q_1$}.
 Both of these can be derived directly.
-This pattern, where the outputs can be directly derived, extends to the {\tt pi2} and {\tt prev} reduction rules and all value marking rules.
+This pattern, where the outputs can be directly derived, extends to the {\tt pi2} and {\tt prev} reduction rules and all value promotion rules.
 
-\paragraph{Reduction of Application.} Let $E = \app {\exv{\fix f x e}} {\exv v}$,
-$E' = [\fix fxe, v/f,x] e$, \\ 
-$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$,
-$Q = \fix f {\tup{x,\roll l}} r$, and \\
+\paragraph{Reduction of Application.} Let \\
+$\begin{array}{cl}
+E &= \app {\exv{\fix f x e}} {\exv v}, \\
+E' &= [\fix fxe, v/f,x] e, \\ 
+I &= \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}, \\
+Q &= \fix f {\tup{x,\roll l}} r, \text{and}
+\end{array}$
 $C = \left[
@@
 We are given $\stepmix E {E'}$,
-$\splitone E A C {\tup{\_,\_,l}.\app{\exv Q}{\exv{\tup{q_1,l}}}}$,
+$\splitone E A C {\tup{\_,\_,l}.\app{\exv Q}{\exv{\tup{q_1,l}}}}$, \\
 $\splitone {E'} A {[I,i_1/f,x]c} {l.[Q,q_1/f,x]r}$,
 and $\reduce {[I,i_1/f,x]c} {\tup{i,b}}$.
-From this, we can directly derive $\reduce C {\tup{i,\tup {\tup{},\tup{},\roll b}}}$
-and $\app{\exv Q}{\exv{\tup{q_1,\roll b}}} \equiv [Q,q_1,b/f,x,l] r$.
+From this, we can derive $\reduce C {\tup{i,\tup {\tup{},\tup{},\roll b}}}$
+and \mbox{$\app{\exv Q}{\exv{\tup{q_1,\roll b}}} \equiv [Q,q_1,b/f,x,l] r$} directly.
 This pattern applies to all of the other reduction rules involving substitution,
 namely those for {\tt caseg}, and {\tt letg}.
-Observe that, in our formal version of the language, 
-{\tt caseg} is the only construct where information is added to the boundary value.
 
 \paragraph{Compatibility of {\tt pi1} at \bbtwo.}
-We are given 
-$\steptwo {\pio e} {\pio {e'}}$, 
-$\splittwo {\pio e} A p l {\pio r}$, 
+By the case, we are given 
+$\steptwo {\pio e} {\pio {e'}}$, \\
+\mbox{$\splittwo {\pio e} A p l {\pio r}$}, 
 $\splittwo {\pio {e'}} A {p'} {l'} {\pio {r'}}$,
@@
 Inversion of the first three yields
-$\steptwo e {e'}$, $\splittwo e A p l r$, and $\splittwo {e'} A {p'} {l'} {r'}$.
+$\steptwo e {e'}$, $\splittwo e A p l r$, and \mbox{$\splittwo {e'} A {p'} {l'} {r'}$}.
 Using \ref{lem:stepCorrect} inductively gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r$.
-From this, $[b/l]\pio r \equiv [b'/l']\pio {r}$ can be derived directly.
+From this, \mbox{$[b/l]\pio r \equiv [b'/l']\pio {r}$} can be derived directly.
 This pattern generalizes to all world \bbtwo\ compatibility rules.
@@
 \paragraph{Compatibility of {\tt pi1} at \bbonem.}
-We are given
-$\stepmix {\pio e} {\pio {e'}}$, 
-$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$, \\
-\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$},
-and \\ $\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$.
+By the case, we are given
+$\stepmix {\pio e} {\pio {e'}}$, \\
+\mbox{$\splitone {\pio e} A {\letin{\tup{y,z}}{c}{\tup{\pio {\exv y},\exv z}}} {l.\pio r}$}, \\
+\mbox{$\splitone {\pio {e'}} A {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {l'.\pio {r'}}$}, and \\
+\mbox{$\reduce {\letin{\tup{y,z}}{c'}{\tup{\pio {\exv y},\exv z}}} {\tup{i,b'}}$}.
 Inversion of the first three yields 
 $\stepmix e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, 
-and inversion of the reduction yields $\reduce {c'} {\tup{\tup{i,i_2},b'}}$.
+and inversion of the reduction yields, for some $i_2$, \mbox{$\reduce {c'} {\tup{\tup{i,i_2},b'}}$}.
 Using \ref{lem:stepCorrect} inductively gives $\reduce c {\tup{\tup{i,i_2},b}}$ and $[b/l]r \equiv [b'/l']r$.
@@
 We are given 
-$\stepmix {\pure e} {\pure {e'}}$, \\
+$\stepmix {\pure e} {\pure {e'}}$, 
 $\splitone {\pure e} A {\tup {e,\exv {\tup{}}}} {\_.\exv{\tup{}}}$, 

### 2015-10-17 paper/splitting.tex
formatting

@@
 
-The rules of the three splitting judgements are given in \ref{fig:termSplit}.
+The rules of the three splitting judgements are given in \ref{fig:valueSplit,fig:termSplit}.
 Since the rules are simply recursive on the structure of the term, 
@@
 A precise definition of the partial value splitting relation is given in
-\ref{fig:termSplit}.
+\ref{fig:valueSplit}.
 
@@
 
-% \subsection{Term Splitting at \bbonem}
-% \label{sec:split-one}
+\end{abstrsyn}
 
-% Given the definition of masking, we now describe how to split
-% \bbonem\ terms. The algorithm, given in \cref{fig:termSplit}, proceeds
-% recursively on the structure of~$e$. We derive a few cases here.
+\input{figures/splitting-term}
+\input{figures/splitting-misc}
 
-% \paragraph {Unit.} 
-% As a simple first case, consider splitting the unit value, $\tup{}$
-% which masks to $\mval {\tup{}} {\tup{}}$.
-% %Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
-% Splitting must produce a $c$ and $l.r$ satisfying
-% \[
-% 	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
-% \]
-% Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
-% Splitting $\pure{e}$ follows in similar fashion, except the first 
-% component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.
-
-% \paragraph {Injections.}
-% Now consider splitting the non-terminal $\inl {e}$, where $\diaone
-% {e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
-% \[
-% 	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
-% \]
-% Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
-% splitting must produce a $c'$ and $l.r'$ satisfying
-% \[
-% 	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
-% \]
-% Our algorithm generates a first pass $c'$ that produces the same
-% boundary value as splitting $e$:
-% \[
-% c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
-% \]
-% Splitting the injection also generates the same resumer as splitting
-% the subterm $e$ (that is, $l.r' = l.r$).
-
-% \paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
-% where $\diaone {e_k} {\xi_k} {v_k}$ and 
-% $\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
-% Inductively:
-% \[
-% 	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
-% \]
-% Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
-% and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
-% splitting generates:
-% \[
-% 	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
-% \]
-% The resulting boundary value is the tuple of the boundary values of
-% the subterms, $b = \tup{b_1,b_2}$.  The value is 
-% constructed using
-% \[
-% c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
-% \]
-% and deconstructed in the resumer with a tuple pattern:
-% \[
-% l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
-% \]
-% \paragraph {Case.}
-% Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
-% Here we consider only the case where
-% $\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
-% (The \texttt{inr} case is analogous.)
-
-% \noindent
-% Inductively,
-% \[
-% 	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
-% \]
-% Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
-% and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
-% Therefore:
-% \begin{gather*}
-% 	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
-%   {\tup{i_2,b_2}} \\
-% \text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
-% \end{gather*}
-% (Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
-% When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
-% of $v_1$ into $c_2$ and $l_2.r_2$.)
-
-% The entire case evaluates as
-% $\diaone {\caseof {e_1} {x_2.e_2} {x_3.e_3}} {\xi_1,\xi_2} {v_2}$, so splitting should produce 
-% some $c'$ and
-% $l'.r'$ such that 
-% \[
-%   \reduce {c'} {\tup{i_2,b'}} \text{ and }
-%   [b'/l']r' \equiv q_2
-% \]
-% We choose
-% \[
-% c' =
-% {\left(
-% \talllet{\tup{y_1,z_1}}{c_1}{
-% \tallcase{y_1}
-% {x_2.\letin{\tup{y_2,z_2}}{c_2}{\tup{y_2,\tup{z_1,\inl{z_2}}}}}
-% {x_3.\cdots\ttrpar}
-% }\right)}
-% \]
-% \[
-% l'.r' =
-% \tup{l_1,l_b}.\letin{z}{r_1}{\caseof{l_b}{l_2.[z/x_2]{r_2}}{l_3.\cdots}}
-% \]
-% The term $c'$ reduces to
-% $\tup{i_2,\tup{b_1,\inl{b_2}}}$, so 
-% $b' = \tup{b_1,\inl{b_2}}$. Therefore
-% $[b'/l']r'$ reduces to
-% $[[b_1/l_1]r_1/x_2][b_2/l_2]r_2$, exactly the 
-% $[q_1/x_2][b_2/l_2]r_2$ required.
-
-% More intuitively, $c'$ cases on $i_1$ to run the first branch of first-stage
-% code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
-% {b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
-% the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
-% similarly constrained by the $\texttt{inr}$ case.
-
-% %Recovering first stage divergence is easy, since for $\splitone {e_1} ? {c_1} {l_1.r_1}$, 
-% %$\reduce {e_1} {\inl {v_1}}$ implies $\reduce {c_1} {\tup{\inl {i_1},b_1}}$.
-
-% \paragraph {Function.} 
-% Function introduction has a $\tup{}$ boundary value, since functions
-% are already fully reduced in our semantics.  However, since the body
-% of a function may itself be multistage, splitting must descend into
-% it.  The first pass generated by splitting produces an $i$ that is a new
-% function formed from the first-stage part of the original body.  The
-% resumer is a new function formed out of the second-stage part of the
-% original body.  It is the responsibility of the application site to
-% save the precomputation of the function body and pass it to the
-% resumer version of the function.
-
-% \subsection{Term Splitting at \bbtwo}
-
-% Because world \bbtwo\ terms in \lang\ reduce to monostage residuals
-% (as opposed to partial values), term splitting at world
-% \bbtwo\ (specified by the $\splittwosym$ relation in
-% \cref{fig:termSplit}) is simpler than that of \bbonem.
-
-% Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
-%   \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
-% Note that for any $\diatwo e q$, the residual $q$ must have the form
-% $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
-% nullary or the $q_i$ have binders.
-% For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
-% \[
-% 	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
-% \]
-% Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
-% \[
-% 	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
-% \]
-% Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
-% $l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
-
-% %Again, this holds even for nullary terms and under binders.
-% In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
-% \[
-% \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
-% \]
-% Note how the precomputations of $e_2$ and $e_3$ are pulled out from
-% underneath the binder into a tuple that is always executed in the
-% first stage. This occurs regardless of what branch is eventually
-% executed in the second stage.
-
-% \paragraph{Staging constructs.}
-% Since \texttt{next} and \texttt{prev} do nothing more than convert between 
-% second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
-% their splitting rules are relatively simple.
-% Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
-% \[
-% 	\reduce p b \text{ and } [b/l]r \equiv q
-% \]
-% Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
-% and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
-% \[
-% 	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
-% \]
-% Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
-% Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
-
+\begin{abstrsyn}
 \subsection {Role of World \bbonep}
@@
 Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
-However, we do not yet have a formal characterization of these boundary types that is total for all \lang\ programs.
+However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs.
 \footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
@@
 \end{abstrsyn}
-
-\input{figures/splitting-term}
-\input{figures/splitting-misc}

### 2015-12-15 paper/overview.tex
quick edit pass based on reviews

@@
 innovation---augmenting the tree by storing at each node the size of its left
-subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
-nearly linear-time improvement in $m$.
+subtree. For a sufficiently balanced tree, this approach has 
+(expected) runtime $\Theta(n\log{n} + m\log{n})$, 
+a nearly linear-time improvement in $m$.
 
@@
 Splitting \texttt{qss} yields a two-part program that creates a
-probabilistically balanced, augmented binary search tree (which is isomorphic to
-a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
+probabilistically balanced, augmented binary search tree 
+as an intermediate data structure. In
 particular, its first part (\texttt{qs1}) constructs such a binary search tree
@@
 \emph{all} of the recursive calls, performing these comparisons essentially
-walks the tree, using the rank along with the size data \texttt{i} in order to
+walks the tree, using the rank along with the size data \texttt{i} to
 look up the \texttt{k}th leftmost node in the tree.

### 2015-12-15 paper/semantics.tex
quick edit pass based on reviews

@@
 
-The example code in this paper makes liberal use of $\rmint$s and various
+The example code in this paper uses an extension of the formalized \lang.
+In particular, it makes liberal use of $\rmint$s and various
 functions on these, as well as a function $\pause$ which takes a $\curr\rmint$
@@
 \paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
-Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code where it performs a reduction. 
+Thus, $\steppure e {e'}$ traverses into subterms to find the left-most unevaluated code where it performs a reduction. 
 A ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.
@@
 Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a reduction.
-It also descends into {\tt gr} and {\tt next} blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
+It also descends into {\tt gr} and {\tt next} blocks, 
+using one of the other two step judgements ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
 The value form for \bbonem, called a partial value, comprises units, tuples, functions, {\tt gr} blocks of ground values, and {\tt next} blocks containing \emph{only a stage two variable}.

### 2015-12-15 paper/splitting.tex
One old change from paper submission.
quick edit pass based on reviews

@@
 \label{lem:hoistCorrect}
-If $\lift e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, then $c = c'$, and $l.r \equiv l'.r'$.
+If $\lift e {e'}$, $\splitone e A c {l.r}$, 
+and $\splitone {e'} A {c'} {l'.r'}$, 
+then $c = c'$, 
+and $l.r \equiv l'.\letin y q {r'}$.
 \end{lemma}
@@
 Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
-However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs.
+However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
+though we plan to pursue this in future work.
 \footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 

### 2016-01-01 paper/semantics.tex
fix footnote mark spacing

@@
 we have chosen to statically distinguish between values and general computations,
-using an underline constructor (``$\exv v$'') which explicitly note the parts of a computation that have been reduced to a value.
+using an underline constructor (``$\exv v$'') which explicitly note the parts of a computation that have been reduced to a value.%
 \footnote{In this presentation, the value/computation distinction is encoded intrinsically at the syntactic level.
@@
 In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
-where it is reified into a {\tt let} statement.
+where it is reified into a {\tt let} statement.%
 \footnote{Because a program is a term at \bbtwo, this {\tt prev} always exists.

### 2016-01-01 paper/splitting.tex
fix footnote mark spacing

@@
 However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
-though we plan to pursue this in future work.
+though we plan to pursue this in future work.%
 \footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 

### 2016-01-06 paper/intro.tex
Carlo's comments

@@
 evaluation, splitting is a static program transformation
-(``metastatic'' in partial evaluation terms) and does not require
+(``metastatic'' in partial evaluation terminology) and does not require
 dynamic code generation.

### 2016-01-06 paper/semantics.tex
Carlo's comments

--- 2015-12-15 paper/semantics.tex
@@
 Although \lang\ describes computations that occur in two stages,
-we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
+we find it helpful for the specification of splitting to codify terms using one of three \emph{worlds}.
 A world is essentially a slightly finer classification than a stage.
@@
 and \bbonep\ for \emph{ground} first-stage computations, which may not.
-The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
+The distinction between these two first-stage worlds is necessary 
+for the splitting algorithm to produce efficient outputs
 and will be discussed in \ref{sec:needGround}.
@@
 of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
-of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.
+of worlds \bbtwo\ and \bbonep, respectively, to types at world \bbonem.
 
@@
 \end{lstlisting}
-Depending on what {\tt P} evaluates to in the first stage, whole term will step to either {\tt next\{0\}} or {\tt next\{1\}}.
+Depending on what {\tt P} evaluates to in the first stage, 
+the whole term will step to either {\tt next\{0\}} or {\tt next\{1\}}.
 In this sense, we can see {\tt case} (at world \bbonem) and {\tt caseg} as the constructs that facilitate

### 2016-01-06 paper/splitting.tex
Carlo's comments

--- 2015-12-15 paper/splitting.tex
@@
 and since transfer of information between the stages occurs in the first pass of evalution,
-we know that the second stage components of $v$ can no longer depend on its first stage components.
+we know that the second-stage components of $v$ can no longer depend on its first-stage components.
 Analogously, this must also hold for the output of splitting, 
@@
 though we plan to pursue this in future work.
-\footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
+\footnote{Observe that in the outputs of splitting we could have omitted the {\tt roll} tag from functions or 
 replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.

### 2016-01-06 paper/splitting.tex

@@
 However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
-though we plan to pursue this in future work.
+though we plan to pursue this in future work.%
 \footnote{Observe that in the outputs of splitting we could have omitted the {\tt roll} tag from functions or 

### 2016-01-06 paper/paper.tex
format fiddling
changed paper style
various revisions
mostly formatting
formatting
quick edit pass based on reviews
cleaning and authors

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{llncs}
 
+\usepackage{titlesec}  
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{stmaryrd}
@@
 
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem*{theorem*}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{remark}[theorem]{Remark}
+%\newtheorem{theorem}{Theorem}[section]
+%\newtheorem{theorem*}{Theorem}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{definition}[theorem]{Definition}
+%\newtheorem{remark}[theorem]{Remark}
 
 \input{macros}
-\input{macros-ref}
-\input{remarks}
+%\input{remarks}
 
+\DeclareMathSymbol{\Gamma}{\mathalpha}{operators}{0}
 \definecolor{stageOneColor}{RGB}{120,30,0}
-\definecolor{stageTwoColor}{RGB}{0,0,127}
+\definecolor{stageTwoColor}{RGB}{0,30,127}
 \definecolor{stageMonoColor}{RGB}{0,120,0}
@@
                  {$}{{$\fut$}}1}
+\begin{comment}$\end{comment}
+\lstset{aboveskip=3pt,belowskip=3pt}
 
@@
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
 %\doi{nnnnnnn.nnnnnnn}
 
+%
 % Uncomment one of the following two, if you are not going for the 
@@
 \title{Automatically Splitting a Two-Stage Lambda Calculus}
-\subtitle{(Supplementary Material in Appendix)}
-
-\authorinfo{~}%Nicolas Feltman \and Carlo Angiuli \and Umut A. Acar \and Kayvon Fatahalian}
-           {~}%Carnegie Mellon University}
-           {~}%\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}
+%
+\titlerunning{Automatic Splitting} 
+%
+\author{Nicolas Feltman\inst{1} \and Carlo Angiuli\inst{1}
+\and Umut Acar\inst{1,2} \and Kayvon Fatahalian\inst{1}}
+%
+\authorrunning{Feltman et al.} % abbreviated author list (for running head)
+%
+%%%% list of authors for the TOC (use if author list has to be modified)
+\tocauthor{Nicolas Feltman, Carlo Angiuli, Umut Acar, and Kayvon Fatahalian}
+%
+\institute{Carnegie Mellon University \and Inria Paris-Rocquencourt}
 
@@
 
-%\input{umut}
-
 %\acks
@@
 
+\newpage
 \bibliography{paper}
-\bibliographystyle{abbrvnat}
+\bibliographystyle{splncs03}
 
@@
 
-\newpage\appendix
-\input{appendix}
+%\newpage\appendix
+%\input{appendix}
 

### 2016-01-06 paper/abstract.tex
fixed hyphenation of stage

@@
 calculus \lang\ with a $\fut$ modality denoting computation at a later
-stage, and a $\curr$ modality noting purely first stage code.  Most
+stage, and a $\curr$ modality noting purely first-stage code.  Most
 notably, the algorithm splits mixed-stage recursive and higher-order

### 2016-01-06 paper/relatedwork.tex
fixed hyphenation of stage

@@
 
-However unlike splitting, where the goal is to emit code where first stage results
-are computed once and then reused in multiple invocations of second
-stage execution, the second (dynamic) function produced by binding time separation
+However unlike splitting, where the goal is to emit code where first-stage results
+are computed once and then reused in multiple invocations of second-stage execution, 
+the second (dynamic) function produced by binding time separation
 does not use the results of the first; instead,
@@
 %specialization phase of partial evaluation.
-it has access to the first stage inputs and recomputes all required
-first stage computations. As noted by Knoblock and
+it has access to the first-stage inputs and recomputes all required
+first-stage computations. As noted by Knoblock and
 Ruf\,\cite{knoblock96}, it may be possible to modify the program

### 2016-01-06 paper/semantics.tex
cleaning and authors
fixed hyphenation of stage

@@
 of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
-of worlds \bbtwo\ and \bbonep, respectively, to types at world \bbonem.
+of worlds \bbtwo\ and \bbonep, respectively, as types at world \bbonem.
 
@@
 A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
-To be a residual, a term must have no first stage subterms (equivalently, no {\tt prev}s), even within the body of a function or branches of a case.
-This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.
+To be a residual, a term must have no first-stage subterms (equivalently, no {\tt prev}s), even within the body of a function or branches of a case.
+This implies that $\stepsym \bbtwo$ must proceed underneath second-stage binders.
 
@@
 In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
-representing the second stage binders under which we are evaluating.
+representing the second-stage binders under which we are evaluating.
 Note how the progress theorem for world \bbonem\ states that every well-typed term must take 

### 2016-01-06 paper/splitting.tex
fixed hyphenation of stage

@@
 The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
-Because hoisting steps are nothing but rearrangement of second stage code,
+Because hoisting steps are nothing but rearrangement of second-stage code,
 this lemma can use the strong requirement of identical combined terms.
@@
 Since direct staged term evaluation (\ref{sec:semantics})
-reduces all first stage terms to value forms without any remaining stage two work,
+reduces all first-stage terms to value forms without any remaining stage two work,
 the distinction between \bbonem\ and \bbonep\ is unnecessary.
-In contrast, when performing program transformations before the first stage inputs are known,
+In contrast, when performing program transformations before the first-stage inputs are known,
 it is valuable to form a clear distinction between ground and mixed terms.

### 2016-01-08 paper/abstract.tex
various little changes

@@
 programs in a language featuring recursion and higher-order features
-enables non-trivial algorithmic transformations that asymptotically
-improve code efficiency by nearly a linear factor and also
+enables non-trivial algorithmic transformations that
+improve code efficiency by up to a linear factor and also
 facilitates modular expression of staged programs.

### 2016-01-08 paper/intro.tex
various little changes

@@
 %
-In this scenario, a multivariate function is applied many times with
-one argument fixed, so it is advantageous to find the computations in
-$f$ that depend on $x$ and then stage program execution so they are
-performed only once at the beginning of execution.
+Observe that this would be wasteful if $f(x,y_i)$ does a significant 
+amount of work that does not depend on $y_i$.
+In such a case, it would be advantageous to find the computations in
+$f$ that depend only on $x$ and then stage execution of $f$ so they are
+performed only once at the beginning of $F$.
 
@@
 \end{lstlisting}
-eliminating the need to compute $m$ times those parts of $f(v,-)$
+This eliminates the need to compute $m$ times those parts of $f(v,-)$
 which do not depend on the second argument.

### 2016-01-08 paper/overview.tex
Carlo's comments
various little changes

@@
 on a list~\texttt{l}. To this end, we can use the quickselect
-algorithm~\cite{Hoare:1961}, which given a list~\texttt{l} and a rank
-\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
+algorithm~\cite{Hoare:1961}, which given a list~\texttt{l} and an integer
+\texttt{k}, returns the element of \texttt{l} with rank \texttt{k} (i.e. the $k$th-largest element).
 As implemented in an ML-like language in \ref{fig:qs-unstaged},
@@
 If the index is out of range, a default value of \texttt{0} is returned.
-Assuming that the input is uniformly
-randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
-runs in expected $O(n)$ time.
+Assuming that the input list, with size $n$, is uniformly
+randomly ordered (which can be achieved by pre-permuting it), \texttt{qs}
+runs in expected $\Theta(n)$ time.
 %
@@
 %
-Unfortunately, this approach requires $O(n \cdot m)$ time.
+Unfortunately, this approach requires $\Theta(n \cdot m)$ time.
 
@@
 A first cut at a solution might be to first sort \texttt{l} into
-\texttt{s} and then perform lookups in \texttt{s}.
-Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
+a sorted list \texttt{s} and then perform lookups in \texttt{s}.
+Unfortunately, lookups in \texttt{s} are also $\Theta(n)$ on average, since
 \texttt{s} is a list, leading to no improvement in efficiency.
@@
 innovation---augmenting the tree by storing at each node the size of its left
-subtree. For a sufficiently balanced tree, this approach has 
-(expected) runtime $\Theta(n\log{n} + m\log{n})$, 
+subtree. For a sufficiently balanced tree, 
+which one would get with a random choice of pivot, this approach has 
+expected runtime $\Theta(n\log{n} + m\log{n})$, 
 a nearly linear-time improvement in $m$.
@@
 second-stage code. Additionally, $\texttt{gr}$ (\emph{ground}) annotations 
-mark certain first stage components as being purely first-stage, rather than mixed-stage.
+mark certain first-stage components as being purely first-stage, rather than mixed-stage.
 %

### 2016-01-08 paper/abstract.tex
changes to front

@@
 enables non-trivial algorithmic transformations that
-improve code efficiency by up to a linear factor and also
-facilitates modular expression of staged programs.
+improve code efficiency and also facilitates modular 
+expression of staged programs.
 

### 2016-01-08 paper/examples.tex
mostly formatting
quick edit pass based on reviews
changes to front

@@
 %
-The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
+The body of \texttt{dot} is an \rmint\ term at world \bbtwo\ containing an
 \rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
@@
 This is an instance where stage-splitting is more practical than partial
-evaluation---if $l$ is large, partially evaluating \texttt{quickselect} requires
+evaluation; if $l$ is large, partially evaluating \texttt{quickselect} requires
 runtime generation of a huge amount of code simultaneously encoding the tree and
@@
 \texttt{qs2} performs expected $\Theta(\log n)$ work.  This results in a net
-speedup over standard quickselect if we perform $\Omega(n / \log n)$ queries on
+speedup over standard quickselect if we perform $\omega(\log n)$ queries on
 the same list.
@@
 1`datatype object = ...`
-2`atsignnext{
-  type coord = int * int
-  type color = int
-}`
+2`atsignnext{  type coord = int * int
+        type color = int        }`
 1`type pipeline = object * $`2`coord`1` -> `$2`color`

### 2016-01-08 paper/overview.tex
changes to front

@@
 We can attempt to improve on this algorithm by factoring out computations shared
-between these calls to \texttt{qs}. In particular, the comparisons performed by
-\texttt{part} depend only on \texttt{l}, although different calls to
-\texttt{part} are made by each call to \texttt{qs}.
-
+between these calls to \texttt{qs}, 
+namely the parts that depend only on \texttt{l}.
 A first cut at a solution might be to first sort \texttt{l} into
-a sorted list \texttt{s} and then perform lookups in \texttt{s}.
-Unfortunately, lookups in \texttt{s} are also $\Theta(n)$ on average, since
-\texttt{s} is a list, leading to no improvement in efficiency.
+a sorted list \texttt{s}, at cost $\Theta(n \log n)$, and then perform lookups in \texttt{s}.
+Unfortunately, lookups in \texttt{s} are $\Theta(n)$ on average, 
+since \texttt{s} is a list.
+Thus, the total cost is $\Theta(n \log n + mn)$, which is no improvement.
 
 A more sophisticated solution is to construct a binary search
-tree out of \texttt{l} and then simply look for the \texttt{k}th
-leftmost element of that tree. To do this efficiently requires one more
-innovation---augmenting the tree by storing at each node the size of its left
-subtree. For a sufficiently balanced tree, 
-which one would get with a random choice of pivot, this approach has 
-expected runtime $\Theta(n\log{n} + m\log{n})$, 
-a nearly linear-time improvement in $m$.
+tree out of \texttt{l}, again at cost $\Theta(n\log{n})$, and then simply look for the \texttt{k}th
+leftmost element of that tree. 
+Doing lookups efficiently, in $\Theta(\log{n})$ time, 
+requires one more innovation---augmenting the tree 
+by storing at each node the size of its left subtree. 
+Thus in total this approach has expected runtime $\Theta(n\log{n} + m\log{n})$.
 
-In both cases, we replace our simple solution with a two-part algorithm which
-first processes the input into a lookup data structure, and then subsequently
-performs fast lookups. Improving algorithms in this way is highly non-trivial,
-as it requires reasoning about intricate algorithmic concerns, and implementing
-more complex algorithms and data structures.
+Is this method, wherein we precompute a data structure, better than the direct $\Theta(mn)$ method from above? 
+The answer depends on the relationship between the number of lookups $m$ and the size of the list $n$.
+If $m$ is constant, then the direct method is superior for sufficiently large $n$.
+This is because the precomputed method does unnecessary work 
+sorting parts of the list where there are no query points.
+However, if the number of queries grows with the size of the list, 
+specficially in $\omega(\log n)$,
+then the precomputed method will be asymptotically faster.
 
-In this paper, we present a splitting algorithm which automatically transforms
-staged programs into multi-part programs. In certain cases, this algorithm even
-improves input algorithms in the way described above!
+Rewriting algorithms in this way---to 
+precompute some intermediate results that depend on constant (or slowly-varying) inputs---is 
+highly non-trivial,
+as it requires implementing more complex data structures and algorithms.  
+In this paper, we present a splitting algorithm which does much of this task automatically.
 

### 2016-01-08 paper/splitting.tex
mostly compression

@@
 where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
-and ``$\equiv$'' indicates a monostage equivalence, which we cover later.
+and ``$\equiv$'' indicates a monostage equivalence, which is defined in \ref{fig:equiv}.
 
@@
 
+\end{abstrsyn}
+
+\input{figures/splitting-values}
+
+\begin{abstrsyn}
+
 \subsection{World \bbonem\ Term Splitting}
@@
 
+\end{abstrsyn}
+
+\input{figures/splitting-term}
+\input{figures/splitting-misc}
+
+\begin{abstrsyn}
+
 \subsection{Example Cases}
@@
 from the proof of \ref{lem:stepCorrect}.
-The equivalence relation ``$\equiv$'' is defined in \ref{fig:equiv}.
 
@@
 
-\end{abstrsyn}
-
-\input{figures/splitting-term}
-\input{figures/splitting-misc}
-
-\begin{abstrsyn}
 \subsection {Role of World \bbonep}

### 2016-01-08 paper/intro.tex
mostly compression
small edits

@@
 \input{figures/quickselect}
-Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
+Consider a function $F$ which, given $x, y_1,\dots,y_m$, computes
 $f(x,y_i)$ for each $y_i$:
@@
 %
-Observe that this would be wasteful if $f(x,y_i)$ does a significant 
+Observe that this implementation would be wasteful if $f(x,y_i)$ does a significant 
 amount of work that does not depend on $y_i$.

### 2016-01-08 paper/overview.tex
small edits

@@
 We can attempt to improve on this algorithm by factoring out computations shared
-between these calls to \texttt{qs}, 
-namely the parts that depend only on \texttt{l}.
-A first cut at a solution might be to first sort \texttt{l} into
-a sorted list \texttt{s}, at cost $\Theta(n \log n)$, and then perform lookups in \texttt{s}.
-Unfortunately, lookups in \texttt{s} are $\Theta(n)$ on average, 
-since \texttt{s} is a list.
-Thus, the total cost is $\Theta(n \log n + mn)$, which is no improvement.
-
-A more sophisticated solution is to construct a binary search
-tree out of \texttt{l}, again at cost $\Theta(n\log{n})$, and then simply look for the \texttt{k}th
+between these calls to \texttt{qs}.
+In particular, we can construct a binary search
+tree out of \texttt{l}, at cost $\Theta(n\log{n})$, and then simply look for the \texttt{k}th
 leftmost element of that tree. 
@@
 
-Is this method, wherein we precompute a data structure, better than the direct $\Theta(mn)$ method from above? 
+Is this method, wherein we precompute a data structure, better than the direct $\Theta(mn)$ method? 
 The answer depends on the relationship between the number of lookups $m$ and the size of the list $n$.
@@
 Rewriting algorithms in this way---to 
-precompute some intermediate results that depend on constant (or slowly-varying) inputs---is 
-highly non-trivial,
+precompute some intermediate results that depend on constant (or infrequently-varying) inputs---is non-trivial,
 as it requires implementing more complex data structures and algorithms.  
@@
 %
-Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
+Lastly, it records~\texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
 (the pivot/head of the list) in the boundary structure, because those

### 2016-01-08 paper/semantics.tex
mostly compression
small edits

@@
 It is sensible to extend $\pause$ to all base types and to products and sums
-thereof. This is related to the notion of {\em mobility} in \cite{murphy05} and
-{\em stability} in \cite{krishnaswami13}.}
+thereof. This is related to the notion of {\em mobility} (\cite{murphy05}).}
 
@@
 
-We can also directly observe the reuse of first-stage computations across
-multiple second-stage computations. To run many order statistics queries
-$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
+Moreover, the reuse of first-stage computations across
+multiple second-stage computations can even be encoded within \lang. 
+The following program runs many order statistics queries
+$\texttt{k1},\dots,\texttt{km}$ on the same list:
 \begin{lstlisting}
-2`let fun qs k = prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
-2`in (qs k1,...,qs km)`
+2`prev{`1` let val list = `3`gr{`1`[7,4,2,5,9,...,3]`3`}`1` in 
+next{ `2`let fun qs k = prev{`1`qss (list,next{`2`k`1`})2`}`
+      2`in (qs k1,...,qs km)`1`}`2`}`
 \end{lstlisting}

### 2016-01-09 paper/appendix.tex
line ending
working on composition examples

@@
 splitting algorithm.
-
-\subsection {Mixed Map Combinator}
-
-In \ref{sec:exampleQS}, we showed the output of splitting quickselect
-applied to one list and $m$ queries.  Here, we generalize to mapping
-quickselect onto a list of queries. Note that we have to define a
-datatype for integer lists at the second stage, which is given by
-\texttt{list2}.
-\begin{lstlisting} 
-1`atsignnext{`2`
-  datatype list2 = Empty2 | Cons2 of int * list2
-`1`}
-
-map : (^list * $`2`int`1` -> $`2`int`1`)
-      -> ^list * $`2`list2`1` -> $`2`list2`1`
-fun map f (l, q) = next {`2`
-  let 
-  fun m Empty2 = Empty2
-    | m (Cons2(h,t)) =
-                    Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
-  in m prev{`1`q`2`}`1`}
-
-val mapqss = map qss`
-\end{lstlisting}
-The \texttt{map} function splits into the following two functions:
-\begin{lstlisting} 
-2`datatype list2 = Empty2 | Cons2 of int * list2
-
-`1`fun map1 f = (fn (l,()) => ((),#2 (f (l,()))),())
-`2`fun map2 (f,()) ((l,q), p : tree) =
-  let 
-  fun m Empty2 = Empty2
-    | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
-  in m q
-
-`1`val (mapqss1,()) = map1 qs1
-`2`val mapqss2 = map2 (qs2, ())`
-\end{lstlisting}
-%
-Indeed, \texttt{qs1}, which is passed to \texttt{map1}, is evaluated only in the
-first pass, whereas \texttt{qs2}, which is passed to \texttt{map2}, is evaluated
-in the second pass, once per query \texttt{k}.
 
@@
 datatype letter2 = A2 | B2 | C2
-datatype string2 = EmptyS2 | ConsS2 of letter2 * string2
+datatype string2 = EmptyS2 
+				 | ConsS2 of letter2 * string2
 `1`}

### 2016-01-09 paper/examples.tex
working on composition examples

@@
 Our algorithm splits \texttt{dot} into the two functions:
-%
 \begin{lstlisting} 
@@
 1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`gr{`1`e`3`}`1` : ^int) = 
-  if `3`gr{`1`e == 0`3`}`1` then 
-    next{`2`literalone`1`}
-  else if `3`gr{`1`(e mod 2) == 0`3`}`1` then
-    exp(next{`2`b*b`1`},`3`gr{`1`e/literaltwo`3`}`1`)
-  else 
-    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`gr{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
+ if      `3`gr{`1`e == 0`3`}`1`       then next{`2`literalone`1`}
+ else if `3`gr{`1`e mod 2 == 0`3`}`1` then exp(next{`2`b*b`1`},`3`gr{`1`e/literaltwo`3`}`1`)
+ else                     next{`2`b*prev{`1`exp(next{`2`b*b`1`},`3`gr{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
 \end{lstlisting}
@@
 1`fun exp1 (b : unit, e : int) =
-  if e==0 then
-    ((), Zero)
-  else if (e mod 2)==0 then 
-    ((), Even (#2 (exp1 ((), e/2))))
-  else 
-    ((), Odd (#2 (exp1 ((), (e-1)/2))))`
+  if      e == 0       then ((), Zero)
+  else if e mod 2 == 0 then ((), Even (#2 (exp1 ((), e/2))))
+  else                      ((), Odd  (#2 (exp1 ((), (e-1)/2))))`
 
@@
 runtime generation of a huge amount of code simultaneously encoding the tree and
-traversal algorithm. (Avoiding the code blowup by not expanding some calls to
+traversal algorithm. (Avoiding the code blowup, by not expanding some calls to
 \texttt{part}, would result in duplicating first-stage computations.)
@@
 With \texttt{gr} annotations, however, this can be prevented.
+%
 
-To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
-list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
-the second pass, evaluate \texttt{qs2} on that tree and each rank.
+\subsection {Mixed Map Combinator}
+
+[working on this section!]
+
+In \ref{sec:exampleQS}, we showed the output of splitting quickselect
+applied to one list and $m$ queries.  Here, we generalize to mapping
+quickselect onto a list of queries. Note that we have to define a
+datatype for integer lists at the second stage, which is given by
+\texttt{list2}.
 \begin{lstlisting} 
-1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
-in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
+1`atsignnext{`2`
+  datatype list2 = Empty2 | Cons2 of int * list2
+`1`}
+
+map : (^list * $`2`int`1` -> $`2`int`1`)
+      -> ^list * $`2`list2`1` -> $`2`list2`1`
+fun map f (l, q) = next {`2`
+  let 
+  fun m Empty2 = Empty2
+    | m (Cons2(h,t)) =
+                    Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
+  in m prev{`1`q`2`}`1`}
+
+val mapqss = map qss`
+\end{lstlisting}
+The \texttt{map} function splits into the following two functions:
+\begin{lstlisting} 
+2`datatype list2 = Empty2 | Cons2 of int * list2
+
+`1`fun map1 f = (fn (l,()) => ((),#2 (f (l,()))),())
+`2`fun map2 (f,()) ((l,q), p : tree) =
+  let 
+  fun m Empty2 = Empty2
+    | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
+  in m q
+
+`1`val (mapqss1,()) = map1 qs1
+`2`val mapqss2 = map2 (qs2, ())`
 \end{lstlisting}
 %
+Indeed, \texttt{qs1}, which is passed to \texttt{map1}, is evaluated only in the
+first pass, whereas \texttt{qs2}, which is passed to \texttt{map2}, is evaluated
+in the second pass, once per query \texttt{k}.
 
@@
 
-Programs featuring staged structure are particularly important in
-real-time computer graphics since modern graphics architectures
-\emph{require} that graphics computations be organized into a pipeline
+Composable staged programs are particularly important the domain of real-time graphics.
+This need arises because modern graphics architectures actually
+\emph{require} that graphics computations be structured as a pipeline
 of stages which perform increasingly fine-grained computations (e.g.,
-per object, per screen region, per pixel). Many computations in later
+per-object, per-screen region, per-pixel). Many computations in later
 stages use the results from a single computation in an earlier
@@
 
-While performant, manually partitioning programs by stage results in
-complex code where invariants must hold across different stages, and
+The standard way to program these graphics pipelines is to define one program (usually called a \emph{shader}) per stage.
+In other words, the programmer is expected to manually split their multi-stage programs.
+Yet manually partitioning programs by stage results in
+complex code where invariants must hold across different stages and
 local changes to the logic of one stage may require changes to that of
-upstream stages other stages to ``plumb'' its inputs through the
-pipeline.  Thus, manual separation of programs by stage breaks key
-programming abstractions such as composition and modularity.  Graphics
-researchers therefore have suggested using mixed-staged programs
-\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express logic related to
-a single application concern conveniently as a single staged function
-(referred to as a pipeline program)\,\cite{Foley:2011}.
+upstream stages to communicate required inputs through the pipeline. 
+This harms composition and modularity.  
+Graphics researchers therefore have suggested using mixed-staged programs
+\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express graphics program logic 
+for the entire pipeline as a single staged function\,\cite{Foley:2011}.
 
-Pipeline programs can be directly expressed as staged functions in
-\lang\ with splitting used to generate monostaged code required for
-each graphics pipeline stage. Composition of such functions in
-\lang\ allows more complex pipelines to be expressed in a modular
-programming style that is becoming increasingly desirable as graphics
-programs increase in complexity\,\cite{Foley:2011,He:2014}.
-
-A simple example of composition of staged functions (representing graphics
-pipeline programs) is given below.  The code defines a pipeline as a
-staged function that takes an object definition in the first stage
-(\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second
-stage and emits the color of the object at the specified pixel. The
-code defines a pipeline combinator \texttt{shade} which modulates the
-results of the object's reflectance at a given pixel with the results
-of a function that computes the object's albedo (surface color).
+As a language, \lang\ is well suited for specification of such functions,
+and we give a simple example below.
+In it, we consider a graphics pipeline program to be a
+staged function that takes an object definition in the first stage (\texttt{object})
+and a pixel coordinate in the second stage ($\fut$\texttt{coord}), 
+and emits the color of the object at the specified pixel ($\fut$\texttt{color}). 
+Given two such multistage functions, we then define a combinator
+that multiplies their results pointwise in the second 
+stage.\footnote{Although this is a general pointwise multiplication combinator, 
+the variable names suggest a possible interpretations of the inputs and output:
+the first input function calculates the albedo of an object (a measure of how much light it reflects),
+the second input calculates the object's incoming light,
+and so the output (the product of these terms) is a function that calculates 
+the outgoing light from an object.}
 %
@@
 2`atsignnext{  type coord = int * int
-        type color = int        }`
+        type color = ...        }`
 1`type pipeline = object * $`2`coord`1` -> `$2`color`
@@
 %
-Whereas prior systems provide combinators like \texttt{par} as
-language features\,\cite{Foley:2011} to support composition, \lang\ is
-powerful enough to define those combinators in-language.
+A key strength of \lang\ is that it allows users to define combinators like this one as higher order functions.
+Prior work\,\cite{Foley:2011} does not support higher order functions.

### 2016-01-09 paper/semantics.tex
working on composition examples

@@
 \end{lstlisting}
-
 Moreover, the reuse of first-stage computations across
@@
 2`prev{`1` let val list = `3`gr{`1`[7,4,2,5,9,...,3]`3`}`1` in 
-next{ `2`let fun qs k = prev{`1`qss (list,next{`2`k`1`})2`}`
-      2`in (qs k1,...,qs km)`1`}`2`}`
+next{ `2`let fun lookup k = prev{`1`qss (list,next{`2`k`1`})2`}`
+      2`in (lookup k1,...,lookup km)`1`}`2`}`
 \end{lstlisting}
-To see that the partial application of \texttt{qss} is only evaluated once,
-observe that after the first stage of evaluation, the residual has the form:
-\begin{lstlisting}
-2`let fun qs k = (let val y = ... in y)
-2`in (qs k1,...,qs km)`
-\end{lstlisting}
-where $y$ is bound to a large nest of case statements open on $k$.
+Observe that this code evaluates \texttt{qss} only once
+and subsitutes its result into the body of the function \texttt{lookup},
+which is then called many times in the second stage.

### 2016-01-09 paper/examples.tex
added mixed map and cut cut cut

@@
 prior work on pass separation for imperative languages. \texttt{dot} takes the
-dot product of two three-dimensional integer vectors, where the first two
+dot product of two three-dimensional vectors, where the first two
 coordinates are first-stage, and the last coordinate is second-stage. 
@@
 1`type vec = ^int * ^int * $`2`int`
-
-1`dot : vec * vec -> $int
+1`// dot : vec * vec -> $int
 fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
@@
 of the first stage, then multiply and add.
-
 Our algorithm splits \texttt{dot} into the two functions:
@@
 precomputation, and \texttt{dot2} adds that precomputation to the products of
-the final coordinates. This is exactly what is done in \cite{knoblock96}, except
-that they write the precomputation into a mutable cache.
+the final coordinates. This is exactly what is done in \cite{knoblock96}, though
+they write the precomputation into a mutable cache.
 
@@
 \begin{lstlisting} 
-datatype tree = Leaf
-              | Branch of int * tree * int * tree
+datatype tree = Leaf | Branch of int * tree * int * tree
 1`datatype list = Empty | Cons of int * list
@@
 fun part ((p,l):int*list) : (int*list*list) = 
-  case l of 
-    Empty => (0,Empty, Empty) 
+  case l of Empty => (0,Empty, Empty) 
   | Cons (h,t) => 
       let val (n,left,right) = part (p,t) in 
-      if h<p 
-      then (n+1,Cons(h,left),right) 
+      if h<p then (n+1,Cons(h,left),right) 
              else (n,left,Cons(h,right))
@@
 fun qs1 (l : list, k : unit) = 
-  ((), case l of
-    Empty => Leaf
-  | Cons (h,t) => 
-      Branch (
+  ((), case l of Empty => Leaf
+       | Cons (h,t) => Branch (
            let val (n,left,right) = part (h,t) in
@@
 2`fun qs2 (((), k : int), p : tree) = 
-  case p of
-    Leaf => 0
+  case p of Leaf => 0
   | Branch (n,left,h,right) =>
@@
 
-Thus, \texttt{qs1} performs $\Theta(n \log n)$ work per call, whereas
-\texttt{qs2} performs expected $\Theta(\log n)$ work.  This results in a net
-speedup over standard quickselect if we perform $\omega(\log n)$ queries on
-the same list.
-
 Note that the recursive \texttt{part} function is defined within a \texttt{gr}
@@
 With \texttt{gr} annotations, however, this can be prevented.
-%
 
-\subsection {Mixed Map Combinator}
+As discussed in \ref{sec:overview}, \texttt{qs1} performs $\Theta(n \log n)$ expected work per call, 
+whereas \texttt{qs2} performs $\Theta(\log n)$ expected work.  
+This results in a net speedup over standard quickselect if we 
+perform many (specifically, $\omega(\log n)$) queries on
+the same list---precisely the topic of our next example.
 
-[working on this section!]
+\subsection {Mixed-Stage Map Combinator}
 
-In \ref{sec:exampleQS}, we showed the output of splitting quickselect
-applied to one list and $m$ queries.  Here, we generalize to mapping
-quickselect onto a list of queries. Note that we have to define a
-datatype for integer lists at the second stage, which is given by
-\texttt{list2}.
+As a lambda calculus, one of the strengths of \lang\ is that it
+can express combinators as higher order functions.
+In this example, we consider just such a combinator: \texttt{tmap},
+which turns a function of type
+$\curr \mathtt{list} * \fut \mathtt{int} \to \fut \mathtt{int}$
+into one of type
+$\curr \mathtt{list} * \fut \mathtt{list2} \to \fut \mathtt{list2}$,
+by mapping over the second argument.\footnote{
+\lang\ doesn't have a way to ``share'' datatype declarations between stages, 
+so we define \texttt{list2} to be a list of integers at the second stage.}
 \begin{lstlisting} 
-1`atsignnext{`2`
-  datatype list2 = Empty2 | Cons2 of int * list2
-`1`}
+1`atsignnext{`2` datatype list2 = Empty2 | Cons2 of int * list2 `1`}
 
-map : (^list * $`2`int`1` -> $`2`int`1`)
-      -> ^list * $`2`list2`1` -> $`2`list2`1`
-fun map f (l, q) = next {`2`
-  let 
-  fun m Empty2 = Empty2
-    | m (Cons2(h,t)) =
-                    Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
+fun tmap (f : ^list * $`2`int`1` -> $`2`int`1`) (l : ^list, q : $`2`listliteraltwo`1`) = 
+  next{`2` let fun m Empty2       = Empty2
+              | m (Cons2(h,t)) = Cons2(prev{`1`f (l,next{`2`h`1`})`2`}, m t)
         in m prev{`1`q`2`}`1`}
+val mapqss = tmap qss`
+\end{lstlisting}
+Importantly, \texttt{tmap f} performs the first-stage part of \texttt{f} once
+and second-stage part of \texttt{f} many times.
+This was discussed in the context of partial evaluation in \ref{sec:topLevel}, 
+but it is especially clear when we look at the output of splitting:
+\begin{lstlisting} 
+1`fun tmap1 f = (fn (l,()) => ((),#2 (f (l,()))),())
+val (mapqss1,()) = tmap1 qs1
 
-val mapqss = map qss`
-\end{lstlisting}
-The \texttt{map} function splits into the following two functions:
-\begin{lstlisting} 
-2`datatype list2 = Empty2 | Cons2 of int * list2
-
-`1`fun map1 f = (fn (l,()) => ((),#2 (f (l,()))),())
-`2`fun map2 (f,()) ((l,q), p : tree) =
-  let 
-  fun m Empty2 = Empty2
+`2`datatype list2 = Empty2 | Cons2 of int * list2
+fun tmap2 (f,()) ((l,q), p : tree) =
+  let fun m Empty2 = Empty2
         | m (Cons2(h,t)) = Cons2(f ((l,h), p), m t) 
   in m q
-
-`1`val (mapqss1,()) = map1 qs1
-`2`val mapqss2 = map2 (qs2, ())`
+val mapqss2 = tmap2 (qs2, ())`
 \end{lstlisting}
 %
-Indeed, \texttt{qs1}, which is passed to \texttt{map1}, is evaluated only in the
-first pass, whereas \texttt{qs2}, which is passed to \texttt{map2}, is evaluated
-in the second pass, once per query \texttt{k}.
+Indeed, observe that the argument \texttt{f} of \texttt{qs1} is called only once,
+whereas the corresponding argument \texttt{f} in \texttt{qs2} is evaluated once per element in \texttt{q}.
 
@@
 of stages which perform increasingly fine-grained computations (e.g.,
-per-object, per-screen region, per-pixel). Many computations in later
-stages use the results from a single computation in an earlier
-stage\,\cite{OpenGL4Spec}.
+per-object, per-screen region, per-pixel), where computations in later
+stages use the results of an earlier
+stage multiple times\,\cite{OpenGL4Spec}.
 
-The standard way to program these graphics pipelines is to define one program (usually called a \emph{shader}) per stage.
-In other words, the programmer is expected to manually split their multi-stage programs.
-Yet manually partitioning programs by stage results in
-complex code where invariants must hold across different stages and
-local changes to the logic of one stage may require changes to that of
-upstream stages to communicate required inputs through the pipeline. 
+The standard way to program these graphics pipelines is 
+to define one program (usually called a \emph{shader}) per stage.
+In other words, the programmer is expected to write their 
+multi-stage programs in an already split form.
+This requirement results in complex code where invariants must hold across different stages and
+local changes to the logic of one stage may require changes to that of upstream stages. 
 This harms composition and modularity.  
-Graphics researchers therefore have suggested using mixed-staged programs
-\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express graphics program logic 
-for the entire pipeline as a single staged function\,\cite{Foley:2011}.
+Graphics researchers therefore have suggested using mechanisms like our stages
+\,\cite{Proudfoot:2001,Foley:2011,He:2014} to express graphics program logic,
+including representing entire pipeline as a single multi-stage function\,\cite{Foley:2011}.
 
@@
 stage.\footnote{Although this is a general pointwise multiplication combinator, 
-the variable names suggest a possible interpretations of the inputs and output:
+the variable names suggest a possible interpretation of the inputs and output:
 the first input function calculates the albedo of an object (a measure of how much light it reflects),
@@
 %
-A key strength of \lang\ is that it allows users to define combinators like this one as higher order functions.
-Prior work\,\cite{Foley:2011} does not support higher order functions.
+Prior work\,\cite{Foley:2011} lacks the ability to define such combinators,
+because it does not support higher order functions.

### 2016-01-09 paper/implementation.tex
added mixed map and cut cut cut

@@
 
-\subsection {Proof Implementation}
-
+\begin{abstrsyn}
 We have encoded the type system, dynamic semantics, splitting algorithm, and output semantics in Twelf,
 as well as all of the theorems of \ref{sec:semantics,sec:splitting}, and their proofs. 
-Details can be found in the supplementary materials.
-
-\subsection {Example Implementation}
-
-\begin{abstrsyn}
-We have a Standard ML implementation of the \lang\ language, a staged
+We also have a Standard ML implementation of the \lang\ language, a staged
 interpreter, the splitting algorithm, and an interpreter for split programs.
@@
 these additional features when convenient.
-
-We allow staging annotations around declarations. For example,
+Our expanded syntax allows staging annotations around declarations. For example,
 \begin{lstlisting}

### 2016-01-09 paper/relatedwork.tex
added mixed map and cut cut cut

@@
 performing computations that depend only on the static
-inputs~\cite{jones96}.  The roots of partial evaluation go back to
-Kleene's $s$-$m$-$n$ theorem~\cite{Kleene52},
-%Kleene did not intend to improve efficiency, however.  Lombardi
-%appears to be the first person to use the term in his work on
-%incremental computation~\cite{Lombardi67}.
-%There has since been an enormous body of work on partial evaluation,
-and we refer the reader to the book by Jones, Gommard, and Sestoft for
+inputs~\cite{jones96}.  
+We refer the reader to the book by Jones, Gommard, and Sestoft for
 a comprehensive discussion of partial evaluation work until the early
@@
 changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
-Thus, incremental computation can be thought of as a staged model,
-where changes in inputs come over a period of time in stages.
 However, unlike splitting, incremental computation does not require
@@
 incurring linear time cost at each step of the algorithm (as opposed
-to the logarithmic result produced by splitting).  Unlike splitting,
+to the logarithmic result produced by splitting).  Moreover,
 incremental computation techniques must also maintain sophisticated

### 2016-01-09 paper/examples.tex
small fixes

@@
 %
-Indeed, observe that the argument \texttt{f} of \texttt{qs1} is called only once,
-whereas the corresponding argument \texttt{f} in \texttt{qs2} is evaluated once per element in \texttt{q}.
+Indeed, observe that the argument \texttt{f} (for which \texttt{qs1} is later substituted) of \texttt{tmap1} is called only once,
+whereas the corresponding argument \texttt{f} (for which \texttt{qs2} is later substituted) in \texttt{tmap2} is evaluated once 
+per element in the query list \texttt{q}.
 
